### 2017-01-24 attic/LTI_DMI.tex
initial commit

\documentclass[a4paper, 11 pt]{article}
\usepackage{amsmath, amssymb, color, setspace, graphicx, dsfont, pdfpages, float, wrapfig, indentfirst}
\usepackage[left=0.5in, right=0.5in]{geometry}
\usepackage[font=small,labelfont=bf]{caption}
\begin{document}

\begin{equation}
  \begin{split}
    \dot{x} &= Ax + Bu \\
    y &= Cx
  \end{split}
\end{equation}
Let $\mathcal{A}$ be the set of matrices with eigenvalues $\l_{i} = \lambda_{i}$ and constants (determined by the transfer function) $k_{j} = c_{j}$. 
Define $\epsilon_{i} = |\lambda_{i}^{*} - \lambda_{i}|$ and $\epislon_{j} = |k^{*}_{j} - k_{j}| : j > i$
Let $\mathcal{L}$ be the set of matrices where $\sum_{i}^{n} \alpha_{i}\epsilon_{i} < \epsilon_{0}$
\begin{equation}
  P \in \mathcal{P} \text{ if } PB = B \text{ and } CP^{-1} = C \text{ and } PAP^{-1} \in \mathcal{A}
\end{equation}

For $C = [1 \ 0]$ and $B = [1 \ 1]^{T}$, $P_{p} = \left[\begin{matrix} 1 & 0 \\ p & 1-p \end{matrix}\right]$

$\left[\begin{matrix}1 & 0 \\ 0 & 0 \end{matrix}\right] P_{p}AP^{-1}_{p} + \left[\begin{matrix}0 & 0 \\ 0 & 1 \end{matrix}\right] P_{q}AP^{-1}_{q} \ \notin \ \mathcal{L}$

$\lambda_{1,2} = \frac{1}{2} \Bigg( \lambda_{1} + \lambda_{2} + x - \bar{x} \pm \Big( (\lambda_{1}-\lambda_{2})^{2} - (x-\bar{x}) \big( (x + 3\bar{x}) + 2(\lambda_{1} + \lambda_{2}) + \frac{4(\lambda_{1}-\bar{x})(\lambda_{2}-\bar{x})}{k + \bar{x}} \big) \Big)^{1/2} \Bigg)$

$\gimel$

\end{document}

### 2017-01-24 attic/LTI_system_notes.tex
initial commit

\documentclass[a4paper, 11 pt]{article}
\usepackage{amsmath, amssymb, color, setspace, graphicx, dsfont, pdfpages, float, wrapfig, indentfirst}
\usepackage[left=0.5in, right=0.5in]{geometry}
\usepackage[font=small,labelfont=bf]{caption}
\begin{document}

\section{What are LTI systems?}

A linear time invariant (LTI) dynamical system is a system of linear differential equations that describes a physical process. LTI systems are used in electrical and control engineering to model a myriad of phenomena including circuits.

One can use this methodology, under a set of assumptions, to reverse engineer a mechanism from impulse experiments (input/output data). The idea is that given a black box, and it's experimental manipulation, can we describe the internal mechanisms?

The inpulse data is summarized by the transfer function $G(s) = C( sI - A )^{-1} B + D$, a Laplace transform of the state space equation of the system. While the transfer function description of a system can be unique, the state space representation is not always unique.

The state space representation is as follows:

\begin{equation}
  \begin{split}
    \dot{x} =& Ax + Bu \\
y =& Cx + Du
\end{split}
\end{equation}

Where $x \in \mathbb{R}^{n}, \ y \in \mathbb{R}^{q}, \ u \in \mathbb{R}^{p}, \ A \in \mathbb{R}^{n \times n}, B \in \mathbb{R}^{n \times p}, C \in \mathbb{R}^{q \times n}, D \in \mathbb{R}^{q \times p}$.

The state space representation is considered an internal description of the system, whereas the transfer function is an external description. $\vec{x}$ is generally considered to be a state of the system, $\vec{u}$ to be the input, and $\vec{y}$ to be the observed output.

Multiple systems can interact in serial,

\begin{equation}
  \begin{split}
    \dot{x}_{1} &= A_{1}x_{1} + B_{1}u \\
    \dot{x}_{2} &= A_{2}x_{2} + B_{2}C_{1}x_{1} \\
    y &= Cx_{2}
\end{split}
\end{equation}

or in parallel as well:
\begin{equation}
  \begin{split}
    \dot{x}_{1} &= A_{1}x_{1} + B_{1}u \\
    \dot{x}_{2} &= A_{2}x_{2} + B_{2}u \\
    y &= C_{1}x_{1} + C_{2}x_{2}
\end{split}
\end{equation}

Let $P \in \mathbb{R}^{n \times n}$ be a non-singular matrix, and let $\dot{\overline{x}} = Px$, then
\begin{equation}
  \begin{split}
  \dot{\overline{x}} &= \overline{A}\overline{x} + \overline{B}u \\
  y &= \overline{C}\overline{x} + \overline{D}u
  \end{split}
\end{equation}
be an algebraically equivalent (AE) ``realization'' of the system, where $\overline{A} = PAP^{-1}$, $\overline{B} = PB$, $\overline{C} = CP^{-1}$, and $\overline{D} = D$.

If multiple realizations of a system are algebraically equivalent, then they must also have identical transfer functions $G(s) = \overline{G}(s)$. However, the opposite relation is not necessarily true. Multiple systems can have identical transfer functions and not have algebraic equivlaence. In fact, two systems with identical transfer functions can even have different dimensions.

Two LTI systems $\{A, B, C, D\}$ and $\{\overline{A}, \overline{B}, \overline{C}, \overline{D} \}$ have ``Zero-State Equivalence (ZSE)'' if they have the same transfer function. (AE $\Rightarrow$ ZSE).

\section{What biological question are we asking?}
How do gene regulatory networks (GRNs) change through evolutionary time? Specifically, under constant selection and environmental pressures, how does a maximally fit population's GRN change and what molecular and population parameters are significant to the process? How frequent does gene network rewiring occur in the absense of genetic drift or adaptation? Is rewiring typically compensatory and ususally follow the fixation of a deleterious allele, or can networks neutrally reorganize? How does the size, average degree, function, peiotropy, etc of a network influence its likelihood to drift? If networks can drift, how much topological heterogeneity can a well-mixed population tolerate, if any? Does topoligical heterogeneity and/or the size of the neutral genotype-set, if consequences of network organization, constrain or entail exaptation and evolvability? Are some network organizations (maybe higher dimensional realizations?) able to more rapidly evolve to construct novel phenotypes and meet the demands of changes in selection pressures?

\section{How can we mathematically model this question?}
In some respect an organism can be thought of as a black box that responds to some set of inputs (it's environment and other initial conditions) and outputs a phenotype. The phenotype is then evaluated by natural selection. The black box metaphor holds, because much of the details of what happens to construct a phenotype are unimportant as far as selection is concered, so long as the phenotype reliably performs its function. If multiple internal mechanisms can map the same set of inputs to the same set of outputs, and these mechanisms are mutationally nearby (in genotype space), then evolution may drift from mechanism to mechanism.

In this simple model we can say that $B\vec{u}$ is a list of initial GRN protein concentrations determined by the more general initial environmental conditions $\vec{u}$. $A$ represents all the genetic interactions of a particular GRN, and $C\vec{x}$ is an organism's phenotype. Fitness could be calculated by comparing a subset of the phenotype to an optimum. Maybe, $f(\cdot) = e^{- \int_{a}^{b} \Vert G(s) - G^{*}(s) \Vert ds}$ Alternatively, fitness scores could simply assess whether or not a phenotype is within some range or breaks some threshold by some time point. Or possbily the simplest: any phenotype not exactly equal to $G(s)$ is lethal, and $G(s)$ is perfectly fit.

Next, we will have to define mutation and recombination. Mutation needs to add and remove genetic interactions and recombination needs to shuffle genes during sexual reproduction. We could adapt methods from Lynch 2007 and Lynch and Hagner 2015 to model the probability of a TFBS appearing or disappearing due to mutational pressure. Every time a new binding site is gained or lost, the values within the A matrix are either modified or rows and columns are added removed following a specific set of rules. I am not sure if it would be easier to maintain a matrix size (say $m \times m$) where $m$ is arbitrarily large, with  most entries being zero, or if a matrix should only have as many dimensions as the number of active TFs. Recombination should then shuffle rows of the $A$ matrix randomly, as each row represents the regulatory region of a given gene (assuming only cis-regulation, and that the regulatory element is small enough to not break up during recombination).

To study some of the above posed questions, we can simulate (or analytically determine) the rates of GRN change, and how frequently these changes lead to significant rewiring or speciation. We would want to know the dynamics of a brownian motion over the set of all realizations.
Another interesting question would be to determine the size of the neutral genotype space as well as the number of connections from the neutral genotype space to the non-neutral genotype space. Maybe higher dimensional realizations of a GRN will have significantly more connections to non-neutral genotypes and therefore be more ``evolvable'' or exapted to novel environments?

Let $M \in \mathbb{R}^{n \times n}$ be a least dimensional realization of $A$, and let $M'$ bean $(n+m \times n+m)$ matrix where $M'_{(i,j)} = M_{(i,j)} \ \text{for } i,j \leq n, \ \text{and } 0 \ \text{otherwise}$. Let $P$ be any $(n+m \times n+m)$ matrix with rank $n$. Then for any matrix $B$ such that $BP = PM'$, $B$ is a realization of $A$.

\section{List of papers with some notes.}

R. E. Kalman, “Mathematical description of linear dynamical systems,” Journal of the Society for Industrial and Applied Mathematics, Series A: Control, vol. 1, no. 2, pp. 152–192, 1963.
This is a good overview written by Kalman. Probably the most important paper on this list.

Minimal state-space realization in linear system theory: an overview. B. De Schutter. 2000.
Another, more recent, overview that I found helpful to understand controllability and observability.

The Inverse Problem of Stationary Covariance Generation. BDO Anderson. 1969.

Equivalence of Linear Time- Invariant Dynamical Systems. BDO Anderson, BW Necomb, RE Kalman. 1966.

B. Ho and R. E. Kalman, “Effective construction of linear state-variable models from input/output functions,” Proceedings of the 3rd Annual Allerton Conference on Circuit and System Theory, vol. 1, no. 2, p. 449459, 1965.

Dynamic Structure Functions for the Reverse Engineering of LTI Networks. J Goncalves. 2007.
I have not read this paper yet, but apparently it outlines a method for recovering the state space equation from specific sets of measurements.

New Developments in Systems Theory Relevant to Biology. RE Kalman. 1968.
Not technical but kind of interesting.

Old and New Directions of Research in Systems Theory. RE Kalman. 2010?

Minimality and Observability of Group Systems. HA Loeliger, GD Forney Jr., T Mittelholzer, MD Trott. 1994. 
Another paper I just found. It looks like it might be useful.

Partial Realization of Descriptor Systems. P Benner and VI Sokolov. 2006. 
I think this is about recovering a finite realization from infinite systems. Maybe it would be interesting to think about the similarity between transfer functions though? If selection only cares about a specific interval of the transfer function, say $G(10)$, then maybe two partial transfer functions will be effecrively equivlaent?

There are also the previous papers I sent on identifiability and distinguishability. In the original Bellman and Astrom paper where they introduce ``structural identifiability'' they show that a linear dynamical system is identifiabile if and only if it is also controllable and observable.

\end{document}

### 2017-01-24 attic/LTI_systems_notes_2.tex
initial commit

\documentclass[a4paper, 11 pt]{article}
\usepackage{amsmath, amssymb, color, setspace, graphicx, dsfont, pdfpages, float, wrapfig, indentfirst}
\usepackage[left=0.5in, right=0.5in]{geometry}
\usepackage[font=small,labelfont=bf]{caption}
\begin{document}

\section{Transfer Functions}

\section{GRN}

Let $G^{*}(s) = \frac{(s-k)}{(s - \lambda_{1})(s - \lambda_{2})}$ be the optimal phenotype. This transfer function can be realized by a GRN composed of 2 or more genes. If $k$ is equal to $\lambda$ a one gene GRN can completely realize the dynamics.

The phenotype of an organism is,
G(s) = \begin{bmatrix} 1 & 0 \end{bmatrix} \Big(\begin{bmatrix} s & 0 \\ 0 & s \end{bmatrix} \right] - \begin{bmatrix} a_{11} & a_{12} \\ a_{21} & a_{22} \end{bmatrix} \Big)^{-1} \left[ \begin{array}{c} 1 \\ 1 \end{array} \right]

The dynamics of the optimal GRN in the time domain (via the inverse Laplace transform) is,

$f^{*}(t) = \mathcal{L}^{-1}(G^{*}(s)) = \frac{e^{\lambda_{1}t}(\lambda_{1} - k) - e^{\lambda_{2}t}(\lambda_{2} - k)}{(\lambda_{1} - \lambda_{2})}$

These dynamics can be realized by an infinite number of $\geq 2$ gene systems ($in \mathbb{R}$). The $2$ dimensional realizations will all have the form below where $x$ can be any real number. If the transfer function is of the form $G^{*}(s) = \frac{1}{(s - \lambda_{1})}$, then $\lambda_{2} = k$ can also be any real number.

\[
\begin{bmatrix}
  x  & \lambda_{1} + \lambda_{2} - k - x \\
  \frac{-(x - \lambda_{1})(x - \lambda_{2})}{\lambda_{1} + \lambda_{2} - k - x} & \lambda_{1} + \lambda_{2} - x
\end{bmatrix}
\]

The fitness of an individual organism is computed as
\begin{equation}
  \phi(f(t)) = e^{-\sqrt{\int_{0}^{\infty} (f^{*}(t) - f(t))^{2}dt}}
\end{equation}
Where as individual realization is $f(t) = \frac{e^{\gamma_{1}t}(\gamma_{1} - h) - e^{\gamma_{2}t}(\gamma_{2} - h)}{(\gamma_{1} - \gamma_{2})}$
If we look at the difference (instead of the squared difference) between optimal and actual, and assume that the real parts of the eigenvalues are $\leq 0$, we get, 
\begin{equation}
  \varphi(f(t)) = e^{-\int_{0}^{\infty} (f^{*}(t) - f(t))dt} = e^{-\left|\frac{k}{\lambda_{1}\lambda_{2}} - \frac{h}{\gamma_{1}\gamma_{2}}\right|}
\end{equation}

\begin{equation}
  \frac{\partial(\text{det}(A))}{\partial A_{ij}} = \lambda_{1}\lambda_{2} \text{Tr}\left[A^{-1}\frac{\partial A}{\partial A_{ij}}\right]
\end{equation}
And, $\gamma_{1}\gamma_{2} \approx \lambda_{1}\lambda_{2} + \text{Tr}\left[\text{adj}(A)\frac{\partial A}{\partial A_{ij}}\right]$ = $\lambda_{1}\lambda_{2} + (\text{adj}(A))^{T}_{ij}$
\begin{equation}
  \nabla G = \left[ \begin{array}{c} (s - a_{22})(s + a_{12} - a_{22}) \\ (s - a_{22})(s + a_{21} - a_{11}) \\ a_{12}(s + a_{12} - a_{22}) \\ a_{12}(s + a_{21} - a_{11}) \end{array} \right] \times \big((s-\lambda_{1})(s-\lambda_{2})\big)^{-2}
\end{equation}

\begin{equation}
  \nabla^{2}G = \\
    \!\begin{aligned}
&
\left[\begin{matrix}
    2(s - a_{22})^{2}(s + a_{12} - a_{22}) & (s - a_{22})(2a_{21}(s+ a_{12} - a_{22}) + (s-\lambda_{1})(s-\lambda_{2})) \\
    \cdot & 2a_{21}(s - a_{22})(s+a_{21} - a_{11})^{2} \\
    \cdot & \cdot \\
    \cdot & \cdot \\
  \end{matrix}\right.\\
&\qquad\qquad
\left.\begin{matrix}

  {} 2a_{12}(s - a_{22})(s + a_{12} - a_{22}) & a_{12}(2a_{21}(s + a_{12} - a_{22}) + (s - \lambda_{1})(s - \lambda_{2})) \\
  {} (s-a_{22})(2a_{12}(s + a_{21} - a_{11}) + (s-\lambda_{1})(s-\lambda_{2}))  & (s + a_{21} - a_{11})(2a_{12}a_{21} + (s-\lambda_{1})(s-\lambda_{2})) \\
  {} 2(a_{12})^{2}(s + a_{12} - a_{22}) & a_{12}(2a_{12}(s + a_{21} - a_{11}) + (s-\lambda_{1})(s-\lambda_{2})) \\
  {} \cdot & 2a_{12}(s - a_{11})(s + a_{21} - a_{11}) 
\end{matrix}\right] \\ \\
&\qquad\qquad
  \times \big((s-\lambda_{1})(s-\lambda_{2})\big)^{-3}
\end{aligned}
\end{equation}

\end{document}

### 2017-01-24 attic/eugene_notes.tex
initial commit

\documentclass[a4paper, 11 pt]{article}
\usepackage{amsmath, amssymb, color, setspace, graphicx, dsfont, pdfpages, float, wrapfig, indentfirst, multicol}
\usepackage[left=0.5in, right=0.5in]{geometry}
\usepackage[font=small,labelfont=bf]{caption}

\newcommand\bigZero{\scalebox{2}{$0$}}
\newcommand\bigLambda{\scalebox{2}{$\Lambda$}}
\newcommand\bigI{\scalebox{2}{$I$}}
\newcommand\bigPhi{\scalebox{2}{$\Phi$}}
\newcommand\bigPsi{\scalebox{2}{$\Psi$}}

\begin{document}
  \title{A Framework for the Evolution of Genetic Networks}
  \section{Introduction}

    It is commonly taught that an organism's genome contains most of the heritable material that natural selection filters, and that an organism's phenotype directly determines its evolutionary fitness. Between genotype and phenotype is an often complicated and poorly understood molecular machinery -- and it is a major goal common to many disciplines within the life sciences to elucidate its form, function, and evolution.

    It is becoming increasingly evident that in order to understand evolutionary processes, we need to precisely understand how the molecular machinery that maps genotypes to phenotypes both functions and evolves. This complicated molecular machinery can be modelled as a dynamical system. Thus, an idealized study of evolution would benefit from knowledge of genomes, phenotypes, and importantly detailed descriptions of genotype-phenotype maps. Movement in this direction is ongoing, as researhers have begun to study the evolution of empirically inspired computational and mathematical models of gene regulatory networks (GRNs) and metabolic networks that incorporate empirical data such as sequences and expression patterns. If we allow the reasonable assumption that the genotype-phenotype map can be represented as a system of differential equations, we can immediately discuss its evolution in a much more mechanistic, yet general, manner.

  \section{Linear Genetic Networks}
    \begin{equation}
      \begin{split}
        \dot{x} &= A x + B u \\
        y &= C x
      \end{split}
    \end{equation}

    \begin{equation}
      \begin{split}
        T(s) :&= C (sI - A)^{-1} B \\
        T(s) &= \bar{T}(s) \iff \\ 
        CA^{k}B &= \bar{C} \bar{A}^{k} \bar{B} \ \forall \ 0 \leq k \leq n-1
      \end{split}
    \end{equation}

   \section{Parameterization}

      \begin{equation}
        \begin{bmatrix} \bigPhi & \bigPsi \\ \\ \bigZero & \begin{bmatrix} Z_{m} & 0 \\ 0 & 0_{\infty} \end{bmatrix} \end{bmatrix} \begin{bmatrix} \bigLambda_{n} & \bigZero \\ \\ \bigZero & \bigI_{\infty} \end{bmatrix} \begin{bmatrix} \bigPhi^{-1} & \bigZero \\ \\ \bigZero & \bigI_{\infty} \end{bmatrix}
      \end{equation}

  \section{Notes}

    We are interested in a perfectly adapted, large population, evolving in a static environment for an infinite number of generations. Given these conditions, througout evolutionary time, the population will realize different, yet symmetric GRN (gene regulatory network) topologies. GRNs are symmetric if they are input/output equivlaent such that they produce identical phenotypes (with respect to selection) when active in indentical environments.

\end{document}

### 2017-01-24 attic/lti_notes_3.tex
initial commit

\documentclass[a4paper, 11 pt]{article}
\usepackage{amsmath, amssymb, color, setspace, graphicx, dsfont, pdfpages, float, wrapfig, indentfirst, epstopdf, inputenc, multicol}
\usepackage[left=0.5in, right=0.5in]{geometry}
\usepackage[font=small,labelfont=bf]{caption}
\begin{document}

\textbf{Zero State Definition:}

If a system is in the zero-state then the zero-input will yield a null response.

\textbf{Zero-State Response Definition:}

The zero-state response of a system to $\delta(t-\xi)$ is, 
\begin{equation}
  \begin{split}
  \delta(t-\xi) = \int_{t_{0}}^{t} f(t, \xi^{'})\delta(\xi^{'}-\xi)d\xi^{'} \\
   = f(t, \xi) \text{ for } t \geq \xi \geq t_{0} \\
   = 0 \text{ for } \xi > t \geq t_{0}
 \end{split}
\end{equation}

Therefore the zero-state response of a system to input $u$ is, 
\begin{equation}
  A(u) = \int_{t_{0}}^{t}f(t,\xi)u(\xi)d\xi \ \ t_{0} \leq \xi \leq t
\end{equation}

A system is zero-state time invariant if and only if its impulse response $f(t,\xi)$ is of the form $f(t-\xi)$.

\textbf{Proof:}

    Let $f(t,\xi) = w(\tau, t)$

    where $\tau \triangleq (t -\xi)$

    The zero-state response of a system to $\delta(t-\xi)$ is $f(t,\xi)$

    thus the zero-state response of the system to $\delta(t -(\xi + \lambda))$ where $\lambda$ is an arbitrary shift is $f(t, \xi + \lambda)$ or $w(\tau -\lambda,t)$

    this implies $w(\tau - \lambda, t) = w(\tau - \lambda, t - \lambda) \ \forall t \forall \tau \forall \lambda$

\textbf{Transfer Function Definition:}
\begin{equation}
  G(s) \triangleq \int_{-\infty}^{\infty} e^{-st}f(t)dt
\end{equation}

If $y$ is the zero-state response to $u$ then,

\begin{equation}
  Y(s) = G(s)U(s)
\end{equation}

The Transfer Function gives incomplete information about the zero-input response and fully characterizes the zero-state response.

The Transfer Function can also be viewed as the steady state response of a system to input $e^{st}$ divided by $e^{st}$:
$ G(s) \triangleq \frac{\text{steady state response of a system to }e^{st}}{e^{st}}$

Zero-State Equivalent Linear Dynamical Systems.

\begin{multicols}{2}
\begin{equation}
    \Sigma  
    \left \{ 
      \begin{array}{ll}
        \dot{x} = Ax + Bu \\
        y = Cx
      \end{array}
      \right.
\end{equation}

  \begin{equation}
  \bar{\Sigma}
    \left \{ 
      \begin{array}{ll}
        \dot{\bar{x}} = \bar{A}\bar{x} + \bar{B}u \\
        y = \bar{C}\bar{x}
      \end{array}
      \right. 
\end{equation}
\end{multicols}

\textbf{Definition:} two dynamical systems $\Sigma$ and $\bar{\Sigma}$ are algebraically equivalent if (where $\bar{x} = Px$),

  (a) $\bar{A} = PAP^{-1}$

  (b) $\bar{B} = PB$

  (c) $\bar{C} = CP^{-1}$

  Where $P$ is a nonsingular square matrix of rank $n$. If two systems are algebraically equivalent then they are also zero-state equivalent. Two systems that are zero-state equivalent are not necessarily algebraically equivalent and can be in different dimensions.

  Definition: two dynamical systems are zero-state equivalent if they have the same transfer function $G(s)$.

  The transfer function, $G(s) = C(sI-A)^{-1}B$, is the Laplace transform of the system's time domain dynamics.

  $G(s) = \bar{G}(s) \iff$

  $CA^{i}B = \bar{C}\bar{A}^{i}\bar{B}$ for all $i \geq 0$.

  \textbf{Proof:}

  $G(s) = C(sI-A)^{-1}B$

  $=Cs^{-1}(I-s^{-1}A)^{-1}B$

$=Cs^{-1} \left( \sum_{i=0}^{\infty}(s^{-1}A)^{i}\right)B$

$=\sum_{i=0}^{\infty} CA^{i}Bs^{i+1}$

$\sum_{i=0}^{\infty} CA^{i}Bs^{-(i+1)} = \sum_{i=0}^{\infty} \bar{C}\bar{A}^{i}\bar{B}s^{-(i+1)}$

$\iff CA^{i}B = \bar{C}\bar{A}^{i}\bar{B} \ \forall i$

Note: the controllability matrix $\mathcal{C}$ is defined as:
\begin{equation}
  \begin{bmatrix}
      B \ AB \ ... \ A^{n-1}B
  \end{bmatrix}
\end{equation}
And the observability matrix $\mathcal{O}$ is defined as:
\begin{equation}
    \begin{align}
  \begin{bmatrix}
      C \\ CA \\ \vdots \\ CA^{n-1}
  \end{bmatrix}
    \end{align}
\end{equation}
\end{document}

### 2017-01-24 attic/lti_notes_4.tex
initial commit

\documentclass{seminar}
\usepackage{hyperref, amsmath, amssymb, color, setspace, graphicx, dsfont, pdfpages, float, wrapfig, indentfirst, epstopdf, inputenc, multicol}
%\usepackage[left=0.5in, right=0.5in]{geometry}
%\usepackage[font=small,labelfont=bf]{caption}
\begin{document}
\begin{slide}
  \centering
  \section*{A Mathematical Model of Gene Regulatory Network Evolution}

  \textbf{Josh Schiffman}

  December 16, 2016
\end{slide}
  \begin{slide}
    \section*{Background}
    \begin{itemize}
      \item How do we mathematically model Developmental Systems Drift (DSD) and Gene Regulatory Network (GRN) evolution?
        \begin{itemize}
          \item First, we need a good analytic description of an organism -- something that includes and distinguishes among genotype, GRN, and phenotype. 
          \item Second, these descriptions should be as precise as possible. 
          \item Third, we need to describe the evolutionary and population dynamics of this model.
        \end{itemize}
      \item How do we apply this model?
    \end{itemize}
  \end{slide}

  \begin{slide}
    \section*{An organism is described by its molecular time-dynamics.}
      Molecular (say protein or mRNA) concentrations are represented by $\vec{x}$, the gene regulatory interactions by $\alpha, \beta, \gamma,$ and $\delta,$ and the environmental/initial inputs by $B$.  
      \begin{equation*}
        \begin{bmatrix} \dot{x}_{1} \\ \dot{x}_{2} \end{bmatrix} = \begin{bmatrix} \alpha & \beta \\ \gamma & \delta \end{bmatrix} \begin{bmatrix} x_{1} \\ x_{2} \end{bmatrix} + \begin{bmatrix} B_{1} \\ B_{2} \end{bmatrix} u
      \end{equation*}
    Each row vector $\begin{bmatrix} \alpha & \beta \end{bmatrix}$ of the GRN matrix is the promoter region of a gene, such that, $\alpha$ describes the how much gene-1 self regulates and $\beta$ describes how much gene-2 regulates gene-1. 
  \end{slide}

  \begin{slide}
    \section*{The phenotype is just a subset of an organism's proteins directly visible to selection}
    \begin{equation*}
      y = \begin{bmatrix} C_{1} & C_{2} \end{bmatrix} \begin{bmatrix} x_{1} \\ x_{2} \end{bmatrix}
    \end{equation*}
    For instance, if $C = \begin{bmatrix} 1 & 0 \end{bmatrix}$, the phenotype is only described by the dynamics of gene-1. This means that selection can only ``see'' what happens to the proteins produced by gene-1, whereas the dynamics of gene-2 are invisible to selection (and thus not formally defined as part of the phenotype).

      This does not mean, however, that gene-2 is unnecessary with respect to the phenotype!
  \end{slide}

  \begin{slide}
    \section*{For example, consider a GRN with oscillating protein concentrations.}
    In this 2 gene system, neither gene-1 nor gene-2 self regulate. Gene-2 up-regulates gene-1, and gene-2 is down-regulated by gene-1:
      \begin{equation*}
        \begin{bmatrix} \dot{x}_{1} \\ \dot{x}_{2} \end{bmatrix} = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} \begin{bmatrix} x_{1} \\ x_{2} \end{bmatrix} + \begin{bmatrix} 1 \\ 1 \end{bmatrix} u
      \end{equation*}
      Furthermore, only the dynamics of gene-1 are important:
      \begin{equation*}
        y = \begin{bmatrix} 1 & 0 \end{bmatrix} \begin{bmatrix} x_{1} \\ x_{2} \end{bmatrix}
      \end{equation*}
  \end{slide}

  \begin{slide}
    \begin{figure}{\bf{2-Gene GRN Oscillating Expression Dynamics}}
      \centering
      \caption{The dynamics of gene-1 are, $y = \cos(t) + \sin(t)$}
      \includegraphics[width=\textwidth, height=0.7\textheight]{sin_cos_t}
    \end{figure}
    A single-gene system cannot oscillate. 
  \end{slide}

  \begin{slide}
    \section*{Expression dynamic complexity is limited by the number of genes in a network.}
    $f(t)$ describes protein concentrations at time $t$, $\lambda_{j}$ is the $j$th eigenvalue of the GRN and $\xi$ is a constant. 
      \begin{equation*}
        f(t) = \sum_{j=1}^{n} e^{\lambda_{j}t} \xi_{j}
      \end{equation*}
      The oscillating dynamics can be rewritten in this form too:
      \begin{equation*}
        f(t) = e^{it}\frac{(i-1)}{2i} + e^{-it}\frac{(i+1)}{2i} = \cos(t) + \sin(t)
      \end{equation*}
      Sometimes it's (visually) easier to study the Laplace transform of the system. 
      \begin{equation*}
        \mathcal{L}(f(t)) := G(s) = \sum_{j}^{n} \frac{c_{j} b_{j}}{(s- \lambda_{j})}
      \end{equation*}

  \end{slide}

  \begin{slide}
    \textbf{Is the GRN architecture for a given phenotype unique? If no, how many GRN topologies yield equivalent dynamics?}

    Let two systems $\Sigma$ and $\bar{\Sigma}$ be defined as,
    \begin{multicols}{2}
\begin{equation*}
    \Sigma  
    \left \{ 
      \begin{array}{ll}
        \dot{x} = Ax + Bu \\
        y = Cx
      \end{array}
      \right.
\end{equation*}

  \begin{equation*}
  \bar{\Sigma}
    \left \{ 
      \begin{array}{ll}
        \dot{\bar{x}} = \bar{A}\bar{x} + \bar{B}u \\
        y = \bar{C}\bar{x}
      \end{array}
      \right. 
\end{equation*}
\end{multicols}

    Where $x \in \mathbb{R}^{n}, \ y \in \mathbb{R}^{l}, \ u \in \mathbb{R}, \ A, \bar{A} \in \mathbb{R}^{n \times n}, \ B, \bar{B} \in \mathbb{R}^{n \times l}, \ C, \bar{C} \in \mathbb{R}^{l \times n}$. 
  \end{slide}

  \begin{slide}
    The systems $\Sigma$ and $\bar{\Sigma}$ are ``algebraically equivalent'' (and have identical dynamics) if,
    \begin{equation}
      \begin{split}
        \bar{x} =& Px \\
        \bar{A} =& PAP^{-1} \\
        \bar{B} =& PB \\
        \bar{C} =& CP
      \end{split}
    \end{equation}
    Where $P$ is any invertible matrix.

    This is only true if both GRNs are composed of the exact same number of genes.

    Can two systems have identical dynamics, yet differ in the number of genes?
  \end{slide}

  \begin{slide}
    Two systems, even in different dimensions (!!), exhibit identical dynamics (and are ``zero-state equivalent'') if and only if their transfer functions are identical $G(s) = \bar{G}(s)$.

    Remember, $G(s) = \mathcal{L}(f) = C(sI-A)^{-1}B$

    \begin{equation*}
      \begin{split}
        G(s) &= \bar{G}(s) \iff \\
        CA^{i}B &= \bar{C}\bar{A}^{i}\bar{B} \ \forall \ i \geq 0
      \end{split}
    \end{equation*}
  \textbf{Proof:}
    \begin{equation*}
      \begin{split}
        &G(s) = C(sI-A)^{-1}B \\ 
        &= Cs^{-1}(I-s^{-1}A)^{-1}B \\
        &=Cs^{-1} \left( \sum_{i=0}^{\infty}(s^{-1}A)^{i}\right)B \\
        &= \sum_{i=0}^{\infty} CA^{i}Bs^{i+1} \\
        \sum_{i=0}^{\infty} CA^{i}Bs^{-(i+1)} &= \sum_{i=0}^{\infty} \bar{C}\bar{A}^{i}\bar{B}s^{-(i+1)} \\
        \iff CA^{i}B &= \bar{C}\bar{A}^{i}\bar{B} \ \forall i 
      \end{split}
    \end{equation*}
  \end{slide}

  \begin{slide}
    \section*{Wait... how does any of this make biological sense?}
    We only care about systems where $\bar{C} = C$ and $\bar{B} = B$.

    Any two systems that are ``algebraically equivalent'' where $PB = B$ and $CP = C$ ($P$ is a stabilizer subgroup of the invertible matrices), have identical inputs and identical phenotypes, but have different \textit{mechanisms}, since it's possible that $A \neq PAP^{-1}$.

    The \textbf{set of all 2-gene GRNs} with the dynamics $G(s) = \frac{(s-k)}{(s-\lambda_{1})(s-\lambda_{2})}$ is, 
      \begin{equation*}
        \begin{bmatrix} 1 & 0 \\ r & 1-r \end{bmatrix} \begin{bmatrix} \lambda_{1} & \lambda_{2} - k \\ 0 & \lambda_{2} \end{bmatrix} \begin{bmatrix} 1 & 0 \\ \frac{r}{r-1} & \frac{-1}{r-1} \end{bmatrix}
      \end{equation*}
      $\forall \ r \neq 1 \in \mathbb{R}$
  \end{slide}

  \begin{slide}
    \section*{The set of all GRNs, in any dimension, with identical dynamics is...}
    \begin{equation*}
    P \left[
      \begin{array}{cc}
        A &  0  \\
        0 & \Lambda 
      \end{array}\right] P^{-1} : PB = B \text{ and } 
    \end{equation*}
    \begin{equation*}
      CP = C,  \text{ and } \Lambda = \begin{bmatrix} 
      \lambda_{n+1} &  & 0 \\
      & \ddots & \\
      0 &        & \lambda_{n+m} 
    \end{bmatrix}
    \end{equation*}
    Where some elements in $P$ and all eigenvalues in $\Lambda$ are free to take any value.
    ($B$ and $C$ can be augmented with $0$s).
  \end{slide}

  \begin{slide}
    \section*{DMIs}
      Define reproduction as 
      \begin{equation*}
        \begin{bmatrix} 1 & 0 \\ 0 & 0 \end{bmatrix} A + \begin{bmatrix} 0 & 0 \\ 0 & 1 \end{bmatrix} \bar{A}
      \end{equation*}
      Let phenotypic error be $\epsilon_0 := \big( \epsilon_{1} + \epsilon_{2} + \epsilon_{k} \big)$, where $\epsilon_{1,2} := \vert \lambda_{1,2} - \lambda^{H}_{1,2} \vert$ and $\epsilon_{k} := \vert k - k^{H} \vert$.

      $\lambda^{H}_{1,2} = \frac{1}{2} \Bigg( \lambda_{1} + \lambda_{2} + \zeta - \bar{\zeta} \pm \Big( (\lambda_{1} - \lambda_{2})^{2} - (\zeta - \bar{\zeta}) \big( ( \zeta + 3\bar{\zeta}) + 2 (\lambda_{1} + \lambda_{2}) + \frac{4(\lambda_{1} - \bar{\zeta})(\lambda_{2} - \bar{\zeta})}{k+\bar{\zeta}}\big) \Big)^{\frac{1}{2}} \Bigg)$

      $k^{H} = \zeta - \bar{\zeta}$.

      Let GRNs that survive have $\epsilon_{0} < \delta$.
  \end{slide}

  \begin{slide}
    \textbf{Hybrid expression dynamics deviate significantly}
   \\\\ 
    Both $\begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix}$ and $\begin{bmatrix} -1 & 2 \\ -1 & 1 \end{bmatrix}$ have identical phenotypes, however their hybrids do not. 
      \begin{figure}
        \hspace*{-3.7cm}
        \includegraphics[width=1.7\textwidth,height=0.6\textheight]{hybrid_osc_2n}
      \end{figure}
  \end{slide}

  \begin{slide}
    $\begin{bmatrix} 0 & -6 \\ 1 & -5 \end{bmatrix}$ and $\begin{bmatrix} -12 & 6 \\ -15 & 7 \end{bmatrix}$
       \begin{figure}
        %\hspace*{-3.7cm}
        \includegraphics[width=0.8\textwidth,height=0.7\textheight]{hybrid_321_fixed}
      \end{figure}

  \end{slide}

  \begin{slide}
    \section*{Evolution as Brownian Motion}

    If $\mathcal{B}$ is an $\eta$-dimensional (where $\eta$ consists of the free parameters in $P$ and eigenvalues $\lambda_{n+1}$ through $\lambda_{n+m}$) Brownian Motion with covariance matrix $\sigma^{2} I$ then $\mathbb{E}\left[ \Vert \mathcal{B} \Vert^{2} \right] = \eta \sigma^{2} t$, however, $A = F(\mathcal{B})$. 
  \end{slide}

  \begin{slide}
    \textbf{How to model gene addition and deletion.}
      \begin{itemize}
        \item Gene addition: add row and column to the GRN matrix; augment $B$ and $C$ with zeros. 
        \item Gene deletion: more complicated. What is the probability that, 
          \begin{itemize}
            \item a set of dimensions (genes) we don't care about. 
            \item a set of dimensions only affects the other genes in the network ``a little'' bit (the GRN is mostly lower-triangular).
              \begin{multicols}{2} \begin{equation*}
                Z = \begin{bmatrix} \bigstar &\vert \epsilon \\ \hdots & \hdots \end{bmatrix}
              \end{equation*}
              \begin{equation*}
                \Vert G(s) - \bar{G}(s) \Vert \leq \delta
              \end{equation*}
          \end{multicols}
        \item if the phenotypic effect of these extra dimensions diminishes as gene networks evolve to be larger, pherhaps there will be an equillibrium level GRN complexity. (note: empirical network connectedness and robustness). 
          \end{itemize}
      \end{itemize}
  \end{slide}
  \begin{slide}
    \section*{Thank you!}
    \textbf{Peter Ralph \\ Sergey Nuzhdin \\ Erik Lundgren}
  \end{slide}
\end{document}

### 2017-01-24 attic/peter_visit_notes.tex
initial commit

\documentclass[a4paper, 11 pt]{article}
\usepackage{amsmath, amssymb, color, setspace, graphicx, dsfont, pdfpages, float, wrapfig, indentfirst, multicol}
\usepackage[left=0.5in, right=0.5in]{geometry}
\usepackage[font=small,labelfont=bf]{caption}

\newcommand\bigZero{\scalebox{2}{$0$}}
\newcommand\bigGamma{\scalebox{2}{$\Gamma$}}
\newcommand\bigA{\scalebox{2}{$A$}}
\begin{document}

  \section{Integral Formulation}
  \begin{equation}
    \begin{split}
      \dot{x} &= Ax + Bu \\
      &= e^{-tA}x \\
      \rightarrow \dot{\gamma} &= e^{-tA} \dot{x} -A e^{-tA}x \\
      &= e^{-tA}(Ax + Bu) - Ae^{-tA}x \\
      &= e^{-tA}Bu \\
      \rightarrow \gamma(t) &= \gamma(0) + \int_{0}^{t} e^{-sA}Bu(s) ds \\
      x(t) &= e^{tA} (x(0) + \int_{0}^{t} e^{-sA}Bu(s) ds \\
      \text{If } AB=\bar{A}B &\text{ and } CA=C\bar{A} \\
      \text{ then } y(t) &= \bar{y}(t)
    \end{split}
  \end{equation}

  \section{Null Space Parameterization}
  \begin{equation}
    \begin{split}
      AB=\bar{A}B \\
      CA = C\bar{A} \text{ then } \\
      (A-\bar{A})B=0 \text{ and } C(A-\bar{A}) = 0 \\
      \rightarrow \text{Let } Z=\bar{A}-A \\
      \text{Let } \{v_{i}\} \text{ span the right null space of } C \\
      \rightarrow Cv_{i} = 0 \forall i \\
      \{w_{j}\} \text{ span the left null space of } C \\
      w_{j}B = 0 \forall j \\
      \iff Z=\sum_{ij} e_{ij}v_{i}w_{j}^{T} \text{ for some } e_{ij} \in \mathbb{R} \\
      (\text{Also } A^{k}B=\bar{A}^kB)
    \end{split}
  \end{equation}

  \begin{equation}
    \begin{split}
      \bar{A} = PAP^{-1} \\
      (A^{k} - PA^{k}P^{-1})B=0=(A^{k}-(I-G)A^{k}(I+G+G^{2}+ \hdots))B \\
      =(-A^{k}(G + G^{2} + \hdots) + GA^{k}(I + G + G^{2} + \hdots))B \\
      = (GA^{k}-(I-G)A^{k}(G+G^{2} + \hdots))B \\
      P=I-G \\
      P^{-1} = \sum_{n \geq 0} G^{n} = I + G + G^{2} + \hdots \\
    \end{split}
  \end{equation}

  \section{Model Notes}
  \begin{itemize}
    \item $A_{ij} \rightarrow$
      \begin{itemize}
        \item parameterization $\rightarrow$
          \begin{itemize}
            \item $e$'s independently
            \item $A_{ij}$'s independently
            \item $A_{ij}$'s jointly
          \end{itemize}
        \item bias towards $0$? Speed?
          \begin{itemize}
            \item none
            \item Ornstein-Uhlenbeck Process
            \item stays at $0$
          \end{itemize}
      \end{itemize}
    \item Dimension of $A$
      \begin{itemize}
        \item up $\rightarrow$ add rows/columns (either all $0$ or not?)
        \item down $\rightarrow$ delete if all $0$, use Kalman decomposition algorithm, if ``unnecessary'', or other ideas.
      \end{itemize}
    \item Recombination
    \item Fitness
  \end{itemize}

  \section{Parameterization}
    \begin{equation}
      \begin{split}
        G(s) = C(sI-A)^{-1}B : A \in \mathbb{R}^{n \times n}, \ C \in \mathbb{R}^{k \times n}, \ B \in \mathbb{R}^{n \times l} \\
        \text{If } A = \sum^{n}_{i=1} \lambda_{i}u_{i}v_{i}^{T} = U \Lambda V^{T}, \ \Lambda \text{ is diagonal and } V^{T}U=I \\
        \text{Then } (sI-A)^{-1} = \sum_{i = 1}^{n} \big( \frac{ 1 }{ s-\lambda_{i} } \big) u_{i}v_{i}^{T} \\
        =U(sI-\Lambda)^{-1}V^{T} \\
        \text{So } G(s) = CU(sI-\Lambda)^{-1}V^{T}B \\
        \text{If } k=l=1 \rightarrow c=CU, \ b=V^{T}B \\
        \rightarrow G(s)=\sum_{i} \frac{ c_{i} b_{i} }{s-\lambda_{i}} = \frac{ \sum_{i} c_{i} b_{i} \prod_{j \neq i} (s-\lambda_{j}) }{ \prod_{i} (s-\lambda_{i}) }
      \end{split}
    \end{equation}

   \textbf{Claim:}

   Let $\{A, B, C \}$ be a system, and let $\Lambda=PAP^{-1}$ be a diagonalization of $A$.

   Let $PB=\tilde{B}$ and $PC=\tilde{C}$

   If $B \in \mathbb{R}^{1 \times n}$ and $C \in \mathbb{R}^{n \times 1}$ then all equivalent systems are of the form,

   $\tilde{A} = Q \tilde{\Lambda} Q^{-1}$ where $\tilde{\Lambda}$ is diagonal. Also if $\tilde{B}_{i}$ and $\tilde{C}_{i}$ are both $\neq 0$ then $\tilde{\Lambda}_{ii}=\Lambda_{ii}$.

   And $QB=B$ and $CQ=C$.

   \begin{itemize}
     \item Is this true?
     \item is this the minimal parameterization?
     \item how to generalize this to $k, l > 1$?
   \end{itemize}

   \section{Proof}
   Define $\mathcal{C}$ as the controllability matrix and $\mathcal{O}$ as the observability matrix.

   \begin{equation}
     \mathcal{C} := \begin{bmatrix} \tilde{B} & \tilde{\Lambda} \tilde{B} & \hdots & \tilde{\Lambda}^{n-1} \tilde{B} \end{bmatrix}
   \end{equation}

   \begin{equation}
     \mathcal{O} := \begin{bmatrix} \tilde{C} \\ \tilde{C} \tilde{\Lambda}  \\ \vdots \\ \tilde{C} \tilde{\Lambda}^{n-1} \end{bmatrix}
   \end{equation}

    If a system is both controllable and observable it is a minimal realization. A system is controllable and observable if and only if both rank($\mathcal{C}$) $=$ rank($\mathcal{O}$) $= n$.

    \begin{equation}
      \mathcal{C} = \underbrace{\begin{bmatrix} \tilde{B}_{1} & & 0 \\ & \ddots & \\ 0 & & \tilde{B}_{n} \end{bmatrix}}_{\beta} \underbrace{\begin{bmatrix} 1 & \lambda_{1} & \hdots & \lambda_{1}^{n-1} \\ \vdots & \vdots & \vdots & \vdots \\ 1 & \lambda_{n} & \hdots & \lambda_{n}^{n-1} \end{bmatrix}}_{L}
    \end{equation}

    By the Sylvester Rank Inequality: $\text{rank}(\beta) + \text{rank}(L) - n \leq \text{rank}(\beta L) \leq \min\{\text{rank}(\beta) , \text{rank}(L)\}$.

    Therefore rank$(\mathcal{C}) =$ rank$(\beta L) = \text{rank}(\beta)$, and rank$(\mathcal{O}) = \text{rank}(L^{T} K) = \text{rank}(K)$.

    The $\text{rank}(L) = n$ as long as all of the eigenvalues are unique. The $\text{rank}(\beta) = n$ and $\text{rank}(K) = n$ if and only if $\tilde{B}_{i} \neq 0 \ \forall i$ and $\tilde{C}_{i} \neq 0 \ \forall i$, respectively.

    \begin{equation}
      \mathcal{O} = \underbrace{\begin{bmatrix} 1 & \hdots & 1 \\ \lambda_{1} & \hdots & \lambda_{n} \\ \vdots & \vdots & \vdots \\ \lambda_{1}^{n-1} & \hdots & \lambda_{n}^{n-1} \end{bmatrix}}_{L^{T}} \underbrace{\begin{bmatrix} \tilde{C}_{1} & & 0 \\ & \ddots & \\ 0 & & \tilde{C}_{n} \end{bmatrix}}_{K}
    \end{equation}

   \begin{equation}
     G(s) = \sum_{i}^{n} \frac{\tilde{C}_{i} \tilde{B}_{i}}{(s-\lambda_{i})}
   \end{equation}

   \section{2-degree TF gene deletion}
   \begin{equation}
     G(s) = \frac{B_{1} C_{1}}{s-\lambda_{1}} + \frac{B_{2} C_{2}}{s-\lambda_{2}} = C \Big(sI - \mathcal{S}_{(\vec{r})} \begin{bmatrix} G & 0 \\ 0 & \Gamma \end{bmatrix}  \mathcal{S}^{-1}_{(\vec{r})} \Big)^{-1} B^{T}
   \end{equation}
   \begin{equation}
     \mathcal{S}_{(\vec{r})} = \begin{bmatrix} 1 & 0 & 0 & \hdots & 0 \\ r_{1,1} & 1-r_{1,1} & r_{1,2} & \hdots & r_{1,n} \\ r_{2,1} & -r_{2,1} & r_{2,2} & \hdots & r_{2,n} \\ r_{3,1} & -r_{3,1} & r_{3,2} & \hdots & r_{3,n} \\ \vdots & \vdots & \vdots & \vdots & \vdots \\ r_{n,1} & -r_{n,1} & r_{n,2} & \hdots & r_{n,n} \end{bmatrix} 
   \end{equation}
       \begin{equation}
      \begin{bmatrix} G & 0 \\ 0 & \Gamma \end{bmatrix} =  \begin{bmatrix} \lambda_{1} & \lambda_{2} - k & 0 & 0 & \hdots & 0 \\ 0 & \lambda_{2} & 0 & 0 & \hdots & 0 \\ 0 & 0 & \gamma_{1} & 0 & \hdots & 0 \\ \vdots & \vdots & 0 & \ddots & 0 & \vdots \\ 0 & 0 & \vdots & 0 & \ddots & 0 \\ 0 & 0 & 0 & 0 & \hdots & \gamma_{n-2} \end{bmatrix}
   \end{equation}

   For which set of $\vec{r}$ does $\vec{\epsilon}_{j} = \vec{0}$?

   \begin{equation}
     \begin{bmatrix} && \bigA && \vec{\epsilon_{1}} & \cdots & \vec{\epsilon_{n-2}} \\ &&&& \\ && \cdots && & \bigGamma && \end{bmatrix} 
   \end{equation}

  \section{Gene Deletion}
             What is the probability that,

             \begin{equation*}
                \Vert G(s) - G^{*}(s) \Vert \leq \delta
              \end{equation*}

              Where $G^{*}(s)$ is the transfer function after removing a row and corresponding column.

              For example we care about the $\bigstar$ region of the GRN matrix:

              \begin{equation*}
                Z = \begin{bmatrix} \bigstar &\vert \epsilon \\ \hdots & \hdots \end{bmatrix}
              \end{equation*}

              What is the probability that there exists a set of dimensions that, 
              \begin{itemize}
                \item we don't care about them (w.r.t. $C$). 
                \item and they only affect the other genes a little bit.
              \end{itemize}

              If this increases dimensions fast enough then there will be an equilibrium level of network complexity. (Note: this is also related to robustness).

              As GRNs increase in size, their mutational target size and therefore genetic load may also increase, pushing the size of GRNs downward. We could also look at organisms that closely approximate the optimal molecular dynamics. For example, if the optimal dynamics are represented by the transfer function ($G^{m}(s)$) of polynomial degree $m$, and a smaller degree TF (say $G^{m-1}(s)$) closely approximates the dynamics such that $\Vert G^{m}(s) - G^{m-1}(s) \Vert \leq \delta$, and the difference in fitness due to genetic load between a population using the $m$-degree TF vs. the $(m-1)$-degree polynomial is $\epsilon$. The smaller $\frac{\delta}{\epsilon}$, the more likely a population will risk using a smaller TF to approximate the optimal one. (This isn't always true though, because it is possible for a smaller degree polynomial to have a larger mutational target than the larger one, if the smaller one is using a very non-minimal representation of the GRN (the $A$ matrix).)

              I think there are going to be a lot of oppositional forces acting on GRN size. I need to spend more time thinking about the additional ones. (Would an infinite dimensional GRN be completely immune to the fitness cost of mutation?).

  \section{Brownian Motion}

    If $Q_{t} = I + \sum_{j} B_{j}(t)w_{j}w{j}^{T}$ and

    $\tilde{\Lambda}_{m,t} = W_{m}(t)$

    Then $\sum_{ij, kl} = \frac{d}{dt} \text{cov} [\tilde{A}_{ij}(t), \tilde{A}_{kl}(t)] \vert_{t=0}$

    If $\mathcal{B}$ is an $\eta$-dimensional (where $\eta$ consists of the free parameters in $P$ and eigenvalues $\lambda_{n+1}$ through $\lambda_{n+m}$) Brownian Motion with covariance matrix $\sigma^{2} I$ then $\mathbb{E}\left[ \Vert \mathcal{B} \Vert^{2} \right] = \eta \sigma^{2} t$, however, $A = F(\mathcal{B})$.

\end{document}

### 2017-01-25 kalman_walk.tex
starting writeup

--- 2017-01-24 attic/LTI_system_notes.tex
@@
-\documentclass[a4paper, 11 pt]{article}
-\usepackage{amsmath, amssymb, color, setspace, graphicx, dsfont, pdfpages, float, wrapfig, indentfirst}
-\usepackage[left=0.5in, right=0.5in]{geometry}
-\usepackage[font=small,labelfont=bf]{caption}
+\documentclass[11 pt]{article}
+\usepackage{amsmath, amssymb, color}
+\usepackage{graphicx}
+\usepackage{fullpage}
+\usepackage[hidelinks]{hyperref}
+
 \begin{document}
 
-\section{What are LTI systems?}
+% Outline:
+% 
+% - What the problems are:
+% 
+%     * evolution of perfectly fit populations
+% 
+% - Motivations: Can you infer function from structure? And...
+% 
+%     * motifs
+%     * speciation
+%     * network size
+%     * rube-goldberginess
+%     * redundancy, robustness
+%     * evolvability related to minimality?
+% 
+% - Give a concrete, simple example.
+% 
+% - Modeling framework (LTI systems)
+%
+%     * y = "phenotype"
+%     * x = "internal state" or "kryptotype"?
+%     * A = "regulatory matrix" or "genetic network"?
+%     * Kalman decomposition
+% 
+% - Evolution of free parameters in the Kalman decomposition
+%
+%     * Brownian motion on free parameters
+%     * how mutations map into this space
+%     * dimension: duplication, deletion, recruitment of new things
+% 
+% - (Compute things)
+% 
+% - (Allow non-fit stuff too.)
 
-A linear time invariant (LTI) dynamical system is a system of linear differential equations that describes a physical process.
-LTI systems are used in electrical and control engineering to model a myriad of phenomena including circuits.
+\newcommand{\plr}[1]{{\color{blue}\it #1}}
+% \newcommand{\ddt}{\frac{d}{dt}}
+\newcommand{\ddt}{\dot}
 
-One can use this methodology, under a set of assumptions, to reverse engineer a mechanism from impulse experiments (input/output data).
-The idea is that given a black box, and it's experimental manipulation, can we describe the internal mechanisms?
+\section{Background}
 
-The inpulse data is summarized by the transfer function $G(s) = C( sI - A )^{-1} B + D$, a Laplace transform of the state space equation of the system.
-While the transfer function description of a system can be unique, the state space representation is not always unique.
+\plr{Review of relevant other stuff.}
 
-The state space representation is as follows:
+A linear time invariant (LTI) dynamical system is a system of linear
+differential equations that describes a physical process. 
+LTI systems are used in electrical and control engineering to model a myriad of phenomena including
+circuits. 
+One can use this methodology, under a set of assumptions, to reverse engineer a
+mechanism from impulse experiments (input/output data). The idea is that given
+a black box, and it's experimental manipulation, can we describe the internal
+mechanisms?
 
-\begin{equation}
-  \begin{split}
-    \dot{x} =& Ax + Bu \\
-y =& Cx + Du
-\end{split}
-\end{equation}
+\section{The problem}
 
-Where $x \in \mathbb{R}^{n}, \ y \in \mathbb{R}^{q}, \ u \in \mathbb{R}^{p}, \ A \in \mathbb{R}^{n \times n}, B \in \mathbb{R}^{n \times p}, C \in \mathbb{R}^{q \times n}, D \in \mathbb{R}^{q \times p}$.
+Organisms often have to respond to their environments,
+and the way this happens is that external input triggers a cascade of molecular signals
+whose eventual result is a reaction,
+whose approprateness for the situation determines fitness.
+\plr{A similar situation occurs in development (...?).}
+However, there are many ways to skin a cat:
+especially given a number of possible molecular intermediaries,
+there may be a large number of ways to construct such a molecular signalling system
+that has precisely the \emph{same} input-output relationships,
+and hence the same function.
+Therefore, any mutation that changes one such system to another, equivalent version
+will be neutral,
+and may well become polymorphic in a population if there are no ill effects in heterozygotes.
+Certainly, mutations that move a system away from optimal functioning will also appear,
+but in very large populations will not rise to high frequency.
+\plr{Deleterious things are quite important but somehow justify starting with the neutral ones.}
 
-The state space representation is considered an internal description of the system, whereas the transfer function is an external description.
-$\vec{x}$ is generally considered to be a state of the system, $\vec{u}$ to be the input, and $\vec{y}$ to be the observed output.
+\plr{
+    More stuff on why we care about this:  
+Provide neutral model; can address questions.  
+Main point of paper.
+}
 
-Multiple systems can interact in serial,
+\paragraph{Example: Oscillator.}
+Motivate the system:
+\begin{align*}
+    A &= \left[\begin{matrix} 
+        0 & 1 \\ 
+       -1 & 0 
+    \end{matrix}\right]
+\end{align*}
+as a special case of the below.
 
-\begin{equation}
-  \begin{split}
-    \dot{x}_{1} &= A_{1}x_{1} + B_{1}u \\
-    \dot{x}_{2} &= A_{2}x_{2} + B_{2}C_{1}x_{1} \\
-    y &= Cx_{2}
-\end{split}
-\end{equation}
+\paragraph{Example: two-component systems.}
+As a more general example,
+suppose that a cell produces two molecular species, $S_1$ and $S_2$ 
+whose production are both stimulated by the presence of an external ``input'' substance.
+\plr{call them TF?}
+Suppose that the two species self-regulate with strengths $\lambda_1$ and $\lambda_2$, respectively,
+and the second species also regulates the first species with strength $\gamma$.
+\plr{Rephrase in terms of binding to promoter sequences?}
+However, only the time course of the concentration of
+the first one of these species determines the fitness of the organism.
+(As happens for instance in \plr{XXX}.)
+Write $x(t) = (x_1(t),x_2(t))$ for the concentrations of the two species at time $t$,
+and write $u(t)$ for the concentration at time $t$ of the input substance.
+Then, if the rates at which each species are produced
+are linear functions of the concentrations
+\plr{(look up way this is usually said in the literature.)},
+then the dynamics of the system are given by
+\begin{align*}
+    \ddt x_1(t) &= \lambda_1 x_1(t) + \gamma x_2(t) + u(t) \\
+    \ddt x_2(t) &= \lambda_2 x_1(t) + u(t) ,
+\end{align*}
+where $\ddt x$ denotes the time derivative.
+The initial conditions, $x_1(0)$ and $x_2(0)$, 
+and the input $u(t)$ then determine the concentrations through time.
+If we record the regulatory coefficients in the matrix:
+\begin{align*}
+    A &= \left[\begin{matrix} 
+        \lambda_1 & \gamma \\ 
+       0 & \lambda_2  
+    \end{matrix}\right],
+\end{align*}
+and define the column vector $B = [1,1]^T$,
+then in matrix notation the dynamics are
+\begin{align*}
+    \ddt x(t) &= A x(t) + B u(t) .
+\end{align*}
+If these are all transcription factors,
+and we suppose their binding motifs are fixed,
+then the $i^\text{th}$ row of $A$ is determined by the $i^\text{th}$ promoter sequence.
 
-or in parallel as well:
-\begin{equation}
-  \begin{split}
-    \dot{x}_{1} &= A_{1}x_{1} + B_{1}u \\
-    \dot{x}_{2} &= A_{2}x_{2} + B_{2}u \\
-    y &= C_{1}x_{1} + C_{2}x_{2}
-\end{split}
-\end{equation}
+\plr{Note that $B$ and $C$ are fixed here.}
 
-Let $P \in \mathbb{R}^{n \times n}$ be a non-singular matrix, and let $\dot{\overline{x}} = Px$, then
-\begin{equation}
-  \begin{split}
-  \dot{\overline{x}} &= \overline{A}\overline{x} + \overline{B}u \\
-  y &= \overline{C}\overline{x} + \overline{D}u
-  \end{split}
-\end{equation}
-be an algebraically equivalent (AE) ``realization'' of the system, where $\overline{A} = PAP^{-1}$, $\overline{B} = PB$, $\overline{C} = CP^{-1}$, and $\overline{D} = D$.
+Since the system is linear, the state of the system at any time
+is the superposition of its responses to all previous inputs.
+This implies that 
+if $G(t)$ is the transient response of the system to a unit impulse,
+$t$ units of time later,
+and $x(0) = 0$,
+then the time course of the concentration of the thing we care about, $x_1(t)$, 
+can be written as
+\begin{align} \label{eqn:ex_convolution}
+    x_1(t) &= \int_0^t G(t-s) u(s) ds .
+\end{align}
+However, it turns out that 
+\plr{look for simple demontration this is true}
+\begin{align*}
+    P_{p} = \left[\begin{matrix} 
+        1 & 0 \\ 
+        p & 1-p 
+    \end{matrix}\right],
+\end{align*}
+then for any $p \neq 1$, 
+if we replace the regulatory matrix $A$ with
+\begin{align*}
+    A(p) = P^{-1} A P,
+\end{align*}
+then the response of $x_1(t)$ to any particular input $u(t)$ is identical to the original system.
+(However, $x_2$ may well be different!)
 
-If multiple realizations of a system are algebraically equivalent, then they must also have identical transfer functions $G(s) = \overline{G}(s)$.
-However, the opposite relation is not necessarily true.
-Multiple systems can have identical transfer functions and not have algebraic equivlaence.
-In fact, two systems with identical transfer functions can even have different dimensions.
+For instance, setting $p=-1$,
+the system with
+\begin{align*}
+    A(-1) = \left[\begin{matrix} 
+        \lambda_1 + \gamma/2 & \gamma/2 \\ 
+        \lambda_2-\lambda_1-\gamma/2 & \lambda_2 - \gamma/2
+    \end{matrix}\right],
+\end{align*}
+looks very different, but gives the same input-output relationship.
+Although these systems are equivalent,
+hybrids between them may not be:
+\plr{do example}.
 
-Two LTI systems $\{A, B, C, D\}$ and $\{\overline{A}, \overline{B}, \overline{C}, \overline{D} \}$ have ``Zero-State Equivalence (ZSE)'' if they have the same transfer function.
-(AE $\Rightarrow$ ZSE).
+\section{Regulatory networks as linear, time-invariant systems}
+
+\plr{Make sure to have strongly worded disclaimer somewhere
+that we know real networks aren't linear
+but they are locally.
+Plus, it works in engineering.}
+
+% - General description of the LTI process in biological context
+%     * y = "phenotype"
+%     * x = "internal state" or "kryptotype"?
+%     * A = "regulatory matrix" or "genetic network"?
+% - Why transfer function determines input-output relationship
+% - Kalman decomposition in unique form
+
+We will now lay out the model in more general terms.
+Suppose that the \emph{internal state} of the system
+is parameterized by the concentrations of a collection of $n$ molecular species,
+$S_1, \ldots, S_n$,
+and the vector of concentrations at time $t$ we denote $x(t)=(x_1(t),\ldots,x_n(t))$.
+There are also $m$ ``input'' species, whose concentrations are determined
+exogenously to the system,
+and are denoted $u(t) = (u_1(t),\ldots,u_m(t))$,
+and $\ell$ ``output'' species, whose concentrations are denoted
+$y(t) = (y_1(t),\ldots,y_\ell(t))$.
+The output is merely a linear function of the internal state:
+\begin{align*}
+    y_i(t) = \sum_j C_{ij} x_i(t).
+\end{align*}
+Since $y$ is what natural selection acts on, we refer to it as the \emph{phenotype},
+and sometimes in contrast refer to $x$ as the \emph{kryptotype},
+as it is ``hidden'' from direct selection.
+\plr{Maybe.}
+The rate at which the $i^\text{th}$ species is produced
+is a weighted sum of the concentrations of the other species
+as well as the input:
+\begin{align*}
+    \ddt x_i(t) = \sum_j A_{ij} x_j(t) + \sum_k B_{ik} u_k(t) .
+\end{align*}
+In matrix notation, this is written more concisely as
+\begin{align*}
+    \ddt x(t) &= A x(t) + B u(t) \\
+    y(t) &= C x(t) .
+\end{align*}
+
+\plr{Everything we say I think is assuming $x(0)=0)$.  
+Figure this out. ``Zero-state equivalence.''}
+
+Given an initial condition and an input, 
+it is possible to write the solution $x(t)$
+as a convolution with an input-response kernel as in the example above
+(equation \eqref{eqn:ex_convolution}).
+An alternative way of describing the input-output relationship,
+more common in engineering,
+is instead to find the \emph{transfer function}
+\begin{align*}
+    H(z) = C(z I - A)^{-1} B,
+\end{align*}
+which is more mathematically readable \plr{if you know what it means}.
+This can be thought of as the system's response to input at ``frequency'' $z$.
+The fact that the transfer function uniquely determines the system's
+input-output relationship (i.e., the mapping $u \mapsto y$)
+follows from the fact that if $\tilde y(z) = \int_0^\infty e^{-zt} y(t) dt$
+is the Laplace transform of $x$,
+and $\tilde u(z)$ is the Laplace transform of $u$,
+then
+\begin{align*}
+    \tilde y(z) = H(z) \tilde u(z) .
+\end{align*}
+(This happens because the Laplace transform takes convolutions to products,
+and $H$ is the Laplace transform of the kernel $G$.)
+
+\plr{Should we remind what dimensions everything is?}
+
+The state space representation is considered an internal description of the
+system, whereas the transfer function is an external description. 
+``Realizations'' are something.
+
+State explicitly that same transfer function iff same input-output relationship.
+
+One way to get equivalent things is change of basis (``algebraic equivlance'').
+If multiple realizations of a system are algebraically equivalent, then they
+must also have identical transfer functions $G(s) = \overline{G}(s)$. However,
+the opposite relation is not necessarily true. Multiple systems can have
+identical transfer functions and not have algebraic equivlaence. In fact, two
+systems with identical transfer functions can even have different dimensions.
+
+But, that's not all: Kalman decomposition.  
+Theorem that all equivalent systems are of this form; 
+and representation is unique.
+
+Talk about minimal dimension: point out that are degrees of freedom 
+even in minimal dimension with B and C fixed.
+
+Give an example.
 
 \section{What biological question are we asking?}
+
 How do gene regulatory networks (GRNs) change through evolutionary time?
-Specifically, under constant selection and environmental pressures, how does a maximally fit population's GRN change and what molecular and population parameters are significant to the process?
-How frequent does gene network rewiring occur in the absense of genetic drift or adaptation?
-Is rewiring typically compensatory and ususally follow the fixation of a deleterious allele, or can networks neutrally reorganize?
-How does the size, average degree, function, peiotropy, etc of a network influence its likelihood to drift?
-If networks can drift, how much topological heterogeneity can a well-mixed population tolerate, if any?
-Does topoligical heterogeneity and/or the size of the neutral genotype-set, if consequences of network organization, constrain or entail exaptation and evolvability?
-Are some network organizations (maybe higher dimensional realizations?) able to more rapidly evolve to construct novel phenotypes and meet the demands of changes in selection pressures?
+Specifically, under constant selection and environmental pressures, how does a
+maximally fit population's GRN change and what molecular and population
+parameters are significant to the process? How frequent does gene network
+rewiring occur in the absense of genetic drift or adaptation? Is rewiring
+typically compensatory and ususally follow the fixation of a deleterious
+allele, or can networks neutrally reorganize? How does the size, average
+degree, function, peiotropy, etc of a network influence its likelihood to
+drift? If networks can drift, how much topological heterogeneity can a
+well-mixed population tolerate, if any? Does topoligical heterogeneity and/or
+the size of the neutral genotype-set, if consequences of network organization,
+constrain or entail exaptation and evolvability? Are some network organizations
+(maybe higher dimensional realizations?) able to more rapidly evolve to
+construct novel phenotypes and meet the demands of changes in selection
+pressures?
 
 \section{How can we mathematically model this question?}
-In some respect an organism can be thought of as a black box that responds to some set of inputs (it's environment and other initial conditions) and outputs a phenotype.
-The phenotype is then evaluated by natural selection.
-The black box metaphor holds, because much of the details of what happens to construct a phenotype are unimportant as far as selection is concered, so long as the phenotype reliably performs its function.
-If multiple internal mechanisms can map the same set of inputs to the same set of outputs, and these mechanisms are mutationally nearby (in genotype space), then evolution may drift from mechanism to mechanism.
 
-In this simple model we can say that $B\vec{u}$ is a list of initial GRN protein concentrations determined by the more general initial environmental conditions $\vec{u}$.
-$A$ represents all the genetic interactions of a particular GRN, and $C\vec{x}$ is an organism's phenotype.
-Fitness could be calculated by comparing a subset of the phenotype to an optimum.
-Maybe, $f(\cdot) = e^{- \int_{a}^{b} \Vert G(s) - G^{*}(s) \Vert ds}$ Alternatively, fitness scores could simply assess whether or not a phenotype is within some range or breaks some threshold by some time point.
-Or possbily the simplest: any phenotype not exactly equal to $G(s)$ is lethal, and $G(s)$ is perfectly fit.
+In some respect an organism can be thought of as a black box that responds to
+some set of inputs (it's environment and other initial conditions) and outputs
+a phenotype. The phenotype is then evaluated by natural selection. The black
+box metaphor holds, because much of the details of what happens to construct a
+phenotype are unimportant as far as selection is concered, so long as the
+phenotype reliably performs its function. If multiple internal mechanisms can
+map the same set of inputs to the same set of outputs, and these mechanisms are
+mutationally nearby (in genotype space), then evolution may drift from
+mechanism to mechanism.
 
-Next, we will have to define mutation and recombination.
-Mutation needs to add and remove genetic interactions and recombination needs to shuffle genes during sexual reproduction.
-We could adapt methods from Lynch 2007 and Lynch and Hagner 2015 to model the probability of a TFBS appearing or disappearing due to mutational pressure.
-Every time a new binding site is gained or lost, the values within the A matrix are either modified or rows and columns are added removed following a specific set of rules.
-I am not sure if it would be easier to maintain a matrix size (say $m \times m$) where $m$ is arbitrarily large, with  most entries being zero, or if a matrix should only have as many dimensions as the number of active TFs.
-Recombination should then shuffle rows of the $A$ matrix randomly, as each row represents the regulatory region of a given gene (assuming only cis-regulation, and that the regulatory element is small enough to not break up during recombination).
+In this simple model we can say that $B\vec{u}$ is a list of initial GRN
+protein concentrations determined by the more general initial environmental
+conditions $\vec{u}$. $A$ represents all the genetic interactions of a
+particular GRN, and $C\vec{x}$ is an organism's phenotype. Fitness could be
+calculated by comparing a subset of the phenotype to an optimum. Maybe,
+$f(\cdot) = e^{- \int_{a}^{b} \Vert G(s) - G^{*}(s) \Vert ds}$ Alternatively,
+fitness scores could simply assess whether or not a phenotype is within some
+range or breaks some threshold by some time point. Or possbily the simplest:
+any phenotype not exactly equal to $G(s)$ is lethal, and $G(s)$ is perfectly
+fit.
 
-To study some of the above posed questions, we can simulate (or analytically determine) the rates of GRN change, and how frequently these changes lead to significant rewiring or speciation.
-We would want to know the dynamics of a brownian motion over the set of all realizations.
-Another interesting question would be to determine the size of the neutral genotype space as well as the number of connections from the neutral genotype space to the non-neutral genotype space.
-Maybe higher dimensional realizations of a GRN will have significantly more connections to non-neutral genotypes and therefore be more ``evolvable'' or exapted to novel environments?
+Next, we will have to define mutation and recombination. Mutation needs to add
+and remove genetic interactions and recombination needs to shuffle genes during
+sexual reproduction. We could adapt methods from Lynch 2007 and Lynch and
+Hagner 2015 to model the probability of a TFBS appearing or disappearing due to
+mutational pressure. Every time a new binding site is gained or lost, the
+values within the A matrix are either modified or rows and columns are added
+removed following a specific set of rules. I am not sure if it would be easier
+to maintain a matrix size (say $m \times m$) where $m$ is arbitrarily large,
+with  most entries being zero, or if a matrix should only have as many
+dimensions as the number of active TFs. Recombination should then shuffle rows
+of the $A$ matrix randomly, as each row represents the regulatory region of a
+given gene (assuming only cis-regulation, and that the regulatory element is
+small enough to not break up during recombination).
 
-Let $M \in \mathbb{R}^{n \times n}$ be a least dimensional realization of $A$, and let $M'$ bean $(n+m \times n+m)$ matrix where $M'_{(i,j)} = M_{(i,j)} \ \text{for } i,j \leq n, \ \text{and } 0 \ \text{otherwise}$.
-Let $P$ be any $(n+m \times n+m)$ matrix with rank $n$.
-Then for any matrix $B$ such that $BP = PM'$, $B$ is a realization of $A$.
+To study some of the above posed questions, we can simulate (or analytically
+determine) the rates of GRN change, and how frequently these changes lead to
+significant rewiring or speciation. We would want to know the dynamics of a
+brownian motion over the set of all realizations.  Another interesting question
+would be to determine the size of the neutral genotype space as well as the
+number of connections from the neutral genotype space to the non-neutral
+genotype space. Maybe higher dimensional realizations of a GRN will have
+significantly more connections to non-neutral genotypes and therefore be more
+``evolvable'' or exapted to novel environments?
+
+Let $M \in \mathbb{R}^{n \times n}$ be a least dimensional realization of $A$,
+and let $M'$ bean $(n+m \times n+m)$ matrix where $M'_{(i,j)} = M_{(i,j)} \
+\text{for } i,j \leq n, \ \text{and } 0 \ \text{otherwise}$. Let $P$ be any
+$(n+m \times n+m)$ matrix with rank $n$. Then for any matrix $B$ such that $BP
+= PM'$, $B$ is a realization of $A$.
 

### 2017-01-26 kalman_walk.tex
josh edits to intro

@@
 \documentclass[11 pt]{article}
-\usepackage{amsmath, amssymb, color}
+\usepackage{amsmath, amssymb, color, xcolor}
 \usepackage{graphicx}
@@
 %
+% - Generalize ideas to nonlinear systems
+%
+%   * compute isomorphic or dynamically equivallent nonlinear systems of the feed-forward motif or of the three-node adaptation motif. 
+% 
 % - (Allow non-fit stuff too.)
 
+% is y the phenotype or H(z) the phenotype? Do we need new jargon anyway?
+
 \newcommand{\plr}[1]{{\color{blue}\it #1}}
+\newcommand{\jss}[1]{{\color{olive}\it #1}}
 % \newcommand{\ddt}{\frac{d}{dt}}
@@
 \plr{Review of relevant other stuff.}
+
+\jss{Introduction}
+
+It is commonly taught that an organism's genome contains the heritable material 
+that natural selection filters and that an organism's phenotype directly determines its
+evolutionary fitness. Between genotype and phenotype is an often complicated and poorly
+undrstood molecular machinery -- and it is a major goal common to many different disciplines
+within the life sciences to elucidate its form, function, and evolution. These aims are
+delicately intertwined and a comphrehensive understanding of a system's evolution requires
+an understanding of its function and vice versa.
+
+The molecular machinery, interacting with the environment, and bridging genotype to phenotype
+can be mathematically described as a dynamical system (or a system of differential equations).
+Thus, an idealized study of evolution would benefit from knowledge of genomes, phenotypes,
+and importantly detailed descriptions of genotype-phenotype maps. Movement in this direction
+is ongoing, as researhers have begun to study the evolution of empirically inspired
+computational and mathematical models of gene regulatory networks (GRNs) and metabolic
+networks that incorporate empirical data such as sequences and expression patterns. If we
+allow the reasonable assumption that the genotype-phenotype map can be represented as a
+system of differential equations, we can immediately discuss its evolution and function
+in a much more mechanistic, yet general, manner.
+
+In some fields that seek to fit parametric models to experimental data, such as control
+theory, chemical engineering, and statistics, it is well known that mathematical models can
+fundamentally be \emph{unidentifiable} and/or \emph{indistinguishable} -- meaning that there
+can be uncertaintity about an inferred model's parameters or even its claims about causal
+structure, even with access to complete and perfect data. Models with different parameter
+schemes, or even different mechanics can be equally accurate, but still not \emph{actually}
+agree with what is being modelled. Although this may frusturate the occasional engineer or
+scientist, viewed from another angle, the concepts of unidentifiability and
+indistinguishability can provide a starting point for thinking about externally equivalent
+systems -- systems that evolution can explore, so long as the parameters and structures can
+be realized biologically.
+
+The set of all systems -- that is all model structures and all parameter schemes -- that
+are externally, or phenomenologically, equivalent, biologically realizable, and mutationally accessible,
+
+What are typically referred to as ``design principles'' or some sort of clear structure-function relationships are commonly saught
+
+Seeking to attribute functionality, uninformed by evolution, can lead to spurious claims
+(i.e. see Graur).
 
@@
 \begin{align*}
-    A(p) = P^{-1} A P,
+    A(p) = P A P^{-1},
 \end{align*}
@@
 In matrix notation, this is written more concisely as
-\begin{align*}
+\begin{align} \label{eqn:lti_system}
     \ddt x(t) &= A x(t) + B u(t) \\
     y(t) &= C x(t) .
-\end{align*}
+\end{align}
 
@@
 is instead to find the \emph{transfer function}
-\begin{align*}
+\begin{align} \label{eqn:transfer_fn}
     H(z) = C(z I - A)^{-1} B,
-\end{align*}
+\end{align}
 which is more mathematically readable \plr{if you know what it means}.
@@
 
-One way to get equivalent things is change of basis (``algebraic equivlance'').
-If multiple realizations of a system are algebraically equivalent, then they
-must also have identical transfer functions $G(s) = \overline{G}(s)$. However,
-the opposite relation is not necessarily true. Multiple systems can have
-identical transfer functions and not have algebraic equivlaence. In fact, two
-systems with identical transfer functions can even have different dimensions.
+\begin{definition}[Phenomenological equivalence of systems]
+    Let $(x(t),y(t))$ and $(\bar x(t),\bar y(t))$ be the solutions to \eqref{eqn:lti_system}
+    with coefficient matrices $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ respectively,
+    and both $x(0)$ and $\bar x(0)$ are zero. 
+    The systems defined by $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ are
+    \textbf{phenomenologically equivalent} if
+    \begin{align*}
+        y(t) = \bar y(t) \qquad \text{for all} t \ge 0.
+    \end{align*}
+    Equivalently, this occurs if and only if
+    \begin{align*}
+        H(z) = \bar H(z)  \qquad \text{for all} z \ge 0,
+    \end{align*}
+    where $H$ and $\bar H$ are the transfer functions of the two systems.
+\end{definition}
 
+One way to find other systems equvalent to a given one
+is by change of coordinates (``algebraic equivalance''):
+if $T$ is an invertible matrix, then the systems $(A,B,C)$ and $(TAT^{-1},TB,CT^{-1})$
+have the same dyanamics because their transfer functions are equal:
+\begin{align*}
+    CT^{-1}( zI - TAT^{-1})^{-1}TB
+    =
+    CT^{-1}T( zI - A)^{-1}T^{-1}TB
+    =
+    C( zI - A)^{-1}B .
+\end{align*}
+However, the converse is not necessarily true: 
+systems can have identical transfer functions without being changes of coordinates of each other.
+In fact, systems with identical transfer functions can involve interactions between different
+numbers of molecular species.
+
+The general problem of how to characterize the set of equivalent
 But, that's not all: Kalman decomposition.  
@@
 even in minimal dimension with B and C fixed.
+
+B and C fixed.
 

### 2017-01-27 kalman_walk.tex
josh edited intro

@@
 
+\newcommand{\plr}[1]{{\color{blue}\it #1}}
+\newcommand{\jss}[1]{{\color{olive}\it #1}}
+% \newcommand{\ddt}{\frac{d}{dt}}
+\newcommand{\ddt}{\dot}
+\newcommand{\ro}{{ro}}
+\newcommand{\nro}{{\bar{r}o}}
+\newcommand{\rno}{{r\bar{o}}}
+\newcommand{\nrno}{{\bar{r}\bar{o}}}
+\newcommand{\reachable}{\mathcal{R}}
+\newcommand{\unobservable}{\bar{\mathcal{O}}}
+\newcommand{\R}{\mathbb{R}}
+\newcommand{\E}{\mathbb{E}}
+\renewcommand{\P}{\mathbb{P}}
+
+\DeclareMathOperator{\spn}{span}
+
+\newtheorem{theorem}{Theorem}
+\newtheorem{lemma}{Lemma}
+\newtheorem{definition}{Definition}
+\newtheorem{example}{Example}
+
 \begin{document}
+
+$\unobservable$
 
@@
 %
+%     * Give or refer to earlier simple description of population dynamics which makes the system evolve.
+%     * What do we want to know from this model:
+%
+%       - How fast do the internal dynamics change?
+%       - Stationary distribution on network complexity/shape/properties
+%       - How fast does incompatibility accumulate?
+%
 %     * Brownian motion on free parameters
@@
 % is y the phenotype or H(z) the phenotype? Do we need new jargon anyway? 
-
-\newcommand{\plr}[1]{{\color{blue}\it #1}}
-\newcommand{\jss}[1]{{\color{olive}\it #1}}
-% \newcommand{\ddt}{\frac{d}{dt}}
-\newcommand{\ddt}{\dot}
+% 
+% Deleterious mutations: can we describe heterosis?
 
@@
 can be mathematically described as a dynamical system (or a system of differential equations).
-Thus, an idealized study of evolution would benefit from knowledge of genomes, phenotypes,
-and importantly detailed descriptions of genotype-phenotype maps. Movement in this direction
-is ongoing, as researhers have begun to study the evolution of empirically inspired
-computational and mathematical models of gene regulatory networks (GRNs) and metabolic
-networks that incorporate empirical data such as sequences and expression patterns. If we
-allow the reasonable assumption that the genotype-phenotype map can be represented as a
-system of differential equations, we can immediately discuss its evolution and function
-in a much more mechanistic, yet general, manner.
+Previously it has been argued that any idealized study of evolution is incomplete without
+a mathematically sufficient description of the genotype, phenotype, and transformation from
+one to the other. Movement in this direction is ongoing, as researhers have begun to study 
+the evolution of empirically inspired computational and mathematical models of gene regulatory networks (GRNs) and metabolic networks that incorporate empirical data such as sequences and expression patterns.
+If we allow the reasonable assumption that the genotype-phenotype map can be represented as a system of differential equations, we can immediately discuss its evolution and function in a much more mechanistic, yet general, manner.
 
 In some fields that seek to fit parametric models to experimental data, such as control
-theory, chemical engineering, and statistics, it is well known that mathematical models can
-fundamentally be \emph{unidentifiable} and/or \emph{indistinguishable} -- meaning that there
-can be uncertaintity about an inferred model's parameters or even its claims about causal
-structure, even with access to complete and perfect data. Models with different parameter
-schemes, or even different mechanics can be equally accurate, but still not \emph{actually}
-agree with what is being modelled. Although this may frusturate the occasional engineer or
-scientist, viewed from another angle, the concepts of unidentifiability and
-indistinguishability can provide a starting point for thinking about externally equivalent
-systems -- systems that evolution can explore, so long as the parameters and structures can
-be realized biologically.
+theory, chemical engineering, and statistics, it is well known that mathematical models
+can fundamentally be \emph{unidentifiable} and/or \emph{indistinguishable} -- meaning that 
+there can be uncertaintity about an inferred model's parameters or even its claims about
+causal structure, even with access to complete and perfect data. Models with different 
+parameter schemes, or even different mechanics can be equally accurate, but still not
+\emph{actually} agree with what is being modelled. In control theory, where electrical 
+circuits and mechanical systems are often the focus, it is understood that there can be an 
+infinite number of ``realizations,'' or ways to reverse engineer the dynamics of a black box,
+even if all possible input and output experiments on the black box are performed. In chemical
+engineering, those who study chemical reaction networks sometimes refer to the fundamental
+unidentifiability of these networks as ``the fundamental dogma of chemical kinetics.''
+Although this may frusturate the occasional engineer or scientist, viewed from another angle,
+the concepts of unidentifiability and indistinguishability can provide a starting point for
+thinking about externally equivalent systems -- systems that evolution can explore, so long
+as the parameters and structures can be realized biologically. In fact, evolutionary
+biologists who study homology and analogy are very familiar with such functional symmetries; macroscopically identical phenotypes in closely related species can be divergent at the molecular level.
 
+In this paper we propose a framework to study the evolution of biological systems.
+To begin, we focus on the evolution of an idealized population.
+We consider the evolution of a perfectly adapted, large population, evolving in a static environment for an infinite number of generations.
+Under these ideal circumstances, we expect to observe a ``conservation of phenotype,'' where the population explores the manifold of phenotypically-invariant (or symmetric) genetic and developmental architecutres.
+We would like to understand which parameters influence the distribution of a population along the manifold of phenotypically invariant genetic systems.
+Further, we can show how dispersion along this manifold contributes to speciation and evolvability.
+
+\plr{phenomenologically or dynamically equivalent?}
 The set of all systems -- that is all model structures and all parameter schemes -- that
@@
 
-\paragraph{Example: Oscillator.}
+\begin{example}[Example: Oscillator.]
 Motivate the system:
@@
 
-\paragraph{Example: two-component systems.}
+\end{example}
+
+\begin{example}[Example: two-component systems.] \label{ex:2x2}
 As a more general example,
@@
 
-\plr{Note that $B$ and $C$ are fixed here.}
+\end{example}
 
@@
 follows from the fact that if $\tilde y(z) = \int_0^\infty e^{-zt} y(t) dt$
-is the Laplace transform of $x$,
+is the Laplace transform of $y$,
 and $\tilde u(z)$ is the Laplace transform of $u$,
@@
 and $H$ is the Laplace transform of the kernel $G$.)
+Concretely,
 
@@
 
-State explicitly that same transfer function iff same input-output relationship.
-
 \begin{definition}[Phenomenological equivalence of systems]
@@
     The systems defined by $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ are
-    \textbf{phenomenologically equivalent} if
+    \textbf{phenomenologically equivalent} 
+    if
     \begin{align*}
-        y(t) = \bar y(t) \qquad \text{for all} t \ge 0.
+        y(t) = \bar y(t) \qquad \text{for all} \; t \ge 0.
     \end{align*}
@@
     \begin{align*}
-        H(z) = \bar H(z)  \qquad \text{for all} z \ge 0,
+        H(z) = \bar H(z)  \qquad \text{for all} \; z \ge 0,
     \end{align*}
@@
 
-The general problem of how to characterize the set of equivalent
-But, that's not all: Kalman decomposition.  
-Theorem that all equivalent systems are of this form; 
-and representation is unique.
+The set of all systems phenomenologically equivalent to a given system $(A,B,C)$ 
+is elegantly described using the Kalman decomposition,
+which also clarifies the system dynamics? tells us a lot about how it works? \plr{or something}
+To motivate this, first note that the the input $u(t)$ only directly pushes the system
+in directions lying in the span of the columns of $B$.
+As a result, different combinations of input can 
+move the system in any direction that lies in the \emph{reachable subspace},
+which we denote by $\reachable$,
+and is defined to be the closure of $\spn(B)$ under applying $A$
+(or equivalently, the span of $B, AB, A^2B, \ldots A^{n-1}B$).
+Analogously to this, we define
+the \emph{observable subspace}, $\mathcal{O}$,
+to be the closure of $\spn(C^T)$ under applying $A$.
+(Or: $\unobservable$ is the largest $A$-invariant subspace
+contained in the null space of $C$;
+and $\reachable$ is the largest $A$-invariant subspace contained in the image of $B$.)
 
-Talk about minimal dimension: point out that are degrees of freedom 
-even in minimal dimension with B and C fixed.
+If we define
+\begin{enumerate}
+    \item The columns of $P_\rno$ are an orthonormal basis for $\reachable \cap \unobservable$.
+    \item The columns of $P_\ro$ are an orthonormal basis of
+        the complement of $\reachable \cap \unobservable$ in $\reachable$.
+    \item The columns of $P_\nro$ are an orthonormal basis of
+        the complement of $\reachable \cap \unobservable$ in $\unobservable$.
+    \item The columns of $P_\nrno$ are an orthonormal basis of
+        the the remainder of $\R^n$.
+\end{enumerate}
+If we then define
+\begin{align*}
+    P &= 
+    \left[ \begin{array}{c|c|c|c}
+        P_\rno & P_\ro & P_\nro & P_\nrno
+    \end{array} \right] ,
+\end{align*}
+then
+\begin{align*}
+    P^T P
+    &=
+    \left[ \begin{array}{c|c|c|c}
+        I & 0 & 0 & 0 \\
+        \hline
+        0 & I & U & 0 \\
+        \hline
+        0 & V & I & 0 \\
+        \hline
+        0 & 0 & 0 & I 
+    \end{array} \right] .
+\end{align*}
+\plr{Check this.  Can we get $U=V=0$?}
 
-B and C fixed.
+The following theorem can be found in SOME REFERENCE.
 
-Give an example.
+\begin{theorem}[Kalman decomposition] \label{thm:kalman}
+        For any system $(A,B,C)$ with corresponding Kalman basis matrix $P$,
+        the transformed system $(PAP^{-1},PB,CP^{-1})$  has the following form:
+        \begin{align*}
+            \widehat A = PAP^{-1}
+            &=
+            \left[ \begin{array}{cccc}
+                A_{\rno} & A_{\rno,\ro} & A_{\rno,\nrno} & A_{\rno,\nro} \\
+                0 & A_{\ro} & 0 & A_{\ro,\nro} \\
+                0 & 0 & A_{\nrno} & A_{\nrno,\nro} \\
+                0 & 0 & 0 & A_{\nro}
+            \end{array} \right] ,
+        \end{align*}
+        and
+        \begin{align*}
+            \widehat B = PB
+            &=
+            \left[ \begin{array}{cccc}
+                B_{\rno} \\
+                B_{\ro} \\
+                0 \\
+                0 
+            \end{array} \right] ,
+        \end{align*}
+        and
+        \begin{align*}
+            \widehat C = CP^{-1}
+            &=
+            \left[ \begin{array}{cccc}
+                0 & C_{\ro} & C_{\nrno} & 0 
+            \end{array} \right] .
+        \end{align*}
+        The transfer function of both systems is given by
+        \begin{align*}
+            H(z) = C_{\ro} ( zI - A_{\ro} )^{-1} B_{\ro} .
+        \end{align*}
+\end{theorem}
+
+In the latter case, we say that the system is \emph{minimal} 
+-- there is no equivalent system with a smaller number of species.
+Note that this says that any two equivalent minimal systems
+are changes of basis of each other.
+
+Since any system can be put into this form,
+and once in this form, its transfer function is determined only by 
+$C_{\ro}$, $A_{\ro}$, and $B_\ro$,
+therefore, the set of all equivalent systems are parameterized by
+the dimension $n$,
+the choice of basis ($P$),
+the remaining submatrices in $\widehat A$, $\widehat B$, and $\widehat C$
+(which are unconstrained),
+and a invertible transformation of $\spn(P_{\ro})$, which we call $T_\ro$.
+
+\begin{theorem}[Parameterization of equivalent systems]
+    Let $(A,B,C)$ be a minimal system.
+    \begin{itemize}
+        \item[(a)]
+            Every equivalent system is of the form given in Theorem \ref{thm:kalman},
+            i.e., can be specified by choosing a dimension, $n$;
+            submatrices in $\widehat A$, $\widehat B$, and $\widehat C$ 
+            except for $A_\ro=A$, $B_\ro=B$, and $C_\ro=C$;
+            and choosing an invertible matrix $P$.
+
+        \item[(b)]
+            \plr{conjecture:}
+            The parameterization is unique
+            if $P$ is futhermore chosen so that 
+            each $P_x$ other than $P_\ro$ is a projection matrix,
+            and that 
+            \begin{align*}
+                0
+                =
+                P_x^T P_y
+            \end{align*}
+            for all $(x,y)$ except $(\ro,\nrno)$.
+
+    \end{itemize} 
+\end{theorem}
+
+\plr{Another way of saying it: pick the $\reachable$ and $\unobservable$ subspaces,
+that must intersect in something of the minimal dimension;
+then let $P$ be the appropriate basis?}
+
+In some situations we may be interested in only ``network rewiring'',
+where $A$ changes while $B$ and $C$ do not.
+For instance, 
+if all non-regulatory functions of each molecule are strongly constrained,
+then $C$ cannot change.
+Likewise, if responses of each molecule to the external inputs are not changed by evolution,
+then $B$ does not change.
+
+\begin{lemma}
+    Let $(A,B,C)$ be a system.
+    Every matrix $\widehat A$ such that $(\widehat A,B,C)$ is equivalent to $(A,B,C)$
+    can be specified in the form given in Theorem \ref{thm:kalman},
+    with \plr{them having the same $\ro$ bit}
+    and satisfying
+    \begin{align*}
+        P^{-1} \widehat B &= B \\
+        \widehat C P &= C
+    \end{align*}
+\end{lemma}
+
+It is remarkable to note that even with the relationships 
+between environment, cryptotype, and phenotype constant,
+and in the minimal dimension,
+there are still almost always degrees of freedom.
+These correspond to distinct genetic networks
+that perform indistinguishable functions.
+For example, the equivalent systems of Example \ref{ex:2x2} above
+are minimal, and share common $B$ and $C$ matrices.
+
+\plr{Maybe say that stability to noise can be determined by $A_\nro$ (and others?),
+while complexity of the internal, unobserved dynamics
+are somehow given by the difference between the dimension of $y$ and the dimension of $A_\ro$.}
 

### 2017-01-27 kalman_walk.tex
.

@@
 %
-% Deleterious mutations: can we describe heterosis?
 
@@
     \begin{align*}
-        P^{-1} \widehat B &= B \\
+        \widehat B &= P B \\
         \widehat C P &= C
@@
 are somehow given by the difference between the dimension of $y$ and the dimension of $A_\ro$.}
+
+\plr{General argument why Brownian motion on set of equivalent things is maybe ok.}
+
+\begin{example}
+    The system in Example \ref{ex:2x2} is minimal, so $A=A_\ro$,
+    and the only free parameter is $P$.
+    Suppose also that the effects of the input and output are strongly constrained,
+    so that as in that example,
+    $B$ and $C$ cannot change,
+    and neutral evolution moves through the set of of $P$ such that $PB=B$ and $CP=C$.
+    \plr{Clarify we mean with no gene duplication and only neutral drift.}
+    How will this system change over evolutionary time in the manner described \plr{above}?
+    An unbiased Brownian motion on this set 
+    can be written in terms of a one-dimensional Brownian motion $\beta_\tau$ 
+    with $\beta_0=0$ as
+    \begin{align*}
+        P(\tau) = \left[
+            \begin{array}{cc}
+                1 & 0 \\
+                e^{\beta_\tau}-1 & e^{\beta_\tau}
+            \end{array}
+            \right] .
+    \end{align*}
+    Therefore, the state of the system after (evolutionary) time $\tau$
+    is $(P(\tau)AP(\tau)^{-1},B,C)$.
+\end{example}
 
@@
 
+\section{Notes on what to do when allowing deleterious mutations}
+
+Epistasis, coming from mutations deleterious in some networks but not in others
+(come from incompatibilities).
+
+Describe heterosis.
+
 \end{document}

### 2017-01-29 kalman_walk.tex
edits

@@
 \usepackage[hidelinks]{hyperref}
+\usepackage{tikz}
+\usetikzlibrary{arrows.meta}
+\usepackage{caption, subcaption}
 
@@
 
-$\unobservable$
+%$\unobservable$
 
@@
 % 
+\title{Mathematical Gene Network Evolution}
+\maketitle
+
+\begin{abstract}
+Abstract.
+\end{abstract}
 
@@
 
-\begin{example}[Example: Oscillator.]
+\begin{example}[Example: Oscillator.] \label{ex:osc}
 Motivate the system:
@@
 \end{align*}
+
+\begin{tikzpicture}
+\begin{scope}[every node/.style={circle,thick,draw}]
+  \node (A) at (0,0) {$x_{1}$};
+    \node[dashed] (B) at (3,0) {$x_{2}$};
+    \node[shape=rectangle, dashed] (U) at (1.5,-2) {$u(t)$};
+\end{scope}
+
+\begin{scope}[>={Stealth[black]},
+              every node/.style={fill=white,circle},
+              every edge/.style={draw=black, thick}]
+    \path [->, >=Rectangle] (A) edge[bend left] node {\tiny $-1$} (B);
+    \path [->] (B) edge[bend left] node {\tiny $1$} (A); 
+    \path[->] (U) edge[dashed] node {\tiny $1$} (A);
+    \path[->] (U) edge[dashed] node {\tiny $1$} (B);
+\end{scope}
+\begin{scope}[>={Stealth[black]},
+              every edge/.style={draw=black, thick}]
+    %\path [->] (A) edge[loop left] node {\tiny $\lambda_{1}$} (A);
+    %\path [->] (B) edge[loop left] node {\tiny $\lambda_{2}$} (B);
+\end{scope}
+
+\end{tikzpicture}
+
 as a special case of the below.
@@
 \plr{do example}.
+
+\begin{example}[Hyrbid Inviability]
+  Let, \begin{align*}
+    A(0) =\left[\begin{matrix}
+      0 & 1 \\
+      -1 & 0
+    \end{matrix}\right]
+    \text{and }
+    A(2) = \left[\begin{matrix}
+      2 & -1 \\
+      5 & -2
+    \end{matrix}\right]
+  \end{align*}
+  from Example \ref{ex:osc} ($B=\begin{bmatrix} 1 \\ 1 \end{bmatrix}$ and $C=\begin{bmatrix} 1 & 0 \end{bmatrix}$). 
+If hybrids are formed by exchanging genes (swapping rows), the hybrid's gene network is,
+  \begin{align*}
+    \text{hybrid network } := \mathcal{H}(A(0), A(2)) = \left[\begin{matrix}
+      2 & -1 \\
+      -1 & 0
+    \end{matrix}\right] ,
+  \end{align*}
+  \jss{ Or should it be: \begin{align*}
+    \text{hybrid network } := \text{mean}(A(0), A(2)) = \left[\begin{matrix}
+      1 & 0 \\
+      2 & -1
+    \end{matrix}\right] ,
+  \end{align*} }
+  It's transfer function will be $H(z, \mathcal{H}) = \frac{z-2}{z^2 - 3z +1}$, which produces very different dynamics than its parents, $H(z, A(0)) = H(z, A(2)) = \frac{z+1}{z^2 + 1}$.
+The time dynamcs of molecular species $1$ in the hybrid is $y^{(\mathcal{H})} =e^{t} \cosh(\sqrt{2} t)$ \jss{ which is approx.
+$\approx e^{t}$}, with the concentration of species $1$ rapidly heading towards infinity.
+The dynamics of species $1$ in the parental networks is $y = \sin(t) + \cos(t)$, where it continuously oscillates.
+
+  \jss{The transfer function $H(z, \mathcal{H})$, by averaging networks, is $\frac{1}{z-1}$ and the time dynamics are: $y^{(\mathcal{H})} = e^{t}$.}
+\end{example}
+
+\begin{figure}
+  \centering
+  \begin{subfigure}{0.5\textwidth}
+    \centering
+\begin{tikzpicture}
+\begin{scope}[every node/.style={circle,thick,draw}]
+  \node (A) at (0,0) {$x_{1}$};
+    \node[dashed] (B) at (0,-4) {$x_{2}$};
+    \node[shape=rectangle, dashed] (U) at (3,-2) {$u(t)$};
+\end{scope}
+
+\begin{scope}[>={Stealth[black]},
+              every node/.style={fill=white,circle},
+              every edge/.style={draw=black, thick}]
+    %\path [->] (A) edge[bend left] node {$c$} (B);
+    \path [->] (B) edge node {\tiny $\gamma$} (A); 
+    \path[->] (U) edge[dashed] node {\tiny $1$} (A);
+    \path[->] (U) edge[dashed] node {\tiny $1$} (B);
+\end{scope}
+\begin{scope}[>={Stealth[black]},
+              every edge/.style={draw=black, thick}]
+    \path [->] (A) edge[loop left] node {\tiny $\lambda_{1}$} (A);
+    \path [->] (B) edge[loop left] node {\tiny $\lambda_{2}$} (B);
+\end{scope}
+
+\end{tikzpicture}
+    \caption{Gene Regulatory Network $A(0)$.}
+  \end{subfigure}%
+  \begin{subfigure}{0.5\textwidth}
+    \centering
+\begin{tikzpicture}
+\begin{scope}[every node/.style={circle,thick,draw}]
+  \node (A) at (0,0) {$x_{1}$};
+    \node[dashed] (B) at (0,-4) {$x_{2}$};
+    \node[shape=rectangle, dashed] (U) at (3,-2) {$u(t)$};
+\end{scope}
+
+\begin{scope}[>={Stealth[black]},
+              every node/.style={fill=white,circle},
+              every edge/.style={draw=black, thick}]
+    \path [->, sloped] (A) edge[bend left] node {\tiny $(\lambda_{2} - \lambda_{1} - \frac{\gamma}{2})$} (B);
+    \path [->, sloped] (B) edge[bend left] node {\tiny $\frac{\gamma}{2}$} (A); 
+    \path[->] (U) edge[dashed] node {\tiny $1$} (A);
+    \path[->] (U) edge[dashed] node {\tiny $1$} (B);
+\end{scope}
+\begin{scope}[>={Stealth[black]},
+              every edge/.style={draw=black, thick}]
+    \path [->] (A) edge[loop left] node {\tiny $(\lambda_{1} + \frac{\gamma}{2})$} (A);
+    \path [->] (B) edge[loop left] node {\tiny $(\lambda_{2}-\frac{\gamma}{2})$} (B);
+\end{scope}
+
+\end{tikzpicture}
+    \caption{Gene Regulatory Network $A(-1)$.}
+  \end{subfigure}
+  \caption{Neutral rewiring of gene network $A(p)$ from Example \ref{ex:2x2}. Not only are the regulatory coefficients different between $A(0)$ and $A(-1)$, there is also a new regulatory connection.}
+\end{figure}
 

### 2017-01-30 kalman_walk.tex
biblio stuff.
notes

@@
 \usepackage{tikz}
-\usetikzlibrary{arrows.meta}
+\usetikzlibrary{arrows.meta, shapes}
 \usepackage{caption, subcaption}
+\usepackage{natbib} % gives us \citet: Author (year) and \citep: (Author; year)
 
@@
 
-%$\unobservable$
+\paragraph{To do, soon:}
+
+\begin{enumerate}
+    \item Add biology to introduction. (keep an eye out for concrete numbers we can use!)
+
+    \item Do some order-of-magnitude calculation to figure out how fast systems should drift.
+
+    \item Write out simple examples in biological words.
+
+    \item Rephrase hybrid inviability example in biological terms.  Discuss F1's (or heterozygotes) versus F2's.
+
+    \item Clarify our assumptions around $x(0)$.
+
+    \item Note we don't include $D$ for simplicity but don't.
+
+    \item Figure out how to specify $P$ for Kalman decomposition.
+
+    \item Prove or find citation for uniqueness of minimal system up to basis change.
+
+    \item Work examples (minimal; not minimal; with/without B,C fixed) of Kalman decomposition and description of moving along the walk.
+
+    \item Write down and decide between options for parameterizing the random walk on the neutral manifold:
+        use Kalman decomposition, and parameterize Brownian motion so that it has constant speed;
+        OR consider Josh's parameteriziation like this;
+        OR just say we do Brownian motion on the manifold (more or less implicit characterization).
+
+    \item Consider options for fitness function: Euclidean? Sign of eigenvalues?  Time domain or frequency?
+        First try on simple concrete examples.
+
+    \item Describe model leading to constraints on regulation so we have a stationary distribution.
+        Ask questions of stationary distribution: robustness, connectivity, xxx?
+
+    \item Write something in discussion about nonlinear systems.  Do an example if there is a feasible one.
+
+\end{enumerate}
 
@@
 
+%%% Questions:
+
 % is y the phenotype or H(z) the phenotype? Do we need new jargon anyway?
-% 
+
 \title{Mathematical Gene Network Evolution}
@@
 
+    \begin{figure}
+        \begin{center}
             \begin{tikzpicture}
@@
                 \node[dashed] (B) at (3,0) {$x_{2}$};
-    \node[shape=rectangle, dashed] (U) at (1.5,-2) {$u(t)$};
+                \node[shape=rectangle] (U) at (1.5,2) {$u$};
+                \node[shape=rectangle] (y) at (1.5,-2) {$y$};
             \end{scope}
@@
                 \path[->] (U) edge[dashed] node {\tiny $1$} (B);
+                \path[->] (A) edge[dashed,bend right] node {\tiny $1$} (y);
             \end{scope}
@@
             \end{tikzpicture}
+        \end{center}
+        \caption{
+            Diagram of Example \ref{ex:osc} in the text.  
+            \plr{explain what arrows mean if nec}
+            \label{fig:oscillator_diagram}}
+    \end{figure}
 
@@
 
+Distribution of dominance coefficients.
+
+\bibliographystyle{plainnat}
+\bibliography{krefs}
+
 \end{document}

### 2017-01-31 kalman_walk.tex
added more references and changed the title.
a few more things to the future list

@@
         Ask questions of stationary distribution: robustness, connectivity, xxx?
+
+    \item Figure out coherent model for changing dimension: do coefficients stay stuck at zero? In what parameterization? 
+        Do we need to allow deleterious variation to get deletion?
 
@@
 
-Epistasis, coming from mutations deleterious in some networks but not in others
+\begin{enumerate}
+
+    \item Epistasis, coming from mutations deleterious in some networks but not in others
 (come from incompatibilities).
 
-Describe heterosis.
+    \item Describe heterosis.
 
-Distribution of dominance coefficients.
+    \item Distribution of dominance coefficients.
+
+    \item Evolvability: easier for bigger networks to adapt to a new environment?
+
+    \item Do short-term responses to strong directional selection have deleterious epistatic consequences in other traits?
+
+\end{enumerate}
 

### 2017-01-31 kalman_walk.tex
assigned tasks

@@
 \usepackage{natbib} % gives us \citet: Author (year) and \citep: (Author; year)
-
+\usepackage{authblk}
 \newcommand{\plr}[1]{{\color{blue}\it #1}}
@@
 \begin{enumerate}
-    \item Add biology to introduction. (keep an eye out for concrete numbers we can use!)
+
+    \item \emph{(josh)} Add biology to introduction. (keep an eye out for concrete numbers we can use!)
 
     \item Do some order-of-magnitude calculation to figure out how fast systems should drift.
+        \begin{itemize}
+            \item \emph{(peter)} Work out the math for how to do this.
+            \item \emph{(josh)} Find real parameters to put into this.
+        \end{itemize}
 
-    \item Write out simple examples in biological words.
+    \item \emph{(josh)} Write out simple examples in biological words.
 
-    \item Rephrase hybrid inviability example in biological terms.  Discuss F1's (or heterozygotes) versus F2's.
+    \item \emph{(josh)} Rephrase hybrid inviability example in biological terms.  Discuss F1's (or heterozygotes) versus F2's.
 
-    \item Clarify our assumptions around $x(0)$.
+    \item \emph{(josh)} Clarify our assumptions around $x(0)$.
 
-    \item Note we don't include $D$ for simplicity but don't.
+    \item \emph{(josh)} Note we don't include $D$ for simplicity but don't.
 
-    \item Figure out how to specify $P$ for Kalman decomposition.
+    \item \emph{(peter)} Figure out how to specify $P$ for Kalman decomposition.
 
-    \item Prove or find citation for uniqueness of minimal system up to basis change.
+    \item \emph{(josh,peter)} Prove or find citation for uniqueness of minimal system up to basis change.
 
     \item Work examples (minimal; not minimal; with/without B,C fixed) of Kalman decomposition and description of moving along the walk.
+        \begin{itemize}
+            \item \emph{(josh)} Work out parameterization.
+            \item \emph{(peter)} Describe Brownian motion.
+        \end{itemize}
 
-    \item Write down and decide between options for parameterizing the random walk on the neutral manifold:
+    \item \emph{(peter)} Write down and decide between options for parameterizing the random walk on the neutral manifold:
         use Kalman decomposition, and parameterize Brownian motion so that it has constant speed;
@@
 
+    \item \emph{(peter)} Describe model leading to constraints on regulation so we have a stationary distribution.
+        Ask questions of stationary distribution: robustness, connectivity, xxx?
+
     \item Consider options for fitness function: Euclidean? Sign of eigenvalues?  Time domain or frequency?
         First try on simple concrete examples.
-
-    \item Describe model leading to constraints on regulation so we have a stationary distribution.
-        Ask questions of stationary distribution: robustness, connectivity, xxx?
 
@@
     \item Write something in discussion about nonlinear systems.  Do an example if there is a feasible one.
+
+    \item Simulate.
 
@@
 
-\title{Mathematical Gene Network Evolution}
+\title{The Evolution of Phenotypically Invariant Genetic Networks}
+\author[1]{Joshua S. Schiffman}
+\author[1,2]{Peter L. Ralph}
+\affil[1]{University of Southern California, Los Angeles, California}
+\affil[2]{University of Oregon, Eugene, Oregon}
+\affil[ ]{\texttt{jsschiff@usc.edu} \qquad \texttt{plr@uoregon.edu}}
 \maketitle
@@
                 \node[dashed] (B) at (3,0) {$x_{2}$};
-                \node[shape=rectangle] (U) at (1.5,2) {$u$};
-                \node[shape=rectangle] (y) at (1.5,-2) {$y$};
+                \node[shape=rectangle] (U) at (1.5,2) {input};
+                \node[shape=rectangle] (y) at (1.5,-2) {output};
             \end{scope}
@@
   \node (A) at (0,0) {$x_{1}$};
-    \node[dashed] (B) at (0,-4) {$x_{2}$};
-    \node[shape=rectangle, dashed] (U) at (3,-2) {$u(t)$};
+    \node[dashed] (B) at (4,0) {$x_{2}$};
+    \node[shape=rectangle] (U) at (2,2) {input};
+    \node[shape=rectangle] (y) at (2,-2) {output};
 \end{scope}
@@
     \path[->] (U) edge[dashed] node {\tiny $1$} (B);
+    \path[->] (A) edge[dashed, bend right] node {\tiny $1$} (y);
 \end{scope}
@@
     \path [->] (A) edge[loop left] node {\tiny $\lambda_{1}$} (A);
-    \path [->] (B) edge[loop left] node {\tiny $\lambda_{2}$} (B);
+    \path [->] (B) edge[loop right] node {\tiny $\lambda_{2}$} (B);
 \end{scope}
@@
     \path [->] (A) edge[loop left] node {\tiny $(\lambda_{1} + \frac{\gamma}{2})$} (A);
-    \path [->] (B) edge[loop left] node {\tiny $(\lambda_{2}-\frac{\gamma}{2})$} (B);
+    \path [->] (B) edge[loop right] node {\tiny $(\lambda_{2}-\frac{\gamma}{2})$} (B);
 \end{scope}

### 2017-01-31 kalman_walk.tex
added the abstract and a definition for speciation.

--- 2017-01-30 kalman_walk.tex
@@
 \begin{abstract}
-Abstract.
+Bridging the gulf between an organism's genome and phenotype is a poorly understood and complex molecular machinery.
+Progress in a suite of biological subdisciplines is stalled by our general lack of understanding of this molecular machinery: with respect to both its function and evolution.
+There does exist a growing body of experiment and data on the evolutionary histories and molecular characterizations of particular gene regulatory networks, as well as thoughtfull verbal and conceptual models.
+However, as Hardy and Weinberg taught us over a century ago, verbal theories are often insufficient, if not downright misleading.
+This is especially pertinent given the staggering complexity and scope of contemporary research programmes.
+This outlook necessitates the advancement of conceptual frameworks of such precision, only mathematics will suffice.
+Here we outine an analytical theory to study the evolution of biological systems, borrowing insight and methods from control engineering, systems identificaiton, and dynamical systems theory.
+Presently, we focus on the neutral evolution of genetic regulatory networks.
+That is, we analytically describe the set of all linear gene networks (of any size) that produce identical phenotypes -- and the evolutionary paths connecting them.
+In the idealized case of a perfectly adapted population, constant selection, and a static environment, we observe evolution under the ``conservation of phenotype'' as a Brownian motion over phenotypically-invariant network-space.
+This analysis provides insight to the mechanisms and parameters important for understanding developmental systems drift, network rewiring, evolvability, epistasis, and speciation, as well as the tenuous connection between network architecture and function.
+
 \end{abstract}
@@
 
+\begin{definition}[Biological Species Concept]
+  Let $\mathcal{A}$ be the set of all phenotypically symmetric gene networks with transfer funtion $H(z)$. 
+  Choose $A$ and $\widetilde A$, both in $\mathcal{A}$. 
+  If $\mathcal{H}(A, \widetilde A) \notin \mathcal{A}$ then $A$ and $\widetilde A$ are distinct species. 
+\end{definition}
+
 \begin{figure}

### 2017-02-02 kalman_walk.tex
added abstract and definition for speciation.

@@
 \begin{abstract}
-Bridging the gulf between an organism's genome and phenotype is a poorly understood and complex molecular machinery.
-Progress in a suite of biological subdisciplines is stalled by our general lack of understanding of this molecular machinery: with respect to both its function and evolution.
-There does exist a growing body of experiment and data on the evolutionary histories and molecular characterizations of particular gene regulatory networks, as well as thoughtfull verbal and conceptual models.
-However, as Hardy and Weinberg taught us over a century ago, verbal theories are often insufficient, if not downright misleading.
-This is especially pertinent given the staggering complexity and scope of contemporary research programmes.
-This outlook necessitates the advancement of conceptual frameworks of such precision, only mathematics will suffice.
-Here we outine an analytical theory to study the evolution of biological systems, borrowing insight and methods from control engineering, systems identificaiton, and dynamical systems theory.
-Presently, we focus on the neutral evolution of genetic regulatory networks.
-That is, we analytically describe the set of all linear gene networks (of any size) that produce identical phenotypes -- and the evolutionary paths connecting them.
-In the idealized case of a perfectly adapted population, constant selection, and a static environment, we observe evolution under the ``conservation of phenotype'' as a Brownian motion over phenotypically-invariant network-space.
-This analysis provides insight to the mechanisms and parameters important for understanding developmental systems drift, network rewiring, evolvability, epistasis, and speciation, as well as the tenuous connection between network architecture and function.
-
+I will outline an analytical theory to study the evolution of biological systems such as gene regulatory networks, borrowing insight and tools from control engineering, systems identification, and dynamical systems theory.
+I will describe a null model of regulatory network evolution by analytically describing the set of all linear gene networks (of any size) that produce identical phenotypes -- and the evolutionary paths connecting them.
+In the idealized case of a perfectly adapted population, constant selection, and a static environment, we observe neutral evolution as a random walk over the phenotypically- invariant network-space.
+Under neutral conditions, this model can provide descriptions of expected network size and connectivity under mutation-selection equilibrium, estimate the rate of regulatory rewiring, and the rates at which Dobzhansky-Muller incompatibilities arise in reproductively isolated populations.
+This analysis provides insight into the mechanisms and parameters important for understanding developmental systems drift, network rewiring, evolvability, epistasis, and speciation, as well as the tenuous connection between network architecture and function.
 \end{abstract}
@@
 \jss{Introduction}
+
+Bridging the gulf between an organism's genome and phenotype is a poorly understood and complex molecular machinery.
+Progress in a suite of biological subdisciplines is stalled by our general lack of understanding of this molecular machinery: with respect to both its function and evolution.
+There does exist a growing body of experiment and data on the evolutionary histories and molecular characterizations of particular gene regulatory networks, as well as thoughtful verbal and conceptual models.
+However, as Hardy and Weinberg taught us over a century ago, verbal theories are often insufficient, if not downright misleading.
+This is especially pertinent given the staggering complexity and scope of contemporary research programs.
+This outlook necessitates the advancement of conceptual frameworks of such precision, only mathematics will suffice.
+Here we outline an analytical theory to study the evolution of biological systems, borrowing insight and methods from control engineering, systems identification, and dynamical systems theory.
+Presently, we focus on the neutral evolution of genetic regulatory networks.
+That is, we analytically describe the set of all linear gene networks (of any size) that produce identical phenotypes -- and the evolutionary paths connecting them.
+In the idealized case of a perfectly adapted population, constant selection, and a static environment, we observe evolution under the ``conservation of phenotype'' as a Brownian motion over phenotypically-invariant network-space.
+This analysis provides insight to the mechanisms and parameters important for understanding developmental systems drift, network rewiring, evolvability, epistasis, and speciation, as well as the tenuous connection between network architecture and function.
 

### 2017-02-02 kalman_walk.tex

@@
 
+    \item \emph{(peter)} Describe model leading to constraints on regulation so we have a stationary distribution.
+        Ask questions of stationary distribution: robustness, connectivity, xxx?
+
     \item Consider options for fitness function: Euclidean? Sign of eigenvalues?  Time domain or frequency?
@@
 
-    \item Describe model leading to constraints on regulation so we have a stationary distribution.
-        Ask questions of stationary distribution: robustness, connectivity, xxx?
+    \item Figure out coherent model for changing dimension: do coefficients stay stuck at zero? In what parameterization? 
+        Do we need to allow deleterious variation to get deletion?
 
     \item Write something in discussion about nonlinear systems.  Do an example if there is a feasible one.
+
+    \item Simulate.
 

### 2017-02-03 kalman_walk.tex
started to add citations to the intro.

@@
 \usepackage{fullpage}
-\usepackage[hidelinks]{hyperref}
+\usepackage[backref, hidelinks, colorlinks=true, citecolor=blue!60!black!100]{hyperref}
 \usepackage{tikz}
@@
 Progress in a suite of biological subdisciplines is stalled by our general lack of understanding of this molecular machinery: with respect to both its function and evolution.
-There does exist a growing body of experiment and data on the evolutionary histories and molecular characterizations of particular gene regulatory networks, as well as thoughtful verbal and conceptual models.
-However, as Hardy and Weinberg taught us over a century ago, verbal theories are often insufficient, if not downright misleading.
+There does exist a growing body of experiment and data on the evolutionary histories and molecular characterizations of particular gene regulatory networks \citep{jaeger2011gap, davidson2006gene, israel2016comparative}, as well as thoughtful verbal and conceptual models \citep{true2001developmental, gwagner1}.
+However, as Hardy and Weinberg taught us over a century ago, verbal theories are often insufficient, if not downright misleading \citep{hardy1908mendelian, weinberg1908vererbungsgesetze}.
 This is especially pertinent given the staggering complexity and scope of contemporary research programs.
 This outlook necessitates the advancement of conceptual frameworks of such precision, only mathematics will suffice.
-Here we outline an analytical theory to study the evolution of biological systems, borrowing insight and methods from control engineering, systems identification, and dynamical systems theory.
-Presently, we focus on the neutral evolution of genetic regulatory networks.
+Previously it has been suggested that any idealized study of evolution is incomplete without a mathematically sufficient description of the genotype, phenotype, and transformation from one to the other \citep{Lewontin1974genetic}.%Here we outline an analytical theory to study the evolution of biological systems, borrowing insight and methods from control engineering, systems identification, and dynamical systems theory.
+
+%Presently, we focus on the neutral evolution of genetic regulatory networks.
 That is, we analytically describe the set of all linear gene networks (of any size) that produce identical phenotypes -- and the evolutionary paths connecting them.
@@
 
-It is commonly taught that an organism's genome contains the heritable material 
-that natural selection filters and that an organism's phenotype directly determines its
-evolutionary fitness. Between genotype and phenotype is an often complicated and poorly
-undrstood molecular machinery -- and it is a major goal common to many different disciplines
-within the life sciences to elucidate its form, function, and evolution. These aims are
-delicately intertwined and a comphrehensive understanding of a system's evolution requires
-an understanding of its function and vice versa.
+%It is commonly taught that an organism's genome contains the heritable material 
+%that natural selection filters and that an organism's phenotype directly determines its
+%evolutionary fitness. Between genotype and phenotype is an often complicated and poorly
+%undrstood molecular machinery -- and it is a major goal common to many different disciplines
+%within the life sciences to elucidate its form, function, and evolution. These aims are
+%delicately intertwined and a comphrehensive understanding of a system's evolution requires
+%an understanding of its function and vice versa.
 
 The molecular machinery, interacting with the environment, and bridging genotype to phenotype
-can be mathematically described as a dynamical system (or a system of differential equations).
-Previously it has been argued that any idealized study of evolution is incomplete without
-a mathematically sufficient description of the genotype, phenotype, and transformation from
-one to the other. Movement in this direction is ongoing, as researhers have begun to study 
-the evolution of empirically inspired computational and mathematical models of gene regulatory networks (GRNs) and metabolic networks that incorporate empirical data such as sequences and expression patterns.
+can be mathematically described as a dynamical system -- or a system of differential equations\citep{jaeger2015comet}.
+ Movement in this direction is ongoing, as researchers have begun to study 
+the evolution of both abstract \citep{wagner1994evolution, wagner1996does,  siegal2002waddington, bergman2003evolutionary} and empirically inspired computational and mathematical models of gene regulatory networks (GRNs) \citep{mjolsness1991connectionist, jaeger2004dynamic, maria1, vitaly1, vitaly2, crombach2016gap, wotton2015quantitative}.
 If we allow the reasonable assumption that the genotype-phenotype map can be represented as a system of differential equations, we can immediately discuss its evolution and function in a much more mechanistic, yet general, manner.
@@
 there can be uncertaintity about an inferred model's parameters or even its claims about
-causal structure, even with access to complete and perfect data. Models with different 
+causal structure, even with access to complete and perfect data \citep{bellman1970structural, grewal1976identifiability, walter1984structural}. Models with different 
 parameter schemes, or even different mechanics can be equally accurate, but still not
@@
 infinite number of ``realizations,'' or ways to reverse engineer the dynamics of a black box,
-even if all possible input and output experiments on the black box are performed. In chemical
+even if all possible input and output experiments on the black box are performed \citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. In chemical
 engineering, those who study chemical reaction networks sometimes refer to the fundamental
-unidentifiability of these networks as ``the fundamental dogma of chemical kinetics.''
+unidentifiability of these networks as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}.
+In computer science, this is framed as the relationship among processes that simulate one another \citep{van2004equivalence}.
 Although this may frusturate the occasional engineer or scientist, viewed from another angle,
@@
 as the parameters and structures can be realized biologically. In fact, evolutionary
-biologists who study homology and analogy are very familiar with such functional symmetries; macroscopically identical phenotypes in closely related species can be divergent at the molecular level.
+biologists who study homology and analogy are very familiar with such functional symmetries; macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level \citep{true2001developmental, tsong2006evolution, hare2008sepsid, vierstra2014mouse, stergachis2014conservation, taylor2016diverse, matsui2015regulatory}.
 
@@
 
-What are typically referred to as ``design principles'' or some sort of clear structure-function relationships are commonly saught
+What are typically referred to as ``design principles'' or some sort of clear structure-function relationships are commonly saught \citep{milo2002network, alon2007network, ma2009defining}.
 
 Seeking to attribute functionality, uninformed by evolution, can lead to spurious claims
-(i.e. see Graur).
+(i.e. see Graur \citep{lynch2007evolution, lynch2007frailty, graur2013immortality}).
 
@@
 \plr{A similar situation occurs in development (...?).}
-However, there are many ways to skin a cat:
+However, there are many ways to skin a cat \jss{(there are also many ways to develop a cat)}:
 especially given a number of possible molecular intermediaries,

### 2017-02-03 kalman_walk.tex
added a citation; minor edits

@@
 \usepackage{authblk}
+
 \newcommand{\plr}[1]{{\color{blue}\it #1}}
@@
 There does exist a growing body of experiment and data on the evolutionary histories and molecular characterizations of particular gene regulatory networks \citep{jaeger2011gap, davidson2006gene, israel2016comparative}, as well as thoughtful verbal and conceptual models \citep{true2001developmental, gwagner1}.
-However, as Hardy and Weinberg taught us over a century ago, verbal theories are often insufficient, if not downright misleading \citep{hardy1908mendelian, weinberg1908vererbungsgesetze}.
+However, as Hardy and Weinberg taught us over a century ago, verbal theories are often insufficient, if not downright misleading \citep{hardy1908mendelian, weinberg1908vererbungsgesetze, servedio2014not}.
 This is especially pertinent given the staggering complexity and scope of contemporary research programs.
@@
 Plus, it works in engineering.}
+
+\jss{Maybe we should discuss the Hartman-Grobman Theorem?}
 

### 2017-02-06 kalman_walk.tex
added citation for minimal realization proof.

@@
 \usepackage{fullpage}
-\usepackage[backref, hidelinks, colorlinks=true, citecolor=blue!60!black!100]{hyperref}
+\usepackage[backref=page, hidelinks, colorlinks=true, citecolor=blue!60!black!100]{hyperref}
 \usepackage{tikz}
@@
 
-    \item \emph{(josh)} Add biology to introduction. (keep an eye out for concrete numbers we can use!)
+  \item \emph{(josh)} Add biology to introduction. (keep an eye out for concrete numbers we can use!) \jss{added some relevant biology citations to introduction. Still needs a lot of refinement.}
 
@@
 
-    \item \emph{(josh,peter)} Prove or find citation for uniqueness of minimal system up to basis change.
+    \item \emph{(josh \jss{\checkmark},peter)} Prove or find citation for uniqueness of minimal system up to basis change.
+\jss{See \citep[Chapter 10.6, Theorem 6.9: Fundamental Theorem of Linear Realization Theory.]{kalman1969topics}}
 

### 2017-02-07 kalman_walk.tex
added some notes on parameters/calculations helpful for computing DSD rates.

@@
 
-  \item \emph{(josh)} Add biology to introduction. (keep an eye out for concrete numbers we can use!) \jss{added some relevant biology citations to introduction. Still needs a lot of refinement.}
+  \item \emph{(josh)} Add biology to introduction.
+(keep an eye out for concrete numbers we can use!) \jss{added some relevant biology citations to introduction.
+Still needs a lot of refinement.} \ref{sec:intro}
 
@@
             \item \emph{(josh)} Find real parameters to put into this. 
+              \jss{added some notes for TFBS gain and loss calculations from Lynch} \ref{sec:param}
         \end{itemize}
@@
 
-\section{Background}
+\section{Background} \label{sec:intro}
 
@@
 
-\section{What biological question are we asking?}
+\section{Estimates of network rewiring and developmenal systems drift.} \label{sec:param}
+
+\jss{Parameter estimates for DSD $\downarrow$}
+
+What parameters (and other relevant factors) do we need from the literature and experiment? 
+  \begin{itemize}
+    \item mutation rate
+    \item Transcription factor binding site appearence and loss
+    \begin{itemize}
+      \item In \citep{lynch2007evolution} TF binding site gain ($u_{g}$) and loss ($u_{l}$) is estimated to be,
+        \begin{align*}
+          u_{g} &= \frac{(L-n+1)nu}{4^{n}} \\
+          u_{l} &= nu
+        \end{align*}
+      Where $L$ is the length of the promoter, $n$ is the number of nucleotides in a TFBS, and $u$ is the mutation rate per base pair per generation. 
+    \item \citet{lynch2007evolution} further states that in prokaryotes $L$ will usually be $\leq 100$ and in eukaryotes $L$ can range from $10^{3}$ to $10^{6}$.
+He also cites \citep{harbison2004transcriptional} that TFBSs range from $5-10$bps in length.
+    \item in \citep{lynch2014evolutionary}, they estimate
+        \begin{align*}
+          \tilde{P}(m) = C \left[ 3^{\ell - m}  \left( \begin{array}{c} \ell \\ m \end{array} \right) \right] e^{2N_{e}s(m)}  ,
+        \end{align*}
+        where $\tilde{P}(m)$ is the stationary distribution of a TFBS with match (and strength) $m$.
+$C$ is the normalizer constant, $m$ is the number of matches, $\ell$ is the length of the TFBS motif.
+The term on the right is the ratio of fixation probabilities.
+        Or in their own words, there are two interpretations: 
+        \begin{quote} ``(i) For a single TFBS, it represents the long-term proportion of evolutionary time spent in the various matching states and (ii) for a set of different TFBSs under the same selective constraints, it represents the expected distribution of states at any point in time.'' \end{quote}
+      \item \citet{tuugrul2015dynamics, anderson2015intermolecular} also may be helpful.
+    \end{itemize}
+    \item rate of both \emph{trans-} and \emph{cis-} regulatory evolution. (see Sergey's papers).
+      i.e. \citep{nuzhdin2004common, fear2016buffering, graze2009regulatory, graze2012allelic, genissel2008cis}
+    \item The PWN-score landscape (does binding strength increase linearly, exponentially, have a threshold, etc.) with respect to SNPs.
+    \item Others? 
+  \end{itemize}
+
+\section{(Old Notes) What biological question are we asking?}
 
@@
 
-\section{How can we mathematically model this question?}
+\section{(Old Notes) How can we mathematically model this question?}
 
@@
 
-\section{List of papers with some notes.}
+\section{(Old Notes) List of papers with some notes.}
 

### 2017-02-07 kalman_walk.tex
added a section on deleterious stuff and a temporary fitness function.

@@
 \usepackage{amsmath, amssymb, color, xcolor}
-\usepackage{graphicx}
+\usepackage{graphicx, wrapfig, float, caption}
 \usepackage{fullpage}
@@
 \renewcommand{\P}{\mathbb{P}}
+\newcommand{\pda}{\frac{\partial}{\partial A_{ij}}}
 
@@
 
+\section{Including Deleterious Mutations and a Fitness Function}
+
+  \begin{minipage}{0.5\textwidth}
+  At the moment we focus strictly on the case where the minimal realization is in $2$ dimensions and both of its eigenvalues are real and negative (so a special case of the system from Example \ref{ex:2x2}).
+\jss{(Note: this is not to say the ideas below will not be generalizable).}
+
+    The system of interest has the form,
+  \begin{align*}
+    \Sigma := \left\{ A = \begin{bmatrix} A_{11} & A_{12} \\ A_{21} & A_{22} \end{bmatrix}, B = \begin{bmatrix} 1 \\ 1 \end{bmatrix}, C = \begin{bmatrix} 1 & 0 \end{bmatrix} \right\}
+  \end{align*}
+  \end{minipage}%
+  \begin{minipage}{0.5\textwidth}
+    \centering
+    \includegraphics[width=0.95\textwidth]{figure_2x2mutation01}%
+      \captionof{figure}{The transfer function for $A = \left[ \protect\begin{array}{cc} -3 & 1 \\ 0 & -1 \protect\end{array} \right]$ and all eight $\mu(\Sigma, A_{ij})$ mutations, where $\upsilon = \pm 0.5$}\label{fig:mut}%
+  \end{minipage}
+
+Define the fitness function to be,
+
+\begin{align}\label{eq:fitness}
+  \Phi(\bar{\Sigma}) := \left\{ \begin{array}{ll}   e^{- | H(0) - \tilde{H}(0) | = - \left| \frac{k}{\lambda_{1} \lambda_{2}} - \frac{\tilde{k}}{\tilde{\lambda}_{1} \tilde{\lambda}_{2}} \right|}    &\text{if } \lambda_{1,2} < 0 \text{ and } \in \mathbb{R}, \ \det(A) \neq 0 \\
+  0   &\text{if } \lambda_{1,2} \geq 0 \ \text{ or } \ \notin \mathbb{R}, \ \det(A) = 0 \end{array} \right.
+\end{align}
+and let $\upsilon$ be the magnitude and direction (sign) of a mutation on regulatory interaction strength.
+The fitness function (\ref{eq:fitness}) may me appropriate if mutations only slightly alter the dynamics of the system, as in Figure \ref{fig:mut}.
+
+Further let $\mu(\Sigma, A_{ij})$ be resulting gene regulatory network following a mutation of matrix element $A_{ij}$ to system $\Sigma$ (which contains matrix $A$).
+
+  \begin{align*}
+    \Phi(\mu(\Sigma, A_{ij} \upsilon)) = \exp\left(- \left| \frac{A_{22}-A_{12}}{\lambda_{1} \lambda_{2}} - \frac{k + \upsilon \pda (A_{22}-A_{12})}{\lambda_{1} \lambda_{2} + \text{adj}(A)^{T}_{ij} \upsilon} \right|\right)   .
+  \end{align*}
+For instance, the fitness of an organism with a change of one of its entries $A_{ij}$ (by $\upsilon$) is listed,
+  \begin{align*}
+    \Phi(\mu(\Sigma, A_{11})) &= \exp \left(- \left| \frac{A_{22}-A_{12}}{\det(A)} - \frac{A_{22}-A_{12}}{\det(A) + \upsilon A_{22}} \right| \right) \\
+    \Phi(\mu(\Sigma, A_{12})) &= \exp \left(- \left| \frac{A_{22}-A_{12}}{\det(A)} - \frac{A_{22}-A_{12} - \upsilon}{\det(A) - \upsilon A_{21}} \right| \right) \\
+    \Phi(\mu(\Sigma, A_{21})) &= \exp \left(- \left| \frac{A_{22}-A_{12}}{\det(A)} - \frac{A_{22}-A_{12}}{\det(A) - \upsilon A_{12}} \right| \right) \\
+    \Phi(\mu(\Sigma, A_{22})) &= \exp \left(- \left| \frac{A_{22}-A_{12}}{\det(A)} - \frac{A_{22}-A_{12} + \upsilon}{\det(A) + \upsilon A_{11}} \right| \right)
+  \end{align*}
+
 \section{Notes on what to do when allowing deleterious mutations}

### 2017-02-09 kalman_walk.tex
rewrote a fitness function.

@@
 \usepackage{amsmath, amssymb, color, xcolor}
-\usepackage{graphicx, wrapfig, float, caption}
+\usepackage{graphicx, wrapfig, float, caption, dsfont}
 \usepackage{fullpage}
@@
 \newcommand{\pda}{\frac{\partial}{\partial A_{ij}}}
+\newcommand{\ind}{\mathds{1}}
 
@@
     \item Consider options for fitness function: Euclidean? Sign of eigenvalues?  Time domain or frequency?
-        First try on simple concrete examples.
+      First try on simple concrete examples. \jss{see \ref{sec:del}}
 
@@
 
-\section{Including Deleterious Mutations and a Fitness Function}
+\section{Including Deleterious Mutations and a Fitness Function}\label{sec:del}
 
@@
 \begin{align}\label{eq:fitness}
-  \Phi(\bar{\Sigma}) := \left\{ \begin{array}{ll}   e^{- | H(0) - \tilde{H}(0) | = - \left| \frac{k}{\lambda_{1} \lambda_{2}} - \frac{\tilde{k}}{\tilde{\lambda}_{1} \tilde{\lambda}_{2}} \right|}    &\text{if } \lambda_{1,2} < 0 \text{ and } \in \mathbb{R}, \ \det(A) \neq 0 \\
+  \Phi(\bar{\Sigma}) := \left\{ \begin{array}{ll}   e^{- \left| H(0) - \tilde{H}(0) \right| = - \left| \frac{k}{\lambda_{1} \lambda_{2}} - \frac{\tilde{k}}{\tilde{\lambda}_{1} \tilde{\lambda}_{2}} \right|}    &\text{if } \lambda_{1,2} < 0 \text{ and } \in \mathbb{R}, \ \det(A) \neq 0 \\
   0   &\text{if } \lambda_{1,2} \geq 0 \ \text{ or } \ \notin \mathbb{R}, \ \det(A) = 0 \end{array} \right.
@@
 
-Further let $\mu(\Sigma, A_{ij})$ be resulting gene regulatory network following a mutation of matrix element $A_{ij}$ to system $\Sigma$ (which contains matrix $A$).
+Further let $\mu(\Sigma, A_{ij})$ be the resulting gene regulatory network following a mutation of matrix element $A_{ij}$ to system $\Sigma$ (which contains matrix $A$).
 
@@
 
+ The transfer function at $z=0$ for any GRN with mutations is, 
+  \begin{align*}
+    H(0, \begin{bmatrix} A_{11} + \upsilon & A_{12} + \varepsilon \\ A_{21} + \delta & A_{22} + \omega \end{bmatrix}) = \frac{A_{22} - A_{12} + \omega - \varepsilon}{ \det(A) + \omega A_{11} - \delta A_{12} - \varepsilon A_{21} + \upsilon A_{22} + \omega \upsilon - \varepsilon \delta }
+  \end{align*}
+
+%  \begin{align*}
+%    \frac{\left(A_{22}-A_{12}\right) + \left( \displaystyle\sum_{\jmath = 1}^{\ind_{\omega}} \omega_{\jmath} - \displaystyle\sum_{\jmath = 1}^{\ind_{\varepsilon}} \varepsilon_{\jmath} \right)}{\displaystyle\det(A) + \displaystyle\sum_{\jmath=1}^{\ind_{\omega}} \omega_{\jmath} A_{11} - \sum_{\jmath=1}^{\ind_{\delta}} \delta_{\jmath} A_{12} -  \sum_{\jmath=1}^{\ind_{\varepsilon}} \varepsilon_{\jmath} A_{21} + \sum_{\jmath=1}^{\ind_{\upsilon}} \upsilon_{\jmath} A_{22} + \sum_{\jmath=1}^{\ind_{\omega}} \omega_{\jmath} \sum_{\jmath=1}^{\ind_{\upsilon}} \upsilon_{\jmath} - \sum_{\jmath=1}^{\ind_{\varepsilon}} \varepsilon_{\jmath} \sum_{\jmath=1}^{\ind_{\delta}} \delta_{\jmath} }
+%  \end{align*}
+
+  If we want to know how new mutations fix in the population, we can explicitly show all mutational paths that could compensate for fixed mutations at sites $A_{ij}$.
+In a population, not at a phenotypic optimum, we are interested in the combination of mutations with varying magnitudes and locations that are most likely to fix.
+
+  \begin{align*}
+    \frac{\left(A_{22}-A_{12}\right) + \left( \displaystyle\sum_{\jmath = 1}^{\ind_{\omega}} \omega_{\jmath} - \displaystyle\sum_{\jmath = 1}^{\ind_{\varepsilon}} \varepsilon_{\jmath} \right)}{\det(A) + \displaystyle\sum_{\mu=\{\omega , \delta , \varepsilon , \upsilon \}}^{\ind_{\mu}} \mu_{\jmath} (-1)^{i+j} A_{ij} + \sum_{\jmath=1}^{\ind_{\omega}} \omega_{\jmath} \sum_{\jmath=1}^{\ind_{\upsilon}} \upsilon_{\jmath} - \sum_{\jmath=1}^{\ind_{\varepsilon}} \varepsilon_{\jmath} \sum_{\jmath=1}^{\ind_{\delta}} \delta_{\jmath} }
+  \end{align*}
+
 \section{Notes on what to do when allowing deleterious mutations}

### 2017-02-28 kalman_walk.tex
removed redundant equations.
added bio examples, including GRN to Metabolic Network systems.

@@
  Movement in this direction is ongoing, as researchers have begun to study 
-the evolution of both abstract \citep{wagner1994evolution, wagner1996does,  siegal2002waddington, bergman2003evolutionary} and empirically inspired computational and mathematical models of gene regulatory networks (GRNs) \citep{mjolsness1991connectionist, jaeger2004dynamic, maria1, vitaly1, vitaly2, crombach2016gap, wotton2015quantitative}.
+the evolution of both abstract \citep{wagner1994evolution, wagner1996does,  siegal2002waddington, bergman2003evolutionary} and empirically inspired computational and mathematical models of gene regulatory networks (GRNs) \citep{mjolsness1991connectionist, jaeger2004dynamic, maria1, vitaly1, vitaly2, crombach2016gap, wotton2015quantitative, chertkova2017insilico}.
 If we allow the reasonable assumption that the genotype-phenotype map can be represented as a system of differential equations, we can immediately discuss its evolution and function in a much more mechanistic, yet general, manner.
@@
 
-Further let $\mu(\Sigma, A_{ij})$ be the resulting gene regulatory network following a mutation of matrix element $A_{ij}$ to system $\Sigma$ (which contains matrix $A$).
-
-  \begin{align*}
-    \Phi(\mu(\Sigma, A_{ij} \upsilon)) = \exp\left(- \left| \frac{A_{22}-A_{12}}{\lambda_{1} \lambda_{2}} - \frac{k + \upsilon \pda (A_{22}-A_{12})}{\lambda_{1} \lambda_{2} + \text{adj}(A)^{T}_{ij} \upsilon} \right|\right)   .
-  \end{align*}
-For instance, the fitness of an organism with a change of one of its entries $A_{ij}$ (by $\upsilon$) is listed,
-  \begin{align*}
-    \Phi(\mu(\Sigma, A_{11})) &= \exp \left(- \left| \frac{A_{22}-A_{12}}{\det(A)} - \frac{A_{22}-A_{12}}{\det(A) + \upsilon A_{22}} \right| \right) \\
-    \Phi(\mu(\Sigma, A_{12})) &= \exp \left(- \left| \frac{A_{22}-A_{12}}{\det(A)} - \frac{A_{22}-A_{12} - \upsilon}{\det(A) - \upsilon A_{21}} \right| \right) \\
-    \Phi(\mu(\Sigma, A_{21})) &= \exp \left(- \left| \frac{A_{22}-A_{12}}{\det(A)} - \frac{A_{22}-A_{12}}{\det(A) - \upsilon A_{12}} \right| \right) \\
-    \Phi(\mu(\Sigma, A_{22})) &= \exp \left(- \left| \frac{A_{22}-A_{12}}{\det(A)} - \frac{A_{22}-A_{12} + \upsilon}{\det(A) + \upsilon A_{11}} \right| \right)
-  \end{align*}
-
+%Further let $\mu(\Sigma, A_{ij})$ be the resulting gene regulatory network following a mutation of matrix element $A_{ij}$ to system $\Sigma$ (which contains matrix $A$).
+%
+%  \begin{align*}
+%    \Phi(\mu(\Sigma, A_{ij} \upsilon)) = \exp\left(- \left| \frac{A_{22}-A_{12}}{\lambda_{1} \lambda_{2}} - \frac{k + \upsilon \pda (A_{22}-A_{12})}{\lambda_{1} \lambda_{2} + \text{adj}(A)^{T}_{ij} \upsilon} \right|\right)   .
+%  \end{align*}
+%For instance, the fitness of an organism with a change of one of its entries $A_{ij}$ (by $\upsilon$) is listed,
+%  \begin{align*}
+%    \Phi(\mu(\Sigma, A_{11})) &= \exp \left(- \left| \frac{A_{22}-A_{12}}{\det(A)} - \frac{A_{22}-A_{12}}{\det(A) + \upsilon A_{22}} \right| \right) \\
+%    \Phi(\mu(\Sigma, A_{12})) &= \exp \left(- \left| \frac{A_{22}-A_{12}}{\det(A)} - \frac{A_{22}-A_{12} - \upsilon}{\det(A) - \upsilon A_{21}} \right| \right) \\
+%    \Phi(\mu(\Sigma, A_{21})) &= \exp \left(- \left| \frac{A_{22}-A_{12}}{\det(A)} - \frac{A_{22}-A_{12}}{\det(A) - \upsilon A_{12}} \right| \right) \\
+%    \Phi(\mu(\Sigma, A_{22})) &= \exp \left(- \left| \frac{A_{22}-A_{12}}{\det(A)} - \frac{A_{22}-A_{12} + \upsilon}{\det(A) + \upsilon A_{11}} \right| \right)
+%  \end{align*}
+%
  The transfer function at $z=0$ for any GRN with mutations is, 
@@
 
+\section{Biological Examples}
+
+  \begin{example}
+    Negative feedback and Positive feedback systems:
+    \begin{align*}
+      T(z) = \frac{H(z)}{1 + G(z)H(z)} \\
+      T(z) = \frac{H(z)}{1 - G(z)H(z)} \\
+    \end{align*}
+  \end{example}
+
+  \begin{example}
+    Systems in serial ($u \rightarrow \Sigma_{1} \{A_{1},B_{1},C_{1} \} \rightarrow \Sigma_{2} \{A_{2},B_{2},C_{2} \} \rightarrow y$):
+    \begin{align*}
+      \begin{bmatrix} \dot{x}_{1} \\ \dot{x}_{2} \end{bmatrix} &= \left[ \begin{array}{c|c}
+        A_{1} & 0 \\
+          \hline
+          B_{2}C_{2} & A_{2} \\
+        \end{array} \right] \begin{bmatrix} \dot{x}_{1} \\ \dot{x}_{2} \end{bmatrix}  + \begin{bmatrix} B_{1} \\ 0 \end{bmatrix} u \\
+          y &= \begin{bmatrix} 0 & C_{2} \end{bmatrix} \begin{bmatrix} \dot{x}_{1} \\ \dot{x}_{2} \end{bmatrix} .
+    \end{align*}
+
+  Systems in serial could be applied to model a gene network $A_{1}$ and the subsequent degradation rates of its outputs at rate $A_{2}$.
+  \end{example}
+
+  \begin{example}
+  \begin{align*}
+    \dot{x} &= \begin{bmatrix} + & - \\ 0 & + \end{bmatrix} \vec{x} + \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \vec{u} \\
+      y(t) &= \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \vec{x} \\
+        E(t) &= \begin{bmatrix} -y_{1}(t) & 0 & 0 \\ y_{1}(t) & -y_{2}(t) & 0 \\ -\sigma_{1} y_{1}(t) & (1 - \sigma_{2})y_{2}(t) & 1 \end{bmatrix} \vec{\zeta} + \begin{bmatrix} u_{1} \\ u_{2} \\ \epsilon \end{bmatrix} .
+  \end{align*}
+    Where $u_{1}$ is input of \emph{lactose}, $u_{2}$ is the input of \emph{glucose}, $y_{1}$ is the enzyme \emph{lactase} and $y_{2}$ is enzyme \emph{``glucase''}.
+$E(t)$ is metabolism of the organism, converting sugars to energy as a function of enzyme concentrations.
+Lactase converts lactose to glucose, and the enzyme is produced at energy cost $\sigma_{1}$; glucase converts glucose to energy and is produced at cost $\sigma_{2}$.
+Fitness is a function of the total energy of the system, given inputs, possibly: $\int_{0}^{T}E(t)_{3} dt$ after some time $T$ has passed.
+Or perhaps, once the total metabolic energy produced passes some threshold, the organism undergoes fission.
+Thus, fitness selects for maximizing energy as quickly as possible.
+  \end{example}
+
+  \begin{example} 
+    Gene Regulatory and Metabolic Network Cascade:
+
+    Gene Regulatory Network
+    \begin{align*}
+      \dot{x}(t) &= A x(t) + B_{1} u(t) \\
+      y(t) &= C_{1} x(t)
+    \end{align*}
+      Metabolism
+      \begin{align*}
+      \dot{\zeta}(t) &= M\big(y(t), \sigma \big) \zeta(t) + B_{2} \begin{bmatrix} u(t) \\ v(t) \end{bmatrix} \\
+        m(t) &= C_{2} \zeta(t) \\
+      b(t) &= C_{3} \zeta(t)
+      \end{align*}
+      Fitness
+      \begin{align*}
+      \Phi(m(t), p(t))
+    \end{align*}
+
+    The GRN produces time varying proteins involved in the metabolism.
+The metabilic network (MN), is a function of protein concentrations $y(t)$ and production costs $\sigma$.
+The metabolic network produces energy $m(t)$ and toxic byproducts $b(t)$.
+Fitness $\Phi$, or maybe something more concrete, like fission time and/or cell death, is a function of both energy $m(t)$ and toxins $b(t)$.
+  \end{example}
 \section{Notes on what to do when allowing deleterious mutations}

### 2017-03-19 biosystem_examples.tex
added an example of yeast galactose metabolism.

\documentclass[11 pt]{article}
\usepackage{amsmath, amssymb, color, xcolor}
\usepackage{graphicx, wrapfig, float, caption, dsfont}
\usepackage{fullpage}
\usepackage[backref=page, hidelinks, colorlinks=true, citecolor=blue!60!black!100]{hyperref}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, shapes}
\usepackage{caption, subcaption}
\usepackage{natbib} % gives us \citet: Author (year) and \citep: (Author; year)
\usepackage{authblk}

\newcommand{\plr}[1]{{\color{blue}\it #1}}
\newcommand{\jss}[1]{{\color{olive}\it #1}}
% \newcommand{\ddt}{\frac{d}{dt}}
\newcommand{\ddt}{\dot}
\newcommand{\ro}{{ro}}
\newcommand{\nro}{{\bar{r}o}}
\newcommand{\rno}{{r\bar{o}}}
\newcommand{\nrno}{{\bar{r}\bar{o}}}
\newcommand{\reachable}{\mathcal{R}}
\newcommand{\unobservable}{\bar{\mathcal{O}}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\pda}{\frac{\partial}{\partial A_{ij}}}
\newcommand{\ind}{\mathds{1}}

\DeclareMathOperator{\spn}{span}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}

\begin{document}

  \section{\emph{S. cerevisiae} galactose catabolism network}

  Yeast galactose catabolism is accomplished via a complex system of molecules including enzymes, transporters, and transcription factors. The GAL2 gene encodes a transporter than imports galactose into the cell cytoplasm. GAL1, GAL7, and GAL10 encode components of enzymes that act sequentially to metabolize the imported galactose. Furthermore, in \emph{S. cerevisiae} the enzyme encoding genes, GAL1, GAL7, and GAL10 are encoded within the same regulon (on the same chromosome and regulated by the same promoter), whereas GAL2 is encoded on a separate chromosome. The transcription factor produced by GAL4 binds to promoter of the regulon, inducing the expression of the enzymes. Another gene, MIG1, which is induced by the presence of glucose, encodes a repressor for the GAL regulon. MIG1 acts by repressing the GAL enzymes, as well as down regulating some of the GAL transcription factors, by repressing GAL4 and GAL3, but not acting on GAL80. MIG1 also represses the transporter GAL2.

   \begin{figure}
        \begin{center}
            \begin{tikzpicture}
            \begin{scope}[every node/.style={circle,thick,draw}]
                \node (GAL3) at (2,0) {GAL3};
                \node (GAL4) at (4,0) {GAL4};
                \node[dashed] (BTM) at (4,2) {BTM};
                \node (GAL80) at (6,0) {GAL80};
                \node (MIG1) at (0,0) {MIG1};
                \node[shape=rectangle] (GAL2) at (0,-2) {GAL2};
                \node[shape=rectangle] (enzymes) at (4,-2) {GAL1-GAL7-GAL10};
                \node[dashed] (galactose) at (-2, 2) {galactose};
                \node[dashed] (glucose) at (0,2) {glucose};
            \end{scope}

            \begin{scope}[>={Stealth[black]},
                          every edge/.style={draw=black, thick}]
                 \path[->] (GAL4) edge[left] node {} (GAL3);
                 \path[->] (GAL4) edge[bend left] node {} (GAL80);
                 \path[->, >=Rectangle] (GAL3) edge[bend left=50] node {} (GAL80);
                 \path[->, >=Rectangle] (GAL80) edge[bend left] node {} (GAL4);
                 \path[->] (GAL4) edge[bend right] node {} (enzymes);
                 \path[->] (GAL4) edge[left] node {} (GAL2);
                 \path[->, >=Rectangle] (MIG1) edge[bend left] node {} (GAL3);
                 \path[->, >=Rectangle] (MIG1) edge[bend left=50] node {} (GAL4);
                 \path[->, >=Rectangle] (MIG1) edge node {} (enzymes);
                 \path[->, >=Rectangle] (MIG1) edge node {} (GAL2);
                 \path[->] (glucose) edge node {} (MIG1);
                 \path[->] (galactose) edge[bend right=70] node {} (GAL3);
                 \path[->] (GAL2) edge node {} (galactose);
                 \path[->] (BTM) edge node {} (GAL4);
                 \path[->] (BTM) edge[bend right] node {} (MIG1);
            \end{scope}
            \end{tikzpicture}
        \end{center}
        \caption{
          Yeast galactose catabolism transcription network.}
    \end{figure}

    \begin{align*}
      \dot{x}(t) &= \begin{bmatrix} 0 & + & 0 & - \\ 0 & 0 & - & - \\ - & + & 0 & 0 \\ 0 & 0 & 0 & 0 \end{bmatrix} \begin{bmatrix} \text{GAL3} \\ \text{GAL4} \\ \text{GAL80} \\ \text{MIG1} \end{bmatrix}(t) + \begin{bmatrix} + & 0 & 0 \\ 0 & 0 & + \\ 0 & 0 & 0 \\ 0 & + & + \end{bmatrix} \begin{bmatrix} \text{galactose} \\ \text{glucose} \\ \text{BTM} \end{bmatrix}(t) \\
        y(t) &= \begin{bmatrix} 0 & + & 0 & - \end{bmatrix} \vec{x}(t)
    \end{align*}
    In the above parameterization, $y(t)$, or the cumulative impact of transcriptional regulation on the expression of the GAL regulon, will either tend towards positive or negative infinity depending on the ratio of galactose to glucose, and the impact of the basal transcriptional machinery (BTM) on the components of the system.

    \subsection{Notes}
    \begin{itemize}
      \item The interaction strength between GAL4 and GAL80 can be altered via mutation. Furthermore, a mutation in GAL4 that decreases affinity can be compensated by a mutation in GAL80 \citep{li2010alterations, salmeron1990gal4, adhikari2014perturbation}.
      \item The galactose metablic network shows tremendous regulatory variation among different species of yeast \citep{lavoie2009rearrangements, martchenko2007transcriptional, dalal2016transcriptional}.
    \end{itemize}

\bibliographystyle{plainnat}
\bibliography{krefs}

\end{document}

### 2017-03-19 biosystem_examples.tex
added a note

@@
 Furthermore, a mutation in GAL4 that decreases affinity can be compensated by a mutation in GAL80 \citep{li2010alterations, salmeron1990gal4, adhikari2014perturbation}.
-      \item The galactose metablic network shows tremendous regulatory variation among different species of yeast \citep{lavoie2009rearrangements, martchenko2007transcriptional, dalal2016transcriptional}.
+      \item The galactose metablic network shows tremendous regulatory variation among different species of yeast \citep{lavoie2009rearrangements, martchenko2007transcriptional, dalal2016transcriptional, christensen2011unique, hartl2007induction, alam2013aspergillus}.
+      \item GAL3 (a TF) is a paralog of GAL1 (enzyme), yet GAL3 no longer has any galactokinase activity. GAL1, however, can weakly regulate in place of GAL3.
     \end{itemize}

### 2017-03-23 biosystem_examples.tex
added notes from the white board today.

@@
     \end{align*}
+
     In the above parameterization, $y(t)$, or the cumulative impact of transcriptional regulation on the expression of the GAL regulon, will either tend towards positive or negative infinity depending on the ratio of galactose to glucose, and the impact of the basal transcriptional machinery (BTM) on the components of the system.
+    \subsection{concrete example}
+    \begin{align*}
+      \dot{x}(t) &= \begin{bmatrix} 0 & 1 & 0 & -1 \\ 0 & 0 & -1 & -1 \\ -1 & 1 & 0 & 0 \\ 0 & 0 & 0 & 0 \end{bmatrix} \begin{bmatrix} \text{GAL3} \\ \text{GAL4} \\ \text{GAL80} \\ \text{MIG1} \end{bmatrix}(t) + \begin{bmatrix} 1 & 0 & 0 \\ 0 & 0 & 1 \\ 0 & 0 & 0 \\ 0 & 1 & 1 \end{bmatrix} \begin{bmatrix} \text{galactose} \\ \text{glucose} \\ \text{BTM} \end{bmatrix}(t) \\
+        y(t) &= \begin{bmatrix} 0 & 1 & 0 & -1 \end{bmatrix} \vec{x}(t)
+    \end{align*}
+    \begin{align*}
+      H(z) = \begin{bmatrix} 1 \\ z^{2} + z + 1 \\ z + 1 \end{bmatrix} \frac{1}{z^{3} + z -1}
+    \end{align*}
 
+    \begin{align*}
+      CV &= C \\
+      VB &= B \\
+      V :&= \begin{bmatrix} 1 & 0 & a & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & b & 0 \\ 0 & 0 & 0 & 1 \end{bmatrix} , V^{-1} = \begin{bmatrix} 1 & 0 & \frac{-a}{b} & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & \frac{1}{b} & 0 \\ 0 & 0 & 0 & 1\end{bmatrix} \\
+        A(a,b) :&= VAV^{-1} = \begin{bmatrix} -a & a+1 & \frac{a^{2}}{b} & -1 \\ 0 & 0 & \frac{-1}{b} & -1 \\ -b & b & a & 0 \\ 0 & 0 & 0 & 0 \end{bmatrix}
+    \end{align*}
+    The GAL regulatory network in \emph{S.
+cerevisiae} is modelled here.
+The activating transcription factor GAL4, which is regulated by the presence of galactose and two other TFs, binds to the promoter region of the GAL regulon -- a cluster of genes regulated by the same promoter encoding 3 enzymes (GAL1, GAL7, and GAL10), for the metabolism of the sugar galactose.
+The repressing transcription factor MIG1, which is activated by the presence of glucose and other proteins omitted here, also binds to the regulon, preventing the expression of the galactose metabolizing enzymes.
+This network is one of the most studied gene regulatory networks in yeast, and experiments have already demonstrated significant transcriptional variation among different species and genuses of yeast.
+
+    In the present model, the $A$ matrix encodes the interactions among GAL3, GAL4, GAL80, and MIG1.
+$B$ interprets the input: the quantity of glucose, galactose, and the basal transcription (BTM) rates impact of TF concentrations.
+$C$ represents the promoter region of the GAL regulon and is the sum of the influence of GAL4 (activating) and MIG1 (repressing).
+
+    $A(a,b)$ represent alternative network structures that produce identical outputs, with identical input and output transformations -- meaning the regulatory contributions of the sugars and the basasl machinery, as well as the promoter of the GAL regulon are held constant.
+These interactions may be realized in nature, if the regulatory differences predicted by $A(a,b)$ are biologically/mutationally realizable.
+Mutations influencing the interactions between GAL4 and GAL80 have been experimentally demonstrated \citep{li2010alterations}.
+
+    Furthermore, our fitness function is restricting the network to realize the transfer function $H(z)$ exactly.
+We see this as justifiable, as experiment has shown the output of a metabolic regulatory network directly influences organismal growth rates.
+In \citet{dekel2005optimality} \emph{E.
+coli} that over or under express lactase, governed by the \emph{lac} operon, in a given sugar environment, their fitness (growth rate) suffers.
     \subsection{Notes}
@@
       \item GAL3 (a TF) is a paralog of GAL1 (enzyme), yet GAL3 no longer has any galactokinase activity. GAL1, however, can weakly regulate in place of GAL3.
+      \item RTG1-RTG3 regulates the Gal regulon in other species of yeast. 
     \end{itemize}
+
+    \section{Cell Cycle Control}
+      The cell cycle, and the initiation of cellular processes in general, is often triggered by the simultaneous presence and absence of specific molecules.
+In yeast, traversal through the cell cycle is largely determined by the presence of a class of proteins called cyclins.
+Cyclin concentration levels oscillate throughout the cell cycle, and after reaching a threshold interact with CDKs, triggering a transition to another phase of cellular life.
+Although the underlying mechanics controllying cell cycle phase transitions is extremly complex, involving many steps including transcriptional regulation and enzyme phosphorylation, several lines of evidence suggest that the simple oscillating expression patterns of transcription factors is important for cell cycle control \citep{orlando2008global, simon2001serial}.
+      \begin{quote}
+        We propose that periodic transcription is an emergent property of a transcription factor network that can function as a cell-cycle oscillator independently of, and in tandem with, the CDK oscillator.
+\citep{orlando2008global}
+      \end{quote}
+
+      \subsection{Notes}
+      \begin{itemize}
+        \item According to \citet{cross2011evolution}, the network topology and dynamic properties of cell cycle control in both yeast and mammals is highly conserved, however the proteins themselves are highly diverged -- even having completely unrelated amino acid sequences.
+        \item Mammalian G1-S regulation is more complex than yeast: the mammalian system contains addition feedback loops.
+Furthermore, CDKs ``play complementary roles'' as both CDK4 and CDK6 both bind cylicn D and CDK1 and CDK2 both bind to cyclins, A, B and E.
+The binding is not very specific as there is significant cross-binding among the CDKs and cyclins \citep{cross2011evolution}.
+        \item ``Proteins involved in DNA replication are conserved from yeast to mammals, suggesting that the mechanism was established at an early stage of eukaryotic evolution.
+In spite of this common origin, recent findings have revealed surprising variations in how replication initiation is controlled, implying that a conserved mechanism has not necessarily resulted in regulatory conservation'' \citep{kearsey2003enigmatic}.
+        \item Putative DMI may be gene involved in cell cycle control?
+Researchers identified regions \emph{hms1} and \emph{hms2} -- or hybrid male sterility 1/2 -- in \emph{Mimulus}.
+These genomic regions contain genes known to be involved in cell cycle control \citep{sweigart2015evidence}.
+      \end{itemize}
+    \section{Early Development and Pattern Formation in \emph{Drosophila}}
+    \begin{itemize}
+      \item DSD in gap gene network lead to DMI in \emph{D. melanogaster} and \emph{D. santomea} (see D. Matute work). 
+    \end{itemize}
+    \section{Carbohydrate metabolic network in \emph{S. cerevisiae}}
+
+    In Flux Balance Analysis, a metabolic reaction at equilibirum is represented as, 
+    \begin{align*}
+      Sv = 0
+    \end{align*}
+    Where, $S$ is an $n \times m$ stochiometric matrix and $v$ is a vector of fluxes through the network.
+Each column of $S$ is a reaction, and each row is a unique compound.
+Any vector in the nullspace of $S$ can represent fluxes through the network in steady state.
+Although flux is a function of enzyme activity, expression level, and substrate concentrations, we may be able to model each element of $v$ as simply enzyme expression levels.
+Thus, the output of a GRN $y$ can be input as the flux vector for the metabolic network.
+      \begin{itemize}
+        \item Integration of Flux Balance Metabloism and TRN models (see B Palsson work).
+        \item Could the diversity in wine taste profiles partially be a consequence of DSD in the metabolic reaction network? 
+      \end{itemize}
+    \section{Circadian clock}
+      \begin{itemize}
+        \item The Intelligent Clock and the Rube Goldberg Clock \citep{sancar2008intelligent}.
+      \end{itemize}
+    \section{Chemotaxis in prokaryotes}
 

### 2017-03-29 kalman_walk.tex
added bio examples, including GRN to Metabolic Network systems. Added reference link.
added some questions about adaptation and mutational robustness.

@@
 
-    \item \emph{(josh)} Write out simple examples in biological words.
+      \item \emph{(josh)} Write out simple examples in biological words. \ref{sec:examples}
 
@@
 
+\section{Adaptation}
+
+Suppose a population is perfectly adapted to its environment -- it expresses a metabolic enzyme in response to environmental sugar availability at a rate that balances the energetic costs and benefits of the enzyme's synthesis and utility.
+Then, the population suddenly finds itself in a novel environment and exhibits suboptimal expression rates.
+How does the population adapt?
+Are some network mechanisms mutationally closer or farther from the desired new network connectivity?
+
+Let $A_{\alpha}(p)$ be the $p$th system realization perfectly adapted to the first environment ($\alpha$), and let $A_{\beta}(q)$ be the $q$th system realization perfectly adapted to the second environment ($\beta$).
+\begin{align*}
+  \{p,q : \min \sum_{i,j}^{n} | A_{\alpha}(p)_{i,j} - A_{\beta}(q)_{i,j} | \}
+\end{align*}
+In an asexual population, if there is variation in network topology, the $A_{\alpha}$ network closest to the organization of a $A_{\beta}$ network might be the fastest to adapt and outcompete all of its neighbors.
+
+\section{Mutational Robustness}
+
+Are any network realizations $A(p)$ more mutationally robust than others?
+If $s_{i,j}$ is the fitness cost of mutating element $A_{i,j}$ and $\mathbb{E}_{p}(s) = \frac{1}{n}\displaystyle\sum_{i,j}^{n} s_{i,j}$ is the mutationaly susceptibility of realization $p$, is there a $p$ such that $\mathbb{E}(s)$ is minimal?
+Will the population find this network orgnization?
+
 \section{(Old Notes) What biological question are we asking?}
@@
 
-\section{Biological Examples}
+\section{Biological Examples}\label{sec:examples}
 
@@
   \end{example}
+
+  \begin{example}
+    Rough yeast colony morphology phenotype / Simplified Ras pathway.
+Rough colony morphology is a complex phenotypic trait influenced by many different genetic variants.
+Ultimately it appears that colony roughness is a function of RNA Polymerase II activity, downstream of Ras and Gpr1, and acting on FLO11.
+If RNA pol II activity breaks a threshold level, FLO11 is expressed and colony roughness is observed.
+
+    \begin{align*}
+      \dot{x} &= \begin{bmatrix} 0 & 0 & 0 & 0 & 0 & 0 \\ + & 0 & \zeta & 0 & 0 & 0 \\ + & \eta & 0 & 0 & 0 & 0 \\ 0 & + & 0 & 0 & 0 & 0 \\ 0 & + & 0 & 0 & 0 & 0 \\ 0 & 0 & - & 0 & 0 & 0 \end{bmatrix} \begin{bmatrix} \text{Cyr1} \\ \text{Tpk1} \\ \text{Tpk3} \\ \text{Flo8} \\ \text{Mga1} \\ \text{Sfl1} \end{bmatrix} + \begin{bmatrix} + & +  & 0 \\ 0 & 0 & 0 \\ 0 & 0 & 0 \\ 0 & 0 & 0 \\ 0 & 0 & 0 \\ 0 & 0 & + \end{bmatrix} \begin{bmatrix} \text{Gpa2} & \text{Ras} & \text{Sfl1} \end{bmatrix} \\
+        y &= \text{cumulative impact on RNA Pol II activity } = \begin{bmatrix} 0 & 0 & 0 & + & + & - \end{bmatrix} \vec{x}
+    \end{align*}
+  \end{example}
 \section{Notes on what to do when allowing deleterious mutations}

### 2017-03-29 kalman_walk.tex
added questions about rube goldgergness and network ratchets

@@
 
+\section{Network Ratchet}
+
+If a network evolves to include more molecules (such as transcription factors) and thus becomes higher dimensional, can the network also remove these originally superfluous interactions or will removing these typically lead to catastrophic network dysfunction after some amount of evolutionary time and system drift has occured?
+
+\section{Network Minimality versus Rube Goldbergness}
+
+Are there evolutionary differences between realizing a transfer function minimally (with a reachable and observable system) or non-minimally (a system with unreachable and/or unobservable subspaces)?
+Can reducible systems be more or less mutationally robust?
+Are minimal systems energetically more efficient?
+Are reducible systems more or less evolvable?
+
 \section{(Old Notes) What biological question are we asking?}

### 2017-05-02 brownian_motion.tex
notes on a RW

\documentclass[11 pt]{article}
\usepackage{amsmath, amssymb, color, xcolor}
\usepackage{graphicx, wrapfig, float, caption, dsfont, bbm}
\usepackage{fullpage}
\usepackage[backref=page, hidelinks, colorlinks=true, citecolor=blue!60!black!100]{hyperref}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, shapes}
\usepackage{caption, subcaption}
\usepackage{natbib} % gives us \citet: Author (year) and \citep: (Author; year)
\usepackage{authblk}

\newcommand{\plr}[1]{{\color{blue}\it #1}}
\newcommand{\jss}[1]{{\color{olive}\it #1}}
% \newcommand{\ddt}{\frac{d}{dt}}
\newcommand{\ddt}{\dot}
\newcommand{\ro}{{ro}}
\newcommand{\nro}{{\bar{r}o}}
\newcommand{\rno}{{r\bar{o}}}
\newcommand{\nrno}{{\bar{r}\bar{o}}}
\newcommand{\reachable}{\mathcal{R}}
\newcommand{\unobservable}{\bar{\mathcal{O}}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\pda}{\frac{\partial}{\partial A_{ij}}}
\newcommand{\ind}{\mathds{1}}

\newcommand{\A}{\mathcal{A}}
\newcommand{\diag}{\text{diag}}
\newcommand{\1}{\mathbbm{1}}

\DeclareMathOperator{\spn}{span}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}

\begin{document}

  \section{Systems Drift as a Random Process}
  \subsection{LE Systems}
    An LE System: Linear Evolving (TI) System (in evolutionary state $\tau$), 
    \begin{align*}
      \Sigma(\tau) = \left\{ \begin{array}{cc} \dot{x} =& A(\tau)x + Bu \\
      y =& Cx \end{array} \right.
    \end{align*}
    \begin{align*}
      \mathcal{L}(\Sigma(\tau)) := H(z, \tau) = C \left(zI - A(\tau) \right)^{-1} B = \frac{(z-\gamma)}{(z-\lambda_{1})(z-\lambda_{2})} \quad \forall \tau \neq 1 \in \mathbb{R}
    \end{align*}
 $\Sigma(\tau)$ is a Linear Time Invariant dynamical system, with transfer function (the Laplace transform of it's time-domain dynamics) $H(z)$. We are interested in a random process over the manifold defined by the set of all systems in the minimal and same dimension that share identical transfer functions.

  \subsection{Simple random walk in one dimension.}
  Let $A(\tau)$ describe the gene network interaction topology in evolutionary state $\tau$. Let $V(\tau)$ be a subset of the general linear group that preserves the input and output matrices in the system $\Sigma := \{ A, B, C \}$. That is, $V$ is invertible and $VB = B$ and $CV = C$ for all $V$.

  \begin{align*}
    A(0) :&= \begin{bmatrix} \lambda_{1} & \lambda_{2} - \gamma \\ 0 & \lambda_{2} \end{bmatrix} \qquad B^{T} = \begin{bmatrix} 1 & 1 \end{bmatrix} \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix} \\
    V(\tau) :&= \begin{bmatrix} 1 & 0 \\ \tau & 1-\tau \end{bmatrix} \\
    A(\tau) :&= V(\tau)A(0)V^{-1}(\tau)
  \end{align*}

  Let $\Delta_{\tau}^{\pm}$ be the rate at which the topology of the gene network $A$ changes, given it's evolutionary state $\tau$, and the direction $\pm \tau$ it is moving in. If evolution is a random walk, let the probability of $A(\tau) \rightarrow A(\tau + 1)$ be $p(\tau) \mu$ and the probability of $A(\tau) \rightarrow A(\tau - 1)$ be $q(\tau) \mu$, where $\mu$ is the rate of transcription factor binding site (TFBS) change. The probability that the evolutionary state does not change is $(1-\mu)$. 
  \begin{align*}
    \Delta^{\pm}_{\tau} :&= \left\lVert \frac{\partial}{\partial^{\pm} \tau} \text{vec}\left[A(\tau)\right] \right\rVert^{-1} \\
    p(\tau) :&= \frac{\Delta^{+}_{\tau}}{\Delta^{+}_{\tau} + \Delta^{-}_{\tau}} \\
    q(\tau) :&= \frac{\Delta^{-}_{\tau}}{\Delta^{+}_{\tau} + \Delta^{-}_{\tau}}
  \end{align*}

  \begin{align*}
    A(\tau) = p(\tau) A(\tau - 1) \mu + q(\tau) A (\tau + 1) \mu + A(\tau) (1 - \mu)
  \end{align*}

\bibliographystyle{plainnat}
\bibliography{krefs}

\end{document}

### 2017-05-09 examples/two-diml-minimal.Rmd
start of doc for doing brownian motion

---
title: "Brownian motion on a 2D minimal system"
author: "Peter Ralph and Josh Shiffman"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
```

# Model

Consider the dynamical system given by
$$\begin{aligned}
      \dot{x}(t) &= Ax(t) + Bu(t) \\
          y(t) &= Cx(t) 
\end{aligned}$$
where
$$\begin{aligned}
    B^{T} = \begin{bmatrix} 1 & 1 \end{bmatrix} \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix} .
\end{aligned}$$
and $A=A(\tau)$ depends on a free parameter,
$$\begin{aligned}
    A_0 &= \begin{bmatrix} \lambda_{1} & \lambda_{2} - \gamma \\ 0 & \lambda_{2} \end{bmatrix} \\
    V_\tau &= \begin{bmatrix} 1 & 0 \\ \tau & 1-\tau \end{bmatrix} \\
    A_\tau &= V_\tau A_0 V^{-1}_\tau
\end{aligned}$$
These systems are input-output equivalent (although may have different noise properties).

# Brownian motion

Now suppose that $A$ is changing due to mutation, 
and that it takes the value $A(t)$ after $t$ units of evolutionary time.
We know that $A(t)$ must equal $A_\tau$ for some $\tau$, 
and hence we can define $A(t) = A_{W(t)}$, 
where $W(t)$ is a random process on the free parameter;
but what are the appropriate dynamics for $W$?
Suppose that genetic drift acts uniformly on the entries of $A$, 
so that ...

# Toy example

As a toy example, suppose that we want to describe a process through $\R^2$
on the set
$$\begin{aligned}
    x^2 + y^2 = 1,
\end{aligned}$$
i.e., the unit circle.
This is parameterized by
$$\begin{aligned}
    x(\theta) &= \cos(\theta) \\
    y(\theta) &= \sin(\theta) .
\end{aligned}$$

The dynamics we want are limits of the following discrete dynamics.
Let $\epsilon$ denote the step size,
and suppose that at rate $\eta$ steps of size $\epsilon$ 
in a uniformly random direction $\phi$ are proposed;
but the probability that such a step is accepted is Kimura's expression:
$$\begin{aligned}
    q(x',y';x,y) = \frac{1 - e^{-2 (f(x',y')-f(x,y))}}{1 - e^{-2 N_e (f(x',y')-f(x,y))}}
\end{aligned}$$
where $(x',y')$ is the new (proposed) location, and
$$\begin{aligned}
    f(x,y) = (\sqrt{x^2+y^2}-1)^2
\end{aligned}$$
is the fitness function.
Note that $q(x,y;x,y) = 1$ 
and that 
$$\begin{aligned}
    \partial_u q(u,v;x,y) 
    &= 
        ( \partial_u f(u,v) ) 
        \left( \partial_z \frac{1 - e^{-2 (z-f(x,y))}}{1 - e^{-2 N_e (z-f(x,y))}} \vert_{z=f(x',y')} \right) \\
    &= 
        \left( \frac{2 x (\sqrt{x^2+y^2}-1)}{\sqrt{x^2+y^2}} \right) 
        \left( \frac{ 2 e^{-2 (f(x',y')-f(x,y))} (1 - e^{-2 N_e (f(x',y')-f(x,y))}) - 2 N_e e^{-2 (f(x',y')-f(x,y))} }{(1 - e^{-2 N_e (f(x',y')-f(x,y))})^2} \right)
\end{aligned}$$
and so that
$$\begin{aligned}
    \partial_u q(u,v;x,y) \vert_{u=x, v=y}
    &= 
        \left( \frac{2 x (\sqrt{x^2+y^2}-1)}{\sqrt{x^2+y^2}} \right) 
        \left( \frac{ - 2 N_e  }{(1 - e^{-2 N_e (f(x',y')-f(x,y))})^2} \right)
\end{aligned}$$

This has the result of multiplying jump rates by the acceptance probabilities;
so equivalently, if the process is at $(x,y)$,
then it jumps to 
$$\begin{aligned}
    (x',y') = (x+\epsilon \cos \phi, y + \epsilon \sin \phi)
\end{aligned}$$
at rate
$$\begin{aligned}
    \eta q(x',y';x,y) 
    &\approx

\end{aligned}$$

### 2017-05-25 RG_machines.tex
added notes on network size drift from our meeting

\documentclass[11 pt]{article}
\usepackage{amsmath, amssymb, color, xcolor}
\usepackage{graphicx, wrapfig, float, caption, dsfont, bbm}
\usepackage{fullpage}
\usepackage[backref=page, hidelinks, colorlinks=true, citecolor=blue!60!black!100]{hyperref}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, shapes}
\usepackage{caption, subcaption}
\usepackage{natbib} % gives us \citet: Author (year) and \citep: (Author; year)
\usepackage{authblk}

\newcommand{\plr}[1]{{\color{blue}\it #1}}
\newcommand{\jss}[1]{{\color{olive}\it #1}}
% \newcommand{\ddt}{\frac{d}{dt}}
\newcommand{\ddt}{\dot}
\newcommand{\ro}{{ro}}
\newcommand{\nro}{{\bar{r}o}}
\newcommand{\rno}{{r\bar{o}}}
\newcommand{\nrno}{{\bar{r}\bar{o}}}
\newcommand{\reachable}{\mathcal{R}}
\newcommand{\unobservable}{\bar{\mathcal{O}}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\pda}{\frac{\partial}{\partial A_{ij}}}
\newcommand{\ind}{\mathds{1}}

\newcommand{\A}{\mathcal{A}}
\newcommand{\diag}{\text{diag}}
\newcommand{\1}{\mathbbm{1}}

\DeclareMathOperator{\spn}{span}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}

\begin{document}

  \begin{itemize}
    \item Do gene networks drift into uneccesarily complex configurations? If the optimal dynamics of a system can be realized by $g$ genes and/or started in a state with only $g$ genes, during evolution, will the system be composed of $> g$ genes. How many more genes than necessary? 
    \item Are there forces constraining gene network size other than a fitness cost associated with the expense of uncessary energy? For instance, will genes in a non-minimal gene network contribute less overall to the system's output, diminising the cost of their removal (on average)? 
    \item as the population explores neutral network space, how often will the population be on or near a point on the manifold where a gene becomes unccessary? Does this probability go up significantly as unneccsary dimensionality goes up?
    \item Does the unncessary complexity of some gene networks confer evolvability advantages? 
  \end{itemize}
Let, 
  \begin{align*}
    \mathcal{F}_{n} = \left\{ A : C_{n}(zI-A)^{-1} B_{n} = H(z) \right\} \subseteq \mathbb{R}^{n \times n}
  \end{align*}
  be an $n$ dimensional manifold, where $A$ is gene network (not necessarily minimal), and $H(z)$ is a description of the phenotype in the Laplace domain. In the non-minimal case, $C_{k}$ and $B_{k}$ are augmented with zeroes.
  \begin{align*}
    C_{n} &= \left[ \begin{array}{cc} C_{k}  \\ \hline 0 \end{array} \right], C_{i} \neq 0 \\
      B_{n} &= \left[ \begin{array}{ccc} B_{k} & \vert & 0 \end{array} \right], B_{i} \neq 0
  \end{align*}
Let, 
  \begin{align*}
        P_{i} : \mathcal{F}_{n} \rightarrow \Omega_{n-1}
  \end{align*}
  be a projection from an $n \times n$ network to an $(n-1) \times (n-1)$ network. Biologically, this is a gene deletion or removal from the system.

  Let $G_{n,i}$ be the set of networks with identical phenotypes following a deletion. 
  \begin{align*}
        G_{n,i} = \left\{ f \in \mathcal{F}_{n} : P_{i}f \in \mathcal{F}_{n-1} \right\} \\
        G_{n} = \cup_{i} G_{n,i} \\
        d(f) = \# \left\{ i : P_{i}f \in \mathcal{F}_{n-1} \right\} 
  \end{align*}
$d(f)$ is the number of different genes that can potentially be deleted to end up in a phenotypically identical space.

Gene duplication or recruitment is defined by, 
  \begin{align*}
        U_{i}(x) \mathcal{F}_{n-1} \rightarrow \mathcal{F}_{n} \\
        \left\{ U_{i}(x) : x \in \mathbb{R}^{n} \right\} \subseteq G_{n,i} 
  \end{align*}
  where, $x$ is in the $i$th row and $0$ in the $i$th column, except along the diagonal.

  How much of a $d_{n}$-dimensional manifold is near one of $(n-k)(d_{n-1}+n)$-dimenional sub-manifolds? 
  \begin{align*}
    (n-k)\varepsilon^{d_{n}-d_{n-1} - n}
  \end{align*}

  Alternatively, what is the hitting time of $U(x)$?

\end{document}

### 2017-05-28 RG_machines.tex
corrections to notes

@@
   \end{itemize}
-Let, 
+
+Let $\Omega_n$ be the set of all allowable networks: this is some bounded subset of $\R^{n \times n}$,
+and let
   \begin{align*}
-    \mathcal{F}_{n} = \left\{ A : C_{n}(zI-A)^{-1} B_{n} = H(z) \right\} \subseteq \mathbb{R}^{n \times n}
+    \mathcal{F}_{n} = \left\{ A : C_{n}(zI-A)^{-1} B_{n} = H(z) \right\} \subseteq \Omega_n
   \end{align*}
-  be an $n$ dimensional manifold, where $A$ is gene network (not necessarily minimal), and $H(z)$ is a description of the phenotype in the Laplace domain.
-In the non-minimal case, $C_{k}$ and $B_{k}$ are augmented with zeroes.
+be the $d_n$-dimensional manifold of equivalent systems, 
+where $A$ is gene network (not necessarily minimal), and $H(z)$ is a description of the phenotype in the Laplace domain. In the non-minimal case, $C_{k}$ and $B_{k}$ are augmented with zeroes.
   \begin{align*}
@@
   \end{align*}
-  be a projection from an $n \times n$ network to an $(n-1) \times (n-1)$ network. Biologically, this is a gene deletion or removal from the system.
+be the projection from an $n \times n$ network to an $(n-1) \times (n-1)$ network
+obtained by removing the $i^\text{th}$ row and column. 
+Biologically, this is a gene deletion or removal from the system.
 
@@
 
-Gene duplication or recruitment is defined by, 
+Our basic question is: for a given $\epsilon$,
+how much of $\mathcal{F}_n$ is within $\epsilon$ of $G_n$?
+The exact answer will depend on the details, but maybe we see how this changes with $n$.
+
+One way of getting at this is to look in the other direction.
+For instance, we know immediately a set of networks in which a gene can be deleted with no effect:
+it is those networks that just had a gene added.
+Given a network $A$ we can add a new gene without changing its phenotype
+by inserting an arbitrary row $x$ into the $i^\text{th}$ slot
+and a corresponding column of zeros.
+We denote this operation $U_i(x)$, and:
   \begin{align*}
         U_{i}(x) \mathcal{F}_{n-1} \rightarrow \mathcal{F}_{n} \\
-        \left\{ U_{i}(x) : x \in \mathbb{R}^{n} \right\} \subseteq G_{n,i} 
+        \left\{ U_{i}(x) f : x \in \mathbb{R}^{n} \; f \in \mathcal{F}_n \right\} \subseteq G_{n,i} 
   \end{align*}
-  where, $x$ is in the $i$th row and $0$ in the $i$th column, except along the diagonal.
+(Note that gene duplication would do something more specific
+if we assume the new gene copy does the same thing as the old one.)
+This implies that $d_n \ge d_{n-1} + n$.
 
-  How much of a $d_{n}$-dimensional manifold is near one of $(n-k)(d_{n-1}+n)$-dimenional sub-manifolds? 
+Now we want to answer one of two questions:
+
+How much of a $d_{n}$-dimensional manifold is within $\varepsilon$ of one of $(n-k)$ sub-manifolds, each of dimension $(d_{n-1}+n)$?
+This scales like
 \begin{align*}

### 2017-06-01 handnotes.tex
handnotes contains a simple  explanantion of the model.

--- 2017-02-28 kalman_walk.tex
@@
-\documentclass[11 pt]{article}
+\documentclass[9 pt]{article}
 \usepackage{amsmath, amssymb, color, xcolor}
-\usepackage{graphicx, wrapfig, float, caption, dsfont}
+\usepackage{graphicx, wrapfig, float, caption, dsfont, bbm}
 \usepackage{fullpage}
@@
 \usepackage{authblk}
+
+\usepackage{multicol}
 
@@
 
+\newcommand{\A}{\mathcal{A}}
+\newcommand{\diag}{\text{diag}}
+\newcommand{\1}{\mathbbm{1}}
+
 \DeclareMathOperator{\spn}{span}
@@
 
-\paragraph{To do, soon:}
+\section*{The Evolution of Phenotypically Invariant Genetic Networks}
+Joshua S. Schiffman and Peter L. Ralph
 
-\begin{enumerate}
+\begin{multicols}{2}
 
-  \item \emph{(josh)} Add biology to introduction.
-(keep an eye out for concrete numbers we can use!) \jss{added some relevant biology citations to introduction.
-Still needs a lot of refinement.} \ref{sec:intro}
-
-    \item Do some order-of-magnitude calculation to figure out how fast systems should drift.
-        \begin{itemize}
-            \item \emph{(peter)} Work out the math for how to do this.
-            \item \emph{(josh)} Find real parameters to put into this. 
-              \jss{added some notes for TFBS gain and loss calculations from Lynch} \ref{sec:param}
-        \end{itemize}
-
-    \item \emph{(josh)} Write out simple examples in biological words.
-
-    \item \emph{(josh)} Rephrase hybrid inviability example in biological terms.  Discuss F1's (or heterozygotes) versus F2's.
-
-    \item \emph{(josh)} Clarify our assumptions around $x(0)$.
-
-    \item \emph{(josh)} Note we don't include $D$ for simplicity but don't.
-
-    \item \emph{(peter)} Figure out how to specify $P$ for Kalman decomposition.
-
-    \item \emph{(josh \jss{\checkmark},peter)} Prove or find citation for uniqueness of minimal system up to basis change.
-\jss{See \citep[Chapter 10.6, Theorem 6.9: Fundamental Theorem of Linear Realization Theory.]{kalman1969topics}}
-
-    \item Work examples (minimal; not minimal; with/without B,C fixed) of Kalman decomposition and description of moving along the walk.
-        \begin{itemize}
-            \item \emph{(josh)} Work out parameterization.
-            \item \emph{(peter)} Describe Brownian motion.
-        \end{itemize}
-
-    \item \emph{(peter)} Write down and decide between options for parameterizing the random walk on the neutral manifold:
-        use Kalman decomposition, and parameterize Brownian motion so that it has constant speed;
-        OR consider Josh's parameteriziation like this;
-        OR just say we do Brownian motion on the manifold (more or less implicit characterization).
-
-    \item \emph{(peter)} Describe model leading to constraints on regulation so we have a stationary distribution.
-        Ask questions of stationary distribution: robustness, connectivity, xxx?
-
-    \item Consider options for fitness function: Euclidean? Sign of eigenvalues?  Time domain or frequency?
-      First try on simple concrete examples. \jss{see \ref{sec:del}}
-
-    \item Figure out coherent model for changing dimension: do coefficients stay stuck at zero? In what parameterization? 
-        Do we need to allow deleterious variation to get deletion?
-
-    \item Write something in discussion about nonlinear systems.  Do an example if there is a feasible one.
-
-    \item Simulate.
-
-\end{enumerate}
-
-% Outline:
-% 
-% - What the problems are:
-% 
-%     * evolution of perfectly fit populations
-% 
-% - Motivations: Can you infer function from structure? And...
-% 
-%     * motifs
-%     * speciation
-%     * network size
-%     * rube-goldberginess
-%     * redundancy, robustness
-%     * evolvability related to minimality?
-% 
-% - Give a concrete, simple example.
-% 
-% - Modeling framework (LTI systems)
-%
-%     * y = "phenotype"
-%     * x = "internal state" or "kryptotype"?
-%     * A = "regulatory matrix" or "genetic network"?
-%     * Kalman decomposition
-% 
-% - Evolution of free parameters in the Kalman decomposition
-%
-%     * Give or refer to earlier simple description of population dynamics which makes the system evolve.
-%     * What do we want to know from this model:
-%
-%       - How fast do the internal dynamics change?
-%       - Stationary distribution on network complexity/shape/properties
-%       - How fast does incompatibility accumulate?
-%
-%     * Brownian motion on free parameters
-%     * how mutations map into this space
-%     * dimension: duplication, deletion, recruitment of new things
-% 
-% - (Compute things)
-%
-% - Generalize ideas to nonlinear systems
-%
-%   * compute isomorphic or dynamically equivallent nonlinear systems of the feed-forward motif or of the three-node adaptation motif. 
-% 
-% - (Allow non-fit stuff too.)
-
-%%% Questions:
-
-% is y the phenotype or H(z) the phenotype? Do we need new jargon anyway?
-
-\title{The Evolution of Phenotypically Invariant Genetic Networks}
-\author[1]{Joshua S. Schiffman}
-\author[1,2]{Peter L. Ralph}
-\affil[1]{University of Southern California, Los Angeles, California}
-\affil[2]{University of Oregon, Eugene, Oregon}
-\affil[ ]{\texttt{jsschiff@usc.edu} \qquad \texttt{plr@uoregon.edu}}
-\maketitle
-
-\begin{abstract}
+\section*{Abstract}
 I will outline an analytical theory to study the evolution of biological systems such as gene regulatory networks, borrowing insight and tools from control engineering, systems identification, and dynamical systems theory.
@@
 This analysis provides insight into the mechanisms and parameters important for understanding developmental systems drift, network rewiring, evolvability, epistasis, and speciation, as well as the tenuous connection between network architecture and function.
-\end{abstract}
 
-\section{Background} \label{sec:intro}
-
-\plr{Review of relevant other stuff.}
-
-\jss{Introduction}
+\section*{Introduction}
 
@@
 
-\plr{phenomenologically or dynamically equivalent?}
-The set of all systems -- that is all model structures and all parameter schemes -- that
-are externally, or phenomenologically, equivalent, biologically realizable, and mutationally accessible,
+\section*{The Model I: Gene Networks as Linear Systems}
 
-What are typically referred to as ``design principles'' or some sort of clear structure-function relationships are commonly saught \citep{milo2002network, alon2007network, ma2009defining}.
+  Organisms' phenotypes are constructred by complex gene by gene and gene by environment interactions.
+Although a very simplified description, here we define the phenotype to be the spatio-temporal molecular dynamics directly under natural selection.
+Basically the \emph{what}, \emph{when}, and \emph{where}, of an organism's molecules that are physiologically or otherwise important and relevant to survival.
+Thus we say that some function $f(t)$ is a phenotype where,
+  \begin{align}
+    f(t) = \int_{0}^{\infty} h(t) u(t) dt  ,
+  \end{align}
+  and $h(t)$ is the \emph{impulse response} or \emph{kernel} of the system and $u(t)$ is the \emph{input} function.
+The input can be interpreted as the environment, or as initial conditions, or otherwise, depending on the biological specifics under study.
 
-Seeking to attribute functionality, uninformed by evolution, can lead to spurious claims
-(i.e. see Graur \citep{lynch2007evolution, lynch2007frailty, graur2013immortality}).
+  Essentially the phenotype $f(t)$ is a consequence of an organism's specific gene by gene interactions, given by $h(t)$, reacting further with the local environment, given by $u(t)$.
 
-A linear time invariant (LTI) dynamical system is a system of linear
-differential equations that describes a physical process. 
-LTI systems are used in electrical and control engineering to model a myriad of phenomena including
-circuits. 
-One can use this methodology, under a set of assumptions, to reverse engineer a
-mechanism from impulse experiments (input/output data). The idea is that given
-a black box, and it's experimental manipulation, can we describe the internal
-mechanisms?
+  We describe the impulse response as, 
+  \begin{align}
+    h(t) = C e^{A t} B
+  \end{align}
+  where $A$ is a gene network -- a square matrix, and $B$ filters and translates the input to the system.
+The form of $B$ determines precisely how the state of the external environment influences the internal gene network.
+$C$ filters and translates the dynamics of the system and precesily determines the output, that is, what is visible to selection.
 
-\section{The problem}
+  Generally $A$ can be any real $n \times n$ matrix, $B$ any $n \times \ell$, and $C$ any $\ell \times n$ dimensional matrix.
+However, for simplicity in exposition (and without loss of generality) we set $\ell = 1$, so that $B$ and $C$ are simply vectors of length $n$.
 
-Organisms often have to respond to their environments,
-and the way this happens is that external input triggers a cascade of molecular signals
-whose eventual result is a reaction,
-whose approprateness for the situation determines fitness.
-\plr{A similar situation occurs in development (...?).}
-However, there are many ways to skin a cat \jss{(there are also many ways to develop a cat)}:
-especially given a number of possible molecular intermediaries,
-there may be a large number of ways to construct such a molecular signalling system
-that has precisely the \emph{same} input-output relationships,
-and hence the same function.
-Therefore, any mutation that changes one such system to another, equivalent version
-will be neutral,
-and may well become polymorphic in a population if there are no ill effects in heterozygotes.
-Certainly, mutations that move a system away from optimal functioning will also appear,
-but in very large populations will not rise to high frequency.
-\plr{Deleterious things are quite important but somehow justify starting with the neutral ones.}
+  Above, $t$ refers to time.
+Although $f(t)$ describes the phenotype given an input, $h(t)$ describes the phenotype subject only to an impulse -- an input present initially and absent immediately thereafter.
+Typically, a system $\Sigma$ is defines as,
+  \begin{align}
+    \Sigma = \left\{ \begin{array}{ll} \dot{x}(t) &= A x(t) + B u(t) \\ y(t) &= C x(t) \end{array} \right.
+  \end{align}
+  Variables have the same identities as described above and $x(t)$ is a vector of molecule concentrations at time $t$.
+Therefore the molecular concentrations at a specific time are completely determined by the environmental input and gene by gene interactions.
+Lastly, a portion and/or combination of these molecules, $y(t)$, are ``observed'' by selection.
 
-\plr{
-    More stuff on why we care about this:  
-Provide neutral model; can address questions.  
-Main point of paper.
-}
+  \section*{The Model II: Linear Evolutionary Systems}
 
-\begin{example}[Example: Oscillator.] \label{ex:osc}
-Motivate the system:
-\begin{align*}
-    A &= \left[\begin{matrix} 
-        0 & 1 \\ 
-       -1 & 0 
-    \end{matrix}\right]
-\end{align*}
+  Systems with identical external dynamics do not necessarily have identical internal dynamics.
+Any linear and minimal system -- minimal, informally meaning that the system's external dynamics are achieved with the fewest possible number of internal components -- has identical external dynamics up to a change of coordinates.
+  \begin{align}
+    h(t) &= C e^{A t} B \\
+    &= C V^{-1} e^{V A V^{-1} t} V B \\
+    &= C V^{-1} V e^{A t} V^{-1} V B \\
+    &= C e^{A t} B
+  \end{align}
+  Two systems, $\Sigma = \{ A, B, C \}$, and $\bar{\Sigma} = \{\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} \}$, have the same dynamics if they are related by a change of coordinates.
 
-    \begin{figure}
-        \begin{center}
-            \begin{tikzpicture}
-            \begin{scope}[every node/.style={circle,thick,draw}]
-                \node (A) at (0,0) {$x_{1}$};
-                \node[dashed] (B) at (3,0) {$x_{2}$};
-                \node[shape=rectangle] (U) at (1.5,2) {input};
-                \node[shape=rectangle] (y) at (1.5,-2) {output};
-            \end{scope}
+  Although systems may not be identifiable up to a change of coordinates, at present we are primarily interested in a subset of these systems.
+That is, systems that not only have equivalent external dynamics, but also equivlanet input relationships and under equivalent selective pressures.
+Formally, systems related by a change of coordinates that leave $B$ and $C$ invariant:
+  \begin{align}
+    VB &= B \implies \bar{B} = B \\
+    CV &= C \implies \bar{C} = C
+  \end{align}
+  In other words systems with varying genetic architectures yet identical selection pressures, environment, and phenotype.
 
-            \begin{scope}[>={Stealth[black]},
-                          every node/.style={fill=white,circle},
-                          every edge/.style={draw=black, thick}]
-                \path [->, >=Rectangle] (A) edge[bend left] node {\tiny $-1$} (B);
-                \path [->] (B) edge[bend left] node {\tiny $1$} (A); 
-                \path[->] (U) edge[dashed] node {\tiny $1$} (A);
-                \path[->] (U) edge[dashed] node {\tiny $1$} (B);
-                \path[->] (A) edge[dashed,bend right] node {\tiny $1$} (y);
-            \end{scope}
-            \begin{scope}[>={Stealth[black]},
-                          every edge/.style={draw=black, thick}]
-                %\path [->] (A) edge[loop left] node {\tiny $\lambda_{1}$} (A);
-                %\path [->] (B) edge[loop left] node {\tiny $\lambda_{2}$} (B);
-            \end{scope}
-
-            \end{tikzpicture}
-        \end{center}
-        \caption{
-            Diagram of Example \ref{ex:osc} in the text.  
-            \plr{explain what arrows mean if nec}
-            \label{fig:oscillator_diagram}}
-    \end{figure}
-
-as a special case of the below.
-
-\end{example}
-
-\begin{example}[Example: two-component systems.] \label{ex:2x2}
-As a more general example,
-suppose that a cell produces two molecular species, $S_1$ and $S_2$ 
-whose production are both stimulated by the presence of an external ``input'' substance.
-\plr{call them TF?}
-Suppose that the two species self-regulate with strengths $\lambda_1$ and $\lambda_2$, respectively,
-and the second species also regulates the first species with strength $\gamma$.
-\plr{Rephrase in terms of binding to promoter sequences?}
-However, only the time course of the concentration of
-the first one of these species determines the fitness of the organism.
-(As happens for instance in \plr{XXX}.)
-Write $x(t) = (x_1(t),x_2(t))$ for the concentrations of the two species at time $t$,
-and write $u(t)$ for the concentration at time $t$ of the input substance.
-Then, if the rates at which each species are produced
-are linear functions of the concentrations
-\plr{(look up way this is usually said in the literature.)},
-then the dynamics of the system are given by
-\begin{align*}
-    \ddt x_1(t) &= \lambda_1 x_1(t) + \gamma x_2(t) + u(t) \\
-    \ddt x_2(t) &= \lambda_2 x_1(t) + u(t) ,
-\end{align*}
-where $\ddt x$ denotes the time derivative.
-The initial conditions, $x_1(0)$ and $x_2(0)$, 
-and the input $u(t)$ then determine the concentrations through time.
-If we record the regulatory coefficients in the matrix:
-\begin{align*}
-    A &= \left[\begin{matrix} 
-        \lambda_1 & \gamma \\ 
-       0 & \lambda_2  
-    \end{matrix}\right],
-\end{align*}
-and define the column vector $B = [1,1]^T$,
-then in matrix notation the dynamics are
-\begin{align*}
-    \ddt x(t) &= A x(t) + B u(t) .
-\end{align*}
-If these are all transcription factors,
-and we suppose their binding motifs are fixed,
-then the $i^\text{th}$ row of $A$ is determined by the $i^\text{th}$ promoter sequence.
-
-\end{example}
-
-Since the system is linear, the state of the system at any time
-is the superposition of its responses to all previous inputs.
-This implies that 
-if $G(t)$ is the transient response of the system to a unit impulse,
-$t$ units of time later,
-and $x(0) = 0$,
-then the time course of the concentration of the thing we care about, $x_1(t)$, 
-can be written as
-\begin{align} \label{eqn:ex_convolution}
-    x_1(t) &= \int_0^t G(t-s) u(s) ds .
+  Define $V(\tau)$ as the change of coordinates matrix that preserves $B$ and $C$, with $\tau$ a vector of free parameters. The set of \emph{all} phenotypically invariant (minimal) gene networks is, 
+  \begin{align}
+    A(\tau) = V(\tau) A(0) V^{-1}(\tau) ,
   \end{align}
-However, it turns out that 
-\plr{look for simple demontration this is true}
-\begin{align*}
-    P_{p} = \left[\begin{matrix} 
-        1 & 0 \\ 
-        p & 1-p 
-    \end{matrix}\right],
-\end{align*}
-then for any $p \neq 1$, 
-if we replace the regulatory matrix $A$ with
-\begin{align*}
-    A(p) = P A P^{-1},
-\end{align*}
-then the response of $x_1(t)$ to any particular input $u(t)$ is identical to the original system.
-(However, $x_2$ may well be different!)
-
-For instance, setting $p=-1$,
-the system with
-\begin{align*}
-    A(-1) = \left[\begin{matrix} 
-        \lambda_1 + \gamma/2 & \gamma/2 \\ 
-        \lambda_2-\lambda_1-\gamma/2 & \lambda_2 - \gamma/2
-    \end{matrix}\right],
-\end{align*}
-looks very different, but gives the same input-output relationship.
-Although these systems are equivalent,
-hybrids between them may not be:
-\plr{do example}.
-
-\begin{example}[Hyrbid Inviability]
-  Let, \begin{align*}
-    A(0) =\left[\begin{matrix}
-      0 & 1 \\
-      -1 & 0
-    \end{matrix}\right]
-    \text{and }
-    A(2) = \left[\begin{matrix}
-      2 & -1 \\
-      5 & -2
-    \end{matrix}\right]
-  \end{align*}
-  from Example \ref{ex:osc} ($B=\begin{bmatrix} 1 \\ 1 \end{bmatrix}$ and $C=\begin{bmatrix} 1 & 0 \end{bmatrix}$). 
-If hybrids are formed by exchanging genes (swapping rows), the hybrid's gene network is,
-  \begin{align*}
-    \text{hybrid network } := \mathcal{H}(A(0), A(2)) = \left[\begin{matrix}
-      2 & -1 \\
-      -1 & 0
-    \end{matrix}\right] ,
-  \end{align*}
-  \jss{ Or should it be: \begin{align*}
-    \text{hybrid network } := \text{mean}(A(0), A(2)) = \left[\begin{matrix}
-      1 & 0 \\
-      2 & -1
-    \end{matrix}\right] ,
-  \end{align*} }
-  It's transfer function will be $H(z, \mathcal{H}) = \frac{z-2}{z^2 - 3z +1}$, which produces very different dynamics than its parents, $H(z, A(0)) = H(z, A(2)) = \frac{z+1}{z^2 + 1}$.
-The time dynamcs of molecular species $1$ in the hybrid is $y^{(\mathcal{H})} =e^{t} \cosh(\sqrt{2} t)$ \jss{ which is approx.
-$\approx e^{t}$}, with the concentration of species $1$ rapidly heading towards infinity.
-The dynamics of species $1$ in the parental networks is $y = \sin(t) + \cos(t)$, where it continuously oscillates.
-
-  \jss{The transfer function $H(z, \mathcal{H})$, by averaging networks, is $\frac{1}{z-1}$ and the time dynamics are: $y^{(\mathcal{H})} = e^{t}$.}
-\end{example}
-
-\begin{definition}[Biological Species Concept]
-  Let $\mathcal{A}$ be the set of all phenotypically symmetric gene networks with transfer funtion $H(z)$. 
-  Choose $A$ and $\widetilde A$, both in $\mathcal{A}$. 
-  If $\mathcal{H}(A, \widetilde A) \notin \mathcal{A}$ then $A$ and $\widetilde A$ are distinct species. 
-\end{definition}
-
-\begin{figure}
-  \centering
-  \begin{subfigure}{0.5\textwidth}
-    \centering
-\begin{tikzpicture}
-\begin{scope}[every node/.style={circle,thick,draw}]
-  \node (A) at (0,0) {$x_{1}$};
-    \node[dashed] (B) at (4,0) {$x_{2}$};
-    \node[shape=rectangle] (U) at (2,2) {input};
-    \node[shape=rectangle] (y) at (2,-2) {output};
-\end{scope}
-
-\begin{scope}[>={Stealth[black]},
-              every node/.style={fill=white,circle},
-              every edge/.style={draw=black, thick}]
-    %\path [->] (A) edge[bend left] node {$c$} (B);
-    \path [->] (B) edge node {\tiny $\gamma$} (A); 
-    \path[->] (U) edge[dashed] node {\tiny $1$} (A);
-    \path[->] (U) edge[dashed] node {\tiny $1$} (B);
-    \path[->] (A) edge[dashed, bend right] node {\tiny $1$} (y);
-\end{scope}
-\begin{scope}[>={Stealth[black]},
-              every edge/.style={draw=black, thick}]
-    \path [->] (A) edge[loop left] node {\tiny $\lambda_{1}$} (A);
-    \path [->] (B) edge[loop right] node {\tiny $\lambda_{2}$} (B);
-\end{scope}
-
-\end{tikzpicture}
-    \caption{Gene Regulatory Network $A(0)$.}
-  \end{subfigure}%
-  \begin{subfigure}{0.5\textwidth}
-    \centering
-\begin{tikzpicture}
-\begin{scope}[every node/.style={circle,thick,draw}]
-  \node (A) at (0,0) {$x_{1}$};
-    \node[dashed] (B) at (4,0) {$x_{2}$};
-    \node[shape=rectangle] (U) at (2,2) {input};
-    \node[shape=rectangle] (y) at (2,-2) {output};
-\end{scope}
-
-\begin{scope}[>={Stealth[black]},
-              every node/.style={fill=white,circle},
-              every edge/.style={draw=black, thick}]
-    \path [->, sloped] (A) edge[bend left] node {\tiny $(\lambda_{2} - \lambda_{1} - \frac{\gamma}{2})$} (B);
-    \path [->, sloped] (B) edge[bend left] node {\tiny $\frac{\gamma}{2}$} (A); 
-    \path[->] (U) edge[dashed] node {\tiny $1$} (A);
-    \path[->] (U) edge[dashed] node {\tiny $1$} (B);
-    \path[->] (A) edge[dashed, bend right] node {\tiny $1$} (y);
-\end{scope}
-\begin{scope}[>={Stealth[black]},
-              every edge/.style={draw=black, thick}]
-    \path [->] (A) edge[loop left] node {\tiny $(\lambda_{1} + \frac{\gamma}{2})$} (A);
-    \path [->] (B) edge[loop right] node {\tiny $(\lambda_{2}-\frac{\gamma}{2})$} (B);
-\end{scope}
-
-\end{tikzpicture}
-    \caption{Gene Regulatory Network $A(-1)$.}
-  \end{subfigure}
-  \caption{Neutral rewiring of gene network $A(p)$ from Example \ref{ex:2x2}. Not only are the regulatory coefficients different between $A(0)$ and $A(-1)$, there is also a new regulatory connection.}
-\end{figure}
-
-\section{Regulatory networks as linear, time-invariant systems}
-
-\plr{Make sure to have strongly worded disclaimer somewhere
-that we know real networks aren't linear
-but they are locally.
-Plus, it works in engineering.}
-
-\jss{Maybe we should discuss the Hartman-Grobman Theorem?}
-
-% - General description of the LTI process in biological context
-%     * y = "phenotype"
-%     * x = "internal state" or "kryptotype"?
-%     * A = "regulatory matrix" or "genetic network"?
-% - Why transfer function determines input-output relationship
-% - Kalman decomposition in unique form
-
-We will now lay out the model in more general terms.
-Suppose that the \emph{internal state} of the system
-is parameterized by the concentrations of a collection of $n$ molecular species,
-$S_1, \ldots, S_n$,
-and the vector of concentrations at time $t$ we denote $x(t)=(x_1(t),\ldots,x_n(t))$.
-There are also $m$ ``input'' species, whose concentrations are determined
-exogenously to the system,
-and are denoted $u(t) = (u_1(t),\ldots,u_m(t))$,
-and $\ell$ ``output'' species, whose concentrations are denoted
-$y(t) = (y_1(t),\ldots,y_\ell(t))$.
-The output is merely a linear function of the internal state:
-\begin{align*}
-    y_i(t) = \sum_j C_{ij} x_i(t).
-\end{align*}
-Since $y$ is what natural selection acts on, we refer to it as the \emph{phenotype},
-and sometimes in contrast refer to $x$ as the \emph{kryptotype},
-as it is ``hidden'' from direct selection.
-\plr{Maybe.}
-The rate at which the $i^\text{th}$ species is produced
-is a weighted sum of the concentrations of the other species
-as well as the input:
-\begin{align*}
-    \ddt x_i(t) = \sum_j A_{ij} x_j(t) + \sum_k B_{ik} u_k(t) .
-\end{align*}
-In matrix notation, this is written more concisely as
-\begin{align} \label{eqn:lti_system}
-    \ddt x(t) &= A x(t) + B u(t) \\
-    y(t) &= C x(t) .
+  and a Linear Evolutionary System is, 
+  \begin{align}
+    \Sigma(\tau) = \left\{ \begin{array}{ll} \dot{x}(t) &= A(\tau) x(t) + B u(t) \\ y(t) &= C x(t) \end{array} \right .
   \end{align}
 
-\plr{Everything we say I think is assuming $x(0)=0)$.  
-Figure this out. ``Zero-state equivalence.''}
+  Evolution thus proceeds as a random walk in phenotypically invariant network space.
 
-Given an initial condition and an input, 
-it is possible to write the solution $x(t)$
-as a convolution with an input-response kernel as in the example above
-(equation \eqref{eqn:ex_convolution}).
-An alternative way of describing the input-output relationship,
-more common in engineering,
-is instead to find the \emph{transfer function}
-\begin{align} \label{eqn:transfer_fn}
-    H(z) = C(z I - A)^{-1} B,
+  \begin{align}
+    A(\tau) \xrightarrow{T} A(\tau + \epsilon)
   \end{align}
-which is more mathematically readable \plr{if you know what it means}.
-This can be thought of as the system's response to input at ``frequency'' $z$.
-The fact that the transfer function uniquely determines the system's
-input-output relationship (i.e., the mapping $u \mapsto y$)
-follows from the fact that if $\tilde y(z) = \int_0^\infty e^{-zt} y(t) dt$
-is the Laplace transform of $y$,
-and $\tilde u(z)$ is the Laplace transform of $u$,
-then
-\begin{align*}
-    \tilde y(z) = H(z) \tilde u(z) .
-\end{align*}
-(This happens because the Laplace transform takes convolutions to products,
-and $H$ is the Laplace transform of the kernel $G$.)
-Concretely,
 
-\plr{Should we remind what dimensions everything is?}
+  \section*{Phenotypic Invariance}
 
-The state space representation is considered an internal description of the
-system, whereas the transfer function is an external description. 
-``Realizations'' are something.
+  \section*{Examples}
+    \subsection*{Cell Cycle Control or Circadian Rhytm}
 
-\begin{definition}[Phenomenological equivalence of systems]
-    Let $(x(t),y(t))$ and $(\bar x(t),\bar y(t))$ be the solutions to \eqref{eqn:lti_system}
-    with coefficient matrices $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ respectively,
-    and both $x(0)$ and $\bar x(0)$ are zero. 
-    The systems defined by $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ are
-    \textbf{phenomenologically equivalent} 
-    if
-    \begin{align*}
-        y(t) = \bar y(t) \qquad \text{for all} \; t \ge 0.
-    \end{align*}
-    Equivalently, this occurs if and only if
-    \begin{align*}
-        H(z) = \bar H(z)  \qquad \text{for all} \; z \ge 0,
-    \end{align*}
-    where $H$ and $\bar H$ are the transfer functions of the two systems.
-\end{definition}
+    \subsection*{Metabolic Network}
 
-One way to find other systems equvalent to a given one
-is by change of coordinates (``algebraic equivalance''):
-if $T$ is an invertible matrix, then the systems $(A,B,C)$ and $(TAT^{-1},TB,CT^{-1})$
-have the same dyanamics because their transfer functions are equal:
-\begin{align*}
-    CT^{-1}( zI - TAT^{-1})^{-1}TB
-    =
-    CT^{-1}T( zI - A)^{-1}T^{-1}TB
-    =
-    C( zI - A)^{-1}B .
-\end{align*}
-However, the converse is not necessarily true: 
-systems can have identical transfer functions without being changes of coordinates of each other.
-In fact, systems with identical transfer functions can involve interactions between different
-numbers of molecular species.
+    \subsection*{Gap Gene Network}
 
-The set of all systems phenomenologically equivalent to a given system $(A,B,C)$ 
-is elegantly described using the Kalman decomposition,
-which also clarifies the system dynamics? tells us a lot about how it works? \plr{or something}
-To motivate this, first note that the the input $u(t)$ only directly pushes the system
-in directions lying in the span of the columns of $B$.
-As a result, different combinations of input can 
-move the system in any direction that lies in the \emph{reachable subspace},
-which we denote by $\reachable$,
-and is defined to be the closure of $\spn(B)$ under applying $A$
-(or equivalently, the span of $B, AB, A^2B, \ldots A^{n-1}B$).
-Analogously to this, we define
-the \emph{observable subspace}, $\mathcal{O}$,
-to be the closure of $\spn(C^T)$ under applying $A$.
-(Or: $\unobservable$ is the largest $A$-invariant subspace
-contained in the null space of $C$;
-and $\reachable$ is the largest $A$-invariant subspace contained in the image of $B$.)
-
-If we define
-\begin{enumerate}
-    \item The columns of $P_\rno$ are an orthonormal basis for $\reachable \cap \unobservable$.
-    \item The columns of $P_\ro$ are an orthonormal basis of
-        the complement of $\reachable \cap \unobservable$ in $\reachable$.
-    \item The columns of $P_\nro$ are an orthonormal basis of
-        the complement of $\reachable \cap \unobservable$ in $\unobservable$.
-    \item The columns of $P_\nrno$ are an orthonormal basis of
-        the the remainder of $\R^n$.
-\end{enumerate}
-If we then define
-\begin{align*}
-    P &= 
-    \left[ \begin{array}{c|c|c|c}
-        P_\rno & P_\ro & P_\nro & P_\nrno
-    \end{array} \right] ,
-\end{align*}
-then
-\begin{align*}
-    P^T P
-    &=
-    \left[ \begin{array}{c|c|c|c}
-        I & 0 & 0 & 0 \\
-        \hline
-        0 & I & U & 0 \\
-        \hline
-        0 & V & I & 0 \\
-        \hline
-        0 & 0 & 0 & I 
-    \end{array} \right] .
-\end{align*}
-\plr{Check this.  Can we get $U=V=0$?}
-
-The following theorem can be found in SOME REFERENCE.
-
-\begin{theorem}[Kalman decomposition] \label{thm:kalman}
-        For any system $(A,B,C)$ with corresponding Kalman basis matrix $P$,
-        the transformed system $(PAP^{-1},PB,CP^{-1})$  has the following form:
-        \begin{align*}
-            \widehat A = PAP^{-1}
-            &=
-            \left[ \begin{array}{cccc}
-                A_{\rno} & A_{\rno,\ro} & A_{\rno,\nrno} & A_{\rno,\nro} \\
-                0 & A_{\ro} & 0 & A_{\ro,\nro} \\
-                0 & 0 & A_{\nrno} & A_{\nrno,\nro} \\
-                0 & 0 & 0 & A_{\nro}
-            \end{array} \right] ,
-        \end{align*}
-        and
-        \begin{align*}
-            \widehat B = PB
-            &=
-            \left[ \begin{array}{cccc}
-                B_{\rno} \\
-                B_{\ro} \\
-                0 \\
-                0 
-            \end{array} \right] ,
-        \end{align*}
-        and
-        \begin{align*}
-            \widehat C = CP^{-1}
-            &=
-            \left[ \begin{array}{cccc}
-                0 & C_{\ro} & C_{\nrno} & 0 
-            \end{array} \right] .
-        \end{align*}
-        The transfer function of both systems is given by
-        \begin{align*}
-            H(z) = C_{\ro} ( zI - A_{\ro} )^{-1} B_{\ro} .
-        \end{align*}
-\end{theorem}
-
-In the latter case, we say that the system is \emph{minimal} 
--- there is no equivalent system with a smaller number of species.
-Note that this says that any two equivalent minimal systems
-are changes of basis of each other.
-
-Since any system can be put into this form,
-and once in this form, its transfer function is determined only by 
-$C_{\ro}$, $A_{\ro}$, and $B_\ro$,
-therefore, the set of all equivalent systems are parameterized by
-the dimension $n$,
-the choice of basis ($P$),
-the remaining submatrices in $\widehat A$, $\widehat B$, and $\widehat C$
-(which are unconstrained),
-and a invertible transformation of $\spn(P_{\ro})$, which we call $T_\ro$.
-
-\begin{theorem}[Parameterization of equivalent systems]
-    Let $(A,B,C)$ be a minimal system.
-    \begin{itemize}
-        \item[(a)]
-            Every equivalent system is of the form given in Theorem \ref{thm:kalman},
-            i.e., can be specified by choosing a dimension, $n$;
-            submatrices in $\widehat A$, $\widehat B$, and $\widehat C$ 
-            except for $A_\ro=A$, $B_\ro=B$, and $C_\ro=C$;
-            and choosing an invertible matrix $P$.
-
-        \item[(b)]
-            \plr{conjecture:}
-            The parameterization is unique
-            if $P$ is futhermore chosen so that 
-            each $P_x$ other than $P_\ro$ is a projection matrix,
-            and that 
-            \begin{align*}
-                0
-                =
-                P_x^T P_y
-            \end{align*}
-            for all $(x,y)$ except $(\ro,\nrno)$.
-
-    \end{itemize} 
-\end{theorem}
-
-\plr{Another way of saying it: pick the $\reachable$ and $\unobservable$ subspaces,
-that must intersect in something of the minimal dimension;
-then let $P$ be the appropriate basis?}
-
-In some situations we may be interested in only ``network rewiring'',
-where $A$ changes while $B$ and $C$ do not.
-For instance, 
-if all non-regulatory functions of each molecule are strongly constrained,
-then $C$ cannot change.
-Likewise, if responses of each molecule to the external inputs are not changed by evolution,
-then $B$ does not change.
-
-\begin{lemma}
-    Let $(A,B,C)$ be a system.
-    Every matrix $\widehat A$ such that $(\widehat A,B,C)$ is equivalent to $(A,B,C)$
-    can be specified in the form given in Theorem \ref{thm:kalman},
-    with \plr{them having the same $\ro$ bit}
-    and satisfying
-    \begin{align*}
-        \widehat B &= P B \\
-        \widehat C P &= C
-    \end{align*}
-\end{lemma}
-
-It is remarkable to note that even with the relationships 
-between environment, cryptotype, and phenotype constant,
-and in the minimal dimension,
-there are still almost always degrees of freedom.
-These correspond to distinct genetic networks
-that perform indistinguishable functions.
-For example, the equivalent systems of Example \ref{ex:2x2} above
-are minimal, and share common $B$ and $C$ matrices.
-
-\plr{Maybe say that stability to noise can be determined by $A_\nro$ (and others?),
-while complexity of the internal, unobserved dynamics
-are somehow given by the difference between the dimension of $y$ and the dimension of $A_\ro$.}
-
-\plr{General argument why Brownian motion on set of equivalent things is maybe ok.}
-
-\begin{example}
-    The system in Example \ref{ex:2x2} is minimal, so $A=A_\ro$,
-    and the only free parameter is $P$.
-    Suppose also that the effects of the input and output are strongly constrained,
-    so that as in that example,
-    $B$ and $C$ cannot change,
-    and neutral evolution moves through the set of of $P$ such that $PB=B$ and $CP=C$.
-    \plr{Clarify we mean with no gene duplication and only neutral drift.}
-    How will this system change over evolutionary time in the manner described \plr{above}?
-    An unbiased Brownian motion on this set 
-    can be written in terms of a one-dimensional Brownian motion $\beta_\tau$ 
-    with $\beta_0=0$ as
-    \begin{align*}
-        P(\tau) = \left[
-            \begin{array}{cc}
-                1 & 0 \\
-                e^{\beta_\tau}-1 & e^{\beta_\tau}
-            \end{array}
-            \right] .
-    \end{align*}
-    Therefore, the state of the system after (evolutionary) time $\tau$
-    is $(P(\tau)AP(\tau)^{-1},B,C)$.
-\end{example}
-
-\section{Estimates of network rewiring and developmenal systems drift.} \label{sec:param}
-
-\jss{Parameter estimates for DSD $\downarrow$}
-
-What parameters (and other relevant factors) do we need from the literature and experiment? 
-  \begin{itemize}
-    \item mutation rate
-    \item Transcription factor binding site appearence and loss
-    \begin{itemize}
-      \item In \citep{lynch2007evolution} TF binding site gain ($u_{g}$) and loss ($u_{l}$) is estimated to be,
-        \begin{align*}
-          u_{g} &= \frac{(L-n+1)nu}{4^{n}} \\
-          u_{l} &= nu
-        \end{align*}
-      Where $L$ is the length of the promoter, $n$ is the number of nucleotides in a TFBS, and $u$ is the mutation rate per base pair per generation. 
-    \item \citet{lynch2007evolution} further states that in prokaryotes $L$ will usually be $\leq 100$ and in eukaryotes $L$ can range from $10^{3}$ to $10^{6}$.
-He also cites \citep{harbison2004transcriptional} that TFBSs range from $5-10$bps in length.
-    \item in \citep{lynch2014evolutionary}, they estimate
-        \begin{align*}
-          \tilde{P}(m) = C \left[ 3^{\ell - m}  \left( \begin{array}{c} \ell \\ m \end{array} \right) \right] e^{2N_{e}s(m)}  ,
-        \end{align*}
-        where $\tilde{P}(m)$ is the stationary distribution of a TFBS with match (and strength) $m$.
-$C$ is the normalizer constant, $m$ is the number of matches, $\ell$ is the length of the TFBS motif.
-The term on the right is the ratio of fixation probabilities.
-        Or in their own words, there are two interpretations: 
-        \begin{quote} ``(i) For a single TFBS, it represents the long-term proportion of evolutionary time spent in the various matching states and (ii) for a set of different TFBSs under the same selective constraints, it represents the expected distribution of states at any point in time.'' \end{quote}
-      \item \citet{tuugrul2015dynamics, anderson2015intermolecular} also may be helpful.
-    \end{itemize}
-    \item rate of both \emph{trans-} and \emph{cis-} regulatory evolution. (see Sergey's papers).
-      i.e. \citep{nuzhdin2004common, fear2016buffering, graze2009regulatory, graze2012allelic, genissel2008cis}
-    \item The PWN-score landscape (does binding strength increase linearly, exponentially, have a threshold, etc.) with respect to SNPs.
-    \item Others? 
-  \end{itemize}
-
-\section{(Old Notes) What biological question are we asking?}
-
-How do gene regulatory networks (GRNs) change through evolutionary time?
-Specifically, under constant selection and environmental pressures, how does a
-maximally fit population's GRN change and what molecular and population
-parameters are significant to the process? How frequent does gene network
-rewiring occur in the absense of genetic drift or adaptation? Is rewiring
-typically compensatory and ususally follow the fixation of a deleterious
-allele, or can networks neutrally reorganize? How does the size, average
-degree, function, peiotropy, etc of a network influence its likelihood to
-drift? If networks can drift, how much topological heterogeneity can a
-well-mixed population tolerate, if any? Does topoligical heterogeneity and/or
-the size of the neutral genotype-set, if consequences of network organization,
-constrain or entail exaptation and evolvability? Are some network organizations
-(maybe higher dimensional realizations?) able to more rapidly evolve to
-construct novel phenotypes and meet the demands of changes in selection
-pressures?
-
-\section{(Old Notes) How can we mathematically model this question?}
-
-In some respect an organism can be thought of as a black box that responds to
-some set of inputs (it's environment and other initial conditions) and outputs
-a phenotype. The phenotype is then evaluated by natural selection. The black
-box metaphor holds, because much of the details of what happens to construct a
-phenotype are unimportant as far as selection is concered, so long as the
-phenotype reliably performs its function. If multiple internal mechanisms can
-map the same set of inputs to the same set of outputs, and these mechanisms are
-mutationally nearby (in genotype space), then evolution may drift from
-mechanism to mechanism.
-
-In this simple model we can say that $B\vec{u}$ is a list of initial GRN
-protein concentrations determined by the more general initial environmental
-conditions $\vec{u}$. $A$ represents all the genetic interactions of a
-particular GRN, and $C\vec{x}$ is an organism's phenotype. Fitness could be
-calculated by comparing a subset of the phenotype to an optimum. Maybe,
-$f(\cdot) = e^{- \int_{a}^{b} \Vert G(s) - G^{*}(s) \Vert ds}$ Alternatively,
-fitness scores could simply assess whether or not a phenotype is within some
-range or breaks some threshold by some time point. Or possbily the simplest:
-any phenotype not exactly equal to $G(s)$ is lethal, and $G(s)$ is perfectly
-fit.
-
-Next, we will have to define mutation and recombination. Mutation needs to add
-and remove genetic interactions and recombination needs to shuffle genes during
-sexual reproduction. We could adapt methods from Lynch 2007 and Lynch and
-Hagner 2015 to model the probability of a TFBS appearing or disappearing due to
-mutational pressure. Every time a new binding site is gained or lost, the
-values within the A matrix are either modified or rows and columns are added
-removed following a specific set of rules. I am not sure if it would be easier
-to maintain a matrix size (say $m \times m$) where $m$ is arbitrarily large,
-with  most entries being zero, or if a matrix should only have as many
-dimensions as the number of active TFs. Recombination should then shuffle rows
-of the $A$ matrix randomly, as each row represents the regulatory region of a
-given gene (assuming only cis-regulation, and that the regulatory element is
-small enough to not break up during recombination).
-
-To study some of the above posed questions, we can simulate (or analytically
-determine) the rates of GRN change, and how frequently these changes lead to
-significant rewiring or speciation. We would want to know the dynamics of a
-brownian motion over the set of all realizations.  Another interesting question
-would be to determine the size of the neutral genotype space as well as the
-number of connections from the neutral genotype space to the non-neutral
-genotype space. Maybe higher dimensional realizations of a GRN will have
-significantly more connections to non-neutral genotypes and therefore be more
-``evolvable'' or exapted to novel environments?
-
-Let $M \in \mathbb{R}^{n \times n}$ be a least dimensional realization of $A$,
-and let $M'$ bean $(n+m \times n+m)$ matrix where $M'_{(i,j)} = M_{(i,j)} \
-\text{for } i,j \leq n, \ \text{and } 0 \ \text{otherwise}$. Let $P$ be any
-$(n+m \times n+m)$ matrix with rank $n$. Then for any matrix $B$ such that $BP
-= PM'$, $B$ is a realization of $A$.
-
-\section{(Old Notes) List of papers with some notes.}
-
-R. E. Kalman, “Mathematical description of linear dynamical systems,” Journal of the Society for Industrial and Applied Mathematics, Series A: Control, vol. 1, no. 2, pp. 152–192, 1963.
-This is a good overview written by Kalman. Probably the most important paper on this list.
-
-Minimal state-space realization in linear system theory: an overview. B. De Schutter. 2000.
-Another, more recent, overview that I found helpful to understand controllability and observability.
-
-The Inverse Problem of Stationary Covariance Generation. BDO Anderson. 1969.
-
-Equivalence of Linear Time- Invariant Dynamical Systems. BDO Anderson, BW Necomb, RE Kalman. 1966.
-
-B.
-Ho and R.
-E.
-Kalman, “Effective construction of linear state-variable models from input/output functions,” Proceedings of the 3rd Annual Allerton Conference on Circuit and System Theory, vol.
-1, no.
-2, p.
-449459, 1965.
-
-Dynamic Structure Functions for the Reverse Engineering of LTI Networks. J Goncalves. 2007.
-I have not read this paper yet, but apparently it outlines a method for recovering the state space equation from specific sets of measurements.
-
-New Developments in Systems Theory Relevant to Biology. RE Kalman. 1968.
-Not technical but kind of interesting.
-
-Old and New Directions of Research in Systems Theory. RE Kalman. 2010?
-
-Minimality and Observability of Group Systems. HA Loeliger, GD Forney Jr., T Mittelholzer, MD Trott. 1994. 
-Another paper I just found. It looks like it might be useful.
-
-Partial Realization of Descriptor Systems. P Benner and VI Sokolov. 2006. 
-I think this is about recovering a finite realization from infinite systems.
-Maybe it would be interesting to think about the similarity between transfer functions though?
-If selection only cares about a specific interval of the transfer function, say $G(10)$, then maybe two partial transfer functions will be effecrively equivlaent?
-
-There are also the previous papers I sent on identifiability and distinguishability.
-In the original Bellman and Astrom paper where they introduce ``structural identifiability'' they show that a linear dynamical system is identifiabile if and only if it is also controllable and observable.
-
-\section{Including Deleterious Mutations and a Fitness Function}\label{sec:del}
-
-  \begin{minipage}{0.5\textwidth}
-  At the moment we focus strictly on the case where the minimal realization is in $2$ dimensions and both of its eigenvalues are real and negative (so a special case of the system from Example \ref{ex:2x2}).
-\jss{(Note: this is not to say the ideas below will not be generalizable).}
-
-    The system of interest has the form,
-  \begin{align*}
-    \Sigma := \left\{ A = \begin{bmatrix} A_{11} & A_{12} \\ A_{21} & A_{22} \end{bmatrix}, B = \begin{bmatrix} 1 \\ 1 \end{bmatrix}, C = \begin{bmatrix} 1 & 0 \end{bmatrix} \right\}
-  \end{align*}
-  \end{minipage}%
-  \begin{minipage}{0.5\textwidth}
-    \centering
-    \includegraphics[width=0.95\textwidth]{figure_2x2mutation01}%
-      \captionof{figure}{The transfer function for $A = \left[ \protect\begin{array}{cc} -3 & 1 \\ 0 & -1 \protect\end{array} \right]$ and all eight $\mu(\Sigma, A_{ij})$ mutations, where $\upsilon = \pm 0.5$}\label{fig:mut}%
-  \end{minipage}
-
-Define the fitness function to be,
-
-\begin{align}\label{eq:fitness}
-  \Phi(\bar{\Sigma}) := \left\{ \begin{array}{ll}   e^{- \left| H(0) - \tilde{H}(0) \right| = - \left| \frac{k}{\lambda_{1} \lambda_{2}} - \frac{\tilde{k}}{\tilde{\lambda}_{1} \tilde{\lambda}_{2}} \right|}    &\text{if } \lambda_{1,2} < 0 \text{ and } \in \mathbb{R}, \ \det(A) \neq 0 \\
-  0   &\text{if } \lambda_{1,2} \geq 0 \ \text{ or } \ \notin \mathbb{R}, \ \det(A) = 0 \end{array} \right.
-\end{align}
-and let $\upsilon$ be the magnitude and direction (sign) of a mutation on regulatory interaction strength.
-The fitness function (\ref{eq:fitness}) may me appropriate if mutations only slightly alter the dynamics of the system, as in Figure \ref{fig:mut}.
-
-%Further let $\mu(\Sigma, A_{ij})$ be the resulting gene regulatory network following a mutation of matrix element $A_{ij}$ to system $\Sigma$ (which contains matrix $A$).
-%
-%  \begin{align*}
-%    \Phi(\mu(\Sigma, A_{ij} \upsilon)) = \exp\left(- \left| \frac{A_{22}-A_{12}}{\lambda_{1} \lambda_{2}} - \frac{k + \upsilon \pda (A_{22}-A_{12})}{\lambda_{1} \lambda_{2} + \text{adj}(A)^{T}_{ij} \upsilon} \right|\right)   .
-%  \end{align*}
-%For instance, the fitness of an organism with a change of one of its entries $A_{ij}$ (by $\upsilon$) is listed,
-%  \begin{align*}
-%    \Phi(\mu(\Sigma, A_{11})) &= \exp \left(- \left| \frac{A_{22}-A_{12}}{\det(A)} - \frac{A_{22}-A_{12}}{\det(A) + \upsilon A_{22}} \right| \right) \\
-%    \Phi(\mu(\Sigma, A_{12})) &= \exp \left(- \left| \frac{A_{22}-A_{12}}{\det(A)} - \frac{A_{22}-A_{12} - \upsilon}{\det(A) - \upsilon A_{21}} \right| \right) \\
-%    \Phi(\mu(\Sigma, A_{21})) &= \exp \left(- \left| \frac{A_{22}-A_{12}}{\det(A)} - \frac{A_{22}-A_{12}}{\det(A) - \upsilon A_{12}} \right| \right) \\
-%    \Phi(\mu(\Sigma, A_{22})) &= \exp \left(- \left| \frac{A_{22}-A_{12}}{\det(A)} - \frac{A_{22}-A_{12} + \upsilon}{\det(A) + \upsilon A_{11}} \right| \right)
-%  \end{align*}
-%
- The transfer function at $z=0$ for any GRN with mutations is, 
-  \begin{align*}
-    H(0, \begin{bmatrix} A_{11} + \upsilon & A_{12} + \varepsilon \\ A_{21} + \delta & A_{22} + \omega \end{bmatrix}) = \frac{A_{22} - A_{12} + \omega - \varepsilon}{ \det(A) + \omega A_{11} - \delta A_{12} - \varepsilon A_{21} + \upsilon A_{22} + \omega \upsilon - \varepsilon \delta }
-  \end{align*}
-
-%  \begin{align*}
-%    \frac{\left(A_{22}-A_{12}\right) + \left( \displaystyle\sum_{\jmath = 1}^{\ind_{\omega}} \omega_{\jmath} - \displaystyle\sum_{\jmath = 1}^{\ind_{\varepsilon}} \varepsilon_{\jmath} \right)}{\displaystyle\det(A) + \displaystyle\sum_{\jmath=1}^{\ind_{\omega}} \omega_{\jmath} A_{11} - \sum_{\jmath=1}^{\ind_{\delta}} \delta_{\jmath} A_{12} -  \sum_{\jmath=1}^{\ind_{\varepsilon}} \varepsilon_{\jmath} A_{21} + \sum_{\jmath=1}^{\ind_{\upsilon}} \upsilon_{\jmath} A_{22} + \sum_{\jmath=1}^{\ind_{\omega}} \omega_{\jmath} \sum_{\jmath=1}^{\ind_{\upsilon}} \upsilon_{\jmath} - \sum_{\jmath=1}^{\ind_{\varepsilon}} \varepsilon_{\jmath} \sum_{\jmath=1}^{\ind_{\delta}} \delta_{\jmath} }
-%  \end{align*}
-
-  If we want to know how new mutations fix in the population, we can explicitly show all mutational paths that could compensate for fixed mutations at sites $A_{ij}$.
-In a population, not at a phenotypic optimum, we are interested in the combination of mutations with varying magnitudes and locations that are most likely to fix.
-
-  \begin{align*}
-    \frac{\left(A_{22}-A_{12}\right) + \left( \displaystyle\sum_{\jmath = 1}^{\ind_{\omega}} \omega_{\jmath} - \displaystyle\sum_{\jmath = 1}^{\ind_{\varepsilon}} \varepsilon_{\jmath} \right)}{\det(A) + \displaystyle\sum_{\mu=\{\omega , \delta , \varepsilon , \upsilon \}}^{\ind_{\mu}} \mu_{\jmath} (-1)^{i+j} A_{ij} + \sum_{\jmath=1}^{\ind_{\omega}} \omega_{\jmath} \sum_{\jmath=1}^{\ind_{\upsilon}} \upsilon_{\jmath} - \sum_{\jmath=1}^{\ind_{\varepsilon}} \varepsilon_{\jmath} \sum_{\jmath=1}^{\ind_{\delta}} \delta_{\jmath} }
-  \end{align*}
-
-\section{Biological Examples}
-
-  \begin{example}
-    Negative feedback and Positive feedback systems:
-    \begin{align*}
-      T(z) = \frac{H(z)}{1 + G(z)H(z)} \\
-      T(z) = \frac{H(z)}{1 - G(z)H(z)} \\
-    \end{align*}
-  \end{example}
-
-  \begin{example}
-    Systems in serial ($u \rightarrow \Sigma_{1} \{A_{1},B_{1},C_{1} \} \rightarrow \Sigma_{2} \{A_{2},B_{2},C_{2} \} \rightarrow y$):
-    \begin{align*}
-      \begin{bmatrix} \dot{x}_{1} \\ \dot{x}_{2} \end{bmatrix} &= \left[ \begin{array}{c|c}
-        A_{1} & 0 \\
-          \hline
-          B_{2}C_{2} & A_{2} \\
-        \end{array} \right] \begin{bmatrix} \dot{x}_{1} \\ \dot{x}_{2} \end{bmatrix}  + \begin{bmatrix} B_{1} \\ 0 \end{bmatrix} u \\
-          y &= \begin{bmatrix} 0 & C_{2} \end{bmatrix} \begin{bmatrix} \dot{x}_{1} \\ \dot{x}_{2} \end{bmatrix} .
-    \end{align*}
-
-  Systems in serial could be applied to model a gene network $A_{1}$ and the subsequent degradation rates of its outputs at rate $A_{2}$.
-  \end{example}
-
-  \begin{example}
-  \begin{align*}
-    \dot{x} &= \begin{bmatrix} + & - \\ 0 & + \end{bmatrix} \vec{x} + \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \vec{u} \\
-      y(t) &= \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \vec{x} \\
-        E(t) &= \begin{bmatrix} -y_{1}(t) & 0 & 0 \\ y_{1}(t) & -y_{2}(t) & 0 \\ -\sigma_{1} y_{1}(t) & (1 - \sigma_{2})y_{2}(t) & 1 \end{bmatrix} \vec{\zeta} + \begin{bmatrix} u_{1} \\ u_{2} \\ \epsilon \end{bmatrix} .
-  \end{align*}
-    Where $u_{1}$ is input of \emph{lactose}, $u_{2}$ is the input of \emph{glucose}, $y_{1}$ is the enzyme \emph{lactase} and $y_{2}$ is enzyme \emph{``glucase''}.
-$E(t)$ is metabolism of the organism, converting sugars to energy as a function of enzyme concentrations.
-Lactase converts lactose to glucose, and the enzyme is produced at energy cost $\sigma_{1}$; glucase converts glucose to energy and is produced at cost $\sigma_{2}$.
-Fitness is a function of the total energy of the system, given inputs, possibly: $\int_{0}^{T}E(t)_{3} dt$ after some time $T$ has passed.
-Or perhaps, once the total metabolic energy produced passes some threshold, the organism undergoes fission.
-Thus, fitness selects for maximizing energy as quickly as possible.
-  \end{example}
-
-  \begin{example} 
-    Gene Regulatory and Metabolic Network Cascade:
-
-    Gene Regulatory Network
-    \begin{align*}
-      \dot{x}(t) &= A x(t) + B_{1} u(t) \\
-      y(t) &= C_{1} x(t)
-    \end{align*}
-      Metabolism
-      \begin{align*}
-      \dot{\zeta}(t) &= M\big(y(t), \sigma \big) \zeta(t) + B_{2} \begin{bmatrix} u(t) \\ v(t) \end{bmatrix} \\
-        m(t) &= C_{2} \zeta(t) \\
-      b(t) &= C_{3} \zeta(t)
-      \end{align*}
-      Fitness
-      \begin{align*}
-      \Phi(m(t), p(t))
-    \end{align*}
-
-    The GRN produces time varying proteins involved in the metabolism.
-The metabilic network (MN), is a function of protein concentrations $y(t)$ and production costs $\sigma$.
-The metabolic network produces energy $m(t)$ and toxic byproducts $b(t)$.
-Fitness $\Phi$, or maybe something more concrete, like fission time and/or cell death, is a function of both energy $m(t)$ and toxins $b(t)$.
-  \end{example}
-\section{Notes on what to do when allowing deleterious mutations}
-
-\begin{enumerate}
-
-    \item Epistasis, coming from mutations deleterious in some networks but not in others
-(come from incompatibilities).
-
-    \item Describe heterosis.
-
-    \item Distribution of dominance coefficients.
-
-    \item Evolvability: easier for bigger networks to adapt to a new environment?
-
-    \item Do short-term responses to strong directional selection have deleterious epistatic consequences in other traits?
-
-\end{enumerate}
+    \section*{Discussion}
 
 \bibliographystyle{plainnat}
-\bibliography{krefs}
-
+%\bibliography{krefs}
+\end{multicols}
 \end{document}

### 2017-06-01 handnotes.tex
handnotes edit

@@
   \begin{align}
-    A(\tau) = V(\tau) A(0) V^{-1}(\tau) ,
+    A(\tau) = V(\tau) A(\tau_{0}) V^{-1}(\tau) ,
   \end{align}
@@
   \end{align}
+  After evolutionary time $T$, the population's gene network architecture evolves from $A(\tau)$ to $A(\tau + \epsilon)$ with a probability inversely proportional to the magnitude of $\epsilon$ and proportional to the magnitude of $T$.
 
   \section*{Phenotypic Invariance}
+    First we focus on a simple evolutionary scenario: a large population, perfectly adapted, and in a constant environment.
+In this circumstance we expect phenotype to be conserved throughout evolutionary time.
+As such we should only expect the phenotype to change as a consequence of genetic drift (small effective population size), adaptation to new selective and/or environemental pressures.
+These phenotypic variations should yield distinct signatures.
+Adaptative changes will change the optimal impulse response function $h(t) \xrightarrow{adaptation} h'(t)$.
+Genetic drift registers as an increase in the intrapopulation variation in $h(t)$.
+
+    Presently we ask two questions, (1) holding $f(t)$, $h(t)$, and $u(t)$ constant for evolutionary time $T$, how much do we expect gene network organization to drift, and (2) does this contribute to speciation, primarily via the fixation of reproductive incompatibilities?
+
+    \subsection*{Not all minimal gene networks can drift}
+      If a gene network is minimal and all the molecular species involved in the network are under selection, such that $C$ is the $n \times n$ identity matrix  ($C = I_{n}$), the only acceptable  change of coordinate matrix is the identify matrix.
+
+     \begin{align}
+        C \lor B &= I \\
+        IV^{-1} &= I \\
+        \iff V &= I
+      \end{align}
+
+      \subsection*{All non-minimal gene networks can drift}
+
+      Despite the existence of a unique genetic architecture in the minimal case, there still exists an infinite number of systems with larger networks that have identical external dynamics. 
+      \begin{align}
+        h(t) &= \widehat{h}(t) \iff \\ 
+        CA^{k}B &= \widehat{C} \widehat{A}^{k} \widehat{C} \qquad \text{for } k = 0, 1, \dots
+      \end{align}
+      Any two systems with equivalent impulse responses will have equivalent phenotypes.
+
+      \subsection*{Speciation via Reproductive Incompatibility}
+
+      Define reproduction in diploids as first, the recombination of unlinked genes to make gametes, and second, as the averaging of two individual parental gametes to produce an offspring.
+Assuming parental populations are both phenotypically identical and genetically homogenous within each population, first generation hybrids (F1s) can be computed by averaging the two parental gene networks.
+Second generation hybrids (F2s) can be computed by first swapping the genes between the two parental gene networks, and next averaging these hybrid gametes.
+      \begin{align*}
+        A_{(F1)}(\tau_{i}, \tau_{j}) &= \frac{1}{2}\left(A(\tau_{i}) + A(\tau_{j}) \right) \\
+        h_{(F1)}(t) &= C e^{\frac{t}{2}\left(A(\tau_{i}) + A(\tau_{j}) \right)} B
+      \end{align*}
+
+      \begin{align*}
+        G_{(F2)}(r) = Q(r)A(\tau_{i}) + \left( I-Q(r) \right)A(\tau_{j})
+      \end{align*}
+
+      \begin{align*}
+        h_{(F2,r, r')}(t) = C e^{\frac{t}{2} \left( G(r) + G(r') \right)} B
+      \end{align*}
+
+      \begin{align*}
+        \Phi(h'(t) ) = e^{- \int_{0}^{\infty} \left\lVert h(t) - h'(t) \right\rVert dt}
+      \end{align*}
 

### 2017-06-05 handnotes.tex
added figures and text to handnotes

@@
 \documentclass[9 pt]{article}
-\usepackage{amsmath, amssymb, color, xcolor}
+\usepackage{amsmath, amssymb, color, xcolor, amsthm}
 \usepackage{graphicx, wrapfig, float, caption, dsfont, bbm}
@@
 
-  Organisms' phenotypes are constructred by complex gene by gene and gene by environment interactions.
-Although a very simplified description, here we define the phenotype to be the spatio-temporal molecular dynamics directly under natural selection.
-Basically the \emph{what}, \emph{when}, and \emph{where}, of an organism's molecules that are physiologically or otherwise important and relevant to survival.
+  Organisms' phenotypes are constructred by gene by gene by environment interactions.
+Here we simply define the phenotype to be the organismal temporal molecular dynamics directly under natural selection.
+The \emph{what}, \emph{when}, and \emph{where}, of an organism's molecules that are physiologically or otherwise relevant for survival.
 Thus we say that some function $f(t)$ is a phenotype where,
@@
   \end{align}
-  and $h(t)$ is the \emph{impulse response} or \emph{kernel} of the system and $u(t)$ is the \emph{input} function.
-The input can be interpreted as the environment, or as initial conditions, or otherwise, depending on the biological specifics under study.
+  and $h(t)$ is the \emph{impulse response} of the system and $u(t)$ is the \emph{input} function, both functions of time $t$.
+The input can be interpreted as the environment, as initial conditions, or otherwise, depending on the biological specifics under study.
 
@@
 
-  Above, $t$ refers to time.
   Although $f(t)$ describes the phenotype given an input, $h(t)$ describes the phenotype subject only to an impulse -- an input present initially and absent immediately thereafter.
-Typically, a system $\Sigma$ is defines as,
+Typically, a system $\Sigma$, is defined as,
   \begin{align}
@@
   Variables have the same identities as described above and $x(t)$ is a vector of molecule concentrations at time $t$.
-Therefore the molecular concentrations at a specific time are completely determined by the environmental input and gene by gene interactions.
+Therefore the molecular concentrations at a specific time are completely determined by the input and gene by gene interactions.
 Lastly, a portion and/or combination of these molecules, $y(t)$, are ``observed'' by selection.
+
+  \begin{example}[Oscillating Gene Network: Circadian Rhythm]
+    \begin{align*}
+      A = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} , \qquad B = \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix}
+    \end{align*}
+    \begin{align*}
+      \Sigma = \left \{ \begin{array}{ll} \dot{x}(t) &= \begin{bmatrix} 
+        0 & 1 \\ 
+       -1 & 0 
+        \end{bmatrix} x (t) + \begin{bmatrix} 1 \\ 1 \end{bmatrix} u(t) \\ 
+          y(t) &= \begin{bmatrix} 1 & 0 \end{bmatrix} x(t) \end{array} \right.
+     \end{align*}
+
+      \begin{align*}
+        h(t) = \sin(t) + \cos(t)
+      \end{align*}
+
+    \begin{figure}[H]
+        \begin{center}
+            \begin{tikzpicture}
+            \begin{scope}[every node/.style={circle,thick,draw}]
+                \node (A) at (0,0) {$x_{1}$};
+                \node[dashed] (B) at (3,0) {$x_{2}$};
+                \node[shape=rectangle] (U) at (1.5,2) {input};
+                \node[shape=rectangle] (y) at (1.5,-2) {output};
+            \end{scope}
+
+            \begin{scope}[>={Stealth[black]},
+                          every node/.style={fill=white,circle},
+                          every edge/.style={draw=black, thick}]
+                \path [->, >=Rectangle] (A) edge[bend left] node {\tiny $-1$} (B);
+                \path [->] (B) edge[bend left] node {\tiny $1$} (A); 
+                \path[->] (U) edge[dashed] node {\tiny $1$} (A);
+                \path[->] (U) edge[dashed] node {\tiny $1$} (B);
+                \path[->] (A) edge[dashed,bend right] node {\tiny $1$} (y);
+            \end{scope}
+            \begin{scope}[>={Stealth[black]},
+                          every edge/.style={draw=black, thick}]
+                %\path [->] (A) edge[loop left] node {\tiny $\lambda_{1}$} (A);
+                %\path [->] (B) edge[loop left] node {\tiny $\lambda_{2}$} (B);
+            \end{scope}
+
+            \end{tikzpicture}
+        \end{center}
+        \caption{
+            Diagram of Example \ref{ex:osc} in the text.
+            \plr{explain what arrows mean if nec}
+            \label{fig:oscillator_diagram}}
+    \end{figure}
+    \begin{figure}[H]
+      \centering
+      \includegraphics[width=0.5\textwidth]{osc_impulse}
+    \end{figure}
+  \end{example}
 
@@
 
-  Although systems may not be identifiable up to a change of coordinates, at present we are primarily interested in a subset of these systems.
-That is, systems that not only have equivalent external dynamics, but also equivlanet input relationships and under equivalent selective pressures.
-Formally, systems related by a change of coordinates that leave $B$ and $C$ invariant:
+  Although systems may not be identifiable beyond a change of coordinates, at present we are primarily interested in a subset of these systems.
+That is, systems that not only have equivalent external dynamics, but also equivalent input and output relationships.
+Formally, this means systems related by a change of coordinates (any invertible matrix $V$) that leaves $B$ and $C$ invariant:
   \begin{align}
@@
 
-  Define $V(\tau)$ as the change of coordinates matrix that preserves $B$ and $C$, with $\tau$ a vector of free parameters. The set of \emph{all} phenotypically invariant (minimal) gene networks is, 
+  Define $V(\tau)$ as the parameterized change of coordinates matrix that preserves $B$ and $C$, with $\tau$ a vector of free parameters.
+The set of \emph{all} phenotypically invariant (minimal) gene networks is,
   \begin{align}
-    A(\tau) = V(\tau) A(\tau_{0}) V^{-1}(\tau) ,
+    A(\tau) = V(\tau) A(0) V^{-1}(\tau) ,
   \end{align}
-  and a Linear Evolutionary System is, 
+  and a \emph{Linear Evolutionary System} is, 
   \begin{align}
@@
   \end{align}
-
+  \begin{example}[External Equivalence does not imply internal equivalence]
+    Gene 1 dynamics (blue) are equivalent for network architectures $A(0)$ and $A(2)$, however the internal dynamics containing gene 2 (orange) are very different.
+  \begin{figure}[H]
+    \centering
+    \includegraphics[width=0.5\textwidth]{osc_A0_A2_both_compare}
+    %\caption{Gene 1 dynamics (blue) are equivalent for network architectures $A(0)$ and $A(2)$, however the internal dynamics containing gene 2 (orange) are very different.}
+  \end{figure}
+  \end{example}
   Evolution thus proceeds as a random walk in phenotypically invariant network space.
@@
 
+  \begin{example}[All Phenotypically Equivalent Oscillators]
+    \begin{align*}
+      A(\tau) &= \begin{bmatrix} 1 & 0 \\ \tau & 1-\tau \end{bmatrix} \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 0 \\ \frac{\tau}{\tau-1} & \frac{-1}{\tau-1} \end{bmatrix} \\
+        B &= \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix} \\
+          h(t) &= \sin(t) + \cos(t) \qquad \forall \ \tau \neq 1
+    \end{align*}
+    \begin{figure}[H]
+    \centering
+\begin{tikzpicture}
+\begin{scope}[every node/.style={circle,thick,draw}]
+  \node (A) at (0,0) {$x_{1}$};
+    \node[dashed] (B) at (4,0) {$x_{2}$};
+    \node[shape=rectangle] (U) at (2,2) {input};
+    \node[shape=rectangle] (y) at (2,-2) {output};
+\end{scope}
+
+\begin{scope}[>={Stealth[black]},
+              every node/.style={fill=white,circle},
+              every edge/.style={draw=black, thick}]
+    \path [->, sloped] (A) edge[bend left] node {\tiny $2 \tau + (\tau-1)^{-1}$} (B);
+    \path [->, sloped] (B) edge[bend left] node {\tiny $-(\tau-1)^{-1}$} (A); 
+    \path[->] (U) edge[dashed] node {\tiny $1$} (A);
+    \path[->] (U) edge[dashed] node {\tiny $1$} (B);
+    \path[->] (A) edge[dashed, bend right] node {\tiny $1$} (y);
+\end{scope}
+\begin{scope}[>={Stealth[black]},
+              every edge/.style={draw=black, thick}]
+    \path [->] (A) edge[loop left] node[sloped, anchor=center, above] {\tiny $1 + (\tau-1)^{-1}$} (A);
+    \path [->] (B) edge[loop right] node[sloped, anchor=center, above] {\tiny $-1 - (\tau-1)^{-1}$} (B);
+\end{scope}
+
+\end{tikzpicture}
+    \end{figure}
+  \end{example}
   \section*{Phenotypic Invariance}
@@
 
-    \subsection*{Not all minimal gene networks can drift}
+    \begin{example}[Not all minimal gene networks can drift]
       If a gene network is minimal and all the molecular species involved in the network are under selection, such that $C$ is the $n \times n$ identity matrix  ($C = I_{n}$), the only acceptable  change of coordinate matrix is the identify matrix.
@@
       \end{align}
-
-      \subsection*{All non-minimal gene networks can drift}
+      \end{example}
+      \begin{example}[All Non-Minimal Gene Networks can Drift]
 
@@
       Any two systems with equivalent impulse responses will have equivalent phenotypes.
+      \end{example}
+
+      \plr{add Kalman decomposition stuff here}
 
@@
 
-      Define reproduction in diploids as first, the recombination of unlinked genes to make gametes, and second, as the averaging of two individual parental gametes to produce an offspring.
+      %Define reproduction in diploids as first, the recombination of unlinked genes to make gametes, and second, as the averaging of two individual parental gametes to produce an offspring.
 Assuming parental populations are both phenotypically identical and genetically homogenous within each population, first generation hybrids (F1s) can be computed by averaging the two parental gene networks.
 Second generation hybrids (F2s) can be computed by first swapping the genes between the two parental gene networks, and next averaging these hybrid gametes.
-      \begin{align*}
-        A_{(F1)}(\tau_{i}, \tau_{j}) &= \frac{1}{2}\left(A(\tau_{i}) + A(\tau_{j}) \right) \\
-        h_{(F1)}(t) &= C e^{\frac{t}{2}\left(A(\tau_{i}) + A(\tau_{j}) \right)} B
-      \end{align*}
 
-      \begin{align*}
-        G_{(F2)}(r) = Q(r)A(\tau_{i}) + \left( I-Q(r) \right)A(\tau_{j})
-      \end{align*}
+      A diploid organism's gene network is simply the average of both of its gene network copies; one from each parent.
+Further, each haploid parental gene network copy is formed via meiosis -- by swapping independent genes randomly.
+Assuming two distinct but genetically homogeneous populations evolving in allopatry meet and form hybrids, the first generation hybrids (F1s) gene network dynamics will be determined by the average of the parental haplotypes.
+The second generation hybrids (F2s), however, will be the product of a meiosis between parental haplotypes followed by an averaging of gametes.
 
-      \begin{align*}
-        h_{(F2,r, r')}(t) = C e^{\frac{t}{2} \left( G(r) + G(r') \right)} B
-      \end{align*}
+      Specifically, $F_{1}$ is the first generation hybrid gene network architecture formed by mating (averaging) $A(\tau)$ and $A(\hat{\tau})$, 
+      \begin{align}
+        F_{1}(\tau, \hat{\tau}) &= \frac{ A(\tau) + A(\hat{\tau}) }{2}  .
+     %   h(t, F_{1}) &= C e^{\left( F_{1}\left( \tau, \hat{\tau} \right) t \right)} B
+      \end{align}
 
-      \begin{align*}
-        \Phi(h'(t) ) = e^{- \int_{0}^{\infty} \left\lVert h(t) - h'(t) \right\rVert dt}
-      \end{align*}
+      %\begin{align*}
+       % G\left(r, \tau, \hat{\tau}\right) = Q(r) A( \tau ) + \left( I - Q(r) \right) A( \hat{\tau} )
+      %\end{align*}
+      and $F_{2}$ is the second generation hybrid gene network architecture formed by gametes $G(i, \tau, \hat{\tau} )$ and $G(j, \tau, \hat{\tau})$.
+Where each $G(\cdot)$ is formed by randomly swapping rows between $A(\tau)$ and $A(\hat{\tau})$, such that the $i$th  gene comes from $A(\tau)$ ($i$ and $j$ are orthogonal vectors, each element $0$ or $1$, and $i \neq j$).
+      \begin{align}
+        F_{2}(i, j) = \frac{G (i, \tau, \hat{\tau} ) + G( j, \tau, \hat{\tau} )}{2}
+      \end{align}
 
-  \section*{Examples}
-    \subsection*{Cell Cycle Control or Circadian Rhytm}
+      %\begin{align*}
+      %  h\left(t, F_{2}\left( r, r' \right)\right) = C e^{\left( F_{2}\left(r, r' \right) t \right)} B
+      %\end{align*}
+
+      The fitness of an organism can be computed by comparing its impulse response with the optimal response,
+      \begin{align}
+        \Phi\left( \widehat{h}(t) \right) = \exp \left\{- \int_{0}^{\infty} \left\lVert h(t) - \widehat{h}(t) \right\rVert dt \right\}  .
+      \end{align}
+      Therefore a hybrid's fitness can be computed by comparing its impulse response with that of its parents.
+
+      \begin{example}[Reproductive Incompatibility in an Oscillating Gene Network]
+        DMI examples...
+      \end{example}
+
+      \begin{example}[Not all Networks can Host Incompatibilities]
+        convex sets cant have DMIs
+      \end{example}
+
+  \section*{Additionl Examples}
 
@@
     \section*{Discussion}
-
+      \plr{mention $B$ part of genetic architecture}
 \bibliographystyle{plainnat}
-%\bibliography{krefs}
+\bibliography{krefs}
 \end{multicols}

### 2017-06-06 handnotes.tex
added DMI figures to handnnotes.

@@
 \usepackage{caption, subcaption}
-\usepackage{natbib} % gives us \citet: Author (year) and \citep: (Author; year)
+\usepackage[super]{natbib} % gives us \citet: Author (year) and \citep: (Author; year)
 \usepackage{authblk}
@@
 
-\section*{The Evolution of Phenotypically Invariant Genetic Networks}
-Joshua S. Schiffman and Peter L. Ralph
+\section*{The evolution of phenotypically invariant gene networks with implications for speciation, adaptation, and neutral variation}
+{\centering
+Joshua S. Schiffman$^{\dagger}$ \qquad Peter L. Ralph$^{\dagger \ddagger}$ \\
+$^{\dagger}$University of Southern California, Los Angeles, California \qquad $^{\ddagger}$University of Oregon, Eugene, Oregon \\
+\texttt{jsschiff@usc.edu} \qquad \texttt{plr@uoregon.edu}
+\\
+}
 
@@
 \section*{Abstract}
+  \textbf{
 I will outline an analytical theory to study the evolution of biological systems such as gene regulatory networks, borrowing insight and tools from control engineering, systems identification, and dynamical systems theory.
@@
 This analysis provides insight into the mechanisms and parameters important for understanding developmental systems drift, network rewiring, evolvability, epistasis, and speciation, as well as the tenuous connection between network architecture and function.
-
+}
 \section*{Introduction}
@@
 Progress in a suite of biological subdisciplines is stalled by our general lack of understanding of this molecular machinery: with respect to both its function and evolution.
-There does exist a growing body of experiment and data on the evolutionary histories and molecular characterizations of particular gene regulatory networks \citep{jaeger2011gap, davidson2006gene, israel2016comparative}, as well as thoughtful verbal and conceptual models \citep{true2001developmental, gwagner1}.
+There does exist a growing body of experiment and data on the evolutionary histories and molecular characterizations of particular gene regulatory networks \citep{jaeger2011gap, davidson2006gene, israel2016comparative}, as well as thoughtful verbal and conceptual models \citep{true2001developmental, gwagner1, weiss2000phenogenetic, edelman2001degeneracy}.
 However, as Hardy and Weinberg taught us over a century ago, verbal theories are often insufficient, if not downright misleading \citep{hardy1908mendelian, weinberg1908vererbungsgesetze, servedio2014not}.
@@
 
-      \begin{example}[Reproductive Incompatibility in an Oscillating Gene Network]
+      \begin{example}[F1 Reproductive Incompatibility in an Oscillating Gene Network]
         DMI examples...
+        \begin{figure}[H]
+          \includegraphics[width=0.5\textwidth]{A2_A3_osci_F1_hyb}
+        \end{figure}
+      \end{example}
+
+      \begin{example}[F2 Reproductive Incompatibility in an Oscillating Gene Network]
+        F2 versus parental expression dynamics.
+
+        \begin{figure}[H]
+          \begin{tabular}{cc}
+            \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_1}
+            & \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_2}
+           \\ \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_3}
+            & \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_4}
+            \\ \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_5}
+            & \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_6}
+          \end{tabular}
+        \end{figure}
       \end{example}
@@
         convex sets cant have DMIs
+
+        \begin{align*}
+          h(t) = 2 e^{- \theta t}
+        \end{align*}
+        Any non-minimal system with rows summing to $\theta$ is PI.
+Further, these systems are closed under averaging (mating) and row swapping (meiosis), leaving all hybrids optimally fit.
+The set of gene matrices is affine and therefore convex.
       \end{example}
@@
       \plr{mention $B$ part of genetic architecture}
-\bibliographystyle{plainnat}
+
+      \tiny
+\bibliographystyle{unsrt}
 \bibliography{krefs}

### 2017-06-06 handnotes.tex
edited variables. added some notes to random walk section.

@@
 The \emph{what}, \emph{when}, and \emph{where}, of an organism's molecules that are physiologically or otherwise relevant for survival.
-Thus we say that some function $f(t)$ is a phenotype where,
+Thus we say that some function $\phi(t)$ is a phenotype where,
   \begin{align}
-    f(t) = \int_{0}^{\infty} h(t) u(t) dt  ,
+    \phi(t) = \int_{0}^{\infty} h(t) u(t) dt  ,
   \end{align}
@@
 
-  Essentially the phenotype $f(t)$ is a consequence of an organism's specific gene by gene interactions, given by $h(t)$, reacting further with the local environment, given by $u(t)$.
+  Essentially the phenotype $\phi(t)$ is a consequence of an organism's specific gene by gene interactions, given by $h(t)$, reacting further with the local environment, given by $u(t)$.
 
@@
 
-  Although $f(t)$ describes the phenotype given an input, $h(t)$ describes the phenotype subject only to an impulse -- an input present initially and absent immediately thereafter.
+  Although $\phi(t)$ describes the phenotype given an input, $h(t)$ describes the phenotype subject only to an impulse -- an input present initially and absent immediately thereafter.
 Typically, a system $\Sigma$, is defined as,
   \begin{align}
-    \Sigma = \left\{ \begin{array}{ll} \dot{x}(t) &= A x(t) + B u(t) \\ y(t) &= C x(t) \end{array} \right.
+    \Sigma = \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right.
   \end{align}
-  Variables have the same identities as described above and $x(t)$ is a vector of molecule concentrations at time $t$.
+  Variables have the same identities as described above and $\kappa(t)$ is a vector of molecule concentrations at time $t$.
 Therefore the molecular concentrations at a specific time are completely determined by the input and gene by gene interactions.
-Lastly, a portion and/or combination of these molecules, $y(t)$, are ``observed'' by selection.
+Lastly, a portion and/or combination of these molecules, $\phi(t)$, are ``observed'' by selection (this is in constrast to $\kappa(t)$ -- the \emph{kryptotype} -- as it is ``hidden'' from direct selection).
 
@@
     \begin{align*}
-      \Sigma = \left \{ \begin{array}{ll} \dot{x}(t) &= \begin{bmatrix} 
+      \Sigma = \left \{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 
         0 & 1 \\ 
        -1 & 0 
-        \end{bmatrix} x (t) + \begin{bmatrix} 1 \\ 1 \end{bmatrix} u(t) \\ 
-          y(t) &= \begin{bmatrix} 1 & 0 \end{bmatrix} x(t) \end{array} \right.
+        \end{bmatrix} \kappa(t) + \begin{bmatrix} 1 \\ 1 \end{bmatrix} u(t) \\ 
+          \phi(t) &= \begin{bmatrix} 1 & 0 \end{bmatrix} \kappa(t) \end{array} \right.
      \end{align*}
@@
             \begin{scope}[every node/.style={circle,thick,draw}]
-                \node (A) at (0,0) {$x_{1}$};
-                \node[dashed] (B) at (3,0) {$x_{2}$};
+                \node (A) at (0,0) {$\kappa_{1}$};
+                \node[dashed] (B) at (3,0) {$\kappa_{2}$};
                 \node[shape=rectangle] (U) at (1.5,2) {input};
@@
   \begin{align}
-    \Sigma(\tau) = \left\{ \begin{array}{ll} \dot{x}(t) &= A(\tau) x(t) + B u(t) \\ y(t) &= C x(t) \end{array} \right .
+    \Sigma(\tau) = \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A(\tau) \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right .
   \end{align}
@@
   After evolutionary time $T$, the population's gene network architecture evolves from $A(\tau)$ to $A(\tau + \epsilon)$ with a probability inversely proportional to the magnitude of $\epsilon$ and proportional to the magnitude of $T$.
-
+\jss{Change in $\tau$ is tracking movement of population mean, not ``macro'' evolutionary change, as interpreted previously.}
+  \begin{align}
+    \Delta_{\tau} &= \left\lVert \frac{d}{d \tau} \text{vec} \left[A\left(\tau\right)\right] \right\rVert^{-1} \\
+    A(\tau) &\xrightarrow{\mu} A(\tau \pm \mu \Delta_{\tau})
+  \end{align}
   \begin{example}[All Phenotypically Equivalent Oscillators]
@@
 \begin{scope}[every node/.style={circle,thick,draw}]
-  \node (A) at (0,0) {$x_{1}$};
-    \node[dashed] (B) at (4,0) {$x_{2}$};
+  \node (A) at (0,0) {$\kappa_{1}$};
+    \node[dashed] (B) at (4,0) {$\kappa_{2}$};
     \node[shape=rectangle] (U) at (2,2) {input};
@@
 
-    Presently we ask two questions, (1) holding $f(t)$, $h(t)$, and $u(t)$ constant for evolutionary time $T$, how much do we expect gene network organization to drift, and (2) does this contribute to speciation, primarily via the fixation of reproductive incompatibilities?
+    Presently we ask two questions, (1) holding $\phi(t)$, $h(t)$, and $u(t)$ constant for evolutionary time $T$, how much do we expect gene network organization to drift, and (2) does this contribute to speciation, primarily via the fixation of reproductive incompatibilities?
 
@@
         h(t) &= \widehat{h}(t) \iff \\ 
-        CA^{k}B &= \widehat{C} \widehat{A}^{k} \widehat{C} \qquad \text{for } k = 0, 1, \dots
+        CA^{\jmath}B &= \widehat{C} \widehat{A}^{\jmath} \widehat{C} \qquad \text{for } \jmath = 0, 1, \dots
       \end{align}
@@
       \begin{align}
-        \Phi\left( \widehat{h}(t) \right) = \exp \left\{- \int_{0}^{\infty} \left\lVert h(t) - \widehat{h}(t) \right\rVert dt \right\}  .
+        \mathcal{F}\left( \widehat{h}(t) \right) = \exp \left\{- \int_{0}^{\infty} \left\lVert h(t) - \widehat{h}(t) \right\rVert dt \right\}  .
       \end{align}

### 2017-06-06 handnotes.tex
added GAL network notes to the end (not sure if necessary for this version)

@@
     \subsection*{Metabolic Network}
+  \begin{align*}
+      \dot{x}(t) &= \begin{bmatrix} 0 & 1 & 0 & -1 \\ 0 & 0 & -1 & -1 \\ -1 & 1 & 0 & 0 \\ 0 & 0 & 0 & 0 \end{bmatrix} \begin{bmatrix} \text{GAL3} \\ \text{GAL4} \\ \text{GAL80} \\ \text{MIG1} \end{bmatrix}(t) + \begin{bmatrix} 1 & 0 & 0 \\ 0 & 0 & 1 \\ 0 & 0 & 0 \\ 0 & 1 & 1 \end{bmatrix} \begin{bmatrix} \text{galactose} \\ \text{glucose} \\ \text{BTM} \end{bmatrix}(t) \\
+        y(t) &= \begin{bmatrix} 0 & 1 & 0 & -1 \end{bmatrix} \vec{x}(t)
+    \end{align*}
+    \begin{align*}
+      H(z) = \begin{bmatrix} 1 \\ z^{2} + z + 1 \\ z + 1 \end{bmatrix} \frac{1}{z^{3} + z -1}
+    \end{align*}
+
+    \begin{align*}
+      CV &= C \\
+      VB &= B \\
+      V :&= \begin{bmatrix} 1 & 0 & a & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & b & 0 \\ 0 & 0 & 0 & 1 \end{bmatrix} , V^{-1} = \begin{bmatrix} 1 & 0 & \frac{-a}{b} & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & \frac{1}{b} & 0 \\ 0 & 0 & 0 & 1\end{bmatrix} \\
+        A(a,b) :&= VAV^{-1} = \begin{bmatrix} -a & a+1 & \frac{a^{2}}{b} & -1 \\ 0 & 0 & \frac{-1}{b} & -1 \\ -b & b & a & 0 \\ 0 & 0 & 0 & 0 \end{bmatrix}
+    \end{align*}
+    The GAL regulatory network in \emph{S.
+cerevisiae} is modelled here.
+The activating transcription factor GAL4, which is regulated by the presence of galactose and two other TFs, binds to the promoter region of the GAL regulon -- a cluster of genes regulated by the same promoter encoding 3 enzymes (GAL1, GAL7, and GAL10), for the metabolism of the sugar galactose.
+The repressing transcription factor MIG1, which is activated by the presence of glucose and other proteins omitted here, also binds to the regulon, preventing the expression of the galactose metabolizing enzymes.
+This network is one of the most studied gene regulatory networks in yeast, and experiments have already demonstrated significant transcriptional variation among different species and genuses of yeast.
+
+    In the present model, the $A$ matrix encodes the interactions among GAL3, GAL4, GAL80, and MIG1.
+$B$ interprets the input: the quantity of glucose, galactose, and the basal transcription (BTM) rates impact of TF concentrations.
+$C$ represents the promoter region of the GAL regulon and is the sum of the influence of GAL4 (activating) and MIG1 (repressing).
+
+    $A(a,b)$ represent alternative network structures that produce identical outputs, with identical input and output transformations -- meaning the regulatory contributions of the sugars and the basasl machinery, as well as the promoter of the GAL regulon are held constant.
+These interactions may be realized in nature, if the regulatory differences predicted by $A(a,b)$ are biologically/mutationally realizable.
+Mutations influencing the interactions between GAL4 and GAL80 have been experimentally demonstrated \citep{li2010alterations}.
 

### 2017-06-06 handnotes.tex
added: Quant Bio Spec Concept

@@
 
+      \begin{example}[Quantitative Biological Species Concept]
+        Two populations are said to be different species if they produce low fitness offspring.
+
+        F1s created by crossing externally equivlaent oscillators $A(0)$ and $A(2)$ have an $h_{F_{1}}(t) = e^{t}$, in contrast to both parents with $h_{F_{0}}(t) = \sin(t) + \cos(t)$.
+The hybrid phenotype is significantly different (it does not oscillate and increases infinitely) despite the phenotypic equivalence of the parents.
+
+        \begin{align*}
+          \mathcal{F} \left(h_{F_{0}}\right) &= 1 \\
+          \mathcal{F}\left(h_{F_{1}}\right) &= 0
+        \end{align*}
+      \end{example}
+
       \begin{example}[F1 Reproductive Incompatibility in an Oscillating Gene Network]

### 2017-06-07 handnotes.tex
minor edits to Model II, added citations to molecular examples

@@
 
+  The literature is filled with detailed observations of molecular systems and their diversity.
+There are examples of significant diversity in the networks underlying processes such as circadian rhythm \citep{sancar2008intelligent}, cell cycle control \citep{cross2011evolution, kearsey2003enigmatic}, pattern formation, and metabolism  \citep{lavoie2009rearrangements, martchenko2007transcriptional, dalal2016transcriptional, christensen2011unique, hartl2007induction, alam2013aspergillus}.
+Despite a symmetry in functionality or phenotype these systems often differ, sometimes substationaly, at the molecular level.
+How many different mechanisms have the same function?
+We urge the reader to consider the metaphorical black box.
+
   Systems with identical external dynamics do not necessarily have identical internal dynamics.
@@
   \end{align}
+  That is, all (linear and minimal) mechanisms capable of producing the same phenotype can be realized by a unique choice of $\tau$ in $\Sigma(\tau)$.
   \begin{example}[External Equivalence does not imply internal equivalence]
@@
   \begin{align*}
-      \dot{x}(t) &= \begin{bmatrix} 0 & 1 & 0 & -1 \\ 0 & 0 & -1 & -1 \\ -1 & 1 & 0 & 0 \\ 0 & 0 & 0 & 0 \end{bmatrix} \begin{bmatrix} \text{GAL3} \\ \text{GAL4} \\ \text{GAL80} \\ \text{MIG1} \end{bmatrix}(t) + \begin{bmatrix} 1 & 0 & 0 \\ 0 & 0 & 1 \\ 0 & 0 & 0 \\ 0 & 1 & 1 \end{bmatrix} \begin{bmatrix} \text{galactose} \\ \text{glucose} \\ \text{BTM} \end{bmatrix}(t) \\
+    \dot{x}(t) &= \begin{bmatrix} 0 & 1 & 0 & -1 \\ 0 & 0 & -1 & -1 \\ -1 & 1 & 0 & 0 \\ 0 & 0 & 0 & 0 \end{bmatrix} \begin{bmatrix} \text{GAL3} \\ \text{GAL4} \\ \text{GAL80} \\ \text{MIG1} \end{bmatrix}(t) \\ &+ \begin{bmatrix} 1 & 0 & 0 \\ 0 & 0 & 1 \\ 0 & 0 & 0 \\ 0 & 1 & 1 \end{bmatrix} \begin{bmatrix} \text{galactose} \\ \text{glucose} \\ \text{BTM} \end{bmatrix}(t) \\
         y(t) &= \begin{bmatrix} 0 & 1 & 0 & -1 \end{bmatrix} \vec{x}(t)
@@
     \begin{align*}
-      H(z) = \begin{bmatrix} 1 \\ z^{2} + z + 1 \\ z + 1 \end{bmatrix} \frac{1}{z^{3} + z -1}
+      H(z) &= \begin{bmatrix} 1 \\ z^{2} + z + 1 \\ z + 1 \end{bmatrix} \frac{1}{z^{3} + z -1}
     \end{align*}

### 2017-06-07 attic/draftnotes.tex
added to attic

\documentclass[11 pt]{article}
\usepackage{amsmath, amssymb, color, xcolor}
\usepackage{graphicx, wrapfig, float, caption, dsfont, bbm}
\usepackage{fullpage}
\usepackage[backref=page, hidelinks, colorlinks=true, citecolor=blue!60!black!100]{hyperref}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, shapes}
\usepackage{caption, subcaption}
\usepackage{natbib} % gives us \citet: Author (year) and \citep: (Author; year)
\usepackage{authblk}

\newcommand{\plr}[1]{{\color{blue}\it #1}}
\newcommand{\jss}[1]{{\color{olive}\it #1}}
% \newcommand{\ddt}{\frac{d}{dt}}
\newcommand{\ddt}{\dot}
\newcommand{\ro}{{ro}}
\newcommand{\nro}{{\bar{r}o}}
\newcommand{\rno}{{r\bar{o}}}
\newcommand{\nrno}{{\bar{r}\bar{o}}}
\newcommand{\reachable}{\mathcal{R}}
\newcommand{\unobservable}{\bar{\mathcal{O}}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\pda}{\frac{\partial}{\partial A_{ij}}}
\newcommand{\ind}{\mathds{1}}

\newcommand{\A}{\mathcal{A}}
\newcommand{\diag}{\text{diag}}
\newcommand{\1}{\mathbbm{1}}

\DeclareMathOperator{\spn}{span}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}

\begin{document}

  \section{Draft}

  In order to survive -- that is to aquire enough energy to replicate -- an organism needs to execute at least one of a large number of potential survival strategies. The set of potential strategies includes which molecules to metabolize and which behaviors to exhibit in response to stimuli, among many others. To play at least one of these strategies an organism must organize its phenotype in a highly specific way. Presumably this is largely accomplished via the coordinated expression of evolutionarily pliable gene networks. We can say that a given environment determines a set of strategies, and an organism's fitness is determined by its phenotype's ability to execute one or more of such strategies effectively. An environment with two different available types of sugar can be viewed as (at minimum)  a two strategy environment. The successful organism can play one strategy or the other, or both strategies, either metabolizing one or both of the sugars. In this simplified world, adaptation includes the processes of improving existing, as well as finding novel strategies to play (although the energy required to forge novel strategies may well be too high if is not precipitated by environmental or other change rendering current strategies ineffective). It should also be noted that depending on the strategy, more than one type of phenotype may be equally suitable (\emph{e.g.} different metabolic networks processing the same metabolite may be equally energetically efficient).

  We aim to understand how populations evolve. To begin, we focus on the simplest case: the evolution of a large and perfectly adapted population in a constant one strategy, one phenotype environment for an indefinite number of generations. Under these conditions we expect a conservation of phenotype as no adaptation is possible and genetic drift will be very rare. Despite this conservation at the phenotypic level we can still expect significant evolutionary change at the genetic level, in most cases. In more complex scenarios (with varying phenotypes) we would observe conservation of strategy, or more broadly (with varying strategies), conservation of fitness.

   Ordinarily, biologists distinguish only phenotype from genotype, however, here we will introduce (and justify the necessity of) an organism's \emph{kryptotype}. If one imagines an organism to be a \emph{black box}, taking its environment as input and outputting a phenotype, its kryptotype can be thought of as the mechanism inside the black box, as its specifics are hidden directly from selection. We will discuss the properties of kryptotypes, and primarily focus on the non-uniqueness and diversity of the kryptotype -- how the kryptotype can evolve and influence a myriad of evolutionary and biological phenomena such as speciation and evolvability.

  In order to model this mathematically, we represent a phenotype as the temporal expression patterns of a set of molecular species, most commonly proteins and mRNA. Can an organism express the enzymes and other relevent moleclar species for metabolism efficiently, given the presence of a metabolite? That is the phenotype is a given number of protein expressing genes and their regulatory interactions. In this model, we only focus on the evolution of regulation, and ignore mutations in protein function. For mathematical tractability these regulatory interactions are represented as a linear system with evolutionarily pliable regulatory coefficients.

  \subsection{An organism is defined by its molecular time-dynamics}
   Molecular (say protein or mRNA) concentrations are represented by $\vec{x}$, the gene regulatory interactions by $\alpha, \beta, \gamma,$ and $\delta,$ and the environmental/initial inputs by $B$.  
      \begin{equation*}
        \begin{bmatrix} \dot{x}_{1} \\ \dot{x}_{2} \end{bmatrix} = \begin{bmatrix} \alpha & \beta \\ \gamma & \delta \end{bmatrix} \begin{bmatrix} x_{1} \\ x_{2} \end{bmatrix} + \begin{bmatrix} B_{1} \\ B_{2} \end{bmatrix} u
      \end{equation*}
    Each row vector $\begin{bmatrix} \alpha & \beta \end{bmatrix}$ of the GRN matrix is the promoter region of a gene, such that, $\alpha$ describes the how much gene-1 self regulates and $\beta$ describes how much gene-2 regulates gene-1. 
  \subsection{The phenotype is just a subset of an organism's proteins directly visible to selection}
    \begin{equation*}
      y = \begin{bmatrix} C_{1} & C_{2} \end{bmatrix} \begin{bmatrix} x_{1} \\ x_{2} \end{bmatrix}
    \end{equation*}
    For instance, if $C = \begin{bmatrix} 1 & 0 \end{bmatrix}$, the phenotype is only described by the dynamics of gene-1. This means that selection can only ``see'' what happens to the proteins produced by gene-1, whereas the dynamics of gene-2 are invisible to selection (and thus not formally defined as part of the phenotype).

      This does not mean, however, that gene-2 is unnecessary with respect to the phenotype!

        \begin{example}
          For example, consider a GRN with oscillating protein concentrations.
    In this 2 gene system, neither gene-1 nor gene-2 self regulate. Gene-2 up-regulates gene-1, and gene-2 is down-regulated by gene-1:
      \begin{equation*}
        \begin{bmatrix} \dot{x}_{1} \\ \dot{x}_{2} \end{bmatrix} = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} \begin{bmatrix} x_{1} \\ x_{2} \end{bmatrix} + \begin{bmatrix} 1 \\ 1 \end{bmatrix} u
      \end{equation*}
      Furthermore, only the dynamics of gene-1 are important:
      \begin{equation*}
        y = \begin{bmatrix} 1 & 0 \end{bmatrix} \begin{bmatrix} x_{1} \\ x_{2} \end{bmatrix}
      \end{equation*}
    The dynamics of gene-1 are, $y = \cos(t) + \sin(t)$.

    A single-gene system cannot oscillate.
  \end{example}
    \subsection{Expression dynamic complexity is limited by the number of genes in a network.}
    $f(t)$ describes protein concentrations at time $t$, $\lambda_{j}$ is the $j$th eigenvalue of the GRN and $\xi_{j}$ is a variable.
      \begin{equation*}
        f(t) = \sum_{j=1}^{n} e^{\lambda_{j}t} \xi_{j}
      \end{equation*}
      Sometimes it's (visually) easier to study the Laplace transform of the system. 
      \begin{equation*}
        \mathcal{L}(f(t)) := H(z) = \sum_{j}^{n} \frac{c_{j} b_{j}}{(z- \lambda_{j})}
      \end{equation*}

  \subsection{Model details -- LE Systems}
      An LE System: Linear Evolving (TI) System (in evolutionary state $\tau$), 
    \begin{align*}
      \Sigma(\tau) = \left\{ \begin{array}{cc} \dot{x} =& A(\tau)x + Bu \\
      y =& Cx \end{array} \right.
    \end{align*}
    \begin{align*}
      \mathcal{L}(\Sigma(\tau)) := H(z) = C \left(zI - A(\tau) \right)^{-1} B
    \end{align*}

      \begin{definition}[Phenomenological equivalence of systems]
    Let $(x(t),y(t))$ and $(\bar x(t),\bar y(t))$ be the solutions to \eqref{eqn:lti_system}
    with coefficient matrices $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ respectively,
    and both $x(0)$ and $\bar x(0)$ are zero. 
    The systems defined by $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ are
    \textbf{phenomenologically equivalent} 
    if
    \begin{align*}
        y(t) = \bar y(t) \qquad \text{for all} \; t \ge 0.
    \end{align*}
    Equivalently, this occurs if and only if
    \begin{align*}
        H(z) = \bar H(z)  \qquad \text{for all} \; z \ge 0,
    \end{align*}
    where $H$ and $\bar H$ are the transfer functions of the two systems.
\end{definition}

 \subsection{Gene Network Diversity}
      First we need to describe the phase space. How many different gene networks can do exactly the same thing? Rather, can we describe the set of phenotypically invariant gene regulatory networks -- the set of all gene networks that produce identical outputs, when given identical inputs. In the linear case we can analytically describe all the networks in this space,
      \begin{align*}
        \mathcal{A} := \{ VAV^{-1}, CV, VB : V \in GL_{n}(\mathbb{R}) \}
      \end{align*}
      for the minimal dimension, and for the general (including non-minimal) case as,
      \begin{align*}
        CA^{k}B = \widehat{C} \widehat{A}^{k} \widehat{B} \qquad .
      \end{align*}
      where $A$ is a decription of gene network topology, and $B$ and $C$ transform the system intputs to and outputs from the network, respectively.

  \subsection{Constant Selection Through Time}

    A large population, perfectly adapted to it's environment, under constant selection, will be phenotypically invariant for all generations (given the circumstances outlined above). Despite a conservation of phenotype, variation at the genetic and interemediate molecular levels is possible and expected through the course of evolution, as shown below. Under these constraints, phenotypic variance can only be the consequence of genetic drift (stochastic evolutionary forces overpowering selective ones due to population size) or adaptation to a novel phenotype for a different strategy (signatures of each will be discussed).

      For convenience, we hold $B$ (input transformation) and $C$ (output transformation) constant, considering only systems with varying internal mechanics.
      \begin{align*}
        P(\vec{r}) \in \{ GL_{n}(\mathbb{R}) : CP = C \text{and } BP = B \} \\
        A(\vec{r}) := P(\vec{r}) A P^{-1}(\vec{r})
      \end{align*}

      Where $A := A(0)$ be the initial and minimal configuration of a population's gene network.
      Where $\vec{r}$ is the unique vector of free variables associated with any instantiation of $P$.
      Elements of $\vec{r}$  can be restricted to a biologically reasonable range, $s \leq r_{k} \leq l$ to prevent unrealistic topologies.
      From this persepctive, the values of $A(r)$ can be written as the $n^{2}$ vector $\vec{a}(r) = \text{vec}(A(r))$. In this space, evolution acts as a random walk on $r$. 
      Furthermore, let $\Delta a(r) := \lVert \frac{d}{dr} a(r) \rVert^{-1}$. $\Delta a(r)$ describes the average magnitude of neutral genetic change required to reach a phenotypically invariant gene network. Systems should neutrally drift and rewire at rates inversely proportional to this value, as at high values of $\Delta$, more mutations are required to rewire neutrally -- that is the fitness valley is wider. Given enough time, we expect evolution to traverse the space for all of $r$, albeit at different rates and probabilities.

      Perhaps the rate of phenotypically invariant evolution near state $\tau$ can be expressed as,
      \begin{align*}
        \Delta_{r} :&= \left\lVert \frac{d}{d \tau} \text{vec}\left[A(\tau)\right] \right\rVert^{-1} \\
      \end{align*}
      Thus, $\mu \Delta_{\tau}$ is the rate at which the kryptotype evolves near $\tau$ with mutation rate $\mu$.
      \jss{Maybe need to distinguish between differentiating from the left and right of $\tau$ since we want to know the rate and direction of evolution. i.e. $\partial_{\pm} f(A(\tau))$}

      Therefore the system will evolve from $A(tau) \rightarrow A(\tau \pm \mu \Delta_{\tau})$ stochastically through generational time.

      \subsection{Gene Network Growth and Decay}

      Consider the minimal 2-dimensional system $A_{[2]}(\tau) := P(\tau) \begin{bmatrix} \lambda_{1} & \lambda_{2} - k \\ 0 & \lambda_{2} \end{bmatrix} P^{-1}(\tau)$. With probability $\mu_{g}$ (network growth), a new regulatory gene is added to the system. 
        \begin{align*}
          A_{[2]}(\tau) \xrightarrow{\mu_{g}} A_{[3]}(\vec{\tau}) \\
          A_{[3]}(0) = \begin{bmatrix} A_{[2]}(\tau) & \begin{array}{cc} 0 \\ 0 \end{array} \\ \begin{array}{cc} 1 & 1 \end{array} & \gamma \end{bmatrix}
        \end{align*}
        After some time has passed, the GRN will reorganize to, 
        \begin{align*}
          P_{[3]}(\vec{r})A_{[3]}(0)P_{[3]}^{-1}(\vec{r}) \\
          P_{[3]}B_{[3]} = B_{[3]} \\
          C_{[3]}P_{[3]} = C_{[3]} \\
          P_{[3]}(\vec{r}) := \begin{bmatrix} 1 & 0 & 0 \\ 1 & 1-r_{1} & r_{2} \\ r_{3} & -r_{3} & r_{4} \end{bmatrix} \\ 
        B_{[3]} := \begin{bmatrix} B_{[2]} \\ 0 \end{bmatrix} \\ 
        C_{[3]} := \begin{bmatrix} C_{[2]} & 0 \end{bmatrix}
        \end{align*}

        After some time, if the higher dimensional GRN is of the following form, and a submatrix of it, $\alpha \approx A(\tau)$, with probability $\mu_{d}$, the GRN can decay in size. 
        \begin{align*}
          A_{[3]}(\vec{r}) = \begin{bmatrix} \alpha_{[2]} & \begin{array}{cc} c_{1} \\ c_{2} \end{array} \\ \begin{array}{cc} c_{3} & c_{4} \end{array} & c_{5} \end{bmatrix}
        \end{align*}

        \begin{align*}
            v_{0} := \text{vec}(A_{2}(\tau)) &= \left(\begin{array}{c} \frac{\tau}{\tau - 1} - 1\\ \frac{\tau \left(3 \tau - 2\right)}{\tau - 1} - \tau \\ - \frac{1}{\tau - 1} \\ -\frac{3 \tau - 2}{\tau - 1} \end{array}\right) \\
          v_{1} := \text{vec}(\alpha_{[2]}(\vec{r})) &= \left(\begin{array}{c} -\frac{r_{4}}{r_{2} r_{3} - r_{4} \left(r_{1} - 1\right)}\\ \frac{\left(2 r_{1} - 1\right) \left(r_{1} + r_{2}\right)}{r_{1} - 1} + \frac{r_{2} r_{3} \left(\gamma - r_{2} + 2\right) - r_{1} r_{2} r_{3} \left(\gamma + 3\right)}{\left(r_{2} r_{3} - r_{4}\ \left(r - 1\right)\right)\ \left(r - 1\right)}\\ \frac{r_{4}}{r_{2} r_{3} - r_{4}\ \left(r_{1} - 1\right)}\\ \frac{\left(r_{4} + r_{2}\ r_{3}\right)\ \left(\gamma_{1} + 3\right) - r_{4}\ \left(\gamma - r_{2} + 2\right)}{r_{4} - r_{1}\ r_{4} + r_{2}\ r_{3}} - 3 \end{array}\right) 
        \end{align*}
        If, 
        \begin{align*}
          \lVert v_{0} - v_{1} \rVert < \delta \\
        \end{align*}
        Then, 
        \begin{align*}
          A_{[3]} \xrightarrow{\mu_{d}} A_{[2]}
        \end{align*}

        Or perhaps, what is the expected difference after time?
        \begin{align*}
          \mathbb{E}_{t} \left[ \lVert v_{0} - v_{1} \rVert \right] = x_{t}
        \end{align*}
        Does $x_{t}$ increase with time?

        \begin{align*}
          \text{Tr} \left[ \text{submatrix}_{33} \left(A_{[3]}\right) \right] = \frac{r_{2} \left(-(\lambda_{1} + \lambda_{2}) r_{3} + r_{4} + \gamma r_{3}\right)}{r_{4} - r_{1}r_{4} + r_{2}r_{3}} + (\lambda_{1} + \lambda_{2})
        \end{align*}

        The difference between the trace of the sub-matrix in $A_{[3]}$ and any matrix in $A_{[2]}$ is the term on the left. Since the trace determines if the sub-matrix has the same eigenvalues as a minimal system, if that number grows with time since duplication, then we will see a ratchet. 
\bibliographystyle{plainnat}
\bibliography{krefs}

\end{document}

### 2017-06-07 attic/kalman_walk_edit.tex
added to attic

--- 2017-02-28 kalman_walk.tex
@@
 \usepackage{amsmath, amssymb, color, xcolor}
-\usepackage{graphicx, wrapfig, float, caption, dsfont}
+\usepackage{graphicx}
 \usepackage{fullpage}
-\usepackage[backref=page, hidelinks, colorlinks=true, citecolor=blue!60!black!100]{hyperref}
+\usepackage[hidelinks]{hyperref}
 \usepackage{tikz}
-\usetikzlibrary{arrows.meta, shapes}
+\usetikzlibrary{arrows.meta}
 \usepackage{caption, subcaption}
-\usepackage{natbib} % gives us \citet: Author (year) and \citep: (Author; year)
-\usepackage{authblk}
 
@@
 \renewcommand{\P}{\mathbb{P}}
-\newcommand{\pda}{\frac{\partial}{\partial A_{ij}}}
-\newcommand{\ind}{\mathds{1}}
 
@@
 
-\paragraph{To do, soon:}
-
-\begin{enumerate}
-
-  \item \emph{(josh)} Add biology to introduction.
-(keep an eye out for concrete numbers we can use!) \jss{added some relevant biology citations to introduction.
-Still needs a lot of refinement.} \ref{sec:intro}
-
-    \item Do some order-of-magnitude calculation to figure out how fast systems should drift.
-        \begin{itemize}
-            \item \emph{(peter)} Work out the math for how to do this.
-            \item \emph{(josh)} Find real parameters to put into this. 
-              \jss{added some notes for TFBS gain and loss calculations from Lynch} \ref{sec:param}
-        \end{itemize}
-
-    \item \emph{(josh)} Write out simple examples in biological words.
-
-    \item \emph{(josh)} Rephrase hybrid inviability example in biological terms.  Discuss F1's (or heterozygotes) versus F2's.
-
-    \item \emph{(josh)} Clarify our assumptions around $x(0)$.
-
-    \item \emph{(josh)} Note we don't include $D$ for simplicity but don't.
-
-    \item \emph{(peter)} Figure out how to specify $P$ for Kalman decomposition.
-
-    \item \emph{(josh \jss{\checkmark},peter)} Prove or find citation for uniqueness of minimal system up to basis change.
-\jss{See \citep[Chapter 10.6, Theorem 6.9: Fundamental Theorem of Linear Realization Theory.]{kalman1969topics}}
-
-    \item Work examples (minimal; not minimal; with/without B,C fixed) of Kalman decomposition and description of moving along the walk.
-        \begin{itemize}
-            \item \emph{(josh)} Work out parameterization.
-            \item \emph{(peter)} Describe Brownian motion.
-        \end{itemize}
-
-    \item \emph{(peter)} Write down and decide between options for parameterizing the random walk on the neutral manifold:
-        use Kalman decomposition, and parameterize Brownian motion so that it has constant speed;
-        OR consider Josh's parameteriziation like this;
-        OR just say we do Brownian motion on the manifold (more or less implicit characterization).
-
-    \item \emph{(peter)} Describe model leading to constraints on regulation so we have a stationary distribution.
-        Ask questions of stationary distribution: robustness, connectivity, xxx?
-
-    \item Consider options for fitness function: Euclidean? Sign of eigenvalues?  Time domain or frequency?
-      First try on simple concrete examples. \jss{see \ref{sec:del}}
-
-    \item Figure out coherent model for changing dimension: do coefficients stay stuck at zero? In what parameterization? 
-        Do we need to allow deleterious variation to get deletion?
-
-    \item Write something in discussion about nonlinear systems.  Do an example if there is a feasible one.
-
-    \item Simulate.
-
-\end{enumerate}
+%$\unobservable$
 
@@
 
-%%% Questions:
-
 % is y the phenotype or H(z) the phenotype? Do we need new jargon anyway? 
-
-\title{The Evolution of Phenotypically Invariant Genetic Networks}
-\author[1]{Joshua S. Schiffman}
-\author[1,2]{Peter L. Ralph}
-\affil[1]{University of Southern California, Los Angeles, California}
-\affil[2]{University of Oregon, Eugene, Oregon}
-\affil[ ]{\texttt{jsschiff@usc.edu} \qquad \texttt{plr@uoregon.edu}}
+% 
+\title{Mathematical Gene Network Evolution}
 \maketitle
@@
 \begin{abstract}
-I will outline an analytical theory to study the evolution of biological systems such as gene regulatory networks, borrowing insight and tools from control engineering, systems identification, and dynamical systems theory.
-I will describe a null model of regulatory network evolution by analytically describing the set of all linear gene networks (of any size) that produce identical phenotypes -- and the evolutionary paths connecting them.
-In the idealized case of a perfectly adapted population, constant selection, and a static environment, we observe neutral evolution as a random walk over the phenotypically- invariant network-space.
-Under neutral conditions, this model can provide descriptions of expected network size and connectivity under mutation-selection equilibrium, estimate the rate of regulatory rewiring, and the rates at which Dobzhansky-Muller incompatibilities arise in reproductively isolated populations.
-This analysis provides insight into the mechanisms and parameters important for understanding developmental systems drift, network rewiring, evolvability, epistasis, and speciation, as well as the tenuous connection between network architecture and function.
+Abstract.
 \end{abstract}
 
-\section{Background} \label{sec:intro}
+\section{Background}
 
@@
 
-Bridging the gulf between an organism's genome and phenotype is a poorly understood and complex molecular machinery.
-Progress in a suite of biological subdisciplines is stalled by our general lack of understanding of this molecular machinery: with respect to both its function and evolution.
-There does exist a growing body of experiment and data on the evolutionary histories and molecular characterizations of particular gene regulatory networks \citep{jaeger2011gap, davidson2006gene, israel2016comparative}, as well as thoughtful verbal and conceptual models \citep{true2001developmental, gwagner1}.
-However, as Hardy and Weinberg taught us over a century ago, verbal theories are often insufficient, if not downright misleading \citep{hardy1908mendelian, weinberg1908vererbungsgesetze, servedio2014not}.
-This is especially pertinent given the staggering complexity and scope of contemporary research programs.
-This outlook necessitates the advancement of conceptual frameworks of such precision, only mathematics will suffice.
-Previously it has been suggested that any idealized study of evolution is incomplete without a mathematically sufficient description of the genotype, phenotype, and transformation from one to the other \citep{Lewontin1974genetic}.%Here we outline an analytical theory to study the evolution of biological systems, borrowing insight and methods from control engineering, systems identification, and dynamical systems theory.
-
-%Presently, we focus on the neutral evolution of genetic regulatory networks.
-That is, we analytically describe the set of all linear gene networks (of any size) that produce identical phenotypes -- and the evolutionary paths connecting them.
-In the idealized case of a perfectly adapted population, constant selection, and a static environment, we observe evolution under the ``conservation of phenotype'' as a Brownian motion over phenotypically-invariant network-space.
-This analysis provides insight to the mechanisms and parameters important for understanding developmental systems drift, network rewiring, evolvability, epistasis, and speciation, as well as the tenuous connection between network architecture and function.
-
-%It is commonly taught that an organism's genome contains the heritable material 
-%that natural selection filters and that an organism's phenotype directly determines its
-%evolutionary fitness. Between genotype and phenotype is an often complicated and poorly
-%undrstood molecular machinery -- and it is a major goal common to many different disciplines
-%within the life sciences to elucidate its form, function, and evolution. These aims are
-%delicately intertwined and a comphrehensive understanding of a system's evolution requires
-%an understanding of its function and vice versa.
+It is commonly taught that an organism's genome contains the heritable material 
+that natural selection filters and that an organism's phenotype directly determines its
+evolutionary fitness. Between genotype and phenotype is an often complicated and poorly
+undrstood molecular machinery -- and it is a major goal common to many different disciplines
+within the life sciences to elucidate its form, function, and evolution. These aims are
+delicately intertwined and a comphrehensive understanding of a system's evolution requires
+an understanding of its function and vice versa.
 
 The molecular machinery, interacting with the environment, and bridging genotype to phenotype
-can be mathematically described as a dynamical system -- or a system of differential equations\citep{jaeger2015comet}.
- Movement in this direction is ongoing, as researchers have begun to study 
-the evolution of both abstract \citep{wagner1994evolution, wagner1996does,  siegal2002waddington, bergman2003evolutionary} and empirically inspired computational and mathematical models of gene regulatory networks (GRNs) \citep{mjolsness1991connectionist, jaeger2004dynamic, maria1, vitaly1, vitaly2, crombach2016gap, wotton2015quantitative, chertkova2017insilico}.
+can be mathematically described as a dynamical system (or a system of differential equations).
+Previously it has been argued that any idealized study of evolution is incomplete without
+a mathematically sufficient description of the genotype, phenotype, and transformation from
+one to the other. Movement in this direction is ongoing, as researhers have begun to study 
+the evolution of empirically inspired computational and mathematical models of gene regulatory networks (GRNs) and metabolic networks that incorporate empirical data such as sequences and expression patterns.
 If we allow the reasonable assumption that the genotype-phenotype map can be represented as a system of differential equations, we can immediately discuss its evolution and function in a much more mechanistic, yet general, manner.
@@
 there can be uncertaintity about an inferred model's parameters or even its claims about
-causal structure, even with access to complete and perfect data \citep{bellman1970structural, grewal1976identifiability, walter1984structural}. Models with different 
+causal structure, even with access to complete and perfect data. Models with different 
 parameter schemes, or even different mechanics can be equally accurate, but still not
@@
 infinite number of ``realizations,'' or ways to reverse engineer the dynamics of a black box,
-even if all possible input and output experiments on the black box are performed \citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. In chemical
+even if all possible input and output experiments on the black box are performed. In chemical
 engineering, those who study chemical reaction networks sometimes refer to the fundamental
-unidentifiability of these networks as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}.
-In computer science, this is framed as the relationship among processes that simulate one another \citep{van2004equivalence}.
+unidentifiability of these networks as ``the fundamental dogma of chemical kinetics.''
 Although this may frusturate the occasional engineer or scientist, viewed from another angle,
@@
 as the parameters and structures can be realized biologically. In fact, evolutionary
-biologists who study homology and analogy are very familiar with such functional symmetries; macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level \citep{true2001developmental, tsong2006evolution, hare2008sepsid, vierstra2014mouse, stergachis2014conservation, taylor2016diverse, matsui2015regulatory}.
+biologists who study homology and analogy are very familiar with such functional symmetries; macroscopically identical phenotypes in closely related species can be divergent at the molecular level.
 
@@
 
-What are typically referred to as ``design principles'' or some sort of clear structure-function relationships are commonly saught \citep{milo2002network, alon2007network, ma2009defining}.
+What are typically referred to as ``design principles'' or some sort of clear structure-function relationships are commonly saught
 
 Seeking to attribute functionality, uninformed by evolution, can lead to spurious claims
-(i.e. see Graur \citep{lynch2007evolution, lynch2007frailty, graur2013immortality}).
+(i.e. see Graur).
 
@@
 \plr{A similar situation occurs in development (...?).}
-However, there are many ways to skin a cat \jss{(there are also many ways to develop a cat)}:
+However, there are many ways to skin a cat:
 especially given a number of possible molecular intermediaries,
@@
 \end{align*}
-
-    \begin{figure}
-        \begin{center}
-            \begin{tikzpicture}
-            \begin{scope}[every node/.style={circle,thick,draw}]
-                \node (A) at (0,0) {$x_{1}$};
-                \node[dashed] (B) at (3,0) {$x_{2}$};
-                \node[shape=rectangle] (U) at (1.5,2) {input};
-                \node[shape=rectangle] (y) at (1.5,-2) {output};
-            \end{scope}
-
-            \begin{scope}[>={Stealth[black]},
-                          every node/.style={fill=white,circle},
-                          every edge/.style={draw=black, thick}]
-                \path [->, >=Rectangle] (A) edge[bend left] node {\tiny $-1$} (B);
-                \path [->] (B) edge[bend left] node {\tiny $1$} (A); 
-                \path[->] (U) edge[dashed] node {\tiny $1$} (A);
-                \path[->] (U) edge[dashed] node {\tiny $1$} (B);
-                \path[->] (A) edge[dashed,bend right] node {\tiny $1$} (y);
-            \end{scope}
-            \begin{scope}[>={Stealth[black]},
-                          every edge/.style={draw=black, thick}]
-                %\path [->] (A) edge[loop left] node {\tiny $\lambda_{1}$} (A);
-                %\path [->] (B) edge[loop left] node {\tiny $\lambda_{2}$} (B);
-            \end{scope}
-
-            \end{tikzpicture}
-        \end{center}
-        \caption{
-            Diagram of Example \ref{ex:osc} in the text.  
-            \plr{explain what arrows mean if nec}
-            \label{fig:oscillator_diagram}}
-    \end{figure}
-
 as a special case of the below.
@@
   \end{align*}
-  from Example \ref{ex:osc} ($B=\begin{bmatrix} 1 \\ 1 \end{bmatrix}$ and $C=\begin{bmatrix} 1 & 0 \end{bmatrix}$). 
-If hybrids are formed by exchanging genes (swapping rows), the hybrid's gene network is,
+  from Example \ref{ex:osc} ($B=[1 ,1]^{T}$ and $C=[1 0]$). 
+If hybrids are formed by exchanining genes (swapping rows), they hybrid,
   \begin{align*}
-    \text{hybrid network } := \mathcal{H}(A(0), A(2)) = \left[\begin{matrix}
+    \mathcal{H}(A(0), A(2)) = \left[\begin{matrix}
       2 & -1 \\
@@
   \end{align*}
-  \jss{ Or should it be: \begin{align*}
-    \text{hybrid network } := \text{mean}(A(0), A(2)) = \left[\begin{matrix}
-      1 & 0 \\
-      2 & -1
-    \end{matrix}\right] ,
-  \end{align*} }
   It's transfer function will be $H(z, \mathcal{H}) = \frac{z-2}{z^2 - 3z +1}$, which produces very different dynamics than its parents, $H(z, A(0)) = H(z, A(2)) = \frac{z+1}{z^2 + 1}$.
-The time dynamcs of molecular species $1$ in the hybrid is $y^{(\mathcal{H})} =e^{t} \cosh(\sqrt{2} t)$ \jss{ which is approx.
-$\approx e^{t}$}, with the concentration of species $1$ rapidly heading towards infinity.
-The dynamics of species $1$ in the parental networks is $y = \sin(t) + \cos(t)$, where it continuously oscillates.
-
-  \jss{The transfer function $H(z, \mathcal{H})$, by averaging networks, is $\frac{1}{z-1}$ and the time dynamics are: $y^{(\mathcal{H})} = e^{t}$.}
+The time dynamcs of molecular species $1$ in the hybrid is $y' =\mathrm{e}^{t}\, \cosh\!\left(\sqrt{2}\, t\right)$, with species $1$ quickly heading towards infinity.
+The dynamics of species $1$ in the parental networks is, $y = \sin(t) + \cos(t)$, where species $1$ continuously oscillates.
 \end{example}
-
-\begin{definition}[Biological Species Concept]
-  Let $\mathcal{A}$ be the set of all phenotypically symmetric gene networks with transfer funtion $H(z)$. 
-  Choose $A$ and $\widetilde A$, both in $\mathcal{A}$. 
-  If $\mathcal{H}(A, \widetilde A) \notin \mathcal{A}$ then $A$ and $\widetilde A$ are distinct species. 
-\end{definition}
 
@@
   \node (A) at (0,0) {$x_{1}$};
-    \node[dashed] (B) at (4,0) {$x_{2}$};
-    \node[shape=rectangle] (U) at (2,2) {input};
-    \node[shape=rectangle] (y) at (2,-2) {output};
+    \node[dashed] (B) at (0,-4) {$x_{2}$};
+    \node[shape=rectangle, dashed] (U) at (3,-2) {$u(t)$};
 \end{scope}
@@
     \path[->] (U) edge[dashed] node {\tiny $1$} (B);
-    \path[->] (A) edge[dashed, bend right] node {\tiny $1$} (y);
 \end{scope}
@@
     \path [->] (A) edge[loop left] node {\tiny $\lambda_{1}$} (A);
-    \path [->] (B) edge[loop right] node {\tiny $\lambda_{2}$} (B);
+    \path [->] (B) edge[loop left] node {\tiny $\lambda_{2}$} (B);
 \end{scope}
@@
     \path [->] (A) edge[loop left] node {\tiny $(\lambda_{1} + \frac{\gamma}{2})$} (A);
-    \path [->] (B) edge[loop right] node {\tiny $(\lambda_{2}-\frac{\gamma}{2})$} (B);
+    \path [->] (B) edge[loop left] node {\tiny $(\lambda_{2}-\frac{\gamma}{2})$} (B);
 \end{scope}
@@
 Plus, it works in engineering.}
-
-\jss{Maybe we should discuss the Hartman-Grobman Theorem?}
 
@@
 
-\section{Estimates of network rewiring and developmenal systems drift.} \label{sec:param}
-
-\jss{Parameter estimates for DSD $\downarrow$}
-
-What parameters (and other relevant factors) do we need from the literature and experiment? 
-  \begin{itemize}
-    \item mutation rate
-    \item Transcription factor binding site appearence and loss
-    \begin{itemize}
-      \item In \citep{lynch2007evolution} TF binding site gain ($u_{g}$) and loss ($u_{l}$) is estimated to be,
-        \begin{align*}
-          u_{g} &= \frac{(L-n+1)nu}{4^{n}} \\
-          u_{l} &= nu
-        \end{align*}
-      Where $L$ is the length of the promoter, $n$ is the number of nucleotides in a TFBS, and $u$ is the mutation rate per base pair per generation. 
-    \item \citet{lynch2007evolution} further states that in prokaryotes $L$ will usually be $\leq 100$ and in eukaryotes $L$ can range from $10^{3}$ to $10^{6}$.
-He also cites \citep{harbison2004transcriptional} that TFBSs range from $5-10$bps in length.
-    \item in \citep{lynch2014evolutionary}, they estimate
-        \begin{align*}
-          \tilde{P}(m) = C \left[ 3^{\ell - m}  \left( \begin{array}{c} \ell \\ m \end{array} \right) \right] e^{2N_{e}s(m)}  ,
-        \end{align*}
-        where $\tilde{P}(m)$ is the stationary distribution of a TFBS with match (and strength) $m$.
-$C$ is the normalizer constant, $m$ is the number of matches, $\ell$ is the length of the TFBS motif.
-The term on the right is the ratio of fixation probabilities.
-        Or in their own words, there are two interpretations: 
-        \begin{quote} ``(i) For a single TFBS, it represents the long-term proportion of evolutionary time spent in the various matching states and (ii) for a set of different TFBSs under the same selective constraints, it represents the expected distribution of states at any point in time.'' \end{quote}
-      \item \citet{tuugrul2015dynamics, anderson2015intermolecular} also may be helpful.
-    \end{itemize}
-    \item rate of both \emph{trans-} and \emph{cis-} regulatory evolution. (see Sergey's papers).
-      i.e. \citep{nuzhdin2004common, fear2016buffering, graze2009regulatory, graze2012allelic, genissel2008cis}
-    \item The PWN-score landscape (does binding strength increase linearly, exponentially, have a threshold, etc.) with respect to SNPs.
-    \item Others? 
-  \end{itemize}
-
-\section{(Old Notes) What biological question are we asking?}
+\section{What biological question are we asking?}
 
@@
 
-\section{(Old Notes) How can we mathematically model this question?}
+\section{How can we mathematically model this question?}
 
@@
 
-\section{(Old Notes) List of papers with some notes.}
+\section{List of papers with some notes.}
 
@@
 
-\section{Including Deleterious Mutations and a Fitness Function}\label{sec:del}
-
-  \begin{minipage}{0.5\textwidth}
-  At the moment we focus strictly on the case where the minimal realization is in $2$ dimensions and both of its eigenvalues are real and negative (so a special case of the system from Example \ref{ex:2x2}).
-\jss{(Note: this is not to say the ideas below will not be generalizable).}
-
-    The system of interest has the form,
-  \begin{align*}
-    \Sigma := \left\{ A = \begin{bmatrix} A_{11} & A_{12} \\ A_{21} & A_{22} \end{bmatrix}, B = \begin{bmatrix} 1 \\ 1 \end{bmatrix}, C = \begin{bmatrix} 1 & 0 \end{bmatrix} \right\}
-  \end{align*}
-  \end{minipage}%
-  \begin{minipage}{0.5\textwidth}
-    \centering
-    \includegraphics[width=0.95\textwidth]{figure_2x2mutation01}%
-      \captionof{figure}{The transfer function for $A = \left[ \protect\begin{array}{cc} -3 & 1 \\ 0 & -1 \protect\end{array} \right]$ and all eight $\mu(\Sigma, A_{ij})$ mutations, where $\upsilon = \pm 0.5$}\label{fig:mut}%
-  \end{minipage}
-
-Define the fitness function to be,
-
-\begin{align}\label{eq:fitness}
-  \Phi(\bar{\Sigma}) := \left\{ \begin{array}{ll}   e^{- \left| H(0) - \tilde{H}(0) \right| = - \left| \frac{k}{\lambda_{1} \lambda_{2}} - \frac{\tilde{k}}{\tilde{\lambda}_{1} \tilde{\lambda}_{2}} \right|}    &\text{if } \lambda_{1,2} < 0 \text{ and } \in \mathbb{R}, \ \det(A) \neq 0 \\
-  0   &\text{if } \lambda_{1,2} \geq 0 \ \text{ or } \ \notin \mathbb{R}, \ \det(A) = 0 \end{array} \right.
-\end{align}
-and let $\upsilon$ be the magnitude and direction (sign) of a mutation on regulatory interaction strength.
-The fitness function (\ref{eq:fitness}) may me appropriate if mutations only slightly alter the dynamics of the system, as in Figure \ref{fig:mut}.
-
-%Further let $\mu(\Sigma, A_{ij})$ be the resulting gene regulatory network following a mutation of matrix element $A_{ij}$ to system $\Sigma$ (which contains matrix $A$).
-%
-%  \begin{align*}
-%    \Phi(\mu(\Sigma, A_{ij} \upsilon)) = \exp\left(- \left| \frac{A_{22}-A_{12}}{\lambda_{1} \lambda_{2}} - \frac{k + \upsilon \pda (A_{22}-A_{12})}{\lambda_{1} \lambda_{2} + \text{adj}(A)^{T}_{ij} \upsilon} \right|\right)   .
-%  \end{align*}
-%For instance, the fitness of an organism with a change of one of its entries $A_{ij}$ (by $\upsilon$) is listed,
-%  \begin{align*}
-%    \Phi(\mu(\Sigma, A_{11})) &= \exp \left(- \left| \frac{A_{22}-A_{12}}{\det(A)} - \frac{A_{22}-A_{12}}{\det(A) + \upsilon A_{22}} \right| \right) \\
-%    \Phi(\mu(\Sigma, A_{12})) &= \exp \left(- \left| \frac{A_{22}-A_{12}}{\det(A)} - \frac{A_{22}-A_{12} - \upsilon}{\det(A) - \upsilon A_{21}} \right| \right) \\
-%    \Phi(\mu(\Sigma, A_{21})) &= \exp \left(- \left| \frac{A_{22}-A_{12}}{\det(A)} - \frac{A_{22}-A_{12}}{\det(A) - \upsilon A_{12}} \right| \right) \\
-%    \Phi(\mu(\Sigma, A_{22})) &= \exp \left(- \left| \frac{A_{22}-A_{12}}{\det(A)} - \frac{A_{22}-A_{12} + \upsilon}{\det(A) + \upsilon A_{11}} \right| \right)
-%  \end{align*}
-%
- The transfer function at $z=0$ for any GRN with mutations is, 
-  \begin{align*}
-    H(0, \begin{bmatrix} A_{11} + \upsilon & A_{12} + \varepsilon \\ A_{21} + \delta & A_{22} + \omega \end{bmatrix}) = \frac{A_{22} - A_{12} + \omega - \varepsilon}{ \det(A) + \omega A_{11} - \delta A_{12} - \varepsilon A_{21} + \upsilon A_{22} + \omega \upsilon - \varepsilon \delta }
-  \end{align*}
-
-%  \begin{align*}
-%    \frac{\left(A_{22}-A_{12}\right) + \left( \displaystyle\sum_{\jmath = 1}^{\ind_{\omega}} \omega_{\jmath} - \displaystyle\sum_{\jmath = 1}^{\ind_{\varepsilon}} \varepsilon_{\jmath} \right)}{\displaystyle\det(A) + \displaystyle\sum_{\jmath=1}^{\ind_{\omega}} \omega_{\jmath} A_{11} - \sum_{\jmath=1}^{\ind_{\delta}} \delta_{\jmath} A_{12} -  \sum_{\jmath=1}^{\ind_{\varepsilon}} \varepsilon_{\jmath} A_{21} + \sum_{\jmath=1}^{\ind_{\upsilon}} \upsilon_{\jmath} A_{22} + \sum_{\jmath=1}^{\ind_{\omega}} \omega_{\jmath} \sum_{\jmath=1}^{\ind_{\upsilon}} \upsilon_{\jmath} - \sum_{\jmath=1}^{\ind_{\varepsilon}} \varepsilon_{\jmath} \sum_{\jmath=1}^{\ind_{\delta}} \delta_{\jmath} }
-%  \end{align*}
-
-  If we want to know how new mutations fix in the population, we can explicitly show all mutational paths that could compensate for fixed mutations at sites $A_{ij}$.
-In a population, not at a phenotypic optimum, we are interested in the combination of mutations with varying magnitudes and locations that are most likely to fix.
-
-  \begin{align*}
-    \frac{\left(A_{22}-A_{12}\right) + \left( \displaystyle\sum_{\jmath = 1}^{\ind_{\omega}} \omega_{\jmath} - \displaystyle\sum_{\jmath = 1}^{\ind_{\varepsilon}} \varepsilon_{\jmath} \right)}{\det(A) + \displaystyle\sum_{\mu=\{\omega , \delta , \varepsilon , \upsilon \}}^{\ind_{\mu}} \mu_{\jmath} (-1)^{i+j} A_{ij} + \sum_{\jmath=1}^{\ind_{\omega}} \omega_{\jmath} \sum_{\jmath=1}^{\ind_{\upsilon}} \upsilon_{\jmath} - \sum_{\jmath=1}^{\ind_{\varepsilon}} \varepsilon_{\jmath} \sum_{\jmath=1}^{\ind_{\delta}} \delta_{\jmath} }
-  \end{align*}
-
-\section{Biological Examples}
-
-  \begin{example}
-    Negative feedback and Positive feedback systems:
-    \begin{align*}
-      T(z) = \frac{H(z)}{1 + G(z)H(z)} \\
-      T(z) = \frac{H(z)}{1 - G(z)H(z)} \\
-    \end{align*}
-  \end{example}
-
-  \begin{example}
-    Systems in serial ($u \rightarrow \Sigma_{1} \{A_{1},B_{1},C_{1} \} \rightarrow \Sigma_{2} \{A_{2},B_{2},C_{2} \} \rightarrow y$):
-    \begin{align*}
-      \begin{bmatrix} \dot{x}_{1} \\ \dot{x}_{2} \end{bmatrix} &= \left[ \begin{array}{c|c}
-        A_{1} & 0 \\
-          \hline
-          B_{2}C_{2} & A_{2} \\
-        \end{array} \right] \begin{bmatrix} \dot{x}_{1} \\ \dot{x}_{2} \end{bmatrix}  + \begin{bmatrix} B_{1} \\ 0 \end{bmatrix} u \\
-          y &= \begin{bmatrix} 0 & C_{2} \end{bmatrix} \begin{bmatrix} \dot{x}_{1} \\ \dot{x}_{2} \end{bmatrix} .
-    \end{align*}
-
-  Systems in serial could be applied to model a gene network $A_{1}$ and the subsequent degradation rates of its outputs at rate $A_{2}$.
-  \end{example}
-
-  \begin{example}
-  \begin{align*}
-    \dot{x} &= \begin{bmatrix} + & - \\ 0 & + \end{bmatrix} \vec{x} + \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \vec{u} \\
-      y(t) &= \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \vec{x} \\
-        E(t) &= \begin{bmatrix} -y_{1}(t) & 0 & 0 \\ y_{1}(t) & -y_{2}(t) & 0 \\ -\sigma_{1} y_{1}(t) & (1 - \sigma_{2})y_{2}(t) & 1 \end{bmatrix} \vec{\zeta} + \begin{bmatrix} u_{1} \\ u_{2} \\ \epsilon \end{bmatrix} .
-  \end{align*}
-    Where $u_{1}$ is input of \emph{lactose}, $u_{2}$ is the input of \emph{glucose}, $y_{1}$ is the enzyme \emph{lactase} and $y_{2}$ is enzyme \emph{``glucase''}.
-$E(t)$ is metabolism of the organism, converting sugars to energy as a function of enzyme concentrations.
-Lactase converts lactose to glucose, and the enzyme is produced at energy cost $\sigma_{1}$; glucase converts glucose to energy and is produced at cost $\sigma_{2}$.
-Fitness is a function of the total energy of the system, given inputs, possibly: $\int_{0}^{T}E(t)_{3} dt$ after some time $T$ has passed.
-Or perhaps, once the total metabolic energy produced passes some threshold, the organism undergoes fission.
-Thus, fitness selects for maximizing energy as quickly as possible.
-  \end{example}
-
-  \begin{example} 
-    Gene Regulatory and Metabolic Network Cascade:
-
-    Gene Regulatory Network
-    \begin{align*}
-      \dot{x}(t) &= A x(t) + B_{1} u(t) \\
-      y(t) &= C_{1} x(t)
-    \end{align*}
-      Metabolism
-      \begin{align*}
-      \dot{\zeta}(t) &= M\big(y(t), \sigma \big) \zeta(t) + B_{2} \begin{bmatrix} u(t) \\ v(t) \end{bmatrix} \\
-        m(t) &= C_{2} \zeta(t) \\
-      b(t) &= C_{3} \zeta(t)
-      \end{align*}
-      Fitness
-      \begin{align*}
-      \Phi(m(t), p(t))
-    \end{align*}
-
-    The GRN produces time varying proteins involved in the metabolism.
-The metabilic network (MN), is a function of protein concentrations $y(t)$ and production costs $\sigma$.
-The metabolic network produces energy $m(t)$ and toxic byproducts $b(t)$.
-Fitness $\Phi$, or maybe something more concrete, like fission time and/or cell death, is a function of both energy $m(t)$ and toxins $b(t)$.
-  \end{example}
 \section{Notes on what to do when allowing deleterious mutations}
 
-\begin{enumerate}
-
-    \item Epistasis, coming from mutations deleterious in some networks but not in others
+Epistasis, coming from mutations deleterious in some networks but not in others
 (come from incompatibilities).
 
-    \item Describe heterosis.
-
-    \item Distribution of dominance coefficients.
-
-    \item Evolvability: easier for bigger networks to adapt to a new environment?
-
-    \item Do short-term responses to strong directional selection have deleterious epistatic consequences in other traits?
-
-\end{enumerate}
-
-\bibliographystyle{plainnat}
-\bibliography{krefs}
+Describe heterosis.
 

### 2017-06-07 attic/may17pres.tex
added to attic

--- 2017-02-28 kalman_walk.tex
@@
-\documentclass[11 pt]{article}
+\documentclass{seminar}
 \usepackage{amsmath, amssymb, color, xcolor}
@@
 \usepackage{tikz}
-\usetikzlibrary{arrows.meta, shapes}
+\usetikzlibrary{arrows.meta, shapes, backgrounds}
 \usepackage{caption, subcaption}
@@
 
+\pagecolor{black}
+\color{white}
+\slideframe{none}
 \begin{document}
 
-\paragraph{To do, soon:}
+  \begin{slide}
+    \centering
+    \section*{The Evolution of Phenotypically Invariant Gene Networks}
+      \subsection*{Josh Schiffman}
+      \subsection*{May 10, 2017}
+  \end{slide}
 
-\begin{enumerate}
+  \begin{slide}
+    \section*{Outline}
+      \begin{itemize}
+        \item How can we model the genotype to phenotype map, and its evolution?
+        \item Introduction to Linear Systems (LTI and LE Systems). 
+        \item Developmental Systems Drift
+        \item Gene Network Evolution 
+        \item Biological Examples
+        \item Speciation
+      \end{itemize}
+  \end{slide}
 
-  \item \emph{(josh)} Add biology to introduction.
-(keep an eye out for concrete numbers we can use!) \jss{added some relevant biology citations to introduction.
-Still needs a lot of refinement.} \ref{sec:intro}
-
-    \item Do some order-of-magnitude calculation to figure out how fast systems should drift.
-        \begin{itemize}
-            \item \emph{(peter)} Work out the math for how to do this.
-            \item \emph{(josh)} Find real parameters to put into this. 
-              \jss{added some notes for TFBS gain and loss calculations from Lynch} \ref{sec:param}
-        \end{itemize}
-
-    \item \emph{(josh)} Write out simple examples in biological words.
-
-    \item \emph{(josh)} Rephrase hybrid inviability example in biological terms.  Discuss F1's (or heterozygotes) versus F2's.
-
-    \item \emph{(josh)} Clarify our assumptions around $x(0)$.
-
-    \item \emph{(josh)} Note we don't include $D$ for simplicity but don't.
-
-    \item \emph{(peter)} Figure out how to specify $P$ for Kalman decomposition.
-
-    \item \emph{(josh \jss{\checkmark},peter)} Prove or find citation for uniqueness of minimal system up to basis change.
-\jss{See \citep[Chapter 10.6, Theorem 6.9: Fundamental Theorem of Linear Realization Theory.]{kalman1969topics}}
-
-    \item Work examples (minimal; not minimal; with/without B,C fixed) of Kalman decomposition and description of moving along the walk.
-        \begin{itemize}
-            \item \emph{(josh)} Work out parameterization.
-            \item \emph{(peter)} Describe Brownian motion.
-        \end{itemize}
-
-    \item \emph{(peter)} Write down and decide between options for parameterizing the random walk on the neutral manifold:
-        use Kalman decomposition, and parameterize Brownian motion so that it has constant speed;
-        OR consider Josh's parameteriziation like this;
-        OR just say we do Brownian motion on the manifold (more or less implicit characterization).
-
-    \item \emph{(peter)} Describe model leading to constraints on regulation so we have a stationary distribution.
-        Ask questions of stationary distribution: robustness, connectivity, xxx?
-
-    \item Consider options for fitness function: Euclidean? Sign of eigenvalues?  Time domain or frequency?
-      First try on simple concrete examples. \jss{see \ref{sec:del}}
-
-    \item Figure out coherent model for changing dimension: do coefficients stay stuck at zero? In what parameterization? 
-        Do we need to allow deleterious variation to get deletion?
-
-    \item Write something in discussion about nonlinear systems.  Do an example if there is a feasible one.
-
-    \item Simulate.
-
-\end{enumerate}
-
-% Outline:
-% 
-% - What the problems are:
-% 
-%     * evolution of perfectly fit populations
-% 
-% - Motivations: Can you infer function from structure? And...
-% 
-%     * motifs
-%     * speciation
-%     * network size
-%     * rube-goldberginess
-%     * redundancy, robustness
-%     * evolvability related to minimality?
-% 
-% - Give a concrete, simple example.
-% 
-% - Modeling framework (LTI systems)
-%
-%     * y = "phenotype"
-%     * x = "internal state" or "kryptotype"?
-%     * A = "regulatory matrix" or "genetic network"?
-%     * Kalman decomposition
-% 
-% - Evolution of free parameters in the Kalman decomposition
-%
-%     * Give or refer to earlier simple description of population dynamics which makes the system evolve.
-%     * What do we want to know from this model:
-%
-%       - How fast do the internal dynamics change?
-%       - Stationary distribution on network complexity/shape/properties
-%       - How fast does incompatibility accumulate?
-%
-%     * Brownian motion on free parameters
-%     * how mutations map into this space
-%     * dimension: duplication, deletion, recruitment of new things
-% 
-% - (Compute things)
-%
-% - Generalize ideas to nonlinear systems
-%
-%   * compute isomorphic or dynamically equivallent nonlinear systems of the feed-forward motif or of the three-node adaptation motif. 
-% 
-% - (Allow non-fit stuff too.)
-
-%%% Questions:
-
-% is y the phenotype or H(z) the phenotype? Do we need new jargon anyway?
-
-\title{The Evolution of Phenotypically Invariant Genetic Networks}
-\author[1]{Joshua S. Schiffman}
-\author[1,2]{Peter L. Ralph}
-\affil[1]{University of Southern California, Los Angeles, California}
-\affil[2]{University of Oregon, Eugene, Oregon}
-\affil[ ]{\texttt{jsschiff@usc.edu} \qquad \texttt{plr@uoregon.edu}}
-\maketitle
-
-\begin{abstract}
-I will outline an analytical theory to study the evolution of biological systems such as gene regulatory networks, borrowing insight and tools from control engineering, systems identification, and dynamical systems theory.
-I will describe a null model of regulatory network evolution by analytically describing the set of all linear gene networks (of any size) that produce identical phenotypes -- and the evolutionary paths connecting them.
-In the idealized case of a perfectly adapted population, constant selection, and a static environment, we observe neutral evolution as a random walk over the phenotypically- invariant network-space.
-Under neutral conditions, this model can provide descriptions of expected network size and connectivity under mutation-selection equilibrium, estimate the rate of regulatory rewiring, and the rates at which Dobzhansky-Muller incompatibilities arise in reproductively isolated populations.
-This analysis provides insight into the mechanisms and parameters important for understanding developmental systems drift, network rewiring, evolvability, epistasis, and speciation, as well as the tenuous connection between network architecture and function.
-\end{abstract}
-
-\section{Background} \label{sec:intro}
-
-\plr{Review of relevant other stuff.}
-
-\jss{Introduction}
-
-Bridging the gulf between an organism's genome and phenotype is a poorly understood and complex molecular machinery.
-Progress in a suite of biological subdisciplines is stalled by our general lack of understanding of this molecular machinery: with respect to both its function and evolution.
-There does exist a growing body of experiment and data on the evolutionary histories and molecular characterizations of particular gene regulatory networks \citep{jaeger2011gap, davidson2006gene, israel2016comparative}, as well as thoughtful verbal and conceptual models \citep{true2001developmental, gwagner1}.
-However, as Hardy and Weinberg taught us over a century ago, verbal theories are often insufficient, if not downright misleading \citep{hardy1908mendelian, weinberg1908vererbungsgesetze, servedio2014not}.
-This is especially pertinent given the staggering complexity and scope of contemporary research programs.
-This outlook necessitates the advancement of conceptual frameworks of such precision, only mathematics will suffice.
-Previously it has been suggested that any idealized study of evolution is incomplete without a mathematically sufficient description of the genotype, phenotype, and transformation from one to the other \citep{Lewontin1974genetic}.%Here we outline an analytical theory to study the evolution of biological systems, borrowing insight and methods from control engineering, systems identification, and dynamical systems theory.
-
-%Presently, we focus on the neutral evolution of genetic regulatory networks.
-That is, we analytically describe the set of all linear gene networks (of any size) that produce identical phenotypes -- and the evolutionary paths connecting them.
-In the idealized case of a perfectly adapted population, constant selection, and a static environment, we observe evolution under the ``conservation of phenotype'' as a Brownian motion over phenotypically-invariant network-space.
-This analysis provides insight to the mechanisms and parameters important for understanding developmental systems drift, network rewiring, evolvability, epistasis, and speciation, as well as the tenuous connection between network architecture and function.
-
-%It is commonly taught that an organism's genome contains the heritable material 
-%that natural selection filters and that an organism's phenotype directly determines its
-%evolutionary fitness. Between genotype and phenotype is an often complicated and poorly
-%undrstood molecular machinery -- and it is a major goal common to many different disciplines
-%within the life sciences to elucidate its form, function, and evolution. These aims are
-%delicately intertwined and a comphrehensive understanding of a system's evolution requires
-%an understanding of its function and vice versa.
-
-The molecular machinery, interacting with the environment, and bridging genotype to phenotype
-can be mathematically described as a dynamical system -- or a system of differential equations\citep{jaeger2015comet}.
- Movement in this direction is ongoing, as researchers have begun to study 
-the evolution of both abstract \citep{wagner1994evolution, wagner1996does,  siegal2002waddington, bergman2003evolutionary} and empirically inspired computational and mathematical models of gene regulatory networks (GRNs) \citep{mjolsness1991connectionist, jaeger2004dynamic, maria1, vitaly1, vitaly2, crombach2016gap, wotton2015quantitative, chertkova2017insilico}.
-If we allow the reasonable assumption that the genotype-phenotype map can be represented as a system of differential equations, we can immediately discuss its evolution and function in a much more mechanistic, yet general, manner.
-
-In some fields that seek to fit parametric models to experimental data, such as control
-theory, chemical engineering, and statistics, it is well known that mathematical models
-can fundamentally be \emph{unidentifiable} and/or \emph{indistinguishable} -- meaning that 
-there can be uncertaintity about an inferred model's parameters or even its claims about
-causal structure, even with access to complete and perfect data \citep{bellman1970structural, grewal1976identifiability, walter1984structural}. Models with different 
-parameter schemes, or even different mechanics can be equally accurate, but still not
-\emph{actually} agree with what is being modelled. In control theory, where electrical 
-circuits and mechanical systems are often the focus, it is understood that there can be an 
-infinite number of ``realizations,'' or ways to reverse engineer the dynamics of a black box,
-even if all possible input and output experiments on the black box are performed \citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. In chemical
-engineering, those who study chemical reaction networks sometimes refer to the fundamental
-unidentifiability of these networks as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}.
-In computer science, this is framed as the relationship among processes that simulate one another \citep{van2004equivalence}.
-Although this may frusturate the occasional engineer or scientist, viewed from another angle,
-the concepts of unidentifiability and indistinguishability can provide a starting point for
-thinking about externally equivalent systems -- systems that evolution can explore, so long
-as the parameters and structures can be realized biologically. In fact, evolutionary
-biologists who study homology and analogy are very familiar with such functional symmetries; macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level \citep{true2001developmental, tsong2006evolution, hare2008sepsid, vierstra2014mouse, stergachis2014conservation, taylor2016diverse, matsui2015regulatory}.
-
-In this paper we propose a framework to study the evolution of biological systems.
-To begin, we focus on the evolution of an idealized population.
-We consider the evolution of a perfectly adapted, large population, evolving in a static environment for an infinite number of generations.
-Under these ideal circumstances, we expect to observe a ``conservation of phenotype,'' where the population explores the manifold of phenotypically-invariant (or symmetric) genetic and developmental architecutres.
-We would like to understand which parameters influence the distribution of a population along the manifold of phenotypically invariant genetic systems.
-Further, we can show how dispersion along this manifold contributes to speciation and evolvability.
-
-\plr{phenomenologically or dynamically equivalent?}
-The set of all systems -- that is all model structures and all parameter schemes -- that
-are externally, or phenomenologically, equivalent, biologically realizable, and mutationally accessible,
-
-What are typically referred to as ``design principles'' or some sort of clear structure-function relationships are commonly saught \citep{milo2002network, alon2007network, ma2009defining}.
-
-Seeking to attribute functionality, uninformed by evolution, can lead to spurious claims
-(i.e. see Graur \citep{lynch2007evolution, lynch2007frailty, graur2013immortality}).
-
-A linear time invariant (LTI) dynamical system is a system of linear
-differential equations that describes a physical process. 
-LTI systems are used in electrical and control engineering to model a myriad of phenomena including
-circuits. 
-One can use this methodology, under a set of assumptions, to reverse engineer a
-mechanism from impulse experiments (input/output data). The idea is that given
-a black box, and it's experimental manipulation, can we describe the internal
-mechanisms?
-
-\section{The problem}
-
-Organisms often have to respond to their environments,
-and the way this happens is that external input triggers a cascade of molecular signals
-whose eventual result is a reaction,
-whose approprateness for the situation determines fitness.
-\plr{A similar situation occurs in development (...?).}
-However, there are many ways to skin a cat \jss{(there are also many ways to develop a cat)}:
-especially given a number of possible molecular intermediaries,
-there may be a large number of ways to construct such a molecular signalling system
-that has precisely the \emph{same} input-output relationships,
-and hence the same function.
-Therefore, any mutation that changes one such system to another, equivalent version
-will be neutral,
-and may well become polymorphic in a population if there are no ill effects in heterozygotes.
-Certainly, mutations that move a system away from optimal functioning will also appear,
-but in very large populations will not rise to high frequency.
-\plr{Deleterious things are quite important but somehow justify starting with the neutral ones.}
-
-\plr{
-    More stuff on why we care about this:  
-Provide neutral model; can address questions.  
-Main point of paper.
-}
-
-\begin{example}[Example: Oscillator.] \label{ex:osc}
-Motivate the system:
-\begin{align*}
-    A &= \left[\begin{matrix} 
-        0 & 1 \\ 
-       -1 & 0 
-    \end{matrix}\right]
-\end{align*}
-
+  \begin{slide}
        \begin{figure}
         \begin{center}
-            \begin{tikzpicture}
+          \begin{tikzpicture}[background rectangle/.style={fill=white}, show background rectangle]
             \begin{scope}[every node/.style={circle,thick,draw}]
-                \node (A) at (0,0) {$x_{1}$};
-                \node[dashed] (B) at (3,0) {$x_{2}$};
-                \node[shape=rectangle] (U) at (1.5,2) {input};
-                \node[shape=rectangle] (y) at (1.5,-2) {output};
+              \node[black] (A) at (0,0) {$x_{1}$};
+                \node[dashed, black] (B) at (3,0) {$x_{2}$};
+                \node[shape=rectangle, black] (U) at (1.5,2) {input};
+                \node[shape=rectangle, black] (y) at (1.5,-2) {output};
             \end{scope}
@@
     \end{figure}
+  \end{slide}
 
-as a special case of the below.
-
-\end{example}
-
-\begin{example}[Example: two-component systems.] \label{ex:2x2}
-As a more general example,
-suppose that a cell produces two molecular species, $S_1$ and $S_2$ 
-whose production are both stimulated by the presence of an external ``input'' substance.
-\plr{call them TF?}
-Suppose that the two species self-regulate with strengths $\lambda_1$ and $\lambda_2$, respectively,
-and the second species also regulates the first species with strength $\gamma$.
-\plr{Rephrase in terms of binding to promoter sequences?}
-However, only the time course of the concentration of
-the first one of these species determines the fitness of the organism.
-(As happens for instance in \plr{XXX}.)
-Write $x(t) = (x_1(t),x_2(t))$ for the concentrations of the two species at time $t$,
-and write $u(t)$ for the concentration at time $t$ of the input substance.
-Then, if the rates at which each species are produced
-are linear functions of the concentrations
-\plr{(look up way this is usually said in the literature.)},
-then the dynamics of the system are given by
-\begin{align*}
-    \ddt x_1(t) &= \lambda_1 x_1(t) + \gamma x_2(t) + u(t) \\
-    \ddt x_2(t) &= \lambda_2 x_1(t) + u(t) ,
-\end{align*}
-where $\ddt x$ denotes the time derivative.
-The initial conditions, $x_1(0)$ and $x_2(0)$, 
-and the input $u(t)$ then determine the concentrations through time.
-If we record the regulatory coefficients in the matrix:
-\begin{align*}
-    A &= \left[\begin{matrix} 
-        \lambda_1 & \gamma \\ 
-       0 & \lambda_2  
-    \end{matrix}\right],
-\end{align*}
-and define the column vector $B = [1,1]^T$,
-then in matrix notation the dynamics are
-\begin{align*}
-    \ddt x(t) &= A x(t) + B u(t) .
-\end{align*}
-If these are all transcription factors,
-and we suppose their binding motifs are fixed,
-then the $i^\text{th}$ row of $A$ is determined by the $i^\text{th}$ promoter sequence.
-
-\end{example}
-
-Since the system is linear, the state of the system at any time
-is the superposition of its responses to all previous inputs.
-This implies that 
-if $G(t)$ is the transient response of the system to a unit impulse,
-$t$ units of time later,
-and $x(0) = 0$,
-then the time course of the concentration of the thing we care about, $x_1(t)$, 
-can be written as
-\begin{align} \label{eqn:ex_convolution}
-    x_1(t) &= \int_0^t G(t-s) u(s) ds .
-\end{align}
-However, it turns out that 
-\plr{look for simple demontration this is true}
-\begin{align*}
-    P_{p} = \left[\begin{matrix} 
-        1 & 0 \\ 
-        p & 1-p 
-    \end{matrix}\right],
-\end{align*}
-then for any $p \neq 1$, 
-if we replace the regulatory matrix $A$ with
-\begin{align*}
-    A(p) = P A P^{-1},
-\end{align*}
-then the response of $x_1(t)$ to any particular input $u(t)$ is identical to the original system.
-(However, $x_2$ may well be different!)
-
-For instance, setting $p=-1$,
-the system with
-\begin{align*}
-    A(-1) = \left[\begin{matrix} 
-        \lambda_1 + \gamma/2 & \gamma/2 \\ 
-        \lambda_2-\lambda_1-\gamma/2 & \lambda_2 - \gamma/2
-    \end{matrix}\right],
-\end{align*}
-looks very different, but gives the same input-output relationship.
-Although these systems are equivalent,
-hybrids between them may not be:
-\plr{do example}.
-
-\begin{example}[Hyrbid Inviability]
-  Let, \begin{align*}
-    A(0) =\left[\begin{matrix}
-      0 & 1 \\
-      -1 & 0
-    \end{matrix}\right]
-    \text{and }
-    A(2) = \left[\begin{matrix}
-      2 & -1 \\
-      5 & -2
-    \end{matrix}\right]
-  \end{align*}
-  from Example \ref{ex:osc} ($B=\begin{bmatrix} 1 \\ 1 \end{bmatrix}$ and $C=\begin{bmatrix} 1 & 0 \end{bmatrix}$). 
-If hybrids are formed by exchanging genes (swapping rows), the hybrid's gene network is,
-  \begin{align*}
-    \text{hybrid network } := \mathcal{H}(A(0), A(2)) = \left[\begin{matrix}
-      2 & -1 \\
-      -1 & 0
-    \end{matrix}\right] ,
-  \end{align*}
-  \jss{ Or should it be: \begin{align*}
-    \text{hybrid network } := \text{mean}(A(0), A(2)) = \left[\begin{matrix}
-      1 & 0 \\
-      2 & -1
-    \end{matrix}\right] ,
-  \end{align*} }
-  It's transfer function will be $H(z, \mathcal{H}) = \frac{z-2}{z^2 - 3z +1}$, which produces very different dynamics than its parents, $H(z, A(0)) = H(z, A(2)) = \frac{z+1}{z^2 + 1}$.
-The time dynamcs of molecular species $1$ in the hybrid is $y^{(\mathcal{H})} =e^{t} \cosh(\sqrt{2} t)$ \jss{ which is approx.
-$\approx e^{t}$}, with the concentration of species $1$ rapidly heading towards infinity.
-The dynamics of species $1$ in the parental networks is $y = \sin(t) + \cos(t)$, where it continuously oscillates.
-
-  \jss{The transfer function $H(z, \mathcal{H})$, by averaging networks, is $\frac{1}{z-1}$ and the time dynamics are: $y^{(\mathcal{H})} = e^{t}$.}
-\end{example}
-
-\begin{definition}[Biological Species Concept]
-  Let $\mathcal{A}$ be the set of all phenotypically symmetric gene networks with transfer funtion $H(z)$. 
-  Choose $A$ and $\widetilde A$, both in $\mathcal{A}$. 
-  If $\mathcal{H}(A, \widetilde A) \notin \mathcal{A}$ then $A$ and $\widetilde A$ are distinct species. 
-\end{definition}
+  \begin{slide}
 
@@
   \centering
-  \begin{subfigure}{0.5\textwidth}
+  \begin{subfigure}{0.25\slideheight}
     \centering
-\begin{tikzpicture}
+\begin{tikzpicture}[background rectangle/.style={fill=white}, show background rectangle]
 \begin{scope}[every node/.style={circle,thick,draw}]
-  \node (A) at (0,0) {$x_{1}$};
-    \node[dashed] (B) at (4,0) {$x_{2}$};
-    \node[shape=rectangle] (U) at (2,2) {input};
-    \node[shape=rectangle] (y) at (2,-2) {output};
+  \node[black] (A) at (0,0) {$x_{1}$};
+    \node[dashed, black] (B) at (4,0) {$x_{2}$};
+    \node[shape=rectangle, black] (U) at (2,2) {input};
+    \node[shape=rectangle, black] (y) at (2,-2) {output};
 \end{scope}
@@
   \end{subfigure}%
-  \begin{subfigure}{0.5\textwidth}
+  \begin{subfigure}{0.25\slideheight}
     \centering
-\begin{tikzpicture}
+\begin{tikzpicture}[background rectangle/.style={fill=white}, show background rectangle]
 \begin{scope}[every node/.style={circle,thick,draw}]
-  \node (A) at (0,0) {$x_{1}$};
-    \node[dashed] (B) at (4,0) {$x_{2}$};
-    \node[shape=rectangle] (U) at (2,2) {input};
-    \node[shape=rectangle] (y) at (2,-2) {output};
+  \node[black] (A) at (0,0) {$x_{1}$};
+    \node[dashed, black] (B) at (4,0) {$x_{2}$};
+    \node[shape=rectangle, black] (U) at (2,2) {input};
+    \node[shape=rectangle, black] (y) at (2,-2) {output};
 \end{scope}
@@
 
-\section{Regulatory networks as linear, time-invariant systems}
-
-\plr{Make sure to have strongly worded disclaimer somewhere
-that we know real networks aren't linear
-but they are locally.
-Plus, it works in engineering.}
-
-\jss{Maybe we should discuss the Hartman-Grobman Theorem?}
-
-% - General description of the LTI process in biological context
-%     * y = "phenotype"
-%     * x = "internal state" or "kryptotype"?
-%     * A = "regulatory matrix" or "genetic network"?
-% - Why transfer function determines input-output relationship
-% - Kalman decomposition in unique form
-
-We will now lay out the model in more general terms.
-Suppose that the \emph{internal state} of the system
-is parameterized by the concentrations of a collection of $n$ molecular species,
-$S_1, \ldots, S_n$,
-and the vector of concentrations at time $t$ we denote $x(t)=(x_1(t),\ldots,x_n(t))$.
-There are also $m$ ``input'' species, whose concentrations are determined
-exogenously to the system,
-and are denoted $u(t) = (u_1(t),\ldots,u_m(t))$,
-and $\ell$ ``output'' species, whose concentrations are denoted
-$y(t) = (y_1(t),\ldots,y_\ell(t))$.
-The output is merely a linear function of the internal state:
-\begin{align*}
-    y_i(t) = \sum_j C_{ij} x_i(t).
-\end{align*}
-Since $y$ is what natural selection acts on, we refer to it as the \emph{phenotype},
-and sometimes in contrast refer to $x$ as the \emph{kryptotype},
-as it is ``hidden'' from direct selection.
-\plr{Maybe.}
-The rate at which the $i^\text{th}$ species is produced
-is a weighted sum of the concentrations of the other species
-as well as the input:
-\begin{align*}
-    \ddt x_i(t) = \sum_j A_{ij} x_j(t) + \sum_k B_{ik} u_k(t) .
-\end{align*}
-In matrix notation, this is written more concisely as
-\begin{align} \label{eqn:lti_system}
-    \ddt x(t) &= A x(t) + B u(t) \\
-    y(t) &= C x(t) .
-\end{align}
-
-\plr{Everything we say I think is assuming $x(0)=0)$.  
-Figure this out. ``Zero-state equivalence.''}
-
-Given an initial condition and an input, 
-it is possible to write the solution $x(t)$
-as a convolution with an input-response kernel as in the example above
-(equation \eqref{eqn:ex_convolution}).
-An alternative way of describing the input-output relationship,
-more common in engineering,
-is instead to find the \emph{transfer function}
-\begin{align} \label{eqn:transfer_fn}
-    H(z) = C(z I - A)^{-1} B,
-\end{align}
-which is more mathematically readable \plr{if you know what it means}.
-This can be thought of as the system's response to input at ``frequency'' $z$.
-The fact that the transfer function uniquely determines the system's
-input-output relationship (i.e., the mapping $u \mapsto y$)
-follows from the fact that if $\tilde y(z) = \int_0^\infty e^{-zt} y(t) dt$
-is the Laplace transform of $y$,
-and $\tilde u(z)$ is the Laplace transform of $u$,
-then
-\begin{align*}
-    \tilde y(z) = H(z) \tilde u(z) .
-\end{align*}
-(This happens because the Laplace transform takes convolutions to products,
-and $H$ is the Laplace transform of the kernel $G$.)
-Concretely,
-
-\plr{Should we remind what dimensions everything is?}
-
-The state space representation is considered an internal description of the
-system, whereas the transfer function is an external description. 
-``Realizations'' are something.
-
-\begin{definition}[Phenomenological equivalence of systems]
-    Let $(x(t),y(t))$ and $(\bar x(t),\bar y(t))$ be the solutions to \eqref{eqn:lti_system}
-    with coefficient matrices $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ respectively,
-    and both $x(0)$ and $\bar x(0)$ are zero. 
-    The systems defined by $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ are
-    \textbf{phenomenologically equivalent} 
-    if
-    \begin{align*}
-        y(t) = \bar y(t) \qquad \text{for all} \; t \ge 0.
-    \end{align*}
-    Equivalently, this occurs if and only if
-    \begin{align*}
-        H(z) = \bar H(z)  \qquad \text{for all} \; z \ge 0,
-    \end{align*}
-    where $H$ and $\bar H$ are the transfer functions of the two systems.
-\end{definition}
-
-One way to find other systems equvalent to a given one
-is by change of coordinates (``algebraic equivalance''):
-if $T$ is an invertible matrix, then the systems $(A,B,C)$ and $(TAT^{-1},TB,CT^{-1})$
-have the same dyanamics because their transfer functions are equal:
-\begin{align*}
-    CT^{-1}( zI - TAT^{-1})^{-1}TB
-    =
-    CT^{-1}T( zI - A)^{-1}T^{-1}TB
-    =
-    C( zI - A)^{-1}B .
-\end{align*}
-However, the converse is not necessarily true: 
-systems can have identical transfer functions without being changes of coordinates of each other.
-In fact, systems with identical transfer functions can involve interactions between different
-numbers of molecular species.
-
-The set of all systems phenomenologically equivalent to a given system $(A,B,C)$ 
-is elegantly described using the Kalman decomposition,
-which also clarifies the system dynamics? tells us a lot about how it works? \plr{or something}
-To motivate this, first note that the the input $u(t)$ only directly pushes the system
-in directions lying in the span of the columns of $B$.
-As a result, different combinations of input can 
-move the system in any direction that lies in the \emph{reachable subspace},
-which we denote by $\reachable$,
-and is defined to be the closure of $\spn(B)$ under applying $A$
-(or equivalently, the span of $B, AB, A^2B, \ldots A^{n-1}B$).
-Analogously to this, we define
-the \emph{observable subspace}, $\mathcal{O}$,
-to be the closure of $\spn(C^T)$ under applying $A$.
-(Or: $\unobservable$ is the largest $A$-invariant subspace
-contained in the null space of $C$;
-and $\reachable$ is the largest $A$-invariant subspace contained in the image of $B$.)
-
-If we define
-\begin{enumerate}
-    \item The columns of $P_\rno$ are an orthonormal basis for $\reachable \cap \unobservable$.
-    \item The columns of $P_\ro$ are an orthonormal basis of
-        the complement of $\reachable \cap \unobservable$ in $\reachable$.
-    \item The columns of $P_\nro$ are an orthonormal basis of
-        the complement of $\reachable \cap \unobservable$ in $\unobservable$.
-    \item The columns of $P_\nrno$ are an orthonormal basis of
-        the the remainder of $\R^n$.
-\end{enumerate}
-If we then define
-\begin{align*}
-    P &= 
-    \left[ \begin{array}{c|c|c|c}
-        P_\rno & P_\ro & P_\nro & P_\nrno
-    \end{array} \right] ,
-\end{align*}
-then
-\begin{align*}
-    P^T P
-    &=
-    \left[ \begin{array}{c|c|c|c}
-        I & 0 & 0 & 0 \\
-        \hline
-        0 & I & U & 0 \\
-        \hline
-        0 & V & I & 0 \\
-        \hline
-        0 & 0 & 0 & I 
-    \end{array} \right] .
-\end{align*}
-\plr{Check this.  Can we get $U=V=0$?}
-
-The following theorem can be found in SOME REFERENCE.
-
-\begin{theorem}[Kalman decomposition] \label{thm:kalman}
-        For any system $(A,B,C)$ with corresponding Kalman basis matrix $P$,
-        the transformed system $(PAP^{-1},PB,CP^{-1})$  has the following form:
-        \begin{align*}
-            \widehat A = PAP^{-1}
-            &=
-            \left[ \begin{array}{cccc}
-                A_{\rno} & A_{\rno,\ro} & A_{\rno,\nrno} & A_{\rno,\nro} \\
-                0 & A_{\ro} & 0 & A_{\ro,\nro} \\
-                0 & 0 & A_{\nrno} & A_{\nrno,\nro} \\
-                0 & 0 & 0 & A_{\nro}
-            \end{array} \right] ,
-        \end{align*}
-        and
-        \begin{align*}
-            \widehat B = PB
-            &=
-            \left[ \begin{array}{cccc}
-                B_{\rno} \\
-                B_{\ro} \\
-                0 \\
-                0 
-            \end{array} \right] ,
-        \end{align*}
-        and
-        \begin{align*}
-            \widehat C = CP^{-1}
-            &=
-            \left[ \begin{array}{cccc}
-                0 & C_{\ro} & C_{\nrno} & 0 
-            \end{array} \right] .
-        \end{align*}
-        The transfer function of both systems is given by
-        \begin{align*}
-            H(z) = C_{\ro} ( zI - A_{\ro} )^{-1} B_{\ro} .
-        \end{align*}
-\end{theorem}
-
-In the latter case, we say that the system is \emph{minimal} 
--- there is no equivalent system with a smaller number of species.
-Note that this says that any two equivalent minimal systems
-are changes of basis of each other.
-
-Since any system can be put into this form,
-and once in this form, its transfer function is determined only by 
-$C_{\ro}$, $A_{\ro}$, and $B_\ro$,
-therefore, the set of all equivalent systems are parameterized by
-the dimension $n$,
-the choice of basis ($P$),
-the remaining submatrices in $\widehat A$, $\widehat B$, and $\widehat C$
-(which are unconstrained),
-and a invertible transformation of $\spn(P_{\ro})$, which we call $T_\ro$.
-
-\begin{theorem}[Parameterization of equivalent systems]
-    Let $(A,B,C)$ be a minimal system.
-    \begin{itemize}
-        \item[(a)]
-            Every equivalent system is of the form given in Theorem \ref{thm:kalman},
-            i.e., can be specified by choosing a dimension, $n$;
-            submatrices in $\widehat A$, $\widehat B$, and $\widehat C$ 
-            except for $A_\ro=A$, $B_\ro=B$, and $C_\ro=C$;
-            and choosing an invertible matrix $P$.
-
-        \item[(b)]
-            \plr{conjecture:}
-            The parameterization is unique
-            if $P$ is futhermore chosen so that 
-            each $P_x$ other than $P_\ro$ is a projection matrix,
-            and that 
-            \begin{align*}
-                0
-                =
-                P_x^T P_y
-            \end{align*}
-            for all $(x,y)$ except $(\ro,\nrno)$.
-
-    \end{itemize} 
-\end{theorem}
-
-\plr{Another way of saying it: pick the $\reachable$ and $\unobservable$ subspaces,
-that must intersect in something of the minimal dimension;
-then let $P$ be the appropriate basis?}
-
-In some situations we may be interested in only ``network rewiring'',
-where $A$ changes while $B$ and $C$ do not.
-For instance, 
-if all non-regulatory functions of each molecule are strongly constrained,
-then $C$ cannot change.
-Likewise, if responses of each molecule to the external inputs are not changed by evolution,
-then $B$ does not change.
-
-\begin{lemma}
-    Let $(A,B,C)$ be a system.
-    Every matrix $\widehat A$ such that $(\widehat A,B,C)$ is equivalent to $(A,B,C)$
-    can be specified in the form given in Theorem \ref{thm:kalman},
-    with \plr{them having the same $\ro$ bit}
-    and satisfying
-    \begin{align*}
-        \widehat B &= P B \\
-        \widehat C P &= C
-    \end{align*}
-\end{lemma}
-
-It is remarkable to note that even with the relationships 
-between environment, cryptotype, and phenotype constant,
-and in the minimal dimension,
-there are still almost always degrees of freedom.
-These correspond to distinct genetic networks
-that perform indistinguishable functions.
-For example, the equivalent systems of Example \ref{ex:2x2} above
-are minimal, and share common $B$ and $C$ matrices.
-
-\plr{Maybe say that stability to noise can be determined by $A_\nro$ (and others?),
-while complexity of the internal, unobserved dynamics
-are somehow given by the difference between the dimension of $y$ and the dimension of $A_\ro$.}
-
-\plr{General argument why Brownian motion on set of equivalent things is maybe ok.}
-
-\begin{example}
-    The system in Example \ref{ex:2x2} is minimal, so $A=A_\ro$,
-    and the only free parameter is $P$.
-    Suppose also that the effects of the input and output are strongly constrained,
-    so that as in that example,
-    $B$ and $C$ cannot change,
-    and neutral evolution moves through the set of of $P$ such that $PB=B$ and $CP=C$.
-    \plr{Clarify we mean with no gene duplication and only neutral drift.}
-    How will this system change over evolutionary time in the manner described \plr{above}?
-    An unbiased Brownian motion on this set 
-    can be written in terms of a one-dimensional Brownian motion $\beta_\tau$ 
-    with $\beta_0=0$ as
-    \begin{align*}
-        P(\tau) = \left[
-            \begin{array}{cc}
-                1 & 0 \\
-                e^{\beta_\tau}-1 & e^{\beta_\tau}
-            \end{array}
-            \right] .
-    \end{align*}
-    Therefore, the state of the system after (evolutionary) time $\tau$
-    is $(P(\tau)AP(\tau)^{-1},B,C)$.
-\end{example}
-
-\section{Estimates of network rewiring and developmenal systems drift.} \label{sec:param}
-
-\jss{Parameter estimates for DSD $\downarrow$}
-
-What parameters (and other relevant factors) do we need from the literature and experiment? 
-  \begin{itemize}
-    \item mutation rate
-    \item Transcription factor binding site appearence and loss
-    \begin{itemize}
-      \item In \citep{lynch2007evolution} TF binding site gain ($u_{g}$) and loss ($u_{l}$) is estimated to be,
-        \begin{align*}
-          u_{g} &= \frac{(L-n+1)nu}{4^{n}} \\
-          u_{l} &= nu
-        \end{align*}
-      Where $L$ is the length of the promoter, $n$ is the number of nucleotides in a TFBS, and $u$ is the mutation rate per base pair per generation. 
-    \item \citet{lynch2007evolution} further states that in prokaryotes $L$ will usually be $\leq 100$ and in eukaryotes $L$ can range from $10^{3}$ to $10^{6}$.
-He also cites \citep{harbison2004transcriptional} that TFBSs range from $5-10$bps in length.
-    \item in \citep{lynch2014evolutionary}, they estimate
-        \begin{align*}
-          \tilde{P}(m) = C \left[ 3^{\ell - m}  \left( \begin{array}{c} \ell \\ m \end{array} \right) \right] e^{2N_{e}s(m)}  ,
-        \end{align*}
-        where $\tilde{P}(m)$ is the stationary distribution of a TFBS with match (and strength) $m$.
-$C$ is the normalizer constant, $m$ is the number of matches, $\ell$ is the length of the TFBS motif.
-The term on the right is the ratio of fixation probabilities.
-        Or in their own words, there are two interpretations: 
-        \begin{quote} ``(i) For a single TFBS, it represents the long-term proportion of evolutionary time spent in the various matching states and (ii) for a set of different TFBSs under the same selective constraints, it represents the expected distribution of states at any point in time.'' \end{quote}
-      \item \citet{tuugrul2015dynamics, anderson2015intermolecular} also may be helpful.
-    \end{itemize}
-    \item rate of both \emph{trans-} and \emph{cis-} regulatory evolution. (see Sergey's papers).
-      i.e. \citep{nuzhdin2004common, fear2016buffering, graze2009regulatory, graze2012allelic, genissel2008cis}
-    \item The PWN-score landscape (does binding strength increase linearly, exponentially, have a threshold, etc.) with respect to SNPs.
-    \item Others? 
-  \end{itemize}
-
-\section{(Old Notes) What biological question are we asking?}
-
-How do gene regulatory networks (GRNs) change through evolutionary time?
-Specifically, under constant selection and environmental pressures, how does a
-maximally fit population's GRN change and what molecular and population
-parameters are significant to the process? How frequent does gene network
-rewiring occur in the absense of genetic drift or adaptation? Is rewiring
-typically compensatory and ususally follow the fixation of a deleterious
-allele, or can networks neutrally reorganize? How does the size, average
-degree, function, peiotropy, etc of a network influence its likelihood to
-drift? If networks can drift, how much topological heterogeneity can a
-well-mixed population tolerate, if any? Does topoligical heterogeneity and/or
-the size of the neutral genotype-set, if consequences of network organization,
-constrain or entail exaptation and evolvability? Are some network organizations
-(maybe higher dimensional realizations?) able to more rapidly evolve to
-construct novel phenotypes and meet the demands of changes in selection
-pressures?
-
-\section{(Old Notes) How can we mathematically model this question?}
-
-In some respect an organism can be thought of as a black box that responds to
-some set of inputs (it's environment and other initial conditions) and outputs
-a phenotype. The phenotype is then evaluated by natural selection. The black
-box metaphor holds, because much of the details of what happens to construct a
-phenotype are unimportant as far as selection is concered, so long as the
-phenotype reliably performs its function. If multiple internal mechanisms can
-map the same set of inputs to the same set of outputs, and these mechanisms are
-mutationally nearby (in genotype space), then evolution may drift from
-mechanism to mechanism.
-
-In this simple model we can say that $B\vec{u}$ is a list of initial GRN
-protein concentrations determined by the more general initial environmental
-conditions $\vec{u}$. $A$ represents all the genetic interactions of a
-particular GRN, and $C\vec{x}$ is an organism's phenotype. Fitness could be
-calculated by comparing a subset of the phenotype to an optimum. Maybe,
-$f(\cdot) = e^{- \int_{a}^{b} \Vert G(s) - G^{*}(s) \Vert ds}$ Alternatively,
-fitness scores could simply assess whether or not a phenotype is within some
-range or breaks some threshold by some time point. Or possbily the simplest:
-any phenotype not exactly equal to $G(s)$ is lethal, and $G(s)$ is perfectly
-fit.
-
-Next, we will have to define mutation and recombination. Mutation needs to add
-and remove genetic interactions and recombination needs to shuffle genes during
-sexual reproduction. We could adapt methods from Lynch 2007 and Lynch and
-Hagner 2015 to model the probability of a TFBS appearing or disappearing due to
-mutational pressure. Every time a new binding site is gained or lost, the
-values within the A matrix are either modified or rows and columns are added
-removed following a specific set of rules. I am not sure if it would be easier
-to maintain a matrix size (say $m \times m$) where $m$ is arbitrarily large,
-with  most entries being zero, or if a matrix should only have as many
-dimensions as the number of active TFs. Recombination should then shuffle rows
-of the $A$ matrix randomly, as each row represents the regulatory region of a
-given gene (assuming only cis-regulation, and that the regulatory element is
-small enough to not break up during recombination).
-
-To study some of the above posed questions, we can simulate (or analytically
-determine) the rates of GRN change, and how frequently these changes lead to
-significant rewiring or speciation. We would want to know the dynamics of a
-brownian motion over the set of all realizations.  Another interesting question
-would be to determine the size of the neutral genotype space as well as the
-number of connections from the neutral genotype space to the non-neutral
-genotype space. Maybe higher dimensional realizations of a GRN will have
-significantly more connections to non-neutral genotypes and therefore be more
-``evolvable'' or exapted to novel environments?
-
-Let $M \in \mathbb{R}^{n \times n}$ be a least dimensional realization of $A$,
-and let $M'$ bean $(n+m \times n+m)$ matrix where $M'_{(i,j)} = M_{(i,j)} \
-\text{for } i,j \leq n, \ \text{and } 0 \ \text{otherwise}$. Let $P$ be any
-$(n+m \times n+m)$ matrix with rank $n$. Then for any matrix $B$ such that $BP
-= PM'$, $B$ is a realization of $A$.
-
-\section{(Old Notes) List of papers with some notes.}
-
-R. E. Kalman, “Mathematical description of linear dynamical systems,” Journal of the Society for Industrial and Applied Mathematics, Series A: Control, vol. 1, no. 2, pp. 152–192, 1963.
-This is a good overview written by Kalman. Probably the most important paper on this list.
-
-Minimal state-space realization in linear system theory: an overview. B. De Schutter. 2000.
-Another, more recent, overview that I found helpful to understand controllability and observability.
-
-The Inverse Problem of Stationary Covariance Generation. BDO Anderson. 1969.
-
-Equivalence of Linear Time- Invariant Dynamical Systems. BDO Anderson, BW Necomb, RE Kalman. 1966.
-
-B.
-Ho and R.
-E.
-Kalman, “Effective construction of linear state-variable models from input/output functions,” Proceedings of the 3rd Annual Allerton Conference on Circuit and System Theory, vol.
-1, no.
-2, p.
-449459, 1965.
-
-Dynamic Structure Functions for the Reverse Engineering of LTI Networks. J Goncalves. 2007.
-I have not read this paper yet, but apparently it outlines a method for recovering the state space equation from specific sets of measurements.
-
-New Developments in Systems Theory Relevant to Biology. RE Kalman. 1968.
-Not technical but kind of interesting.
-
-Old and New Directions of Research in Systems Theory. RE Kalman. 2010?
-
-Minimality and Observability of Group Systems. HA Loeliger, GD Forney Jr., T Mittelholzer, MD Trott. 1994. 
-Another paper I just found. It looks like it might be useful.
-
-Partial Realization of Descriptor Systems. P Benner and VI Sokolov. 2006. 
-I think this is about recovering a finite realization from infinite systems.
-Maybe it would be interesting to think about the similarity between transfer functions though?
-If selection only cares about a specific interval of the transfer function, say $G(10)$, then maybe two partial transfer functions will be effecrively equivlaent?
-
-There are also the previous papers I sent on identifiability and distinguishability.
-In the original Bellman and Astrom paper where they introduce ``structural identifiability'' they show that a linear dynamical system is identifiabile if and only if it is also controllable and observable.
-
-\section{Including Deleterious Mutations and a Fitness Function}\label{sec:del}
-
-  \begin{minipage}{0.5\textwidth}
-  At the moment we focus strictly on the case where the minimal realization is in $2$ dimensions and both of its eigenvalues are real and negative (so a special case of the system from Example \ref{ex:2x2}).
-\jss{(Note: this is not to say the ideas below will not be generalizable).}
-
-    The system of interest has the form,
-  \begin{align*}
-    \Sigma := \left\{ A = \begin{bmatrix} A_{11} & A_{12} \\ A_{21} & A_{22} \end{bmatrix}, B = \begin{bmatrix} 1 \\ 1 \end{bmatrix}, C = \begin{bmatrix} 1 & 0 \end{bmatrix} \right\}
-  \end{align*}
-  \end{minipage}%
-  \begin{minipage}{0.5\textwidth}
-    \centering
-    \includegraphics[width=0.95\textwidth]{figure_2x2mutation01}%
-      \captionof{figure}{The transfer function for $A = \left[ \protect\begin{array}{cc} -3 & 1 \\ 0 & -1 \protect\end{array} \right]$ and all eight $\mu(\Sigma, A_{ij})$ mutations, where $\upsilon = \pm 0.5$}\label{fig:mut}%
-  \end{minipage}
-
-Define the fitness function to be,
-
-\begin{align}\label{eq:fitness}
-  \Phi(\bar{\Sigma}) := \left\{ \begin{array}{ll}   e^{- \left| H(0) - \tilde{H}(0) \right| = - \left| \frac{k}{\lambda_{1} \lambda_{2}} - \frac{\tilde{k}}{\tilde{\lambda}_{1} \tilde{\lambda}_{2}} \right|}    &\text{if } \lambda_{1,2} < 0 \text{ and } \in \mathbb{R}, \ \det(A) \neq 0 \\
-  0   &\text{if } \lambda_{1,2} \geq 0 \ \text{ or } \ \notin \mathbb{R}, \ \det(A) = 0 \end{array} \right.
-\end{align}
-and let $\upsilon$ be the magnitude and direction (sign) of a mutation on regulatory interaction strength.
-The fitness function (\ref{eq:fitness}) may me appropriate if mutations only slightly alter the dynamics of the system, as in Figure \ref{fig:mut}.
-
-%Further let $\mu(\Sigma, A_{ij})$ be the resulting gene regulatory network following a mutation of matrix element $A_{ij}$ to system $\Sigma$ (which contains matrix $A$).
-%
-%  \begin{align*}
-%    \Phi(\mu(\Sigma, A_{ij} \upsilon)) = \exp\left(- \left| \frac{A_{22}-A_{12}}{\lambda_{1} \lambda_{2}} - \frac{k + \upsilon \pda (A_{22}-A_{12})}{\lambda_{1} \lambda_{2} + \text{adj}(A)^{T}_{ij} \upsilon} \right|\right)   .
-%  \end{align*}
-%For instance, the fitness of an organism with a change of one of its entries $A_{ij}$ (by $\upsilon$) is listed,
-%  \begin{align*}
-%    \Phi(\mu(\Sigma, A_{11})) &= \exp \left(- \left| \frac{A_{22}-A_{12}}{\det(A)} - \frac{A_{22}-A_{12}}{\det(A) + \upsilon A_{22}} \right| \right) \\
-%    \Phi(\mu(\Sigma, A_{12})) &= \exp \left(- \left| \frac{A_{22}-A_{12}}{\det(A)} - \frac{A_{22}-A_{12} - \upsilon}{\det(A) - \upsilon A_{21}} \right| \right) \\
-%    \Phi(\mu(\Sigma, A_{21})) &= \exp \left(- \left| \frac{A_{22}-A_{12}}{\det(A)} - \frac{A_{22}-A_{12}}{\det(A) - \upsilon A_{12}} \right| \right) \\
-%    \Phi(\mu(\Sigma, A_{22})) &= \exp \left(- \left| \frac{A_{22}-A_{12}}{\det(A)} - \frac{A_{22}-A_{12} + \upsilon}{\det(A) + \upsilon A_{11}} \right| \right)
-%  \end{align*}
-%
- The transfer function at $z=0$ for any GRN with mutations is, 
-  \begin{align*}
-    H(0, \begin{bmatrix} A_{11} + \upsilon & A_{12} + \varepsilon \\ A_{21} + \delta & A_{22} + \omega \end{bmatrix}) = \frac{A_{22} - A_{12} + \omega - \varepsilon}{ \det(A) + \omega A_{11} - \delta A_{12} - \varepsilon A_{21} + \upsilon A_{22} + \omega \upsilon - \varepsilon \delta }
-  \end{align*}
-
-%  \begin{align*}
-%    \frac{\left(A_{22}-A_{12}\right) + \left( \displaystyle\sum_{\jmath = 1}^{\ind_{\omega}} \omega_{\jmath} - \displaystyle\sum_{\jmath = 1}^{\ind_{\varepsilon}} \varepsilon_{\jmath} \right)}{\displaystyle\det(A) + \displaystyle\sum_{\jmath=1}^{\ind_{\omega}} \omega_{\jmath} A_{11} - \sum_{\jmath=1}^{\ind_{\delta}} \delta_{\jmath} A_{12} -  \sum_{\jmath=1}^{\ind_{\varepsilon}} \varepsilon_{\jmath} A_{21} + \sum_{\jmath=1}^{\ind_{\upsilon}} \upsilon_{\jmath} A_{22} + \sum_{\jmath=1}^{\ind_{\omega}} \omega_{\jmath} \sum_{\jmath=1}^{\ind_{\upsilon}} \upsilon_{\jmath} - \sum_{\jmath=1}^{\ind_{\varepsilon}} \varepsilon_{\jmath} \sum_{\jmath=1}^{\ind_{\delta}} \delta_{\jmath} }
-%  \end{align*}
-
-  If we want to know how new mutations fix in the population, we can explicitly show all mutational paths that could compensate for fixed mutations at sites $A_{ij}$.
-In a population, not at a phenotypic optimum, we are interested in the combination of mutations with varying magnitudes and locations that are most likely to fix.
-
-  \begin{align*}
-    \frac{\left(A_{22}-A_{12}\right) + \left( \displaystyle\sum_{\jmath = 1}^{\ind_{\omega}} \omega_{\jmath} - \displaystyle\sum_{\jmath = 1}^{\ind_{\varepsilon}} \varepsilon_{\jmath} \right)}{\det(A) + \displaystyle\sum_{\mu=\{\omega , \delta , \varepsilon , \upsilon \}}^{\ind_{\mu}} \mu_{\jmath} (-1)^{i+j} A_{ij} + \sum_{\jmath=1}^{\ind_{\omega}} \omega_{\jmath} \sum_{\jmath=1}^{\ind_{\upsilon}} \upsilon_{\jmath} - \sum_{\jmath=1}^{\ind_{\varepsilon}} \varepsilon_{\jmath} \sum_{\jmath=1}^{\ind_{\delta}} \delta_{\jmath} }
-  \end{align*}
-
-\section{Biological Examples}
-
-  \begin{example}
-    Negative feedback and Positive feedback systems:
-    \begin{align*}
-      T(z) = \frac{H(z)}{1 + G(z)H(z)} \\
-      T(z) = \frac{H(z)}{1 - G(z)H(z)} \\
-    \end{align*}
-  \end{example}
-
-  \begin{example}
-    Systems in serial ($u \rightarrow \Sigma_{1} \{A_{1},B_{1},C_{1} \} \rightarrow \Sigma_{2} \{A_{2},B_{2},C_{2} \} \rightarrow y$):
-    \begin{align*}
-      \begin{bmatrix} \dot{x}_{1} \\ \dot{x}_{2} \end{bmatrix} &= \left[ \begin{array}{c|c}
-        A_{1} & 0 \\
-          \hline
-          B_{2}C_{2} & A_{2} \\
-        \end{array} \right] \begin{bmatrix} \dot{x}_{1} \\ \dot{x}_{2} \end{bmatrix}  + \begin{bmatrix} B_{1} \\ 0 \end{bmatrix} u \\
-          y &= \begin{bmatrix} 0 & C_{2} \end{bmatrix} \begin{bmatrix} \dot{x}_{1} \\ \dot{x}_{2} \end{bmatrix} .
-    \end{align*}
-
-  Systems in serial could be applied to model a gene network $A_{1}$ and the subsequent degradation rates of its outputs at rate $A_{2}$.
-  \end{example}
-
-  \begin{example}
-  \begin{align*}
-    \dot{x} &= \begin{bmatrix} + & - \\ 0 & + \end{bmatrix} \vec{x} + \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \vec{u} \\
-      y(t) &= \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \vec{x} \\
-        E(t) &= \begin{bmatrix} -y_{1}(t) & 0 & 0 \\ y_{1}(t) & -y_{2}(t) & 0 \\ -\sigma_{1} y_{1}(t) & (1 - \sigma_{2})y_{2}(t) & 1 \end{bmatrix} \vec{\zeta} + \begin{bmatrix} u_{1} \\ u_{2} \\ \epsilon \end{bmatrix} .
-  \end{align*}
-    Where $u_{1}$ is input of \emph{lactose}, $u_{2}$ is the input of \emph{glucose}, $y_{1}$ is the enzyme \emph{lactase} and $y_{2}$ is enzyme \emph{``glucase''}.
-$E(t)$ is metabolism of the organism, converting sugars to energy as a function of enzyme concentrations.
-Lactase converts lactose to glucose, and the enzyme is produced at energy cost $\sigma_{1}$; glucase converts glucose to energy and is produced at cost $\sigma_{2}$.
-Fitness is a function of the total energy of the system, given inputs, possibly: $\int_{0}^{T}E(t)_{3} dt$ after some time $T$ has passed.
-Or perhaps, once the total metabolic energy produced passes some threshold, the organism undergoes fission.
-Thus, fitness selects for maximizing energy as quickly as possible.
-  \end{example}
-
-  \begin{example} 
-    Gene Regulatory and Metabolic Network Cascade:
-
-    Gene Regulatory Network
-    \begin{align*}
-      \dot{x}(t) &= A x(t) + B_{1} u(t) \\
-      y(t) &= C_{1} x(t)
-    \end{align*}
-      Metabolism
-      \begin{align*}
-      \dot{\zeta}(t) &= M\big(y(t), \sigma \big) \zeta(t) + B_{2} \begin{bmatrix} u(t) \\ v(t) \end{bmatrix} \\
-        m(t) &= C_{2} \zeta(t) \\
-      b(t) &= C_{3} \zeta(t)
-      \end{align*}
-      Fitness
-      \begin{align*}
-      \Phi(m(t), p(t))
-    \end{align*}
-
-    The GRN produces time varying proteins involved in the metabolism.
-The metabilic network (MN), is a function of protein concentrations $y(t)$ and production costs $\sigma$.
-The metabolic network produces energy $m(t)$ and toxic byproducts $b(t)$.
-Fitness $\Phi$, or maybe something more concrete, like fission time and/or cell death, is a function of both energy $m(t)$ and toxins $b(t)$.
-  \end{example}
-\section{Notes on what to do when allowing deleterious mutations}
-
-\begin{enumerate}
-
-    \item Epistasis, coming from mutations deleterious in some networks but not in others
-(come from incompatibilities).
-
-    \item Describe heterosis.
-
-    \item Distribution of dominance coefficients.
-
-    \item Evolvability: easier for bigger networks to adapt to a new environment?
-
-    \item Do short-term responses to strong directional selection have deleterious epistatic consequences in other traits?
-
-\end{enumerate}
-
-\bibliographystyle{plainnat}
-\bibliography{krefs}
-
+\end{slide}
 \end{document}

### 2017-06-07 attic/model_notes.tex
added to attic

\documentclass[11 pt]{article}
\usepackage{amsmath, amssymb, color, xcolor}
\usepackage{graphicx, wrapfig, float, caption, dsfont, bbm}
\usepackage{fullpage}
\usepackage[backref=page, hidelinks, colorlinks=true, citecolor=blue!60!black!100]{hyperref}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, shapes}
\usepackage{caption, subcaption}
\usepackage{natbib} % gives us \citet: Author (year) and \citep: (Author; year)
\usepackage{authblk}

\newcommand{\plr}[1]{{\color{blue}\it #1}}
\newcommand{\jss}[1]{{\color{olive}\it #1}}
% \newcommand{\ddt}{\frac{d}{dt}}
\newcommand{\ddt}{\dot}
\newcommand{\ro}{{ro}}
\newcommand{\nro}{{\bar{r}o}}
\newcommand{\rno}{{r\bar{o}}}
\newcommand{\nrno}{{\bar{r}\bar{o}}}
\newcommand{\reachable}{\mathcal{R}}
\newcommand{\unobservable}{\bar{\mathcal{O}}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\pda}{\frac{\partial}{\partial A_{ij}}}
\newcommand{\ind}{\mathds{1}}

\newcommand{\A}{\mathcal{A}}
\newcommand{\diag}{\text{diag}}
\newcommand{\1}{\mathbbm{1}}

\DeclareMathOperator{\spn}{span}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}

\begin{document}

\section{Introduction}

%In this paper we propose a framework to study the evolution of biological systems. To begin, we focus on the evolution of an idealized population. We consider the evolution of a perfectly adapted, large population, evolving in a static environment for an infinite number of generations. Under these ideal circumstances, we expect to observe a ``conservation of phenotype,'' where the population explores the manifold of phenotypically-invariant (or symmetric) genetic and developmental architecutres. We would like to understand which parameters influence the distribution of a population along the manifold of phenotypically invariant genetic systems. Further, we can show how dispersion along this manifold contributes to speciation and evolvability. 
\section{Notes and Ideas}

    We aim to answer the following questions and demonstrate:
      \begin{itemize}
        \item What does evolution, under phenotypic conservation, look like?
        \item How many different molecular mechanisms can realize the same phenotype? 
        \item How does evolution traverse the manifold of phenotypically identical mechanisms? 
        \item What parameters influence the qualitative aspects of neutral gene network evolution?
        \item How neutral evolution can cause speciation via reproductive isolation in F1s and F2s. 
        \item How neutral evolution influences robustness and evolvability.
        \item How these processes contribute to biological complexity (i.e. is there a ``network ratchet''? How strongly -- if at all -- does selection favor minimal molecular mechanisms?).
        \item Can we distinguish between adaptation and systems drift?
      \end{itemize}

\section{Model}
    \begin{itemize}
      \item Under the application of constant population and molecular parameters \jss{(forces?)}, such as selection, (a sufficiently large) population size, (a sufficiently small) mutation rate, stable ecology, \emph{etc.}, phenotype is invariant through evolution.
      \item Genes $\rightarrow$ Phenotype $\rightarrow$ Strategy $\rightarrow$ Fitness $\rightarrow$ Genes
    \end{itemize}

      \jss{note: add demonstration of degrees of freedom. Specifically show that if $B=C=I$ and the dimension is minimal, then there are no degrees of freedom, any $CV=C$ and $VB=B$ entails $V=I$.
      Maybe also try to do an example where speciation/hybrid incompatibility is impossible? So if the BSC concept is that, the set of high fitness GRNs is closed under the action of $V$, maybe there exits a phenotype that is also closed under averaging and recombination -- rendering incompatibility impossible -- even if very trivial it could be interesting.}

(Evolution can be described as a mapping from genes to phenotype to strategy to fitness back to genes.)

To obtain maximal fitness, an organism needs to correctly execute at least one or more of potentially an ifinite number of survival strategies. These strategies include choosing which molecule(s) to metabolize, which behavior(s) to exhibit, and which niches to compete in. Additionally, a strategy's fitness is context dependent. The organism needs to organize its phenotype in a highly specific way to play any one or more of these strategies. Presumably this is achieved by the coordinated expression of evolutionarily pliable gene networks. 
As such, we can approximately say that evolutionary fitness presupposes a termporally fluctuating set (potentially infinite) of gene-permutations with expression levels. For instance, fitness may discrimate based on an organism's ability to aquire and metabolize energy, but is ambivalent as to the source. If there are two distinct sources of energy, the combination and expression patterns of at least two (or a combination of the two) gene networks may be deemed fit. Despite this complexity, at present, we focus on the case where only one specific combination of genes and their expression levels are suitable at a time -- that is a one strategy only environment.

In this scenario, once a large population reaches a phenotype of optimal fitness, it will maintain this phenotype for infinitely long, or until a change in selection (i.e. environmental or ecological change). Here we study the null model of molecular evolution without selective flux.

    \subsection{LE Systems}
    An LE System: Linear Evolving (TI) System (in evolutionary state $\tau$), 
    \begin{align*}
      \Sigma(\tau) = \left\{ \begin{array}{cc} \dot{x} =& A(\tau)x + Bu \\
      y =& Cx \end{array} \right.
    \end{align*}
    \begin{align*}
      \mathcal{L}(\Sigma(\tau)) := H(z, \tau) = C \left(zI - A(\tau) \right)^{-1} B
    \end{align*}
    \subsection{Gene Network Diversity}
      First we need to describe the phase space. How many different gene networks can do exactly the same thing? Rather, can we describe the set of phenotypically invariant gene regulatory networks -- the set of all gene networks that produce identical outputs, when given identical inputs. In the linear case we can analytically describe all the networks in this space,
      \begin{align*}
        \mathcal{A} := \{ VAV^{-1}, CV, VB : V \in GL_{n}(\mathbb{R}) \}
      \end{align*}
      for the minimal dimension, and for the general (including non-minimal) case as,
      \begin{align*}
        CA^{k}B = \widehat{C} \widehat{A}^{k} \widehat{B} \qquad .
      \end{align*}
      where $A$ is a decription of gene network topology, and $B$ and $C$ transform the system intputs to and outputs from the network, respectively.

  \subsection{Constant Selection Through Time}

    A large population, perfectly adapted to it's environment, under constant selection, will be phenotypically invariant for all generations (given the circumstances outlined above). Despite a conservation of phenotype, variation at the genetic and interemediate molecular levels is possible and expected through the course of evolution, as shown below. Under these constraints, phenotypic variance can only be the consequence of genetic drift (stochastic evolutionary forces overpowering selective ones due to population size) or adaptation to a novel phenotype for a different strategy (signatures of each will be discussed).

      For convenience, we hold $B$ (input transformation) and $C$ (output transformation) constant, considering only systems with varying internal mechanics.
      \begin{align*}
        P(\vec{r}) \in \{ GL_{n}(\mathbb{R}) : CP = C \text{and } BP = B \} \\
        A(\vec{r}) := P(\vec{r}) A P^{-1}(\vec{r})
      \end{align*}

      Where $A := A(0)$ be the initial and minimal configuration of a population's gene network.
      Where $\vec{r}$ is the unique vector of free variables associated with any instantiation of $P$.
      Elements of $\vec{r}$  can be restricted to a biologically reasonable range, $s \leq r_{k} \leq l$ to prevent unrealistic topologies.
      From this persepctive, the values of $A(r)$ can be written as the $n^{2}$ vector $\vec{a}(r) = \text{vec}(A(r))$. In this space, evolution acts as a random walk on $r$. 
      Furthermore, let $\Delta a(r) := \lVert \frac{d}{dr} a(r) \rVert^{-1}$. $\Delta a(r)$ describes the average magnitude of neutral genetic change required to reach a phenotypically invariant gene network. Systems should neutrally drift and rewire at rates inversely proportional to this value, as at high values of $\Delta$, more mutations are required to rewire neutrally -- that is the fitness valley is wider. Given enough time, we expect evolution to traverse the space for all of $r$, albeit at different rates and probabilities.

      Perhaps the rate of phenotypically invariant evolution near state $\tau$ can be expressed as,
      \begin{align*}
        \kappa :&= \mu \Delta_{r} = \mu \left\lVert \frac{d}{d r} \text{vec}\left[A(r)\right] \right\rVert^{-1} \\
        r &= f(\tau). 
      \end{align*}
      Where $\kappa$ is the speed the kryptotype evolves.
      \jss{Maybe need to distinguish between differentiating from the left and right of $\tau$ since we want to know the rate and direction of evolution. i.e. $\partial_{\pm} f(A(\tau))$}

      \subsection{Reproduction}
      \begin{align*}
        \text{Meiosis } &\rightarrow \begin{bmatrix} 1 & 0 \\ 0 & 0 \end{bmatrix} A_{\text{genome }1} + \begin{bmatrix} 0 & 0 \\ 0 & 1 \end{bmatrix} A_{\text{genome }2} \\
          &\rightarrow M = \text{diag}(\text{Bernoulli}(0.5)_{i}), \qquad M A_{1} + (I-M)A_{2} \\
        \text{Reproduction } &\rightarrow \frac{1}{2}(A_{\text{parent }1} + A_{\text{parent }2})
      \end{align*}
      So F1 and F2 hybrids would have the GRNs,
      \begin{align*}
        \text{F}1 &\rightarrow \frac{1}{2}(A(r) + A(\rho)) \\
        \text{F}2 &\rightarrow \frac{1}{2} \big( M A(r) + ( I - M ) A(\rho) + \widehat{M} A(r) + ( I - \widehat{M} ) A(\rho) \big)
      \end{align*}

      \subsection{Recombination -- Idea}

      Let each chromosome $\mathcal{C}$ be swapped with probability $0.5$, so the meiosis matrix $\mathcal{M}$ will have an identity matrix of appropriate dimension for each chromosome. Let intra-chromosomal recombination occur with probability $\omega \left(\gg 0.5 \right)$, breaking up the chromosomal identity matrices:
        \begin{align*}
          \mathbbm{1}_{x_{i}} &= \left\{ \begin{array}{cc} \mathcal{C}_{i} \qquad &\text{if Bernoulli}_{i}(0.5) = 1 \\ 0 \qquad &\text{if Bernoulli}_{i}(0.5) = 0 \end{array} \right. \\
            \mathcal{M} &= \begin{bmatrix} \mathbbm{1}_{x_{1}} &  & \\ & \ddots & \\ & & \mathbbm{1}_{x_{\ell}} \end{bmatrix} \\
              \mathcal{C}_{i} &= \text{diag}\left(\text{Bernoulli}\left( \omega \right) \right) \\
            A_{(\text{gamete})} &= \mathcal{M} A_{\text{genome }1} + \left(I - \mathcal{M} \right) A_{\text{genome }2}
        \end{align*}
        This formulation may be useful to see how deleterious epistatic fitness effects might not appear in a population until many generations into the future.

        After how many generations, do we expect to observe all gamete combinations. What frequency will each gamete combination occur at? 
      \subsection{Hybridization}
       F1 hybrids between $A(0)$ and $A(r)$, for the parental phenotype $H(z) = \displaystyle\frac{z-k}{\left(z-\lambda_{1}\right)\left(z-\lambda_{2}\right)}$,
      \begin{align*}
        A\left(\mathcal{H}(\, 0,r)\, \right) = \frac{1}{2} \left( A(0) + A(r) \right) = \frac{2}{(r-1)} \left[\begin{array}{cc} r\, \left(2\, \lambda_{1} - k + \lambda_{2}\right) - 2\, \lambda_{1} & - \left(k - \lambda_{2}\right)\, \left(r - 2\right)\\ \left(\lambda_{1} - k\right)\, r^2 + \left(\lambda_{2} - \lambda_{1}\right)\, r & r\, \left(k + \lambda_{2}\right) - 2\, \lambda_{2} \end{array}\right]
      \end{align*}
      F2 hybrids' gene networks are formed from the gametes of the F1s. So gametes ($G_{i}$) are formed during meiosis by swapping genes (rows) and then averaging the two new genomes in the F2 hybrid $h_{ij}$ as follows,
      \begin{align*}
        G_{1} :&= A(0) \\
        G_{2} := \text{diag}(1,0) A(0) + \text{diag}(0,1) A(r) &= \left[\begin{array}{cc} \lambda_{1} & \lambda_{2} - k\\ \lambda_{1}\, r - \frac{r\, \left(r\, \left(k - \lambda_{2}\right) + \lambda_{2}\, \left(r - 1\right)\right)}{r - 1} & \frac{r\, \left(k - \lambda_{2}\right) + \lambda_{2}\, \left(r - 1\right)}{r - 1} \end{array}\right] \\
          G_{3} := \text{diag}(0,1) A(0) + \text{diag}(1,0) A(r) &= \left[\begin{array}{cc} \lambda_{1}\, r - \frac{r\, \left(r\, \left(k - \lambda_{2}\right) + \lambda_{2}\, \left(r - 1\right)\right)}{r - 1} & \frac{r\, \left(k - \lambda_{2}\right) + \lambda_{2}\, \left(r - 1\right)}{r - 1}\\ 0 & \lambda_{2} \end{array}\right] \\
            G_{4} :&= A(r) \\
            h_{ij} :&= \frac{1}{2} \left( G_{i} + G_{j} \right)
      \end{align*}
      \jss{How does the size of a GRN contribute to hybrid incompatibility? As $n$ increases, the probability of an $A(r) \times A(r)$ cross in an F2, or even a completely $A(r)$ gamete forming during meiosis, drops rapidly.}

     \subsection{Fitness and Incompatibility}
     Hybrid Incompatibility via reproductive isolation can happen in a variety of ways. In general, if two phenotypically invariant gene networks are highly diverged such that in $A(r)$ and $A(\rho)$, $|r-\rho| \gg 0$, or if a major rewring occurs, such that in the previous example $r$ crosses from greater than to less than $1$, or vice versa, or if in special cases, the network's expression patterns are especially sensitive to perturbation.

     However, there may be other mechanisms that lead to hybrid dysfunction. Perhaps, even a minor change in phenotype, due to negatively epistatic networks interacting, across many different networks can significantly alter a hybrid's fitness. Imagine two populations, slightly diverged, such that the first is homozygous for the $A(r)$ gene network, and the second is homozygous for the $A(r+1)$ network topology. Qualitative expression patterns in most hybrids might mirror those of the parents almost exactly -- but not quite. If the fitness cost is small, say $s=0.01$, the fitness of the hybrid would be $\Phi(h) = e^{-0.01} = 0.99$, or only $0.01$ less than each of it's parents. However, if each other hybrid gene network is equally and slightly perturbed, and the organism has $200$ networks (a lower limit estimate in humans), and fitness is multiplicitive, the fitness of a hybrid between the two populations would be $\Phi(h_{(\text{all})}) = \left( e^{-0.01} \right)^{200} \approx 0.14$, despite the fact that each individual network is only slighly less fit than the optimum. 
     \begin{align*}
        \Phi(\cdot) := \prod_{\gamma = 1}^{G} e^{-s_{\gamma}}
     \end{align*}

     Types of DMIs,
    \begin{itemize}
      \item ``parameter bifurcation'' or maybe ``rewiring'' (i.e. when $r$ jumps over $1$ in $P(r)$, as $P(r) \notin \text{GL}_{2}(\mathbb{R}$ because it is not invertible. 
      \item large neutral ``binding shifts'' or ``regulatory coefficient divergence'' when $| r - \rho | \gg 0$ for ($P(r)$ and $P(\rho)$.
      \item Dimension divergence -- If $A \in \text{GL}_{n}(\mathbb{R})$ and $\widehat{A} \in \text{GL}_{m}(\mathbb{R})$ such that $ n \neq m$.
    \end{itemize}
        These rates may be very different. If a population finds itself near its ``rewiring/bifurcation'' point $\beta$, such that $r=\beta + \varepsilon$, speciation may happen very rapidly, whereas speciation via binding shifts may take significantly more time, if possible at all, depending on the fitness function and the biological likelihood of such shitfts. For instance, $A(r)$ with respect to a theoretically possible $A(\rho)$ may exhibit a signifcant binding shift, however the coefficients set by $\rho$ may be to extreme to be realized in a biological system (\emph{e.g.} a gene regulates another greater than $\sim 10^{6}$ times the rate that itself is regulated. While these two systems may produce incomaptible hybrids, $A(\rho)$ is said to be an extremeley unlike biological manifestation of $\A$.
        Understanding the rates at which populations move in $\A$, and at which point in $\A$ gene flow was diminished among populations will allow us to predict the likelihood and rates of reproductive isolation. If $\Delta_{\tau} : \Delta_{\tau} := \mu \left\lVert \frac{d}{d \tau}\text{vec}\left[A(\tau)\right]\right\rVert^{-1}$, is very small around $r$, then binding shifts should occur rapidly. Perhaps rewiring will occur at a rate proportional to $\Delta^{2}_{\beta +\varepsilon}$ (as $\Delta_{\beta +\varepsilon} \rightarrow 0$ as $\varepsilon \downarrow 0$)?

        It may also be possible that evolution tends to favor gene networks with small regulatory coefficients. At least in the above example network, $\lVert A(r) \rVert$ is smallest as $r$ is close $\beta$. If this is the case, systems drift may be very rare, and may only/typically be due to rewiring and not shifting. Small regulatory coefficients may be favored if larger coefficients require the selective maintence of many more transcription factor binding sites in each promoter (this  may not be  a problem in populations with sufficiently small mutation rates?)?

   \section{Model Notes Version 2}

   \subsection{Evolutionarily Constant Input and Output Transformations}
    If $\Sigma_{\min}$ is a minimal system with transfer function $H(z)$, and $P$ is any invertible matrix that preserves $B$ and $C$, 
    \begin{align*}
      \Sigma_{\min} &= \{A, \ B, \ C \} \\
      P \in \mathcal{P} :&= \{ \text{GL}_{n} : PB = B, \ CP = C \text{ and } B,C \in \Sigma_{\min} \} \\
      A(\tau) \in \A :&= \{P A P^{-1} : P \in \mathcal{P} \text{ and } A \in \Sigma_{\min} \}
    \end{align*}
    $A(\tau)$ is a potential minimal gene network topology that also satisfies $H(z)$.

    If,
    \begin{align*} 
      B \lor C^{T} = I  &\implies \\
      \mathcal{P} = \{ I \} &\implies \\
      \left\vert \A \right\vert &= 1
    \end{align*}\label{eq:k}
   then there is only one network topology of $A$ (as $\A$ only contains one element).

   Not all of the molecules and dynamics of a system, even if the system is minimal, may be directly visible to selection. As such, we distinguish between the \emph{phenotype} and the \emph{kryptotype} (\emph{pheno- ``to show''} and \emph{krypto- ``to hide''}), as selection acts directly on the phenotype molecules and only indirectly (if at all) on the molecules in the kryptotype. In \ref{eq:k} if $C=I$, the dynamics of each molecule in the system are visible to selection, and there exists only one minimal gene network organization that can precisely satisfy natural selection. It is important to note, however, that despite the uniqueness of a $A$ for the minimal case, there are an infinite number of gene network topologies with equivalent phenotypic behavior for the non-minimal case. As any gene network $\widehat{A}$ that satisfies, 
   \begin{align*}
     CA^{k}B &= \widehat{C} \widehat{A}^{k} \widehat{B} \qquad \text{for } k = 0, 1, \dots
   \end{align*}
 will be selectively indistinguishable. \footnote{If the system under study is the one in \ref{eq:k}, then for biological congruity, it probably makes sense to restrict $\widehat{B}$ and $\widehat{C}^{T}$ to the form $\left[ \begin{array}{c} I \\ \hline 0 \end{array} \right]$.} Thus in principle, under the minimal set of assumptions of this model, there are always multiple gene network topologies (and in different dimensions) that can biologically realize a given phenotype.

   To further understand neutral gene nework evolution, we need to understand how the kryptotype evolves and is contrained. Not all kryptotpes may be epistatically compatible and could lead to negative epistasis in complex traits as well as low fitness hybrids between reproducitvely isolated populations.

   \subsection{DMIs}
   Just as there are some minimal systems that cannot drift (without sacrificing fitness), there may also be systems incapable of hosting \emph{Dobzhansky-Muller Incompatibilities} (\emph{DMIs}) despite drifting.
   In the non-minimal set of systems $\A$ below,
   \begin{align*}
     H(z) :&= \frac{2}{z+3} , \ A \in \mathbb{R}^{2 \times 2}, \ C = \begin{bmatrix} 1 & 1 \end{bmatrix}, \ B = \begin{bmatrix} 1 & 1 \end{bmatrix}^{T} \\
      \A :&= \left\{ \begin{array}{cc} \left[ \begin{array}{cc} a_{11} & a_{12} \\ a_{21} & a_{22} \end{array} \right] : \begin{array}{cc} a_{11} + a_{12} &= -3 \\
   a_{21} + a_{22} &= -3 \end{array} \end{array} \right\}
   \end{align*}
   All reproductive and meiotic combinations of any element in $\A$ are also in $\A$. Thus, in the above example $\A$ is closed under recombination and reproduction. (Note: One can see that this system could easily tolerate gene duplications and evolve into an $n \times n$ system, as the phenotype is constant so long as the $n$ elements in each row sum to $-3$).

   The neutral manifold of gene network topologies for the system above is an affine, and therefore convex set. A convex set, by definition, is closed under averaging, thus any F1 hybrid among drifted networks over a convex manifold will not suffer from any incompatibilities.

   Despite the above system's immunity to negative epistasis following systems drift, many molecular systems, even minimal ones, can suffer from negative epistasis and harbor DMIs.
   For instance, the system outlined below, which oscillates in two parental populations, can be deleteriously expressed as a simple ON switch in certain F1 hybrids.

      Parental gene networks have topologies,
      $A(0) =\left[\begin{matrix}
      0 & 1 \\
      -1 & 0
    \end{matrix}\right]$
     and
    $A(2) = \left[\begin{matrix}
      2 & -1 \\
      5 & -2
    \end{matrix}\right]$ with impulse response dynamics, $f(t) = \sin(t) + \cos(t)$, can hybridize to form,
     $\mathcal{H}\left( A(0), A(2) \right) = \left[\begin{matrix}
      1 & 0 \\
      2 & -1
     \end{matrix}\right]$ which has impulse response dynamics, $f(t) = e^{t}$.

     \section{Version 3: Model Notes}

     \subsection{Outline/Premises}

      An organism is constructed by an elaborate process including complex environment-by-gene, and gene-by-gene interactions. We refer to the (heritable) characteristics of an organism that influence its chance of replication as its phenotype. In otherwords the phenotype is defined as heritable characteristics that natural selection \emph{observes} and filters. Traditionally, biologists distinguish only phenotype from genotype, however, here we will introduce (and justify the necessity of) an organism's \emph{kryptotype}. If one crudely imagines an organism to be a \emph{black box}, taking its environment as input and outputting a phenotype, its kryptotype can be thought of as the mechanism inside the blackbox as its specifics are hidden from selection. We will discuss the properties of kryptotypes, and primarily focus on the non-uniqueness and diversity of the kryptotype -- how the kryptotype can evolve and influence a myriad of evolutionary and biological phenomena such as speciation and evolvability.

     \subsection{Description of LTI and LE Systems}

     \subsection{Examples of LE Systems: kryptotype (do multiplie exist?), DMIs, Evolvability}

     \subsection{Applications to ``real'' systems: Gal Metabolism, Gap Development, Cell Cycle Control, etc.}

     \subsection{Conclusions and Future work}

     \section{Notes for May presentation}
       \begin{align*}
         \Sigma(0) := \left\{ \begin{array}{ll}
         \dot{x} &= \left[ \begin{array}{cc} 0 & 1 \\ -1 & 0 \end{array} \right] x + \left[ \begin{array}{c} 1 \\ 1 \end{array}\right] u \\
           y &= \left[ \begin{array}{cc} 1 & 0 \end{array} \right] x
         \end{array} \right.
    \end{align*}

    \begin{align*}
      \Sigma(0) := \left\{ \begin{array}{ll}
      \dot{x}_{1} &= +x_{2} + u \\
      \dot{x}_{2} &= -x_{1} + u \\
      y &= x_{1}
      \end{array} \right.
    \end{align*}

       \begin{align*}
         \Sigma(2) :&= \left\{ \begin{array}{ll}
         \dot{x} &= \left[ \begin{array}{cc} 2 & -1 \\ 5 & -2 \end{array} \right] x + \left[ \begin{array}{c} 1 \\ 1 \end{array}\right] u \\
           y &= \left[ \begin{array}{cc} 1 & 0 \end{array} \right] x
         \end{array} \right. \\ \\
         \Sigma(2) :&= \left\{ \begin{array}{ll}
        \dot{x}_{1} &= 2 x_{1} + - x_{2} + u \\
        \dot{x}_{2} &= 5 x_{1} + -2 x_{2} + u \\
      y &= x_{1}
      \end{array} \right.
    \end{align*}
    \begin{align*}
      \mathcal{L}(\Sigma(\tau)) := H(z, \tau) = C \left(zI - A(\tau) \right)^{-1} B
    \end{align*}

    \begin{align*}
      f(t) &= \left( e^{\lambda_{1} t} \left(\lambda_{1} - k\right) - e^{\lambda_{2}t}\left(\lambda_{2} - k \right) \right) \frac{1}{\lambda_{1}-\lambda_{2}} \\
        H(z) &= \mathcal{L}\left(f(t)\right) \\ \\
        H(z) &= \frac{z - \lambda_{2}}{\left(z -\lambda_{1} \right)\left(z - \lambda_{2} \right)} \\
        &= C( zI - A)^{-1}B
    \end{align*}

    \begin{align*}
        H(z) &= CP^{-1}( zI - PAP^{-1})^{-1}PB \\
        &= CP^{-1}P( zI - A)^{-1}P^{-1}PB \\
        &= C( zI - A)^{-1}B \\ \\
      PB &= B \\
      CP &= C
    \end{align*}

    \begin{align*}
      P(\tau) :&= \begin{bmatrix} 1 & 0 \\ \tau & 1-\tau \end{bmatrix} \qquad \tau \neq 1 \\ \\
        A(\tau) :&= P(\tau)AP^{-1}(\tau)
    \end{align*}

    \begin{figure}
    \centering
\begin{tikzpicture}
\begin{scope}[every node/.style={circle,thick,draw}]
  \node (A) at (0,0) {$x_{1}$};
    \node[dashed] (B) at (4,0) {$x_{2}$};
    \node[shape=rectangle] (U) at (2,2) {input};
    \node[shape=rectangle] (y) at (2,-2) {output};
\end{scope}

\begin{scope}[>={Stealth[black]},
              every node/.style={fill=white,circle},
              every edge/.style={draw=black, thick}]
    \path [->, sloped] (A) edge[bend left] node {$5$} (B);
    \path [->, sloped, >=Rectangle] (B) edge[bend left] node {$-1$} (A); 
    \path[->] (U) edge[dashed] node {$1$} (A);
    \path[->] (U) edge[dashed] node {$1$} (B);
    \path[->] (A) edge[dashed, bend right] node {$1$} (y);
\end{scope}
\begin{scope}[>={Stealth[black]},
              every edge/.style={draw=black, thick}]
    \path [->] (A) edge[loop left] node {$2$} (A);
    \path [->,>=Rectangle] (B) edge[loop right] node {$-2$} (B);
\end{scope}

\end{tikzpicture}
    \end{figure}

    \begin{figure}
    \centering
\begin{tikzpicture}
\begin{scope}[every node/.style={circle,thick,draw}]
  \node (A) at (0,0) {$x_{1}$};
    \node[dashed] (B) at (4,0) {$x_{2}$};
    \node[shape=rectangle] (U) at (2,2) {input};
    \node[shape=rectangle] (y) at (2,-2) {output};
\end{scope}

\begin{scope}[>={Stealth[black]},
              every node/.style={fill=white,circle},
              every edge/.style={draw=black, thick}]
    \path [->, sloped] (A) edge[bend left] node {\tiny $2 \tau + (\tau-1)^{-1}$} (B);
    \path [->, sloped] (B) edge[bend left] node {\tiny $-(\tau-1)^{-1}$} (A); 
    \path[->] (U) edge[dashed] node {\tiny $1$} (A);
    \path[->] (U) edge[dashed] node {\tiny $1$} (B);
    \path[->] (A) edge[dashed, bend right] node {$1$} (y);
\end{scope}
\begin{scope}[>={Stealth[black]},
              every edge/.style={draw=black, thick}]
    \path [->] (A) edge[loop left] node {\tiny $1 + (\tau-1)^{-1}$} (A);
    \path [->] (B) edge[loop right] node {\tiny $-1 - (\tau-1)^{-1}$} (B);
\end{scope}

\end{tikzpicture}
    \end{figure}
    \begin{align*}
      A(\tau) = \left[ \begin{array}{cc} \lambda_{1} - \frac{\tau \left(k-\lambda_{2}\right)}{\tau - 1} & \frac{k-\lambda_{2}}{\tau - 1} \\ \lambda_{1} \tau - \frac{\tau \left( \tau \left(k-\lambda_{2}\right) + \lambda_{2} \left(\tau-1\right)\right)}{\tau - 1} & - \frac{\lambda_{2} - k \tau}{\tau - 1} \end{array} \right]
    \end{align*}

     \begin{align*}
       \mu \Delta_{\tau} \\
       \Delta_{\tau} :&= \left\lVert \frac{d}{d \tau} \text{vec}\left[A(\tau)\right] \right\rVert^{-1} \\
      \end{align*}

      \begin{align*}
      \Delta\text{Tr} = \frac{r_{2}\left(r_{4} + r_{3} \left(\gamma-k\right)\right)}{r_{4}\left(1-r_{1}\right) + r_{2}r_{3}} 
      \end{align*}

\bibliographystyle{plainnat}
\bibliography{krefs}

\end{document}

### 2017-06-07 kalman_walk.tex
added to attic

@@
 
+\section{Algorithm for Gene Network Growth}
+
+Let $A^(d)$ be a gene network with index $(d)$; $(d)$ is the difference between the specific realization's dimension and the dimension of a minimal realization of the requisite transfer function, such that $A^{(0)}$ is a minimal realization, $A^{(1)}$ has one superfluous dimension and $A^{(-1)}$ is one short.
+
+Algorithm for constructing $A^{(d+1)}$ from $A^{(d)}$,
+
+\begin{align*}
+  A^{(d+1)} = \left[ 
+  \begin{array}{c|c}
+    A^{(d)} & 0 \\
+    r_{1} \dots r_{n+d} & r_{n+d+1}
+  \end{array}
+  \right].
+  \end{align*}
+
+  Two systems have equivalent external dynamics iff their transfer functions are identical. Two transfer functions are identical iff $CA^{k}B = \bar{C}\bar{A}^{k}\bar{B} \qquad k = 0, 1, \dots$.
+
+  If system $\bar{A} := A^{(1)}$ is an extension of the minimal system $A^{(0)}$ then it must be of the form,
+
+  \begin{align*}
+    \left[ \begin{array}{c|c} C & 0 \end{array} \right]
+      \left[ \begin{array}{cc} V & \Psi \\ \Phi & \Gamma \end{array} \right]
+        \left[ \begin{array}{cc} \Lambda & 0 \\ 0 & I \end{array} \right]
+          \left[ \begin{array}{cc} V^{-1} & \Omega \\ \Theta & I \end{array} \right]
+            \left[ \begin{array}{cc} B \\ \hline 0 \end{array} \right] \\
+              \Phi \Theta = 0  \\
+              \bar{C} \bar{A}^{k} \bar{B} = (CV\Lambda^{k}V^{-1}B + C \Psi \Theta B).
+  \end{align*}
+
 \section{Adaptation}
@@
 
-\section{(Old Notes) List of papers with some notes.}
-
-R. E. Kalman, “Mathematical description of linear dynamical systems,” Journal of the Society for Industrial and Applied Mathematics, Series A: Control, vol. 1, no. 2, pp. 152–192, 1963.
-This is a good overview written by Kalman. Probably the most important paper on this list.
-
-Minimal state-space realization in linear system theory: an overview. B. De Schutter. 2000.
-Another, more recent, overview that I found helpful to understand controllability and observability.
-
-The Inverse Problem of Stationary Covariance Generation. BDO Anderson. 1969.
-
-Equivalence of Linear Time- Invariant Dynamical Systems. BDO Anderson, BW Necomb, RE Kalman. 1966.
-
-B.
-Ho and R.
-E.
-Kalman, “Effective construction of linear state-variable models from input/output functions,” Proceedings of the 3rd Annual Allerton Conference on Circuit and System Theory, vol.
-1, no.
-2, p.
-449459, 1965.
-
-Dynamic Structure Functions for the Reverse Engineering of LTI Networks. J Goncalves. 2007.
-I have not read this paper yet, but apparently it outlines a method for recovering the state space equation from specific sets of measurements.
-
-New Developments in Systems Theory Relevant to Biology. RE Kalman. 1968.
-Not technical but kind of interesting.
-
-Old and New Directions of Research in Systems Theory. RE Kalman. 2010?
-
-Minimality and Observability of Group Systems. HA Loeliger, GD Forney Jr., T Mittelholzer, MD Trott. 1994. 
-Another paper I just found. It looks like it might be useful.
-
-Partial Realization of Descriptor Systems. P Benner and VI Sokolov. 2006. 
-I think this is about recovering a finite realization from infinite systems.
-Maybe it would be interesting to think about the similarity between transfer functions though?
-If selection only cares about a specific interval of the transfer function, say $G(10)$, then maybe two partial transfer functions will be effecrively equivlaent?
-
-There are also the previous papers I sent on identifiability and distinguishability.
-In the original Bellman and Astrom paper where they introduce ``structural identifiability'' they show that a linear dynamical system is identifiabile if and only if it is also controllable and observable.
-
 \section{Including Deleterious Mutations and a Fitness Function}\label{sec:del}

### 2017-06-07 handnotes.tex
added to attic
added figure F1

@@
 
+      \begin{figure}[H]
+        \centering
+        \includegraphics[width=0.5\textwidth]{A2_A2-1_F2s}
+        \caption{F2s from $A(2)$ and $A(2.1)$.}
+      \end{figure}
+
       \begin{example}[Not all Networks can Host Incompatibilities]
@@
 
+      \jss{Discussion guidelines:
+        \begin{itemize}
+        \item Why is this important/useful?
+        \item What are the assumptions and shortcomings of the research?
+        \item Compare to other studies in the literature. 
+        \item Future directions. 
+        \item Wild speculations?
+        \item Conclusion and overall impact. 
+      \end{itemize}}
+
+  The complexity of biological systems has limited our understanding of their function and evolution.
+Above we outline an approach, a first step, towards untangling this complexity in reference to function and evolution.
+This methodology borrows successfully applied tools from engineering and aims to synthesize these with the concepts and tools of molecular and evolutionary biology.
+Theoretical models in evolution and population genetics often lack the molecular details of physiology or of the genotype-phenotype map.
+Here, we offer a tractable and simple model which includes these missing features.
+Further, we provide, in clear mathematical language, a description of phenomena hitherto only discussed verbally and conceptualy (phenogenetic drift, developmental systems drift, biological degeneracy, \emph{etc.}).
+The tractability and relative simplicity of this exposition enables the interested biologist to work out by hand, if desired, the dynamics of a genetic system, as well as perturbations to the system -- an attribute not likely to be found in less tractable models and simulations.
+
       \tiny

### 2017-06-08 handnotes.tex
peter begun editing

@@
 
-\newcommand{\A}{\mathcal{A}}
+\newcommand{\calA}{\mathcal{A}}
 \newcommand{\diag}{\text{diag}}
@@
   That is, all (linear and minimal) mechanisms capable of producing the same phenotype can be realized by a unique choice of $\tau$ in $\Sigma(\tau)$.
+
+  More generally, \plr{introduce Kalman},
+  we denote by $\calA_n(A_0)$ the set of all $n$-dimensional systems equivalent to $A_0$:
+  \begin{align}
+      \calA_n(A_0) = \left\{
+          A \in R^{n \times n} : C e^{At} B = C e^{A_0 t} B \quad \text{for all}\; t \ge 0 
+          \right\} .
+  \end{align}
+  Usually, the reference system $A_0$ is implicit and we write only $\calA$.
+
   \begin{example}[External Equivalence does not imply internal equivalence]
@@
   \end{example}
-  Evolution thus proceeds as a random walk in phenotypically invariant network space.
+
+  %%%%%%%%%%%%%%%%%%%%%%%%
+  \section*{Genetic drift}
+
+  In the absence of directional evolution by selective pressures,
+  demographic and mutational noise
+  will cause a population to move randomly on (or near) the space of phenotypically equivalent systems.
+  On short to intermediate time scales this is thought to proceed mostly through genetic drift
+  resampling available diversity.
+  The amount and structure of 
+  this standing variation is maintained on long time scales
+  by mutation-selection balance
+  shifts in the phenotypic optimum,
+  diversifying selection, and/or other factors \citep{mut_sel_balance}.
+  We will take a quantitative genetics framework to describe this process.
+
+  If the system is a network of transcription factors,
+  each row of the matrix $A$ represents the regulatory region of a particular transcription factor gene.
+  Regulatory regions are typically kilobases in length,
+  and so are expected to be represented in a population by a moderate number of haplotypes,
+  within which both mutation and recombination generate new diversity at roughly equal rates.
+  Nonetheless, as a first approximation we will assume that within the population,
+  the distribution of additive genetic variation in each entry of $A$ is multivariate Gaussian
+  with covariance matrix $G$.
+  The state of the population is then determined by the population mean and covariance matrix.
+  We then further decompose the population mean system into a position in the set of equivalent, optimal systems 
+  (a projection onto the set $\calA$, which will usually be unique)
+  and the deviation of the mean from that set.
+  We would like to then consider these two components separately,
+  since deviations away from the optima $\calA$ occur within populations,
+  and so should be close to neutral,
+  while two populations whose means drift in different directions along $\calA$
+  may develop incompatibilities.
+
+  Suppose that a large population with effective size $N_e$ has population mean $\bar A = A_0 + Z$,
+  where $A_0$ is the closest point in $\calA$ and $Z$ is the deviation from $\calA$,
+  and suppose the population varies about this point with covariance between $A_{ij}$ and $A_{k\ell}$
+  equal to $G_{ij,k\ell}$.
+
+  Treating $A$ as a vector, the mean per-generation displacement of $\bar A$ is approximately $G (G+E)^{-1} S$
+  where $E$ is the covariance matrix of environmental noise affecting the phenotypes,
+
+  It also seems reasonable to assume that population variation occurs on a smaller scale
+  than the effects of selection -- 
+  otherwise, a substantial portion of new offspring would die due to (recombination) genetic load each generation \citep{genetic_load}.
 
@@
 This methodology borrows successfully applied tools from engineering and aims to synthesize these with the concepts and tools of molecular and evolutionary biology.
+
   Theoretical models in evolution and population genetics often lack the molecular details of physiology or of the genotype-phenotype map.
 Here, we offer a tractable and simple model which includes these missing features.
-Further, we provide, in clear mathematical language, a description of phenomena hitherto only discussed verbally and conceptualy (phenogenetic drift, developmental systems drift, biological degeneracy, \emph{etc.}).
+Further, we provide, in clear mathematical language, a description of phenomena hitherto only discussed verbally and conceptualy (phenogenetic drift \citep{weiss2000phenogenetic}, developmental systems drift \citep{true2001developmental}, biological degeneracy \citep{edelman2001degeneracy}, \emph{etc.}).
 The tractability and relative simplicity of this exposition enables the interested biologist to work out by hand, if desired, the dynamics of a genetic system, as well as perturbations to the system -- an attribute not likely to be found in less tractable models and simulations.
+
+  We have offered a novel interpretation of system identification: to see it as an evolutionarily neutral manifold, and not simply a computational nuissance.
+We have demonstrated a method to analytically determine the set of all phenotypically invarint gene networks; by a simple change of coordinates in the minimal configuration, or more generally by applying the Kalman decomposition in higher dimensions.
+Further, we emphasize that evolution proceeds through this high dimensional space as stochastic coordinate transformation, constrained by sexual reproduction and selection.
+This set is explored over evolutionary time when phenotype is conserved, and can lead to a diverse set of consequences, including the accumulation of Dobzhansky-Muller incompatibilities.
+We emphasize that these incompatibilities are a consequence of recombining different, yet functionally equivalent, mechanisms.
+We also suggest that gene networks may not always use their components parsimoniously as network size tends to ratchet up in the absence of strong selection against extra parts.
+Although unexplored presently, this phenomena may lead to insights on evolvability and developmental innovation.
+Lastly, we show that hybrid gene networks break down as function of genetic distance, and may, in part, explain broad patterns of reproductive isolation among diverse phyla \citep{roux2016shedding}.
+
+  As Richard Levins opined, models in population biology face a tradeoff among precision, realism, and generality \citep{levins1966strategy}.
+As Levins expects, any tractable and general model, such as the present one under discussion, will have limitations.
+Most notable is linearity.
+It is often stated that life is not linear.
+This is often true, however, many of the ideas developed here should be generalizable to nonlinear cases (multilinear systems, say).
+Futher, we see this as a necessary first step in the direction of more life-like nonlinear evolutionary systems theory.
+Depending on an actual biological system's particularities, its (potential) nonlinearity, may buffer or exacerbate effects elucidated in this paper, such as the aquisition of Dobzansky-Muller incompatibilities.
+
+  This theoretical framework can easily be applied to other interesting questions in evolutionary biology not tackled presently: such as the evolution of linkage, the necessity of network complexity (does evolution tend towards Rube Goldberg or \jss{parsimonious} network organization?), evolvability, structure/function inference, and intrapopulation context dependency of mutational effects, as well as many others.
 

### 2017-06-08 handnotes.tex
edited discussion

@@
 
-\newcommand{\calA}{\mathcal{A}}
+\newcommand{\A}{\mathcal{A}}
 \newcommand{\diag}{\text{diag}}
@@
   That is, all (linear and minimal) mechanisms capable of producing the same phenotype can be realized by a unique choice of $\tau$ in $\Sigma(\tau)$.
-
-  More generally, \plr{introduce Kalman},
-  we denote by $\calA_n(A_0)$ the set of all $n$-dimensional systems equivalent to $A_0$:
-  \begin{align}
-      \calA_n(A_0) = \left\{
-          A \in R^{n \times n} : C e^{At} B = C e^{A_0 t} B \quad \text{for all}\; t \ge 0 
-          \right\} .
-  \end{align}
-  Usually, the reference system $A_0$ is implicit and we write only $\calA$.
-
   \begin{example}[External Equivalence does not imply internal equivalence]
@@
   \end{example}
-
-  %%%%%%%%%%%%%%%%%%%%%%%%
-  \section*{Genetic drift}
-
-  In the absence of directional evolution by selective pressures,
-  demographic and mutational noise
-  will cause a population to move randomly on (or near) the space of phenotypically equivalent systems.
-  On short to intermediate time scales this is thought to proceed mostly through genetic drift
-  resampling available diversity.
-  The amount and structure of 
-  this standing variation is maintained on long time scales
-  by mutation-selection balance
-  shifts in the phenotypic optimum,
-  diversifying selection, and/or other factors \citep{mut_sel_balance}.
-  We will take a quantitative genetics framework to describe this process.
-
-  If the system is a network of transcription factors,
-  each row of the matrix $A$ represents the regulatory region of a particular transcription factor gene.
-  Regulatory regions are typically kilobases in length,
-  and so are expected to be represented in a population by a moderate number of haplotypes,
-  within which both mutation and recombination generate new diversity at roughly equal rates.
-  Nonetheless, as a first approximation we will assume that within the population,
-  the distribution of additive genetic variation in each entry of $A$ is multivariate Gaussian
-  with covariance matrix $G$.
-  The state of the population is then determined by the population mean and covariance matrix.
-  We then further decompose the population mean system into a position in the set of equivalent, optimal systems 
-  (a projection onto the set $\calA$, which will usually be unique)
-  and the deviation of the mean from that set.
-  We would like to then consider these two components separately,
-  since deviations away from the optima $\calA$ occur within populations,
-  and so should be close to neutral,
-  while two populations whose means drift in different directions along $\calA$
-  may develop incompatibilities.
-
-  Suppose that a large population with effective size $N_e$ has population mean $\bar A = A_0 + Z$,
-  where $A_0$ is the closest point in $\calA$ and $Z$ is the deviation from $\calA$,
-  and suppose the population varies about this point with covariance between $A_{ij}$ and $A_{k\ell}$
-  equal to $G_{ij,k\ell}$.
-
-  Treating $A$ as a vector, the mean per-generation displacement of $\bar A$ is approximately $G (G+E)^{-1} S$
-  where $E$ is the covariance matrix of environmental noise affecting the phenotypes,
-
-  It also seems reasonable to assume that population variation occurs on a smaller scale
-  than the effects of selection -- 
-  otherwise, a substantial portion of new offspring would die due to (recombination) genetic load each generation \citep{genetic_load}.
+  Evolution thus proceeds as a random walk in phenotypically invariant network space.
 
@@
 
-  We have offered a novel interpretation of system identification: to see it as an evolutionarily neutral manifold, and not simply a computational nuissance.
-We have demonstrated a method to analytically determine the set of all phenotypically invarint gene networks; by a simple change of coordinates in the minimal configuration, or more generally by applying the Kalman decomposition in higher dimensions.
-Further, we emphasize that evolution proceeds through this high dimensional space as stochastic coordinate transformation, constrained by sexual reproduction and selection.
-This set is explored over evolutionary time when phenotype is conserved, and can lead to a diverse set of consequences, including the accumulation of Dobzhansky-Muller incompatibilities.
-We emphasize that these incompatibilities are a consequence of recombining different, yet functionally equivalent, mechanisms.
-We also suggest that gene networks may not always use their components parsimoniously as network size tends to ratchet up in the absence of strong selection against extra parts.
-Although unexplored presently, this phenomena may lead to insights on evolvability and developmental innovation.
-Lastly, we show that hybrid gene networks break down as function of genetic distance, and may, in part, explain broad patterns of reproductive isolation among diverse phyla \citep{roux2016shedding}.
-
   As Richard Levins opined, models in population biology face a tradeoff among precision, realism, and generality \citep{levins1966strategy}.
@@
 
+  We have offered a novel interpretation of system identification: to see it as an evolutionarily neutral manifold, and not simply a computational nuissance.
+We have demonstrated a method to analytically determine the set of all phenotypically invarint gene networks; by a simple change of coordinates in the minimal configuration, or more generally by applying the Kalman decomposition in higher dimensions.
+Further, we emphasize that evolution proceeds through this high dimensional space as stochastic coordinate transformation, constrained by sexual reproduction and selection.
+This set is explored over evolutionary time when phenotype is conserved, and can lead to a diverse set of consequences, including the accumulation of Dobzhansky-Muller incompatibilities.
+We emphasize that these incompatibilities are a consequence of recombining different, yet functionally equivalent, mechanisms.
+We also suggest that gene networks may not always use their components parsimoniously as network size tends to ratchet up in the absence of strong selection against extra parts.
+Although unexplored presently, this phenomena may lead to insights on evolvability and developmental innovation.
+Lastly, we show that hybrid gene networks break down as function of genetic distance, and may, in part, explain broad patterns of reproductive isolation among diverse phyla \citep{roux2016shedding}.
+
       \tiny

### 2017-06-09 handnotes.tex
edited discussion
drift written up

@@
   That is, all (linear and minimal) mechanisms capable of producing the same phenotype can be realized by a unique choice of $\tau$ in $\Sigma(\tau)$.
+
+  More generally, \plr{introduce Kalman},
+  we denote by $\calA_n(A_0)$ the set of all $n$-dimensional systems equivalent to $A_0$:
+  \begin{equation} \label{eqn:equivalence}
+      \begin{aligned}
+      \calA_n(A_0) 
+          &= \left\{
+              A : C e^{At} B = C e^{A_0 t} B \; \text{for}\; t \ge 0 
+          \right\}  \\
+          &= \left\{
+              A : C A^k B = C A_0^k B \; \text{for}\; 1 \le k \le n-1 
+          \right\} .
+      \end{aligned}
+  \end{equation}
+  Equivalence of the two characterizations follows from the Cayley-Hamilton theorem.
+  Usually, the dimension $n$ and the reference system $A_0$ is implicit and we write only $\calA$.
+
   \begin{example}[External Equivalence does not imply internal equivalence]
@@
   \end{example}
-  Evolution thus proceeds as a random walk in phenotypically invariant network space.
+
+  %%%%%%%%%%%%%%%%%%%%%%%%
+  \section*{Intraspecific variation and genetic drift}
+
+  At any given time, there will be a range of network coefficients present in the population
+  due to segregating genetic polymorphism.
+  Over many generations, even if selective pressures do not change,
+  this range of networks will shift 
+  as recombination, mutation, and demographic noise create new alleles and shift allele frequencies.
+  How much variation do we expect to find within a population?
+  Is this range limited by available variation or kept in check by selection?
+  How fast will a population explore the space of equivalent networks?
+
+  The amount and structure of 
+  this standing variation is established over long time scales
+  by many factors, including
+  mutation-selection balance, 
+  shifts in the phenotypic optimum,
+  and/or spatial variation in the optimum \citep{sources_of_standing_var}.
+  Quantitative genetics models of mutation-selection balance 
+  predict precise levels and structure of standing variation \citep{kimura_mutsel,lande_mutsel},
+  but it is unclear how well these predictions match reality \citep{johnson_barton}
+  and how much they are expected to change over time \citep{arnold_changing_G}.
+  However, recent empirical work allows us to estimate at least the rough magnitude of variation.
+  XXX FILL IN DETAILS: 
+  Mutational scans find this distribution of binding site change per nucleotide change in promoters.
+  Intraspecific variation is often in the range of 0.1\%--1\%.
+  This suggests typical regulatory binding strengths within a population might vary by up to roughly X\%.
+  But, does it?
+  \citet{maria_and_sergey} (others?) found evidence that large-effect regulatory mutations are weakly selected against in Drosophila,
+  which suggests that (a) the observed range of variation is somewhat less than this,
+  and (b) the strength of selection on phenotype is sufficient to weakly constrain this variation.
+
+  This suggests that within a population, transcription factor binding strengths 
+  -- the entries of $A$ --
+  vary by around X\%, at least for networks whose function is strongly constrained.
+  Subsequent generations for the most part resample from this diversity,
+  so in a population of effective size $N_e$,
+  simply by the variance of the mean of a random sample,
+  the population mean binding strengths
+  will move a few multiples of $X/\sqrt{N_e}\%$ per generation \citep{lande_drift}.
+  (This could be taken as a definition of $N_e$.)
+  Selection will tend to push this mean towards the optimal set of networks \citep{ou_process},
+  but mean movement parallel to the optimal set (that leaves the phenotype invariant) is unconstrained
+  unless recombination load is substantial \citep{recomb_load}.
+  The action of genetic drift is also strongly determined by covariance between 
+  standing genetic variation in different regulatory coefficients -- known as the $G$ matrix \citep{G_matrix},
+  covariance which may arise due to functional constraints and/or statistical linkage.
+  There may well be functional constraints -- but these are not sufficiently well-known to say anything general about.
+  Linkage will almost certainly lead to covariance
+  if the variation is due to \textit{cis}-regulatory variants,
+  in which case the genetic basis of each \emph{row} of $A$ likely lies within a few kilobases of tightly linked sequence,
+  across which a population may carry only a few common haplotypes.
+  However, covariance due to transiently assembled haplotypes is not expected to be stable over long periods of time --
+  a common \textit{cis}-regulatory haplotype of transcription factor $k$ with particularly strong binding to both $i$ and $j$
+  (leading to positive covariance between $A_{ik}$ and $A_{jk}$)
+  is no more likely to appear than one with strong binding to $i$ but particularly weak binding to $j$ (negative covariance).
+  (Such transient covariances may well increase the variance of the per-generation change in network mean, however \citep{barton_linkage}.)
+  In principle, there may be substantially less variation away from the set of optimal coefficients 
+  than there is along the set, due to the action of selection.
+  If so, this might require substantial epistatic load 
+  -- mortality or reduced fitness of a large proportion of new offspring due to recombination between somewhat incompatible alleles.
+  However, it is unclear if this is likely to occur and XXX we revisit this below.
+
+  It therefore seems reasonable to coarsely model the time evolution of population variation in network coefficents as 
+  (a) a ``cloud'' of width X about the population mean, 
+  which (b) moves as an unbiased Brownian motion through the set of network coefficients that give the optimal phenotype.
+  In fact, the population mean will not produce exactly the optimal phenotype,
+  but it will be convenient to refer to this closest point on the optimal set as ``the population mean''.
+
+  %%%%%%%%%%%%%%%%%%%%%%%%%%%
+  \paragraph{Brownian motion} on the set of equivalent networks:
+  The set $\calA$ is characterized as the solutions to the equations \eqref{eqn:equivalence},
+  and is hence an algebraic variety. 
+  In fact, the Kalman decomposition \plr{XXX above} provides us with an explicit characterization of the set.
+  Above we argued that the population mean set of coefficients $A$ was subject to genetic drift, 
+  with each entry $A_{ij}$ changing with mean square displacement $\sigma^2/\sqrt{N_e}$ per generation.
+  Since selection constrains the population mean to stay near to the set of optimal networks $\calA$,
+  to a good approximation, the population mean moves as Brownian motion in the space of matrices $\R^{n \times n}$
+  but conditioned to stay on the optimal set.
+  This Brownian motion on $\calA$ is a stochastic process driven by the Laplace-Beltrami operator on $\calA$
+  where this makes sense, with special behavior at the singular points (if any \plr{XXX}).
+  Unlike Brownian motion in flat space, this stochastic process can have \emph{bias} --
+  for instance, it is pushed away from regions of negative intrinsic curvature
+  (because there is ``less space'' in such regions).
+
+  \plr{NOT TRUE?:}
+  However, the process always has the property that
+  changes in $A$ accumulate at constant (mean squared) rate:
+  if $\calA$ is locally isomorphic to $\R^m$ around $A_0$
+  (there are $m$ degrees of freedom in the solutions to \eqref{eqn:equivalence}) then
+  \begin{aligned}
+      \E\left[ \|A_t - A_0\|^2 \right] = m \sigma^2 t + O(t^2) .
+  \end{aligned}
 
@@
       The fitness of an organism can be computed by comparing its impulse response with the optimal response,
+      \plr{this is either zero or infinite.  apply a weighting function (see appendix)?}
       \begin{align}
@@
 
+  We have offered a novel interpretation of system identification: to see it as an evolutionarily neutral manifold, and not simply a computational nuissance.
+We have demonstrated a method to analytically determine the set of all phenotypically invarint gene networks; by a simple change of coordinates in the minimal configuration, or more generally by applying the Kalman decomposition in higher dimensions.
+Further, we emphasize that evolution proceeds through this high dimensional space as stochastic coordinate transformation, constrained by sexual reproduction and selection.
+This set is explored over evolutionary time when phenotype is conserved, and can lead to a diverse set of consequences, including the accumulation of Dobzhansky-Muller incompatibilities.
+We emphasize that these incompatibilities are a consequence of recombining different, yet functionally equivalent, mechanisms.
+We also suggest that gene networks may not always use their components parsimoniously as network size tends to ratchet up in the absence of strong selection against extra parts.
+Although unexplored presently, this phenomena may lead to insights on evolvability and developmental innovation.
+Lastly, we show that hybrid gene networks break down as function of genetic distance, and may, in part, explain broad patterns of reproductive isolation among diverse phyla \citep{roux2016shedding}.
+
   As Richard Levins opined, models in population biology face a tradeoff among precision, realism, and generality \citep{levins1966strategy}.
@@
 
-  We have offered a novel interpretation of system identification: to see it as an evolutionarily neutral manifold, and not simply a computational nuissance.
-We have demonstrated a method to analytically determine the set of all phenotypically invarint gene networks; by a simple change of coordinates in the minimal configuration, or more generally by applying the Kalman decomposition in higher dimensions.
-Further, we emphasize that evolution proceeds through this high dimensional space as stochastic coordinate transformation, constrained by sexual reproduction and selection.
-This set is explored over evolutionary time when phenotype is conserved, and can lead to a diverse set of consequences, including the accumulation of Dobzhansky-Muller incompatibilities.
-We emphasize that these incompatibilities are a consequence of recombining different, yet functionally equivalent, mechanisms.
-We also suggest that gene networks may not always use their components parsimoniously as network size tends to ratchet up in the absence of strong selection against extra parts.
-Although unexplored presently, this phenomena may lead to insights on evolvability and developmental innovation.
-Lastly, we show that hybrid gene networks break down as function of genetic distance, and may, in part, explain broad patterns of reproductive isolation among diverse phyla \citep{roux2016shedding}.
-
       \tiny
@@
 \end{multicols}
+
+\normalsize
+\appendix
+
+\section{Differentiating the fitness function}
+
+\plr{In case this is useful...}
+
+Suppose that $\rho(t) \ge 0$ is a weighting function on $[0,\infty)$
+so that fitness is a function of $L^2(\rho)$ distance of the impulse response from optimal:
+\begin{equation}
+    \begin{aligned}
+        D(A) 
+        &:= 
+        \int_0^\infty \rho(t) \left| C e^{At} B - C e^{A_0 t} B \right|^2 dt \\
+        &= 
+        \int_0^\infty \rho(t) \left| C \left( e^{At} - e^{A_0 t} \right) B \right|^2 dt \\
+        &= 
+        \int_0^\infty \rho(t) C \left( e^{At} - e^{A_0 t} \right) B B^T \left( e^{At} - e^{A_0 t} \right)^T C^T dt
+    \end{aligned}
+\end{equation}
+How does this change with $A$?
+Since
+\begin{equation}
+  \begin{aligned}
+      \frac{d}{du} e^{(A+uZ)t} \vert_{u=0}
+      &=
+      \int_0^t e^{As} Z e^{A(t-s)} ds, 
+  \end{aligned}
+\end{equation}
+we have that
+\begin{equation}
+  \begin{aligned}
+      \frac{d}{du} D(A+uZ)\vert_{u=0}
+      &=
+        2 \int_0^\infty \rho(t) C \left( \int_0^t e^{As} Z e^{A(t-s)} ds \right) B B^T \left( e^{At} - e^{A_0 t} \right)^T C^T dt \\
+      &=
+        2 \int_0^\infty \int_0^t \rho(t) C e^{As} Z e^{A(t-s)} B B^T \left( e^{At} - e^{A_0 t} \right)^T C^T ds dt .
+  \end{aligned}
+\end{equation}
+
 \end{document}

### 2017-06-10 handnotes.tex
more notes on computing the gradient

@@
 Suppose that $\rho(t) \ge 0$ is a weighting function on $[0,\infty)$
-so that fitness is a function of $L^2(\rho)$ distance of the impulse response from optimal:
+so that fitness is a function of $L^2(\rho)$ distance of the impulse response from optimal.
+With $A_0$ a representative of the optimal set:
 \begin{equation}
@@
         D(A) 
+        &:= 
+        \int_0^\infty \rho(t) \left| h_A(t) - h_{A_0}(t) \right|^2 dt \\
         &:= 
@@
       &=
-        2 \int_0^\infty \int_0^t \rho(t) C e^{As} Z e^{A(t-s)} B B^T \left( e^{At} - e^{A_0 t} \right)^T C^T ds dt .
+        2 \int_0^\infty \rho(t) C \left( \int_0^t e^{As} Z e^{A(t-s)} ds \right) B \left( h_A(t) - h_{A_0}(t) \right)^T dt 
+  \end{aligned}
+\end{equation}
+
+We could use this to compute the gradient of $D$.
+To do so, it'd be nice to have a way to compute the inner integral above.
+Suppose that we can diagonalize $A = U \Lambda U^{-1}$.
+Then
+\begin{equation} \label{eqn:exp_deriv}
+  \begin{aligned}
+      \int_0^t e^{As} Z e^{A(t-s)} ds 
+      &=
+      \int_0^t U e^{\Lambda s} U^{-1} Z U e^{\Lambda (t-s)} U^{-1} ds 
+  \end{aligned}
+\end{equation}
+Now, notice that
+\begin{equation}
+  \begin{aligned}
+      \int_0^t e^{s \lambda_i} e^{(t-s) \lambda_j} ds
+      &=
+      \frac{ e^{t \lambda_i} - e^{t \lambda_j} }{ \lambda_i - \lambda_j } .
+  \end{aligned}
+\end{equation}
+Therefore, 
+defining
+\begin{equation}
+    X_{ij}(t) = \left( U^{-1} Z U \right)_{ij}
+      \frac{ e^{t \lambda_i} - e^{t \lambda_j} }{ \lambda_i - \lambda_j } 
+\end{equation}
+moving the $U$ and $U^{-1}$ outside the integral and integrating we get that
+\begin{equation}
+  \begin{aligned}
+      \int_0^t e^{As} Z e^{A(t-s)} ds 
+      &=
+      U X(t) U^{-1} .
   \end{aligned}

### 2017-06-12 examples/oscillator.Rmd
oscillator example begun; more appendix notes

--- 2017-05-09 examples/two-diml-minimal.Rmd
@@
 ---
-title: "Brownian motion on a 2D minimal system"
+title: "Brownian motion on an oscillating system"
 author: "Peter Ralph and Josh Shiffman"
@@
 ```{r setup, include=FALSE}
+library(expm)
+library(knitr)
+fig.dim <- 5
+knitr::opts_chunk$set(fig.height=fig.dim,fig.width=2*fig.dim,fig.align='center')
 ```
@@
 $$\begin{aligned}
-    B^{T} = \begin{bmatrix} 1 & 1 \end{bmatrix} \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix} .
+    B^{T} = \begin{bmatrix} 1 & 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix} 
 \end{aligned}$$
-and $A=A(\tau)$ depends on a free parameter,
+and
 $$\begin{aligned}
-    A_0 &= \begin{bmatrix} \lambda_{1} & \lambda_{2} - \gamma \\ 0 & \lambda_{2} \end{bmatrix} \\
-    V_\tau &= \begin{bmatrix} 1 & 0 \\ \tau & 1-\tau \end{bmatrix} \\
-    A_\tau &= V_\tau A_0 V^{-1}_\tau
+    A = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} .
+\end{aligned}$$
+This system has transfer function
+$$\begin{aligned}
+    h(t) &= C e^{tA} B \\
+        &= sin(t) + cos(t) .
+\end{aligned}$$
+The set of equivalent $A$ are given by
+$\{A(\tau): \tau \neq 1 \in \R\}$, where
+$$\begin{aligned}
+    V(\tau) 
+    &= \begin{bmatrix} 1 & 0 \\ \tau & 1-\tau \end{bmatrix} \\
+    A(\tau) 
+    &= V(\tau) A V(\tau)^{-1} \\
+    % &= \begin{bmatrix} 0 & 1 \\ \tau-1 & \tau \end{bmatrix} 
+    %    \begin{bmatrix} 1 & 0 \\ -\tau/(1-\tau) & 1/(1-\tau) \end{bmatrix}  \\
+    &= \begin{bmatrix} -\tau/(1-\tau) & 1/(1-\tau) \\ -(\tau^2 + (1-\tau)^2)/(1-\tau) & \tau/(1-\tau) \end{bmatrix} 
 \end{aligned}$$
 These systems are input-output equivalent (although may have different noise properties).
+
+```{r system_setup}
+A0 <- matrix(c(0,-1,1,0), nrow=2)
+B <- matrix(c(1,1), ncol=1)
+C <- matrix(c(1,0), nrow=1)
+V <- function (tau) { matrix(c(1,tau,0,1-tau), nrow=2) }
+Vinv <- function (tau) { matrix(c(1,-tau/(1-tau),0,1/(1-tau)), nrow=2) }
+At <- function (tau) { V(tau) %*% A0 %*% Vinv(tau) }
+A <- function (tau) { 
+    matrix(c(-tau/(1-tau), -(tau^2+(1-tau)^2)/(1-tau), 
+             1/(1-tau), tau/(1-tau)), nrow=2) 
+}
+h <- function (t,tau=0) { C %*% expm(t*A(tau)) %*% B }
+```
+
+Here is the impulse response function:
+```{r impulse_response_plot}
+tt <- seq(0,2*pi, length.out=100)
+plot(tt, sapply(tt, h), type='l',
+     xlab='t', ylab=expression(h(t)), main='transfer function')
+lines(tt, sin(tt)+cos(tt), col='red', lty=2)
+```
 

### 2017-06-12 handnotes.tex
oscillator example begun; more appendix notes

@@
 \newcommand{\calA}{\mathcal{A}}
+\newcommand{\calH}{\mathcal{H}}
 \newcommand{\diag}{\text{diag}}
@@
 
+\section{Genetic drift with a multivariate trait}
+\label{ss:quant_gen}
+
+For completeness, we provide a brief argument of how the population mean
+moves under genetic drift
+with a quantitative genetics model.
+
+\paragraph{Completing the square}
+First note that 
+\begin{align*}
+    (x-y)^T A (x-y)
+    &=
+    x^T A \left( x - 2y) + y^T A y ,
+\end{align*}
+and so
+\begin{align*}
+    (x-y)^T A (x-y) + x^T B x
+    &=
+    x^T (A + B) \left( x - 2 (A + B)^{-1} A y \right) + y^T A y \\
+    &=
+    \left( x - (A + B)^{-1} A y \right)^T
+    (A + B)
+    \left( x - (A + B)^{-1} A y \right)
+    + \text{(terms that don't depend on $x$)} .
+\end{align*}
+Therefore, if $f(x;\Sigma,y)$ is the density of a Gaussian with mean $y$ and covariance matrix $\Sigma$
+then substituting $A=\Sigma^{-1}$ and $B=U^{-1}$ above,
+\begin{align*}
+    \frac{ f(x;\Sigma,y) f(x;U,0) }{\int_x f(z;\Sigma,y) f(z;U,0) dz}
+    &=
+    f(x; (\Sigma^{-1} + U^{-1})^{-1}, (\Sigma^{-1}+U^{-1})^{-1} \Sigma^{-1} y) .
+\end{align*}
+
+Now suppose that the population is distributed in genotype space
+as a Gaussian with covariance matrix $\Sigma$ and mean $y$.
+Selection has the effect of multiplying this density by the fitness function and renormalizing,
+so that if expected fitness of $x$ is proportional to $f(x;U,z)$
+then the above argument shows that the next generation will be sampled from a Gaussian distribution
+with covariance matrix $(\Sigma^{-1} + U^{-1})^{-1}$ 
+and mean $z + (\Sigma^{-1}+U^{-1})^{-1} \Sigma^{-1} (y-z)$
+Taking a sample of size $N$ to construct the next generation 
+will produce something close to this but with a slightly (stochastically) deviating mean.
+The next generation's mean is drawn from a Gaussian distribution with mean
+with covariance matrix $(\Sigma^{-1} + U^{-1})^{-1}/N$ 
+and mean $z + (\Sigma^{-1}+U^{-1})^{-1} \Sigma^{-1} (y-z)$.
+
+Roughly, what is this doing?
+Suppose that the population mean differs from the optimum by $\epsilon$,
+that $\Sigma = \sigma^2 I$ and $U = s I$.
+Then the population mean gets closer to the optimum on average, moving to
+$\epsilon/(1 + \sigma^2/s)$
+and adds noise of size $\sqrt{s} \sigma/\sqrt{N \sigma^2 + N s}$.
+At equilibrium, these two movements will be of the same order,
+so that $\epsilon$ is of order $(\sigma/\sqrt{N}) \sqrt{1+\sigma^2/s}$.
+
 \section{Differentiating the fitness function}
@@
 \end{equation}
+and, by differentiating this and supposing that $A$ is on the optimal set,
+i.e., $h_A(t)=h_{A_0}(t)$, (so wolog $A=A_0$):
+\begin{equation}
+  \begin{aligned}
+      \calH(Y,Z) &:= \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0+uY+vZ)\vert_{u=v=0} \\
+      &=
+        \int_0^\infty \rho(t) C 
+        \left( \int_0^t e^{A_0 s} Y e^{A_0 (t-s)} ds \right) 
+        B B^T 
+        \left( \int_0^t e^{A_0 s} Z e^{A_0 (t-s)} ds \right)^T
+        C^T dt  .
+  \end{aligned}
+\end{equation}
+Here $\calH$ is the quadratic form underlying the Hamiltonian.
+By defining $\Delta_{ij}$ to be the matrix with a 1 in the $(i,j)$th slot
+and 0 elsewhere,
+the coefficients of the quadratic form is
+\begin{equation}
+    \begin{aligned}
+        H_{ij, k\ell}(A)
+        &:=
+        \calH(\Delta_{ij}, \Delta_{k\ell}) .
+    \end{aligned}
+\end{equation}
 
-We could use this to compute the gradient of $D$.
+We could use this to compute the gradient of $D$,
+or to get the quadratic approximation to $D$ near the optimal set.
 To do so, it'd be nice to have a way to compute the inner integral above.
@@
 \begin{equation}
-    X_{ij}(t) = \left( U^{-1} Z U \right)_{ij}
+    X_{ij}(t,Z) = \left( U^{-1} Z U \right)_{ij}
       \frac{ e^{t \lambda_i} - e^{t \lambda_j} }{ \lambda_i - \lambda_j } 
@@
       &=
-      U X(t) U^{-1} .
+      U X(t,Z) U^{-1} .
   \end{aligned}
@@
 
+Following on from above, we see that if $Z=\Delta_{k \ell}$, then
+\begin{equation}
+  \begin{aligned}
+      X_{ij}^{k\ell}(t) = 
+      \frac{ e^{t \lambda_i} - e^{t \lambda_j} }{ \lambda_i - \lambda_j } 
+      (U^{-1})_{\cdot k} U_{\ell \cdot},
+  \end{aligned}
+\end{equation}
+where $U_{k \cdot}$ is the $k$th row of $U$,
+and so
+\begin{equation}
+    \begin{aligned}
+        H_{ij, k\ell}(A)
+        &=
+        \int_0^\infty
+            \rho(t) C U X^{ij}(t) U^{-1} B B^T (U^{-1})^T X^{k\ell}(t)^T U^T C^T
+        dt .
+    \end{aligned}
+\end{equation}
+This implies that
+\begin{equation}
+    \begin{aligned}
+        D(A_0+\epsilon Z)
+        &\approx \epsilon^2\sum_{ijk\ell} H^{ij,k\ell} Z_{ij} Z_{k\ell} 
+    \end{aligned}
+\end{equation}
+and so
+\begin{equation}
+    \begin{aligned}
+        D(A_0+\epsilon Z)
+        &\approx \epsilon^2\sum_{ijk\ell} H^{ij,k\ell} Z_{ij} Z_{k\ell} 
+    \end{aligned}
+\end{equation}
+
+By section \ref{ss:quant_gen},
+if we set $\Sigma=\sigma^2 I$ and $U=H$,
+then a population at $A_0+Z$ experiences a restoring force of strength
+$(I + \sigma^2 H^{-1})^{-1} Z$ (treating $Z$ as a vector and $H$ as an operator on these).
+If $\sigma^2$ is small compared to $H^{-1}$
+then this is approximately $-\sigma^2 H^{-1} Z$.
+
 \end{document}

### 2017-06-12 promoterdata.tex
added notes on empirical promoter variation and evolution

\documentclass[9 pt]{article}
\usepackage{amsmath, amssymb, color, xcolor, amsthm}
\usepackage{graphicx, wrapfig, float, caption, dsfont, bbm}
\usepackage{fullpage}
\usepackage[backref=page, hidelinks, colorlinks=true, citecolor=blue!60!black!100]{hyperref}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, shapes}
\usepackage{caption, subcaption}
\usepackage{natbib} % gives us \citet: Author (year) and \citep: (Author; year)
\usepackage{authblk}

\usepackage{multicol}

\newcommand{\plr}[1]{{\color{blue}\it #1}}
\newcommand{\jss}[1]{{\color{olive}\it #1}}
% \newcommand{\ddt}{\frac{d}{dt}}
\newcommand{\ddt}{\dot}
\newcommand{\ro}{{ro}}
\newcommand{\nro}{{\bar{r}o}}
\newcommand{\rno}{{r\bar{o}}}
\newcommand{\nrno}{{\bar{r}\bar{o}}}
\newcommand{\reachable}{\mathcal{R}}
\newcommand{\unobservable}{\bar{\mathcal{O}}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\pda}{\frac{\partial}{\partial A_{ij}}}
\newcommand{\ind}{\mathds{1}}

\newcommand{\calA}{\mathcal{A}}
\newcommand{\diag}{\text{diag}}
\newcommand{\1}{\mathbbm{1}}

\DeclareMathOperator{\spn}{span}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}

\begin{document}

  \section{Notes on Empirical \emph{Cis}-Regulatory Module Variation and Evolution}

    \begin{itemize}
      \item \citet{gertz2009analysis} study synthetic yeast promoters. The synthetic promoters contained a random combination of 3-4 TFBSs. Weak MIG1 TFBS has a 6.7-9.0 fold lower affinity than the strong TFBS. Weak sites in concert with a strong site act as a strong site too.

      \item \citet{frankel2011morphological} study the loss of thrichome phenotype in D. sechellia (vs. melanogaster). Identify 6 embryonic enhancers (5 kb) that are diverged. Examine in detail the E enhancer -- two subunits of it, each ~1 kb long, E3 and E6. Say that E3 is conserved, while the E6 region has evolved. Comparing the E6 sechellia sequence with five closely related species, all containing a high density of thrichomes (in contrast to sechellia), they identify 13 substitutions and 1 single nucleotide deletion. There was no polymorphism within D. sechellia within the E6 region (9 sequences). Estimates these substitutions occured at an accelerated rate -- 4.8 times the flanking regions.

        Introduced all sechellia substitutions into melanogaster (and the reverse!). Found that expression changed in the directions as expected in the transgenic lines -- but did not completely rescue phenotypes -- there are likely other regulatory elements involved.

        The temporal expression delay in sechellia could not be captured by a single mutation -- only a combination of the known differences. At least 5 substituitons are necessary. Each cluster of mutations (2-5) reduced trichome expression by 4.6-33.5\%.

        ``none of the D. melanogaster to D. sechellia mutations led to a significant increase in trichome number, and none of the reciprocal mutations led to a decrease in trichome number''

      \item \citet{rebeiz2009stepwise} look at adaptive melaninism in D. melanogaster. 
        Find a minimum of 5 mutations differentiate regulation of dark from light lines. These account for 8 to 40\% of the overal difference in ebony activity.

        Lots of standing variation preceding phenotype divergence: 3 of the identified causative substitutions in one region were found at high frequencies in both Ugandan light lines and in a Kenyan light population. Two of the substitutions were found in all 5 populations sampled. The third subst was found in 4 of the 5 populations. Dark specific substitutions were not observed in other lines; suggesting they were rare variants or evolved de novo.

        Estimate that standing variation in the enhancer happened first, then the fixation of the de novo mutations.

        ``We have shown that the adaptive evolution of melanism in a Ugandan population ofD. melanogaster occurred through multiple, stepwise substitutions in one enhancer of the ebony locus. We suggest that this genetic path of enhancer evolution with multiple substitutions of varying effect sizes, which originate from both standing variation and new mutations and combine to create an allele of large effect, may be a general feature of enhancer evolution in populations. This view is consistent with studies that have demonstrated that substitutions at multi le sites within enhancers are responsible for evolutionary changes in gene expression''

      \item \citet{verlaan2009targeted} study expression variation within humans. 
        cis-regulatory variation is known to account for 2-6\% expression variation in human blood-derived primary cells.

      \item \citet{yan2002allelic} studied CEPH families (human). Fraction of individual variation in heterozygous individuals was 3 to 30\% (differentially expressed alleles). Expressed transcript ratio varied from 1.3 : 1.0 (FBN1) to 4.3 : 1.0 (p73). Many of these expression differences were Mendellian.

      \item \citep{lappalainen2013transcriptome} mapped cis-QTLs to transcriptome variation in human samples from CEPH, FIN, GBR, TSI and the 1000 genome project. Population differences explain 3\% of the variation. Identified 263-4,379 genes with differential expression (or transcript ratios) between population pairs.

        See figure 1b.

    \end{itemize}

\bibliographystyle{plainnat}
\bibliography{krefs}

\end{document}

### 2017-06-13 examples/oscillator.Rmd
.

@@
 
-# Brownian motion
+We can define the dynamics in two ways: extrinsically or intrinsically.
+Extrinsically,
+we look at the dynamics of a population mean in $\R^{2\times 2}$,
+thinking about $A$ as evolving.
+Intrinsically, we move about only on the set of equivalent $A$,
+letting $\tau$ evolve.
+As the population size goes to infinity,
+the extrinsic dynamics should approach the intrinsic dynamics.
 
-Now suppose that $A$ is changing due to mutation, 
-and that it takes the value $A(t)$ after $t$ units of evolutionary time.
-We know that $A(t)$ must equal $A_\tau$ for some $\tau$, 
-and hence we can define $A(t) = A_{W(t)}$, 
-where $W(t)$ is a random process on the free parameter;
-but what are the appropriate dynamics for $W$?
-Suppose that genetic drift acts uniformly on the entries of $A$, 
-so that ...
+# Extrinsic dynamics
 
-# Toy example
-
-As a toy example, suppose that we want to describe a process through $\R^2$
-on the set
-$$\begin{aligned}
-    x^2 + y^2 = 1,
-\end{aligned}$$
-i.e., the unit circle.
-This is parameterized by
-$$\begin{aligned}
-    x(\theta) &= \cos(\theta) \\
-    y(\theta) &= \sin(\theta) .
-\end{aligned}$$
-
-The dynamics we want are limits of the following discrete dynamics.
-Let $\epsilon$ denote the step size,
-and suppose that at rate $\eta$ steps of size $\epsilon$ 
-in a uniformly random direction $\phi$ are proposed;
-but the probability that such a step is accepted is Kimura's expression:
-$$\begin{aligned}
-    q(x',y';x,y) = \frac{1 - e^{-2 (f(x',y')-f(x,y))}}{1 - e^{-2 N_e (f(x',y')-f(x,y))}}
-\end{aligned}$$
-where $(x',y')$ is the new (proposed) location, and
-$$\begin{aligned}
-    f(x,y) = (\sqrt{x^2+y^2}-1)^2
-\end{aligned}$$
-is the fitness function.
-Note that $q(x,y;x,y) = 1$ 
-and that 
-$$\begin{aligned}
-    \partial_u q(u,v;x,y) 
-    &= 
-        ( \partial_u f(u,v) ) 
-        \left( \partial_z \frac{1 - e^{-2 (z-f(x,y))}}{1 - e^{-2 N_e (z-f(x,y))}} \vert_{z=f(x',y')} \right) \\
-    &= 
-        \left( \frac{2 x (\sqrt{x^2+y^2}-1)}{\sqrt{x^2+y^2}} \right) 
-        \left( \frac{ 2 e^{-2 (f(x',y')-f(x,y))} (1 - e^{-2 N_e (f(x',y')-f(x,y))}) - 2 N_e e^{-2 (f(x',y')-f(x,y))} }{(1 - e^{-2 N_e (f(x',y')-f(x,y))})^2} \right)
-\end{aligned}$$
-and so that
-$$\begin{aligned}
-    \partial_u q(u,v;x,y) \vert_{u=x, v=y}
-    &= 
-        \left( \frac{2 x (\sqrt{x^2+y^2}-1)}{\sqrt{x^2+y^2}} \right) 
-        \left( \frac{ - 2 N_e  }{(1 - e^{-2 N_e (f(x',y')-f(x,y))})^2} \right)
-\end{aligned}$$
-
-This has the result of multiplying jump rates by the acceptance probabilities;
-so equivalently, if the process is at $(x,y)$,
-then it jumps to 
-$$\begin{aligned}
-    (x',y') = (x+\epsilon \cos \phi, y + \epsilon \sin \phi)
-\end{aligned}$$
-at rate
-$$\begin{aligned}
-    \eta q(x',y';x,y) 
-    &\approx
-
-\end{aligned}$$
+Let $N$ denote the population size,
+and suppose that at any time the population is spread about from the mean $A$
+as a Gaussian with variance $\sigma^2$, independently in each entry of $A$.
+Then we expect that the population mean takes steps of size $\sigma/\sqrt{N}$
+with a restoring force towards the set of optimal $A$.

### 2017-06-13 examples/oscillator.Rmd
oscillator numerical example

@@
 library(knitr)
+library(parallel)
 fig.dim <- 5
@@
 At <- function (tau) { V(tau) %*% A0 %*% Vinv(tau) }
-A <- function (tau) { 
+A_fn <- function (tau) { 
     matrix(c(-tau/(1-tau), -(tau^2+(1-tau)^2)/(1-tau), 
@@
 }
-h <- function (t,tau=0) { C %*% expm(t*A(tau)) %*% B }
+h <- function (t,tau=0,A=A_fn(tau)) { sapply(t, function (tt) C %*% expm(tt*A) %*% B) }
 ```
@@
 plot(tt, sapply(tt, h), type='l',
-     xlab='t', ylab=expression(h(t)), main='transfer function')
+     xlab='t', ylab=expression(h(t)), main='impulse response function')
 lines(tt, sin(tt)+cos(tt), col='red', lty=2)
@@
 with a restoring force towards the set of optimal $A$.
+
+# Fitness of hybrids
+
+First we need the fitness function.
+This will be $\exp(-D^2/2\sigma^2)$, where $D$ is the distance to optimal:
+$$\begin{aligned}
+  D(A) := \int_0^\infty e^{-t/4\pi} \| h(t,A) - h_0(t) \| dt ,
+\end{aligned}
+where $h_0(t)=\sin(t)+\cos(t)$ is the optimal impulse response function.
+
+```{r distance_fun}
+D <- function (A, upper=100, ...) {
+    f <- function (t) {
+        exp(-t/(4*pi)) * ( h(t,A=A) - cos(t) - sin(t) )^2
+    }
+    integrate(f, lower=0, upper=upper, ...)$value
+}
+```
+
+To check this works, the distance of $A(2)$ to optimal should be zero:
+```{r check_D}
+D(A(2))
+```
+
+We also need to be able to produce F1s and F2s:
+```{r offspring}
+F1 <- function (tau1,tau2) { (A_fn(tau1)+A_fn(tau2))/2 }
+
+gamete <- function (X,Y,rows) { 
+    out <- X
+    out[rows,] <- Y[rows,]
+    return(out)
+}
+
+F2 <- function (tau1, tau2, rows1, rows2) {
+    (gamete(A_fn(tau1), A_fn(tau2), rows1) + gamete(A_fn(tau1), A_fn(tau2), rows2))/2
+}
+
+random_F2 <- function (tau1, tau2) {
+    rows1 <- (rbinom(2,size=1,prob=1/2)>0)
+    rows2 <- (rbinom(2,size=1,prob=1/2)>0)
+
+    F2(tau1, tau2, rows1, rows2)
+}
+
+all_F2s <- function (tau1, tau2) {
+    all_rows <- as.matrix(do.call( expand.grid, list(c(TRUE,FALSE))[rep(1,4)] ))
+    lapply(1:nrow(all_rows), function (k) {
+                x <- all_rows[k,]
+                F2(tau1, tau2, x[1:2], x[3:4])
+            })
+}
+
+```
+
+Here are the phenotypes for $\tau=2$, $\tau=2.01$, and the hybrids:
+```{r phenotypes}
+dtau <- .01
+tt <- seq(0,6*2*pi, length.out=500)
+hh <- sapply( list(A_fn(2), A_fn(2+dtau), F1(2, 2+dtau)), function (A) { sapply(tt, h, A=A) } )
+matplot(tt, hh, type='l', main=sprintf("Parental phenotypes and F1s at dtau=%s", dtau), lty=c(1,1,2), col=1:3, xlab="time", ylab="phenotype")
+legend("topright", lty=c(1,1,2), col=1:3,
+        legend=c(expression(tau==2), expression(tau==2+dtau), "F1"))
+# now the F2s
+hh <- sapply( all_F2s(2, 2+dtau), function (A) { sapply(tt, h, A=A) } )
+matplot(tt, hh, type='l', main="F2s", xlab="time", ylab="phenotype")
+```
+And, the same thing for $\tau=2.1$:
+```{r phenotypes2}
+dtau <- .1
+tt <- seq(0,6*2*pi, length.out=500)
+hh <- sapply( list(A_fn(2), A_fn(2+dtau), F1(2, 2+dtau)), function (A) { sapply(tt, h, A=A) } )
+matplot(tt, hh, type='l', main=sprintf("Parental phenotypes and F1s at dtau=%s", dtau), lty=c(1,1,2), col=1:3, xlab="time", ylab="phenotype")
+legend("topright", lty=c(1,1,2), col=1:3,
+        legend=c(expression(tau==2), expression(tau==2+dtau), "F1"))
+# now the F2s
+hh <- sapply( all_F2s(2, 2+dtau), function (A) { sapply(tt, h, A=A) } )
+matplot(tt, hh, type='l', main="F2s", xlab="time", ylab="phenotype")
+```
+
+These F2s have distances of $D=$:
+```{r F2_dists}
+sort(round(sapply(all_F2s(2, 2+dtau), function (x) D(A=x, upper=100, subdivisions=500)),2))
+```
+
+Now let's look at $D$ as a function of $d\tau$:
+```{r d_fn, cache=TRUE}
+
+dvec <- seq(0, .02, length.out=20)
+D_F1 <- sapply(dvec, function (dtau) {
+            D(A=F1(tau1=2, tau2=2+dtau))
+        } )
+
+D_F2 <- do.call(cbind, mclapply(dvec, function (dtau) {
+            F2s <- all_F2s(2, 2+dtau)
+            sapply(F2s, D, upper=100)
+        }, mc.cores=detectCores() ) )
+```
+
+```{r plot_d_fn}
+layout(t(1:2))
+plot( dvec, D_F1, type='l', xlab="d tau", ylab="phenotype distance (D)", main="distance of F1s")
+plot( dvec, D_F1^(1/4), type='l', main="one-fourth power", xlab="d tau", ylab="D^(1/4)")
+
+plot( dvec, D_F1, type='l', ylim=range(D_F2, D_F1),
+        xlab="dtau", ylab='phenotype distance (D)',
+        main="all hybrids")
+lines(dvec, colMeans(D_F2), lty=2)
+matlines(dvec, t(D_F2), lty=3)
+legend("topleft", lty=c(1,2,3), legend=c("F1", "mean F2", "F2s"))
+
+plot( dvec, sqrt(D_F1), type='l', ylim=sqrt(range(D_F2, D_F1)),
+        xlab="dtau", ylab='phenotype distance (D)',
+        main="sqrt distance")
+lines(dvec, sqrt(colMeans(D_F2)), lty=2)
+matlines(dvec, sqrt(t(D_F2)), lty=3)
+
+```

### 2017-06-13 examples/upswitch.Rmd
upswitch

--- 2017-06-13 examples/oscillator.Rmd
@@
 ---
-title: "Brownian motion on an oscillating system"
+title: "Brownian motion on an upswitch"
 author: "Peter Ralph and Josh Shiffman"
@@
 $$\begin{aligned}
-    B^{T} = \begin{bmatrix} 1 & 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix} 
+    B^{T} = \begin{bmatrix} 1 & 0 \end{bmatrix}, \qquad C = \begin{bmatrix} 0 & 1 \end{bmatrix} 
 \end{aligned}$$
@@
 $$\begin{aligned}
-    A = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} .
+    A = \begin{bmatrix} -1 & 0 \\ 1 & 0 \end{bmatrix} .
 \end{aligned}$$
-This system has transfer function
+This system has impulse response function
 $$\begin{aligned}
     h(t) &= C e^{tA} B \\
-        &= sin(t) + cos(t) .
+        &= 1 - e^{-t} .
 \end{aligned}$$
@@
     V(\tau) 
-    &= \begin{bmatrix} 1 & 0 \\ \tau & 1-\tau \end{bmatrix} \\
+    &= \begin{bmatrix} 1 & \tau \\ 0 & 1 \end{bmatrix} \\
     A(\tau) 
-    &= V(\tau) A V(\tau)^{-1} \\
-    % &= \begin{bmatrix} 0 & 1 \\ \tau-1 & \tau \end{bmatrix} 
-    %    \begin{bmatrix} 1 & 0 \\ -\tau/(1-\tau) & 1/(1-\tau) \end{bmatrix}  \\
-    &= \begin{bmatrix} -\tau/(1-\tau) & 1/(1-\tau) \\ -(\tau^2 + (1-\tau)^2)/(1-\tau) & \tau/(1-\tau) \end{bmatrix} 
+    &= V(\tau) A V(\tau)^{-1} 
 \end{aligned}$$
@@
 ```{r system_setup}
-A0 <- matrix(c(0,-1,1,0), nrow=2)
-B <- matrix(c(1,1), ncol=1)
-C <- matrix(c(1,0), nrow=1)
-V <- function (tau) { matrix(c(1,tau,0,1-tau), nrow=2) }
-Vinv <- function (tau) { matrix(c(1,-tau/(1-tau),0,1/(1-tau)), nrow=2) }
-At <- function (tau) { V(tau) %*% A0 %*% Vinv(tau) }
-A_fn <- function (tau) { 
-    matrix(c(-tau/(1-tau), -(tau^2+(1-tau)^2)/(1-tau), 
-             1/(1-tau), tau/(1-tau)), nrow=2) 
-}
+A0 <- matrix(c(-1,1,0,0), nrow=2)
+B <- matrix(c(1,0), ncol=1)
+C <- matrix(c(0,1), nrow=1)
+V <- function (tau) { matrix(c(1,0,tau,1), nrow=2) }
+Vinv <- function (tau) { matrix(c(1,0,-tau,1), nrow=2) }
+A_fn <- function (tau) { V(tau) %*% A0 %*% Vinv(tau) }
 h <- function (t,tau=0,A=A_fn(tau)) { sapply(t, function (tt) C %*% expm(tt*A) %*% B) }
+optimal_h <- function (t) { 1 - exp(-t) }
 ```
@@
 tt <- seq(0,2*pi, length.out=100)
-plot(tt, sapply(tt, h), type='l',
+plot(tt, h(tt), type='l',
      xlab='t', ylab=expression(h(t)), main='impulse response function')
-lines(tt, sin(tt)+cos(tt), col='red', lty=2)
+lines(tt, h(tt, tau=0.1), col='blue', lty=3)
+lines(tt, optimal_h(tt), col='red', lty=2)
 ```
@@
 \end{aligned}
-where $h_0(t)=\sin(t)+\cos(t)$ is the optimal impulse response function.
+where $h_0(t)$ is the optimal impulse response function.
 
 ```{r distance_fun}
-D <- function (A, upper=100, ...) {
-    f <- function (t) {
-        exp(-t/(4*pi)) * ( h(t,A=A) - cos(t) - sin(t) )^2
+Df <- function (A,t) {
+    exp(-t/(4*pi)) * ( h(t,A=A) - optimal_h(t) )^2
 }
+D <- function (A, upper=10, ...) {
+    f <- function (t) { Df(A,t) }
     integrate(f, lower=0, upper=upper, ...)$value
@@
 ```{r check_D}
-D(A(2))
+D(A_fn(2))
 ```
@@
 hh <- sapply( list(A_fn(2), A_fn(2+dtau), F1(2, 2+dtau)), function (A) { sapply(tt, h, A=A) } )
-matplot(tt, hh, type='l', main=sprintf("Parental phenotypes and F1s at dtau=%s", dtau), lty=c(1,1,2), col=1:3, xlab="time", ylab="phenotype")
+matplot(tt, hh, type='l', main=sprintf("Parental phenotypes and F1s at dtau=%s", dtau), lty=c(1,1,2), col=1:3, xlab="time", ylab="phenotype", log='y')
 legend("topright", lty=c(1,1,2), col=1:3,
@@
 hh <- sapply( all_F2s(2, 2+dtau), function (A) { sapply(tt, h, A=A) } )
-matplot(tt, hh, type='l', main="F2s", xlab="time", ylab="phenotype")
+matplot(tt, hh, type='l', main="F2s", xlab="time", ylab="phenotype", log='y')
 ```
@@
 Now let's look at $D$ as a function of $d\tau$:
-```{r d_fn, cache=TRUE}
+```{r d_fn}
 

### 2017-06-14 examples/oscillator.Rmd
upswitch example

@@
 knitr::opts_chunk$set(fig.height=fig.dim,fig.width=2*fig.dim,fig.align='center')
+do.pdfs <- TRUE
+pdf_copy <- function ( width=6, height=width*knitr::opts_current$get("fig.height")/knitr::opts_current$get("fig.width"), plot.id=NULL,
+                filename) {
+    if (missing(filename)) {
+        file.id <- if (is.null(plot.id)) { "" } else { paste0("_",plot.id) }
+        filename <- knitr::fig_path(paste(file.id,".pdf",sep=""))
+    }
+    cat("pdf version at:",filename)
+    dev.print( file=filename, device=pdf, width=width, height=height, pointsize=10, family="sans")
+}
 ```
@@
 \end{aligned}$$
-This system has transfer function
+This system has impulse response function
 $$\begin{aligned}
@@
 h <- function (t,tau=0,A=A_fn(tau)) { sapply(t, function (tt) C %*% expm(tt*A) %*% B) }
+optimal_h <- function (t) { sin(t)+cos(t) }
 ```
@@
 tt <- seq(0,2*pi, length.out=100)
-plot(tt, sapply(tt, h), type='l',
+plot(tt, h(tt), type='l',
      xlab='t', ylab=expression(h(t)), main='impulse response function')
-lines(tt, sin(tt)+cos(tt), col='red', lty=2)
+lines(tt, h(tt, tau=0.1), col='blue', lty=3)
+lines(tt, optimal_h(tt), col='red', lty=2)
+
+if (do.pdfs) { pdf_copy() }
 ```
@@
 ```{r distance_fun}
-D <- function (A, upper=100, ...) {
-    f <- function (t) {
-        exp(-t/(4*pi)) * ( h(t,A=A) - cos(t) - sin(t) )^2
+Df <- function (A,t) {
+    exp(-t/(4*pi)) * ( h(t,A=A) - optimal_h(t) )^2
 }
+D <- function (A, upper=10, ...) {
+    f <- function (t) { Df(A,t) }
     integrate(f, lower=0, upper=upper, ...)$value
@@
 matplot(tt, hh, type='l', main="F2s", xlab="time", ylab="phenotype")
+
+if (do.pdfs) { pdf_copy() }
 ```
@@
 
+if (do.pdfs) { pdf_copy() }
 ```

### 2017-06-14 examples/upswitch.Rmd
upswitch example

@@
 lines(tt, optimal_h(tt), col='red', lty=2)
+
+if (do.pdfs) { pdf_copy() }
 ```
@@
 matplot(tt, hh, type='l', main="F2s", xlab="time", ylab="phenotype", log='y')
+
+if (do.pdfs) { pdf_copy() }
 ```

### 2017-06-16 handnotes.tex
minor edits
added words to examples, edited wrt to plr notes, changed plots.

@@
 \documentclass[9 pt]{article}
-\usepackage{amsmath, amssymb, color, xcolor, amsthm}
+\usepackage{amsmath, amssymb, color, xcolor}
 \usepackage{graphicx, wrapfig, float, caption, dsfont, bbm}
@@
 \usepackage{caption, subcaption}
-\usepackage[super]{natbib} % gives us \citet: Author (year) and \citep: (Author; year)
+\usepackage{natbib} % gives us \citet: Author (year) and \citep: (Author; year)
 \usepackage{authblk}
@@
 
-\begin{multicols}{2}
+%\begin{multicols}{2}
 
-\section*{Abstract}
-  \textbf{
+\begin{abstract}
+ % \textbf{
 I will outline an analytical theory to study the evolution of biological systems such as gene regulatory networks, borrowing insight and tools from control engineering, systems identification, and dynamical systems theory.
@@
 This analysis provides insight into the mechanisms and parameters important for understanding developmental systems drift, network rewiring, evolvability, epistasis, and speciation, as well as the tenuous connection between network architecture and function.
-}
+%}
+\end{abstract}
 \section*{Introduction}
+
+  \jss{add stuff on Haldane's Rule and Hybrid Vigor; note: F1s diverge to the 4th power, and F2s quadratically}
 
@@
 This is especially pertinent given the staggering complexity and scope of contemporary research programs.
-This outlook necessitates the advancement of conceptual frameworks of such precision, only mathematics will suffice.
+This outlook necessitates the advancement of conceptual frameworks of such precision, only mathematics will suffice, as models allow the development of concrete numerical predictions.
 Previously it has been suggested that any idealized study of evolution is incomplete without a mathematically sufficient description of the genotype, phenotype, and transformation from one to the other \citep{Lewontin1974genetic}.%Here we outline an analytical theory to study the evolution of biological systems, borrowing insight and methods from control engineering, systems identification, and dynamical systems theory.
@@
  Movement in this direction is ongoing, as researchers have begun to study 
-the evolution of both abstract \citep{wagner1994evolution, wagner1996does,  siegal2002waddington, bergman2003evolutionary} and empirically inspired computational and mathematical models of gene regulatory networks (GRNs) \citep{mjolsness1991connectionist, jaeger2004dynamic, maria1, vitaly1, vitaly2, crombach2016gap, wotton2015quantitative, chertkova2017insilico}.
+the evolution of both abstract \citep{wagner1994evolution, wagner1996does,  siegal2002waddington, bergman2003evolutionary, draghi2015robustness} and empirically inspired computational and mathematical models of gene regulatory networks (GRNs) \citep{mjolsness1991connectionist, jaeger2004dynamic, maria1, vitaly1, vitaly2, crombach2016gap, wotton2015quantitative, chertkova2017insilico}.
 If we allow the reasonable assumption that the genotype-phenotype map can be represented as a system of differential equations, we can immediately discuss its evolution and function in a much more mechanistic, yet general, manner.
@@
 circuits and mechanical systems are often the focus, it is understood that there can be an 
-infinite number of ``realizations,'' or ways to reverse engineer the dynamics of a black box,
+infinite number of ``realizations'', or ways to reverse engineer the dynamics of a black box,
 even if all possible input and output experiments on the black box are performed \citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. In chemical
@@
 as the parameters and structures can be realized biologically. In fact, evolutionary
-biologists who study homology and analogy are very familiar with such functional symmetries; macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level \citep{true2001developmental, tsong2006evolution, hare2008sepsid, vierstra2014mouse, stergachis2014conservation, taylor2016diverse, matsui2015regulatory}.
+biologists who study convergent versus parallel evolution, homology, and analogy are very familiar with such functional symmetries; macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level \citep{true2001developmental, tsong2006evolution, hare2008sepsid, vierstra2014mouse, stergachis2014conservation, taylor2016diverse, matsui2015regulatory}.
 
   In this paper we propose a framework to study the evolution of biological systems.
+We expect wide applicability to this approach (\emph{i.e.} non-neutral evolution), however presently, we focus solely on neutral evolution.
 To begin, we focus on the evolution of an idealized population.
-We consider the evolution of a perfectly adapted, large population, evolving in a static environment for an infinite number of generations.
-Under these ideal circumstances, we expect to observe a ``conservation of phenotype,'' where the population explores the manifold of phenotypically-invariant (or symmetric) genetic and developmental architecutres.
-We would like to understand which parameters influence the distribution of a population along the manifold of phenotypically invariant genetic systems.
-Further, we can show how dispersion along this manifold contributes to speciation and evolvability.
+We consider the neutral evolution of a large population, evolving in a static environment.
+Under these circumstances we expect the population explores the manifold of phenotypically invariant (or symmetric) gene regulatory networks.
 
-\section*{The Model I: Gene Networks as Linear Systems}
+  Presently we ask two questions, (1) holding the phenotype and environment constant for evolutionary time, how much do we expect gene network organization to drift, and (2) does this contribute to speciation, primarily via the fixation of reproductive incompatibilities?
+
+\section*{Gene Networks as Linear Dynamical Systems}
 
@@
 The input can be interpreted as the environment, as initial conditions, or otherwise, depending on the biological specifics under study.
+The phenotype is a convolution of both the impulse response and the input, as allowed by our assumption of linearity.
 
@@
   \end{align}
-  where $A$ is a gene network -- a square matrix, and $B$ filters and translates the input to the system.
+  where $A$ is a gene regulatory network (although this model can be generalized to other biological networks) -- a square matrix, and $B$ filters and translates the input to the system.
 The form of $B$ determines precisely how the state of the external environment influences the internal gene network.
@@
   Generally $A$ can be any real $n \times n$ matrix, $B$ any $n \times \ell$, and $C$ any $\ell \times n$ dimensional matrix.
-However, for simplicity in exposition (and without loss of generality) we set $\ell = 1$, so that $B$ and $C$ are simply vectors of length $n$.
+Each $i$th row of $A$ describes the \emph{cis}-regulatory module for gene $i$, and each $j$th entry, the specific regulatory influence of gene $j$ on gene $i$.
 
@@
 
-  \begin{example}[Oscillating Gene Network: Circadian Rhythm]
+  \begin{example}[Oscillating Gene Network: Cell Cycle Control]\label{ex:oscillator}
+
+    Cellular division is a complicated phenomena, governed by many different processes, however it is agreed that its rhythm is partially controlled by periodic (oscillating) gene transcription \citep{orlando2008global}.
+Consider a simplified model of oscillating gene transcription.
+In the present framework periodic expression requires at minimum two interacting genes.
+
+    Suppose gene-2 up-regulates the transcription of gene-1 and that gene-1 down-regulates the transcription of gene-2 with equal magnitude (of $1$) and relative to each of their concentrations, denoted by $\kappa_{1}$ and $\kappa_{2}$.
+Furthermore, suppose that only the dynamics of gene-1 are consequential to the cell cycle (perhaps the amount of gene-1 activates another downstream gene network).
+Lastly suppose that the production of both genes is stimulated by an impulse of a molecule present immediately after division.
+
+    If the rate each of these genes is expressed is a linear function of their concentrations, the dynamics of the system are given by
     \begin{align*}
-      A = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} , \qquad B = \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix}
+      \dot{\kappa_{1}}(t) &= \kappa_{2}(t) + u(t) \\
+      \dot{\kappa_{2}}(t) &= - \kappa_{1}(t) + u(t)
     \end{align*}
+    where $\dot{\kappa}$ denotes the time derivative.
+The initial conditions $\kappa_{1}(0)$ and $\kappa_{2}(0)$, and the input $u(t)$ then determiine the concentrations through time.
+If we record the regulatory coefficients in the matrix
     \begin{align*}
-      \Sigma = \left \{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 
-        0 & 1 \\ 
-       -1 & 0 
-        \end{bmatrix} \kappa(t) + \begin{bmatrix} 1 \\ 1 \end{bmatrix} u(t) \\ 
-          \phi(t) &= \begin{bmatrix} 1 & 0 \end{bmatrix} \kappa(t) \end{array} \right.
+      A &= \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} ,
     \end{align*}
+    and define the column vector $B = \begin{bmatrix} 1 & 1 \end{bmatrix}^{T}$, then in matrix notation the dynamics are
+      \begin{align*}
+      \dot{\kappa} &= A \kappa(t) + B u(t) .
+      \end{align*}
+      Since only the dynamics of gene-1 are directly relevant to biological function, the dynamics of interest are given by
+      \begin{align*}
+      \phi(t) &= C \kappa (t)
+      \end{align*}
+      where the row vector $C$ is defined as $C = \begin{bmatrix} 1 & 0 \end{bmatrix}$.
+(Note: if the dynamics of both genes were physiologically relevant to the cell cycle, we would set $C$ to be the identity matrix).
+  %  \begin{align*}
+  %    A = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} , \qquad B = \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix}
+  %  \end{align*}
+ %       The oscillatory system $\Sigma$ is thus given as
+ %   \begin{align*}
+ %     \Sigma = \left \{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 
+ %       0 & 1 \\ 
+ %      -1 & 0 
+ %       \end{bmatrix} \kappa(t) + \begin{bmatrix} 1 \\ 1 \end{bmatrix} u(t) \\ 
+ %         \phi(t) &= \begin{bmatrix} 1 & 0 \end{bmatrix} \kappa(t) \end{array} \right.
+ %    \end{align*}
+
+      Since the input is simply an impulse, its phenotype is equivlanet to its impulse response
 
       \begin{align*}
-        h(t) = \sin(t) + \cos(t)
+        \phi(t) = h(t) = \sin(t) + \cos(t)  .
       \end{align*}
@@
     \begin{figure}[H]
-        \begin{center}
+       % \begin{center} 
+      \centering
+         \begin{tabular}{cc}
             \begin{tikzpicture}
@@
 
-            \end{tikzpicture}
-        \end{center}
-        \caption{
-            Diagram of Example \ref{ex:osc} in the text.
-            \plr{explain what arrows mean if nec}
-            \label{fig:oscillator_diagram}}
+            \end{tikzpicture} &
+       % \end{center}
+      %  \caption{
+      %      Diagram of Example \ref{ex:osc} in the text.
+      %      \plr{explain what arrows mean if nec}
+      %      \label{fig:oscillator_diagram}}
+   % \end{figure}
+   % \begin{figure}[H]
+    %  \centering
+      \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{osc_impulse}
+   \end{tabular}
+  %  \end{center}
+      \caption{(Left) Graphical representation of the cell cycle control gene network, and (right) plot of the phenotype $\phi(t)$ against time $t$.}
     \end{figure}
-    \begin{figure}[H]
-      \centering
-      \includegraphics[width=0.5\textwidth]{osc_impulse}
-    \end{figure}
+      We return to the evolution of such a system below.
   \end{example}
@@
     h(t) &= C e^{A t} B \\
+    &= C V^{-1} V e^{A t} V^{-1} V B \\
     &= C V^{-1} e^{V A V^{-1} t} V B \\
-    &= C V^{-1} V e^{A t} V^{-1} V B \\
-    &= C e^{A t} B
+    &= \bar{C} e^{\bar{A} t} \bar{B}
   \end{align}
@@
 
-  \begin{example}[External Equivalence does not imply internal equivalence]
-    Gene 1 dynamics (blue) are equivalent for network architectures $A(0)$ and $A(2)$, however the internal dynamics containing gene 2 (orange) are very different.
+  \begin{example}[All Phenotypically Equivalent Cell Cycle Control Networks]
+
+    Let
+    \begin{align*}
+      A(0) = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix}, \qquad B = \begin{bmatrix} 1 & 1 \end{bmatrix}^{T}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix},
+    \end{align*}
+    with $V(\tau)$ preserving both $B$ and $C$, then
+    the set of all two-gene regulatory networks phenotypically equivalent to the cell cycle control network in Example \ref{ex:oscillator} are given by
+    \begin{align*}
+      A(\tau) &= \begin{bmatrix} 1 & 0 \\ \tau & 1-\tau \end{bmatrix} \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 0 \\ \frac{\tau}{\tau-1} & \frac{-1}{\tau-1} \end{bmatrix}  \qquad \forall \ \tau \neq 1 %\\
+     %   B &= \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix} \\
+       %   h(t) &= \sin(t) + \cos(t) \qquad \forall \ \tau \neq 1
+    \end{align*}
     \begin{figure}[H]
     \centering
-    \includegraphics[width=0.5\textwidth]{osc_A0_A2_both_compare}
-    %\caption{Gene 1 dynamics (blue) are equivalent for network architectures $A(0)$ and $A(2)$, however the internal dynamics containing gene 2 (orange) are very different.}
+\begin{tikzpicture}
+\begin{scope}[every node/.style={circle,thick,draw}]
+  \node (A) at (0,0) {$\kappa_{1}$};
+    \node[dashed] (B) at (4,0) {$\kappa_{2}$};
+    \node[shape=rectangle] (U) at (2,2) {input};
+    \node[shape=rectangle] (y) at (2,-2) {output};
+\end{scope}
+
+\begin{scope}[>={Stealth[black]},
+              every node/.style={fill=white,circle},
+              every edge/.style={draw=black, thick}]
+    \path [->, sloped] (A) edge[bend left] node {\tiny $2 \tau + (\tau-1)^{-1}$} (B);
+    \path [->, sloped] (B) edge[bend left] node {\tiny $-(\tau-1)^{-1}$} (A); 
+    \path[->] (U) edge[dashed] node {\tiny $1$} (A);
+    \path[->] (U) edge[dashed] node {\tiny $1$} (B);
+    \path[->] (A) edge[dashed, bend right] node {\tiny $1$} (y);
+\end{scope}
+\begin{scope}[>={Stealth[black]},
+              every edge/.style={draw=black, thick}]
+    \path [->] (A) edge[loop left] node[sloped, anchor=center, above] {\tiny $1 + (\tau-1)^{-1}$} (A);
+    \path [->] (B) edge[loop right] node[sloped, anchor=center, above] {\tiny $-1 - (\tau-1)^{-1}$} (B);
+\end{scope}
+
+\end{tikzpicture}
+    \caption{A graphical depiction of the set of all externally equivalent cell cycle control networks, $A(\tau)$. $\tau$ can take any valye}
+    \end{figure}
+ % \end{example}
+
+ % \begin{example}[External Equivalence does not imply internal equivalence]
+   Despite the phenotypic equivalence of all instantiations of $A(\tau)$, the internal dynamics, or kryptotypes, vary as a function of $\tau$. 
+    Gene-1 dynamics (blue) are equivalent for network architectures $A(0)$ and $A(2)$, however the dynamics of gene-2 (orange) differ with $\tau$.
+  \begin{figure}[H]
+    \centering
+    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{osc_A0_A2_both_compare}
+    \caption{Gene-1 (blue) and gene-2 (orange) dynamics for $A(0)$ (top) and $A(2)$ (bottom).
+Both (top and bottom) gene-1 dynamics are given by $ \kappa_{1} = \sin(t) + \cos(t)$, and gene-2 by $\kappa_{2} = \cos(t) - \sin(t)$ (top) and $\kappa'_{2} = \cos(t) + 3 \sin(t)$ (bottom).}
   \end{figure}
@@
   \section*{Intraspecific variation and genetic drift}
+
+First we focus on a simple evolutionary scenario -- \emph{phenotypically invariant evolution}: a large population, perfectly adapted, and in a constant environment.
+In this circumstance we expect phenotype to be conserved throughout evolutionary time.
+As such we should only expect the phenotype to change as a consequence of genetic drift (small effective population size), adaptation to new selective and/or environemental pressures.
+These phenotypic variations should yield distinct signatures.
+Adaptative changes will change the optimal impulse response function $h(t) \xrightarrow{adaptation} h'(t)$.
+Genetic drift registers as an increase in the intrapopulation variation in $h(t)$.
+ \jss{added the phenotypic invariance stuff here (above) -- not sure if it should put elsewhere yet}
 
@@
   (there are $m$ degrees of freedom in the solutions to \eqref{eqn:equivalence}) then
-  \begin{aligned}
+  \begin{align}
       \E\left[ \|A_t - A_0\|^2 \right] = m \sigma^2 t + O(t^2) .
-  \end{aligned}
+  \end{align}
 
@@
   \end{align}
-  \begin{example}[All Phenotypically Equivalent Oscillators]
-    \begin{align*}
-      A(\tau) &= \begin{bmatrix} 1 & 0 \\ \tau & 1-\tau \end{bmatrix} \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 0 \\ \frac{\tau}{\tau-1} & \frac{-1}{\tau-1} \end{bmatrix} \\
-        B &= \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix} \\
-          h(t) &= \sin(t) + \cos(t) \qquad \forall \ \tau \neq 1
-    \end{align*}
-    \begin{figure}[H]
-    \centering
-\begin{tikzpicture}
-\begin{scope}[every node/.style={circle,thick,draw}]
-  \node (A) at (0,0) {$\kappa_{1}$};
-    \node[dashed] (B) at (4,0) {$\kappa_{2}$};
-    \node[shape=rectangle] (U) at (2,2) {input};
-    \node[shape=rectangle] (y) at (2,-2) {output};
-\end{scope}
-
-\begin{scope}[>={Stealth[black]},
-              every node/.style={fill=white,circle},
-              every edge/.style={draw=black, thick}]
-    \path [->, sloped] (A) edge[bend left] node {\tiny $2 \tau + (\tau-1)^{-1}$} (B);
-    \path [->, sloped] (B) edge[bend left] node {\tiny $-(\tau-1)^{-1}$} (A); 
-    \path[->] (U) edge[dashed] node {\tiny $1$} (A);
-    \path[->] (U) edge[dashed] node {\tiny $1$} (B);
-    \path[->] (A) edge[dashed, bend right] node {\tiny $1$} (y);
-\end{scope}
-\begin{scope}[>={Stealth[black]},
-              every edge/.style={draw=black, thick}]
-    \path [->] (A) edge[loop left] node[sloped, anchor=center, above] {\tiny $1 + (\tau-1)^{-1}$} (A);
-    \path [->] (B) edge[loop right] node[sloped, anchor=center, above] {\tiny $-1 - (\tau-1)^{-1}$} (B);
-\end{scope}
-
-\end{tikzpicture}
-    \end{figure}
-  \end{example}
-  \section*{Phenotypic Invariance}
-    First we focus on a simple evolutionary scenario: a large population, perfectly adapted, and in a constant environment.
-In this circumstance we expect phenotype to be conserved throughout evolutionary time.
-As such we should only expect the phenotype to change as a consequence of genetic drift (small effective population size), adaptation to new selective and/or environemental pressures.
-These phenotypic variations should yield distinct signatures.
-Adaptative changes will change the optimal impulse response function $h(t) \xrightarrow{adaptation} h'(t)$.
-Genetic drift registers as an increase in the intrapopulation variation in $h(t)$.
-
-    Presently we ask two questions, (1) holding $\phi(t)$, $h(t)$, and $u(t)$ constant for evolutionary time $T$, how much do we expect gene network organization to drift, and (2) does this contribute to speciation, primarily via the fixation of reproductive incompatibilities?
+ % \section*{Phenotypic Invariance}
 
@@
 
-     \begin{align}
-        C \lor B &= I \\
-        IV^{-1} &= I \\
-        \iff V &= I
-      \end{align}
+   %  \begin{align}
+   %     C  = I \lor B = I \\
+   %     IV^{-1} &= I \\
+   %     \iff V &= I
+   %   \end{align}
       \end{example}
@@
 
-      A diploid organism's gene network is simply the average of both of its gene network copies; one from each parent.
-Further, each haploid parental gene network copy is formed via meiosis -- by swapping independent genes randomly.
+     An F1's genetic regulation is a consequence of both of its genomes, where genes and regulatory sequences from both parents are equally present.
+Thus we say that a diploid organism's gene network is simply the average of both of its gene network copies; one from each parent.
+Further, each haploid parental gene network copy is formed via meiosis -- by swapping independent genes (specifically their \emph{cis}-regulatory modules) randomly.
 Assuming two distinct but genetically homogeneous populations evolving in allopatry meet and form hybrids, the first generation hybrids (F1s) gene network dynamics will be determined by the average of the parental haplotypes.
@@
       %\end{align*}
-      and $F_{2}$ is the second generation hybrid gene network architecture formed by gametes $G(i, \tau, \hat{\tau} )$ and $G(j, \tau, \hat{\tau})$.
-Where each $G(\cdot)$ is formed by randomly swapping rows between $A(\tau)$ and $A(\hat{\tau})$, such that the $i$th  gene comes from $A(\tau)$ ($i$ and $j$ are orthogonal vectors, each element $0$ or $1$, and $i \neq j$).
+      and $F_{2}$ is the second generation hybrid gene network architecture formed by gametes $G(i)$ and $G(j)$.
+Where each $G$ is formed by randomly swapping rows between $A(\tau)$ and $A(\hat{\tau})$, such that the $i$th  gene comes from $A(\tau)$ ($i$ and $j$ are orthogonal vectors, each element $0$ or $1$, and $i \neq j$).
       \begin{align}
-        F_{2}(i, j) = \frac{G (i, \tau, \hat{\tau} ) + G( j, \tau, \hat{\tau} )}{2}
+        F_{2}(i, j) = \frac{G (i) + G(j)}{2}
       \end{align}
@@
       \begin{align}
-        \mathcal{F}\left( \widehat{h}(t) \right) = \exp \left\{- \int_{0}^{\infty} \left\lVert h(t) - \widehat{h}(t) \right\rVert dt \right\}  .
+        \mathcal{F}\left( \widehat{\phi}(t) \right) = \exp \left\{- \int_{0}^{\infty} \left\lVert \phi(t) - \widehat{\phi}(t) \right\rVert dt \right\}  .
       \end{align}
@@
 
-      \begin{example}[Quantitative Biological Species Concept]
-        Two populations are said to be different species if they produce low fitness offspring.
+      \begin{definition}[Reproductive Incompatibility]
+        According to Mayr's \emph{Biological Species Concept}, two populations can be considered different species if they are reproductively isolated, meaning crosses between them produce low fitness offspring.
 
-        F1s created by crossing externally equivlaent oscillators $A(0)$ and $A(2)$ have an $h_{F_{1}}(t) = e^{t}$, in contrast to both parents with $h_{F_{0}}(t) = \sin(t) + \cos(t)$.
+        We quantitatively describe the degree of incompatibility between two populations $P_{1}$ and $P_{2}$ as
+        \begin{align*}
+          \mathcal{I} = \frac{2 \left\langle \mathcal{F} \left(\phi_{F_{1}}\right) \right\rangle}{\left\langle \mathcal{F} \left(\phi_{P_{1}}\right) \right\rangle +  \left\langle \mathcal{F} \left(\phi_{P_{2}}\right) \right\rangle} ,
+        \end{align*}
+        where angled brackets imply averaging.
+
+        F1s created by crossing phenotypically equivalent oscillators $A(0)$ and $A(2)$ have an $\phi_{F_{1}}(t) = e^{t}$, in contrast to both parents with $\phi_{P_{1}}(t) = \phi_{P_{2}}(t) = \sin(t) + \cos(t)$.
 The hybrid phenotype is significantly different (it does not oscillate and increases infinitely) despite the phenotypic equivalence of the parents.
@@
         \begin{align*}
-          \mathcal{F} \left(h_{F_{0}}\right) &= 1 \\
-          \mathcal{F}\left(h_{F_{1}}\right) &= 0
+          \mathcal{F} \left(\phi_{P_{1}}\right) = \mathcal{F} \left( \phi_{P_{2}} \right) &= 1 \\
+          \mathcal{F}\left(\phi_{F_{1}}\right) &= 0 \\
+          \mathcal{I} &= 0
         \end{align*}
-      \end{example}
 
-      \begin{example}[F1 Reproductive Incompatibility in an Oscillating Gene Network]
-        DMI examples...
+        Thus if populations $1$ and $2$ are homogenous $A(0)$ and $A(2)$, respectively, we say that they are completely incompatible as $\mathcal{I} = 0$.
+      \end{definition}
       \begin{figure}[H]
-          \includegraphics[width=0.5\textwidth]{A2_A3_osci_F1_hyb}
+        \centering
+        \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{expF1}
+        \caption{F1 hybrid (orange) and parental (blue) phenotypic oscillator dynamics for an $A(0)$ by $A(2)$ cross. The hybrid fails to oscillate and exhibits qualitatively different dynamics.}
+      \end{figure}
+ %     \begin{example}[F1 Reproductive Incompatibility in an Oscillating Gene Network]
+ %       DMI examples...
+ %       \begin{figure}[H]
+ %         \includegraphics[width=0.5\textwidth]{A2_A3_osci_F1_hyb}
+ %       \end{figure}
+ %     \end{example}
+
+      \begin{example}[Hybrid Incompatibility in an Oscillating Gene Network]
+
+        Here we compare the phenotypes for F2 hybrids formed by crossing oscillators $A(2)$ with $A(2.01)$, $A(2.1)$, and $A(2.5)$ ($B$ and $C$ are the same as above).
+Each $A$ has is phenotypically identical ($\phi(t) = \sin(t) + \cos(t)$), however some of the hybrids exhibit markedly different dynamics.
+These differences tend to increase with time, quadratically ($x^{2}$) in F2s, and quartically ($x^{4}$) in F1s.
+
+%        \begin{align*}
+ %         A(2) = \begin{bmatrix} 2 & -1 \\ 5 & -2 \end{bmatrix} &\qquad A(2.01) = \begin{bmatrix} 1.9901 & -0.9901 \\ 5.0101 & -1.9901 \end{bmatrix} \\
+  %          A(2.1) = \begin{bmatrix} 1.9091 & -0.9091 \\ 5.1091 & -1.9091 \end{bmatrix} &\qquad A(2.5) = \begin{bmatrix} 1.6667 & -0.6667 \\ 5.6667 & -1.6667 \end{bmatrix}
+   %     \end{align*}
+
+        \begin{alignat*}{2}
+          A(2) &= \left[\begin{array}{cc} 2 & -1 \\[6pt] 5 & -2 \end{array}\right] &&\qquad A(2.01) = \left[\begin{array}{cc} 2 -\frac{1}{101} & -1 + \frac{1}{101} \\[6pt] 5 + \frac{1}{99} & -2 + \frac{1}{101} \end{array}\right] \\
+            A(2.1) &= \left[\begin{array}{cc} 2 - \frac{1}{11} & -1 + \frac{1}{11} \\[6pt] 5 + \frac{6}{55} & -2 + \frac{1}{11} \end{array}\right] &&\qquad A(2.5) = \left[\begin{array}{cc} 2 - \frac{1}{3} & -1 + \frac{1}{3} \\[6pt] 5 + \frac{2}{3} & -2 + \frac{1}{3} \end{array}\right]
+        \end{alignat*}
+        F1 gene regulatory networks are formed by averaging the two parental $A(\tau)$ matrices; F2s are formed by first recombining parental matrices followed by an averaging. 
+  %      \begin{figure}[H]
+  %        \begin{tabular}{cc}
+  %          \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_1}
+  %          & \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_2}
+  %         \\ \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_3}
+  %          & \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_4}
+  %          \\ \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_5}
+  %          & \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_6}
+  %        \end{tabular}
+  %      \end{figure}
+
+  %    \begin{figure}[H]
+  %      \centering
+  %      \includegraphics[width=0.5\textwidth]{A2_A2-1_F2s}
+  %      \caption{F2s from $A(2)$ and $A(2.1)$.}
+  %    \end{figure}
+  %      \begin{figure}[H]
+  %        \begin{tabular}{cc}
+  %        \includegraphics[width=0.5\textwidth]{F2s-small} \\
+  %        \includegraphics[width=0.5\textwidth]{F2s} \\
+  %        \includegraphics[width=0.5\textwidth]{F2s-large}
+  %        \end{tabular}
+  %        \caption{F2s from $A(2)$ and $A(2.1)$.}
+  %      \end{figure}
+        \begin{figure}[H]
+          \centering
+          \begin{tabular}{cc}
+            \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{F1_comparison} & 
+            \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{F2_comparison}
+          \end{tabular}
+          \caption{F1 (left) and F2 (right) hybrids crossing $A(2)$ with $A(2.01)$ (top), $A(2.1)$ (middle), and $A(2.5)$ (bottom).
+          Note the difference in scale on the $y$-axis.
+F2 hybrids display more phenotypic divergence than F1s, on average.
+Further, some F2s completely fail to oscillate, as seen in an $A(2.5)$ F2 (light blue).}
         \end{figure}
     \end{example}
-
-      \begin{example}[F2 Reproductive Incompatibility in an Oscillating Gene Network]
-        F2 versus parental expression dynamics.
-
-        \begin{figure}[H]
-          \begin{tabular}{cc}
-            \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_1}
-            & \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_2}
-           \\ \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_3}
-            & \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_4}
-            \\ \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_5}
-            & \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_6}
-          \end{tabular}
-        \end{figure}
-      \end{example}
-
-      \begin{figure}[H]
-        \centering
-        \includegraphics[width=0.5\textwidth]{A2_A2-1_F2s}
-        \caption{F2s from $A(2)$ and $A(2.1)$.}
-      \end{figure}
-
       \begin{example}[Not all Networks can Host Incompatibilities]
@@
 Here, we offer a tractable and simple model which includes these missing features.
-Further, we provide, in clear mathematical language, a description of phenomena hitherto only discussed verbally and conceptualy (phenogenetic drift \citep{weiss2000phenogenetic}, developmental systems drift \citep{true2001developmental}, biological degeneracy \citep{edelman2001degeneracy}, \emph{etc.}).
+Further, we provide, in clear mathematical language, an analytical description of phenomena hitherto only discussed verbally and conceptualy (phenogenetic drift \citep{weiss2000phenogenetic}, developmental systems drift \citep{true2001developmental}, biological degeneracy \citep{edelman2001degeneracy}, \emph{etc.}).
 The tractability and relative simplicity of this exposition enables the interested biologist to work out by hand, if desired, the dynamics of a genetic system, as well as perturbations to the system -- an attribute not likely to be found in less tractable models and simulations.
 
-  We have offered a novel interpretation of system identification: to see it as an evolutionarily neutral manifold, and not simply a computational nuissance.
+  We have suggested an interpretation of system identification: to see it as an evolutionarily neutral manifold, and not simply a computational nuissance.
 We have demonstrated a method to analytically determine the set of all phenotypically invarint gene networks; by a simple change of coordinates in the minimal configuration, or more generally by applying the Kalman decomposition in higher dimensions.
@@
 
-      \tiny
-\bibliographystyle{unsrt}
+  \jss{literature comparison:}
+
+  Over the last several years, several different computational approaches have been applied to study reproductive incompatibility and speciation.
+\citet{tulchinsky } simulated the evolution of a transcription factor and its binding site using a thermodynamic model.
+Their simulations suggest that the language by which a transcription factor recognizes its binding site can change, and potentially lead to hybird incomaptibility when allopatric populations employ divergent readout languages.
+This study, despite looking at gene regulation, does not analyze overall gene network architecture -- as we do here -- it only looks at the expression level of a single gene.
+Furthermore, they report reproducitve isolation primarily following direcitonal selection for a change in expression levels in each allopatric population; the evidecne for reproductive isolation following balancing selection is much weaker.
+Johnson and Porter 2000 did not observe any hybrid fitness declines under stabilizing selection -- only under directional selection.
+  Khatari et al, Tulchinsky et al, and Porter et al, all study hybrid incompatibility from a transcription factor/binding site interaction perspective, not from an overal network architecture perspective.
+Palmer and Feldman only see hybrid incompatibility in constant environments if the parental populations are relatively poorly adapted initially.
+Otherwise hybrids between two allopatric populations have fairly high fitnesses.
+
+\section*{Acknowledgements}
+We would like to thank Sergey Nuzhdin, Erik Lundgren, and Hossein Asgharian for valuable discussion.
+
+\bibliographystyle{plainnat}
 \bibliography{krefs}
-\end{multicols}
+%\end{multicols}
 
@@
     &=
-    x^T A \left( x - 2y) + y^T A y ,
+    x^T A \left( x - 2y \right) + y^T A y ,
 \end{align*}

### 2017-06-16 handnotes.tex
added LTI description from kalman_walk

@@
 The \emph{what}, \emph{when}, and \emph{where}, of an organism's molecules that are physiologically or otherwise relevant for survival.
-Thus we say that some function $\phi(t)$ is a phenotype where,
+  \jss{  Thus we say that some function $\phi(t)$ is a phenotype where, 
   \begin{align}
@@
 Lastly, a portion and/or combination of these molecules, $\phi(t)$, are ``observed'' by selection (this is in constrast to $\kappa(t)$ -- the \emph{kryptotype} -- as it is ``hidden'' from direct selection).
+}
+
+We will now lay out the model in more general terms.
+Suppose that the \emph{internal state} of the system
+is parameterized by the concentrations of a collection of $n$ molecular species,
+$S_1, \ldots, S_n$,
+and the vector of concentrations at time $t$ we denote $\kappa(t)=(\kappa_1(t),\ldots,\kappa_n(t))$.
+There are also $m$ ``input'' species, whose concentrations are determined
+exogenously to the system,
+and are denoted $u(t) = (u_1(t),\ldots,u_m(t))$,
+and $\ell$ ``output'' species, whose concentrations are denoted
+$\phi(t) = (\phi_1(t),\ldots,\phi_\ell(t))$.
+The output is merely a linear function of the internal state:
+\begin{align*}
+    \phi_i(t) = \sum_j C_{ij} \kappa_i(t).
+\end{align*}
+Since $\phi$ is what natural selection acts on, we refer to it as the \emph{phenotype},
+and sometimes in contrast refer to $\kappa$ as the \emph{kryptotype},
+as it is ``hidden'' from direct selection.
+The rate at which the $i^\text{th}$ species is produced
+is a weighted sum of the concentrations of the other species
+as well as the input:
+\begin{align*}
+    \ddt \kappa_i(t) = \sum_j A_{ij} \kappa_j(t) + \sum_r B_{ir} u_r(t) .
+\end{align*}
+In matrix notation, this is written more concisely as
+\begin{align} \label{eqn:lti_system}
+    \ddt \kappa(t) &= A \kappa(t) + B u(t) \\
+    \phi(t) &= C \kappa(t) .
+\end{align}
 

### 2017-06-20 handnotes.tex
conceptual figure

@@
 \end{abstract}
+
+Additional ideas to consider adding:
+\begin{itemize}
+    \item Haldane's rule (easy point in discussion: makes F1s look like F2s)
+    \item add point about F1: quartic and F2: quadratic to results and maybe abstract/discussion
+    \item hybrid vigor (need to do calculation)
+    \item discuss linearity, linearization, and canalization in introduction
+    \item obtain estimates of variation in $A$ from thermodynamic occupancy model
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%
 \section*{Introduction}
-
-  \jss{add stuff on Haldane's Rule and Hybrid Vigor; note: F1s diverge to the 4th power, and F2s quadratically}
 
@@
 
+  \plr{rework this: we assume optimal phenotype is fixed and population is large but polymorphic and not strictly on the optimum}
+
 First we focus on a simple evolutionary scenario -- \emph{phenotypically invariant evolution}: a large population, perfectly adapted, and in a constant environment.
@@
   How fast will a population explore the space of equivalent networks?
+  To answer these questions, we need to know two things:
+  the strength of stabilizing selection on the phenotype,
+  and the amount (and structure) of heritable variation in the genotype.
+  Both are hard to measure, somewhat ambiguous, and have values that may depend strongly on the context.
+  We aim to get order-of-magnitude estimates.
 
-  The amount and structure of 
-  this standing variation is established over long time scales
+  Specifically, we want to know how much the entries of $A$ vary within a population.
+  The coefficient $A_{ij}$ measures how much the rate of net production of $i$ changes
+  per change in concentration of $j$.
+  It is generally thought that regulatory sequence change contributes much more to inter- and intraspecific variation
+  than does coding sequence change affecting molecular structure \citep{schmidt2010vertebrate}.
+  In the context of transcription factor networks this may be affected 
+  not only by the binding strength of $j$ to the promoter region of $i$
+  but also the effects of other transcription factors (e.g., cooperativity)
+  and local chromatin accessibliity \citep{stefflova2013cooperativity}.
+  (Recall that although these are best modeled through nonlinear terms,
+  by linearizing we essentially consider first-order effects.)
+  Furthermore, the mutational target size for variation in $A$ may be much larger than the kilobase or so of a typical promoter sequence.
+  On the other hand, a diverse set of buffering mechanisms are thought to contribute to phenotypic stability
+  in the presence of substantial molecular noise \citep{canalization,buffering},
+  suggesting that substantial variation in the micro-scale dynamics we consider here
+  may be necessary to produce relevant phenotypic effects downstream.
+  \plr{replace "micro-scale" with sthg else or discuss earlier}
+
+  The amount and structure of this standing variation is established over long time scales
   by many factors, including
@@
   shifts in the phenotypic optimum,
-  and/or spatial variation in the optimum \citep{sources_of_standing_var}.
+  and/or spatial variation in the optimum \citep{hansen1996translating}.
   Quantitative genetics models of mutation-selection balance 
-  predict precise levels and structure of standing variation \citep{kimura_mutsel,lande_mutsel},
+  predict precise levels and structure of standing variation \citep{kimura_mutsel,lande_mutsel,lande1981models},
   but it is unclear how well these predictions match reality \citep{johnson_barton}
   and how much they are expected to change over time \citep{arnold_changing_G}.
-  However, recent empirical work allows us to estimate at least the rough magnitude of variation.
+  However, empirical work allows us to estimate at least the rough magnitude of variation.
   XXX FILL IN DETAILS: 
@@
 \section*{Acknowledgements}
-We would like to thank Sergey Nuzhdin, Erik Lundgren, and Hossein Asgharian for valuable discussion.
+We would like to thank Sergey Nuzhdin, Stevan Arnold, Erik Lundgren, and Hossein Asgharian for valuable discussion.
 
@@
 moves under genetic drift
-with a quantitative genetics model.
+with a quantitative genetics model,
+as in \citet{lande1981models} or \citet{hansen1996translating}.
+These ignore details of the underlying genetic basis,
+but developing a more accurate model is beyond the scope of this paper.
 
@@
 \section{Differentiating the fitness function}
-
-\plr{In case this is useful...}
 
@@
 then this is approximately $-\sigma^2 H^{-1} Z$.
+This suggests that the population mean follows an Ornstein-Uhlenbeck process,
+as described (in different terms) in \citet{hansen1996translating}.
 

### 2017-06-20 promoterdata.tex
conceptual figure

@@
 \usepackage[backref=page, hidelinks, colorlinks=true, citecolor=blue!60!black!100]{hyperref}
-\usepackage{tikz}
-\usetikzlibrary{arrows.meta, shapes}
 \usepackage{caption, subcaption}
 \usepackage{natbib} % gives us \citet: Author (year) and \citep: (Author; year)
-\usepackage{authblk}
-
-\usepackage{multicol}
-
-\newcommand{\plr}[1]{{\color{blue}\it #1}}
-\newcommand{\jss}[1]{{\color{olive}\it #1}}
-% \newcommand{\ddt}{\frac{d}{dt}}
-\newcommand{\ddt}{\dot}
-\newcommand{\ro}{{ro}}
-\newcommand{\nro}{{\bar{r}o}}
-\newcommand{\rno}{{r\bar{o}}}
-\newcommand{\nrno}{{\bar{r}\bar{o}}}
-\newcommand{\reachable}{\mathcal{R}}
-\newcommand{\unobservable}{\bar{\mathcal{O}}}
-\newcommand{\R}{\mathbb{R}}
-\newcommand{\E}{\mathbb{E}}
-\renewcommand{\P}{\mathbb{P}}
-\newcommand{\pda}{\frac{\partial}{\partial A_{ij}}}
-\newcommand{\ind}{\mathds{1}}
-
-\newcommand{\calA}{\mathcal{A}}
-\newcommand{\diag}{\text{diag}}
-\newcommand{\1}{\mathbbm{1}}
-
-\DeclareMathOperator{\spn}{span}
-
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}{Lemma}
-\newtheorem{definition}{Definition}
-\newtheorem{example}{Example}
 
 \begin{document}
+
+\section{Nonlinearity in TF networks}
+
+A common model is to use the Hill equation:
+if $j$ upregulates $i$ then
+\begin{align}
+    \dot x_i(t) = \beta \frac{x_j(t)^n}{ K_A^n + x_j(t)^n}
+\end{align}
+or if it downregulates,
+\begin{align}
+    \dot x_i(t) = \beta (1-\frac{x_j(t)^n}{ K_A^n + x_j(t)^n} ).
+\end{align}
+Citation: Alon's book? gertz2008analysis?
 
@@
 \begin{itemize}
+
+    \item \citet{kasowski2010variation} use ChIP-seq to survey variation in binding sites between 10 human lymphoblastoid cell lines:
+        RNA PolII (not a TF like we are considering) and NFkB (p65, a regulator of immune responses, which does have binding motifs);
+        individuals differed at 25\% and 7.5\% of the sites respectively. 
+        Also found 32\% binding sites having significant difference in PolII occupancy between human/chimp
+        \begin{itemize}
+            \item Differences explained by SNPs, structural variants, and other things (like \citet{shi2016evaluating})
+            \item ``Significant differences in binding'' have SD of +/- 1 log2 difference 
+            \item Weak connection of binding differences to expression differences:
+                ``We examined the effect of binding variation on gene expression by generation of deep RNASeq
+                data from each cell line (9) and comparison with binding data (Fig. 3A, Fig. S9A). A
+                significant correlation was observed (Spearman correlation coefficients of 0.475 and 0.461 for
+                NFκB and PolII, respectively) (Fig. 3B, Fig. S9B, Table S9), suggesting an influence of binding
+                differences on mRNA abundance''
+            \item "Since humans and chimpanzees exhibit 5–10\% differences in gene expression (15)"
+        \end{itemize}
+
+    \item \citet{schmidt2010vertebrate} use ChIP-seq to compare TF occupancy (for CEBPA and HNF4A) between five vertebrates (human, mouse, dog, opossum, and chicken), 
+        and find (a) conservation of binding motif; (b) little sharing of actual binding sites; and (c) a fair amount of compensatory changes
+        \begin{itemize}
+
+            \item "Among the binding events lost in one lineage, only half are recovered by another binding event within 10 kilobases"
+
+            \item "CEBPA and HNF4A were selected as representative transcription factors within the liver-specific regulatory
+                network because both are conserved and constitutively expressed with well-characterized target genes (10, 11). In addition,
+                they represent distinct TF classes, and the DNA binding domains of each factor's orthologs are nearly identical among the study species "
+
+            \item "About 250 direct functional HNF4A target genes have recently been identified"
+
+            \item both have consensus motifs of about 5 of 10 bp
+
+            \item "only between 6-8 \% of the genomic regions occupied by CEBPA in opossum liver align with CEBPA binding events also found in
+                mouse, dog, and/or human liver. This divergence was even greater in chicken, which shared only 2\% of CEBPA binding with human"
+        \end{itemize}
+
+    \item \citet{shi2016evaluating} looked at variation in TF binding in heterozygotes in ENCODE data
+        and found that differential binding is often explained by SNPs in the TF's motif (19.3\%) or potential partner TFs' motifs (9.4\%);
+        variation in chromatin accessibility; or histone modifications (NOT just SNPs).
+        \begin{itemize}
+            \item shifts of motif score of 0.05--0.1 were responsible for differential occupancy
+            \item "While altered canonical TFBSs for the ChIP'd TFs were frequently observed (19.3\%), most ASB SNVs did not overlap with the primary TF motif."
+        \end{itemize}
+
+    \item \citet{stefflova2013cooperativity} In five Mus (sub-)species, found that
+        ``Although individual mutations in bound sequence motifs can influence TF binding, most binding differences occur in the absence of nearby sequence variations.''
+        \begin{itemize}
+            \item "Clustered TF binding appears to result in large part
+                from indirect cooperativity to open chromatin regions, as opposed
+                to direct TF-TF protein interactions (Kaplan et al., 2011; Miller
+                and Widom, 2003; Mirny, 2010). For binding sites within a
+                nucleosome-length distance, each TF contributes partially to a
+                competitive displacement of specific nucleosomes by indirect
+                collaboration with other TFs, mutually aiding each others’ binding
+                to DNA. TFs within a cluster can have different regulatory roles
+                depending on their motif strength and ability to compete with
+                nucleosomes (Zinzen et al., 2009)." 
+        \end{itemize}
+
+    \item \citet{tugrul2015dyanmics} do theory on the gain and loss of TFBS.  
+        Conclusion: more than 10bp unlikely; de novo very hard but made possible ad the scale of eukaryotic species divergence by pre-existing relics, cooperativity, etc.
+        Euk TFBS typically 6--12bp.
+
   \item \citet{gertz2009analysis} study synthetic yeast promoters.

### 2017-06-20 handnotes.tex
more words about variation

@@
-\documentclass[9 pt]{article}
+\documentclass{article}
 \usepackage{amsmath, amssymb, color, xcolor}
@@
   How fast will a population explore the space of equivalent networks?
-  To answer these questions, we need to know two things:
+  To answer these questions, we need to know two things known at best only roughly \citep{felsenstein1988phylogenies} --
   the strength of stabilizing selection on the phenotype,
   and the amount (and structure) of heritable variation in the genotype.
-  Both are hard to measure, somewhat ambiguous, and have values that may depend strongly on the context.
   We aim to get order-of-magnitude estimates.
 
-  Specifically, we want to know how much the entries of $A$ vary within a population.
+  \plr{amount of heritable variation}
+  We quantify (roughly) the amount of heritable variation
+  by $\sigma^2$, the genetic variance present in a population in a typical entry of $A$.
   The coefficient $A_{ij}$ measures how much the rate of net production of $i$ changes
@@
   and local chromatin accessibliity \citep{stefflova2013cooperativity}.
+  For this reason, 
+  the mutational target size for variation in $A_{ij}$ may be much larger than the dozens of base pairs
+  typically implicated in the handful of binding sites for transcription factor $j$ of a typical promoter region.
   (Recall that although these are best modeled through nonlinear terms,
   by linearizing we essentially consider first-order effects.)
-  Furthermore, the mutational target size for variation in $A$ may be much larger than the kilobase or so of a typical promoter sequence.
   On the other hand, a diverse set of buffering mechanisms are thought to contribute to phenotypic stability
@@
   \plr{replace "micro-scale" with sthg else or discuss earlier}
+  Differences in $A_{ij}$ due to a sequence change are hard to measure 
+  -- the closest available data we could find either related to variation in transcription factor binding site occupancy
+  or in expression levels (e.g., cis-eQTL).
+  The former may be an overestimate, 
+  since it does not capture buffering effects (if for instance only one site of many needs be occupied for transcription to begin)
+  and the latter probably measures changes in steady-state concentration (our $x_i$) rather than the rate of change.
+  \citet{kasowski2010variation} found differential occupancy in 7.5\%
+  of binding sites of a transcription factor (p65) between human individuals.
+  \citet{verlaan2009targeted} showed that cis-regulatory variation
+  accounts for around 2--6\% of expression variation in human blood-derived primary cells,
+  while \citep{lappalainen2013transcriptome} found that human population variation 
+  explained about 3\% of expression variation.
+  \plr{Get some data from at least one other species in here!}
+  Taken together, this suggests that variation in the entries of $A$
+  may be on the scale of 1\% between individuals of a population --
+  doubtless varying substantially between species and between genes.
 
+  \plr{merge this above}
   The amount and structure of this standing variation is established over long time scales
@@
   However, empirical work allows us to estimate at least the rough magnitude of variation.
-  XXX FILL IN DETAILS: 
-  Mutational scans find this distribution of binding site change per nucleotide change in promoters.
-  Intraspecific variation is often in the range of 0.1\%--1\%.
-  This suggests typical regulatory binding strengths within a population might vary by up to roughly X\%.
-  But, does it?
-  \citet{maria_and_sergey} (others?) found evidence that large-effect regulatory mutations are weakly selected against in Drosophila,
-  which suggests that (a) the observed range of variation is somewhat less than this,
-  and (b) the strength of selection on phenotype is sufficient to weakly constrain this variation.
 
-  This suggests that within a population, transcription factor binding strengths 
-  -- the entries of $A$ --
-  vary by around X\%, at least for networks whose function is strongly constrained.
+  \plr{strength of selection}
+  It seems certain that selection is not so strong that intrapopulation variation
+  is strongly deleterious,
+  so that if $u$ is the typical scale on which selection acts
+  -- e.g., a phenotype that differs from the optimum by $x$ has fitness of $\exp(-x^2/2u^2)$ --
+  then $u > \sigma$.
+  However,
+  a range of studies \plr{find them} have found evidence for weak stabilizing selection
+  on regulatory SNPs and cis-eQTL.
+  For instance, 
+  \citet{maria_and_sergey} (others?) found evidence that large-effect regulatory mutations are weakly selected against in Drosophila.
+  This suggests that 
+  the strength of selection on phenotype is sufficient to weakly constrain regulatory variation,
+  so that perhaps $\sigma$ and $u$ are relatively close.
+  This is as would be expected if available variation is held in check by mutation--selection balance
+  rather than genetic drift.
+  A conservative estimate would be that $u = 5 \sigma$;
+  taking $\sigma=.01$ as above, this suggests that changes in phenotype of 5\% are sufficient
+  to effect a noticable drop in fitness.
+  \plr{BUT $\sigma$ IS VARIATION IN $A$ NOT PHENOTYPE}
+
+  We have guessed that within a population the entries of $A$
+  vary by a factor of $\sigma=0.01$, at least for networks whose function is strongly constrained.
   Subsequent generations for the most part resample from this diversity,
@@
   the population mean binding strengths
-  will move a few multiples of $X/\sqrt{N_e}\%$ per generation \citep{lande_drift}.
+  will move a few multiples of $\sigma/\sqrt{N_e}\%$ per generation \citep{lande_drift}.
   (This could be taken as a definition of $N_e$.)
@@
   Over the last several years, several different computational approaches have been applied to study reproductive incompatibility and speciation.
-\citet{tulchinsky } simulated the evolution of a transcription factor and its binding site using a thermodynamic model.
+\citet{tulchinsky} simulated the evolution of a transcription factor and its binding site using a thermodynamic model.
 Their simulations suggest that the language by which a transcription factor recognizes its binding site can change, and potentially lead to hybird incomaptibility when allopatric populations employ divergent readout languages.

### 2017-06-20 promoterdata.tex
more words about variation

@@
-\documentclass[9 pt]{article}
+\documentclass{article}
 \usepackage{amsmath, amssymb, color, xcolor, amsthm}
@@
 
+Generally, there are three sorts of data on variation:
+\begin{enumerate}
+    \item how binding site occupancy varies in a population
+    \item how promoter SNPs affect expression (or phenotype) by swapping them out
+    \item how cis-eQTL affect expression by population scans, esp in heterozygotes
+\end{enumerate}
+
+\emph{Orphaned, unconvincing paragraph:}
+  In \textit{Drosophila}, \citet{frankel2011morphological} 
+  found that clusters of 2--5 mutations in enhancers
+  reduced expression \plr{of what?} by 4.6--33.5\%,
+  and \citet{rebeiz2009stepwise} found that around five cis-regulatory SNPs
+  cause differences of 8--40\% of \textit{ebony} expression differences
+  between light and dark lines;
+  while in human (ENCODE) data, \citet{shi2016evaluating} found that shifts in binding site motif score
+  of 5\%--10\% are responsible for differential occupancy.
+  These latter observations suggest that changes of several base pairs
+  can shift occupancy and/or expression by tens of percent;
+  if a typical promoter contains 100 bases relevant for binding of a particular transcription factor,
+  in which region 1\% of the population contains a SNP,
+  this suggests population expression variance due to such factors of around 0.1--1\%.
+
 \begin{itemize}
@@
 
-    \item \citet{schmidt2010vertebrate} use ChIP-seq to compare TF occupancy (for CEBPA and HNF4A) between five vertebrates (human, mouse, dog, opossum, and chicken), 
+    \item \citet{schmidt2010fivevertebrate} use ChIP-seq to compare TF occupancy (for CEBPA and HNF4A) between five vertebrates (human, mouse, dog, opossum, and chicken), 
         and find (a) conservation of binding motif; (b) little sharing of actual binding sites; and (c) a fair amount of compensatory changes
@@
 
-    \item \citet{tugrul2015dyanmics} do theory on the gain and loss of TFBS.  
+    \item \citet{tugrul2015dynamics} do theory on the gain and loss of TFBS.  
         Conclusion: more than 10bp unlikely; de novo very hard but made possible ad the scale of eukaryotic species divergence by pre-existing relics, cooperativity, etc.
@@
   \item \citet{rebeiz2009stepwise} look at adaptive melaninism in D. melanogaster. 
-    Find a minimum of 5 mutations differentiate regulation of dark from light lines. These account for 8 to 40\% of the overal difference in ebony activity.
+      Find a minimum of 5 mutations differentiate regulation of dark from light lines.
+These account for 8 to 40\% of the overal difference in ebony activity.
+\emph{(But, this is adaptive -- different selection in different places.)}
 

### 2017-06-20 attic/model_notes.tex
minor editing

@@
       \Sigma(0) := \left\{ \begin{array}{ll}
-         \dot{x} &= \left[ \begin{array}{cc} 0 & 1 \\ -1 & 0 \end{array} \right] x + \left[ \begin{array}{c} 1 \\ 1 \end{array}\right] u \\
-           y &= \left[ \begin{array}{cc} 1 & 0 \end{array} \right] x
-         \end{array} \right.
-    \end{align*}
-
-    \begin{align*}
-      \Sigma(0) := \left\{ \begin{array}{ll}
       \dot{x}_{1} &= +x_{2} + u \\
@@
     \end{align*}
-
+\newpage
        \begin{align*}
-         \Sigma(2) :&= \left\{ \begin{array}{ll}
-         \dot{x} &= \left[ \begin{array}{cc} 2 & -1 \\ 5 & -2 \end{array} \right] x + \left[ \begin{array}{c} 1 \\ 1 \end{array}\right] u \\
-           y &= \left[ \begin{array}{cc} 1 & 0 \end{array} \right] x
+         \Sigma &= \left\{ \begin{array}{ll}
+           \dot{\kappa}(t) &= \left[ \begin{array}{cc} 0 & 1 \\ -1 & 0 \end{array} \right] \kappa(t) + \left[ \begin{array}{c} 1 \\ 1 \end{array}\right] u(t) \\
+             \phi(t) &= \left[ \begin{array}{cc} 1 & 0 \end{array} \right] \kappa(t)
          \end{array} \right. \\ \\
-         \Sigma(2) :&= \left\{ \begin{array}{ll}
-        \dot{x}_{1} &= 2 x_{1} + - x_{2} + u \\
-        \dot{x}_{2} &= 5 x_{1} + -2 x_{2} + u \\
-      y &= x_{1}
-      \end{array} \right.
+         \widehat{\Sigma} &= \left\{ \begin{array}{ll}
+           \dot{\kappa}(t) &= \left[ \begin{array}{cc} 2 & -1 \\ 5 & -2 \end{array} \right] \kappa(t) + \left[ \begin{array}{c} 1 \\ 1 \end{array}\right] u(t) \\
+             \phi(t) &= \left[ \begin{array}{cc} 1 & 0 \end{array} \right] \kappa(t)
+         \end{array} \right. \\ \\
+         \phi(t) &= \sin(t) + \cos(t)
+        % \Sigma(2) :&= \left\{ \begin{array}{ll}
+       % \dot{x}_{1} &= 2 x_{1} + - x_{2} + u \\
+       % \dot{x}_{2} &= 5 x_{1} + -2 x_{2} + u \\
+     % y &= x_{1}
+     % \end{array} \right.
     \end{align*}

### 2017-06-20 handnotes.tex
minor editing

@@
 
-  Organisms' phenotypes are constructred by gene by gene by environment interactions.
+  \jss{Organisms' phenotypes are constructred by gene by gene by environment interactions.
 Here we simply define the phenotype to be the organismal temporal molecular dynamics directly under natural selection.
 The \emph{what}, \emph{when}, and \emph{where}, of an organism's molecules that are physiologically or otherwise relevant for survival.
-  \jss{  Thus we say that some function $\phi(t)$ is a phenotype where, 
+   Thus we say that some function $\phi(t)$ is a phenotype where, 
   \begin{align}

### 2017-07-02 handnotes.tex
rescaled F2 hybrid figure
edited talk
made a poster

@@
 I will outline an analytical theory to study the evolution of biological systems such as gene regulatory networks, borrowing insight and tools from control engineering, systems identification, and dynamical systems theory.
-I will describe a null model of regulatory network evolution by analytically describing the set of all linear gene networks (of any size) that produce identical phenotypes -- and the evolutionary paths connecting them.
-In the idealized case of a perfectly adapted population, constant selection, and a static environment, we observe neutral evolution as a random walk over the phenotypically- invariant network-space.
-Under neutral conditions, this model can provide descriptions of expected network size and connectivity under mutation-selection equilibrium, estimate the rate of regulatory rewiring, and the rates at which Dobzhansky-Muller incompatibilities arise in reproductively isolated populations.
-This analysis provides insight into the mechanisms and parameters important for understanding developmental systems drift, network rewiring, evolvability, epistasis, and speciation, as well as the tenuous connection between network architecture and function.
+I will describe a null model of regulatory network evolution by analytically describing the set of all linear gene networks (of any size) that produce identical phenotypes – and the evolutionary paths connecting them.
+In the idealized case of a perfectly adapted population, constant selection, and a static environment, we observe neutral evolution as a random walk over the phenotypically invariant network-space.
+Under neutral conditions, this model can describe the rates at which Dobzhansky-Muller incompatibilities arise in reproductively isolated populations.
+This analysis provides insight into the mechanisms and parameters important for understanding developmental systems drift, network rewiring, and speciation.
 %}
@@
       \begin{align}
-        \mathcal{F}\left( \widehat{\phi}(t) \right) = \exp \left\{- \int_{0}^{\infty} \left\lVert \phi(t) - \widehat{\phi}(t) \right\rVert dt \right\}  .
+        \mathcal{F}\left( \widehat{\phi}(t) \right) = \exp \left\{- \frac{1}{\sigma} \int_{0}^{\infty} \left\lVert \phi(t) - \widehat{\phi}(t) \right\rVert^{2} dt  \right\}  .
       \end{align}
@@
             \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{F1_comparison} & 
-            \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{F2_comparison}
+            \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{F2s_comparison2}
           \end{tabular}

### 2017-07-27 handnotes.tex
better apx explanation
edited paper -- added Haldane's rule to discussion, added stuff on rate of speciation, and changed the description of the linear model. Also added Kalman to the appendix.

@@
 
-\section*{The evolution of phenotypically invariant gene networks with implications for speciation, adaptation, and neutral variation}
+\section*{Gene regulatory network drift and speciation occurs rapidly under neutrality}
 {\centering
@@
  % \textbf{
-I will outline an analytical theory to study the evolution of biological systems such as gene regulatory networks, borrowing insight and tools from control engineering, systems identification, and dynamical systems theory.
-I will describe a null model of regulatory network evolution by analytically describing the set of all linear gene networks (of any size) that produce identical phenotypes – and the evolutionary paths connecting them.
-In the idealized case of a perfectly adapted population, constant selection, and a static environment, we observe neutral evolution as a random walk over the phenotypically invariant network-space.
+We introduce an analytical theory to study the evolution of biological systems such as gene regulatory networks, applying insight and tools from control engineering, systems identification, and dynamical systems theory.
+We describe a null model of regulatory network evolution by analytically describing the set of all linear gene networks (of any size) that produce identical phenotypes – and the evolutionary paths connecting them.
+Under constant selection and environmental stasis, neutral evolution proceeds as a random walk through phenotypically identical network-space.
 Under neutral conditions, this model can describe the rates at which Dobzhansky-Muller incompatibilities arise in reproductively isolated populations.
-This analysis provides insight into the mechanisms and parameters important for understanding developmental systems drift, network rewiring, and speciation.
+This analysis provides insight into the mechanisms and parameters important for understanding developmental systems drift and speciation.
 %}
@@
 
-  In this paper we propose a framework to study the evolution of biological systems.
-We expect wide applicability to this approach (\emph{i.e.} non-neutral evolution), however presently, we focus solely on neutral evolution.
-To begin, we focus on the evolution of an idealized population.
-We consider the neutral evolution of a large population, evolving in a static environment.
-Under these circumstances we expect the population explores the manifold of phenotypically invariant (or symmetric) gene regulatory networks.
-
-  Presently we ask two questions, (1) holding the phenotype and environment constant for evolutionary time, how much do we expect gene network organization to drift, and (2) does this contribute to speciation, primarily via the fixation of reproductive incompatibilities?
+  In this paper we outline a theoretical framework to study the evolution of biological systems.
+We expect wide applicability to this approach (\emph{i.e.} non-neutral evolution), however presently, we focus solely on neutral evolution, that is where phenotype is conserved over evolutionary time.
+
+  We derive an analytical description of the set of all linear biological systems with identical phenotypes -- that is we decribe the set of all gene network architectures that yield identical phenotypes, and show that all biological systems can, in principal, can undergo systems drift.
+In the neutral case, this set desribes a manifold that evolution explores leaving phenotype invariant with respect to mutation, and predicts that if two populations become reproductively isolated, hybrid incompatibility can occur, despite the absence of adaptation, directional selection, or environmental change.
+Speciation typically occurs on timesclaes approximately on the order of $N_{e}$ generations, where $N_{e}$ is the effective population size.
 
@@
 
-  \jss{Organisms' phenotypes are constructred by gene by gene by environment interactions.
+%  \jss{Organisms' phenotypes are constructred by gene by gene by environment interactions.
 Here we simply define the phenotype to be the organismal temporal molecular dynamics directly under natural selection.
 The \emph{what}, \emph{when}, and \emph{where}, of an organism's molecules that are physiologically or otherwise relevant for survival.
-   Thus we say that some function $\phi(t)$ is a phenotype where, 
-  \begin{align}
-    \phi(t) = \int_{0}^{\infty} h(t) u(t) dt  ,
-  \end{align}
-  and $h(t)$ is the \emph{impulse response} of the system and $u(t)$ is the \emph{input} function, both functions of time $t$.
+%   Thus we say that some function $\phi(t)$ is a phenotype where, 
+%  \begin{align}
+%    \phi(t) = \int_{0}^{\infty} h(t) u(t) dt  ,
+%  \end{align}
+%  and $h(t)$ is the \emph{impulse response} of the system and $u(t)$ is the \emph{input} function, both functions of time $t$.
 The input can be interpreted as the environment, as initial conditions, or otherwise, depending on the biological specifics under study.
 The phenotype is a convolution of both the impulse response and the input, as allowed by our assumption of linearity.
-
-  Essentially the phenotype $\phi(t)$ is a consequence of an organism's specific gene by gene interactions, given by $h(t)$, reacting further with the local environment, given by $u(t)$.
-
-  We describe the impulse response as, 
-  \begin{align}
-    h(t) = C e^{A t} B
-  \end{align}
-  where $A$ is a gene regulatory network (although this model can be generalized to other biological networks) -- a square matrix, and $B$ filters and translates the input to the system.
+%
+%  Essentially the phenotype $\phi(t)$ is a consequence of an organism's specific gene by gene interactions, given by $h(t)$, reacting further with the local environment, given by $u(t)$. 
+%
+%  We describe the impulse response as, 
+%  \begin{align}
+%    h(t) = C e^{A t} B
+%  \end{align}
+%  where $A$ is a gene regulatory network (although this model can be generalized to other biological networks) -- a square matrix, and $B$ filters and translates the input to the system.
 The form of $B$ determines precisely how the state of the external environment influences the internal gene network.
 $C$ filters and translates the dynamics of the system and precesily determines the output, that is, what is visible to selection.
-
-  Generally $A$ can be any real $n \times n$ matrix, $B$ any $n \times \ell$, and $C$ any $\ell \times n$ dimensional matrix.
+%
+%  Generally $A$ can be any real $n \times n$ matrix, $B$ any $n \times \ell$, and $C$ any $\ell \times n$ dimensional matrix.
 Each $i$th row of $A$ describes the \emph{cis}-regulatory module for gene $i$, and each $j$th entry, the specific regulatory influence of gene $j$ on gene $i$.
-
-  Although $\phi(t)$ describes the phenotype given an input, $h(t)$ describes the phenotype subject only to an impulse -- an input present initially and absent immediately thereafter.
+%
+%  Although $\phi(t)$ describes the phenotype given an input, $h(t)$ describes the phenotype subject only to an impulse -- an input present initially and absent immediately thereafter.
 Typically, a system $\Sigma$, is defined as,
-  \begin{align}
-    \Sigma = \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right.
-  \end{align}
-  Variables have the same identities as described above and $\kappa(t)$ is a vector of molecule concentrations at time $t$.
+%  \begin{align}
+%    \Sigma = \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right.
+%  \end{align}
+%  Variables have the same identities as described above and $\kappa(t)$ is a vector of molecule concentrations at time $t$.
 Therefore the molecular concentrations at a specific time are completely determined by the input and gene by gene interactions.
 Lastly, a portion and/or combination of these molecules, $\phi(t)$, are ``observed'' by selection (this is in constrast to $\kappa(t)$ -- the \emph{kryptotype} -- as it is ``hidden'' from direct selection).
-}
-
-We will now lay out the model in more general terms.
-Suppose that the \emph{internal state} of the system
-is parameterized by the concentrations of a collection of $n$ molecular species,
-$S_1, \ldots, S_n$,
-and the vector of concentrations at time $t$ we denote $\kappa(t)=(\kappa_1(t),\ldots,\kappa_n(t))$.
-There are also $m$ ``input'' species, whose concentrations are determined
-exogenously to the system,
-and are denoted $u(t) = (u_1(t),\ldots,u_m(t))$,
-and $\ell$ ``output'' species, whose concentrations are denoted
-$\phi(t) = (\phi_1(t),\ldots,\phi_\ell(t))$.
-The output is merely a linear function of the internal state:
-\begin{align*}
-    \phi_i(t) = \sum_j C_{ij} \kappa_i(t).
-\end{align*}
-Since $\phi$ is what natural selection acts on, we refer to it as the \emph{phenotype},
-and sometimes in contrast refer to $\kappa$ as the \emph{kryptotype},
-as it is ``hidden'' from direct selection.
-The rate at which the $i^\text{th}$ species is produced
-is a weighted sum of the concentrations of the other species
-as well as the input:
-\begin{align*}
-    \ddt \kappa_i(t) = \sum_j A_{ij} \kappa_j(t) + \sum_r B_{ir} u_r(t) .
-\end{align*}
-In matrix notation, this is written more concisely as
-\begin{align} \label{eqn:lti_system}
-    \ddt \kappa(t) &= A \kappa(t) + B u(t) \\
-    \phi(t) &= C \kappa(t) .
+%}
+Here we outline a method to model biological systems, such as gene regulatory networks, as linear dynamical systems.
+We define an organism's phenotype $\phi(t)$ to be the molecular time-dynamics directly relevant to survival.
+These dynamics are the result of the interconnections of a gene regulatory network (or some other biological system) $A$ (an $n \times n$ matrix) and an environmental input $u(t)$.
+Such a system $\mathcal{S}$, is composed of two equaitons,
+  \begin{align}
+    \mathcal{S} := \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right.
   \end{align}
+ Where the \emph{kryptotype} $\kappa(t)$, is a list of a system's molecular concentrations at time $t$ -- or more generally the systems \emph{internal dynamics}.
+As some system dynamics may not be visible nor relevant to selection, we distinguish the \emph{kryptotype} (as it is hidden) from the \emph{phenotype} -- which is the systems \emph{external dynamics}.
+  The change in internal molecular concentrations as a function of time $\dot{\kappa}(t)$ is simply a function of the systems current state (or kryptotype), the system architecture $A$ -- such as a gene regulatory network, the environmental input $u(t)$, and $B$ (an $n \times l$ matrix) -- how the system processes its environment.
+$C$ (an $l \times n$ matrix) filters the dynamics relevant to survival.
+  Thus the phenotype is simply a convolution of the system organization and the environment, 
+  \begin{align}
+    \phi(t) = \int_{0}^{t} C e^{A t} B u(t) dt ,
+  \end{align}
+  where we sometimes refer to $h(t) := Ce^{A t}B$ as the \emph{impulse response} of the system.
+As $A$ specifies gene regulatory network architecture, the $i^\text{th}$ row of $A$ can be interpreted as the \emph{cis}-regultory element (or promoter) for gene $i$.
+
+%We will now lay out the model in more general terms.
+%Suppose that the \emph{internal state} of the system
+%is parameterized by the concentrations of a collection of $n$ molecular species,
+%$S_1, \ldots, S_n$,
+%and the vector of concentrations at time $t$ we denote $\kappa(t)=(\kappa_1(t),\ldots,\kappa_n(t))$.
+%There are also $m$ ``input'' species, whose concentrations are determined
+%exogenously to the system,
+%and are denoted $u(t) = (u_1(t),\ldots,u_m(t))$,
+%and $\ell$ ``output'' species, whose concentrations are denoted
+%$\phi(t) = (\phi_1(t),\ldots,\phi_\ell(t))$.
+%The output is merely a linear function of the internal state:
+%\begin{align*}
+%    \phi_i(t) = \sum_j C_{ij} \kappa_i(t).
+%\end{align*}
+%Since $\phi$ is what natural selection acts on, we refer to it as the \emph{phenotype},
+%and sometimes in contrast refer to $\kappa$ as the \emph{kryptotype},
+%as it is ``hidden'' from direct selection.
+%The rate at which the $i^\text{th}$ species is produced
+%is a weighted sum of the concentrations of the other species
+%as well as the input:
+%\begin{align*}
+%    \ddt \kappa_i(t) = \sum_j A_{ij} \kappa_j(t) + \sum_r B_{ir} u_r(t) .
+%\end{align*}
+%In matrix notation, this is written more concisely as
+%\begin{align} \label{eqn:lti_system}
+%    \ddt \kappa(t) &= A \kappa(t) + B u(t) \\
+%    \phi(t) &= C \kappa(t) .
+%\end{align}
 
@@
 
-  \section*{The Model II: Linear Evolutionary Systems}
+  \section*{Linear Evolutionary Systems}
 
@@
 
-  Systems with identical external dynamics do not necessarily have identical internal dynamics.
-Any linear and minimal system -- minimal, informally meaning that the system's external dynamics are achieved with the fewest possible number of internal components -- has identical external dynamics up to a change of coordinates.
+  Gene regulatory networks with identical phenotypes (external dynamics) do not necessarily have identical kryptotypes (internal dynamics).
+Any linear and minimal system (a gene network) -- minimal, informally meaning that the system's phenotype is achieved with the fewest possible number of genes -- has an identical phenotype up to a change of coordinates.
   \begin{align}
@@
   \end{align}
-  Two systems, $\Sigma = \{ A, B, C \}$, and $\bar{\Sigma} = \{\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} \}$, have the same dynamics if they are related by a change of coordinates.
+  Two biological systems, $\mathcal{S} = \{ A, B, C \}$, and $\bar{\mathcal{S}} = \{\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} \}$, have the same phenotype if they are related by a change of coordinates.
 
@@
   \end{align}
-  In other words systems with varying genetic architectures yet identical selection pressures, environment, and phenotype.
+  In other words systems with varying gene regulatory network architectures yet identical selection pressures, environment, and phenotype.
 
@@
   \begin{align}
-    \Sigma(\tau) = \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A(\tau) \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right .
+    \mathcal{S}(\tau) := \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A(\tau) \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right .
   \end{align}
-  That is, all (linear and minimal) mechanisms capable of producing the same phenotype can be realized by a unique choice of $\tau$ in $\Sigma(\tau)$.
+  That is, all (linear and minimal) mechanisms capable of producing the same phenotype can be realized by a unique choice of $\tau$ in $\mathcal{S}(\tau)$.
 
@@
   Usually, the dimension $n$ and the reference system $A_0$ is implicit and we write only $\calA$.
+
+  Regardless of minimality, two systems, even in different dimensions, can have identical external dynamics if they are in $\mathcal{A}$.
+This set can be completely parameterized using the \emph{Kalman Decomposition}; the set of all linear gene regulatory networks with equivalent phenotypes can be precisely defined.
+Note that this implies that there is always more than one possible gene regulatory network architecture per phenotype, however we can analytically describe the complete set.
 
@@
 
-  \plr{rework this: we assume optimal phenotype is fixed and population is large but polymorphic and not strictly on the optimum}
-
-First we focus on a simple evolutionary scenario -- \emph{phenotypically invariant evolution}: a large population, perfectly adapted, and in a constant environment.
+%  \plr{rework this: we assume optimal phenotype is fixed and population is large but polymorphic and not strictly on the optimum}
+
+%First we focus on a simple evolutionary scenario -- \emph{phenotypically invariant evolution}: a large population, perfectly adapted, and in a constant environment.
 In this circumstance we expect phenotype to be conserved throughout evolutionary time.
@@
 Genetic drift registers as an increase in the intrapopulation variation in $h(t)$.
- \jss{added the phenotypic invariance stuff here (above) -- not sure if it should put elsewhere yet}
+% \jss{added the phenotypic invariance stuff here (above) -- not sure if it should put elsewhere yet}
 
@@
   \begin{align}
-      \E\left[ \|A_t - A_0\|^2 \right] = m \sigma^2 t + O(t^2) .
+      \E\left[ \|A_t - A_0\|^2 \right] = m \sigma^2 T + O(T^2) .
   \end{align}
 
-  \begin{align}
-    A(\tau) \xrightarrow{T} A(\tau + \epsilon)
-  \end{align}
-  After evolutionary time $T$, the population's gene network architecture evolves from $A(\tau)$ to $A(\tau + \epsilon)$ with a probability inversely proportional to the magnitude of $\epsilon$ and proportional to the magnitude of $T$.
-\jss{Change in $\tau$ is tracking movement of population mean, not ``macro'' evolutionary change, as interpreted previously.}
-  \begin{align}
-    \Delta_{\tau} &= \left\lVert \frac{d}{d \tau} \text{vec} \left[A\left(\tau\right)\right] \right\rVert^{-1} \\
-    A(\tau) &\xrightarrow{\mu} A(\tau \pm \mu \Delta_{\tau})
-  \end{align}
+  %\begin{align}
+  %  \Delta_{\tau} &= \left\lVert \frac{d}{d \tau} \text{vec} \left[A\left(\tau\right)\right] \right\rVert^{-1} \\
+  %  A(\tau) &\xrightarrow{\mu} A(\tau \pm \mu \Delta_{\tau})
+  %\end{align}
  % \section*{Phenotypic Invariance}
 
-    \begin{example}[Not all minimal gene networks can drift]
-      If a gene network is minimal and all the molecular species involved in the network are under selection, such that $C$ is the $n \times n$ identity matrix  ($C = I_{n}$), the only acceptable  change of coordinate matrix is the identify matrix.
+%    \begin{example}[Not all minimal gene networks can drift]
+ %     If a gene network is minimal and all the molecular species involved in the network are under selection, such that $C$ is the $n \times n$ identity matrix  ($C = I_{n}$), the only acceptable  change of coordinate matrix is the identify matrix.
 
@@
    %   \end{align}
-      \end{example}
-      \begin{example}[All Non-Minimal Gene Networks can Drift]
-
-      Despite the existence of a unique genetic architecture in the minimal case, there still exists an infinite number of systems with larger networks that have identical external dynamics. 
-      \begin{align}
-        h(t) &= \widehat{h}(t) \iff \\ 
-        CA^{\jmath}B &= \widehat{C} \widehat{A}^{\jmath} \widehat{C} \qquad \text{for } \jmath = 0, 1, \dots
-      \end{align}
-      Any two systems with equivalent impulse responses will have equivalent phenotypes.
-      \end{example}
+ %     \end{example}
+  %    \begin{example}[All Non-Minimal Gene Networks can Drift]
+  %    
+  %    Despite the existence of a unique genetic architecture in the minimal case, there still exists an infinite number of systems with larger networks that have identical external dynamics. 
+  %    \begin{align}
+  %      h(t) &= \widehat{h}(t) \iff \\ 
+  %      CA^{\jmath}B &= \widehat{C} \widehat{A}^{\jmath} \widehat{C} \qquad \text{for } \jmath = 0, 1, \dots
+  %    \end{align}
+  %    Any two systems with equivalent impulse responses will have equivalent phenotypes.
+  %    \end{example}
 
@@
         \end{figure}
+
+        \jss{$A(2)$ and $A(2.1)$ differ in regulatory interaction strengths by $5.09\%$.
+If intrapopulation regulatory variation is approximately $5\%$, then this level of divergence is expected after $\sqrt{\frac{t}{N_{e}} = 1}$ generations.}
     \end{example}
-      \begin{example}[Not all Networks can Host Incompatibilities]
-        convex sets cant have DMIs
-
-        \begin{align*}
-          h(t) = 2 e^{- \theta t}
-        \end{align*}
-        Any non-minimal system with rows summing to $\theta$ is PI.
+ %     \begin{example}[Not all Networks can Host Incompatibilities]
+ %       convex sets cant have DMIs
+
+ %       \begin{align*}
+ %         h(t) = 2 e^{- \theta t}
+ %       \end{align*}
+ %       Any non-minimal system with rows summing to $\theta$ is PI.
 Further, these systems are closed under averaging (mating) and row swapping (meiosis), leaving all hybrids optimally fit.
 The set of gene matrices is affine and therefore convex.
-      \end{example}
-
-  \section*{Additionl Examples}
-
-    \subsection*{Metabolic Network}
-  \begin{align*}
-    \dot{x}(t) &= \begin{bmatrix} 0 & 1 & 0 & -1 \\ 0 & 0 & -1 & -1 \\ -1 & 1 & 0 & 0 \\ 0 & 0 & 0 & 0 \end{bmatrix} \begin{bmatrix} \text{GAL3} \\ \text{GAL4} \\ \text{GAL80} \\ \text{MIG1} \end{bmatrix}(t) \\ &+ \begin{bmatrix} 1 & 0 & 0 \\ 0 & 0 & 1 \\ 0 & 0 & 0 \\ 0 & 1 & 1 \end{bmatrix} \begin{bmatrix} \text{galactose} \\ \text{glucose} \\ \text{BTM} \end{bmatrix}(t) \\
-        y(t) &= \begin{bmatrix} 0 & 1 & 0 & -1 \end{bmatrix} \vec{x}(t)
-    \end{align*}
-    \begin{align*}
-      H(z) &= \begin{bmatrix} 1 \\ z^{2} + z + 1 \\ z + 1 \end{bmatrix} \frac{1}{z^{3} + z -1}
-    \end{align*}
-
-    \begin{align*}
-      CV &= C \\
-      VB &= B \\
-      V :&= \begin{bmatrix} 1 & 0 & a & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & b & 0 \\ 0 & 0 & 0 & 1 \end{bmatrix} , V^{-1} = \begin{bmatrix} 1 & 0 & \frac{-a}{b} & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & \frac{1}{b} & 0 \\ 0 & 0 & 0 & 1\end{bmatrix} \\
-        A(a,b) :&= VAV^{-1} = \begin{bmatrix} -a & a+1 & \frac{a^{2}}{b} & -1 \\ 0 & 0 & \frac{-1}{b} & -1 \\ -b & b & a & 0 \\ 0 & 0 & 0 & 0 \end{bmatrix}
-    \end{align*}
-    The GAL regulatory network in \emph{S.
+ %     \end{example}
+
+\section*{The rate of speciation under neutrality in allopatric populations}
+   In general terms, if fitness $\mathcal{F}$ is some function $g$ of phenotypic divergence from an optimal scaled by the strength of selection $\beta$,
+   \begin{align*}
+      \mathcal{F} = g \left( \frac{\phi - \phi_{0}}{\beta} \right)
+    \end{align*}
+    And a population $P$ is normally distributed around a mean gene network architecture $A(\tau)$ and $\sigma^{2}$ is the intrapopulation regulatory variance,
+    \begin{align*}
+      P_{T} \sim \mathcal{N} \left(A(\tau), T \frac{\sigma^{2}}{N_{e}}\right)
+    \end{align*}
+    the population mean architecture will move at rate
+    \begin{align*}
+      \Delta \tau = \sigma \sqrt{\frac{T}{N_{e}}}
+    \end{align*}
+    where $T$ is time measured in generations. 
+    USING GENERAL MATH FROM MULTIVARIATE TAYLOR SERIES....
+we know that the phenotype of a first generation $F_{1}$ hybrid diverges from that of its parents (assuming both parents are at an optimum and phenotypically equivalent) quartically as a function of $\Delta \tau$.
+    Thus the squared difference of it's phenotype will be,
+    \begin{align*}
+      \Delta \Phi^{2}_{1} = c_{1}^{2} \sigma^{4} \left( \frac{T}{N_{e}} \right)^{2}
+    \end{align*}
+    and $F_{1}$ fitness will decline linearly with respect to $\frac{T}{N_{e}}$. ARGUMENT FOR WHY $\sigma^{2} c_{1} \sim \beta$ HERE.
+       \begin{align*}
+         \Delta \mathcal{F}_{1} = \left( \frac{\Delta \Phi_{1}}{\beta} = \frac{c_{1} \sigma^{2} \frac{T}{N_{e}}}{\beta} \approx \frac{T}{N_{e}} \right)
+    \end{align*}
+     It follows FROM THE SAME MATH ABOUT THE MULTIVAR TAYLOR that the phenotypes of second generation $F_{2}$ hybrid crosses diverges quadratically as a function of $\Delta \tau$, 
+    \begin{align*}
+      \Delta \Phi^{2}_{2} &= c_{2}^{2} \left( \sigma \sqrt{\frac{T}{N_{e}}} \right)^{2} = c_{2}^{2} \frac{T}{N_{e}} \\
+      \Delta \mathcal{F}_{2} &= \left( \frac{\Delta \Phi_{2}}{\beta} = \frac{c_{2} \sigma \sqrt{\frac{T}{N_{e}}}}{\beta} \approx \sqrt{\frac{T}{N_{e}}} \right)
+    \end{align*}
+    and that $F_{2}$ fitness drops rapidly -- at approximately rate $\sqrt{\frac{T}{N_{e}}}$. ARGUMENT FOR WHY $c_{2} \sigma \sim \beta$ HERE.
+
+    This suggests that reproductively isolated populations can speciate rapidly -- on timescales much less than $N_{e}$ generations, depending on the specifics of the fitness function.
+
+  \section*{Additional Examples}
+
+%    \subsection*{Metabolic Network}
+%  \begin{align*}
+%    \dot{x}(t) &= \begin{bmatrix} 0 & 1 & 0 & -1 \\ 0 & 0 & -1 & -1 \\ -1 & 1 & 0 & 0 \\ 0 & 0 & 0 & 0 \end{bmatrix} \begin{bmatrix} \text{GAL3} \\ \text{GAL4} \\ \text{GAL80} \\ \text{MIG1} \end{bmatrix}(t) \\ &+ \begin{bmatrix} 1 & 0 & 0 \\ 0 & 0 & 1 \\ 0 & 0 & 0 \\ 0 & 1 & 1 \end{bmatrix} \begin{bmatrix} \text{galactose} \\ \text{glucose} \\ \text{BTM} \end{bmatrix}(t) \\
+%        y(t) &= \begin{bmatrix} 0 & 1 & 0 & -1 \end{bmatrix} \vec{x}(t)
+%    \end{align*}
+%    \begin{align*}
+%      H(z) &= \begin{bmatrix} 1 \\ z^{2} + z + 1 \\ z + 1 \end{bmatrix} \frac{1}{z^{3} + z -1}
+%    \end{align*}
+%
+%    \begin{align*}
+%      CV &= C \\
+%      VB &= B \\
+%      V :&= \begin{bmatrix} 1 & 0 & a & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & b & 0 \\ 0 & 0 & 0 & 1 \end{bmatrix} , V^{-1} = \begin{bmatrix} 1 & 0 & \frac{-a}{b} & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & \frac{1}{b} & 0 \\ 0 & 0 & 0 & 1\end{bmatrix} \\
+%        A(a,b) :&= VAV^{-1} = \begin{bmatrix} -a & a+1 & \frac{a^{2}}{b} & -1 \\ 0 & 0 & \frac{-1}{b} & -1 \\ -b & b & a & 0 \\ 0 & 0 & 0 & 0 \end{bmatrix}
+%    \end{align*}
+%    The GAL regulatory network in \emph{S.
 cerevisiae} is modelled here.
@@
 This network is one of the most studied gene regulatory networks in yeast, and experiments have already demonstrated significant transcriptional variation among different species and genuses of yeast.
-
-    In the present model, the $A$ matrix encodes the interactions among GAL3, GAL4, GAL80, and MIG1.
+%
+%    In the present model, the $A$ matrix encodes the interactions among GAL3, GAL4, GAL80, and MIG1.
 $B$ interprets the input: the quantity of glucose, galactose, and the basal transcription (BTM) rates impact of TF concentrations.
 $C$ represents the promoter region of the GAL regulon and is the sum of the influence of GAL4 (activating) and MIG1 (repressing).
-
-    $A(a,b)$ represent alternative network structures that produce identical outputs, with identical input and output transformations -- meaning the regulatory contributions of the sugars and the basasl machinery, as well as the promoter of the GAL regulon are held constant.
+%
+%    $A(a,b)$ represent alternative network structures that produce identical outputs, with identical input and output transformations -- meaning the regulatory contributions of the sugars and the basasl machinery, as well as the promoter of the GAL regulon are held constant.
 These interactions may be realized in nature, if the regulatory differences predicted by $A(a,b)$ are biologically/mutationally realizable.
 Mutations influencing the interactions between GAL4 and GAL80 have been experimentally demonstrated \citep{li2010alterations}.
-
-    \subsection*{Gap Gene Network}
+%
+%    \subsection*{Gap Gene Network}
 
@@
 We emphasize that these incompatibilities are a consequence of recombining different, yet functionally equivalent, mechanisms.
+
+  Furthermore, using a quantitative genetic approach, we estimated that a genetically variable population will drift in neutral system space at a rate determined by its intrapopulation variation and its effective population size.
+Because mechanistically distinct yet phenotypically equivalent biological systems can fail to produce viable hybrids, we predict allopatric populations to speciate at a rate on the order of $N_{e}$ under reasonable population genetic parameter estimates.
+Additionally we see second-generation hybrid fitness plummet much faster than that of first-generation hybrids.
+This insight is a consequence of combinding our mechanistic model with a quantitative genetic one: we observe that $F_{1}$ phenotypes diverge quartically, and $F_{2}$ phenotypess quadratically, with evolutionary time.
+This result is also consistent with Haldane's rule; that if only one hybrid sex is inviable or sterile it is likely the heterogametic sex.
+The consistency comes from gene networks localized to the sex chromosomes functioning as an $F_{2}$ hybrid cross within a diploid $F_{1}$ heterogamete as there is only one sex chromosome.
+
   We also suggest that gene networks may not always use their components parsimoniously as network size tends to ratchet up in the absence of strong selection against extra parts.
@@
 
-  This theoretical framework can easily be applied to other interesting questions in evolutionary biology not tackled presently: such as the evolution of linkage, the necessity of network complexity (does evolution tend towards Rube Goldberg or \jss{parsimonious} network organization?), evolvability, structure/function inference, and intrapopulation context dependency of mutational effects, as well as many others.
+  This theoretical framework can easily be applied to other interesting questions in evolutionary biology not tackled presently: such as the evolution of linkage, the necessity of network complexity (does evolution tend towards Rube Goldberg or parsimonious network organization?), evolvability, structure/function inference, and intrapopulation context dependency of mutational effects, as well as many others.
 
@@
 \appendix
+
+\section{Kalman Decomposition}
+\begin{definition}[Phenotypic equivalence of systems]
+    Let $(\kappa(t),\phi(t))$ and $(\bar \kappa(t),\bar \phi(t))$ be the solutions to \eqref{eqn:lti_system}
+    with coefficient matrices $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ respectively,
+    and both $\kappa(0)$ and $\bar \kappa(0)$ are zero. 
+    The systems defined by $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ are
+    \textbf{phenotypically equivalent} 
+    if
+    \begin{align*}
+        \phi(t) = \bar \phi(t) \qquad \text{for all} \; t \ge 0.
+    \end{align*}
+    Equivalently, this occurs if and only if
+    \begin{align*}
+        h(t) = \bar h(t)  \qquad \text{for all} \; t \ge 0,
+    \end{align*}
+    where $h$ and $\bar h$ are the impulse responses of the two systems.
+\end{definition}
+
+One way to find other systems equvalent to a given one
+is by change of coordinates (``algebraic equivalance''):
+if $V$ is an invertible matrix, then the systems $(A,B,C)$ and $(VAV^{-1},VB,CV^{-1})$
+have the same dyanamics because their transfer functions are equal:
+\begin{align*}
+    CV^{-1}( zI - VAV^{-1})^{-1}VB
+    =
+    CV^{-1}V( zI - A)^{-1}V^{-1}VB
+    =
+    C( zI - A)^{-1}B .
+\end{align*}
+However, the converse is not necessarily true: 
+systems can have identical transfer functions without being changes of coordinates of each other.
+In fact, systems with identical transfer functions can involve interactions between different
+numbers of molecular species.
+
+The set of all systems phenotypically equivalent to a given system $(A,B,C)$ 
+is elegantly described using the Kalman decomposition,
+which also clarifies the system dynamics? tells us a lot about how it works? \plr{or something}
+To motivate this, first note that the the input $u(t)$ only directly pushes the system
+in directions lying in the span of the columns of $B$.
+As a result, different combinations of input can 
+move the system in any direction that lies in the \emph{reachable subspace},
+which we denote by $\reachable$,
+and is defined to be the closure of $\spn(B)$ under applying $A$
+(or equivalently, the span of $B, AB, A^2B, \ldots A^{n-1}B$).
+Analogously to this, we define
+the \emph{observable subspace}, $\mathcal{O}$,
+to be the closure of $\spn(C^T)$ under applying $A$.
+(Or: $\unobservable$ is the largest $A$-invariant subspace
+contained in the null space of $C$;
+and $\reachable$ is the largest $A$-invariant subspace contained in the image of $B$.)
+
+If we define
+\begin{enumerate}
+    \item The columns of $P_\rno$ are an orthonormal basis for $\reachable \cap \unobservable$.
+    \item The columns of $P_\ro$ are an orthonormal basis of
+        the complement of $\reachable \cap \unobservable$ in $\reachable$.
+    \item The columns of $P_\nro$ are an orthonormal basis of
+        the complement of $\reachable \cap \unobservable$ in $\unobservable$.
+    \item The columns of $P_\nrno$ are an orthonormal basis of
+        the the remainder of $\R^n$.
+\end{enumerate}
+If we then define
+\begin{align*}
+    P &= 
+    \left[ \begin{array}{c|c|c|c}
+        P_\rno & P_\ro & P_\nro & P_\nrno
+    \end{array} \right] ,
+\end{align*}
+then
+\begin{align*}
+    P^T P
+    &=
+    \left[ \begin{array}{c|c|c|c}
+        I & 0 & 0 & 0 \\
+        \hline
+        0 & I & U & 0 \\
+        \hline
+        0 & V & I & 0 \\
+        \hline
+        0 & 0 & 0 & I 
+    \end{array} \right] .
+\end{align*}
+\plr{Check this.  Can we get $U=V=0$?}
+
+The following theorem can be found in SOME REFERENCE.
+
+\begin{theorem}[Kalman decomposition] \label{thm:kalman}
+        For any system $(A,B,C)$ with corresponding Kalman basis matrix $P$,
+        the transformed system $(PAP^{-1},PB,CP^{-1})$  has the following form:
+        \begin{align*}
+            \widehat A = PAP^{-1}
+            &=
+            \left[ \begin{array}{cccc}
+                A_{\rno} & A_{\rno,\ro} & A_{\rno,\nrno} & A_{\rno,\nro} \\
+                0 & A_{\ro} & 0 & A_{\ro,\nro} \\
+                0 & 0 & A_{\nrno} & A_{\nrno,\nro} \\
+                0 & 0 & 0 & A_{\nro}
+            \end{array} \right] ,
+        \end{align*}
+        and
+        \begin{align*}
+            \widehat B = PB
+            &=
+            \left[ \begin{array}{cccc}
+                B_{\rno} \\
+                B_{\ro} \\
+                0 \\
+                0 
+            \end{array} \right] ,
+        \end{align*}
+        and
+        \begin{align*}
+            \widehat C = CP^{-1}
+            &=
+            \left[ \begin{array}{cccc}
+                0 & C_{\ro} & C_{\nrno} & 0 
+            \end{array} \right] .
+        \end{align*}
+        The transfer function of both systems is given by
+        \begin{align*}
+            H(z) = C_{\ro} ( zI - A_{\ro} )^{-1} B_{\ro} .
+        \end{align*}
+\end{theorem}
+
+In the latter case, we say that the system is \emph{minimal} 
+-- there is no equivalent system with a smaller number of species.
+Note that this says that any two equivalent minimal systems
+are changes of basis of each other.
+
+Since any system can be put into this form,
+and once in this form, its transfer function is determined only by 
+$C_{\ro}$, $A_{\ro}$, and $B_\ro$,
+therefore, the set of all equivalent systems are parameterized by
+the dimension $n$,
+the choice of basis ($P$),
+the remaining submatrices in $\widehat A$, $\widehat B$, and $\widehat C$
+(which are unconstrained),
+and a invertible transformation of $\spn(P_{\ro})$, which we call $T_\ro$.
+
+\begin{theorem}[Parameterization of equivalent systems]
+    Let $(A,B,C)$ be a minimal system.
+    \begin{itemize}
+        \item[(a)]
+            Every equivalent system is of the form given in Theorem \ref{thm:kalman},
+            i.e., can be specified by choosing a dimension, $n$;
+            submatrices in $\widehat A$, $\widehat B$, and $\widehat C$ 
+            except for $A_\ro=A$, $B_\ro=B$, and $C_\ro=C$;
+            and choosing an invertible matrix $P$.
+
+        \item[(b)]
+            \plr{conjecture:}
+            The parameterization is unique
+            if $P$ is futhermore chosen so that 
+            each $P_x$ other than $P_\ro$ is a projection matrix,
+            and that 
+            \begin{align*}
+                0
+                =
+                P_x^T P_y
+            \end{align*}
+            for all $(x,y)$ except $(\ro,\nrno)$.
+
+    \end{itemize} 
+\end{theorem}
+
+\plr{Another way of saying it: pick the $\reachable$ and $\unobservable$ subspaces,
+that must intersect in something of the minimal dimension;
+then let $P$ be the appropriate basis?}
+
+In some situations we may be interested in only ``network rewiring'',
+where $A$ changes while $B$ and $C$ do not.
+For instance, 
+if all non-regulatory functions of each molecule are strongly constrained,
+then $C$ cannot change.
+Likewise, if responses of each molecule to the external inputs are not changed by evolution,
+then $B$ does not change.
 

### 2017-07-28 handnotes.tex
started a metabolic example

@@
 
+  \subsection*{Metabolic network}
+
+  Consider an organism that can metabolize two different sugars $s_{1}$ and $s_{2}$ (present at logarithmic concentrations $u_{1}$ and $u_{2}$ respectively), with enzymes $e_{1}$ and $e_{2}$ (with log concentration denoted as $\phi_{1}$ and $\phi_{2}$).
+Further suppose that one of the sugars $s_{2}$ is the preferred energy source (perhaps it contains significantly more energy than the other sugar or is otherwise more efficiently metabolized).
+The organism may have a gene regulatory network $\mathcal{S}$ that can depoly a situation specific metabolic strategy.
+That is depending on both $u{2}$ and $u_{1}$ the organism synthesize an appropriate $\phi_{1}$ and $\phi_{2}$.
+Furthermore, consider this system to contain at least two transcription factors, whose log concentrations are given by $\kappa_{1}$, $\kappa_{2}$, $\dots \kappa_{n}$.
+
+  Minimally such a system may have the architecture,
+  \begin{align*}
+    \mathcal{S} = \left\{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 0 & -1 \\ 0 & 0 \end{bmatrix} \begin{bmatrix} \kappa_{1} \\ \kappa_{2} \end{bmatrix} (t) + \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \begin{bmatrix} u_{1} \\ u_{2} \end{bmatrix} (t) \\[11pt]
+      \phi(t) &= \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \begin{bmatrix} \kappa_{1} \\ \kappa_{2} \end{bmatrix} (t) \end{array} \right.
+  \end{align*}
+  $\mathcal{S}$ is minimal as its reachability and observability matrices both have rank $=2$, $\text{rank}(\mathcal{R}) = \text{rank}(\mathcal{O}) = 2 = n \implies \min$.
+
+  The impulse response matrix of this system is, 
+  \begin{align*}
+    h(t) = \begin{bmatrix} 1 & - t \\ 0 & 1 \end{bmatrix}  .
+  \end{align*}
+
+  The phenotype is, 
+  \begin{align*}
+    \phi(t) = \exp \left\{ \kappa(0) + \int_{0}^{t} h(t) u(t) dt \right\} 
+  \end{align*}
+  ($\kappa(0)$ can be set to something like $\left[ -10, -10 \right]$, assuming the transient transcription factor and enzyme concentrations in the organism are typically quite low).
+
+  We can see that changing coordinates on this system will find new system architecutres, as shown before, however, we can also apply the Kalman decomposition to find systems in different dimensions -- that is systems that employ more than $2$ transcription factors, yet have the same external dynamics/phenotype.
+
+  DO KALMAN EXAMPLE HERE.
+
 %    \subsection*{Metabolic Network}

### 2017-07-31 handnotes.tex
metabolic example with Kalman decomp

@@
 The organism may have a gene regulatory network $\mathcal{S}$ that can depoly a situation specific metabolic strategy.
-That is depending on both $u{2}$ and $u_{1}$ the organism synthesize an appropriate $\phi_{1}$ and $\phi_{2}$.
+That is depending on both $u_{2}$ and $u_{1}$ the organism will synthesize an appropriate $\phi_{1}$ and $\phi_{2}$.
 Furthermore, consider this system to contain at least two transcription factors, whose log concentrations are given by $\kappa_{1}$, $\kappa_{2}$, $\dots \kappa_{n}$.
@@
 
-  We can see that changing coordinates on this system will find new system architecutres, as shown before, however, we can also apply the Kalman decomposition to find systems in different dimensions -- that is systems that employ more than $2$ transcription factors, yet have the same external dynamics/phenotype.
-
-  DO KALMAN EXAMPLE HERE.
-
+  We can see that changing coordinates on this system (with any $V \in \mathbb{R}^{2 \times 2}$) will find new system architecutres, as shown before, however, we can also apply the Kalman decomposition to find systems in different dimensions -- that is systems that employ more than $2$ transcription factors, yet have the same external dynamics/phenotype.
+This may happen if a system is co-opted from another, or may be a consequence of gene duplication and deletion.
+
+  $\widehat{\mathcal{S}}$ is $4$-dimensional biological system,  \begin{align*}
+    \widehat{\mathcal{S}} = \left\{ \begin{array}{ll}
+      \widehat{A} = V \begin{bmatrix} X_{1} & X_{2} \\ 0 & A_{\min} \end{bmatrix} V^{-1}; \quad \widehat{B} = V \begin{bmatrix} X_{3} \\ B_{\min} \end{bmatrix}; \quad \widehat{C} = \begin{bmatrix} 0 & C_{\min} \end{bmatrix} V^{-1} \end{array} \right\}
+  \end{align*}
+     where $V$ is any invertible $4 \times 4$ matrix and $X_{j}$ is any $2 \times 2$ matrix.
+
+     So for example, some system can be wired as follows, and still be input-output equivalent to the minimal metabolic system $A_{\min}$:
+     \begin{align*}
+       A' &= \begin{bmatrix}
+         1.6923  &  1.5385 &  -2.6154 &  -2 \\
+         0.8462  &  0.7692 &  -1.3077 &  -1 \\
+         1.2692  &  1.1538 &  -1.9615 &  -1.5 \\
+         0.4231  &  0.3846 &  -0.6538 &  -0.5
+       \end{bmatrix} \\
+       B' &= \begin{bmatrix} 4 & 4 \\ 2 & 2 \\ 3 & 3 \\ 1 & 3 \end{bmatrix} \\
+         C' &= \begin{bmatrix}  0.2692  &  0.1538  &  0.0385 &  -0.5 \\
+         -0.4231 &  -0.3846 &   0.6538  &  0.5 \end{bmatrix}
+     \end{align*}
+
+     Despite the present example consisting of a minimal $2 \times 2$ and a non-minimal $4 \times 4$ system, any $n$-dimensional system can be constructed using this method -- applying a change of coordinates to the Kalman decomposition -- to construct a mechanistically different system with identical phenotypic dynamics.
+Depending on the specifics of a system being modeled, one may have to take care to restrict the free parameter values and network architecutres to be biologically appropriate.
 %    \subsection*{Metabolic Network}

### 2017-07-31 handnotes.tex
metabolic example with Kalman decomp

@@
 
-  \section*{Additional Examples}
-
-  \subsection*{Metabolic network}
+  \begin{example}[Metabolic network]
 
@@
   \begin{align*}
-    \mathcal{S} = \left\{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 0 & -1 \\ 0 & 0 \end{bmatrix} \begin{bmatrix} \kappa_{1} \\ \kappa_{2} \end{bmatrix} (t) + \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \begin{bmatrix} u_{1} \\ u_{2} \end{bmatrix} (t) \\[11pt]
+    \mathcal{S}_{\min} = \left\{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 0 & -1 \\ 0 & 0 \end{bmatrix} \begin{bmatrix} \kappa_{1} \\ \kappa_{2} \end{bmatrix} (t) + \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \begin{bmatrix} u_{1} \\ u_{2} \end{bmatrix} (t) \\[11pt]
       \phi(t) &= \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \begin{bmatrix} \kappa_{1} \\ \kappa_{2} \end{bmatrix} (t) \end{array} \right.
   \end{align*}
-  $\mathcal{S}$ is minimal as its reachability and observability matrices both have rank $=2$, $\text{rank}(\mathcal{R}) = \text{rank}(\mathcal{O}) = 2 = n \implies \min$.
+    $\mathcal{S}_{\min}$ is minimal as its reachability and observability matrices both have rank $=2$, $\text{rank}(\mathcal{R}) = \text{rank}(\mathcal{O}) = 2 = n \implies \min$.
 
@@
 
-  We can see that changing coordinates on this system (with any $V \in \mathbb{R}^{2 \times 2}$) will find new system architecutres, as shown before, however, we can also apply the Kalman decomposition to find systems in different dimensions -- that is systems that employ more than $2$ transcription factors, yet have the same external dynamics/phenotype.
+  We can see that changing coordinates on this system (with any invertible $V \in \mathbb{R}^{2 \times 2}$) will find new system architecutres, as shown before, however, we can also apply the Kalman decomposition to find systems in different dimensions -- that is systems that employ more than $2$ transcription factors, yet have the same external dynamics/phenotype.
 This may happen if a system is co-opted from another, or may be a consequence of gene duplication and deletion.
@@
 
-     So for example, some system can be wired as follows, and still be input-output equivalent to the minimal metabolic system $A_{\min}$:
+     So for example, some system can be wired as follows, and still be input-output equivalent to the minimal metabolic system $\mathcal{S}_{\min}$:
      \begin{align*}
@@
 Depending on the specifics of a system being modeled, one may have to take care to restrict the free parameter values and network architecutres to be biologically appropriate.
+  \end{example}
 %    \subsection*{Metabolic Network}

### 2017-07-31 handnotes.tex
edited abstract

@@
 \begin{abstract}
- % \textbf{
-We introduce an analytical theory to study the evolution of biological systems such as gene regulatory networks, applying insight and tools from control engineering, systems identification, and dynamical systems theory.
-We describe a null model of regulatory network evolution by analytically describing the set of all linear gene networks (of any size) that produce identical phenotypes – and the evolutionary paths connecting them.
-Under constant selection and environmental stasis, neutral evolution proceeds as a random walk through phenotypically identical network-space.
-Under neutral conditions, this model can describe the rates at which Dobzhansky-Muller incompatibilities arise in reproductively isolated populations.
-This analysis provides insight into the mechanisms and parameters important for understanding developmental systems drift and speciation.
-%}
+Here we introduce an analytical theory to study the evolution of biological systems, such as gene regulatory networks.
+The evolutionary conservation of phenotype under selective and environmental stasis does not necessitate conservation of the underlying mechanism, as distinct molecular pathways can realize identical phenotypes.
+Here we give an exact expression for the set of all linear mechanisms with identical phenotypes.
+We employ a quantiative genetic approach to model evolution under neutrality as a random process over the set of all phenotype-invariant mechanisms: only mutational tweaks to the pathway that leave the phenotype invariant are optimally fit.
+We show that there is never a unique linear system architecture for any phenotype and that the evolutionary exploration of these distinct and mutationally connected mechanisms can lead to the rapid accumulation of hybrid incompatibilities between allopatric populations and thus lead to the rapid formation of new species -- in fewer generations than there are breeding individuals in a population.
 \end{abstract}

### 2017-07-31 handnotes.tex
edited abstract

@@
 The evolutionary conservation of phenotype under selective and environmental stasis does not necessitate conservation of the underlying mechanism, as distinct molecular pathways can realize identical phenotypes.
-Here we give an exact expression for the set of all linear mechanisms with identical phenotypes.
+Here we give an exact expression for the set of all linear mechanisms with identical phenotypes, and expect evolution under neutrality to explore this set.
 We employ a quantiative genetic approach to model evolution under neutrality as a random process over the set of all phenotype-invariant mechanisms: only mutational tweaks to the pathway that leave the phenotype invariant are optimally fit.
@@
 
-Additional ideas to consider adding:
+  \plr{Additional ideas to consider adding:
 \begin{itemize}
-    \item Haldane's rule (easy point in discussion: makes F1s look like F2s)
-    \item add point about F1: quartic and F2: quadratic to results and maybe abstract/discussion
+    %\item Haldane's rule (easy point in discussion: makes F1s look like F2s)
+    %\item add point about F1: quartic and F2: quadratic to results and maybe abstract/discussion
     \item hybrid vigor (need to do calculation)
     \item discuss linearity, linearization, and canalization in introduction
-    \item obtain estimates of variation in $A$ from thermodynamic occupancy model
-\end{itemize}
+    %\item obtain estimates of variation in $A$ from thermodynamic occupancy model
+\end{itemize}}
 

### 2017-08-01 handnotes.tex
edited abstract and spelling

@@
 Here we give an exact expression for the set of all linear mechanisms with identical phenotypes, and expect evolution under neutrality to explore this set.
-We employ a quantiative genetic approach to model evolution under neutrality as a random process over the set of all phenotype-invariant mechanisms: only mutational tweaks to the pathway that leave the phenotype invariant are optimally fit.
+We employ a quantitative genetic approach to model evolution under neutrality as a random process over the set of all phenotype-invariant mechanisms: only mutational tweaks to the pathway that leave the phenotype invariant are optimally fit.
 We show that there is never a unique linear system architecture for any phenotype and that the evolutionary exploration of these distinct and mutationally connected mechanisms can lead to the rapid accumulation of hybrid incompatibilities between allopatric populations and thus lead to the rapid formation of new species -- in fewer generations than there are breeding individuals in a population.
@@
 can fundamentally be \emph{unidentifiable} and/or \emph{indistinguishable} -- meaning that 
-there can be uncertaintity about an inferred model's parameters or even its claims about
+there can be uncertainty about an inferred model's parameters or even its claims about
 causal structure, even with access to complete and perfect data \citep{bellman1970structural, grewal1976identifiability, walter1984structural}. Models with different 
@@
 In computer science, this is framed as the relationship among processes that simulate one another \citep{van2004equivalence}.
-Although this may frusturate the occasional engineer or scientist, viewed from another angle,
+Although this may frustrate the occasional engineer or scientist, viewed from another angle,
 the concepts of unidentifiability and indistinguishability can provide a starting point for
@@
 
-  We derive an analytical description of the set of all linear biological systems with identical phenotypes -- that is we decribe the set of all gene network architectures that yield identical phenotypes, and show that all biological systems can, in principal, can undergo systems drift.
-In the neutral case, this set desribes a manifold that evolution explores leaving phenotype invariant with respect to mutation, and predicts that if two populations become reproductively isolated, hybrid incompatibility can occur, despite the absence of adaptation, directional selection, or environmental change.
-Speciation typically occurs on timesclaes approximately on the order of $N_{e}$ generations, where $N_{e}$ is the effective population size.
+  We derive an analytical description of the set of all linear biological systems with identical phenotypes -- that is we describe the set of all gene network architectures that yield identical phenotypes, and show that all biological systems can, in principal, can undergo systems drift.
+In the neutral case, this set describes a manifold that evolution explores leaving phenotype invariant with respect to mutation, and predicts that if two populations become reproductively isolated, hybrid incompatibility can occur, despite the absence of adaptation, directional selection, or environmental change.
+Speciation typically occurs on timescales approximately on the order of $N_{e}$ generations, where $N_{e}$ is the effective population size.
 
@@
 These dynamics are the result of the interconnections of a gene regulatory network (or some other biological system) $A$ (an $n \times n$ matrix) and an environmental input $u(t)$.
-Such a system $\mathcal{S}$, is composed of two equaitons,
+Such a system $\mathcal{S}$, is composed of two equations,
   \begin{align}
@@
   where we sometimes refer to $h(t) := Ce^{A t}B$ as the \emph{impulse response} of the system.
-As $A$ specifies gene regulatory network architecture, the $i^\text{th}$ row of $A$ can be interpreted as the \emph{cis}-regultory element (or promoter) for gene $i$.
+As $A$ specifies gene regulatory network architecture, the $i^\text{th}$ row of $A$ can be interpreted as the \emph{cis}-regulatory element (or promoter) for gene $i$.
 
@@
     where $\dot{\kappa}$ denotes the time derivative.
-The initial conditions $\kappa_{1}(0)$ and $\kappa_{2}(0)$, and the input $u(t)$ then determiine the concentrations through time.
+The initial conditions $\kappa_{1}(0)$ and $\kappa_{2}(0)$, and the input $u(t)$ then determine the concentrations through time.
 If we record the regulatory coefficients in the matrix
@@
 
-      Since the input is simply an impulse, its phenotype is equivlanet to its impulse response
+      Since the input is simply an impulse, its phenotype is equivalent to its impulse response
 
@@
 There are examples of significant diversity in the networks underlying processes such as circadian rhythm \citep{sancar2008intelligent}, cell cycle control \citep{cross2011evolution, kearsey2003enigmatic}, pattern formation, and metabolism  \citep{lavoie2009rearrangements, martchenko2007transcriptional, dalal2016transcriptional, christensen2011unique, hartl2007induction, alam2013aspergillus}.
-Despite a symmetry in functionality or phenotype these systems often differ, sometimes substationaly, at the molecular level.
+Despite a symmetry in functionality or phenotype these systems often differ, sometimes substantially, at the molecular level.
 How many different mechanisms have the same function?
@@
 \end{tikzpicture}
-    \caption{A graphical depiction of the set of all externally equivalent cell cycle control networks, $A(\tau)$. $\tau$ can take any valye}
+    \caption{A graphical depiction of the set of all externally equivalent cell cycle control networks, $A(\tau)$. $\tau$ can take any value}
     \end{figure}
@@
   but also the effects of other transcription factors (e.g., cooperativity)
-  and local chromatin accessibliity \citep{stefflova2013cooperativity}.
+  and local chromatin accessibility \citep{stefflova2013cooperativity}.
   For this reason, 
@@
   \plr{strength of selection}
-  It seems certain that selection is not so strong that intrapopulation variation
+  It seems certain that selection is not so strong that intra-population variation
   is strongly deleterious,
@@
   taking $\sigma=.01$ as above, this suggests that changes in phenotype of 5\% are sufficient
-  to effect a noticable drop in fitness.
+  to effect a noticeable drop in fitness.
   \plr{BUT $\sigma$ IS VARIATION IN $A$ NOT PHENOTYPE}
@@
 
-  It therefore seems reasonable to coarsely model the time evolution of population variation in network coefficents as 
+  It therefore seems reasonable to coarsely model the time evolution of population variation in network coefficients as 
   (a) a ``cloud'' of width X about the population mean, 
@@
 
-        Thus if populations $1$ and $2$ are homogenous $A(0)$ and $A(2)$, respectively, we say that they are completely incompatible as $\mathcal{I} = 0$.
+        Thus if populations $1$ and $2$ are homogeneous $A(0)$ and $A(2)$, respectively, we say that they are completely incompatible as $\mathcal{I} = 0$.
       \end{definition}
@@
         \jss{$A(2)$ and $A(2.1)$ differ in regulatory interaction strengths by $5.09\%$.
-If intrapopulation regulatory variation is approximately $5\%$, then this level of divergence is expected after $\sqrt{\frac{t}{N_{e}} = 1}$ generations.}
+If intra-population regulatory variation is approximately $5\%$, then this level of divergence is expected after $\sqrt{\frac{t}{N_{e}} = 1}$ generations.}
     \end{example}
@@
     \end{align*}
-    And a population $P$ is normally distributed around a mean gene network architecture $A(\tau)$ and $\sigma^{2}$ is the intrapopulation regulatory variance,
+    And a population $P$ is normally distributed around a mean gene network architecture $A(\tau)$ and $\sigma^{2}$ is the intra-population regulatory variance,
     \begin{align*}
@@
 Further suppose that one of the sugars $s_{2}$ is the preferred energy source (perhaps it contains significantly more energy than the other sugar or is otherwise more efficiently metabolized).
-The organism may have a gene regulatory network $\mathcal{S}$ that can depoly a situation specific metabolic strategy.
+The organism may have a gene regulatory network $\mathcal{S}$ that can deploy a situation specific metabolic strategy.
 That is depending on both $u_{2}$ and $u_{1}$ the organism will synthesize an appropriate $\phi_{1}$ and $\phi_{2}$.
@@
 
-  We can see that changing coordinates on this system (with any invertible $V \in \mathbb{R}^{2 \times 2}$) will find new system architecutres, as shown before, however, we can also apply the Kalman decomposition to find systems in different dimensions -- that is systems that employ more than $2$ transcription factors, yet have the same external dynamics/phenotype.
+  We can see that changing coordinates on this system (with any invertible $V \in \mathbb{R}^{2 \times 2}$) will find new system architectures, as shown before, however, we can also apply the Kalman decomposition to find systems in different dimensions -- that is systems that employ more than $2$ transcription factors, yet have the same external dynamics/phenotype.
 This may happen if a system is co-opted from another, or may be a consequence of gene duplication and deletion.
@@
      Despite the present example consisting of a minimal $2 \times 2$ and a non-minimal $4 \times 4$ system, any $n$-dimensional system can be constructed using this method -- applying a change of coordinates to the Kalman decomposition -- to construct a mechanistically different system with identical phenotypic dynamics.
-Depending on the specifics of a system being modeled, one may have to take care to restrict the free parameter values and network architecutres to be biologically appropriate.
+Depending on the specifics of a system being modeled, one may have to take care to restrict the free parameter values and network architectures to be biologically appropriate.
   \end{example}
@@
 Here, we offer a tractable and simple model which includes these missing features.
-Further, we provide, in clear mathematical language, an analytical description of phenomena hitherto only discussed verbally and conceptualy (phenogenetic drift \citep{weiss2000phenogenetic}, developmental systems drift \citep{true2001developmental}, biological degeneracy \citep{edelman2001degeneracy}, \emph{etc.}).
+Further, we provide, in clear mathematical language, an analytical description of phenomena hitherto only discussed verbally and conceptually (phenogenetic drift \citep{weiss2000phenogenetic}, developmental systems drift \citep{true2001developmental}, biological degeneracy \citep{edelman2001degeneracy}, \emph{etc.}).
 The tractability and relative simplicity of this exposition enables the interested biologist to work out by hand, if desired, the dynamics of a genetic system, as well as perturbations to the system -- an attribute not likely to be found in less tractable models and simulations.
 
-  We have suggested an interpretation of system identification: to see it as an evolutionarily neutral manifold, and not simply a computational nuissance.
-We have demonstrated a method to analytically determine the set of all phenotypically invarint gene networks; by a simple change of coordinates in the minimal configuration, or more generally by applying the Kalman decomposition in higher dimensions.
+  We have suggested an interpretation of system identification: to see it as an evolutionarily neutral manifold, and not simply a computational nuisance.
+We have demonstrated a method to analytically determine the set of all phenotypically invariant gene networks; by a simple change of coordinates in the minimal configuration, or more generally by applying the Kalman decomposition in higher dimensions.
 Further, we emphasize that evolution proceeds through this high dimensional space as stochastic coordinate transformation, constrained by sexual reproduction and selection.
@@
 
-  Furthermore, using a quantitative genetic approach, we estimated that a genetically variable population will drift in neutral system space at a rate determined by its intrapopulation variation and its effective population size.
+  Furthermore, using a quantitative genetic approach, we estimated that a genetically variable population will drift in neutral system space at a rate determined by its intra-population variation and its effective population size.
 Because mechanistically distinct yet phenotypically equivalent biological systems can fail to produce viable hybrids, we predict allopatric populations to speciate at a rate on the order of $N_{e}$ under reasonable population genetic parameter estimates.
 Additionally we see second-generation hybrid fitness plummet much faster than that of first-generation hybrids.
-This insight is a consequence of combinding our mechanistic model with a quantitative genetic one: we observe that $F_{1}$ phenotypes diverge quartically, and $F_{2}$ phenotypess quadratically, with evolutionary time.
+This insight is a consequence of combining our mechanistic model with a quantitative genetic one: we observe that $F_{1}$ phenotypes diverge quartically, and $F_{2}$ phenotypes quadratically, with evolutionary time.
 This result is also consistent with Haldane's rule; that if only one hybrid sex is inviable or sterile it is likely the heterogametic sex.
@@
 
-  As Richard Levins opined, models in population biology face a tradeoff among precision, realism, and generality \citep{levins1966strategy}.
+  As Richard Levins opined, models in population biology face a trade-off among precision, realism, and generality \citep{levins1966strategy}.
 As Levins expects, any tractable and general model, such as the present one under discussion, will have limitations.
@@
 It is often stated that life is not linear.
-This is often true, however, many of the ideas developed here should be generalizable to nonlinear cases (multilinear systems, say).
-Futher, we see this as a necessary first step in the direction of more life-like nonlinear evolutionary systems theory.
-Depending on an actual biological system's particularities, its (potential) nonlinearity, may buffer or exacerbate effects elucidated in this paper, such as the aquisition of Dobzansky-Muller incompatibilities.
-
-  This theoretical framework can easily be applied to other interesting questions in evolutionary biology not tackled presently: such as the evolution of linkage, the necessity of network complexity (does evolution tend towards Rube Goldberg or parsimonious network organization?), evolvability, structure/function inference, and intrapopulation context dependency of mutational effects, as well as many others.
+This is often true, however, many of the ideas developed here should be generalizable to nonlinear cases (multi-linear systems, say).
+Further, we see this as a necessary first step in the direction of more life-like nonlinear evolutionary systems theory.
+Depending on an actual biological system's particularities, its (potential) non-linearity, may buffer or exacerbate effects elucidated in this paper, such as the acquisition of Dobzansky-Muller incompatibilities.
+
+  This theoretical framework can easily be applied to other interesting questions in evolutionary biology not tackled presently: such as the evolution of linkage, the necessity of network complexity (does evolution tend towards Rube Goldberg or parsimonious network organization?), evolvability, structure/function inference, and intra-population context dependency of mutational effects, as well as many others.
 
@@
 \citet{tulchinsky} simulated the evolution of a transcription factor and its binding site using a thermodynamic model.
-Their simulations suggest that the language by which a transcription factor recognizes its binding site can change, and potentially lead to hybird incomaptibility when allopatric populations employ divergent readout languages.
+Their simulations suggest that the language by which a transcription factor recognizes its binding site can change, and potentially lead to hybrid incompatibility when allopatric populations employ divergent readout languages.
 This study, despite looking at gene regulation, does not analyze overall gene network architecture -- as we do here -- it only looks at the expression level of a single gene.
-Furthermore, they report reproducitve isolation primarily following direcitonal selection for a change in expression levels in each allopatric population; the evidecne for reproductive isolation following balancing selection is much weaker.
+Furthermore, they report reproductive isolation primarily following directional selection for a change in expression levels in each allopatric population; the evidence for reproductive isolation following balancing selection is much weaker.
 Johnson and Porter 2000 did not observe any hybrid fitness declines under stabilizing selection -- only under directional selection.
-  Khatari et al, Tulchinsky et al, and Porter et al, all study hybrid incompatibility from a transcription factor/binding site interaction perspective, not from an overal network architecture perspective.
+  Khatari et al, Tulchinsky et al, and Porter et al, all study hybrid incompatibility from a transcription factor/binding site interaction perspective, not from an overall network architecture perspective.
 Palmer and Feldman only see hybrid incompatibility in constant environments if the parental populations are relatively poorly adapted initially.
@@
 
-One way to find other systems equvalent to a given one
-is by change of coordinates (``algebraic equivalance''):
+One way to find other systems equivalent to a given one
+is by change of coordinates (``algebraic equivalence''):
 if $V$ is an invertible matrix, then the systems $(A,B,C)$ and $(VAV^{-1},VB,CV^{-1})$
-have the same dyanamics because their transfer functions are equal:
+have the same dynamics because their transfer functions are equal:
 \begin{align*}
@@
 which also clarifies the system dynamics? tells us a lot about how it works? \plr{or something}
-To motivate this, first note that the the input $u(t)$ only directly pushes the system
+To motivate this, first note that the input $u(t)$ only directly pushes the system
 in directions lying in the span of the columns of $B$.
@@
     \item The columns of $P_\nrno$ are an orthonormal basis of
-        the the remainder of $\R^n$.
+        the remainder of $\R^n$.
 \end{enumerate}
@@
             The parameterization is unique
-            if $P$ is futhermore chosen so that 
+            if $P$ is furthermore chosen so that 
             each $P_x$ other than $P_\ro$ is a projection matrix,

### 2017-08-01 handnotes.tex
pushed missing figured.
edited the title

@@
 
-\section*{Gene regulatory network drift and speciation occurs rapidly under neutrality}
+\section*{Evolutionary conservation of phenotype does not entail conservation of the underlying molecular mechanism leading to the rapid formation of new species}
 {\centering
@@
 \begin{abstract}
-Here we introduce an analytical theory to study the evolution of biological systems, such as gene regulatory networks.
+We introduce an analytical theory to study the evolution of biological systems, such as gene regulatory networks.
 The evolutionary conservation of phenotype under selective and environmental stasis does not necessitate conservation of the underlying mechanism, as distinct molecular pathways can realize identical phenotypes.

### 2017-08-02 kalman_walk_paper.tex
first pass through with comments

--- 2017-08-01 handnotes.tex
@@
 Joshua S. Schiffman$^{\dagger}$ \qquad Peter L. Ralph$^{\dagger \ddagger}$ \\
-$^{\dagger}$University of Southern California, Los Angeles, California \qquad $^{\ddagger}$University of Oregon, Eugene, Oregon \\
-\texttt{jsschiff@usc.edu} \qquad \texttt{plr@uoregon.edu}
+$^{\dagger}$University of Southern California, Los Angeles, California \qquad 
+$^{\ddagger}$University of Oregon, Eugene, Oregon \\
+\texttt{jsschiff@usc.edu} \qquad 
+\texttt{plr@uoregon.edu}
 \\
 }
+
+% Other title ideas:
+% More than one way to grow a cat: regulatory network redundancy and speciation
+% 
+% Neutral network drift leads to incompatibilities on the same time scale as genetic drift
+% 
+% More than one way to grow a cat: can network drift lead to speciation?
+% 
+% How fast does network drift create incompatibilities?
+% 
+% Beyond the snowball: a quantitative model of incompatibility accumulation
+% 
+% Systems drift and speciation: more than one way to grow a cat
+% 
+% An explicit model of neutral regulatory network evolution, with applications
+% to the rate of accumulation of hybrid incompatibility
+%
 
@@
 
+\plr{Note: in the \LaTeX source I'm putting in semantic linebreaks, so it's easy to edit and move around phrases and ideas.}
+
 %%%%%%%%%%%%%%%%%%%%%%
@@
 Progress in a suite of biological subdisciplines is stalled by our general lack of understanding of this molecular machinery: with respect to both its function and evolution. 
-There does exist a growing body of experiment and data on the evolutionary histories and molecular characterizations of particular gene regulatory networks \citep{jaeger2011gap, davidson2006gene, israel2016comparative}, as well as thoughtful verbal and conceptual models \citep{true2001developmental, gwagner1, weiss2000phenogenetic, edelman2001degeneracy}.
+\plr{Saying "stalled" reflects negatively on other fields - rephrase?}
+There does exist a growing body of data on the evolutionary histories and molecular characterizations of particular gene regulatory networks \citep{jaeger2011gap, davidson2006gene, israel2016comparative}, as well as thoughtful verbal and conceptual models \citep{true2001developmental, gwagner1, weiss2000phenogenetic, edelman2001degeneracy}.
+\plr{I don't think the HW example will speak to our audience.  Also, better to say why math is good rather than why not-math is bad.}
 However, as Hardy and Weinberg taught us over a century ago, verbal theories are often insufficient, if not downright misleading \citep{hardy1908mendelian, weinberg1908vererbungsgesetze, servedio2014not}.
 This is especially pertinent given the staggering complexity and scope of contemporary research programs. 
-This outlook necessitates the advancement of conceptual frameworks of such precision, only mathematics will suffice, as models allow the development of concrete numerical predictions.
-Previously it has been suggested that any idealized study of evolution is incomplete without a mathematically sufficient description of the genotype, phenotype, and transformation from one to the other \citep{Lewontin1974genetic}.%Here we outline an analytical theory to study the evolution of biological systems, borrowing insight and methods from control engineering, systems identification, and dynamical systems theory.
+This outlook necessitates the advancement of conceptual frameworks of such precision, 
+only mathematics will suffice, as models allow the development of concrete numerical predictions. 
+Previously it has been suggested that any idealized study of evolution is incomplete 
+without a mathematically sufficient description of the genotype, phenotype, and transformation from one to the other \citep{Lewontin1974genetic}. 
+\plr{probably some more recent Wagner papers in this line as well?}
+%Here we outline an analytical theory to study the evolution of biological systems, borrowing insight and methods from control engineering, systems identification, and dynamical systems theory.
 
@@
 The molecular machinery, interacting with the environment, and bridging genotype to phenotype
-can be mathematically described as a dynamical system -- or a system of differential equations\citep{jaeger2015comet}.
+can be mathematically described as a dynamical system -- or a system of differential equations \citep{jaeger2015comet}.
  Movement in this direction is ongoing, as researchers have begun to study 
@@
 there can be uncertainty about an inferred model's parameters or even its claims about
-causal structure, even with access to complete and perfect data \citep{bellman1970structural, grewal1976identifiability, walter1984structural}. Models with different 
-parameter schemes, or even different mechanics can be equally accurate, but still not
-\emph{actually} agree with what is being modelled. In control theory, where electrical 
-circuits and mechanical systems are often the focus, it is understood that there can be an 
-infinite number of ``realizations'', or ways to reverse engineer the dynamics of a black box,
-even if all possible input and output experiments on the black box are performed \citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. In chemical
-engineering, those who study chemical reaction networks sometimes refer to the fundamental
+causal structure, even with access to complete and perfect data \citep{bellman1970structural, grewal1976identifiability, walter1984structural}. 
+\plr{not clear what you mean by "actually"}
+Models with different parameter schemes, or even different mechanics can be equally accurate, 
+but still not \emph{actually} agree with what is being modelled. 
+In control theory, where electrical circuits and mechanical systems are often the focus, 
+it is understood that there can be an infinite number of ``realizations'', 
+or ways to reverse engineer the dynamics of a black box,
+even if all possible input and output experiments on the black box are performed \citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. 
+In chemical engineering, those who study chemical reaction networks sometimes refer to the fundamental
 unidentifiability of these networks as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}. 
 In computer science, this is framed as the relationship among processes that simulate one another \citep{van2004equivalence}.  
+\plr{here you are jumping to evolution -- haven't said yet why evolution "can explore" these -- needs to say explicitly what we mean by neutral here}
 Although this may frustrate the occasional engineer or scientist, viewed from another angle,
@@
 as the parameters and structures can be realized biologically. In fact, evolutionary
-biologists who study convergent versus parallel evolution, homology, and analogy are very familiar with such functional symmetries; macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level \citep{true2001developmental, tsong2006evolution, hare2008sepsid, vierstra2014mouse, stergachis2014conservation, taylor2016diverse, matsui2015regulatory}.
+biologists who study convergent versus parallel evolution, homology, and analogy are very familiar with such functional symmetries; 
+macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level 
+\citep{true2001developmental, tsong2006evolution, hare2008sepsid, vierstra2014mouse, stergachis2014conservation, taylor2016diverse, matsui2015regulatory}.
 
 In this paper we outline a theoretical framework to study the evolution of biological systems. 
-We expect wide applicability to this approach (\emph{i.e.} non-neutral evolution), however presently, we focus solely on neutral evolution, that is where phenotype is conserved over evolutionary time.
-
-  We derive an analytical description of the set of all linear biological systems with identical phenotypes -- that is we describe the set of all gene network architectures that yield identical phenotypes, and show that all biological systems can, in principal, can undergo systems drift.
-In the neutral case, this set describes a manifold that evolution explores leaving phenotype invariant with respect to mutation, and predicts that if two populations become reproductively isolated, hybrid incompatibility can occur, despite the absence of adaptation, directional selection, or environmental change.
+\plr{rather than "we expect" maybe say that the framework could be applied to non-neutral evolution}
+We expect wide applicability to this approach (\emph{i.e.} non-neutral evolution), 
+however presently, we focus solely on neutral evolution, 
+\plr{here's the definition of neutral, but it's not quite right?}
+that is where phenotype is conserved over evolutionary time.
+
+\plr{We don't "derive" it, we point out that Kalman already derived it.}
+We derive an analytical description of the set of all linear biological systems with identical phenotypes 
+\plr{We need something here saying what we mean by "gene network architecture".  Also, I don't think we claim anything for "all biological systems".}
+-- that is we describe the set of all gene network architectures that yield identical phenotypes, and show that all biological systems can, in principal, can undergo systems drift. 
+In the neutral case, this set describes a manifold that evolution explores leaving phenotype invariant with respect to mutation, 
+and predicts that if two populations become reproductively isolated, hybrid incompatibility can occur, despite the absence of adaptation, directional selection, or environmental change. 
 Speciation typically occurs on timescales approximately on the order of $N_{e}$ generations, where $N_{e}$ is the effective population size.
@@
 \section*{Gene Networks as Linear Dynamical Systems}
+
+\plr{Below we mix general language with language specific to GRNs, like switching between "internal state" and "transcription factor concentration.  
+Rather than keep making comments like "(or some other system)",
+we could say up top that this applies to other situations, but to make it concrete we'll talk about regulatory networks.}
 
@@
 Here we outline a method to model biological systems, such as gene regulatory networks, as linear dynamical systems. 
-We define an organism's phenotype $\phi(t)$ to be the molecular time-dynamics directly relevant to survival.
-These dynamics are the result of the interconnections of a gene regulatory network (or some other biological system) $A$ (an $n \times n$ matrix) and an environmental input $u(t)$.
+We define an organism's ``phenotype'' to be the time dynamics of molecular concentrations directly relevant to survival, 
+\plr{tried to say it more precisely. better phrase than "component of phenotype"?}
+and denote by $\phi_k(t)$ the concentration of the $k^\text{th}$ component of phenotype at time $t$.
+These dynamics are the result of the interconnections of 
+a gene regulatory network (or some other biological system) $A$ (an $n \times n$ matrix) 
+and an environmental input $u(t)$.
+\plr{say what $u$ is (a vector)}
+\plr{say in words what $\kappa$ is before the equation.
 Such a system $\mathcal{S}$, is composed of two equations,
@@
 The change in internal molecular concentrations as a function of time $\dot{\kappa}(t)$ is simply a function of the systems current state (or kryptotype), the system architecture $A$ -- such as a gene regulatory network, the environmental input $u(t)$, and $B$ (an $n \times l$ matrix) -- how the system processes its environment.
+\plr{Could say instead of "filters" that e.g. $C_{ij}$ is the amount by which the $j$th input $u_j$ affects the production of the $i$th transcription factor concentration.}
 $C$ (an $l \times n$ matrix) filters the dynamics relevant to survival.
+\plr{This "Thus" needs expanding, e.g., "the solution to this equation is unique and given by" with a reference.}
 Thus the phenotype is simply a convolution of the system organization and the environment, 
+\plr{You've missed out $\kappa(0)$. Also, you can't have $t$ be both a limit of the integral and the variable being integrated over -- check this and get it right.}
   \begin{align}
@@
 where we sometimes refer to $h(t) := Ce^{A t}B$ as the \emph{impulse response} of the system.
+
+\plr{How about expanding this minor comment into a short paragraph with more interpretation of not just $A$ but also $B$ and $C$? Should say why. 
+Include talking through why $A_{ij}>0$ means that $j$ "upregulates" $i$.}
 As $A$ specifies gene regulatory network architecture, the $i^\text{th}$ row of $A$ can be interpreted as the \emph{cis}-regulatory element (or promoter) for gene $i$.
@@
 
+\plr{you haven't explained $u(t)$ above. and, isn't it more natural in this set-up to have $B = [1,0]$ than $[1,1]$?}
     If the rate each of these genes is expressed is a linear function of their concentrations, the dynamics of the system are given by
@@
       where the row vector $C$ is defined as $C = \begin{bmatrix} 1 & 0 \end{bmatrix}$. 
-(Note: if the dynamics of both genes were physiologically relevant to the cell cycle, we would set $C$ to be the identity matrix).
+      (If the dynamics of both genes were physiologically relevant to the cell cycle, we would set $C$ to be the identity matrix).
   %  \begin{align*}
@@
       Since the input is simply an impulse, its phenotype is equivalent to its impulse response
-
       \begin{align*}
@@
   \section*{Linear Evolutionary Systems}
-
+\plr{instead title this section "Systems drift for linear systems"?}
+
+\plr{This paragraph feels like the introduction.}
 The literature is filled with detailed observations of molecular systems and their diversity. 
-There are examples of significant diversity in the networks underlying processes such as circadian rhythm \citep{sancar2008intelligent}, cell cycle control \citep{cross2011evolution, kearsey2003enigmatic}, pattern formation, and metabolism  \citep{lavoie2009rearrangements, martchenko2007transcriptional, dalal2016transcriptional, christensen2011unique, hartl2007induction, alam2013aspergillus}.
+\plr{Diversity doesn't imply systems drift -- only if the diverse systems are homologous.}
+There are examples of significant diversity in the networks underlying processes such as 
+circadian rhythm \citep{sancar2008intelligent}, 
+cell cycle control \citep{cross2011evolution, kearsey2003enigmatic}, 
+pattern formation \plr{cite?}, 
+and metabolism \citep{lavoie2009rearrangements, martchenko2007transcriptional, dalal2016transcriptional, christensen2011unique, hartl2007induction, alam2013aspergillus}. 
 Despite a symmetry in functionality or phenotype these systems often differ, sometimes substantially, at the molecular level. 
@@
 Gene regulatory networks with identical phenotypes (external dynamics) do not necessarily have identical kryptotypes (internal dynamics). 
-Any linear and minimal system (a gene network) -- minimal, informally meaning that the system's phenotype is achieved with the fewest possible number of genes -- has an identical phenotype up to a change of coordinates.
+Any linear and minimal system (a gene network) 
+-- minimal, informally meaning that the system's phenotype is achieved with the fewest possible number of genes 
+-- has an identical phenotype up to a change of coordinates. 
+\plr{"has an identical phenotype" is not what you mean to say here}
   \begin{align}
@@
   \end{align}
-  Two biological systems, $\mathcal{S} = \{ A, B, C \}$, and $\bar{\mathcal{S}} = \{\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} \}$, have the same phenotype if they are related by a change of coordinates.
+\plr{what are we calling these -- not "biological system" -- that's too broad}
+Two biological systems, $\mathcal{S} = \{ A, B, C \}$, and $\bar{\mathcal{S}} = \{\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} \}$, 
+have the same phenotype if they are related by a change of coordinates. 
+\plr{is this the place to say "and only if, in the minimal dimension"?}
 
@@
 That is, systems that not only have equivalent external dynamics, but also equivalent input and output relationships. 
+\plr{clarify that you mean actually having the same $B$ and $C$. (at first I read "input and output relationship" as the mapping $u \to \phi$),
+    i.e. behaviour across many possible inputs.}
 Formally, this means systems related by a change of coordinates (any invertible matrix $V$) that leaves $B$ and $C$ invariant:
@@
   \end{align}
+\plr{this sentence lacks an object (or some sentence part?)}
 In other words systems with varying gene regulatory network architectures yet identical selection pressures, environment, and phenotype.
 
+\plr{I don't think it makes sense to write things in general as a function of $\tau$,
+    since in general the set of "free parameters" is complicated and we don't actually specify it.
+    Better, maybe define $\mathcal{V}$ to be the set of all $V$ that preserve $B$ and $C$, then define $A(V)$ instead of $A(\tau)$...}
 Define $V(\tau)$ as the parameterized change of coordinates matrix that preserves $B$ and $C$, with $\tau$ a vector of free parameters.
@@
 and a \emph{Linear Evolutionary System} is, 
+\plr{why is this "evolutionary"? It is the thing that we're thinking about evolution of, but that doesn't make it evolutionary.}
   \begin{align}
@@
   \end{align}
+\plr{again, lacking an object}
 That is, all (linear and minimal) mechanisms capable of producing the same phenotype can be realized by a unique choice of $\tau$ in $\mathcal{S}(\tau)$.
@@
 Equivalence of the two characterizations follows from the Cayley-Hamilton theorem.
+\plr{To make sense of this we need to say how $C$ and $B$ change with $n$.}
 Usually, the dimension $n$ and the reference system $A_0$ is implicit and we write only $\calA$.
 
+\plr{should say more about this decomposition here.}
 Regardless of minimality, two systems, even in different dimensions, can have identical external dynamics if they are in $\mathcal{A}$.
@@
 
+\plr{this is same as above, so reference instead of writing out again?}
     Let
@@
     \centering
+% Nice tikz!!!
 \begin{tikzpicture}
@@
 \end{tikzpicture}
-    \caption{A graphical depiction of the set of all externally equivalent cell cycle control networks, $A(\tau)$. $\tau$ can take any value}
+    \caption{A graphical depiction of the set of all externally equivalent cell cycle control networks, $A(\tau)$. $\tau$ can take any value \plr{sentence fragment}}
     \end{figure}
@@
 
+\plr{I think we should first put the section about quantitative drift in, and then the following section that tries to get at real values of parameters.}
+
   %%%%%%%%%%%%%%%%%%%%%%%%
   \section*{Intraspecific variation and genetic drift}
+
+\plr{Below we need to separate variation in $A$ from variation in phenotype.}
 
@@
 
-  At any given time, there will be a range of network coefficients present in the population
-  due to segregating genetic polymorphism.
+  At any given time, there will be a range of regulatory coefficients present in the population
+  due to segregating genetic variants.
   Over many generations, even if selective pressures do not change,
@@
   How fast will a population explore the space of equivalent networks?
-  To answer these questions, we need to know two things known at best only roughly \citep{felsenstein1988phylogenies} --
+  To answer these questions, we need to know 
   the strength of stabilizing selection on the phenotype,
   and the amount (and structure) of heritable variation in the genotype.
-  We aim to get order-of-magnitude estimates.
-
-  \plr{amount of heritable variation}
+  These are known at best only roughly \citep{felsenstein1988phylogenies},
+  so we aim for order-of-magnitude estimates.
+
+  %% amount of heritable variation
   We quantify (roughly) the amount of heritable variation
@@
   In the context of transcription factor networks this may be affected 
-  not only by the binding strength of $j$ to the promoter region of $i$
+  not only by the binding strength of molecule $j$ to the promoter region of gene $i$
   but also the effects of other transcription factors (e.g., cooperativity)
@@
   the mutational target size for variation in $A_{ij}$ may be much larger than the dozens of base pairs
-  typically implicated in the handful of binding sites for transcription factor $j$ of a typical promoter region.
-  (Recall that although these are best modeled through nonlinear terms,
-  by linearizing we essentially consider first-order effects.)
+  typically implicated in the handful of binding sites for transcription factor $j$ of a typical promoter region,
+  and single variants may affect many entries of $A$ simultaneously.
+  % (Recall that although these are best modeled through nonlinear terms,
+  % by linearizing we essentially consider first-order effects.)
   On the other hand, a diverse set of buffering mechanisms are thought to contribute to phenotypic stability
@@
   \plr{replace "micro-scale" with sthg else or discuss earlier}
-  Differences in $A_{ij}$ due to a sequence change are hard to measure 
-  -- the closest available data we could find either related to variation in transcription factor binding site occupancy
-  or in expression levels (e.g., cis-eQTL).
-  The former may be an overestimate, 
-  since it does not capture buffering effects (if for instance only one site of many needs be occupied for transcription to begin)
-  and the latter probably measures changes in steady-state concentration (our $x_i$) rather than the rate of change.
-  \citet{kasowski2010variation} found differential occupancy in 7.5\%
-  of binding sites of a transcription factor (p65) between human individuals.
-  \citet{verlaan2009targeted} showed that cis-regulatory variation
-  accounts for around 2--6\% of expression variation in human blood-derived primary cells,
-  while \citep{lappalainen2013transcriptome} found that human population variation 
-  explained about 3\% of expression variation.
-  \plr{Get some data from at least one other species in here!}
-  Taken together, this suggests that variation in the entries of $A$
-  may be on the scale of 1\% between individuals of a population --
-  doubtless varying substantially between species and between genes.
-
-  \plr{merge this above}
+
+  %% empirical data on sigma
   The amount and structure of this standing variation is established over long time scales
@@
   However, empirical work allows us to estimate at least the rough magnitude of variation.
-
-  \plr{strength of selection}
-  It seems certain that selection is not so strong that intra-population variation
+  Differences in $A_{ij}$ due to a sequence change are hard to measure,
+  but variation in both transcription factor binding site occupancy
+  and expression levels (e.g., cis-eQTL) have been measured in various systems.
+  However, variation in binding site occupancy may overestimate variation in $A$,
+  since it does not capture buffering effects (if for instance only one site of many needs be occupied for transcription to begin),
+  and variation in expression levels measures changes in steady-state concentration (our $\kappa_i$) rather than the \emph{rate} of change.
+  Nonetheless,
+  \citet{kasowski2010variation} found differential occupancy in 7.5\%
+  of binding sites of a transcription factor (p65) between human individuals.
+  \citet{verlaan2009targeted} showed that cis-regulatory variation
+  accounts for around 2--6\% of expression variation in human blood-derived primary cells,
+  while \citep{lappalainen2013transcriptome} found that human population variation 
+  explained about 3\% of expression variation.
+  \plr{Get some data from at least one other species in here!}
+  Taken together, this suggests that variation in the entries of $A$
+  may be on the order of 1\% between individuals of a population --
+  doubtless varying substantially between species and between genes.
+
+  %% argument for strength of selection
+  It seems certain that selection in most species is not so strong that intra-population variation
   is strongly deleterious,
-  so that if $u$ is the typical scale on which selection acts
-  -- e.g., a phenotype that differs from the optimum by $x$ has fitness of $\exp(-x^2/2u^2)$ --
+  so that if $u$ is the typical scale on which selection acts,
   then $u > \sigma$.  
+  \plr{do we call this $\beta$ or $u$?}
   However,
@@
   For instance, 
-  \citet{maria_and_sergey} (others?) found evidence that large-effect regulatory mutations are weakly selected against in Drosophila.
+  \citet{maria_and_sergey} \plr{(others?)} found evidence that large-effect regulatory mutations are weakly selected against in Drosophila.
   This suggests that 
@@
 
-  We have guessed that within a population the entries of $A$
-  vary by a factor of $\sigma=0.01$, at least for networks whose function is strongly constrained.
+  We have guessed that within a population, 
+  the entries of $A$ vary by around 1\%, 
+  at least for networks whose function is strongly constrained.
   Subsequent generations for the most part resample from this diversity,
-  so in a population of effective size $N_e$,
-  simply by the variance of the mean of a random sample,
+  so in a population of effective size $N_e$, with variation in $A$ of order $\sigma$,
   the population mean binding strengths
-  will move a few multiples of $\sigma/\sqrt{N_e}\%$ per generation \citep{lande_drift}.
-  (This could be taken as a definition of $N_e$.)
+  will move a distance of order $\sigma/\sqrt{N_e}$ per generation,
+  simply because this is the the variance of the mean of a random sample \citep{lande_drift}.
+  % (This could be taken as a definition of $N_e$.)
   Selection will tend to push this mean towards the optimal set of networks \citep{ou_process},
   but mean movement parallel to the optimal set (that leaves the phenotype invariant) is unconstrained
-  unless recombination load is substantial \citep{recomb_load}.
+  (although perhaps complicated by recombination load \citep{recomb_load}).
+    \plr{recombination load won't slow it down actually I don't think}
   The action of genetic drift is also strongly determined by covariance between 
@@
 
-  It therefore seems reasonable to coarsely model the time evolution of population variation in network coefficients as 
+  It therefore seems reasonable to coarsely model the time evolution of population variation in regulatory coefficients as 
   (a) a ``cloud'' of width X about the population mean, 
@@
   \paragraph{Brownian motion} on the set of equivalent networks:
+    \plr{revisit.  should this describe motion \emph{on} the set, or motion \emph{near} it, as in writeup with todd?}
   The set $\calA$ is characterized as the solutions to the equations \eqref{eqn:equivalence},
@@
 
-      \plr{add Kalman decomposition stuff here}
-
       \subsection*{Speciation via Reproductive Incompatibility}
@@
 
+\plr{what does "equally present" mean?}
  An F1's genetic regulation is a consequence of both of its genomes, where genes and regulatory sequences from both parents are equally present. 
+\plr{should have a stronger justification for the assertion that $F_1$ is the arithmetic average of the parents than "Thus we say"}
 Thus we say that a diploid organism's gene network is simply the average of both of its gene network copies; one from each parent. 
-Further, each haploid parental gene network copy is formed via meiosis -- by swapping independent genes (specifically their \emph{cis}-regulatory modules) randomly.
-Assuming two distinct but genetically homogeneous populations evolving in allopatry meet and form hybrids, the first generation hybrids (F1s) gene network dynamics will be determined by the average of the parental haplotypes.
+Further, each haploid parental gene network copy is formed via meiosis 
+-- by swapping independent genes (specifically their \emph{cis}-regulatory modules) randomly. 
+Assuming two distinct but genetically homogeneous populations evolving in allopatry meet and form hybrids, 
+the first generation hybrids (F1s) gene network dynamics will be determined by the average of the parental haplotypes. 
 The second generation hybrids (F2s), however, will be the product of a meiosis between parental haplotypes followed by an averaging of gametes. 
-
+\plr{To agree with the previous section we need to incorporate within-pop variation in this argument.}
+
+\plr{I don't think we say that $A$ is a "gene network architecture".}
 Specifically, $F_{1}$ is the first generation hybrid gene network architecture formed by mating (averaging) $A(\tau)$ and $A(\hat{\tau})$, 
+\plr{mating is not averaging}
 \begin{align}
@@
 \end{align}
-
 %\begin{align*}
@@
 and $F_{2}$ is the second generation hybrid gene network architecture formed by gametes $G(i)$ and $G(j)$. 
+\plr{sentences don't begin with "Where", even in math. Also, there isn't "the" $F_2$ - there are many possible.}
 Where each $G$ is formed by randomly swapping rows between $A(\tau)$ and $A(\hat{\tau})$, such that the $i$th  gene comes from $A(\tau)$ ($i$ and $j$ are orthogonal vectors, each element $0$ or $1$, and $i \neq j$).
@@
 \begin{definition}[Reproductive Incompatibility]
+\plr{This part doesn't need a definition - we don't need to discuss what is or isn't a "species".}
     According to Mayr's \emph{Biological Species Concept}, two populations can be considered different species if they are reproductively isolated, meaning crosses between them produce low fitness offspring.
@@
 \end{align*}
-        where angled brackets imply averaging.
-
-        F1s created by crossing phenotypically equivalent oscillators $A(0)$ and $A(2)$ have an $\phi_{F_{1}}(t) = e^{t}$, in contrast to both parents with $\phi_{P_{1}}(t) = \phi_{P_{2}}(t) = \sin(t) + \cos(t)$.
+where angled brackets imply averaging.  \plr{averaging over what? And, this is as others define - cite.}
+
+\plr{This is not part of the definition of $\mathcal{I}$:}
+F1s created by crossing phenotypically equivalent oscillators $A(0)$ and $A(2)$ have a phenotype of $\phi_{F_{1}}(t) = e^{t}$, 
+in contrast to both parents, who have $\phi_{P_{1}}(t) = \phi_{P_{2}}(t) = \sin(t) + \cos(t)$. 
 The hybrid phenotype is significantly different (it does not oscillate and increases infinitely) despite the phenotypic equivalence of the parents.
-
+\plr{don't need these to be big equations.}
 \begin{align*}
@@
 \end{definition}
+
   \begin{figure}[H]
@@
 Here we compare the phenotypes for F2 hybrids formed by crossing oscillators $A(2)$ with $A(2.01)$, $A(2.1)$, and $A(2.5)$ ($B$ and $C$ are the same as above). 
-Each $A$ has is phenotypically identical ($\phi(t) = \sin(t) + \cos(t)$), however some of the hybrids exhibit markedly different dynamics.
+Each $A$ is phenotypically identical ($\phi(t) = \sin(t) + \cos(t)$), however some of the hybrids exhibit markedly different dynamics. 
+\plr{I don't think this is the right place to introduce this?}
 These differences tend to increase with time, quadratically ($x^{2}$) in F2s, and quartically ($x^{4}$) in F1s.
@@
 \end{alignat*}
+\plr{cut this sentence, was already said:}
 F1 gene regulatory networks are formed by averaging the two parental $A(\tau)$ matrices; F2s are formed by first recombining parental matrices followed by an averaging.
+
 %      \begin{figure}[H]
@@
 %      \end{figure}
+
 \begin{figure}[H]
@@
   \end{tabular}
-          \caption{F1 (left) and F2 (right) hybrids crossing $A(2)$ with $A(2.01)$ (top), $A(2.1)$ (middle), and $A(2.5)$ (bottom).
+  \caption{
+    $F_1$ (left) and $F_2$ (right) hybrids crossing $A(2)$ with $A(2.01)$ (top), $A(2.1)$ (middle), and $A(2.5)$ (bottom).
     Note the difference in scale on the $y$-axis.
 F2 hybrids display more phenotypic divergence than F1s, on average.
-Further, some F2s completely fail to oscillate, as seen in an $A(2.5)$ F2 (light blue).}
+Further, some F2s completely fail to oscillate, as seen in an $A(2.5)$ F2 (light blue).
+    \plr{make axis labels bigger}
+  }
 \end{figure}
@@
 \section*{The rate of speciation under neutrality in allopatric populations}
+
+\plr{separate out here (or above) the difference in genotype of hybrids from the optimum set, the difference in phenotype, and the difference in fitness.}
+
 In general terms, if fitness $\mathcal{F}$ is some function $g$ of phenotypic divergence from an optimal scaled by the strength of selection $\beta$,
@@
 \end{align*}
+\plr{sentences don't begin with "And"}
 And a population $P$ is normally distributed around a mean gene network architecture $A(\tau)$ and $\sigma^{2}$ is the intra-population regulatory variance,
+\plr{what is $T$?}
 \begin{align*}
@@
 the population mean architecture will move at rate
+\plr{what is $\Delta \tau$? Also, if we reframe in terms of $V$ not $\tau$ above will need to reframe this.}
 \begin{align*}
@@
 
+%%%%%%%%%%%%%%%
 \begin{example}[Metabolic network]
 
-  Consider an organism that can metabolize two different sugars $s_{1}$ and $s_{2}$ (present at logarithmic concentrations $u_{1}$ and $u_{2}$ respectively), with enzymes $e_{1}$ and $e_{2}$ (with log concentration denoted as $\phi_{1}$ and $\phi_{2}$).
+\plr{Concentrations are "logarithmic", you mean that $u_1(t) = \log(\text{concentration of molcule 1 at time t})$.}
+Consider an organism that can metabolize two different sugars $s_{1}$ and $s_{2}$ (present at logarithmic concentrations $u_{1}$ and $u_{2}$ respectively), 
+with enzymes $e_{1}$ and $e_{2}$ (with log concentration denoted as $\phi_{1}$ and $\phi_{2}$). 
 Further suppose that one of the sugars $s_{2}$ is the preferred energy source (perhaps it contains significantly more energy than the other sugar or is otherwise more efficiently metabolized). 
 The organism may have a gene regulatory network $\mathcal{S}$ that can deploy a situation specific metabolic strategy. 
+\plr{I think you argue below that the optimal response is a complete switch from one enzyme to the other in the presence of the second sugar? Should say so here.}
 That is depending on both $u_{2}$ and $u_{1}$ the organism will synthesize an appropriate $\phi_{1}$ and $\phi_{2}$. 
@@
 The phenotype is, 
+\plr{this is wrong, fix it}
 \begin{align*}
@@
 
-  We can see that changing coordinates on this system (with any invertible $V \in \mathbb{R}^{2 \times 2}$) will find new system architectures, as shown before, however, we can also apply the Kalman decomposition to find systems in different dimensions -- that is systems that employ more than $2$ transcription factors, yet have the same external dynamics/phenotype.
+We can see that changing coordinates on this system (with any invertible $V \in \mathbb{R}^{2 \times 2}$) will find new system architectures, 
+as shown before, however, we can also apply the Kalman decomposition to find systems in different dimensions 
+-- that is systems that employ more than $2$ transcription factors, yet have the same external dynamics/phenotype. 
+\plr{Ah -- this is much simpler if we let $B$ and $C$ change as well!}
 This may happen if a system is co-opted from another, or may be a consequence of gene duplication and deletion.
 
-  $\widehat{\mathcal{S}}$ is $4$-dimensional biological system,  \begin{align*}
+\plr{where did $\widehat{\mathcal{S}}$ come from? need a lead-in here.}
+$\widehat{\mathcal{S}}$ is $4$-dimensional biological system,  
+\begin{align*}
     \widehat{\mathcal{S}} = \left\{ \begin{array}{ll}
@@
 
+\plr{the point here is not clear.}
 So for example, some system can be wired as follows, and still be input-output equivalent to the minimal metabolic system $\mathcal{S}_{\min}$:
@@
 
-     Despite the present example consisting of a minimal $2 \times 2$ and a non-minimal $4 \times 4$ system, any $n$-dimensional system can be constructed using this method -- applying a change of coordinates to the Kalman decomposition -- to construct a mechanistically different system with identical phenotypic dynamics.
+Despite the present example consisting of a minimal $2 \times 2$ and a non-minimal $4 \times 4$ system, 
+any $n$-dimensional system can be constructed using this method 
+-- applying a change of coordinates to the Kalman decomposition 
+-- to construct a mechanistically different system with identical phenotypic dynamics. 
 Depending on the specifics of a system being modeled, one may have to take care to restrict the free parameter values and network architectures to be biologically appropriate. 
+\plr{what are you thinking of as being a potential problem here? say so explicitly.}
+
 \end{example}
+
 %    \subsection*{Metabolic Network}
@@
 
+%%%%%%%%%%%%%%%%%%%%%
 \section*{Discussion}
+
 \plr{mention $B$ part of genetic architecture}
@@
 Here, we offer a tractable and simple model which includes these missing features. 
-Further, we provide, in clear mathematical language, an analytical description of phenomena hitherto only discussed verbally and conceptually (phenogenetic drift \citep{weiss2000phenogenetic}, developmental systems drift \citep{true2001developmental}, biological degeneracy \citep{edelman2001degeneracy}, \emph{etc.}).
-The tractability and relative simplicity of this exposition enables the interested biologist to work out by hand, if desired, the dynamics of a genetic system, as well as perturbations to the system -- an attribute not likely to be found in less tractable models and simulations.
+\plr{this is a claim we are the first to do something.  best not to make these claims (and we aren't the first to make math models of these)}
+Further, we provide, in clear mathematical language, an analytical description of phenomena hitherto only discussed verbally and conceptually 
+(phenogenetic drift \citep{weiss2000phenogenetic}, developmental systems drift \citep{true2001developmental}, biological degeneracy \citep{edelman2001degeneracy}, \emph{etc.}). 
+The tractability and relative simplicity of this exposition enables the interested biologist to work out by hand, 
+if desired, 
+\plr{probably not desired}
+the dynamics of a genetic system, as well as perturbations to the system -- an attribute not likely to be found in less tractable models and simulations.
 
 We have suggested an interpretation of system identification: to see it as an evolutionarily neutral manifold, and not simply a computational nuisance. 
-We have demonstrated a method to analytically determine the set of all phenotypically invariant gene networks; by a simple change of coordinates in the minimal configuration, or more generally by applying the Kalman decomposition in higher dimensions.
+We have demonstrated a method to analytically determine the set of all phenotypically invariant gene networks; 
+by a simple change of coordinates in the minimal configuration, or more generally by applying the Kalman decomposition in higher dimensions. 
 Further, we emphasize that evolution proceeds through this high dimensional space as stochastic coordinate transformation, constrained by sexual reproduction and selection. 
-This set is explored over evolutionary time when phenotype is conserved, and can lead to a diverse set of consequences, including the accumulation of Dobzhansky-Muller incompatibilities.
+This set is explored over evolutionary time when phenotype is conserved, and can lead to a diverse set of consequences, 
+including the accumulation of Dobzhansky-Muller incompatibilities. 
+\plr{check original Bateson/DM papers to see if this accords with those defns}
 We emphasize that these incompatibilities are a consequence of recombining different, yet functionally equivalent, mechanisms.
 
-  Furthermore, using a quantitative genetic approach, we estimated that a genetically variable population will drift in neutral system space at a rate determined by its intra-population variation and its effective population size.
-Because mechanistically distinct yet phenotypically equivalent biological systems can fail to produce viable hybrids, we predict allopatric populations to speciate at a rate on the order of $N_{e}$ under reasonable population genetic parameter estimates.
+Furthermore, using a quantitative genetic approach, we estimated that a genetically variable population will drift in neutral system space 
+at a rate determined by its intra-population variation and its effective population size. 
+\plr{replace "speciate" with "accumulate genetic incompatibilities"}
+Because mechanistically distinct yet phenotypically equivalent biological systems can fail to produce viable hybrids, we predict allopatric populations to speciate 
+at a rate on the order of $N_{e}$ under reasonable population genetic parameter estimates. 
 Additionally we see second-generation hybrid fitness plummet much faster than that of first-generation hybrids. 
-This insight is a consequence of combining our mechanistic model with a quantitative genetic one: we observe that $F_{1}$ phenotypes diverge quartically, and $F_{2}$ phenotypes quadratically, with evolutionary time.
+\plr{refer to Turelli here}
+This is a consequence of combining our mechanistic model with a quantitative genetic one: 
+we observe that $F_{1}$ phenotypes diverge quartically, and $F_{2}$ phenotypes quadratically, with evolutionary time. 
+\plr{awkward phrasing}
 This result is also consistent with Haldane's rule; that if only one hybrid sex is inviable or sterile it is likely the heterogametic sex. 
@@
 
-  We also suggest that gene networks may not always use their components parsimoniously as network size tends to ratchet up in the absence of strong selection against extra parts.
+We also suggest that gene networks may not always use their components parsimoniously 
+as network size tends to ratchet up in the absence of strong selection against extra parts. 
+\plr{probably shouldn't claim this is "unexplored"}
 Although unexplored presently, this phenomena may lead to insights on evolvability and developmental innovation. 
@@
 
+\plr{this paragraph isn't very clear}
 As Richard Levins opined, models in population biology face a trade-off among precision, realism, and generality \citep{levins1966strategy}.
@@
 Johnson and Porter 2000 did not observe any hybrid fitness declines under stabilizing selection -- only under directional selection.
+\plr{citations:}
 Khatari et al, Tulchinsky et al, and Porter et al, all study hybrid incompatibility from a transcription factor/binding site interaction perspective, not from an overall network architecture perspective.

### 2017-08-03 kalman_walk_paper.tex
moved comments to margin
edits.

@@
 
+% for comments in margins or not
+\newif\ifmargincomments
+\margincommentstrue
+% \margincommentstrue
+\ifmargincomments
+    \usepackage{todonotes}
+    \usepackage[left=1cm,right=6.5cm,top=3cm,bottom=3cm,nohead,nofoot,marginparwidth=6cm]{geometry}
+    \newcommand{\plr}[1]{\todo[color=blue!25]{#1}}
+    % inline comment version
+    \newcommand{\plri}[1]{{\color{blue}\it #1}}
+\else
     \newcommand{\plr}[1]{{\color{blue}\it #1}}
+    \newcommand{\plri}[1]{\plr{#1}}
+\fi
+
 \newcommand{\jss}[1]{{\color{olive}\it #1}}
@@
 \newcommand{\1}{\mathbbm{1}}
+
+% fitness as a fn of distance
+\newcommand{\fit}{\mathcal{F}}
+% fitness as a fn of A
+\newcommand{\fitx}{\mathcal{f}}
+% set of optimal coefficients
+\newcommand{\optx}{\mathcal{X}}
+% optimal phenotype
+\newcommand{\optph}{\Phi_0}
+% distance in phenotype space
+\newcommand{\dph}{d}
 
@@
 
-  \plr{Additional ideas to consider adding:
+\plr{Additional ideas to consider adding:}
 \begin{itemize}
@@
     %\item obtain estimates of variation in $A$ from thermodynamic occupancy model
-\end{itemize}}
-
-\plr{Note: in the \LaTeX source I'm putting in semantic linebreaks, so it's easy to edit and move around phrases and ideas.}
+\end{itemize}
+
+\plri{Note: in the \LaTeX source I'm putting in semantic linebreaks, so it's easy to edit and move around phrases and ideas.}
+
+\plri{Need to come up with a consistent term for ``the $A_{ij}$''s. -- ``regulatory coefficients''? ``genotype''?}
 
@@
 \plr{say what $u$ is (a vector)}
-\plr{say in words what $\kappa$ is before the equation.
+\plr{say in words what $\kappa$ is before the equation.}
 Such a system $\mathcal{S}$, is composed of two equations,
@@
 \end{tikzpicture}
-    \caption{A graphical depiction of the set of all externally equivalent cell cycle control networks, $A(\tau)$. $\tau$ can take any value \plr{sentence fragment}}
+    \caption{A graphical depiction of the set of all externally equivalent cell cycle control networks, $A(\tau)$. 
+        $\tau$ can take any value \plri{sentence fragment}}
     \end{figure}
@@
 
+\plri{Missing: a statement about when there are neutral directions, and how many.}
+
 \plr{I think we should first put the section about quantitative drift in, and then the following section that tries to get at real values of parameters.}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%
-  \section*{Intraspecific variation and genetic drift}
-
-\plr{Below we need to separate variation in $A$ from variation in phenotype.}
-
-%  \plr{rework this: we assume optimal phenotype is fixed and population is large but polymorphic and not strictly on the optimum}
-
-%First we focus on a simple evolutionary scenario -- \emph{phenotypically invariant evolution}: a large population, perfectly adapted, and in a constant environment.
-In this circumstance we expect phenotype to be conserved throughout evolutionary time.
-As such we should only expect the phenotype to change as a consequence of genetic drift (small effective population size), adaptation to new selective and/or environemental pressures.
-These phenotypic variations should yield distinct signatures.
-Adaptative changes will change the optimal impulse response function $h(t) \xrightarrow{adaptation} h'(t)$.
-Genetic drift registers as an increase in the intrapopulation variation in $h(t)$.
-% \jss{added the phenotypic invariance stuff here (above) -- not sure if it should put elsewhere yet}
+\section*{Systems drift and the accumulation of incompatibilities}
 
@@
 How fast will a population explore the space of equivalent networks?
-  To answer these questions, we need to know 
+In this section we explore informally a general model for this situation,
+in which a population drifts stochastically near a set of equivalent, optimal systems.
+We work with a population of effective size $N_e$.
+
+Since the argument applies more generally, we abstract the model above.
+Suppose that a set $x$ of coefficients that determine a system (this is $A$ above),
+produce a phenotype $\Phi(x)$ (the time course of $\phi(t)$).
+There is an optimal phenotype $\optph$,
+and a set $\optx$ of ``optimal'' coefficients that produce this phenotype.
+Fitness depends on distance to the optimal phenotype --
+we will write the ``distance'' between phenotypes $\phi$ and $\psi$ as $\dph(\phi, \psi)$,
+so that the fitness of an organism with coefficents $x$ is $\fitx(x) = \fit(\dph(\Phi(x),\optph)/\beta)$,
+where $\beta$ controls the scale on which fitness drops.
+We will assume that the map $\Phi$ and the distance function $\dph()$ are smooth,
+and that the optimal set $\optx$ is locally isomorphic to $\R^m$.
+\plr{say that better}
+
+\paragraph{Offspring.}
+Individuals are diploid; we assume that each haploid genome determines a set of coefficients,
+and the individual's coefficients are the average of her two haploid values (no dominance).
+This implies that the diploid population variance, $\sigma^2$, is one-half the haploid variance.
+Each new gamete is produced from the parent's two haploid copies;
+for simplicity we assume that the gamete inherits a random choice of one of the two parental copies,
+and so $\sigma$ remains constant, up to a $1/N_e$ term.
+A more general model including segregation variance \citep{barton_infinitesimal}
+would result in the same conclusions.
+\plri{but put this in the appendix?}
+
+\paragraph{System drift}
+% plr how fast does it drift
+If the variation within a population of some coefficient
+has standard deviation $\sigma$,
+then since subsequent generations resample from this diversity,
+the population mean coefficient will move a random distance of size $\sigma/\sqrt{N_e}$ per generation,
+simply because this is the the standard deviation of the mean of a random sample \citep{lande_drift}.
+% (This could be taken as a definition of $N_e$.)
+Selection will tend to restrain this motion,
+but mean movement along the optimal set $\optx$ is unconstrained.
+% (although perhaps complicated by recombination load \citep{recomb_load})
+The value of $\sigma$ in particular directions in coefficient space 
+depends on constraints imposed by selection and 
+the covariance between genetic variation between different coefficients -- known as the $G$ matrix \citep{G_matrix}.
+% covariance which may arise due to functional constraints and/or statistical linkage.
+% There may well be functional constraints -- but these are not sufficiently well-known to say anything general about.
+For instance, linkage will lead to covariance
+if the variation is due to \textit{cis}-regulatory variants,
+in which case the genetic basis of each \emph{row} of $A$ likely lies within a few kilobases of tightly linked sequence,
+across which a population may carry only a few common haplotypes.
+However, covariance due to transiently assembled haplotypes is not expected to be stable over long periods of time --
+a common \textit{cis}-regulatory haplotype of transcription factor $k$ with particularly strong binding to both $i$ and $j$
+(leading to positive covariance between $A_{ik}$ and $A_{jk}$)
+is no more likely to appear than one with strong binding to $i$ but particularly weak binding to $j$ (negative covariance).
+(Such transient covariances may well increase the variance of the per-generation change in network mean, however \citep{barton_linkage}.)
+
+% It therefore seems reasonable to coarsely model the time evolution of population variation in regulatory coefficients as 
+% (a) a ``cloud'' of width X about the population mean, 
+% which (b) moves as an unbiased Brownian motion through the set of network coefficients that give the optimal phenotype.
+% In fact, the population mean will not produce exactly the optimal phenotype,
+% but it will be convenient to refer to this closest point on the optimal set as ``the population mean''.
+
+To obtain a general quantitative picture, we need to know 
+$\sigma_N$ and $\sigma_S$, the standard deviations of coefficient variation along and perpendicular to $\optx$ respectively;
+$\gamma$, how fast phenotype changes moving away from $\optx$;
+and $\beta$, how fast fitness drops with distance from the optimal phenotype.
+(Concretely, $\gamma$ is the derivative of $\dph(\Phi(x+u z), \Phi(x))$ with respect to $u$ for $x \in \optx$ and $z$ perpendicular to the tangent subspace at $x$.)
+With these parameters, a typical individual will have a fitness of around $\fit(\sigma_S \gamma / \beta)$.
+
+\paragraph{Hybridization}
+\plr{I put F1s and F2s here, but with less explanation than below. Merge somehow.}
+By the arguments above, 
+over $T$ generations the population mean will move a distance of order $\sqrt{\sigma_N/T}$ along $\optx$.
+A $F_1$ hybrid between the two populations
+has one haploid genome from each,
+so will be distributed as the mean of two randomly chosen haploids from the two populations,
+and so will have mean at the midpoint between their means,
+and variance equal to $\sigma$.
+A $F_2$ hybrid will be homozygous for one parental allele on averge at half of the loci in the genome,
+so the distribution of $F_2$s will have mean at the average of the two populations,
+as before,
+but variance equal to $\sigma^2 + z^2/2$, where $z$ is the distance between the parental populations.
+These are depicted in figure \ref{fig:conceptual_fig}.
+
+\begin{figure}
+\centering
+\includegraphics{figures/conceptual_fig}
+\caption{
+    \label{fig:conceptual_fig}
+    WRITE CAPTION
+    \plri{The figure is not quite right now -- should show all types of F2 in purple.  but then it will be messy.  ideas?}
+}
+\end{figure}
+
+\plri{improve figure by putting labels on from the following}
+\plr{I think we can set $\beta=1$ by defining $\fit$ appropriately?}
+Suppose that two populations have drifted independently to differ by $z$,
+and that $z$ is of the same order as $\sigma$ but is smaller than $\gamma$,
+so that for $x_0$ a point nearby on $\optx$,
+\begin{align*}
+    \Phi(x_0 + x) = \Phi(x_0) + D \Phi(x_0) \cdot x + O(\|\gamma x\|^2),
+\end{align*}
+and so $d(\Phi(x_0+x), \Phi(x_0)) \approx  \gamma x$.
+The mean $F_1$ is the average of the parental means,
+and since the first-order terms in the Taylor series vanish,
+has phenotype differing from the optimum by a distance of order $(\gamma z)^2$.
+\plri{(Proof: let the parental means be $x_1$ and $x_2$; then as $z=x_2-x_1$,
+    $d(\Phi((x_1+x_2)/2), \Phi(x_1)) = D \Phi(x_1) \cdot z/2 + O(\gamma z)^2$,
+    but expanding about $x_2$ instead we get
+    $d(\Phi((x_1+x_2)/2), \Phi(x_2)) = - D \Phi(x_2) \cdot z/2 + O(\gamma \|z\|)^2$.
+    Since $\Phi(x_1) = \Phi(x_2)$ and $D \Phi(x_1) - D \Phi(x_2) = O(\|z\|^2)$,
+    we see that $D \Phi(x_1) \cdot z = O(\gamma z)^2$.}
+The mean $F_2$ is the same,
+but is a distance away from the optimal set of order $z$.
+In summary, the typical fitness of an individual in the original population is
+$\fit_0 = \fit(\sigma_s \gamma / \beta)$;
+of an $F_1$ is
+$\fit_1 = \fit((\sigma_s \gamma + \gamma z^2) / \beta)$;
+and of an $F_2$ is
+$\fit_2 = \fit(\gamma z / \beta)$.
+
+If $\fit(x) = e^{-x^2} \approx 1 - x^2$,
+then ...
+\plri{reparameterize to define $\fit_1$ as the \emph{drop} in fitness?}
+
+%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection*{Parameter estimates for real systems}
+
+To translate these results into real predictions, we need to know
 the strength of stabilizing selection on the phenotype,
@@
 at least for networks whose function is strongly constrained.
-  Subsequent generations for the most part resample from this diversity,
-  so in a population of effective size $N_e$, with variation in $A$ of order $\sigma$,
-  the population mean binding strengths
-  will move a distance of order $\sigma/\sqrt{N_e}$ per generation,
-  simply because this is the the variance of the mean of a random sample \citep{lande_drift}.
-  % (This could be taken as a definition of $N_e$.)
-  Selection will tend to push this mean towards the optimal set of networks \citep{ou_process},
-  but mean movement parallel to the optimal set (that leaves the phenotype invariant) is unconstrained
-  (although perhaps complicated by recombination load \citep{recomb_load}).
-    \plr{recombination load won't slow it down actually I don't think}
-  The action of genetic drift is also strongly determined by covariance between 
-  standing genetic variation in different regulatory coefficients -- known as the $G$ matrix \citep{G_matrix},
-  covariance which may arise due to functional constraints and/or statistical linkage.
-  There may well be functional constraints -- but these are not sufficiently well-known to say anything general about.
-  Linkage will almost certainly lead to covariance
-  if the variation is due to \textit{cis}-regulatory variants,
-  in which case the genetic basis of each \emph{row} of $A$ likely lies within a few kilobases of tightly linked sequence,
-  across which a population may carry only a few common haplotypes.
-  However, covariance due to transiently assembled haplotypes is not expected to be stable over long periods of time --
-  a common \textit{cis}-regulatory haplotype of transcription factor $k$ with particularly strong binding to both $i$ and $j$
-  (leading to positive covariance between $A_{ik}$ and $A_{jk}$)
-  is no more likely to appear than one with strong binding to $i$ but particularly weak binding to $j$ (negative covariance).
-  (Such transient covariances may well increase the variance of the per-generation change in network mean, however \citep{barton_linkage}.)
-  In principle, there may be substantially less variation away from the set of optimal coefficients 
-  than there is along the set, due to the action of selection.
-  If so, this might require substantial epistatic load 
-  -- mortality or reduced fitness of a large proportion of new offspring due to recombination between somewhat incompatible alleles.
-  However, it is unclear if this is likely to occur and XXX we revisit this below.
-
-  It therefore seems reasonable to coarsely model the time evolution of population variation in regulatory coefficients as 
-  (a) a ``cloud'' of width X about the population mean, 
-  which (b) moves as an unbiased Brownian motion through the set of network coefficients that give the optimal phenotype.
-  In fact, the population mean will not produce exactly the optimal phenotype,
-  but it will be convenient to refer to this closest point on the optimal set as ``the population mean''.
-
-  %%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \paragraph{Brownian motion} on the set of equivalent networks:
-    \plr{revisit.  should this describe motion \emph{on} the set, or motion \emph{near} it, as in writeup with todd?}
-  The set $\calA$ is characterized as the solutions to the equations \eqref{eqn:equivalence},
-  and is hence an algebraic variety. 
-  In fact, the Kalman decomposition \plr{XXX above} provides us with an explicit characterization of the set.
-  Above we argued that the population mean set of coefficients $A$ was subject to genetic drift, 
-  with each entry $A_{ij}$ changing with mean square displacement $\sigma^2/\sqrt{N_e}$ per generation.
-  Since selection constrains the population mean to stay near to the set of optimal networks $\calA$,
-  to a good approximation, the population mean moves as Brownian motion in the space of matrices $\R^{n \times n}$
-  but conditioned to stay on the optimal set.
-  This Brownian motion on $\calA$ is a stochastic process driven by the Laplace-Beltrami operator on $\calA$
-  where this makes sense, with special behavior at the singular points (if any \plr{XXX}).
-  Unlike Brownian motion in flat space, this stochastic process can have \emph{bias} --
-  for instance, it is pushed away from regions of negative intrinsic curvature
-  (because there is ``less space'' in such regions).
-
-  \plr{NOT TRUE?:}
-  However, the process always has the property that
-  changes in $A$ accumulate at constant (mean squared) rate:
-  if $\calA$ is locally isomorphic to $\R^m$ around $A_0$
-  (there are $m$ degrees of freedom in the solutions to \eqref{eqn:equivalence}) then
-  \begin{align}
-      \E\left[ \|A_t - A_0\|^2 \right] = m \sigma^2 T + O(T^2) .
-  \end{align}
-
-  %\begin{align}
-  %  \Delta_{\tau} &= \left\lVert \frac{d}{d \tau} \text{vec} \left[A\left(\tau\right)\right] \right\rVert^{-1} \\
-  %  A(\tau) &\xrightarrow{\mu} A(\tau \pm \mu \Delta_{\tau})
-  %\end{align}
- % \section*{Phenotypic Invariance}
-
-%    \begin{example}[Not all minimal gene networks can drift]
- %     If a gene network is minimal and all the molecular species involved in the network are under selection, such that $C$ is the $n \times n$ identity matrix  ($C = I_{n}$), the only acceptable  change of coordinate matrix is the identify matrix.
-
-   %  \begin{align}
-   %     C  = I \lor B = I \\
-   %     IV^{-1} &= I \\
-   %     \iff V &= I
-   %   \end{align}
- %     \end{example}
-  %    \begin{example}[All Non-Minimal Gene Networks can Drift]
-  %    
-  %    Despite the existence of a unique genetic architecture in the minimal case, there still exists an infinite number of systems with larger networks that have identical external dynamics. 
-  %    \begin{align}
-  %      h(t) &= \widehat{h}(t) \iff \\ 
-  %      CA^{\jmath}B &= \widehat{C} \widehat{A}^{\jmath} \widehat{C} \qquad \text{for } \jmath = 0, 1, \dots
-  %    \end{align}
-  %    Any two systems with equivalent impulse responses will have equivalent phenotypes.
-  %    \end{example}
-
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection*{Speciation via Reproductive Incompatibility}
@@
 Further, some F2s completely fail to oscillate, as seen in an $A(2.5)$ F2 (light blue).
-    \plr{make axis labels bigger}
+    \plri{make axis labels bigger}
   }
@@
 Suppose that the population mean differs from the optimum by $\epsilon$,
-that $\Sigma = \sigma^2 I$ and $U = s I$.
+that $\Sigma = \sigma^2 I$ and $U = I/\beta^2$ (so, stablizing selection happening on a distance scale of $\beta$).
 Then the population mean gets closer to the optimum on average, moving to
-$\epsilon/(1 + \sigma^2/s)$
-and adds noise of size $\sqrt{s} \sigma/\sqrt{N \sigma^2 + N s}$.
+$\epsilon/(1 + \sigma^2 \beta^2)$
+and adds noise of size $(1/\beta) \sigma/\sqrt{N \sigma^2 + N 1/\beta^2}$.
 At equilibrium, these two movements will be of the same order,
-so that $\epsilon$ is of order $(\sigma/\sqrt{N}) \sqrt{1+\sigma^2/s}$.
+so that $\epsilon$ is of order $(\sigma/\sqrt{N}) \sqrt{1+\sigma^2 \beta^2}$.
 

### 2017-08-03 figures/landscape_figure.R

@@ -4,7 +4,13 @@ dx <- 0.25
 xx <- yy <- seq(1/sqrt(2)-dx,1/sqrt(2)+dx,length.out=100)
 z <- (sqrt(outer(xx^2, yy^2, "+")) - 1.0)^2
 
-do.plot <- function (p1, p2, add=FALSE) {
+plot_pop <- function (x, sigma, nreps=100, pch=20, cex=1, ...) {
+    y <- matrix(rnorm(length(x)*nreps, mean=rep(x,each=nreps), sd=sigma), nrow=nreps)
+    points(y, pch=pch, cex=cex, ...)
+    return(invisible(y))
+}
+
+do.plot <- function (p1, p2, sigma, add=FALSE) {
     F1 <- (p1+p2)/2
     F2 <- rbind( c(p1[1],p2[2]), c(p2[1],p1[2]) )
     proj <- function (xy) { th <- atan2(x=xy[1],y=xy[2]); c(cos(th), sin(th)) }
@@ -19,6 +25,10 @@ do.plot <- function (p1, p2, add=FALSE) {
              lty=2, col='purple')
     segments(x0=F1[1], y0=F1[2], x1=proj(F1)[1], y1=proj(F1)[2], col='red')
     segments(x0=F2[2,1], y0=F2[2,2], x1=proj(F2[2,])[1], y1=proj(F2[2,])[2], col='purple')
+    plot_pop(p1, sigma, col=adjustcolor('black', 0.2))
+    plot_pop(p2, sigma, col=adjustcolor('black', 0.2))
+    plot_pop(F1, sigma, col=adjustcolor('red', 0.2))
+    plot_pop(F2[2,], sigma, col=adjustcolor('purple', 0.2))
     points(rbind(p1,p2,F1,F2[2,]), cex=2, pch=20,
            col=c("black","black","red","purple") )
     if (!add) legend("topright", pch=20, pt.cex=2, bg='white',
@@ -36,8 +46,8 @@ p4 <- c(0.85, sqrt(1-0.85^2))
 pdf(file="conceptual_fig.pdf", width=3, height=3, pointsize=10)
 par(mar=c(1,1,1,1))
 
-do.plot(p1, p2)
+do.plot(p1, p2, sigma=.012)
 
-do.plot(p3, p4, add=TRUE)
+do.plot(p3, p4, sigma=.012, add=TRUE)
 
 dev.off()

### 2017-08-04 kalman_walk_paper.tex
rewrote quant gen section

@@
 % fitness as a fn of A
-\newcommand{\fitx}{\mathcal{f}}
+\newcommand{\fitx}{\mathcal{F}}
 % set of optimal coefficients
@@
 \newcommand{\dph}{d}
+% incompatibility
+\newcommand{\Incompat}{\mathcal{I}}
 
@@
 
-Since the argument applies more generally, we abstract the model above.
 Suppose that a set $x$ of coefficients that determine a system (this is $A$ above),
@@
 we will write the ``distance'' between phenotypes $\phi$ and $\psi$ as $\dph(\phi, \psi)$,
-so that the fitness of an organism with coefficents $x$ is $\fitx(x) = \fit(\dph(\Phi(x),\optph)/\beta)$,
-where $\beta$ controls the scale on which fitness drops.
-We will assume that the map $\Phi$ and the distance function $\dph()$ are smooth,
+measured so that the fitness of an organism with coefficents $x$ is $\fitx(x) = \exp(-\dph(\Phi(x),\optph)^2)$.
+We will assume that the map $\Phi$ is smooth
 and that the optimal set $\optx$ is locally isomorphic to $\R^m$.
@@
 A more general model including segregation variance \citep{barton_infinitesimal}
-would result in the same conclusions.
+would result in the same qualitative conclusions.
 \plri{but put this in the appendix?}
@@
 % (although perhaps complicated by recombination load \citep{recomb_load})
-The value of $\sigma$ in particular directions in coefficient space 
-depends on constraints imposed by selection and 
-the covariance between genetic variation between different coefficients -- known as the $G$ matrix \citep{G_matrix}.
+The amount of variance in particular directions in coefficient space 
+depend on constraints imposed by selection and 
+the covariance between genetic variation between different coefficients (the $G$ matrix \citep{G_matrix}).
 % covariance which may arise due to functional constraints and/or statistical linkage.
 % There may well be functional constraints -- but these are not sufficiently well-known to say anything general about.
-For instance, linkage will lead to covariance
+For instance,
 if the variation is due to \textit{cis}-regulatory variants,
-in which case the genetic basis of each \emph{row} of $A$ likely lies within a few kilobases of tightly linked sequence,
+the genetic basis of each \emph{row} of $A$ likely lies within a few kilobases of tightly linked sequence,
 across which a population may carry only a few common haplotypes.
@@
 To obtain a general quantitative picture, we need to know 
-$\sigma_N$ and $\sigma_S$, the standard deviations of coefficient variation along and perpendicular to $\optx$ respectively;
-$\gamma$, how fast phenotype changes moving away from $\optx$;
-and $\beta$, how fast fitness drops with distance from the optimal phenotype.
-(Concretely, $\gamma$ is the derivative of $\dph(\Phi(x+u z), \Phi(x))$ with respect to $u$ for $x \in \optx$ and $z$ perpendicular to the tangent subspace at $x$.)
-With these parameters, a typical individual will have a fitness of around $\fit(\sigma_S \gamma / \beta)$.
+$\sigma_N$ and $\sigma_S$, the standard deviations of coefficient variation along and perpendicular to $\optx$ respectively,
+and $\gamma$, the scale on which phenotype changes moving away from $\optx$.
+Concretely, $\gamma$ is the inverse of the derivative of $\dph(\Phi(x+u z), \Phi(x))$ 
+with respect to $u$ for $x \in \optx$ and $z$ perpendicular to the tangent space at $x$.
+With these parameters, a typical individual will have a fitness of around $\exp(-(\sigma_S/\gamma)^2)$.
 
 \paragraph{Hybridization}
-\plr{I put F1s and F2s here, but with less explanation than below. Merge somehow.}
-By the arguments above, 
-over $T$ generations the population mean will move a distance of order $\sqrt{\sigma_N/T}$ along $\optx$.
-A $F_1$ hybrid between the two populations
-has one haploid genome from each,
-so will be distributed as the mean of two randomly chosen haploids from the two populations,
-and so will have mean at the midpoint between their means,
+\plr{I put F1s and F2s here, and $\Incompat$, but with less explanation than below. Merge somehow.}
+By the arguments above, the means of two allopatric populations separated for $T$ generations
+will be a distance of order $2\sigma_N \sqrt{T/N_e}$ apart along $\optx$.
+A population of $F_1$ hybrids has one haploid genome from each,
+whose coefficients are averaged,
+and so will have mean system coefficients at the midpoint between their means,
 and variance equal to $\sigma$.
-A $F_2$ hybrid will be homozygous for one parental allele on averge at half of the loci in the genome,
+Each $F_2$ hybrid will be homozygous for one parental allele on averge at half of the loci in the genome,
 so the distribution of $F_2$s will have mean at the average of the two populations,
@@
     \label{fig:conceptual_fig}
-    WRITE CAPTION
-    \plri{The figure is not quite right now -- should show all types of F2 in purple.  but then it will be messy.  ideas?}
+    A conceptual figure of the fitness consequences of hybridization:
+    axes represent system coefficients (i.e., entries of $A$);
+    the line of optimal system coefficients is down in black;
+    dotted lines give phenotypic distances to the optimum.
+    Two pairs of parental populations are shown in black, along the optimum;
+    a hypothetical population of $F_1$s are shown for each in red,
+    and the distribution of one type of $F_2$ is shown in purple
+    (other types of $F_2$ are not shown).
+    Solid lines depict the distance of the $F_2$ to optimum.
+    \plri{Should show all types of F2? would be messy.}
 }
@@
 \plri{improve figure by putting labels on from the following}
-\plr{I think we can set $\beta=1$ by defining $\fit$ appropriately?}
 Suppose that two populations have drifted independently to differ by $z$,
-and that $z$ is of the same order as $\sigma$ but is smaller than $\gamma$,
-so that for $x_0$ a point nearby on $\optx$,
-\begin{align*}
-    \Phi(x_0 + x) = \Phi(x_0) + D \Phi(x_0) \cdot x + O(\|\gamma x\|^2),
-\end{align*}
-and so $d(\Phi(x_0+x), \Phi(x_0)) \approx  \gamma x$.
+and that $z$ is of the same order as $\sigma$ but is smaller than $\gamma$.
 The mean $F_1$ is the average of the parental means,
 and since the first-order terms in the Taylor series vanish,
-has phenotype differing from the optimum by a distance of order $(\gamma z)^2$.
-\plri{(Proof: let the parental means be $x_1$ and $x_2$; then as $z=x_2-x_1$,
-    $d(\Phi((x_1+x_2)/2), \Phi(x_1)) = D \Phi(x_1) \cdot z/2 + O(\gamma z)^2$,
-    but expanding about $x_2$ instead we get
-    $d(\Phi((x_1+x_2)/2), \Phi(x_2)) = - D \Phi(x_2) \cdot z/2 + O(\gamma \|z\|)^2$.
-    Since $\Phi(x_1) = \Phi(x_2)$ and $D \Phi(x_1) - D \Phi(x_2) = O(\|z\|^2)$,
-    we see that $D \Phi(x_1) \cdot z = O(\gamma z)^2$.}
+has phenotype differing from the optimum by a distance of order $\|z\|^2$ 
+(see appendix \ref{apx:away_from_opt}).
 The mean $F_2$ is the same,
-but is a distance away from the optimal set of order $z$.
-In summary, the typical fitness of an individual in the original population is
-$\fit_0 = \fit(\sigma_s \gamma / \beta)$;
+but the standard deviation is of order $z$,
+so that up to lower order terms, 
+while the typical fitness of an individual in the original population is
+$\fit_0 = \exp(-(\sigma_s/\gamma)^2)$;
 of an $F_1$ is
-$\fit_1 = \fit((\sigma_s \gamma + \gamma z^2) / \beta)$;
+$\fit_1/\fit_0 = \exp(-(c_1 \sigma_N^2 T/N_e)^2)$;
 and of an $F_2$ is
-$\fit_2 = \fit(\gamma z / \beta)$.
-
-If $\fit(x) = e^{-x^2} \approx 1 - x^2$,
-then ...
-\plri{reparameterize to define $\fit_1$ as the \emph{drop} in fitness?}
+$\fit_2/\fit_0 = \exp(- T/(N_e\gamma^2))$.
 
@@
 \begin{align}
-    \mathcal{F}\left( \widehat{\phi}(t) \right) = \exp \left\{- \frac{1}{\sigma} \int_{0}^{\infty} \left\lVert \phi(t) - \widehat{\phi}(t) \right\rVert^{2} dt  \right\}  .
+    \fit\left( \widehat{\phi}(t) \right) = \exp \left\{- \frac{1}{\sigma} \int_{0}^{\infty} \left\lVert \phi(t) - \widehat{\phi}(t) \right\rVert^{2} dt  \right\}  .
 \end{align}
@@
 \begin{align*}
-      \mathcal{I} = \frac{2 \left\langle \mathcal{F} \left(\phi_{F_{1}}\right) \right\rangle}{\left\langle \mathcal{F} \left(\phi_{P_{1}}\right) \right\rangle +  \left\langle \mathcal{F} \left(\phi_{P_{2}}\right) \right\rangle} ,
+      \Incompat = \frac{2 \left\langle \fit \left(\phi_{F_{1}}\right) \right\rangle}{\left\langle \fit \left(\phi_{P_{1}}\right) \right\rangle +  \left\langle \fit \left(\phi_{P_{2}}\right) \right\rangle} ,
 \end{align*}
@@
 \begin{align*}
-  \mathcal{F} \left(\phi_{P_{1}}\right) = \mathcal{F} \left( \phi_{P_{2}} \right) &= 1 \\
-  \mathcal{F}\left(\phi_{F_{1}}\right) &= 0 \\
+  \fit \left(\phi_{P_{1}}\right) = \fit \left( \phi_{P_{2}} \right) &= 1 \\
+  \fit\left(\phi_{F_{1}}\right) &= 0 \\
   \mathcal{I} &= 0
@@
 
-In general terms, if fitness $\mathcal{F}$ is some function $g$ of phenotypic divergence from an optimal scaled by the strength of selection $\beta$,
-\begin{align*}
-  \mathcal{F} = g \left( \frac{\phi - \phi_{0}}{\beta} \right)
+In general terms, if fitness $\fit$ is some function $g$ of phenotypic divergence from an optimal scaled by the strength of selection $\beta$,
+\begin{align*}
+  \fit = g \left( \frac{\phi - \phi_{0}}{\beta} \right)
 \end{align*}
@@
    \begin{align*}
-     \Delta \mathcal{F}_{1} = \left( \frac{\Delta \Phi_{1}}{\beta} = \frac{c_{1} \sigma^{2} \frac{T}{N_{e}}}{\beta} \approx \frac{T}{N_{e}} \right)
+     \Delta \fit_{1} = \left( \frac{\Delta \Phi_{1}}{\beta} = \frac{c_{1} \sigma^{2} \frac{T}{N_{e}}}{\beta} \approx \frac{T}{N_{e}} \right)
 \end{align*}
@@
   \Delta \Phi^{2}_{2} &= c_{2}^{2} \left( \sigma \sqrt{\frac{T}{N_{e}}} \right)^{2} = c_{2}^{2} \frac{T}{N_{e}} \\
-  \Delta \mathcal{F}_{2} &= \left( \frac{\Delta \Phi_{2}}{\beta} = \frac{c_{2} \sigma \sqrt{\frac{T}{N_{e}}}}{\beta} \approx \sqrt{\frac{T}{N_{e}}} \right)
+  \Delta \fit_{2} &= \left( \frac{\Delta \Phi_{2}}{\beta} = \frac{c_{2} \sigma \sqrt{\frac{T}{N_{e}}}}{\beta} \approx \sqrt{\frac{T}{N_{e}}} \right)
 \end{align*}
@@
 
+\section{Away from the optimum}
+\label{apx:away_from_opt}
+
+Let two points  on $\optx$ be $x_1$ and $x_2$, let $\bar x = (x_1+x_2)/2$, and let $z=(x_2 - x_1)/2$.
+Then with $D \Phi$ and $D^2 \Phi$ the first and second derivatives of $\Phi$, respectively,
+then Taylor expanding about $x_1$ and $x_2$ finds that
+\begin{align*}
+    \Phi(\bar x) 
+    &= \Phi(x_1) + D\Phi(x_1) \cdot z + \frac{1}{2} z^T D^2 \Phi(x_1) z + O(\|z\|^3) \\
+    &= \Phi(x_2) - D\Phi(x_2) \cdot z + \frac{1}{2} z^T D^2 \Phi(x_2) z + O(\|z\|^3) .
+\end{align*}
+Now, since $\Phi(x_1) = \Phi(x_2) = \optph$ and
+\begin{align*}
+    D\Phi(x_2) = D\Phi(x_1) + 2 z^T D^2 \Phi(x_1) + O(\|z\|^2), \quad \text{and} \\
+    D^2\Phi(x_2) = D^2\Phi(x_1) + O(\|z\|), \quad \text{and} \\
+\end{align*}
+adding together the two equations above and dividing by two gets that
+\begin{align*}
+    \Phi(\bar x) 
+    &= \optph - \frac{3}{2} z^T D^2 \Phi(x_1) z + O(\|z\|^3) .
+\end{align*}
+
 \section{Differentiating the fitness function}
+\label{apx:H_calc}
 

### 2017-08-04 kalman_walk_paper.tex
removed handnotes files (moved to kalman_walk_paper) and added notes on neutral directions to kalman_walk_paper

@@
 
+\subsection{Neutral directions from the Kalman decomposition}
+
+The Kalman decomposition above says that any system $(A,B,C)$ can be decomposed into
+$(P, \hat A, \hat B, \hat C)$ so that
+$$\begin{aligned}
+    A &= P^{-1} \hat A P  \\
+    B &= P^{-1} \hat B  \\
+    C &= \hat C P ,
+\end{aligned}$$
+and we know precisely how we can change these to preserve the transfer function:
+\begin{enumerate}
+    \item $P \to P + \epsilon Q$ as long as the result is still invertible,
+    \item $\hat A \to A + \epsilon X$ as long as $X$ is zero in the correct places,
+    \item $\hat B \to B + \epsilon Y$ as long as $Y$ is zero in the correct places,
+    \item $\hat C \to C + \epsilon Y$ as long as $Z$ is zero in the correct places.
+\end{enumerate}
+By taking $\epsilon \to 0$, these tell us the local directions we can move a system $(A,B,C)$ in.
+All statements below are up to first order in $\epsilon$,
+omitting terms of order $\epsilon^2$.
+
+First, since $(P + \epsilon Q)^{-1} = P^{-1} + \epsilon P^{-1} Q P^{-1}$,
+modifying $P \to P + \epsilon Q$ changes
+$$\begin{aligned}
+    A 
+        &\to A + \epsilon P^{-1} \hat A Q - \epsilon P^{-1} Q P^{-1} \hat A P \\
+        &= A + \epsilon \left(A P^{-1} Q - P^{-1} Q A\right) , \\
+    B
+        &\to B - \epsilon P^{-1} Q B \\
+    C
+        &\to C + \epsilon C P^{-1} Q .
+\end{aligned}$$
+
+Then, $\hat A \to A + \epsilon X$  does
+$$\begin{aligned}
+    A \to A + \epsilon P^{-1} X P 
+\end{aligned}$$
+and $\hat B \to B + \epsilon Y$ does
+$$\begin{aligned}
+    B \to B + \epsilon P^{-1} Y
+\end{aligned}$$
+and $\hat C \to C + \epsilon Z$ does
+$$\begin{aligned}
+    C \to C + \epsilon Z P .
+\end{aligned}$$
+
+This gives us an upper bound on the number of degrees of freedom.
+However, some of these may be redundant.
+For instance, changing $P$ in the direction of 
+a $Q$ that satisfies both $A P^{-1} Q = P^{-1} Q A$ and $C P^{-1} Q = 0$
+is equivalent to changing $B$ by $Y = QB$.
+
 \section{Genetic drift with a multivariate trait}

### 2017-08-04 kalman_walk_paper.tex
another note

@@
 \end{aligned}$$
+Since $P$ is invertible and $Q$ can be anything (if $\epsilon$ is small enough),
+this allows changes in the direction of an arbitrary $W$:
+$$\begin{aligned}
+    A 
+        &= A + \epsilon \left(A W - W A\right) , \\
+    B
+        &\to B - \epsilon W B \\
+    C
+        &\to C + \epsilon C W .
+\end{aligned}$$
 
@@
 \end{aligned}$$
+These degrees of freedom depend on $P$, but both through the change of basis $P$
+and the form of allowable $X$, $Y$, and $Z$.
 
 This gives us an upper bound on the number of degrees of freedom.
+\plri{maybe? depending on if choice of $P$ affects anything.}
 However, some of these may be redundant.

### 2017-08-04 kalman_walk_paper.tex
another small note

@@
 \end{aligned}$$
-These degrees of freedom depend on $P$, but both through the change of basis $P$
-and the form of allowable $X$, $Y$, and $Z$.
-
-This gives us an upper bound on the number of degrees of freedom.
-\plri{maybe? depending on if choice of $P$ affects anything.}
+These degrees of freedom look like they depend on $P$, 
+which is not unique,
+but for any two choices of $P$ there are corresponding choices of $X$
+that give the same actual change in $A$ (and likewise for $Y$ and $Z$).
+
+Therefore, this gives us an upper bound on the number of degrees of freedom,
+in terms of the dimensions of the blocks in the Kalman decomposition ($n_\ro$ etc)
+and the dimensions of $B$ and $C$ ($n_B$ and $n_C$ respectively):
+namely, for $W$, $X$, $Y$, and $Z$ respectively:
+$$\begin{aligned}
+    n^2 
+    + (n_{\rno} + n_\ro n_\nro + n_\nrno(n_\nrno + n_\nro) + n_\nro^2)
+    + n_B n_\rno
+    + n_C n_\nrno .
+\end{aligned}$$
 However, some of these may be redundant.

### 2017-08-09 kalman_walk_paper.tex
edited and replied to some comments.

@@
     \newcommand{\plr}[1]{\todo[color=blue!25]{#1}}
+    \newcommand{\js}[1]{\todo[color=green!25]{#1}}
     % inline comment version
@@
 
-%\begin{multicols}{2}
-
 \begin{abstract}
-Here we introduce an analytical theory to study the evolution of biological systems, such as gene regulatory networks.
+We introduce an analytical theory to study the evolution of biological systems, such as gene regulatory networks.
 The evolutionary conservation of phenotype under selective and environmental stasis does not necessitate conservation of the underlying mechanism, as distinct molecular pathways can realize identical phenotypes.
@@
 %%%%%%%%%%%%%%%%%%%%%%
+%\begin{multicols}{2}
 \section*{Introduction}
@@
 Progress in a suite of biological subdisciplines is stalled by our general lack of understanding of this molecular machinery: with respect to both its function and evolution. 
-\plr{Saying "stalled" reflects negatively on other fields - rephrase?}
+  \plr{Saying "stalled" reflects negatively on other fields - rephrase?} \js{I don't think it reflects negatively. Even if it did, as long as it's a true statement it should be fine.}.
 There does exist a growing body of data on the evolutionary histories and molecular characterizations of particular gene regulatory networks \citep{jaeger2011gap, davidson2006gene, israel2016comparative}, as well as thoughtful verbal and conceptual models \citep{true2001developmental, gwagner1, weiss2000phenogenetic, edelman2001degeneracy}.
-\plr{I don't think the HW example will speak to our audience.  Also, better to say why math is good rather than why not-math is bad.}
-However, as Hardy and Weinberg taught us over a century ago, verbal theories are often insufficient, if not downright misleading \citep{hardy1908mendelian, weinberg1908vererbungsgesetze, servedio2014not}.
+  \plr{I don't think the HW example will speak to our audience.
+Also, better to say why math is good rather than why not-math is bad.} \js{deleted HW.
+I don't think we need to write an anodyne -- the current sentence is true enough.}
+However, verbal theories are often insufficient, if not downright misleading \citep{servedio2014not}. 
 This is especially pertinent given the staggering complexity and scope of contemporary research programs. 
@@
 only mathematics will suffice, as models allow the development of concrete numerical predictions. 
-Previously it has been suggested that any idealized study of evolution is incomplete 
-without a mathematically sufficient description of the genotype, phenotype, and transformation from one to the other \citep{Lewontin1974genetic}. 
+%Previously it has been suggested that any idealized study of evolution is incomplete 
+%without a mathematically sufficient description of the genotype, phenotype, and transformation from one to the other \citep{Lewontin1974genetic}.
 \plr{probably some more recent Wagner papers in this line as well?}
@@
 causal structure, even with access to complete and perfect data \citep{bellman1970structural, grewal1976identifiability, walter1984structural}. 
-\plr{not clear what you mean by "actually"}
+%\plr{not clear what you mean by "actually"}\js{I think we discussed it's clear enough for an intro?}
 Models with different parameter schemes, or even different mechanics can be equally accurate, 
@@
 In this paper we outline a theoretical framework to study the evolution of biological systems. 
-\plr{rather than "we expect" maybe say that the framework could be applied to non-neutral evolution}
-We expect wide applicability to this approach (\emph{i.e.} non-neutral evolution), 
-however presently, we focus solely on neutral evolution, 
-\plr{here's the definition of neutral, but it's not quite right?}
+  \plr{rather than "we expect" maybe say that the framework could be applied to non-neutral evolution}\js{fixed}
+Presently, we focus solely on a neutral scenario, 
 that is where phenotype is conserved over evolutionary time. 
-
-\plr{We don't "derive" it, we point out that Kalman already derived it.}
-We derive an analytical description of the set of all linear biological systems with identical phenotypes 
-\plr{We need something here saying what we mean by "gene network architecture".  Also, I don't think we claim anything for "all biological systems".}
--- that is we describe the set of all gene network architectures that yield identical phenotypes, and show that all biological systems can, in principal, can undergo systems drift. 
+however, this framework could be applied to a wider set of evolutionary scenarios. 
+\plr{here's the definition of neutral, but it's not quite right?} \js{how's that?}
+
+We present an analytical description of the set of all linear biological systems with identical phenotypes 
+\plr{We need something here saying what we mean by "gene network architecture".  Also, I don't think we claim anything for "all biological systems".} \js{added the word ``linear.''}
+-- that is we describe the set of all gene network architectures that yield identical phenotypes, and show that all linear biological systems can, in principal, can undergo systems drift. 
 In the neutral case, this set describes a manifold that evolution explores leaving phenotype invariant with respect to mutation, 
@@
 %}
-Here we outline a method to model biological systems, such as gene regulatory networks, as linear dynamical systems. 
-We define an organism's ``phenotype'' to be the time dynamics of molecular concentrations directly relevant to survival, 
-\plr{tried to say it more precisely. better phrase than "component of phenotype"?}
-and denote by $\phi_k(t)$ the concentration of the $k^\text{th}$ component of phenotype at time $t$.
+\js{Will rewrite this section again.} Here we outline a method to model biological systems, such as gene regulatory networks, as linear dynamical systems. 
+We define an organism's ``phenotype'' to be the temporal dynamics of molecular concentrations directly relevant to survival, 
+and denote by $\phi_k(t)$ the concentration of the $k^\text{th}$ element of the phenotype at time $t$, given an input; for instance $\phi_k(t)$ could describe the concentrations of a particular enzyme.
 These dynamics are the result of the interconnections of 
 a gene regulatory network (or some other biological system) $A$ (an $n \times n$ matrix) 
-and an environmental input $u(t)$.
-\plr{say what $u$ is (a vector)}
-\plr{say in words what $\kappa$ is before the equation.}
+and an environmental input vector $u(t)$, and the \emph{kryptotype} $\kappa(t)$, is a list of a system's molecular concentrations at time $t$ -- or more generally the systems \emph{internal dynamics}.
 Such a system $\mathcal{S}$, is composed of two equations,
@@
   \end{align}
-Where the \emph{kryptotype} $\kappa(t)$, is a list of a system's molecular concentrations at time $t$ -- or more generally the systems \emph{internal dynamics}.
-As some system dynamics may not be visible nor relevant to selection, we distinguish the \emph{kryptotype} (as it is hidden) from the \emph{phenotype} -- which is the systems \emph{external dynamics}.
-The change in internal molecular concentrations as a function of time $\dot{\kappa}(t)$ is simply a function of the systems current state (or kryptotype), the system architecture $A$ -- such as a gene regulatory network, the environmental input $u(t)$, and $B$ (an $n \times l$ matrix) -- how the system processes its environment.
-\plr{Could say instead of "filters" that e.g. $C_{ij}$ is the amount by which the $j$th input $u_j$ affects the production of the $i$th transcription factor concentration.}
-$C$ (an $l \times n$ matrix) filters the dynamics relevant to survival.
+As some system dynamics may not be visible nor relevant to selection, we distinguish the \emph{kryptotype} (as it is hidden) from the \emph{phenotype} -- which is the system's \emph{external dynamics}.
+The change in internal molecular concentrations as a function of time $\dot{\kappa}(t)$ is simply a function of the system's current state (the kryptotype), the system architecture $(A,B,C)$ -- such as the wiring of a gene regulatory network, and the environmental input $u(t)$ to the system.
+The architecture of $A$ determines exactly how the internal molecular concentrations regulate one another's expression as $A_{ij}$ is the magnitude at which transcription factor $\kappa_j(t)$ regulates transcription factor $\kappa_i(t)$.
+If $A_{ij} > 0$, we say that $\kappa_j$ upregulate $\kappa_i$, and if $A_{ij} < 0$, we say that $\kappa_j$ down-regulates $\kappa_i$.
+Furthermore, we interpret the $i^\text{th}$ row of $A$ to be the \emph{cis}-regulatory element (or promoter) for gene $i$.
+The architecture of $B$ determines how the system processes the environmental intput, and the structure of $C$ converts the internal dynamics to external dynamics.
+That is if $\mathcal{S}$ is a metabolic system, $C_{ij}$ is the amount at which the $j$th input metabolite $u_j$ affects the production of the $i$th enzyme concentration.
+
 \plr{This "Thus" needs expanding, e.g., "the solution to this equation is unique and given by" with a reference.}
 Thus the phenotype is simply a convolution of the system organization and the environment, 
-\plr{You've missed out $\kappa(0)$. Also, you can't have $t$ be both a limit of the integral and the variable being integrated over -- check this and get it right.}
   \begin{align}
-    \phi(t) = \int_{0}^{t} C e^{A t} B u(t) dt ,
+    \phi(t) = C e^{A t} \kappa(0) + \int_{0}^{t} C e^{A (t-s)} B u(s) ds ,
   \end{align}
-where we sometimes refer to $h(t) := Ce^{A t}B$ as the \emph{impulse response} of the system.
-
-\plr{How about expanding this minor comment into a short paragraph with more interpretation of not just $A$ but also $B$ and $C$? Should say why. 
-Include talking through why $A_{ij}>0$ means that $j$ "upregulates" $i$.}
-As $A$ specifies gene regulatory network architecture, the $i^\text{th}$ row of $A$ can be interpreted as the \emph{cis}-regulatory element (or promoter) for gene $i$.
-
+where we refer to $h(t) := Ce^{A t}B$ as the system's \emph{impulse response}. 
 %We will now lay out the model in more general terms.
@@
 
-  \begin{example}[Oscillating Gene Network: Cell Cycle Control]\label{ex:oscillator}
-
-    Cellular division is a complicated phenomena, governed by many different processes, however it is agreed that its rhythm is partially controlled by periodic (oscillating) gene transcription \citep{orlando2008global}.
-Consider a simplified model of oscillating gene transcription.
-In the present framework periodic expression requires at minimum two interacting genes.
-
-    Suppose gene-2 up-regulates the transcription of gene-1 and that gene-1 down-regulates the transcription of gene-2 with equal magnitude (of $1$) and relative to each of their concentrations, denoted by $\kappa_{1}$ and $\kappa_{2}$.
-Furthermore, suppose that only the dynamics of gene-1 are consequential to the cell cycle (perhaps the amount of gene-1 activates another downstream gene network).
-Lastly suppose that the production of both genes is stimulated by an impulse of a molecule present immediately after division.
-
-\plr{you haven't explained $u(t)$ above. and, isn't it more natural in this set-up to have $B = [1,0]$ than $[1,1]$?}
-    If the rate each of these genes is expressed is a linear function of their concentrations, the dynamics of the system are given by
-    \begin{align*}
-      \dot{\kappa_{1}}(t) &= \kappa_{2}(t) + u(t) \\
-      \dot{\kappa_{2}}(t) &= - \kappa_{1}(t) + u(t)
-    \end{align*}
-    where $\dot{\kappa}$ denotes the time derivative.
-The initial conditions $\kappa_{1}(0)$ and $\kappa_{2}(0)$, and the input $u(t)$ then determine the concentrations through time.
-If we record the regulatory coefficients in the matrix
-    \begin{align*}
-      A &= \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} ,
-    \end{align*}
-    and define the column vector $B = \begin{bmatrix} 1 & 1 \end{bmatrix}^{T}$, then in matrix notation the dynamics are
-      \begin{align*}
-      \dot{\kappa} &= A \kappa(t) + B u(t) .
-      \end{align*}
-      Since only the dynamics of gene-1 are directly relevant to biological function, the dynamics of interest are given by
-      \begin{align*}
-      \phi(t) &= C \kappa (t)
-      \end{align*}
-      where the row vector $C$ is defined as $C = \begin{bmatrix} 1 & 0 \end{bmatrix}$. 
-      (If the dynamics of both genes were physiologically relevant to the cell cycle, we would set $C$ to be the identity matrix).
-  %  \begin{align*}
-  %    A = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} , \qquad B = \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix}
-  %  \end{align*}
- %       The oscillatory system $\Sigma$ is thus given as
- %   \begin{align*}
- %     \Sigma = \left \{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 
- %       0 & 1 \\ 
- %      -1 & 0 
- %       \end{bmatrix} \kappa(t) + \begin{bmatrix} 1 \\ 1 \end{bmatrix} u(t) \\ 
- %         \phi(t) &= \begin{bmatrix} 1 & 0 \end{bmatrix} \kappa(t) \end{array} \right.
- %    \end{align*}
-
-      Since the input is simply an impulse, its phenotype is equivalent to its impulse response
-      \begin{align*}
-        \phi(t) = h(t) = \sin(t) + \cos(t)  .
-      \end{align*}
-
-    \begin{figure}[H]
-       % \begin{center} 
-      \centering
-         \begin{tabular}{cc}
-            \begin{tikzpicture}
-            \begin{scope}[every node/.style={circle,thick,draw}]
-                \node (A) at (0,0) {$\kappa_{1}$};
-                \node[dashed] (B) at (3,0) {$\kappa_{2}$};
-                \node[shape=rectangle] (U) at (1.5,2) {input};
-                \node[shape=rectangle] (y) at (1.5,-2) {output};
-            \end{scope}
-
-            \begin{scope}[>={Stealth[black]},
-                          every node/.style={fill=white,circle},
-                          every edge/.style={draw=black, thick}]
-                \path [->, >=Rectangle] (A) edge[bend left] node {\tiny $-1$} (B);
-                \path [->] (B) edge[bend left] node {\tiny $1$} (A); 
-                \path[->] (U) edge[dashed] node {\tiny $1$} (A);
-                \path[->] (U) edge[dashed] node {\tiny $1$} (B);
-                \path[->] (A) edge[dashed,bend right] node {\tiny $1$} (y);
-            \end{scope}
-            \begin{scope}[>={Stealth[black]},
-                          every edge/.style={draw=black, thick}]
-                %\path [->] (A) edge[loop left] node {\tiny $\lambda_{1}$} (A);
-                %\path [->] (B) edge[loop left] node {\tiny $\lambda_{2}$} (B);
-            \end{scope}
-
-            \end{tikzpicture} &
-       % \end{center}
-      %  \caption{
-      %      Diagram of Example \ref{ex:osc} in the text.
-      %      \plr{explain what arrows mean if nec}
-      %      \label{fig:oscillator_diagram}}
-   % \end{figure}
-   % \begin{figure}[H]
-    %  \centering
-      \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{osc_impulse}
-   \end{tabular}
-  %  \end{center}
-      \caption{(Left) Graphical representation of the cell cycle control gene network, and (right) plot of the phenotype $\phi(t)$ against time $t$.}
-    \end{figure}
-      We return to the evolution of such a system below.
-  \end{example}
-
-  \section*{Linear Evolutionary Systems}
-\plr{instead title this section "Systems drift for linear systems"?}
-
-\plr{This paragraph feels like the introduction.}
-The literature is filled with detailed observations of molecular systems and their diversity. 
-\plr{Diversity doesn't imply systems drift -- only if the diverse systems are homologous.}
-There are examples of significant diversity in the networks underlying processes such as 
-circadian rhythm \citep{sancar2008intelligent}, 
-cell cycle control \citep{cross2011evolution, kearsey2003enigmatic}, 
-pattern formation \plr{cite?}, 
-and metabolism \citep{lavoie2009rearrangements, martchenko2007transcriptional, dalal2016transcriptional, christensen2011unique, hartl2007induction, alam2013aspergillus}. 
-Despite a symmetry in functionality or phenotype these systems often differ, sometimes substantially, at the molecular level. 
+  \section*{Gene Network Drift}
+
+%\plr{This paragraph feels like the introduction.}
+%The literature is filled with detailed observations of molecular systems and their diversity. 
+%\plr{Diversity doesn't imply systems drift -- only if the diverse systems are homologous.}
+%There are examples of significant diversity in the networks underlying processes such as 
+%circadian rhythm \citep{sancar2008intelligent}, 
+%cell cycle control \citep{cross2011evolution, kearsey2003enigmatic}, 
+%pattern formation \plr{cite?}, 
+%and metabolism \citep{lavoie2009rearrangements, martchenko2007transcriptional, dalal2016transcriptional, christensen2011unique, hartl2007induction, alam2013aspergillus}. 
+Despite a symmetry in functionality or phenotype systems can often differ, sometimes substantially, at the molecular level. 
 How many different mechanisms have the same function?
-We urge the reader to consider the metaphorical black box.
-
-Gene regulatory networks with identical phenotypes (external dynamics) do not necessarily have identical kryptotypes (internal dynamics). 
+
+Gene regulatory networks with identical phenotypes do not necessarily have identical kryptotypes.
 Any linear and minimal system (a gene network) 
 -- minimal, informally meaning that the system's phenotype is achieved with the fewest possible number of genes 
--- has an identical phenotype up to a change of coordinates. 
+-- has an identical impulse response and therefore an identical phenotype given an identical input $u(t)$, up to a change of coordinates. 
 \plr{"has an identical phenotype" is not what you mean to say here}
+\js{edited -- does that work?}
+  \begin{equation}
+    \begin{aligned}
+      h(t) &= C e^{A t} B 
+      = C V^{-1} V e^{A t} V^{-1} V B \\
+      &= C V^{-1} e^{V A V^{-1} t} V B 
+      = \bar{C} e^{\bar{A} t} \bar{B}
+    \end{aligned}
+  \end{equation}
+\plr{what are we calling these -- not "biological system" -- that's too broad}\js{how about ``molecular systems''?}
+Two molecular systems, $\mathcal{S} = \{ A, B, C \}$, and $\bar{\mathcal{S}} = \{\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} \}$, 
+have the same phenotype, for all possible inputs, if they are related by a change of coordinates. 
+\plr{is this the place to say "and only if, in the minimal dimension"?}
+
+%Although systems may not be identifiable beyond a change of coordinates, at present we are primarily interested in a subset of these systems. 
+%That is, systems that not only have equivalent external dynamics, but also equivalent input and output relationships. 
+%\plr{clarify that you mean actually having the same $B$ and $C$. (at first I read "input and output relationship" as the mapping $u \to \phi$),
+%    i.e. behaviour across many possible inputs.}
+%Formally, this means systems related by a change of coordinates (any invertible matrix $V$) that leaves $B$ and $C$ invariant:
+%  \begin{align}
+%    VB &= B \implies \bar{B} = B \\
+%    CV &= C \implies \bar{C} = C
+%  \end{align}
+%\plr{this sentence lacks an object (or some sentence part?)}
+%In other words systems with varying gene regulatory network architectures yet identical selection pressures, environment, and phenotype.
+
+%Define $V(\tau)$ as the parameterized change of coordinates matrix that preserves $B$ and $C$, with $\tau$ a vector of free parameters.
+The set of \emph{all} phenotypically invariant (minimal) gene networks is,
+ % \begin{align}
+ %   A(\tau) = V(\tau) A(0) V^{-1}(\tau) ,
+ % \end{align}
+%and a \emph{Linear Evolutionary System} is, 
+%\plr{why is this "evolutionary"? It is the thing that we're thinking about evolution of, but that doesn't make it evolutionary.}
+Therefore, if a system is minimal, the set of all phenotypically identical systems $\mathcal{S}(V)$ can be parameterized as,
     \begin{align}
-    h(t) &= C e^{A t} B \\
-    &= C V^{-1} V e^{A t} V^{-1} V B \\
-    &= C V^{-1} e^{V A V^{-1} t} V B \\
-    &= \bar{C} e^{\bar{A} t} \bar{B}
+      \mathcal{S}(V) := \left\{ \begin{array}{ll} V\dot{\kappa}(t) &= VA V^{-1} \kappa(t) + VB u(t) \\ \phi(t) &= C V^{-1} \kappa(t) \end{array} \right.
   \end{align}
-\plr{what are we calling these -- not "biological system" -- that's too broad}
-Two biological systems, $\mathcal{S} = \{ A, B, C \}$, and $\bar{\mathcal{S}} = \{\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} \}$, 
-have the same phenotype if they are related by a change of coordinates. 
-\plr{is this the place to say "and only if, in the minimal dimension"?}
-
-Although systems may not be identifiable beyond a change of coordinates, at present we are primarily interested in a subset of these systems. 
-That is, systems that not only have equivalent external dynamics, but also equivalent input and output relationships. 
-\plr{clarify that you mean actually having the same $B$ and $C$. (at first I read "input and output relationship" as the mapping $u \to \phi$),
-    i.e. behaviour across many possible inputs.}
-Formally, this means systems related by a change of coordinates (any invertible matrix $V$) that leaves $B$ and $C$ invariant:
-  \begin{align}
-    VB &= B \implies \bar{B} = B \\
-    CV &= C \implies \bar{C} = C
-  \end{align}
-\plr{this sentence lacks an object (or some sentence part?)}
-In other words systems with varying gene regulatory network architectures yet identical selection pressures, environment, and phenotype.
-
-\plr{I don't think it makes sense to write things in general as a function of $\tau$,
-    since in general the set of "free parameters" is complicated and we don't actually specify it.
-    Better, maybe define $\mathcal{V}$ to be the set of all $V$ that preserve $B$ and $C$, then define $A(V)$ instead of $A(\tau)$...}
-Define $V(\tau)$ as the parameterized change of coordinates matrix that preserves $B$ and $C$, with $\tau$ a vector of free parameters.
-The set of \emph{all} phenotypically invariant (minimal) gene networks is,
-  \begin{align}
-    A(\tau) = V(\tau) A(0) V^{-1}(\tau) ,
-  \end{align}
-and a \emph{Linear Evolutionary System} is, 
-\plr{why is this "evolutionary"? It is the thing that we're thinking about evolution of, but that doesn't make it evolutionary.}
-  \begin{align}
-    \mathcal{S}(\tau) := \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A(\tau) \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right .
-  \end{align}
-\plr{again, lacking an object}
-That is, all (linear and minimal) mechanisms capable of producing the same phenotype can be realized by a unique choice of $\tau$ in $\mathcal{S}(\tau)$.
-
-More generally, \plr{introduce Kalman},
+  where $V$ is any invertible matrix and it's elements are free parameters. 
+More generally,
 we denote by $\calA_n(A_0)$ the set of all $n$-dimensional systems equivalent to $A_0$:
@@
 Regardless of minimality, two systems, even in different dimensions, can have identical external dynamics if they are in $\mathcal{A}$.
-This set can be completely parameterized using the \emph{Kalman Decomposition}; the set of all linear gene regulatory networks with equivalent phenotypes can be precisely defined.
+This set can be completely parameterized using the \emph{Kalman decomposition} \ref{apx:kalman}; the set of all linear gene regulatory networks with equivalent phenotypes can be precisely defined.
 Note that this implies that there is always more than one possible gene regulatory network architecture per phenotype, however we can analytically describe the complete set.
 
-  \begin{example}[All Phenotypically Equivalent Cell Cycle Control Networks]
-
-\plr{this is same as above, so reference instead of writing out again?}
-    Let
-    \begin{align*}
-      A(0) = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix}, \qquad B = \begin{bmatrix} 1 & 1 \end{bmatrix}^{T}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix},
-    \end{align*}
-    with $V(\tau)$ preserving both $B$ and $C$, then
-    the set of all two-gene regulatory networks phenotypically equivalent to the cell cycle control network in Example \ref{ex:oscillator} are given by
-    \begin{align*}
-      A(\tau) &= \begin{bmatrix} 1 & 0 \\ \tau & 1-\tau \end{bmatrix} \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 0 \\ \frac{\tau}{\tau-1} & \frac{-1}{\tau-1} \end{bmatrix}  \qquad \forall \ \tau \neq 1 %\\
-     %   B &= \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix} \\
-       %   h(t) &= \sin(t) + \cos(t) \qquad \forall \ \tau \neq 1
-    \end{align*}
-    \begin{figure}[H]
-    \centering
-% Nice tikz!!!
-\begin{tikzpicture}
-\begin{scope}[every node/.style={circle,thick,draw}]
-  \node (A) at (0,0) {$\kappa_{1}$};
-    \node[dashed] (B) at (4,0) {$\kappa_{2}$};
-    \node[shape=rectangle] (U) at (2,2) {input};
-    \node[shape=rectangle] (y) at (2,-2) {output};
-\end{scope}
-
-\begin{scope}[>={Stealth[black]},
-              every node/.style={fill=white,circle},
-              every edge/.style={draw=black, thick}]
-    \path [->, sloped] (A) edge[bend left] node {\tiny $2 \tau + (\tau-1)^{-1}$} (B);
-    \path [->, sloped] (B) edge[bend left] node {\tiny $-(\tau-1)^{-1}$} (A); 
-    \path[->] (U) edge[dashed] node {\tiny $1$} (A);
-    \path[->] (U) edge[dashed] node {\tiny $1$} (B);
-    \path[->] (A) edge[dashed, bend right] node {\tiny $1$} (y);
-\end{scope}
-\begin{scope}[>={Stealth[black]},
-              every edge/.style={draw=black, thick}]
-    \path [->] (A) edge[loop left] node[sloped, anchor=center, above] {\tiny $1 + (\tau-1)^{-1}$} (A);
-    \path [->] (B) edge[loop right] node[sloped, anchor=center, above] {\tiny $-1 - (\tau-1)^{-1}$} (B);
-\end{scope}
-
-\end{tikzpicture}
-    \caption{A graphical depiction of the set of all externally equivalent cell cycle control networks, $A(\tau)$. 
-        $\tau$ can take any value \plri{sentence fragment}}
-    \end{figure}
- % \end{example}
-
- % \begin{example}[External Equivalence does not imply internal equivalence]
-   Despite the phenotypic equivalence of all instantiations of $A(\tau)$, the internal dynamics, or kryptotypes, vary as a function of $\tau$. 
-    Gene-1 dynamics (blue) are equivalent for network architectures $A(0)$ and $A(2)$, however the dynamics of gene-2 (orange) differ with $\tau$.
-  \begin{figure}[H]
-    \centering
-    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{osc_A0_A2_both_compare}
-    \caption{Gene-1 (blue) and gene-2 (orange) dynamics for $A(0)$ (top) and $A(2)$ (bottom).
-Both (top and bottom) gene-1 dynamics are given by $ \kappa_{1} = \sin(t) + \cos(t)$, and gene-2 by $\kappa_{2} = \cos(t) - \sin(t)$ (top) and $\kappa'_{2} = \cos(t) + 3 \sin(t)$ (bottom).}
-  \end{figure}
-\end{example}
-
-\plri{Missing: a statement about when there are neutral directions, and how many.}
-
-\plr{I think we should first put the section about quantitative drift in, and then the following section that tries to get at real values of parameters.}
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\subsection*{Speciation}
+
+%Define reproduction in diploids as first, the recombination of unlinked genes to make gametes, and second, as the averaging of two individual parental gametes to produce an offspring.
+Assuming parental populations are both phenotypically identical and genetically homogenous within each population, first generation hybrids (F1s) can be computed by averaging the two parental gene networks.
+Second generation hybrids (F2s) can be computed by first swapping the genes between the two parental gene networks, and next averaging these hybrid gametes.
+
+%Speciation, or the formation of new species, is often a consequence of reprdouctive incompatibility among populations -- usually via hybrid sterility or inviability.
+To discuss speciaiton we first need to describe how reproduction is modelled.
+ %A diploid first generation hybird's ($F_1$'s) genetic regulation is a consequence of its two genomes, as a copy of genes and regulatory sequences from each parent is present.
+%\plr{should have a stronger justification for the assertion that $F_1$ is the arithmetic average of the parents than "Thus we say"}
+%A diploid organism contains two gene network copies: one from each parent.
+As such, a diploid's system dynamics is computed by averaging the coefficients of both systems.
+To reproduce, each organism passes on a recombined haploid gene network to its offspring.
+Haploid gene networks swap system matrix rows randomly between its own two networks.
+%\plr{I don't think we say that $A$ is a "gene network architecture".}
+%Therefore the phenotypic dynamics typical of an $F_{1}$ cross between two allopatric populations will be determined simply by averaging it's two parental genomes.
+However, in second generation hybrid crosses $F_2$ between allopatric populations, first new haploid systems will be formed by recombination -- in the process shuffling and combining regulatory coefficients from allopatric populations -- and then brought together to form a diploid.
+
+%\begin{align}
+%    F_{1}(\tau, \hat{\tau}) &= \frac{ A(\tau) + A(\hat{\tau}) }{2}  .
+%   h(t, F_{1}) &= C e^{\left( F_{1}\left( \tau, \hat{\tau} \right) t \right)} B
+%\end{align}
+%\begin{align*}
+% G\left(r, \tau, \hat{\tau}\right) = Q(r) A( \tau ) + \left( I - Q(r) \right) A( \hat{\tau} )
+%\end{align*}
+%\begin{align}
+%    F_{2}(i, j) = \frac{G (i) + G(j)}{2}
+%\end{align}
+
+  %\begin{align*}
+  %  h\left(t, F_{2}\left( r, r' \right)\right) = C e^{\left( F_{2}\left(r, r' \right) t \right)} B
+  %\end{align*}
+
+%Fitness $\fit$ is a function (usually a Guassian) of an organism's weighted phenotypic distance from an optimum,
+%  $\fit\left( \widehat{\phi}(t) \right) = \exp \left(- \int_{0}^{t} \frac{\rho(s)}{\beta} \left\lVert \phi(s) - \widehat{\phi}(s) \right\rVert^{2} ds  \right)$
+%where $\rho(t)$ is some weighting function and $\beta$ is a parameter.
+
+%We quantitatively describe the degree of incompatibility between two populations $P_{1}$ and $P_{2}$ as
+%\begin{align*}
+%      \Incompat = \frac{2 \left\langle \fit \left(\phi_{F_{1}}\right) \right\rangle}{\left\langle \fit \left(\phi_{P_{1}}\right) \right\rangle +  \left\langle \fit \left(\phi_{P_{2}}\right) \right\rangle} ,
+%\end{align*}
+%where angled brackets imply averaging.  \plr{averaging over what? And, this is as others define - cite.}
+
+%\plr{This is not part of the definition of $\mathcal{I}$:}
+%\js{should we delete this example ($A(0) \times A(2)$ hybrids)? I like how clean it is -- but I'm not sure it's that important in this context. It's a good proof of principle though.}
+%\jss{$F_1$s created by crossing phenotypically equivalent oscillators $A(0)$ and $A(2)$ have a phenotype of $\phi_{F_{1}}(t) = e^{t}$, 
+%in contrast to both parents, who have $\phi_{P_{1}}(t) = \phi_{P_{2}}(t) = \sin(t) + \cos(t)$. 
+%The hybrid phenotype is significantly different (it does not oscillate and increases infinitely) despite the phenotypic equivalence of the parents.
+%%\plr{don't need these to be big equations.}
+%%\begin{align*}
+%%  \fit \left(\phi_{P_{1}}\right) = \fit \left( \phi_{P_{2}} \right) &= 1 \\
+%%  \fit\left(\phi_{F_{1}}\right) &= 0 \\
+%%  \mathcal{I} &= 0
+%%\end{align*}
+%
+%%Thus if populations $1$ and $2$ are homogeneous $A(0)$ and $A(2)$, respectively, we say that they are completely incompatible as $\mathcal{I} = 0$.
+%
+%  \begin{figure}[H]
+%    \centering
+%    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{expF1}
+%    \caption{$F_1$ hybrid (orange) and parental (blue) phenotypic oscillator dynamics for an $A(0)$ by $A(2)$ cross. The hybrid fails to oscillate and exhibits qualitatively different dynamics.}
+%  \end{figure}
+%  }
+%%     \begin{example}[F1 Reproductive Incompatibility in an Oscillating Gene Network]
+%%       DMI examples...
+%%       \begin{figure}[H]
+%%         \includegraphics[width=0.5\textwidth]{A2_A3_osci_F1_hyb}
+%%       \end{figure}
+%%     \end{example}
+%
+%%     \begin{example}[Not all Networks can Host Incompatibilities]
+%%       convex sets cant have DMIs
+%
+%%       \begin{align*}
+%%         h(t) = 2 e^{- \theta t}
+%%       \end{align*}
+%%       Any non-minimal system with rows summing to $\theta$ is PI.
+Further, these systems are closed under averaging (mating) and row swapping (meiosis), leaving all hybrids optimally fit.
+The set of gene matrices is affine and therefore convex.
+%%     \end{example}
+%
+%
+%\plri{Missing: a statement about when there are neutral directions, and how many.}
+%
+%\plr{I think we should first put the section about quantitative drift in, and then the following section that tries to get at real values of parameters.}
+%
+%
+%
 %%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+\paragraph{Sexual reproduction}
+A diploid first generation hybird's ($F_1$) genetic regulation is a consequence of its two genomes, as a copy of genes and regulatory sequences from each parent is present.
+\plr{should have a stronger justification for the assertion that $F_1$ is the arithmetic average of the parents than "Thus we say"}
+A diploid organism contains two gene network copies: one from each parent.
+As such, a diploid's system dynamics is computed by averaging the coefficients of both systems.
+To reproduce, each organism passes on a recombined haploid gene network to its offspring.
+Haploid gene networks swap system matrix rows randomly between its own two networks.
+Therefore the phenotypic dynamics typical of an $F_{1}$ cross between two allopatric populations will be determined simply by averaging it's two parental genomes.
+However, in second generation hybrid crosses $F_2$ between allopatric populations, first new haploid systems will be formed by recombination -- in the process shuffling and combining regulatory coefficients from allopatric populations -- and then brought together to form a diploid.
+
 \paragraph{Hybridization}
-\plr{I put F1s and F2s here, and $\Incompat$, but with less explanation than below. Merge somehow.}
+\plr{I put F1s and F2s here, and $\Incompat$, but with less explanation than below. Merge somehow.} \js{In progress, see paragraph above.}
 By the arguments above, the means of two allopatric populations separated for $T$ generations
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Speciation via Reproductive Incompatibility}
-
-%Define reproduction in diploids as first, the recombination of unlinked genes to make gametes, and second, as the averaging of two individual parental gametes to produce an offspring.
-Assuming parental populations are both phenotypically identical and genetically homogenous within each population, first generation hybrids (F1s) can be computed by averaging the two parental gene networks.
-Second generation hybrids (F2s) can be computed by first swapping the genes between the two parental gene networks, and next averaging these hybrid gametes.
-
-\plr{what does "equally present" mean?}
- An F1's genetic regulation is a consequence of both of its genomes, where genes and regulatory sequences from both parents are equally present. 
-\plr{should have a stronger justification for the assertion that $F_1$ is the arithmetic average of the parents than "Thus we say"}
-Thus we say that a diploid organism's gene network is simply the average of both of its gene network copies; one from each parent. 
-Further, each haploid parental gene network copy is formed via meiosis 
--- by swapping independent genes (specifically their \emph{cis}-regulatory modules) randomly. 
-Assuming two distinct but genetically homogeneous populations evolving in allopatry meet and form hybrids, 
-the first generation hybrids (F1s) gene network dynamics will be determined by the average of the parental haplotypes. 
-The second generation hybrids (F2s), however, will be the product of a meiosis between parental haplotypes followed by an averaging of gametes. 
-\plr{To agree with the previous section we need to incorporate within-pop variation in this argument.}
-
-\plr{I don't think we say that $A$ is a "gene network architecture".}
-Specifically, $F_{1}$ is the first generation hybrid gene network architecture formed by mating (averaging) $A(\tau)$ and $A(\hat{\tau})$, 
-\plr{mating is not averaging}
-\begin{align}
-    F_{1}(\tau, \hat{\tau}) &= \frac{ A(\tau) + A(\hat{\tau}) }{2}  .
-%   h(t, F_{1}) &= C e^{\left( F_{1}\left( \tau, \hat{\tau} \right) t \right)} B
-\end{align}
+\section*{The rate of speciation under neutrality in allopatric populations}
+
+\plr{separate out here (or above) the difference in genotype of hybrids from the optimum set, the difference in phenotype, and the difference in fitness.}
+
+%In general terms, if fitness $\fit$ is some function $g$ of phenotypic divergence from an optimal scaled by the strength of selection $\beta$,
 %\begin{align*}
-% G\left(r, \tau, \hat{\tau}\right) = Q(r) A( \tau ) + \left( I - Q(r) \right) A( \hat{\tau} )
+%  \fit = g \left( \frac{\phi - \phi_{0}}{\beta} \right)
 %\end{align*}
-and $F_{2}$ is the second generation hybrid gene network architecture formed by gametes $G(i)$ and $G(j)$. 
-\plr{sentences don't begin with "Where", even in math. Also, there isn't "the" $F_2$ - there are many possible.}
-Where each $G$ is formed by randomly swapping rows between $A(\tau)$ and $A(\hat{\tau})$, such that the $i$th  gene comes from $A(\tau)$ ($i$ and $j$ are orthogonal vectors, each element $0$ or $1$, and $i \neq j$).
-\begin{align}
-    F_{2}(i, j) = \frac{G (i) + G(j)}{2}
-\end{align}
-
+%\plr{sentences don't begin with "And"}
+%And a population $P$ is normally distributed around a mean gene network architecture $A(\tau)$ and $\sigma^{2}$ is the intra-population regulatory variance,
+%\plr{what is $T$?}
 %\begin{align*}
-  %  h\left(t, F_{2}\left( r, r' \right)\right) = C e^{\left( F_{2}\left(r, r' \right) t \right)} B
+%  P_{T} \sim \mathcal{N} \left(A(\tau), T \frac{\sigma^{2}}{N_{e}}\right)
 %\end{align*}
-
-The fitness of an organism can be computed by comparing its impulse response with the optimal response,
-\plr{this is either zero or infinite.  apply a weighting function (see appendix)?}
-\begin{align}
-    \fit\left( \widehat{\phi}(t) \right) = \exp \left\{- \frac{1}{\sigma} \int_{0}^{\infty} \left\lVert \phi(t) - \widehat{\phi}(t) \right\rVert^{2} dt  \right\}  .
-\end{align}
-Therefore a hybrid's fitness can be computed by comparing its impulse response with that of its parents.
-
-\begin{definition}[Reproductive Incompatibility]
-\plr{This part doesn't need a definition - we don't need to discuss what is or isn't a "species".}
-    According to Mayr's \emph{Biological Species Concept}, two populations can be considered different species if they are reproductively isolated, meaning crosses between them produce low fitness offspring.
-
-We quantitatively describe the degree of incompatibility between two populations $P_{1}$ and $P_{2}$ as
-\begin{align*}
-      \Incompat = \frac{2 \left\langle \fit \left(\phi_{F_{1}}\right) \right\rangle}{\left\langle \fit \left(\phi_{P_{1}}\right) \right\rangle +  \left\langle \fit \left(\phi_{P_{2}}\right) \right\rangle} ,
-\end{align*}
-where angled brackets imply averaging.  \plr{averaging over what? And, this is as others define - cite.}
-
-\plr{This is not part of the definition of $\mathcal{I}$:}
-F1s created by crossing phenotypically equivalent oscillators $A(0)$ and $A(2)$ have a phenotype of $\phi_{F_{1}}(t) = e^{t}$, 
-in contrast to both parents, who have $\phi_{P_{1}}(t) = \phi_{P_{2}}(t) = \sin(t) + \cos(t)$. 
-The hybrid phenotype is significantly different (it does not oscillate and increases infinitely) despite the phenotypic equivalence of the parents.
-\plr{don't need these to be big equations.}
-\begin{align*}
-  \fit \left(\phi_{P_{1}}\right) = \fit \left( \phi_{P_{2}} \right) &= 1 \\
-  \fit\left(\phi_{F_{1}}\right) &= 0 \\
-  \mathcal{I} &= 0
-\end{align*}
-
-Thus if populations $1$ and $2$ are homogeneous $A(0)$ and $A(2)$, respectively, we say that they are completely incompatible as $\mathcal{I} = 0$.
-\end{definition}
-
+%the population mean architecture will move at rate
+%\plr{what is $\Delta \tau$? Also, if we reframe in terms of $V$ not $\tau$ above will need to reframe this.}
+%\begin{align*}
+%  \Delta \tau = \sigma \sqrt{\frac{T}{N_{e}}}
+%\end{align*}
+%where $T$ is time measured in generations. 
+%USING GENERAL MATH FROM MULTIVARIATE TAYLOR SERIES....
+we know that the phenotype of a first generation $F_{1}$ hybrid diverges from that of its parents (assuming both parents are at an optimum and phenotypically equivalent) quartically as a function of $\Delta \tau$.
+%Thus the squared difference of it's phenotype will be,
+%\begin{align*}
+%  \Delta \Phi^{2}_{1} = c_{1}^{2} \sigma^{4} \left( \frac{T}{N_{e}} \right)^{2}
+%\end{align*}
+%and $F_{1}$ fitness will decline linearly with respect to $\frac{T}{N_{e}}$. ARGUMENT FOR WHY $\sigma^{2} c_{1} \sim \beta$ HERE.
+%   \begin{align*}
+%     \Delta \fit_{1} = \left( \frac{\Delta \Phi_{1}}{\beta} = \frac{c_{1} \sigma^{2} \frac{T}{N_{e}}}{\beta} \approx \frac{T}{N_{e}} \right)
+%\end{align*}
+% It follows FROM THE SAME MATH ABOUT THE MULTIVAR TAYLOR that the phenotypes of second generation $F_{2}$ hybrid crosses diverges quadratically as a function of $\Delta \tau$, 
+%\begin{align*}
+%  \Delta \Phi^{2}_{2} &= c_{2}^{2} \left( \sigma \sqrt{\frac{T}{N_{e}}} \right)^{2} = c_{2}^{2} \frac{T}{N_{e}} \\
+%  \Delta \fit_{2} &= \left( \frac{\Delta \Phi_{2}}{\beta} = \frac{c_{2} \sigma \sqrt{\frac{T}{N_{e}}}}{\beta} \approx \sqrt{\frac{T}{N_{e}}} \right)
+%\end{align*}
+%and that $F_{2}$ fitness drops rapidly -- at approximately rate $\sqrt{\frac{T}{N_{e}}}$. ARGUMENT FOR WHY $c_{2} \sigma \sim \beta$ HERE.
+
+This suggests that reproductively isolated populations can speciate rapidly -- on timescales much less than $N_{e}$ generations, depending on the specifics of the fitness function.
+
+%%%%%%%%%%%%%%%
+
+%    \subsection*{Metabolic Network}
+%  \begin{align*}
+%    \dot{x}(t) &= \begin{bmatrix} 0 & 1 & 0 & -1 \\ 0 & 0 & -1 & -1 \\ -1 & 1 & 0 & 0 \\ 0 & 0 & 0 & 0 \end{bmatrix} \begin{bmatrix} \text{GAL3} \\ \text{GAL4} \\ \text{GAL80} \\ \text{MIG1} \end{bmatrix}(t) \\ &+ \begin{bmatrix} 1 & 0 & 0 \\ 0 & 0 & 1 \\ 0 & 0 & 0 \\ 0 & 1 & 1 \end{bmatrix} \begin{bmatrix} \text{galactose} \\ \text{glucose} \\ \text{BTM} \end{bmatrix}(t) \\
+%        y(t) &= \begin{bmatrix} 0 & 1 & 0 & -1 \end{bmatrix} \vec{x}(t)
+%    \end{align*}
+%    \begin{align*}
+%      H(z) &= \begin{bmatrix} 1 \\ z^{2} + z + 1 \\ z + 1 \end{bmatrix} \frac{1}{z^{3} + z -1}
+%    \end{align*}
+%
+%    \begin{align*}
+%      CV &= C \\
+%      VB &= B \\
+%      V :&= \begin{bmatrix} 1 & 0 & a & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & b & 0 \\ 0 & 0 & 0 & 1 \end{bmatrix} , V^{-1} = \begin{bmatrix} 1 & 0 & \frac{-a}{b} & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & \frac{1}{b} & 0 \\ 0 & 0 & 0 & 1\end{bmatrix} \\
+%        A(a,b) :&= VAV^{-1} = \begin{bmatrix} -a & a+1 & \frac{a^{2}}{b} & -1 \\ 0 & 0 & \frac{-1}{b} & -1 \\ -b & b & a & 0 \\ 0 & 0 & 0 & 0 \end{bmatrix}
+%    \end{align*}
+%    The GAL regulatory network in \emph{S.
+cerevisiae} is modelled here.
+The activating transcription factor GAL4, which is regulated by the presence of galactose and two other TFs, binds to the promoter region of the GAL regulon -- a cluster of genes regulated by the same promoter encoding 3 enzymes (GAL1, GAL7, and GAL10), for the metabolism of the sugar galactose.
+The repressing transcription factor MIG1, which is activated by the presence of glucose and other proteins omitted here, also binds to the regulon, preventing the expression of the galactose metabolizing enzymes.
+This network is one of the most studied gene regulatory networks in yeast, and experiments have already demonstrated significant transcriptional variation among different species and genuses of yeast.
+%
+%    In the present model, the $A$ matrix encodes the interactions among GAL3, GAL4, GAL80, and MIG1.
+$B$ interprets the input: the quantity of glucose, galactose, and the basal transcription (BTM) rates impact of TF concentrations.
+$C$ represents the promoter region of the GAL regulon and is the sum of the influence of GAL4 (activating) and MIG1 (repressing).
+%
+%    $A(a,b)$ represent alternative network structures that produce identical outputs, with identical input and output transformations -- meaning the regulatory contributions of the sugars and the basasl machinery, as well as the promoter of the GAL regulon are held constant.
+These interactions may be realized in nature, if the regulatory differences predicted by $A(a,b)$ are biologically/mutationally realizable.
+Mutations influencing the interactions between GAL4 and GAL80 have been experimentally demonstrated \citep{li2010alterations}.
+%
+%    \subsection*{Gap Gene Network}
+
+%%%%%%%%%%%%%%%%%%%%%
+\section*{Discussion}
+
+%
+%\jss{Discussion guidelines:
+%\begin{itemize}
+%    \item Why is this important/useful?
+%    \item What are the assumptions and shortcomings of the research?
+%    \item Compare to other studies in the literature. 
+%    \item Future directions. 
+%    \item Wild speculations?
+%    \item Conclusion and overall impact. 
+%\end{itemize}}
+%
+
+The complexity of biological systems has limited our understanding of their function and evolution.
+Above we outline an approach, a first step, towards untangling this complexity in reference to function and evolution.
+This methodology borrows successfully applied tools from engineering and aims to synthesize these with the concepts and tools of molecular and evolutionary biology.
+
+Theoretical models in evolution and population genetics often lack the molecular details of physiology or of the genotype-phenotype map. 
+Here, we offer a tractable and simple model which includes these missing features. 
+\plr{this is a claim we are the first to do something.  best not to make these claims (and we aren't the first to make math models of these)}\js{removed the word ``only''}
+Further, we provide, in clear mathematical language, an analytical description of phenomena hitherto discussed verbally and conceptually 
+(phenogenetic drift \citep{weiss2000phenogenetic}, developmental systems drift \citep{true2001developmental}, biological degeneracy \citep{edelman2001degeneracy}, \emph{etc.}). 
+The tractability and relative simplicity of this exposition enables the interested biologist to work out by hand, 
+if desired, 
+\plr{probably not desired}\js{:p}
+the dynamics of a genetic system, as well as perturbations to the system -- an attribute not likely to be found in less tractable models and simulations.
+
+We have suggested an interpretation of system identification: to see it as an evolutionarily neutral manifold, and not simply a computational nuisance. 
+We have demonstrated a method to analytically determine the set of all phenotypically invariant gene networks; 
+by a simple change of coordinates in the minimal configuration, or more generally by applying the Kalman decomposition in higher dimensions. 
+Further, we emphasize that evolution proceeds through this high dimensional space as stochastic coordinate transformation, constrained by sexual reproduction and selection. 
+This set is explored over evolutionary time when phenotype is conserved, and can lead to a diverse set of consequences, 
+including the accumulation of Dobzhansky-Muller incompatibilities. 
+\plr{check original Bateson/DM papers to see if this accords with those defns}\js{I read through Orr's review of those papers, which included excerpts.
+It seems like the DMI definition is very general and this accords.}
+We emphasize that these incompatibilities are a consequence of recombining different, yet functionally equivalent, mechanisms.
+
+Furthermore, using a quantitative genetic approach, we estimated that a genetically variable population will drift in neutral system space 
+at a rate determined by its intra-population variation and its effective population size. 
+Because mechanistically distinct yet phenotypically equivalent biological systems can fail to produce viable hybrids, we predict allopatric populations to accumulate genetic incompatibilities 
+at a rate on the order of $N_{e}$ under reasonable population genetic parameter estimates. 
+Additionally we see second-generation hybrid fitness plummet much faster than that of first-generation hybrids. 
+\plr{refer to Turelli here}
+This is a consequence of combining our mechanistic model with a quantitative genetic one: 
+we observe that $F_{1}$ phenotypes diverge quartically, and $F_{2}$ phenotypes quadratically, with evolutionary time. 
+\plr{awkward phrasing}
+This result is also consistent with Haldane's rule; that if only one hybrid sex is inviable or sterile it is likely the heterogametic sex. 
+The consistency comes from gene networks localized to the sex chromosomes functioning as an $F_{2}$ hybrid cross within a diploid $F_{1}$ heterogamete as there is only one sex chromosome.
+
+We also suggest that gene networks may not always use their components parsimoniously 
+as network size tends to ratchet up in the absence of strong selection against extra parts. 
+\plr{probably shouldn't claim this is "unexplored"} \js{I meant unexplored within this paper.Changed wording. Does that work?}
+Although we leave this question unexplored, this phenomena may lead to insights on evolvability and developmental innovation. 
+Lastly, we show that hybrid gene networks break down as function of genetic distance, and may, in part, explain broad patterns of reproductive isolation among diverse phyla \citep{roux2016shedding}.
+
+\plr{this paragraph isn't very clear} \js{Should I delete it? I guess it's sort of just a vague ``yeah we know this method has shortcomings''.}
+As Richard Levins opined, models in population biology face a trade-off among precision, realism, and generality \citep{levins1966strategy}.
+As Levins expects, any tractable and general model, such as the present one under discussion, will have limitations.
+Most notable is linearity.
+It is often stated that life is not linear.
+This is often true, however, many of the ideas developed here should be generalizable to nonlinear cases (multi-linear systems, say).
+Further, we see this as a necessary first step in the direction of more life-like nonlinear evolutionary systems theory.
+Depending on an actual biological system's particularities, its (potential) non-linearity, may buffer or exacerbate effects elucidated in this paper, such as the acquisition of Dobzansky-Muller incompatibilities.
+
+This theoretical framework can easily be applied to other interesting questions in evolutionary biology not tackled presently: such as the evolution of linkage, the necessity of network complexity (does evolution tend towards Rube Goldberg or parsimonious network organization?), evolvability, structure/function inference, and intra-population context dependency of mutational effects, as well as many others.
+
+%\jss{literature comparison -- I think we should delete}
+%
+%Over the last several years, several different computational approaches have been applied to study reproductive incompatibility and speciation.
+\citet{tulchinsky} simulated the evolution of a transcription factor and its binding site using a thermodynamic model.
+Their simulations suggest that the language by which a transcription factor recognizes its binding site can change, and potentially lead to hybrid incompatibility when allopatric populations employ divergent readout languages.
+This study, despite looking at gene regulation, does not analyze overall gene network architecture -- as we do here -- it only looks at the expression level of a single gene.
+Furthermore, they report reproductive isolation primarily following directional selection for a change in expression levels in each allopatric population; the evidence for reproductive isolation following balancing selection is much weaker.
+Johnson and Porter 2000 did not observe any hybrid fitness declines under stabilizing selection -- only under directional selection.
+%\plr{citations:}
+%Khatari et al, Tulchinsky et al, and Porter et al, all study hybrid incompatibility from a transcription factor/binding site interaction perspective, not from an overall network architecture perspective.
+Palmer and Feldman only see hybrid incompatibility in constant environments if the parental populations are relatively poorly adapted initially.
+Otherwise hybrids between two allopatric populations have fairly high fitnesses.
+
+\section*{Acknowledgements}
+We would like to thank Sergey Nuzhdin, Stevan Arnold, Erik Lundgren, and Hossein Asgharian for valuable discussion.
+
+\bibliographystyle{plainnat}
+\bibliography{krefs}
+%\end{multicols}
+
+\normalsize
+
+\section*{Examples}
+  \begin{example}[Oscillating Gene Network: Cell Cycle Control]\label{ex:oscillator}
+
+    Cellular division is a complicated phenomena, governed by many different processes, however it is agreed that its rhythm is partially controlled by periodic (oscillating) gene transcription \citep{orlando2008global}.
+Consider a simplified model of oscillating gene transcription.
+In the present framework periodic expression requires at minimum two interacting genes.
+
+    Suppose gene-2 up-regulates the transcription of gene-1 and that gene-1 down-regulates the transcription of gene-2 with equal magnitude (of $1$) and relative to each of their concentrations, denoted by $\kappa_{1}$ and $\kappa_{2}$.
+Furthermore, suppose that only the dynamics of gene-1 are consequential to the cell cycle (perhaps the amount of gene-1 activates another downstream gene network).
+Lastly suppose that the production of both genes is stimulated by an impulse of a molecule present immediately after division.
+
+\plr{you haven't explained $u(t)$ above. and, isn't it more natural in this set-up to have $B = [1,0]$ than $[1,1]$?}
+    If the rate each of these genes is expressed is a linear function of their concentrations, the dynamics of the system are given by
+    \begin{align*}
+      \dot{\kappa_{1}}(t) &= \kappa_{2}(t) + u(t) \\
+      \dot{\kappa_{2}}(t) &= - \kappa_{1}(t) + u(t)
+    \end{align*}
+    where $\dot{\kappa}$ denotes the time derivative.
+The initial conditions $\kappa_{1}(0)$ and $\kappa_{2}(0)$, and the input $u(t)$ then determine the concentrations through time.
+If we record the regulatory coefficients in the matrix
+    \begin{align*}
+      A &= \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} ,
+    \end{align*}
+    and define the column vector $B = \begin{bmatrix} 1 & 1 \end{bmatrix}^{T}$, then in matrix notation the dynamics are
+      \begin{align*}
+      \dot{\kappa} &= A \kappa(t) + B u(t) .
+      \end{align*}
+      Since only the dynamics of gene-1 are directly relevant to biological function, the dynamics of interest are given by
+      \begin{align*}
+      \phi(t) &= C \kappa (t)
+      \end{align*}
+      where the row vector $C$ is defined as $C = \begin{bmatrix} 1 & 0 \end{bmatrix}$. 
+      (If the dynamics of both genes were physiologically relevant to the cell cycle, we would set $C$ to be the identity matrix).
+  %  \begin{align*}
+  %    A = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} , \qquad B = \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix}
+  %  \end{align*}
+ %       The oscillatory system $\Sigma$ is thus given as
+ %   \begin{align*}
+ %     \Sigma = \left \{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 
+ %       0 & 1 \\ 
+ %      -1 & 0 
+ %       \end{bmatrix} \kappa(t) + \begin{bmatrix} 1 \\ 1 \end{bmatrix} u(t) \\ 
+ %         \phi(t) &= \begin{bmatrix} 1 & 0 \end{bmatrix} \kappa(t) \end{array} \right.
+ %    \end{align*}
+
+      Since the input is simply an impulse, its phenotype is equivalent to its impulse response
+      \begin{align*}
+        \phi(t) = h(t) = \sin(t) + \cos(t)  .
+      \end{align*}
+
+    \begin{figure}[H]
+       % \begin{center} 
+      \centering
+         \begin{tabular}{cc}
+            \begin{tikzpicture}
+            \begin{scope}[every node/.style={circle,thick,draw}]
+                \node (A) at (0,0) {$\kappa_{1}$};
+                \node[dashed] (B) at (3,0) {$\kappa_{2}$};
+                \node[shape=rectangle] (U) at (1.5,2) {input};
+                \node[shape=rectangle] (y) at (1.5,-2) {output};
+            \end{scope}
+
+            \begin{scope}[>={Stealth[black]},
+                          every node/.style={fill=white,circle},
+                          every edge/.style={draw=black, thick}]
+                \path [->, >=Rectangle] (A) edge[bend left] node {\tiny $-1$} (B);
+                \path [->] (B) edge[bend left] node {\tiny $1$} (A); 
+                \path[->] (U) edge[dashed] node {\tiny $1$} (A);
+                \path[->] (U) edge[dashed] node {\tiny $1$} (B);
+                \path[->] (A) edge[dashed,bend right] node {\tiny $1$} (y);
+            \end{scope}
+            \begin{scope}[>={Stealth[black]},
+                          every edge/.style={draw=black, thick}]
+                %\path [->] (A) edge[loop left] node {\tiny $\lambda_{1}$} (A);
+                %\path [->] (B) edge[loop left] node {\tiny $\lambda_{2}$} (B);
+            \end{scope}
+
+            \end{tikzpicture} &
+       % \end{center}
+      %  \caption{
+      %      Diagram of Example \ref{ex:osc} in the text.
+      %      \plr{explain what arrows mean if nec}
+      %      \label{fig:oscillator_diagram}}
+   % \end{figure}
+   % \begin{figure}[H]
+    %  \centering
+      \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{osc_impulse}
+   \end{tabular}
+  %  \end{center}
+      \caption{(Left) Graphical representation of the cell cycle control gene network, and (right) plot of the phenotype $\phi(t)$ against time $t$.}
+    \end{figure}
+      We return to the evolution of such a system below.
+  \end{example}
+
+  \begin{example}[All Phenotypically Equivalent Cell Cycle Control Networks] \label{ex:all_osc}
+
+%\plr{this is same as above, so reference instead of writing out again?}
+%    Let
+%    \begin{align*}
+%      A(0) = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix}, \qquad B = \begin{bmatrix} 1 & 1 \end{bmatrix}^{T}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix},
+%    \end{align*}
+%    with $V(\tau)$ preserving both $B$ and $C$, then
+    The set of all two-gene regulatory networks phenotypically equivalent to the cell cycle control network in \ref{ex:oscillator} is given by
+    \begin{align*}
+      A(\tau) &= \begin{bmatrix} 1 & 0 \\ \tau & 1-\tau \end{bmatrix} \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 0 \\ \frac{\tau}{\tau-1} & \frac{-1}{\tau-1} \end{bmatrix}  \qquad \forall \ \tau \neq 1 %\\
+     %   B &= \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix} \\
+       %   h(t) &= \sin(t) + \cos(t) \qquad \forall \ \tau \neq 1
+    \end{align*}
     \begin{figure}[H]
     \centering
-    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{expF1}
-    \caption{F1 hybrid (orange) and parental (blue) phenotypic oscillator dynamics for an $A(0)$ by $A(2)$ cross. The hybrid fails to oscillate and exhibits qualitatively different dynamics.}
+% Nice tikz!!!
+\begin{tikzpicture}
+\begin{scope}[every node/.style={circle,thick,draw}]
+  \node (A) at (0,0) {$\kappa_{1}$};
+    \node[dashed] (B) at (4,0) {$\kappa_{2}$};
+    \node[shape=rectangle] (U) at (2,2) {input};
+    \node[shape=rectangle] (y) at (2,-2) {output};
+\end{scope}
+
+\begin{scope}[>={Stealth[black]},
+              every node/.style={fill=white,circle},
+              every edge/.style={draw=black, thick}]
+    \path [->, sloped] (A) edge[bend left] node {\tiny $2 \tau + (\tau-1)^{-1}$} (B);
+    \path [->, sloped] (B) edge[bend left] node {\tiny $-(\tau-1)^{-1}$} (A); 
+    \path[->] (U) edge[dashed] node {\tiny $1$} (A);
+    \path[->] (U) edge[dashed] node {\tiny $1$} (B);
+    \path[->] (A) edge[dashed, bend right] node {\tiny $1$} (y);
+\end{scope}
+\begin{scope}[>={Stealth[black]},
+              every edge/.style={draw=black, thick}]
+    \path [->] (A) edge[loop left] node[sloped, anchor=center, above] {\tiny $1 + (\tau-1)^{-1}$} (A);
+    \path [->] (B) edge[loop right] node[sloped, anchor=center, above] {\tiny $-1 - (\tau-1)^{-1}$} (B);
+\end{scope}
+
+\end{tikzpicture}
+    \caption{A graphical depiction of the set of all externally equivalent cell cycle control networks, $A(\tau)$. 
+        $\tau$ can take any value \plri{sentence fragment}}
     \end{figure}
-%     \begin{example}[F1 Reproductive Incompatibility in an Oscillating Gene Network]
-%       DMI examples...
-%       \begin{figure}[H]
-%         \includegraphics[width=0.5\textwidth]{A2_A3_osci_F1_hyb}
-%       \end{figure}
  % \end{example}
 
-  \begin{example}[Hybrid Incompatibility in an Oscillating Gene Network]
+ % \begin{example}[External Equivalence does not imply internal equivalence]
+   Despite the phenotypic equivalence of all instantiations of $A(\tau)$, the internal dynamics, or kryptotypes, vary as a function of $\tau$. 
+    Gene-1 dynamics (blue) are equivalent for network architectures $A(0)$ and $A(2)$, however the dynamics of gene-2 (orange) differ with $\tau$.
+  \begin{figure}[H]
+    \centering
+    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{osc_A0_A2_both_compare}
+    \caption{Gene-1 (blue) and gene-2 (orange) dynamics for $A(0)$ (top) and $A(2)$ (bottom).
+Both (top and bottom) gene-1 dynamics are given by $ \kappa_{1} = \sin(t) + \cos(t)$, and gene-2 by $\kappa_{2} = \cos(t) - \sin(t)$ (top) and $\kappa'_{2} = \cos(t) + 3 \sin(t)$ (bottom).}
+  \end{figure}
+\end{example}
+
+  \begin{example}[Hybrid Incompatibility in an Oscillating Gene Network] \label{ex:hybrid_osc}
 
@@
 Each $A$ is phenotypically identical ($\phi(t) = \sin(t) + \cos(t)$), however some of the hybrids exhibit markedly different dynamics.
-\plr{I don't think this is the right place to introduce this?}
-These differences tend to increase with time, quadratically ($x^{2}$) in F2s, and quartically ($x^{4}$) in F1s.
 
@@
 \end{alignat*}
-\plr{cut this sentence, was already said:}
-F1 gene regulatory networks are formed by averaging the two parental $A(\tau)$ matrices; F2s are formed by first recombining parental matrices followed by an averaging.
 
@@
 \end{example}
-%     \begin{example}[Not all Networks can Host Incompatibilities]
-%       convex sets cant have DMIs
-
-%       \begin{align*}
-%         h(t) = 2 e^{- \theta t}
-%       \end{align*}
-%       Any non-minimal system with rows summing to $\theta$ is PI.
-Further, these systems are closed under averaging (mating) and row swapping (meiosis), leaving all hybrids optimally fit.
-The set of gene matrices is affine and therefore convex.
-%     \end{example}
-
-\section*{The rate of speciation under neutrality in allopatric populations}
-
-\plr{separate out here (or above) the difference in genotype of hybrids from the optimum set, the difference in phenotype, and the difference in fitness.}
-
-In general terms, if fitness $\fit$ is some function $g$ of phenotypic divergence from an optimal scaled by the strength of selection $\beta$,
-\begin{align*}
-  \fit = g \left( \frac{\phi - \phi_{0}}{\beta} \right)
-\end{align*}
-\plr{sentences don't begin with "And"}
-And a population $P$ is normally distributed around a mean gene network architecture $A(\tau)$ and $\sigma^{2}$ is the intra-population regulatory variance,
-\plr{what is $T$?}
-\begin{align*}
-  P_{T} \sim \mathcal{N} \left(A(\tau), T \frac{\sigma^{2}}{N_{e}}\right)
-\end{align*}
-the population mean architecture will move at rate
-\plr{what is $\Delta \tau$? Also, if we reframe in terms of $V$ not $\tau$ above will need to reframe this.}
-\begin{align*}
-  \Delta \tau = \sigma \sqrt{\frac{T}{N_{e}}}
-\end{align*}
-where $T$ is time measured in generations. 
-USING GENERAL MATH FROM MULTIVARIATE TAYLOR SERIES....
-we know that the phenotype of a first generation $F_{1}$ hybrid diverges from that of its parents (assuming both parents are at an optimum and phenotypically equivalent) quartically as a function of $\Delta \tau$.
-Thus the squared difference of it's phenotype will be,
-\begin{align*}
-  \Delta \Phi^{2}_{1} = c_{1}^{2} \sigma^{4} \left( \frac{T}{N_{e}} \right)^{2}
-\end{align*}
-and $F_{1}$ fitness will decline linearly with respect to $\frac{T}{N_{e}}$. ARGUMENT FOR WHY $\sigma^{2} c_{1} \sim \beta$ HERE.
-   \begin{align*}
-     \Delta \fit_{1} = \left( \frac{\Delta \Phi_{1}}{\beta} = \frac{c_{1} \sigma^{2} \frac{T}{N_{e}}}{\beta} \approx \frac{T}{N_{e}} \right)
-\end{align*}
- It follows FROM THE SAME MATH ABOUT THE MULTIVAR TAYLOR that the phenotypes of second generation $F_{2}$ hybrid crosses diverges quadratically as a function of $\Delta \tau$, 
-\begin{align*}
-  \Delta \Phi^{2}_{2} &= c_{2}^{2} \left( \sigma \sqrt{\frac{T}{N_{e}}} \right)^{2} = c_{2}^{2} \frac{T}{N_{e}} \\
-  \Delta \fit_{2} &= \left( \frac{\Delta \Phi_{2}}{\beta} = \frac{c_{2} \sigma \sqrt{\frac{T}{N_{e}}}}{\beta} \approx \sqrt{\frac{T}{N_{e}}} \right)
-\end{align*}
-and that $F_{2}$ fitness drops rapidly -- at approximately rate $\sqrt{\frac{T}{N_{e}}}$. ARGUMENT FOR WHY $c_{2} \sigma \sim \beta$ HERE.
-
-This suggests that reproductively isolated populations can speciate rapidly -- on timescales much less than $N_{e}$ generations, depending on the specifics of the fitness function.
-
-%%%%%%%%%%%%%%%
-\begin{example}[Metabolic network]
+
+\begin{example}[Metabolic network] \label{ex:metabolic}
 
@@
 The phenotype is, 
-\plr{this is wrong, fix it}
-\begin{align*}
-    \phi(t) = \exp \left\{ \kappa(0) + \int_{0}^{t} h(t) u(t) dt \right\} 
+\begin{align*}
+  \phi(t) = C e^{A t} \kappa(0) + \int_{0}^{t} h(s-t) u(s) ds
 \end{align*}
@@
 
-%    \subsection*{Metabolic Network}
-%  \begin{align*}
-%    \dot{x}(t) &= \begin{bmatrix} 0 & 1 & 0 & -1 \\ 0 & 0 & -1 & -1 \\ -1 & 1 & 0 & 0 \\ 0 & 0 & 0 & 0 \end{bmatrix} \begin{bmatrix} \text{GAL3} \\ \text{GAL4} \\ \text{GAL80} \\ \text{MIG1} \end{bmatrix}(t) \\ &+ \begin{bmatrix} 1 & 0 & 0 \\ 0 & 0 & 1 \\ 0 & 0 & 0 \\ 0 & 1 & 1 \end{bmatrix} \begin{bmatrix} \text{galactose} \\ \text{glucose} \\ \text{BTM} \end{bmatrix}(t) \\
-%        y(t) &= \begin{bmatrix} 0 & 1 & 0 & -1 \end{bmatrix} \vec{x}(t)
-%    \end{align*}
-%    \begin{align*}
-%      H(z) &= \begin{bmatrix} 1 \\ z^{2} + z + 1 \\ z + 1 \end{bmatrix} \frac{1}{z^{3} + z -1}
-%    \end{align*}
-%
-%    \begin{align*}
-%      CV &= C \\
-%      VB &= B \\
-%      V :&= \begin{bmatrix} 1 & 0 & a & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & b & 0 \\ 0 & 0 & 0 & 1 \end{bmatrix} , V^{-1} = \begin{bmatrix} 1 & 0 & \frac{-a}{b} & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & \frac{1}{b} & 0 \\ 0 & 0 & 0 & 1\end{bmatrix} \\
-%        A(a,b) :&= VAV^{-1} = \begin{bmatrix} -a & a+1 & \frac{a^{2}}{b} & -1 \\ 0 & 0 & \frac{-1}{b} & -1 \\ -b & b & a & 0 \\ 0 & 0 & 0 & 0 \end{bmatrix}
-%    \end{align*}
-%    The GAL regulatory network in \emph{S.
-cerevisiae} is modelled here.
-The activating transcription factor GAL4, which is regulated by the presence of galactose and two other TFs, binds to the promoter region of the GAL regulon -- a cluster of genes regulated by the same promoter encoding 3 enzymes (GAL1, GAL7, and GAL10), for the metabolism of the sugar galactose.
-The repressing transcription factor MIG1, which is activated by the presence of glucose and other proteins omitted here, also binds to the regulon, preventing the expression of the galactose metabolizing enzymes.
-This network is one of the most studied gene regulatory networks in yeast, and experiments have already demonstrated significant transcriptional variation among different species and genuses of yeast.
-%
-%    In the present model, the $A$ matrix encodes the interactions among GAL3, GAL4, GAL80, and MIG1.
-$B$ interprets the input: the quantity of glucose, galactose, and the basal transcription (BTM) rates impact of TF concentrations.
-$C$ represents the promoter region of the GAL regulon and is the sum of the influence of GAL4 (activating) and MIG1 (repressing).
-%
-%    $A(a,b)$ represent alternative network structures that produce identical outputs, with identical input and output transformations -- meaning the regulatory contributions of the sugars and the basasl machinery, as well as the promoter of the GAL regulon are held constant.
-These interactions may be realized in nature, if the regulatory differences predicted by $A(a,b)$ are biologically/mutationally realizable.
-Mutations influencing the interactions between GAL4 and GAL80 have been experimentally demonstrated \citep{li2010alterations}.
-%
-%    \subsection*{Gap Gene Network}
-
-%%%%%%%%%%%%%%%%%%%%%
-\section*{Discussion}
-
-\plr{mention $B$ part of genetic architecture}
-
-\jss{Discussion guidelines:
-\begin{itemize}
-    \item Why is this important/useful?
-    \item What are the assumptions and shortcomings of the research?
-    \item Compare to other studies in the literature. 
-    \item Future directions. 
-    \item Wild speculations?
-    \item Conclusion and overall impact. 
-\end{itemize}}
-
-The complexity of biological systems has limited our understanding of their function and evolution.
-Above we outline an approach, a first step, towards untangling this complexity in reference to function and evolution.
-This methodology borrows successfully applied tools from engineering and aims to synthesize these with the concepts and tools of molecular and evolutionary biology.
-
-Theoretical models in evolution and population genetics often lack the molecular details of physiology or of the genotype-phenotype map. 
-Here, we offer a tractable and simple model which includes these missing features. 
-\plr{this is a claim we are the first to do something.  best not to make these claims (and we aren't the first to make math models of these)}
-Further, we provide, in clear mathematical language, an analytical description of phenomena hitherto only discussed verbally and conceptually 
-(phenogenetic drift \citep{weiss2000phenogenetic}, developmental systems drift \citep{true2001developmental}, biological degeneracy \citep{edelman2001degeneracy}, \emph{etc.}). 
-The tractability and relative simplicity of this exposition enables the interested biologist to work out by hand, 
-if desired, 
-\plr{probably not desired}
-the dynamics of a genetic system, as well as perturbations to the system -- an attribute not likely to be found in less tractable models and simulations.
-
-We have suggested an interpretation of system identification: to see it as an evolutionarily neutral manifold, and not simply a computational nuisance. 
-We have demonstrated a method to analytically determine the set of all phenotypically invariant gene networks; 
-by a simple change of coordinates in the minimal configuration, or more generally by applying the Kalman decomposition in higher dimensions. 
-Further, we emphasize that evolution proceeds through this high dimensional space as stochastic coordinate transformation, constrained by sexual reproduction and selection. 
-This set is explored over evolutionary time when phenotype is conserved, and can lead to a diverse set of consequences, 
-including the accumulation of Dobzhansky-Muller incompatibilities. 
-\plr{check original Bateson/DM papers to see if this accords with those defns}
-We emphasize that these incompatibilities are a consequence of recombining different, yet functionally equivalent, mechanisms.
-
-Furthermore, using a quantitative genetic approach, we estimated that a genetically variable population will drift in neutral system space 
-at a rate determined by its intra-population variation and its effective population size. 
-\plr{replace "speciate" with "accumulate genetic incompatibilities"}
-Because mechanistically distinct yet phenotypically equivalent biological systems can fail to produce viable hybrids, we predict allopatric populations to speciate 
-at a rate on the order of $N_{e}$ under reasonable population genetic parameter estimates. 
-Additionally we see second-generation hybrid fitness plummet much faster than that of first-generation hybrids. 
-\plr{refer to Turelli here}
-This is a consequence of combining our mechanistic model with a quantitative genetic one: 
-we observe that $F_{1}$ phenotypes diverge quartically, and $F_{2}$ phenotypes quadratically, with evolutionary time. 
-\plr{awkward phrasing}
-This result is also consistent with Haldane's rule; that if only one hybrid sex is inviable or sterile it is likely the heterogametic sex. 
-The consistency comes from gene networks localized to the sex chromosomes functioning as an $F_{2}$ hybrid cross within a diploid $F_{1}$ heterogamete as there is only one sex chromosome.
-
-We also suggest that gene networks may not always use their components parsimoniously 
-as network size tends to ratchet up in the absence of strong selection against extra parts. 
-\plr{probably shouldn't claim this is "unexplored"}
-Although unexplored presently, this phenomena may lead to insights on evolvability and developmental innovation. 
-Lastly, we show that hybrid gene networks break down as function of genetic distance, and may, in part, explain broad patterns of reproductive isolation among diverse phyla \citep{roux2016shedding}.
-
-\plr{this paragraph isn't very clear}
-As Richard Levins opined, models in population biology face a trade-off among precision, realism, and generality \citep{levins1966strategy}.
-As Levins expects, any tractable and general model, such as the present one under discussion, will have limitations.
-Most notable is linearity.
-It is often stated that life is not linear.
-This is often true, however, many of the ideas developed here should be generalizable to nonlinear cases (multi-linear systems, say).
-Further, we see this as a necessary first step in the direction of more life-like nonlinear evolutionary systems theory.
-Depending on an actual biological system's particularities, its (potential) non-linearity, may buffer or exacerbate effects elucidated in this paper, such as the acquisition of Dobzansky-Muller incompatibilities.
-
-This theoretical framework can easily be applied to other interesting questions in evolutionary biology not tackled presently: such as the evolution of linkage, the necessity of network complexity (does evolution tend towards Rube Goldberg or parsimonious network organization?), evolvability, structure/function inference, and intra-population context dependency of mutational effects, as well as many others.
-
-\jss{literature comparison:}
-
-Over the last several years, several different computational approaches have been applied to study reproductive incompatibility and speciation.
-\citet{tulchinsky} simulated the evolution of a transcription factor and its binding site using a thermodynamic model.
-Their simulations suggest that the language by which a transcription factor recognizes its binding site can change, and potentially lead to hybrid incompatibility when allopatric populations employ divergent readout languages.
-This study, despite looking at gene regulation, does not analyze overall gene network architecture -- as we do here -- it only looks at the expression level of a single gene.
-Furthermore, they report reproductive isolation primarily following directional selection for a change in expression levels in each allopatric population; the evidence for reproductive isolation following balancing selection is much weaker.
-Johnson and Porter 2000 did not observe any hybrid fitness declines under stabilizing selection -- only under directional selection.
-\plr{citations:}
-Khatari et al, Tulchinsky et al, and Porter et al, all study hybrid incompatibility from a transcription factor/binding site interaction perspective, not from an overall network architecture perspective.
-Palmer and Feldman only see hybrid incompatibility in constant environments if the parental populations are relatively poorly adapted initially.
-Otherwise hybrids between two allopatric populations have fairly high fitnesses.
-
-\section*{Acknowledgements}
-We would like to thank Sergey Nuzhdin, Stevan Arnold, Erik Lundgren, and Hossein Asgharian for valuable discussion.
-
-\bibliographystyle{plainnat}
-\bibliography{krefs}
-%\end{multicols}
-
-\normalsize
 \appendix
 
-\section{Kalman Decomposition}
+\section{Kalman Decomposition} \label{apx:kalman}
 \begin{definition}[Phenotypic equivalence of systems]
@@
 
+\section{Linear Evolution-Variant Systems}
+  Let $V(\tau) = \prod_{\ell = 0}^{\tau} (I + Q_\ell)$ (where $Q_\ell$ is a random matrix with norm $< \sigma$), and let $A$ be minimal.
+Then the system architecture after evolutionary time $\tau$ is given by,
+    \begin{align}
+      \mathcal{S}(\tau) := \left\{ \begin{array}{ll} V(\tau)\dot{\kappa}(t) &= V(\tau)A V(\tau)^{-1} \kappa(t) + V(\tau)B u(t) \\ \phi(t) &= C V(\tau)^{-1} \kappa(t) \end{array} \right .
+  \end{align}
 \section{Genetic drift with a multivariate trait}

### 2017-08-11 kalman_walk_paper.tex
more title ideas
started stuff on hybrid vigor

@@
 
-\section*{Gene regulatory network drift and speciation occurs rapidly under neutrality}
+\section*{Rapid speciation despite conservation of phenotype}
 {\centering
@@
 % Other title ideas:
+%
+% More than one way to grow a cat: regulatory network drift and speciation
+% 
+% Rapid speciation despite conservation of phenotype
+%
 % More than one way to grow a cat: regulatory network redundancy and speciation
@@
 % 
+% Evolutionary conservation of phenotype does not entail conservation of the underlying
+% molecular mechanism leading to rapid speciation
+%
+% The evolution of phenotype-invariant gene networks rapidly leads to hybrid incompatibiliy
+%
+% Evolutionary network rewiring can rapidly lead to hybrid incompatibility despite
+% phenotypic conservation
+%
+% Hybrid incompatibilities can evolve rapidly due to developmental systems drift
+%
+% Evolutionary systems theory and speciation
 
@@
 
-\section{Linear Evolution-Variant Systems}
-  Let $V(\tau) = \prod_{\ell = 0}^{\tau} (I + Q_\ell)$ (where $Q_\ell$ is a random matrix with norm $< \sigma$), and let $A$ be minimal.
-Then the system architecture after evolutionary time $\tau$ is given by,
-    \begin{align}
-      \mathcal{S}(\tau) := \left\{ \begin{array}{ll} V(\tau)\dot{\kappa}(t) &= V(\tau)A V(\tau)^{-1} \kappa(t) + V(\tau)B u(t) \\ \phi(t) &= C V(\tau)^{-1} \kappa(t) \end{array} \right .
-  \end{align}
 \section{Genetic drift with a multivariate trait}
@@
 
+\section{Hybrid Vigor}
+
+If a fitness optimum is at the origin $(0,0) \in \R^{2}$, and two populations have drifted away from the optimum by an distance of $r$, then the expected value of their $F_1$ hybrids is,
+
+\begin{align*}
+  \mathbb{E}\left[ \left\lVert \frac{1}{2} ((x_1,y_1) + (x_2,y_2))  \right\rVert \right] = \\
+  \frac{\sqrt{2}}{8}\left(\sqrt{3\, r^2 - {\sin^{-1}\!\left(r\right)}^2 + r^4 - 2\, r\, \sin^{-1}\!\left(r\right)\, \sqrt{1 - r^2}} + \sqrt{{\sin^{-1}\!\left(r\right)}^2 + 5\, r^2 - r^4 + 2\, r\, \sin^{-1}\!\left(r\right)\, \sqrt{1 - r^2}}\right)
+\end{align*}
+
+If $r < \sim 5.5$ then we will observe hybrid vigor, however if $r$ is sufficiently large, the hybrids will have lower fitness, on average.
+
+Some $F_2$s will have perfect fitness, and some will be worse off than both parents.
+\jss{I still need to do the calculations.} For instance if $r=1$ so that parental populations reside on the unit circle, two randomly drifted populations will create $F_1$s a distance of, on average, $0.6684$.
+Some $F_2$s from this population wil be back at the origin (a quarter of the time), others as far away as $\sqrt{2}$ (a quarter of the time), and some will stay a distance of $1$ away (half the time).
+
 \end{document}

### 2017-08-11 kalman_walk_paper.tex
started stuff on hybrid vigor

@@
 Some $F_2$s will have perfect fitness, and some will be worse off than both parents.
-\jss{I still need to do the calculations.} For instance if $r=1$ so that parental populations reside on the unit circle, two randomly drifted populations will create $F_1$s a distance of, on average, $0.6684$.
+\jss{I still need to do the calculations.} For instance if $r=1$ so that parental populations reside on the unit circle circumference, two randomly drifted populations will create $F_1$s a distance of, on average, $0.6684$.
 Some $F_2$s from this population wil be back at the origin (a quarter of the time), others as far away as $\sqrt{2}$ (a quarter of the time), and some will stay a distance of $1$ away (half the time).

### 2017-08-11 kalman_walk_paper.tex
hybrid vigor edit

--- 2017-08-11 kalman_walk_paper.tex
@@
 
-If a fitness optimum is at the origin $(0,0) \in \R^{2}$, and two populations have drifted away from the optimum by an distance of $r$, then the expected value of their $F_1$ hybrids is,
-
-\begin{align*}
-  \mathbb{E}\left[ \left\lVert \frac{1}{2} ((x_1,y_1) + (x_2,y_2))  \right\rVert \right] = \\
-  \frac{\sqrt{2}}{8}\left(\sqrt{3\, r^2 - {\sin^{-1}\!\left(r\right)}^2 + r^4 - 2\, r\, \sin^{-1}\!\left(r\right)\, \sqrt{1 - r^2}} + \sqrt{{\sin^{-1}\!\left(r\right)}^2 + 5\, r^2 - r^4 + 2\, r\, \sin^{-1}\!\left(r\right)\, \sqrt{1 - r^2}}\right)
-\end{align*}
-
-If $r < \sim 5.5$ then we will observe hybrid vigor, however if $r$ is sufficiently large, the hybrids will have lower fitness, on average.
-
-Some $F_2$s will have perfect fitness, and some will be worse off than both parents.
-\jss{I still need to do the calculations.} For instance if $r=1$ so that parental populations reside on the unit circle, two randomly drifted populations will create $F_1$s a distance of, on average, $0.6684$.
-Some $F_2$s from this population wil be back at the origin (a quarter of the time), others as far away as $\sqrt{2}$ (a quarter of the time), and some will stay a distance of $1$ away (half the time).
+If a fitness optimum is at the origin $(0,0) \in \R^{2}$,
+%and two populations have drifted away from the optimum by an distance of $r$, then the expected value of their $F_1$ hybrids is,
+%\begin{align*}
+%  \mathbb{E}\left[ \left\lVert \frac{1}{2} ((x_1,y_1) + (x_2,y_2))  \right\rVert \right] = \\
+%  \frac{\sqrt{2}}{8}\left(\sqrt{3\, r^2 - {\sin^{-1}\!\left(r\right)}^2 + r^4 - 2\, r\, \sin^{-1}\!\left(r\right)\, \sqrt{1 - r^2}} + \sqrt{{\sin^{-1}\!\left(r\right)}^2 + 5\, r^2 - r^4 + 2\, r\, \sin^{-1}\!\left(r\right)\, \sqrt{1 - r^2}}\right)
+%\end{align*} 
+%If $r < \sim 5.5$ then we will observe hybrid vigor, however if $r$ is sufficiently large, the hybrids will have lower fitness, on average. 
+and two populations have drifted away from the optimum by a distance of $r=1$,
+then $F_1$s will have an average fitness of,
+\begin{align*}
+  \mathbb{E} \left[ d(F_1)] \right] &= \int_{0}^{2\pi} \int_{0}^{2\pi} \left\lVert \frac{1}{8 \pi^{2}} \left( \left( \cos(\theta_1), \sin(\theta_1) \right) + \left( \cos(\theta_2), \sin(\theta_2) \right) \right) \right\rVert d\theta_1 d\theta_2 \\
+  &= \cos(\frac{\pi}{2}) = \frac{\sqrt{2}}{2}
+\end{align*}
+Haploid $F_2$s will, on average, be a distance of $1$ half of the time, a distance of $0$ a quarter of the time, and a distance of $\sqrt{2}$ a quarter of the time.
 

### 2017-08-29 kalman_walk_paper.tex
F2 hybrid vigor calc
hybrid vigor
edited system description, qgen, examples, speciation and more.

@@
 
-\newcommand{\calA}{\mathcal{A}}
+\newcommand{\calA}{\mathcal{S}}
 \newcommand{\calH}{\mathcal{H}}
@@
 % Other title ideas:
+%
+% More than one way to grow a cat: speciation despite conservation of phenotype
 %
@@
 \section*{Gene Networks as Linear Dynamical Systems}
-
+%
 \plr{Below we mix general language with language specific to GRNs, like switching between "internal state" and "transcription factor concentration.  
@@
 we could say up top that this applies to other situations, but to make it concrete we'll talk about regulatory networks.}
-
+%
 %  \jss{Organisms' phenotypes are constructred by gene by gene by environment interactions.
@@
 %}
-\js{Will rewrite this section again.} Here we outline a method to model biological systems, such as gene regulatory networks, as linear dynamical systems. 
-We define an organism's ``phenotype'' to be the temporal dynamics of molecular concentrations directly relevant to survival, 
-and denote by $\phi_k(t)$ the concentration of the $k^\text{th}$ element of the phenotype at time $t$, given an input; for instance $\phi_k(t)$ could describe the concentrations of a particular enzyme.
-These dynamics are the result of the interconnections of 
-a gene regulatory network (or some other biological system) $A$ (an $n \times n$ matrix) 
-and an environmental input vector $u(t)$, and the \emph{kryptotype} $\kappa(t)$, is a list of a system's molecular concentrations at time $t$ -- or more generally the systems \emph{internal dynamics}.
-Such a system $\mathcal{S}$, is composed of two equations,
+Organisms' phenotypes are constructred by gene by gene by environment interactions.
+Here we simply define the \emph{phenotype} to be the temporal molecular dynamics directly under natural selection.
+The \emph{what}, \emph{when}, and \emph{how much}, of an organism's molecules that are physiologically or otherwise relevant to survival.
+
+Thus an organism's phenotype $\phi(t)$ -- a vector of molecular concentrations at time $t$ -- is determined both by the structure and organization of a biological system (\emph{e.g.} a gene regualtory network), given by the triple $(A,B,C)$,
+and by its environment $u(t)$.
+
+Such a biological system $\mathcal{S}$ is given by,
   \begin{align}
@@
   \end{align}
-As some system dynamics may not be visible nor relevant to selection, we distinguish the \emph{kryptotype} (as it is hidden) from the \emph{phenotype} -- which is the system's \emph{external dynamics}.
-The change in internal molecular concentrations as a function of time $\dot{\kappa}(t)$ is simply a function of the system's current state (the kryptotype), the system architecture $(A,B,C)$ -- such as the wiring of a gene regulatory network, and the environmental input $u(t)$ to the system.
-The architecture of $A$ determines exactly how the internal molecular concentrations regulate one another's expression as $A_{ij}$ is the magnitude at which transcription factor $\kappa_j(t)$ regulates transcription factor $\kappa_i(t)$.
-If $A_{ij} > 0$, we say that $\kappa_j$ upregulate $\kappa_i$, and if $A_{ij} < 0$, we say that $\kappa_j$ down-regulates $\kappa_i$.
-Furthermore, we interpret the $i^\text{th}$ row of $A$ to be the \emph{cis}-regulatory element (or promoter) for gene $i$.
-The architecture of $B$ determines how the system processes the environmental intput, and the structure of $C$ converts the internal dynamics to external dynamics.
-That is if $\mathcal{S}$ is a metabolic system, $C_{ij}$ is the amount at which the $j$th input metabolite $u_j$ affects the production of the $i$th enzyme concentration.
-
+
+Generally $A$ is any real $n \times n$ matrix, $B$ any $n \times \ell$, and $C$ any $\ell \times n$ dimensional matrix.
+Although many different biological systems can be modeled with this approach, for clarity, we focus on gene regulatory networks.
+%
+Each $i$th row of $A$ describes the \emph{cis}-regulatory module for gene $i$, and each $j$th entry, the specific regulatory influence of gene $j$ on gene $i$. 
+%The $i^\text{th}$ row of $A$ contains the regulatory logic pertaining to gene $i$.
+%
+As such, $A_{ij}$ is the magnitude at which transcription factor $\kappa_j(t)$ regulates transcription factor $\kappa_i(t)$, and if $A_{ij} > 0$, we say that $\kappa_j$ upregulates $\kappa_i$.
+If $A_{ij} < 0$, we say that $\kappa_j$ down-regulates $\kappa_i$.
+
+The form of $B$ determines precisely how the environment influences the organism -- that is $B$ filters and translates the input to the system.
+$C$ filters and translates the dynamics of the system and precesily determines the output, that is, what is visible to selection.
+\emph{E.g.} for a metabolic system, $C_{ij}$ is the amount the $j$th metabolite $u_j$ affects the production of the $i$th enzyme.
+%
+
+  %where $A$ is a gene regulatory network (although this model can be generalized to other biological networks) -- a square matrix, 
+  %and $B$ filters and translates the input to the system.
+The form of $B$ determines precisely how the state of the environment influences the internal gene network.
+$C$ filters and translates the dynamics of the system and precesily determines the output, that is, what is visible to selection.
+%
+%
+Furthermore, whereas the phenotype is a subset of molecules \emph{visible} to selection, the \emph{kryptotype} includes the molecular dynamics \emph{hidden} from selection.
+That is $\kappa(t)$ is a vector of the system's molecular concentrations at time $t$, directly, indrectly, and irrelevant to survival.
+%
+%The structure of $B$ and $C$ determines how systems process their environment and convert krypto- to phenotypes, respectively. 
+%
 \plr{This "Thus" needs expanding, e.g., "the solution to this equation is unique and given by" with a reference.}
-Thus the phenotype is simply a convolution of the system organization and the environment, 
+
+Finally, we can write the phenotype as a convolution of the system organization and the environment,
   \begin{align}
@@
 %\end{align}
-
-  \section*{Gene Network Drift}
+ \begin{example}[Oscillating Gene Network: Cell Cycle Control]\label{ex:oscillator}
+    Cellular division is governed by many different processes, however it is thought that its rhythm is partially controlled by oscillating gene transcription \citep{orlando2008global}.
+Here we consider a simplified model of oscillating gene transcription.
+
+    Suppose gene-2 up-regulates the transcription of gene-1 and that gene-1 down-regulates gene-2 with equal magnitudes, whose concentrations are given by $\kappa_1(t)$ and $\kappa_2(t)$.
+Furthermore, suppose that only the dynamics of gene-1 are consequential to the cell cycle (perhaps the amount of gene-1 activates another downstream gene network).
+Lastly suppose that the production of both genes is stimulated by an impulse of a molecule present immediately after division.
+
+    If the rate each of these genes is expressed is a linear function of their concentrations, the dynamics of the system are given by
+    \begin{align*}
+      \dot{\kappa_{1}}(t) &= \kappa_{2}(t) + u(t) \\
+      \dot{\kappa_{2}}(t) &= - \kappa_{1}(t) + u(t)
+    \end{align*}
+    where $\dot{\kappa}$ denotes the time derivative.
+The initial conditions $\kappa_{1}(0)$ and $\kappa_{2}(0)$, and the input $u(t)$ then determine the concentrations through time.
+If we record the regulatory coefficients in the matrix
+    \begin{align*}
+      A &= \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} ,
+    \end{align*}
+    and define the column vector $B = \begin{bmatrix} 1 & 1 \end{bmatrix}^{T}$, then in matrix notation the dynamics are
+      \begin{align*}
+      \dot{\kappa} &= A \kappa(t) + B u(t) .
+      \end{align*}
+      Since only the dynamics of gene-1 are directly relevant to biological function, the dynamics of interest are given by
+      \begin{align*}
+      \phi(t) &= C \kappa (t)
+      \end{align*}
+      where the row vector $C$ is defined as $C = \begin{bmatrix} 1 & 0 \end{bmatrix}$. 
+  %  \begin{align*}
+  %    A = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} , \qquad B = \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix}
+  %  \end{align*}
+ %       The oscillatory system $\Sigma$ is thus given as
+ %   \begin{align*}
+ %     \Sigma = \left \{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 
+ %       0 & 1 \\ 
+ %      -1 & 0 
+ %       \end{bmatrix} \kappa(t) + \begin{bmatrix} 1 \\ 1 \end{bmatrix} u(t) \\ 
+ %         \phi(t) &= \begin{bmatrix} 1 & 0 \end{bmatrix} \kappa(t) \end{array} \right.
+ %    \end{align*}
+
+      Since the input is simply an impulse, its phenotype is equivalent to its impulse response
+      \begin{align*}
+        \phi(t) = h(t) = \sin(t) + \cos(t)  .
+      \end{align*}
+
+    \begin{figure}[H]
+       % \begin{center} 
+      \centering
+         \begin{tabular}{cc}
+            \begin{tikzpicture}
+            \begin{scope}[every node/.style={circle,thick,draw}]
+                \node (A) at (0,0) {$\kappa_{1}$};
+                \node[dashed] (B) at (3,0) {$\kappa_{2}$};
+                \node[shape=rectangle] (U) at (1.5,2) {input};
+                \node[shape=rectangle] (y) at (1.5,-2) {output};
+            \end{scope}
+
+            \begin{scope}[>={Stealth[black]},
+                          every node/.style={fill=white,circle},
+                          every edge/.style={draw=black, thick}]
+                \path [->, >=Rectangle] (A) edge[bend left] node {\tiny $-1$} (B);
+                \path [->] (B) edge[bend left] node {\tiny $1$} (A); 
+                \path[->] (U) edge[dashed] node {\tiny $1$} (A);
+                \path[->] (U) edge[dashed] node {\tiny $1$} (B);
+                \path[->] (A) edge[dashed,bend right] node {\tiny $1$} (y);
+            \end{scope}
+            \begin{scope}[>={Stealth[black]},
+                          every edge/.style={draw=black, thick}]
+                %\path [->] (A) edge[loop left] node {\tiny $\lambda_{1}$} (A);
+                %\path [->] (B) edge[loop left] node {\tiny $\lambda_{2}$} (B);
+            \end{scope}
+
+            \end{tikzpicture} &
+       % \end{center}
+      %  \caption{
+      %      Diagram of Example \ref{ex:osc} in the text.
+      %      \plr{explain what arrows mean if nec}
+      %      \label{fig:oscillator_diagram}}
+   % \end{figure}
+   % \begin{figure}[H]
+    %  \centering
+      \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{osc_impulse}
+   \end{tabular}
+  %  \end{center}
+      \caption{(Left) Graphical representation of the cell cycle control gene network, and (right) plot of the phenotype $\phi(t)$ against time $t$.}
+    \end{figure}
+      We return to the evolution of such a system below.
+  \end{example}
+
+  \section*{Phenotype-equivalent gene networks}
 
@@
 Any linear and minimal system (a gene network) 
--- minimal, informally meaning that the system's phenotype is achieved with the fewest possible number of genes 
+-- minimal, informally meaning that the system's phenotype is achieved with the fewest possible genes 
 -- has an identical impulse response and therefore an identical phenotype given an identical input $u(t)$, up to a change of coordinates. 
@@
   \end{equation}
-\plr{what are we calling these -- not "biological system" -- that's too broad}\js{how about ``molecular systems''?}
-Two molecular systems, $\mathcal{S} = \{ A, B, C \}$, and $\bar{\mathcal{S}} = \{\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} \}$, 
+\plr{what are we calling these -- not "biological system" -- that's too broad}\js{I see your point, but I don't want to bog down the paper by constantly qualifying statements -- it seemsnatural to refer to biological systems, since these are, afterall dynamical systems.}
+Two biological systems, $\mathcal{S} = \{ A, B, C \}$, and $\bar{\mathcal{S}} = \{\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} \}$, 
 have the same phenotype, for all possible inputs, if they are related by a change of coordinates. 
@@
 Therefore, if a system is minimal, the set of all phenotypically identical systems $\mathcal{S}(V)$ can be parameterized as,
-    \begin{align}
+    \begin{align} \label{eq:min_param}
       \mathcal{S}(V) := \left\{ \begin{array}{ll} V\dot{\kappa}(t) &= VA V^{-1} \kappa(t) + VB u(t) \\ \phi(t) &= C V^{-1} \kappa(t) \end{array} \right.
@@
   where $V$ is any invertible matrix and it's elements are free parameters.
+
+Therefore varying $V$ can tweak the relationships between genes within a regulatory network, yet preserve the phenotype.
+Some gene networks, however, can grown or shrink, perhaps following gene duplications and deletions, and also still preserve their phenotypes.
+These changes cannot be described by \ref{eq:min_param}, but they are ecanpuslated by \ref{eqn:equivalence}, below.
 More generally,
-we denote by $\calA_n(A_0)$ the set of all $n$-dimensional systems equivalent to $A_0$:
+we denote by $\calA_n(\mathcal{S}_0)$ the set of all $n$-dimensional systems equivalent to $\mathcal{S}_0$:
 \begin{equation} \label{eqn:equivalence}
   \begin{aligned}
-  \calA_n(A_0) 
+    \calA_n(\mathcal{S}_0) 
       &= \left\{
-          A : C e^{At} B = C e^{A_0 t} B \; \text{for}\; t \ge 0 
+        (A,B,C) : C e^{At} B = C_0 e^{A_0 t} B_0 \; \text{for}\; t \ge 0 
       \right\}  \\
       &= \left\{
-          A : C A^k B = C A_0^k B \; \text{for}\; 1 \le k \le n-1 
+        (A,B,C) : C A^r B = C_0 A_0^r B_0 \; \text{for}\; 1 \le r \le n-1 
       \right\} .
@@
 
-\plr{should say more about this decomposition here.}
-Regardless of minimality, two systems, even in different dimensions, can have identical external dynamics if they are in $\mathcal{A}$.
-This set can be completely parameterized using the \emph{Kalman decomposition} \ref{apx:kalman}; the set of all linear gene regulatory networks with equivalent phenotypes can be precisely defined.
-Note that this implies that there is always more than one possible gene regulatory network architecture per phenotype, however we can analytically describe the complete set.
+Two systems, even if in different dimensions, can have identical phenotypes if they are in $\calA_n$.
+Further, there is no unique triple $(A,B,C)$ for the mapping $u \mapsto \phi$.
+This implies that the gene regulatory network architecture per phenotype/environment pair is never unique -- that is \emph{all} gene regulatory networks can drift/rewire.
+This set can be precisley defined and completely parameterized using the \emph{Kalman decomposition} \ref{apx:kalman}.
 
@@
 %
+ \begin{example}[All Phenotypically Equivalent Cell Cycle Control Networks] \label{ex:all_osc}
+
+%\plr{this is same as above, so reference instead of writing out again?}
+%    Let
+%    \begin{align*}
+%      A(0) = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix}, \qquad B = \begin{bmatrix} 1 & 1 \end{bmatrix}^{T}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix},
+%    \end{align*}
+%    with $V(\tau)$ preserving both $B$ and $C$, then
+    The set of all two-gene regulatory networks phenotypically equivalent to the cell cycle control network in \ref{ex:oscillator}, where only $A$ can vary, is given by
+    \begin{align*}
+      A(\tau) &= \begin{bmatrix} 1 & 0 \\ \tau & 1-\tau \end{bmatrix} \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 0 \\ \frac{\tau}{\tau-1} & \frac{-1}{\tau-1} \end{bmatrix}  \qquad \forall \ \tau \neq 1 %\\
+     %   B &= \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix} \\
+       %   h(t) &= \sin(t) + \cos(t) \qquad \forall \ \tau \neq 1
+    \end{align*}
+    \begin{figure}[H]
+    \centering
+% Nice tikz!!!
+\begin{tikzpicture}
+\begin{scope}[every node/.style={circle,thick,draw}]
+  \node (A) at (0,0) {$\kappa_{1}$};
+    \node[dashed] (B) at (4,0) {$\kappa_{2}$};
+    \node[shape=rectangle] (U) at (2,2) {input};
+    \node[shape=rectangle] (y) at (2,-2) {output};
+\end{scope}
+
+\begin{scope}[>={Stealth[black]},
+              every node/.style={fill=white,circle},
+              every edge/.style={draw=black, thick}]
+    \path [->, sloped] (A) edge[bend left] node {\tiny $2 \tau + (\tau-1)^{-1}$} (B);
+    \path [->, sloped] (B) edge[bend left] node {\tiny $-(\tau-1)^{-1}$} (A); 
+    \path[->] (U) edge[dashed] node {\tiny $1$} (A);
+    \path[->] (U) edge[dashed] node {\tiny $1$} (B);
+    \path[->] (A) edge[dashed, bend right] node {\tiny $1$} (y);
+\end{scope}
+\begin{scope}[>={Stealth[black]},
+              every edge/.style={draw=black, thick}]
+    \path [->] (A) edge[loop left] node[sloped, anchor=center, above] {\tiny $1 + (\tau-1)^{-1}$} (A);
+    \path [->] (B) edge[loop right] node[sloped, anchor=center, above] {\tiny $-1 - (\tau-1)^{-1}$} (B);
+\end{scope}
+
+\end{tikzpicture}
+    \caption{The set of all phenotype-equivalent cell cycle control networks, $A(\tau)$.
+      $A(\tau) = \left\{ \calA_2 : B = B_0,\ C = C_0\right\}$.}
+    \end{figure}
+ % \end{example}
+
+ % \begin{example}[External Equivalence does not imply internal equivalence]
+   Despite the phenotypic equivalence of all instantiations of $A(\tau)$, the kryptotypes, vary as a function of $\tau$. 
+    Gene-1 dynamics (blue) are equivalent for network architectures $A(0)$ and $A(2)$, however the dynamics of gene-2 (orange) differ with $\tau$.
+  \begin{figure}[H]
+    \centering
+    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{osc_A0_A2_both_compare}
+    \caption{Gene-1 (blue) and gene-2 (orange) dynamics for $A(0)$ (top) and $A(2)$ (bottom).
+Both (top and bottom) gene-1 dynamics are given by $ \kappa_{1} = \sin(t) + \cos(t)$, and gene-2 by $\kappa_{2} = \cos(t) - \sin(t)$ (top) and $\kappa'_{2} = \cos(t) + 3 \sin(t)$ (bottom).}
+  \end{figure}
+\end{example}
+
+\paragraph{Sexual reproduction and speciation}
+A diploid organism contains two gene network copies: one from each parent.
+As such, a diploid's system dynamics is computed by averaging the coefficients of both systems.
+To reproduce, each organism passes on a recombined haploid gene network to its offspring.
+Haploid gene networks swap system matrix rows randomly between its own two networks.
+Therefore the phenotypic dynamics typical of a first generation ($F_{1}$) cross between two allopatric populations will be determined simply by averaging it's two parental genomes.
+However, in second generation hybrid ($F_2$) crosses, first new haploid systems will be formed by recombination -- in the process shuffling and combining regulatory coefficients from allopatric populations -- and then brought together to form a diploid.
+
+  \begin{example}[Hybrid Incompatibility in an Oscillating Gene Network] \label{ex:hybrid_osc}
+
+Here we compare the phenotypes for F2 hybrids formed by crossing oscillators $A(2)$ with $A(2.01)$, $A(2.1)$, and $A(2.5)$ ($B$ and $C$ are the same as above). 
+Each $A$ is phenotypically identical ($\phi(t) = \sin(t) + \cos(t)$), however some of the hybrids exhibit markedly different dynamics.
+
+%        \begin{align*}
+%         A(2) = \begin{bmatrix} 2 & -1 \\ 5 & -2 \end{bmatrix} &\qquad A(2.01) = \begin{bmatrix} 1.9901 & -0.9901 \\ 5.0101 & -1.9901 \end{bmatrix} \\
+%          A(2.1) = \begin{bmatrix} 1.9091 & -0.9091 \\ 5.1091 & -1.9091 \end{bmatrix} &\qquad A(2.5) = \begin{bmatrix} 1.6667 & -0.6667 \\ 5.6667 & -1.6667 \end{bmatrix}
+%     \end{align*}
+
+\begin{alignat*}{2}
+      A(2) &= \left[\begin{array}{cc} 2 & -1 \\[6pt] 5 & -2 \end{array}\right] &&\qquad A(2.01) = \left[\begin{array}{cc} 2 -\frac{1}{101} & -1 + \frac{1}{101} \\[6pt] 5 + \frac{1}{99} & -2 + \frac{1}{101} \end{array}\right] \\
+        A(2.1) &= \left[\begin{array}{cc} 2 - \frac{1}{11} & -1 + \frac{1}{11} \\[6pt] 5 + \frac{6}{55} & -2 + \frac{1}{11} \end{array}\right] &&\qquad A(2.5) = \left[\begin{array}{cc} 2 - \frac{1}{3} & -1 + \frac{1}{3} \\[6pt] 5 + \frac{2}{3} & -2 + \frac{1}{3} \end{array}\right]
+\end{alignat*}
+
+%      \begin{figure}[H]
+%        \begin{tabular}{cc}
+%          \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_1}
+%          & \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_2}
+%         \\ \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_3}
+%          & \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_4}
+%          \\ \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_5}
+%          & \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_6}
+%        \end{tabular}
+%      \end{figure}
+
+%    \begin{figure}[H]
+%      \centering
+%      \includegraphics[width=0.5\textwidth]{A2_A2-1_F2s}
+%      \caption{F2s from $A(2)$ and $A(2.1)$.}
+%    \end{figure}
+%      \begin{figure}[H]
+%        \begin{tabular}{cc}
+%        \includegraphics[width=0.5\textwidth]{F2s-small} \\
+%        \includegraphics[width=0.5\textwidth]{F2s} \\
+%        \includegraphics[width=0.5\textwidth]{F2s-large}
+%        \end{tabular}
+%        \caption{F2s from $A(2)$ and $A(2.1)$.}
+%      \end{figure}
+
+\begin{figure}[H]
+  \centering
+  \begin{tabular}{cc}
+    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{F1_comparison} & 
+    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{F2s_comparison2}
+  \end{tabular}
+  \caption{
+    $F_1$ (left) and $F_2$ (right) hybrids crossing $A(2)$ with $A(2.01)$ (top), $A(2.1)$ (middle), and $A(2.5)$ (bottom).
+    Note the difference in scale on the $y$-axis.
+F2 hybrids display more phenotypic divergence than F1s, on average.
+Further, some F2s completely fail to oscillate, as seen in an $A(2.5)$ F2 (light blue).
+    \plri{make axis labels bigger}
+  }
+\end{figure}
+
+    \jss{$A(2)$ and $A(2.1)$ differ in regulatory interaction strengths by $5.09\%$.
+If intra-population regulatory variation is approximately $5\%$, then this level of divergence is expected after $\sqrt{\frac{t}{N_{e}} = 1}$ generations.}
+\end{example}
 %
@@
 we will write the ``distance'' between phenotypes $\phi$ and $\psi$ as $\dph(\phi, \psi)$,
-measured so that the fitness of an organism with coefficents $x$ is $\fitx(x) = \exp(-\dph(\Phi(x),\optph)^2)$.
+measured so that the fitness of an organism with coefficents $x$ is $\fitx(x) = \exp(-\dph(\Phi(x),\optph)^2)$. \js{need to add $\gamma^{2}$ here?}
 We will assume that the map $\Phi$ is smooth
@@
 then since subsequent generations resample from this diversity,
-the population mean coefficient will move a random distance of size $\sigma/\sqrt{N_e}$ per generation,
+the population mean coefficient will move a random distance of size $\sigma/\sqrt{N_e}$ per generation, \js{since it moves at rate $\sigma \sqrt{T/N_e}$ isn't saying  $\sigma/\sqrt{N_e}$ \emph{per generation} not that clear?}
 simply because this is the the standard deviation of the mean of a random sample \citep{lande_drift}.
@@
 
-\paragraph{Sexual reproduction}
-A diploid first generation hybird's ($F_1$) genetic regulation is a consequence of its two genomes, as a copy of genes and regulatory sequences from each parent is present.
-\plr{should have a stronger justification for the assertion that $F_1$ is the arithmetic average of the parents than "Thus we say"}
-A diploid organism contains two gene network copies: one from each parent.
-As such, a diploid's system dynamics is computed by averaging the coefficients of both systems.
-To reproduce, each organism passes on a recombined haploid gene network to its offspring.
-Haploid gene networks swap system matrix rows randomly between its own two networks.
-Therefore the phenotypic dynamics typical of an $F_{1}$ cross between two allopatric populations will be determined simply by averaging it's two parental genomes.
-However, in second generation hybrid crosses $F_2$ between allopatric populations, first new haploid systems will be formed by recombination -- in the process shuffling and combining regulatory coefficients from allopatric populations -- and then brought together to form a diploid.
-
 \paragraph{Hybridization}
@@
 and so will have mean system coefficients at the midpoint between their means,
-and variance equal to $\sigma$.
+and variance equal to $\sigma$. \js{I think you mean $\sigma^{2}$?}
 Each $F_2$ hybrid will be homozygous for one parental allele on averge at half of the loci in the genome,
@@
 as before,
-but variance equal to $\sigma^2 + z^2/2$, where $z$ is the distance between the parental populations.
+but variance equal to $\sigma^2 + z^2/2$, where $z$ is the distance between the parental populations. \js{Can you clarify why $z^{2}/2$? Is this because it's averaging two distributions?}
 These are depicted in figure \ref{fig:conceptual_fig}.
@@
 
+  \jss{ Assume that selection acts directly against organisms whose regulatory coefficients are not on the optimum manifold.
+Further assume that selection constrains this diversity to be, on average, a distance of $m$ from optimum.
+We would expect $\sigma_S = m$.
+Furtheremore, two individuals within a population may both be on the optimum manifold, but at different points, on average of $\sigma_N$.
+Since individuals a distance of $\sigma_N$ apart will produce $F_2$s off of the manifold of $\mu = \frac{\sigma_N}{2}\sin2\theta$, recombination load may constrain $\sigma_N$ such that $\mu \leq m = \sigma_S$ (where $\theta$ is the slope of the manifold -- as $\theta$ approaches $0$ or $1$, $\sigma_N \rightarrow \infty$.).
+We can estimate $\sigma_N = \frac{2\sigma_S}{\sin2\theta}$.
+  }
 %%%%%%%%%%%%%%%%%%%%%%%%
@@
 is strongly deleterious,
-so that if $u$ is the typical scale on which selection acts,
-then $u > \sigma$.  
-\plr{do we call this $\beta$ or $u$?}
+so that if $\beta$ is the typical scale on which selection acts,
+then $\beta > \sigma$.  
+\plr{do we call this $\beta$ or $u$?} \js{$\beta$, $u$ is input.}
 However,
@@
 the strength of selection on phenotype is sufficient to weakly constrain regulatory variation,
-so that perhaps $\sigma$ and $u$ are relatively close.
+so that perhaps $\sigma$ and $\beta$ are relatively close.
 This is as would be expected if available variation is held in check by mutation--selection balance
 rather than genetic drift.
-A conservative estimate would be that $u = 5 \sigma$;
+A conservative estimate would be that $\beta = 5 \sigma$;
 taking $\sigma=.01$ as above, this suggests that changes in phenotype of 5\% are sufficient
@@
 
-\section*{The rate of speciation under neutrality in allopatric populations}
+\subsection*{The rate of speciation under neutrality in allopatric populations}
 
@@
 
-This suggests that reproductively isolated populations can speciate rapidly -- on timescales much less than $N_{e}$ generations, depending on the specifics of the fitness function.
+$F_2$s between allopatric populations will have an average fitness relative to their parents of,
+  \begin{align}
+    \exp\left(-\left( \frac{2\sigma_{N}}{\beta}\sqrt{\frac{T}{N_e}} \right)^{2} \right)
+  \end{align}
+where $\beta$ is the strength of selection against phenotypic divergence.
+
+If the strength of selection is only a few multiples $\xi_1$ of the standing phenotypic diversity within a population $\sigma_S$, such that $\beta = \xi_1 \sigma_S$, and recombination load constrains the standing kryptotype diversity $\sigma_N$ to be only a few multiples $\xi_2$ of $\sigma_S$, then $F_2$ fitness will be,
+  \begin{align}
+    \exp\left(-\left( \xi \sqrt{\frac{T}{N_e}} \right)^{2} \right)
+  \end{align}
+  where $\xi := 2\xi_2/\xi_1$.
+  This suggests that if a population contains as much phenotypic diversity as allowed by selection and if $\sigma_N$ is on the order of $\sigma_S$, reproductively isolated populations can speciate rapidly -- on timescales shorter than $N_{e}$ generations.
 
@@
 \section*{Examples}
-  \begin{example}[Oscillating Gene Network: Cell Cycle Control]\label{ex:oscillator}
-
-    Cellular division is a complicated phenomena, governed by many different processes, however it is agreed that its rhythm is partially controlled by periodic (oscillating) gene transcription \citep{orlando2008global}.
-Consider a simplified model of oscillating gene transcription.
-In the present framework periodic expression requires at minimum two interacting genes.
-
-    Suppose gene-2 up-regulates the transcription of gene-1 and that gene-1 down-regulates the transcription of gene-2 with equal magnitude (of $1$) and relative to each of their concentrations, denoted by $\kappa_{1}$ and $\kappa_{2}$.
-Furthermore, suppose that only the dynamics of gene-1 are consequential to the cell cycle (perhaps the amount of gene-1 activates another downstream gene network).
-Lastly suppose that the production of both genes is stimulated by an impulse of a molecule present immediately after division.
-
-\plr{you haven't explained $u(t)$ above. and, isn't it more natural in this set-up to have $B = [1,0]$ than $[1,1]$?}
-    If the rate each of these genes is expressed is a linear function of their concentrations, the dynamics of the system are given by
-    \begin{align*}
-      \dot{\kappa_{1}}(t) &= \kappa_{2}(t) + u(t) \\
-      \dot{\kappa_{2}}(t) &= - \kappa_{1}(t) + u(t)
-    \end{align*}
-    where $\dot{\kappa}$ denotes the time derivative.
-The initial conditions $\kappa_{1}(0)$ and $\kappa_{2}(0)$, and the input $u(t)$ then determine the concentrations through time.
-If we record the regulatory coefficients in the matrix
-    \begin{align*}
-      A &= \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} ,
-    \end{align*}
-    and define the column vector $B = \begin{bmatrix} 1 & 1 \end{bmatrix}^{T}$, then in matrix notation the dynamics are
-      \begin{align*}
-      \dot{\kappa} &= A \kappa(t) + B u(t) .
-      \end{align*}
-      Since only the dynamics of gene-1 are directly relevant to biological function, the dynamics of interest are given by
-      \begin{align*}
-      \phi(t) &= C \kappa (t)
-      \end{align*}
-      where the row vector $C$ is defined as $C = \begin{bmatrix} 1 & 0 \end{bmatrix}$. 
-      (If the dynamics of both genes were physiologically relevant to the cell cycle, we would set $C$ to be the identity matrix).
-  %  \begin{align*}
-  %    A = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} , \qquad B = \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix}
-  %  \end{align*}
- %       The oscillatory system $\Sigma$ is thus given as
- %   \begin{align*}
- %     \Sigma = \left \{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 
- %       0 & 1 \\ 
- %      -1 & 0 
- %       \end{bmatrix} \kappa(t) + \begin{bmatrix} 1 \\ 1 \end{bmatrix} u(t) \\ 
- %         \phi(t) &= \begin{bmatrix} 1 & 0 \end{bmatrix} \kappa(t) \end{array} \right.
- %    \end{align*}
-
-      Since the input is simply an impulse, its phenotype is equivalent to its impulse response
-      \begin{align*}
-        \phi(t) = h(t) = \sin(t) + \cos(t)  .
-      \end{align*}
-
-    \begin{figure}[H]
-       % \begin{center} 
-      \centering
-         \begin{tabular}{cc}
-            \begin{tikzpicture}
-            \begin{scope}[every node/.style={circle,thick,draw}]
-                \node (A) at (0,0) {$\kappa_{1}$};
-                \node[dashed] (B) at (3,0) {$\kappa_{2}$};
-                \node[shape=rectangle] (U) at (1.5,2) {input};
-                \node[shape=rectangle] (y) at (1.5,-2) {output};
-            \end{scope}
-
-            \begin{scope}[>={Stealth[black]},
-                          every node/.style={fill=white,circle},
-                          every edge/.style={draw=black, thick}]
-                \path [->, >=Rectangle] (A) edge[bend left] node {\tiny $-1$} (B);
-                \path [->] (B) edge[bend left] node {\tiny $1$} (A); 
-                \path[->] (U) edge[dashed] node {\tiny $1$} (A);
-                \path[->] (U) edge[dashed] node {\tiny $1$} (B);
-                \path[->] (A) edge[dashed,bend right] node {\tiny $1$} (y);
-            \end{scope}
-            \begin{scope}[>={Stealth[black]},
-                          every edge/.style={draw=black, thick}]
-                %\path [->] (A) edge[loop left] node {\tiny $\lambda_{1}$} (A);
-                %\path [->] (B) edge[loop left] node {\tiny $\lambda_{2}$} (B);
-            \end{scope}
-
-            \end{tikzpicture} &
-       % \end{center}
-      %  \caption{
-      %      Diagram of Example \ref{ex:osc} in the text.
-      %      \plr{explain what arrows mean if nec}
-      %      \label{fig:oscillator_diagram}}
-   % \end{figure}
-   % \begin{figure}[H]
-    %  \centering
-      \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{osc_impulse}
-   \end{tabular}
-  %  \end{center}
-      \caption{(Left) Graphical representation of the cell cycle control gene network, and (right) plot of the phenotype $\phi(t)$ against time $t$.}
-    \end{figure}
-      We return to the evolution of such a system below.
-  \end{example}
-
-  \begin{example}[All Phenotypically Equivalent Cell Cycle Control Networks] \label{ex:all_osc}
-
-%\plr{this is same as above, so reference instead of writing out again?}
-%    Let
-%    \begin{align*}
-%      A(0) = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix}, \qquad B = \begin{bmatrix} 1 & 1 \end{bmatrix}^{T}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix},
-%    \end{align*}
-%    with $V(\tau)$ preserving both $B$ and $C$, then
-    The set of all two-gene regulatory networks phenotypically equivalent to the cell cycle control network in \ref{ex:oscillator} is given by
-    \begin{align*}
-      A(\tau) &= \begin{bmatrix} 1 & 0 \\ \tau & 1-\tau \end{bmatrix} \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 0 \\ \frac{\tau}{\tau-1} & \frac{-1}{\tau-1} \end{bmatrix}  \qquad \forall \ \tau \neq 1 %\\
-     %   B &= \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix} \\
-       %   h(t) &= \sin(t) + \cos(t) \qquad \forall \ \tau \neq 1
-    \end{align*}
-    \begin{figure}[H]
-    \centering
-% Nice tikz!!!
-\begin{tikzpicture}
-\begin{scope}[every node/.style={circle,thick,draw}]
-  \node (A) at (0,0) {$\kappa_{1}$};
-    \node[dashed] (B) at (4,0) {$\kappa_{2}$};
-    \node[shape=rectangle] (U) at (2,2) {input};
-    \node[shape=rectangle] (y) at (2,-2) {output};
-\end{scope}
-
-\begin{scope}[>={Stealth[black]},
-              every node/.style={fill=white,circle},
-              every edge/.style={draw=black, thick}]
-    \path [->, sloped] (A) edge[bend left] node {\tiny $2 \tau + (\tau-1)^{-1}$} (B);
-    \path [->, sloped] (B) edge[bend left] node {\tiny $-(\tau-1)^{-1}$} (A); 
-    \path[->] (U) edge[dashed] node {\tiny $1$} (A);
-    \path[->] (U) edge[dashed] node {\tiny $1$} (B);
-    \path[->] (A) edge[dashed, bend right] node {\tiny $1$} (y);
-\end{scope}
-\begin{scope}[>={Stealth[black]},
-              every edge/.style={draw=black, thick}]
-    \path [->] (A) edge[loop left] node[sloped, anchor=center, above] {\tiny $1 + (\tau-1)^{-1}$} (A);
-    \path [->] (B) edge[loop right] node[sloped, anchor=center, above] {\tiny $-1 - (\tau-1)^{-1}$} (B);
-\end{scope}
-
-\end{tikzpicture}
-    \caption{A graphical depiction of the set of all externally equivalent cell cycle control networks, $A(\tau)$. 
-        $\tau$ can take any value \plri{sentence fragment}}
-    \end{figure}
- % \end{example}
-
- % \begin{example}[External Equivalence does not imply internal equivalence]
-   Despite the phenotypic equivalence of all instantiations of $A(\tau)$, the internal dynamics, or kryptotypes, vary as a function of $\tau$. 
-    Gene-1 dynamics (blue) are equivalent for network architectures $A(0)$ and $A(2)$, however the dynamics of gene-2 (orange) differ with $\tau$.
-  \begin{figure}[H]
-    \centering
-    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{osc_A0_A2_both_compare}
-    \caption{Gene-1 (blue) and gene-2 (orange) dynamics for $A(0)$ (top) and $A(2)$ (bottom).
-Both (top and bottom) gene-1 dynamics are given by $ \kappa_{1} = \sin(t) + \cos(t)$, and gene-2 by $\kappa_{2} = \cos(t) - \sin(t)$ (top) and $\kappa'_{2} = \cos(t) + 3 \sin(t)$ (bottom).}
-  \end{figure}
-\end{example}
-
-  \begin{example}[Hybrid Incompatibility in an Oscillating Gene Network] \label{ex:hybrid_osc}
-
-Here we compare the phenotypes for F2 hybrids formed by crossing oscillators $A(2)$ with $A(2.01)$, $A(2.1)$, and $A(2.5)$ ($B$ and $C$ are the same as above). 
-Each $A$ is phenotypically identical ($\phi(t) = \sin(t) + \cos(t)$), however some of the hybrids exhibit markedly different dynamics.
-
-%        \begin{align*}
-%         A(2) = \begin{bmatrix} 2 & -1 \\ 5 & -2 \end{bmatrix} &\qquad A(2.01) = \begin{bmatrix} 1.9901 & -0.9901 \\ 5.0101 & -1.9901 \end{bmatrix} \\
-%          A(2.1) = \begin{bmatrix} 1.9091 & -0.9091 \\ 5.1091 & -1.9091 \end{bmatrix} &\qquad A(2.5) = \begin{bmatrix} 1.6667 & -0.6667 \\ 5.6667 & -1.6667 \end{bmatrix}
-%     \end{align*}
-
-\begin{alignat*}{2}
-      A(2) &= \left[\begin{array}{cc} 2 & -1 \\[6pt] 5 & -2 \end{array}\right] &&\qquad A(2.01) = \left[\begin{array}{cc} 2 -\frac{1}{101} & -1 + \frac{1}{101} \\[6pt] 5 + \frac{1}{99} & -2 + \frac{1}{101} \end{array}\right] \\
-        A(2.1) &= \left[\begin{array}{cc} 2 - \frac{1}{11} & -1 + \frac{1}{11} \\[6pt] 5 + \frac{6}{55} & -2 + \frac{1}{11} \end{array}\right] &&\qquad A(2.5) = \left[\begin{array}{cc} 2 - \frac{1}{3} & -1 + \frac{1}{3} \\[6pt] 5 + \frac{2}{3} & -2 + \frac{1}{3} \end{array}\right]
-\end{alignat*}
-
-%      \begin{figure}[H]
-%        \begin{tabular}{cc}
-%          \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_1}
-%          & \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_2}
-%         \\ \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_3}
-%          & \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_4}
-%          \\ \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_5}
-%          & \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_6}
-%        \end{tabular}
-%      \end{figure}
-
-%    \begin{figure}[H]
-%      \centering
-%      \includegraphics[width=0.5\textwidth]{A2_A2-1_F2s}
-%      \caption{F2s from $A(2)$ and $A(2.1)$.}
-%    \end{figure}
-%      \begin{figure}[H]
-%        \begin{tabular}{cc}
-%        \includegraphics[width=0.5\textwidth]{F2s-small} \\
-%        \includegraphics[width=0.5\textwidth]{F2s} \\
-%        \includegraphics[width=0.5\textwidth]{F2s-large}
-%        \end{tabular}
-%        \caption{F2s from $A(2)$ and $A(2.1)$.}
-%      \end{figure}
-
-\begin{figure}[H]
-  \centering
-  \begin{tabular}{cc}
-    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{F1_comparison} & 
-    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{F2s_comparison2}
-  \end{tabular}
-  \caption{
-    $F_1$ (left) and $F_2$ (right) hybrids crossing $A(2)$ with $A(2.01)$ (top), $A(2.1)$ (middle), and $A(2.5)$ (bottom).
-    Note the difference in scale on the $y$-axis.
-F2 hybrids display more phenotypic divergence than F1s, on average.
-Further, some F2s completely fail to oscillate, as seen in an $A(2.5)$ F2 (light blue).
-    \plri{make axis labels bigger}
-  }
-\end{figure}
-
-    \jss{$A(2)$ and $A(2.1)$ differ in regulatory interaction strengths by $5.09\%$.
-If intra-population regulatory variation is approximately $5\%$, then this level of divergence is expected after $\sqrt{\frac{t}{N_{e}} = 1}$ generations.}
-\end{example}
 
@@
 
-\section{Hybrid Vigor}
+\section{Hybrid Vigor: Unit Circle approach}
 
@@
   \begin{align*}
-  \mathbb{E} \left[ d(F_1)] \right] &= \int_{0}^{2\pi} \int_{0}^{2\pi} \left\lVert \frac{1}{8 \pi^{2}} \left( \left( \cos(\theta_1), \sin(\theta_1) \right) + \left( \cos(\theta_2), \sin(\theta_2) \right) \right) \right\rVert d\theta_1 d\theta_2 \\
-  &= \cos(\frac{\pi}{2}) = \frac{\sqrt{2}}{2}
-\end{align*}
-Haploid $F_2$s will, on average, be a distance of $1$ half of the time, a distance of $0$ a quarter of the time, and a distance of $\sqrt{2}$ a quarter of the time.
-
+    \mathbb{E} \left[ d(F_1) \right] &= \int_{0}^{2\pi} \frac{1}{4 \pi} \left\lVert (1,0) + (\cos\theta,\sin\theta) \right\rVert d\theta \\
+    &= \frac{2}{\pi}
+  \end{align*}
+  %\begin{align*}
+  %  \mathbb{E} \left[ d(F_1)] \right] &= \int_{0}^{2\pi} \int_{0}^{2\pi} \frac{1}{8 \pi^{2}}\left\lVert  \left( \left( \cos(\theta_1), \sin(\theta_1) \right) , \left( \cos(\theta_2), \sin(\theta_2) \right) \right) \right\rVert d\theta_1 d\theta_2 \\
+  %  &= \cos\left(\frac{\pi}{2}\right) = \frac{\sqrt{2}}{2}
+  %\end{align*}
+%  \begin{align*}
+%    \mathbb{E} \left[ d(F_1)] \right] &\approx \sqrt{\int_{0}^{2\pi} \frac{1}{4 \pi}\left\lVert  \left( \cos(\pi), \sin(\pi) \right) , \left( \cos(\theta), \sin(\theta) \right) \right\rVert^2 d\theta} \\
+%    &= \sqrt{\frac{1}{2}}
+%  \end{align*}
+%Haploid $F_2$s will, on average, be a distance of $1$ half of the time, a distance of $0$ a quarter of the time, and a distance of $\sqrt{2}$ a quarter of the time.
+
+  \begin{align*}
+    \mathbb{E} \left[ d(F_2) \right] &= \int_{0}^{2 \pi} \int_{0}^{2 \pi} \frac{1}{4 \pi^{2}} \sum_{i}^{2} \sum_{j}^{2} \sum_{k}^{2} \sum_{l}^{2} \frac{1}{32} \left\lVert \left( \cos(\theta_i) + \cos(\theta_k), \sin(\theta_j) + \sin(\theta_l) \right) \right\rVert d\theta_1 d\theta_2 \\ &= 0.6705
+  \end{align*}
+
+  \begin{align*}
+    \mathbb{E}\left[ d(F_{2_{\text{hap}}}) \right] &= \sqrt{\int_{0}^{2 \pi} \int_{0}^{2 \pi} \frac{1}{8 \pi^{2}} \cos(\theta_1)^{2} + \sin(\theta_2)^{2} d\theta_1 d\theta_2} \\
+    &= \cos \left( \frac{\pi}{2} \right) = \frac{\sqrt{2}}{2}
+  \end{align*}
+
+%  There are $16$ diploid $F_2$s, two have $F_0$ genotypes, two have $F_1$ genotpyes, eight contrain one $F_0$ and one $F_2$ haploid genotype (referred to as $H$; $\fit(H) = \sqrt{\frac{3}{8}}$), two are homozygous for an $F_2$ genotype (referred to as $W$; $\fit(W) = \frac{\sqrt{2}}{2}$), and finally two are heterozygous $F_2$s (referred to as $L$; $\fit(L) = \frac{1}{2}$.)
+  %The expected average distance of diploid $F_2$s is $0.6705$.
+
+  \section{Hybrid Vigor: Probability}
+
+  \begin{align*}
+    X &\sim \mathcal{N} (x, \sigma^{2}) \\
+    Y &\sim \mathcal{N} (y, \sigma^{2}) \\
+    x &\sim \mathcal{N} (0, \eta^{2}) \\
+    y &\sim \mathcal{N}(0, \eta^{2}) \\
+    Z &= \frac{X+Y}{2} \rightarrow \mathbb{E} \left[ \lvert z \rvert^{2} \right] \\
+    Z &= \frac{x+y+U+V}{2} \\
+    \mathbb{E} \left[ \lvert X \rvert^{2} \right] &= \mathbb{E} \left[ \lvert x + U \rvert^{2} \right] = \mathbb{E} \left[ (x+U)^{T} (x+U) \right] \\
+    &= \mathbb{E} \left[ x^{T}x + U^{T}x + x^{T}U + U^{T}T \right] \\
+    &= \mathbb{E} \left[x^{T}x\right] + \mathbb{E} \left[U^{T}U\right] \\
+    &= d\eta^{2} + d\sigma^{2} \qquad d=\text{dimension} \\
+    X &\sim \mathcal{N}(0, \sigma^{2} + \eta^{2}) \\
+    Z &\sim \mathcal{N}\left(0, \frac{1}{4}(2 \eta^{2} + 2\sigma^{2})\right) \\
+    &\rightarrow \mathbb{E}\left[\lvert Z \rvert^{2} \right] \\
+    &= \frac{d}{2} \left(\eta^{2} + \sigma^{2} \right) \\
+    &= \frac{1}{2} \mathbb{E} \left[ \lvert X \rvert^{2} \right]
+  \end{align*}
 \end{document}

### 2017-08-29 kalman_walk_paper.tex
edited system description, qgen, examples, speciation and more.

@@
 
-\section*{Gene Networks as Linear Dynamical Systems}
+\section*{Gene networks as linear dynamical systems}
 %
@@
 %}
+\paragraph{Systems theory}
 Organisms' phenotypes are constructred by gene by gene by environment interactions.
@@
 %\end{align}
- \begin{example}[Oscillating Gene Network: Cell Cycle Control]\label{ex:oscillator}
+ \begin{example}[Oscillating gene network: cell cycle control]\label{ex:oscillator}
     Cellular division is governed by many different processes, however it is thought that its rhythm is partially controlled by oscillating gene transcription \citep{orlando2008global}.
@@
 
-  \section*{Phenotype-equivalent gene networks}
+  \paragraph{Phenotypically equivalent gene networks}
 
@@
 Gene regulatory networks with identical phenotypes do not necessarily have identical kryptotypes.
-Any linear and minimal system (a gene network) 
--- minimal, informally meaning that the system's phenotype is achieved with the fewest possible genes 
--- has an identical impulse response and therefore an identical phenotype given an identical input $u(t)$, up to a change of coordinates. 
+Any linear and minimal gene network 
+-- minimal, informally meaning that the system's phenotype is achieved with the fewest possible genes $(n=m)$
+-- has an identical impulse response and thus phenotype given identical inputs $u(t)$, up to a change of coordinates. 
 \plr{"has an identical phenotype" is not what you mean to say here}
@@
 %\plr{why is this "evolutionary"? It is the thing that we're thinking about evolution of, but that doesn't make it evolutionary.}
-Therefore, if a system is minimal, the set of all phenotypically identical systems $\mathcal{S}(V)$ can be parameterized as,
-    \begin{align} \label{eq:min_param}
-      \mathcal{S}(V) := \left\{ \begin{array}{ll} V\dot{\kappa}(t) &= VA V^{-1} \kappa(t) + VB u(t) \\ \phi(t) &= C V^{-1} \kappa(t) \end{array} \right.
-  \end{align}
+Therefore the set of all phenotypically identical (and minimal) systems can be parameterized as, $\mathcal{S}_{m}(V) := \{VA_0V^{-1}, VB_0, C_0V \}$,
+  %  \begin{align} \label{eq:min_param}
+  %    \mathcal{S}(V) := \left\{ \begin{array}{ll} V\dot{\kappa}(t) &= VA V^{-1} \kappa(t) + VB u(t) \\ \phi(t) &= C V^{-1} \kappa(t) \end{array} \right.
+  %\end{align}
   where $V$ is any invertible matrix and it's elements are free parameters.
@@
 Some gene networks, however, can grown or shrink, perhaps following gene duplications and deletions, and also still preserve their phenotypes.
-These changes cannot be described by \ref{eq:min_param}, but they are ecanpuslated by \ref{eqn:equivalence}, below.
+These changes cannot be described by $S(V)$, but are ecanpuslated by \ref{eqn:equivalence}, below.
+
 More generally,
@@
 There is an optimal phenotype $\optph$,
-and a set $\optx$ of ``optimal'' coefficients that produce this phenotype.
+and a set $\optx$ of ``optimal'' coefficients that produce this phenotype ($\calA_n$ above).
 Fitness depends on distance to the optimal phenotype --
@@
 
-\subsection*{The rate of speciation under neutrality in allopatric populations}
+\paragraph{The rate of speciation under neutrality in allopatric populations}
 
@@
 
-$F_2$s between allopatric populations will have an average fitness relative to their parents of,
+$F_2$s between allopatric populations will suffer an average fitness drop relative to their parents of,
   \begin{align}
-    \exp\left(-\left( \frac{2\sigma_{N}}{\beta}\sqrt{\frac{T}{N_e}} \right)^{2} \right)
+    \exp\left(-\left( \frac{2\sigma_{N}}{\beta}\sqrt{\frac{T}{N_e}} \right)^{2} \right) .
   \end{align}
-where $\beta$ is the strength of selection against phenotypic divergence.
-
-If the strength of selection is only a few multiples $\xi_1$ of the standing phenotypic diversity within a population $\sigma_S$, such that $\beta = \xi_1 \sigma_S$, and recombination load constrains the standing kryptotype diversity $\sigma_N$ to be only a few multiples $\xi_2$ of $\sigma_S$, then $F_2$ fitness will be,
+
+If the strength of selection is only a few multiples $\xi_1$ of the standing phenotypic diversity within a population $\sigma_S$, such that $\beta = \xi_1 \sigma_S$, and recombination load constrains the standing phenotype-neutral diversity $\sigma_N$ to be only a few multiples $\xi_2$ of $\sigma_S$, then $F_2$ fitness drop will be,
   \begin{align}
-    \exp\left(-\left( \xi \sqrt{\frac{T}{N_e}} \right)^{2} \right)
+    \exp\left(- \xi^{2} \frac{T}{N_e} \right)
   \end{align}
   where $\xi := 2\xi_2/\xi_1$.
-  This suggests that if a population contains as much phenotypic diversity as allowed by selection and if $\sigma_N$ is on the order of $\sigma_S$, reproductively isolated populations can speciate rapidly -- on timescales shorter than $N_{e}$ generations.
+  This suggests that if a population contains as much phenotypic diversity as allowed by selection and if $\sigma_N$ is on the order of $\sigma_S$, reproductively isolated populations can speciate rapidly -- possibly on timescales shorter than $N_{e}$ generations.
 

### 2017-08-29 kalman_walk_paper.tex
edited system description, qgen, examples, speciation and more.

@@
   \end{equation}
-\plr{what are we calling these -- not "biological system" -- that's too broad}\js{I see your point, but I don't want to bog down the paper by constantly qualifying statements -- it seemsnatural to refer to biological systems, since these are, afterall dynamical systems.}
-Two biological systems, $\mathcal{S} = \{ A, B, C \}$, and $\bar{\mathcal{S}} = \{\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} \}$, 
+Two gene networks, $\mathcal{S} = \{ A, B, C \}$, and $\bar{\mathcal{S}} = \{\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} \}$, 
 have the same phenotype, for all possible inputs, if they are related by a change of coordinates. 

### 2017-08-31 kalman_walk_paper.tex
edited speciation section

@@
 Furthermore, whereas the phenotype is a subset of molecules \emph{visible} to selection, the \emph{kryptotype} includes the molecular dynamics \emph{hidden} from selection.
-That is $\kappa(t)$ is a vector of the system's molecular concentrations at time $t$, directly, indrectly, and irrelevant to survival.
+That is $\kappa(t)$ is a vector of the system's molecular concentrations at time $t$.
 %
@@
 Any linear and minimal gene network 
--- minimal, informally meaning that the system's phenotype is achieved with the fewest possible genes $(n=m)$
+-- minimal, informally meaning that the system's phenotype is achieved with the fewest possible genes $(n=\min)$
 -- has an identical impulse response and thus phenotype given identical inputs $u(t)$, up to a change of coordinates. 
@@
 %\plr{why is this "evolutionary"? It is the thing that we're thinking about evolution of, but that doesn't make it evolutionary.}
-Therefore the set of all phenotypically identical (and minimal) systems can be parameterized as, $\mathcal{S}_{m}(V) := \{VA_0V^{-1}, VB_0, C_0V \}$,
+Therefore the set of all phenotypically identical (and minimal) systems can be parameterized as, $\mathcal{S}_{\min}(V) := \{VA_0V^{-1}, VB_0, C_0V \}$,
   %  \begin{align} \label{eq:min_param}
@@
 \end{equation}
+  \begin{align}
+    \mathcal{K} := \bigcup_{n=\min}^{\infty} \calA_n(\calA_0)
+  \end{align}
 Equivalence of the two characterizations follows from the Cayley-Hamilton theorem.
@@
 
-Two systems, even if in different dimensions, can have identical phenotypes if they are in $\calA_n$.
+Two systems, even if in different dimensions, will have identical phenotypes if they are in $\mathcal{K}$.
 Further, there is no unique triple $(A,B,C)$ for the mapping $u \mapsto \phi$.
@@
 This set can be precisley defined and completely parameterized using the \emph{Kalman decomposition} \ref{apx:kalman}.
+
+A large, adapted population will move around in $\mathcal{K}$ given environmental and selective stasis.
+Even if there is a conservation of phenotype through evolutionary time, the molecular mechanism underpinning the phenotype is not constrained and likely will not be conserved.
 
@@
 \end{example}
+
+\paragraph{Kryptotypic variation and hybrid incompatibility}
+Phenotypically identical yet kryptotypically variant gene networks, when brought together and/or recombined during sexual reproduction, can lead to phenotypic inviability.
+
+For instance, the phenotypic dynamics of a diploid organism are a function of its two haploid genomes $x$ and $y$: $P(x,y) = \left(x + y\right)/2$.
+Additionally, during meiosis, a diploid recombines its two genomes to form a new one: $\left(x \times y \right) \rightarrow z$; by randomly swapping matrix rows between two systems $\mathcal{S}$ and $\mathcal{S}'$.
+
+Even if two genomes $x$ and $y$ are both in $\mathcal{K}$, their diploid systems $P(x,y)$, and their recombinant systems $P(z,z)$, may not be in $\mathcal{K}$.
+This is because $\mathcal{K}$ is not mathematically required to be closed under averaging and recombination.
+
+If sexual recombination among systems drawn from $\mathcal{K}$ yields systems with divergent phenotypes, populations containing significant diversity in $\mathcal{K}$ will carry a large genetic load, and isolated populations may fail to produce hybrids with viable phenotypes.
 

### 2017-09-09 kalman_walk_paper.tex
edited speciation rate envelope calculation. added figure of phenotypic divergence to the speciation example. edited discussion, and minor edits to other sections.

@@
 \documentclass{article}
-\usepackage{amsmath, amssymb, color, xcolor}
+\usepackage{amsmath, amssymb, color, xcolor, amsthm}
 \usepackage{graphicx, wrapfig, float, caption, dsfont, bbm}
@@
 
-\newcommand{\calA}{\mathcal{S}}
+\newcommand{\calA}{\mathcal{A}}
 \newcommand{\calH}{\mathcal{H}}
@@
 \newcommand{\1}{\mathbbm{1}}
+
+\newcommand{\Sys}{\mathcal{S}}
 
@@
 \section*{Introduction}
-
 Bridging the gulf between an organism's genome and phenotype is a poorly understood and complex molecular machinery. 
@@
 %Here we outline an analytical theory to study the evolution of biological systems, borrowing insight and methods from control engineering, systems identification, and dynamical systems theory. 
-
+%
 %Presently, we focus on the neutral evolution of genetic regulatory networks.
@@
 This analysis provides insight to the mechanisms and parameters important for understanding developmental systems drift, network rewiring, evolvability, epistasis, and speciation, as well as the tenuous connection between network architecture and function.
-
+%
+%
 %It is commonly taught that an organism's genome contains the heritable material 
@@
 Speciation typically occurs on timescales approximately on the order of $N_{e}$ generations, where $N_{e}$ is the effective population size.
-
+%
 \section*{Gene networks as linear dynamical systems}
@@
 Some gene networks, however, can grown or shrink, perhaps following gene duplications and deletions, and also still preserve their phenotypes.
-These changes cannot be described by $S(V)$, but are ecanpuslated by \ref{eqn:equivalence}, below.
+These changes cannot be described by $\mathcal{S}(V)$, but are ecanpuslated by \ref{eqn:equivalence}, below.
 
 More generally,
-we denote by $\calA_n(\mathcal{S}_0)$ the set of all $n$-dimensional systems equivalent to $\mathcal{S}_0$:
+we denote by $\calA_n(\Sys_0)$ the set of all $n$-dimensional systems equivalent to $\Sys_0$:
 \begin{equation} \label{eqn:equivalence}
   \begin{aligned}
-    \calA_n(\mathcal{S}_0) 
+    \calA_n(\Sys_0) 
       &= \left\{
@@
   \begin{align}
-    \mathcal{K} := \bigcup_{n=\min}^{\infty} \calA_n(\calA_0)
+    \mathcal{K} := \bigcup_{n=\min}^{\infty} \calA_n(\Sys_0)
   \end{align}
@@
 
-For instance, the phenotypic dynamics of a diploid organism are a function of its two haploid genomes $x$ and $y$: $P(x,y) = \left(x + y\right)/2$.
-Additionally, during meiosis, a diploid recombines its two genomes to form a new one: $\left(x \times y \right) \rightarrow z$; by randomly swapping matrix rows between two systems $\mathcal{S}$ and $\mathcal{S}'$.
-
-Even if two genomes $x$ and $y$ are both in $\mathcal{K}$, their diploid systems $P(x,y)$, and their recombinant systems $P(z,z)$, may not be in $\mathcal{K}$.
-This is because $\mathcal{K}$ is not mathematically required to be closed under averaging and recombination.
+For instance, diploid system coefficients are the average of two haploid genomes $\Sys$ and $\Sys'$: $\Sys_{\text{diploid}} = \left(\Sys + \Sys'\right)/2$.
+Additionally, during meiosis, a diploid recombines its two genomes to form a new one: $\left(\Sys \times \Sys' \right) \rightarrow \Sys''$; by randomly swapping matrix rows between two systems $\mathcal{S}$ and $\mathcal{S}'$.
+
+Even if two genomes $\Sys$ and $\Sys'$ are both in $\mathcal{K}$, their diploid and recombinant systems may not be in $\mathcal{K}$.
+This is because $\mathcal{K}$ is not required to be closed under averaging or recombination.
 
 If sexual recombination among systems drawn from $\mathcal{K}$ yields systems with divergent phenotypes, populations containing significant diversity in $\mathcal{K}$ will carry a large genetic load, and isolated populations may fail to produce hybrids with viable phenotypes.
-
-\paragraph{Sexual reproduction and speciation}
-A diploid organism contains two gene network copies: one from each parent.
+%
+%\paragraph{Sexual reproduction and speciation}
+%A diploid organism contains two gene network copies: one from each parent.
 As such, a diploid's system dynamics is computed by averaging the coefficients of both systems.
@@
 Haploid gene networks swap system matrix rows randomly between its own two networks.
-Therefore the phenotypic dynamics typical of a first generation ($F_{1}$) cross between two allopatric populations will be determined simply by averaging it's two parental genomes.
+%Therefore the phenotypic dynamics typical of a first generation ($F_{1}$) cross between two allopatric populations will be determined simply by averaging it's two parental genomes.
 However, in second generation hybrid ($F_2$) crosses, first new haploid systems will be formed by recombination -- in the process shuffling and combining regulatory coefficients from allopatric populations -- and then brought together to form a diploid.
-
+%
+%
   \begin{example}[Hybrid Incompatibility in an Oscillating Gene Network] \label{ex:hybrid_osc}
@@
 
-\begin{alignat*}{2}
-      A(2) &= \left[\begin{array}{cc} 2 & -1 \\[6pt] 5 & -2 \end{array}\right] &&\qquad A(2.01) = \left[\begin{array}{cc} 2 -\frac{1}{101} & -1 + \frac{1}{101} \\[6pt] 5 + \frac{1}{99} & -2 + \frac{1}{101} \end{array}\right] \\
-        A(2.1) &= \left[\begin{array}{cc} 2 - \frac{1}{11} & -1 + \frac{1}{11} \\[6pt] 5 + \frac{6}{55} & -2 + \frac{1}{11} \end{array}\right] &&\qquad A(2.5) = \left[\begin{array}{cc} 2 - \frac{1}{3} & -1 + \frac{1}{3} \\[6pt] 5 + \frac{2}{3} & -2 + \frac{1}{3} \end{array}\right]
-\end{alignat*}
+%\begin{alignat*}{2}
+%      A(2) &= \left[\begin{array}{cc} 2 & -1 \\[6pt] 5 & -2 \end{array}\right] &&\qquad A(2.01) = \left[\begin{array}{cc} 2 -\frac{1}{101} & -1 + \frac{1}{101} \\[6pt] 5 + \frac{1}{99} & -2 + \frac{1}{101} \end{array}\right] \\
+%        A(2.1) &= \left[\begin{array}{cc} 2 - \frac{1}{11} & -1 + \frac{1}{11} \\[6pt] 5 + \frac{6}{55} & -2 + \frac{1}{11} \end{array}\right] &&\qquad A(2.5) = \left[\begin{array}{cc} 2 - \frac{1}{3} & -1 + \frac{1}{3} \\[6pt] 5 + \frac{2}{3} & -2 + \frac{1}{3} \end{array}\right]
+%\end{alignat*}
 
@@
   \centering
-  \begin{tabular}{cc}
-    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{F1_comparison} & 
-    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{F2s_comparison2}
+  \begin{tabular}{ccc}
+    \includegraphics[width=0.25\textwidth, height=0.125\paperheight]{F1_comparison} & 
+    \includegraphics[width=0.25\textwidth, height=0.125\paperheight]{F2s_comparison2} &
+    \includegraphics[width=0.25\textwidth, height=0.125\paperheight]{figures/quartic_quadratic2}
   \end{tabular}
@@
 \end{figure}
-
-    \jss{$A(2)$ and $A(2.1)$ differ in regulatory interaction strengths by $5.09\%$.
-If intra-population regulatory variation is approximately $5\%$, then this level of divergence is expected after $\sqrt{\frac{t}{N_{e}} = 1}$ generations.}
 \end{example}
@@
 \section*{Systems drift and the accumulation of incompatibilities}
+Thus far we have shown that many distinct molecular mechanisms can realize identical phenotypes
+and that these mechanisms may fail to produce viable hybrids.
+Here we discuss the rate at which evolution shifts molecular mechanisms
+and ask whether this process is fast enough to be a significant driver of speciation.
+To do so, we explore a general quantitative genetic model in which a population drifts stochastically
+near a set of equivalent and optimal systems. We work with a population of effective size $N_e$.
 
@@
 as recombination, mutation, and demographic noise create new alleles and shift allele frequencies.
-How much variation do we expect to find within a population?
-Is this range limited by available variation or kept in check by selection?
-How fast will a population explore the space of equivalent networks?
-In this section we explore informally a general model for this situation,
-in which a population drifts stochastically near a set of equivalent, optimal systems.
-We work with a population of effective size $N_e$.
-
-Suppose that a set $x$ of coefficients that determine a system (this is $A$ above),
+%How much variation do we expect to find within a population?
+%Is this range limited by available variation or kept in check by selection?
+%How fast will a population explore the space of equivalent networks?
+%In this section we explore informally a general model for this situation,
+%in which a population drifts stochastically near a set of equivalent, optimal systems.
+%We work with a population of effective size $N_e$.
+\js{commented out questions here b/c we don't answer them directly -- what do you think?}
+
+Suppose that a set $x$ of coefficients that determine a system (this is $\Sys$ above),
 produce a phenotype $\Phi(x)$ (the time course of $\phi(t)$).
 There is an optimal phenotype $\optph$,
-and a set $\optx$ of ``optimal'' coefficients that produce this phenotype ($\calA_n$ above).
+and a set $\optx$ of ``optimal'' coefficients that produce this phenotype.
 Fitness depends on distance to the optimal phenotype --
@@
 \paragraph{Offspring.}
+\js{Not sure if this section is redundant, since reproduction is described above for the LTI model.}
 Individuals are diploid; we assume that each haploid genome determines a set of coefficients,
@@
 then since subsequent generations resample from this diversity,
-the population mean coefficient will move a random distance of size $\sigma/\sqrt{N_e}$ per generation, \js{since it moves at rate $\sigma \sqrt{T/N_e}$ isn't saying  $\sigma/\sqrt{N_e}$ \emph{per generation} not that clear?}
+the population mean coefficient will move a random distance of size $\sigma/\sqrt{N_e}$ per generation, 
+\js{since it moves at rate $\sigma \sqrt{T/N_e}$ isn't saying  $\sigma/\sqrt{N_e}$ \emph{per generation} not that clear?}
 simply because this is the the standard deviation of the mean of a random sample \citep{lande_drift}.
@@
 \paragraph{Hybridization}
-\plr{I put F1s and F2s here, and $\Incompat$, but with less explanation than below. Merge somehow.} \js{In progress, see paragraph above.}
+\plr{I put F1s and F2s here, and $\Incompat$, but with less explanation than below. Merge somehow.} 
 By the arguments above, the means of two allopatric populations separated for $T$ generations
@@
 and so will have mean system coefficients at the midpoint between their means,
-and variance equal to $\sigma$. \js{I think you mean $\sigma^{2}$?}
+and variance equal to $\sigma$. 
+\js{I think you mean $\sigma^{2}$?}
 Each $F_2$ hybrid will be homozygous for one parental allele on averge at half of the loci in the genome,
@@
 as before,
-but variance equal to $\sigma^2 + z^2/2$, where $z$ is the distance between the parental populations. \js{Can you clarify why $z^{2}/2$? Is this because it's averaging two distributions?}
+but variance equal to $\sigma^2 + z^2/2$, where $z$ is the distance between the parental populations. 
+\js{Can you clarify why $z^{2}/2$? Is this because it's averaging two distributions?}
 These are depicted in figure \ref{fig:conceptual_fig}.
@@
 
-  \jss{ Assume that selection acts directly against organisms whose regulatory coefficients are not on the optimum manifold.
-Further assume that selection constrains this diversity to be, on average, a distance of $m$ from optimum.
-We would expect $\sigma_S = m$.
-Furtheremore, two individuals within a population may both be on the optimum manifold, but at different points, on average of $\sigma_N$.
-Since individuals a distance of $\sigma_N$ apart will produce $F_2$s off of the manifold of $\mu = \frac{\sigma_N}{2}\sin2\theta$, recombination load may constrain $\sigma_N$ such that $\mu \leq m = \sigma_S$ (where $\theta$ is the slope of the manifold -- as $\theta$ approaches $0$ or $1$, $\sigma_N \rightarrow \infty$.).
-We can estimate $\sigma_N = \frac{2\sigma_S}{\sin2\theta}$.
-  }
-%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Parameter estimates for real systems}
-
-To translate these results into real predictions, we need to know
-the strength of stabilizing selection on the phenotype,
-and the amount (and structure) of heritable variation in the genotype.
-These are known at best only roughly \citep{felsenstein1988phylogenies},
-so we aim for order-of-magnitude estimates.
-
-%% amount of heritable variation
-We quantify (roughly) the amount of heritable variation
-by $\sigma^2$, the genetic variance present in a population in a typical entry of $A$.
-The coefficient $A_{ij}$ measures how much the rate of net production of $i$ changes
-per change in concentration of $j$.
-It is generally thought that regulatory sequence change contributes much more to inter- and intraspecific variation
-than does coding sequence change affecting molecular structure \citep{schmidt2010vertebrate}.
-In the context of transcription factor networks this may be affected 
-not only by the binding strength of molecule $j$ to the promoter region of gene $i$
-but also the effects of other transcription factors (e.g., cooperativity)
-and local chromatin accessibility \citep{stefflova2013cooperativity}.
-For this reason, 
-the mutational target size for variation in $A_{ij}$ may be much larger than the dozens of base pairs
-typically implicated in the handful of binding sites for transcription factor $j$ of a typical promoter region,
-and single variants may affect many entries of $A$ simultaneously.
-% (Recall that although these are best modeled through nonlinear terms,
-% by linearizing we essentially consider first-order effects.)
-On the other hand, a diverse set of buffering mechanisms are thought to contribute to phenotypic stability
-in the presence of substantial molecular noise \citep{canalization,buffering},
-suggesting that substantial variation in the micro-scale dynamics we consider here
-may be necessary to produce relevant phenotypic effects downstream.
-\plr{replace "micro-scale" with sthg else or discuss earlier}
-
-%% empirical data on sigma
-The amount and structure of this standing variation is established over long time scales
-by many factors, including
-mutation-selection balance, 
-shifts in the phenotypic optimum,
-and/or spatial variation in the optimum \citep{hansen1996translating}.
-Quantitative genetics models of mutation-selection balance 
-predict precise levels and structure of standing variation \citep{kimura_mutsel,lande_mutsel,lande1981models},
-but it is unclear how well these predictions match reality \citep{johnson_barton}
-and how much they are expected to change over time \citep{arnold_changing_G}.
-However, empirical work allows us to estimate at least the rough magnitude of variation.
-Differences in $A_{ij}$ due to a sequence change are hard to measure,
-but variation in both transcription factor binding site occupancy
-and expression levels (e.g., cis-eQTL) have been measured in various systems.
-However, variation in binding site occupancy may overestimate variation in $A$,
-since it does not capture buffering effects (if for instance only one site of many needs be occupied for transcription to begin),
-and variation in expression levels measures changes in steady-state concentration (our $\kappa_i$) rather than the \emph{rate} of change.
-Nonetheless,
-\citet{kasowski2010variation} found differential occupancy in 7.5\%
-of binding sites of a transcription factor (p65) between human individuals.
-\citet{verlaan2009targeted} showed that cis-regulatory variation
-accounts for around 2--6\% of expression variation in human blood-derived primary cells,
-while \citep{lappalainen2013transcriptome} found that human population variation 
-explained about 3\% of expression variation.
-\plr{Get some data from at least one other species in here!}
-Taken together, this suggests that variation in the entries of $A$
-may be on the order of 1\% between individuals of a population --
-doubtless varying substantially between species and between genes.
-
-%% argument for strength of selection
-It seems certain that selection in most species is not so strong that intra-population variation
-is strongly deleterious,
-so that if $\beta$ is the typical scale on which selection acts,
-then $\beta > \sigma$.  
-\plr{do we call this $\beta$ or $u$?} \js{$\beta$, $u$ is input.}
-However,
-a range of studies \plr{find them} have found evidence for weak stabilizing selection
-on regulatory SNPs and cis-eQTL.
-For instance, 
-\citet{maria_and_sergey} \plr{(others?)} found evidence that large-effect regulatory mutations are weakly selected against in Drosophila.
-This suggests that 
-the strength of selection on phenotype is sufficient to weakly constrain regulatory variation,
-so that perhaps $\sigma$ and $\beta$ are relatively close.
-This is as would be expected if available variation is held in check by mutation--selection balance
-rather than genetic drift.
-A conservative estimate would be that $\beta = 5 \sigma$;
-taking $\sigma=.01$ as above, this suggests that changes in phenotype of 5\% are sufficient
-to effect a noticeable drop in fitness.
-\plr{BUT $\sigma$ IS VARIATION IN $A$ NOT PHENOTYPE}
-
-We have guessed that within a population, 
-the entries of $A$ vary by around 1\%, 
-at least for networks whose function is strongly constrained.
-
-\paragraph{The rate of speciation under neutrality in allopatric populations}
-
+\paragraph{Speciation rates under neutrality}
+%
 \plr{separate out here (or above) the difference in genotype of hybrids from the optimum set, the difference in phenotype, and the difference in fitness.}
-
 %In general terms, if fitness $\fit$ is some function $g$ of phenotypic divergence from an optimal scaled by the strength of selection $\beta$,
@@
 %and that $F_{2}$ fitness drops rapidly -- at approximately rate $\sqrt{\frac{T}{N_{e}}}$. ARGUMENT FOR WHY $c_{2} \sigma \sim \beta$ HERE. 
-
-$F_2$s between allopatric populations will suffer an average fitness drop relative to their parents of,
+%
+For cases where system structures and $\gamma$ can be inferred,
+specific speciation rates can be estimated.
+For other situations we offer a back of the envelope calculation.
+
+Consider a population that is distributed on $\mathcal{X}$ such that the magnitude of phenotypically
+divergent $\sigma_S$ and neutral $\sigma_N$ diversity are constrained by selection;
+$\sigma_S$ directly and $\sigma_N$ indirectly via recombination load.
+Two reproductively isolated populations will, on average, be a distance of $2 \sigma_N \sqrt{T/N_e}$.
+If we assume that $\sigma_N$ is constrained by a function of $\sigma_S$,
+neutral movement along $\mathcal{X}$ pushes hybrids off of the manifold by $\sigma_S$ per $\sigma_N$,
+Lastly, if fitness is Gaussian: $\fit_2 = \exp\left(- d(x, x')^2/\beta^2\right)$,
+and we assume that the strength of selection is only several multiples of the segregating divergent diversity,
+such that $\sigma_S = c \beta$, we can simplify the expression estimating $F_2$ hybrid fitness drops
+to be,
   \begin{align}
-    \exp\left(-\left( \frac{2\sigma_{N}}{\beta}\sqrt{\frac{T}{N_e}} \right)^{2} \right) .
+    \exp\left(-c^2 \frac{T}{N_e} \right)
   \end{align}
-
-If the strength of selection is only a few multiples $\xi_1$ of the standing phenotypic diversity within a population $\sigma_S$, such that $\beta = \xi_1 \sigma_S$, and recombination load constrains the standing phenotype-neutral diversity $\sigma_N$ to be only a few multiples $\xi_2$ of $\sigma_S$, then $F_2$ fitness drop will be,
-  \begin{align}
-    \exp\left(- \xi^{2} \frac{T}{N_e} \right)
-  \end{align}
-  where $\xi := 2\xi_2/\xi_1$.
-  This suggests that if a population contains as much phenotypic diversity as allowed by selection and if $\sigma_N$ is on the order of $\sigma_S$, reproductively isolated populations can speciate rapidly -- possibly on timescales shorter than $N_{e}$ generations.
+%
+%$F_2$s between allopatric populations will suffer an average fitness drop relative to their parents of,
+%  \begin{align}
+%    \exp\left(-\left( \frac{2\sigma_{N}}{\beta}\sqrt{\frac{T}{N_e}} \right)^{2} \right) .
+%  \end{align}
+
+%If the strength of selection is only a few multiples $\xi_1$ of the standing phenotypic diversity within a population $\sigma_S$, such that $\beta = \xi_1 \sigma_S$, and recombination load constrains the standing phenotype-neutral diversity $\sigma_N$ to be only a few multiples $\xi_2$ of $\sigma_S$, then $F_2$ fitness drop will be,
+%  \begin{align}
+%    \exp\left(- \xi^{2} \frac{T}{N_e} \right)
+%  \end{align}
+%  where $\xi := 2\xi_2/\xi_1$.
+  This suggests that, as an upper bound, if a population contains as much phenotypic diversity as allowed by selection, reproductively isolated populations can speciate rapidly -- possibly on timescales shorter than $N_{e}$ generations.
+%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection*{Parameter estimates for real systems}
+\js{I think we should move this section?}
+To translate these results into real predictions, we need to know
+the strength of stabilizing selection on the phenotype,
+and the amount (and structure) of heritable variation in the genotype.
+These are known at best only roughly \citep{felsenstein1988phylogenies},
+so we aim for order-of-magnitude estimates.
+
+%% amount of heritable variation
+We quantify (roughly) the amount of heritable variation
+by $\sigma^2$, the genetic variance present in a population in a typical entry of $A$.
+The coefficient $A_{ij}$ measures how much the rate of net production of $i$ changes
+per change in concentration of $j$.
+It is generally thought that regulatory sequence change contributes much more to inter- and intraspecific variation
+than does coding sequence change affecting molecular structure \citep{schmidt2010vertebrate}.
+In the context of transcription factor networks this may be affected 
+not only by the binding strength of molecule $j$ to the promoter region of gene $i$
+but also the effects of other transcription factors (e.g., cooperativity)
+and local chromatin accessibility \citep{stefflova2013cooperativity}.
+For this reason, 
+the mutational target size for variation in $A_{ij}$ may be much larger than the dozens of base pairs
+typically implicated in the handful of binding sites for transcription factor $j$ of a typical promoter region,
+and single variants may affect many entries of $A$ simultaneously.
+% (Recall that although these are best modeled through nonlinear terms,
+% by linearizing we essentially consider first-order effects.)
+On the other hand, a diverse set of buffering mechanisms are thought to contribute to phenotypic stability
+in the presence of substantial molecular noise \citep{canalization,buffering},
+suggesting that substantial variation in the micro-scale dynamics we consider here
+may be necessary to produce relevant phenotypic effects downstream.
+\plr{replace "micro-scale" with sthg else or discuss earlier}
+
+%% empirical data on sigma
+The amount and structure of this standing variation is established over long time scales
+by many factors, including
+mutation-selection balance, 
+shifts in the phenotypic optimum,
+and/or spatial variation in the optimum \citep{hansen1996translating}.
+Quantitative genetics models of mutation-selection balance 
+predict precise levels and structure of standing variation \citep{kimura_mutsel,lande_mutsel,lande1981models},
+but it is unclear how well these predictions match reality \citep{johnson_barton}
+and how much they are expected to change over time \citep{arnold_changing_G}.
+However, empirical work allows us to estimate at least the rough magnitude of variation.
+Differences in $A_{ij}$ due to a sequence change are hard to measure,
+but variation in both transcription factor binding site occupancy
+and expression levels (e.g., cis-eQTL) have been measured in various systems.
+However, variation in binding site occupancy may overestimate variation in $A$,
+since it does not capture buffering effects (if for instance only one site of many needs be occupied for transcription to begin),
+and variation in expression levels measures changes in steady-state concentration (our $\kappa_i$) rather than the \emph{rate} of change.
+Nonetheless,
+\citet{kasowski2010variation} found differential occupancy in 7.5\%
+of binding sites of a transcription factor (p65) between human individuals.
+\citet{verlaan2009targeted} showed that cis-regulatory variation
+accounts for around 2--6\% of expression variation in human blood-derived primary cells,
+while \citep{lappalainen2013transcriptome} found that human population variation 
+explained about 3\% of expression variation.
+\plr{Get some data from at least one other species in here!}
+Taken together, this suggests that variation in the entries of $A$
+may be on the order of 1\% between individuals of a population --
+doubtless varying substantially between species and between genes.
+
+%% argument for strength of selection
+It seems certain that selection in most species is not so strong that intra-population variation
+is strongly deleterious,
+so that if $\beta$ is the typical scale on which selection acts,
+then $\beta > \sigma$.  
+\plr{do we call this $\beta$ or $u$?} \js{$\beta$, $u$ is input.}
+However,
+a range of studies \plr{find them} have found evidence for weak stabilizing selection
+on regulatory SNPs and cis-eQTL.
+For instance, 
+\citet{maria_and_sergey} \plr{(others?)} found evidence that large-effect regulatory mutations are weakly selected against in Drosophila.
+This suggests that 
+the strength of selection on phenotype is sufficient to weakly constrain regulatory variation,
+so that perhaps $\sigma$ and $\beta$ are relatively close.
+This is as would be expected if available variation is held in check by mutation--selection balance
+rather than genetic drift.
+A conservative estimate would be that $\beta = 5 \sigma$;
+taking $\sigma=.01$ as above, this suggests that changes in phenotype of 5\% are sufficient
+to effect a noticeable drop in fitness.
+\plr{BUT $\sigma$ IS VARIATION IN $A$ NOT PHENOTYPE}
+
+We have guessed that within a population, 
+the entries of $A$ vary by around 1\%, 
+at least for networks whose function is strongly constrained.
 
@@
 \section*{Discussion}
-
 %
@@
 %
-
 The complexity of biological systems has limited our understanding of their function and evolution.
@@
 by a simple change of coordinates in the minimal configuration, or more generally by applying the Kalman decomposition in higher dimensions. 
-Further, we emphasize that evolution proceeds through this high dimensional space as stochastic coordinate transformation, constrained by sexual reproduction and selection. 
+%Further, we emphasize that evolution proceeds through this high dimensional space as stochastic coordinate transformation, constrained by sexual reproduction and selection. 
 This set is explored over evolutionary time when phenotype is conserved, and can lead to a diverse set of consequences, 
@@
 Because mechanistically distinct yet phenotypically equivalent biological systems can fail to produce viable hybrids, we predict allopatric populations to accumulate genetic incompatibilities 
-at a rate on the order of $N_{e}$ under reasonable population genetic parameter estimates. 
+at a rate on the order of $N_{e}$ if intrapopulation regulatory variation is constrained by selection.
 Additionally we see second-generation hybrid fitness plummet much faster than that of first-generation hybrids. 
 \plr{refer to Turelli here}
-This is a consequence of combining our mechanistic model with a quantitative genetic one: 
-we observe that $F_{1}$ phenotypes diverge quartically, and $F_{2}$ phenotypes quadratically, with evolutionary time. 
-\plr{awkward phrasing}
+%This is a consequence of combining our mechanistic model with a quantitative genetic one: 
+%we observe that $F_{1}$ phenotypes diverge quartically, and $F_{2}$ phenotypes quadratically, with evolutionary time. 
+%\plr{awkward phrasing}
 This result is also consistent with Haldane's rule; that if only one hybrid sex is inviable or sterile it is likely the heterogametic sex. 
@@
 
-We also suggest that gene networks may not always use their components parsimoniously 
-as network size tends to ratchet up in the absence of strong selection against extra parts. 
-\plr{probably shouldn't claim this is "unexplored"} \js{I meant unexplored within this paper.Changed wording. Does that work?}
-Although we leave this question unexplored, this phenomena may lead to insights on evolvability and developmental innovation. 
+%We also suggest that gene networks may not always use their components parsimoniously 
+%as network size tends to ratchet up in the absence of strong selection against extra parts. 
+%\plr{probably shouldn't claim this is "unexplored"} \js{I meant unexplored within this paper.Changed wording. Does that work?}
+%Although we leave this question unexplored, this phenomena may lead to insights on evolvability and developmental innovation. 
 Lastly, we show that hybrid gene networks break down as function of genetic distance, and may, in part, explain broad patterns of reproductive isolation among diverse phyla \citep{roux2016shedding}.
 
-\plr{this paragraph isn't very clear} \js{Should I delete it? I guess it's sort of just a vague ``yeah we know this method has shortcomings''.}
-As Richard Levins opined, models in population biology face a trade-off among precision, realism, and generality \citep{levins1966strategy}.
+%\plr{this paragraph isn't very clear} \js{Should I delete it? I guess it's sort of just a vague ``yeah we know this method has shortcomings''.}
+%As Richard Levins opined, models in population biology face a trade-off among precision, realism, and generality \citep{levins1966strategy}.
 As Levins expects, any tractable and general model, such as the present one under discussion, will have limitations.
@@
 
-This theoretical framework can easily be applied to other interesting questions in evolutionary biology not tackled presently: such as the evolution of linkage, the necessity of network complexity (does evolution tend towards Rube Goldberg or parsimonious network organization?), evolvability, structure/function inference, and intra-population context dependency of mutational effects, as well as many others.
+%This theoretical framework can easily be applied to other interesting questions in evolutionary biology not tackled presently: such as the evolution of linkage, the necessity of network complexity (does evolution tend towards Rube Goldberg or parsimonious network organization?), evolvability, structure/function inference, and intra-population context dependency of mutational effects, as well as many others.
 

### 2017-09-11 kalman_walk_paper.tex
edited examples

@@
  Movement in this direction is ongoing, as researchers have begun to study 
-the evolution of both abstract \citep{wagner1994evolution, wagner1996does,  siegal2002waddington, bergman2003evolutionary, draghi2015robustness} and empirically inspired computational and mathematical models of gene regulatory networks (GRNs) \citep{mjolsness1991connectionist, jaeger2004dynamic, maria1, vitaly1, vitaly2, crombach2016gap, wotton2015quantitative, chertkova2017insilico}.
+the evolution of both abstract \citep{wagner1994evolution, wagner1996does,  siegal2002waddington, bergman2003evolutionary, draghi2015robustness} and empirically inspired computational and mathematical models of gene regulatory networks \citep{mjolsness1991connectionist, jaeger2004dynamic, maria1, vitaly1, vitaly2, crombach2016gap, wotton2015quantitative, chertkova2017insilico}.
 If we allow the reasonable assumption that the genotype-phenotype map can be represented as a system of differential equations, we can immediately discuss its evolution and function in a much more mechanistic, yet general, manner.
@@
 \paragraph{Systems theory}
-Organisms' phenotypes are constructred by gene by gene by environment interactions.
+Organisms' phenotypes are constructred by gene by gene and gene by environment interactions.
 Here we simply define the \emph{phenotype} to be the temporal molecular dynamics directly under natural selection.
@@
 Furthermore, suppose that only the dynamics of gene-1 are consequential to the cell cycle (perhaps the amount of gene-1 activates another downstream gene network).
-Lastly suppose that the production of both genes is stimulated by an impulse of a molecule present immediately after division.
-
-    If the rate each of these genes is expressed is a linear function of their concentrations, the dynamics of the system are given by
-    \begin{align*}
-      \dot{\kappa_{1}}(t) &= \kappa_{2}(t) + u(t) \\
-      \dot{\kappa_{2}}(t) &= - \kappa_{1}(t) + u(t)
-    \end{align*}
-    where $\dot{\kappa}$ denotes the time derivative.
-The initial conditions $\kappa_{1}(0)$ and $\kappa_{2}(0)$, and the input $u(t)$ then determine the concentrations through time.
-If we record the regulatory coefficients in the matrix
-    \begin{align*}
-      A &= \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} ,
-    \end{align*}
-    and define the column vector $B = \begin{bmatrix} 1 & 1 \end{bmatrix}^{T}$, then in matrix notation the dynamics are
-      \begin{align*}
-      \dot{\kappa} &= A \kappa(t) + B u(t) .
-      \end{align*}
-      Since only the dynamics of gene-1 are directly relevant to biological function, the dynamics of interest are given by
-      \begin{align*}
-      \phi(t) &= C \kappa (t)
-      \end{align*}
-      where the row vector $C$ is defined as $C = \begin{bmatrix} 1 & 0 \end{bmatrix}$. 
+Lastly suppose that the production of both genes is equally stimulated by an impulse of a molecule present immediately after division.
+
+    If the rate each of these genes is expressed is a linear function of their concentrations, the dynamics of the system is given by
+    \begin{alignat*}{2}
+      %\left\{ \begin{array}{ll} 
+      \dot{\kappa_{1}}(t) &= &&\kappa_{2}(t) + u(t) \\
+        \dot{\kappa_{2}}(t) &= - &&\kappa_{1}(t) + u(t) \\
+        \phi(t) &= &&\kappa_{1}(t)   .
+      %\end{array} \right.
+    \end{alignat*}
+Equivalently, in matrix form the system regulatory coefficients are given as,
+    $A \!=\!
+\left[\begin{smallmatrix} 0 & 1 \\ -1 & 0 \end{smallmatrix}\right] ,\ B \!
+=\!
+\left[\begin{smallmatrix} 1 \\ 1 \end{smallmatrix}\right] , \ C \!
+=\!
+\left[\begin{smallmatrix} 1 & 0 \end{smallmatrix}\right]$
   %  \begin{align*}
@@
 
-      Since the input is simply an impulse, its phenotype is equivalent to its impulse response
+      Since the input is simply an impulse, its phenotype is equivalent to its impulse response,
       \begin{align*}
@@
                 \node[dashed] (B) at (3,0) {$\kappa_{2}$};
-                \node[shape=rectangle] (U) at (1.5,2) {input};
-                \node[shape=rectangle] (y) at (1.5,-2) {output};
+                \node[shape=rectangle] (U) at (1.5,1.5) {input};
+                \node[shape=rectangle] (y) at (1.5,-1.5) {output};
             \end{scope}
@@
             \end{tikzpicture} &
-       % \end{center}
-      %  \caption{
-      %      Diagram of Example \ref{ex:osc} in the text.
-      %      \plr{explain what arrows mean if nec}
-      %      \label{fig:oscillator_diagram}}
-   % \end{figure}
-   % \begin{figure}[H]
-    %  \centering
-      \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{osc_impulse}
+      \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{osc_impulse}
    \end{tabular}
@@
  \begin{example}[All Phenotypically Equivalent Cell Cycle Control Networks] \label{ex:all_osc}
-
-%\plr{this is same as above, so reference instead of writing out again?}
-%    Let
-%    \begin{align*}
-%      A(0) = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix}, \qquad B = \begin{bmatrix} 1 & 1 \end{bmatrix}^{T}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix},
-%    \end{align*}
-%    with $V(\tau)$ preserving both $B$ and $C$, then
     The set of all two-gene regulatory networks phenotypically equivalent to the cell cycle control network in \ref{ex:oscillator}, where only $A$ can vary, is given by
-    \begin{align*}
-      A(\tau) &= \begin{bmatrix} 1 & 0 \\ \tau & 1-\tau \end{bmatrix} \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 0 \\ \frac{\tau}{\tau-1} & \frac{-1}{\tau-1} \end{bmatrix}  \qquad \forall \ \tau \neq 1 %\\
-     %   B &= \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix} \\
-       %   h(t) &= \sin(t) + \cos(t) \qquad \forall \ \tau \neq 1
+    %\begin{align*}
+    %  A(\tau) &= \begin{bmatrix} 1 & 0 \\ \tau & 1-\tau \end{bmatrix} \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 0 \\ \frac{\tau}{\tau-1} & \frac{-1}{\tau-1} \end{bmatrix}  \qquad \forall \ \tau \neq 1 %\\
+    %\end{align*}
+    \begin{align*}
+      A(\tau) &= \frac{1}{\tau-1} \begin{bmatrix} \tau & -1 \\ 2 \tau(\tau - 1) + 1 &  -\tau \end{bmatrix} \ \forall \ \tau \neq 1
     \end{align*}
@@
     \centering
-    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{osc_A0_A2_both_compare}
-    \caption{Gene-1 (blue) and gene-2 (orange) dynamics for $A(0)$ (top) and $A(2)$ (bottom).
+   % \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{osc_A0_A2_both_compare}
+    %\caption{Gene-1 (blue) and gene-2 (orange) dynamics for $A(0)$ (top) and $A(2)$ (bottom).
 Both (top and bottom) gene-1 dynamics are given by $ \kappa_{1} = \sin(t) + \cos(t)$, and gene-2 by $\kappa_{2} = \cos(t) - \sin(t)$ (top) and $\kappa'_{2} = \cos(t) + 3 \sin(t)$ (bottom).}
+    \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{figures/osc_kryp_compare}
+    \caption{(Gene-1 dynamics (blue) for both systems $A(0)$ and $A(2)$ are identical, however, $A(0)$ gene-2 dynamics (orange) differ from $A(2)$ (green).
+Both gene-1 dynamics are given by $\kappa_{1} = \sin(t) + \cos(t)$, and gene-2 by $\kappa_{2} = \cos(t) - \sin(t)$ ($A(0)$) and $\kappa'_{2} = \cos(t) + 3 \sin(t)$ ($A(2)$).}
   \end{figure}

### 2017-09-11 kalman_walk_paper.tex
edited examples

@@
 If sexual recombination among systems drawn from $\mathcal{K}$ yields systems with divergent phenotypes, populations containing significant diversity in $\mathcal{K}$ will carry a large genetic load, and isolated populations may fail to produce hybrids with viable phenotypes.
+
+\paragraph*{Hybrid incompatibility}
+For example if the two phenotypically equivalent systems from Example \ref{ex:all_osc} hybridize, the offspring's phenotype will fail to oscillate completely. 
+  \begin{figure}[H]
+    \centering
+    \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{expF1}
+    \caption{$F_1$ hybrid (orange) and parental (blue) phenotypic oscillator dynamics for an $A(0)$ by $A(2)$ cross.
+The hybrid fails to oscillate and exhibits qualitatively different dynamics ($\phi(t)_{F_0}\!
+= \!
+\sin(t) \!
++ \!
+\cos(t)$, however $\phi(t)_{F_1} \!
+= \!
+e^t$).
+    }
+  \end{figure}
 %

### 2017-09-11 kalman_walk_paper.tex
edited examples

--- 2017-09-11 kalman_walk_paper.tex
@@
 \newcommand{\Sys}{\mathcal{S}}
+\newcommand{\allS}{\mathcal{K}(h)}
 
@@
       \begin{align*}
-        \phi(t) = h(t) = \sin(t) + \cos(t)  .
+        \phi(t) = h(t) = \sin t + \cos t  .
       \end{align*}
@@
 \end{equation}
-  \begin{align}
-    \mathcal{K} := \bigcup_{n=\min}^{\infty} \calA_n(\Sys_0)
-  \end{align}
 Equivalence of the two characterizations follows from the Cayley-Hamilton theorem.
@@
 Usually, the dimension $n$ and the reference system $A_0$ is implicit and we write only $\calA$.
-
-Two systems, even if in different dimensions, will have identical phenotypes if they are in $\mathcal{K}$.
-Further, there is no unique triple $(A,B,C)$ for the mapping $u \mapsto \phi$.
-This implies that the gene regulatory network architecture per phenotype/environment pair is never unique -- that is \emph{all} gene regulatory networks can drift/rewire.
-This set can be precisley defined and completely parameterized using the \emph{Kalman decomposition} \ref{apx:kalman}.
-
-A large, adapted population will move around in $\mathcal{K}$ given environmental and selective stasis.
+Further, we refer to the set of all linear systems with impulse response $h(t)$, regardless of dimension, to be $\allS$,
+  \begin{align}
+    \allS := \bigcup_{n=\min}^{\infty} \calA_n(\Sys_0)
+  \end{align}
+Two systems, even if in different dimensions, will have identical phenotypes if they are in $\allS$.
+Further, there is no unique triple $(A,B,C)$ for the mapping $h: u \mapsto \phi$.
+This implies that the gene regulatory network architecture per phenotype/environment pair is never unique -- that is under these assumptions, \emph{all} gene regulatory networks can drift/rewire.
+This set can be precisley defined and completely parameterized using the \emph{Kalman decomposition}\ref{apx:kalman}.
+
+A large, adapted population will move around in $\allS$ under environmental and selective stasis.
+In the minimal dimension neutral evolution proceeds as a stochastic coordinate transformation on the elements in $\Sys_{\min}$.
 Even if there is a conservation of phenotype through evolutionary time, the molecular mechanism underpinning the phenotype is not constrained and likely will not be conserved.
@@
     \begin{align*}
-      A(\tau) &= \frac{1}{\tau-1} \begin{bmatrix} \tau & -1 \\ 2 \tau(\tau - 1) + 1 &  -\tau \end{bmatrix} \ \forall \ \tau \neq 1
+      A(\tau) &= \frac{1}{\tau-1} \begin{bmatrix} \tau & -1 \\ 2 \tau(\tau - 1) + 1 &  -\tau \end{bmatrix} \ \text{for} \ \tau \neq 1
     \end{align*}
@@
     \caption{(Gene-1 dynamics (blue) for both systems $A(0)$ and $A(2)$ are identical, however, $A(0)$ gene-2 dynamics (orange) differ from $A(2)$ (green).
-Both gene-1 dynamics are given by $\kappa_{1} = \sin(t) + \cos(t)$, and gene-2 by $\kappa_{2} = \cos(t) - \sin(t)$ ($A(0)$) and $\kappa'_{2} = \cos(t) + 3 \sin(t)$ ($A(2)$).}
+Both gene-1 dynamics are given by $\kappa_{1} = \sin t + \cos t$, and gene-2 by $\kappa_{2} = \cos t - \sin t$ ($A(0)$) and $\kappa'_{2} = \cos t + 3 \sin t$ ($A(2)$).}
   \end{figure}
@@
 
-For instance, diploid system coefficients are the average of two haploid genomes $\Sys$ and $\Sys'$: $\Sys_{\text{diploid}} = \left(\Sys + \Sys'\right)/2$.
-Additionally, during meiosis, a diploid recombines its two genomes to form a new one: $\left(\Sys \times \Sys' \right) \rightarrow \Sys''$; by randomly swapping matrix rows between two systems $\mathcal{S}$ and $\mathcal{S}'$.
-
-Even if two genomes $\Sys$ and $\Sys'$ are both in $\mathcal{K}$, their diploid and recombinant systems may not be in $\mathcal{K}$.
-This is because $\mathcal{K}$ is not required to be closed under averaging or recombination.
-
-If sexual recombination among systems drawn from $\mathcal{K}$ yields systems with divergent phenotypes, populations containing significant diversity in $\mathcal{K}$ will carry a large genetic load, and isolated populations may fail to produce hybrids with viable phenotypes.
+For instance, diploid system coefficients are derived by taking the average of two haploid genomes $\Sys$ and $\Sys'$: $\Sys_{\text{diploid}} \!
+= \!
+\left(\Sys \!
++ \!
+\Sys'\right)\!/2$.
+During meiosis a diploid recombines its two genomes to form a new one (a gamete): $\left(\Sys \!
+\times \!
+\Sys' \right) \rightarrow \Sys''$; by randomly swapping matrix rows between two systems $\mathcal{S}$ and $\mathcal{S}'$.
+
+Even if two genomes ($\Sys$ and $\Sys'$) have the same phenotype, their diploid and recombinant systems may not. This is because $\allS$ is not required to be closed under averaging or recombination.
+
+If sexual recombination among systems drawn from $\allS$ yields systems with divergent phenotypes, populations containing significant diversity in $\allS$ can carry genetic load, and isolated populations may fail to produce hybrids with viable phenotypes.
+
+\paragraph*{Hybrid phenotypic breakdown between oscillators}
+For example, if the two phenotypically equivalent systems $A(0)$ and $A(2)$ from Example \ref{ex:all_osc} hybridize, their offspring's phenotype will fail to oscillate completely. 
+  \begin{figure}[H]
+    \centering
+    \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{expF1}
+    \caption{$F_1$ hybrid (orange) and parental (blue) phenotypic oscillator dynamics for an $A(0)$ by $A(2)$ cross.
+The hybrid $\left(A_{F_1}\!\!=\!\!\tfrac{A\!(\!0\!) \!
++ \!
+A\!(\!2\!)}{2}\right)$ fails to oscillate and exhibits qualitatively different dynamics ($\phi_{F_0}\!
+= \!
+\sin t \!
++ \!
+\cos t$, however $\phi_{F_1} \!
+= \!
+\exp t$).
+    }
+  \end{figure}
 %
@@
 Here we compare the phenotypes for F2 hybrids formed by crossing oscillators $A(2)$ with $A(2.01)$, $A(2.1)$, and $A(2.5)$ ($B$ and $C$ are the same as above). 
-Each $A$ is phenotypically identical ($\phi(t) = \sin(t) + \cos(t)$), however some of the hybrids exhibit markedly different dynamics.
+Each $A$ is phenotypically identical ($\phi(t) = \sin t + \cos t$), however some of the hybrids exhibit markedly different dynamics.
 
@@
 is equivalent to changing $B$ by $Y = QB$.
+
+\section{Sexual reproduction in linear systems}\label{apx:reproduction}
+
+  Recombination: gamete element $A''$ is formed by recombining (randomly swapping rows) between two parental, possibly distinct, system elements $A$ and $A'$ as follows,
+  \begin{align*}
+    A'' = MA + (I-M)A'
+  \end{align*}
+  where $M$ is a diagonal matrix with only $0$s and $1$s on the diagonal.
 

### 2017-09-13 kalman_walk_paper.tex
added an informal description of the kalman decomposition to the main text. Edited metabolic example and recombination section.

@@
 \newcommand{\Sys}{\mathcal{S}}
-\newcommand{\allS}{\mathcal{K}(h)}
+\newcommand{\allS}{\mathcal{K}}
+%\newcommand{\allS}{\mathcal{A}}
 
@@
 Any linear and minimal gene network 
--- minimal, informally meaning that the system's phenotype is achieved with the fewest possible genes $(n=\min)$
+-- minimal, informally meaning that the system's phenotype is achieved with the fewest possible genes ($n=\min$, see \ref{apx:kalman})
 -- has an identical impulse response and thus phenotype given identical inputs $u(t)$, up to a change of coordinates. 
@@
   \end{equation}
-Two gene networks, $\mathcal{S} = \{ A, B, C \}$, and $\bar{\mathcal{S}} = \{\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} \}$, 
+Two gene networks, $\mathcal{S} = ( A, B, C )$, and $\bar{\mathcal{S}} = (\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} )$, 
 have the same phenotype, for all possible inputs, if they are related by a change of coordinates. 
@@
 %\plr{why is this "evolutionary"? It is the thing that we're thinking about evolution of, but that doesn't make it evolutionary.}
-Therefore the set of all phenotypically identical (and minimal) systems can be parameterized as, $\mathcal{S}_{\min}(V) := \{VA_0V^{-1}, VB_0, C_0V \}$,
+Therefore the set of all phenotypically identical (and minimal) systems can be parameterized as, $\mathcal{S}_{\min}(V) = (VA_0V^{-1}, VB_0, C_0V )$,
   %  \begin{align} \label{eq:min_param}
@@
   \end{align}
+  \js{what do you like better for $\allS$?
+$\mathcal{A}$, $\mathcal{A}_{\infty}$, or $\mathcal{K}_{h(t)}$, or something else -- maybe a modification of $\Sys$?
+Using a script of ``A'' was more intuitive when we were leavng $B$ and $C$ invariant.}
 Two systems, even if in different dimensions, will have identical phenotypes if they are in $\allS$.
 Further, there is no unique triple $(A,B,C)$ for the mapping $h: u \mapsto \phi$.
-This implies that the gene regulatory network architecture per phenotype/environment pair is never unique -- that is under these assumptions, \emph{all} gene regulatory networks can drift/rewire.
-This set can be precisley defined and completely parameterized using the \emph{Kalman decomposition}\ref{apx:kalman}.
+This implies that the gene regulatory network architecture per phenotype/environment pair is never unique -- that is under these assumptions, \emph{all} gene regulatory networks can drift/rewire, and still conserve the phenotype.
+This set can be precisley defined and completely parameterized using the \emph{Kalman decomposition} (Appendix \ref{apx:kalman}).
+
+The Kalman decomposition decomposes a system into \emph{reachable} and \emph{observable} components -- that is coefficients that respond to an input and coefficients that produce an observable output.
+A minimal system's components are both reachable and observable.
+In this form, it is possible to add components that respond to an input, but do not influence the output, or components that, in principle, influence the output, but do not respond to any inputs.
+Any system with such components is, by definition, non-minimal, and any system can be modified to include any finite number of extra unreachable and/or unobservable components, without modifying its impulse response.
+Decomposing the system allows for the easy manipulation of ``superfluous''(but not necessarily expendable) components.
+After unreachable/unobservable coefficients are added, changing coordinates ``stirs'' the system coefficients such that the previously superfluous coefficients can become essential to system funciton.
+Whether or not evolution constrains biological systems to be minimal or nearly minimal is an open question. 
+\js{should the sentence of evolution of nearly minimal be in the paper? Or should we save this for our next paper?}
 
@@
    Despite the phenotypic equivalence of all instantiations of $A(\tau)$, the kryptotypes, vary as a function of $\tau$. 
-    Gene-1 dynamics (blue) are equivalent for network architectures $A(0)$ and $A(2)$, however the dynamics of gene-2 (orange) differ with $\tau$.
+    %Gene-1 dynamics (blue) are equivalent for network architectures $A(0)$ and $A(2)$, however the dynamics of gene-2 (orange) differ with $\tau$.
   \begin{figure}[H]
@@
     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{figures/osc_kryp_compare}
-    \caption{(Gene-1 dynamics (blue) for both systems $A(0)$ and $A(2)$ are identical, however, $A(0)$ gene-2 dynamics (orange) differ from $A(2)$ (green).
+    \caption{Gene-1 dynamics (blue) for both systems $A(0)$ and $A(2)$ are identical, however, $A(0)$ gene-2 dynamics (orange) differ from $A(2)$ (green).
 Both gene-1 dynamics are given by $\kappa_{1} = \sin t + \cos t$, and gene-2 by $\kappa_{2} = \cos t - \sin t$ ($A(0)$) and $\kappa'_{2} = \cos t + 3 \sin t$ ($A(2)$).}
@@
 \end{example}
-
-\paragraph{Kryptotypic variation and hybrid incompatibility}
+\begin{example}[Metabolic network] \label{ex:metabolic}
+Consider an organism that can metabolize two different sugars (present at logarithmic concentrations $u_{1}$ and $u_{2}$), 
+with two enzymes (at log concentrations $\phi_{1}$ and $\phi_{2}$).
+Further suppose that the second sugar is prefered,
+%and that the organism has a regulatory network that can deploy situation specific metabolic strategy.
+and that
+depending on $u_{2}$ and $u_{1}$ the organism will synthesize an appropriate $\phi_{1}$ and $\phi_{2}$. 
+Furthermore, assume this system contains at least two transcription factors, whose log concentrations are $\kappa_{1}$, $\kappa_{2}$, $\dots \kappa_{n}$.
+Minimally such a system may have the architecture,
+%\begin{align*}
+%    \mathcal{S}_{\min} = \left\{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 0 & -1 \\ 0 & 0 \end{bmatrix} \begin{bmatrix} \kappa_{1} \\ \kappa_{2} \end{bmatrix} (t) + \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \begin{bmatrix} u_{1} \\ u_{2} \end{bmatrix} (t) \\[11pt]
+%    \phi(t) &= \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \begin{bmatrix} \kappa_{1} \\ \kappa_{2} \end{bmatrix} (t) \end{array} \right.
+%\end{align*}
+  $\Sys_{\min}(U) = (U \left[ \begin{smallmatrix} 0 & -1 \\ 0 & 0 \end{smallmatrix} \right] U^{-1}, U \left[\begin{smallmatrix} 1 & 0 \\ 0 & 1 \end{smallmatrix}\right], \left[\begin{smallmatrix} 1 & 0 \\ 0 & 1 \end{smallmatrix}\right]U^{-1})$.
+%
+%The impulse response of the system is,
+%$h(t) = \left[\begin{smallmatrix} 1 & - t \\ 0 & 1 \end{smallmatrix}\right]$.
+    We can find alternative equivalent systems by changning coordinates ($U \rightarrow U'$) or, more generally, by applying the Kalman decomposition \ref{apx:kalman}.
+To illustrate that phenotypic invariance does not require dimensional invariance, we apply the Kalman decomposition to $\Sys_{\min}$ to find
+    %This may happen if a system is co-opted from another, or may be a consequence of gene duplication and deletion. 
+  $\Sys(\small X_{1-3},V) \! =  \! \left( \! \begin{array}{ll}
+     V \!
+\left[\begin{smallmatrix} X_{1} & X_{2} \\ 0 & A \end{smallmatrix}\right] \!
+V^{-1}, V \!
+\left[\begin{smallmatrix} X_{3} \\ B \end{smallmatrix}\right], \left[\begin{smallmatrix} 0 & C \end{smallmatrix}\right] \!
+V^{-1} \end{array} \!
+\right)
+$, both of which are in $\allS$
+%  \begin{align*}
+%    \Sys(X_{1-3},V) = \left( \begin{array}{ll}
+%     V \left[\begin{smallmatrix} X_{1} & X_{2} \\ 0 & A_{\min} \end{smallmatrix}\right] V^{-1}, V \left[\begin{smallmatrix} X_{3} \\ B_{\min} \end{smallmatrix}\right], \left[\begin{smallmatrix} 0 & C_{\min} \end{smallmatrix}\right] V^{-1} \end{array} \right)
+%    \end{align*}
+  ($V$ can be any $4$-dimensional and invertible matrix, $X_{1-3} \in \R^{2 \time 2}$, and $(A,B,C) \in \Sys_{\min}$).
+%So for example, some system can be wired as follows, and still be input-output equivalent to the minimal metabolic system $\mathcal{S}_{\min}$:
+% \begin{align*}
+%   A' &= \begin{bmatrix}
+%     1.6923  &  1.5385 &  -2.6154 &  -2 \\
+%     0.8462  &  0.7692 &  -1.3077 &  -1 \\
+%     1.2692  &  1.1538 &  -1.9615 &  -1.5 \\
+%     0.4231  &  0.3846 &  -0.6538 &  -0.5
+%   \end{bmatrix} \\
+%   B' &= \begin{bmatrix} 4 & 4 \\ 2 & 2 \\ 3 & 3 \\ 1 & 3 \end{bmatrix} \\
+%     C' &= \begin{bmatrix}  0.2692  &  0.1538  &  0.0385 &  -0.5 \\
+%     -0.4231 &  -0.3846 &   0.6538  &  0.5 \end{bmatrix}
+% \end{align*}
+%
+%Despite the present example consisting of a minimal $2 \times 2$ and a non-minimal $4 \times 4$ system, 
+%any $m$-dimensional system can be constructed using this method 
+%-- applying a change of coordinates to the Kalman decomposition 
+%-- to construct a mechanistically different system with identical phenotypic dynamics. 
+%
+%When applying the Kalman decomposition to real biological networks, one may have to restrict the free parameters, as physiology and biochemistry might not allow all mathematically possible coefficients.
+We note that as there are so many degrees of freedom, it may be difficult to infer system function solely from structure.
+\end{example}
+
+\paragraph{Sexual recombination and hybrid incompatibility}
 Phenotypically identical yet kryptotypically variant gene networks, when brought together and/or recombined during sexual reproduction, can lead to phenotypic inviability.
 
-For instance, diploid system coefficients are derived by taking the average of two haploid genomes $\Sys$ and $\Sys'$: $\Sys_{\text{diploid}} \!
+    For instance, the diploid $\{\Sys, \Sys' \}$ has system architure $\Sys_{\text{diploid}} \!
 = \!
@@
 + \!
-\Sys'\right)\!/2$.
+\Sys'\right)\!/2$, the average of its two haploid systems.
 During meiosis a diploid recombines its two genomes to form a new one (a gamete): $\left(\Sys \!
 \times \!
-\Sys' \right) \rightarrow \Sys''$; by randomly swapping matrix rows between two systems $\mathcal{S}$ and $\mathcal{S}'$.
-
-Even if two genomes ($\Sys$ and $\Sys'$) have the same phenotype, their diploid and recombinant systems may not. This is because $\allS$ is not required to be closed under averaging or recombination.
-
-If sexual recombination among systems drawn from $\allS$ yields systems with divergent phenotypes, populations containing significant diversity in $\allS$ can carry genetic load, and isolated populations may fail to produce hybrids with viable phenotypes.
-
-\paragraph*{Hybrid phenotypic breakdown between oscillators}
-For example, if the two phenotypically equivalent systems $A(0)$ and $A(2)$ from Example \ref{ex:all_osc} hybridize, their offspring's phenotype will fail to oscillate completely. 
+\Sys' \right) \rightarrow \Sys''$; by randomly swapping matrix rows between two systems (Appendix \ref{apx:recombination}).
+Then two gametes join to form an offspring (\emph{e.g.} $\{\Sys'',\Sys\} $).
+
+Even if two systems yield the same phenotype, their diploid and recombinant systems may not. This is because $\allS$ is not required to be closed under averaging or recombination.
+
+If sexual recombination among systems drawn from $\allS$ yields systems with divergent phenotypes, populations containing significant diversity in $\allS$ can carry genetic load, and isolated populations may fail to produce hybrids with viable phenotypes (Figure \ref{fig:expF1}).
+
+%\paragraph*{Hybrid phenotypic breakdown between oscillators}
+%For example, if the two phenotypically equivalent systems $A(0)$ and $A(2)$ from Example \ref{ex:all_osc} hybridize, their offspring's phenotype will fail to oscillate completely. 
   \begin{figure}[H] 
@@
     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{expF1}
-    \caption{$F_1$ hybrid (orange) and parental (blue) phenotypic oscillator dynamics for an $A(0)$ by $A(2)$ cross.
-The hybrid $\left(A_{F_1}\!\!=\!\!\tfrac{A\!(\!0\!) \!
-+ \!
-A\!(\!2\!)}{2}\right)$ fails to oscillate and exhibits qualitatively different dynamics ($\phi_{F_0}\!
-= \!
-\sin t \!
-+ \!
-\cos t$, however $\phi_{F_1} \!
-= \!
-\exp t$).
+    \caption{\textbf{Hybrid phenotypic breakdown} $F_1$ hybrid (orange) and parental (blue) phenotypic oscillator dynamics for an $A(0)$ by $A(2)$ cross from Example \ref{ex:all_osc}. The $F_1$ hybrid
+%$\left(A_{F_1}\!\!=\!\!\tfrac{A\!(\!0\!) \! + \! A\!(\!2\!)}{2}\right)$
+gene expression fails to oscillate and instead grows exponentialy,
+exhibiting qualitatively different dynamics ($\phi_{F_0}\! = \! \sin t \! + \! \cos t$, however $\phi_{F_1} \! = \! \exp t$).
     }
+    \label{fig:expF1}
   \end{figure}
@@
 
-\begin{example}[Metabolic network] \label{ex:metabolic}
-
-\plr{Concentrations are "logarithmic", you mean that $u_1(t) = \log(\text{concentration of molcule 1 at time t})$.}
-Consider an organism that can metabolize two different sugars $s_{1}$ and $s_{2}$ (present at logarithmic concentrations $u_{1}$ and $u_{2}$ respectively), 
-with enzymes $e_{1}$ and $e_{2}$ (with log concentration denoted as $\phi_{1}$ and $\phi_{2}$). 
-Further suppose that one of the sugars $s_{2}$ is the preferred energy source (perhaps it contains significantly more energy than the other sugar or is otherwise more efficiently metabolized). 
-The organism may have a gene regulatory network $\mathcal{S}$ that can deploy a situation specific metabolic strategy. 
-\plr{I think you argue below that the optimal response is a complete switch from one enzyme to the other in the presence of the second sugar? Should say so here.}
-That is depending on both $u_{2}$ and $u_{1}$ the organism will synthesize an appropriate $\phi_{1}$ and $\phi_{2}$. 
-Furthermore, consider this system to contain at least two transcription factors, whose log concentrations are given by $\kappa_{1}$, $\kappa_{2}$, $\dots \kappa_{n}$.
-
-Minimally such a system may have the architecture,
-\begin{align*}
-    \mathcal{S}_{\min} = \left\{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 0 & -1 \\ 0 & 0 \end{bmatrix} \begin{bmatrix} \kappa_{1} \\ \kappa_{2} \end{bmatrix} (t) + \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \begin{bmatrix} u_{1} \\ u_{2} \end{bmatrix} (t) \\[11pt]
-    \phi(t) &= \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \begin{bmatrix} \kappa_{1} \\ \kappa_{2} \end{bmatrix} (t) \end{array} \right.
-\end{align*}
-$\mathcal{S}_{\min}$ is minimal as its reachability and observability matrices both have rank $=2$, $\text{rank}(\mathcal{R}) = \text{rank}(\mathcal{O}) = 2 = n \implies \min$.
-
-The impulse response matrix of this system is, 
-\begin{align*}
-    h(t) = \begin{bmatrix} 1 & - t \\ 0 & 1 \end{bmatrix}  .
-\end{align*}
-
-The phenotype is, 
-\begin{align*}
-  \phi(t) = C e^{A t} \kappa(0) + \int_{0}^{t} h(s-t) u(s) ds
-\end{align*}
-($\kappa(0)$ can be set to something like $\left[ -10, -10 \right]$, assuming the transient transcription factor and enzyme concentrations in the organism are typically quite low).
-
-We can see that changing coordinates on this system (with any invertible $V \in \mathbb{R}^{2 \times 2}$) will find new system architectures, 
-as shown before, however, we can also apply the Kalman decomposition to find systems in different dimensions 
--- that is systems that employ more than $2$ transcription factors, yet have the same external dynamics/phenotype. 
-\plr{Ah -- this is much simpler if we let $B$ and $C$ change as well!}
-This may happen if a system is co-opted from another, or may be a consequence of gene duplication and deletion.
-
-\plr{where did $\widehat{\mathcal{S}}$ come from? need a lead-in here.}
-$\widehat{\mathcal{S}}$ is $4$-dimensional biological system,  
-\begin{align*}
-    \widehat{\mathcal{S}} = \left\{ \begin{array}{ll}
-      \widehat{A} = V \begin{bmatrix} X_{1} & X_{2} \\ 0 & A_{\min} \end{bmatrix} V^{-1}; \quad \widehat{B} = V \begin{bmatrix} X_{3} \\ B_{\min} \end{bmatrix}; \quad \widehat{C} = \begin{bmatrix} 0 & C_{\min} \end{bmatrix} V^{-1} \end{array} \right\}
-\end{align*}
-where $V$ is any invertible $4 \times 4$ matrix and $X_{j}$ is any $2 \times 2$ matrix.
-
-\plr{the point here is not clear.}
-So for example, some system can be wired as follows, and still be input-output equivalent to the minimal metabolic system $\mathcal{S}_{\min}$:
- \begin{align*}
-   A' &= \begin{bmatrix}
-     1.6923  &  1.5385 &  -2.6154 &  -2 \\
-     0.8462  &  0.7692 &  -1.3077 &  -1 \\
-     1.2692  &  1.1538 &  -1.9615 &  -1.5 \\
-     0.4231  &  0.3846 &  -0.6538 &  -0.5
-   \end{bmatrix} \\
-   B' &= \begin{bmatrix} 4 & 4 \\ 2 & 2 \\ 3 & 3 \\ 1 & 3 \end{bmatrix} \\
-     C' &= \begin{bmatrix}  0.2692  &  0.1538  &  0.0385 &  -0.5 \\
-     -0.4231 &  -0.3846 &   0.6538  &  0.5 \end{bmatrix}
- \end{align*}
-
-Despite the present example consisting of a minimal $2 \times 2$ and a non-minimal $4 \times 4$ system, 
-any $n$-dimensional system can be constructed using this method 
--- applying a change of coordinates to the Kalman decomposition 
--- to construct a mechanistically different system with identical phenotypic dynamics. 
-Depending on the specifics of a system being modeled, one may have to take care to restrict the free parameter values and network architectures to be biologically appropriate. 
-\plr{what are you thinking of as being a potential problem here? say so explicitly.}
-
-\end{example}
-
 \appendix
@@
 
-\section{Sexual reproduction in linear systems}\label{apx:reproduction}
-
-  Recombination: gamete element $A''$ is formed by recombining (randomly swapping rows) between two parental, possibly distinct, system elements $A$ and $A'$ as follows,
-  \begin{align*}
-    A'' = MA + (I-M)A'
-  \end{align*}
-  where $M$ is a diagonal matrix with only $0$s and $1$s on the diagonal.
+\section{Meiotic recombination in linear systems}\label{apx:recombination}
+
+  Recombination is performed by taking two analogous system components from $\Sys$ and $\Sys'$ and randomly swapping rows or columns.
+  \emph{E.g.} gamete systems $(A'', B'', C'')$, produced by the diploid $\{\Sys, \Sys'\}$, are formed by recombining (randomly swapping rows or columns) between two, possibly distinct, systems $\Sys = (A,B,C)$ and $\Sys' = (A', B', C')$ such that,
+  \begin{align*}
+    \Sys'' = \left( \begin{array}{ll}
+    A'' &= MA + (I-M)A' , \\
+    B'' &= MB + (I-M)B' , \\
+    C'' &= CM + C'(I-M)
+    \end{array} \right)
+  \end{align*}
+  where $M$ is a diagonal matrix where each diagonal element is a Bernoulli random variable ($M_{ii} = 0$ or $1$ with equal probability, and $M_{ij}=0$ if $i \neq j$).
+If systems are different dimensions the smaller system elements can be augmented with $0$s (\emph{e.g.} $\left[ \begin{smallmatrix} A & 0 \\ 0 & 0 \end{smallmatrix} \right], \left[ \begin{smallmatrix} B \\ 0 \end{smallmatrix} \right], \left[ \begin{smallmatrix} C & 0 \end{smallmatrix}\right]$).
 
@@
 
-\section{Hybrid Vigor: Unit Circle approach}
-
-If a fitness optimum is at the origin $(0,0) \in \R^{2}$,
-%and two populations have drifted away from the optimum by an distance of $r$, then the expected value of their $F_1$ hybrids is,
-%\begin{align*}
-%  \mathbb{E}\left[ \left\lVert \frac{1}{2} ((x_1,y_1) + (x_2,y_2))  \right\rVert \right] = \\
-%  \frac{\sqrt{2}}{8}\left(\sqrt{3\, r^2 - {\sin^{-1}\!\left(r\right)}^2 + r^4 - 2\, r\, \sin^{-1}\!\left(r\right)\, \sqrt{1 - r^2}} + \sqrt{{\sin^{-1}\!\left(r\right)}^2 + 5\, r^2 - r^4 + 2\, r\, \sin^{-1}\!\left(r\right)\, \sqrt{1 - r^2}}\right)
-%\end{align*} 
-%If $r < \sim 5.5$ then we will observe hybrid vigor, however if $r$ is sufficiently large, the hybrids will have lower fitness, on average. 
-and two populations have drifted away from the optimum by a distance of $r=1$,
-then $F_1$s will have an average fitness of,
-  \begin{align*}
-    \mathbb{E} \left[ d(F_1) \right] &= \int_{0}^{2\pi} \frac{1}{4 \pi} \left\lVert (1,0) + (\cos\theta,\sin\theta) \right\rVert d\theta \\
-    &= \frac{2}{\pi}
-  \end{align*}
-  %\begin{align*}
-  %  \mathbb{E} \left[ d(F_1)] \right] &= \int_{0}^{2\pi} \int_{0}^{2\pi} \frac{1}{8 \pi^{2}}\left\lVert  \left( \left( \cos(\theta_1), \sin(\theta_1) \right) , \left( \cos(\theta_2), \sin(\theta_2) \right) \right) \right\rVert d\theta_1 d\theta_2 \\
-  %  &= \cos\left(\frac{\pi}{2}\right) = \frac{\sqrt{2}}{2}
-  %\end{align*}
+%\section{Hybrid Vigor: Unit Circle approach}
+%
+%If a fitness optimum is at the origin $(0,0) \in \R^{2}$,
+%%and two populations have drifted away from the optimum by an distance of $r$, then the expected value of their $F_1$ hybrids is,
+%%\begin{align*}
+%%  \mathbb{E}\left[ \left\lVert \frac{1}{2} ((x_1,y_1) + (x_2,y_2))  \right\rVert \right] = \\
+%%  \frac{\sqrt{2}}{8}\left(\sqrt{3\, r^2 - {\sin^{-1}\!\left(r\right)}^2 + r^4 - 2\, r\, \sin^{-1}\!\left(r\right)\, \sqrt{1 - r^2}} + \sqrt{{\sin^{-1}\!\left(r\right)}^2 + 5\, r^2 - r^4 + 2\, r\, \sin^{-1}\!\left(r\right)\, \sqrt{1 - r^2}}\right)
+%%\end{align*} 
+%%If $r < \sim 5.5$ then we will observe hybrid vigor, however if $r$ is sufficiently large, the hybrids will have lower fitness, on average. 
+%and two populations have drifted away from the optimum by a distance of $r=1$,
+%then $F_1$s will have an average fitness of,
 %  \begin{align*}
-%    \mathbb{E} \left[ d(F_1)] \right] &\approx \sqrt{\int_{0}^{2\pi} \frac{1}{4 \pi}\left\lVert  \left( \cos(\pi), \sin(\pi) \right) , \left( \cos(\theta), \sin(\theta) \right) \right\rVert^2 d\theta} \\
-%    &= \sqrt{\frac{1}{2}}
+%    \mathbb{E} \left[ d(F_1) \right] &= \int_{0}^{2\pi} \frac{1}{4 \pi} \left\lVert (1,0) + (\cos\theta,\sin\theta) \right\rVert d\theta \\
+%    &= \frac{2}{\pi}
 %  \end{align*}
-%Haploid $F_2$s will, on average, be a distance of $1$ half of the time, a distance of $0$ a quarter of the time, and a distance of $\sqrt{2}$ a quarter of the time.
-
-  \begin{align*}
-    \mathbb{E} \left[ d(F_2) \right] &= \int_{0}^{2 \pi} \int_{0}^{2 \pi} \frac{1}{4 \pi^{2}} \sum_{i}^{2} \sum_{j}^{2} \sum_{k}^{2} \sum_{l}^{2} \frac{1}{32} \left\lVert \left( \cos(\theta_i) + \cos(\theta_k), \sin(\theta_j) + \sin(\theta_l) \right) \right\rVert d\theta_1 d\theta_2 \\ &= 0.6705
-  \end{align*}
-
-  \begin{align*}
-    \mathbb{E}\left[ d(F_{2_{\text{hap}}}) \right] &= \sqrt{\int_{0}^{2 \pi} \int_{0}^{2 \pi} \frac{1}{8 \pi^{2}} \cos(\theta_1)^{2} + \sin(\theta_2)^{2} d\theta_1 d\theta_2} \\
-    &= \cos \left( \frac{\pi}{2} \right) = \frac{\sqrt{2}}{2}
-  \end{align*}
-
-%  There are $16$ diploid $F_2$s, two have $F_0$ genotypes, two have $F_1$ genotpyes, eight contrain one $F_0$ and one $F_2$ haploid genotype (referred to as $H$; $\fit(H) = \sqrt{\frac{3}{8}}$), two are homozygous for an $F_2$ genotype (referred to as $W$; $\fit(W) = \frac{\sqrt{2}}{2}$), and finally two are heterozygous $F_2$s (referred to as $L$; $\fit(L) = \frac{1}{2}$.)
-  %The expected average distance of diploid $F_2$s is $0.6705$.
-
-  \section{Hybrid Vigor: Probability}
-
-  \begin{align*}
-    X &\sim \mathcal{N} (x, \sigma^{2}) \\
-    Y &\sim \mathcal{N} (y, \sigma^{2}) \\
-    x &\sim \mathcal{N} (0, \eta^{2}) \\
-    y &\sim \mathcal{N}(0, \eta^{2}) \\
-    Z &= \frac{X+Y}{2} \rightarrow \mathbb{E} \left[ \lvert z \rvert^{2} \right] \\
-    Z &= \frac{x+y+U+V}{2} \\
-    \mathbb{E} \left[ \lvert X \rvert^{2} \right] &= \mathbb{E} \left[ \lvert x + U \rvert^{2} \right] = \mathbb{E} \left[ (x+U)^{T} (x+U) \right] \\
-    &= \mathbb{E} \left[ x^{T}x + U^{T}x + x^{T}U + U^{T}T \right] \\
-    &= \mathbb{E} \left[x^{T}x\right] + \mathbb{E} \left[U^{T}U\right] \\
-    &= d\eta^{2} + d\sigma^{2} \qquad d=\text{dimension} \\
-    X &\sim \mathcal{N}(0, \sigma^{2} + \eta^{2}) \\
-    Z &\sim \mathcal{N}\left(0, \frac{1}{4}(2 \eta^{2} + 2\sigma^{2})\right) \\
-    &\rightarrow \mathbb{E}\left[\lvert Z \rvert^{2} \right] \\
-    &= \frac{d}{2} \left(\eta^{2} + \sigma^{2} \right) \\
-    &= \frac{1}{2} \mathbb{E} \left[ \lvert X \rvert^{2} \right]
-  \end{align*}
+%  %\begin{align*}
+%  %  \mathbb{E} \left[ d(F_1)] \right] &= \int_{0}^{2\pi} \int_{0}^{2\pi} \frac{1}{8 \pi^{2}}\left\lVert  \left( \left( \cos(\theta_1), \sin(\theta_1) \right) , \left( \cos(\theta_2), \sin(\theta_2) \right) \right) \right\rVert d\theta_1 d\theta_2 \\
+%  %  &= \cos\left(\frac{\pi}{2}\right) = \frac{\sqrt{2}}{2}
+%  %\end{align*}
+%%  \begin{align*}
+%%    \mathbb{E} \left[ d(F_1)] \right] &\approx \sqrt{\int_{0}^{2\pi} \frac{1}{4 \pi}\left\lVert  \left( \cos(\pi), \sin(\pi) \right) , \left( \cos(\theta), \sin(\theta) \right) \right\rVert^2 d\theta} \\
+%%    &= \sqrt{\frac{1}{2}}
+%%  \end{align*}
+%%Haploid $F_2$s will, on average, be a distance of $1$ half of the time, a distance of $0$ a quarter of the time, and a distance of $\sqrt{2}$ a quarter of the time. 
+%
+%  \begin{align*}
+%    \mathbb{E} \left[ d(F_2) \right] &= \int_{0}^{2 \pi} \int_{0}^{2 \pi} \frac{1}{4 \pi^{2}} \sum_{i}^{2} \sum_{j}^{2} \sum_{k}^{2} \sum_{l}^{2} \frac{1}{32} \left\lVert \left( \cos(\theta_i) + \cos(\theta_k), \sin(\theta_j) + \sin(\theta_l) \right) \right\rVert d\theta_1 d\theta_2 \\ &= 0.6705
+%  \end{align*}
+%
+%  \begin{align*}
+%    \mathbb{E}\left[ d(F_{2_{\text{hap}}}) \right] &= \sqrt{\int_{0}^{2 \pi} \int_{0}^{2 \pi} \frac{1}{8 \pi^{2}} \cos(\theta_1)^{2} + \sin(\theta_2)^{2} d\theta_1 d\theta_2} \\
+%    &= \cos \left( \frac{\pi}{2} \right) = \frac{\sqrt{2}}{2}
+%  \end{align*}
+%
+%%  There are $16$ diploid $F_2$s, two have $F_0$ genotypes, two have $F_1$ genotpyes, eight contrain one $F_0$ and one $F_2$ haploid genotype (referred to as $H$; $\fit(H) = \sqrt{\frac{3}{8}}$), two are homozygous for an $F_2$ genotype (referred to as $W$; $\fit(W) = \frac{\sqrt{2}}{2}$), and finally two are heterozygous $F_2$s (referred to as $L$; $\fit(L) = \frac{1}{2}$.)
+%  %The expected average distance of diploid $F_2$s is $0.6705$. 
+%
+%  \section{Hybrid Vigor: Probability}
+%
+%  \begin{align*}
+%    X &\sim \mathcal{N} (x, \sigma^{2}) \\
+%    Y &\sim \mathcal{N} (y, \sigma^{2}) \\
+%    x &\sim \mathcal{N} (0, \eta^{2}) \\
+%    y &\sim \mathcal{N}(0, \eta^{2}) \\
+%    Z &= \frac{X+Y}{2} \rightarrow \mathbb{E} \left[ \lvert z \rvert^{2} \right] \\
+%    Z &= \frac{x+y+U+V}{2} \\
+%    \mathbb{E} \left[ \lvert X \rvert^{2} \right] &= \mathbb{E} \left[ \lvert x + U \rvert^{2} \right] = \mathbb{E} \left[ (x+U)^{T} (x+U) \right] \\
+%    &= \mathbb{E} \left[ x^{T}x + U^{T}x + x^{T}U + U^{T}T \right] \\
+%    &= \mathbb{E} \left[x^{T}x\right] + \mathbb{E} \left[U^{T}U\right] \\
+%    &= d\eta^{2} + d\sigma^{2} \qquad d=\text{dimension} \\
+%    X &\sim \mathcal{N}(0, \sigma^{2} + \eta^{2}) \\
+%    Z &\sim \mathcal{N}\left(0, \frac{1}{4}(2 \eta^{2} + 2\sigma^{2})\right) \\
+%    &\rightarrow \mathbb{E}\left[\lvert Z \rvert^{2} \right] \\
+%    &= \frac{d}{2} \left(\eta^{2} + \sigma^{2} \right) \\
+%    &= \frac{1}{2} \mathbb{E} \left[ \lvert X \rvert^{2} \right]
+%  \end{align*}
 \end{document}

### 2017-09-13 kalman_walk_paper.tex
added an informal description of the kalman decomposition to the main text. Edited metabolic example and recombination section.

@@
 %\plr{why is this "evolutionary"? It is the thing that we're thinking about evolution of, but that doesn't make it evolutionary.}
-Therefore the set of all phenotypically identical (and minimal) systems can be parameterized as, $\mathcal{S}_{\min}(V) = (VA_0V^{-1}, VB_0, C_0V )$,
+Therefore the set of all phenotypically identical (and minimal) systems can be parameterized as, $\mathcal{S}_{\min}(V) = (VAV^{-1}, VB, CV )$,
   %  \begin{align} \label{eq:min_param}
@@
 \plr{To make sense of this we need to say how $C$ and $B$ change with $n$.}
-Usually, the dimension $n$ and the reference system $A_0$ is implicit and we write only $\calA$.
+Usually, the dimension $n$ and the reference system $\Sys_0$ is implicit and we write only $\calA$.
 Further, we refer to the set of all linear systems with impulse response $h(t)$, regardless of dimension, to be $\allS$,
@@
 In the minimal dimension neutral evolution proceeds as a stochastic coordinate transformation on the elements in $\Sys_{\min}$.
-Even if there is a conservation of phenotype through evolutionary time, the molecular mechanism underpinning the phenotype is not constrained and likely will not be conserved.
+Even if there is conservation of phenotype through evolutionary time, the molecular mechanism underpinning the phenotype is not constrained and likely will not be conserved.
 
@@
  \begin{example}[All Phenotypically Equivalent Cell Cycle Control Networks] \label{ex:all_osc}
-    The set of all two-gene regulatory networks phenotypically equivalent to the cell cycle control network in \ref{ex:oscillator}, where only $A$ can vary, is given by
+   The set of all two-gene regulatory networks phenotypically equivalent to the cell cycle control network in \ref{ex:oscillator}, where only $A$ can vary, is given by $\Sys (\tau) = (A(\tau), B, C)$, such that,
     %\begin{align*}
@@
 \end{tikzpicture}
-    \caption{The set of all phenotype-equivalent cell cycle control networks, $A(\tau)$.
-      $A(\tau) = \left\{ \calA_2 : B = B_0,\ C = C_0\right\}$.}
+      \caption{$\Sys(\tau)$, the set of all phenotype-equivalent cell cycle control networks $\Sys (\tau) = \left\{ \calA_2 : B = [1\ 1]^{T} ,\ C = [1\ 0] \right\}$.}
     \end{figure}
@@
  % \begin{example}[External Equivalence does not imply internal equivalence]
-   Despite the phenotypic equivalence of all instantiations of $A(\tau)$, the kryptotypes, vary as a function of $\tau$. 
+   Despite the phenotypic equivalence of all instantiations of $\Sys(\tau)$, the kryptotypes, vary as a function of $\tau$. 
     %Gene-1 dynamics (blue) are equivalent for network architectures $A(0)$ and $A(2)$, however the dynamics of gene-2 (orange) differ with $\tau$.
@@
     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{figures/osc_kryp_compare}
-    \caption{Gene-1 dynamics (blue) for both systems $A(0)$ and $A(2)$ are identical, however, $A(0)$ gene-2 dynamics (orange) differ from $A(2)$ (green).
-Both gene-1 dynamics are given by $\kappa_{1} = \sin t + \cos t$, and gene-2 by $\kappa_{2} = \cos t - \sin t$ ($A(0)$) and $\kappa'_{2} = \cos t + 3 \sin t$ ($A(2)$).}
+    \caption{Gene-1 dynamics (blue) for both systems $\Sys(0)$ and $\Sys(2)$ are identical, however, $\Sys(0)$ gene-2 dynamics (orange) differ from $\Sys(2)$ (green).
+Both gene-1 dynamics are given by $\kappa_{1} = \sin t + \cos t$, and gene-2 by $\kappa_{2} = \cos t - \sin t$ ($\Sys(0)$) and $\kappa'_{2} = \cos t + 3 \sin t$ ($\Sys(2)$).}
   \end{figure}
@@
     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{expF1}
-    \caption{\textbf{Hybrid phenotypic breakdown} $F_1$ hybrid (orange) and parental (blue) phenotypic oscillator dynamics for an $A(0)$ by $A(2)$ cross from Example \ref{ex:all_osc}. The $F_1$ hybrid
+    \caption{\textbf{Hybrid phenotypic breakdown} $F_1$ hybrid (orange) and parental (blue) phenotypic cell-cycle control (oscillator) dynamics for an $\Sys(0)$ by $\Sys(2)$ cross from Example \ref{ex:all_osc}.
+The $F_1$ hybrid
 %$\left(A_{F_1}\!\!=\!\!\tfrac{A\!(\!0\!) \! + \! A\!(\!2\!)}{2}\right)$
-gene expression fails to oscillate and instead grows exponentialy,
-exhibiting qualitatively different dynamics ($\phi_{F_0}\! = \! \sin t \! + \! \cos t$, however $\phi_{F_1} \! = \! \exp t$).
+gene expression fails to oscillate and instead grows exponentialy
+($\phi_{F_0}\! = \! \sin t \! + \! \cos t$, however $\phi_{F_1} \! = \! e^t$).
     }
@@
   \begin{example}[Hybrid Incompatibility in an Oscillating Gene Network] \label{ex:hybrid_osc}
-
-Here we compare the phenotypes for F2 hybrids formed by crossing oscillators $A(2)$ with $A(2.01)$, $A(2.1)$, and $A(2.5)$ ($B$ and $C$ are the same as above). 
-Each $A$ is phenotypically identical ($\phi(t) = \sin t + \cos t$), however some of the hybrids exhibit markedly different dynamics.
+    Here we compare the phenotypes for $F_2$ hybrids formed by crossing oscillators with different values for $\tau$ from \ref{ex:all_osc}, $\Sys(2)$ with $\Sys(2.01)$, $\Sys(2.1)$, and $\Sys(2.5)$.
+    Systems have identical phenotypes, however some of the hybrids exhibit markedly different dynamics.
 
@@
   \caption{
-    $F_1$ (left) and $F_2$ (right) hybrids crossing $A(2)$ with $A(2.01)$ (top), $A(2.1)$ (middle), and $A(2.5)$ (bottom).
-    Note the difference in scale on the $y$-axis.
-F2 hybrids display more phenotypic divergence than F1s, on average.
-Further, some F2s completely fail to oscillate, as seen in an $A(2.5)$ F2 (light blue).
+    $F_1$ (left) and $F_2$ (right) hybrids crossing $\Sys(2)$ with $\Sys(2.01)$ (top), $\Sys(2.1)$ (middle), and $\Sys(2.5)$ (bottom).
+    $F_2$ hybrids display more phenotypic divergence than $F_1$s, on average. Further, some $F_2$s completely fail to oscillate, as seen in an $\Sys(2.5)$ $F_2$ (light blue).
     \plri{make axis labels bigger}

### 2017-09-14 kalman_walk_paper.tex
added an informal description of the kalman decomposition to the main text. Edited metabolic example and recombination section.
deleted parameter estimate section. edited some of the quantitative genetics stuff -- I think it needs a bit of work though. There are a lot of different versions of the same argument floating around, whether we are using gamma, alpha, c, xi etc.

@@
 and that
-depending on $u_{2}$ and $u_{1}$ the organism will synthesize an appropriate $\phi_{1}$ and $\phi_{2}$. 
+depending on $u_{1}$ and $u_{2}$ the organism will synthesize an appropriate $\phi_{1}$ and $\phi_{2}$. 
 Furthermore, assume this system contains at least two transcription factors, whose log concentrations are $\kappa_{1}$, $\kappa_{2}$, $\dots \kappa_{n}$.
@@
 %\end{align*}
-  $\Sys_{\min}(U) = (U \left[ \begin{smallmatrix} 0 & -1 \\ 0 & 0 \end{smallmatrix} \right] U^{-1}, U \left[\begin{smallmatrix} 1 & 0 \\ 0 & 1 \end{smallmatrix}\right], \left[\begin{smallmatrix} 1 & 0 \\ 0 & 1 \end{smallmatrix}\right]U^{-1})$.
+  $\Sys_{\min}(U) = (U \left[ \begin{smallmatrix} 0 & -1 \\ 0 & 0 \end{smallmatrix} \right] U^{-1}, U 
+    %\left[\begin{smallmatrix} 1 & 0 \\ 0 & 1 \end{smallmatrix}\right]
+      ,
+     % \left[\begin{smallmatrix} 1 & 0 \\ 0 & 1 \end{smallmatrix}\right]
+        U^{-1})$. 
 %
@@
     %This may happen if a system is co-opted from another, or may be a consequence of gene duplication and deletion. 
-  $\Sys(\small X_{1-3},V) \! =  \! \left( \! \begin{array}{ll}
+  $\Sys(\small D_{1-3},V) \! =  \! \left( \! \begin{array}{ll}
      V \!
-\left[\begin{smallmatrix} X_{1} & X_{2} \\ 0 & A \end{smallmatrix}\right] \!
+\left[\begin{smallmatrix} D_{1} & D_{2} \\ 0 & A \end{smallmatrix}\right] \!
 V^{-1}, V \!
-\left[\begin{smallmatrix} X_{3} \\ B \end{smallmatrix}\right], \left[\begin{smallmatrix} 0 & C \end{smallmatrix}\right] \!
+\left[\begin{smallmatrix} D_{3} \\ B \end{smallmatrix}\right], \left[\begin{smallmatrix} 0 & C \end{smallmatrix}\right] \!
 V^{-1} \end{array} \!
@@
 %    \end{align*}
-  ($V$ can be any $4$-dimensional and invertible matrix, $X_{1-3} \in \R^{2 \time 2}$, and $(A,B,C) \in \Sys_{\min}$).
+  ($V$ can be any $4$-dimensional and invertible matrix, $D_{1-3} \in \R^{2 \time 2}$, and $(A,B,C) \in \Sys_{\min}$).
 %So for example, some system can be wired as follows, and still be input-output equivalent to the minimal metabolic system $\mathcal{S}_{\min}$:
@@
 
-    For instance, the diploid $\{\Sys, \Sys' \}$ has system architure $\Sys_{\text{diploid}} \!
+    For instance, the diploid $\{\Sys, \Sys' \}$ has system architecture $\Sys_{\text{diploid}} \!
 = \!
@@
   \begin{tabular}{ccc}
-    \includegraphics[width=0.25\textwidth, height=0.125\paperheight]{F1_comparison} & 
-    \includegraphics[width=0.25\textwidth, height=0.125\paperheight]{F2s_comparison2} &
-    \includegraphics[width=0.25\textwidth, height=0.125\paperheight]{figures/quartic_quadratic2}
+    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{F1_comparison} & 
+    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{F2s_comparison2} \\
+    \includegraphics[width=0.25\textwidth, height=0.125\paperheight]{figures/f1_quartic2} &
+    \includegraphics[width=0.25\textwidth, height=0.125\paperheight]{figures/f2_quad2} 
   \end{tabular}
@@
 
-\paragraph{Offspring.}
-\js{Not sure if this section is redundant, since reproduction is described above for the LTI model.}
-Individuals are diploid; we assume that each haploid genome determines a set of coefficients,
-and the individual's coefficients are the average of her two haploid values (no dominance).
-This implies that the diploid population variance, $\sigma^2$, is one-half the haploid variance.
-Each new gamete is produced from the parent's two haploid copies;
-for simplicity we assume that the gamete inherits a random choice of one of the two parental copies,
-and so $\sigma$ remains constant, up to a $1/N_e$ term.
-A more general model including segregation variance \citep{barton_infinitesimal}
-would result in the same qualitative conclusions.
-\plri{but put this in the appendix?}
+%\paragraph{Offspring.}
+%\js{Not sure if this section is redundant, since reproduction is described above for the LTI model.}
+%Individuals are diploid; we assume that each haploid genome determines a set of coefficients,
+%and the individual's coefficients are the average of her two haploid values (no dominance).
+%This implies that the diploid population variance, $\sigma^2$, is one-half the haploid variance.
+%Each new gamete is produced from the parent's two haploid copies;
+%for simplicity we assume that the gamete inherits a random choice of one of the two parental copies,
+%and so $\sigma$ remains constant, up to a $1/N_e$ term.
+%A more general model including segregation variance \citep{barton_infinitesimal}
+%would result in the same qualitative conclusions.
+%\plri{but put this in the appendix?}
 
@@
 % plr how fast does it drift
-If the variation within a population of some coefficient
+If the regulatory coefficient population variation
 has standard deviation $\sigma$,
-then since subsequent generations resample from this diversity,
+since subsequent generations resample from this diversity,
 the population mean coefficient will move a random distance of size $\sigma/\sqrt{N_e}$ per generation,
@@
 % There may well be functional constraints -- but these are not sufficiently well-known to say anything general about.
-For instance,
-if the variation is due to \textit{cis}-regulatory variants,
-the genetic basis of each \emph{row} of $A$ likely lies within a few kilobases of tightly linked sequence,
-across which a population may carry only a few common haplotypes.
-However, covariance due to transiently assembled haplotypes is not expected to be stable over long periods of time --
-a common \textit{cis}-regulatory haplotype of transcription factor $k$ with particularly strong binding to both $i$ and $j$
-(leading to positive covariance between $A_{ik}$ and $A_{jk}$)
-is no more likely to appear than one with strong binding to $i$ but particularly weak binding to $j$ (negative covariance).
-(Such transient covariances may well increase the variance of the per-generation change in network mean, however \citep{barton_linkage}.)
+%For instance,
+%if the variation is due to \textit{cis}-regulatory variants,
+%the genetic basis of each \emph{row} of $A$ likely lies within a few kilobases of tightly linked sequence,
+%across which a population may carry only a few common haplotypes.
+%However, covariance due to transiently assembled haplotypes is not expected to be stable over long periods of time --
+%a common \textit{cis}-regulatory haplotype of transcription factor $k$ with particularly strong binding to both $i$ and $j$
+%(leading to positive covariance between $A_{ik}$ and $A_{jk}$)
+%is no more likely to appear than one with strong binding to $i$ but particularly weak binding to $j$ (negative covariance).
+%(Such transient covariances may well increase the variance of the per-generation change in network mean, however \citep{barton_linkage}.)
 
@@
 
-To obtain a general quantitative picture, we need to know 
+%To obtain a general quantitative picture, we need to know 
+We denote
 $\sigma_N$ and $\sigma_S$, the standard deviations of coefficient variation along and perpendicular to $\optx$ respectively,
@@
 \plr{I put F1s and F2s here, and $\Incompat$, but with less explanation than below. Merge somehow.} 
-By the arguments above, the means of two allopatric populations separated for $T$ generations
+The means of two allopatric populations separated for $T$ generations
 will be a distance of order $2\sigma_N \sqrt{T/N_e}$ apart along $\optx$.
@@
 and so will have mean system coefficients at the midpoint between their means,
-and variance equal to $\sigma$. 
-\js{I think you mean $\sigma^{2}$?}
+and variance equal to $\sigma^2$. 
+\js{I think you mean $\sigma^{2}$? I changed it.}
 Each $F_2$ hybrid will be homozygous for one parental allele on averge at half of the loci in the genome,
@@
 
+If we prefer not to make any assumptions about the shape of the fitness function, we can more conservatively state that hybrid phenotypes diverge from their optimal parents at rates $d(\Phi_{F_1}, \Phi_0) = \alpha_1 \sigma_N^2 T/N_e$ and $d(\Phi_{F_2}, \Phi_0) = \alpha_2 \sigma_N \sqrt{T/N_e}$, where $\alpha_1$ and $\alpha_2$ are constants.
+Whether or not hybrid phenotypic divergence is associated with a high fitness cost, $F_2$ phenotypes should diverge much faster than $F_1$s, with a rate proportional to the square root of $T/N_e$, versus linearly.
+\js{I think $\alpha_1$ is your $c_1$ and $\alpha_2$ is a function of $\gamma^{-1}$?}
+
+%\subsection*{Parameter estimates for real systems}
+%\js{I think we should move this section?}
+%To translate these results into real predictions, we need to know
+%the strength of stabilizing selection on the phenotype,
+%and the amount (and structure) of heritable variation in the genotype.
+%These are known at best only roughly \citep{felsenstein1988phylogenies},
+%so we aim for order-of-magnitude estimates.
+
+%% amount of heritable variation
+%We quantify (roughly) the amount of heritable variation
+%by $\sigma^2$, the genetic variance present in a population in a typical entry of $\Sys$.
+%The coefficient $A_{ij}$ measures how much the rate of net production of $i$ changes
+%per change in concentration of $j$.
+%It is generally thought that regulatory sequence change contributes much more to inter- and intraspecific variation
+%than does coding sequence change affecting molecular structure \citep{schmidt2010vertebrate}.
+%In the context of transcription factor networks this may be affected 
+%not only by the binding strength of molecule $j$ to the promoter region of gene $i$
+%but also the effects of other transcription factors (e.g., cooperativity)
+%and local chromatin accessibility \citep{stefflova2013cooperativity}.
+%For this reason, 
+%the mutational target size for variation in $A_{ij}$ may be much larger than the dozens of base pairs
+%typically implicated in the handful of binding sites for transcription factor $j$ of a typical promoter region,
+%and single variants may affect many entries of $\Sys$ simultaneously.
+%% (Recall that although these are best modeled through nonlinear terms,
+%% by linearizing we essentially consider first-order effects.)
+%On the other hand, a diverse set of buffering mechanisms are thought to contribute to phenotypic stability
+%in the presence of substantial molecular noise \citep{canalization,buffering},
+%suggesting that substantial variation in the micro-scale dynamics we consider here
+%may be necessary to produce relevant phenotypic effects downstream.
+%\plr{replace "micro-scale" with sthg else or discuss earlier}
+
+%% empirical data on sigma
+%The amount and structure of this standing variation is established over long time scales
+%by many factors, including
+%mutation-selection balance, 
+%shifts in the phenotypic optimum,
+%and/or spatial variation in the optimum \citep{hansen1996translating}.
+%Quantitative genetics models of mutation-selection balance 
+%predict precise levels and structure of standing variation \citep{kimura_mutsel,lande_mutsel,lande1981models},
+%but it is unclear how well these predictions match reality \citep{johnson_barton}
+%and how much they are expected to change over time \citep{arnold_changing_G}.
+%However, empirical work allows us to estimate at least the rough magnitude of variation.
+%Differences in $A_{ij}$ due to a sequence change are hard to measure,
+%but variation in both transcription factor binding site occupancy
+%and expression levels (e.g., cis-eQTL) have been measured in various systems.
+%However, variation in binding site occupancy may overestimate variation in $\Sys$,
+%since it does not capture buffering effects (if for instance only one site of many needs be occupied for transcription to begin),
+%and variation in expression levels measures changes in steady-state concentration (our $\kappa_i$) rather than the \emph{rate} of change.
+%%Nonetheless,
+%%\citet{kasowski2010variation} found differential occupancy in 7.5\%
+%%of binding sites of a transcription factor (p65) between human individuals.
+%%\citet{verlaan2009targeted} showed that cis-regulatory variation
+%%accounts for around 2--6\% of expression variation in human blood-derived primary cells,
+%%while \citep{lappalainen2013transcriptome} found that human population variation 
+%%explained about 3\% of expression variation.
+%%\plr{Get some data from at least one other species in here!}
+%%Taken together, this suggests that variation in the entries of $\Sys$
+%%may be on the order of 1\% between individuals of a population --
+%%doubtless varying substantially between species and between genes.
+%%
+%Given the empirical data on transcription factor binding and gene expression
+%intrapopulation variation \citet{kasowski2010variation, verlaan2009targetedm lappalainen2013transcriptome}, we conservatively estimate that variation in the regulatory coefficients in $\Sys$ may be on the order of 1\% between individuals of a population --
+%doubtless varying substantially between species and between genes.
+%
+%%% argument for strength of selection
+%%It seems certain that selection in most species is not so strong that intra-population variation
+%%is strongly deleterious,
+%%so that if $\beta$ is the typical scale on which selection acts,
+%%then $\beta > \sigma$.  
+%%However,
+%%a range of studies \plr{find them} have found evidence for weak stabilizing selection
+%%on regulatory SNPs and cis-eQTL.
+%For instance, 
+%\citet{maria_and_sergey} \plr{(others?)} found evidence that large-effect regulatory mutations are weakly selected against in Drosophila.
+%This suggests that 
+%the strength of selection on phenotype is sufficient to weakly constrain regulatory variation,
+%so that perhaps $\sigma$ and $\beta$ are relatively close.
+%This is as would be expected if available variation is held in check by mutation--selection balance
+%rather than genetic drift.
+%A conservative estimate would be that $\beta = 5 \sigma$;
+%taking $\sigma=.01$ as above, this suggests that changes in phenotype of 5\% are sufficient
+%to effect a noticeable drop in fitness.
+%\plr{BUT $\sigma$ IS VARIATION IN $A$ NOT PHENOTYPE}
+%
+%We have guessed that within a population, 
+%the entries of $\Sys$ vary by around 1\%, 
+%at least for networks whose function is strongly constrained.
+%
+%
 \paragraph{Speciation rates under neutrality}
-%
-\plr{separate out here (or above) the difference in genotype of hybrids from the optimum set, the difference in phenotype, and the difference in fitness.}
-%In general terms, if fitness $\fit$ is some function $g$ of phenotypic divergence from an optimal scaled by the strength of selection $\beta$,
-%\begin{align*}
-%  \fit = g \left( \frac{\phi - \phi_{0}}{\beta} \right)
-%\end{align*}
-%\plr{sentences don't begin with "And"}
-%And a population $P$ is normally distributed around a mean gene network architecture $A(\tau)$ and $\sigma^{2}$ is the intra-population regulatory variance,
-%\plr{what is $T$?}
-%\begin{align*}
-%  P_{T} \sim \mathcal{N} \left(A(\tau), T \frac{\sigma^{2}}{N_{e}}\right)
-%\end{align*}
-%the population mean architecture will move at rate
-%\plr{what is $\Delta \tau$? Also, if we reframe in terms of $V$ not $\tau$ above will need to reframe this.}
-%\begin{align*}
-%  \Delta \tau = \sigma \sqrt{\frac{T}{N_{e}}}
-%\end{align*}
-%where $T$ is time measured in generations. 
-%USING GENERAL MATH FROM MULTIVARIATE TAYLOR SERIES....
-we know that the phenotype of a first generation $F_{1}$ hybrid diverges from that of its parents (assuming both parents are at an optimum and phenotypically equivalent) quartically as a function of $\Delta \tau$.
-%Thus the squared difference of it's phenotype will be,
-%\begin{align*}
-%  \Delta \Phi^{2}_{1} = c_{1}^{2} \sigma^{4} \left( \frac{T}{N_{e}} \right)^{2}
-%\end{align*}
-%and $F_{1}$ fitness will decline linearly with respect to $\frac{T}{N_{e}}$. ARGUMENT FOR WHY $\sigma^{2} c_{1} \sim \beta$ HERE.
-%   \begin{align*}
-%     \Delta \fit_{1} = \left( \frac{\Delta \Phi_{1}}{\beta} = \frac{c_{1} \sigma^{2} \frac{T}{N_{e}}}{\beta} \approx \frac{T}{N_{e}} \right)
-%\end{align*}
-% It follows FROM THE SAME MATH ABOUT THE MULTIVAR TAYLOR that the phenotypes of second generation $F_{2}$ hybrid crosses diverges quadratically as a function of $\Delta \tau$, 
-%\begin{align*}
-%  \Delta \Phi^{2}_{2} &= c_{2}^{2} \left( \sigma \sqrt{\frac{T}{N_{e}}} \right)^{2} = c_{2}^{2} \frac{T}{N_{e}} \\
-%  \Delta \fit_{2} &= \left( \frac{\Delta \Phi_{2}}{\beta} = \frac{c_{2} \sigma \sqrt{\frac{T}{N_{e}}}}{\beta} \approx \sqrt{\frac{T}{N_{e}}} \right)
-%\end{align*}
-%and that $F_{2}$ fitness drops rapidly -- at approximately rate $\sqrt{\frac{T}{N_{e}}}$. ARGUMENT FOR WHY $c_{2} \sigma \sim \beta$ HERE. 
-%
 For cases where system structures and $\gamma$ can be inferred,
@@
 %  where $\xi := 2\xi_2/\xi_1$.
-  This suggests that, as an upper bound, if a population contains as much phenotypic diversity as allowed by selection, reproductively isolated populations can speciate rapidly -- possibly on timescales shorter than $N_{e}$ generations.
+  This suggests that, as an upper bound, if a population contains as much phenotypic diversity as allowed by selection, reproductively isolated populations can speciate rapidly -- possibly on timescales shorter than $N_{e}$ generations (It is likely that, on average, $c \leq 2/3$).
 %%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Parameter estimates for real systems}
-\js{I think we should move this section?}
-To translate these results into real predictions, we need to know
-the strength of stabilizing selection on the phenotype,
-and the amount (and structure) of heritable variation in the genotype.
-These are known at best only roughly \citep{felsenstein1988phylogenies},
-so we aim for order-of-magnitude estimates.
-
-%% amount of heritable variation
-We quantify (roughly) the amount of heritable variation
-by $\sigma^2$, the genetic variance present in a population in a typical entry of $A$.
-The coefficient $A_{ij}$ measures how much the rate of net production of $i$ changes
-per change in concentration of $j$.
-It is generally thought that regulatory sequence change contributes much more to inter- and intraspecific variation
-than does coding sequence change affecting molecular structure \citep{schmidt2010vertebrate}.
-In the context of transcription factor networks this may be affected 
-not only by the binding strength of molecule $j$ to the promoter region of gene $i$
-but also the effects of other transcription factors (e.g., cooperativity)
-and local chromatin accessibility \citep{stefflova2013cooperativity}.
-For this reason, 
-the mutational target size for variation in $A_{ij}$ may be much larger than the dozens of base pairs
-typically implicated in the handful of binding sites for transcription factor $j$ of a typical promoter region,
-and single variants may affect many entries of $A$ simultaneously.
-% (Recall that although these are best modeled through nonlinear terms,
-% by linearizing we essentially consider first-order effects.)
-On the other hand, a diverse set of buffering mechanisms are thought to contribute to phenotypic stability
-in the presence of substantial molecular noise \citep{canalization,buffering},
-suggesting that substantial variation in the micro-scale dynamics we consider here
-may be necessary to produce relevant phenotypic effects downstream.
-\plr{replace "micro-scale" with sthg else or discuss earlier}
-
-%% empirical data on sigma
-The amount and structure of this standing variation is established over long time scales
-by many factors, including
-mutation-selection balance, 
-shifts in the phenotypic optimum,
-and/or spatial variation in the optimum \citep{hansen1996translating}.
-Quantitative genetics models of mutation-selection balance 
-predict precise levels and structure of standing variation \citep{kimura_mutsel,lande_mutsel,lande1981models},
-but it is unclear how well these predictions match reality \citep{johnson_barton}
-and how much they are expected to change over time \citep{arnold_changing_G}.
-However, empirical work allows us to estimate at least the rough magnitude of variation.
-Differences in $A_{ij}$ due to a sequence change are hard to measure,
-but variation in both transcription factor binding site occupancy
-and expression levels (e.g., cis-eQTL) have been measured in various systems.
-However, variation in binding site occupancy may overestimate variation in $A$,
-since it does not capture buffering effects (if for instance only one site of many needs be occupied for transcription to begin),
-and variation in expression levels measures changes in steady-state concentration (our $\kappa_i$) rather than the \emph{rate} of change.
-Nonetheless,
-\citet{kasowski2010variation} found differential occupancy in 7.5\%
-of binding sites of a transcription factor (p65) between human individuals.
-\citet{verlaan2009targeted} showed that cis-regulatory variation
-accounts for around 2--6\% of expression variation in human blood-derived primary cells,
-while \citep{lappalainen2013transcriptome} found that human population variation 
-explained about 3\% of expression variation.
-\plr{Get some data from at least one other species in here!}
-Taken together, this suggests that variation in the entries of $A$
-may be on the order of 1\% between individuals of a population --
-doubtless varying substantially between species and between genes.
-
-%% argument for strength of selection
-It seems certain that selection in most species is not so strong that intra-population variation
-is strongly deleterious,
-so that if $\beta$ is the typical scale on which selection acts,
-then $\beta > \sigma$.  
-\plr{do we call this $\beta$ or $u$?} \js{$\beta$, $u$ is input.}
-However,
-a range of studies \plr{find them} have found evidence for weak stabilizing selection
-on regulatory SNPs and cis-eQTL.
-For instance, 
-\citet{maria_and_sergey} \plr{(others?)} found evidence that large-effect regulatory mutations are weakly selected against in Drosophila.
-This suggests that 
-the strength of selection on phenotype is sufficient to weakly constrain regulatory variation,
-so that perhaps $\sigma$ and $\beta$ are relatively close.
-This is as would be expected if available variation is held in check by mutation--selection balance
-rather than genetic drift.
-A conservative estimate would be that $\beta = 5 \sigma$;
-taking $\sigma=.01$ as above, this suggests that changes in phenotype of 5\% are sufficient
-to effect a noticeable drop in fitness.
-\plr{BUT $\sigma$ IS VARIATION IN $A$ NOT PHENOTYPE}
-
-We have guessed that within a population, 
-the entries of $A$ vary by around 1\%, 
-at least for networks whose function is strongly constrained.
-
 %%%%%%%%%%%%%%%

### 2017-09-27 kalman_walk_paper.tex
edited kalman walk and figure 6

@@
 \newcommand{\Sys}{\mathcal{S}}
-\newcommand{\allS}{\mathcal{K}}
+\newcommand{\allS}{\mathcal{A}}
 %\newcommand{\allS}{\mathcal{A}}
@@
     %\item add point about F1: quartic and F2: quadratic to results and maybe abstract/discussion
-    \item hybrid vigor (need to do calculation)
+    %\item hybrid vigor (need to do calculation)
     \item discuss linearity, linearization, and canalization in introduction
@@
 
-\plri{Note: in the \LaTeX source I'm putting in semantic linebreaks, so it's easy to edit and move around phrases and ideas.}
-
-\plri{Need to come up with a consistent term for ``the $A_{ij}$''s. -- ``regulatory coefficients''? ``genotype''?}
+%\plri{Note: in the \LaTeX source I'm putting in semantic linebreaks, so it's easy to edit and move around phrases and ideas.}
+
+%\plri{Need to come up with a consistent term for ``the $A_{ij}$''s. -- ``regulatory coefficients''? ``genotype''?}
 
@@
 Progress in a suite of biological subdisciplines is stalled by our general lack of understanding of this molecular machinery: with respect to both its function and evolution. 
-  \plr{Saying "stalled" reflects negatively on other fields - rephrase?} \js{I don't think it reflects negatively. Even if it did, as long as it's a true statement it should be fine.}.
 There does exist a growing body of data on the evolutionary histories and molecular characterizations of particular gene regulatory networks \citep{jaeger2011gap, davidson2006gene, israel2016comparative}, as well as thoughtful verbal and conceptual models \citep{true2001developmental, gwagner1, weiss2000phenogenetic, edelman2001degeneracy}.
-  \plr{I don't think the HW example will speak to our audience.
-Also, better to say why math is good rather than why not-math is bad.} \js{deleted HW.
-I don't think we need to write an anodyne -- the current sentence is true enough.}
 However, verbal theories are often insufficient, if not downright misleading \citep{servedio2014not}. 
@@
 In computer science, this is framed as the relationship among processes that simulate one another \citep{van2004equivalence}.  
-\plr{here you are jumping to evolution -- haven't said yet why evolution "can explore" these -- needs to say explicitly what we mean by neutral here}
 Although this may frustrate the occasional engineer or scientist, viewed from another angle,
 the concepts of unidentifiability and indistinguishability can provide a starting point for
-thinking about externally equivalent systems -- systems that evolution can explore, so long
-as the parameters and structures can be realized biologically. In fact, evolutionary
-biologists who study convergent versus parallel evolution, homology, and analogy are very familiar with such functional symmetries; 
+thinking about externally equivalent systems
+-- systems that evolution can explore, so long as the parameters and structures can be realized biologically.
+In fact, evolutionary
+biologists who study convergent versus parallel evolution, and developmental homology and analogy, are very familiar with such functional symmetries; 
 macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level 
@@
 
-In this paper we outline a theoretical framework to study the evolution of biological systems. 
-  \plr{rather than "we expect" maybe say that the framework could be applied to non-neutral evolution}\js{fixed}
-Presently, we focus solely on a neutral scenario, 
+In this paper we outline a theoretical framework to study the evolution of biological systems, such as gene regulatory networks. 
+We focus primarily on the neutral scenario: 
 that is where phenotype is conserved over evolutionary time. 
-however, this framework could be applied to a wider set of evolutionary scenarios. 
-\plr{here's the definition of neutral, but it's not quite right?} \js{how's that?}
-
+%however, this framework could be applied to a wider set of evolutionary scenarios. 
 We present an analytical description of the set of all linear biological systems with identical phenotypes 
-\plr{We need something here saying what we mean by "gene network architecture".  Also, I don't think we claim anything for "all biological systems".} \js{added the word ``linear.''}
--- that is we describe the set of all gene network architectures that yield identical phenotypes, and show that all linear biological systems can, in principal, can undergo systems drift. 
-In the neutral case, this set describes a manifold that evolution explores leaving phenotype invariant with respect to mutation, 
-and predicts that if two populations become reproductively isolated, hybrid incompatibility can occur, despite the absence of adaptation, directional selection, or environmental change. 
-Speciation typically occurs on timescales approximately on the order of $N_{e}$ generations, where $N_{e}$ is the effective population size.
+-- that is we describe the set of all gene network architectures that yield identical phenotypes in equivalent environments, and show that all linear biological systems can, in principal, can undergo systems drift and rewire.
+Under neutrality, a population will drift through the set of all possible phenotypically equivalent linear gene networks.
+Consequentially, two phenotypically equivlanet yet allopatrically isolated populations, will likely produce incompatibile hybrids, despite the absence of adaptation, directional selection, or environmental change.
+Under these conditions, speciation typically occurs on timescales approximately on the order of $N_{e}$ generations, where $N_{e}$ is the effective population size.
 %
 \section*{Gene networks as linear dynamical systems}
-%
-\plr{Below we mix general language with language specific to GRNs, like switching between "internal state" and "transcription factor concentration.  
-Rather than keep making comments like "(or some other system)",
-we could say up top that this applies to other situations, but to make it concrete we'll talk about regulatory networks.}
-%
 %  \jss{Organisms' phenotypes are constructred by gene by gene by environment interactions.
@@
 Such a biological system $\mathcal{S}$ is given by,
-  \begin{align}
+  \begin{align}\label{eqn:system}
     \mathcal{S} := \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right.
@@
 How many different mechanisms have the same function? 
-
 Gene regulatory networks with identical phenotypes do not necessarily have identical kryptotypes.
-Any linear and minimal gene network 
--- minimal, informally meaning that the system's phenotype is achieved with the fewest possible genes ($n=\min$, see \ref{apx:kalman})
--- has an identical impulse response and thus phenotype given identical inputs $u(t)$, up to a change of coordinates. 
-\plr{"has an identical phenotype" is not what you mean to say here}
-\js{edited -- does that work?}
-  \begin{equation}
-    \begin{aligned}
-      h(t) &= C e^{A t} B 
-      = C V^{-1} V e^{A t} V^{-1} V B \\
-      &= C V^{-1} e^{V A V^{-1} t} V B 
-      = \bar{C} e^{\bar{A} t} \bar{B}
-    \end{aligned}
-  \end{equation}
-Two gene networks, $\mathcal{S} = ( A, B, C )$, and $\bar{\mathcal{S}} = (\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} )$, 
-have the same phenotype, for all possible inputs, if they are related by a change of coordinates. 
-\plr{is this the place to say "and only if, in the minimal dimension"?}
-
+%
+%Any linear and minimal gene network 
+%-- minimal, informally meaning that the system's phenotype is achieved with the fewest possible genes ($n=\min$, see \ref{apx:kalman})
+%-- has an identical impulse response and thus phenotype given identical inputs $u(t)$, up to a change of coordinates. 
+%  \begin{equation}
+%    \begin{aligned}
+%      h(t) &= C e^{A t} B 
+%      = C V^{-1} V e^{A t} V^{-1} V B \\
+%      &= C V^{-1} e^{V A V^{-1} t} V B 
+%      = \bar{C} e^{\bar{A} t} \bar{B}
+%    \end{aligned}
+%  \end{equation}
+%Two gene networks, $\mathcal{S} = ( A, B, C )$, and $\bar{\mathcal{S}} = (\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} )$, 
+%have the same phenotype, for all possible inputs, if and only if they are related by a change of coordinates. 
+%
 %Although systems may not be identifiable beyond a change of coordinates, at present we are primarily interested in a subset of these systems. 
@@
 %In other words systems with varying gene regulatory network architectures yet identical selection pressures, environment, and phenotype.
-
+%
 %Define $V(\tau)$ as the parameterized change of coordinates matrix that preserves $B$ and $C$, with $\tau$ a vector of free parameters.
@@
 %\plr{why is this "evolutionary"? It is the thing that we're thinking about evolution of, but that doesn't make it evolutionary.}
-Therefore the set of all phenotypically identical (and minimal) systems can be parameterized as, $\mathcal{S}_{\min}(V) = (VAV^{-1}, VB, CV )$,
+%
+%Therefore the set of all phenotypically identical (and minimal) systems can be parameterized as, $\mathcal{S}_{\min}(V) = (VAV^{-1}, VB, CV )$,
   %  \begin{align} \label{eq:min_param}
@@
   %\end{align}
-  where $V$ is any invertible matrix and it's elements are free parameters.
-
-Therefore varying $V$ can tweak the relationships between genes within a regulatory network, yet preserve the phenotype.
+ % where $V$ is any invertible matrix and it's elements are free parameters. 
+%
+%Therefore varying $V$ can tweak the relationships between genes within a regulatory network, yet preserve the phenotype.
 Some gene networks, however, can grown or shrink, perhaps following gene duplications and deletions, and also still preserve their phenotypes.
 These changes cannot be described by $\mathcal{S}(V)$, but are ecanpuslated by \ref{eqn:equivalence}, below.
-
-More generally,
-we denote by $\calA_n(\Sys_0)$ the set of all $n$-dimensional systems equivalent to $\Sys_0$:
+%The Kalman decomposition decomposes a system into \emph{reachable} and \emph{observable} components -- that is coefficients that respond to an input and coefficients that produce an observable output.
+A minimal system's components are both reachable and observable.
+In this form, it is possible to add components that respond to an input, but do not influence the output, or components that, in principle, influence the output, but do not respond to any inputs.
+Any system with such components is, by definition, non-minimal, and any system can be modified to include any finite number of extra unreachable and/or unobservable components, without modifying its impulse response.
+Decomposing the system allows for the easy manipulation of ``superfluous''(but not necessarily expendable) components.
+After unreachable/unobservable coefficients are added, changing coordinates ``stirs'' the system coefficients such that the previously superfluous coefficients can become essential to system funciton.
+%Whether or not evolution constrains biological systems to be minimal or nearly minimal is an open question.
+
+\begin{definition}[Phenotypic equivalence of systems]
+    Let $(\kappa(t),\phi(t))$ and $(\bar \kappa(t),\bar \phi(t))$ be the solutions to \eqref{eqn:system}
+    with coefficient matrices $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ respectively,
+    and both $\kappa(0)$ and $\bar \kappa(0)$ are zero. 
+    The systems defined by $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ are
+    \textbf{phenotypically equivalent} 
+    if
+    \begin{align*}
+        \phi(t) = \bar \phi(t) \qquad \text{for all} \; t \ge 0.
+    \end{align*}
+    Equivalently, this occurs if and only if
+    \begin{align*}
+        h(t) = \bar h(t)  \qquad \text{for all} \; t \ge 0,
+    \end{align*}
+    where $h$ and $\bar h$ are the impulse responses of the two systems.
+\end{definition}
+
+One way to find other systems equivalent to a given one
+is by change of coordinates:
+if $V$ is an invertible matrix, then the systems $(A,B,C)$ and $(VAV^{-1},VB,CV^{-1})$
+have the same dynamics because their impulse responses are equal:
+  \begin{equation}
+    \begin{aligned}
+      h(t) &= C e^{A t} B 
+      = C V^{-1} V e^{A t} V^{-1} V B \\
+      &= C V^{-1} e^{V A V^{-1} t} V B 
+      = \bar{C} e^{\bar{A} t} \bar{B}
+    \end{aligned}
+  \end{equation}
+%\begin{align*}
+%    CV^{-1}( zI - VAV^{-1})^{-1}VB
+%    =
+%    CV^{-1}V( zI - A)^{-1}V^{-1}VB
+%    =
+%    C( zI - A)^{-1}B .
+%\end{align*}
+However, the converse is not necessarily true: 
+systems can have identical impulse responses without being changes of coordinates of each other.
+In fact, systems with identical impulse responses can involve interactions between different
+numbers of molecular species, and thus be in different dimensions.
+
+We refer to $\calA(\Sys_0)$ as the set of all systems equivalent to $\Sys_0$, regardless of dimension:
 \begin{equation} \label{eqn:equivalence}
   \begin{aligned}
-    \calA_n(\Sys_0) 
+    \calA(\Sys_0) 
       &= \left\{
@@
 Equivalence of the two characterizations follows from the Cayley-Hamilton theorem.
-\plr{To make sense of this we need to say how $C$ and $B$ change with $n$.}
-Usually, the dimension $n$ and the reference system $\Sys_0$ is implicit and we write only $\calA$.
-Further, we refer to the set of all linear systems with impulse response $h(t)$, regardless of dimension, to be $\allS$,
-  \begin{align}
-    \allS := \bigcup_{n=\min}^{\infty} \calA_n(\Sys_0)
-  \end{align}
-  \js{what do you like better for $\allS$?
-$\mathcal{A}$, $\mathcal{A}_{\infty}$, or $\mathcal{K}_{h(t)}$, or something else -- maybe a modification of $\Sys$?
-Using a script of ``A'' was more intuitive when we were leavng $B$ and $C$ invariant.}
-Two systems, even if in different dimensions, will have identical phenotypes if they are in $\allS$.
-Further, there is no unique triple $(A,B,C)$ for the mapping $h: u \mapsto \phi$.
-This implies that the gene regulatory network architecture per phenotype/environment pair is never unique -- that is under these assumptions, \emph{all} gene regulatory networks can drift/rewire, and still conserve the phenotype.
-This set can be precisley defined and completely parameterized using the \emph{Kalman decomposition} (Appendix \ref{apx:kalman}).
-
-The Kalman decomposition decomposes a system into \emph{reachable} and \emph{observable} components -- that is coefficients that respond to an input and coefficients that produce an observable output.
-A minimal system's components are both reachable and observable.
-In this form, it is possible to add components that respond to an input, but do not influence the output, or components that, in principle, influence the output, but do not respond to any inputs.
-Any system with such components is, by definition, non-minimal, and any system can be modified to include any finite number of extra unreachable and/or unobservable components, without modifying its impulse response.
-Decomposing the system allows for the easy manipulation of ``superfluous''(but not necessarily expendable) components.
-After unreachable/unobservable coefficients are added, changing coordinates ``stirs'' the system coefficients such that the previously superfluous coefficients can become essential to system funciton.
-Whether or not evolution constrains biological systems to be minimal or nearly minimal is an open question. 
-\js{should the sentence of evolution of nearly minimal be in the paper? Or should we save this for our next paper?}
-
-A large, adapted population will move around in $\allS$ under environmental and selective stasis.
-In the minimal dimension neutral evolution proceeds as a stochastic coordinate transformation on the elements in $\Sys_{\min}$.
-Even if there is conservation of phenotype through evolutionary time, the molecular mechanism underpinning the phenotype is not constrained and likely will not be conserved.
+
+The set of all systems phenotypically equivalent to a given system $(A,B,C)$ 
+is elegantly described using the Kalman decomposition,
+which also clarifies the system dynamics.
+To motivate this, first note that the input $u(t)$ only directly pushes the system
+in directions lying in the span of the columns of $B$.
+As a result, different combinations of input can 
+move the system in any direction that lies in the \emph{reachable subspace},
+which we denote by $\reachable$,
+and is defined to be the closure of $\spn(B)$ under applying $A$
+(or equivalently, the span of $B, AB, A^2B, \ldots A^{n-1}B$).
+Analogously to this, we define
+the \emph{observable subspace}, $\mathcal{O}$,
+to be the closure of $\spn(C^T)$ under applying $A$.
+(Or: $\unobservable$ is the largest $A$-invariant subspace
+contained in the null space of $C$;
+and $\reachable$ is the largest $A$-invariant subspace contained in the image of $B$.)
+
+If we define
+\begin{enumerate}
+    \item The columns of $P_\rno$ are an orthonormal basis for $\reachable \cap \unobservable$.
+    \item The columns of $P_\ro$ are an orthonormal basis of
+        the complement of $\reachable \cap \unobservable$ in $\reachable$.
+    \item The columns of $P_\nro$ are an orthonormal basis of
+        the complement of $\reachable \cap \unobservable$ in $\unobservable$.
+    \item The columns of $P_\nrno$ are an orthonormal basis of
+        the remainder of $\R^n$.
+\end{enumerate}
+If we then define
+\begin{align*}
+    P &= 
+    \left[ \begin{array}{c|c|c|c}
+        P_\rno & P_\ro & P_\nro & P_\nrno
+    \end{array} \right] ,
+\end{align*}
+then
+\begin{align*}
+    P^T P
+    &=
+    \left[ \begin{array}{c|c|c|c}
+        I & 0 & 0 & 0 \\
+        \hline
+        0 & I & U & 0 \\
+        \hline
+        0 & V & I & 0 \\
+        \hline
+        0 & 0 & 0 & I 
+    \end{array} \right] .
+\end{align*}
+\plr{Check this.  Can we get $U=V=0$?}
+
+The following theorem can be found in SOME REFERENCE.
+
+\begin{theorem}[Kalman decomposition] \label{thm:kalman}
+        For any system $(A,B,C)$ with corresponding Kalman basis matrix $P$,
+        the transformed system $(PAP^{-1},PB,CP^{-1})$  has the following form:
+        \begin{align*}
+            \widehat A = PAP^{-1}
+            &=
+            \left[ \begin{array}{cccc}
+                A_{\rno} & A_{\rno,\ro} & A_{\rno,\nrno} & A_{\rno,\nro} \\
+                0 & A_{\ro} & 0 & A_{\ro,\nro} \\
+                0 & 0 & A_{\nrno} & A_{\nrno,\nro} \\
+                0 & 0 & 0 & A_{\nro}
+            \end{array} \right] ,
+        \end{align*}
+        and
+        \begin{align*}
+            \widehat B = PB
+            &=
+            \left[ \begin{array}{cccc}
+                B_{\rno} \\
+                B_{\ro} \\
+                0 \\
+                0 
+            \end{array} \right] ,
+        \end{align*}
+        and
+        \begin{align*}
+            \widehat C = CP^{-1}
+            &=
+            \left[ \begin{array}{cccc}
+                0 & C_{\ro} & C_{\nrno} & 0 
+            \end{array} \right] .
+        \end{align*}
+        The impulse response of both systems is given by
+        \begin{align*}
+          h(t) = C_{\ro} e^{A_{\ro} t} B_{\ro} .
+        \end{align*}
+\end{theorem}
+
+In the latter case, we say that the system is \emph{minimal} 
+-- there is no equivalent system with a smaller number of molecular species.
+Note that this says that any two equivalent minimal systems
+are changes of coordinates of each other.
+
+Since any system can be put into this form,
+and once in this form, its impulse response is determined only by 
+$C_{\ro}$, $A_{\ro}$, and $B_\ro$,
+therefore, the set of all equivalent systems are parameterized by
+the dimension $n$,
+the choice of basis ($P$),
+the remaining submatrices in $\widehat A$, $\widehat B$, and $\widehat C$
+(which are unconstrained),
+and a invertible transformation of $\spn(P_{\ro})$, which we call $T_\ro$.
+
+\begin{theorem}[Parameterization of equivalent systems]
+    Let $(A,B,C)$ be a minimal system.
+    \begin{itemize}
+        \item[(a)]
+            Every equivalent system is of the form given in Theorem \ref{thm:kalman},
+            i.e., can be specified by choosing a dimension, $n$;
+            submatrices in $\widehat A$, $\widehat B$, and $\widehat C$ 
+            except for $A_\ro=A$, $B_\ro=B$, and $C_\ro=C$;
+            and choosing an invertible matrix $P$.
+
+        \item[(b)]
+            \plr{conjecture:}
+            The parameterization is unique
+            if $P$ is furthermore chosen so that 
+            each $P_x$ other than $P_\ro$ is a projection matrix,
+            and that 
+            \begin{align*}
+                0
+                =
+                P_x^T P_y
+            \end{align*}
+            for all $(x,y)$ except $(\ro,\nrno)$.
+
+    \end{itemize} 
+\end{theorem}
+
+%Usually, the dimension $n$ and the reference system $\Sys_0$ is implicit and we write only $\calA$.
+%Further, we refer to the set of all linear systems with impulse response $h(t)$, regardless of dimension, to be $\allS$,
+%  \begin{align}
+%    \allS := \bigcup_{n=\min}^{\infty} \calA_n(\Sys_0)
+%  \end{align}
+It is remarkable to note that even with the relationships between environment, kryptotype, and phenotype constant, and in the minimal dimension, there are still almost always degrees of freedom.
+These correspond to distinct genetic networks that perform indistinguishable functions.
+For example, the equivalent systems in Example \ref{ex:all_osc} are minimal, and share common B and C matrices.
+
+It is also remarkable to note that there is no unique triple $(A,B,C)$ for the mapping $h: u \mapsto \phi$.
+This implies that the gene regulatory network architecture per phenotype/environment pair is never unique -- that is under these assumptions, there is always more than one possible gene regulatory network architecture per phenotype.
+%This set can be precisley defined and completely parameterized using the \emph{Kalman decomposition} (Appendix \ref{apx:kalman}).
+
+For the remainder of the paper, we interpret $\allS$ as the phenotypially neutral landscape, wherein a large population will drift under environmental and selective stasis.
+Even if the phenotype is constrained and remains constant through evolutionary time, the molecular mechanism underpinning it is not constrained and likely will not be conserved.
 
@@
 %    \end{align*}
-  ($V$ can be any $4$-dimensional and invertible matrix, $D_{1-3} \in \R^{2 \time 2}$, and $(A,B,C) \in \Sys_{\min}$).
+  ($V$ can be any $4$-dimensional and invertible matrix, $D_{1-3} \in \R^{2 \times 2}$, and $(A,B,C) \in \Sys_{\min}$).
 %So for example, some system can be wired as follows, and still be input-output equivalent to the minimal metabolic system $\mathcal{S}_{\min}$:
@@
 
-    For instance, the diploid $\{\Sys, \Sys' \}$ has system architecture $\Sys_{\text{diploid}} \!
-= \!
-\left(\Sys \!
-+ \!
-\Sys'\right)\!/2$, the average of its two haploid systems.
-During meiosis a diploid recombines its two genomes to form a new one (a gamete): $\left(\Sys \!
+    For instance, the diploid $\{\Sys, \Sys' \}$ has system architecture $\Sys_{\text{diploid}} \! = \! \left(\Sys \! + \! \Sys'\right)\!/2$, the average of its two haploid systems.
+    %During meiosis a diploid recombines its two genomes to form a new one (a gamete): $\left(\Sys \!
 \times \!
@@
 Then two gametes join to form an offspring (\emph{e.g.} $\{\Sys'',\Sys\} $).
+
+     During meiosis, a diploid recombines its two genomes to form a gamete by randomly swapping rows or columns between its two systems.
+  \emph{E.g.} gamete systems $(A'', B'', C'')$, produced by the diploid $\{\Sys, \Sys'\}$, are formed by recombining two systems $\Sys = (A,B,C)$ and $\Sys' = (A', B', C')$  to form $\Sys''$ such that,
+  %\begin{align*}
+   $A'' = MA + (I-M)A'$, 
+   $B'' = MB + (I-M)B'$, and
+   $C'' = CM + C'(I-M)$.
+ % \end{align*}
+  $M$ is a diagonal matrix where each diagonal element is a Bernoulli random variable ($M_{ii} = 0$ or $1$ with equal probability, and $M_{ij}=0$ if $i \neq j$).
+If systems are different dimensions the smaller system elements can be augmented with $0$s (\emph{e.g.} $\left[ \begin{smallmatrix} A & 0 \\ 0 & 0 \end{smallmatrix} \right], \left[ \begin{smallmatrix} B \\ 0 \end{smallmatrix} \right], \left[ \begin{smallmatrix} C & 0 \end{smallmatrix}\right]$).
 
@@
 
-\begin{figure}[H]
+    \begin{figure}[H] \label{fig:hybs}
   \centering
@@
     \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{F1_comparison} &
-    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{F2s_comparison2} \\
-    \includegraphics[width=0.25\textwidth, height=0.125\paperheight]{figures/f1_quartic2} &
-    \includegraphics[width=0.25\textwidth, height=0.125\paperheight]{figures/f2_quad2} 
+    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{F2s_comparison2}
   \end{tabular}
@@
   }
+\end{figure}
+
+  \begin{figure}[H]
+    \centering
+    \begin{tabular}{cc}
+    \includegraphics[width=0.25\textwidth, height=0.125\paperheight]{figures/f1_quartic2} &
+    \includegraphics[width=0.25\textwidth, height=0.125\paperheight]{figures/f2_quad2}
+    \end{tabular}
+    \caption{(Left) Squared $F_1$ hybrid phenotypes from Figure \ref{fig:hybs} diverge quartically, and (right) $F_2$ hybrid phenotypes diverge quadratically as a function of time in allopatry.}
   \end{figure}
@@
 %We work with a population of effective size $N_e$.
-\js{commented out questions here b/c we don't answer them directly -- what do you think?}
 
@@
 we will write the ``distance'' between phenotypes $\phi$ and $\psi$ as $\dph(\phi, \psi)$,
-measured so that the fitness of an organism with coefficents $x$ is $\fitx(x) = \exp(-\dph(\Phi(x),\optph)^2)$. \js{need to add $\gamma^{2}$ here?}
+measured so that the fitness of an organism with coefficents $x$ is $\fitx(x) = \exp(-\dph(\Phi(x),\optph)^2)$. 
 We will assume that the map $\Phi$ is smooth
@@
 the population mean coefficient will move a random distance of size $\sigma/\sqrt{N_e}$ per generation,
-\js{since it moves at rate $\sigma \sqrt{T/N_e}$ isn't saying  $\sigma/\sqrt{N_e}$ \emph{per generation} not that clear?}
 simply because this is the the standard deviation of the mean of a random sample \citep{lande_drift}.
@@
 \paragraph{Hybridization}
-\plr{I put F1s and F2s here, and $\Incompat$, but with less explanation than below. Merge somehow.} 
 The means of two allopatric populations separated for $T$ generations
@@
 and variance equal to $\sigma^2$. 
-\js{I think you mean $\sigma^{2}$? I changed it.}
 Each $F_2$ hybrid will be homozygous for one parental allele on averge at half of the loci in the genome,
@@
 but variance equal to $\sigma^2 + z^2/2$, where $z$ is the distance between the parental populations. 
-\js{Can you clarify why $z^{2}/2$? Is this because it's averaging two distributions?}
 These are depicted in figure \ref{fig:conceptual_fig}.
@@
 
-If we prefer not to make any assumptions about the shape of the fitness function, we can more conservatively state that hybrid phenotypes diverge from their optimal parents at rates $d(\Phi_{F_1}, \Phi_0) = \alpha_1 \sigma_N^2 T/N_e$ and $d(\Phi_{F_2}, \Phi_0) = \alpha_2 \sigma_N \sqrt{T/N_e}$, where $\alpha_1$ and $\alpha_2$ are constants.
+If we prefer not to make any assumptions about the shape of the fitness function, we can more conservatively state that hybrid phenotypes diverge from their optimal parents at rates $d(\Phi_{F_1}, \Phi_0) = c_1 \sigma_N^2 T/N_e$ and $d(\Phi_{F_2}, \Phi_0) = \gamma^{-1} \sigma_N \sqrt{T/N_e}$.
 Whether or not hybrid phenotypic divergence is associated with a high fitness cost, $F_2$ phenotypes should diverge much faster than $F_1$s, with a rate proportional to the square root of $T/N_e$, versus linearly.
-\js{I think $\alpha_1$ is your $c_1$ and $\alpha_2$ is a function of $\gamma^{-1}$?}
 
@@
 Here, we offer a tractable and simple model which includes these missing features. 
-\plr{this is a claim we are the first to do something.  best not to make these claims (and we aren't the first to make math models of these)}\js{removed the word ``only''}
 Further, we provide, in clear mathematical language, an analytical description of phenomena hitherto discussed verbally and conceptually 
@@
 if desired, 
-\plr{probably not desired}\js{:p}
 the dynamics of a genetic system, as well as perturbations to the system -- an attribute not likely to be found in less tractable models and simulations.

### 2017-10-03 kalman_walk_paper.tex
edited envelope calc

@@
 \usepackage{amsmath, amssymb, color, xcolor, amsthm}
-\usepackage{graphicx, wrapfig, float, caption, dsfont, bbm}
+\usepackage{graphicx, wrapfig, float, caption, dsfont, bbm, xfrac}
 \usepackage{fullpage}
@@
 
-\begin{figure}
+\begin{figure}[H]
 \centering
@@
 
-If we prefer not to make any assumptions about the shape of the fitness function, we can more conservatively state that hybrid phenotypes diverge from their optimal parents at rates $d(\Phi_{F_1}, \Phi_0) = c_1 \sigma_N^2 T/N_e$ and $d(\Phi_{F_2}, \Phi_0) = \gamma^{-1} \sigma_N \sqrt{T/N_e}$.
-Whether or not hybrid phenotypic divergence is associated with a high fitness cost, $F_2$ phenotypes should diverge much faster than $F_1$s, with a rate proportional to the square root of $T/N_e$, versus linearly.
+Note that this follows directly from the fact that $F_1$ and $F_2$ hybrid phenotypes diverge from $F_0$ phenotypes, linearly and inverse-quadratically, with respect to time (more precisely $T/N_e$).
+As such, removing assumptions about the form of the fitness function, this model still predicts hybrid $F_2$ phenotypes to diverge much faster than $F_1$s, initially.
 
@@
 \paragraph{Speciation rates under neutrality}
-For cases where system structures and $\gamma$ can be inferred,
-specific speciation rates can be estimated.
-For other situations we offer a back of the envelope calculation.
-
-Consider a population that is distributed on $\mathcal{X}$ such that the magnitude of phenotypically
-divergent $\sigma_S$ and neutral $\sigma_N$ diversity are constrained by selection;
-$\sigma_S$ directly and $\sigma_N$ indirectly via recombination load.
-Two reproductively isolated populations will, on average, be a distance of $2 \sigma_N \sqrt{T/N_e}$.
-If we assume that $\sigma_N$ is constrained by a function of $\sigma_S$,
-neutral movement along $\mathcal{X}$ pushes hybrids off of the manifold by $\sigma_S$ per $\sigma_N$,
-Lastly, if fitness is Gaussian: $\fit_2 = \exp\left(- d(x, x')^2/\beta^2\right)$,
-and we assume that the strength of selection is only several multiples of the segregating divergent diversity,
-such that $\sigma_S = c \beta$, we can simplify the expression estimating $F_2$ hybrid fitness drops
-to be,
-  \begin{align}
-    \exp\left(-c^2 \frac{T}{N_e} \right)
-  \end{align}
+%For cases where system structures and $\gamma$ can be inferred,
+%specific speciation rates can be estimated.
+%For other situations we offer a back of the envelope calculation.
+%
+%Consider a population that is distributed on $\mathcal{X}$ such that the magnitude of phenotypically
+%divergent $\sigma_S$ and neutral $\sigma_N$ diversity are constrained by selection;
+%$\sigma_S$ directly, such that $\sigma_S = \gamma$, and $\sigma_N$ indirectly via recombination load.
+%Two reproductively isolated populations will, on average, be a distance of $2 \sigma_N \sqrt{T/N_e}$.
+%If we assume that $\sigma_N$ is constrained by a function of $\sigma_S$,
+%neutral movement along $\mathcal{X}$ pushes hybrids off of the manifold by $\sigma_S$ per $\sigma_N$,
+%Lastly, if fitness is Gaussian: $\fit_2 = \exp\left(- d(x, x')^2/\beta^2\right)$,
+%and we assume that the strength of selection is only several multiples of the segregating divergent diversity,
+%such that $\sigma_S = c \beta$, we can simplify the expression estimating $F_2$ hybrid fitness drops
+%to be,
+%
 %
@@
 %  where $\xi := 2\xi_2/\xi_1$.
-  This suggests that, as an upper bound, if a population contains as much phenotypic diversity as allowed by selection, reproductively isolated populations can speciate rapidly -- possibly on timescales shorter than $N_{e}$ generations (It is likely that, on average, $c \leq 2/3$).
+  If we assume that perpendicular variation segregating within a population is constrained by selection such that $\sigma_S/\gamma = \xi$, and that neutral variation is on the order of perpendicular variation, such that $\sigma_N = \sigma_S$, then the typical distance between two individuals drawn from separate allopatric populations will be $z = 2 \sigma_N \sqrt{T/N_e}$.
+We will expect a drop in hybrid fitness to be,
+  \begin{align}
+  \fit_2/\fit_0 = \exp \left(- 4\xi^2 \frac{T}{N_e} \right) .
+  \end{align}
+  Hybrid fitness depression will be significant when $z > \gamma$. 
+  If $\xi$ is in $[\sfrac{1}{5},1]$, speciation will occur on the order of $N_e$ generations, and possibly much faster.
+Note, however, that if intrapopulation variation is very small such that $\sigma_S \ll \gamma$, speciation rates will be slower.
 %%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%
-
-%    \subsection*{Metabolic Network}
-%  \begin{align*}
-%    \dot{x}(t) &= \begin{bmatrix} 0 & 1 & 0 & -1 \\ 0 & 0 & -1 & -1 \\ -1 & 1 & 0 & 0 \\ 0 & 0 & 0 & 0 \end{bmatrix} \begin{bmatrix} \text{GAL3} \\ \text{GAL4} \\ \text{GAL80} \\ \text{MIG1} \end{bmatrix}(t) \\ &+ \begin{bmatrix} 1 & 0 & 0 \\ 0 & 0 & 1 \\ 0 & 0 & 0 \\ 0 & 1 & 1 \end{bmatrix} \begin{bmatrix} \text{galactose} \\ \text{glucose} \\ \text{BTM} \end{bmatrix}(t) \\
-%        y(t) &= \begin{bmatrix} 0 & 1 & 0 & -1 \end{bmatrix} \vec{x}(t)
-%    \end{align*}
-%    \begin{align*}
-%      H(z) &= \begin{bmatrix} 1 \\ z^{2} + z + 1 \\ z + 1 \end{bmatrix} \frac{1}{z^{3} + z -1}
-%    \end{align*}
-%
-%    \begin{align*}
-%      CV &= C \\
-%      VB &= B \\
-%      V :&= \begin{bmatrix} 1 & 0 & a & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & b & 0 \\ 0 & 0 & 0 & 1 \end{bmatrix} , V^{-1} = \begin{bmatrix} 1 & 0 & \frac{-a}{b} & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & \frac{1}{b} & 0 \\ 0 & 0 & 0 & 1\end{bmatrix} \\
-%        A(a,b) :&= VAV^{-1} = \begin{bmatrix} -a & a+1 & \frac{a^{2}}{b} & -1 \\ 0 & 0 & \frac{-1}{b} & -1 \\ -b & b & a & 0 \\ 0 & 0 & 0 & 0 \end{bmatrix}
-%    \end{align*}
-%    The GAL regulatory network in \emph{S.
-cerevisiae} is modelled here.
-The activating transcription factor GAL4, which is regulated by the presence of galactose and two other TFs, binds to the promoter region of the GAL regulon -- a cluster of genes regulated by the same promoter encoding 3 enzymes (GAL1, GAL7, and GAL10), for the metabolism of the sugar galactose.
-The repressing transcription factor MIG1, which is activated by the presence of glucose and other proteins omitted here, also binds to the regulon, preventing the expression of the galactose metabolizing enzymes.
-This network is one of the most studied gene regulatory networks in yeast, and experiments have already demonstrated significant transcriptional variation among different species and genuses of yeast.
-%
-%    In the present model, the $A$ matrix encodes the interactions among GAL3, GAL4, GAL80, and MIG1.
-$B$ interprets the input: the quantity of glucose, galactose, and the basal transcription (BTM) rates impact of TF concentrations.
-$C$ represents the promoter region of the GAL regulon and is the sum of the influence of GAL4 (activating) and MIG1 (repressing).
-%
-%    $A(a,b)$ represent alternative network structures that produce identical outputs, with identical input and output transformations -- meaning the regulatory contributions of the sugars and the basasl machinery, as well as the promoter of the GAL regulon are held constant.
-These interactions may be realized in nature, if the regulatory differences predicted by $A(a,b)$ are biologically/mutationally realizable.
-Mutations influencing the interactions between GAL4 and GAL80 have been experimentally demonstrated \citep{li2010alterations}.
-%
-%    \subsection*{Gap Gene Network}
-
 %%%%%%%%%%%%%%%%%%%%%
@@
 %Although we leave this question unexplored, this phenomena may lead to insights on evolvability and developmental innovation. 
-Lastly, we show that hybrid gene networks break down as function of genetic distance, and may, in part, explain broad patterns of reproductive isolation among diverse phyla \citep{roux2016shedding}.
+Lastly, we show that hybrid gene networks, under neutral processes, break down as function of genetic distance, and may, in part, explain broad patterns of reproductive isolation among diverse phyla \citep{roux2016shedding}.
 

### 2017-10-03 kalman_walk_paper.tex
edited envelope calc

@@
 %  where $\xi := 2\xi_2/\xi_1$.
-  If we assume that perpendicular variation segregating within a population is constrained by selection such that $\sigma_S/\gamma = \xi$, and that neutral variation is on the order of perpendicular variation, such that $\sigma_N = \sigma_S$, then the typical distance between two individuals drawn from separate allopatric populations will be $z = 2 \sigma_N \sqrt{T/N_e}$.
-We will expect a drop in hybrid fitness to be,
+  Assume that perpendicular variation segregating within a population is constrained by selection such that $\sigma_S/\gamma = \xi$, and that neutral variation is on the order of perpendicular variation, such that $\sigma_N = \sigma_S$.
+Since the typical distance between two individuals drawn from separate allopatric populations will be $z = 2 \sigma_N \sqrt{T/N_e}$, we expect a drop in hybrid fitness of
   \begin{align}

### 2017-10-03 kalman_walk_paper.tex
edited envelope calc and kalman

@@
 %
-%Any linear and minimal gene network 
-%-- minimal, informally meaning that the system's phenotype is achieved with the fewest possible genes ($n=\min$, see \ref{apx:kalman})
-%-- has an identical impulse response and thus phenotype given identical inputs $u(t)$, up to a change of coordinates. 
-%  \begin{equation}
-%    \begin{aligned}
-%      h(t) &= C e^{A t} B 
-%      = C V^{-1} V e^{A t} V^{-1} V B \\
-%      &= C V^{-1} e^{V A V^{-1} t} V B 
-%      = \bar{C} e^{\bar{A} t} \bar{B}
-%    \end{aligned}
-%  \end{equation}
-%Two gene networks, $\mathcal{S} = ( A, B, C )$, and $\bar{\mathcal{S}} = (\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} )$, 
-%have the same phenotype, for all possible inputs, if and only if they are related by a change of coordinates. 
+Any linear and minimal gene network 
+-- minimal, informally meaning that the system's phenotype is achieved with the fewest possible genes ($n=\min$, see \ref{apx:kalman})
+-- has an identical impulse response and thus phenotype given identical inputs $u(t)$, up to a change of coordinates. 
+  \begin{equation}
+    \begin{aligned}
+      h(t) &= C e^{A t} B 
+      = C V^{-1} V e^{A t} V^{-1} V B \\
+      &= C V^{-1} e^{V A V^{-1} t} V B 
+      = \bar{C} e^{\bar{A} t} \bar{B}
+    \end{aligned}
+  \end{equation}
+Two gene networks, $\mathcal{S} = ( A, B, C )$, and $\bar{\mathcal{S}} = (\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} )$, 
+have the same phenotype, for all possible inputs, if and only if they are related by a change of coordinates. 
 %
@@
 %
-%Therefore varying $V$ can tweak the relationships between genes within a regulatory network, yet preserve the phenotype.
+Therefore varying $V$ can tweak the relationships between genes within a regulatory network, yet preserve the phenotype.
 Some gene networks, however, can grown or shrink, perhaps following gene duplications and deletions, and also still preserve their phenotypes.
-These changes cannot be described by $\mathcal{S}(V)$, but are ecanpuslated by \ref{eqn:equivalence}, below.
+These changes cannot be described by $\mathcal{S}(V)$.
 %The Kalman decomposition decomposes a system into \emph{reachable} and \emph{observable} components -- that is coefficients that respond to an input and coefficients that produce an observable output.
@@
 
-\begin{definition}[Phenotypic equivalence of systems]
-    Let $(\kappa(t),\phi(t))$ and $(\bar \kappa(t),\bar \phi(t))$ be the solutions to \eqref{eqn:system}
-    with coefficient matrices $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ respectively,
-    and both $\kappa(0)$ and $\bar \kappa(0)$ are zero. 
-    The systems defined by $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ are
-    \textbf{phenotypically equivalent} 
-    if
-    \begin{align*}
-        \phi(t) = \bar \phi(t) \qquad \text{for all} \; t \ge 0.
-    \end{align*}
-    Equivalently, this occurs if and only if
-    \begin{align*}
-        h(t) = \bar h(t)  \qquad \text{for all} \; t \ge 0,
-    \end{align*}
-    where $h$ and $\bar h$ are the impulse responses of the two systems.
-\end{definition}
-
-One way to find other systems equivalent to a given one
-is by change of coordinates:
-if $V$ is an invertible matrix, then the systems $(A,B,C)$ and $(VAV^{-1},VB,CV^{-1})$
-have the same dynamics because their impulse responses are equal:
-  \begin{equation}
-    \begin{aligned}
-      h(t) &= C e^{A t} B 
-      = C V^{-1} V e^{A t} V^{-1} V B \\
-      &= C V^{-1} e^{V A V^{-1} t} V B 
-      = \bar{C} e^{\bar{A} t} \bar{B}
-    \end{aligned}
-  \end{equation}
+%\begin{definition}[Phenotypic equivalence of systems]
+ %   Let $(\kappa(t),\phi(t))$ and $(\bar \kappa(t),\bar \phi(t))$ be the solutions to \eqref{eqn:system}
+  %  with coefficient matrices $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ respectively,
+   % and both $\kappa(0)$ and $\bar \kappa(0)$ are zero. 
+   % The systems defined by $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ are
+   % \textbf{phenotypically equivalent} 
+   % if
+   % \begin{align*}
+   %     \phi(t) = \bar \phi(t) \qquad \text{for all} \; t \ge 0.
+   % \end{align*}
+   % Equivalently, this occurs if and only if
+   % \begin{align*}
+   %     h(t) = \bar h(t)  \qquad \text{for all} \; t \ge 0,
+   % \end{align*}
+   % where $h$ and $\bar h$ are the impulse responses of the two systems.
+%\end{definition}
+
+%One way to find other systems equivalent to a given one
+%is by change of coordinates:
+%if $V$ is an invertible matrix, then the systems $(A,B,C)$ and $(VAV^{-1},VB,CV^{-1})$
+%have the same dynamics because their impulse responses are equal:
+%  \begin{equation}
+%    \begin{aligned}
+%      h(t) &= C e^{A t} B 
+%      = C V^{-1} V e^{A t} V^{-1} V B \\
+%      &= C V^{-1} e^{V A V^{-1} t} V B 
+%      = \bar{C} e^{\bar{A} t} \bar{B}
+%    \end{aligned}
+%  \end{equation}
 %\begin{align*}
@@
 %\end{align*}
-However, the converse is not necessarily true: 
-systems can have identical impulse responses without being changes of coordinates of each other.
-In fact, systems with identical impulse responses can involve interactions between different
-numbers of molecular species, and thus be in different dimensions.
+%However, the converse is not necessarily true: 
+%systems can have identical impulse responses without being changes of coordinates of each other.
+%In fact, systems with identical impulse responses can involve interactions between different
+%numbers of molecular species, and thus be in different dimensions.
 

### 2017-10-12 kalman_walk_paper.tex
edits from overleaf

--- 2017-10-03 kalman_walk_paper.tex
@@
 \newif\ifmargincomments
-\margincommentstrue
+\margincommentsfalse
 % \margincommentstrue
@@
     \newcommand{\plr}[1]{{\color{blue}\it #1}}
+    \newcommand{\js}[1]{{\color{green}\it #1}}
     \newcommand{\plri}[1]{\plr{#1}}
@@
 \begin{abstract}
-We introduce an analytical theory to study the evolution of biological systems, such as gene regulatory networks.
-The evolutionary conservation of phenotype under selective and environmental stasis does not necessitate conservation of the underlying mechanism, as distinct molecular pathways can realize identical phenotypes.
-Here we give an exact expression for the set of all linear mechanisms with identical phenotypes, and expect evolution under neutrality to explore this set.
-We employ a quantitative genetic approach to model evolution under neutrality as a random process over the set of all phenotype-invariant mechanisms: only mutational tweaks to the pathway that leave the phenotype invariant are optimally fit.
-We show that there is never a unique linear system architecture for any phenotype and that the evolutionary exploration of these distinct and mutationally connected mechanisms can lead to the rapid accumulation of hybrid incompatibilities between allopatric populations and thus lead to the rapid formation of new species -- in fewer generations than there are breeding individuals in a population.
+It is known that even if a species' phenotype has remained unchanged over evolutionary time,
+the underlying mechanism may have changed,
+since distinct molecular pathways can realize identical phenotypes.
+In this paper, we use quantitative genetics and linear systems theory to study
+how a gene network underlying a conserved phenotype
+evolves as genetic drift of small mutational tweaks to the molecular pathways
+cause a population to explore the set of mechanisms with identical phenotypes.
+In this setting we treat organisms as ``black boxes'' for which the environment provides input
+and the phenotype is the output,
+and there exists an exact characterization of the set of all mechanisms that give the same input--output relationship. 
+We show that in this situation, there is never a unique architecture for any phenotype
+and that the evolutionary exploration of these distinct and mutationally connected mechanisms
+can lead to the rapid accumulation of hybrid incompatibilities between allopatric populations.
+We estimate that in reasonably numerous species,
+this process could thus the formation of new species
+over a number of generations proportional to the effective population size.
 \end{abstract}
@@
 \begin{itemize}
+    \item speciation literature
     %\item Haldane's rule (easy point in discussion: makes F1s look like F2s)
@@
 \section*{Introduction}
-Bridging the gulf between an organism's genome and phenotype is a poorly understood and complex molecular machinery. 
-Progress in a suite of biological subdisciplines is stalled by our general lack of understanding of this molecular machinery: with respect to both its function and evolution. 
-There does exist a growing body of data on the evolutionary histories and molecular characterizations of particular gene regulatory networks \citep{jaeger2011gap, davidson2006gene, israel2016comparative}, as well as thoughtful verbal and conceptual models \citep{true2001developmental, gwagner1, weiss2000phenogenetic, edelman2001degeneracy}.
-However, verbal theories are often insufficient, if not downright misleading \citep{servedio2014not}. 
-This is especially pertinent given the staggering complexity and scope of contemporary research programs. 
-This outlook necessitates the advancement of conceptual frameworks of such precision, 
-only mathematics will suffice, as models allow the development of concrete numerical predictions. 
-%Previously it has been suggested that any idealized study of evolution is incomplete 
-%without a mathematically sufficient description of the genotype, phenotype, and transformation from one to the other \citep{Lewontin1974genetic}.
-\plr{probably some more recent Wagner papers in this line as well?}
-%Here we outline an analytical theory to study the evolution of biological systems, borrowing insight and methods from control engineering, systems identification, and dynamical systems theory. 
-%
-%Presently, we focus on the neutral evolution of genetic regulatory networks.
-That is, we analytically describe the set of all linear gene networks (of any size) that produce identical phenotypes -- and the evolutionary paths connecting them.
-In the idealized case of a perfectly adapted population, constant selection, and a static environment, we observe evolution under the ``conservation of phenotype'' as a Brownian motion over phenotypically-invariant network-space.
-This analysis provides insight to the mechanisms and parameters important for understanding developmental systems drift, network rewiring, evolvability, epistasis, and speciation, as well as the tenuous connection between network architecture and function.
-%
-%
-%It is commonly taught that an organism's genome contains the heritable material 
-%that natural selection filters and that an organism's phenotype directly determines its
-%evolutionary fitness. Between genotype and phenotype is an often complicated and poorly
-%undrstood molecular machinery -- and it is a major goal common to many different disciplines
-%within the life sciences to elucidate its form, function, and evolution. These aims are
-%delicately intertwined and a comphrehensive understanding of a system's evolution requires
-%an understanding of its function and vice versa.
-
-The molecular machinery, interacting with the environment, and bridging genotype to phenotype
-can be mathematically described as a dynamical system -- or a system of differential equations \citep{jaeger2015comet}.
+
+A complex molecular machinery translates an organism's genome into the characteristics on which natural selection acts,
+her phenotype.
+% Bridging the gulf between an organism's genome and phenotype is a poorly understood and complex molecular machinery. 
+Attaining a general understanding of the functioning and evolution of this molecular machinery
+is an overarching goal of many subdisciplines of biology.
+For example, 
+there is a growing body of data on evolutionary histories and molecular characterizations of particular gene regulatory networks
+\citep{jaeger2011gap, davidson2006gene, israel2016comparative}, 
+as well as thoughtful verbal and conceptual models \citep{true2001developmental, gwagner1, weiss2000phenogenetic, edelman2001degeneracy}. 
+Mathematical models of both particular regulatory networks
+and the evolution of such systems in general
+can provide guidance where intuition fails,
+and thus has the potential to discover general principles in the organization of biological systems 
+and provide concrete numerical predictions \citep{servedio2014not}.
+
+The dynamics of the molecular machinery and its interactions with the environment
+can be mathematically described in various ways as a dynamical system \citep{jaeger2015comet}.
 Movement in this direction is ongoing, as researchers have begun to study 
-the evolution of both abstract \citep{wagner1994evolution, wagner1996does,  siegal2002waddington, bergman2003evolutionary, draghi2015robustness} and empirically inspired computational and mathematical models of gene regulatory networks \citep{mjolsness1991connectionist, jaeger2004dynamic, maria1, vitaly1, vitaly2, crombach2016gap, wotton2015quantitative, chertkova2017insilico}.
-If we allow the reasonable assumption that the genotype-phenotype map can be represented as a system of differential equations, we can immediately discuss its evolution and function in a much more mechanistic, yet general, manner.
-
-In some fields that seek to fit parametric models to experimental data, such as control
-theory, chemical engineering, and statistics, it is well known that mathematical models
-can fundamentally be \emph{unidentifiable} and/or \emph{indistinguishable} -- meaning that 
+the evolution of both abstract \citep{wagner1994evolution, wagner1996does, siegal2002waddington, bergman2003evolutionary, draghi2015robustness} 
+and empirically inspired computational and mathematical models of gene regulatory networks,
+\citep[e.g.][]{mjolsness1991connectionist, jaeger2004dynamic, maria1, vitaly1, vitaly2, crombach2016gap, wotton2015quantitative, chertkova2017insilico}.
+% If we allow the reasonable assumption that the genotype-phenotype map 
+% can be represented as a system of differential equations \citep{jaeger2015comet}, 
+% we can immediately discuss its evolution and function in a much more mechanistic, yet general, manner. 
+It is well known that in many contexts mathematical models
+can fundamentally be \emph{nonidentifiable} and/or \emph{indistinguishable} -- meaning that 
 there can be uncertainty about an inferred model's parameters or even its claims about
 causal structure, even with access to complete and perfect data \citep{bellman1970structural, grewal1976identifiability, walter1984structural}. 
-%\plr{not clear what you mean by "actually"}\js{I think we discussed it's clear enough for an intro?}
-Models with different parameter schemes, or even different mechanics can be equally accurate, 
-but still not \emph{actually} agree with what is being modelled. 
+Models with different parameter schemes, or even different mechanics 
+can equally accurately predict the observed behavior of a given physical system, 
+but still not actually reflect the internal dynamics of the system.
 In control theory, where electrical circuits and mechanical systems are often the focus, 
 it is understood that there can be an infinite number of ``realizations'', 
-or ways to reverse engineer the dynamics of a black box,
-even if all possible input and output experiments on the black box are performed \citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. 
-In chemical engineering, those who study chemical reaction networks sometimes refer to the fundamental
-unidentifiability of these networks as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}. 
+or ways to reverse engineer the dynamics of a ``black box'',
+even if all possible input and output experiments on the ``black box'' are performed 
+\citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. 
+The fundamental nonidentifiability of chemical reaction networks is sometimes referred to as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}. 
 In computer science, this is framed as the relationship among processes that simulate one another \citep{van2004equivalence}.
-Although this may frustrate the occasional engineer or scientist, viewed from another angle,
-the concepts of unidentifiability and indistinguishability can provide a starting point for
-thinking about externally equivalent systems
+Finally,
+the field of \emph{inverse problems} studies those cases where,
+even if a one-to-one mapping between model and behavior is possible in theory,
+even tiny amounts of noise can make inference problems nonidentifiable in practice.
+
+Although nonidentifiability may frustrate the occasional engineer or scientist, viewed from another angle,
+these concepts can provide a starting point for thinking about externally equivalent systems
 -- systems that evolution can explore, so long as the parameters and structures can be realized biologically.
-In fact, evolutionary
-biologists who study convergent versus parallel evolution, and developmental homology and analogy, are very familiar with such functional symmetries; 
+These functional symmetries manifest in convergent and parallel evolution, 
+as well as \emph{developmental systems drift}: the observation that
 macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level 
-\citep{true2001developmental, tsong2006evolution, hare2008sepsid, vierstra2014mouse, stergachis2014conservation, taylor2016diverse, matsui2015regulatory}.
-
-In this paper we outline a theoretical framework to study the evolution of biological systems, such as gene regulatory networks. 
-We focus primarily on the neutral scenario: 
-that is where phenotype is conserved over evolutionary time. 
-%however, this framework could be applied to a wider set of evolutionary scenarios. 
-We present an analytical description of the set of all linear biological systems with identical phenotypes 
--- that is we describe the set of all gene network architectures that yield identical phenotypes in equivalent environments, and show that all linear biological systems can, in principal, can undergo systems drift and rewire.
-Under neutrality, a population will drift through the set of all possible phenotypically equivalent linear gene networks.
-Consequentially, two phenotypically equivlanet yet allopatrically isolated populations, will likely produce incompatibile hybrids, despite the absence of adaptation, directional selection, or environmental change.
-Under these conditions, speciation typically occurs on timescales approximately on the order of $N_{e}$ generations, where $N_{e}$ is the effective population size.
-%
-\section*{Gene networks as linear dynamical systems}
-%  \jss{Organisms' phenotypes are constructred by gene by gene by environment interactions.
-Here we simply define the phenotype to be the organismal temporal molecular dynamics directly under natural selection.
-The \emph{what}, \emph{when}, and \emph{where}, of an organism's molecules that are physiologically or otherwise relevant for survival.
-%   Thus we say that some function $\phi(t)$ is a phenotype where, 
-%  \begin{align}
-%    \phi(t) = \int_{0}^{\infty} h(t) u(t) dt  ,
-%  \end{align}
-%  and $h(t)$ is the \emph{impulse response} of the system and $u(t)$ is the \emph{input} function, both functions of time $t$.
-The input can be interpreted as the environment, as initial conditions, or otherwise, depending on the biological specifics under study.
-The phenotype is a convolution of both the impulse response and the input, as allowed by our assumption of linearity.
-%
-%  Essentially the phenotype $\phi(t)$ is a consequence of an organism's specific gene by gene interactions, given by $h(t)$, reacting further with the local environment, given by $u(t)$. 
-%
-%  We describe the impulse response as, 
-%  \begin{align}
-%    h(t) = C e^{A t} B
-%  \end{align}
-%  where $A$ is a gene regulatory network (although this model can be generalized to other biological networks) -- a square matrix, and $B$ filters and translates the input to the system.
-The form of $B$ determines precisely how the state of the external environment influences the internal gene network.
-$C$ filters and translates the dynamics of the system and precesily determines the output, that is, what is visible to selection.
-%
-%  Generally $A$ can be any real $n \times n$ matrix, $B$ any $n \times \ell$, and $C$ any $\ell \times n$ dimensional matrix.
-Each $i$th row of $A$ describes the \emph{cis}-regulatory module for gene $i$, and each $j$th entry, the specific regulatory influence of gene $j$ on gene $i$.
-%
-%  Although $\phi(t)$ describes the phenotype given an input, $h(t)$ describes the phenotype subject only to an impulse -- an input present initially and absent immediately thereafter.
-Typically, a system $\Sigma$, is defined as,
-%  \begin{align}
-%    \Sigma = \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right.
-%  \end{align}
-%  Variables have the same identities as described above and $\kappa(t)$ is a vector of molecule concentrations at time $t$.
-Therefore the molecular concentrations at a specific time are completely determined by the input and gene by gene interactions.
-Lastly, a portion and/or combination of these molecules, $\phi(t)$, are ``observed'' by selection (this is in constrast to $\kappa(t)$ -- the \emph{kryptotype} -- as it is ``hidden'' from direct selection).
-%}
-\paragraph{Systems theory}
-Organisms' phenotypes are constructred by gene by gene and gene by environment interactions.
-Here we simply define the \emph{phenotype} to be the temporal molecular dynamics directly under natural selection.
-The \emph{what}, \emph{when}, and \emph{how much}, of an organism's molecules that are physiologically or otherwise relevant to survival.
-
-Thus an organism's phenotype $\phi(t)$ -- a vector of molecular concentrations at time $t$ -- is determined both by the structure and organization of a biological system (\emph{e.g.} a gene regualtory network), given by the triple $(A,B,C)$,
-and by its environment $u(t)$.
-
-Such a biological system $\mathcal{S}$ is given by,
-  \begin{align}\label{eqn:system}
-    \mathcal{S} := \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right.
-  \end{align}
-
-Generally $A$ is any real $n \times n$ matrix, $B$ any $n \times \ell$, and $C$ any $\ell \times n$ dimensional matrix.
-Although many different biological systems can be modeled with this approach, for clarity, we focus on gene regulatory networks.
-%
-Each $i$th row of $A$ describes the \emph{cis}-regulatory module for gene $i$, and each $j$th entry, the specific regulatory influence of gene $j$ on gene $i$. 
-%The $i^\text{th}$ row of $A$ contains the regulatory logic pertaining to gene $i$.
-%
-As such, $A_{ij}$ is the magnitude at which transcription factor $\kappa_j(t)$ regulates transcription factor $\kappa_i(t)$, and if $A_{ij} > 0$, we say that $\kappa_j$ upregulates $\kappa_i$.
-If $A_{ij} < 0$, we say that $\kappa_j$ down-regulates $\kappa_i$.
-
-The form of $B$ determines precisely how the environment influences the organism -- that is $B$ filters and translates the input to the system.
-$C$ filters and translates the dynamics of the system and precesily determines the output, that is, what is visible to selection.
-\emph{E.g.} for a metabolic system, $C_{ij}$ is the amount the $j$th metabolite $u_j$ affects the production of the $i$th enzyme.
-%
-
-  %where $A$ is a gene regulatory network (although this model can be generalized to other biological networks) -- a square matrix, 
-  %and $B$ filters and translates the input to the system.
-The form of $B$ determines precisely how the state of the environment influences the internal gene network.
-$C$ filters and translates the dynamics of the system and precesily determines the output, that is, what is visible to selection.
-%
-%
-Furthermore, whereas the phenotype is a subset of molecules \emph{visible} to selection, the \emph{kryptotype} includes the molecular dynamics \emph{hidden} from selection.
-That is $\kappa(t)$ is a vector of the system's molecular concentrations at time $t$.
-%
-%The structure of $B$ and $C$ determines how systems process their environment and convert krypto- to phenotypes, respectively. 
-%
-\plr{This "Thus" needs expanding, e.g., "the solution to this equation is unique and given by" with a reference.}
-
-Finally, we can write the phenotype as a convolution of the system organization and the environment,
+\citep{true2001developmental, tsong2006evolution, hare2008sepsid, vierstra2014mouse, stergachis2014conservation}.
+% are these good here? {taylor2016diverse, matsui2015regulatory}.
+
+EDIT IN THESE REFS?
+The literature is filled with detailed observations of molecular systems and their diversity. 
+\plr{Diversity doesn't imply systems drift -- only if the diverse systems are homologous.}
+There are examples of significant diversity in the networks underlying processes such as 
+circadian rhythm \citep{sancar2008intelligent}, 
+cell cycle control \citep{cross2011evolution, kearsey2003enigmatic}, 
+pattern formation \plr{cite?}, 
+and metabolism \citep{lavoie2009rearrangements, martchenko2007transcriptional, dalal2016transcriptional, christensen2011unique, hartl2007induction, alam2013aspergillus}.
+
+In this paper we outline a theoretical framework to study the evolution of biological systems
+that can be described by systems of differential equations, such as gene regulatory networks.
+We study the scenario
+where the optimal phenotype remains constant over evolutionary time,
+so that despite strong stabilizing selection for phenotype,
+neutral drift in the underlying genotype is possible.
+Results from systems theory provide
+an analytical description of the set of 
+all linear gene network architectures that yield identical phenotypes,
+which gives concrete expectations for how any such
+system can, in principal, undergo systems drift and rewire. 
+Even with stabilization selection, a population will explore the set of all possible phenotypically equivalent gene networks.
+Consequentially, two populations isolated for a sufficiently long period of time, 
+will likely produce inviable hybrids, despite the absence of adaptation, directional selection, or environmental change.
+% Under these conditions, speciation typically occurs on timescales approximately on the order of $N_{e}$ generations, where $N_{e}$ is the effective population size.
+
+\section*{Methods}
+
+We study an abstract model of temporal dynamics of the concentrations of a collection of $n$ coregulating molecules
+within an organism, that may also be affected by temporally varying signals from the environment.
+% Organisms' phenotypes are constructed by gene by gene and gene by environment interactions. 
+%Here we simply define the \emph{phenotype} to be those aspects of the dynamics directly under natural selection
+%-- the \emph{what}, \emph{when}, and \emph{how much}, of an organism's molecules that are physiologically or otherwise relevant to survival.
+We write $\kappa(t)$ for the vector of $n$ molecular concentrations at time $t$.
+There are also $m$ ``inputs'' determined exogenously to the system, denoted $u(t)$,
+and $\ell$ ``outputs'', denoted $\phi(t)$.
+The output is merely a linear function of the internal state:
+$\phi_i(t) = \sum_j C_{ij} \kappa_i(t)$
+for some matrix $C$.
+Since $\phi$ is what natural selection acts on, we refer to it as the \emph{phenotype}
+(meaning the ``visible'' aspects of the organism),
+and in contrast refer to $\kappa$ as the \emph{kryptotype},
+as it is ``hidden'' from direct selection.
+Although $\phi$ may depend on all entries of $\kappa$,
+it is usually of lower dimension than $\kappa$,
+and we tend to think of it as the subset of molecules relevant for survival.
+The dynamics are determined by
+the matrix of regulatory coefficients, $A$;
+a time-varying vector of inputs $u(t)$,
+and a matrix $B$ that encodes the effect of each entry of $u$ on the elements of the kryptotype.
+%Thus an organism's phenotype $\phi(t)$ -- a vector of molecular concentrations at time $t$ -- is determined both by the structure and organization of a biological system (\emph{e.g.} a gene regulatory network), given by the triple $(A,B,C)$,
+%and by its environment $u(t)$.
+The rate at which the $i^\text{th}$ concentration changes
+is a weighted sum of the concentrations of the other concentrations
+as well as the input:
+\begin{equation}\label{eqn:system}
+   \begin{aligned}
+    \dot{\kappa}(t) &= A \kappa(t) + B u(t) \\
+    \phi(t) &= C \kappa(t) .
+  \end{aligned} 
+\end{equation}
+Furthermore, we always assume that $\kappa(0) = 0$,
+so that the kryptotype measures deviations from initial concentrations. 
+Here $A$ can be any $n \times n$ matrix, $B$ a $n \times m$, and $C$ any $\ell \times n$ dimensional matrix,
+with usually $\ell$ and $m$ less than $n$.
+We think of the system as the triple $\mathcal{S} = (A,B,C)$,
+which translates (time-varying) $m$-dimensional input $u(t)$ into the $\ell$-dimensional output $\phi(t)$.
+Under quite general assumptions,
+we can write the phenotype as
   \begin{align}
@@
   \end{align}
-where we refer to $h(t) := Ce^{A t}B$ as the system's \emph{impulse response}. 
-%We will now lay out the model in more general terms.
-%Suppose that the \emph{internal state} of the system
-%is parameterized by the concentrations of a collection of $n$ molecular species,
-%$S_1, \ldots, S_n$,
-%and the vector of concentrations at time $t$ we denote $\kappa(t)=(\kappa_1(t),\ldots,\kappa_n(t))$.
-%There are also $m$ ``input'' species, whose concentrations are determined
-%exogenously to the system,
-%and are denoted $u(t) = (u_1(t),\ldots,u_m(t))$,
-%and $\ell$ ``output'' species, whose concentrations are denoted
-%$\phi(t) = (\phi_1(t),\ldots,\phi_\ell(t))$.
-%The output is merely a linear function of the internal state:
-%\begin{align*}
-%    \phi_i(t) = \sum_j C_{ij} \kappa_i(t).
-%\end{align*}
-%Since $\phi$ is what natural selection acts on, we refer to it as the \emph{phenotype},
-%and sometimes in contrast refer to $\kappa$ as the \emph{kryptotype},
-%as it is ``hidden'' from direct selection.
-%The rate at which the $i^\text{th}$ species is produced
-%is a weighted sum of the concentrations of the other species
-%as well as the input:
-%\begin{align*}
-%    \ddt \kappa_i(t) = \sum_j A_{ij} \kappa_j(t) + \sum_r B_{ir} u_r(t) .
-%\end{align*}
-%In matrix notation, this is written more concisely as
-%\begin{align} \label{eqn:lti_system}
-%    \ddt \kappa(t) &= A \kappa(t) + B u(t) \\
-%    \phi(t) &= C \kappa(t) .
-%\end{align}
- \begin{example}[Oscillating gene network: cell cycle control]\label{ex:oscillator}
-    Cellular division is governed by many different processes, however it is thought that its rhythm is partially controlled by oscillating gene transcription \citep{orlando2008global}.
-Here we consider a simplified model of oscillating gene transcription.
-
-    Suppose gene-2 up-regulates the transcription of gene-1 and that gene-1 down-regulates gene-2 with equal magnitudes, whose concentrations are given by $\kappa_1(t)$ and $\kappa_2(t)$.
-Furthermore, suppose that only the dynamics of gene-1 are consequential to the cell cycle (perhaps the amount of gene-1 activates another downstream gene network).
-Lastly suppose that the production of both genes is equally stimulated by an impulse of a molecule present immediately after division.
-
-    If the rate each of these genes is expressed is a linear function of their concentrations, the dynamics of the system is given by
-    \begin{alignat*}{2}
-      %\left\{ \begin{array}{ll} 
-      \dot{\kappa_{1}}(t) &= &&\kappa_{2}(t) + u(t) \\
-        \dot{\kappa_{2}}(t) &= - &&\kappa_{1}(t) + u(t) \\
-        \phi(t) &= &&\kappa_{1}(t)   .
-      %\end{array} \right.
-    \end{alignat*}
-Equivalently, in matrix form the system regulatory coefficients are given as,
-    $A \!=\!
-\left[\begin{smallmatrix} 0 & 1 \\ -1 & 0 \end{smallmatrix}\right] ,\ B \!
-=\!
-\left[\begin{smallmatrix} 1 \\ 1 \end{smallmatrix}\right] , \ C \!
-=\!
-\left[\begin{smallmatrix} 1 & 0 \end{smallmatrix}\right]$
+which is a convolution of the input $u(t)$ with the system's \emph{impulse response},
+which we denote as $h(t) := Ce^{A t}B$.
+
+Although many different biological systems can be modeled with this approach, for clarity, we focus on gene regulatory networks.
+In this interpretation, $A_{ij}$ determines how the $j^\text{th}$ transcription factor regulates the $i^\text{th}$ transcription factor.
+If $A_{ij} > 0$, then $\kappa_j$ upregulates $\kappa_i$, while if $A_{ij} < 0$, then $\kappa_j$ downregulates $\kappa_i$.
+The $i$th row of $A$ is therefore determined by genetic features such as
+the strength of $j$-binding sites in the promoter of gene $i$,
+factors affecting chromatin accessibility near gene $i$,
+or basal transcription machinery activity.
+The form of $B$ determines how the environment influences transcription factor expression levels,
+and $C$ might be the rate of production of a downstream enzyme
+(although other arrangements could be made).
+% \emph{E.g.} for a metabolic system, $C_{ij}$ is the amount the $j$th metabolite $u_j$ affects the production of the $i$th enzyme.
+
+{Linearity} also ``zero-state equivalence''
+Here we have assumed that the system is linear,
+and begins from the ``zero'' state ($\kappa(0)=0$).
+Of course, neither of these are necessarily true for real systems,
+but the dynamics of most nonlinear systems can be approximated locally by a linear systems near most points. 
+Furthermore, the ease of analyzing linear systems makes this an attractive place to start.
+
+\begin{example}[An oscillator]\label{ex:oscillator}
+For illustration, we consider an extremely simplified model of oscillating gene transcription,
+as for instance is found in cell cycle control or the circadian rhythm.
+% Cellular division is governed by many different processes, however it is thought that its rhythm is partially controlled by oscillating gene transcription \citep{orlando2008global}.
+Suppose there are two genes, 
+whose transcript concentrations are given by $\kappa_1(t)$ and $\kappa_2(t)$, 
+and that gene-2 upregulates gene-1 and that gene-1 downregulates gene-2 with equal strength.
+Furthermore, suppose that only the dynamics of gene-1 are consequential to the oscillator 
+(perhaps the amount of gene-1 activates another downstream gene network). 
+Lastly suppose that the production of both genes is equally upregulated by an exogenous signal.
+The dynamics of the system are described by
+    \begin{align*}
+      \dot{\kappa_{1}}(t) &= \kappa_{2}(t) + u(t) \\
+        \dot{\kappa_{2}}(t) &= - \kappa_{1}(t) + u(t) \\
+        \phi(t) &= \kappa_{1}(t)   .
+    \end{align*}
+In matrix form the system regulatory coefficients are given as,
+$A \!=\! \left[\begin{smallmatrix} 0 & 1 \\ -1 & 0 \end{smallmatrix}\right]$, 
+$B \! =\! \left[\begin{smallmatrix} 1 \\ 1 \end{smallmatrix}\right]$,
+and $C \!=\! \left[\begin{smallmatrix} 1 & 0 \end{smallmatrix}\right]$.
   %  \begin{align*}
@@
 
-      Since the input is simply an impulse, its phenotype is equivalent to its impulse response,
+      Suppose the input is an impulse at time zero (a delta function), and so its phenotype is equal to its impulse response,
       \begin{align*}
@@
       \end{align*}
-
+      The system and its dynamics are referred to in Figure \ref{fig:oscillator}. We return to the evolution of such a system below.   
     \begin{figure}[H]
@@
   %  \end{center}
-      \caption{(Left) Graphical representation of the cell cycle control gene network, and (right) plot of the phenotype $\phi(t)$ against time $t$.}
+      \caption{(Left) Graphical representation of the cell cycle control gene network, and (right) plot of the phenotype $\phi(t)$ against time $t$.} \label{fig:oscillator}
     \end{figure}
-      We return to the evolution of such a system below.
   \end{example}
 
-  \paragraph{Phenotypically equivalent gene networks}
-
-%\plr{This paragraph feels like the introduction.}
-%The literature is filled with detailed observations of molecular systems and their diversity. 
-%\plr{Diversity doesn't imply systems drift -- only if the diverse systems are homologous.}
-%There are examples of significant diversity in the networks underlying processes such as 
-%circadian rhythm \citep{sancar2008intelligent}, 
-%cell cycle control \citep{cross2011evolution, kearsey2003enigmatic}, 
-%pattern formation \plr{cite?}, 
-%and metabolism \citep{lavoie2009rearrangements, martchenko2007transcriptional, dalal2016transcriptional, christensen2011unique, hartl2007induction, alam2013aspergillus}. 
-Despite a symmetry in functionality or phenotype systems can often differ, sometimes substantially, at the molecular level. 
-How many different mechanisms have the same function? 
-Gene regulatory networks with identical phenotypes do not necessarily have identical kryptotypes.
-%
-%Any linear and minimal gene network 
-%-- minimal, informally meaning that the system's phenotype is achieved with the fewest possible genes ($n=\min$, see \ref{apx:kalman})
-%-- has an identical impulse response and thus phenotype given identical inputs $u(t)$, up to a change of coordinates. 
-%  \begin{equation}
-%    \begin{aligned}
-%      h(t) &= C e^{A t} B 
-%      = C V^{-1} V e^{A t} V^{-1} V B \\
-%      &= C V^{-1} e^{V A V^{-1} t} V B 
-%      = \bar{C} e^{\bar{A} t} \bar{B}
-%    \end{aligned}
-%  \end{equation}
-%Two gene networks, $\mathcal{S} = ( A, B, C )$, and $\bar{\mathcal{S}} = (\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} )$, 
-%have the same phenotype, for all possible inputs, if and only if they are related by a change of coordinates. 
-%
-%Although systems may not be identifiable beyond a change of coordinates, at present we are primarily interested in a subset of these systems. 
-%That is, systems that not only have equivalent external dynamics, but also equivalent input and output relationships. 
-%\plr{clarify that you mean actually having the same $B$ and $C$. (at first I read "input and output relationship" as the mapping $u \to \phi$),
-%    i.e. behaviour across many possible inputs.}
-%Formally, this means systems related by a change of coordinates (any invertible matrix $V$) that leaves $B$ and $C$ invariant:
-%  \begin{align}
-%    VB &= B \implies \bar{B} = B \\
-%    CV &= C \implies \bar{C} = C
-%  \end{align}
-%\plr{this sentence lacks an object (or some sentence part?)}
-%In other words systems with varying gene regulatory network architectures yet identical selection pressures, environment, and phenotype.
-%
-%Define $V(\tau)$ as the parameterized change of coordinates matrix that preserves $B$ and $C$, with $\tau$ a vector of free parameters.
-The set of \emph{all} phenotypically invariant (minimal) gene networks is,
- % \begin{align}
- %   A(\tau) = V(\tau) A(0) V^{-1}(\tau) ,
- % \end{align}
-%and a \emph{Linear Evolutionary System} is, 
-%\plr{why is this "evolutionary"? It is the thing that we're thinking about evolution of, but that doesn't make it evolutionary.}
-%
-%Therefore the set of all phenotypically identical (and minimal) systems can be parameterized as, $\mathcal{S}_{\min}(V) = (VAV^{-1}, VB, CV )$,
-  %  \begin{align} \label{eq:min_param}
-  %    \mathcal{S}(V) := \left\{ \begin{array}{ll} V\dot{\kappa}(t) &= VA V^{-1} \kappa(t) + VB u(t) \\ \phi(t) &= C V^{-1} \kappa(t) \end{array} \right.
-  %\end{align}
- % where $V$ is any invertible matrix and it's elements are free parameters. 
-%
+\subsection*{Equivalent gene networks}
+
+As reviewed above,
+some systems with identical phenotypes are known to differ, sometimes substantially, at the molecular level; 
+systems with identical phenotypes do not necessarily have identical kryptotypes.
+How many different mechanisms perform the same function?
+
 %Therefore varying $V$ can tweak the relationships between genes within a regulatory network, yet preserve the phenotype.
@@
 
-\begin{definition}[Phenotypic equivalence of systems]
-    Let $(\kappa(t),\phi(t))$ and $(\bar \kappa(t),\bar \phi(t))$ be the solutions to \eqref{eqn:system}
-    with coefficient matrices $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ respectively,
-    and both $\kappa(0)$ and $\bar \kappa(0)$ are zero. 
-    The systems defined by $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ are
+Two systems are equivalent if they produce the same phenotype given the same input,
+i.e., have the same input--output relationship.
+For any input $u(t)$,
+suppose that $(\kappa(t),\phi(t))$ and $(\bar \kappa(t),\bar \phi(t))$ are the solutions to equation \eqref{eqn:system}
+with coefficient matrices $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ respectively.
+% (recall that $\kappa(0) = \bar \kappa(0) = 0$). 
+We say that
+the systems defined by $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ are
 \textbf{phenotypically equivalent} 
@@
 \end{align*}
-    Equivalently, this occurs if and only if
-    \begin{align*}
-        h(t) = \bar h(t)  \qquad \text{for all} \; t \ge 0,
-    \end{align*}
-    where $h$ and $\bar h$ are the impulse responses of the two systems.
-\end{definition}
-
-One way to find other systems equivalent to a given one
+This occurs if and only if their impulse response functions are the same:
+if $h(t) = \bar h(t)$ for all $t \ge 0$.
+
+One way to find other systems phenotypically equivalent to a given one
 is by change of coordinates:
 if $V$ is an invertible matrix, then the systems $(A,B,C)$ and $(VAV^{-1},VB,CV^{-1})$
-have the same dynamics because their impulse responses are equal:
+are phenotypically equivalent because their impulse response functions are equal:
   \begin{equation}
@@
       &= C V^{-1} e^{V A V^{-1} t} V B 
-      = \bar{C} e^{\bar{A} t} \bar{B}
+      = \bar{C} e^{\bar{A} t} \bar{B} .
     \end{aligned}
   \end{equation}
-%\begin{align*}
-%    CV^{-1}( zI - VAV^{-1})^{-1}VB
-%    =
-%    CV^{-1}V( zI - A)^{-1}V^{-1}VB
-%    =
-%    C( zI - A)^{-1}B .
-%\end{align*}
-However, the converse is not necessarily true: 
-systems can have identical impulse responses without being changes of coordinates of each other.
+However, not all phenotypically equivalent systems are of this form:
+systems can have identical impulse responses without being coordinate changes of each other.
 In fact, systems with identical impulse responses can involve interactions between different
-numbers of molecular species, and thus be in different dimensions.
+numbers of molecules, and thus 
+have kryptotypes in
+different dimensions altogether.
+
+This implies that most systems have at least $n^2$ degrees of freedom,
+where recall $n$ is the number of components of the kryptotype vector.
+This is because for an arbitrary $n \times n$ matrix $Z$,
+taking $V$ to be the identity matrix plus a small perturbation in the direction of $Z$
+above implies that
+moving $A$ in the direction of $ZA - AZ$
+% (e.g., $\bar A = A + \epsilon(ZA-AZ)$)
+while also moving $B$ in the direction of $ZB$ 
+and $C$ in the direction of $-CZ$
+will leave the phenotype unchanged.
+If $A$ is invertible, for instance,
+then any such $Z$ will result in a different system.
+
+TALK ABOUT MINIMALITY
 
@@
 
+What about nonlinearity?
+
+Note that the drift above could actually be caused by the optimal phenotype shifting.
+
 %\plr{this paragraph isn't very clear} \js{Should I delete it? I guess it's sort of just a vague ``yeah we know this method has shortcomings''.}

### 2017-10-13 kalman_walk_paper.tex
minor edits from meeting

@@
 
-TALK ABOUT MINIMALITY
+It turns out that in general, there are more degrees of freedom,
+except if the system is \emph{minimal} -- meaning, informally, that it uses the smallest possible number of components
+to achieve the desired dynamics.
+Results in system theory show that any system can be realized in a particular minimal dimension
+(the dimension of the kryptotype, $n_\text{min}$),
+and that any two phenotypically equivalent systems of dimension $n_\text{min}$ are related by a change of coordinates.
+
+If the system is not minimal, then the Kalman decomposition tells us XXX
 
@@
         (A,B,C) : C e^{At} B = C_0 e^{A_0 t} B_0 \; \text{for}\; t \ge 0 
-      \right\}  \\
-      &= \left\{
-        (A,B,C) : C A^r B = C_0 A_0^r B_0 \; \text{for}\; 1 \le r \le n-1 
       \right\}  .
@@
 \end{equation}
-Equivalence of the two characterizations follows from the Cayley-Hamilton theorem.
 

### 2017-10-20 kalman_walk_paper.tex
more edits from overleaf

@@
 
-\newcommand{\calA}{\mathcal{A}}
 \newcommand{\calH}{\mathcal{H}}
@@
 
+% the triple (A,B,C) as 'system' (not using)
 \newcommand{\Sys}{\mathcal{S}}
-\newcommand{\allS}{\mathcal{A}}
-%\newcommand{\allS}{\mathcal{A}}
+% neutral set of all systems
+\newcommand{\allS}{\mathcal{N}}
 
@@
     %\item obtain estimates of variation in $A$ from thermodynamic occupancy model
+    \item ``On the origin of species not by means of natural selection''
 \end{itemize}
@@
 
-{Linearity} also ``zero-state equivalence''
+% {Linearity} also ``zero-state equivalence''
 Here we have assumed that the system is linear,
@@
                 \node[dashed] (B) at (3,0) {$\kappa_{2}$};
-                \node[shape=rectangle] (U) at (1.5,1.5) {input};
-                \node[shape=rectangle] (y) at (1.5,-1.5) {output};
+                \node[shape=rectangle] (U) at (1.5,1.5) {input ($u(t)$)};
+                \node[shape=rectangle] (y) at (1.5,-1.5) {output ($\phi(t)$)};
             \end{scope}
@@
 
-%Therefore varying $V$ can tweak the relationships between genes within a regulatory network, yet preserve the phenotype.
-Some gene networks, however, can grown or shrink, perhaps following gene duplications and deletions, and also still preserve their phenotypes.
-These changes cannot be described by $\mathcal{S}(V)$, but are ecanpuslated by \ref{eqn:equivalence}, below.
-%The Kalman decomposition decomposes a system into \emph{reachable} and \emph{observable} components -- that is coefficients that respond to an input and coefficients that produce an observable output.
-A minimal system's components are both reachable and observable.
-In this form, it is possible to add components that respond to an input, but do not influence the output, or components that, in principle, influence the output, but do not respond to any inputs.
-Any system with such components is, by definition, non-minimal, and any system can be modified to include any finite number of extra unreachable and/or unobservable components, without modifying its impulse response.
-Decomposing the system allows for the easy manipulation of ``superfluous''(but not necessarily expendable) components.
-After unreachable/unobservable coefficients are added, changing coordinates ``stirs'' the system coefficients such that the previously superfluous coefficients can become essential to system funciton.
-%Whether or not evolution constrains biological systems to be minimal or nearly minimal is an open question.
-
 Two systems are equivalent if they produce the same phenotype given the same input,
 i.e., have the same input--output relationship.
-For any input $u(t)$,
-suppose that $(\kappa(t),\phi(t))$ and $(\bar \kappa(t),\bar \phi(t))$ are the solutions to equation \eqref{eqn:system}
-with coefficient matrices $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ respectively.
-% (recall that $\kappa(0) = \bar \kappa(0) = 0$). 
 We say that
@@
 \textbf{phenotypically equivalent} 
-if
-\begin{align*}
-      \phi(t) = \bar \phi(t) \qquad \text{for all} \; t \ge 0.
-\end{align*}
-This occurs if and only if their impulse response functions are the same:
-if $h(t) = \bar h(t)$ for all $t \ge 0$.
+if their impulse response functions are the same:
+$h(t) = \bar h(t)$ for all $t \ge 0$.
+This implies that for any acceptable input $u(t)$,
+if $(\kappa_u(t),\phi_u(t))$ and $(\bar \kappa_u(t),\bar \phi_u(t))$ are the solutions to equation \eqref{eqn:system}
+of these two systems, respectively, then
+\begin{align*}
+      \phi_u(t) = \bar \phi_u(t) \qquad \text{for all} \; t \ge 0.
+\end{align*}
+% (recall that $\kappa(0) = \bar \kappa(0) = 0$).
 
@@
       &= C V^{-1} e^{V A V^{-1} t} V B 
-      = \bar{C} e^{\bar{A} t} \bar{B} .
+      = \bar{C} e^{\bar{A} t} \bar{B} = \bar h(t).
     \end{aligned}
@@
 
-If the system is not minimal, then the Kalman decomposition tells us XXX
-
-We refer to $\calA(\Sys_0)$ as the set of all systems equivalent to $\Sys_0$, regardless of dimension:
+Some gene networks, however, can grown or shrink, perhaps following gene duplications and deletions, 
+and also still preserve their phenotypes.
+More generally, even if the system is not minimal, 
+results from systems theory % -- the Kalman decomposition --
+explicitly describe the set of all phenotypically equivalent systems.
+% Could define script S here for 'system' if we want: $\Sys$
+We refer to $\allS(A_0,B_0,C_0)$ as the set of all systems phenotypically equivalent
+to the system defined by $(A_0, B_0, C_0)$.
+Concretely, this is
 \begin{equation} \label{eqn:equivalence}
   \begin{aligned}
-    \calA(\Sys_0) 
+    \allS(A_0, B_0, C_0) 
       &= \left\{
@@
 \end{equation}
-
-The set of all systems phenotypically equivalent to a given system $(A,B,C)$ 
-is elegantly described using the Kalman decomposition,
-which also clarifies the system dynamics.
+These systems need not have the same kryptotypic dimension $n$,
+but must have the same input and output dimensions ($\ell$ and $m$, respectively).
+
+The Kalman decomposition, which we now describe informally, elegantly characterizes this set
+\citep{kalman1963mathematical,kalman1969topics,anderson1966equivalence}.
 To motivate this, first note that the input $u(t)$ only directly pushes the system
-in directions lying in the span of the columns of $B$.
+in certain directions (those lying in the span of the columns of $B$).
 As a result, different combinations of input can 
-move the system in any direction that lies in the \emph{reachable subspace},
-which we denote by $\reachable$,
-and is defined to be the closure of $\spn(B)$ under applying $A$
-(or equivalently, the span of $B, AB, A^2B, \ldots A^{n-1}B$).
-Analogously to this, we define
-the \emph{observable subspace}, $\mathcal{O}$,
-to be the closure of $\spn(C^T)$ under applying $A$.
-(Or: $\unobservable$ is the largest $A$-invariant subspace
-contained in the null space of $C$;
-and $\reachable$ is the largest $A$-invariant subspace contained in the image of $B$.)
-
-If we define
-\begin{enumerate}
-    \item The columns of $P_\rno$ are an orthonormal basis for $\reachable \cap \unobservable$.
-    \item The columns of $P_\ro$ are an orthonormal basis of
-        the complement of $\reachable \cap \unobservable$ in $\reachable$.
-    \item The columns of $P_\nro$ are an orthonormal basis of
-        the complement of $\reachable \cap \unobservable$ in $\unobservable$.
-    \item The columns of $P_\nrno$ are an orthonormal basis of
-        the remainder of $\R^n$.
-\end{enumerate}
-If we then define
-\begin{align*}
-    P &= 
-    \left[ \begin{array}{c|c|c|c}
-        P_\rno & P_\ro & P_\nro & P_\nrno
-    \end{array} \right] ,
-\end{align*}
-then
-\begin{align*}
-    P^T P
-    &=
-    \left[ \begin{array}{c|c|c|c}
-        I & 0 & 0 & 0 \\
-        \hline
-        0 & I & U & 0 \\
-        \hline
-        0 & V & I & 0 \\
-        \hline
-        0 & 0 & 0 & I 
-    \end{array} \right] .
-\end{align*}
-\plr{Check this.  Can we get $U=V=0$?}
-
-The following theorem can be found in SOME REFERENCE.
-
-\begin{theorem}[Kalman decomposition] \label{thm:kalman}
-        For any system $(A,B,C)$ with corresponding Kalman basis matrix $P$,
+move the system in any direction that lies in what is known as the \emph{reachable subspace}.
+Analogously, we can only observe motion of the system in certain directions
+(those lying in the span of the columns of $C$),
+and so can only infer motion in what is known as the \emph{observable subspace}.
+The Kalman decomposition then classifies each direction in kryptotype space
+as either reachable or unreachable, and as either observable or unobservable.
+Only the components that are both reachable and observable determine the system's phenotype --
+that is, components that respond to an input and components that produce an observable output. 
+% it is possible to add components that respond to an input, 
+% but do not influence the output, or components that, in principle, influence the output, but do not respond to any inputs.
+
+Concretely, the \textbf{Kalman decomposition} of a system $(A,B,C)$  
+gives a change of basis $P$ such that
 the transformed system $(PAP^{-1},PB,CP^{-1})$  has the following form:
 \begin{align*}
-            \widehat A = PAP^{-1}
+       PAP^{-1}
        &=
@@
 \begin{align*}
-            \widehat B = PB
+     PB
      &=
@@
          0 
-            \end{array} \right] ,
-        \end{align*}
-        and
-        \begin{align*}
-            \widehat C = CP^{-1}
+   \end{array} \right] 
+   &
+   (CP^{-1})^T
    &=
    \left[ \begin{array}{cccc}
-                0 & C_{\ro} & C_{\nrno} & 0 
+       0 \\
+       C_{\ro}^T \\
+       C_{\nrno}^T \\
+       0 
    \end{array} \right] .
 \end{align*}
-        The impulse response of both systems is given by
-        \begin{align*}
-          h(t) = C_{\ro} e^{A_{\ro} t} B_{\ro} .
-        \end{align*}
-\end{theorem}
-
-In the latter case, we say that the system is \emph{minimal} 
--- there is no equivalent system with a smaller number of molecular species.
-Note that this says that any two equivalent minimal systems
-are changes of coordinates of each other.
-
-Since any system can be put into this form,
-and once in this form, its impulse response is determined only by 
-$C_{\ro}$, $A_{\ro}$, and $B_\ro$,
-therefore, the set of all equivalent systems are parameterized by
-the dimension $n$,
-the choice of basis ($P$),
-the remaining submatrices in $\widehat A$, $\widehat B$, and $\widehat C$
-(which are unconstrained),
-and a invertible transformation of $\spn(P_{\ro})$, which we call $T_\ro$.
-
-\begin{theorem}[Parameterization of equivalent systems]
-    Let $(A,B,C)$ be a minimal system.
-    \begin{itemize}
-        \item[(a)]
-            Every equivalent system is of the form given in Theorem \ref{thm:kalman},
-            i.e., can be specified by choosing a dimension, $n$;
-            submatrices in $\widehat A$, $\widehat B$, and $\widehat C$ 
-            except for $A_\ro=A$, $B_\ro=B$, and $C_\ro=C$;
-            and choosing an invertible matrix $P$.
-
-        \item[(b)]
-            \plr{conjecture:}
-            The parameterization is unique
-            if $P$ is furthermore chosen so that 
-            each $P_x$ other than $P_\ro$ is a projection matrix,
-            and that 
-            \begin{align*}
-                0
-                =
-                P_x^T P_y
-            \end{align*}
-            for all $(x,y)$ except $(\ro,\nrno)$.
-
-    \end{itemize} 
-\end{theorem}
+% and
+% \begin{align*}
+%    CP^{-1}
+%    &=
+%    \left[ \begin{array}{cccc}
+%        0 & C_{\ro} & C_{\nrno} & 0 
+%    \end{array} \right] .
+%\end{align*}
+The impulse response of the system is given by
+\begin{align*}
+      h(t) = C_{\ro} e^{A_{\ro} t} B_{\ro},
+\end{align*}
+and therefore, the system is phenotypically equivalent to the \emph{minimal} system $(A_{\ro}, B_{\ro}, C_{\ro})$.
+(Here the subscript $\ro$ refers to the both \emph{reachable and observable} subspace,
+while $\nrno$ refers to the \emph{unreachable and unobservable} subspace,
+and similarly for $\nro$ and $\rno$.)
+
+Any two minimal systems are related by a change of coordinates,
+and so the minimal subsystems obtained by the Kalman decomposition
+are unique up to a change of coordinates.
+In particular, this implies that there is no equivalent system with a smaller number of kryptotypic dimensions
+than the dimension of the minimal system.
+It is also remarkable to note that the gene regulatory network architecture to achieve a given input--output map is never unique --
+both the change of basis used to obtain the decomposition
+and, once in this form, all submatrices other than $A_{\ro}$, $B_{\ro}$, and $C_{\ro}$ can be changed without affecting the phenotype,
+and so represent degrees of freedom.
+
+\emph{Note on implementation:}
+The \emph{reachable subspace},
+which we denote by $\reachable$,
+is defined to be the closure of $\spn(B)$ under applying $A$,
+% (or equivalently, the span of $B, AB, A^2B, \ldots A^{n-1}B$).
+% Analogously to this, we define
+and the \emph{unobservable subspace}, 
+%$\mathcal{O}$, is the closure of $\spn(C^T)$ under applying $A$.
+denoted $\unobservable$, is the largest $A$-invariant subspace
+contained in the null space of $C$.
+% and $\reachable$ is the largest $A$-invariant subspace contained in the image of $B$.)
+The four subspaces, $\rno$, $\ro$, $\nrno$, and $nro$
+are defined from these by intersections and orthogonal complements.
 
@@
 %  \end{align}
-It is remarkable to note that even with the relationships between environment, kryptotype, and phenotype constant, and in the minimal dimension, there are still almost always degrees of freedom.
+
+For the remainder of the paper, we interpret $\allS$ as the phenotypically neutral landscape, 
+wherein a large population will drift under environmental and selective stasis. 
+Even if the phenotype is constrained and remains constant through evolutionary time, 
+the molecular mechanism underpinning it is not constrained and likely will not be conserved.
+
+Finally, note that if $B$ and $C$ are held constant --
+i.e., if the relationships between environment, kryptotype, and phenotype do not change --
+there are \emph{still} usually degrees of freedom. 
 These correspond to distinct genetic networks that perform indistinguishable functions. 
-For example, the equivalent systems in Example \ref{ex:all_osc} are minimal, and share common B and C matrices.
-
-It is also remarkable to note that there is no unique triple $(A,B,C)$ for the mapping $h: u \mapsto \phi$.
-This implies that the gene regulatory network architecture per phenotype/environment pair is never unique -- that is under these assumptions, there is always more than one possible gene regulatory network architecture per phenotype.
-%This set can be precisley defined and completely parameterized using the \emph{Kalman decomposition} (Appendix \ref{apx:kalman}).
-
-For the remainder of the paper, we interpret $\allS$ as the phenotypially neutral landscape, wherein a large population will drift under environmental and selective stasis.
-Even if the phenotype is constrained and remains constant through evolutionary time, the molecular mechanism underpinning it is not constrained and likely will not be conserved.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\subsection*{Speciation}
-
-%Define reproduction in diploids as first, the recombination of unlinked genes to make gametes, and second, as the averaging of two individual parental gametes to produce an offspring.
-Assuming parental populations are both phenotypically identical and genetically homogenous within each population, first generation hybrids (F1s) can be computed by averaging the two parental gene networks.
-Second generation hybrids (F2s) can be computed by first swapping the genes between the two parental gene networks, and next averaging these hybrid gametes.
-
-%Speciation, or the formation of new species, is often a consequence of reprdouctive incompatibility among populations -- usually via hybrid sterility or inviability.
-To discuss speciaiton we first need to describe how reproduction is modelled.
- %A diploid first generation hybird's ($F_1$'s) genetic regulation is a consequence of its two genomes, as a copy of genes and regulatory sequences from each parent is present.
-%\plr{should have a stronger justification for the assertion that $F_1$ is the arithmetic average of the parents than "Thus we say"}
-%A diploid organism contains two gene network copies: one from each parent.
-As such, a diploid's system dynamics is computed by averaging the coefficients of both systems.
-To reproduce, each organism passes on a recombined haploid gene network to its offspring.
-Haploid gene networks swap system matrix rows randomly between its own two networks.
-%\plr{I don't think we say that $A$ is a "gene network architecture".}
-%Therefore the phenotypic dynamics typical of an $F_{1}$ cross between two allopatric populations will be determined simply by averaging it's two parental genomes.
-However, in second generation hybrid crosses $F_2$ between allopatric populations, first new haploid systems will be formed by recombination -- in the process shuffling and combining regulatory coefficients from allopatric populations -- and then brought together to form a diploid.
-
-%\begin{align}
-%    F_{1}(\tau, \hat{\tau}) &= \frac{ A(\tau) + A(\hat{\tau}) }{2}  .
-%   h(t, F_{1}) &= C e^{\left( F_{1}\left( \tau, \hat{\tau} \right) t \right)} B
-%\end{align}
-%\begin{align*}
-% G\left(r, \tau, \hat{\tau}\right) = Q(r) A( \tau ) + \left( I - Q(r) \right) A( \hat{\tau} )
-%\end{align*}
-%\begin{align}
-%    F_{2}(i, j) = \frac{G (i) + G(j)}{2}
-%\end{align}
-
-  %\begin{align*}
-  %  h\left(t, F_{2}\left( r, r' \right)\right) = C e^{\left( F_{2}\left(r, r' \right) t \right)} B
-  %\end{align*}
-
-%Fitness $\fit$ is a function (usually a Guassian) of an organism's weighted phenotypic distance from an optimum,
-%  $\fit\left( \widehat{\phi}(t) \right) = \exp \left(- \int_{0}^{t} \frac{\rho(s)}{\beta} \left\lVert \phi(s) - \widehat{\phi}(s) \right\rVert^{2} ds  \right)$
-%where $\rho(t)$ is some weighting function and $\beta$ is a parameter.
-
-%We quantitatively describe the degree of incompatibility between two populations $P_{1}$ and $P_{2}$ as
-%\begin{align*}
-%      \Incompat = \frac{2 \left\langle \fit \left(\phi_{F_{1}}\right) \right\rangle}{\left\langle \fit \left(\phi_{P_{1}}\right) \right\rangle +  \left\langle \fit \left(\phi_{P_{2}}\right) \right\rangle} ,
-%\end{align*}
-%where angled brackets imply averaging.  \plr{averaging over what? And, this is as others define - cite.}
-
-%\plr{This is not part of the definition of $\mathcal{I}$:}
-%\js{should we delete this example ($A(0) \times A(2)$ hybrids)? I like how clean it is -- but I'm not sure it's that important in this context. It's a good proof of principle though.}
-%\jss{$F_1$s created by crossing phenotypically equivalent oscillators $A(0)$ and $A(2)$ have a phenotype of $\phi_{F_{1}}(t) = e^{t}$, 
-%in contrast to both parents, who have $\phi_{P_{1}}(t) = \phi_{P_{2}}(t) = \sin(t) + \cos(t)$. 
-%The hybrid phenotype is significantly different (it does not oscillate and increases infinitely) despite the phenotypic equivalence of the parents.
-%%\plr{don't need these to be big equations.}
-%%\begin{align*}
-%%  \fit \left(\phi_{P_{1}}\right) = \fit \left( \phi_{P_{2}} \right) &= 1 \\
-%%  \fit\left(\phi_{F_{1}}\right) &= 0 \\
-%%  \mathcal{I} &= 0
-%%\end{align*}
-%
-%%Thus if populations $1$ and $2$ are homogeneous $A(0)$ and $A(2)$, respectively, we say that they are completely incompatible as $\mathcal{I} = 0$.
-%
-%  \begin{figure}[H]
-%    \centering
-%    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{expF1}
-%    \caption{$F_1$ hybrid (orange) and parental (blue) phenotypic oscillator dynamics for an $A(0)$ by $A(2)$ cross. The hybrid fails to oscillate and exhibits qualitatively different dynamics.}
-%  \end{figure}
-%  }
-%%     \begin{example}[F1 Reproductive Incompatibility in an Oscillating Gene Network]
-%%       DMI examples...
-%%       \begin{figure}[H]
-%%         \includegraphics[width=0.5\textwidth]{A2_A3_osci_F1_hyb}
-%%       \end{figure}
-%%     \end{example}
-%
-%%     \begin{example}[Not all Networks can Host Incompatibilities]
-%%       convex sets cant have DMIs
-%
-%%       \begin{align*}
-%%         h(t) = 2 e^{- \theta t}
-%%       \end{align*}
-%%       Any non-minimal system with rows summing to $\theta$ is PI.
-Further, these systems are closed under averaging (mating) and row swapping (meiosis), leaving all hybrids optimally fit.
-The set of gene matrices is affine and therefore convex.
-%%     \end{example}
-%
-%
-%\plri{Missing: a statement about when there are neutral directions, and how many.}
-%
-%\plr{I think we should first put the section about quantitative drift in, and then the following section that tries to get at real values of parameters.}
-%
-%
- \begin{example}[All Phenotypically Equivalent Cell Cycle Control Networks] \label{ex:all_osc}
-   The set of all two-gene regulatory networks phenotypically equivalent to the cell cycle control network in \ref{ex:oscillator}, where only $A$ can vary, is given by $\Sys (\tau) = (A(\tau), B, C)$, such that,
-    %\begin{align*}
-    %  A(\tau) &= \begin{bmatrix} 1 & 0 \\ \tau & 1-\tau \end{bmatrix} \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 0 \\ \frac{\tau}{\tau-1} & \frac{-1}{\tau-1} \end{bmatrix}  \qquad \forall \ \tau \neq 1 %\\
-    %\end{align*}
-    \begin{align*}
-      A(\tau) &= \frac{1}{\tau-1} \begin{bmatrix} \tau & -1 \\ 2 \tau(\tau - 1) + 1 &  -\tau \end{bmatrix} \ \text{for} \ \tau \neq 1
-    \end{align*}
+The following example \ref{ex:all_osc} gives the set of minimal systems equivalent to the oscillator of example \ref{ex:oscillator},
+that all share common $B$ and $C$ matrices.
+
+\begin{example}[All Phenotypically Equivalent Oscillators] \label{ex:all_osc}
+The oscillator of example \ref{ex:oscillator} is minimal, and so any equivalent system is a change of coordinates
+by an invertible matrix $V$.
+If we further require $B$ and $C$ to be invariant then we need $VB=B$ and $CV=C$.
+Solving these equations,
+we find that a one-parameter family $(A(\tau), B, C)$ describes the set of all two-gene systems
+phenotypically equivalent to the oscillator,
+where
+    \begin{align*}
+      A(\tau) &= \frac{1}{\tau-1} \begin{bmatrix} \tau & -1 \\ 2 \tau(\tau - 1) + 1 &  -\tau \end{bmatrix} \ \text{for} \ \tau \neq 1 .
+    \end{align*}
+The resulting set of systems,
+and their dynamics, are depicted in Figure \ref{fig:all_osc}
+\end{example}
+
     \begin{figure}[H]
     \centering
-% Nice tikz!!!
+\begin{tabular}{cc}
 \begin{tikzpicture}
@@
 \end{scope}
-
-\end{tikzpicture}
-      \caption{$\Sys(\tau)$, the set of all phenotype-equivalent cell cycle control networks $\Sys (\tau) = \left\{ \calA_2 : B = [1\ 1]^{T} ,\ C = [1\ 0] \right\}$.}
+\end{tikzpicture} &	\includegraphics[width=0.5\textwidth, height=0.125\paperheight]{figures/osc_kryp_compare}
+    \end{tabular}
+% \caption{}
+      \caption{
+      (Left) $A(\tau)$, the set of all phenotype-equivalent cell cycle control networks.
+      (Right) Gene-1 dynamics (blue) for both systems $A(0)$ and $A(2)$ are identical, however, $A(0)$ gene-2 dynamics (orange) differ from $A(2)$ (green).
+      % Both gene-1 dynamics are given by $\kappa_{1}(t) = \sin t + \cos t$, and gene-2 by $\kappa_{2}(t) = \cos t - \sin t$ ($A(0)$) and $\kappa'_{2}(t) = \cos t + 3 \sin t$ ($A(2)$).
+      } 
+    \label{fig:all_osc}
 \end{figure}
- % \end{example}
-
- % \begin{example}[External Equivalence does not imply internal equivalence]
-   Despite the phenotypic equivalence of all instantiations of $\Sys(\tau)$, the kryptotypes, vary as a function of $\tau$. 
-    %Gene-1 dynamics (blue) are equivalent for network architectures $A(0)$ and $A(2)$, however the dynamics of gene-2 (orange) differ with $\tau$.
+
+\paragraph{Sexual reproduction and recombination}
+Parents with phenotypically equivalent, yet different, gene networks may produce offspring with dramatically different phenotypes. 
+If the phenotypes are significantly divergent then the offspring may be inviable or otherwise have very low fitnesses, 
+despite both parents being well adapted. If this is consistent for the entire population, we would consider them to be separate species.
+
+Diploid organisms have two copies of the genome, each of which encodes a set of system coefficients.
+We assume that a diploid who has inherited systems $(A', B', C')$ and $(A'', B'', C'')$ from her mother and father, respectively,
+has phenotype determined by the system that averages these two,
+$((A'+A'')/2, (B'+B'')/2, (C'+C'')/2)$.
+
+Each of the copies of the genome an organism inherits from her parents are generated from their two copies of the genome by meiosis,
+in which the diploid parent recombines her two genomes.
+We will assume that each coefficient (i.e., entry of $A$, $B$ or $C$)
+is determined by a single nonrecombining locus,
+so that each coefficient in the system produced by meiosis is an independent random choice
+of the two parental coefficients.
+With these definitions, an $F_1$ offspring of two individuals
+carries one system copy from each parent,
+and an $F_2$ is the offspring of two independently formed $F_1$ individuals.
+If the parents are from distinct populations,
+these are first- and second-generation hybrids, respectively.
+
+This is a simplification:
+since the $i^\text{th}$ row of $A$ summarizes how each gene regulates gene $i$,
+and hence is determined by the promoter region of gene $i$,
+we would actually expect the elements of a row of $A$ to tend to be inherited together.
+Similarly, we expect in practice heritable variation in each coefficient
+to be determined by more than one locus --
+but this may be a reasonable approximation.
+
+The offspring of two systems with the same phenotype may not have the same phenotype as the parents -- in other words $\allS$, the set of all systems phenotypically equivalent to a given one, is not, in general, closed under averaging or recombination.
+Next we discuss how this fact can contribute to hybrid incompatibility and genetic load.
+If sexual recombination among systems drawn from $\allS$ yields systems with divergent phenotypes, populations containing significant diversity in $\allS$ can carry genetic load, and isolated populations may fail to produce hybrids with viable phenotypes.
+
 \begin{figure}[H]
 \centering
-   % \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{osc_A0_A2_both_compare}
-    %\caption{Gene-1 (blue) and gene-2 (orange) dynamics for $A(0)$ (top) and $A(2)$ (bottom).
-Both (top and bottom) gene-1 dynamics are given by $ \kappa_{1} = \sin(t) + \cos(t)$, and gene-2 by $\kappa_{2} = \cos(t) - \sin(t)$ (top) and $\kappa'_{2} = \cos(t) + 3 \sin(t)$ (bottom).}
-    \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{figures/osc_kryp_compare}
-    \caption{Gene-1 dynamics (blue) for both systems $\Sys(0)$ and $\Sys(2)$ are identical, however, $\Sys(0)$ gene-2 dynamics (orange) differ from $\Sys(2)$ (green).
-Both gene-1 dynamics are given by $\kappa_{1} = \sin t + \cos t$, and gene-2 by $\kappa_{2} = \cos t - \sin t$ ($\Sys(0)$) and $\kappa'_{2} = \cos t + 3 \sin t$ ($\Sys(2)$).}
+\includegraphics{figures/conceptual_fig}
+\caption{
+    \label{fig:conceptual_fig}
+    A conceptual figure of the fitness consequences of hybridization:
+    axes represent system coefficients (i.e., entries of $A$);
+    the line of optimal system coefficients is down in black;
+    dotted lines give phenotypic distances to the optimum.
+    Two pairs of parental populations are shown in black, along the optimum;
+    a hypothetical population of $F_1$s are shown for each in red,
+    and the distribution of one type of $F_2$ is shown in purple
+    (other types of $F_2$ are not shown).
+    Solid lines depict the distance of the $F_2$ to optimum.
+}
 \end{figure}
-\end{example}
-\begin{example}[Metabolic network] \label{ex:metabolic}
-Consider an organism that can metabolize two different sugars (present at logarithmic concentrations $u_{1}$ and $u_{2}$), 
-with two enzymes (at log concentrations $\phi_{1}$ and $\phi_{2}$).
-Further suppose that the second sugar is prefered,
-%and that the organism has a regulatory network that can deploy situation specific metabolic strategy.
-and that
-depending on $u_{1}$ and $u_{2}$ the organism will synthesize an appropriate $\phi_{1}$ and $\phi_{2}$. 
-Furthermore, assume this system contains at least two transcription factors, whose log concentrations are $\kappa_{1}$, $\kappa_{2}$, $\dots \kappa_{n}$.
-Minimally such a system may have the architecture,
-%\begin{align*}
-%    \mathcal{S}_{\min} = \left\{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 0 & -1 \\ 0 & 0 \end{bmatrix} \begin{bmatrix} \kappa_{1} \\ \kappa_{2} \end{bmatrix} (t) + \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \begin{bmatrix} u_{1} \\ u_{2} \end{bmatrix} (t) \\[11pt]
-%    \phi(t) &= \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \begin{bmatrix} \kappa_{1} \\ \kappa_{2} \end{bmatrix} (t) \end{array} \right.
-%\end{align*}
-  $\Sys_{\min}(U) = (U \left[ \begin{smallmatrix} 0 & -1 \\ 0 & 0 \end{smallmatrix} \right] U^{-1}, U 
-    %\left[\begin{smallmatrix} 1 & 0 \\ 0 & 1 \end{smallmatrix}\right]
-      ,
-     % \left[\begin{smallmatrix} 1 & 0 \\ 0 & 1 \end{smallmatrix}\right]
-        U^{-1})$. 
-%
-%The impulse response of the system is,
-%$h(t) = \left[\begin{smallmatrix} 1 & - t \\ 0 & 1 \end{smallmatrix}\right]$.
-    We can find alternative equivalent systems by changning coordinates ($U \rightarrow U'$) or, more generally, by applying the Kalman decomposition \ref{apx:kalman}.
-To illustrate that phenotypic invariance does not require dimensional invariance, we apply the Kalman decomposition to $\Sys_{\min}$ to find
-    %This may happen if a system is co-opted from another, or may be a consequence of gene duplication and deletion. 
-  $\Sys(\small D_{1-3},V) \! =  \! \left( \! \begin{array}{ll}
-     V \!
-\left[\begin{smallmatrix} D_{1} & D_{2} \\ 0 & A \end{smallmatrix}\right] \!
-V^{-1}, V \!
-\left[\begin{smallmatrix} D_{3} \\ B \end{smallmatrix}\right], \left[\begin{smallmatrix} 0 & C \end{smallmatrix}\right] \!
-V^{-1} \end{array} \!
-\right)
-$, both of which are in $\allS$
-%  \begin{align*}
-%    \Sys(X_{1-3},V) = \left( \begin{array}{ll}
-%     V \left[\begin{smallmatrix} X_{1} & X_{2} \\ 0 & A_{\min} \end{smallmatrix}\right] V^{-1}, V \left[\begin{smallmatrix} X_{3} \\ B_{\min} \end{smallmatrix}\right], \left[\begin{smallmatrix} 0 & C_{\min} \end{smallmatrix}\right] V^{-1} \end{array} \right)
-%    \end{align*}
-  ($V$ can be any $4$-dimensional and invertible matrix, $D_{1-3} \in \R^{2 \times 2}$, and $(A,B,C) \in \Sys_{\min}$).
-%So for example, some system can be wired as follows, and still be input-output equivalent to the minimal metabolic system $\mathcal{S}_{\min}$:
-% \begin{align*}
-%   A' &= \begin{bmatrix}
-%     1.6923  &  1.5385 &  -2.6154 &  -2 \\
-%     0.8462  &  0.7692 &  -1.3077 &  -1 \\
-%     1.2692  &  1.1538 &  -1.9615 &  -1.5 \\
-%     0.4231  &  0.3846 &  -0.6538 &  -0.5
-%   \end{bmatrix} \\
-%   B' &= \begin{bmatrix} 4 & 4 \\ 2 & 2 \\ 3 & 3 \\ 1 & 3 \end{bmatrix} \\
-%     C' &= \begin{bmatrix}  0.2692  &  0.1538  &  0.0385 &  -0.5 \\
-%     -0.4231 &  -0.3846 &   0.6538  &  0.5 \end{bmatrix}
-% \end{align*}
-%
-%Despite the present example consisting of a minimal $2 \times 2$ and a non-minimal $4 \times 4$ system, 
-%any $m$-dimensional system can be constructed using this method 
-%-- applying a change of coordinates to the Kalman decomposition 
-%-- to construct a mechanistically different system with identical phenotypic dynamics. 
-%
-%When applying the Kalman decomposition to real biological networks, one may have to restrict the free parameters, as physiology and biochemistry might not allow all mathematically possible coefficients.
-We note that as there are so many degrees of freedom, it may be difficult to infer system function solely from structure.
-\end{example}
-
-\paragraph{Sexual recombination and hybrid incompatibility}
-Phenotypically identical yet kryptotypically variant gene networks, when brought together and/or recombined during sexual reproduction, can lead to phenotypic inviability.
-
-    For instance, the diploid $\{\Sys, \Sys' \}$ has system architecture $\Sys_{\text{diploid}} \! = \! \left(\Sys \! + \! \Sys'\right)\!/2$, the average of its two haploid systems.
-    %During meiosis a diploid recombines its two genomes to form a new one (a gamete): $\left(\Sys \!
-\times \!
-\Sys' \right) \rightarrow \Sys''$; by randomly swapping matrix rows between two systems (Appendix \ref{apx:recombination}).
-Then two gametes join to form an offspring (\emph{e.g.} $\{\Sys'',\Sys\} $).
-
-     During meiosis, a diploid recombines its two genomes to form a gamete by randomly swapping rows or columns between its two systems.
-  \emph{E.g.} gamete systems $(A'', B'', C'')$, produced by the diploid $\{\Sys, \Sys'\}$, are formed by recombining two systems $\Sys = (A,B,C)$ and $\Sys' = (A', B', C')$  to form $\Sys''$ such that,
-  %\begin{align*}
-   $A'' = MA + (I-M)A'$, 
-   $B'' = MB + (I-M)B'$, and
-   $C'' = CM + C'(I-M)$.
- % \end{align*}
-  $M$ is a diagonal matrix where each diagonal element is a Bernoulli random variable ($M_{ii} = 0$ or $1$ with equal probability, and $M_{ij}=0$ if $i \neq j$).
-If systems are different dimensions the smaller system elements can be augmented with $0$s (\emph{e.g.} $\left[ \begin{smallmatrix} A & 0 \\ 0 & 0 \end{smallmatrix} \right], \left[ \begin{smallmatrix} B \\ 0 \end{smallmatrix} \right], \left[ \begin{smallmatrix} C & 0 \end{smallmatrix}\right]$).
-
-Even if two systems yield the same phenotype, their diploid and recombinant systems may not. This is because $\allS$ is not required to be closed under averaging or recombination.
-
-If sexual recombination among systems drawn from $\allS$ yields systems with divergent phenotypes, populations containing significant diversity in $\allS$ can carry genetic load, and isolated populations may fail to produce hybrids with viable phenotypes (Figure \ref{fig:expF1}).
-
-%\paragraph*{Hybrid phenotypic breakdown between oscillators}
-%For example, if the two phenotypically equivalent systems $A(0)$ and $A(2)$ from Example \ref{ex:all_osc} hybridize, their offspring's phenotype will fail to oscillate completely. 
-  \begin{figure}[H] 
-    \centering
-    \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{expF1}
-    \caption{\textbf{Hybrid phenotypic breakdown} $F_1$ hybrid (orange) and parental (blue) phenotypic cell-cycle control (oscillator) dynamics for an $\Sys(0)$ by $\Sys(2)$ cross from Example \ref{ex:all_osc}.
+
+\paragraph{Hybrid incompatibility}
+Two parents with the optimal phenotype can produce offspring whose phenotype is suboptimal
+if the parents have different underlying systems.
+This leads to the question: 
+How quickly does the hybrid phenotype break down with increasing distance between the parents?
+To quantify this, we will measure how far a system's phenotype is from optimal
+using a weighted difference between impulse response functions.
+Suppose that $\rho(t)$ is a nonnegative, smooth, square-integrable weighting function,
+suppose that $h_0(t)$ is the \emph{optimal} impulse response function 
+and define the ``distance to optimum'' of another impulse response function
+to be
+\begin{align}
+	D(h) = \left( \int_0^\infty \rho(t) \|h(t) - h_0(t)\|^2 dt \right)^{1/2} .
+\end{align}
+Consider reproduction between a parent with system $(A, B, C)$ 
+and another displaced by distance $\epsilon$ in the direction $(X,Y,Z)$,
+i.e., having  system $(A + \epsilon X, B + \epsilon Y, C + \epsilon Z)$.
+We assume both are ``perfectly adapted'' systems, 
+i.e., having impulse response function $h_0(t)$,
+and their offspring has impulse response function $h_\epsilon(t)$.
+A Taylor expansion of $D(h_\epsilon)$ in $\epsilon$ is explicitly worked out in Appendix \ref{apx:H_calc},
+and shows that the phenotype of an $F_1$ hybrid between these two is at distance proportional to $\epsilon^2$ from optimal,
+while $F_2$ hybrids are at distance proportional to $\epsilon$.
+This is because an $F_1$ hybrid has one copy of each parental system,
+and therefore lies directly between the parental systems (see Figure \ref{fig:conceptual_fig}) --
+the parents both lie in $\allS$, which is the valley defined by $D$,
+and so their midpoint only differs from optimal due to curvature of $\allS$.
+In contrast, an $F_2$ hybrid may be homozygous for one parental type in some coefficients
+and homozygous for the other parental type in others;
+this means that each coefficient of an $F_2$ may be equal to either one of the parents,
+or intermediate between the two;
+this means that possible $F_2$ systems may be as far from the optimal set, $\allS$,
+as the distance between the parents.
+The precise rate at which the phenotype of a hybrid diverges depends on the geometry --
+in Figure \ref{fig:conceptual_fig}, this is depicted as the angle of the black line (the optimal set)
+with respect to the coordinates.
+
+\begin{example}[Hybrid Incompatibility in the Oscillator] \label{ex:hybrid_osc}
+Offspring of two equivalent systems from Example \ref{ex:all_osc}
+can easily fail to oscillate.
+For instance, the $F_1$ offspring between homozygous parents at $\tau=0$ and $\tau=2$
+has phenotype $\phi_{F_1}(t) = e^t$, rather than $\phi(t) = \sin t + \cos t$.
+However, the coefficients of these two parental systems differ substantially,
+probably more than would be realistically observed between diverging populations.
+In figure \ref{fig:hybs} we compare the phenotypes for $F_1$ and $F_2$ hybrids between more similar parents,
+and see increasingly divergent phenotypes as the difference between the parental systems increases.
+(In this example, the coefficients of $A(2+\epsilon)$ differ from those of $A(2)$ by an average factor of $1+\epsilon/2$;
+such small differences could plausibly be caused by changes to promoter sequences.)
+This divergence is quantified in Figure \ref{fig:osc_incompat},
+which shows that mean distance to optimum phenotype of the $F_1$ and $F_2$ hybrid offspring between $A(2)$ and $A(2+\epsilon)$
+increases with $\epsilon^2$ and $\epsilon$, respectively.
+
+The coefficients in $A$ -- i.e., the regulatory coefficients --
+differ between parents by only a few percent (around 0.5\% for $\tau=2.01$ and 5\% for $\tau=2.1$).
+This is well within the amount of regulatory coefficient variation we expect to find segregating within real populations
+(discussed further below). 
+For these small values of $\epsilon$, hybrid phenotypes remain relatively stable, 
+consistent with the idea that natural selection will allow such intrapopulation variation.
+
+%%% figure of tau=0 and tau=2 hybrid:
+%  \begin{figure}[H] 
+%    \centering
+%    \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{expF1}
+%    \caption{\textbf{Hybrid phenotypic breakdown} $F_1$ hybrid (orange) and parental (blue) phenotypic cell-cycle control (oscillator) dynamics %for an $\Sys(0)$ by $\Sys(2)$ cross from Example \ref{ex:all_osc}.
 The $F_1$ hybrid
 %$\left(A_{F_1}\!\!=\!\!\tfrac{A\!(\!0\!) \! + \! A\!(\!2\!)}{2}\right)$
-gene expression fails to oscillate and instead grows exponentialy
-($\phi_{F_0}\! = \! \sin t \! + \! \cos t$, however $\phi_{F_1} \! = \! e^t$).
-    }
-    \label{fig:expF1}
-  \end{figure}
-%
-%\paragraph{Sexual reproduction and speciation}
-%A diploid organism contains two gene network copies: one from each parent.
-As such, a diploid's system dynamics is computed by averaging the coefficients of both systems.
-To reproduce, each organism passes on a recombined haploid gene network to its offspring.
-Haploid gene networks swap system matrix rows randomly between its own two networks.
-%Therefore the phenotypic dynamics typical of a first generation ($F_{1}$) cross between two allopatric populations will be determined simply by averaging it's two parental genomes.
-However, in second generation hybrid ($F_2$) crosses, first new haploid systems will be formed by recombination -- in the process shuffling and combining regulatory coefficients from allopatric populations -- and then brought together to form a diploid.
-%
-%
-  \begin{example}[Hybrid Incompatibility in an Oscillating Gene Network] \label{ex:hybrid_osc}
-    Here we compare the phenotypes for $F_2$ hybrids formed by crossing oscillators with different values for $\tau$ from \ref{ex:all_osc}, $\Sys(2)$ with $\Sys(2.01)$, $\Sys(2.1)$, and $\Sys(2.5)$.
-    Systems have identical phenotypes, however some of the hybrids exhibit markedly different dynamics.
-
-%        \begin{align*}
-%         A(2) = \begin{bmatrix} 2 & -1 \\ 5 & -2 \end{bmatrix} &\qquad A(2.01) = \begin{bmatrix} 1.9901 & -0.9901 \\ 5.0101 & -1.9901 \end{bmatrix} \\
-%          A(2.1) = \begin{bmatrix} 1.9091 & -0.9091 \\ 5.1091 & -1.9091 \end{bmatrix} &\qquad A(2.5) = \begin{bmatrix} 1.6667 & -0.6667 \\ 5.6667 & -1.6667 \end{bmatrix}
-%     \end{align*}
-
-%\begin{alignat*}{2}
-%      A(2) &= \left[\begin{array}{cc} 2 & -1 \\[6pt] 5 & -2 \end{array}\right] &&\qquad A(2.01) = \left[\begin{array}{cc} 2 -\frac{1}{101} & -1 + \frac{1}{101} \\[6pt] 5 + \frac{1}{99} & -2 + \frac{1}{101} \end{array}\right] \\
-%        A(2.1) &= \left[\begin{array}{cc} 2 - \frac{1}{11} & -1 + \frac{1}{11} \\[6pt] 5 + \frac{6}{55} & -2 + \frac{1}{11} \end{array}\right] &&\qquad A(2.5) = \left[\begin{array}{cc} 2 - \frac{1}{3} & -1 + \frac{1}{3} \\[6pt] 5 + \frac{2}{3} & -2 + \frac{1}{3} \end{array}\right]
-%\end{alignat*}
-
-%      \begin{figure}[H]
-%        \begin{tabular}{cc}
-%          \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_1}
-%          & \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_2}
-%         \\ \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_3}
-%          & \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_4}
-%          \\ \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_5}
-%          & \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_6}
-%        \end{tabular}
+%gene expression fails to oscillate and instead grows exponentialy
+%($\phi_{F_0}\! = \! \sin t \! + \! \cos t$, however $\phi_{F_1} \! = \! e^t$).
+%    }
+%    \label{fig:expF1}
 %  \end{figure}
 
-%    \begin{figure}[H]
-%      \centering
-%      \includegraphics[width=0.5\textwidth]{A2_A2-1_F2s}
-%      \caption{F2s from $A(2)$ and $A(2.1)$.}
-%    \end{figure}
-%      \begin{figure}[H]
-%        \begin{tabular}{cc}
-%        \includegraphics[width=0.5\textwidth]{F2s-small} \\
-%        \includegraphics[width=0.5\textwidth]{F2s} \\
-%        \includegraphics[width=0.5\textwidth]{F2s-large}
-%        \end{tabular}
-%        \caption{F2s from $A(2)$ and $A(2.1)$.}
-%      \end{figure}
-
-    \begin{figure}[H] \label{fig:hybs}
+\begin{figure}[H]
   \centering
@@
   \caption{
-    $F_1$ (left) and $F_2$ (right) hybrids crossing $\Sys(2)$ with $\Sys(2.01)$ (top), $\Sys(2.1)$ (middle), and $\Sys(2.5)$ (bottom).
-    $F_2$ hybrids display more phenotypic divergence than $F_1$s, on average. Further, some $F_2$s completely fail to oscillate, as seen in an $\Sys(2.5)$ $F_2$ (light blue).
+  \textbf{(left)} Phenotypes of $F_1$ hybrids between an $A(2)$ parent and, top-to-bottom, an $A(2.01)$, an $A(2.1)$, and $A(2.5)$ parent.
+    Parental phenotypes ($\sin t + \cos t$) are shown in blue, and hybrid phenotypes in orange.
+    \textbf{(right)} Phenotypes of $F_2$ hybrids between the same set of parents,
+    with parental phenotype again in blue.
+    Different lines correspond to different $F_2$s produced by recombination;
+    note that some completely fail to oscillate.
     \plri{make axis labels bigger}
-  }
+  } \label{fig:hybs}
 \end{figure}
@@
   \begin{figure}[H]
+  \label{fig:osc_incompat}
     \centering
     \begin{tabular}{cc}
-    \includegraphics[width=0.25\textwidth, height=0.125\paperheight]{figures/f1_quartic2} &
-    \includegraphics[width=0.25\textwidth, height=0.125\paperheight]{figures/f2_quad2}
+    \includegraphics[width=0.4\textwidth]{figures/f1_quartic2} &
+    \includegraphics[width=0.4\textwidth]{figures/f2_quad2}
     \end{tabular}
-    \caption{(Left) Squared $F_1$ hybrid phenotypes from Figure \ref{fig:hybs} diverge quartically, and (right) $F_2$ hybrid phenotypes diverge quadratically as a function of time in allopatry.}
+    \caption{
+    \textbf{(left)} Phenotype distance to optimum, $D$, using $\rho(t) = \exp(-t/4\pi)$
+    for $F_1$ hybrids between an $A(2)$ and an $A(2+\epsilon)$ parent.
+    \textbf{(right)} Same, but for $F_2$ offspring (one line per type of $F_2$).
+    \plri{make this $D$ not $D^2$ and say which line is the overall average $F_2$.}
+    } \label{fig:osc_incompat}
   \end{figure}
 \end{example}
-%
+
 %%%%%%%%%%%%%%%%%%%%%%%%
 \section*{Systems drift and the accumulation of incompatibilities}
+
 Thus far we have shown that many distinct molecular mechanisms can realize identical phenotypes
@@
 near a set of equivalent and optimal systems. We work with a population of effective size $N_e$. 
+\plri{note that this is more general than LTI model}
 
@@
 These are depicted in figure \ref{fig:conceptual_fig}.
-
-\begin{figure}[H]
-\centering
-\includegraphics{figures/conceptual_fig}
-\caption{
-    \label{fig:conceptual_fig}
-    A conceptual figure of the fitness consequences of hybridization:
-    axes represent system coefficients (i.e., entries of $A$);
-    the line of optimal system coefficients is down in black;
-    dotted lines give phenotypic distances to the optimum.
-    Two pairs of parental populations are shown in black, along the optimum;
-    a hypothetical population of $F_1$s are shown for each in red,
-    and the distribution of one type of $F_2$ is shown in purple
-    (other types of $F_2$ are not shown).
-    Solid lines depict the distance of the $F_2$ to optimum.
-    \plri{Should show all types of F2? would be messy.}
-}
-\end{figure}
 
@@
 %at least for networks whose function is strongly constrained.
-%
-%
+
 \paragraph{Speciation rates under neutrality}
@@
 Note, however, that if intrapopulation variation is very small such that $\sigma_S \ll \gamma$, speciation rates will be slower.
+
 %%%%%%%%%%%%%%%%%%%%%%%%
@@
 We emphasize that these incompatibilities are a consequence of recombining different, yet functionally equivalent, mechanisms.
+\plr{See refs in Barton 2010 paper.}
 
@@
 \section*{Examples}
+\begin{example}[Metabolic network] \label{ex:metabolic}
+Consider an organism that can metabolize two different sugars (present at logarithmic concentrations $u_{1}$ and $u_{2}$), 
+with two enzymes (at log concentrations $\phi_{1}$ and $\phi_{2}$).
+Further suppose that the second sugar is prefered,
+%and that the organism has a regulatory network that can deploy situation specific metabolic strategy.
+and that
+depending on $u_{1}$ and $u_{2}$ the organism will synthesize an appropriate $\phi_{1}$ and $\phi_{2}$. 
+Furthermore, assume this system contains at least two transcription factors, whose log concentrations are $\kappa_{1}$, $\kappa_{2}$, $\dots \kappa_{n}$.
+Minimally such a system may have the architecture,
+%\begin{align*}
+%    \mathcal{S}_{\min} = \left\{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 0 & -1 \\ 0 & 0 \end{bmatrix} \begin{bmatrix} \kappa_{1} \\ \kappa_{2} \end{bmatrix} (t) + \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \begin{bmatrix} u_{1} \\ u_{2} \end{bmatrix} (t) \\[11pt]
+%    \phi(t) &= \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \begin{bmatrix} \kappa_{1} \\ \kappa_{2} \end{bmatrix} (t) \end{array} \right.
+%\end{align*}
+  $\Sys_{\min}(U) = (U \left[ \begin{smallmatrix} 0 & -1 \\ 0 & 0 \end{smallmatrix} \right] U^{-1}, U 
+    %\left[\begin{smallmatrix} 1 & 0 \\ 0 & 1 \end{smallmatrix}\right]
+      ,
+     % \left[\begin{smallmatrix} 1 & 0 \\ 0 & 1 \end{smallmatrix}\right]
+        U^{-1})$. 
+%
+%The impulse response of the system is,
+%$h(t) = \left[\begin{smallmatrix} 1 & - t \\ 0 & 1 \end{smallmatrix}\right]$.
+    We can find alternative equivalent systems by changning coordinates ($U \rightarrow U'$) or, more generally, by applying the Kalman decomposition \ref{apx:kalman}.
+To illustrate that phenotypic invariance does not require dimensional invariance, we apply the Kalman decomposition to $\Sys_{\min}$ to find
+    %This may happen if a system is co-opted from another, or may be a consequence of gene duplication and deletion. 
+  $\Sys(\small D_{1-3},V) \! =  \! \left( \! \begin{array}{ll}
+     V \!
+\left[\begin{smallmatrix} D_{1} & D_{2} \\ 0 & A \end{smallmatrix}\right] \!
+V^{-1}, V \!
+\left[\begin{smallmatrix} D_{3} \\ B \end{smallmatrix}\right], \left[\begin{smallmatrix} 0 & C \end{smallmatrix}\right] \!
+V^{-1} \end{array} \!
+\right)
+$, both of which are in $\allS$
+%  \begin{align*}
+%    \Sys(X_{1-3},V) = \left( \begin{array}{ll}
+%     V \left[\begin{smallmatrix} X_{1} & X_{2} \\ 0 & A_{\min} \end{smallmatrix}\right] V^{-1}, V \left[\begin{smallmatrix} X_{3} \\ B_{\min} \end{smallmatrix}\right], \left[\begin{smallmatrix} 0 & C_{\min} \end{smallmatrix}\right] V^{-1} \end{array} \right)
+%    \end{align*}
+  ($V$ can be any $4$-dimensional and invertible matrix, $D_{1-3} \in \R^{2 \times 2}$, and $(A,B,C) \in \Sys_{\min}$).
+%So for example, some system can be wired as follows, and still be input-output equivalent to the minimal metabolic system $\mathcal{S}_{\min}$:
+% \begin{align*}
+%   A' &= \begin{bmatrix}
+%     1.6923  &  1.5385 &  -2.6154 &  -2 \\
+%     0.8462  &  0.7692 &  -1.3077 &  -1 \\
+%     1.2692  &  1.1538 &  -1.9615 &  -1.5 \\
+%     0.4231  &  0.3846 &  -0.6538 &  -0.5
+%   \end{bmatrix} \\
+%   B' &= \begin{bmatrix} 4 & 4 \\ 2 & 2 \\ 3 & 3 \\ 1 & 3 \end{bmatrix} \\
+%     C' &= \begin{bmatrix}  0.2692  &  0.1538  &  0.0385 &  -0.5 \\
+%     -0.4231 &  -0.3846 &   0.6538  &  0.5 \end{bmatrix}
+% \end{align*}
+%
+%Despite the present example consisting of a minimal $2 \times 2$ and a non-minimal $4 \times 4$ system, 
+%any $m$-dimensional system can be constructed using this method 
+%-- applying a change of coordinates to the Kalman decomposition 
+%-- to construct a mechanistically different system with identical phenotypic dynamics. 
+%
+%When applying the Kalman decomposition to real biological networks, one may have to restrict the free parameters, as physiology and biochemistry might not allow all mathematically possible coefficients.
+We note that as there are so many degrees of freedom, it may be difficult to infer system function solely from structure.
+\end{example}
 
@@
 \label{apx:H_calc}
+
+\plr{Fix this calculation to include derivatives wrt to $B$ and $C$ also.}
 

### 2017-10-21 kalman_walk_paper.tex
rewrote apx about limiting OU process

@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Genetic drift with a multivariate trait}
@@
 
-For completeness, we provide a brief argument of how the population mean
-moves under genetic drift
+For completeness, we provide a brief exposition of how a population 
+evolves due to genetic drift
 with a quantitative genetics model,
@@
 
+Suppose that the population is distributed in trait space
+as a Gaussian with covariance matrix $\Sigma$ and mean $\mu$,
+whose density we write as $f(\cdot;\Sigma,\mu)$.
+Selection has the effect of multiplying this density by the fitness function and renormalizing,
+so that if expected fitness of $x$ is proportional to $f(x;U,0)$,
+then the distribution post-selection
+has density at $x$ proportional to $f(x;\Sigma,\mu) f(x;U,0)$.
+By the computation below (``Completing the square''),
+the result is a Gaussian distribution
+with covariance matrix $(\Sigma^{-1} + U^{-1})^{-1}$ 
+and mean $(\Sigma^{-1}+U^{-1})^{-1} \Sigma^{-1} \mu$.
+Importantly, if $U$ is not invertible, then by $U^{-1}$ 
+we mean $(U^T U)^{-1} U^T$, Moore-Penrose pseudoinverse of $U$,
+so that $U U^{-1}$ is equal to the projection matrix
+onto the span of the columns of $U$.
+
+After selection, we have reproduction:
+suppose this occurs as in the infinitesimal model \citep{infinitesimal},
+so that each offspring of parents with traits $x$ and $y$
+is drawn independently from a Gaussian distribution 
+with mean $(x+y)/2$ and covariance matrix $N$.
+Here, $N$ is the contribution of ``segregation variance''.
+If $\widetilde \Sigma = (\Sigma^{-1} + U^{-1})^{-1}$ 
+is the covariance matrix of the parents post-selection,
+then the distribution of offspring will again be Gaussian,
+with mean equal to that of the parents
+and covariance matrix $\widetilde \Sigma/2 + N$.
+
+In summary, a generation under this model modifies the mean ($\mu$)
+and covariance matrix ($\Sigma$) of a population as follows:
+\begin{align*}
+    \mu &\mapsto \mu' = (\Sigma^{-1}+U^{-1})^{-1} \Sigma^{-1} \mu \\
+    \Sigma &\mapsto \Sigma' = \frac{1}{2} (\Sigma^{-1} + U^{-1})^{-1} + N .
+\end{align*}
+What measures are stable under this transformation?
+The condition $\mu = \mu'$ reduces to $\Sigma U^{-1} \mu = 0$;
+if we assume $N$ and therefore $\Sigma$ are of full rank,
+then this happens if and only if $\mu$ is not in the span of the columns of $U$,
+i.e., if $\mu$ lies in a neutral direction.
+The condition $\Sigma' = \Sigma$ can also be solved,
+at least numerically.
+After rearrangement, it reduces to
+$\Sigma U^{-1} \Sigma + (I/2 - N U^{-1}) \Sigma = N$.
+We can find a more explicit description
+if we assume that $x^T U x = \sum_{i=1}^k x_i^2$,
+i.e., that selection only cares about the first $k$ coordinates,
+and then with no interactions between traits.
+If so, the condition $\Sigma' = \Sigma$
+can be written in block form as
+\begin{align*}
+    \begin{bmatrix}
+        \Sigma_{11}^2 + (I/2 - N_{11}) \Sigma_{11}
+        & \Sigma_{11} \Sigma_{12} + (I/2 - N_{11})\Sigma_{12} \\
+        \Sigma_{12}^T \Sigma_{11} + \Sigma_{12}^T/2 - N_{12}^T \Sigma_{11}
+        & \Sigma_{22} - N_{12}^T \Sigma_{12} 
+    \end{bmatrix}
+    &=
+    \begin{bmatrix}
+        N_{11} & N_{12} \\
+        N_{12}^T & N_{22}
+    \end{bmatrix} .
+\end{align*}
+The first equation, $\Sigma_{11}^2 + (I/2 - N_{11}) \Sigma_{11} = N_{11}$,
+can be solved with the quadratic formula: 
+$$\Sigma_{11} = (N_{11} - I/2 + Q)/2$$
+for any $Q$ that commutes with $N_{11}$
+and is a solution to $Q^2 = (N_{11} - I/2)^2 + 4 N_{11}$.
+Since we need $\Sigma$ to be positive definite,
+we take the solution with positive eigenvalues.
+Given $\Sigma_{11}$, the remaining components are
+\begin{align*}
+    \Sigma_{12} 
+    &= 
+    \left( \Sigma_{11} + I/2 - N_{11} \right)^{-1} N_{12} \\
+    \Sigma_{22}
+    &=
+    N_{12}^T \Sigma_{12} + N_{22} .
+\end{align*}
+Importantly, the mean $\mu$ does not affect 
+either how the covariance matrix moves,
+or its stable shape.
+
+Above we have described the \emph{expected} motion of the mean and covariance..
+However, random resampling will introduce noise.
+Suppose that a population of $N$ individuals
+behaves approximately as described above.
+By the above,
+we may expect that the covariance matrix stays close to a constant value $\Sigma$,
+computed from $N$ and $U$ as above,
+so that we need only consider motion of the mean, $\mu$.
+Since we take a sample of size $N$ to construct the next generation,
+the next generation's mean is drawn from a Gaussian distribution with mean $\mu'$
+and covariance matrix $\Sigma/N$.
+Defining $\Gamma = (I - (I + \Sigma U^{-1})^{-1})$,
+this can be written as
+\begin{align*}
+    \mu' - \mu = \Gamma \mu + \epsilon/\sqrt{N} ,
+\end{align*}
+where $\epsilon$ is a multivariate Gaussian with mean zero and covariance matrix $\Sigma$.
+Let $\mu(k)$ denote the mean in the $k^\text{th}$ generation,
+and suppose that $\mu$ differs from optimal by something of order $1/N$:
+if $\nu(t) = N \mu(tN)$,
+then the previous equation implies that as $N \to \infty$, 
+in the limit $\nu$ solves the It\^{o} equation
+\begin{align*}
+    d \nu(t) = \Gamma \nu(t) dt + \Sigma^{1/2} dW(t) ,
+\end{align*}
+where now $W(t)$ is a multivariate white noise.
+This has an explicit solution as a multivariate Ornstein-Uhlenbeck process:
+\begin{align*}
+    \nu(t) = e^{-t \Gamma} \nu_0 + \int_0^t e^{-(t-s) \Gamma} \Sigma^{1/2} dW(s).
+\end{align*}
+% Note that after some algebra, $\Gamma = \Sigma U^{-1} (I - \Sigma U^{-1})^{-1}$.
+The asymptotic variance of this process in the direction $z$ is
+\begin{align} \label{eqn:limiting_cov}
+    \lim_{t \to \infty} \var[\nu(t) \cdot z]
+    &=
+    \int_0^\infty z^T e^{-s \Gamma} \Sigma e^{-s \Gamma} z ds ,
+\end{align}
+which is infinite iff $\Gamma z = 0$,
+which occurs iff $Uz=0$.
+In other words, population mean trait values lie away from the optimal set
+by a Gaussian displacement of order $1/N$
+with a covariance matrix given by equation \eqref{eqn:limiting_cov}.
+
+\plri{Now write what this means intuitively.}
+
 \paragraph{Completing the square}
@@
     (A + B)
-    \left( x - (A + B)^{-1} A y \right)
+    \left( x - (A + B)^{-1} A y \right) \\
+    &\qquad{}
     + \text{(terms that don't depend on $x$)} .
 \end{align*}
-Therefore, if $f(x;\Sigma,y)$ is the density of a Gaussian with mean $y$ and covariance matrix $\Sigma$
-then substituting $A=\Sigma^{-1}$ and $B=U^{-1}$ above,
-\begin{align*}
-    \frac{ f(x;\Sigma,y) f(x;U,0) }{\int_x f(z;\Sigma,y) f(z;U,0) dz}
+Therefore, 
+by substituting $A=\Sigma^{-1}$ and $B=U^{-1}$,
+\begin{align*}
+    \frac{ f(x;\Sigma,y) f(x;U,0) }{\int f(z;\Sigma,y) f(z;U,0) dz}
     &=
@@
 \end{align*}
-
-Now suppose that the population is distributed in genotype space
-as a Gaussian with covariance matrix $\Sigma$ and mean $y$.
-Selection has the effect of multiplying this density by the fitness function and renormalizing,
-so that if expected fitness of $x$ is proportional to $f(x;U,z)$
-then the above argument shows that the next generation will be sampled from a Gaussian distribution
-with covariance matrix $(\Sigma^{-1} + U^{-1})^{-1}$ 
-and mean $z + (\Sigma^{-1}+U^{-1})^{-1} \Sigma^{-1} (y-z)$
-Taking a sample of size $N$ to construct the next generation 
-will produce something close to this but with a slightly (stochastically) deviating mean.
-The next generation's mean is drawn from a Gaussian distribution with mean
-with covariance matrix $(\Sigma^{-1} + U^{-1})^{-1}/N$ 
-and mean $z + (\Sigma^{-1}+U^{-1})^{-1} \Sigma^{-1} (y-z)$.
-
-Roughly, what is this doing?
-Suppose that the population mean differs from the optimum by $\epsilon$,
-that $\Sigma = \sigma^2 I$ and $U = I/\beta^2$ (so, stablizing selection happening on a distance scale of $\beta$).
-Then the population mean gets closer to the optimum on average, moving to
-$\epsilon/(1 + \sigma^2 \beta^2)$
-and adds noise of size $(1/\beta) \sigma/\sqrt{N \sigma^2 + N 1/\beta^2}$.
-At equilibrium, these two movements will be of the same order,
-so that $\epsilon$ is of order $(\sigma/\sqrt{N}) \sqrt{1+\sigma^2 \beta^2}$.
 

### 2017-10-21 kalman_walk_paper.tex
added argument about evolution of seg variance

@@
 is drawn independently from a Gaussian distribution 
-with mean $(x+y)/2$ and covariance matrix $N$.
-Here, $N$ is the contribution of ``segregation variance''.
+with mean $(x+y)/2$ and covariance matrix $R$.
+Here, $R$ is the contribution of ``segregation variance''.
 If $\widetilde \Sigma = (\Sigma^{-1} + U^{-1})^{-1}$ 
@@
 with mean equal to that of the parents
-and covariance matrix $\widetilde \Sigma/2 + N$.
+and covariance matrix $\widetilde \Sigma/2 + R$.
 
@@
     \mu &\mapsto \mu' = (\Sigma^{-1}+U^{-1})^{-1} \Sigma^{-1} \mu \\
-    \Sigma &\mapsto \Sigma' = \frac{1}{2} (\Sigma^{-1} + U^{-1})^{-1} + N .
+    \Sigma &\mapsto \Sigma' = \frac{1}{2} (\Sigma^{-1} + U^{-1})^{-1} + R .
 \end{align*}
@@
 The condition $\mu = \mu'$ reduces to $\Sigma U^{-1} \mu = 0$;
-if we assume $N$ and therefore $\Sigma$ are of full rank,
+if we assume $R$ and therefore $\Sigma$ are of full rank,
 then this happens if and only if $\mu$ is not in the span of the columns of $U$,
@@
 After rearrangement, it reduces to
-$\Sigma U^{-1} \Sigma + (I/2 - N U^{-1}) \Sigma = N$.
+$\Sigma U^{-1} \Sigma + (I/2 - R U^{-1}) \Sigma = R$.
 We can find a more explicit description
@@
     \begin{bmatrix}
-        \Sigma_{11}^2 + (I/2 - N_{11}) \Sigma_{11}
-        & \Sigma_{11} \Sigma_{12} + (I/2 - N_{11})\Sigma_{12} \\
-        \Sigma_{12}^T \Sigma_{11} + \Sigma_{12}^T/2 - N_{12}^T \Sigma_{11}
-        & \Sigma_{22} - N_{12}^T \Sigma_{12} 
+        \Sigma_{11}^2 + (I/2 - R_{11}) \Sigma_{11}
+        & \Sigma_{11} \Sigma_{12} + (I/2 - R_{11})\Sigma_{12} \\
+        \Sigma_{12}^T \Sigma_{11} + \Sigma_{12}^T/2 - R_{12}^T \Sigma_{11}
+        & \Sigma_{22} - R_{12}^T \Sigma_{12} 
     \end{bmatrix}
@@
     \begin{bmatrix}
-        N_{11} & N_{12} \\
-        N_{12}^T & N_{22}
+        R_{11} & R_{12} \\
+        R_{12}^T & R_{22}
     \end{bmatrix} .
 \end{align*}
-The first equation, $\Sigma_{11}^2 + (I/2 - N_{11}) \Sigma_{11} = N_{11}$,
+The first equation, $\Sigma_{11}^2 + (I/2 - R_{11}) \Sigma_{11} = R_{11}$,
 can be solved with the quadratic formula: 
-$$\Sigma_{11} = (N_{11} - I/2 + Q)/2$$
-for any $Q$ that commutes with $N_{11}$
-and is a solution to $Q^2 = (N_{11} - I/2)^2 + 4 N_{11}$.
+$$\Sigma_{11} = (R_{11} - I/2 + Q)/2$$
+for any $Q$ that commutes with $R_{11}$
+and is a solution to $Q^2 = (R_{11} - I/2)^2 + 4 R_{11}$.
 Since we need $\Sigma$ to be positive definite,
@@
     &= 
-    \left( \Sigma_{11} + I/2 - N_{11} \right)^{-1} N_{12} \\
+    \left( \Sigma_{11} + I/2 - R_{11} \right)^{-1} R_{12} \\
     \Sigma_{22}
     &=
-    N_{12}^T \Sigma_{12} + N_{22} .
+    R_{12}^T \Sigma_{12} + R_{22} .
 \end{align*}
@@
 we may expect that the covariance matrix stays close to a constant value $\Sigma$,
-computed from $N$ and $U$ as above,
+computed from $R$ and $U$ as above,
 so that we need only consider motion of the mean, $\mu$.
@@
 
+%%%%%%%%%%%
+\subsection*{Evolution of segregation covariance}
+
+The description above does not completely describe how two diverging populations interact,
+since the amount of \emph{segregation variance}, quantified by $R$,
+will not stay constant.
+To get an idea of how this might change,
+suppose that a multivariate trait is determined by $L$ unlinked, biallelic loci,
+and that the $i^\mathrm{th}$ locus has two alleles with additive effects $\pm x_i$,
+so that begin homozygous for the $+$ allele contributes $+2x_i$ to the trait.
+For simplicity, we will neglect the effects of selection.
+\plri{fixup the below to be actually multivariate}
+If the $+$ allele at locus $i$ is at frequency $p_i$ in a population,
+then the mean and genetic variance of the trait in a diploid population with random mating is
+\begin{align*}
+    m &= 2 \sum_i p_i x_i \\
+    s^2 &= 2 \sum_i p_i (1-p_i) x_i^2 .
+\end{align*}
+
+Segregation variance between two parents depends on the loci at which either are heterozygous,
+and each locus contributes independently
+since alleles are additive.
+If the alleles are at Hardy-Weinberg proportions,
+then since segregation is a fair coin flip, a heterozygous locus contributes $x_i^2/4$ to the variance,
+and so the \emph{mean} segregation variance, averaging across parents, is
+% two parents, prob 2p(1-p) for prob the locus is heterozygous, variance 0.5(1-0.5)x^2.
+\begin{align*}
+    R_0(p) = \frac{1}{2} \sum_i p_i (1-p_i) x_i^2 .
+\end{align*}
+
+On the other hand,
+if the second parent came from a distinct population
+with frequencies $q_i$
+(an $F_1$ hybrid),
+this would be
+\begin{align*}
+    R_1(p,q) &= \frac{1}{4} \sum_i p_i^2 (1-p_i)^2 x_i^2 
+                + \frac{1}{4} \sum_i q_i^2 (1-q_i)^2 x_i^2 \\
+                &= (R_0(p) + R_0(q))/2 .
+\end{align*}
+If we assume that the populations are at equilibrium, $R_0(p) \approx R_0(q)$,
+and so $R_1(p,q) \approx R_0(p)$.
+
+Now consider an $F_2$ hybrid, where both parents are $F_1$
+and so each heterozygous at locus $i$ with probability $p_i (1-q_i) + (1-p_i) q_i$.
+Then
+\begin{align*}
+    R_2(p,q) &= \frac{1}{4} \sum_i \left(p_i (1-q_i) + (1-p_i) q_i \right) x_i^2  .
+\end{align*}
+Suppose that the two populations are slightly drifted from each other,
+with frequency difference $p_i-q_i = 2\epsilon_i$.
+Then,
+$$
+    p (1-q) + p (1-q)
+    &=
+    (u+\epsilon) (1-u+\epsilon) + (u-\epsilon) (1-u-\epsilon) \\
+    &=
+    2 u (1-u)
+    % + \epsilon (u + (1-u) - u - (1-u))
+    + 2 \epsilon^2 .
+$$
+If the frequencies have evolved neutrally in unconnected, Wright-Fisher populations 
+of effective size $N$ for $t$ generations from a common ancestor with allele frequency $u$,
+then $\epsilon$ has mean zero and variance roughly $u (1-u) t/N$.
+Still assuming the populations are at stationarity,
+so that $R_0$ is constant between the two,
+and taking the frequencies $p_i$ as a proxy for the ancestral frequencies $u_i$,
+this implies that we expect
+\begin{align*}
+    R_2 
+    &\approx 
+    R_0 + \frac{1}{2} \sum_i p_i (1-p_i) x_i^2 t/N \\
+    &=
+    \left( 1 + \frac{t}{N} \right) R_0 .
+\end{align*}
+On the other hand, the expected squared difference in trait \emph{means} here is
+\begin{align} \label{eqn:var_dx}
+    4 \sum_i p_i (1-p_i) x_i^2 t / N = 8 R_0 t /N .
+\end{align}
+This implies that under this model,
+segregation variance in $F_2$s between two populations
+is roughly increased by a factor of $1/8$ of the difference between their means.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Away from the optimum}

### 2017-10-23 kalman_walk_paper.tex
note on error in apx
minor latex corrections

@@
 \renewcommand{\P}{\mathbb{P}}
+\newcommand{\var}{\mathop{\mbox{Var}}}
+\newcommand{\cov}{\mathop{\mbox{Cov}}}
 \newcommand{\pda}{\frac{\partial}{\partial A_{ij}}}
@@
 then the mean and genetic variance of the trait in a diploid population with random mating is
+\plr{the mean should be $\sum_i x_i (2p_i-1)$}
 \begin{align*}
@@
 Then,
-$$
+\begin{align*}
     p (1-q) + p (1-q)
@@
     + 2 \epsilon^2 .
-$$
+\end{align*}
 If the frequencies have evolved neutrally in unconnected, Wright-Fisher populations 

### 2017-10-25 kalman_walk_paper.tex
added in more derivatives to apx

@@
 \newcommand{\cov}{\mathop{\mbox{Cov}}}
+\newcommand{\tr}{\mathop{\mbox{tr}}} % trace
 \newcommand{\pda}{\frac{\partial}{\partial A_{ij}}}
@@
 Selection has the effect of multiplying this density by the fitness function and renormalizing,
-so that if expected fitness of $x$ is proportional to $f(x;U,0)$,
+so that if expected fitness of $x$ is proportional to $\exp(-\|Lx\|^2/2)$,
 then the distribution post-selection
-has density at $x$ proportional to $f(x;\Sigma,\mu) f(x;U,0)$.
+has density at $x$ proportional to $f(x;\Sigma,\mu) \exp(-\|Lx\|^2/2)$.
 By the computation below (``Completing the square''),
 the result is a Gaussian distribution
-with covariance matrix $(\Sigma^{-1} + U^{-1})^{-1}$ 
-and mean $(\Sigma^{-1}+U^{-1})^{-1} \Sigma^{-1} \mu$.
-Importantly, if $U$ is not invertible, then by $U^{-1}$ 
-we mean $(U^T U)^{-1} U^T$, Moore-Penrose pseudoinverse of $U$,
-so that $U U^{-1}$ is equal to the projection matrix
-onto the span of the columns of $U$.
+with covariance matrix $(\Sigma^{-1} + L^T L)^{-1}$ 
+and mean $(\Sigma^{-1}+L^T L)^{-1} \Sigma^{-1} \mu$.
+% Importantly, if $U$ is not invertible, then by $U^{-1}$ 
+% we mean $(U^T U)^{-1} U^T$, Moore-Penrose pseudoinverse of $U$,
+% so that $U U^{-1}$ is equal to the projection matrix
+% onto the span of the columns of $U$.
 
@@
 Here, $R$ is the contribution of ``segregation variance''.
-If $\widetilde \Sigma = (\Sigma^{-1} + U^{-1})^{-1}$ 
+If $\widetilde \Sigma = (\Sigma^{-1} + L^T L)^{-1}$ 
 is the covariance matrix of the parents post-selection,
@@
 \begin{align*}
-    \mu &\mapsto \mu' = (\Sigma^{-1}+U^{-1})^{-1} \Sigma^{-1} \mu \\
-    \Sigma &\mapsto \Sigma' = \frac{1}{2} (\Sigma^{-1} + U^{-1})^{-1} + R .
+    \mu &\mapsto \mu' = (\Sigma^{-1}+L^T L)^{-1} \Sigma^{-1} \mu \\
+    \Sigma &\mapsto \Sigma' = \frac{1}{2} (\Sigma^{-1} + L^T L)^{-1} + R .
 \end{align*}
 What measures are stable under this transformation?
-The condition $\mu = \mu'$ reduces to $\Sigma U^{-1} \mu = 0$;
+The condition $\mu = \mu'$ reduces to $\Sigma L^T L \mu = 0$;
 if we assume $R$ and therefore $\Sigma$ are of full rank,
-then this happens if and only if $\mu$ is not in the span of the columns of $U$,
+then this happens if and only if $\mu$ is in the null space of $L$,
 i.e., if $\mu$ lies in a neutral direction.
@@
 After rearrangement, it reduces to
-$\Sigma U^{-1} \Sigma + (I/2 - R U^{-1}) \Sigma = R$.
+$\Sigma L^T L \Sigma + (I/2 - R L^T L) \Sigma = R$.
 We can find a more explicit description
-if we assume that $x^T U x = \sum_{i=1}^k x_i^2$,
+if we assume that $x^T L^T L x = \sum_{i=1}^k x_i^2$,
 i.e., that selection only cares about the first $k$ coordinates,
@@
 we may expect that the covariance matrix stays close to a constant value $\Sigma$,
-computed from $R$ and $U$ as above,
+computed from $R$ and $L$ as above,
 so that we need only consider motion of the mean, $\mu$.
@@
 and covariance matrix $\Sigma/N$.
-Defining $\Gamma = (I - (I + \Sigma U^{-1})^{-1})$,
+Defining $\Gamma = (I - (I + \Sigma L^T L)^{-1})$,
 this can be written as
@@
 \end{align*}
-% Note that after some algebra, $\Gamma = \Sigma U^{-1} (I - \Sigma U^{-1})^{-1}$.
+% Note that after some algebra, $\Gamma = \Sigma L^T L (I - \Sigma L^T L)^{-1}$.
 The asymptotic variance of this process in the direction $z$ is
@@
 which is infinite iff $\Gamma z = 0$,
-which occurs iff $Uz=0$.
+which occurs iff $Lz=0$.
 In other words, population mean trait values lie away from the optimal set
@@
 Therefore, 
-by substituting $A=\Sigma^{-1}$ and $B=U^{-1}$,
-\begin{align*}
-    \frac{ f(x;\Sigma,y) f(x;U,0) }{\int f(z;\Sigma,y) f(z;U,0) dz}
-    &=
-    f(x; (\Sigma^{-1} + U^{-1})^{-1}, (\Sigma^{-1}+U^{-1})^{-1} \Sigma^{-1} y) .
+by substituting $A=\Sigma^{-1}$ and $B=L^T L$,
+\begin{align*}
+    \frac{ f(x;\Sigma,y) \exp(-x^T L^T L x / 2)}{\int f(z;\Sigma,y) \exp(-z^T L^T L z / 2) dz}
+    &=
+    f(x; (\Sigma^{-1} + L^T L)^{-1}, (\Sigma^{-1}+L^T L)^{-1} \Sigma^{-1} y) .
 \end{align*}
@@
 
-\plr{Fix this calculation to include derivatives wrt to $B$ and $C$ also.}
-
 Suppose that $\rho(t) \ge 0$ is a weighting function on $[0,\infty)$
 so that fitness is a function of $L^2(\rho)$ distance of the impulse response from optimal.
-With $A_0$ a representative of the optimal set:
+With $h_0(t) = C_0 e^{tA_0} B_0$ a representative of the optimal set:
 \begin{equation}
     \begin{aligned}
-        D(A) 
+        D(A, B, C) 
         &:= 
-        \int_0^\infty \rho(t) \left| h_A(t) - h_{A_0}(t) \right|^2 dt \\
+        \int_0^\infty \rho(t) \left| h_A(t) - h_0(t) \right|^2 dt \\
         &:= 
-        \int_0^\infty \rho(t) \left| C e^{At} B - C e^{A_0 t} B \right|^2 dt \\
-        &= 
-        \int_0^\infty \rho(t) \left| C \left( e^{At} - e^{A_0 t} \right) B \right|^2 dt \\
-        &= 
-        \int_0^\infty \rho(t) C \left( e^{At} - e^{A_0 t} \right) B B^T \left( e^{At} - e^{A_0 t} \right)^T C^T dt
+        \int_0^\infty \rho(t) \left| C e^{At} B - C_0 e^{A_0 t} B_0 \right|^2 dt \\
+        &= 
+        \int_0^\infty \rho(t) \tr\left\{
+            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)^T
+            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)
+        \right\} dt \\
+        &= 
+        \int_0^\infty \rho(t) \tr\left\{
+            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)
+            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)^T
+        \right\} dt  .
+        % &= 
+        % \int_0^\infty \rho(t)  \tr\left\{
+        %         C e^{At} B B^T e^{A^T t} C^T 
+        %         - C_0 e^{A_0t} B_0 B^T e^{A^T t} C^T 
+        %         \right. \\ &\qquad \qquad \left. {}
+        %         - C e^{At} B B_0^T e^{A_0^T t} C_0^T 
+        %         + C_0 e^{A_0t} B_0 B_0^T e^{A_0^T t} C_0^T 
+        % \right\} dt \\
+        % \int_0^\infty \rho(t) \left| C \left( e^{At} - e^{A_0 t} \right) B \right|^2 dt \\
+        % &= 
+        % \int_0^\infty \rho(t) C \left( e^{At} - e^{A_0 t} \right) B B^T \left( e^{At} - e^{A_0 t} \right)^T C^T dt
     \end{aligned}
 \end{equation}
-How does this change with $A$?
+How does this change as we perturb about $(A_0, B_0, C_0)$?
+First we differentiate with respect to $A$, keeping $B=B_0$ and $C=C_0$ fixed.
 Since
@@
   \begin{aligned}
-      \frac{d}{du} D(A+uZ)\vert_{u=0}
-      &=
-        2 \int_0^\infty \rho(t) C \left( \int_0^t e^{As} Z e^{A(t-s)} ds \right) B B^T \left( e^{At} - e^{A_0 t} \right)^T C^T dt \\
-      &=
-        2 \int_0^\infty \rho(t) C \left( \int_0^t e^{As} Z e^{A(t-s)} ds \right) B \left( h_A(t) - h_{A_0}(t) \right)^T dt 
+      \frac{d}{du} D(A+uZ,B_0,C_0)\vert_{u=0}
+      &=
+        2 \int_0^\infty \rho(t) \tr\left\{ C_0 \left( \int_0^t e^{As} Z e^{A(t-s)} ds \right) B_0 B_0^T \left( e^{At} - e^{A_0 t} \right)^T C_0^T \right\} dt \\
+      &=
+        2 \int_0^\infty \rho(t) \tr\left\{ C_0 \left( \int_0^t e^{As} Z e^{A(t-s)} ds \right) B_0 \left( h_A(t) - h_0(t) \right)^T \right\} dt 
   \end{aligned}
@@
 and, by differentiating this and supposing that $A$ is on the optimal set,
-i.e., $h_A(t)=h_{A_0}(t)$, (so wolog $A=A_0$):
+i.e., $h_A(t)=h_0(t)$, (so wolog $A=A_0$):
 \begin{equation}
   \begin{aligned}
-      \calH(Y,Z) &:= \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0+uY+vZ)\vert_{u=v=0} \\
-      &=
-        \int_0^\infty \rho(t) C 
+      \calH^{A,A}(Y,Z) 
+      &:= 
+      \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0+uY+vZ,B_0,C_0)\vert_{u=v=0} \\
+      &=
+        \int_0^\infty \rho(t) \tr\left\{ C_0 
         \left( \int_0^t e^{A_0 s} Y e^{A_0 (t-s)} ds \right) 
-        B B^T 
+        B_0 B_0^T 
         \left( \int_0^t e^{A_0 s} Z e^{A_0 (t-s)} ds \right)^T
-        C^T dt  .
+        C_0^T \right\} dt  .
   \end{aligned}
 \end{equation}
-Here $\calH$ is the quadratic form underlying the Hamiltonian.
+
+The function $\calH$ will define a quadratic form.
+To illustrate the use of this, suppose that $B$ and $C$ are fixed.
+% Here $\calH$ is the quadratic form underlying the Hamiltonian.
 By defining $\Delta_{ij}$ to be the matrix with a 1 in the $(i,j)$th slot
@@
         \int_0^\infty
-            \rho(t) C U X^{ij}(t) U^{-1} B B^T (U^{-1})^T X^{k\ell}(t)^T U^T C^T
+            \rho(t) \tr\left\{ C U X^{ij}(t) U^{-1} B B^T (U^{-1})^T X^{k\ell}(t)^T U^T C^T \right\}
         dt .
@@
 as described (in different terms) in \citet{hansen1996translating}.
+
+More generally, $B$ and $C$ may also change.
+To extend this we need the remaining second derivatives of $D$.
+First, in $B$:
+\begin{equation}
+    \begin{aligned}
+        \mathcal{H}^{B,B}(Y,Z)
+        &:= 
+        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0,B_0+uY+vZ,C_0)\vert_{u=v=0} \\
+        &=
+        \frac{1}{2} \int_0^\infty \rho(t) \tr\left\{ 
+        C_0 e^{t A_0}
+        \frac{d}{du} \frac{d}{dv} 
+        (uY+vZ)
+        (uY+vZ)^T
+        \vert_{u=v=0} 
+        e^{t A_0^T} C_0^T 
+        \right\} dt  \\
+        &=
+        \frac{1}{2} \int_0^\infty \rho(t) \tr\left\{ 
+        C_0 e^{t A_0}
+        \left( Y Z^T + Z Y^T \right)
+        e^{t A_0^T} C_0^T 
+        \right\} dt  .
+  \end{aligned}
+\end{equation}
+Next, in $C$:
+\begin{equation}
+    \begin{aligned}
+        \mathcal{H}^{B,B}(Y,Z)
+        &:= 
+        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0,B_0,C_0+uY+vZ)\vert_{u=v=0} \\
+        &=
+        \frac{1}{2} \int_0^\infty \rho(t) \tr\left\{ 
+        B_0 e^{t A_0^T}
+        \frac{d}{du} \frac{d}{dv} 
+        (uY+vZ)^T
+        (uY+vZ)
+        \vert_{u=v=0} 
+        e^{t A_0} B_0
+        \right\} dt  \\
+        &=
+        \frac{1}{2} \int_0^\infty \rho(t) \tr\left\{ 
+        B_0 e^{t A_0^T}
+        \left( Y Z^T + Z Y^T \right)
+        e^{t A_0} B_0
+        \right\} dt  .
+  \end{aligned}
+\end{equation}
+Now, the mixed derivatives in $B$ and $C$:
+\begin{equation}
+    \begin{aligned}
+        \mathcal{H}^{B,C}(Y,Z)
+        &:= 
+        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0,B_0+uY,C_0+vZ)\vert_{u=v=0} \\
+        &=
+        \int_0^\infty \rho(t) \tr\left\{ 
+        Y e^{t A_0^T} C_0^T Z e^{t A_0} B_0
+        \right\} dt  .
+  \end{aligned}
+\end{equation}
+In $A$ and $B$
+\begin{equation}
+    \begin{aligned}
+        \mathcal{H}^{A,B}(Y,Z)
+        &:= 
+        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0+uY,B_0+vZ,C_0)\vert_{u=v=0} \\
+        &=
+        \int_0^\infty \rho(t) \tr\left\{ 
+        C_0 \left(\int_0^t e^{s A_0} Y e^{(t-s) A_0} ds \right) B_0
+        Z^T e^{t A_0} C_0
+        \right\} dt  ,
+  \end{aligned}
+\end{equation}
+and finally in $A$ and $C$:
+\begin{equation}
+    \begin{aligned}
+        \mathcal{H}^{A,C}(Y,Z)
+        &:= 
+        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0+uY,B_0,C_0+vZ)\vert_{u=v=0} \\
+        &=
+        \int_0^\infty \rho(t) \tr\left\{ 
+        C_0 \left(\int_0^t e^{s A_0} Y e^{(t-s) A_0} ds \right) B_0
+        B_0 e^{t A_0} Z
+        \right\} dt  .
+  \end{aligned}
+\end{equation}
 

### 2017-10-27 kalman_walk_paper.tex
added overleaf edits

--- 2017-10-23 kalman_walk_paper.tex
@@
 \newcommand{\ind}{\mathds{1}}
+\newcommand{\grad}{\nabla}
 
@@
        C_{\ro}^T \\
-       C_{\nrno}^T \\
-       0 
+       0 \\
+       C_{\nro}^T
    \end{array} \right] .
@@
 \begin{align}
+\label{eqn:distance}
 	D(h) = \left( \int_0^\infty \rho(t) \|h(t) - h_0(t)\|^2 dt \right)^{1/2} .
@@
 %%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Systems drift and the accumulation of incompatibilities}
+\section*{System drift and the accumulation of incompatibilities}
 
@@
 and that these mechanisms may fail to produce viable hybrids.
-Here we discuss the rate at which evolution shifts molecular mechanisms
-and ask whether this process is fast enough to be a significant driver of speciation.
-To do so, we explore a general quantitative genetic model in which a population drifts stochastically
-near a set of equivalent and optimal systems. We work with a population of effective size $N_e$. 
-\plri{note that this is more general than LTI model}
-
-At any given time, there will be a range of regulatory coefficients present in the population
-due to segregating genetic variants.
-Over many generations, even if selective pressures do not change,
-this range of networks will shift 
-as recombination, mutation, and demographic noise create new alleles and shift allele frequencies.
-%How much variation do we expect to find within a population?
-%Is this range limited by available variation or kept in check by selection?
-%How fast will a population explore the space of equivalent networks?
-%In this section we explore informally a general model for this situation,
-%in which a population drifts stochastically near a set of equivalent, optimal systems.
-%We work with a population of effective size $N_e$.
-
-Suppose that a set $x$ of coefficients that determine a system (this is $\Sys$ above),
-produce a phenotype $\Phi(x)$ (the time course of $\phi(t)$).
-There is an optimal phenotype $\optph$,
-and a set $\optx$ of ``optimal'' coefficients that produce this phenotype.
-Fitness depends on distance to the optimal phenotype --
-we will write the ``distance'' between phenotypes $\phi$ and $\psi$ as $\dph(\phi, \psi)$,
-measured so that the fitness of an organism with coefficents $x$ is $\fitx(x) = \exp(-\dph(\Phi(x),\optph)^2)$. 
-We will assume that the map $\Phi$ is smooth
-and that the optimal set $\optx$ is locally isomorphic to $\R^m$.
-\plr{say that better}
-
-%\paragraph{Offspring.}
-%\js{Not sure if this section is redundant, since reproduction is described above for the LTI model.}
-%Individuals are diploid; we assume that each haploid genome determines a set of coefficients,
-%and the individual's coefficients are the average of her two haploid values (no dominance).
-%This implies that the diploid population variance, $\sigma^2$, is one-half the haploid variance.
-%Each new gamete is produced from the parent's two haploid copies;
-%for simplicity we assume that the gamete inherits a random choice of one of the two parental copies,
-%and so $\sigma$ remains constant, up to a $1/N_e$ term.
-%A more general model including segregation variance \citep{barton_infinitesimal}
-%would result in the same qualitative conclusions.
-%\plri{but put this in the appendix?}
+This begs the question: does evolution shift molecular mechanisms
+fast enough to be a significant driver of speciation?
+To approach this question,
+we explore a general quantitative genetic model in which a population drifts stochastically
+near a set of equivalent and optimal systems
+due to the action of recombination, mutation, and demographic noise.
+Although this is motivated by the results on linear systems above,
+the quantitative genetics calculations are more general,
+and only depend on the presence of genetic variation and a continuous set of phenotypically equivalent systems.
+
+We will suppose that each organism's phenotype is determined by her vector of coefficients, denoted by $x=(x_1, x_2, \ldots, x_L)$,
+and that the corresponding fitness is determined by the distance of her phenotype to optimum.
+The optimum phenotype is unique, but is realized by many distinct $x$ -- those falling in the ``optimal set'' $\allS$.
+The phenotypic distance to optimum of an organism with coefficients $x$ is denoted $D(x)$.
+In the results above, $x = (A,B,C)$ and $D(x)$ is given by equation \eqref{eqn:distance}.
+determines a system (this is a system $(A,B,C)$ above) 
+with fitness determined by its phenotypic distance $D(x)$ from an optimum (this is analogous to $\allS$ above). 
+Concretely, we define the fitness at $x$ to be $\exp(-D(x)^2)$.
+% this assumption does not entail a loss of generality since $D$ is arbitrary.
+We will assume that in the region of interest, the map $D$ is smooth
+and that we can locally approximate the optimal set $\allS$ as a quadratic surface.
+As above, an individual's coefficients are given by averaging her parentally inherited coefficients and adding random noise due to segregation. 
+Concretely, we use the \emph{infinitesimal model} for reproduction \citep{barton_infinitesimal} --
+the offspring of parents at $x$ and $x'$ will have coefficients $(x+x')/2 + \varepsilon$,
+where $\varepsilon$ is a Gaussian displacement due to random assortment of parental alleles.
 
@@
 % plr how fast does it drift
+We work with a randomly mating population of effective size $N_e$. 
 If the regulatory coefficient population variation
-has standard deviation $\sigma$,
+has standard deviation $\sigma$ in a particular direction,
 since subsequent generations resample from this diversity,
@@
 Selection will tend to restrain this motion,
-but mean movement along the optimal set $\optx$ is unconstrained.
+but movement along the optimal set $\allS$ is unconstrained,
+and so we expect the population mean to drift along the optimal set like a particle diffusing.
 % (although perhaps complicated by recombination load \citep{recomb_load})
 The amount of variance in particular directions in coefficient space 
-depend on constraints imposed by selection and 
-the covariance between genetic variation between different coefficients (the $G$ matrix \citep{G_matrix}).
+depends on constraints imposed by selection and 
+correlations between the genetic variation underlying different coefficients (the $G$ matrix \citep{G_matrix}).
+It therefore seems reasonable to coarsely model the time evolution of population variation in regulatory coefficients as 
+a ``cloud'' of width $\sigma$ about the population mean, 
+which moves as an unbiased Brownian motion through the set of network coefficients that give the optimal phenotype.
+
 % covariance which may arise due to functional constraints and/or statistical linkage.
@@
 
-% It therefore seems reasonable to coarsely model the time evolution of population variation in regulatory coefficients as 
-% (a) a ``cloud'' of width X about the population mean, 
-% which (b) moves as an unbiased Brownian motion through the set of network coefficients that give the optimal phenotype.
-% In fact, the population mean will not produce exactly the optimal phenotype,
-% but it will be convenient to refer to this closest point on the optimal set as ``the population mean''.
-
-%To obtain a general quantitative picture, we need to know 
-We denote
-$\sigma_N$ and $\sigma_S$, the standard deviations of coefficient variation along and perpendicular to $\optx$ respectively,
-and $\gamma$, the scale on which phenotype changes moving away from $\optx$.
-Concretely, $\gamma$ is the inverse of the derivative of $\dph(\Phi(x+u z), \Phi(x))$ 
-with respect to $u$ for $x \in \optx$ and $z$ perpendicular to the tangent space at $x$.
+There will in general be different amounts of variation in different directions;
+to keep the discussion intuitive, we only discuss $\sigma_N$, the amount of variation in ``neutral'' directions
+(i.e., directions along $\allS$),
+and $\sigma_S$, the amount of variation in ``selected'' directions (perpendicular to $\allS$).
+The other relevant scale we denote by $\gamma$,
+which the scale on which distance to phenotypic optimum changes as $x$ moves away from the optimal set, $\allS$.
+Concretely, $\gamma$ is
+%the inverse of the derivative of 
+$1/(\frac{d}{du}D(x+uz))$ 
+with respect to $u$ where $x$ is optimal and $z$ is a ``selected'' direction perpendicular to $\allS$.
 With these parameters, a typical individual will have a fitness of around $\exp(-(\sigma_S/\gamma)^2)$.
+Of course, there are in general many possible neutral and selected directions;
+we take these values to be representative of the possible directions.
 
 \paragraph{Hybridization}
-The means of two allopatric populations separated for $T$ generations
-will be a distance of order $2\sigma_N \sqrt{T/N_e}$ apart along $\optx$.
+The means of two allopatric populations each of effective size $N_e$ separated for $T$ generations
+will be a distance roughly of order $2\sigma_N \sqrt{T/N_e}$ apart along $\optx$.
+(Consult figure \ref{fig:conceptual_fig} for a conceptual diagram.)
 A population of $F_1$ hybrids has one haploid genome from each,
 whose coefficients are averaged,
-and so will have mean system coefficients at the midpoint between their means,
-and variance equal to $\sigma^2$. 
-Each $F_2$ hybrid will be homozygous for one parental allele on averge at half of the loci in the genome,
+and so will have mean system coefficients at the midpoint between their means.
+Each $F_2$ hybrid will be homozygous for one parental allele on average at half of the loci in the genome,
 so the distribution of $F_2$s will have mean at the average of the two populations,
-as before,
-but variance equal to $\sigma^2 + z^2/2$, where $z$ is the distance between the parental populations. 
-These are depicted in figure \ref{fig:conceptual_fig}.
+but will have higher variance.
+Concretely, we expect the population of $F_1$s to have variance $\sigma^2_S$ in the selected direction
+(the same as within each parental population),
+but the population of $F_2$s will have variance of order $\sigma^2_S + \sigma^2_H T/N_e$,
+where $\sigma^2_H$ is a factor that depends on the genetic basis of the coefficients.
+\plr{say something about $\sigma_H$?}
+
+What are the fitness consequences?
+If a population has a Gaussian distribution of trait values with variance $\sigma^2$
+and fitness decays as $\exp(-x^2/2\gamma)$ away from the population mean,
+then the population mean fitness is 
+\begin{align*}
+   \int \frac{1}{\sigma \sqrt{2\pi}} e^{-\frac{x^2}{2\sigma^2}} e^{-\frac{x^2}{2\gamma^2}} dx
+   &=
+   \sqrt{\frac{1}{1 + (\sigma/\gamma)^2}}  \le \frac{1}{\sigma/\gamma} .
+\end{align*}
+If $\sigma$ is much smaller than $\gamma$, the fitness consequences of variation
+(i.e., genetic load) are small,
+but if $\sigma$ is of order $\gamma$, then mean fitness is inversely proportional to trait variance
+(i.e., $\gamma/\sigma$).
+This implies that we expect $\sigma_S < \gamma$
+but that once $T$ is of order $N_e (\gamma / \sigma_H)^2$, there will be loss of fitness
+in $F_2$ hybrids.
+Said another way,
+if we write $\mathcal{F}(T)$ as the mean fitness of an $F_2$ hybrid between two populations separated by $T$ generations,
+then fitness drops with the square root of $T$, measured in units of $N_e$ generations:
+\begin{align*}
+   \mathcal{F}(T) \le \frac{1}{\sigma_H/\gamma}\sqrt{\frac{N_e}{T}} .
+\end{align*}
+
+The fitness consequences of system drift above were due to increased \emph{variance} in $F_2$ hybrids.
+If the optimal set $\allS$ is curved (which it will be in general),
+then the mean $F_1$ offspring will fall away from the optimum slightly,
+leading to an additional fitness cost in hybrids.
+However,
+as shown in appendix \ref{apx:away_from_opt} and depicted in figure \ref{fig:osc_incompat},
+the deviation of the mean from the optimum grows proportionally to $T/N_e$.
+
+\begin{align*}
+\int_{-\infty}^\infty \frac{1}{\sigma\sqrt{2\pi}} e^{-\frac{(x-\mu)^2}{2\sigma^2}} e^{-\frac{x^2}{2\gamma^2}} dx
+&=
+\sqrt{\frac{1}{1 + \sigma^2/\gamma^2}} \exp\left\{-\frac{\mu^2}{\gamma^2}\left(\frac{1}{1 + \sigma^2/\gamma^2}\right)\right\} \\
+% &\approx
+% \frac{1}{\sigma/\gamma}\left(1 - \frac{\mu^2}{\gamma^2}\right) .
+\end{align*}
+With $\sigma = c_\sigma \gamma \sqrt{T/N_e}$
+and $\mu = c_\mu \gamma (T/N_e)^2$,
+this is
+$$
+\sqrt{\frac{1}{1 + c_\sigma^2 T/N_e}} \exp\left\{-c_\mu^2 \frac{T^2}{N_e^2} \left(\frac{1}{1 + c_\sigma^2 T/N_e}\right)\right\} \\
+$$
 
@@
 $\fit_2/\fit_0 = \exp(- T/(N_e\gamma^2))$.
-
-Note that this follows directly from the fact that $F_1$ and $F_2$ hybrid phenotypes diverge from $F_0$ phenotypes, linearly and inverse-quadratically, with respect to time (more precisely $T/N_e$).
-As such, removing assumptions about the form of the fitness function, this model still predicts hybrid $F_2$ phenotypes to diverge much faster than $F_1$s, initially.
 
@@
 
+%%%%%%%%%%%%
+\plri{insert commented-out discussion of what the actual genetic basis is}
+
 %%%%%%%%%%%%%%%%%%%%%%%%
@@
 Lastly, we show that hybrid gene networks, under neutral processes, break down as function of genetic distance, and may, in part, explain broad patterns of reproductive isolation among diverse phyla \citep{roux2016shedding}.
+
+Note that this follows directly from the fact that $F_1$ and $F_2$ hybrid phenotypes diverge from $F_0$ phenotypes, linearly and inverse-quadratically, with respect to time (more precisely $T/N_e$).
+As such, removing assumptions about the form of the fitness function, this model still predicts hybrid $F_2$ phenotypes to diverge much faster than $F_1$s, initially.
+So even if fitness isn't as above maybe you can see this in phenotypes.
 
@@
 as in \citet{lande1981models} or \citet{hansen1996translating}.
-These ignore details of the underlying genetic basis,
+These do not directly model underlying genetic basis,
 but developing a more accurate model is beyond the scope of this paper.
@@
 \paragraph{Completing the square}
-First note that 
+First note that if $A$ is symmetric,
 \begin{align*}
@@
 \end{align*}
-and so
+and so if $B$ is also symmetric and $A+B$ is invertible,
 \begin{align*}
@@
     &\qquad{}
-    + \text{(terms that don't depend on $x$)} .
-\end{align*}
-Therefore, 
+    + y^T A y - y^T A^T (A+B)^{-1} A y .
+\end{align*}
+Since the second line doesn't depend on $x$,
 by substituting $A=\Sigma^{-1}$ and $B=U^{-1}$,
@@
 
-Let two points  on $\optx$ be $x_1$ and $x_2$, let $\bar x = (x_1+x_2)/2$, and let $z=(x_2 - x_1)/2$.
-Then with $D \Phi$ and $D^2 \Phi$ the first and second derivatives of $\Phi$, respectively,
-then Taylor expanding about $x_1$ and $x_2$ finds that
-\begin{align*}
-    \Phi(\bar x) 
-    &= \Phi(x_1) + D\Phi(x_1) \cdot z + \frac{1}{2} z^T D^2 \Phi(x_1) z + O(\|z\|^3) \\
-    &= \Phi(x_2) - D\Phi(x_2) \cdot z + \frac{1}{2} z^T D^2 \Phi(x_2) z + O(\|z\|^3) .
-\end{align*}
-Now, since $\Phi(x_1) = \Phi(x_2) = \optph$ and
-\begin{align*}
-    D\Phi(x_2) = D\Phi(x_1) + 2 z^T D^2 \Phi(x_1) + O(\|z\|^2), \quad \text{and} \\
-    D^2\Phi(x_2) = D^2\Phi(x_1) + O(\|z\|), \quad \text{and} \\
+Let two points on $\allS$ be $x_1$ and $x_2$, let $\bar x = (x_1+x_2)/2$, and let $z=(x_2 - x_1)/2$.
+Then with $\grad D$ and $\grad^2 D$ the first and second derivatives of $D$, respectively,
+Taylor expanding about $x_1$ and $x_2$ finds that
+\begin{align*}
+    D(\bar x) 
+    &= D(x_1) + \grad D(x_1) \cdot z + \frac{1}{2} z^T \grad^2  D(x_1) z + O(\|z\|^3) \\
+    &=  D(x_2) - \grad D(x_2) \cdot z + \frac{1}{2} z^T \grad^2  D(x_2) z + O(\|z\|^3) .
+\end{align*}
+Now, since $ D(x_1) =  D(x_2) = \grad D(x_1) = \grad D(x_2) = 0$ and
+\begin{align*}
+    \grad D(x_2) = \grad D(x_1) + 2 z^T \grad^2  D(x_1) + O(\|z\|^2), \quad \text{and} \\
+    \grad^2 D(x_2) = \grad^2 D(x_1) + O(\|z\|), 
 \end{align*}
@@
 \begin{align*}
-    \Phi(\bar x) 
-    &= \optph - \frac{3}{2} z^T D^2 \Phi(x_1) z + O(\|z\|^3) .
-\end{align*}
-
+     D(\bar x) 
+    &= \optph - \frac{3}{2} z^T \grad^2  D(x_1) z + O(\|z\|^3) .
+\end{align*}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Gaussian load}
+
+Suppose that a population has a Gaussian distribution in trait space
+with mean $\mu$ and covariance matrix $\Sigma$,
+and that fitness of an individual at $x$ is $\exp(-\|Lx\|^2/2)$.
+Then, completing the square as above with $A=\Sigma^{-1}$, $y=\mu$, and $B=L^T L$,
+and defining $Q = (\Sigma^{-1} + L^T L)^{-1}$,
+\begin{align*}
+    &
+    \frac{1}{\sqrt{2 \pi}^n \det(\Sigma)^{1/2}} 
+        \int e^{-\frac{1}{2} x^T \Sigma^{-1} x} e^{-\frac{1}{2} x^T L^T L x} dx \\
+    &\qquad =
+    \frac{1}{\sqrt{2 \pi}^n \det(\Sigma)^{1/2}} 
+        \int e^{-\frac{1}{2}(x-Q \Sigma^{-1} \mu)^T Q^{-1} (x-Q \Sigma^{-1} \mu)} dx 
+        \\ &\qquad \qquad {}
+        \times e^{\mu^T \left( I - \Sigma^{-1} Q \right) \Sigma^{-1} \mu} \\
+    &\qquad =
+    \sqrt{\frac{\det(Q)}{\det(\Sigma)}}
+        \exp\left\{\mu^T \left( I - \Sigma^{-1} Q \right) \Sigma^{-1} \mu\right\} \\
+    &\qquad =
+    \sqrt{\frac{1}{\det(\Sigma)\det(\Sigma^{-1}+L^TL)}}
+        \exp\left\{\mu^T \left( I - (I + L^T L \Sigma)^{-1} \right) \Sigma^{-1} \mu\right\} \\
+\end{align*}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Differentiating the fitness function}
@@
 \plr{Fix this calculation to include derivatives wrt to $B$ and $C$ also.}
+
+\plr{Add refs to sensitivity analysis.}
 
@@
 %  \end{align*}
+
 \end{document}

### 2017-11-07 examples/util.R
function to compute fitness diff using eigen expansion (not currently correct)

@@ -1,0 +1,72 @@
+library(expm)
+
+system_diff <- function (A0, B0, C0, A1, B1, C1, g = 1/(4*pi)) {
+    # Evaluate
+    # \int_0^\infty e^{-tg} |C0 e^{t A0} B0 - C1 e^{t A1} B1|^2 dt
+    n0 <- nrow(A0)
+    n1 <- nrow(A1)
+    eig0 <- eigen(A0, symmetric=TRUE)
+    eig1 <- eigen(A1, symmetric=TRUE)
+    # this is the list of C v v^T B, where v is an eigenvector
+    z0 <- lapply(1:ncol(eig0$vectors), function (k) { C0 %*% eig0$vectors[,k] %*% crossprod(eig0$vectors[,k], B0) })
+    z1 <- lapply(1:ncol(eig1$vectors), function (k) { C1 %*% eig1$vectors[,k] %*% crossprod(eig1$vectors[,k], B1) })
+    ZZ <- get_zmat(z0, z0)
+    ZY <- get_zmat(z0, z1)
+    YY <- get_zmat(z1, z1)
+    LL <- 1/(g - outer(eig0$values, eig0$values, "+"))
+    LN <- 1/(g - outer(eig0$values, eig1$values, "+"))
+    NN <- 1/(g - outer(eig1$values, eig1$values, "+"))
+    return(sum(ZZ*LL) + sum(ZY*LN) + sum(YY*NN))
+}
+
+get_zmat <- function (z0, z1) {
+    # this is the matrix of tr(z[i]^T z[j])
+    n0 <- length(z0)
+    n1 <- length(z1)
+    ZZ <- matrix(0, nrow=n0, ncol=n1)
+    for (i in 1:n0) {
+        for (j in 1:n1) {
+            ZZ[i,j] <- sum(diag(crossprod(z0[[i]], z1[[j]])^2))
+        }
+    }
+    return(ZZ)
+}
+
+system_diff_slow <- function (A0, B0, C0, A1, B1, C1, g = 1/(4*pi), upper=10) {
+
+    h <- function (t,A,B,C) { sapply(t, function (tt) C %*% expm(tt*A) %*% B) }
+
+    f <- function (t) {
+        exp(-t/(4*pi)) * sum(( h(t,A0,B0,C0) - h(t,A1,B1,C1) )^2)
+    }
+
+    integrate(f, lower=0, upper=upper)$value
+}
+
+if (TRUE) {
+
[…]

### 2017-11-10 peturb_Kalman.r
code for peturbation of Kalman matrix

@@ -1,0 +1,80 @@
+sys_gen2 <- function (s=sigma) 
+{
+  A <- matrix(c(0,-1,1,0), nrow=2);
+  B <- matrix(c(1,1), ncol=1);
+  C <- matrix(c(1,0), nrow=1);
+  A00 <- matrix(c(0,0,0,0), nrow=2);
+  B00 <- matrix(c(0,0), ncol=1);
+  C00 <- matrix(c(0,0), nrow=1);
+  #A14 <- matrix(c(rnorm(4,mean=0, sd=1)),nrow=2);
+  A12 <- matrix(c(rnorm(4,mean=0, sd=s)),nrow=2);
+  Arno <- matrix(c(rnorm(4,mean=0, sd=s)),nrow=2);
+  A13 <- matrix(c(rnorm(4,mean=0, sd=s)),nrow=2);
+  #A24 <- matrix(c(rnorm(4,mean=0, sd=1)),nrow=2);
+  Anrno <- matrix(c(rnorm(4,mean=0, sd=s)),nrow=2);
+  
+  p <- rnorm(25,mean=0, sd=s);
+  P <- matrix(c(1,1-p[1],1-p[2],p[3],p[4],p[5],
+                0,p[1],p[2],-p[3],-p[4],-p[5],
+                0,0,p[7],p[8],p[9],p[10],
+                0,0,p[12],p[13],p[14],p[15],
+                0,0,p[17],p[18],p[19],p[20],
+                0,0,p[22],p[23],p[24],p[25]), nrow=6);
+  invP <- solve(P);
+  
+  AK <- (rbind(cbind(A,A00,A00),
+               cbind(A12,Arno,A13),
+               cbind(A00,A00, Anrno)));
+  BK <- matrix(c(1,1,0,0,0,0),ncol=1);
+  CK <- matrix(c(1,0,0,0,0,0),nrow=1);
+  KalC <- CK%*%P;
+  KalB <- P%*%BK;
+  KalA <- P%*%AK%*%invP
+
+output <- matrix(0,nrow=1000,ncol=2)
+for (i in 1:1000)
+{
+nam0 <- paste("new_KalA", i, sep = ""); 
+
+mA12 <- matrix(c(rnorm(4, mean=0, sd=0.01)),nrow=2,ncol=2)
+mArno <- matrix(c(rnorm(4, mean=0, sd=0.01)),nrow=2,ncol=2)
+mA13 <- matrix(c(rnorm(4, mean=0, sd=0.01)),nrow=2,ncol=2)
[…]

### 2017-11-10 kalman_walk_paper.tex
put computational exploration in a subdir

@@
 \end{align*}
-Since the second line doesn't depend on $x$,
-by substituting $A=\Sigma^{-1}$ and $B=U^{-1}$,
-\begin{align*}
-    \frac{ f(x;\Sigma,y) f(x;U,0) }{\int f(z;\Sigma,y) f(z;U,0) dz}
-    &=
-    f(x; (\Sigma^{-1} + U^{-1})^{-1}, (\Sigma^{-1}+U^{-1})^{-1} \Sigma^{-1} y) .
+Therefore, 
+by substituting $A=\Sigma^{-1}$ and $B=L^T L$,
+\begin{align*}
+    \frac{ f(x;\Sigma,y) \exp(-x^T L^T L x / 2)}{\int f(z;\Sigma,y) \exp(-z^T L^T L z / 2) dz}
+    &=
+    f(x; (\Sigma^{-1} + L^T L)^{-1}, (\Sigma^{-1}+L^T L)^{-1} \Sigma^{-1} y) .
 \end{align*}
@@
 
-\plr{Fix this calculation to include derivatives wrt to $B$ and $C$ also.}
-
 \plr{Add refs to sensitivity analysis.}
@@
 so that fitness is a function of $L^2(\rho)$ distance of the impulse response from optimal.
-With $A_0$ a representative of the optimal set:
+With $h_0(t) = C_0 e^{tA_0} B_0$ a representative of the optimal set:
 \begin{equation}
     \begin{aligned}
-        D(A) 
+        D(A, B, C) 
         &:= 
-        \int_0^\infty \rho(t) \left| h_A(t) - h_{A_0}(t) \right|^2 dt \\
+        \int_0^\infty \rho(t) \left| h_A(t) - h_0(t) \right|^2 dt \\
         &:= 
-        \int_0^\infty \rho(t) \left| C e^{At} B - C e^{A_0 t} B \right|^2 dt \\
-        &= 
-        \int_0^\infty \rho(t) \left| C \left( e^{At} - e^{A_0 t} \right) B \right|^2 dt \\
-        &= 
-        \int_0^\infty \rho(t) C \left( e^{At} - e^{A_0 t} \right) B B^T \left( e^{At} - e^{A_0 t} \right)^T C^T dt
+        \int_0^\infty \rho(t) \left| C e^{At} B - C_0 e^{A_0 t} B_0 \right|^2 dt \\
+        &= 
+        \int_0^\infty \rho(t) \tr\left\{
+            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)^T
+            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)
+        \right\} dt \\
+        &= 
+        \int_0^\infty \rho(t) \tr\left\{
+            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)
+            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)^T
+        \right\} dt  .
+        % &= 
+        % \int_0^\infty \rho(t)  \tr\left\{
+        %         C e^{At} B B^T e^{A^T t} C^T 
+        %         - C_0 e^{A_0t} B_0 B^T e^{A^T t} C^T 
+        %         \right. \\ &\qquad \qquad \left. {}
+        %         - C e^{At} B B_0^T e^{A_0^T t} C_0^T 
+        %         + C_0 e^{A_0t} B_0 B_0^T e^{A_0^T t} C_0^T 
+        % \right\} dt \\
+        % \int_0^\infty \rho(t) \left| C \left( e^{At} - e^{A_0 t} \right) B \right|^2 dt \\
+        % &= 
+        % \int_0^\infty \rho(t) C \left( e^{At} - e^{A_0 t} \right) B B^T \left( e^{At} - e^{A_0 t} \right)^T C^T dt
     \end{aligned}
 \end{equation}
-How does this change with $A$?
+How does this change as we perturb about $(A_0, B_0, C_0)$?
+First we differentiate with respect to $A$, keeping $B=B_0$ and $C=C_0$ fixed.
 Since

### 2017-11-10 examples/util.R

@@ -32,7 +32,7 @@ get_zmat <- function (z0, z1) {
     return(ZZ)
 }
 
-system_diff_slow <- function (A0, B0, C0, A1, B1, C1, g = 1/(4*pi), upper=10) {
+system_diff_slow <- function (A0, B0, C0, A1, B1, C1, g = 1/(4*pi), upper=100) {
 
     h <- function (t,A,B,C) { sapply(t, function (tt) C %*% expm(tt*A) %*% B) }
 

### 2017-11-10 sims/perturb_Kalman.r
perturbation code

@@ -1,3 +1,44 @@
+library(MASS)
+library(Matrix)
+library(expm)
+
+sigma <- 1
+#####################################
+h <- function (t,M) { sapply(t, function (tt) CK %*% expm(tt*M) %*% BK) }
+optimal_h <- function (t) { sin(t)+cos(t) }
+###################################
+Df <- function (A,t) {
+  exp(-t/(4*pi)) * ( h(t,M=A) - optimal_h(t) )^2
+}
+####################################
+D <- function (A, upper=10, ...) {
+  f <- function (t) { Df(A,t) }
+  integrate(f, lower=0, upper=upper, ...)$value
+}
+
+##################################
+recombine <- function (X,Y) 
+{
+  n <- dim(X)[1];
+  G <- matrix(0,nrow=n,ncol=n);
+  r <- matrix(c(rbinom(n^2,1,0.5)),nrow=n);
+  for (i in 1:n) 
+  {
+    for (j in 1:n)
+    {
+      if (r[i,j]==0)
+      {
+        G[i,j] = X[i,j]
+      }
+      else if (r[i,j] == 1)
+      {
+        G[i,j] = Y[i,j]
+      }
+    }
+  }
+  G
+}
+#####################################
 sys_gen2 <- function (s=sigma) 
 {
   A <- matrix(c(0,-1,1,0), nrow=2);
@@ -30,51 +71,55 @@ sys_gen2 <- function (s=sigma)
   KalC <- CK%*%P;
   KalB <- P%*%BK;
   KalA <- P%*%AK%*%invP
-
-output <- matrix(0,nrow=1000,ncol=2)
-for (i in 1:1000)
-{
-nam0 <- paste("new_KalA", i, sep = ""); 
-
-mA12 <- matrix(c(rnorm(4, mean=0, sd=0.01)),nrow=2,ncol=2)
-mArno <- matrix(c(rnorm(4, mean=0, sd=0.01)),nrow=2,ncol=2)
-mA13 <- matrix(c(rnorm(4, mean=0, sd=0.01)),nrow=2,ncol=2)
-mAnrno <- matrix(c(rnorm(4, mean=0, sd=0.01)),nrow=2,ncol=2)
-mp <- rnorm(25,mean=0,sd=0.01)
-
-new_A12 <- A12 + A12*mA12
-new_Arno <- Arno + Arno*mArno
-new_A13 <- A13 + A13*mA13
-new_Anrno <- Anrno #+ Anrno*mAnrno
[…]

### 2017-11-10 sims/perturb_Kalman.r
added tests and refactored

@@ -4,119 +4,124 @@ library(expm)
 
 sigma <- 1
 #####################################
-h <- function (t,M) { sapply(t, function (tt) CK %*% expm(tt*M) %*% BK) }
+h <- function (t, M, BK, CK) { sapply(t, function (tt) CK %*% expm(tt*M) %*% BK) }
 optimal_h <- function (t) { sin(t)+cos(t) }
 ###################################
-Df <- function (A,t) {
-  exp(-t/(4*pi)) * ( h(t,M=A) - optimal_h(t) )^2
+Df <- function (A,t, BK, CK) {
+  exp(-t/(4*pi)) * ( h(t, M=A, BK=BK, CK=CK) - optimal_h(t) )^2
 }
 ####################################
-D <- function (A, upper=10, ...) {
-  f <- function (t) { Df(A,t) }
+D <- function (A, BK, CK, upper=10, ...) {
+  f <- function (t) { Df(A,t, BK, CK) }
   integrate(f, lower=0, upper=upper, ...)$value
 }
 
 ##################################
-recombine <- function (X,Y) 
+recombine<- function (X,Y) 
 {
   n <- dim(X)[1];
-  G <- matrix(0,nrow=n,ncol=n);
   r <- matrix(c(rbinom(n^2,1,0.5)),nrow=n);
-  for (i in 1:n) 
-  {
-    for (j in 1:n)
-    {
-      if (r[i,j]==0)
-      {
-        G[i,j] = X[i,j]
-      }
-      else if (r[i,j] == 1)
-      {
-        G[i,j] = Y[i,j]
-      }
-    }
-  }
-  G
+  G <- X * r + Y * (1-r)
+  return(G)
 }
-#####################################
-sys_gen2 <- function (s=sigma) 
-{
-  A <- matrix(c(0,-1,1,0), nrow=2);
-  B <- matrix(c(1,1), ncol=1);
-  C <- matrix(c(1,0), nrow=1);
-  A00 <- matrix(c(0,0,0,0), nrow=2);
-  B00 <- matrix(c(0,0), ncol=1);
-  C00 <- matrix(c(0,0), nrow=1);
-  #A14 <- matrix(c(rnorm(4,mean=0, sd=1)),nrow=2);
-  A12 <- matrix(c(rnorm(4,mean=0, sd=s)),nrow=2);
[…]

### 2017-11-10 sims/test_perturb_Kalman.r

@@ -1,0 +1,52 @@
+source("perturb_Kalman.r")
+
+set.seed(23)
+
+check_system <- function (sys) {
+    obs_D <- abs(D(sys$A, sys$B, sys$C))
+    if (obs_D > 1e-7) {
+        cat("Error: D=", obs_D, "\n")
+        return(1)
+    }
+    return(0)
+}
+
+# check that D(optimal) = 0
+optimal_system <- list(
+  A = matrix(c(0,-1,1,0), nrow=2),
+  B = matrix(c(1,1), ncol=1),
+  C = matrix(c(1,0), nrow=1) )
+check_system(optimal_system)
+
+num_errors <- 0
+for (sigma in c(0.001, 0.1, 1.0)) {
+    for (eps in c(0.001, 0.1, 1.0)) {
+        cat(sprintf("Testing with sigma=%f and eps=%f\n", sigma, eps))
+        AL <- random_oscillator(sigma)
+        random_system <- assemble_oscillator(AL)
+
+        cat("check that random_oscillator returns an oscillator\n")
+        num_errors <- num_errors + check_system(random_system)
+
+        cat("check that random_oscillator does not change B and C\n")
+        num_errors <- num_errors + !(all(abs(random_system$B - c(1,1,rep(0,length(random_system$B)-2))) < 1e-16))
+        num_errors <- num_errors + !(all(abs(random_system$C - c(1,0,rep(0,length(random_system$B)-2))) < 1e-16))
+
+        pAL <- perturb_oscillator(AL, eps=eps)
+        perturbed_system <- assemble_oscillator(pAL)
+        num_errors <- num_errors + !(all(sapply(AL, dim) == sapply(pAL, dim)))
+
+        cat("check that perturb_oscillator returns an oscillator\n")
+        num_errors <- num_errors + check_system(perturbed_system)
+
+        cat("check that perturb_oscillator does not change B and C\n")
+        num_errors <- num_errors + !(all(abs(perturbed_system$B - c(1,1,rep(0,length(perturbed_system$B)-2))) < 1e-16))
+        num_errors <- num_errors + !(all(abs(perturbed_system$C - c(1,0,rep(0,length(perturbed_system$B)-2))) < 1e-16))
+    }
+}
+
+if (num_errors>0) {
+    stop("Failed with ", num_errors, " errors.")
+} else {
+    cat("All good.\n")
+}

### 2017-11-10 kalman_walk_paper.tex
overleaf edits

@@
 %
+% Rapid speciation despite conservation of phenotype
+%
+% How fast does network drift create incompatibilities?
+%
 % More than one way to grow a cat: speciation despite conservation of phenotype
@@
 %
-% Rapid speciation despite conservation of phenotype
-%
 % More than one way to grow a cat: regulatory network redundancy and speciation
@@
 % More than one way to grow a cat: can network drift lead to speciation?
-%
-% How fast does network drift create incompatibilities?
 %
@@
 and metabolism \citep{lavoie2009rearrangements, martchenko2007transcriptional, dalal2016transcriptional, christensen2011unique, hartl2007induction, alam2013aspergillus}.
+
+AND THESE?
+Over the last several years, several different computational approaches have been applied to study reproductive incompatibility and speciation.
+\citet{tulchinsky} simulated the evolution of a transcription factor and its binding site using a thermodynamic model.
+Their simulations suggest that the language by which a transcription factor recognizes its binding site can change, and potentially lead to hybrid incompatibility when allopatric populations employ divergent readout languages.
+This study, despite looking at gene regulation, does not analyze overall gene network architecture -- as we do here -- it only looks at the expression level of a single gene.
+Furthermore, they report reproductive isolation primarily following directional selection for a change in expression levels in each allopatric population; the evidence for reproductive isolation following balancing selection is much weaker.
+Johnson and Porter 2000 did not observe any hybrid fitness declines under stabilizing selection -- only under directional selection.
+Khatari et al, Tulchinsky et al, and Porter et al, all study hybrid incompatibility from a transcription factor/binding site interaction perspective, not from an overall network architecture perspective.
+Palmer and Feldman only see hybrid incompatibility in constant environments if the parental populations are relatively poorly adapted initially.
+Otherwise hybrids between two allopatric populations have fairly high fitnesses.
+MAKE SURE TO REF DMI LIT SOMEWHERE and Barton's paper.
 
@@
 but will have higher variance.
+The variance of $F_2$s can be shown to increase linearly with the square of the distance between
+parental population means
+under models of both simple and polygenic traits.
+This is suggested by figure \ref{fig:conceptual_fig} and shown in Appendix \ref{apx:seg_cov}.
 Concretely, we expect the population of $F_1$s to have variance $\sigma^2_S$ in the selected direction
 (the same as within each parental population),
-but the population of $F_2$s will have variance of order $\sigma^2_S + \sigma^2_H T/N_e$,
-where $\sigma^2_H$ is a factor that depends on the genetic basis of the coefficients.
-\plr{say something about $\sigma_H$?}
+but the population of $F_2$s will have variance of order $\sigma^2_S + 4 \omega \sigma^2_N T/N_e$,
+where $\omega$ is a factor that depends on the genetic basis of the coefficients.
+In a model of $p$ traits in which the optimal set $\allS$ has dimension $q$,
+% if each trait is controlled by a single locus, $\omega = XXX$,
+using the polygenic model of appendix \ref{apx:seg_cov}, $\omega = (p-q)/8$.
+If each trait is controlled by a single locus, as in figure \ref{fig:conceptual_fig},
+the value is similar.
 
 What are the fitness consequences?
+A population of $F_2$s will begin to be substantially less fit than the parentals
+once they differ from the optimum by a distance of order $\gamma$,
+i.e., once $\sqrt{4 \omega T/N_e} \approx \gamma / \sigma_N$.
+This implies that hybrid incompatibility among $F_2$s should appear on a time scale of 
+$N_e \gamma / \sigma_N$ generations.
+The $F_1$s will not suffer fitness consequences until the hybrid mean is further than $\gamma$ from the optimum;
+as shown in appendix \ref{apx:away_from_opt} (and suggested by figure \ref{fig:conceptual_fig},
+this deviation of the mean from optimum grows with the square of the distance between the parental populations,
+and so we expect fitness costs in $F_1$s to appear on a time scale of $N_e^2$ generations.
+
 If a population has a Gaussian distribution of trait values with variance $\sigma^2$
@@
 \end{align*}
-With $\sigma = c_\sigma \gamma \sqrt{T/N_e}$
-and $\mu = c_\mu \gamma (T/N_e)^2$,
-this is
+With $\sigma/\gamma = c_\sigma \sqrt{T/N_e}$
+and $\mu/\gamma = c_\mu T/N_e$,
+this is (check!!)
 $$
@@
 $$
-
-\plri{improve figure by putting labels on from the following}
-Suppose that two populations have drifted independently to differ by $z$,
-and that $z$ is of the same order as $\sigma$ but is smaller than $\gamma$.
-The mean $F_1$ is the average of the parental means,
-and since the first-order terms in the Taylor series vanish,
-has phenotype differing from the optimum by a distance of order $\|z\|^2$ 
-(see appendix \ref{apx:away_from_opt}).
-The mean $F_2$ is the same,
-but the standard deviation is of order $z$,
-so that up to lower order terms, 
-while the typical fitness of an individual in the original population is
-$\fit_0 = \exp(-(\sigma_s/\gamma)^2)$;
-of an $F_1$ is
-$\fit_1/\fit_0 = \exp(-(c_1 \sigma_N^2 T/N_e)^2)$;
-and of an $F_2$ is
-$\fit_2/\fit_0 = \exp(- T/(N_e\gamma^2))$.
 
@@
 
+If $\sigma_N \approx \sigma_S \approx \gamma$ and variance in $F_2$s increases linearly with distance betw parents,
+then fitness starts to drop at $T/4N_e$.
+
 %%%%%%%%%%%%
@@
 \section*{Discussion}
-%
-%\jss{Discussion guidelines:
-%\begin{itemize}
-%    \item Why is this important/useful?
-%    \item What are the assumptions and shortcomings of the research?
-%    \item Compare to other studies in the literature. 
-%    \item Future directions. 
-%    \item Wild speculations?
-%    \item Conclusion and overall impact. 
-%\end{itemize}}
-%
+
+%%% OUTLINE
+% - Summarize methods and main result on rate of speciation.
+% - What empirical evidence is there that this is actually happening? 
+%   * incompats due to misregulation (Noor, Witkopp...)
+%   * numerous species that are visually indistinguishable (skinks, ferns)
+%   * maybe also ref phylogenetic lit
+% - What other evidence could be collected?  Note could be just phenotype not fitness. (cite Matute)
+% - Is linearity an important assumption?
+% - We know that divergent selection and genetic conflict can easily lead to speciation; we've shown neutral drift can maybe surprisingly easily do so also; but it remains unclear the relative contributions of these.
+Note that although drift may be weaker in any one network there are lots of networks.
+Could motivate this with Orr's statement.
+Perhaps tie in spandrels.
+
+% we do cool stuff
 The complexity of biological systems has limited our understanding of their function and evolution.
@@
 
+% no, really, it's cool
 Theoretical models in evolution and population genetics often lack the molecular details of physiology or of the genotype-phenotype map. 
@@
 
+% it says something about speciation and DMIs
 We have suggested an interpretation of system identification: to see it as an evolutionarily neutral manifold, and not simply a computational nuisance. 
@@
 
+% summary and consequences of main result on speed of speciation
 Furthermore, using a quantitative genetic approach, we estimated that a genetically variable population will drift in neutral system space 
@@
 The consistency comes from gene networks localized to the sex chromosomes functioning as an $F_{2}$ hybrid cross within a diploid $F_{1}$ heterogamete as there is only one sex chromosome.
-
 %We also suggest that gene networks may not always use their components parsimoniously 
@@
 
+Do we see evidence of this?  Well, maybe: misregulation in hybrids in Witkopp, Noor, Barbash, etc,
+even with expression levels in parents being the same.
+
+% consequence: prediction for phenotype (not just fitness)
 Note that this follows directly from the fact that $F_1$ and $F_2$ hybrid phenotypes diverge from $F_0$ phenotypes, linearly and inverse-quadratically, with respect to time (more precisely $T/N_e$).
@@
 
+% nonlinearity
 What about nonlinearity?
-
-Note that the drift above could actually be caused by the optimal phenotype shifting.
-
-%\plr{this paragraph isn't very clear} \js{Should I delete it? I guess it's sort of just a vague ``yeah we know this method has shortcomings''.}
-%As Richard Levins opined, models in population biology face a trade-off among precision, realism, and generality \citep{levins1966strategy}.
+As Richard Levins opined, models in population biology face a trade-off among precision, realism, and generality \citep{levins1966strategy}.
 As Levins expects, any tractable and general model, such as the present one under discussion, will have limitations.
@@
 
+% selection moving around optimal set
+Note that the drift above could actually be caused by the optimal phenotype shifting.
+
 %This theoretical framework can easily be applied to other interesting questions in evolutionary biology not tackled presently: such as the evolution of linkage, the necessity of network complexity (does evolution tend towards Rube Goldberg or parsimonious network organization?), evolvability, structure/function inference, and intra-population context dependency of mutational effects, as well as many others.
 
-%\jss{literature comparison -- I think we should delete}
-%
-%Over the last several years, several different computational approaches have been applied to study reproductive incompatibility and speciation.
-\citet{tulchinsky} simulated the evolution of a transcription factor and its binding site using a thermodynamic model.
-Their simulations suggest that the language by which a transcription factor recognizes its binding site can change, and potentially lead to hybrid incompatibility when allopatric populations employ divergent readout languages.
-This study, despite looking at gene regulation, does not analyze overall gene network architecture -- as we do here -- it only looks at the expression level of a single gene.
-Furthermore, they report reproductive isolation primarily following directional selection for a change in expression levels in each allopatric population; the evidence for reproductive isolation following balancing selection is much weaker.
-Johnson and Porter 2000 did not observe any hybrid fitness declines under stabilizing selection -- only under directional selection.
-%\plr{citations:}
-%Khatari et al, Tulchinsky et al, and Porter et al, all study hybrid incompatibility from a transcription factor/binding site interaction perspective, not from an overall network architecture perspective.
-Palmer and Feldman only see hybrid incompatibility in constant environments if the parental populations are relatively poorly adapted initially.
-Otherwise hybrids between two allopatric populations have fairly high fitnesses.
+% J and P didn't get speciation in a neutral  model
+Discuss differences to Johnson and Porter.
+
+What is Ne?  In a metapopulation depends on migr rate: classic Turelli paper.
 
@@
 \subsection*{Evolution of segregation covariance}
+\label{apx:seg_cov}
 

### 2017-11-10 kalman_walk_paper.tex
more edits on overleaf

@@
 \usepackage{authblk}
-
+\usepackage{lineno}
 \usepackage{multicol}
@@
 \begin{document}
+\linenumbers
 
@@
 This implies that hybrid incompatibility among $F_2$s should appear on a time scale of 
-$N_e \gamma / \sigma_N$ generations.
+$N_e (\gamma / \sigma_N)^2 / (4 \omega)$ generations.
 The $F_1$s will not suffer fitness consequences until the hybrid mean is further than $\gamma$ from the optimum;
-as shown in appendix \ref{apx:away_from_opt} (and suggested by figure \ref{fig:conceptual_fig},
+as shown in appendix \ref{apx:away_from_opt} (and suggested by figure \ref{fig:conceptual_fig}),
 this deviation of the mean from optimum grows with the square of the distance between the parental populations,
@@
 
-If a population has a Gaussian distribution of trait values with variance $\sigma^2$
-and fitness decays as $\exp(-x^2/2\gamma)$ away from the population mean,
-then the population mean fitness is 
-\begin{align*}
-   \int \frac{1}{\sigma \sqrt{2\pi}} e^{-\frac{x^2}{2\sigma^2}} e^{-\frac{x^2}{2\gamma^2}} dx
-   &=
-   \sqrt{\frac{1}{1 + (\sigma/\gamma)^2}}  \le \frac{1}{\sigma/\gamma} .
-\end{align*}
-If $\sigma$ is much smaller than $\gamma$, the fitness consequences of variation
-(i.e., genetic load) are small,
-but if $\sigma$ is of order $\gamma$, then mean fitness is inversely proportional to trait variance
-(i.e., $\gamma/\sigma$).
-This implies that we expect $\sigma_S < \gamma$
-but that once $T$ is of order $N_e (\gamma / \sigma_H)^2$, there will be loss of fitness
-in $F_2$ hybrids.
-Said another way,
-if we write $\mathcal{F}(T)$ as the mean fitness of an $F_2$ hybrid between two populations separated by $T$ generations,
-then fitness drops with the square root of $T$, measured in units of $N_e$ generations:
-\begin{align*}
-   \mathcal{F}(T) \le \frac{1}{\sigma_H/\gamma}\sqrt{\frac{N_e}{T}} .
-\end{align*}
-
-The fitness consequences of system drift above were due to increased \emph{variance} in $F_2$ hybrids.
-If the optimal set $\allS$ is curved (which it will be in general),
-then the mean $F_1$ offspring will fall away from the optimum slightly,
-leading to an additional fitness cost in hybrids.
-However,
-as shown in appendix \ref{apx:away_from_opt} and depicted in figure \ref{fig:osc_incompat},
-the deviation of the mean from the optimum grows proportionally to $T/N_e$.
-
-\begin{align*}
+% If a population has a Gaussian distribution of trait values with variance $\sigma^2$
+% and fitness decays as $\exp(-x^2/2\gamma)$ away from the population mean,
+% then the population mean fitness is 
+% \begin{align*}
+%    \int \frac{1}{\sigma \sqrt{2\pi}} e^{-\frac{x^2}{2\sigma^2}} e^{-\frac{x^2}{2\gamma^2}} dx
+%    &=
+%    \sqrt{\frac{1}{1 + (\sigma/\gamma)^2}}  \le \frac{1}{\sigma/\gamma} .
+% \end{align*}
+% If $\sigma$ is much smaller than $\gamma$, the fitness consequences of variation
+% (i.e., genetic load) are small,
+% but if $\sigma$ is of order $\gamma$, then mean fitness is inversely proportional to trait variance
+% (i.e., $\gamma/\sigma$).
+% This implies that we expect $\sigma_S < \gamma$
+% but that once $T$ is of order $N_e (\gamma / \sigma_H)^2$, there will be loss of fitness
+% in $F_2$ hybrids.
+% Said another way,
+% if we write $\mathcal{F}(T)$ as the mean fitness of an $F_2$ hybrid between two populations separated by $T$ generations,
+% then fitness drops with the square root of $T$, measured in units of $N_e$ generations:
+% \begin{align*}
+%    \mathcal{F}(T) \le \frac{1}{\sigma_H/\gamma}\sqrt{\frac{N_e}{T}} .
+% \end{align*}
+
+For a more concrete prediction, suppose that the distribution among hybrids is Gaussian.
+A population whose trait distribution is Gaussian with mean $\mu$ and variance $\sigma$,
+has mean fitness 
+\begin{align} \label{eqn:gauss_fit}
 \int_{-\infty}^\infty \frac{1}{\sigma\sqrt{2\pi}} e^{-\frac{(x-\mu)^2}{2\sigma^2}} e^{-\frac{x^2}{2\gamma^2}} dx
 &=
-\sqrt{\frac{1}{1 + \sigma^2/\gamma^2}} \exp\left\{-\frac{\mu^2}{\gamma^2}\left(\frac{1}{1 + \sigma^2/\gamma^2}\right)\right\} \\
+\sqrt{\frac{1}{1 + \sigma^2/\gamma^2}} \exp\left\{-\frac{\mu^2}{\gamma^2}\left(\frac{1}{1 + \sigma^2/\gamma^2}\right)\right\} .
 % &\approx
 % \frac{1}{\sigma/\gamma}\left(1 - \frac{\mu^2}{\gamma^2}\right) .
-\end{align*}
-With $\sigma/\gamma = c_\sigma \sqrt{T/N_e}$
-and $\mu/\gamma = c_\mu T/N_e$,
-this is (check!!)
-$$
-\sqrt{\frac{1}{1 + c_\sigma^2 T/N_e}} \exp\left\{-c_\mu^2 \frac{T^2}{N_e^2} \left(\frac{1}{1 + c_\sigma^2 T/N_e}\right)\right\} \\
-$$
-
-%\subsection*{Parameter estimates for real systems}
+\end{align}
+This assumes a single trait, for simplicity; the multivariate case is done in appendix \ref{apx:gauss_load}.
+A population of $F_2$s will have, as above, variance $\sigma^2 = \sigma^2_S + 4 \omega \sigma^2_N T/N_e$.
+The mean diverges with the square of the distance between the parentals
+with a speed depending on the local geometry of the optimal set
+(calculated in appendix \ref{apx:H_calc}), so we set $\mu = c_\mu \gamma T/N_e$.
+The mean fitness in parental populations is as in equation \ref{eqn:gauss_fit} with $\mu=0$ and $\sigma = \sigma_S$.
+This implies that if we define $\fit_2(T)$ to be the mean relative fitness among $F_2$ hybrids
+between two populations separated by $T$ generations, 
+(i.e., the mean fitness divided by the mean fitness of the parents)
+then
+\begin{align} \label{eqn:fit2}
+\fit_2(T) = 
+  \left(1 + \frac{4 \omega (\sigma_N/\gamma)^2}{1 + (\sigma_S/\gamma)^2} \frac{T}{N_e} \right)^{-1/2} 
+     \exp\left\{-\left(c_\mu \frac{T}{N_e}\right)^2 \left(\frac{1}{1 + (\sigma_S/\gamma)^2 + 4 \omega (\sigma_N/\gamma)^2 T/N_e}\right)\right\} .
+\end{align}
+We discuss the implications of this expression in the next section.
+
+\plr{note: should be a plot of the components of this against $T/N_e$ later}
+
+%%%%%%%%%%
+\paragraph{Speciation rates under neutrality}
+\plr{continue editing here}
+(say something about here we interpret the big math expression in words)
+From equation \eqref{eqn:fit2} we can observe that
+time is always scaled in units of $N_e$ generations,
+the population standard deviations are always scaled by $\gamma$,
+and the most important term is the rate of accumulation of segregation variance,
+$4 \omega (\sigma_N/\gamma)^2$.
+All else being equal, this process will lead to speciation more quickly in smaller populations
+and in populations with more neutral genetic variation (larger $\sigma_N$).
+These parameters are related -- larger populations generally have more genetic variation --
+but since these details depend on the situation, we leave these separate.
+
+argue sometimes $\sigma_N \approx \sigma_S \approx \gamma$, especially if $N_e$ is big
+(``large species'')
+
+and that sometimes $\sigma_N \approx \sigma_S \ll \gamma$, but only if $N_e$ is small
+(``isolated population'')
+
+and that sometimes maaybe $\sigma_N \gg \sigma_S \approx \gamma$, if $N_e$ is big and there's a lot of recombination load
+(``species complex'')
+
+but not likely to be other combinations; make plot of \eqref{eqn:fit2} against $T/N_e$ (or even $T$?) in these three situations.
+
+\subsection*{Parameter estimates for real systems}
 %\js{I think we should move this section?}
@@
 
-\paragraph{Speciation rates under neutrality}
-%For cases where system structures and $\gamma$ can be inferred,
-%specific speciation rates can be estimated.
-%For other situations we offer a back of the envelope calculation.
-%
-%Consider a population that is distributed on $\mathcal{X}$ such that the magnitude of phenotypically
-%divergent $\sigma_S$ and neutral $\sigma_N$ diversity are constrained by selection;
-%$\sigma_S$ directly, such that $\sigma_S = \gamma$, and $\sigma_N$ indirectly via recombination load.
-%Two reproductively isolated populations will, on average, be a distance of $2 \sigma_N \sqrt{T/N_e}$.
-%If we assume that $\sigma_N$ is constrained by a function of $\sigma_S$,
-%neutral movement along $\mathcal{X}$ pushes hybrids off of the manifold by $\sigma_S$ per $\sigma_N$,
-%Lastly, if fitness is Gaussian: $\fit_2 = \exp\left(- d(x, x')^2/\beta^2\right)$,
-%and we assume that the strength of selection is only several multiples of the segregating divergent diversity,
-%such that $\sigma_S = c \beta$, we can simplify the expression estimating $F_2$ hybrid fitness drops
-%to be,
-%
-%
-%$F_2$s between allopatric populations will suffer an average fitness drop relative to their parents of,
-%  \begin{align}
-%    \exp\left(-\left( \frac{2\sigma_{N}}{\beta}\sqrt{\frac{T}{N_e}} \right)^{2} \right) .
-%  \end{align}
-
-%If the strength of selection is only a few multiples $\xi_1$ of the standing phenotypic diversity within a population $\sigma_S$, such that $\beta = \xi_1 \sigma_S$, and recombination load constrains the standing phenotype-neutral diversity $\sigma_N$ to be only a few multiples $\xi_2$ of $\sigma_S$, then $F_2$ fitness drop will be,
-%  \begin{align}
-%    \exp\left(- \xi^{2} \frac{T}{N_e} \right)
-%  \end{align}
-%  where $\xi := 2\xi_2/\xi_1$.
-  Assume that perpendicular variation segregating within a population is constrained by selection such that $\sigma_S/\gamma = \xi$, and that neutral variation is on the order of perpendicular variation, such that $\sigma_N = \sigma_S$.
-Since the typical distance between two individuals drawn from separate allopatric populations will be $z = 2 \sigma_N \sqrt{T/N_e}$, we expect a drop in hybrid fitness of
-  \begin{align}
-  \fit_2/\fit_0 = \exp \left(- 4\xi^2 \frac{T}{N_e} \right) .
-  \end{align}
-  Hybrid fitness depression will be significant when $z > \gamma$. 
-  If $\xi$ is in $[\sfrac{1}{5},1]$, speciation will occur on the order of $N_e$ generations, and possibly much faster.
-Note, however, that if intrapopulation variation is very small such that $\sigma_S \ll \gamma$, speciation rates will be slower.
-
-If $\sigma_N \approx \sigma_S \approx \gamma$ and variance in $F_2$s increases linearly with distance betw parents,
-then fitness starts to drop at $T/4N_e$.
-
 %%%%%%%%%%%%
@@
 \section{Gaussian load}
+\label{apx:gauss_load}
 

### 2017-11-11 speciation_rates.R
added plots on speciation rates

@@ -1,0 +1,37 @@
+
+f <- function (t, a, b, c_mu=1/20) {
+    # a = sigma_N^2 / gamma^2
+    # b = sigma_S^2 / gamma^2
+    # omega = 1
+    f1(t, a, b, c_mu) * f2(t, a, b, c_mu)
+}
+
+f2 <- function (t, a, b, c_mu=1/20) {
+    1/sqrt(1 + 4 * a * t / (1+b))
+}
+
+f1 <- function (t, a, b, c_mu=1/20) {
+    exp(-(c_mu * t)^2 / (1 + b + 4 * b * t))
+}
+
+plotf <- function (main, ...) {
+    tt <- seq(0, 10, length.out=200)
+    plot(tt, f(tt, ...), type='l', ylim=c(0,1), main=main,
+            xlab=expression(T/N[e]), ylab="F2 fitness")
+    lines(tt, f1(tt, ...), col='red', lty=3)
+    lines(tt, f2(tt, ...), col='green', lty=3)
+}
+
+pdf(file="speciation_rates.pdf", width=7, height=3, pointsize=10)
+
+layout(t(1:3))
+plotf(a=1, b=1, main="large population")
+plotf(a=.1, b=.1, main="small population")
+plotf(a=10, b=1, main="metapopulation")
+
+legend("topright", lty=c(1,3,3), col=c('black', 'red', 'green'),
+        legend=c("total fitness", "mean", "variance"))
+
+dev.off()
+
+

### 2017-11-11 kalman_walk_paper.tex
overleaf edits

@@
 then
+\plr{figure out how dimensions come in here}
 \begin{align} \label{eqn:fit2}
 \fit_2(T) = 
-  \left(1 + \frac{4 \omega (\sigma_N/\gamma)^2}{1 + (\sigma_S/\gamma)^2} \frac{T}{N_e} \right)^{-1/2} 
+  \left(1 + \frac{4 \omega (\sigma_N/\gamma)^2}{(1 + (\sigma_S/\gamma)^2)} \frac{T}{N_e} \right)^{-1/2} 
      \exp\left\{-\left(c_\mu \frac{T}{N_e}\right)^2 \left(\frac{1}{1 + (\sigma_S/\gamma)^2 + 4 \omega (\sigma_N/\gamma)^2 T/N_e}\right)\right\} .
@@
 \paragraph{Speciation rates under neutrality}
-\plr{continue editing here}
-(say something about here we interpret the big math expression in words)
+Above, equation \eqref{eqn:fit2} shows how fast hybrids become inviable 
+as the time that the parental populations are isolated increases;
+what does this tell us about speciation rates under neutrality?
 From equation \eqref{eqn:fit2} we can observe that
@@
 $4 \omega (\sigma_N/\gamma)^2$.
+In particular, $\omega$, the rate that hybrid variance increases with separation between parentals,
+is proportional to $p-q$, the number of ways mutation can move the system without changing the phenotype,
+i.e., the dimensionality of the optimal set.
+This implies that less constrained systems should drift faster,
+and so lead to speciation more quickly, given the same strength of selection ($\gamma$).
+\plr{revisit}
 All else being equal, this process will lead to speciation more quickly in smaller populations
@@
 
-argue sometimes $\sigma_N \approx \sigma_S \approx \gamma$, especially if $N_e$ is big
-(``large species'')
-
-and that sometimes $\sigma_N \approx \sigma_S \ll \gamma$, but only if $N_e$ is small
-(``isolated population'')
-
-and that sometimes maaybe $\sigma_N \gg \sigma_S \approx \gamma$, if $N_e$ is big and there's a lot of recombination load
-(``species complex'')
+% simgaN = sigmaS = gamma, large population
+Suppose in a large, genetically diverse population,
+the amount of heritable variation in the neutral and selected directions are roughly equal 
+($\sigma_N \approx \sigma_S$)
+but the overall amount of variation is (weakly) constrained by selection
+($\sigma_N \approx \gamma$).
+If so, then the first term of equation \eqref{eqn:fit2} is
+$1/\sqrt{1 + 2 \omega T/N_e} \approx 1 - \omega T/N_e$.
+If also $\omega=1$, then, for instance, after $0.1 N_e$ generations
+the average $F_2$ fitness has dropped by 10\% relative to the parentals.
+
+%  $\sigma_N \approx \sigma_S \ll \gamma$, ``isolated population''
+Consider instead a much smaller, isolated population
+whose genetic variation is primarily constrained by genetic drift,
+so that $\sigma_N \approx \sigma_S \ll \gamma$.
+Setting $a = (\sigma_N/\gamma)^2$ to be small,
+the fitness of $F_2$s is $\fit_2 \le 1/\sqrt{1 + 4 \omega a^2 T/N_e} \approx 1 - 2 \omega a T/N_e$.
+Since in small populations at mutation-drift equilibrium
+the amount of genetic variance ($\sigma_N$) is proportional to $N_e$,
+so this predicts the rate of decrease of hybrid fitness to be \emph{independent} of population size
+for small populations.
+\plr{look up what $\sigma_N^2$ is at mut-drift equilib}
+
+% $\sigma_N \gg \sigma_S \approx \gamma$, if $N_e$ is big and there's a lot of recombination load (``species complex'')
+In the other direction, consider large metapopulations (or isolated ``species complexes'')
+among which heritable variation is strongly constrained by selection
+(i.e., there is substantial recombination load),
+so that $\sigma_S \approx \gamma$ but $\sigma_N/\gamma$ is large.
+Then the fitness of $F_2$s is $\fit_2 \le 1/\sqrt{1 + 2 \omega a^2 T/N_e} \approx 1 - \omega a T/N_e$,
+and could be extremely rapid if $a$ is large.
 
 but not likely to be other combinations; make plot of \eqref{eqn:fit2} against $T/N_e$ (or even $T$?) in these three situations.
+
+\begin{figure}
+\begin{center}
+\includegraphics{speciation_rates}
+\caption{
+Mean drop if $F_2$ fitness relative to parental species,
+with $\omega=1$ and 
+\textbf{(A)} $\sigma_N = \sigma_S = \gamma$
+\textbf{(B)} $\sigma_N = \sigma_S = 0.1 \gamma$
+\textbf{(C)} $0.1 \sigma_N = \sigma_S = \gamma$.
+\label{fig:speciation_rates}}
+\end{center}
+\end{figure}
 
@@
 
+\jss{talk about speciation rate in small population results as compared to Katari and Goldstein (I think). they look at the influence of $N_e$ on speciation rate.}
 \section*{Acknowledgements}

### 2017-11-13 speciation_rates.R
another row to speciation rates fig

@@ -14,12 +14,14 @@ f1 <- function (t, a, b, c_mu=1/20) {
     exp(-(c_mu * t)^2 / (1 + b + 4 * b * t))
 }
 
-plotf <- function (main, ...) {
-    tt <- seq(0, 10, length.out=200)
-    plot(tt, f(tt, ...), type='l', ylim=c(0,1), main=main,
-            xlab=expression(T/N[e]), ylab="F2 fitness")
-    lines(tt, f1(tt, ...), col='red', lty=3)
-    lines(tt, f2(tt, ...), col='green', lty=3)
+plotf <- function (main, Ne=1.0, tmax=10,
+        ...) {
+    tt <- seq(0, tmax/Ne, length.out=200)
+    plot(tt*Ne, f(tt, ...), type='l', ylim=c(0,1), main=main,
+            xlab=if (Ne==1) { expression(T/N[e])} else { expression(T) }, 
+            ylab="F2 fitness")
+    lines(tt*Ne, f1(tt, ...), col='red', lty=3)
+    lines(tt*Ne, f2(tt, ...), col='green', lty=3)
 }
 
 pdf(file="speciation_rates.pdf", width=7, height=3, pointsize=10)
@@ -28,6 +30,25 @@ layout(t(1:3))
 plotf(a=1, b=1, main="large population")
 plotf(a=.1, b=.1, main="small population")
 plotf(a=10, b=1, main="metapopulation")
+
+legend("topright", lty=c(1,3,3), col=c('black', 'red', 'green'),
+        legend=c("total fitness", "mean", "variance"))
+
+dev.off()
+
+
+
+pdf(file="speciation_rates_2.pdf", width=7, height=6, pointsize=10)
+
+layout(matrix(1:6, nrow=2))
+plotf(a=1, b=1, main="large population")
+plotf(a=1, b=1, main="large population, Ne=1e5", Ne=1e5, tmax=1e4)
+
+plotf(a=.1, b=.1, main="small population")
+plotf(a=.1, b=.1, main="small population, Ne=1e3", Ne=1e3, tmax=1e4)
+
+plotf(a=10, b=1, main="metapopulation")
+plotf(a=10, b=1, main="metapopulation, Ne=1e7", Ne=1e7, tmax=1e4)
 
 legend("topright", lty=c(1,3,3), col=c('black', 'red', 'green'),
         legend=c("total fitness", "mean", "variance"))

### 2017-11-13 speciation_rates.R
updated fig

@@ -32,7 +32,7 @@ plotf(a=.1, b=.1, main="small population")
 plotf(a=10, b=1, main="metapopulation")
 
 legend("topright", lty=c(1,3,3), col=c('black', 'red', 'green'),
-        legend=c("total fitness", "mean", "variance"))
+        legend=c("total fitness", "shift of mean", "increase of variance"))
 
 dev.off()
 
@@ -51,7 +51,7 @@ plotf(a=10, b=1, main="metapopulation")
 plotf(a=10, b=1, main="metapopulation, Ne=1e7", Ne=1e7, tmax=1e4)
 
 legend("topright", lty=c(1,3,3), col=c('black', 'red', 'green'),
-        legend=c("total fitness", "mean", "variance"))
+        legend=c("total fitness", "shift of mean", "increase of variance"))
 
 dev.off()
 

### 2017-11-14 kalman_walk_paper.tex
changes to perturb Kalman

@@
 This is suggested by figure \ref{fig:conceptual_fig} and shown in Appendix \ref{apx:seg_cov}.
+\plr{connect to Barton etc polygenic adaptation lit}
 Concretely, we expect the population of $F_1$s to have variance $\sigma^2_S$ in the selected direction
@@
 
-\plr{note: should be a plot of the components of this against $T/N_e$ later}
-
 %%%%%%%%%%
@@
 Setting $a = (\sigma_N/\gamma)^2$ to be small,
-the fitness of $F_2$s is $\fit_2 \le 1/\sqrt{1 + 4 \omega a^2 T/N_e} \approx 1 - 2 \omega a T/N_e$.
-Since in small populations at mutation-drift equilibrium
-the amount of genetic variance ($\sigma_N$) is proportional to $N_e$,
-so this predicts the rate of decrease of hybrid fitness to be \emph{independent} of population size
+the fitness of $F_2$s is $\fit_2 \le 1/\sqrt{1 + 4 \omega a T/N_e} \approx 1 - 2 \omega a T/N_e$.
+Hybrid fitness seems to drop more slowly in this case in figure \ref{fig:speciation_rates},
+but since time is scaled by $N_e$, so speciation may occur \emph{faster} than in a large population.
+However, at least in some models \citep{bartonXXX}, in small populations at mutation-drift equilibrium
+the amount of genetic variance ($\sigma_N^2$) is proportional to $N_e$,
+which would compensate for this difference, 
+perhaps even predicting the rate of decrease of hybrid fitness to be \emph{independent} of population size
 for small populations.
-\plr{look up what $\sigma_N^2$ is at mut-drift equilib}
 
@@
 so that $\sigma_S \approx \gamma$ but $\sigma_N/\gamma$ is large.
-Then the fitness of $F_2$s is $\fit_2 \le 1/\sqrt{1 + 2 \omega a^2 T/N_e} \approx 1 - \omega a T/N_e$,
+Then the fitness of $F_2$s is $\fit_2 \le 1/\sqrt{1 + 2 \omega a T/N_e} \approx 1 - \omega a T/N_e$,
 and could be extremely rapid if $a$ is large.
 
-but not likely to be other combinations; make plot of \eqref{eqn:fit2} against $T/N_e$ (or even $T$?) in these three situations.
-
-\begin{figure}
+For instance, in a population of one million organisms that has 10 generations per year (a drosophilid species, perhaps)
+under the ``large population'' scenario of Figure \ref{fig:speciation_rates}A,
+system drift would lead to a substantial fitness drop of around 10\% in $F_2$ hybrids in only 10,000 years.
+This drop may be enough to induce evolutionary reinforcement of reproductive isolation.
+If one thousand of these organisms is isolated (perhaps on an island, as in Figure~\ref{fig:speciation_rates}B),
+then a similar drop could occur in around 120 years.
+On the other hand, if the population is one of several of similar size
+that have recently come into secondary contact after population re-expansion,
+the situation may be similar to that of Figure~\ref{fig:speciation_rates}C with $N_e = 10^6$,
+the same drop could occur after 1,100 years.
+However, hyperdiverse populations of this type may not be stable on these time scales.
+
+\begin{figure}[H]
+\label{fig:speciation_rates}
 \begin{center}
@@
 with $\omega=1$ and 
-\textbf{(A)} $\sigma_N = \sigma_S = \gamma$
-\textbf{(B)} $\sigma_N = \sigma_S = 0.1 \gamma$
-\textbf{(C)} $0.1 \sigma_N = \sigma_S = \gamma$.
+\textbf{(A)} $\sigma_N^2 = \sigma_S^2 = \gamma^2$
+\textbf{(B)} $\sigma_N^2 = \sigma_S^2 = 0.1 \gamma^2$
+\textbf{(C)} $0.1 \sigma_N^2 = \sigma_S^2 = \gamma^2$.
+The total fitness is from equation \eqref{eqn:fit2},
+and is the product of the ``shift of mean'' and ``increase of variance'' terms
+(the exponential and the square root, respectively).
+Note that, as we assume, the contribution of the shift in mean
+is small relative to that of increased variance.
 \label{fig:speciation_rates}}
@@
 
-\subsection*{Parameter estimates for real systems}
-%\js{I think we should move this section?}
-%To translate these results into real predictions, we need to know
-%the strength of stabilizing selection on the phenotype,
-%and the amount (and structure) of heritable variation in the genotype.
+\subsection*{Genetic variation in empirical regulatory systems}
+
+Above we show how system variation and drift can contribute to gene network incompatibility.
+What is known about the key quantity above, the amount of heritable variation in real regulatory networks?
 %These are known at best only roughly \citep{felsenstein1988phylogenies},
@@
 
-%% amount of heritable variation
-%We quantify (roughly) the amount of heritable variation
-%by $\sigma^2$, the genetic variance present in a population in a typical entry of $\Sys$.
-%The coefficient $A_{ij}$ measures how much the rate of net production of $i$ changes
-%per change in concentration of $j$.
-%It is generally thought that regulatory sequence change contributes much more to inter- and intraspecific variation
-%than does coding sequence change affecting molecular structure \citep{schmidt2010vertebrate}.
-%In the context of transcription factor networks this may be affected 
-%not only by the binding strength of molecule $j$ to the promoter region of gene $i$
-%but also the effects of other transcription factors (e.g., cooperativity)
-%and local chromatin accessibility \citep{stefflova2013cooperativity}.
-%For this reason, 
-%the mutational target size for variation in $A_{ij}$ may be much larger than the dozens of base pairs
-%typically implicated in the handful of binding sites for transcription factor $j$ of a typical promoter region,
-%and single variants may affect many entries of $\Sys$ simultaneously.
-%% (Recall that although these are best modeled through nonlinear terms,
-%% by linearizing we essentially consider first-order effects.)
-%On the other hand, a diverse set of buffering mechanisms are thought to contribute to phenotypic stability
-%in the presence of substantial molecular noise \citep{canalization,buffering},
-%suggesting that substantial variation in the micro-scale dynamics we consider here
-%may be necessary to produce relevant phenotypic effects downstream.
-%\plr{replace "micro-scale" with sthg else or discuss earlier}
+%% What is A and what determines its genetic variance.
+The coefficient $A_{ij}$ from the system \eqref{eqn:system} measures how much the rate of net production of $i$ changes
+per change in concentration of $j$.
+It is generally thought that regulatory sequence change contributes much more to inter- and intraspecific variation
+than does coding sequence change affecting molecular structure \citep{schmidt2010vertebrate}.
+In the context of transcription factor networks this may be affected 
+not only by the binding strength of molecule $j$ to the promoter region of gene $i$
+but also the effects of other transcription factors (e.g., cooperativity)
+and local chromatin accessibility \citep{stefflova2013cooperativity}.
+For this reason, 
+the mutational target size for variation in $A_{ij}$ may be much larger than the dozens of base pairs
+typically implicated in the handful of binding sites for transcription factor $j$ of a typical promoter region,
+and single variants may affect many entries of $\Sys$ simultaneously.
+% (Recall that although these are best modeled through nonlinear terms,
+% by linearizing we essentially consider first-order effects.)
+On the other hand, a diverse set of buffering mechanisms are thought to contribute to phenotypic stability
+in the presence of substantial molecular noise \citep{canalization,buffering},
+suggesting that substantial variation in the micro-scale dynamics we consider here
+may be necessary to produce relevant phenotypic effects downstream.
+\plr{replace "micro-scale" with sthg else or discuss earlier}
 
 %% empirical data on sigma
-%The amount and structure of this standing variation is established over long time scales
-%by many factors, including
-%mutation-selection balance, 
-%shifts in the phenotypic optimum,
-%and/or spatial variation in the optimum \citep{hansen1996translating}.
-%Quantitative genetics models of mutation-selection balance 
-%predict precise levels and structure of standing variation \citep{kimura_mutsel,lande_mutsel,lande1981models},
-%but it is unclear how well these predictions match reality \citep{johnson_barton}
-%and how much they are expected to change over time \citep{arnold_changing_G}.
-%However, empirical work allows us to estimate at least the rough magnitude of variation.
-%Differences in $A_{ij}$ due to a sequence change are hard to measure,
-%but variation in both transcription factor binding site occupancy
-%and expression levels (e.g., cis-eQTL) have been measured in various systems.
-%However, variation in binding site occupancy may overestimate variation in $\Sys$,
-%since it does not capture buffering effects (if for instance only one site of many needs be occupied for transcription to begin),
-%and variation in expression levels measures changes in steady-state concentration (our $\kappa_i$) rather than the \emph{rate} of change.
-%%Nonetheless,
-%%\citet{kasowski2010variation} found differential occupancy in 7.5\%
-%%of binding sites of a transcription factor (p65) between human individuals.
-%%\citet{verlaan2009targeted} showed that cis-regulatory variation
-%%accounts for around 2--6\% of expression variation in human blood-derived primary cells,
-%%while \citep{lappalainen2013transcriptome} found that human population variation 
-%%explained about 3\% of expression variation.
+The amount and structure of this standing variation is established over long time scales
+by many factors, including
+mutation-selection balance, 
+shifts in the phenotypic optimum,
+and/or spatial variation in the optimum \citep{hansen1996translating}.
+Quantitative genetics models of mutation-selection balance 
+predict precise levels and structure of standing variation \citep{kimura_mutsel,lande_mutsel,lande1981models},
+but it is unclear how well these predictions match reality \citep{johnson_barton}
+and how much they are expected to change over time \citep{arnold_changing_G}.
+However, empirical work allows us to estimate at least the rough magnitude of variation.
+
+Differences in $A_{ij}$ due to a sequence change are hard to measure,
+but variation in both transcription factor binding site occupancy
+and expression levels (e.g., cis-eQTL) have been measured in various systems.
+However, variation in binding site occupancy may overestimate variation in $\Sys$,
+since it does not capture buffering effects (if for instance only one site of many needs be occupied for transcription to begin),
+and variation in expression levels measures changes in steady-state concentration (our $\kappa_i$) rather than the \emph{rate} of change.
+Nonetheless,
+\citet{kasowski2010variation} found differential occupancy in 7.5\%
+of binding sites of a transcription factor (p65) between human individuals.
+\citet{verlaan2009targeted} showed that cis-regulatory variation
+accounts for around 2--6\% of expression variation in human blood-derived primary cells,
+while \citep{lappalainen2013transcriptome} found that human population variation 
+explained about 3\% of expression variation.
 %%\plr{Get some data from at least one other species in here!}
-%%Taken together, this suggests that variation in the entries of $\Sys$
-%%may be on the order of 1\% between individuals of a population --
-%%doubtless varying substantially between species and between genes.
-%%
-%Given the empirical data on transcription factor binding and gene expression
-%intrapopulation variation \citet{kasowski2010variation, verlaan2009targetedm lappalainen2013transcriptome}, we conservatively estimate that variation in the regulatory coefficients in $\Sys$ may be on the order of 1\% between individuals of a population --
-%doubtless varying substantially between species and between genes.
-%
-%%% argument for strength of selection
-%%It seems certain that selection in most species is not so strong that intra-population variation
-%%is strongly deleterious,
-%%so that if $\beta$ is the typical scale on which selection acts,
-%%then $\beta > \sigma$.  
-%%However,
-%%a range of studies \plr{find them} have found evidence for weak stabilizing selection
-%%on regulatory SNPs and cis-eQTL.
-%For instance, 
-%\citet{maria_and_sergey} \plr{(others?)} found evidence that large-effect regulatory mutations are weakly selected against in Drosophila.
-%This suggests that 
-%the strength of selection on phenotype is sufficient to weakly constrain regulatory variation,
-%so that perhaps $\sigma$ and $\beta$ are relatively close.
-%This is as would be expected if available variation is held in check by mutation--selection balance
-%rather than genetic drift.
-%A conservative estimate would be that $\beta = 5 \sigma$;
-%taking $\sigma=.01$ as above, this suggests that changes in phenotype of 5\% are sufficient
-%to effect a noticeable drop in fitness.
-%\plr{BUT $\sigma$ IS VARIATION IN $A$ NOT PHENOTYPE}
-%
-%We have guessed that within a population, 
-%the entries of $\Sys$ vary by around 1\%, 
-%at least for networks whose function is strongly constrained.
-
-%%%%%%%%%%%%
-\plri{insert commented-out discussion of what the actual genetic basis is}
+Taken together, this suggests that variation in the entries of $\Sys$
+may be on the order of 1\% between individuals of a population --
+doubtless varying substantially between species and between genes.
+
+What matters is $\sigma$ relative to $\gamma$ but this is harder to know.
 
@@
 
+Do we talk about the $G$ matrix and mutational correlation somewhere?
+
 %This theoretical framework can easily be applied to other interesting questions in evolutionary biology not tackled presently: such as the evolution of linkage, the necessity of network complexity (does evolution tend towards Rube Goldberg or parsimonious network organization?), evolvability, structure/function inference, and intra-population context dependency of mutational effects, as well as many others.
+
+Note that islands or bottlenecks leave large $\sigma_N$ but small $N_e$ so may go fast.
 
@@
 \jss{talk about speciation rate in small population results as compared to Katari and Goldstein (I think). they look at the influence of $N_e$ on speciation rate.}
+
+This has implications for genetic load also!!
+
+hominids; neanderthal load.
+
 \section*{Acknowledgements}

### 2017-11-14 sims/perturb_Kalman.r

@@ -2,7 +2,6 @@ library(MASS)
 library(Matrix)
 library(expm)
 
-sigma <- 1
 #####################################
 h <- function (t, M, BK, CK) { sapply(t, function (tt) CK %*% expm(tt*M) %*% BK) }
 optimal_h <- function (t) { sin(t)+cos(t) }
@@ -25,15 +24,17 @@ recombine<- function (X,Y)
   return(G)
 }
 
-make_P <- function (p) {
+make_P <- function (p, perturb=FALSE) {
   # Make the matrix P from a vector of entries
   stopifnot(length(p)==25)
-  P <- matrix(c(1,1-p[1],1-p[2],p[3],p[4],p[5],
-                0,p[1],p[2],-p[3],-p[4],-p[5],
-                0,0,p[7],p[8],p[9],p[10],
-                0,0,p[12],p[13],p[14],p[15],
-                0,0,p[17],p[18],p[19],p[20],
-                0,0,p[22],p[23],p[24],p[25]), nrow=6);
+  x <- if(perturb){0} else {1}
+  P <- matrix(c(x,x,x,0,0,0,
+                0,x+p[1],p[2],p[3],p[4],p[5],
+                0,0,x+p[7],p[8],p[9],p[10],
+                0,0,p[12],x+p[13],p[14],p[15],
+                0,0,p[17],p[18],x+p[19],p[20],
+                0,0,p[22],p[23],p[24],x+p[25]), nrow=6);
+  P[2:6,1] <- P[2:6,1] - P[2:6,2]
   return(P)
 }
 
@@ -90,11 +91,11 @@ perturb_oscillator <- function(AL, eps) {
     mAnrno <- matrix(c(rnorm(4, mean=0, sd=eps)),nrow=2,ncol=2)
     mp <- rnorm(25,mean=0,sd=eps)
     
-    new_A12 <- AL$A12 + AL$A12*mA12
-    new_Arno <- AL$Arno + AL$Arno*mArno
-    new_A13 <- AL$A13 + AL$A13*mA13
-    new_Anrno <- AL$Anrno + AL$Anrno*mAnrno
-    mP <- AL$P + make_P(mp)
+    new_A12 <- AL$A12 * (1 + mA12)
+    new_Arno <- AL$Arno * (1 + mArno)
+    new_A13 <- AL$A13 * (1 + mA13)
+    new_Anrno <- AL$Anrno * (1 + mAnrno)
[…]

### 2017-11-14 sims/test_perturb_Kalman.r

@@ -1,6 +1,7 @@
 source("perturb_Kalman.r")
 
 set.seed(23)
+num_errors <- 0
 
 check_system <- function (sys) {
     obs_D <- abs(D(sys$A, sys$B, sys$C))
@@ -11,14 +12,17 @@ check_system <- function (sys) {
     return(0)
 }
 
-# check that D(optimal) = 0
+cat("check that D(optimal) = 0\n")
 optimal_system <- list(
   A = matrix(c(0,-1,1,0), nrow=2),
   B = matrix(c(1,1), ncol=1),
   C = matrix(c(1,0), nrow=1) )
-check_system(optimal_system)
+num_errors <- num_errors + check_system(optimal_system)
 
-num_errors <- 0
+cat("check perturbation functions\n")
+stopifnot(all(make_P(rep(0,25), perturb=TRUE)==0))
+
+
 for (sigma in c(0.001, 0.1, 1.0)) {
     for (eps in c(0.001, 0.1, 1.0)) {
         cat(sprintf("Testing with sigma=%f and eps=%f\n", sigma, eps))

### 2017-11-14 kalman_walk_paper.tex
overleaf edits

@@
 
+%%% CITE SOME OF THESE
+%% empirical data on sigma
+% The amount and structure of this standing variation is established over long time scales
+% by many factors, including mutation-selection balance, 
+% shifts in the phenotypic optimum, and/or spatial variation in the optimum \citep{hansen1996translating}.
+% Quantitative genetics models of mutation-selection balance 
+% predict precise levels and structure of standing variation \citep{kimura_mutsel,lande_mutsel,lande1981models},
+% but it is unclear how well these predictions match reality \citep{johnson_barton}
+% and how much they are expected to change over time \citep{arnold_changing_G}.
+
 What are the fitness consequences?
@@
 
-Above we show how system variation and drift can contribute to gene network incompatibility.
 What is known about the key quantity above, the amount of heritable variation in real regulatory networks?
@@
 %so we aim for order-of-magnitude estimates.
-
-%% What is A and what determines its genetic variance.
 The coefficient $A_{ij}$ from the system \eqref{eqn:system} measures how much the rate of net production of $i$ changes
@@
 % by linearizing we essentially consider first-order effects.)
-On the other hand, a diverse set of buffering mechanisms are thought to contribute to phenotypic stability
-in the presence of substantial molecular noise \citep{canalization,buffering},
-suggesting that substantial variation in the micro-scale dynamics we consider here
-may be necessary to produce relevant phenotypic effects downstream.
-\plr{replace "micro-scale" with sthg else or discuss earlier}
-
-%% empirical data on sigma
-The amount and structure of this standing variation is established over long time scales
-by many factors, including
-mutation-selection balance, 
-shifts in the phenotypic optimum,
-and/or spatial variation in the optimum \citep{hansen1996translating}.
-Quantitative genetics models of mutation-selection balance 
-predict precise levels and structure of standing variation \citep{kimura_mutsel,lande_mutsel,lande1981models},
-but it is unclear how well these predictions match reality \citep{johnson_barton}
-and how much they are expected to change over time \citep{arnold_changing_G}.
-However, empirical work allows us to estimate at least the rough magnitude of variation.
-
-Differences in $A_{ij}$ due to a sequence change are hard to measure,
-but variation in both transcription factor binding site occupancy
-and expression levels (e.g., cis-eQTL) have been measured in various systems.
-However, variation in binding site occupancy may overestimate variation in $\Sys$,
+% On the other hand, a diverse set of buffering mechanisms are thought to contribute to phenotypic stability
+% in the presence of substantial molecular noise \citep{canalization,buffering},
+% suggesting that substantial variation in the micro-scale dynamics we consider here
+% may be necessary to produce relevant phenotypic effects downstream.
+
+Variation in binding site occupancy may overestimate variation in $A$,
 since it does not capture buffering effects (if for instance only one site of many needs be occupied for transcription to begin),
@@
 Taken together, this suggests that variation in the entries of $\Sys$
-may be on the order of 1\% between individuals of a population --
+may be on the order of at least a few percent between individuals of a population --
 doubtless varying substantially between species and between genes.
 
-What matters is $\sigma$ relative to $\gamma$ but this is harder to know.
+The impact of regulatory variation ($\sigma$) above
+depends only on its magnitude relative to selection ($\gamma$),
+but it seems reasonable that these two quantities are of the same magnitude.
 
@@
 Further, we see this as a necessary first step in the direction of more life-like nonlinear evolutionary systems theory.
-Depending on an actual biological system's particularities, its (potential) non-linearity, may buffer or exacerbate effects elucidated in this paper, such as the acquisition of Dobzansky-Muller incompatibilities.
+Depending on an actual biological system's particularities, its (potential) non-linearity, may buffer or exacerbate effects elucidated in this paper, such as the acquisition of Dobzhansky-Muller incompatibilities.
 

### 2017-11-14 kalman_walk_paper.tex
note on dimensions

@@
 \end{align}
+If each of the $q$ selected directions acts independently,
+the drop in fitness will be $\fit_2(T)^q$
+(the expression for the correlated cases is given in Appendix \ref{apx:gauss_load}).
 We discuss the implications of this expression in the next section.
@@
 $4 \omega (\sigma_N/\gamma)^2$.
-In particular, $\omega$, the rate that hybrid variance increases with separation between parentals,
-is proportional to $p-q$, the number of ways mutation can move the system without changing the phenotype,
-i.e., the dimensionality of the optimal set.
-This implies that less constrained systems should drift faster,
-and so lead to speciation more quickly, given the same strength of selection ($\gamma$).
-\plr{revisit}
 All else being equal, this process will lead to speciation more quickly in smaller populations
@@
 but since these details depend on the situation, we leave these separate.
+
+How does this prediction depend on the system size and constraint?
+If there are $p$ trait dimensions, constrained in $q$ dimensions,
+and if $\omega$ is proportional to $p-q$,
+% the rate that hybrid variance increases with separation between parentals, $\omega$,
+% is in some cases proportional to $p-q$, 
+then the rate that $F_2$ fitness drops is, roughly,
+$(1 + 4 (p-q) C T/n_e)^{-q/2} \propto q (p-q)$, where $C$ is a constant.
+This suggests that both degree of constraint and number of available neutral directions
+affect the speed of accumulation of incompatibilities.
+However, note that in real systems, it is likely that $\gamma$ also depends on $p$ and $q$.
+\plr{revisit with sims}
 
@@
 
-Suppose that a population has a Gaussian distribution in trait space
+Suppose that a population has a Gaussian distribution in $d$-dimensional trait space
 with mean $\mu$ and covariance matrix $\Sigma$,
@@
         \exp\left\{\mu^T \left( I - (I + L^T L \Sigma)^{-1} \right) \Sigma^{-1} \mu\right\} \\
+\end{align*}
+
+Now suppose that $\Sigma = \sigma^2 I$ and $L = I/\gamma$.
+Then,
+\begin{align*}
+    \sqrt{\frac{1}{\det(\Sigma)\det(\Sigma^{-1}+L^TL)}}
+    &=
+    \sqrt{\frac{1}{\sigma^{2d} (1/\sigma^2 + 1/\gamma^2)^d}} \\
+    &=
+    \frac{1}{(1+ \left(\sigma/\gamma\right)^2)^{d/2}} .
+\end{align*}
+Also, 
+\begin{align*}
+        \left( I - (I + L^T L \Sigma)^{-1} \right) \Sigma^{-1} 
+        &=
+        \frac{1}{\sigma^2}\left( 1 - (1 + (\sigma/\gamma)^2)^{-1} \right) I \\
+        &=
+        \frac{1}{\gamma^2}\frac{1}{(1 + (\sigma/\gamma)^2)} I \\
 \end{align*}

### 2017-11-16 kalman_walk_paper.tex
added Haldane's rule. edited the discussion.

@@
 \usepackage{multicol}
+\usepackage[bottom]{footmisc}
 
@@
 macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level 
-\citep{true2001developmental, tsong2006evolution, hare2008sepsid, vierstra2014mouse, stergachis2014conservation}.
-% are these good here? {taylor2016diverse, matsui2015regulatory}.
+\citep{true2001developmental, tanay2005conservation, tsong2006evolution, hare2008sepsid, vierstra2014mouse,  dalal2016transcriptional, dalal2017transcription}.
+% are these good here? {taylor2016diverse, matsui2015regulatory, stergachis2014conservation}.
 
@@
 pattern formation \plr{cite?}, 
-and metabolism \citep{lavoie2009rearrangements, martchenko2007transcriptional, dalal2016transcriptional, christensen2011unique, hartl2007induction, alam2013aspergillus}.
+and metabolism \citep{lavoie2009rearrangements, martchenko2007transcriptional, christensen2011unique, hartl2007induction, alam2013aspergillus}.
 
@@
 
+\paragraph{A new mechanism underlying Haldane's rule}
+The faster rate of fitness decay observed in $F_2$ relative to $F_1$ hybrids is consistent with Haldane's rule: the observation that if only one hybrid sex is sterile or inviable it is likely the heterogametic sex (e.g.
+the male XY) \citep{haldane1922sex}.
+If the regulatory coefficients for one system are shared between the sex and one or more autosomal chromosomes, and dosage compensation mechanisms are present (e.g.
+the homogametic sex silences or inactivates the expression of one of her two copies), $F_1$ males are effectively equivalent to purely autosomal-system $F_2$s.
+This is because incompatibilities between loci on the sex chromome with autosomal loci in $F_1$ males are less likely to be balanced.
+\footnote{For example, in a 2 gene network where the first gene resides on an autosome and the second gene on the X chromosome, the male with system $\left( \left[ \begin{smallmatrix} a_1 & a_2 \end{smallmatrix} \right], \left[ \begin{smallmatrix} a_1 & a_2 \\ X_1 & X_2 \end{smallmatrix} \right] \right)$ hybridizing with the homozygous female $\left( \left[\begin{smallmatrix} \bar a_1 & \bar a_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right] \right)$ will produce an $F_1$ male offspring with dynamics given by $ \left[\begin{smallmatrix} 0.5(a_1 + \bar a_1) & 0.5(a_2 + \bar a_2) \\ X_1 & X_2 \end{smallmatrix}\right]$.
+If both genes resided on the autosomes this system would only be possible in an $F_2$ cross.}
+
+As far as we know, this mechanism for Haldane's rule is novel and distinct from the currently favored ``dominance'', ``faster-X'', and ``faster-male'' theories \citep{orr1997haldane, coyne1998evolutionary}.
+In fact, this mechanism makes different predictions than the dominance theory as formulated by \citet{turelli1995dominance}.
+\citet{turelli1995dominance} employ a population genetic model which strictly requires alleles causing hybrid incompatibility on the X chromosome not to exhibit additive effects, that is the \emph{dominance coefficient} (typically $h$ in classic population genetics) must not be equal to $\frac{1}{2}$.
+In their model, if alleles are strictly additive, that is neither dominant nor recessive, then male and female $F_1$ hybrids will drop in fitness at the exact same rates, and therefore defy Haldane's rule.
+Furthermore, the dominance theory struggles with dosage compensation and X-inactivation \citep{watson2012haldane}; the present model, however, complies with Haldane's rule with X-inactivation explicitly included.
+
 %%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-% we do cool stuff
 The complexity of biological systems has limited our understanding of their function and evolution.
@@
 This methodology borrows successfully applied tools from engineering and aims to synthesize these with the concepts and tools of molecular and evolutionary biology.
-
-% no, really, it's cool
 Theoretical models in evolution and population genetics often lack the molecular details of physiology or of the genotype-phenotype map. 
@@
 Further, we provide, in clear mathematical language, an analytical description of phenomena hitherto discussed verbally and conceptually 
-(phenogenetic drift \citep{weiss2000phenogenetic}, developmental systems drift \citep{true2001developmental}, biological degeneracy \citep{edelman2001degeneracy}, \emph{etc.}). 
+(phenogenetic drift \citep{weiss2000phenogenetic}, developmental system drift \citep{true2001developmental}, biological degeneracy \citep{edelman2001degeneracy}). 
 The tractability and relative simplicity of this exposition enables the interested biologist to work out by hand, 
@@
 % it says something about speciation and DMIs
-We have suggested an interpretation of system identification: to see it as an evolutionarily neutral manifold, and not simply a computational nuisance. 
-We have demonstrated a method to analytically determine the set of all phenotypically invariant gene networks; 
-by a simple change of coordinates in the minimal configuration, or more generally by applying the Kalman decomposition in higher dimensions. 
+We have suggested an interpretation of system nonidentifiability: to see it as an evolutionarily neutral manifold, and not simply a computational nuisance. 
+We have demonstrated a method to analytically determine the set of all phenotypically equivalent gene networks; 
+by a simple change of coordinates in the minimal configuration, or more generally by applying the Kalman decomposition.
 %Further, we emphasize that evolution proceeds through this high dimensional space as stochastic coordinate transformation, constrained by sexual reproduction and selection. 
 This set is explored over evolutionary time when phenotype is conserved, and can lead to a diverse set of consequences, 
-including the accumulation of Dobzhansky-Muller incompatibilities. 
-\plr{check original Bateson/DM papers to see if this accords with those defns}\js{I read through Orr's review of those papers, which included excerpts.
-It seems like the DMI definition is very general and this accords.}
-We emphasize that these incompatibilities are a consequence of recombining different, yet functionally equivalent, mechanisms.
+including an increase in hybrid inviability.
+We emphasize that here, hybrid inviability is a consequence of recombining different, yet functionally equivalent, mechanisms.
 \plr{See refs in Barton 2010 paper.}
@@
 % summary and consequences of main result on speed of speciation
-Furthermore, using a quantitative genetic approach, we estimated that a genetically variable population will drift in neutral system space 
-at a rate determined by its intra-population variation and its effective population size. 
-Because mechanistically distinct yet phenotypically equivalent biological systems can fail to produce viable hybrids, we predict allopatric populations to accumulate genetic incompatibilities 
-at a rate on the order of $N_{e}$ if intrapopulation regulatory variation is constrained by selection.
-Additionally we see second-generation hybrid fitness plummet much faster than that of first-generation hybrids. 
-\plr{refer to Turelli here}
-%This is a consequence of combining our mechanistic model with a quantitative genetic one: 
-%we observe that $F_{1}$ phenotypes diverge quartically, and $F_{2}$ phenotypes quadratically, with evolutionary time. 
-%\plr{awkward phrasing}
-This result is also consistent with Haldane's rule; that if only one hybrid sex is inviable or sterile it is likely the heterogametic sex. 
-The consistency comes from gene networks localized to the sex chromosomes functioning as an $F_{2}$ hybrid cross within a diploid $F_{1}$ heterogamete as there is only one sex chromosome.
-%We also suggest that gene networks may not always use their components parsimoniously 
-%as network size tends to ratchet up in the absence of strong selection against extra parts. 
-%\plr{probably shouldn't claim this is "unexplored"} \js{I meant unexplored within this paper.Changed wording. Does that work?}
-%Although we leave this question unexplored, this phenomena may lead to insights on evolvability and developmental innovation. 
-Lastly, we show that hybrid gene networks, under neutral processes, break down as function of genetic distance, and may, in part, explain broad patterns of reproductive isolation among diverse phyla \citep{roux2016shedding}.
-
-Do we see evidence of this?  Well, maybe: misregulation in hybrids in Witkopp, Noor, Barbash, etc,
-even with expression levels in parents being the same.
+After demonstrating that it is, in principle, possible for phenotypically indistinguishable populations to speciate, we next probed whether system drift is possible under neutrality and whether the processs is fast enough to be a significant driver of speciation.
+This analysis was carried out by applying methods from quantitative genetics.
+In this framework a population is composed of many organisms, some with slightly differing gene regulatory network architectures.
+Some of this intrapopulation architectural variation is neutral and some is mildly deleterious and contributes to genetic load.
+Since there are always numerous neighboring system architecures with equivalent phenotypes and populations are not genetically homogenous, the typical system architecture within a population will drift.
+The speed at which a population will drift per generation, and thus the rate at which hybrid inviability appears, is a function of a population's size and of how much intrapopulation variation is present.
+Above, with the input of a few population parameters, we predict the rate at which isolated populations' hybrids become inviable.
+Our model suggests, that for a biologically reasonable range of parameter values, system drift can be a significant, if not rapid, driver of speciation, under neutrality.
+
+Additionally we see $F_2$ hybrid fitnesses plummet much faster than that of $F_1$ hybrids. 
+%\plr{refer to Turelli here}
+This is due to recombination shuffling system architectures leaving the hybrid system homozygous for one parent at one particular locus and homozygous for the other parent at another incompatible locus.
+If part of a gene regulatory network resides on the X chromosome, then $F_1$ males (XY) will effectivley look like $F_2$s at those locations.
+This observation is consistent with Haldane's rule; that if only one hybrid sex is inviable or sterile it is likely to be the heterogametic sex.
+
+While sumarizing speciation research \citet{orr2004speciation} remarked that ``speciation results from positive Darwinian selection within species'' thus maintaining one of the ``central tenets of the modern synthesis''.
+While it is known that divergent selection and genetic conflict, as well as other mechanisms, can precipitate speciation, we have shown here that, surprisingly, neutral drfit may also significantly drive species formation.
+%, although it is at present still unclear the relative contribution of each process. 
+System drift induced hybrid breakdown occurs rapidly even in the absence of postive Darwinian selection, suggesting that species originate not necessarily by means of natural selection \citep{darwin1859origin}.
+While the incompatibility of any one given network within an organism may be small, we note that organsisms are made up of many different networks, and the cumulative impact of system drift across all of these may be substantial.
+Further, even in populations experiencing directional selection, networks underlying conserved parts of the phenotype can still experience neutral drift and harbor incompatibilities.
+
+Despite the results of this inquiry, in practice it remains unclear whether system drift or other speciation mechanisms predominate, as hybrid incompatibilities due to both selection and conflict have been uncovered and hint at general processes \citep{orr2004speciation}.
+It also seems reasonable that systems drift could be common as there is evidence (1) that the set of molecular mechanisms underlying a trait is degenerate (illustrated in this paper), (2) that there is significant regulatory variation segregating within populations, (3) that functionally equivalent systems, even in closesly related species, can be mechanistically divergent \citep{dalal2017transcription}, (4) transcription in hybrids between closely related species with conserved transcriptional patterns can be divergent \citep{haerty2006gene, maheshwari2012cis, coolon2014tempo, michalak2004association}, and that (5) gene misregulation can cause dysfunction (by definition).
+
+Lastly, we show that hybrid gene networks, under neutral processes, break down as a function of genetic distance, and may, in part, explain broad patterns of reproductive isolation among diverse phyla \citep{roux2016shedding, hedges2015tree}.
+
+Additionally, if we ignore the fitness function used in this paper, our model still predicts that $F_1$ and $F_2$ phenotypes diverge linearly and at the square root of time with parentals, respectivey.
+This has implications for experimentation, as one could plot the change in phenotype (or gene expression) as a function of divergence, to test the predictions of this model.
+
+%Do we see evidence of this?  Well, maybe: misregulation in hybrids in Witkopp, Noor, Barbash, etc,
+%even with expression levels in parents being the same.
 
 % consequence: prediction for phenotype (not just fitness)
-Note that this follows directly from the fact that $F_1$ and $F_2$ hybrid phenotypes diverge from $F_0$ phenotypes, linearly and inverse-quadratically, with respect to time (more precisely $T/N_e$).
+%Note that this follows directly from the fact that $F_1$ and $F_2$ hybrid phenotypes diverge from $F_0$ phenotypes, linearly and inverse-quadratically, with respect to time (more precisely $T/N_e$).
 As such, removing assumptions about the form of the fitness function, this model still predicts hybrid $F_2$ phenotypes to diverge much faster than $F_1$s, initially.
-So even if fitness isn't as above maybe you can see this in phenotypes.
+%So even if fitness isn't as above maybe you can see this in phenotypes.
 
@@
 It is often stated that life is not linear.
-This is often true, however, many of the ideas developed here should be generalizable to nonlinear cases (multi-linear systems, say).
-Further, we see this as a necessary first step in the direction of more life-like nonlinear evolutionary systems theory.
-Depending on an actual biological system's particularities, its (potential) non-linearity, may buffer or exacerbate effects elucidated in this paper, such as the acquisition of Dobzhansky-Muller incompatibilities.
+This is often true, 
+%however, many of the ideas developed here should be generalizable to nonlinear cases (multi-linear systems, say). 
+however, we see this as a necessary first step in the direction of a more life-like (nonlinear, perhaps) evolutionary systems theory.
+Depending on an actual biological system's particularities, its non-linearity, may buffer or exacerbate effects elucidated in this paper.
 
 % selection moving around optimal set
-Note that the drift above could actually be caused by the optimal phenotype shifting.
+%Note that the drift above could actually be caused by the optimal phenotype shifting.
+%\js{I don't think we need to address optimal phenotype shifting}
 

### 2017-11-16 kalman_walk_paper.tex
added Haldane's rule. edited the discussion.

@@
 This is because incompatibilities between loci on the sex chromome with autosomal loci in $F_1$ males are less likely to be balanced.
-\footnote{For example, in a 2 gene network where the first gene resides on an autosome and the second gene on the X chromosome, the male with system $\left( \left[ \begin{smallmatrix} a_1 & a_2 \end{smallmatrix} \right], \left[ \begin{smallmatrix} a_1 & a_2 \\ X_1 & X_2 \end{smallmatrix} \right] \right)$ hybridizing with the homozygous female $\left( \left[\begin{smallmatrix} \bar a_1 & \bar a_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right] \right)$ will produce an $F_1$ male offspring with dynamics given by $ \left[\begin{smallmatrix} 0.5(a_1 + \bar a_1) & 0.5(a_2 + \bar a_2) \\ X_1 & X_2 \end{smallmatrix}\right]$.
+\footnote{For example, in a 2 gene network where the first gene resides on an autosome and the second gene on the X chromosome, the male with system $\left( \left[ \begin{smallmatrix} a_1 & a_2 \end{smallmatrix} \right], \left[ \begin{smallmatrix} a_1 & a_2 \\ X_1 & X_2 \end{smallmatrix} \right] \right)$ hybridizing with the homozygous female $\left( \left[\begin{smallmatrix} \bar a_1 & \bar a_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right] \right)$ will produce an $F_1$ male offspring with dynamics given by $ \left[\begin{smallmatrix} 0.5(a_1 + \bar a_1) & 0.5(a_2 + \bar a_2) \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$.
 If both genes resided on the autosomes this system would only be possible in an $F_2$ cross.}

### 2017-11-16 kalman_walk_paper.tex
added Haldane's rule. edited the discussion.

@@
 %
-% Systems drift and speciation: more than one way to grow a cat
+% System drift and speciation: more than one way to grow a cat
 %
@@
 %
-% Hybrid incompatibilities can evolve rapidly due to developmental systems drift
+% Hybrid incompatibilities can evolve rapidly due to developmental system drift
 %
@@
 These functional symmetries manifest in convergent and parallel evolution, 
-as well as \emph{developmental systems drift}: the observation that
+as well as \emph{developmental system drift}: the observation that
 macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level 
@@
 The literature is filled with detailed observations of molecular systems and their diversity. 
-\plr{Diversity doesn't imply systems drift -- only if the diverse systems are homologous.}
+\plr{Diversity doesn't imply system drift -- only if the diverse systems are homologous.}
 There are examples of significant diversity in the networks underlying processes such as 
@@
 which gives concrete expectations for how any such
-system can, in principal, undergo systems drift and rewire. 
+system can, in principal, undergo system drift and rewire. 
 Even with stabilization selection, a population will explore the set of all possible phenotypically equivalent gene networks.
@@
 
+% nonlinearity
+%What about nonlinearity?
+As Richard Levins opined, models in population biology face a trade-off among precision, realism, and generality \citep{levins1966strategy}.
+As Levins expects, any tractable and general model, such as the present one under discussion, will have limitations.
+Most notable is linearity.
+It is often stated that life is not linear.
+This is often true, 
+%however, many of the ideas developed here should be generalizable to nonlinear cases (multi-linear systems, say). 
+however, we see this as a necessary first step in the direction of a more life-like (nonlinear, perhaps) evolutionary systems theory.
+Depending on an actual biological system's particularities, its non-linearity, may buffer or exacerbate effects elucidated in this paper.
+
 % summary and consequences of main result on speed of speciation
@@
 
+Lastly, we show that hybrid gene networks, under neutral processes, break down as a function of genetic distance, and may, in part, explain broad patterns of reproductive isolation among diverse phyla \citep{roux2016shedding, hedges2015tree}.
+
 While sumarizing speciation research \citet{orr2004speciation} remarked that ``speciation results from positive Darwinian selection within species'' thus maintaining one of the ``central tenets of the modern synthesis''.
@@
 System drift induced hybrid breakdown occurs rapidly even in the absence of postive Darwinian selection, suggesting that species originate not necessarily by means of natural selection \citep{darwin1859origin}.
-While the incompatibility of any one given network within an organism may be small, we note that organsisms are made up of many different networks, and the cumulative impact of system drift across all of these may be substantial.
+While the incompatibility of any one given network within an organism may be small, we note that organsisms are made up of many different networks, and that the cumulative impact of system drift across all of these may be substantial.
 Further, even in populations experiencing directional selection, networks underlying conserved parts of the phenotype can still experience neutral drift and harbor incompatibilities.
@@
 Despite the results of this inquiry, in practice it remains unclear whether system drift or other speciation mechanisms predominate, as hybrid incompatibilities due to both selection and conflict have been uncovered and hint at general processes \citep{orr2004speciation}.
-It also seems reasonable that systems drift could be common as there is evidence (1) that the set of molecular mechanisms underlying a trait is degenerate (illustrated in this paper), (2) that there is significant regulatory variation segregating within populations, (3) that functionally equivalent systems, even in closesly related species, can be mechanistically divergent \citep{dalal2017transcription}, (4) transcription in hybrids between closely related species with conserved transcriptional patterns can be divergent \citep{haerty2006gene, maheshwari2012cis, coolon2014tempo, michalak2004association}, and that (5) gene misregulation can cause dysfunction (by definition).
-
-Lastly, we show that hybrid gene networks, under neutral processes, break down as a function of genetic distance, and may, in part, explain broad patterns of reproductive isolation among diverse phyla \citep{roux2016shedding, hedges2015tree}.
+It also seems reasonable that system drift could be common as there is evidence (1) that the set of molecular mechanisms underlying a trait is degenerate (illustrated in this paper), (2) that there is significant regulatory variation segregating within populations, (3) that functionally equivalent systems, even in closesly related species, can be mechanistically divergent \citep{dalal2017transcription}, (4) transcription in hybrids between closely related species with conserved transcriptional patterns can be divergent \citep{haerty2006gene, maheshwari2012cis, coolon2014tempo, michalak2004association}, that (5) gene misregulation can cause dysfunction (by definition), and (6) that numerous cryptic species complexes exist (e.g.
+sun skinks \citep{barley2013challenge}).
 
@@
 %So even if fitness isn't as above maybe you can see this in phenotypes.
-
-% nonlinearity
-What about nonlinearity?
-As Richard Levins opined, models in population biology face a trade-off among precision, realism, and generality \citep{levins1966strategy}.
-As Levins expects, any tractable and general model, such as the present one under discussion, will have limitations.
-Most notable is linearity.
-It is often stated that life is not linear.
-This is often true, 
-%however, many of the ideas developed here should be generalizable to nonlinear cases (multi-linear systems, say). 
-however, we see this as a necessary first step in the direction of a more life-like (nonlinear, perhaps) evolutionary systems theory.
-Depending on an actual biological system's particularities, its non-linearity, may buffer or exacerbate effects elucidated in this paper.
 

### 2017-11-17 kalman_walk_paper.tex
overleaf edits

@@
 over a number of generations proportional to the effective population size.
+This model also provides a natural explanation for Haldane's rule
+that appears mechanistically distinct from other proposed mechanisms.
 \end{abstract}
@@
 
-\paragraph{A new mechanism underlying Haldane's rule}
-The faster rate of fitness decay observed in $F_2$ relative to $F_1$ hybrids is consistent with Haldane's rule: the observation that if only one hybrid sex is sterile or inviable it is likely the heterogametic sex (e.g.
-the male XY) \citep{haldane1922sex}.
-If the regulatory coefficients for one system are shared between the sex and one or more autosomal chromosomes, and dosage compensation mechanisms are present (e.g.
-the homogametic sex silences or inactivates the expression of one of her two copies), $F_1$ males are effectively equivalent to purely autosomal-system $F_2$s.
-This is because incompatibilities between loci on the sex chromome with autosomal loci in $F_1$ males are less likely to be balanced.
-\footnote{For example, in a 2 gene network where the first gene resides on an autosome and the second gene on the X chromosome, the male with system $\left( \left[ \begin{smallmatrix} a_1 & a_2 \end{smallmatrix} \right], \left[ \begin{smallmatrix} a_1 & a_2 \\ X_1 & X_2 \end{smallmatrix} \right] \right)$ hybridizing with the homozygous female $\left( \left[\begin{smallmatrix} \bar a_1 & \bar a_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right] \right)$ will produce an $F_1$ male offspring with dynamics given by $ \left[\begin{smallmatrix} 0.5(a_1 + \bar a_1) & 0.5(a_2 + \bar a_2) \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$.
-If both genes resided on the autosomes this system would only be possible in an $F_2$ cross.}
-
-As far as we know, this mechanism for Haldane's rule is novel and distinct from the currently favored ``dominance'', ``faster-X'', and ``faster-male'' theories \citep{orr1997haldane, coyne1998evolutionary}.
-In fact, this mechanism makes different predictions than the dominance theory as formulated by \citet{turelli1995dominance}.
+\paragraph{Haldane's rule}
+This model naturally predicts
+Haldane's rule, the observation that if only one hybrid sex is sterile or inviable 
+it is likely the heterogametic sex (e.g., the male in XY sex determination systems) \citep{haldane1922sex}.
+For example, consider an XY species with a two-gene network where the first gene resides on an autosome and the second gene on the X chromosome.
+A male whose pair of haplotypes are 
+$\left( \left[ \begin{smallmatrix} A_1 & A_2 \\ \cdot & \cdot \end{smallmatrix} \right], \left[ \begin{smallmatrix} A_1 & A_2 \\ X_1 & X_2 \end{smallmatrix} \right] \right)$
+has phenotype determined by $A = \left[ \begin{smallmatrix} A_1 & A_2 \\ X_1/2 & X_2/2 \end{smallmatrix} \right]$,
+while a female homozygous for the haplotype
+$\left[\begin{smallmatrix} \bar A_1 & \bar A_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$,
+thanks to X-inactivation, has phenotype determined by
+$A = \left[\begin{smallmatrix} \bar A_1 & \bar A_2 \\ \bar X_1/2 & \bar X_2/2 \end{smallmatrix}\right]$.
+An $F_1$ male offspring of these two will have phenotype determined by
+$\frac{1}{2} \left[\begin{smallmatrix} A_1 + \bar A_1 & A_2 + \bar A_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$.
+If both genes resided on the autosomes this system would only be possible in an $F_2$ cross.
+More generally, 
+if the regulatory coefficients for a system are shared between the sex and one or more autosomal chromosomes, 
+$F_1$ males are effectively equivalent to purely autosomal-system $F_2$s.
+
+This mechanism is apparently distinct from the currently favored ``dominance'', ``faster-X'', and ``faster-male'' theories \citep{orr1997haldane, coyne1998evolutionary} to explain Haldane's rule,
+and makes several different predictions than the dominance theory as formulated by \citet{turelli1995dominance}.
+\plr{what are they?}
 \citet{turelli1995dominance} employ a population genetic model which strictly requires alleles causing hybrid incompatibility on the X chromosome not to exhibit additive effects, that is the \emph{dominance coefficient} (typically $h$ in classic population genetics) must not be equal to $\frac{1}{2}$.

### 2017-11-20 kalman_walk_paper.tex
discussion edits

@@
 %Further, we emphasize that evolution proceeds through this high dimensional space as stochastic coordinate transformation, constrained by sexual reproduction and selection. 
-This set is explored over evolutionary time when phenotype is conserved, and can lead to a diverse set of consequences, 
-including an increase in hybrid inviability.
+%% summary and consequences of main result on speed of speciation
+%After demonstrating that it is, in principle, possible for phenotypically indistinguishable populations to speciate, we next probed whether system drift is possible under neutrality and whether the processs is fast enough to be a significant driver of speciation.
+This analysis was carried out by applying methods from quantitative genetics.
+In this framework a population is composed of many organisms, some with slightly differing gene regulatory network architectures.
+Some of this intrapopulation architectural variation is neutral and some is mildly deleterious and contributes to genetic load.
+Since there are always numerous neighboring system architecures with equivalent phenotypes and populations are not genetically homogenous, the typical system architecture within a population will drift.
+This set is explored by a population over evolutionary time while phenotype is conserved, and can lead to speciation via an increase in hybrid inviability.
+The speed at which a population will drift through this set, and thus accrue hybrid incompatibilities, is a function of effective population size and genetic variation. 
+With the input of a range of biologically reasonable parameter values, our model predicts that system drift will be a significant -- and likely a rapid -- driver of speciation.
 We emphasize that here, hybrid inviability is a consequence of recombining different, yet functionally equivalent, mechanisms. 
 \plr{See refs in Barton 2010 paper.}
+
+Consistent with empirical observation, we see that the fitnesses of $F_2$ hybrids drop at a much faster rate than those of $F_1$s (as $F_1$ and $F_2$ phenotypes diverge linearly and at the square root of time relative to parentals, respectivey).
+%\plr{refer to Turelli here}
+In this model, this is due to recombination shuffling system architectures, leaving the $F_2$ system homozygous for one parent at one particular locus and homozygous for the other parent at another possibly incompatible locus.
+Interestingly, if part of a gene regulatory network resides on the sex chromosomes (X), then heterogametic $F_1$s (typically males) will effectivley look like $F_2$s.
+This observation is consistent with, and may in fact explain, Haldane's rule: that if only one hybrid sex is inviable or sterile it is likely to be the heterogametic sex.
 
@@
 
-% summary and consequences of main result on speed of speciation
-After demonstrating that it is, in principle, possible for phenotypically indistinguishable populations to speciate, we next probed whether system drift is possible under neutrality and whether the processs is fast enough to be a significant driver of speciation.
-This analysis was carried out by applying methods from quantitative genetics.
-In this framework a population is composed of many organisms, some with slightly differing gene regulatory network architectures.
-Some of this intrapopulation architectural variation is neutral and some is mildly deleterious and contributes to genetic load.
-Since there are always numerous neighboring system architecures with equivalent phenotypes and populations are not genetically homogenous, the typical system architecture within a population will drift.
-The speed at which a population will drift per generation, and thus the rate at which hybrid inviability appears, is a function of a population's size and of how much intrapopulation variation is present.
-Above, with the input of a few population parameters, we predict the rate at which isolated populations' hybrids become inviable.
-Our model suggests, that for a biologically reasonable range of parameter values, system drift can be a significant, if not rapid, driver of speciation, under neutrality.
-
-Additionally we see $F_2$ hybrid fitnesses plummet much faster than that of $F_1$ hybrids. 
-%\plr{refer to Turelli here}
-This is due to recombination shuffling system architectures leaving the hybrid system homozygous for one parent at one particular locus and homozygous for the other parent at another incompatible locus.
-If part of a gene regulatory network resides on the X chromosome, then $F_1$ males (XY) will effectivley look like $F_2$s at those locations.
-This observation is consistent with Haldane's rule; that if only one hybrid sex is inviable or sterile it is likely to be the heterogametic sex.
-
-Lastly, we show that hybrid gene networks, under neutral processes, break down as a function of genetic distance, and may, in part, explain broad patterns of reproductive isolation among diverse phyla \citep{roux2016shedding, hedges2015tree}.
-
-While sumarizing speciation research \citet{orr2004speciation} remarked that ``speciation results from positive Darwinian selection within species'' thus maintaining one of the ``central tenets of the modern synthesis''.
+Our results show that hybrids, under neutral processes, break down as a function of genetic distance, and despite the present model's myriad simplifications and assumptions, 
+it may, in part, explain the observed broad patterns of reproductive isolation among diverse phyla \citep{roux2016shedding, hedges2015tree}.
+\js{maybe also cite the phylogentics lit here?
+Like the speciation clock stuff.}
+
+Although, as classically formulated, the Dobzhansky-Muller model of hybrid incompatibility is agnostic to the relative importance of neutral versus selective genetic substitutions \citep{coyne1998evolutionary}, \citet{orr2004speciation} remarks that hybrid incompatibility data suggests that ``speciation results from positive Darwinian selection within species'' thus maintaining one of the ``central tenets of the modern synthesis''.
+%In practice it remains unclear whether system drift or other speciation mechanisms predominate, 
+%as hybrid incompatibilities due to both selection and conflict have been uncovered and hint at general processes \citep{orr2004speciation}. 
 While it is known that divergent selection and genetic conflict, as well as other mechanisms, can precipitate speciation, we have shown here that, surprisingly, neutral drfit may also significantly drive species formation.
-%, although it is at present still unclear the relative contribution of each process. 
-System drift induced hybrid breakdown occurs rapidly even in the absence of postive Darwinian selection, suggesting that species originate not necessarily by means of natural selection \citep{darwin1859origin}.
+This seems reasonable as there is evidence that the set of molecular mechanisms underlying a trait is degenerate, that there is significant regulatory variation segregating within populations, that functionally equivalent systems, even in closesly related species, can be mechanistically divergent \citep{dalal2017transcription}, transcription in hybrids between closely related species with conserved transcriptional patterns can be divergent \citep{haerty2006gene, maheshwari2012cis, coolon2014tempo, michalak2004association}, that gene misregulation can cause dysfunction (by definition), and that numerous cryptic species complexes exist (e.g.
+sun skinks \citep{barley2013challenge}).
+
+If system drift induced hybrid breakdown occurs as rapidly as our model sugests, even in the absence of postive Darwinian selection, this would imply that species originate not necessarily by means of natural selection \citep{darwin1859origin}.
 While the incompatibility of any one given network within an organism may be small, we note that organsisms are made up of many different networks, and that the cumulative impact of system drift across all of these may be substantial.
 Further, even in populations experiencing directional selection, networks underlying conserved parts of the phenotype can still experience neutral drift and harbor incompatibilities.
-
-Despite the results of this inquiry, in practice it remains unclear whether system drift or other speciation mechanisms predominate, as hybrid incompatibilities due to both selection and conflict have been uncovered and hint at general processes \citep{orr2004speciation}.
-It also seems reasonable that system drift could be common as there is evidence (1) that the set of molecular mechanisms underlying a trait is degenerate (illustrated in this paper), (2) that there is significant regulatory variation segregating within populations, (3) that functionally equivalent systems, even in closesly related species, can be mechanistically divergent \citep{dalal2017transcription}, (4) transcription in hybrids between closely related species with conserved transcriptional patterns can be divergent \citep{haerty2006gene, maheshwari2012cis, coolon2014tempo, michalak2004association}, that (5) gene misregulation can cause dysfunction (by definition), and (6) that numerous cryptic species complexes exist (e.g.
-sun skinks \citep{barley2013challenge}).
-
-Additionally, if we ignore the fitness function used in this paper, our model still predicts that $F_1$ and $F_2$ phenotypes diverge linearly and at the square root of time with parentals, respectivey.
-This has implications for experimentation, as one could plot the change in phenotype (or gene expression) as a function of divergence, to test the predictions of this model.
 

### 2017-11-20 kalman_walk_paper.tex
discussion edits

@@
 Here, we offer a tractable and simple model which includes these missing features.
+
 Further, we provide, in clear mathematical language, an analytical description of phenomena hitherto discussed verbally and conceptually 
@@
 
+System drift may be a significant driver of speciation as there is evidence that the set of molecular mechanisms underlying a trait is degenerate, that there is significant regulatory variation segregating within populations, that functionally equivalent systems, even in closesly related species, can be mechanistically divergent \citep{dalal2017transcription}, transcription in hybrids between closely related species with conserved transcriptional patterns can be divergent \citep{haerty2006gene, maheshwari2012cis, coolon2014tempo, michalak2004association}, that gene misregulation can cause dysfunction (by definition), and that numerous cryptic species complexes exist (e.g.
+sun skinks \citep{barley2013challenge}).
+
 % nonlinearity
@@
 
+Although, as classically formulated, the Dobzhansky-Muller model of hybrid incompatibility is agnostic to the relative importance of neutral versus selective genetic substitutions \citep{coyne1998evolutionary}, \citet{orr2004speciation} remarks that hybrid incompatibility data suggests that ``speciation results from positive Darwinian selection within species'' thus maintaining one of the ``central tenets of the modern synthesis''.
+%In practice it remains unclear whether system drift or other speciation mechanisms predominate, 
+%as hybrid incompatibilities due to both selection and conflict have been uncovered and hint at general processes \citep{orr2004speciation}. 
+While it is known that divergent selection and genetic conflict, as well as other mechanisms, can precipitate speciation, we have shown here that, surprisingly, neutral drfit may also significantly drive species formation.
+
+If system drift induced hybrid breakdown occurs as rapidly as our model sugests, even in the absence of postive Darwinian selection, this would imply that species originate not necessarily by means of natural selection \citep{darwin1859origin}.
+While the incompatibility of any one given network within an organism may be small, we note that organsisms are made up of many different networks, and that the cumulative impact of system drift across all of these may be substantial.
+Further, even in populations experiencing directional selection, networks underlying conserved parts of the phenotype can still experience neutral drift and harbor incompatibilities.
 Our results show that hybrids, under neutral processes, break down as a function of genetic distance, and despite the present model's myriad simplifications and assumptions, 
@@
 Like the speciation clock stuff.}
-
-Although, as classically formulated, the Dobzhansky-Muller model of hybrid incompatibility is agnostic to the relative importance of neutral versus selective genetic substitutions \citep{coyne1998evolutionary}, \citet{orr2004speciation} remarks that hybrid incompatibility data suggests that ``speciation results from positive Darwinian selection within species'' thus maintaining one of the ``central tenets of the modern synthesis''.
-%In practice it remains unclear whether system drift or other speciation mechanisms predominate, 
-%as hybrid incompatibilities due to both selection and conflict have been uncovered and hint at general processes \citep{orr2004speciation}. 
-While it is known that divergent selection and genetic conflict, as well as other mechanisms, can precipitate speciation, we have shown here that, surprisingly, neutral drfit may also significantly drive species formation.
-This seems reasonable as there is evidence that the set of molecular mechanisms underlying a trait is degenerate, that there is significant regulatory variation segregating within populations, that functionally equivalent systems, even in closesly related species, can be mechanistically divergent \citep{dalal2017transcription}, transcription in hybrids between closely related species with conserved transcriptional patterns can be divergent \citep{haerty2006gene, maheshwari2012cis, coolon2014tempo, michalak2004association}, that gene misregulation can cause dysfunction (by definition), and that numerous cryptic species complexes exist (e.g.
-sun skinks \citep{barley2013challenge}).
-
-If system drift induced hybrid breakdown occurs as rapidly as our model sugests, even in the absence of postive Darwinian selection, this would imply that species originate not necessarily by means of natural selection \citep{darwin1859origin}.
-While the incompatibility of any one given network within an organism may be small, we note that organsisms are made up of many different networks, and that the cumulative impact of system drift across all of these may be substantial.
-Further, even in populations experiencing directional selection, networks underlying conserved parts of the phenotype can still experience neutral drift and harbor incompatibilities.
 

### 2017-11-20 kalman_walk_paper.tex
discussion edits

@@
 
-If system drift induced hybrid breakdown occurs as rapidly as our model sugests, even in the absence of postive Darwinian selection, this would imply that species originate not necessarily by means of natural selection \citep{darwin1859origin}.
+If system drift induced hybrid incompatibility occurs as rapidly as our model sugests,
+%even in the absence of postive Darwinian selection, 
+this would imply that species originate not necessarily by means of natural selection \citep{darwin1859origin}.
 While the incompatibility of any one given network within an organism may be small, we note that organsisms are made up of many different networks, and that the cumulative impact of system drift across all of these may be substantial.

### 2017-11-21 kalman_walk_paper.tex
discussion edits from overleaf

@@
 % - Summarize methods and main result on rate of speciation.
+% - Consistent with F1/F2 divergence and Haldane's rule
 % - What empirical evidence is there that this is actually happening? 
@@
 
-The complexity of biological systems has limited our understanding of their function and evolution.
-Above we outline an approach, a first step, towards untangling this complexity in reference to function and evolution.
-This methodology borrows successfully applied tools from engineering and aims to synthesize these with the concepts and tools of molecular and evolutionary biology.
-Theoretical models in evolution and population genetics often lack the molecular details of physiology or of the genotype-phenotype map. 
-Here, we offer a tractable and simple model which includes these missing features.
-
-Further, we provide, in clear mathematical language, an analytical description of phenomena hitherto discussed verbally and conceptually 
-(phenogenetic drift \citep{weiss2000phenogenetic}, developmental system drift \citep{true2001developmental}, biological degeneracy \citep{edelman2001degeneracy}). 
-The tractability and relative simplicity of this exposition enables the interested biologist to work out by hand, 
-if desired, 
-the dynamics of a genetic system, as well as perturbations to the system -- an attribute not likely to be found in less tractable models and simulations.
-
-% it says something about speciation and DMIs
-We have suggested an interpretation of system nonidentifiability: to see it as an evolutionarily neutral manifold, and not simply a computational nuisance. 
-We have demonstrated a method to analytically determine the set of all phenotypically equivalent gene networks; 
-by a simple change of coordinates in the minimal configuration, or more generally by applying the Kalman decomposition.
-%Further, we emphasize that evolution proceeds through this high dimensional space as stochastic coordinate transformation, constrained by sexual reproduction and selection. 
-%% summary and consequences of main result on speed of speciation
-%After demonstrating that it is, in principle, possible for phenotypically indistinguishable populations to speciate, we next probed whether system drift is possible under neutrality and whether the processs is fast enough to be a significant driver of speciation.
-This analysis was carried out by applying methods from quantitative genetics.
-In this framework a population is composed of many organisms, some with slightly differing gene regulatory network architectures.
-Some of this intrapopulation architectural variation is neutral and some is mildly deleterious and contributes to genetic load.
-Since there are always numerous neighboring system architecures with equivalent phenotypes and populations are not genetically homogenous, the typical system architecture within a population will drift.
-This set is explored by a population over evolutionary time while phenotype is conserved, and can lead to speciation via an increase in hybrid inviability.
-The speed at which a population will drift through this set, and thus accrue hybrid incompatibilities, is a function of effective population size and genetic variation. 
-With the input of a range of biologically reasonable parameter values, our model predicts that system drift will be a significant -- and likely a rapid -- driver of speciation.
-We emphasize that here, hybrid inviability is a consequence of recombining different, yet functionally equivalent, mechanisms. 
-\plr{See refs in Barton 2010 paper.}
-
-Consistent with empirical observation, we see that the fitnesses of $F_2$ hybrids drop at a much faster rate than those of $F_1$s (as $F_1$ and $F_2$ phenotypes diverge linearly and at the square root of time relative to parentals, respectivey).
-%\plr{refer to Turelli here}
-In this model, this is due to recombination shuffling system architectures, leaving the $F_2$ system homozygous for one parent at one particular locus and homozygous for the other parent at another possibly incompatible locus.
-Interestingly, if part of a gene regulatory network resides on the sex chromosomes (X), then heterogametic $F_1$s (typically males) will effectivley look like $F_2$s.
-This observation is consistent with, and may in fact explain, Haldane's rule: that if only one hybrid sex is inviable or sterile it is likely to be the heterogametic sex.
-
-System drift may be a significant driver of speciation as there is evidence that the set of molecular mechanisms underlying a trait is degenerate, that there is significant regulatory variation segregating within populations, that functionally equivalent systems, even in closesly related species, can be mechanistically divergent \citep{dalal2017transcription}, transcription in hybrids between closely related species with conserved transcriptional patterns can be divergent \citep{haerty2006gene, maheshwari2012cis, coolon2014tempo, michalak2004association}, that gene misregulation can cause dysfunction (by definition), and that numerous cryptic species complexes exist (e.g.
-sun skinks \citep{barley2013challenge}).
+Above we synthesize models from molecular and evolutionary biology with tools from control theory
+to study the evolution of a molecularly mechanistic model of the genotype-phenotype map under stabilizing selection.
+This model allows us to describe analytically the processes described verbally as
+phenogenetic drift \citep{weiss2000phenogenetic} or developmental system drift \citep{true2001developmental}.
+In this context, the Kalman decomposition \citep{kalman1963mathematical} implies that nearly all systems are nonidentifiable,
+and gives an analytical description of all phenotypically equivalent gene networks.
+System nonidentifiability, rather than being a computational nuisance,
+implies the existence of axes of genetic variation that are not constrained by selection.
+The independent movement of separated populations along these axes
+can lead to reduction in hybrid viability, and hence speciation,
+at a speed that depends on effective population size and amount of genetic variation. 
+In this model, at biologically reasonable parameter values,
+system drift is be a significant -- and often rapid -- driver of speciation.
+This may at first be surprising because
+hybrid inviability appears as a consequence of recombining different, yet functionally equivalent, mechanisms.
+
+Consistent with empirical observation \js{cite}, we see that the fitnesses of $F_2$ hybrids drop at a much faster rate than those of $F_1$s,
+since $F_1$ and $F_2$ phenotypes diverge linearly and at the square root of time relative to parentals, respectively.
+\plr{refer to Turelli here}
+% In this model, this is due to recombination shuffling system architectures, leaving the $F_2$ system homozygous for one parent at one particular locus and homozygous for the other parent at another possibly incompatible locus.
+Another natural consequence of the model is Haldane's rule,
+that if only one hybrid sex is inviable or sterile it is likely to be the heterogametic sex.
+This occurs because if the genes underlying a regulatory network are distributed among both autosomes and the sex chromosome, 
+then heterogametic $F_1$s show variation similar to that seen in $F_2$s.
+
+Is there evidence that this is actually occurring?
+System drift has been inferred across the tree of life \citep{johnson2017rewiring},
+and there is often significant regulatory variation segregating within populations \citep{XXX}.
+\js{cite sergey?}
+Transcription in hybrids between closely related species with conserved transcriptional patterns can also be divergent 
+\citep{haerty2006gene, maheshwari2012cis, coolon2014tempo, michalak2004association}.
+Furthermore, in cryptic species complexes (e.g., sun skinks \citep{barley2013challenge}),
+strongly reproductively incompatible species may be nearly morphologically indistinguishable.
+\plr{check the skinks have RI!!}
+Of course, selective pressures have likely not been constant over evolutionary time
+in any of these examples,
+but the process we describe here will still occur.
+This model therefore may provide a neutral ``null'' model of gene network evolution \citep{koonin2016splendor}.
+\plr{revisit}
 

### 2017-11-21 kalman_walk_paper.tex
more discussion edits

@@
 This model allows us to describe analytically the processes described verbally as
-phenogenetic drift \citep{weiss2000phenogenetic} or developmental system drift \citep{true2001developmental}.
+phenogenetic drift \citep{weiss2000phenogenetic} and developmental system drift \citep{true2001developmental}.
 In this context, the Kalman decomposition \citep{kalman1963mathematical} implies that nearly all systems are nonidentifiable,
@@
 In this model, at biologically reasonable parameter values,
-system drift is be a significant -- and often rapid -- driver of speciation.
+system drift is a significant -- and often rapid -- driver of speciation.
 This may at first be surprising because
@@
 Another natural consequence of the model is Haldane's rule,
-that if only one hybrid sex is inviable or sterile it is likely to be the heterogametic sex.
+that if only one $F_1$ hybrid sex is inviable or sterile it is likely to be the heterogametic sex.
 This occurs because if the genes underlying a regulatory network are distributed among both autosomes and the sex chromosome, 
@@
 but the process we describe here will still occur.
-This model therefore may provide a neutral ``null'' model of gene network evolution \citep{koonin2016splendor}.
+This model therefore may provide a neutral ``null'' model of gene network evolution, a strategy advocated for by \citet{koonin2016splendor}.
 \plr{revisit}
@@
 
-Although, as classically formulated, the Dobzhansky-Muller model of hybrid incompatibility is agnostic to the relative importance of neutral versus selective genetic substitutions \citep{coyne1998evolutionary}, \citet{orr2004speciation} remarks that hybrid incompatibility data suggests that ``speciation results from positive Darwinian selection within species'' thus maintaining one of the ``central tenets of the modern synthesis''.
+Although, as classically formulated, the Dobzhansky-Muller model of hybrid incompatibility is agnostic to the relative importance of neutral versus selective genetic substitutions \citep{coyne1998evolutionary} (as the neutral theory did not yet exist), \citet{orr2004speciation} remarks that hybrid incompatibility data suggests that ``speciation results from positive Darwinian selection within species'' thus maintaining one of the ``central tenets of the modern synthesis''.
+Reviewing both theoretical \citep{nei1983models} and empirical results on speciation, \citet{seehausen2014genomics} suggest that genetic drift is too slow to drive speciation and instead interpret genomic conflict to be a more significant driver.
+
+\emph{Ecological speciation} as espoused by Schluter (and sometimes referred to as \emph{parallel speciation} \citep{schluter1995parallel}) suggests that reproductive incompatibility is a byproduct of adaptation -- and stands in stark contrast to the model presented here -- as it specifically predicts that two independent and initially identical populations evolving under stabilizing selection will \emph{not} harbor incompatibilities \citep{schluter2009evidence}.
+This is not to say that our model does not predict that ecological divergence can drive speciaiton -- if adaptation manifests as a significant difference in optimal (and thus evolved) phenotypes ($\phi \neq \bar \phi$) between initially identical populations, hybrids between the two will surely be less fit in both parental environments, however, this type of incompatibility will be function/network specific and not necessarily entail general steriliy or developmental inviability (so long as the particular networks underlying these processes did not adapt to novel ecologies) in the absense of system drift.
+\jss{should I write out an example including the oscillator?
+Such as, if natural selection shifts the frequencies between adapting populations, hybrids will exhibit neither optimal frequncy of oscillation, however under the ecological speciation model, we would not expect to find incompatibilities in any other of the organism's gene networks.}
+
 %In practice it remains unclear whether system drift or other speciation mechanisms predominate, 
@@
 Our results show that hybrids, under neutral processes, break down as a function of genetic distance, and despite the present model's myriad simplifications and assumptions, 
-it may, in part, explain the observed broad patterns of reproductive isolation among diverse phyla \citep{roux2016shedding, hedges2015tree}.
-\js{maybe also cite the phylogentics lit here?
-Like the speciation clock stuff.}
+it may, in part, explain the observed broad patterns of reproductive isolation, such as the observed consistent relationship between molecular divergence and genetic isolation across diverse taxa and ecologies by \citet{roux2016shedding}, and the suggestivley nonadaptive, clocklike speciation patterns observed by \citet{hedges2015tree} (as diversification is not observed to be accelerated following mass extinctions).
 

### 2017-11-21 kalman_walk_paper.tex
minor Haldane edits

@@
 \paragraph{Haldane's rule}
+\jss{Maybe something like dosage compensation drift could also lead to Haldane's rule.
+That is, if the dosage compensation can either be: upregulate the X by $2$ in male flies, versus in another species: downregulate the X in females by $1/2$, and all possible combinations inbetween?}
 This model naturally predicts
@@
 $\left( \left[ \begin{smallmatrix} A_1 & A_2 \\ \cdot & \cdot \end{smallmatrix} \right], \left[ \begin{smallmatrix} A_1 & A_2 \\ X_1 & X_2 \end{smallmatrix} \right] \right)$
-has phenotype determined by $A = \left[ \begin{smallmatrix} A_1 & A_2 \\ X_1/2 & X_2/2 \end{smallmatrix} \right]$,
+  has phenotype determined by $A = \left[ \begin{smallmatrix} A_1 & A_2 \\ X_1 & X_2 \end{smallmatrix} \right]$, thanks to \emph{Drosophila}-like dosage compensation mechanism (the X is twice upregulated in heterogametes relative to homogametes),
 while a female homozygous for the haplotype
-$\left[\begin{smallmatrix} \bar A_1 & \bar A_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$,
-thanks to X-inactivation, has phenotype determined by
-$A = \left[\begin{smallmatrix} \bar A_1 & \bar A_2 \\ \bar X_1/2 & \bar X_2/2 \end{smallmatrix}\right]$.
+$\left[\begin{smallmatrix} \bar A_1 & \bar A_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$, has phenotype determined by
+$A = \left[\begin{smallmatrix} \bar A_1 & \bar A_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$.
 An $F_1$ male offspring of these two will have phenotype determined by
-$\frac{1}{2} \left[\begin{smallmatrix} A_1 + \bar A_1 & A_2 + \bar A_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$.
+$\left[\begin{smallmatrix} (A_1 + \bar A_1)/2 & (A_2 + \bar A_2)/2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$.
 If both genes resided on the autosomes this system would only be possible in an $F_2$ cross.
@@
 In their model, if alleles are strictly additive, that is neither dominant nor recessive, then male and female $F_1$ hybrids will drop in fitness at the exact same rates, and therefore defy Haldane's rule.
-Furthermore, the dominance theory struggles with dosage compensation and X-inactivation \citep{watson2012haldane}; the present model, however, complies with Haldane's rule with X-inactivation explicitly included.
 

### 2017-11-28 kalman_walk_paper.tex
discussion edits

@@
 
-Above we synthesize models from molecular and evolutionary biology with tools from control theory
-to study the evolution of a molecularly mechanistic model of the genotype-phenotype map under stabilizing selection.
-This model allows us to describe analytically the processes described verbally as
+Above we synthesize concepts and tools from molecular and evolutionary biology with tools from control theory
+to study the evolution of a mechanistic model of the molecular genotype-phenotype map under stabilizing selection.
+This model allows us to analytically describe the processes discussed verbally as
 phenogenetic drift \citep{weiss2000phenogenetic} and developmental system drift \citep{true2001developmental}.
@@
 Is there evidence that this is actually occurring?
-System drift has been inferred across the tree of life \citep{johnson2017rewiring},
+System drift and network rewiring has been inferred across the tree of life \citep{dalal2017transcription, johnson2017rewiring},
 and there is often significant regulatory variation segregating within populations \citep{XXX}.
@@
 \plr{check the skinks have RI!!}
-Of course, selective pressures have likely not been constant over evolutionary time
-in any of these examples,
-but the process we describe here will still occur.
-This model therefore may provide a neutral ``null'' model of gene network evolution, a strategy advocated for by \citet{koonin2016splendor}.
+%Of course, selective pressures have likely not been constant over evolutionary time
+%in any of these examples,
+%but the process we describe here will still occur.
+%This model therefore may provide a neutral ``null'' model of gene network evolution, a strategy advocated for by \citet{koonin2016splendor}.
 \plr{revisit}
-
-% nonlinearity
-%What about nonlinearity?
-As Richard Levins opined, models in population biology face a trade-off among precision, realism, and generality \citep{levins1966strategy}.
-As Levins expects, any tractable and general model, such as the present one under discussion, will have limitations.
-Most notable is linearity.
-It is often stated that life is not linear.
-This is often true, 
-%however, many of the ideas developed here should be generalizable to nonlinear cases (multi-linear systems, say). 
-however, we see this as a necessary first step in the direction of a more life-like (nonlinear, perhaps) evolutionary systems theory.
-Depending on an actual biological system's particularities, its non-linearity, may buffer or exacerbate effects elucidated in this paper.
-
-Although, as classically formulated, the Dobzhansky-Muller model of hybrid incompatibility is agnostic to the relative importance of neutral versus selective genetic substitutions \citep{coyne1998evolutionary} (as the neutral theory did not yet exist), \citet{orr2004speciation} remarks that hybrid incompatibility data suggests that ``speciation results from positive Darwinian selection within species'' thus maintaining one of the ``central tenets of the modern synthesis''.
-Reviewing both theoretical \citep{nei1983models} and empirical results on speciation, \citet{seehausen2014genomics} suggest that genetic drift is too slow to drive speciation and instead interpret genomic conflict to be a more significant driver.
-
-\emph{Ecological speciation} as espoused by Schluter (and sometimes referred to as \emph{parallel speciation} \citep{schluter1995parallel}) suggests that reproductive incompatibility is a byproduct of adaptation -- and stands in stark contrast to the model presented here -- as it specifically predicts that two independent and initially identical populations evolving under stabilizing selection will \emph{not} harbor incompatibilities \citep{schluter2009evidence}.
-This is not to say that our model does not predict that ecological divergence can drive speciaiton -- if adaptation manifests as a significant difference in optimal (and thus evolved) phenotypes ($\phi \neq \bar \phi$) between initially identical populations, hybrids between the two will surely be less fit in both parental environments, however, this type of incompatibility will be function/network specific and not necessarily entail general steriliy or developmental inviability (so long as the particular networks underlying these processes did not adapt to novel ecologies) in the absense of system drift.
-\jss{should I write out an example including the oscillator?
-Such as, if natural selection shifts the frequencies between adapting populations, hybrids will exhibit neither optimal frequncy of oscillation, however under the ecological speciation model, we would not expect to find incompatibilities in any other of the organism's gene networks.}
-
+Of course the basic assumptions of this model will be violated in practice (constant selective pressures, etc.), however, this model can function as a ``neutral null'' description of gene network evolution (a strategy advocated for in \citet{lynch2007frailty, koonin2016splendor}).
+If this model succeeds in describing actual system evolution, it can be inferred that the mechanism underlying species formation does not require the inclusion of adaptation or changes in selection to explain the emergence of hybrid incompatibility.
+
+\paragraph{The origin of species not by means of natural selection?}
+%\plr{(or, "Neutral speciation?")}
+% this is neutral, in contrast to other models. although some folks advocate nonadaptive speciation.
+% other mechanisms also probly important
+% but this one may be strong since happens in all networks at once
+%
+%The model we develop here is perhaps a plausible mechanism for strictly \emph{neutral} speciation.
+%This is perhaps surprising, as speciation is often thought to simply be a byproduct of adaptative change \citep{schluter2009evidence, orr2004speciation}
+%%as much of the speciation literature to date
+%%has focused on positive selection and/or genetic conflict,
+%or due to genetic conflict,
+%and since previous authors have argued that neutral processes should be too slow
+%\citep{nei1983models,seehausen2014genomics}.
+%In most cases to date where the precise molecular mechanism underlying hybrid incompatibility is known,
+%the genes responsible are thought to have been subject 
+%to recent positive selection \citep{examples} 
+%and/or genetic conflict \citep{examples}.
+%\plr{get examples from \citet{orr2004speciation}}
+%%However, incompatibility due to more diffuse system breakdown
+%%(e.g., widespread misregulation \citep{coolon2014tempo} \plr{check that} \jss{I don't think this is the right citation})
+%%may be likely much more difficult to uncover.
+%% \citet{orr2004speciation} remarks that hybrid incompatibility data suggests that ``speciation results from positive Darwinian selection within species''. 
+%On the other hand, the Dobzhansky--Muller model of hybrid incompatibility 
+%is agnostic to the origin of the genetic substitutions \citep{coyne1998evolutionary} \plr{cite bateson, etc},
+%and other plausible mechanisms for the origin of Dobzhanzky-Muller incompatibilities by neutral genetic drift
+%have been proposed \citep[e.g.][]{lynch_force}.
+Although, as classically formulated, 
+the Dobzhansky-Muller model of hybrid incompatibility is agnostic to the relative importance of neutral versus selective genetic substitutions \citep{coyne1998evolutionary},
+and plausible mechanisms for the origin of Dobzhansky--Muller incompatibilities by neutral genetic drift have been proposed \citep{lynch_force} 
+and under stabilizing selection \citep{fierst},
+previous authors have argued that neutral processes are likely too slow to be a significant driver of speciation \citep{nei1983models,seehausen2014genomics}.
+Using simulations, \citet{porter2002speciation} demonstrated the accumulation of hybrid incompatibilities under directional, but not stabilizing selection, and \citet{palmer2009dynamics} observed the appearence of incompatibilities in suboptimally adapted populations in a constant environment.
+This, in light of the few known incompatibilities, has lead some to conclude that hybrid incompatibility is typically a byproduct of positive selection \citep{orr2004speciation, schluter2009evidence} 
+or a conseuqence of genetic conflict \citep{examples}.
+In contrast, the model we develop here, suggests that even under strictly neutral conditions,  and in the absence of adaptation, system drift can lead to speciation and perhaps at a rate fast enough to play a major role in species formation across the tree of life.
+%In contrast, the model we develop here, is perhaps a plausible mechanism for strictly \emph{neutral} speciation. 
+%
 %In practice it remains unclear whether system drift or other speciation mechanisms predominate, 
 %as hybrid incompatibilities due to both selection and conflict have been uncovered and hint at general processes \citep{orr2004speciation}. 
-While it is known that divergent selection and genetic conflict, as well as other mechanisms, can precipitate speciation, we have shown here that, surprisingly, neutral drfit may also significantly drive species formation.
-
-If system drift induced hybrid incompatibility occurs as rapidly as our model sugests,
-%even in the absence of postive Darwinian selection, 
-this would imply that species originate not necessarily by means of natural selection \citep{darwin1859origin}.
-While the incompatibility of any one given network within an organism may be small, we note that organsisms are made up of many different networks, and that the cumulative impact of system drift across all of these may be substantial.
-Further, even in populations experiencing directional selection, networks underlying conserved parts of the phenotype can still experience neutral drift and harbor incompatibilities.
+%While it is known that divergent selection and genetic conflict, as well as other mechanisms, can precipitate speciation, we have shown here that, surprisingly, neutral drift may also significantly drive species formation.
 Our results show that hybrids, under neutral processes, break down as a function of genetic distance, and despite the present model's myriad simplifications and assumptions, 
 it may, in part, explain the observed broad patterns of reproductive isolation, such as the observed consistent relationship between molecular divergence and genetic isolation across diverse taxa and ecologies by \citet{roux2016shedding}, and the suggestivley nonadaptive, clocklike speciation patterns observed by \citet{hedges2015tree} (as diversification is not observed to be accelerated following mass extinctions).
-
-%Do we see evidence of this?  Well, maybe: misregulation in hybrids in Witkopp, Noor, Barbash, etc,
-%even with expression levels in parents being the same.
-
-% consequence: prediction for phenotype (not just fitness)
-%Note that this follows directly from the fact that $F_1$ and $F_2$ hybrid phenotypes diverge from $F_0$ phenotypes, linearly and inverse-quadratically, with respect to time (more precisely $T/N_e$).
-As such, removing assumptions about the form of the fitness function, this model still predicts hybrid $F_2$ phenotypes to diverge much faster than $F_1$s, initially.
-%So even if fitness isn't as above maybe you can see this in phenotypes.
-
-% selection moving around optimal set
-%Note that the drift above could actually be caused by the optimal phenotype shifting.
-%\js{I don't think we need to address optimal phenotype shifting}
+While the incompatibility of any one given network within an organism may be small, we note that organisms are made up of many different networks, and that the cumulative impact of system drift across all of these may be substantial.
+Further, even in populations experiencing directional selection, networks underlying conserved parts of the phenotype can still experience neutral drift and harbor incompatibilities.
+
+\paragraph{Nonlinearity}
+Of course, real regulatory networks are not linear dynamical systems.
+Most notably, physiological limits put upper bounds on expression levels,
+implying saturating response curves \citep{michalis_mentin}.
+It remains to be seen how well these results carry over into real systems,
+but the fact that most nonlinear systems can be locally approximated by a linear one
+suggests qualitatively similar results.
+Furthermore, nonidentifiability (which implies the existence of neutral directions)
+is often found in practice in moderately complex models of biological systems. \plr{need citation}
+
+\paragraph{Peter: Quant gen discussion}
+Maybe compare to \citet{fierst}?
 
@@
 
-%This theoretical framework can easily be applied to other interesting questions in evolutionary biology not tackled presently: such as the evolution of linkage, the necessity of network complexity (does evolution tend towards Rube Goldberg or parsimonious network organization?), evolvability, structure/function inference, and intra-population context dependency of mutational effects, as well as many others.
-
 Note that islands or bottlenecks leave large $\sigma_N$ but small $N_e$ so may go fast.
 
-% J and P didn't get speciation in a neutral  model
-Discuss differences to Johnson and Porter.
-
 What is Ne?  In a metapopulation depends on migr rate: classic Turelli paper.
-
-\jss{talk about speciation rate in small population results as compared to Katari and Goldstein (I think). they look at the influence of $N_e$ on speciation rate.}
 

### 2017-11-28 kalman_walk_paper.tex
edited intro and discussion

@@
 A complex molecular machinery translates an organism's genome into the characteristics on which natural selection acts,
-her phenotype.
+the phenotype.
 % Bridging the gulf between an organism's genome and phenotype is a poorly understood and complex molecular machinery. 
-Attaining a general understanding of the functioning and evolution of this molecular machinery
-is an overarching goal of many subdisciplines of biology.
+%Attaining a general understanding of the functioning and evolution of this molecular machinery
+It is an overarching goal of many biological subdisciplines to attain a general understanding of the function and evolution of this molecular machinery. 
 For example, 
-there is a growing body of data on evolutionary histories and molecular characterizations of particular gene regulatory networks
+there is a growing body of data on the evolutionary histories and molecular characterizations of particular gene regulatory networks
 \citep{jaeger2011gap, davidson2006gene, israel2016comparative}, 
@@
 and thus has the potential to discover general principles in the organization of biological systems 
-and provide concrete numerical predictions \citep{servedio2014not}.
+as well as provide concrete numerical predictions \citep{servedio2014not}.
 
 The dynamics of the molecular machinery and its interactions with the environment
-can be mathematically described in various ways as a dynamical system \citep{jaeger2015comet}.
+can be mathematically described as a dynamical system \citep{jaeger2015comet}.
 Movement in this direction is ongoing, as researchers have begun to study 
@@
 there can be uncertainty about an inferred model's parameters or even its claims about
-causal structure, even with access to complete and perfect data \citep{bellman1970structural, grewal1976identifiability, walter1984structural}. 
+causal structure, despite access to complete and perfect data \citep{bellman1970structural, grewal1976identifiability, walter1984structural}. 
 Models with different parameter schemes, or even different mechanics 
-can equally accurately predict the observed behavior of a given physical system, 
-but still not actually reflect the internal dynamics of the system.
+can make equally accurate predictions,
+but still not actually reflect the internal dynamics of the system being modelled.
 In control theory, where electrical circuits and mechanical systems are often the focus, 
@@
 even if a one-to-one mapping between model and behavior is possible in theory,
-even tiny amounts of noise can make inference problems nonidentifiable in practice.
+  even tiny amounts of noise can make inference problems nonidentifiable in practice \jss{cite?}.
 
@@
 macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level 
-\citep{true2001developmental, tanay2005conservation, tsong2006evolution, hare2008sepsid, vierstra2014mouse,  dalal2016transcriptional, dalal2017transcription}.
-% are these good here? {taylor2016diverse, matsui2015regulatory, stergachis2014conservation}.
-
-EDIT IN THESE REFS?
-The literature is filled with detailed observations of molecular systems and their diversity. 
-\plr{Diversity doesn't imply system drift -- only if the diverse systems are homologous.}
-There are examples of significant diversity in the networks underlying processes such as 
-circadian rhythm \citep{sancar2008intelligent}, 
-cell cycle control \citep{cross2011evolution, kearsey2003enigmatic}, 
-pattern formation \plr{cite?}, 
-and metabolism \citep{lavoie2009rearrangements, martchenko2007transcriptional, christensen2011unique, hartl2007induction, alam2013aspergillus}.
-
-AND THESE?
-Over the last several years, several different computational approaches have been applied to study reproductive incompatibility and speciation.
-\citet{tulchinsky} simulated the evolution of a transcription factor and its binding site using a thermodynamic model.
+\citep{true2001developmental, tanay2005conservation, tsong2006evolution, hare2008sepsid, vierstra2014mouse, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription}.
+  \jss{should hare2008sepsid be here? It's sequence but not ``system'' drift}
+% are these good here? {taylor2016diverse -- maybe not, matsui2015regulatory -- I think this relevant, stergachis2014conservation}.
+%
+%AND THESE?
+  %% Over the last several years, different computational models have been applied to study reproductive incompatibility and speciation in models of gene regulation \citep{porter2002speciation, tulchinsky, palmer2009dynamics, tulchinsky2014hybrid, khatari2015simple}.
+  %\citet{tulchinsky} simulated the evolution of a transcription factor and its binding site using a thermodynamic model.
 Their simulations suggest that the language by which a transcription factor recognizes its binding site can change, and potentially lead to hybrid incompatibility when allopatric populations employ divergent readout languages.
@@
 Johnson and Porter 2000 did not observe any hybrid fitness declines under stabilizing selection -- only under directional selection.
-Khatari et al, Tulchinsky et al, and Porter et al, all study hybrid incompatibility from a transcription factor/binding site interaction perspective, not from an overall network architecture perspective.
+%Khatari et al, Tulchinsky et al, and Porter et al, all study hybrid incompatibility from a transcription factor/binding site interaction perspective, not from an overall network architecture perspective.
 Palmer and Feldman only see hybrid incompatibility in constant environments if the parental populations are relatively poorly adapted initially.
 Otherwise hybrids between two allopatric populations have fairly high fitnesses.
-MAKE SURE TO REF DMI LIT SOMEWHERE and Barton's paper.
-
-In this paper we outline a theoretical framework to study the evolution of biological systems
-that can be described by systems of differential equations, such as gene regulatory networks.
-We study the scenario
-where the optimal phenotype remains constant over evolutionary time,
-so that despite strong stabilizing selection for phenotype,
-neutral drift in the underlying genotype is possible.
-Results from systems theory provide
+
+In this paper we outline a theoretical framework to study the evolution of biological systems, such as gene regulatory networks.
+We study the evolution of an optimally adapted population subject to stabilizing selection for phenotype -- that is the population evolves under constant selective and environmental pressures.
+Even if the phenotype remains stable over evolutionary time, the underlying mechanism might not remain so,
+%Even when stabilizing selection preserves phenotype over evolutionary time, the underlying mechanism may change, 
+as many distinct (and mutationally connected) molecular pathways can realize identical phenotypes.
+%where the optimal phenotype remains constant over evolutionary time,
+%so that despite strong stabilizing selection for phenotype,
+%neutral drift in the underlying genotype is possible.
+Applying results from system theory, we present
 an analytical description of the set of 
-all linear gene network architectures that yield identical phenotypes,
-which gives concrete expectations for how any such
-system can, in principal, undergo system drift and rewire. 
-Even with stabilization selection, a population will explore the set of all possible phenotypically equivalent gene networks.
+all linear gene network architectures that yield identical phenotypes, and show that any biological system, in principal, can undergo system drift, as there is never a unique system architecture per phenotype.
+%which gives concrete expectations for how any such
+%system can, in principal, undergo system drift and rewire.
+Even under stabilizing selection, a population can explore the set of all possible phenotypically equivalent gene networks.
+Phenotypically equivalent gene networks are not necessarily compatible with one another, thus system drift can rapidly lead to the accumulation of hybrid incompatibilities analogous to those in the Dobzhansky-Muller model.
 Consequentially, two populations isolated for a sufficiently long period of time, 
@@
 
+REFERENCE: Barton's paper (PETER).
+
 \section*{Methods}
 
-We study an abstract model of temporal dynamics of the concentrations of a collection of $n$ coregulating molecules
-within an organism, that may also be affected by temporally varying signals from the environment.
-% Organisms' phenotypes are constructed by gene by gene and gene by environment interactions. 
-%Here we simply define the \emph{phenotype} to be those aspects of the dynamics directly under natural selection
-%-- the \emph{what}, \emph{when}, and \emph{how much}, of an organism's molecules that are physiologically or otherwise relevant to survival.
+Here we construct as abstract model to study biological systems, such as gene regulatory networks.
+The behavior (i.e.
+temporal dynamics) of such a system is determined both by a collection of $n$ coregulating moleculues -- such as transcription factors, as well as external or environmental inputs.
+%We study an abstract model of temporal dynamics of the concentrations of a collection of $n$ coregulating molecules
+%within an organism, that may also be affected by temporally varying signals from the environment.
+%% Organisms' phenotypes are constructed by gene by gene and gene by environment interactions. 
+%%Here we simply define the \emph{phenotype} to be those aspects of the dynamics directly under natural selection
+%%-- the \emph{what}, \emph{when}, and \emph{how much}, of an organism's molecules that are physiologically or otherwise relevant to survival.
 We write $\kappa(t)$ for the vector of $n$ molecular concentrations at time $t$.
-There are also $m$ ``inputs'' determined exogenously to the system, denoted $u(t)$,
-and $\ell$ ``outputs'', denoted $\phi(t)$.
+The $m$ ``inputs'' determined exogenously to the system are denoted $u(t)$,
+and the $\ell$ ``outputs'' are denoted $\phi(t)$.
 The output is merely a linear function of the internal state:
@@
 so that the kryptotype measures deviations from initial concentrations. 
-Here $A$ can be any $n \times n$ matrix, $B$ a $n \times m$, and $C$ any $\ell \times n$ dimensional matrix,
+Here $A$ can be any $n \times n$ matrix, $B$ an $n \times m$, and $C$ any $\ell \times n$ dimensional matrix,
 with usually $\ell$ and $m$ less than $n$.
-We think of the system as the triple $\mathcal{S} = (A,B,C)$,
+We think of the system as the triple $(A,B,C)$,
 which translates (time-varying) $m$-dimensional input $u(t)$ into the $\ell$-dimensional output $\phi(t)$.
@@
 If $A_{ij} > 0$, then $\kappa_j$ upregulates $\kappa_i$, while if $A_{ij} < 0$, then $\kappa_j$ downregulates $\kappa_i$.
-The $i$th row of $A$ is therefore determined by genetic features such as
+The $i^\text{th}$ row of $A$ is therefore determined by genetic features such as
 the strength of $j$-binding sites in the promoter of gene $i$,
@@
 \plr{revisit}
-Of course the basic assumptions of this model will be violated in practice (constant selective pressures, etc.), however, this model can function as a ``neutral null'' description of gene network evolution (a strategy advocated for in \citet{lynch2007frailty, koonin2016splendor}).
+Of course the basic assumptions of this model will be violated in practice (constant selective pressures, etc.), however, this model can function as a ``neutral null'' description of gene network evolution (a strategy advocated for in \citet{lynch2007frailty, fay2008evaluating, koonin2016splendor}).
 If this model succeeds in describing actual system evolution, it can be inferred that the mechanism underlying species formation does not require the inclusion of adaptation or changes in selection to explain the emergence of hybrid incompatibility.
@@
 or a conseuqence of genetic conflict \citep{examples}.
-In contrast, the model we develop here, suggests that even under strictly neutral conditions,  and in the absence of adaptation, system drift can lead to speciation and perhaps at a rate fast enough to play a major role in species formation across the tree of life.
+In contrast, the model we develop here, suggests that even under strictly neutral conditions, system drift can lead to speciation, perhaps at a rate fast enough to play a role in species formation across the tree of life.
 %In contrast, the model we develop here, is perhaps a plausible mechanism for strictly \emph{neutral} speciation. 

### 2017-11-28 kalman_walk_paper.tex

@@
 
-Consistent with empirical observation \js{cite}, we see that the fitnesses of $F_2$ hybrids drop at a much faster rate than those of $F_1$s,
+Consistent with empirical observation of \emph{hybrid breakdown} \citep[e.g.][]{plotner2017chlorosis}), we see that the fitnesses of $F_2$ hybrids drop at a much faster rate than those of $F_1$s,
 since $F_1$ and $F_2$ phenotypes diverge linearly and at the square root of time relative to parentals, respectively.
@@
 Furthermore, in cryptic species complexes (e.g., sun skinks \citep{barley2013challenge}),
-strongly reproductively incompatible species may be nearly morphologically indistinguishable.
-\plr{check the skinks have RI!!}
+%strongly reproductively incompatible species
+genetically distinct species
+may be nearly morphologically indistinguishable.
+\plr{check the skinks have RI!!} \jss{the paper just says highly ``genetically distinct''; nothing is stated about RI}
 %Of course, selective pressures have likely not been constant over evolutionary time
@@
 This, in light of the few known incompatibilities, has lead some to conclude that hybrid incompatibility is typically a byproduct of positive selection \citep{orr2004speciation, schluter2009evidence} 
-or a conseuqence of genetic conflict \citep{examples}.
+or a conseuqence of genetic conflict \citep{presgraves2010molecular, crespi2013conflictual}.
 In contrast, the model we develop here, suggests that even under strictly neutral conditions, system drift can lead to speciation, perhaps at a rate fast enough to play a role in species formation across the tree of life.

### 2017-11-28 sims/perturb_Kalman.r
sim code

@@ -145,10 +145,55 @@ sys_gen2 <- function (s=sigma, eps=0.05, nreps=100)
 #########################
 
 pert_sim <- function(rgens){
-  pheno_divergence <- sys_gen2(s=0.1, eps=0.05, nreps=100)
+  pheno_divergence <- sys_gen2(s=0.01, eps=0.1, nreps=100)
   for(i in 2:rgens)
   {
-    ph0 <- sys_gen2(s=0.01,eps=0.05,nreps=100)
+    ph0 <- sys_gen2(s=0.01,eps=0.1,nreps=100)
     pheno_divergence <- rbind(pheno_divergence, ph0)
   }
+  return(pheno_divergence)
 }
+
+####################################
+V <- function (tau) {
+  matrix(c(1,tau,0,1-tau), nrow=2) 
+}
+
+min_perturb <- function(tau=0){
+  initA <- V(tau)%*%A%*%solve(V(tau))
+  output <- matrix(0,nrow=100,ncol=2)
+  colnames(output) <- c("sys_dist", "pheno_dist")
+  for (i in 1:100){
+    eps2 <- i/200
+    new_A <- V(tau-eps2)%*%A%*%solve(V(tau-eps2))
+    dist <- norm(initA - new_A, "F")
+    avg_div <- 0
+    for (j in 1:100) 
+    { 
+      Z1 <- recombine(initA, new_A)
+      Z2 <- recombine(initA, new_A)
+      Z3 <- (Z1 + Z2)/2
+      
+      DZ <- D(Z3, BK=B, CK=C)
+      avg_div <- avg_div + DZ/100
+    }
+    output[i,] <- c(dist,avg_div)
+  }
+  return(output)
+}
+
+#####################
+# COMPARE minimal vs non-minimal system #
+
+min0 <- min_perturb(tau=0)
+nonmin0 <- sys_gen2(s=0.01, eps=0.1, nreps=100)
+
+pdf("~/kalman_walk/sims/2d_vs_6d_oscillator_tau0.pdf")
+plot(min0, col="blue")
+lines(lowess(min0), col="blue")
+points(nonmin0, col="black")
+lines(lowess(nonmin0), col="black")
+legend(0.1,4,legend=c("2D minimal system", "6D system"), col=c("blue", "black"), pch=1)
+dev.off()
+
+##############

### 2017-11-28 sims/perturb_Kalman.r
minor code rearrangement

@@ -101,7 +101,7 @@ perturb_oscillator <- function(AL, eps) {
 }
 
 #####################################
-sys_gen2 <- function (s=sigma, eps=0.05, nreps=100) 
+sys_gen2 <- function (s, eps=0.05, nreps=100) 
 {
   KalA_list <- random_oscillator(s=s)
   KalA <- assemble_oscillator(KalA_list)
@@ -116,27 +116,29 @@ sys_gen2 <- function (s=sigma, eps=0.05, nreps=100)
     dist <- norm(KalA$A-new_KalA$A,"F")
     
     avg_div <- 0
-    jj <- 0
-    for (j in 1:100) 
+    jj <- 0 # number of errors
+    max_errors <- 10
+    j <- 0
+    while (j < 100) {
     { 
+      j <- j + 1
       Z1 <- recombine(KalA$A, new_KalA$A)
       Z2 <- recombine(KalA$A, new_KalA$A)
       Z3 <- (Z1 + Z2)/2
       DZ <- tryCatch(D(Z3, BK=KalA$B, CK=KalA$C), error=function(e) NULL)
-      if (is.null(DZ)==1)
+      if (is.null(DZ))
       {
         j = j-1
         jj = jj+1
-        if(jj==10)
+        if(jj==max_errors)
         {
           break
         }
+      } else {
+        avg_div <- avg_div + DZ
       }
-      if (is.null(DZ)==0) 
-        {
-          avg_div <- avg_div + DZ/100
-        }
     }
+    avg_div <- avg_div / j
     output[i,] <- c(dist,avg_div)
   }
   return(output)
@@ -181,19 +183,3 @@ min_perturb <- function(tau=0){
   }
   return(output)
 }
-
-#####################
-# COMPARE minimal vs non-minimal system #
-
-min0 <- min_perturb(tau=0)
-nonmin0 <- sys_gen2(s=0.01, eps=0.1, nreps=100)
-
-pdf("~/kalman_walk/sims/2d_vs_6d_oscillator_tau0.pdf")
-plot(min0, col="blue")
-lines(lowess(min0), col="blue")
-points(nonmin0, col="black")
-lines(lowess(nonmin0), col="black")
-legend(0.1,4,legend=c("2D minimal system", "6D system"), col=c("blue", "black"), pch=1)
-dev.off()
-
-##############

### 2017-11-28 sims/perturb_Kalman.r
minor sim code edits

@@ -83,6 +83,30 @@ assemble_oscillator <- function (AL) {
   return(list(A=KalA, B=BK, C=CK))
 }
 
+assemble_oscillator4d <- function (AL) {
+    # AL should be a list with components:
+    # A
+    # A12
+    # Arno
+    # Anrno
+    # A13
+    # B
+    # C
+    # P
+  A00 <- matrix(c(0,0,0,0), nrow=2);
+  B00 <- matrix(c(0,0), ncol=1);
+  C00 <- matrix(c(0,0), nrow=1);
+  AK <- (rbind(cbind(AL$A, A00),
+               cbind(AL$A12, AL$Arno)));
+  # pad B and C with zeros
+  BK <- matrix(c(AL$B, rep(0, 2)), ncol=1)
+  CK <- matrix(c(AL$C, rep(0, 2)), nrow=1)
+  invP <- solve(AL$P[1:4,1:4]);
+  KalC <- CK%*%AL$P[1:4,1:4];
+  KalB <- AL$P[1:4,1:4]%*%BK;
+  KalA <- AL$P[1:4,1:4]%*%AK%*%invP
+  return(list(A=KalA, B=BK, C=CK))
+}
 # Perturb an oscillator of the form above
 perturb_oscillator <- function(AL, eps) {
     mA12 <- matrix(c(rnorm(4, mean=0, sd=eps)),nrow=2,ncol=2)
@@ -101,7 +125,7 @@ perturb_oscillator <- function(AL, eps) {
 }
 
 #####################################
-sys_gen2 <- function (s=sigma, eps=0.05, nreps=100) 
+sys_gen6D <- function (s=sigma, eps=0.05, nreps=100) 
 {
   KalA_list <- random_oscillator(s=s)
   KalA <- assemble_oscillator(KalA_list)
@@ -114,6 +138,7 @@ sys_gen2 <- function (s=sigma, eps=0.05, nreps=100)
     new_KalA_list <- perturb_oscillator(KalA_list, eps=eps)
     new_KalA <- assemble_oscillator(new_KalA_list)
     dist <- norm(KalA$A-new_KalA$A,"F")
+    #dist <- sum(abs(Kal$A - new_KalA$A))/36
     
     avg_div <- 0
     jj <- 0
@@ -142,6 +167,49 @@ sys_gen2 <- function (s=sigma, eps=0.05, nreps=100)
   return(output)
 }
 
+sys_gen4D <- function (s=sigma, eps=0.05, nreps=100) 
+{
[…]

### 2017-11-29 kalman_walk_paper.tex

@@
                 \node (A) at (0,0) {$\kappa_{1}$};
-                \node[dashed] (B) at (3,0) {$\kappa_{2}$};
-                \node[shape=rectangle] (U) at (1.5,1.5) {input ($u(t)$)};
-                \node[shape=rectangle] (y) at (1.5,-1.5) {output ($\phi(t)$)};
+                \node[dashed] (B) at (4,0) {$\kappa_{2}$};
+                \node[shape=rectangle] (U) at (2,2) {input ($u(t)$)};
+                \node[shape=rectangle] (y) at (2,-2) {output ($\phi(t)$)};
             \end{scope}
@@
     \node[dashed] (B) at (4,0) {$\kappa_{2}$};
-    \node[shape=rectangle] (U) at (2,2) {input};
-    \node[shape=rectangle] (y) at (2,-2) {output};
+    \node[shape=rectangle] (U) at (2,2) {input $(u)$};
+    \node[shape=rectangle] (y) at (2,-2) {output $(\phi)$};
 \end{scope}
@@
     \path [->, sloped] (B) edge[bend left] node {\tiny $-(\tau-1)^{-1}$} (A); 
-    \path[->] (U) edge[dashed] node {\tiny $1$} (A);
-    \path[->] (U) edge[dashed] node {\tiny $1$} (B);
-    \path[->] (A) edge[dashed, bend right] node {\tiny $1$} (y);
+    \path[->] (U) edge node {\tiny $1$} (A);
+    \path[->] (U) edge node {\tiny $1$} (B);
+    \path[->] (A) edge[bend right] node {\tiny $1$} (y);
 \end{scope}
@@
     \label{fig:all_osc}
+\end{figure}
+%
+\begin{figure}[H]
+\begin{tikzpicture}
+\begin{scope}[every node/.style={circle,thick,draw}]
+  \node (A) at (0,0) {$\kappa_{1}$};
+    \node (B) at (0,4) {$\kappa_{2}$};
+    \node (C) at (3.464,2) {$\kappa_{3}$};
+    \node[shape=rectangle, rotate=0] (U) at (-2.5,2) {input $(u)$};
+    \node[shape=rectangle] (y) at (3.464,0) {output $(\phi)$};
+\end{scope}
+
+\begin{scope}[>={Stealth[black]},
+              every node/.style={fill=white,circle},
+              every edge/.style={draw=black, thick}]
+    \path [->, sloped] (A) edge[bend left] node {\tiny ?} (B);
+    \path [->, sloped] (B) edge[bend left] node {\tiny ?} (A); 
+    \path [->, sloped] (A) edge[bend right] node {\tiny ?} (C); 
+    \path [->, sloped] (B) edge[bend left] node {\tiny ?} (C); 
+    \path [->, sloped] (C) edge[bend right] node {\tiny ?} (A); 
+    \path [->, sloped] (C) edge[bend left] node {\tiny ?} (B); 
+    \path[->] (U) edge node {\tiny $1$} (A);
+    \path[->] (U) edge node {\tiny $1$} (B);
+    \path[->] (A) edge node {\tiny $1$} (y);
+\end{scope}
+\begin{scope}[>={Stealth[black]},
+              every edge/.style={draw=black, thick}]
+    \path [->] (A) edge[loop below] node[sloped, anchor=center, below] {\tiny ?} (A);
+    \path [->] (B) edge[loop above] node[sloped, anchor=center, above] {\tiny ?} (B);
+    \path [->] (C) edge[loop right] node[sloped, anchor=center, above] {\tiny ?} (C);
+\end{scope}
+\end{tikzpicture}
+  \caption{A possible non-minimal three-gene oscillator phenotypically equivalent to the systems in Examples \ref{ex:oscillator} and \ref{ex:all_osc}.}
+  \label{fig:3D_osc}
 \end{figure}

### 2017-11-29 sims/perturb_Kalman.r
redid plots for F1 oscillators in R

@@ -253,7 +253,6 @@ min_perturb <- function(tau=0){
   }
   return(output)
 }
-<<<<<<< HEAD
 
 #####################
 # COMPARE minimal vs non-minimal system #
@@ -262,7 +261,7 @@ min0 <- min_perturb(tau=0)
 nonmin0 <- sys_gen6D(s=0.01, eps=0.1, nreps=100)
 nonmin0_4d <- sys_gen4D(s=0.01, eps=0.1, nreps=100)
 
-pdf("~/kalman_walk/sims/2d_4d_6d_oscillator_tau0.pdf")
+pdf("~/kalman_walk/sims/coeff_dist_2d_4d_6d_oscillator_tau0.pdf")
 plot(min0, col="blue")
 lines(lowess(min0), col="blue")
 
@@ -276,5 +275,3 @@ legend(0.1,4,legend=c("2D minimal system", "4D system", "6D system"), col=c("blu
 dev.off()
 
 ##############
-=======
->>>>>>> 60e6dce9f2929a8aa50a235e1b778311c6c899fd

### 2017-11-29 kalman_walk_paper.tex
plots

@@
   \begin{figure}[H]
+    \centering
+    \begin{tabular}{ccc}
+      \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F1_hundreth_tau0} &
+      \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F2_hundrethtau0} \\
+      \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F1_tenth_tau0.pdf} &
+      \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F2_tenthtau0.pdf} \\
+      \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F1_half_tau0.pdf} &
+      \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F2_halftau0.pdf}
+    \end{tabular}
+    \end{figure}
+
+  \begin{figure}[H]
   \label{fig:osc_incompat}

### 2017-11-29 examples/F2_osc_plotting.R
plots

@@ -1,0 +1,71 @@
+####################
+
+A <- matrix(c(0,-1,1,0), nrow=2);
+B <- matrix(c(1,1), ncol=1);
+C <- matrix(c(1,0), nrow=1);
+
+V <- function (tau) {
+  matrix(c(1,tau,0,1-tau), nrow=2) 
+}
+
+S <- function(tau) {
+  V(tau)%*%A%*%solve(V(tau))
+}
+
+gamete_gen <- function(tau, eps) {
+gametes <- list()
+count <- 0
+for (i in 0:1)
+{
+  for (j in 0:1)
+  {
+    for (k in 0:1)
+    {
+      for (l in 0:1)
+      {
+        count <- count + 1
+        M <- matrix(c(i,k,j,l), ncol=2, nrow=2)
+        name <- paste("G", count, sep = ""); 
+        assign(name, S(tau)*M + S(tau + eps)*(1 - M))
+        gametes[[name]] <- get(name)
+      }
+    }
+  }
+}
+return(gametes)
+}
+
+progeny_gen <- function(tau, eps){
+  gametes <- gamete_gen(tau, eps)
+  progeny <- list()
+  count2 <- 0
+  for (i in 1:16)
+  {
+    for (j in 1:16)
+    {
+      count2 <- count2 + 1
+      progeny[[count2]] <- (gametes[[i]] + gametes[[j]])/2
+    }
+  }
+    return(progeny)
+}
+
+########
+h <- function (t,tau=0,A=A_fn(tau)) { sapply(t, function (tt) C %*% expm(tt*A) %*% B) }
+optimal_h <- function (t) { sin(t)+cos(t) }
+Here is the impulse response function:
+  
+  tt <- seq(0,2*pi, length.out=100)
+plot(tt, h(tt), type='l',
+     xlab='t', ylab=expression(h(t)), main='impulse response function')
+#lines(tt, h(tt, tau=0.1), col='blue', lty=3)
+#lines(tt, optimal_h(tt), col='red', lty=2)
+######
+
+pdyn <- function(t, A) { sapply(t, function (tt) C %*% expm(tt*A) %*% B) }
+tt <- seq(0,10*pi, length.out=1000)
+plot(tt, pdyn(tt, A=A), type='l',
+     xlab='t', ylab=expression(h(t)), main='Phenotype', xlim=c(0,30), ylim=c(-6,6))
+for (i in 1:256){
+lines(tt, pdyn(tt, A=progeny[[i]]), col=sample(256,1), lty=1)
+}

### 2017-12-01 kalman_walk_paper.tex
general edits to paper. Changed figure 6 (f2 vs f1 divergence)

@@
 \begin{abstract}
-It is known that even if a species' phenotype has remained unchanged over evolutionary time,
-the underlying mechanism may have changed,
-since distinct molecular pathways can realize identical phenotypes.
-In this paper, we use quantitative genetics and linear systems theory to study
-how a gene network underlying a conserved phenotype
-evolves as genetic drift of small mutational tweaks to the molecular pathways
+%It is known that even if a species' phenotype has remained unchanged over evolutionary time,
+%the underlying mechanism may have changed,
+%since distinct molecular pathways can realize identical phenotypes.
+Even if a species' phenotype remains unchanged over evolutionary time, the underlying mechanism may have changed, as distinct molecular pathways can realize identical phenotypes.
+Here we use quantitative genetics and linear system theory to study
+how a gene network underlying a conserved phenotype evolves,
+as the genetic drift of small mutational tweaks to these molecular pathways
 cause a population to explore the set of mechanisms with identical phenotypes.
-In this setting we treat organisms as ``black boxes'' for which the environment provides input
-and the phenotype is the output,
+In this setting we treat an organism as a ``black box''
+for which the environment provides input and the phenotype is the output,
+%which translates some environmental input
+%to an output that is subject to the pressures of natural selection-- the phenotype.
 and there exists an exact characterization of the set of all mechanisms that give the same input--output relationship.
-We show that in this situation, there is never a unique architecture for any phenotype
+%There exists an exact expression for the set of all linear mechanisms with identical phenotypes, 
+%Due to the genetic drift of small mutational tweaks to these molecular pathways, we expect an evolving population to explore this set.
+Within this framework, we show that there is never a unique network architecture for any phenotype
 and that the evolutionary exploration of these distinct and mutationally connected mechanisms
-can lead to the rapid accumulation of hybrid incompatibilities between allopatric populations.
-We estimate that in reasonably numerous species,
-this process could thus the formation of new species
-over a number of generations proportional to the effective population size.
-This model also provides a natural explanation for Haldane's rule
-that appears mechanistically distinct from other proposed mechanisms.
+can lead to the reproductive incompatibility of idependently evolving populations.
+  This evolutionary exploration, which we refer to as system drift, proceeds at a rate proportional to the effective population size ($N_e$) and amount of intrapopulation genetic variation.
+We estimate that %in reasonably numerous species,
+  this process can lead to the formation of new species, perhaps in as few as $0.1 N_e$ generations.
+%over a number of generations proportional to the effective population size.
+This model also naturally yields, what appears to be, a distinct explanation for Haldane's rule, or why heterogametic hybrids tend to be disrupted more often than homogametes during the early stages of speciation.
 \end{abstract}
 
-\plr{Additional ideas to consider adding:}
-\begin{itemize}
-    \item speciation literature
+%\plr{Additional ideas to consider adding:}
+%\begin{itemize}
+%    \item speciation literature
     %\item Haldane's rule (easy point in discussion: makes F1s look like F2s)
@@
     %\item hybrid vigor (need to do calculation)
-    \item discuss linearity, linearization, and canalization in introduction
+ %   \item discuss linearity, linearization, and canalization in introduction
     %\item obtain estimates of variation in $A$ from thermodynamic occupancy model
-    \item ``On the origin of species not by means of natural selection''
-\end{itemize}
+ %   \item ``On the origin of species not by means of natural selection''
+%\end{itemize}
 
@@
 Although nonidentifiability may frustrate the occasional engineer or scientist, viewed from another angle,
-these concepts can provide a starting point for thinking about externally equivalent systems
+this concept can provide a starting point for thinking about externally equivalent systems
 -- systems that evolution can explore, so long as the parameters and structures can be realized biologically.
@@
 \citep{true2001developmental, tanay2005conservation, tsong2006evolution, hare2008sepsid, vierstra2014mouse, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription}.
-  \jss{should hare2008sepsid be here? It's sequence but not ``system'' drift}
+ % \jss{should hare2008sepsid be here? It's sequence but not ``system'' drift}
 % are these good here? {taylor2016diverse -- maybe not, matsui2015regulatory -- I think this relevant, stergachis2014conservation}.
@@
 Even under stabilizing selection, a population can explore the set of all possible phenotypically equivalent gene networks.
-Phenotypically equivalent gene networks are not necessarily compatible with one another, thus system drift can rapidly lead to the accumulation of hybrid incompatibilities analogous to those in the Dobzhansky-Muller model.
-Consequentially, two populations isolated for a sufficiently long period of time, 
-will likely produce inviable hybrids, despite the absence of adaptation, directional selection, or environmental change.
+Phenotypically equivalent gene networks are not necessarily compatible with one another, as such system drift may result in the reproductive incompatibility between populations isolated for a sufficiently long period of time, even in the absence of any sort of adaptative, selective, or environmental change.
+Finally, in some cases we estimate that reproductive incompatibility due to system drift can manifest on timescales on the order of $N_e$ generations.
+
+analogous to those in the Dobzhansky-Muller model. 
+%Consequentially, two populations isolated for a sufficiently long period of time, 
+%will likely produce inviable hybrids, despite the absence of adaptation, directional selection, or environmental change.
 % Under these conditions, speciation typically occurs on timescales approximately on the order of $N_{e}$ generations, where $N_{e}$ is the effective population size.
@@
 
-\section*{Methods}
+\section*{Evolutionary system theory}
 
@@
 The behavior (i.e.
-temporal dynamics) of such a system is determined both by a collection of $n$ coregulating moleculues -- such as transcription factors, as well as external or environmental inputs.
+temporal dynamics) of such a system is determined by a collection of $n$ coregulating moleculues -- such as transcription factors, as well as external or environmental inputs.
 %We study an abstract model of temporal dynamics of the concentrations of a collection of $n$ coregulating molecules
@@
 Furthermore, the ease of analyzing linear systems makes this an attractive place to start.
+To demonstrate this approach, we apply it to construct a simple gene network in Example \ref{ex:oscillator} below.
 
@@
       \centering
-         \begin{tabular}{cc}
+       %  \begin{tabular}{cc}
             \begin{tikzpicture}
@@
                 \node (A) at (0,0) {$\kappa_{1}$};
-                \node[dashed] (B) at (4,0) {$\kappa_{2}$};
-                \node[shape=rectangle] (U) at (2,2) {input ($u(t)$)};
-                \node[shape=rectangle] (y) at (2,-2) {output ($\phi(t)$)};
+                \node (B) at (4,0) {$\kappa_{2}$};
+                \node[shape=rectangle] (U) at (2,2) {input ($u$)};
+                \node[shape=rectangle] (y) at (2,-2) {\begin{tabular}{cc} output ($\phi$) \\  \includegraphics[width=0.15\textwidth, height=0.04\paperheight]{osc_impulse} \end{tabular}};
             \end{scope}
@@
                 \path [->] (B) edge[bend left] node {\tiny $1$} (A); 
-                \path[->] (U) edge[dashed] node {\tiny $1$} (A);
-                \path[->] (U) edge[dashed] node {\tiny $1$} (B);
-                \path[->] (A) edge[dashed,bend right] node {\tiny $1$} (y);
+                \path[->] (U) edge node {\tiny $1$} (A);
+                \path[->] (U) edge node {\tiny $1$} (B);
+                \path[->] (A) edge[bend right] node {\tiny $1$} (y);
             \end{scope}
@@
 
-            \end{tikzpicture} &
-      \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{osc_impulse}
-   \end{tabular}
+            \end{tikzpicture} %&
+   %   \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{osc_impulse}
+   %\end{tabular}
   %  \end{center}
-      \caption{(Left) Graphical representation of the cell cycle control gene network, and (right) plot of the phenotype $\phi(t)$ against time $t$.} \label{fig:oscillator}
+      \caption{
+        %(Left) 
+        Diagram of the gene network in Example \ref{ex:oscillator}
+      %, and (right) plot of the phenotype $\phi(t)$ against time $t$
+      .} \label{fig:oscillator}
     \end{figure}
@@
 
-Some gene networks, however, can grown or shrink, perhaps following gene duplications and deletions, 
+Some gene networks, however, can grow or shrink, perhaps following gene duplications and deletions, 
 and also still preserve their phenotypes.
@@
 These correspond to distinct genetic networks that perform indistinguishable functions. 
-The following example \ref{ex:all_osc} gives the set of minimal systems equivalent to the oscillator of example \ref{ex:oscillator},
+The following example \ref{ex:all_osc} gives the set of minimal systems equivalent to the oscillator of Example \ref{ex:oscillator},
 that all share common $B$ and $C$ matrices.
+The oscillator can also be equivalently realized by a three-gene (or larger) network, and will have even more evolutionary degrees of freedom available, as in e.g., Figure \ref{fig:3D_osc}.
 
@@
   \node (A) at (0,0) {$\kappa_{1}$};
-    \node[dashed] (B) at (4,0) {$\kappa_{2}$};
-    \node[shape=rectangle] (U) at (2,2) {input $(u)$};
-    \node[shape=rectangle] (y) at (2,-2) {output $(\phi)$};
+    \node (B) at (4,0) {$\kappa_{2}$};
+    \node[shape=rectangle] (U) at (2,2) {input ($u$)};
+    \node[shape=rectangle] (y) at (2,-2) {output ($\phi$)};
 \end{scope}
@@
     \node (C) at (3.464,2) {$\kappa_{3}$};
-    \node[shape=rectangle, rotate=0] (U) at (-2.5,2) {input $(u)$};
-    \node[shape=rectangle] (y) at (3.464,0) {output $(\phi)$};
+    \node[shape=rectangle, rotate=0] (U) at (-2.5,2) {input ($u$)};
+    \node[shape=rectangle] (y) at (3.464,0) {output ($\phi$)};
 \end{scope}
@@
               every edge/.style={draw=black, thick}]
-    \path [->, sloped] (A) edge[bend left] node {\tiny ?} (B);
-    \path [->, sloped] (B) edge[bend left] node {\tiny ?} (A); 
-    \path [->, sloped] (A) edge[bend right] node {\tiny ?} (C); 
-    \path [->, sloped] (B) edge[bend left] node {\tiny ?} (C); 
-    \path [->, sloped] (C) edge[bend right] node {\tiny ?} (A); 
-    \path [->, sloped] (C) edge[bend left] node {\tiny ?} (B); 
+    \path [->, sloped] (A) edge[bend left] node {\tiny $2.2$} (B);
+   % \path [->, sloped] (B) edge[bend left] node {\tiny ?} (A); 
+    \path [->, sloped] (A) edge[bend right] node {\tiny $2$} (C); 
+    %\path [->, sloped] (B) edge[bend right] node {\tiny ?} (C); 
+    \path [->, >=Rectangle, sloped] (C) edge[bend right] node {\tiny $-1$} (A); 
+    \path [->, >=Rectangle, sloped] (C) edge[bend right] node {\tiny $-2.2$} (B); 
     \path[->] (U) edge node {\tiny $1$} (A);
@@
               every edge/.style={draw=black, thick}]
-    \path [->] (A) edge[loop below] node[sloped, anchor=center, below] {\tiny ?} (A);
-    \path [->] (B) edge[loop above] node[sloped, anchor=center, above] {\tiny ?} (B);
-    \path [->] (C) edge[loop right] node[sloped, anchor=center, above] {\tiny ?} (C);
+    \path [->] (A) edge[loop below] node[sloped, anchor=center, below] {\tiny $1$} (A);
+    \path [->] (B) edge[loop above] node[sloped, anchor=center, above] {\tiny $1$} (B);
+    \path [->, >=Rectangle] (C) edge[loop right] node[sloped, anchor=center, above] {\tiny $-1$} (C);
 \end{scope}
@@
 \paragraph{Sexual reproduction and recombination}
-Parents with phenotypically equivalent, yet different, gene networks may produce offspring with dramatically different phenotypes. 
-If the phenotypes are significantly divergent then the offspring may be inviable or otherwise have very low fitnesses, 
-despite both parents being well adapted. If this is consistent for the entire population, we would consider them to be separate species.
-
-Diploid organisms have two copies of the genome, each of which encodes a set of system coefficients.
-We assume that a diploid who has inherited systems $(A', B', C')$ and $(A'', B'', C'')$ from her mother and father, respectively,
+Parents with phenotypically equivalent yet differently wired gene networks may produce offspring with dramatically different phenotypes. 
+If the phenotypes are significantly divergent then the offspring may be inviable or otherwise dysfunctional, 
+despite both parents being well adapted.
+If this is consistent for the entire population, we would consider them to be separate species, in accord with the biological species concept \citep{mayr2000biological}.
+
+Diploid organisms have two genomes, each of which encodes a set of system coefficients.
+We assume that a diploid which has inherited systems $(A', B', C')$ and $(A'', B'', C'')$ from each of its parents,
 has phenotype determined by the system that averages these two,
@@
 
-Each of the copies of the genome an organism inherits from her parents are generated from their two copies of the genome by meiosis,
-in which the diploid parent recombines her two genomes.
+Each genome an organism inherits is generated by meiosis,
+in which both of its diploid parents recombine their two genomes.
 We will assume that each coefficient (i.e., entry of $A$, $B$ or $C$)
@@
 so that each coefficient in the system produced by meiosis is an independent random choice
-of the two parental coefficients.
-With these definitions, an $F_1$ offspring of two individuals
-carries one system copy from each parent,
-and an $F_2$ is the offspring of two independently formed $F_1$ individuals.
+between the two parental coefficients.
+With these definitions, an $F_1$ offspring
+carries a system copy from each parent,
+and an $F_2$ is an offspring of two independently formed $F_1$s.
 If the parents are from distinct populations,
-these are first- and second-generation hybrids, respectively.
+these are simply first-- and second--generation hybrids, respectively.
 
@@
 
-The offspring of two systems with the same phenotype may not have the same phenotype as the parents -- in other words $\allS$, the set of all systems phenotypically equivalent to a given one, is not, in general, closed under averaging or recombination.
+Offspring formed from two phenotypically identical systems do not necessarily exhibit the same phenotype as both of its parents -- in other words $\allS$, the set of all systems phenotypically equivalent to a given one, is not, in general, closed under averaging or recombination.
 Next we discuss how this fact can contribute to hybrid incompatibility and genetic load.
@@
 if the parents have different underlying systems.
-This leads to the question: 
-How quickly does the hybrid phenotype break down with increasing distance between the parents?
+%This leads to the question: 
+How quickly do hybrid phenotypes break down as genetic distance between parents increases?
 To quantify this, we will measure how far a system's phenotype is from optimal
@@
 However, the coefficients of these two parental systems differ substantially,
-probably more than would be realistically observed between diverging populations.
+probably more than would be observed between diverging populations.
 In figure \ref{fig:hybs} we compare the phenotypes for $F_1$ and $F_2$ hybrids between more similar parents,
 and see increasingly divergent phenotypes as the difference between the parental systems increases.
-(In this example, the coefficients of $A(2+\epsilon)$ differ from those of $A(2)$ by an average factor of $1+\epsilon/2$;
+(In this example, the coefficients of $A(\epsilon)$ differ from those of $A(0)$ by an average factor of $1+\epsilon/2$;
 such small differences could plausibly be caused by changes to promoter sequences.)
 This divergence is quantified in Figure \ref{fig:osc_incompat},
-which shows that mean distance to optimum phenotype of the $F_1$ and $F_2$ hybrid offspring between $A(2)$ and $A(2+\epsilon)$
+which shows that mean distance to optimum phenotype of the $F_1$ and $F_2$ hybrid offspring between $A(0)$ and $A(\epsilon)$
 increases with $\epsilon^2$ and $\epsilon$, respectively.
@@
 The coefficients in $A$ -- i.e., the regulatory coefficients --
-differ between parents by only a few percent (around 0.5\% for $\tau=2.01$ and 5\% for $\tau=2.1$).
+  differ between parents by only a few percent (around 0.5\% for $\epsilon = - \sfrac{1}{100}$ and 5\% for $\epsilon= - \sfrac{1}{10}$).
 This is well within the amount of regulatory coefficient variation we expect to find segregating within real populations
@@
   consistent with the idea that natural selection will allow such intrapopulation variation.
+For larger values of $\left| \epsilon \right|$ (here $\sfrac{1}{2}$), hybrid phenotypes can become dysfunctional, and depending on context, manifest as hybrid inviability.
 
@@
   \begin{tabular}{ccc}
-    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{F1_comparison} &
-    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{F2s_comparison2}
-  \end{tabular}
-  \caption{
-  \textbf{(left)} Phenotypes of $F_1$ hybrids between an $A(2)$ parent and, top-to-bottom, an $A(2.01)$, an $A(2.1)$, and $A(2.5)$ parent.
-    Parental phenotypes ($\sin t + \cos t$) are shown in blue, and hybrid phenotypes in orange.
-    \textbf{(right)} Phenotypes of $F_2$ hybrids between the same set of parents,
-    with parental phenotype again in blue.
-    Different lines correspond to different $F_2$s produced by recombination;
-    note that some completely fail to oscillate.
-    \plri{make axis labels bigger}
-  } \label{fig:hybs}
-\end{figure}
-
-  \begin{figure}[H]
-    \centering
-    \begin{tabular}{ccc}
+    %\includegraphics[width=0.5\textwidth, height=0.25\paperheight]{F1_comparison} &
+    %\includegraphics[width=0.5\textwidth, height=0.25\paperheight]{F2s_comparison2}
     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F1_hundreth_tau0} &
@@
   \end{tabular}
+  \caption{
+    \textbf{(left)} Phenotypes of $F_1$ hybrids between an $A(0)$ parent and, top-to-bottom, an $A(-\sfrac{1}{100})$, an $A(-\sfrac{1}{10})$, and $A(-\sfrac{1}{2})$ parent.
+    Parental phenotypes ($\sin t + \cos t$) are shown in solid black, and hybrid phenotypes in dashed blue.
+    \textbf{(right)} Phenotypes of all 256 possible $F_2$ hybrids between the same set of parents,
+    with parental phenotype again in black.
+    Different colored lines correspond to different $F_2$s;
+    note that some completely fail to oscillate.
+  } \label{fig:hybs}
 \end{figure}
@@
     \centering
-    \begin{tabular}{cc}
-    \includegraphics[width=0.4\textwidth]{figures/f1_quartic2} &
-    \includegraphics[width=0.4\textwidth]{figures/f2_quad2}
-    \end{tabular}
-    \caption{
-    \textbf{(left)} Phenotype distance to optimum, $D$, using $\rho(t) = \exp(-t/4\pi)$
-    for $F_1$ hybrids between an $A(2)$ and an $A(2+\epsilon)$ parent.
-    \textbf{(right)} Same, but for $F_2$ offspring (one line per type of $F_2$).
-    \plri{make this $D$ not $D^2$ and say which line is the overall average $F_2$.}
-    } \label{fig:osc_incompat}
+    %\begin{tabular}{cc}
+    %\includegraphics[width=0.4\textwidth]{figures/f1_quartic2} &
+    %\includegraphics[width=0.4\textwidth]{figures/f2_quad2}
+    %\end{tabular}
+    \includegraphics[width=0.5\textwidth]{examples/F2_vs_F1_divergence_tau0}
+   % \caption{
+    %\textbf{(left)} 
+    \caption{Phenotype distance to optimum, $D$, using $\rho(t) = \exp(-t/4\pi)$
+    for $F_1$ (red) and $F_2$ (blue) hybrids between an $A(0)$ and an $A(\epsilon)$ parent. Genetic distance is computed as the average coefficient difference between two systems.}
+    %\textbf{(right)} Same, but for $F_2$ offspring (one line per type of $F_2$).
+    %\plri{make this $D$ not $D^2$ and say which line is the overall average $F_2$.
+    %}
+    \label{fig:osc_incompat}
   \end{figure}
@@
 \paragraph{Haldane's rule}
-\jss{Maybe something like dosage compensation drift could also lead to Haldane's rule.
-That is, if the dosage compensation can either be: upregulate the X by $2$ in male flies, versus in another species: downregulate the X in females by $1/2$, and all possible combinations inbetween?}
 This model naturally predicts
@@
 For example, consider an XY species with a two-gene network where the first gene resides on an autosome and the second gene on the X chromosome.
-A male whose pair of haplotypes are 
+A male whose pair of haplotypes is 
 $\left( \left[ \begin{smallmatrix} A_1 & A_2 \\ \cdot & \cdot \end{smallmatrix} \right], \left[ \begin{smallmatrix} A_1 & A_2 \\ X_1 & X_2 \end{smallmatrix} \right] \right)$
@@
 $A = \left[\begin{smallmatrix} \bar A_1 & \bar A_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$.
-An $F_1$ male offspring of these two will have phenotype determined by
+An $F_1$ male offspring of these two will have its phenotype determined by
 $\left[\begin{smallmatrix} (A_1 + \bar A_1)/2 & (A_2 + \bar A_2)/2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$.
@@
 $F_1$ males are effectively equivalent to purely autosomal-system $F_2$s.
-
-This mechanism is apparently distinct from the currently favored ``dominance'', ``faster-X'', and ``faster-male'' theories \citep{orr1997haldane, coyne1998evolutionary} to explain Haldane's rule,
-and makes several different predictions than the dominance theory as formulated by \citet{turelli1995dominance}.
-\plr{what are they?}
-\citet{turelli1995dominance} employ a population genetic model which strictly requires alleles causing hybrid incompatibility on the X chromosome not to exhibit additive effects, that is the \emph{dominance coefficient} (typically $h$ in classic population genetics) must not be equal to $\frac{1}{2}$.
+As such, in this setting, Haldane's rule is mechanistically similar to \emph{hybrid breakdown}, the observation that $F_2$s will often be less fit than $F_1$ hybrid crosses, in the early stages of speciation.
+This mechanism appears to be distinct from the ``dominance'', ``faster-X'', and ``faster-male'' theories \citep{orr1997haldane, coyne1998evolutionary, turelli1995dominance} usually suggested to explain Haldane's rule.
+%and makes several different predictions than the dominance theory as formulated by \citet{turelli1995dominance}.
+%\plr{what are they?}
+%\citet{turelli1995dominance} employ a population genetic model which strictly requires alleles causing hybrid incompatibility on the X chromosome not to exhibit additive effects, that is the \emph{dominance coefficient} (typically $h$ in classic population genetics) must not be equal to $\frac{1}{2}$.
 In their model, if alleles are strictly additive, that is neither dominant nor recessive, then male and female $F_1$ hybrids will drop in fitness at the exact same rates, and therefore defy Haldane's rule.
@@
 
-We will suppose that each organism's phenotype is determined by her vector of coefficients, denoted by $x=(x_1, x_2, \ldots, x_L)$,
-and that the corresponding fitness is determined by the distance of her phenotype to optimum.
+We will suppose that each organism's phenotype is determined by its vector of coefficients, denoted by $x=(x_1, x_2, \ldots, x_L)$,
+and that the corresponding fitness is determined by the distance of its phenotype to optimum.
 The optimum phenotype is unique, but is realized by many distinct $x$ -- those falling in the ``optimal set'' $\allS$.
@@
 and that we can locally approximate the optimal set $\allS$ as a quadratic surface.
-As above, an individual's coefficients are given by averaging her parentally inherited coefficients and adding random noise due to segregation. 
+As above, an individual's coefficients are given by averaging its parentally inherited coefficients and adding random noise due to segregation. 
 Concretely, we use the \emph{infinitesimal model} for reproduction \citep{barton_infinitesimal} --
@@
 
-\section*{Examples}
-\begin{example}[Metabolic network] \label{ex:metabolic}
-Consider an organism that can metabolize two different sugars (present at logarithmic concentrations $u_{1}$ and $u_{2}$), 
-with two enzymes (at log concentrations $\phi_{1}$ and $\phi_{2}$).
-Further suppose that the second sugar is prefered,
-%and that the organism has a regulatory network that can deploy situation specific metabolic strategy.
-and that
-depending on $u_{1}$ and $u_{2}$ the organism will synthesize an appropriate $\phi_{1}$ and $\phi_{2}$. 
-Furthermore, assume this system contains at least two transcription factors, whose log concentrations are $\kappa_{1}$, $\kappa_{2}$, $\dots \kappa_{n}$.
-Minimally such a system may have the architecture,
-%\begin{align*}
-%    \mathcal{S}_{\min} = \left\{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 0 & -1 \\ 0 & 0 \end{bmatrix} \begin{bmatrix} \kappa_{1} \\ \kappa_{2} \end{bmatrix} (t) + \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \begin{bmatrix} u_{1} \\ u_{2} \end{bmatrix} (t) \\[11pt]
-%    \phi(t) &= \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \begin{bmatrix} \kappa_{1} \\ \kappa_{2} \end{bmatrix} (t) \end{array} \right.
-%\end{align*}
-  $\Sys_{\min}(U) = (U \left[ \begin{smallmatrix} 0 & -1 \\ 0 & 0 \end{smallmatrix} \right] U^{-1}, U 
-    %\left[\begin{smallmatrix} 1 & 0 \\ 0 & 1 \end{smallmatrix}\right]
-      ,
-     % \left[\begin{smallmatrix} 1 & 0 \\ 0 & 1 \end{smallmatrix}\right]
-        U^{-1})$. 
-%
-%The impulse response of the system is,
-%$h(t) = \left[\begin{smallmatrix} 1 & - t \\ 0 & 1 \end{smallmatrix}\right]$.
-    We can find alternative equivalent systems by changning coordinates ($U \rightarrow U'$) or, more generally, by applying the Kalman decomposition \ref{apx:kalman}.
+%\section*{Examples}
+%\begin{example}[Metabolic network] \label{ex:metabolic}
+%Consider an organism that can metabolize two different sugars (present at logarithmic concentrations $u_{1}$ and $u_{2}$), 
+%with two enzymes (at log concentrations $\phi_{1}$ and $\phi_{2}$).
+%Further suppose that the second sugar is prefered,
+%%and that the organism has a regulatory network that can deploy situation specific metabolic strategy.
+%and that
+%depending on $u_{1}$ and $u_{2}$ the organism will synthesize an appropriate $\phi_{1}$ and $\phi_{2}$. 
+%Furthermore, assume this system contains at least two transcription factors, whose log concentrations are $\kappa_{1}$, $\kappa_{2}$, $\dots \kappa_{n}$.
+%Minimally such a system may have the architecture,
+%%\begin{align*}
+%%    \mathcal{S}_{\min} = \left\{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 0 & -1 \\ 0 & 0 \end{bmatrix} \begin{bmatrix} \kappa_{1} \\ \kappa_{2} \end{bmatrix} (t) + \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \begin{bmatrix} u_{1} \\ u_{2} \end{bmatrix} (t) \\[11pt]
+%%    \phi(t) &= \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \begin{bmatrix} \kappa_{1} \\ \kappa_{2} \end{bmatrix} (t) \end{array} \right.
+%%\end{align*}
+%  $\Sys_{\min}(U) = (U \left[ \begin{smallmatrix} 0 & -1 \\ 0 & 0 \end{smallmatrix} \right] U^{-1}, U 
+%    %\left[\begin{smallmatrix} 1 & 0 \\ 0 & 1 \end{smallmatrix}\right]
+%      ,
+%     % \left[\begin{smallmatrix} 1 & 0 \\ 0 & 1 \end{smallmatrix}\right]
+%        U^{-1})$. 
+%%
+%%The impulse response of the system is,
+%%$h(t) = \left[\begin{smallmatrix} 1 & - t \\ 0 & 1 \end{smallmatrix}\right]$.
+%    We can find alternative equivalent systems by changning coordinates ($U \rightarrow U'$) or, more generally, by applying the Kalman decomposition \ref{apx:kalman}.
 To illustrate that phenotypic invariance does not require dimensional invariance, we apply the Kalman decomposition to $\Sys_{\min}$ to find
-    %This may happen if a system is co-opted from another, or may be a consequence of gene duplication and deletion. 
-  $\Sys(\small D_{1-3},V) \! =  \! \left( \! \begin{array}{ll}
-     V \!
+%    %This may happen if a system is co-opted from another, or may be a consequence of gene duplication and deletion. 
+%  $\Sys(\small D_{1-3},V) \! =  \! \left( \! \begin{array}{ll}
+%     V \!
 \left[\begin{smallmatrix} D_{1} & D_{2} \\ 0 & A \end{smallmatrix}\right] \!
@@
 \right)
-$, both of which are in $\allS$
+%$, both of which are in $\allS$
+%%  \begin{align*}
+%%    \Sys(X_{1-3},V) = \left( \begin{array}{ll}
+%%     V \left[\begin{smallmatrix} X_{1} & X_{2} \\ 0 & A_{\min} \end{smallmatrix}\right] V^{-1}, V \left[\begin{smallmatrix} X_{3} \\ B_{\min} \end{smallmatrix}\right], \left[\begin{smallmatrix} 0 & C_{\min} \end{smallmatrix}\right] V^{-1} \end{array} \right)
+%%    \end{align*}
+%  ($V$ can be any $4$-dimensional and invertible matrix, $D_{1-3} \in \R^{2 \times 2}$, and $(A,B,C) \in \Sys_{\min}$).
+%%So for example, some system can be wired as follows, and still be input-output equivalent to the minimal metabolic system $\mathcal{S}_{\min}$:
+%% \begin{align*}
+%%   A' &= \begin{bmatrix}
+%%     1.6923  &  1.5385 &  -2.6154 &  -2 \\
+%%     0.8462  &  0.7692 &  -1.3077 &  -1 \\
+%%     1.2692  &  1.1538 &  -1.9615 &  -1.5 \\
+%%     0.4231  &  0.3846 &  -0.6538 &  -0.5
+%%   \end{bmatrix} \\
+%%   B' &= \begin{bmatrix} 4 & 4 \\ 2 & 2 \\ 3 & 3 \\ 1 & 3 \end{bmatrix} \\
+%%     C' &= \begin{bmatrix}  0.2692  &  0.1538  &  0.0385 &  -0.5 \\
+%%     -0.4231 &  -0.3846 &   0.6538  &  0.5 \end{bmatrix}
+%% \end{align*}
+%%
+%%Despite the present example consisting of a minimal $2 \times 2$ and a non-minimal $4 \times 4$ system, 
+%%any $m$-dimensional system can be constructed using this method 
+%%-- applying a change of coordinates to the Kalman decomposition 
+%%-- to construct a mechanistically different system with identical phenotypic dynamics. 
+%%
+%%When applying the Kalman decomposition to real biological networks, one may have to restrict the free parameters, as physiology and biochemistry might not allow all mathematically possible coefficients.
+We note that as there are so many degrees of freedom, it may be difficult to infer system function solely from structure.
+%\end{example}
+
+\appendix
+
+%\section{Kalman Decomposition} \label{apx:kalman}
+%\begin{definition}[Phenotypic equivalence of systems]
+%    Let $(\kappa(t),\phi(t))$ and $(\bar \kappa(t),\bar \phi(t))$ be the solutions to \eqref{eqn:lti_system}
+%    with coefficient matrices $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ respectively,
+%    and both $\kappa(0)$ and $\bar \kappa(0)$ are zero. 
+%    The systems defined by $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ are
+%    \textbf{phenotypically equivalent} 
+%    if
 %    \begin{align*}
-%    \Sys(X_{1-3},V) = \left( \begin{array}{ll}
-%     V \left[\begin{smallmatrix} X_{1} & X_{2} \\ 0 & A_{\min} \end{smallmatrix}\right] V^{-1}, V \left[\begin{smallmatrix} X_{3} \\ B_{\min} \end{smallmatrix}\right], \left[\begin{smallmatrix} 0 & C_{\min} \end{smallmatrix}\right] V^{-1} \end{array} \right)
+%        \phi(t) = \bar \phi(t) \qquad \text{for all} \; t \ge 0.
 %    \end{align*}
-  ($V$ can be any $4$-dimensional and invertible matrix, $D_{1-3} \in \R^{2 \times 2}$, and $(A,B,C) \in \Sys_{\min}$).
-%So for example, some system can be wired as follows, and still be input-output equivalent to the minimal metabolic system $\mathcal{S}_{\min}$:
+%    Equivalently, this occurs if and only if
 %    \begin{align*}
-%   A' &= \begin{bmatrix}
-%     1.6923  &  1.5385 &  -2.6154 &  -2 \\
-%     0.8462  &  0.7692 &  -1.3077 &  -1 \\
-%     1.2692  &  1.1538 &  -1.9615 &  -1.5 \\
-%     0.4231  &  0.3846 &  -0.6538 &  -0.5
-%   \end{bmatrix} \\
-%   B' &= \begin{bmatrix} 4 & 4 \\ 2 & 2 \\ 3 & 3 \\ 1 & 3 \end{bmatrix} \\
-%     C' &= \begin{bmatrix}  0.2692  &  0.1538  &  0.0385 &  -0.5 \\
-%     -0.4231 &  -0.3846 &   0.6538  &  0.5 \end{bmatrix}
+%        h(t) = \bar h(t)  \qquad \text{for all} \; t \ge 0,
 %    \end{align*}
-%
-%Despite the present example consisting of a minimal $2 \times 2$ and a non-minimal $4 \times 4$ system, 
-%any $m$-dimensional system can be constructed using this method 
-%-- applying a change of coordinates to the Kalman decomposition 
-%-- to construct a mechanistically different system with identical phenotypic dynamics. 
-%
-%When applying the Kalman decomposition to real biological networks, one may have to restrict the free parameters, as physiology and biochemistry might not allow all mathematically possible coefficients.
-We note that as there are so many degrees of freedom, it may be difficult to infer system function solely from structure.
-\end{example}
-
-\appendix
-
-\section{Kalman Decomposition} \label{apx:kalman}
-\begin{definition}[Phenotypic equivalence of systems]
-    Let $(\kappa(t),\phi(t))$ and $(\bar \kappa(t),\bar \phi(t))$ be the solutions to \eqref{eqn:lti_system}
-    with coefficient matrices $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ respectively,
-    and both $\kappa(0)$ and $\bar \kappa(0)$ are zero. 
-    The systems defined by $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ are
-    \textbf{phenotypically equivalent} 
-    if
-    \begin{align*}
-        \phi(t) = \bar \phi(t) \qquad \text{for all} \; t \ge 0.
-    \end{align*}
-    Equivalently, this occurs if and only if
-    \begin{align*}
-        h(t) = \bar h(t)  \qquad \text{for all} \; t \ge 0,
-    \end{align*}
-    where $h$ and $\bar h$ are the impulse responses of the two systems.
-\end{definition}
-
-One way to find other systems equivalent to a given one
-is by change of coordinates (``algebraic equivalence''):
-if $V$ is an invertible matrix, then the systems $(A,B,C)$ and $(VAV^{-1},VB,CV^{-1})$
-have the same dynamics because their transfer functions are equal:
-\begin{align*}
-    CV^{-1}( zI - VAV^{-1})^{-1}VB
-    =
-    CV^{-1}V( zI - A)^{-1}V^{-1}VB
-    =
-    C( zI - A)^{-1}B .
-\end{align*}
-However, the converse is not necessarily true: 
-systems can have identical transfer functions without being changes of coordinates of each other.
-In fact, systems with identical transfer functions can involve interactions between different
-numbers of molecular species.
-
-The set of all systems phenotypically equivalent to a given system $(A,B,C)$ 
-is elegantly described using the Kalman decomposition,
-which also clarifies the system dynamics? tells us a lot about how it works? \plr{or something}
-To motivate this, first note that the input $u(t)$ only directly pushes the system
-in directions lying in the span of the columns of $B$.
-As a result, different combinations of input can 
-move the system in any direction that lies in the \emph{reachable subspace},
-which we denote by $\reachable$,
-and is defined to be the closure of $\spn(B)$ under applying $A$
-(or equivalently, the span of $B, AB, A^2B, \ldots A^{n-1}B$).
-Analogously to this, we define
-the \emph{observable subspace}, $\mathcal{O}$,
-to be the closure of $\spn(C^T)$ under applying $A$.
-(Or: $\unobservable$ is the largest $A$-invariant subspace
-contained in the null space of $C$;
-and $\reachable$ is the largest $A$-invariant subspace contained in the image of $B$.)
-
-If we define
-\begin{enumerate}
-    \item The columns of $P_\rno$ are an orthonormal basis for $\reachable \cap \unobservable$.
-    \item The columns of $P_\ro$ are an orthonormal basis of
-        the complement of $\reachable \cap \unobservable$ in $\reachable$.
-    \item The columns of $P_\nro$ are an orthonormal basis of
-        the complement of $\reachable \cap \unobservable$ in $\unobservable$.
-    \item The columns of $P_\nrno$ are an orthonormal basis of
-        the remainder of $\R^n$.
-\end{enumerate}
-If we then define
-\begin{align*}
-    P &= 
-    \left[ \begin{array}{c|c|c|c}
-        P_\rno & P_\ro & P_\nro & P_\nrno
-    \end{array} \right] ,
-\end{align*}
-then
-\begin{align*}
-    P^T P
-    &=
-    \left[ \begin{array}{c|c|c|c}
-        I & 0 & 0 & 0 \\
-        \hline
-        0 & I & U & 0 \\
-        \hline
-        0 & V & I & 0 \\
-        \hline
-        0 & 0 & 0 & I 
-    \end{array} \right] .
-\end{align*}
-\plr{Check this.  Can we get $U=V=0$?}
-
-The following theorem can be found in SOME REFERENCE.
-
-\begin{theorem}[Kalman decomposition] \label{thm:kalman}
-        For any system $(A,B,C)$ with corresponding Kalman basis matrix $P$,
-        the transformed system $(PAP^{-1},PB,CP^{-1})$  has the following form:
-        \begin{align*}
-            \widehat A = PAP^{-1}
-            &=
-            \left[ \begin{array}{cccc}
-                A_{\rno} & A_{\rno,\ro} & A_{\rno,\nrno} & A_{\rno,\nro} \\
-                0 & A_{\ro} & 0 & A_{\ro,\nro} \\
-                0 & 0 & A_{\nrno} & A_{\nrno,\nro} \\
-                0 & 0 & 0 & A_{\nro}
-            \end{array} \right] ,
-        \end{align*}
-        and
-        \begin{align*}
-            \widehat B = PB
-            &=
-            \left[ \begin{array}{cccc}
-                B_{\rno} \\
-                B_{\ro} \\
-                0 \\
-                0 
-            \end{array} \right] ,
-        \end{align*}
-        and
-        \begin{align*}
-            \widehat C = CP^{-1}
-            &=
-            \left[ \begin{array}{cccc}
-                0 & C_{\ro} & C_{\nrno} & 0 
-            \end{array} \right] .
-        \end{align*}
-        The transfer function of both systems is given by
-        \begin{align*}
-            H(z) = C_{\ro} ( zI - A_{\ro} )^{-1} B_{\ro} .
-        \end{align*}
-\end{theorem}
-
-In the latter case, we say that the system is \emph{minimal} 
--- there is no equivalent system with a smaller number of species.
-Note that this says that any two equivalent minimal systems
-are changes of basis of each other.
-
-Since any system can be put into this form,
-and once in this form, its transfer function is determined only by 
-$C_{\ro}$, $A_{\ro}$, and $B_\ro$,
-therefore, the set of all equivalent systems are parameterized by
-the dimension $n$,
-the choice of basis ($P$),
-the remaining submatrices in $\widehat A$, $\widehat B$, and $\widehat C$
-(which are unconstrained),
-and a invertible transformation of $\spn(P_{\ro})$, which we call $T_\ro$.
-
-\begin{theorem}[Parameterization of equivalent systems]
-    Let $(A,B,C)$ be a minimal system.
-    \begin{itemize}
-        \item[(a)]
-            Every equivalent system is of the form given in Theorem \ref{thm:kalman},
-            i.e., can be specified by choosing a dimension, $n$;
-            submatrices in $\widehat A$, $\widehat B$, and $\widehat C$ 
-            except for $A_\ro=A$, $B_\ro=B$, and $C_\ro=C$;
-            and choosing an invertible matrix $P$.
-
-        \item[(b)]
-            \plr{conjecture:}
-            The parameterization is unique
-            if $P$ is furthermore chosen so that 
-            each $P_x$ other than $P_\ro$ is a projection matrix,
-            and that 
-            \begin{align*}
-                0
-                =
-                P_x^T P_y
-            \end{align*}
-            for all $(x,y)$ except $(\ro,\nrno)$.
-
-    \end{itemize} 
-\end{theorem}
-
-\plr{Another way of saying it: pick the $\reachable$ and $\unobservable$ subspaces,
-that must intersect in something of the minimal dimension;
-then let $P$ be the appropriate basis?}
-
-In some situations we may be interested in only ``network rewiring'',
-where $A$ changes while $B$ and $C$ do not.
-For instance, 
-if all non-regulatory functions of each molecule are strongly constrained,
-then $C$ cannot change.
-Likewise, if responses of each molecule to the external inputs are not changed by evolution,
-then $B$ does not change.
-
-\subsection{Neutral directions from the Kalman decomposition}
-
-The Kalman decomposition above says that any system $(A,B,C)$ can be decomposed into
-$(P, \hat A, \hat B, \hat C)$ so that
-$$\begin{aligned}
-    A &= P^{-1} \hat A P  \\
-    B &= P^{-1} \hat B  \\
-    C &= \hat C P ,
-\end{aligned}$$
-and we know precisely how we can change these to preserve the transfer function:
-\begin{enumerate}
-    \item $P \to P + \epsilon Q$ as long as the result is still invertible,
-    \item $\hat A \to A + \epsilon X$ as long as $X$ is zero in the correct places,
-    \item $\hat B \to B + \epsilon Y$ as long as $Y$ is zero in the correct places,
-    \item $\hat C \to C + \epsilon Y$ as long as $Z$ is zero in the correct places.
-\end{enumerate}
-By taking $\epsilon \to 0$, these tell us the local directions we can move a system $(A,B,C)$ in.
-All statements below are up to first order in $\epsilon$,
-omitting terms of order $\epsilon^2$.
-
-First, since $(P + \epsilon Q)^{-1} = P^{-1} + \epsilon P^{-1} Q P^{-1}$,
-modifying $P \to P + \epsilon Q$ changes
-$$\begin{aligned}
-    A 
-        &\to A + \epsilon P^{-1} \hat A Q - \epsilon P^{-1} Q P^{-1} \hat A P \\
-        &= A + \epsilon \left(A P^{-1} Q - P^{-1} Q A\right) , \\
-    B
-        &\to B - \epsilon P^{-1} Q B \\
-    C
-        &\to C + \epsilon C P^{-1} Q .
-\end{aligned}$$
-Since $P$ is invertible and $Q$ can be anything (if $\epsilon$ is small enough),
-this allows changes in the direction of an arbitrary $W$:
-$$\begin{aligned}
-    A 
-        &= A + \epsilon \left(A W - W A\right) , \\
-    B
-        &\to B - \epsilon W B \\
-    C
-        &\to C + \epsilon C W .
-\end{aligned}$$
-
-Then, $\hat A \to A + \epsilon X$  does
-$$\begin{aligned}
-    A \to A + \epsilon P^{-1} X P 
-\end{aligned}$$
-and $\hat B \to B + \epsilon Y$ does
-$$\begin{aligned}
-    B \to B + \epsilon P^{-1} Y
-\end{aligned}$$
-and $\hat C \to C + \epsilon Z$ does
-$$\begin{aligned}
-    C \to C + \epsilon Z P .
-\end{aligned}$$
-These degrees of freedom look like they depend on $P$, 
-which is not unique,
-but for any two choices of $P$ there are corresponding choices of $X$
-that give the same actual change in $A$ (and likewise for $Y$ and $Z$).
-
-Therefore, this gives us an upper bound on the number of degrees of freedom,
-in terms of the dimensions of the blocks in the Kalman decomposition ($n_\ro$ etc)
-and the dimensions of $B$ and $C$ ($n_B$ and $n_C$ respectively):
-namely, for $W$, $X$, $Y$, and $Z$ respectively:
-$$\begin{aligned}
-    n^2 
-    + (n_{\rno} + n_\ro n_\nro + n_\nrno(n_\nrno + n_\nro) + n_\nro^2)
-    + n_B n_\rno
-    + n_C n_\nrno .
-\end{aligned}$$
-However, some of these may be redundant.
-For instance, changing $P$ in the direction of 
-a $Q$ that satisfies both $A P^{-1} Q = P^{-1} Q A$ and $C P^{-1} Q = 0$
-is equivalent to changing $B$ by $Y = QB$.
-
-\section{Meiotic recombination in linear systems}\label{apx:recombination}
-
-  Recombination is performed by taking two analogous system components from $\Sys$ and $\Sys'$ and randomly swapping rows or columns.
-  \emph{E.g.} gamete systems $(A'', B'', C'')$, produced by the diploid $\{\Sys, \Sys'\}$, are formed by recombining (randomly swapping rows or columns) between two, possibly distinct, systems $\Sys = (A,B,C)$ and $\Sys' = (A', B', C')$ such that,
-  \begin{align*}
-    \Sys'' = \left( \begin{array}{ll}
-    A'' &= MA + (I-M)A' , \\
-    B'' &= MB + (I-M)B' , \\
-    C'' &= CM + C'(I-M)
-    \end{array} \right)
-  \end{align*}
-  where $M$ is a diagonal matrix where each diagonal element is a Bernoulli random variable ($M_{ii} = 0$ or $1$ with equal probability, and $M_{ij}=0$ if $i \neq j$).
+%    where $h$ and $\bar h$ are the impulse responses of the two systems.
+%\end{definition}
+%
+%One way to find other systems equivalent to a given one
+%is by change of coordinates (``algebraic equivalence''):
+%if $V$ is an invertible matrix, then the systems $(A,B,C)$ and $(VAV^{-1},VB,CV^{-1})$
+%have the same dynamics because their transfer functions are equal:
+%\begin{align*}
+%    CV^{-1}( zI - VAV^{-1})^{-1}VB
+%    =
+%    CV^{-1}V( zI - A)^{-1}V^{-1}VB
+%    =
+%    C( zI - A)^{-1}B .
+%\end{align*}
+%However, the converse is not necessarily true: 
+%systems can have identical transfer functions without being changes of coordinates of each other.
+%In fact, systems with identical transfer functions can involve interactions between different
+%numbers of molecular species.
+%
+%The set of all systems phenotypically equivalent to a given system $(A,B,C)$ 
+%is elegantly described using the Kalman decomposition,
+%which also clarifies the system dynamics? tells us a lot about how it works? \plr{or something}
+%To motivate this, first note that the input $u(t)$ only directly pushes the system
+%in directions lying in the span of the columns of $B$.
+%As a result, different combinations of input can 
+%move the system in any direction that lies in the \emph{reachable subspace},
+%which we denote by $\reachable$,
+%and is defined to be the closure of $\spn(B)$ under applying $A$
+%(or equivalently, the span of $B, AB, A^2B, \ldots A^{n-1}B$).
+%Analogously to this, we define
+%the \emph{observable subspace}, $\mathcal{O}$,
+%to be the closure of $\spn(C^T)$ under applying $A$.
+%(Or: $\unobservable$ is the largest $A$-invariant subspace
+%contained in the null space of $C$;
+%and $\reachable$ is the largest $A$-invariant subspace contained in the image of $B$.)
+%
+%If we define
+%\begin{enumerate}
+%    \item The columns of $P_\rno$ are an orthonormal basis for $\reachable \cap \unobservable$.
+%    \item The columns of $P_\ro$ are an orthonormal basis of
+%        the complement of $\reachable \cap \unobservable$ in $\reachable$.
+%    \item The columns of $P_\nro$ are an orthonormal basis of
+%        the complement of $\reachable \cap \unobservable$ in $\unobservable$.
+%    \item The columns of $P_\nrno$ are an orthonormal basis of
+%        the remainder of $\R^n$.
+%\end{enumerate}
+%If we then define
+%\begin{align*}
+%    P &= 
+%    \left[ \begin{array}{c|c|c|c}
+%        P_\rno & P_\ro & P_\nro & P_\nrno
+%    \end{array} \right] ,
+%\end{align*}
+%then
+%\begin{align*}
+%    P^T P
+%    &=
+%    \left[ \begin{array}{c|c|c|c}
+%        I & 0 & 0 & 0 \\
+%        \hline
+%        0 & I & U & 0 \\
+%        \hline
+%        0 & V & I & 0 \\
+%        \hline
+%        0 & 0 & 0 & I 
+%    \end{array} \right] .
+%\end{align*}
+%\plr{Check this.  Can we get $U=V=0$?}
+%
+%The following theorem can be found in SOME REFERENCE.
+%
+%\begin{theorem}[Kalman decomposition] \label{thm:kalman}
+%        For any system $(A,B,C)$ with corresponding Kalman basis matrix $P$,
+%        the transformed system $(PAP^{-1},PB,CP^{-1})$  has the following form:
+%        \begin{align*}
+%            \widehat A = PAP^{-1}
+%            &=
+%            \left[ \begin{array}{cccc}
+%                A_{\rno} & A_{\rno,\ro} & A_{\rno,\nrno} & A_{\rno,\nro} \\
+%                0 & A_{\ro} & 0 & A_{\ro,\nro} \\
+%                0 & 0 & A_{\nrno} & A_{\nrno,\nro} \\
+%                0 & 0 & 0 & A_{\nro}
+%            \end{array} \right] ,
+%        \end{align*}
+%        and
+%        \begin{align*}
+%            \widehat B = PB
+%            &=
+%            \left[ \begin{array}{cccc}
+%                B_{\rno} \\
+%                B_{\ro} \\
+%                0 \\
+%                0 
+%            \end{array} \right] ,
+%        \end{align*}
+%        and
+%        \begin{align*}
+%            \widehat C = CP^{-1}
+%            &=
+%            \left[ \begin{array}{cccc}
+%                0 & C_{\ro} & C_{\nrno} & 0 
+%            \end{array} \right] .
+%        \end{align*}
+%        The transfer function of both systems is given by
+%        \begin{align*}
+%            H(z) = C_{\ro} ( zI - A_{\ro} )^{-1} B_{\ro} .
+%        \end{align*}
+%\end{theorem}
+%
+%In the latter case, we say that the system is \emph{minimal} 
+%-- there is no equivalent system with a smaller number of species.
+%Note that this says that any two equivalent minimal systems
+%are changes of basis of each other.
+%
+%Since any system can be put into this form,
+%and once in this form, its transfer function is determined only by 
+%$C_{\ro}$, $A_{\ro}$, and $B_\ro$,
+%therefore, the set of all equivalent systems are parameterized by
+%the dimension $n$,
+%the choice of basis ($P$),
+%the remaining submatrices in $\widehat A$, $\widehat B$, and $\widehat C$
+%(which are unconstrained),
+%and a invertible transformation of $\spn(P_{\ro})$, which we call $T_\ro$.
+%
+%\begin{theorem}[Parameterization of equivalent systems]
+%    Let $(A,B,C)$ be a minimal system.
+%    \begin{itemize}
+%        \item[(a)]
+%            Every equivalent system is of the form given in Theorem \ref{thm:kalman},
+%            i.e., can be specified by choosing a dimension, $n$;
+%            submatrices in $\widehat A$, $\widehat B$, and $\widehat C$ 
+%            except for $A_\ro=A$, $B_\ro=B$, and $C_\ro=C$;
+%            and choosing an invertible matrix $P$.
+%
+%        \item[(b)]
+%            \plr{conjecture:}
+%            The parameterization is unique
+%            if $P$ is furthermore chosen so that 
+%            each $P_x$ other than $P_\ro$ is a projection matrix,
+%            and that 
+%            \begin{align*}
+%                0
+%                =
+%                P_x^T P_y
+%            \end{align*}
+%            for all $(x,y)$ except $(\ro,\nrno)$.
+%
+%    \end{itemize} 
+%\end{theorem}
+%
+%\plr{Another way of saying it: pick the $\reachable$ and $\unobservable$ subspaces,
+%that must intersect in something of the minimal dimension;
+%then let $P$ be the appropriate basis?}
+%
+%In some situations we may be interested in only ``network rewiring'',
+%where $A$ changes while $B$ and $C$ do not.
+%For instance, 
+%if all non-regulatory functions of each molecule are strongly constrained,
+%then $C$ cannot change.
+%Likewise, if responses of each molecule to the external inputs are not changed by evolution,
+%then $B$ does not change.
+%
+%
+%\subsection{Neutral directions from the Kalman decomposition}
+%
+%The Kalman decomposition above says that any system $(A,B,C)$ can be decomposed into
+%$(P, \hat A, \hat B, \hat C)$ so that
+%$$\begin{aligned}
+%    A &= P^{-1} \hat A P  \\
+%    B &= P^{-1} \hat B  \\
+%    C &= \hat C P ,
+%\end{aligned}$$
+%and we know precisely how we can change these to preserve the transfer function:
+%\begin{enumerate}
+%    \item $P \to P + \epsilon Q$ as long as the result is still invertible,
+%    \item $\hat A \to A + \epsilon X$ as long as $X$ is zero in the correct places,
+%    \item $\hat B \to B + \epsilon Y$ as long as $Y$ is zero in the correct places,
+%    \item $\hat C \to C + \epsilon Y$ as long as $Z$ is zero in the correct places.
+%\end{enumerate}
+%By taking $\epsilon \to 0$, these tell us the local directions we can move a system $(A,B,C)$ in.
+%All statements below are up to first order in $\epsilon$,
+%omitting terms of order $\epsilon^2$.
+%
+%First, since $(P + \epsilon Q)^{-1} = P^{-1} + \epsilon P^{-1} Q P^{-1}$,
+%modifying $P \to P + \epsilon Q$ changes
+%$$\begin{aligned}
+%    A 
+%        &\to A + \epsilon P^{-1} \hat A Q - \epsilon P^{-1} Q P^{-1} \hat A P \\
+%        &= A + \epsilon \left(A P^{-1} Q - P^{-1} Q A\right) , \\
+%    B
+%        &\to B - \epsilon P^{-1} Q B \\
+%    C
+%        &\to C + \epsilon C P^{-1} Q .
+%\end{aligned}$$
+%Since $P$ is invertible and $Q$ can be anything (if $\epsilon$ is small enough),
+%this allows changes in the direction of an arbitrary $W$:
+%$$\begin{aligned}
+%    A 
+%        &= A + \epsilon \left(A W - W A\right) , \\
+%    B
+%        &\to B - \epsilon W B \\
+%    C
+%        &\to C + \epsilon C W .
+%\end{aligned}$$
+%
+%Then, $\hat A \to A + \epsilon X$  does
+%$$\begin{aligned}
+%    A \to A + \epsilon P^{-1} X P 
+%\end{aligned}$$
+%and $\hat B \to B + \epsilon Y$ does
+%$$\begin{aligned}
+%    B \to B + \epsilon P^{-1} Y
+%\end{aligned}$$
+%and $\hat C \to C + \epsilon Z$ does
+%$$\begin{aligned}
+%    C \to C + \epsilon Z P .
+%\end{aligned}$$
+%These degrees of freedom look like they depend on $P$, 
+%which is not unique,
+%but for any two choices of $P$ there are corresponding choices of $X$
+%that give the same actual change in $A$ (and likewise for $Y$ and $Z$).
+%
+%
+%Therefore, this gives us an upper bound on the number of degrees of freedom,
+%in terms of the dimensions of the blocks in the Kalman decomposition ($n_\ro$ etc)
+%and the dimensions of $B$ and $C$ ($n_B$ and $n_C$ respectively):
+%namely, for $W$, $X$, $Y$, and $Z$ respectively:
+%$$\begin{aligned}
+%    n^2 
+%    + (n_{\rno} + n_\ro n_\nro + n_\nrno(n_\nrno + n_\nro) + n_\nro^2)
+%    + n_B n_\rno
+%    + n_C n_\nrno .
+%\end{aligned}$$
+%However, some of these may be redundant.
+%For instance, changing $P$ in the direction of 
+%a $Q$ that satisfies both $A P^{-1} Q = P^{-1} Q A$ and $C P^{-1} Q = 0$
+%is equivalent to changing $B$ by $Y = QB$.
+%
+%\section{Meiotic recombination in linear systems}\label{apx:recombination}
+%
+%  Recombination is performed by taking two analogous system components from $\Sys$ and $\Sys'$ and randomly swapping rows or columns.
+%  \emph{E.g.} gamete systems $(A'', B'', C'')$, produced by the diploid $\{\Sys, \Sys'\}$, are formed by recombining (randomly swapping rows or columns) between two, possibly distinct, systems $\Sys = (A,B,C)$ and $\Sys' = (A', B', C')$ such that,
+%  \begin{align*}
+%    \Sys'' = \left( \begin{array}{ll}
+%    A'' &= MA + (I-M)A' , \\
+%    B'' &= MB + (I-M)B' , \\
+%    C'' &= CM + C'(I-M)
+%    \end{array} \right)
+%  \end{align*}
+%  where $M$ is a diagonal matrix where each diagonal element is a Bernoulli random variable ($M_{ii} = 0$ or $1$ with equal probability, and $M_{ij}=0$ if $i \neq j$).
 If systems are different dimensions the smaller system elements can be augmented with $0$s (\emph{e.g.} $\left[ \begin{smallmatrix} A & 0 \\ 0 & 0 \end{smallmatrix} \right], \left[ \begin{smallmatrix} B \\ 0 \end{smallmatrix} \right], \left[ \begin{smallmatrix} C & 0 \end{smallmatrix}\right]$).
-
+%
+%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2017-12-01 kalman_walk_paper.tex
general edits to paper. Changed figure 6 (f2 vs f1 divergence)

@@
 \citep{jaeger2011gap, davidson2006gene, israel2016comparative}, 
-as well as thoughtful verbal and conceptual models \citep{true2001developmental, gwagner1, weiss2000phenogenetic, edelman2001degeneracy}. 
+as well as thoughtful verbal and conceptual models \citep{true2001developmental, gwagner1, weiss2000phenogenetic, edelman2001degeneracy, pavlicev2012model}. 
 Mathematical models of both particular regulatory networks
@@
 macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level 
-\citep{true2001developmental, tanay2005conservation, tsong2006evolution, hare2008sepsid, vierstra2014mouse, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription}.
+\citep{kimura1981possibility, true2001developmental, tanay2005conservation, tsong2006evolution, hare2008sepsid, vierstra2014mouse, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription}.
  % \jss{should hare2008sepsid be here? It's sequence but not ``system'' drift}
@@
 Finally, in some cases we estimate that reproductive incompatibility due to system drift can manifest on timescales on the order of $N_e$ generations.
-
-analogous to those in the Dobzhansky-Muller model. 
 %Consequentially, two populations isolated for a sufficiently long period of time, 
@@
 hybrid inviability appears as a consequence of recombining different, yet functionally equivalent, mechanisms. 
+While the genetic incompatibilities contributing to hybrid inviability described in this paper share similarities with those in the commonly discussed Dobzhansky-Muller model, they may in fact share more in common with what has been described as the ``pathway model'' \citep{lindtke2015genetic}.
 
@@
 since $F_1$ and $F_2$ phenotypes diverge linearly and at the square root of time relative to parentals, respectively.
-\plr{refer to Turelli here}
+%\plr{refer to Turelli here}
 % In this model, this is due to recombination shuffling system architectures, leaving the $F_2$ system homozygous for one parent at one particular locus and homozygous for the other parent at another possibly incompatible locus.
@@
 
-\paragraph{Nonlinearity}
+\paragraph{Nonlinearity and model assumptions}
 Of course, real regulatory networks are not linear dynamical systems.
@@
 
-\paragraph{Peter: Quant gen discussion}
-Maybe compare to \citet{fierst}?
-
-Do we talk about the $G$ matrix and mutational correlation somewhere?
-
-Note that islands or bottlenecks leave large $\sigma_N$ but small $N_e$ so may go fast.
-
-What is Ne?  In a metapopulation depends on migr rate: classic Turelli paper.
-
-This has implications for genetic load also!!
-
-hominids; neanderthal load.
-
+Finally, despite our model's precise separation of phenotype and kryptotype, this relationship in nature may be far more complicated as aspects of the kryptotype may be less ``hidden'' than we currently assume.
+For instance, attributes excluded from the phenotype as modelled here, ignore the potential energy costs associated with execessively large (nonminimal) kryptotypes, as well as the relationship between a specific network architecture and mutational robustness, as well as many other presently neglected phenomena.
+Despite the model's simplifications and potentially nontrivial oversights, it may still roughly reflect and shed light upon speciation patterns as they occur in nature.
+
+%\paragraph{Peter: Quant gen discussion}
+%Maybe compare to \citet{fierst}?
+%
+%Do we talk about the $G$ matrix and mutational correlation somewhere?
+%
+%Note that islands or bottlenecks leave large $\sigma_N$ but small $N_e$ so may go fast.
+%
+%What is Ne?  In a metapopulation depends on migr rate: classic Turelli paper.
+%
+%This has implications for genetic load also!!
+%
+%hominids; neanderthal load.
+%
 \section*{Acknowledgements}

### 2017-12-01 kalman_walk_paper.tex
general edits to paper. Changed figure 6 (f2 vs f1 divergence)

@@
 \citep{jaeger2011gap, davidson2006gene, israel2016comparative}, 
-as well as thoughtful verbal and conceptual models \citep{true2001developmental, gwagner1, weiss2000phenogenetic, edelman2001degeneracy, pavlicev2012model}. 
+as well as thoughtful verbal and conceptual models \citep{true2001developmental, weiss2000phenogenetic, edelman2001degeneracy, pavlicev2012model}. 
 Mathematical models of both particular regulatory networks

### 2017-12-01 kalman_walk_paper.tex
general edits to paper

@@
 typically implicated in the handful of binding sites for transcription factor $j$ of a typical promoter region,
-and single variants may affect many entries of $\Sys$ simultaneously.
+and single variants may affect many entries of $\allS$ simultaneously.
 % (Recall that although these are best modeled through nonlinear terms,
@@
 %%\plr{Get some data from at least one other species in here!}
-Taken together, this suggests that variation in the entries of $\Sys$
+Taken together, this suggests that variation in the entries of $\allS$
 may be on the order of at least a few percent between individuals of a population --

### 2017-12-01 kalman_walk_paper.tex
general edits to paper

@@
 \end{tikzpicture}
-  \caption{A possible non-minimal three-gene oscillator phenotypically equivalent to the systems in Examples \ref{ex:oscillator} and \ref{ex:all_osc}.}
+  \caption{A possible non-minimal three-gene oscillator, phenotypically equivalent to $A(\tau)$, the systems in Examples \ref{ex:oscillator} and \ref{ex:all_osc}.}
   \label{fig:3D_osc}
@@
 
-\paragraph{Hybrid incompatibility}
+\subsection*{Hybrid incompatibility}
 Two parents with the optimal phenotype can produce offspring whose phenotype is suboptimal
@@
 previous authors have argued that neutral processes are likely too slow to be a significant driver of speciation \citep{nei1983models,seehausen2014genomics}.
-Using simulations, \citet{porter2002speciation} demonstrated the accumulation of hybrid incompatibilities under directional, but not stabilizing selection, and \citet{palmer2009dynamics} observed the appearence of incompatibilities in suboptimally adapted populations in a constant environment.
+Using simulations, \citet{porter2002speciation} demonstrated the accumulation of hybrid incompatibilities under directional, but not stabilizing selection, and \citet{palmer2009dynamics} observed the appearence of incompatibilities in populations evolving in a constant environment, only if the populations were suboptimally adapted.
 This, in light of the few known incompatibilities, has lead some to conclude that hybrid incompatibility is typically a byproduct of positive selection \citep{orr2004speciation, schluter2009evidence} 
@@
 Finally, despite our model's precise separation of phenotype and kryptotype, this relationship in nature may be far more complicated as aspects of the kryptotype may be less ``hidden'' than we currently assume.
-For instance, attributes excluded from the phenotype as modelled here, ignore the potential energy costs associated with execessively large (nonminimal) kryptotypes, as well as the relationship between a specific network architecture and mutational robustness, as well as many other presently neglected phenomena.
+For instance, attributes excluded from the phenotype as modelled here, ignore the potential energy costs associated with execessively large (non-minimal) kryptotypes, as well as the relationship between a specific network architecture and mutational robustness, as well as many other presently neglected phenomena.
 Despite the model's simplifications and potentially nontrivial oversights, it may still roughly reflect and shed light upon speciation patterns as they occur in nature.

### 2017-12-01 examples/F2_osc_plotting.R

@@ -1,5 +1,13 @@
 ####################
-
+Df <- function (A,t, BK, CK) {
+  exp(-t/(4*pi)) * ( h(t, M=A, BK=B, CK=C) - optimal_h(t) )^2
+}
+####################################
+D <- function (A, BK, CK, upper=10, ...) {
+  f <- function (t) { Df(A,t, B, C) }
+  integrate(f, lower=0, upper=upper, ...)$value
+}
+#############################
 A <- matrix(c(0,-1,1,0), nrow=2);
 B <- matrix(c(1,1), ncol=1);
 C <- matrix(c(1,0), nrow=1);
@@ -69,3 +77,36 @@ plot(tt, pdyn(tt, A=A), type='l',
 for (i in 1:256){
 lines(tt, pdyn(tt, A=progeny[[i]]), col=sample(256,1), lty=1)
 }
+
+
+##################
+
+f2_divergence <- function(tau, eps) {
+  progeny <- progeny_gen(tau,eps)
+  avg_div <- 0
+  for (i in 1:256){
+    avg_div <- avg_div + D(progeny[[i]])/256
+  }
+  return(avg_div)
+}
+
+d <- seq(0,-1,-0.01)
+df2 <- sapply(d, function(d) f2_divergence(0,d))
+
+##########
+
+f1_divergence <- function(tau, eps){
+  f1div <- D((S(tau) + S(tau+eps))/2)
+  return(f1div)
+}
+
+df1 <- sapply(d, function(d) f1_divergence(0,d))
+
+###############
+gen_d <- sapply(d, function(d) norm(S(0)-S(0+d),"F"))
+
+pdf("~/kalman_walk/examples/F2_vs_F1_divergence_tau0.pdf")
+plot(df1~gen_d, ylim=c(0,15), type = "l", lwd=3, col="red", xlab="Genetic distance", ylab="Phenotypic divergence")
+lines(df2~gen_d, lwd=3, col="blue")
+legend(0.2,10, legend=c("F1 hybrids", "F2 hybrids"), col=c("red", "blue"), lty=1, lwd=3)
+dev.off()

### 2017-12-01 kalman_walk_paper.tex
general edits to paper

@@
 macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level 
-\citep{kimura1981possibility, true2001developmental, tanay2005conservation, tsong2006evolution, hare2008sepsid, vierstra2014mouse, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription}.
+\citep{kimura1981possibility, true2001developmental, tanay2005conservation, tsong2006evolution, hare2008sepsid, lavoie2010evolutionary, vierstra2014mouse, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription}.
  % \jss{should hare2008sepsid be here? It's sequence but not ``system'' drift}
@@
 
-\begin{example}[All Phenotypically Equivalent Oscillators] \label{ex:all_osc}
+\begin{example}[All phenotypically equivalent oscillators] \label{ex:all_osc}
 The oscillator of example \ref{ex:oscillator} is minimal, and so any equivalent system is a change of coordinates
@@
 and there is often significant regulatory variation segregating within populations \citep{XXX}.
-\js{cite sergey?}
 Transcription in hybrids between closely related species with conserved transcriptional patterns can also be divergent 
@@
 \plr{revisit}
-Of course the basic assumptions of this model will be violated in practice (constant selective pressures, etc.), however, this model can function as a ``neutral null'' description of gene network evolution (a strategy advocated for in \citet{lynch2007frailty, fay2008evaluating, koonin2016splendor}).
+Of course the basic assumptions of this model will be violated in practice (constant selective pressures, etc.), however, this model can function as a ``neutral null'' description of gene network evolution (a strategy advocated for by \citet{lynch2007frailty, fay2008evaluating, koonin2016splendor}).
 If this model succeeds in describing actual system evolution, it can be inferred that the mechanism underlying species formation does not require the inclusion of adaptation or changes in selection to explain the emergence of hybrid incompatibility.
@@
 the Dobzhansky-Muller model of hybrid incompatibility is agnostic to the relative importance of neutral versus selective genetic substitutions \citep{coyne1998evolutionary},
-and plausible mechanisms for the origin of Dobzhansky--Muller incompatibilities by neutral genetic drift have been proposed \citep{lynch_force} 
+and plausible mechanisms for the origin of Dobzhansky--Muller incompatibilities by neutral genetic drift have been proposed \citep{lynch2000origin} 
 and under stabilizing selection \citep{fierst},

### 2017-12-04 kalman_walk_paper.tex
added citation
edited intro

@@
 
-\section*{Rapid speciation despite conservation of phenotype}
 {\centering
+{\Large \bf Rapid speciation despite conservation of phenotype} \\
 Joshua S. Schiffman$^{\dagger}$ \qquad Peter L. Ralph$^{\dagger \ddagger}$ \\
@@
 how a gene network underlying a conserved phenotype evolves,
-as the genetic drift of small mutational tweaks to these molecular pathways
+as the genetic drift of small mutational changes to these molecular pathways
 cause a population to explore the set of mechanisms with identical phenotypes.
-In this setting we treat an organism as a ``black box''
+% Treating an organism as a ``black box''
+To do this, we model an organism's internal state as a linear system of differential equations
 for which the environment provides input and the phenotype is the output,
@@
 %to an output that is subject to the pressures of natural selection-- the phenotype.
-and there exists an exact characterization of the set of all mechanisms that give the same input--output relationship.
+in which context there exists 
+an exact characterization of the set of all mechanisms that give the same input--output relationship.
 %There exists an exact expression for the set of all linear mechanisms with identical phenotypes, 
 %Due to the genetic drift of small mutational tweaks to these molecular pathways, we expect an evolving population to explore this set.
-Within this framework, we show that there is never a unique network architecture for any phenotype
-and that the evolutionary exploration of these distinct and mutationally connected mechanisms
+This characterization implies that selectively neutral directions in genotype space should be common
+% there is never a unique network architecture for any phenotype
+and that the evolutionary exploration of these distinct but equivalent mechanisms
 can lead to the reproductive incompatibility of idependently evolving populations.
-  This evolutionary exploration, which we refer to as system drift, proceeds at a rate proportional to the effective population size ($N_e$) and amount of intrapopulation genetic variation.
-We estimate that %in reasonably numerous species,
-  this process can lead to the formation of new species, perhaps in as few as $0.1 N_e$ generations.
+This evolutionary exploration, or \emph{system drift}, 
+proceeds at a rate inversely proportional the amount of intrapopulation genetic variation
+divided by the effective population size ($N_e$).
+At biologically reasonable parameter values
+this process can lead to substantial interpopulation incompatibility,
+and thus speciation, in fewer than $N_e$ generations.
 %over a number of generations proportional to the effective population size.
-This model also naturally yields, what appears to be, a distinct explanation for Haldane's rule, or why heterogametic hybrids tend to be disrupted more often than homogametes during the early stages of speciation.
+This model also naturally predicts Haldane's rule, 
+thus providing another possible explanation
+of why heterogametic hybrids tend to be disrupted more often than homogametes during the early stages of speciation.
 \end{abstract}
+
+\plr{I removed the $0.1 N_e$ because that's only in the "metapopulation" case,
+which is where you've probably already got subpopulations acting like species already.}
 
@@
 
-A complex molecular machinery translates an organism's genome into the characteristics on which natural selection acts,
+It is an overarching goal of many biological subdisciplines 
+to attain a general understanding of the function and evolution of the 
+complex molecular machinery that translates an organism's genome 
+into the characteristics on which natural selection acts,
 the phenotype.
@@
 %Attaining a general understanding of the functioning and evolution of this molecular machinery
-It is an overarching goal of many biological subdisciplines to attain a general understanding of the function and evolution of this molecular machinery. 
 For example, 
@@
 even if a one-to-one mapping between model and behavior is possible in theory,
-  even tiny amounts of noise can make inference problems nonidentifiable in practice \jss{cite?}.
-
-Although nonidentifiability may frustrate the occasional engineer or scientist, viewed from another angle,
+even tiny amounts of noise can make inference problems nonidentifiable in practice \citep{petrov2005well}.
+
+Nonidentifiability is a major barrier to mechanistic understanding of real systems, 
+but viewed from another angle,
 this concept can provide a starting point for thinking about externally equivalent systems
@@
 \citep{kimura1981possibility, true2001developmental, tanay2005conservation, tsong2006evolution, hare2008sepsid, lavoie2010evolutionary, vierstra2014mouse, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription}.
+
+In this paper we outline a theoretical framework to study the evolution of biological systems, such as gene regulatory networks.
+We study the evolution of an optimally adapted population subject to stabilizing selection for phenotype.
+Even if the phenotype remains stable over evolutionary time, the underlying mechanism might not remain so,
+as many distinct (and mutationally connected) molecular pathways can realize identical phenotypes.
+Below, we first apply results from system theory which give
+an analytical description of the set of 
+all linear gene network architectures that yield identical phenotypes.
+Since these phenotypically equivalent gene networks are not necessarily compatible with one another, 
+system drift may result in reproductive incompatibility between populations 
+isolated for a sufficiently long period of time, 
+even in the absence of any sort of adaptative, selective, or environmental change. 
+We then use quantitative genetic theory to estimate how quickly 
+reproductive incompatibility due to system drift will manifest
+\plr{we don't need to summarize the results of the paper in the introduction.  the abstract, and maybe the discussion, but not the introduction.}
+
  % \jss{should hare2008sepsid be here? It's sequence but not ``system'' drift}
@@
 
-In this paper we outline a theoretical framework to study the evolution of biological systems, such as gene regulatory networks.
-We study the evolution of an optimally adapted population subject to stabilizing selection for phenotype -- that is the population evolves under constant selective and environmental pressures.
-Even if the phenotype remains stable over evolutionary time, the underlying mechanism might not remain so,
-%Even when stabilizing selection preserves phenotype over evolutionary time, the underlying mechanism may change, 
-as many distinct (and mutationally connected) molecular pathways can realize identical phenotypes.
-%where the optimal phenotype remains constant over evolutionary time,
-%so that despite strong stabilizing selection for phenotype,
-%neutral drift in the underlying genotype is possible.
-Applying results from system theory, we present
-an analytical description of the set of 
-all linear gene network architectures that yield identical phenotypes, and show that any biological system, in principal, can undergo system drift, as there is never a unique system architecture per phenotype.
-%which gives concrete expectations for how any such
-%system can, in principal, undergo system drift and rewire.
-Even under stabilizing selection, a population can explore the set of all possible phenotypically equivalent gene networks.
-Phenotypically equivalent gene networks are not necessarily compatible with one another, as such system drift may result in the reproductive incompatibility between populations isolated for a sufficiently long period of time, even in the absence of any sort of adaptative, selective, or environmental change.
-Finally, in some cases we estimate that reproductive incompatibility due to system drift can manifest on timescales on the order of $N_e$ generations.
-%Consequentially, two populations isolated for a sufficiently long period of time, 
-%will likely produce inviable hybrids, despite the absence of adaptation, directional selection, or environmental change.
-% Under these conditions, speciation typically occurs on timescales approximately on the order of $N_{e}$ generations, where $N_{e}$ is the effective population size.
+It is not a new observations that there is often more than one way to do the same thing,
+and that this may lead to speciation.
+For instance, \citet{tulchinsky2014hybrid} looked for hybrid incompatibility
+following regulatory sequence drift,
+and \citet{porter2002speciation} studied speciation in populations expressing a simple regulatory cascade.
+Both found rapid speciation following directional selection,
+but only equivocal support for speciation under models of purely neutral drift.
+Our model differs from previous work in that we use linear systems theory
+to both explore a much richer class of regulatory network models,
+and provide analytical expectations in large populations with complex phenotypes
+that would be inaccessible to population simulations.
+\plr{missing any important references here?}
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section*{Evolutionary system theory}
 
-Here we construct as abstract model to study biological systems, such as gene regulatory networks.
-The behavior (i.e.
-temporal dynamics) of such a system is determined by a collection of $n$ coregulating moleculues -- such as transcription factors, as well as external or environmental inputs.
-%We study an abstract model of temporal dynamics of the concentrations of a collection of $n$ coregulating molecules
-%within an organism, that may also be affected by temporally varying signals from the environment.
+We first describe the class of models we study, such as gene regulatory networks. 
+The behavior (i.e., the temporal dynamics) of such a system is determined by a collection of $n$ coregulating moleculues -- such as transcription factors, as well as external or environmental inputs.
+% We model the temporal dynamics of the concentrations of a collection of $n$ coregulating molecules
+% within an organism, that may also be affected by temporally varying signals from the environment.
 %% Organisms' phenotypes are constructed by gene by gene and gene by environment interactions. 

### 2017-12-04 kalman_walk_paper.tex
edited up to bit about inheritance

@@
 
-REFERENCE: Barton's paper (PETER).
+REFERENCE: Barton's paper (PETER). \citep{barton2010natural}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Evolutionary system theory}
-
-We first describe the class of models we study, such as gene regulatory networks. 
-The behavior (i.e., the temporal dynamics) of such a system is determined by a collection of $n$ coregulating moleculues -- such as transcription factors, as well as external or environmental inputs.
+\section*{Results}
+
+We first describe the class of models we study, motivated by gene regulatory networks. 
+The behavior (i.e., the temporal dynamics) of such a system is determined by a collection of $n$ coregulating moleculues 
+-- such as transcription factors -- as well as by external or environmental inputs.
 % We model the temporal dynamics of the concentrations of a collection of $n$ coregulating molecules
@@
 We write $\kappa(t)$ for the vector of $n$ molecular concentrations at time $t$.
-The $m$ ``inputs'' determined exogenously to the system are denoted $u(t)$,
-and the $\ell$ ``outputs'' are denoted $\phi(t)$.
+The vector of $m$ ``inputs'' determined exogenously to the system is denoted $u(t)$,
+and the vector of $\ell$ ``outputs'' is denoted $\phi(t)$.
 The output is merely a linear function of the internal state:
-$\phi_i(t) = \sum_j C_{ij} \kappa_i(t)$
+$\phi_i(t) = \sum_j C_{ij} \kappa_j(t)$
 for some matrix $C$.
@@
 The dynamics are determined by
-the matrix of regulatory coefficients, $A$;
+the matrix of regulatory coefficients, $A$,
 a time-varying vector of inputs $u(t)$,
@@
 The form of $B$ determines how the environment influences transcription factor expression levels,
-and $C$ might be the rate of production of a downstream enzyme
-(although other arrangements could be made).
-% \emph{E.g.} for a metabolic system, $C_{ij}$ is the amount the $j$th metabolite $u_j$ affects the production of the $i$th enzyme.
-
-% {Linearity} also ``zero-state equivalence''
+and $C$ might determine the rate of production of downstream enzymes.
+
+%% {Linearity} also ``zero-state equivalence''
 Here we have assumed that the system is linear,
@@
 % Cellular division is governed by many different processes, however it is thought that its rhythm is partially controlled by oscillating gene transcription \citep{orlando2008global}.
-Suppose there are two genes, 
+There are two genes, 
 whose transcript concentrations are given by $\kappa_1(t)$ and $\kappa_2(t)$, 
-and that gene-2 upregulates gene-1 and that gene-1 downregulates gene-2 with equal strength.
-Furthermore, suppose that only the dynamics of gene-1 are consequential to the oscillator 
+and gene-2 upregulates gene-1, while gene-1 downregulates gene-2 with equal strength.
+Only the dynamics of gene-1 are consequential to the oscillator 
 (perhaps the amount of gene-1 activates another downstream gene network). 
-Lastly suppose that the production of both genes is equally upregulated by an exogenous signal.
+Lastly, both genes are equally upregulated by an exogenous signal.
 The dynamics of the system are described by
@@
  %    \end{align*}
-
 Suppose the input is an impulse at time zero (a delta function), and so its phenotype is equal to its impulse response,
@@
 The system and its dynamics are referred to in Figure \ref{fig:oscillator}. We return to the evolution of such a system below.   
+\end{example}
+
 \begin{figure}[H]
@@
 \end{figure}
-  \end{example}
 
@@
 % (recall that $\kappa(0) = \bar \kappa(0) = 0$). 
+In other words, phenotypically equivalent systems respond identically for \emph{any} input.
 
@@
 and $C$ in the direction of $-CZ$
-will leave the phenotype unchanged.
-If $A$ is invertible, for instance,
+will leave the phenotype unchanged to second order in the size of the perturbation.
+%% I believe the following is necessary and sufficient:
+If the columns of $B$ and the rows of $C$ are not all eigenvectors of $A$,
 then any such $Z$ will result in a different system.
@@
 and that any two phenotypically equivalent systems of dimension $n_\text{min}$ are related by a change of coordinates.
-
-Some gene networks, however, can grow or shrink, perhaps following gene duplications and deletions, 
-and also still preserve their phenotypes.
-More generally, even if the system is not minimal, 
-results from systems theory % -- the Kalman decomposition --
+Since gene networks can grow or shrink following gene duplications and deletions, 
+these additional degrees of freedom can apply in principle to any system.
+
+Even if the system is not minimal, results from systems theory % -- the Kalman decomposition --
 explicitly describe the set of all phenotypically equivalent systems.
@@
 We refer to $\allS(A_0,B_0,C_0)$ as the set of all systems phenotypically equivalent
-to the system defined by $(A_0, B_0, C_0)$.
-Concretely, this is
+to the system defined by $(A_0, B_0, C_0)$:
 \begin{equation} \label{eqn:equivalence}
@@
 Analogously, we can only observe motion of the system in certain directions
-(those lying in the span of the columns of $C$),
+(those lying in the span of the rows of $C$),
 and so can only infer motion in what is known as the \emph{observable subspace}.
@@
 Only the components that are both reachable and observable determine the system's phenotype --
-that is, components that respond to an input and components that produce an observable output. 
+that is, components that both respond to an input and produce an observable output. 
 % it is possible to add components that respond to an input, 
@@
 and therefore, the system is phenotypically equivalent to the \emph{minimal} system $(A_{\ro}, B_{\ro}, C_{\ro})$.
-(Here the subscript $\ro$ refers to the both \emph{reachable and observable} subspace,
-while $\nrno$ refers to the \emph{unreachable and unobservable} subspace,
-and similarly for $\nro$ and $\rno$.)
-
-Any two minimal systems are related by a change of coordinates,
-and so the minimal subsystems obtained by the Kalman decomposition
-are unique up to a change of coordinates.
-In particular, this implies that there is no equivalent system with a smaller number of kryptotypic dimensions
+
+This decomposition is unique up to a change of basis that preserves the block structure.
+In particular, 
+the minimal subsystem obtained by the Kalman decomposition
+is unique up to a change of coordinates.
+This implies that there is no equivalent system with a smaller number of kryptotypic dimensions
 than the dimension of the minimal system.
-It is also remarkable to note that the gene regulatory network architecture to achieve a given input--output map is never unique --
+It is remarkable that the gene regulatory network architecture to achieve a given input--output map is never unique --
 both the change of basis used to obtain the decomposition
@@
 and so represent degrees of freedom.
+(However, some of these subspaces may affect how the system deals with noise.)
 
@@
 % and $\reachable$ is the largest $A$-invariant subspace contained in the image of $B$.)
-The four subspaces, $\rno$, $\ro$, $\nrno$, and $nro$
-are defined from these by intersections and orthogonal complements.
+The four subspaces, $\rno$, $\ro$, $\nrno$, and $\nro$
+are defined from these by intersections and orthogonal complements --
+$\ro$ refers to the both \emph{reachable and observable} subspace,
+while $\nrno$ refers to the \emph{unreachable and unobservable} subspace,
+and similarly for $\nro$ and $\rno$.
 
@@
 
-For the remainder of the paper, we interpret $\allS$ as the phenotypically neutral landscape, 
-wherein a large population will drift under environmental and selective stasis. 
+For the remainder of the paper, we interpret $\allS$ as the neutral set in the fitness landscape, 
+along which a large population will drift under environmental and selective stasis. 
 Even if the phenotype is constrained and remains constant through evolutionary time, 
@@
 there are \emph{still} usually degrees of freedom. 
-These correspond to distinct genetic networks that perform indistinguishable functions. 
 The following example \ref{ex:all_osc} gives the set of minimal systems equivalent to the oscillator of Example \ref{ex:oscillator},
 that all share common $B$ and $C$ matrices.
-The oscillator can also be equivalently realized by a three-gene (or larger) network, and will have even more evolutionary degrees of freedom available, as in e.g., Figure \ref{fig:3D_osc}.
+The oscillator can also be equivalently realized by a three-gene (or larger) network, and will have even more evolutionary degrees of freedom available, 
+as in Figure \ref{fig:3D_osc}.
 
@@
 If we further require $B$ and $C$ to be invariant then we need $VB=B$ and $CV=C$.
-Solving these equations,
-we find that a one-parameter family $(A(\tau), B, C)$ describes the set of all two-gene systems
-phenotypically equivalent to the oscillator,
-where
+Therefore
+the following one-parameter family $(A(\tau), B, C)$ describes the set of all two-gene systems
+phenotypically equivalent to the oscillator:
     \begin{align*}
@@
     \end{align*}
-The resulting set of systems,
-and their dynamics, are depicted in Figure \ref{fig:all_osc}
+The resulting set of systems are depicted in Figure \ref{fig:all_osc}.
 \end{example}
@@
 
-Diploid organisms have two genomes, each of which encodes a set of system coefficients.
-We assume that a diploid which has inherited systems $(A', B', C')$ and $(A'', B'', C'')$ from each of its parents,
+First, we must specify how sexual reproduction acts on these systems.
+Suppose that each of a diploid organisms' two genomes encodes a set of system coefficients.
+We assume that a diploid which has inherited systems $(A', B', C')$ and $(A'', B'', C'')$ from its two parents
 has phenotype determined by the system that averages these two,
@@
 Transcription in hybrids between closely related species with conserved transcriptional patterns can also be divergent 
-\citep{haerty2006gene, maheshwari2012cis, coolon2014tempo, michalak2004association}.
+\citep{haerty2006gene, maheshwari2012cis, coolon2014tempo, michalak2004association, mack2016gene}.
 Furthermore, in cryptic species complexes (e.g., sun skinks \citep{barley2013challenge}),

### 2017-12-04 kalman_walk_paper.tex
note on inheritance

@@
 Each genome an organism inherits is generated by meiosis,
-in which both of its diploid parents recombine their two genomes.
-We will assume that each coefficient (i.e., entry of $A$, $B$ or $C$)
-is determined by a single nonrecombining locus,
-so that each coefficient in the system produced by meiosis is an independent random choice
-between the two parental coefficients.
-With these definitions, an $F_1$ offspring
-carries a system copy from each parent,
+in which both of its diploid parents recombine their two genomes,
+and so an $F_1$ offspring carries one system copy from each parent,
 and an $F_2$ is an offspring of two independently formed $F_1$s.
@@
 
-This is a simplification:
-since the $i^\text{th}$ row of $A$ summarizes how each gene regulates gene $i$,
+Exactly how the coefficients 
+(i.e., entries of $A$, $B$ or $C$)
+of a haploid system inherited by an offspring from her diploid parent
+are determined by the parent's two systems
+depends on the genetic basis of any variation in the coefficients.
+Since the $i^\text{th}$ row of $A$ summarizes how each gene regulates gene $i$,
 and hence is determined by the promoter region of gene $i$,
-we would actually expect the elements of a row of $A$ to tend to be inherited together.
-Similarly, we expect in practice heritable variation in each coefficient
-to be determined by more than one locus --
-but this may be a reasonable approximation.
+the elements of a row of $A$ to tend to be inherited together,
+which will create covariance between entries of the same row.
+It is, however, a quite general observation that the variation seen among recombinant systems
+is proportional to the difference between the two parental systems.
+This is certainly true if each coefficient is determined by a single nonrecombining locus,
+so that each coefficient in the system produced by meiosis is an independent random choice
+between the two parental coefficients.
+On the other hand, if a coefficient is determined by additive contributions of independently segregating loci,
+then the law of total variance applied after conditioning on the parental origin of each allele
+implies that the variance is equal to one-quarter of the squared difference between parental coefficients.
+
+\plr{continue editing here}
 
 Offspring formed from two phenotypically identical systems do not necessarily exhibit the same phenotype as both of its parents -- in other words $\allS$, the set of all systems phenotypically equivalent to a given one, is not, in general, closed under averaging or recombination.
-Next we discuss how this fact can contribute to hybrid incompatibility and genetic load.
 If sexual recombination among systems drawn from $\allS$ yields systems with divergent phenotypes, populations containing significant diversity in $\allS$ can carry genetic load, and isolated populations may fail to produce hybrids with viable phenotypes.
@@
 Suppose that $\rho(t)$ is a nonnegative, smooth, square-integrable weighting function,
-suppose that $h_0(t)$ is the \emph{optimal} impulse response function 
+$h_0(t)$ is the \emph{optimal} impulse response function,
 and define the ``distance to optimum'' of another impulse response function

### 2017-12-05 kalman_decomposition.tex
edited through

--- 2017-08-01 handnotes.tex
@@
-\documentclass{article}
-\usepackage{amsmath, amssymb, color, xcolor}
-\usepackage{graphicx, wrapfig, float, caption, dsfont, bbm}
-\usepackage{fullpage}
-\usepackage[backref=page, hidelinks, colorlinks=true, citecolor=blue!60!black!100]{hyperref}
-\usepackage{tikz}
-\usetikzlibrary{arrows.meta, shapes}
-\usepackage{caption, subcaption}
-\usepackage{natbib} % gives us \citet: Author (year) and \citep: (Author; year)
-\usepackage{authblk}
-
-\usepackage{multicol}
-
-\newcommand{\plr}[1]{{\color{blue}\it #1}}
-\newcommand{\jss}[1]{{\color{olive}\it #1}}
-% \newcommand{\ddt}{\frac{d}{dt}}
-\newcommand{\ddt}{\dot}
-\newcommand{\ro}{{ro}}
-\newcommand{\nro}{{\bar{r}o}}
-\newcommand{\rno}{{r\bar{o}}}
-\newcommand{\nrno}{{\bar{r}\bar{o}}}
-\newcommand{\reachable}{\mathcal{R}}
-\newcommand{\unobservable}{\bar{\mathcal{O}}}
-\newcommand{\R}{\mathbb{R}}
-\newcommand{\E}{\mathbb{E}}
-\renewcommand{\P}{\mathbb{P}}
-\newcommand{\pda}{\frac{\partial}{\partial A_{ij}}}
-\newcommand{\ind}{\mathds{1}}
-
-\newcommand{\calA}{\mathcal{A}}
-\newcommand{\calH}{\mathcal{H}}
-\newcommand{\diag}{\text{diag}}
-\newcommand{\1}{\mathbbm{1}}
-
-\DeclareMathOperator{\spn}{span}
-
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}{Lemma}
-\newtheorem{definition}{Definition}
-\newtheorem{example}{Example}
-
-\begin{document}
-
-\section*{Gene regulatory network drift and speciation occurs rapidly under neutrality}
-{\centering
-Joshua S. Schiffman$^{\dagger}$ \qquad Peter L. Ralph$^{\dagger \ddagger}$ \\
-$^{\dagger}$University of Southern California, Los Angeles, California \qquad $^{\ddagger}$University of Oregon, Eugene, Oregon \\
-\texttt{jsschiff@usc.edu} \qquad \texttt{plr@uoregon.edu}
-\\
-}
-
-%\begin{multicols}{2}
-
-\begin{abstract}
-Here we introduce an analytical theory to study the evolution of biological systems, such as gene regulatory networks.
-The evolutionary conservation of phenotype under selective and environmental stasis does not necessitate conservation of the underlying mechanism, as distinct molecular pathways can realize identical phenotypes.
-Here we give an exact expression for the set of all linear mechanisms with identical phenotypes, and expect evolution under neutrality to explore this set.
-We employ a quantitative genetic approach to model evolution under neutrality as a random process over the set of all phenotype-invariant mechanisms: only mutational tweaks to the pathway that leave the phenotype invariant are optimally fit.
-We show that there is never a unique linear system architecture for any phenotype and that the evolutionary exploration of these distinct and mutationally connected mechanisms can lead to the rapid accumulation of hybrid incompatibilities between allopatric populations and thus lead to the rapid formation of new species -- in fewer generations than there are breeding individuals in a population.
-\end{abstract}
-
-  \plr{Additional ideas to consider adding:
-\begin{itemize}
-    %\item Haldane's rule (easy point in discussion: makes F1s look like F2s)
-    %\item add point about F1: quartic and F2: quadratic to results and maybe abstract/discussion
-    \item hybrid vigor (need to do calculation)
-    \item discuss linearity, linearization, and canalization in introduction
-    %\item obtain estimates of variation in $A$ from thermodynamic occupancy model
-\end{itemize}}
-
-%%%%%%%%%%%%%%%%%%%%%%
-\section*{Introduction}
-
-Bridging the gulf between an organism's genome and phenotype is a poorly understood and complex molecular machinery.
-Progress in a suite of biological subdisciplines is stalled by our general lack of understanding of this molecular machinery: with respect to both its function and evolution.
-There does exist a growing body of experiment and data on the evolutionary histories and molecular characterizations of particular gene regulatory networks \citep{jaeger2011gap, davidson2006gene, israel2016comparative}, as well as thoughtful verbal and conceptual models \citep{true2001developmental, gwagner1, weiss2000phenogenetic, edelman2001degeneracy}.
-However, as Hardy and Weinberg taught us over a century ago, verbal theories are often insufficient, if not downright misleading \citep{hardy1908mendelian, weinberg1908vererbungsgesetze, servedio2014not}.
-This is especially pertinent given the staggering complexity and scope of contemporary research programs.
-This outlook necessitates the advancement of conceptual frameworks of such precision, only mathematics will suffice, as models allow the development of concrete numerical predictions.
-Previously it has been suggested that any idealized study of evolution is incomplete without a mathematically sufficient description of the genotype, phenotype, and transformation from one to the other \citep{Lewontin1974genetic}.%Here we outline an analytical theory to study the evolution of biological systems, borrowing insight and methods from control engineering, systems identification, and dynamical systems theory.
-
-%Presently, we focus on the neutral evolution of genetic regulatory networks.
-That is, we analytically describe the set of all linear gene networks (of any size) that produce identical phenotypes -- and the evolutionary paths connecting them.
-In the idealized case of a perfectly adapted population, constant selection, and a static environment, we observe evolution under the ``conservation of phenotype'' as a Brownian motion over phenotypically-invariant network-space.
-This analysis provides insight to the mechanisms and parameters important for understanding developmental systems drift, network rewiring, evolvability, epistasis, and speciation, as well as the tenuous connection between network architecture and function.
-
-%It is commonly taught that an organism's genome contains the heritable material 
-%that natural selection filters and that an organism's phenotype directly determines its
-%evolutionary fitness. Between genotype and phenotype is an often complicated and poorly
-%undrstood molecular machinery -- and it is a major goal common to many different disciplines
-%within the life sciences to elucidate its form, function, and evolution. These aims are
-%delicately intertwined and a comphrehensive understanding of a system's evolution requires
-%an understanding of its function and vice versa.
-
-The molecular machinery, interacting with the environment, and bridging genotype to phenotype
-can be mathematically described as a dynamical system -- or a system of differential equations\citep{jaeger2015comet}.
- Movement in this direction is ongoing, as researchers have begun to study 
-the evolution of both abstract \citep{wagner1994evolution, wagner1996does,  siegal2002waddington, bergman2003evolutionary, draghi2015robustness} and empirically inspired computational and mathematical models of gene regulatory networks (GRNs) \citep{mjolsness1991connectionist, jaeger2004dynamic, maria1, vitaly1, vitaly2, crombach2016gap, wotton2015quantitative, chertkova2017insilico}.
-If we allow the reasonable assumption that the genotype-phenotype map can be represented as a system of differential equations, we can immediately discuss its evolution and function in a much more mechanistic, yet general, manner.
-
-In some fields that seek to fit parametric models to experimental data, such as control
-theory, chemical engineering, and statistics, it is well known that mathematical models
-can fundamentally be \emph{unidentifiable} and/or \emph{indistinguishable} -- meaning that 
-there can be uncertainty about an inferred model's parameters or even its claims about
-causal structure, even with access to complete and perfect data \citep{bellman1970structural, grewal1976identifiability, walter1984structural}. Models with different 
-parameter schemes, or even different mechanics can be equally accurate, but still not
-\emph{actually} agree with what is being modelled. In control theory, where electrical 
-circuits and mechanical systems are often the focus, it is understood that there can be an 
-infinite number of ``realizations'', or ways to reverse engineer the dynamics of a black box,
-even if all possible input and output experiments on the black box are performed \citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. In chemical
-engineering, those who study chemical reaction networks sometimes refer to the fundamental
-unidentifiability of these networks as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}.
-In computer science, this is framed as the relationship among processes that simulate one another \citep{van2004equivalence}.
-Although this may frustrate the occasional engineer or scientist, viewed from another angle,
-the concepts of unidentifiability and indistinguishability can provide a starting point for
-thinking about externally equivalent systems -- systems that evolution can explore, so long
-as the parameters and structures can be realized biologically. In fact, evolutionary
-biologists who study convergent versus parallel evolution, homology, and analogy are very familiar with such functional symmetries; macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level \citep{true2001developmental, tsong2006evolution, hare2008sepsid, vierstra2014mouse, stergachis2014conservation, taylor2016diverse, matsui2015regulatory}.
-
-  In this paper we outline a theoretical framework to study the evolution of biological systems.
-We expect wide applicability to this approach (\emph{i.e.} non-neutral evolution), however presently, we focus solely on neutral evolution, that is where phenotype is conserved over evolutionary time.
-
-  We derive an analytical description of the set of all linear biological systems with identical phenotypes -- that is we describe the set of all gene network architectures that yield identical phenotypes, and show that all biological systems can, in principal, can undergo systems drift.
-In the neutral case, this set describes a manifold that evolution explores leaving phenotype invariant with respect to mutation, and predicts that if two populations become reproductively isolated, hybrid incompatibility can occur, despite the absence of adaptation, directional selection, or environmental change.
-Speciation typically occurs on timescales approximately on the order of $N_{e}$ generations, where $N_{e}$ is the effective population size.
-
-\section*{Gene Networks as Linear Dynamical Systems}
-
-%  \jss{Organisms' phenotypes are constructred by gene by gene by environment interactions.
-Here we simply define the phenotype to be the organismal temporal molecular dynamics directly under natural selection.
-The \emph{what}, \emph{when}, and \emph{where}, of an organism's molecules that are physiologically or otherwise relevant for survival.
-%   Thus we say that some function $\phi(t)$ is a phenotype where, 
-%  \begin{align}
-%    \phi(t) = \int_{0}^{\infty} h(t) u(t) dt  ,
-%  \end{align}
-%  and $h(t)$ is the \emph{impulse response} of the system and $u(t)$ is the \emph{input} function, both functions of time $t$.
-The input can be interpreted as the environment, as initial conditions, or otherwise, depending on the biological specifics under study.
-The phenotype is a convolution of both the impulse response and the input, as allowed by our assumption of linearity.
-%
-%  Essentially the phenotype $\phi(t)$ is a consequence of an organism's specific gene by gene interactions, given by $h(t)$, reacting further with the local environment, given by $u(t)$. 
-%
-%  We describe the impulse response as, 
-%  \begin{align}
-%    h(t) = C e^{A t} B
-%  \end{align}
-%  where $A$ is a gene regulatory network (although this model can be generalized to other biological networks) -- a square matrix, and $B$ filters and translates the input to the system.
-The form of $B$ determines precisely how the state of the external environment influences the internal gene network.
-$C$ filters and translates the dynamics of the system and precesily determines the output, that is, what is visible to selection.
-%
-%  Generally $A$ can be any real $n \times n$ matrix, $B$ any $n \times \ell$, and $C$ any $\ell \times n$ dimensional matrix.
-Each $i$th row of $A$ describes the \emph{cis}-regulatory module for gene $i$, and each $j$th entry, the specific regulatory influence of gene $j$ on gene $i$.
-%
-%  Although $\phi(t)$ describes the phenotype given an input, $h(t)$ describes the phenotype subject only to an impulse -- an input present initially and absent immediately thereafter.
-Typically, a system $\Sigma$, is defined as,
-%  \begin{align}
-%    \Sigma = \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right.
-%  \end{align}
-%  Variables have the same identities as described above and $\kappa(t)$ is a vector of molecule concentrations at time $t$.
-Therefore the molecular concentrations at a specific time are completely determined by the input and gene by gene interactions.
-Lastly, a portion and/or combination of these molecules, $\phi(t)$, are ``observed'' by selection (this is in constrast to $\kappa(t)$ -- the \emph{kryptotype} -- as it is ``hidden'' from direct selection).
-%}
-Here we outline a method to model biological systems, such as gene regulatory networks, as linear dynamical systems.
-We define an organism's phenotype $\phi(t)$ to be the molecular time-dynamics directly relevant to survival.
-These dynamics are the result of the interconnections of a gene regulatory network (or some other biological system) $A$ (an $n \times n$ matrix) and an environmental input $u(t)$.
-Such a system $\mathcal{S}$, is composed of two equations,
-  \begin{align}
-    \mathcal{S} := \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right.
-  \end{align}
- Where the \emph{kryptotype} $\kappa(t)$, is a list of a system's molecular concentrations at time $t$ -- or more generally the systems \emph{internal dynamics}.
-As some system dynamics may not be visible nor relevant to selection, we distinguish the \emph{kryptotype} (as it is hidden) from the \emph{phenotype} -- which is the systems \emph{external dynamics}.
-  The change in internal molecular concentrations as a function of time $\dot{\kappa}(t)$ is simply a function of the systems current state (or kryptotype), the system architecture $A$ -- such as a gene regulatory network, the environmental input $u(t)$, and $B$ (an $n \times l$ matrix) -- how the system processes its environment.
-$C$ (an $l \times n$ matrix) filters the dynamics relevant to survival.
-  Thus the phenotype is simply a convolution of the system organization and the environment, 
-  \begin{align}
-    \phi(t) = \int_{0}^{t} C e^{A t} B u(t) dt ,
-  \end{align}
-  where we sometimes refer to $h(t) := Ce^{A t}B$ as the \emph{impulse response} of the system.
-As $A$ specifies gene regulatory network architecture, the $i^\text{th}$ row of $A$ can be interpreted as the \emph{cis}-regulatory element (or promoter) for gene $i$.
-
-%We will now lay out the model in more general terms.
-%Suppose that the \emph{internal state} of the system
-%is parameterized by the concentrations of a collection of $n$ molecular species,
-%$S_1, \ldots, S_n$,
-%and the vector of concentrations at time $t$ we denote $\kappa(t)=(\kappa_1(t),\ldots,\kappa_n(t))$.
-%There are also $m$ ``input'' species, whose concentrations are determined
-%exogenously to the system,
-%and are denoted $u(t) = (u_1(t),\ldots,u_m(t))$,
-%and $\ell$ ``output'' species, whose concentrations are denoted
-%$\phi(t) = (\phi_1(t),\ldots,\phi_\ell(t))$.
-%The output is merely a linear function of the internal state:
-%\begin{align*}
-%    \phi_i(t) = \sum_j C_{ij} \kappa_i(t).
-%\end{align*}
-%Since $\phi$ is what natural selection acts on, we refer to it as the \emph{phenotype},
-%and sometimes in contrast refer to $\kappa$ as the \emph{kryptotype},
-%as it is ``hidden'' from direct selection.
-%The rate at which the $i^\text{th}$ species is produced
-%is a weighted sum of the concentrations of the other species
-%as well as the input:
-%\begin{align*}
-%    \ddt \kappa_i(t) = \sum_j A_{ij} \kappa_j(t) + \sum_r B_{ir} u_r(t) .
-%\end{align*}
-%In matrix notation, this is written more concisely as
-%\begin{align} \label{eqn:lti_system}
-%    \ddt \kappa(t) &= A \kappa(t) + B u(t) \\
-%    \phi(t) &= C \kappa(t) .
-%\end{align}
-
-  \begin{example}[Oscillating Gene Network: Cell Cycle Control]\label{ex:oscillator}
-
-    Cellular division is a complicated phenomena, governed by many different processes, however it is agreed that its rhythm is partially controlled by periodic (oscillating) gene transcription \citep{orlando2008global}.
-Consider a simplified model of oscillating gene transcription.
-In the present framework periodic expression requires at minimum two interacting genes.
-
-    Suppose gene-2 up-regulates the transcription of gene-1 and that gene-1 down-regulates the transcription of gene-2 with equal magnitude (of $1$) and relative to each of their concentrations, denoted by $\kappa_{1}$ and $\kappa_{2}$.
-Furthermore, suppose that only the dynamics of gene-1 are consequential to the cell cycle (perhaps the amount of gene-1 activates another downstream gene network).
-Lastly suppose that the production of both genes is stimulated by an impulse of a molecule present immediately after division.
-
-    If the rate each of these genes is expressed is a linear function of their concentrations, the dynamics of the system are given by
-    \begin{align*}
-      \dot{\kappa_{1}}(t) &= \kappa_{2}(t) + u(t) \\
-      \dot{\kappa_{2}}(t) &= - \kappa_{1}(t) + u(t)
-    \end{align*}
-    where $\dot{\kappa}$ denotes the time derivative.
-The initial conditions $\kappa_{1}(0)$ and $\kappa_{2}(0)$, and the input $u(t)$ then determine the concentrations through time.
-If we record the regulatory coefficients in the matrix
-    \begin{align*}
-      A &= \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} ,
-    \end{align*}
-    and define the column vector $B = \begin{bmatrix} 1 & 1 \end{bmatrix}^{T}$, then in matrix notation the dynamics are
-      \begin{align*}
-      \dot{\kappa} &= A \kappa(t) + B u(t) .
-      \end{align*}
-      Since only the dynamics of gene-1 are directly relevant to biological function, the dynamics of interest are given by
-      \begin{align*}
-      \phi(t) &= C \kappa (t)
-      \end{align*}
-      where the row vector $C$ is defined as $C = \begin{bmatrix} 1 & 0 \end{bmatrix}$.
-(Note: if the dynamics of both genes were physiologically relevant to the cell cycle, we would set $C$ to be the identity matrix).
-  %  \begin{align*}
-  %    A = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} , \qquad B = \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix}
-  %  \end{align*}
- %       The oscillatory system $\Sigma$ is thus given as
- %   \begin{align*}
- %     \Sigma = \left \{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 
- %       0 & 1 \\ 
- %      -1 & 0 
- %       \end{bmatrix} \kappa(t) + \begin{bmatrix} 1 \\ 1 \end{bmatrix} u(t) \\ 
- %         \phi(t) &= \begin{bmatrix} 1 & 0 \end{bmatrix} \kappa(t) \end{array} \right.
- %    \end{align*}
-
-      Since the input is simply an impulse, its phenotype is equivalent to its impulse response
-
-      \begin{align*}
-        \phi(t) = h(t) = \sin(t) + \cos(t)  .
-      \end{align*}
-
-    \begin{figure}[H]
-       % \begin{center} 
-      \centering
-         \begin{tabular}{cc}
-            \begin{tikzpicture}
-            \begin{scope}[every node/.style={circle,thick,draw}]
-                \node (A) at (0,0) {$\kappa_{1}$};
-                \node[dashed] (B) at (3,0) {$\kappa_{2}$};
-                \node[shape=rectangle] (U) at (1.5,2) {input};
-                \node[shape=rectangle] (y) at (1.5,-2) {output};
-            \end{scope}
-
-            \begin{scope}[>={Stealth[black]},
-                          every node/.style={fill=white,circle},
-                          every edge/.style={draw=black, thick}]
-                \path [->, >=Rectangle] (A) edge[bend left] node {\tiny $-1$} (B);
-                \path [->] (B) edge[bend left] node {\tiny $1$} (A); 
-                \path[->] (U) edge[dashed] node {\tiny $1$} (A);
-                \path[->] (U) edge[dashed] node {\tiny $1$} (B);
-                \path[->] (A) edge[dashed,bend right] node {\tiny $1$} (y);
-            \end{scope}
-            \begin{scope}[>={Stealth[black]},
-                          every edge/.style={draw=black, thick}]
-                %\path [->] (A) edge[loop left] node {\tiny $\lambda_{1}$} (A);
-                %\path [->] (B) edge[loop left] node {\tiny $\lambda_{2}$} (B);
-            \end{scope}
-
-            \end{tikzpicture} &
-       % \end{center}
-      %  \caption{
-      %      Diagram of Example \ref{ex:osc} in the text.
-      %      \plr{explain what arrows mean if nec}
-      %      \label{fig:oscillator_diagram}}
-   % \end{figure}
-   % \begin{figure}[H]
-    %  \centering
-      \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{osc_impulse}
-   \end{tabular}
-  %  \end{center}
-      \caption{(Left) Graphical representation of the cell cycle control gene network, and (right) plot of the phenotype $\phi(t)$ against time $t$.}
-    \end{figure}
-      We return to the evolution of such a system below.
-  \end{example}
-
-  \section*{Linear Evolutionary Systems}
-
-  The literature is filled with detailed observations of molecular systems and their diversity.
-There are examples of significant diversity in the networks underlying processes such as circadian rhythm \citep{sancar2008intelligent}, cell cycle control \citep{cross2011evolution, kearsey2003enigmatic}, pattern formation, and metabolism  \citep{lavoie2009rearrangements, martchenko2007transcriptional, dalal2016transcriptional, christensen2011unique, hartl2007induction, alam2013aspergillus}.
-Despite a symmetry in functionality or phenotype these systems often differ, sometimes substantially, at the molecular level.
-How many different mechanisms have the same function?
-We urge the reader to consider the metaphorical black box.
-
-  Gene regulatory networks with identical phenotypes (external dynamics) do not necessarily have identical kryptotypes (internal dynamics).
-Any linear and minimal system (a gene network) -- minimal, informally meaning that the system's phenotype is achieved with the fewest possible number of genes -- has an identical phenotype up to a change of coordinates.
-  \begin{align}
-    h(t) &= C e^{A t} B \\
-    &= C V^{-1} V e^{A t} V^{-1} V B \\
-    &= C V^{-1} e^{V A V^{-1} t} V B \\
-    &= \bar{C} e^{\bar{A} t} \bar{B}
-  \end{align}
-  Two biological systems, $\mathcal{S} = \{ A, B, C \}$, and $\bar{\mathcal{S}} = \{\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} \}$, have the same phenotype if they are related by a change of coordinates.
-
-  Although systems may not be identifiable beyond a change of coordinates, at present we are primarily interested in a subset of these systems.
-That is, systems that not only have equivalent external dynamics, but also equivalent input and output relationships.
-Formally, this means systems related by a change of coordinates (any invertible matrix $V$) that leaves $B$ and $C$ invariant:
-  \begin{align}
-    VB &= B \implies \bar{B} = B \\
-    CV &= C \implies \bar{C} = C
-  \end{align}
-  In other words systems with varying gene regulatory network architectures yet identical selection pressures, environment, and phenotype.
-
-  Define $V(\tau)$ as the parameterized change of coordinates matrix that preserves $B$ and $C$, with $\tau$ a vector of free parameters.
-The set of \emph{all} phenotypically invariant (minimal) gene networks is,
-  \begin{align}
-    A(\tau) = V(\tau) A(0) V^{-1}(\tau) ,
-  \end{align}
-  and a \emph{Linear Evolutionary System} is, 
-  \begin{align}
-    \mathcal{S}(\tau) := \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A(\tau) \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right .
-  \end{align}
-  That is, all (linear and minimal) mechanisms capable of producing the same phenotype can be realized by a unique choice of $\tau$ in $\mathcal{S}(\tau)$.
-
-  More generally, \plr{introduce Kalman},
-  we denote by $\calA_n(A_0)$ the set of all $n$-dimensional systems equivalent to $A_0$:
-  \begin{equation} \label{eqn:equivalence}
-      \begin{aligned}
-      \calA_n(A_0) 
-          &= \left\{
-              A : C e^{At} B = C e^{A_0 t} B \; \text{for}\; t \ge 0 
-          \right\}  \\
-          &= \left\{
-              A : C A^k B = C A_0^k B \; \text{for}\; 1 \le k \le n-1 
-          \right\} .
-      \end{aligned}
-  \end{equation}
-  Equivalence of the two characterizations follows from the Cayley-Hamilton theorem.
-  Usually, the dimension $n$ and the reference system $A_0$ is implicit and we write only $\calA$.
-
-  Regardless of minimality, two systems, even in different dimensions, can have identical external dynamics if they are in $\mathcal{A}$.
-This set can be completely parameterized using the \emph{Kalman Decomposition}; the set of all linear gene regulatory networks with equivalent phenotypes can be precisely defined.
-Note that this implies that there is always more than one possible gene regulatory network architecture per phenotype, however we can analytically describe the complete set.
-
-  \begin{example}[All Phenotypically Equivalent Cell Cycle Control Networks]
-
-    Let
-    \begin{align*}
-      A(0) = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix}, \qquad B = \begin{bmatrix} 1 & 1 \end{bmatrix}^{T}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix},
-    \end{align*}
-    with $V(\tau)$ preserving both $B$ and $C$, then
-    the set of all two-gene regulatory networks phenotypically equivalent to the cell cycle control network in Example \ref{ex:oscillator} are given by
-    \begin{align*}
-      A(\tau) &= \begin{bmatrix} 1 & 0 \\ \tau & 1-\tau \end{bmatrix} \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 0 \\ \frac{\tau}{\tau-1} & \frac{-1}{\tau-1} \end{bmatrix}  \qquad \forall \ \tau \neq 1 %\\
-     %   B &= \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix} \\
-       %   h(t) &= \sin(t) + \cos(t) \qquad \forall \ \tau \neq 1
-    \end{align*}
-    \begin{figure}[H]
-    \centering
-\begin{tikzpicture}
-\begin{scope}[every node/.style={circle,thick,draw}]
-  \node (A) at (0,0) {$\kappa_{1}$};
-    \node[dashed] (B) at (4,0) {$\kappa_{2}$};
-    \node[shape=rectangle] (U) at (2,2) {input};
-    \node[shape=rectangle] (y) at (2,-2) {output};
-\end{scope}
-
-\begin{scope}[>={Stealth[black]},
-              every node/.style={fill=white,circle},
-              every edge/.style={draw=black, thick}]
-    \path [->, sloped] (A) edge[bend left] node {\tiny $2 \tau + (\tau-1)^{-1}$} (B);
-    \path [->, sloped] (B) edge[bend left] node {\tiny $-(\tau-1)^{-1}$} (A); 
-    \path[->] (U) edge[dashed] node {\tiny $1$} (A);
-    \path[->] (U) edge[dashed] node {\tiny $1$} (B);
-    \path[->] (A) edge[dashed, bend right] node {\tiny $1$} (y);
-\end{scope}
-\begin{scope}[>={Stealth[black]},
-              every edge/.style={draw=black, thick}]
-    \path [->] (A) edge[loop left] node[sloped, anchor=center, above] {\tiny $1 + (\tau-1)^{-1}$} (A);
-    \path [->] (B) edge[loop right] node[sloped, anchor=center, above] {\tiny $-1 - (\tau-1)^{-1}$} (B);
-\end{scope}
-
-\end{tikzpicture}
-    \caption{A graphical depiction of the set of all externally equivalent cell cycle control networks, $A(\tau)$. $\tau$ can take any value}
-    \end{figure}
- % \end{example}
-
- % \begin{example}[External Equivalence does not imply internal equivalence]
-   Despite the phenotypic equivalence of all instantiations of $A(\tau)$, the internal dynamics, or kryptotypes, vary as a function of $\tau$. 
-    Gene-1 dynamics (blue) are equivalent for network architectures $A(0)$ and $A(2)$, however the dynamics of gene-2 (orange) differ with $\tau$.
-  \begin{figure}[H]
-    \centering
-    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{osc_A0_A2_both_compare}
-    \caption{Gene-1 (blue) and gene-2 (orange) dynamics for $A(0)$ (top) and $A(2)$ (bottom).
-Both (top and bottom) gene-1 dynamics are given by $ \kappa_{1} = \sin(t) + \cos(t)$, and gene-2 by $\kappa_{2} = \cos(t) - \sin(t)$ (top) and $\kappa'_{2} = \cos(t) + 3 \sin(t)$ (bottom).}
-  \end{figure}
-  \end{example}
-
-  %%%%%%%%%%%%%%%%%%%%%%%%
-  \section*{Intraspecific variation and genetic drift}
-
-%  \plr{rework this: we assume optimal phenotype is fixed and population is large but polymorphic and not strictly on the optimum}
-
-%First we focus on a simple evolutionary scenario -- \emph{phenotypically invariant evolution}: a large population, perfectly adapted, and in a constant environment.
-In this circumstance we expect phenotype to be conserved throughout evolutionary time.
-As such we should only expect the phenotype to change as a consequence of genetic drift (small effective population size), adaptation to new selective and/or environemental pressures.
-These phenotypic variations should yield distinct signatures.
-Adaptative changes will change the optimal impulse response function $h(t) \xrightarrow{adaptation} h'(t)$.
-Genetic drift registers as an increase in the intrapopulation variation in $h(t)$.
-% \jss{added the phenotypic invariance stuff here (above) -- not sure if it should put elsewhere yet}
-
-  At any given time, there will be a range of network coefficients present in the population
-  due to segregating genetic polymorphism.
-  Over many generations, even if selective pressures do not change,
-  this range of networks will shift 
-  as recombination, mutation, and demographic noise create new alleles and shift allele frequencies.
-  How much variation do we expect to find within a population?
-  Is this range limited by available variation or kept in check by selection?
-  How fast will a population explore the space of equivalent networks?
-  To answer these questions, we need to know two things known at best only roughly \citep{felsenstein1988phylogenies} --
-  the strength of stabilizing selection on the phenotype,
-  and the amount (and structure) of heritable variation in the genotype.
-  We aim to get order-of-magnitude estimates.
-
-  \plr{amount of heritable variation}
-  We quantify (roughly) the amount of heritable variation
-  by $\sigma^2$, the genetic variance present in a population in a typical entry of $A$.
-  The coefficient $A_{ij}$ measures how much the rate of net production of $i$ changes
-  per change in concentration of $j$.
-  It is generally thought that regulatory sequence change contributes much more to inter- and intraspecific variation
-  than does coding sequence change affecting molecular structure \citep{schmidt2010vertebrate}.
-  In the context of transcription factor networks this may be affected 
-  not only by the binding strength of $j$ to the promoter region of $i$
-  but also the effects of other transcription factors (e.g., cooperativity)
-  and local chromatin accessibility \citep{stefflova2013cooperativity}.
-  For this reason, 
-  the mutational target size for variation in $A_{ij}$ may be much larger than the dozens of base pairs
-  typically implicated in the handful of binding sites for transcription factor $j$ of a typical promoter region.
-  (Recall that although these are best modeled through nonlinear terms,
-  by linearizing we essentially consider first-order effects.)
-  On the other hand, a diverse set of buffering mechanisms are thought to contribute to phenotypic stability
-  in the presence of substantial molecular noise \citep{canalization,buffering},
-  suggesting that substantial variation in the micro-scale dynamics we consider here
-  may be necessary to produce relevant phenotypic effects downstream.
-  \plr{replace "micro-scale" with sthg else or discuss earlier}
-  Differences in $A_{ij}$ due to a sequence change are hard to measure 
-  -- the closest available data we could find either related to variation in transcription factor binding site occupancy
-  or in expression levels (e.g., cis-eQTL).
-  The former may be an overestimate, 
-  since it does not capture buffering effects (if for instance only one site of many needs be occupied for transcription to begin)
-  and the latter probably measures changes in steady-state concentration (our $x_i$) rather than the rate of change.
-  \citet{kasowski2010variation} found differential occupancy in 7.5\%
-  of binding sites of a transcription factor (p65) between human individuals.
-  \citet{verlaan2009targeted} showed that cis-regulatory variation
-  accounts for around 2--6\% of expression variation in human blood-derived primary cells,
-  while \citep{lappalainen2013transcriptome} found that human population variation 
-  explained about 3\% of expression variation.
-  \plr{Get some data from at least one other species in here!}
-  Taken together, this suggests that variation in the entries of $A$
-  may be on the scale of 1\% between individuals of a population --
-  doubtless varying substantially between species and between genes.
-
-  \plr{merge this above}
-  The amount and structure of this standing variation is established over long time scales
-  by many factors, including
-  mutation-selection balance, 
-  shifts in the phenotypic optimum,
-  and/or spatial variation in the optimum \citep{hansen1996translating}.
-  Quantitative genetics models of mutation-selection balance 
-  predict precise levels and structure of standing variation \citep{kimura_mutsel,lande_mutsel,lande1981models},
-  but it is unclear how well these predictions match reality \citep{johnson_barton}
-  and how much they are expected to change over time \citep{arnold_changing_G}.
-  However, empirical work allows us to estimate at least the rough magnitude of variation.
-
-  \plr{strength of selection}
-  It seems certain that selection is not so strong that intra-population variation
-  is strongly deleterious,
-  so that if $u$ is the typical scale on which selection acts
-  -- e.g., a phenotype that differs from the optimum by $x$ has fitness of $\exp(-x^2/2u^2)$ --
-  then $u > \sigma$.
-  However,
-  a range of studies \plr{find them} have found evidence for weak stabilizing selection
-  on regulatory SNPs and cis-eQTL.
-  For instance, 
-  \citet{maria_and_sergey} (others?) found evidence that large-effect regulatory mutations are weakly selected against in Drosophila.
-  This suggests that 
-  the strength of selection on phenotype is sufficient to weakly constrain regulatory variation,
-  so that perhaps $\sigma$ and $u$ are relatively close.
-  This is as would be expected if available variation is held in check by mutation--selection balance
-  rather than genetic drift.
-  A conservative estimate would be that $u = 5 \sigma$;
-  taking $\sigma=.01$ as above, this suggests that changes in phenotype of 5\% are sufficient
-  to effect a noticeable drop in fitness.
-  \plr{BUT $\sigma$ IS VARIATION IN $A$ NOT PHENOTYPE}
-
-  We have guessed that within a population the entries of $A$
-  vary by a factor of $\sigma=0.01$, at least for networks whose function is strongly constrained.
-  Subsequent generations for the most part resample from this diversity,
-  so in a population of effective size $N_e$,
-  simply by the variance of the mean of a random sample,
-  the population mean binding strengths
-  will move a few multiples of $\sigma/\sqrt{N_e}\%$ per generation \citep{lande_drift}.
-  (This could be taken as a definition of $N_e$.)
-  Selection will tend to push this mean towards the optimal set of networks \citep{ou_process},
-  but mean movement parallel to the optimal set (that leaves the phenotype invariant) is unconstrained
-  unless recombination load is substantial \citep{recomb_load}.
-  The action of genetic drift is also strongly determined by covariance between 
-  standing genetic variation in different regulatory coefficients -- known as the $G$ matrix \citep{G_matrix},
-  covariance which may arise due to functional constraints and/or statistical linkage.
-  There may well be functional constraints -- but these are not sufficiently well-known to say anything general about.
-  Linkage will almost certainly lead to covariance
-  if the variation is due to \textit{cis}-regulatory variants,
-  in which case the genetic basis of each \emph{row} of $A$ likely lies within a few kilobases of tightly linked sequence,
-  across which a population may carry only a few common haplotypes.
-  However, covariance due to transiently assembled haplotypes is not expected to be stable over long periods of time --
-  a common \textit{cis}-regulatory haplotype of transcription factor $k$ with particularly strong binding to both $i$ and $j$
-  (leading to positive covariance between $A_{ik}$ and $A_{jk}$)
-  is no more likely to appear than one with strong binding to $i$ but particularly weak binding to $j$ (negative covariance).
-  (Such transient covariances may well increase the variance of the per-generation change in network mean, however \citep{barton_linkage}.)
-  In principle, there may be substantially less variation away from the set of optimal coefficients 
-  than there is along the set, due to the action of selection.
-  If so, this might require substantial epistatic load 
-  -- mortality or reduced fitness of a large proportion of new offspring due to recombination between somewhat incompatible alleles.
-  However, it is unclear if this is likely to occur and XXX we revisit this below.
-
-  It therefore seems reasonable to coarsely model the time evolution of population variation in network coefficients as 
-  (a) a ``cloud'' of width X about the population mean, 
-  which (b) moves as an unbiased Brownian motion through the set of network coefficients that give the optimal phenotype.
-  In fact, the population mean will not produce exactly the optimal phenotype,
-  but it will be convenient to refer to this closest point on the optimal set as ``the population mean''.
-
-  %%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \paragraph{Brownian motion} on the set of equivalent networks:
-  The set $\calA$ is characterized as the solutions to the equations \eqref{eqn:equivalence},
-  and is hence an algebraic variety. 
-  In fact, the Kalman decomposition \plr{XXX above} provides us with an explicit characterization of the set.
-  Above we argued that the population mean set of coefficients $A$ was subject to genetic drift, 
-  with each entry $A_{ij}$ changing with mean square displacement $\sigma^2/\sqrt{N_e}$ per generation.
-  Since selection constrains the population mean to stay near to the set of optimal networks $\calA$,
-  to a good approximation, the population mean moves as Brownian motion in the space of matrices $\R^{n \times n}$
-  but conditioned to stay on the optimal set.
-  This Brownian motion on $\calA$ is a stochastic process driven by the Laplace-Beltrami operator on $\calA$
-  where this makes sense, with special behavior at the singular points (if any \plr{XXX}).
-  Unlike Brownian motion in flat space, this stochastic process can have \emph{bias} --
-  for instance, it is pushed away from regions of negative intrinsic curvature
-  (because there is ``less space'' in such regions).
-
-  \plr{NOT TRUE?:}
-  However, the process always has the property that
-  changes in $A$ accumulate at constant (mean squared) rate:
-  if $\calA$ is locally isomorphic to $\R^m$ around $A_0$
-  (there are $m$ degrees of freedom in the solutions to \eqref{eqn:equivalence}) then
-  \begin{align}
-      \E\left[ \|A_t - A_0\|^2 \right] = m \sigma^2 T + O(T^2) .
-  \end{align}
-
-  %\begin{align}
-  %  \Delta_{\tau} &= \left\lVert \frac{d}{d \tau} \text{vec} \left[A\left(\tau\right)\right] \right\rVert^{-1} \\
-  %  A(\tau) &\xrightarrow{\mu} A(\tau \pm \mu \Delta_{\tau})
-  %\end{align}
- % \section*{Phenotypic Invariance}
-
-%    \begin{example}[Not all minimal gene networks can drift]
- %     If a gene network is minimal and all the molecular species involved in the network are under selection, such that $C$ is the $n \times n$ identity matrix  ($C = I_{n}$), the only acceptable  change of coordinate matrix is the identify matrix.
-
-   %  \begin{align}
-   %     C  = I \lor B = I \\
-   %     IV^{-1} &= I \\
-   %     \iff V &= I
-   %   \end{align}
- %     \end{example}
-  %    \begin{example}[All Non-Minimal Gene Networks can Drift]
-  %    
-  %    Despite the existence of a unique genetic architecture in the minimal case, there still exists an infinite number of systems with larger networks that have identical external dynamics. 
-  %    \begin{align}
-  %      h(t) &= \widehat{h}(t) \iff \\ 
-  %      CA^{\jmath}B &= \widehat{C} \widehat{A}^{\jmath} \widehat{C} \qquad \text{for } \jmath = 0, 1, \dots
-  %    \end{align}
-  %    Any two systems with equivalent impulse responses will have equivalent phenotypes.
-  %    \end{example}
-
-      \plr{add Kalman decomposition stuff here}
-
-      \subsection*{Speciation via Reproductive Incompatibility}
-
-      %Define reproduction in diploids as first, the recombination of unlinked genes to make gametes, and second, as the averaging of two individual parental gametes to produce an offspring.
-Assuming parental populations are both phenotypically identical and genetically homogenous within each population, first generation hybrids (F1s) can be computed by averaging the two parental gene networks.
-Second generation hybrids (F2s) can be computed by first swapping the genes between the two parental gene networks, and next averaging these hybrid gametes.
-
-     An F1's genetic regulation is a consequence of both of its genomes, where genes and regulatory sequences from both parents are equally present.
-Thus we say that a diploid organism's gene network is simply the average of both of its gene network copies; one from each parent.
-Further, each haploid parental gene network copy is formed via meiosis -- by swapping independent genes (specifically their \emph{cis}-regulatory modules) randomly.
-Assuming two distinct but genetically homogeneous populations evolving in allopatry meet and form hybrids, the first generation hybrids (F1s) gene network dynamics will be determined by the average of the parental haplotypes.
-The second generation hybrids (F2s), however, will be the product of a meiosis between parental haplotypes followed by an averaging of gametes.
-
-      Specifically, $F_{1}$ is the first generation hybrid gene network architecture formed by mating (averaging) $A(\tau)$ and $A(\hat{\tau})$, 
-      \begin{align}
-        F_{1}(\tau, \hat{\tau}) &= \frac{ A(\tau) + A(\hat{\tau}) }{2}  .
-     %   h(t, F_{1}) &= C e^{\left( F_{1}\left( \tau, \hat{\tau} \right) t \right)} B
-      \end{align}
-
-      %\begin{align*}
-       % G\left(r, \tau, \hat{\tau}\right) = Q(r) A( \tau ) + \left( I - Q(r) \right) A( \hat{\tau} )
-      %\end{align*}
-      and $F_{2}$ is the second generation hybrid gene network architecture formed by gametes $G(i)$ and $G(j)$.
-Where each $G$ is formed by randomly swapping rows between $A(\tau)$ and $A(\hat{\tau})$, such that the $i$th  gene comes from $A(\tau)$ ($i$ and $j$ are orthogonal vectors, each element $0$ or $1$, and $i \neq j$).
-      \begin{align}
-        F_{2}(i, j) = \frac{G (i) + G(j)}{2}
-      \end{align}
-
-      %\begin{align*}
-      %  h\left(t, F_{2}\left( r, r' \right)\right) = C e^{\left( F_{2}\left(r, r' \right) t \right)} B
-      %\end{align*}
-
-      The fitness of an organism can be computed by comparing its impulse response with the optimal response,
-      \plr{this is either zero or infinite.  apply a weighting function (see appendix)?}
-      \begin{align}
-        \mathcal{F}\left( \widehat{\phi}(t) \right) = \exp \left\{- \frac{1}{\sigma} \int_{0}^{\infty} \left\lVert \phi(t) - \widehat{\phi}(t) \right\rVert^{2} dt  \right\}  .
-      \end{align}
-      Therefore a hybrid's fitness can be computed by comparing its impulse response with that of its parents.
-
-      \begin{definition}[Reproductive Incompatibility]
-        According to Mayr's \emph{Biological Species Concept}, two populations can be considered different species if they are reproductively isolated, meaning crosses between them produce low fitness offspring.
-
-        We quantitatively describe the degree of incompatibility between two populations $P_{1}$ and $P_{2}$ as
-        \begin{align*}
-          \mathcal{I} = \frac{2 \left\langle \mathcal{F} \left(\phi_{F_{1}}\right) \right\rangle}{\left\langle \mathcal{F} \left(\phi_{P_{1}}\right) \right\rangle +  \left\langle \mathcal{F} \left(\phi_{P_{2}}\right) \right\rangle} ,
-        \end{align*}
-        where angled brackets imply averaging.
-
-        F1s created by crossing phenotypically equivalent oscillators $A(0)$ and $A(2)$ have an $\phi_{F_{1}}(t) = e^{t}$, in contrast to both parents with $\phi_{P_{1}}(t) = \phi_{P_{2}}(t) = \sin(t) + \cos(t)$.
-The hybrid phenotype is significantly different (it does not oscillate and increases infinitely) despite the phenotypic equivalence of the parents.
-
-        \begin{align*}
-          \mathcal{F} \left(\phi_{P_{1}}\right) = \mathcal{F} \left( \phi_{P_{2}} \right) &= 1 \\
-          \mathcal{F}\left(\phi_{F_{1}}\right) &= 0 \\
-          \mathcal{I} &= 0
-        \end{align*}
-
-        Thus if populations $1$ and $2$ are homogeneous $A(0)$ and $A(2)$, respectively, we say that they are completely incompatible as $\mathcal{I} = 0$.
-      \end{definition}
-      \begin{figure}[H]
-        \centering
-        \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{expF1}
-        \caption{F1 hybrid (orange) and parental (blue) phenotypic oscillator dynamics for an $A(0)$ by $A(2)$ cross. The hybrid fails to oscillate and exhibits qualitatively different dynamics.}
-      \end{figure}
- %     \begin{example}[F1 Reproductive Incompatibility in an Oscillating Gene Network]
- %       DMI examples...
- %       \begin{figure}[H]
- %         \includegraphics[width=0.5\textwidth]{A2_A3_osci_F1_hyb}
- %       \end{figure}
- %     \end{example}
-
-      \begin{example}[Hybrid Incompatibility in an Oscillating Gene Network]
-
-        Here we compare the phenotypes for F2 hybrids formed by crossing oscillators $A(2)$ with $A(2.01)$, $A(2.1)$, and $A(2.5)$ ($B$ and $C$ are the same as above).
-Each $A$ has is phenotypically identical ($\phi(t) = \sin(t) + \cos(t)$), however some of the hybrids exhibit markedly different dynamics.
-These differences tend to increase with time, quadratically ($x^{2}$) in F2s, and quartically ($x^{4}$) in F1s.
-
-%        \begin{align*}
- %         A(2) = \begin{bmatrix} 2 & -1 \\ 5 & -2 \end{bmatrix} &\qquad A(2.01) = \begin{bmatrix} 1.9901 & -0.9901 \\ 5.0101 & -1.9901 \end{bmatrix} \\
-  %          A(2.1) = \begin{bmatrix} 1.9091 & -0.9091 \\ 5.1091 & -1.9091 \end{bmatrix} &\qquad A(2.5) = \begin{bmatrix} 1.6667 & -0.6667 \\ 5.6667 & -1.6667 \end{bmatrix}
-   %     \end{align*}
-
-        \begin{alignat*}{2}
-          A(2) &= \left[\begin{array}{cc} 2 & -1 \\[6pt] 5 & -2 \end{array}\right] &&\qquad A(2.01) = \left[\begin{array}{cc} 2 -\frac{1}{101} & -1 + \frac{1}{101} \\[6pt] 5 + \frac{1}{99} & -2 + \frac{1}{101} \end{array}\right] \\
-            A(2.1) &= \left[\begin{array}{cc} 2 - \frac{1}{11} & -1 + \frac{1}{11} \\[6pt] 5 + \frac{6}{55} & -2 + \frac{1}{11} \end{array}\right] &&\qquad A(2.5) = \left[\begin{array}{cc} 2 - \frac{1}{3} & -1 + \frac{1}{3} \\[6pt] 5 + \frac{2}{3} & -2 + \frac{1}{3} \end{array}\right]
-        \end{alignat*}
-        F1 gene regulatory networks are formed by averaging the two parental $A(\tau)$ matrices; F2s are formed by first recombining parental matrices followed by an averaging. 
-  %      \begin{figure}[H]
-  %        \begin{tabular}{cc}
-  %          \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_1}
-  %          & \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_2}
-  %         \\ \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_3}
-  %          & \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_4}
-  %          \\ \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_5}
-  %          & \includegraphics[width=0.2\textwidth]{A2_A3_osci_F2_hyb_6}
-  %        \end{tabular}
-  %      \end{figure}
-
-  %    \begin{figure}[H]
-  %      \centering
-  %      \includegraphics[width=0.5\textwidth]{A2_A2-1_F2s}
-  %      \caption{F2s from $A(2)$ and $A(2.1)$.}
-  %    \end{figure}
-  %      \begin{figure}[H]
-  %        \begin{tabular}{cc}
-  %        \includegraphics[width=0.5\textwidth]{F2s-small} \\
-  %        \includegraphics[width=0.5\textwidth]{F2s} \\
-  %        \includegraphics[width=0.5\textwidth]{F2s-large}
-  %        \end{tabular}
-  %        \caption{F2s from $A(2)$ and $A(2.1)$.}
-  %      \end{figure}
-        \begin{figure}[H]
-          \centering
-          \begin{tabular}{cc}
-            \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{F1_comparison} & 
-            \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{F2s_comparison2}
-          \end{tabular}
-          \caption{F1 (left) and F2 (right) hybrids crossing $A(2)$ with $A(2.01)$ (top), $A(2.1)$ (middle), and $A(2.5)$ (bottom).
-          Note the difference in scale on the $y$-axis.
-F2 hybrids display more phenotypic divergence than F1s, on average.
-Further, some F2s completely fail to oscillate, as seen in an $A(2.5)$ F2 (light blue).}
-        \end{figure}
-
-        \jss{$A(2)$ and $A(2.1)$ differ in regulatory interaction strengths by $5.09\%$.
-If intra-population regulatory variation is approximately $5\%$, then this level of divergence is expected after $\sqrt{\frac{t}{N_{e}} = 1}$ generations.}
-    \end{example}
- %     \begin{example}[Not all Networks can Host Incompatibilities]
- %       convex sets cant have DMIs
-
- %       \begin{align*}
- %         h(t) = 2 e^{- \theta t}
- %       \end{align*}
- %       Any non-minimal system with rows summing to $\theta$ is PI.
-Further, these systems are closed under averaging (mating) and row swapping (meiosis), leaving all hybrids optimally fit.
-The set of gene matrices is affine and therefore convex.
- %     \end{example}
-
-\section*{The rate of speciation under neutrality in allopatric populations}
-   In general terms, if fitness $\mathcal{F}$ is some function $g$ of phenotypic divergence from an optimal scaled by the strength of selection $\beta$,
-   \begin{align*}
-      \mathcal{F} = g \left( \frac{\phi - \phi_{0}}{\beta} \right)
-    \end{align*}
-    And a population $P$ is normally distributed around a mean gene network architecture $A(\tau)$ and $\sigma^{2}$ is the intra-population regulatory variance,
-    \begin{align*}
-      P_{T} \sim \mathcal{N} \left(A(\tau), T \frac{\sigma^{2}}{N_{e}}\right)
-    \end{align*}
-    the population mean architecture will move at rate
-    \begin{align*}
-      \Delta \tau = \sigma \sqrt{\frac{T}{N_{e}}}
-    \end{align*}
-    where $T$ is time measured in generations. 
-    USING GENERAL MATH FROM MULTIVARIATE TAYLOR SERIES....
-we know that the phenotype of a first generation $F_{1}$ hybrid diverges from that of its parents (assuming both parents are at an optimum and phenotypically equivalent) quartically as a function of $\Delta \tau$.
-    Thus the squared difference of it's phenotype will be,
-    \begin{align*}
-      \Delta \Phi^{2}_{1} = c_{1}^{2} \sigma^{4} \left( \frac{T}{N_{e}} \right)^{2}
-    \end{align*}
-    and $F_{1}$ fitness will decline linearly with respect to $\frac{T}{N_{e}}$. ARGUMENT FOR WHY $\sigma^{2} c_{1} \sim \beta$ HERE.
-       \begin{align*}
-         \Delta \mathcal{F}_{1} = \left( \frac{\Delta \Phi_{1}}{\beta} = \frac{c_{1} \sigma^{2} \frac{T}{N_{e}}}{\beta} \approx \frac{T}{N_{e}} \right)
-    \end{align*}
-     It follows FROM THE SAME MATH ABOUT THE MULTIVAR TAYLOR that the phenotypes of second generation $F_{2}$ hybrid crosses diverges quadratically as a function of $\Delta \tau$, 
-    \begin{align*}
-      \Delta \Phi^{2}_{2} &= c_{2}^{2} \left( \sigma \sqrt{\frac{T}{N_{e}}} \right)^{2} = c_{2}^{2} \frac{T}{N_{e}} \\
-      \Delta \mathcal{F}_{2} &= \left( \frac{\Delta \Phi_{2}}{\beta} = \frac{c_{2} \sigma \sqrt{\frac{T}{N_{e}}}}{\beta} \approx \sqrt{\frac{T}{N_{e}}} \right)
-    \end{align*}
-    and that $F_{2}$ fitness drops rapidly -- at approximately rate $\sqrt{\frac{T}{N_{e}}}$. ARGUMENT FOR WHY $c_{2} \sigma \sim \beta$ HERE.
-
-    This suggests that reproductively isolated populations can speciate rapidly -- on timescales much less than $N_{e}$ generations, depending on the specifics of the fitness function.
-
-  \begin{example}[Metabolic network]
-
-  Consider an organism that can metabolize two different sugars $s_{1}$ and $s_{2}$ (present at logarithmic concentrations $u_{1}$ and $u_{2}$ respectively), with enzymes $e_{1}$ and $e_{2}$ (with log concentration denoted as $\phi_{1}$ and $\phi_{2}$).
-Further suppose that one of the sugars $s_{2}$ is the preferred energy source (perhaps it contains significantly more energy than the other sugar or is otherwise more efficiently metabolized).
-The organism may have a gene regulatory network $\mathcal{S}$ that can deploy a situation specific metabolic strategy.
-That is depending on both $u_{2}$ and $u_{1}$ the organism will synthesize an appropriate $\phi_{1}$ and $\phi_{2}$.
-Furthermore, consider this system to contain at least two transcription factors, whose log concentrations are given by $\kappa_{1}$, $\kappa_{2}$, $\dots \kappa_{n}$.
-
-  Minimally such a system may have the architecture,
-  \begin{align*}
-    \mathcal{S}_{\min} = \left\{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 0 & -1 \\ 0 & 0 \end{bmatrix} \begin{bmatrix} \kappa_{1} \\ \kappa_{2} \end{bmatrix} (t) + \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \begin{bmatrix} u_{1} \\ u_{2} \end{bmatrix} (t) \\[11pt]
-      \phi(t) &= \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \begin{bmatrix} \kappa_{1} \\ \kappa_{2} \end{bmatrix} (t) \end{array} \right.
-  \end{align*}
-    $\mathcal{S}_{\min}$ is minimal as its reachability and observability matrices both have rank $=2$, $\text{rank}(\mathcal{R}) = \text{rank}(\mathcal{O}) = 2 = n \implies \min$.
-
-  The impulse response matrix of this system is, 
-  \begin{align*}
-    h(t) = \begin{bmatrix} 1 & - t \\ 0 & 1 \end{bmatrix}  .
-  \end{align*}
-
-  The phenotype is, 
-  \begin{align*}
-    \phi(t) = \exp \left\{ \kappa(0) + \int_{0}^{t} h(t) u(t) dt \right\} 
-  \end{align*}
-  ($\kappa(0)$ can be set to something like $\left[ -10, -10 \right]$, assuming the transient transcription factor and enzyme concentrations in the organism are typically quite low).
-
-  We can see that changing coordinates on this system (with any invertible $V \in \mathbb{R}^{2 \times 2}$) will find new system architectures, as shown before, however, we can also apply the Kalman decomposition to find systems in different dimensions -- that is systems that employ more than $2$ transcription factors, yet have the same external dynamics/phenotype.
-This may happen if a system is co-opted from another, or may be a consequence of gene duplication and deletion.
-
-  $\widehat{\mathcal{S}}$ is $4$-dimensional biological system,  \begin{align*}
-    \widehat{\mathcal{S}} = \left\{ \begin{array}{ll}
-      \widehat{A} = V \begin{bmatrix} X_{1} & X_{2} \\ 0 & A_{\min} \end{bmatrix} V^{-1}; \quad \widehat{B} = V \begin{bmatrix} X_{3} \\ B_{\min} \end{bmatrix}; \quad \widehat{C} = \begin{bmatrix} 0 & C_{\min} \end{bmatrix} V^{-1} \end{array} \right\}
-  \end{align*}
-     where $V$ is any invertible $4 \times 4$ matrix and $X_{j}$ is any $2 \times 2$ matrix.
-
-     So for example, some system can be wired as follows, and still be input-output equivalent to the minimal metabolic system $\mathcal{S}_{\min}$:
-     \begin{align*}
-       A' &= \begin{bmatrix}
-         1.6923  &  1.5385 &  -2.6154 &  -2 \\
-         0.8462  &  0.7692 &  -1.3077 &  -1 \\
-         1.2692  &  1.1538 &  -1.9615 &  -1.5 \\
-         0.4231  &  0.3846 &  -0.6538 &  -0.5
-       \end{bmatrix} \\
-       B' &= \begin{bmatrix} 4 & 4 \\ 2 & 2 \\ 3 & 3 \\ 1 & 3 \end{bmatrix} \\
-         C' &= \begin{bmatrix}  0.2692  &  0.1538  &  0.0385 &  -0.5 \\
-         -0.4231 &  -0.3846 &   0.6538  &  0.5 \end{bmatrix}
-     \end{align*}
-
-     Despite the present example consisting of a minimal $2 \times 2$ and a non-minimal $4 \times 4$ system, any $n$-dimensional system can be constructed using this method -- applying a change of coordinates to the Kalman decomposition -- to construct a mechanistically different system with identical phenotypic dynamics.
-Depending on the specifics of a system being modeled, one may have to take care to restrict the free parameter values and network architectures to be biologically appropriate.
-  \end{example}
-%    \subsection*{Metabolic Network}
-%  \begin{align*}
-%    \dot{x}(t) &= \begin{bmatrix} 0 & 1 & 0 & -1 \\ 0 & 0 & -1 & -1 \\ -1 & 1 & 0 & 0 \\ 0 & 0 & 0 & 0 \end{bmatrix} \begin{bmatrix} \text{GAL3} \\ \text{GAL4} \\ \text{GAL80} \\ \text{MIG1} \end{bmatrix}(t) \\ &+ \begin{bmatrix} 1 & 0 & 0 \\ 0 & 0 & 1 \\ 0 & 0 & 0 \\ 0 & 1 & 1 \end{bmatrix} \begin{bmatrix} \text{galactose} \\ \text{glucose} \\ \text{BTM} \end{bmatrix}(t) \\
-%        y(t) &= \begin{bmatrix} 0 & 1 & 0 & -1 \end{bmatrix} \vec{x}(t)
-%    \end{align*}
-%    \begin{align*}
-%      H(z) &= \begin{bmatrix} 1 \\ z^{2} + z + 1 \\ z + 1 \end{bmatrix} \frac{1}{z^{3} + z -1}
-%    \end{align*}
-%
-%    \begin{align*}
-%      CV &= C \\
-%      VB &= B \\
-%      V :&= \begin{bmatrix} 1 & 0 & a & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & b & 0 \\ 0 & 0 & 0 & 1 \end{bmatrix} , V^{-1} = \begin{bmatrix} 1 & 0 & \frac{-a}{b} & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & \frac{1}{b} & 0 \\ 0 & 0 & 0 & 1\end{bmatrix} \\
-%        A(a,b) :&= VAV^{-1} = \begin{bmatrix} -a & a+1 & \frac{a^{2}}{b} & -1 \\ 0 & 0 & \frac{-1}{b} & -1 \\ -b & b & a & 0 \\ 0 & 0 & 0 & 0 \end{bmatrix}
-%    \end{align*}
-%    The GAL regulatory network in \emph{S.
-cerevisiae} is modelled here.
-The activating transcription factor GAL4, which is regulated by the presence of galactose and two other TFs, binds to the promoter region of the GAL regulon -- a cluster of genes regulated by the same promoter encoding 3 enzymes (GAL1, GAL7, and GAL10), for the metabolism of the sugar galactose.
-The repressing transcription factor MIG1, which is activated by the presence of glucose and other proteins omitted here, also binds to the regulon, preventing the expression of the galactose metabolizing enzymes.
-This network is one of the most studied gene regulatory networks in yeast, and experiments have already demonstrated significant transcriptional variation among different species and genuses of yeast.
-%
-%    In the present model, the $A$ matrix encodes the interactions among GAL3, GAL4, GAL80, and MIG1.
-$B$ interprets the input: the quantity of glucose, galactose, and the basal transcription (BTM) rates impact of TF concentrations.
-$C$ represents the promoter region of the GAL regulon and is the sum of the influence of GAL4 (activating) and MIG1 (repressing).
-%
-%    $A(a,b)$ represent alternative network structures that produce identical outputs, with identical input and output transformations -- meaning the regulatory contributions of the sugars and the basasl machinery, as well as the promoter of the GAL regulon are held constant.
-These interactions may be realized in nature, if the regulatory differences predicted by $A(a,b)$ are biologically/mutationally realizable.
-Mutations influencing the interactions between GAL4 and GAL80 have been experimentally demonstrated \citep{li2010alterations}.
-%
-%    \subsection*{Gap Gene Network}
-
-    \section*{Discussion}
-      \plr{mention $B$ part of genetic architecture}
-
-      \jss{Discussion guidelines:
-        \begin{itemize}
-        \item Why is this important/useful?
-        \item What are the assumptions and shortcomings of the research?
-        \item Compare to other studies in the literature. 
-        \item Future directions. 
-        \item Wild speculations?
-        \item Conclusion and overall impact. 
-      \end{itemize}}
-
-  The complexity of biological systems has limited our understanding of their function and evolution.
-Above we outline an approach, a first step, towards untangling this complexity in reference to function and evolution.
-This methodology borrows successfully applied tools from engineering and aims to synthesize these with the concepts and tools of molecular and evolutionary biology.
-
-  Theoretical models in evolution and population genetics often lack the molecular details of physiology or of the genotype-phenotype map.
-Here, we offer a tractable and simple model which includes these missing features.
-Further, we provide, in clear mathematical language, an analytical description of phenomena hitherto only discussed verbally and conceptually (phenogenetic drift \citep{weiss2000phenogenetic}, developmental systems drift \citep{true2001developmental}, biological degeneracy \citep{edelman2001degeneracy}, \emph{etc.}).
-The tractability and relative simplicity of this exposition enables the interested biologist to work out by hand, if desired, the dynamics of a genetic system, as well as perturbations to the system -- an attribute not likely to be found in less tractable models and simulations.
-
-  We have suggested an interpretation of system identification: to see it as an evolutionarily neutral manifold, and not simply a computational nuisance.
-We have demonstrated a method to analytically determine the set of all phenotypically invariant gene networks; by a simple change of coordinates in the minimal configuration, or more generally by applying the Kalman decomposition in higher dimensions.
-Further, we emphasize that evolution proceeds through this high dimensional space as stochastic coordinate transformation, constrained by sexual reproduction and selection.
-This set is explored over evolutionary time when phenotype is conserved, and can lead to a diverse set of consequences, including the accumulation of Dobzhansky-Muller incompatibilities.
-We emphasize that these incompatibilities are a consequence of recombining different, yet functionally equivalent, mechanisms.
-
-  Furthermore, using a quantitative genetic approach, we estimated that a genetically variable population will drift in neutral system space at a rate determined by its intra-population variation and its effective population size.
-Because mechanistically distinct yet phenotypically equivalent biological systems can fail to produce viable hybrids, we predict allopatric populations to speciate at a rate on the order of $N_{e}$ under reasonable population genetic parameter estimates.
-Additionally we see second-generation hybrid fitness plummet much faster than that of first-generation hybrids.
-This insight is a consequence of combining our mechanistic model with a quantitative genetic one: we observe that $F_{1}$ phenotypes diverge quartically, and $F_{2}$ phenotypes quadratically, with evolutionary time.
-This result is also consistent with Haldane's rule; that if only one hybrid sex is inviable or sterile it is likely the heterogametic sex.
-The consistency comes from gene networks localized to the sex chromosomes functioning as an $F_{2}$ hybrid cross within a diploid $F_{1}$ heterogamete as there is only one sex chromosome.
-
-  We also suggest that gene networks may not always use their components parsimoniously as network size tends to ratchet up in the absence of strong selection against extra parts.
-Although unexplored presently, this phenomena may lead to insights on evolvability and developmental innovation.
-Lastly, we show that hybrid gene networks break down as function of genetic distance, and may, in part, explain broad patterns of reproductive isolation among diverse phyla \citep{roux2016shedding}.
-
-  As Richard Levins opined, models in population biology face a trade-off among precision, realism, and generality \citep{levins1966strategy}.
-As Levins expects, any tractable and general model, such as the present one under discussion, will have limitations.
-Most notable is linearity.
-It is often stated that life is not linear.
-This is often true, however, many of the ideas developed here should be generalizable to nonlinear cases (multi-linear systems, say).
-Further, we see this as a necessary first step in the direction of more life-like nonlinear evolutionary systems theory.
-Depending on an actual biological system's particularities, its (potential) non-linearity, may buffer or exacerbate effects elucidated in this paper, such as the acquisition of Dobzansky-Muller incompatibilities.
-
-  This theoretical framework can easily be applied to other interesting questions in evolutionary biology not tackled presently: such as the evolution of linkage, the necessity of network complexity (does evolution tend towards Rube Goldberg or parsimonious network organization?), evolvability, structure/function inference, and intra-population context dependency of mutational effects, as well as many others.
-
-  \jss{literature comparison:}
-
-  Over the last several years, several different computational approaches have been applied to study reproductive incompatibility and speciation.
-\citet{tulchinsky} simulated the evolution of a transcription factor and its binding site using a thermodynamic model.
-Their simulations suggest that the language by which a transcription factor recognizes its binding site can change, and potentially lead to hybrid incompatibility when allopatric populations employ divergent readout languages.
-This study, despite looking at gene regulation, does not analyze overall gene network architecture -- as we do here -- it only looks at the expression level of a single gene.
-Furthermore, they report reproductive isolation primarily following directional selection for a change in expression levels in each allopatric population; the evidence for reproductive isolation following balancing selection is much weaker.
-Johnson and Porter 2000 did not observe any hybrid fitness declines under stabilizing selection -- only under directional selection.
-  Khatari et al, Tulchinsky et al, and Porter et al, all study hybrid incompatibility from a transcription factor/binding site interaction perspective, not from an overall network architecture perspective.
-Palmer and Feldman only see hybrid incompatibility in constant environments if the parental populations are relatively poorly adapted initially.
-Otherwise hybrids between two allopatric populations have fairly high fitnesses.
-
-\section*{Acknowledgements}
-We would like to thank Sergey Nuzhdin, Stevan Arnold, Erik Lundgren, and Hossein Asgharian for valuable discussion.
-
-\bibliographystyle{plainnat}
-\bibliography{krefs}
-%\end{multicols}
-
-\normalsize
-\appendix
-
-\section{Kalman Decomposition}
+\section{Kalman Decomposition} \label{apx:kalman}
 \begin{definition}[Phenotypic equivalence of systems]
@@
 
-\section{Genetic drift with a multivariate trait}
-\label{ss:quant_gen}
+\subsection{Neutral directions from the Kalman decomposition}
 
-For completeness, we provide a brief argument of how the population mean
-moves under genetic drift
-with a quantitative genetics model,
-as in \citet{lande1981models} or \citet{hansen1996translating}.
-These ignore details of the underlying genetic basis,
-but developing a more accurate model is beyond the scope of this paper.
+The Kalman decomposition above says that any system $(A,B,C)$ can be decomposed into
+$(P, \hat A, \hat B, \hat C)$ so that
+$$\begin{aligned}
+    A &= P^{-1} \hat A P  \\
+    B &= P^{-1} \hat B  \\
+    C &= \hat C P ,
+\end{aligned}$$
+and we know precisely how we can change these to preserve the transfer function:
+\begin{enumerate}
+    \item $P \to P + \epsilon Q$ as long as the result is still invertible,
+    \item $\hat A \to A + \epsilon X$ as long as $X$ is zero in the correct places,
+    \item $\hat B \to B + \epsilon Y$ as long as $Y$ is zero in the correct places,
+    \item $\hat C \to C + \epsilon Y$ as long as $Z$ is zero in the correct places.
+\end{enumerate}
+By taking $\epsilon \to 0$, these tell us the local directions we can move a system $(A,B,C)$ in.
+All statements below are up to first order in $\epsilon$,
+omitting terms of order $\epsilon^2$.
 
-\paragraph{Completing the square}
-First note that 
+First, since $(P + \epsilon Q)^{-1} = P^{-1} + \epsilon P^{-1} Q P^{-1}$,
+modifying $P \to P + \epsilon Q$ changes
+$$\begin{aligned}
+    A 
+        &\to A + \epsilon P^{-1} \hat A Q - \epsilon P^{-1} Q P^{-1} \hat A P \\
+        &= A + \epsilon \left(A P^{-1} Q - P^{-1} Q A\right) , \\
+    B
+        &\to B - \epsilon P^{-1} Q B \\
+    C
+        &\to C + \epsilon C P^{-1} Q .
+\end{aligned}$$
+Since $P$ is invertible and $Q$ can be anything (if $\epsilon$ is small enough),
+this allows changes in the direction of an arbitrary $W$:
+$$\begin{aligned}
+    A 
+        &= A + \epsilon \left(A W - W A\right) , \\
+    B
+        &\to B - \epsilon W B \\
+    C
+        &\to C + \epsilon C W .
+\end{aligned}$$
+
+Then, $\hat A \to A + \epsilon X$  does
+$$\begin{aligned}
+    A \to A + \epsilon P^{-1} X P 
+\end{aligned}$$
+and $\hat B \to B + \epsilon Y$ does
+$$\begin{aligned}
+    B \to B + \epsilon P^{-1} Y
+\end{aligned}$$
+and $\hat C \to C + \epsilon Z$ does
+$$\begin{aligned}
+    C \to C + \epsilon Z P .
+\end{aligned}$$
+These degrees of freedom look like they depend on $P$, 
+which is not unique,
+but for any two choices of $P$ there are corresponding choices of $X$
+that give the same actual change in $A$ (and likewise for $Y$ and $Z$).
+
+Therefore, this gives us an upper bound on the number of degrees of freedom,
+in terms of the dimensions of the blocks in the Kalman decomposition ($n_\ro$ etc)
+and the dimensions of $B$ and $C$ ($n_B$ and $n_C$ respectively):
+namely, for $W$, $X$, $Y$, and $Z$ respectively:
+$$\begin{aligned}
+    n^2 
+    + (n_{\rno} + n_\ro n_\nro + n_\nrno(n_\nrno + n_\nro) + n_\nro^2)
+    + n_B n_\rno
+    + n_C n_\nrno .
+\end{aligned}$$
+However, some of these may be redundant.
+For instance, changing $P$ in the direction of 
+a $Q$ that satisfies both $A P^{-1} Q = P^{-1} Q A$ and $C P^{-1} Q = 0$
+is equivalent to changing $B$ by $Y = QB$.
+
+\section{Meiotic recombination in linear systems}\label{apx:recombination}
+
+  Recombination is performed by taking two analogous system components from $\Sys$ and $\Sys'$ and randomly swapping rows or columns.
+  \emph{E.g.} gamete systems $(A'', B'', C'')$, produced by the diploid $\{\Sys, \Sys'\}$, are formed by recombining (randomly swapping rows or columns) between two, possibly distinct, systems $\Sys = (A,B,C)$ and $\Sys' = (A', B', C')$ such that,
   \begin{align*}
-    (x-y)^T A (x-y)
-    &=
-    x^T A \left( x - 2y \right) + y^T A y ,
+    \Sys'' = \left( \begin{array}{ll}
+    A'' &= MA + (I-M)A' , \\
+    B'' &= MB + (I-M)B' , \\
+    C'' &= CM + C'(I-M)
+    \end{array} \right)
   \end{align*}
-and so
-\begin{align*}
-    (x-y)^T A (x-y) + x^T B x
-    &=
-    x^T (A + B) \left( x - 2 (A + B)^{-1} A y \right) + y^T A y \\
-    &=
-    \left( x - (A + B)^{-1} A y \right)^T
-    (A + B)
-    \left( x - (A + B)^{-1} A y \right)
-    + \text{(terms that don't depend on $x$)} .
-\end{align*}
-Therefore, if $f(x;\Sigma,y)$ is the density of a Gaussian with mean $y$ and covariance matrix $\Sigma$
-then substituting $A=\Sigma^{-1}$ and $B=U^{-1}$ above,
-\begin{align*}
-    \frac{ f(x;\Sigma,y) f(x;U,0) }{\int_x f(z;\Sigma,y) f(z;U,0) dz}
-    &=
-    f(x; (\Sigma^{-1} + U^{-1})^{-1}, (\Sigma^{-1}+U^{-1})^{-1} \Sigma^{-1} y) .
-\end{align*}
-
-Now suppose that the population is distributed in genotype space
-as a Gaussian with covariance matrix $\Sigma$ and mean $y$.
-Selection has the effect of multiplying this density by the fitness function and renormalizing,
-so that if expected fitness of $x$ is proportional to $f(x;U,z)$
-then the above argument shows that the next generation will be sampled from a Gaussian distribution
-with covariance matrix $(\Sigma^{-1} + U^{-1})^{-1}$ 
-and mean $z + (\Sigma^{-1}+U^{-1})^{-1} \Sigma^{-1} (y-z)$
-Taking a sample of size $N$ to construct the next generation 
-will produce something close to this but with a slightly (stochastically) deviating mean.
-The next generation's mean is drawn from a Gaussian distribution with mean
-with covariance matrix $(\Sigma^{-1} + U^{-1})^{-1}/N$ 
-and mean $z + (\Sigma^{-1}+U^{-1})^{-1} \Sigma^{-1} (y-z)$.
-
-Roughly, what is this doing?
-Suppose that the population mean differs from the optimum by $\epsilon$,
-that $\Sigma = \sigma^2 I$ and $U = s I$.
-Then the population mean gets closer to the optimum on average, moving to
-$\epsilon/(1 + \sigma^2/s)$
-and adds noise of size $\sqrt{s} \sigma/\sqrt{N \sigma^2 + N s}$.
-At equilibrium, these two movements will be of the same order,
-so that $\epsilon$ is of order $(\sigma/\sqrt{N}) \sqrt{1+\sigma^2/s}$.
-
-\section{Differentiating the fitness function}
-
-Suppose that $\rho(t) \ge 0$ is a weighting function on $[0,\infty)$
-so that fitness is a function of $L^2(\rho)$ distance of the impulse response from optimal.
-With $A_0$ a representative of the optimal set:
-\begin{equation}
-    \begin{aligned}
-        D(A) 
-        &:= 
-        \int_0^\infty \rho(t) \left| h_A(t) - h_{A_0}(t) \right|^2 dt \\
-        &:= 
-        \int_0^\infty \rho(t) \left| C e^{At} B - C e^{A_0 t} B \right|^2 dt \\
-        &= 
-        \int_0^\infty \rho(t) \left| C \left( e^{At} - e^{A_0 t} \right) B \right|^2 dt \\
-        &= 
-        \int_0^\infty \rho(t) C \left( e^{At} - e^{A_0 t} \right) B B^T \left( e^{At} - e^{A_0 t} \right)^T C^T dt
-    \end{aligned}
-\end{equation}
-How does this change with $A$?
-Since
-\begin{equation}
-  \begin{aligned}
-      \frac{d}{du} e^{(A+uZ)t} \vert_{u=0}
-      &=
-      \int_0^t e^{As} Z e^{A(t-s)} ds, 
-  \end{aligned}
-\end{equation}
-we have that
-\begin{equation}
-  \begin{aligned}
-      \frac{d}{du} D(A+uZ)\vert_{u=0}
-      &=
-        2 \int_0^\infty \rho(t) C \left( \int_0^t e^{As} Z e^{A(t-s)} ds \right) B B^T \left( e^{At} - e^{A_0 t} \right)^T C^T dt \\
-      &=
-        2 \int_0^\infty \rho(t) C \left( \int_0^t e^{As} Z e^{A(t-s)} ds \right) B \left( h_A(t) - h_{A_0}(t) \right)^T dt 
-  \end{aligned}
-\end{equation}
-and, by differentiating this and supposing that $A$ is on the optimal set,
-i.e., $h_A(t)=h_{A_0}(t)$, (so wolog $A=A_0$):
-\begin{equation}
-  \begin{aligned}
-      \calH(Y,Z) &:= \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0+uY+vZ)\vert_{u=v=0} \\
-      &=
-        \int_0^\infty \rho(t) C 
-        \left( \int_0^t e^{A_0 s} Y e^{A_0 (t-s)} ds \right) 
-        B B^T 
-        \left( \int_0^t e^{A_0 s} Z e^{A_0 (t-s)} ds \right)^T
-        C^T dt  .
-  \end{aligned}
-\end{equation}
-Here $\calH$ is the quadratic form underlying the Hamiltonian.
-By defining $\Delta_{ij}$ to be the matrix with a 1 in the $(i,j)$th slot
-and 0 elsewhere,
-the coefficients of the quadratic form is
-\begin{equation}
-    \begin{aligned}
-        H_{ij, k\ell}(A)
-        &:=
-        \calH(\Delta_{ij}, \Delta_{k\ell}) .
-    \end{aligned}
-\end{equation}
-
-We could use this to compute the gradient of $D$,
-or to get the quadratic approximation to $D$ near the optimal set.
-To do so, it'd be nice to have a way to compute the inner integral above.
-Suppose that we can diagonalize $A = U \Lambda U^{-1}$.
-Then
-\begin{equation} \label{eqn:exp_deriv}
-  \begin{aligned}
-      \int_0^t e^{As} Z e^{A(t-s)} ds 
-      &=
-      \int_0^t U e^{\Lambda s} U^{-1} Z U e^{\Lambda (t-s)} U^{-1} ds 
-  \end{aligned}
-\end{equation}
-Now, notice that
-\begin{equation}
-  \begin{aligned}
-      \int_0^t e^{s \lambda_i} e^{(t-s) \lambda_j} ds
-      &=
-      \frac{ e^{t \lambda_i} - e^{t \lambda_j} }{ \lambda_i - \lambda_j } .
-  \end{aligned}
-\end{equation}
-Therefore, 
-defining
-\begin{equation}
-    X_{ij}(t,Z) = \left( U^{-1} Z U \right)_{ij}
-      \frac{ e^{t \lambda_i} - e^{t \lambda_j} }{ \lambda_i - \lambda_j } 
-\end{equation}
-moving the $U$ and $U^{-1}$ outside the integral and integrating we get that
-\begin{equation}
-  \begin{aligned}
-      \int_0^t e^{As} Z e^{A(t-s)} ds 
-      &=
-      U X(t,Z) U^{-1} .
-  \end{aligned}
-\end{equation}
-
-Following on from above, we see that if $Z=\Delta_{k \ell}$, then
-\begin{equation}
-  \begin{aligned}
-      X_{ij}^{k\ell}(t) = 
-      \frac{ e^{t \lambda_i} - e^{t \lambda_j} }{ \lambda_i - \lambda_j } 
-      (U^{-1})_{\cdot k} U_{\ell \cdot},
-  \end{aligned}
-\end{equation}
-where $U_{k \cdot}$ is the $k$th row of $U$,
-and so
-\begin{equation}
-    \begin{aligned}
-        H_{ij, k\ell}(A)
-        &=
-        \int_0^\infty
-            \rho(t) C U X^{ij}(t) U^{-1} B B^T (U^{-1})^T X^{k\ell}(t)^T U^T C^T
-        dt .
-    \end{aligned}
-\end{equation}
-This implies that
-\begin{equation}
-    \begin{aligned}
-        D(A_0+\epsilon Z)
-        &\approx \epsilon^2\sum_{ijk\ell} H^{ij,k\ell} Z_{ij} Z_{k\ell} 
-    \end{aligned}
-\end{equation}
-and so
-\begin{equation}
-    \begin{aligned}
-        D(A_0+\epsilon Z)
-        &\approx \epsilon^2\sum_{ijk\ell} H^{ij,k\ell} Z_{ij} Z_{k\ell} 
-    \end{aligned}
-\end{equation}
-
-By section \ref{ss:quant_gen},
-if we set $\Sigma=\sigma^2 I$ and $U=H$,
-then a population at $A_0+Z$ experiences a restoring force of strength
-$(I + \sigma^2 H^{-1})^{-1} Z$ (treating $Z$ as a vector and $H$ as an operator on these).
-If $\sigma^2$ is small compared to $H^{-1}$
-then this is approximately $-\sigma^2 H^{-1} Z$.
-This suggests that the population mean follows an Ornstein-Uhlenbeck process,
-as described (in different terms) in \citet{hansen1996translating}.
-
-\end{document}
+  where $M$ is a diagonal matrix where each diagonal element is a Bernoulli random variable ($M_{ii} = 0$ or $1$ with equal probability, and $M_{ij}=0$ if $i \neq j$).
+If systems are different dimensions the smaller system elements can be augmented with $0$s (\emph{e.g.} $\left[ \begin{smallmatrix} A & 0 \\ 0 & 0 \end{smallmatrix} \right], \left[ \begin{smallmatrix} B \\ 0 \end{smallmatrix} \right], \left[ \begin{smallmatrix} C & 0 \end{smallmatrix}\right]$).

### 2017-12-05 kalman_walk_paper.tex
edited through

@@
 reproductive incompatibility due to system drift will manifest
-\plr{we don't need to summarize the results of the paper in the introduction.  the abstract, and maybe the discussion, but not the introduction.}
 
@@
 depends on the genetic basis of any variation in the coefficients.
+Thanks to the randomness of meiotic segregation,
+the result is random to the extent that each parent is heterozygous
+for alleles that affect the coefficients.
 Since the $i^\text{th}$ row of $A$ summarizes how each gene regulates gene $i$,
@@
 implies that the variance is equal to one-quarter of the squared difference between parental coefficients.
-
-\plr{continue editing here}
 
@@
 How quickly do hybrid phenotypes break down as genetic distance between parents increases?
-To quantify this, we will measure how far a system's phenotype is from optimal
+We will quanitfy how far a system's phenotype is from optimal
 using a weighted difference between impulse response functions.
 Suppose that $\rho(t)$ is a nonnegative, smooth, square-integrable weighting function,
-$h_0(t)$ is the \emph{optimal} impulse response function,
+$h_0(t)$ is the \emph{optimal} impulse response function
 and define the ``distance to optimum'' of another impulse response function
@@
 as the distance between the parents.
-The precise rate at which the phenotype of a hybrid diverges depends on the geometry --
-in Figure \ref{fig:conceptual_fig}, this is depicted as the angle of the black line (the optimal set)
-with respect to the coordinates.
+The precise rate at which the phenotype of a hybrid diverges depends on the geometry
+of the optimal set $\allS$ relative to segregating genetic variation.
+% in Figure \ref{fig:conceptual_fig}, this is depicted as the angle of the black line (the optimal set) with respect to the coordinates.
 
@@
 
-The coefficients in $A$ -- i.e., the regulatory coefficients --
-  differ between parents by only a few percent (around 0.5\% for $\epsilon = - \sfrac{1}{100}$ and 5\% for $\epsilon= - \sfrac{1}{10}$).
-This is well within the amount of regulatory coefficient variation we expect to find segregating within real populations
-(discussed further below). 
-For these small values of $\epsilon$, hybrid phenotypes remain relatively stable, 
-  consistent with the idea that natural selection will allow such intrapopulation variation.
+%% PLR: it says this above already.
+% The coefficients in $A$ -- i.e., the regulatory coefficients --
+%   differ between parents by only a few percent (around 0.5\% for $\epsilon = - \sfrac{1}{100}$ and 5\% for $\epsilon= - \sfrac{1}{10}$).
+% This is well within the amount of regulatory coefficient variation we expect to find segregating within real populations
+% (discussed further below). 
+% For these small values of $\epsilon$, hybrid phenotypes remain relatively stable, 
+%   consistent with the idea that natural selection will allow such intrapopulation variation.
 For larger values of $\left| \epsilon \right|$ (here $\sfrac{1}{2}$), hybrid phenotypes can become dysfunctional, and depending on context, manifest as hybrid inviability.
@@
   \caption{
-    \textbf{(left)} Phenotypes of $F_1$ hybrids between an $A(0)$ parent and, top-to-bottom, an $A(-\sfrac{1}{100})$, an $A(-\sfrac{1}{10})$, and $A(-\sfrac{1}{2})$ parent.
+    \textbf{(left)} Phenotypes of $F_1$ hybrids between an $A(0)$ parent and, top-to-bottom, an $A(-\sfrac{1}{100})$, an $A(-\sfrac{1}{10})$, and $A(-\sfrac{1}{2})$ parent;
+    parental coefficients differ by around 0.5\%, 5\%, and 25\% respectively.
+    \plr{These figures are fine; but reparameterize so that the values of $\tau$ are positive.}
     Parental phenotypes ($\sin t + \cos t$) are shown in solid black, and hybrid phenotypes in dashed blue.
-    \textbf{(right)} Phenotypes of all 256 possible $F_2$ hybrids between the same set of parents,
+    \textbf{(right)} Phenotypes of all possible $F_2$ hybrids between the same set of parents,
     with parental phenotype again in black.
+    %% PLR: There are at most 128 possible F2s, not 256; but many are equivalent; 
+    %%   it only matters if the offspring is homo/het/homozygous, so there are 3^4 = 81 possible expression patterns.
     Different colored lines correspond to different $F_2$s;
@@
 $\left( \left[ \begin{smallmatrix} A_1 & A_2 \\ \cdot & \cdot \end{smallmatrix} \right], \left[ \begin{smallmatrix} A_1 & A_2 \\ X_1 & X_2 \end{smallmatrix} \right] \right)$
-  has phenotype determined by $A = \left[ \begin{smallmatrix} A_1 & A_2 \\ X_1 & X_2 \end{smallmatrix} \right]$, thanks to \emph{Drosophila}-like dosage compensation mechanism (the X is twice upregulated in heterogametes relative to homogametes),
+has phenotype determined by $A = \left[ \begin{smallmatrix} A_1 & A_2 \\ X_1 & X_2 \end{smallmatrix} \right]$, 
+if dosage compensation upregulates heterogametes by a factor of 2 relative to homogametes,
+% as in \emph{Drosophila}
 while a female homozygous for the haplotype
@@
 $F_1$ males are effectively equivalent to purely autosomal-system $F_2$s. 
-As such, in this setting, Haldane's rule is mechanistically similar to \emph{hybrid breakdown}, the observation that $F_2$s will often be less fit than $F_1$ hybrid crosses, in the early stages of speciation.
-This mechanism appears to be distinct from the ``dominance'', ``faster-X'', and ``faster-male'' theories \citep{orr1997haldane, coyne1998evolutionary, turelli1995dominance} usually suggested to explain Haldane's rule.
+In this setting, Haldane's rule is mechanistically similar to \emph{hybrid breakdown}, the observation that $F_2$s will often be less fit than $F_1$ hybrid crosses, in the early stages of speciation.
+This mechanism relies on the genetic variation of a system being shared across the sex chromosomes and the autosomes,
+and appears to be distinct from the ``dominance'', ``faster-X'', and ``faster-male'' theories \citep{orr1997haldane, coyne1998evolutionary, turelli1995dominance} usually suggested to explain Haldane's rule.
 %and makes several different predictions than the dominance theory as formulated by \citet{turelli1995dominance}.
@@
 
-Thus far we have shown that many distinct molecular mechanisms can realize identical phenotypes
+Thus far we have seen that many distinct molecular mechanisms can realize identical phenotypes
 and that these mechanisms may fail to produce viable hybrids.
@@
 
-We will suppose that each organism's phenotype is determined by its vector of coefficients, denoted by $x=(x_1, x_2, \ldots, x_L)$,
+We will suppose that each organism's phenotype is determined by its vector of coefficients, denoted by $x=(x_1, x_2, \ldots, x_p)$,
 and that the corresponding fitness is determined by the distance of its phenotype to optimum.
@@
 In the results above, $x = (A,B,C)$ and $D(x)$ is given by equation \eqref{eqn:distance}.
-determines a system (this is a system $(A,B,C)$ above) 
-with fitness determined by its phenotypic distance $D(x)$ from an optimum (this is analogous to $\allS$ above). 
-Concretely, we define the fitness at $x$ to be $\exp(-D(x)^2)$.
+The fitness of an organism with coefficients $x$ will be $\exp(-D(x)^2)$.
 % this assumption does not entail a loss of generality since $D$ is arbitrary.
-We will assume that in the region of interest, the map $D$ is smooth
+We assume that in the region of interest, the map $D$ is smooth
 and that we can locally approximate the optimal set $\allS$ as a quadratic surface.
-As above, an individual's coefficients are given by averaging its parentally inherited coefficients and adding random noise due to segregation. 
-Concretely, we use the \emph{infinitesimal model} for reproduction \citep{barton_infinitesimal} --
+As above, an individual's coefficients are given by averaging its parentally inherited coefficients and adding random noise due to segregation and possibly new mutation.
+Concretely, we use the \emph{infinitesimal model} for reproduction \citep{barton2016infinitesimal} --
 the offspring of parents at $x$ and $x'$ will have coefficients $(x+x')/2 + \varepsilon$,
-where $\varepsilon$ is a Gaussian displacement due to random assortment of parental alleles.
+where $\varepsilon$ is a random Gaussian displacement due to random assortment of parental alleles.
 
@@
 We work with a randomly mating population of effective size $N_e$. 
-If the regulatory coefficient population variation
-has standard deviation $\sigma$ in a particular direction,
+If the population variation has standard deviation $\sigma$ in a particular direction,
 since subsequent generations resample from this diversity,
 the population mean coefficient will move a random distance of size $\sigma/\sqrt{N_e}$ per generation,
-simply because this is the the standard deviation of the mean of a random sample \citep{lande_drift}.
+simply because this is the the standard deviation of the mean of a random sample \citep{lande1981models}.
 % (This could be taken as a definition of $N_e$.)
@@
 depends on constraints imposed by selection and 
-correlations between the genetic variation underlying different coefficients (the $G$ matrix \citep{G_matrix}).
+correlations between the genetic variation underlying different coefficients 
+(the $G$ matrix \citep{arnold2008understanding}).
 It therefore seems reasonable to coarsely model the time evolution of population variation in regulatory coefficients as 
@@
 
+Next, we calculate with some simplifying assumptions to give the general idea;
+multivariate derivations appear in Appendix \ref{ss:quant_gen}.
 There will in general be different amounts of variation in different directions;
@@
 $1/(\frac{d}{du}D(x+uz))$ 
-with respect to $u$ where $x$ is optimal and $z$ is a ``selected'' direction perpendicular to $\allS$.
+where $x$ is optimal and $z$ is a ``selected'' direction perpendicular to $\allS$.
 With these parameters, a typical individual will have a fitness of around $\exp(-(\sigma_S/\gamma)^2)$.
 Of course, there are in general many possible neutral and selected directions;
-we take these values to be representative of the possible directions.
+we take $\gamma$ to be an appropriate average over possible directions.
 
@@
 and so will have mean system coefficients at the midpoint between their means.
-Each $F_2$ hybrid will be homozygous for one parental allele on average at half of the loci in the genome,
-so the distribution of $F_2$s will have mean at the average of the two populations,
+The distribution of $F_2$s will have mean at the average of the two populations,
 but will have higher variance.
@@
 This is suggested by figure \ref{fig:conceptual_fig} and shown in Appendix \ref{apx:seg_cov}.
-\plr{connect to Barton etc polygenic adaptation lit}
 Concretely, we expect the population of $F_1$s to have variance $\sigma^2_S$ in the selected direction
@@
 where $\omega$ is a factor that depends on the genetic basis of the coefficients.
-In a model of $p$ traits in which the optimal set $\allS$ has dimension $q$,
-% if each trait is controlled by a single locus, $\omega = XXX$,
-using the polygenic model of appendix \ref{apx:seg_cov}, $\omega = (p-q)/8$.
+If the optimal set $\allS$ has dimension $q$,
+using the polygenic model of appendix \ref{apx:seg_cov}, 
+$\omega$ is proportional to the number of degrees of freedom: $\omega = (p-q)/8$.
 If each trait is controlled by a single locus, as in figure \ref{fig:conceptual_fig},
@@
 i.e., once $\sqrt{4 \omega T/N_e} \approx \gamma / \sigma_N$.
-This implies that hybrid incompatibility among $F_2$s should appear on a time scale of 
-$N_e (\gamma / \sigma_N)^2 / (4 \omega)$ generations.
+This implies that hybrid incompatibility among $F_2$s should appear much slower --
+on a time scale of $N_e (\gamma / \sigma_N)^2 / (4 \omega)$ generations.
 The $F_1$s will not suffer fitness consequences until the hybrid mean is further than $\gamma$ from the optimum;
@@
 \end{align}
-This assumes a single trait, for simplicity; the multivariate case is done in appendix \ref{apx:gauss_load}.
+This assumes a single trait, for simplicity. %; the multivariate case is done in appendix \ref{apx:gauss_load}.
 A population of $F_2$s will have, as above, variance $\sigma^2 = \sigma^2_S + 4 \omega \sigma^2_N T/N_e$.
-The mean diverges with the square of the distance between the parentals
-with a speed depending on the local geometry of the optimal set
-(calculated in appendix \ref{apx:H_calc}), so we set $\mu = c_\mu \gamma T/N_e$.
+The mean diverges with the square of the distance between the parentals, so we set $\mu = c_\mu \gamma T/N_e$,
+where $c_\mu$ is a constant depending on the local geometry of the optimal set.
 The mean fitness in parental populations is as in equation \ref{eqn:gauss_fit} with $\mu=0$ and $\sigma = \sigma_S$.
@@
 then
-\plr{figure out how dimensions come in here}
 \begin{align} \label{eqn:fit2}
@@
 If each of the $q$ selected directions acts independently,
-the drop in fitness will be $\fit_2(T)^q$
-(the expression for the correlated cases is given in Appendix \ref{apx:gauss_load}).
+the drop in fitness will be $\fit_2(T)^q$;
+the expression for the correlated, multivariate case is given in Appendix \ref{apx:gauss_load}.
 We discuss the implications of this expression in the next section.
@@
 \paragraph{Speciation rates under neutrality}
-Above, equation \eqref{eqn:fit2} shows how fast hybrids become inviable 
+Equation \eqref{eqn:fit2} describes how fast hybrids become inviable 
 as the time that the parental populations are isolated increases;
 what does this tell us about speciation rates under neutrality?
-From equation \eqref{eqn:fit2} we can observe that
+From equation \eqref{eqn:fit2} we observe that
 time is always scaled in units of $N_e$ generations,
@@
 then the rate that $F_2$ fitness drops is, roughly,
-$(1 + 4 (p-q) C T/n_e)^{-q/2} \propto q (p-q)$, where $C$ is a constant.
-This suggests that both degree of constraint and number of available neutral directions
-affect the speed of accumulation of incompatibilities.
+$(1 + 4 (p-q) C T/N_e)^{-q/2} \propto q (p-q)$, where $C$ is a constant.
+Both degree of constraint and number of available neutral directions
+affect the speed of accumulation of incompatibilities --
+more unconstrained directions allows faster system drift,
+but more constrained directions implies greater fitness consequences of hybridization.
 However, note that in real systems, it is likely that $\gamma$ also depends on $p$ and $q$.
-\plr{revisit with sims}
+
+Now we will interpret equation \eqref{eqn:fit2} in three situations plausible for different species,
+depicting how hybrid fitness drops as a function of $T/N_e$ in Figure \ref{fig:speciation_rates}.
+In all cases, the fitness drop for $F_1$ hybrids is much smaller than that of $F_2$s,
+so we work only with the first (square-root) term in equation \eqref{eqn:fit2}.
 
@@
 but since time is scaled by $N_e$, so speciation may occur \emph{faster} than in a large population.
-However, at least in some models \citep{bartonXXX}, in small populations at mutation-drift equilibrium
+However, at least in some models \citep{lynch1986phenotypic}, in small populations at mutation-drift equilibrium
 the amount of genetic variance ($\sigma_N^2$) is proportional to $N_e$,
+\plr{that citation is Lynch \& Hill 1986, "phenotypic evolution by neutral mutation"}
 which would compensate for this difference, 
@@
 % $\sigma_N \gg \sigma_S \approx \gamma$, if $N_e$ is big and there's a lot of recombination load (``species complex'')
-In the other direction, consider large metapopulations (or isolated ``species complexes'')
+In the other direction, consider large metapopulations (or a ``species complex'')
 among which heritable variation is strongly constrained by selection
@@
 
-For instance, in a population of one million organisms that has 10 generations per year (a drosophilid species, perhaps)
+For instance, between two populations of one million organisms that has 10 generations per year (a drosophilid species, perhaps)
 under the ``large population'' scenario of Figure \ref{fig:speciation_rates}A,
@@
 the situation may be similar to that of Figure~\ref{fig:speciation_rates}C with $N_e = 10^6$,
-the same drop could occur after 1,100 years.
-However, hyperdiverse populations of this type may not be stable on these time scales.
+and so the same drop could occur after 1,100 years.
+(However, hyperdiverse populations of this last type may not be stable on these time scales.)
 
@@
 \caption{
-Mean drop if $F_2$ fitness relative to parental species,
+Mean drop of $F_1$ and $F_2$ fitness relative to parental species,
 with $\omega=1$ and 
@@
 \textbf{(C)} $0.1 \sigma_N^2 = \sigma_S^2 = \gamma^2$.
-The total fitness is from equation \eqref{eqn:fit2},
-and is the product of the ``shift of mean'' and ``increase of variance'' terms
-(the exponential and the square root, respectively).
-Note that, as we assume, the contribution of the shift in mean
+The $F_2$ fitness is from equation \eqref{eqn:fit2},
+and the $F_1$ fitness is determined only by the exponential term of that equation.
 is small relative to that of increased variance.
@@
 Variation in binding site occupancy may overestimate variation in $A$,
-since it does not capture buffering effects (if for instance only one site of many needs be occupied for transcription to begin),
-and variation in expression levels measures changes in steady-state concentration (our $\kappa_i$) rather than the \emph{rate} of change.
-Nonetheless,
+since it does not capture buffering effects (if for instance only one site of many needs to be occupied for transcription to begin),
+and variation in expression level measures changes in steady-state concentration (our $\kappa_i$) rather than the \emph{rate} of change.
+Nonetheless, these measures likely give us an idea of the scale of variability.
 \citet{kasowski2010variation} found differential occupancy in 7.5\%
@@
 accounts for around 2--6\% of expression variation in human blood-derived primary cells,
-while \citep{lappalainen2013transcriptome} found that human population variation 
+while \citet{lappalainen2013transcriptome} found that human population variation 
 explained about 3\% of expression variation.
 %%\plr{Get some data from at least one other species in here!}
-Taken together, this suggests that variation in the entries of $\allS$
+Taken together, this suggests that variation in the entries of $A$
 may be on the order of at least a few percent between individuals of a population --
 doubtless varying substantially between species and between genes.
-
-The impact of regulatory variation ($\sigma$) above
-depends only on its magnitude relative to selection ($\gamma$),
-but it seems reasonable that these two quantities are of the same magnitude.
+\plr{JOSH: insert a citation or two to something non-human?}
 
@@
 phenogenetic drift \citep{weiss2000phenogenetic} and developmental system drift \citep{true2001developmental}.
-In this context, the Kalman decomposition \citep{kalman1963mathematical} implies that nearly all systems are nonidentifiable,
-and gives an analytical description of all phenotypically equivalent gene networks.
-System nonidentifiability, rather than being a computational nuisance,
-implies the existence of axes of genetic variation that are not constrained by selection.
+In this context, the Kalman decomposition \citep{kalman1963mathematical}
+gives an analytical description of all phenotypically equivalent gene networks.
+It also implies that nearly all systems are nonidentifiable,
+and hence the general existence of axes of genetic variation that are not constrained by selection.
 The independent movement of separated populations along these axes
@@
 In this model, at biologically reasonable parameter values,
-system drift is a significant -- and often rapid -- driver of speciation.
+system drift is a significant -- and possibly rapid -- driver of speciation.
 This may at first be surprising because
 hybrid inviability appears as a consequence of recombining different, yet functionally equivalent, mechanisms.
-While the genetic incompatibilities contributing to hybrid inviability described in this paper share similarities with those in the commonly discussed Dobzhansky-Muller model, they may in fact share more in common with what has been described as the ``pathway model'' \citep{lindtke2015genetic}.
-
-Consistent with empirical observation of \emph{hybrid breakdown} \citep[e.g.][]{plotner2017chlorosis}), we see that the fitnesses of $F_2$ hybrids drop at a much faster rate than those of $F_1$s,
-since $F_1$ and $F_2$ phenotypes diverge linearly and at the square root of time relative to parentals, respectively.
+
+Consistent with empirical observation of \emph{hybrid breakdown} \citep[e.g.][]{plotner2017chlorosis}, 
+we see that the fitnesses of $F_2$ hybrids drop at a much faster rate than those of $F_1$s.
+% since $F_1$ and $F_2$ phenotypes diverge linearly and at the square root of time relative to parentals, respectively.
 %\plr{refer to Turelli here}
@@
 System drift and network rewiring has been inferred across the tree of life \citep{dalal2017transcription, johnson2017rewiring},
-and there is often significant regulatory variation segregating within populations \citep{XXX}.
+and there is often significant regulatory variation segregating within populations.
 Transcription in hybrids between closely related species with conserved transcriptional patterns can also be divergent 
@@
 may be nearly morphologically indistinguishable.
-\plr{check the skinks have RI!!} \jss{the paper just says highly ``genetically distinct''; nothing is stated about RI}
-%Of course, selective pressures have likely not been constant over evolutionary time
-%in any of these examples,
-%but the process we describe here will still occur.
-%This model therefore may provide a neutral ``null'' model of gene network evolution, a strategy advocated for by \citet{koonin2016splendor}.
-\plr{revisit}
-Of course the basic assumptions of this model will be violated in practice (constant selective pressures, etc.), however, this model can function as a ``neutral null'' description of gene network evolution (a strategy advocated for by \citet{lynch2007frailty, fay2008evaluating, koonin2016splendor}).
-If this model succeeds in describing actual system evolution, it can be inferred that the mechanism underlying species formation does not require the inclusion of adaptation or changes in selection to explain the emergence of hybrid incompatibility.
 
 \paragraph{The origin of species not by means of natural selection?}
-%\plr{(or, "Neutral speciation?")}
 % this is neutral, in contrast to other models. although some folks advocate nonadaptive speciation.
@@
 % but this one may be strong since happens in all networks at once
-%
-%The model we develop here is perhaps a plausible mechanism for strictly \emph{neutral} speciation.
-%This is perhaps surprising, as speciation is often thought to simply be a byproduct of adaptative change \citep{schluter2009evidence, orr2004speciation}
-%%as much of the speciation literature to date
-%%has focused on positive selection and/or genetic conflict,
-%or due to genetic conflict,
-%and since previous authors have argued that neutral processes should be too slow
-%\citep{nei1983models,seehausen2014genomics}.
-%In most cases to date where the precise molecular mechanism underlying hybrid incompatibility is known,
-%the genes responsible are thought to have been subject 
-%to recent positive selection \citep{examples} 
-%and/or genetic conflict \citep{examples}.
-%\plr{get examples from \citet{orr2004speciation}}
-%%However, incompatibility due to more diffuse system breakdown
-%%(e.g., widespread misregulation \citep{coolon2014tempo} \plr{check that} \jss{I don't think this is the right citation})
-%%may be likely much more difficult to uncover.
-%% \citet{orr2004speciation} remarks that hybrid incompatibility data suggests that ``speciation results from positive Darwinian selection within species''. 
-%On the other hand, the Dobzhansky--Muller model of hybrid incompatibility 
-%is agnostic to the origin of the genetic substitutions \citep{coyne1998evolutionary} \plr{cite bateson, etc},
-%and other plausible mechanisms for the origin of Dobzhanzky-Muller incompatibilities by neutral genetic drift
-%have been proposed \citep[e.g.][]{lynch_force}.
 Although, as classically formulated, 
@@
 previous authors have argued that neutral processes are likely too slow to be a significant driver of speciation \citep{nei1983models,seehausen2014genomics}.
-Using simulations, \citet{porter2002speciation} demonstrated the accumulation of hybrid incompatibilities under directional, but not stabilizing selection, and \citet{palmer2009dynamics} observed the appearence of incompatibilities in populations evolving in a constant environment, only if the populations were suboptimally adapted.
+(Also note that the mechanism underlying hybrid inviability here 
+resembles the Dobzhansky-Muller model,
+but is more similar to the ``pathway model'' of \citet{lindtke2015genetic}.)
+Using simulations, \citet{porter2002speciation} demonstrated the accumulation of hybrid incompatibilities under directional, but not stabilizing selection, and \citet{palmer2009dynamics} observed the appearance of incompatibilities in populations evolving in a constant environment, only if the populations were suboptimally adapted.
 This, in light of the few known incompatibilities, has lead some to conclude that hybrid incompatibility is typically a byproduct of positive selection \citep{orr2004speciation, schluter2009evidence} 
-or a conseuqence of genetic conflict \citep{presgraves2010molecular, crespi2013conflictual}.
-In contrast, the model we develop here, suggests that even under strictly neutral conditions, system drift can lead to speciation, perhaps at a rate fast enough to play a role in species formation across the tree of life.
-%In contrast, the model we develop here, is perhaps a plausible mechanism for strictly \emph{neutral} speciation. 
-%
-%In practice it remains unclear whether system drift or other speciation mechanisms predominate, 
-%as hybrid incompatibilities due to both selection and conflict have been uncovered and hint at general processes \citep{orr2004speciation}. 
-%While it is known that divergent selection and genetic conflict, as well as other mechanisms, can precipitate speciation, we have shown here that, surprisingly, neutral drift may also significantly drive species formation.
-Our results show that hybrids, under neutral processes, break down as a function of genetic distance, and despite the present model's myriad simplifications and assumptions, 
-it may, in part, explain the observed broad patterns of reproductive isolation, such as the observed consistent relationship between molecular divergence and genetic isolation across diverse taxa and ecologies by \citet{roux2016shedding}, and the suggestivley nonadaptive, clocklike speciation patterns observed by \citet{hedges2015tree} (as diversification is not observed to be accelerated following mass extinctions).
-While the incompatibility of any one given network within an organism may be small, we note that organisms are made up of many different networks, and that the cumulative impact of system drift across all of these may be substantial.
-Further, even in populations experiencing directional selection, networks underlying conserved parts of the phenotype can still experience neutral drift and harbor incompatibilities.
+or a consequence of genetic conflict \citep{presgraves2010molecular, crespi2013conflictual}.
+In contrast, the model we develop here suggests that even under strictly neutral conditions, 
+system drift can lead to speciation, at a rate fast enough 
+to play a substantial role in species formation across the tree of life. 
+Our results show that hybrids, under neutral processes, break down as a function of genetic distance,
+and so may, in part, explain broad patterns
+such as the relationship between molecular divergence and genetic isolation 
+% across diverse taxa and ecologies 
+seen by \citet{roux2016shedding}, 
+and the clocklike speciation patterns observed by \citet{hedges2015tree}.
+% (as diversification is not observed to be accelerated following mass extinctions).
+
+All of these forces -- adaptive shifts, conflict, and network drift --
+are plausible drivers of speciation, and may even interact.
+For instance, reinforcement due to local adaptation could in some situations provide a selective pressure
+that speeds up system drift.
+It remains to be seen how the relative strengths of these forces compare.
+Furthermore,
+while the fitness consequences of incompatibility in any one given network may be small, 
+the cumulative impact of system drift across the many different networks an organism relies on may be substantial.
+
+Of course the basic assumptions of this model will be violated in practice (constant selective pressures, etc.), however, this model can function as a ``neutral null'' description of gene network evolution (a strategy advocated for by \citet{lynch2007frailty, fay2008evaluating, koonin2016splendor}).
+If this model succeeds in describing actual system evolution, it can be inferred that the mechanism underlying species formation does not require the inclusion of adaptation or changes in selection to explain the emergence of hybrid incompatibility.
 
@@
 Most notably, physiological limits put upper bounds on expression levels,
-implying saturating response curves \citep{michalis_mentin}.
+implying saturating response curves.
 It remains to be seen how well these results carry over into real systems,
 but the fact that most nonlinear systems can be locally approximated by a linear one
-suggests qualitatively similar results.
+suggests our qualitative results may hold more generally.
 Furthermore, nonidentifiability (which implies the existence of neutral directions)
-is often found in practice in moderately complex models of biological systems. \plr{need citation}
+is often found in practice in moderately complex models of biological systems. 
+\plr{Josh, add citation to sensitivity analysis}
 
 Finally, despite our model's precise separation of phenotype and kryptotype, this relationship in nature may be far more complicated as aspects of the kryptotype may be less ``hidden'' than we currently assume.
-For instance, attributes excluded from the phenotype as modelled here, ignore the potential energy costs associated with execessively large (non-minimal) kryptotypes, as well as the relationship between a specific network architecture and mutational robustness, as well as many other presently neglected phenomena.
-Despite the model's simplifications and potentially nontrivial oversights, it may still roughly reflect and shed light upon speciation patterns as they occur in nature.
-
-%\paragraph{Peter: Quant gen discussion}
+For instance, attributes excluded from the phenotype as modelled here, ignore the potential energy costs associated with excessively large (non-minimal) kryptotypes, as well as the relationship between a specific network architecture and robustness to mutational, transcriptional, or environmental noise.
+% Despite the model's simplifications and potentially nontrivial oversights, it may still roughly reflect and shed light upon speciation patterns as they occur in nature.
+More precise modeling will require better mechanistic understanding not only of biological systems,
+but also the nature of selective pressures
+and genetic variation in these systems.
+
+\paragraph{Peter to add quant gen discussion} about non-normality.
+
 %Maybe compare to \citet{fierst}?
@@
 %hominids; neanderthal load.
-%
+
 \section*{Acknowledgements}
 We would like to thank Sergey Nuzhdin, Stevan Arnold, Erik Lundgren, and Hossein Asgharian for valuable discussion.
+
+\plr{Josh: please fix capitalization in the references.  To do this, anything that needs to be capitalized gets put in curly braces, which tells bibtex to render the capitalization 'as-is'.}
 
@@
 \normalsize
+\appendix
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Local expansion of the fitness surface}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Away from the optimum}
+\label{apx:away_from_opt}
+
+Let two points on $\allS$ be $x_1$ and $x_2$, let $\bar x = (x_1+x_2)/2$, and let $z=(x_2 - x_1)/2$.
+Then with $\grad D$ and $\grad^2 D$ the first and second derivatives of $D$, respectively,
+Taylor expanding about $x_1$ and $x_2$ finds that
+\begin{align*}
+    D(\bar x) 
+    &= D(x_1) + \grad D(x_1) \cdot z + \frac{1}{2} z^T \grad^2  D(x_1) z + O(\|z\|^3) \\
+    &=  D(x_2) - \grad D(x_2) \cdot z + \frac{1}{2} z^T \grad^2  D(x_2) z + O(\|z\|^3) .
+\end{align*}
+Now, since $ D(x_1) =  D(x_2) = \grad D(x_1) = \grad D(x_2) = 0$ and
+\begin{align*}
+    \grad D(x_2) = \grad D(x_1) + 2 z^T \grad^2  D(x_1) + O(\|z\|^2), \quad \text{and} \\
+    \grad^2 D(x_2) = \grad^2 D(x_1) + O(\|z\|), 
+\end{align*}
+adding together the two equations above and dividing by two gets that
+\begin{align*}
+     D(\bar x) 
+    &= \optph - \frac{3}{2} z^T \grad^2  D(x_1) z + O(\|z\|^3) .
+\end{align*}
+
+\subsection{Differentiating the fitness function}
+\label{apx:H_calc}
+
+Suppose that $\rho(t) \ge 0$ is a weighting function on $[0,\infty)$
+so that fitness is a function of $L^2(\rho)$ distance of the impulse response from optimal.
+With $h_0(t) = C_0 e^{tA_0} B_0$ a representative of the optimal set:
+\begin{equation}
+    \begin{aligned}
+        D(A, B, C) 
+        &:= 
+        \int_0^\infty \rho(t) \left| h_A(t) - h_0(t) \right|^2 dt \\
+        &:= 
+        \int_0^\infty \rho(t) \left| C e^{At} B - C_0 e^{A_0 t} B_0 \right|^2 dt \\
+        &= 
+        \int_0^\infty \rho(t) \tr\left\{
+            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)^T
+            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)
+        \right\} dt \\
+        &= 
+        \int_0^\infty \rho(t) \tr\left\{
+            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)
+            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)^T
+        \right\} dt  .
+        % &= 
+        % \int_0^\infty \rho(t)  \tr\left\{
+        %         C e^{At} B B^T e^{A^T t} C^T 
+        %         - C_0 e^{A_0t} B_0 B^T e^{A^T t} C^T 
+        %         \right. \\ &\qquad \qquad \left. {}
+        %         - C e^{At} B B_0^T e^{A_0^T t} C_0^T 
+        %         + C_0 e^{A_0t} B_0 B_0^T e^{A_0^T t} C_0^T 
+        % \right\} dt \\
+        % \int_0^\infty \rho(t) \left| C \left( e^{At} - e^{A_0 t} \right) B \right|^2 dt \\
+        % &= 
+        % \int_0^\infty \rho(t) C \left( e^{At} - e^{A_0 t} \right) B B^T \left( e^{At} - e^{A_0 t} \right)^T C^T dt
+    \end{aligned}
+\end{equation}
+How does this change as we perturb about $(A_0, B_0, C_0)$?
+First we differentiate with respect to $A$, keeping $B=B_0$ and $C=C_0$ fixed.
+Since
+\begin{equation}
+  \begin{aligned}
+      \frac{d}{du} e^{(A+uZ)t} \vert_{u=0}
+      &=
+      \int_0^t e^{As} Z e^{A(t-s)} ds, 
+  \end{aligned}
+\end{equation}
+we have that
+\begin{equation}
+  \begin{aligned}
+      \frac{d}{du} D(A+uZ,B_0,C_0)\vert_{u=0}
+      &=
+        2 \int_0^\infty \rho(t) \tr\left\{ C_0 \left( \int_0^t e^{As} Z e^{A(t-s)} ds \right) B_0 B_0^T \left( e^{At} - e^{A_0 t} \right)^T C_0^T \right\} dt \\
+      &=
+        2 \int_0^\infty \rho(t) \tr\left\{ C_0 \left( \int_0^t e^{As} Z e^{A(t-s)} ds \right) B_0 \left( h_A(t) - h_0(t) \right)^T \right\} dt 
+  \end{aligned}
+\end{equation}
+and, by differentiating this and supposing that $A$ is on the optimal set,
+i.e., $h_A(t)=h_0(t)$, (so wolog $A=A_0$):
+\begin{equation}
+  \begin{aligned}
+      \calH^{A,A}(Y,Z) 
+      &:= 
+      \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0+uY+vZ,B_0,C_0)\vert_{u=v=0} \\
+      &=
+        \int_0^\infty \rho(t) \tr\left\{ C_0 
+        \left( \int_0^t e^{A_0 s} Y e^{A_0 (t-s)} ds \right) 
+        B_0 B_0^T 
+        \left( \int_0^t e^{A_0 s} Z e^{A_0 (t-s)} ds \right)^T
+        C_0^T \right\} dt  .
+  \end{aligned}
+\end{equation}
+
+The function $\calH$ will define a quadratic form.
+To illustrate the use of this, suppose that $B$ and $C$ are fixed.
+% Here $\calH$ is the quadratic form underlying the Hamiltonian.
+By defining $\Delta_{ij}$ to be the matrix with a 1 in the $(i,j)$th slot
+and 0 elsewhere,
+the coefficients of the quadratic form is
+\begin{equation}
+    \begin{aligned}
+        H_{ij, k\ell}(A)
+        &:=
+        \calH(\Delta_{ij}, \Delta_{k\ell}) .
+    \end{aligned}
+\end{equation}
+
+We could use this to compute the gradient of $D$,
+or to get the quadratic approximation to $D$ near the optimal set.
+To do so, it'd be nice to have a way to compute the inner integral above.
+Suppose that we can diagonalize $A = U \Lambda U^{-1}$.
+Then
+\begin{equation} \label{eqn:exp_deriv}
+  \begin{aligned}
+      \int_0^t e^{As} Z e^{A(t-s)} ds 
+      &=
+      \int_0^t U e^{\Lambda s} U^{-1} Z U e^{\Lambda (t-s)} U^{-1} ds 
+  \end{aligned}
+\end{equation}
+Now, notice that
+\begin{equation}
+  \begin{aligned}
+      \int_0^t e^{s \lambda_i} e^{(t-s) \lambda_j} ds
+      &=
+      \frac{ e^{t \lambda_i} - e^{t \lambda_j} }{ \lambda_i - \lambda_j } .
+  \end{aligned}
+\end{equation}
+Therefore, 
+defining
+\begin{equation}
+    X_{ij}(t,Z) = \left( U^{-1} Z U \right)_{ij}
+      \frac{ e^{t \lambda_i} - e^{t \lambda_j} }{ \lambda_i - \lambda_j } 
+\end{equation}
+moving the $U$ and $U^{-1}$ outside the integral and integrating we get that
+\begin{equation}
+  \begin{aligned}
+      \int_0^t e^{As} Z e^{A(t-s)} ds 
+      &=
+      U X(t,Z) U^{-1} .
+  \end{aligned}
+\end{equation}
+
+Following on from above, we see that if $Z=\Delta_{k \ell}$, then
+\begin{equation}
+  \begin{aligned}
+      X_{ij}^{k\ell}(t) = 
+      \frac{ e^{t \lambda_i} - e^{t \lambda_j} }{ \lambda_i - \lambda_j } 
+      (U^{-1})_{\cdot k} U_{\ell \cdot},
+  \end{aligned}
+\end{equation}
+where $U_{k \cdot}$ is the $k$th row of $U$,
+and so
+\begin{equation}
+    \begin{aligned}
+        H_{ij, k\ell}(A)
+        &=
+        \int_0^\infty
+            \rho(t) \tr\left\{ C U X^{ij}(t) U^{-1} B B^T (U^{-1})^T X^{k\ell}(t)^T U^T C^T \right\}
+        dt .
+    \end{aligned}
+\end{equation}
+This implies that
+\begin{equation}
+    \begin{aligned}
+        D(A_0+\epsilon Z)
+        &\approx \epsilon^2\sum_{ijk\ell} H^{ij,k\ell} Z_{ij} Z_{k\ell} 
+    \end{aligned}
+\end{equation}
+and so
+\begin{equation}
+    \begin{aligned}
+        D(A_0+\epsilon Z)
+        &\approx \epsilon^2\sum_{ijk\ell} H^{ij,k\ell} Z_{ij} Z_{k\ell} 
+    \end{aligned}
+\end{equation}
+
+By section \ref{ss:quant_gen},
+if we set $\Sigma=\sigma^2 I$ and $U=H$,
+then a population at $A_0+Z$ experiences a restoring force of strength
+$(I + \sigma^2 H^{-1})^{-1} Z$ (treating $Z$ as a vector and $H$ as an operator on these).
+If $\sigma^2$ is small compared to $H^{-1}$
+then this is approximately $-\sigma^2 H^{-1} Z$.
+This suggests that the population mean follows an Ornstein-Uhlenbeck process,
+as described (in different terms) in \citet{hansen1996translating}.
+
+More generally, $B$ and $C$ may also change.
+To extend this we need the remaining second derivatives of $D$.
+First, in $B$:
+\begin{equation}
+    \begin{aligned}
+        \mathcal{H}^{B,B}(Y,Z)
+        &:= 
+        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0,B_0+uY+vZ,C_0)\vert_{u=v=0} \\
+        &=
+        \frac{1}{2} \int_0^\infty \rho(t) \tr\left\{ 
+        C_0 e^{t A_0}
+        \frac{d}{du} \frac{d}{dv} 
+        (uY+vZ)
+        (uY+vZ)^T
+        \vert_{u=v=0} 
+        e^{t A_0^T} C_0^T 
+        \right\} dt  \\
+        &=
+        \frac{1}{2} \int_0^\infty \rho(t) \tr\left\{ 
+        C_0 e^{t A_0}
+        \left( Y Z^T + Z Y^T \right)
+        e^{t A_0^T} C_0^T 
+        \right\} dt  .
+  \end{aligned}
+\end{equation}
+Next, in $C$:
+\begin{equation}
+    \begin{aligned}
+        \mathcal{H}^{B,B}(Y,Z)
+        &:= 
+        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0,B_0,C_0+uY+vZ)\vert_{u=v=0} \\
+        &=
+        \frac{1}{2} \int_0^\infty \rho(t) \tr\left\{ 
+        B_0 e^{t A_0^T}
+        \frac{d}{du} \frac{d}{dv} 
+        (uY+vZ)^T
+        (uY+vZ)
+        \vert_{u=v=0} 
+        e^{t A_0} B_0
+        \right\} dt  \\
+        &=
+        \frac{1}{2} \int_0^\infty \rho(t) \tr\left\{ 
+        B_0 e^{t A_0^T}
+        \left( Y Z^T + Z Y^T \right)
+        e^{t A_0} B_0
+        \right\} dt  .
+  \end{aligned}
+\end{equation}
+Now, the mixed derivatives in $B$ and $C$:
+\begin{equation}
+    \begin{aligned}
+        \mathcal{H}^{B,C}(Y,Z)
+        &:= 
+        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0,B_0+uY,C_0+vZ)\vert_{u=v=0} \\
+        &=
+        \int_0^\infty \rho(t) \tr\left\{ 
+        Y e^{t A_0^T} C_0^T Z e^{t A_0} B_0
+        \right\} dt  .
+  \end{aligned}
+\end{equation}
+In $A$ and $B$
+\begin{equation}
+    \begin{aligned}
+        \mathcal{H}^{A,B}(Y,Z)
+        &:= 
+        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0+uY,B_0+vZ,C_0)\vert_{u=v=0} \\
+        &=
+        \int_0^\infty \rho(t) \tr\left\{ 
+        C_0 \left(\int_0^t e^{s A_0} Y e^{(t-s) A_0} ds \right) B_0
+        Z^T e^{t A_0} C_0
+        \right\} dt  ,
+  \end{aligned}
+\end{equation}
+and finally in $A$ and $C$:
+\begin{equation}
+    \begin{aligned}
+        \mathcal{H}^{A,C}(Y,Z)
+        &:= 
+        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0+uY,B_0,C_0+vZ)\vert_{u=v=0} \\
+        &=
+        \int_0^\infty \rho(t) \tr\left\{ 
+        C_0 \left(\int_0^t e^{s A_0} Y e^{(t-s) A_0} ds \right) B_0
+        B_0 e^{t A_0} Z
+        \right\} dt  .
+  \end{aligned}
+\end{equation}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Genetic drift with a multivariate trait}
+\label{ss:quant_gen}
+
+For completeness, we provide a brief exposition of how a population 
+evolves due to genetic drift
+with a quantitative genetics model,
+as in \citet{lande1981models} or \citet{hansen1996translating}.
+These do not directly model underlying genetic basis,
+but developing a more accurate model is beyond the scope of this paper.
+
+Suppose that the population is distributed in trait space
+as a Gaussian with covariance matrix $\Sigma$ and mean $\mu$,
+whose density we write as $f(\cdot;\Sigma,\mu)$.
+Selection has the effect of multiplying this density by the fitness function and renormalizing,
+so that if expected fitness of $x$ is proportional to $\exp(-\|Lx\|^2/2)$,
+then the distribution post-selection
+has density at $x$ proportional to $f(x;\Sigma,\mu) \exp(-\|Lx\|^2/2)$.
+By the computation below (``Completing the square''),
+the result is a Gaussian distribution
+with covariance matrix $(\Sigma^{-1} + L^T L)^{-1}$ 
+and mean $(\Sigma^{-1}+L^T L)^{-1} \Sigma^{-1} \mu$.
+% Importantly, if $U$ is not invertible, then by $U^{-1}$ 
+% we mean $(U^T U)^{-1} U^T$, Moore-Penrose pseudoinverse of $U$,
+% so that $U U^{-1}$ is equal to the projection matrix
+% onto the span of the columns of $U$.
+
+After selection, we have reproduction:
+suppose this occurs as in the infinitesimal model \citep{barton2016infinitesimal},
+so that each offspring of parents with traits $x$ and $y$
+is drawn independently from a Gaussian distribution 
+with mean $(x+y)/2$ and covariance matrix $R$.
+Here, $R$ is the contribution of ``segregation variance'',
+i.e., random choices of parental alleles.
+If $\widetilde \Sigma = (\Sigma^{-1} + L^T L)^{-1}$ 
+is the covariance matrix of the parents post-selection,
+then the distribution of offspring will again be Gaussian,
+with mean equal to that of the parents
+and covariance matrix $\widetilde \Sigma/2 + R$.
+
+In summary, a generation under this model modifies the mean ($\mu$)
+and covariance matrix ($\Sigma$) of a population as follows:
+\begin{align*}
+    \mu &\mapsto \mu' = (\Sigma^{-1}+L^T L)^{-1} \Sigma^{-1} \mu \\
+    \Sigma &\mapsto \Sigma' = \frac{1}{2} (\Sigma^{-1} + L^T L)^{-1} + R .
+\end{align*}
+What measures are stable under this transformation?
+The condition $\mu = \mu'$ reduces to $\Sigma L^T L \mu = 0$;
+if we assume $R$ and therefore $\Sigma$ are of full rank,
+then this happens if and only if $\mu$ is in the null space of $L$,
+i.e., if $\mu$ lies in a neutral direction.
+The condition $\Sigma' = \Sigma$ can also be solved,
+at least numerically.
+After rearrangement, it reduces to
+$\Sigma L^T L \Sigma + (I/2 - R L^T L) \Sigma = R$.
+%% I DON'T THINK WE USE THE FOLLOWING:
+% We can find a more explicit description
+% if we assume that $x^T L^T L x = \sum_{i=1}^k x_i^2$,
+% i.e., that selection only cares about the first $k$ coordinates,
+% and then with no interactions between traits.
+% If so, the condition $\Sigma' = \Sigma$
+% can be written in block form as
+% \begin{align*}
+%     \begin{bmatrix}
+%         \Sigma_{11}^2 + (I/2 - R_{11}) \Sigma_{11}
+%         & \Sigma_{11} \Sigma_{12} + (I/2 - R_{11})\Sigma_{12} \\
+%         \Sigma_{12}^T \Sigma_{11} + \Sigma_{12}^T/2 - R_{12}^T \Sigma_{11}
+%         & \Sigma_{22} - R_{12}^T \Sigma_{12} 
+%     \end{bmatrix}
+%     &=
+%     \begin{bmatrix}
+%         R_{11} & R_{12} \\
+%         R_{12}^T & R_{22}
+%     \end{bmatrix} .
+% \end{align*}
+% The first equation, $\Sigma_{11}^2 + (I/2 - R_{11}) \Sigma_{11} = R_{11}$,
+% can be solved with the quadratic formula: 
+% $$\Sigma_{11} = (R_{11} - I/2 + Q)/2$$
+% for any $Q$ that commutes with $R_{11}$
+% and is a solution to $Q^2 = (R_{11} - I/2)^2 + 4 R_{11}$.
+% Since we need $\Sigma$ to be positive definite,
+% we take the solution with positive eigenvalues.
+% Given $\Sigma_{11}$, the remaining components are
+% \begin{align*}
+%     \Sigma_{12} 
+%     &= 
+%     \left( \Sigma_{11} + I/2 - R_{11} \right)^{-1} R_{12} \\
+%     \Sigma_{22}
+%     &=
+%     R_{12}^T \Sigma_{12} + R_{22} .
+% \end{align*}
+Importantly, the mean $\mu$ does not affect 
+either how the covariance matrix moves,
+or its stable shape.
+
+Above we have described the \emph{expected} motion of the mean and covariance.
+However, random resampling will introduce noise.
+Suppose that a population of $N$ individuals
+behaves approximately as described above.
+By the above,
+we may expect that the covariance matrix stays close to a constant value $\Sigma$,
+computed from $R$ and $L$ as above,
+so that we need only consider motion of the mean, $\mu$.
+Since we take a sample of size $N$ to construct the next generation,
+the next generation's mean is drawn from a Gaussian distribution with mean $\mu'$
+and covariance matrix $\Sigma/N$.
+Defining $\Gamma = (I - (I + \Sigma L^T L)^{-1})$,
+this can be written as
+\begin{align*}
+    \mu' - \mu = \Gamma \mu + \epsilon/\sqrt{N} ,
+\end{align*}
+where $\epsilon$ is a multivariate Gaussian with mean zero and covariance matrix $\Sigma$.
+Let $\mu(k)$ denote the mean in the $k^\text{th}$ generation,
+and suppose that $\mu$ differs from optimal by something of order $1/\sqrt{N}$:
+if $\nu(t) = \sqrt{N} \mu(t\sqrt{N})$ is the rescaled process,
+then the previous equation implies that as $N \to \infty$, 
+in the limit $\nu$ solves the It\^{o} equation
+\begin{align*}
+    d \nu(t) = \Gamma \nu(t) dt + \Sigma^{1/2} dW(t) ,
+\end{align*}
+where now $W(t)$ is a multivariate white noise.
+This has an explicit solution as a multivariate Ornstein-Uhlenbeck process:
+\begin{align*}
+    \nu(t) = e^{-t \Gamma} \nu(0) + \int_0^t e^{-(t-s) \Gamma} \Sigma^{1/2} dW(s).
+\end{align*}
+% Note that after some algebra, $\Gamma = \Sigma L^T L (I - \Sigma L^T L)^{-1}$.
+The asymptotic variance of this process in the direction $z$ is
+\begin{align} \label{eqn:limiting_cov}
+    \lim_{t \to \infty} \var[\nu(t) \cdot z]
+    &=
+    \int_0^\infty z^T e^{-s \Gamma} \Sigma e^{-s \Gamma} z ds ,
+\end{align}
+which is infinite if and only if $\Gamma z = 0$,
+which occurs if and only if $Lz=0$.
+This implies that at equilibrium, 
+population mean trait values lie away from the optimal set
+by a Gaussian displacement of order $1/\sqrt{N}$
+with a covariance matrix given by equation \eqref{eqn:limiting_cov}.
+
+\paragraph{Completing the square}
+First note that if $A$ is symmetric,
+\begin{align*}
+    (x-y)^T A (x-y)
+    &=
+    x^T A \left( x - 2y \right) + y^T A y ,
+\end{align*}
+and so if $B$ is also symmetric and $A+B$ is invertible,
+\begin{align*}
+    (x-y)^T A (x-y) + x^T B x
+    &=
+    x^T (A + B) \left( x - 2 (A + B)^{-1} A y \right) + y^T A y \\
+    &=
+    \left( x - (A + B)^{-1} A y \right)^T
+    (A + B)
+    \left( x - (A + B)^{-1} A y \right) \\
+    &\qquad{}
+    + y^T A y - y^T A^T (A+B)^{-1} A y .
+\end{align*}
+Therefore, 
+by substituting $A=\Sigma^{-1}$ and $B=L^T L$,
+\begin{align*}
+    \frac{ f(x;\Sigma,y) \exp(-x^T L^T L x / 2)}{\int f(z;\Sigma,y) \exp(-z^T L^T L z / 2) dz}
+    &=
+    f(x; (\Sigma^{-1} + L^T L)^{-1}, (\Sigma^{-1}+L^T L)^{-1} \Sigma^{-1} y) .
+\end{align*}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Gaussian load}
+\label{apx:gauss_load}
+
+Suppose that a population has a Gaussian distribution in $d$-dimensional trait space
+with mean $\mu$ and covariance matrix $\Sigma$,
+and that fitness of an individual at $x$ is $\exp(-\|Lx\|^2/2)$.
+Then, completing the square as above with $A=\Sigma^{-1}$, $y=\mu$, and $B=L^T L$,
+and defining $Q = (\Sigma^{-1} + L^T L)^{-1}$,
+\begin{align*}
+    &
+    \frac{1}{\sqrt{2 \pi}^n \det(\Sigma)^{1/2}} 
+        \int e^{-\frac{1}{2} x^T \Sigma^{-1} x} e^{-\frac{1}{2} x^T L^T L x} dx \\
+    &\qquad =
+    \frac{1}{\sqrt{2 \pi}^n \det(\Sigma)^{1/2}} 
+        \int e^{-\frac{1}{2}(x-Q \Sigma^{-1} \mu)^T Q^{-1} (x-Q \Sigma^{-1} \mu)} dx 
+        \\ &\qquad \qquad {}
+        \times e^{\mu^T \left( I - \Sigma^{-1} Q \right) \Sigma^{-1} \mu} \\
+    &\qquad =
+    \sqrt{\frac{\det(Q)}{\det(\Sigma)}}
+        \exp\left\{\mu^T \left( I - \Sigma^{-1} Q \right) \Sigma^{-1} \mu\right\} \\
+    &\qquad =
+    \sqrt{\frac{1}{\det(\Sigma)\det(\Sigma^{-1}+L^TL)}}
+        \exp\left\{\mu^T \left( I - (I + L^T L \Sigma)^{-1} \right) \Sigma^{-1} \mu\right\} \\
+\end{align*}
+
+Now suppose that $\Sigma = \sigma^2 I$ and $L = I/\gamma$.
+Then,
+\begin{align*}
+    \sqrt{\frac{1}{\det(\Sigma)\det(\Sigma^{-1}+L^TL)}}
+    &=
+    \sqrt{\frac{1}{\sigma^{2d} (1/\sigma^2 + 1/\gamma^2)^d}} \\
+    &=
+    \frac{1}{(1+ \left(\sigma/\gamma\right)^2)^{d/2}} .
+\end{align*}
+Also, 
+\begin{align*}
+        \left( I - (I + L^T L \Sigma)^{-1} \right) \Sigma^{-1} 
+        &=
+        \frac{1}{\sigma^2}\left( 1 - (1 + (\sigma/\gamma)^2)^{-1} \right) I \\
+        &=
+        \frac{1}{\gamma^2}\frac{1}{(1 + (\sigma/\gamma)^2)} I \\
+\end{align*}
+
+%%%%%%%%%%%
+\section*{Evolution of segregation covariance}
+\label{apx:seg_cov}
+
+The description above does not describe how \emph{two} diverging populations interact,
+since the amount of \emph{segregation variance}, quantified by $R$,
+will not stay constant.
+To get an idea of how this might change,
+suppose that a trait is determined by $L$ unlinked, biallelic loci,
+and that the $i^\mathrm{th}$ locus has two alleles with additive effects $\pm x_i$,
+so that being homozygous for the $+$ allele contributes $+2x_i$ to the trait.
+For simplicity, we will neglect the effects of selection.
+If the $+$ allele at locus $i$ is at frequency $p_i$ in a population,
+then the mean and genetic variance of the trait in a diploid population with random mating is
+\begin{align*}
+    m &= 2 \sum_i (2 p_i - 1) x_i \\
+    s^2 &= 4 \sum_i p_i (1-p_i) x_i^2 .
+\end{align*}
+
+Segregation variance between two parents depends on the loci at which either are heterozygous,
+and each locus contributes independently
+since alleles are additive.
+If the alleles are at Hardy-Weinberg proportions,
+then since segregation acts like a fair coin flip, a heterozygous locus contributes $x_i^2/4$ to the variance,
+and so the \emph{mean} segregation variance, averaging across parents, is
+% two parents, prob 2p(1-p) for prob the locus is heterozygous, variance 0.5(1-0.5)x^2.
+\begin{align*}
+    R_0(p) = \sum_i p_i (1-p_i) x_i^2 .
+\end{align*}
+
+On the other hand,
+if the second parent came from a distinct population
+with frequencies $q_i$
+(an $F_1$ hybrid),
+this would be
+\begin{align*}
+    R_1(p,q) &= \frac{1}{2} \sum_i p_i^2 (1-p_i)^2 x_i^2 
+                + \frac{1}{2} \sum_i q_i^2 (1-q_i)^2 x_i^2 \\
+                &= (R_0(p) + R_0(q))/2 .
+\end{align*}
+If we assume that the populations are at equilibrium, $R_0(p) \approx R_0(q)$,
+and so $R_1(p,q) \approx R_0(p)$.
+
+Now consider an $F_2$ hybrid, where both parents are $F_1$
+and so each heterozygous at locus $i$ with probability $p_i (1-q_i) + (1-p_i) q_i$.
+Then
+\begin{align*}
+    R_2(p,q) &= \frac{1}{2} \sum_i \left(p_i (1-q_i) + (1-p_i) q_i \right) x_i^2  .
+\end{align*}
+Suppose that the two populations are slightly drifted from each other,
+with frequency difference $p_i-q_i = 2\epsilon_i$.
+Then,
+\begin{align*}
+    p (1-q) + p (1-q)
+    &=
+    (u+\epsilon) (1-u+\epsilon) + (u-\epsilon) (1-u-\epsilon) \\
+    &=
+    2 u (1-u)
+    % + \epsilon (u + (1-u) - u - (1-u))
+    + 2 \epsilon^2 .
+\end{align*}
+If the frequencies have evolved neutrally in unconnected, Wright-Fisher populations 
+of effective size $N$ for $t$ generations from a common ancestor with allele frequency $u$,
+then $\epsilon$ has mean zero and variance roughly $2 u (1-u) t/N$.
+Still assuming the populations are at stationarity,
+so that $R_0$ is constant between the two,
+and taking the frequencies $p_i$ as a proxy for the ancestral frequencies $u_i$,
+this implies that we expect
+\begin{align*}
+    R_2 
+    &\approx 
+    R_0 + 2 \sum_i p_i (1-p_i) x_i^2 t/N \\
+    &=
+    \left( 1 + \frac{2t}{N} \right) R_0 .
+\end{align*}
+On the other hand, the expected squared difference in trait \emph{means} between two such populations is
+\begin{align} \label{eqn:var_dx}
+    8 \sum_i p_i (1-p_i) x_i^2 t / N = 8 R_0 t /N .
+\end{align}
+This implies that under this model,
+segregation variance in $F_2$s between two populations
+is roughly increased by a factor of $1/8$ of the difference between their means.
+
+\end{document}
 
@@
 %\end{example}
-
-\appendix
-
-%\section{Kalman Decomposition} \label{apx:kalman}
-%\begin{definition}[Phenotypic equivalence of systems]
-%    Let $(\kappa(t),\phi(t))$ and $(\bar \kappa(t),\bar \phi(t))$ be the solutions to \eqref{eqn:lti_system}
-%    with coefficient matrices $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ respectively,
-%    and both $\kappa(0)$ and $\bar \kappa(0)$ are zero. 
-%    The systems defined by $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ are
-%    \textbf{phenotypically equivalent} 
-%    if
-%    \begin{align*}
-%        \phi(t) = \bar \phi(t) \qquad \text{for all} \; t \ge 0.
-%    \end{align*}
-%    Equivalently, this occurs if and only if
-%    \begin{align*}
-%        h(t) = \bar h(t)  \qquad \text{for all} \; t \ge 0,
-%    \end{align*}
-%    where $h$ and $\bar h$ are the impulse responses of the two systems.
-%\end{definition}
-%
-%One way to find other systems equivalent to a given one
-%is by change of coordinates (``algebraic equivalence''):
-%if $V$ is an invertible matrix, then the systems $(A,B,C)$ and $(VAV^{-1},VB,CV^{-1})$
-%have the same dynamics because their transfer functions are equal:
-%\begin{align*}
-%    CV^{-1}( zI - VAV^{-1})^{-1}VB
-%    =
-%    CV^{-1}V( zI - A)^{-1}V^{-1}VB
-%    =
-%    C( zI - A)^{-1}B .
-%\end{align*}
-%However, the converse is not necessarily true: 
-%systems can have identical transfer functions without being changes of coordinates of each other.
-%In fact, systems with identical transfer functions can involve interactions between different
-%numbers of molecular species.
-%
-%The set of all systems phenotypically equivalent to a given system $(A,B,C)$ 
-%is elegantly described using the Kalman decomposition,
-%which also clarifies the system dynamics? tells us a lot about how it works? \plr{or something}
-%To motivate this, first note that the input $u(t)$ only directly pushes the system
-%in directions lying in the span of the columns of $B$.
-%As a result, different combinations of input can 
-%move the system in any direction that lies in the \emph{reachable subspace},
-%which we denote by $\reachable$,
-%and is defined to be the closure of $\spn(B)$ under applying $A$
-%(or equivalently, the span of $B, AB, A^2B, \ldots A^{n-1}B$).
-%Analogously to this, we define
-%the \emph{observable subspace}, $\mathcal{O}$,
-%to be the closure of $\spn(C^T)$ under applying $A$.
-%(Or: $\unobservable$ is the largest $A$-invariant subspace
-%contained in the null space of $C$;
-%and $\reachable$ is the largest $A$-invariant subspace contained in the image of $B$.)
-%
-%If we define
-%\begin{enumerate}
-%    \item The columns of $P_\rno$ are an orthonormal basis for $\reachable \cap \unobservable$.
-%    \item The columns of $P_\ro$ are an orthonormal basis of
-%        the complement of $\reachable \cap \unobservable$ in $\reachable$.
-%    \item The columns of $P_\nro$ are an orthonormal basis of
-%        the complement of $\reachable \cap \unobservable$ in $\unobservable$.
-%    \item The columns of $P_\nrno$ are an orthonormal basis of
-%        the remainder of $\R^n$.
-%\end{enumerate}
-%If we then define
-%\begin{align*}
-%    P &= 
-%    \left[ \begin{array}{c|c|c|c}
-%        P_\rno & P_\ro & P_\nro & P_\nrno
-%    \end{array} \right] ,
-%\end{align*}
-%then
-%\begin{align*}
-%    P^T P
-%    &=
-%    \left[ \begin{array}{c|c|c|c}
-%        I & 0 & 0 & 0 \\
-%        \hline
-%        0 & I & U & 0 \\
-%        \hline
-%        0 & V & I & 0 \\
-%        \hline
-%        0 & 0 & 0 & I 
-%    \end{array} \right] .
-%\end{align*}
-%\plr{Check this.  Can we get $U=V=0$?}
-%
-%The following theorem can be found in SOME REFERENCE.
-%
-%\begin{theorem}[Kalman decomposition] \label{thm:kalman}
-%        For any system $(A,B,C)$ with corresponding Kalman basis matrix $P$,
-%        the transformed system $(PAP^{-1},PB,CP^{-1})$  has the following form:
-%        \begin{align*}
-%            \widehat A = PAP^{-1}
-%            &=
-%            \left[ \begin{array}{cccc}
-%                A_{\rno} & A_{\rno,\ro} & A_{\rno,\nrno} & A_{\rno,\nro} \\
-%                0 & A_{\ro} & 0 & A_{\ro,\nro} \\
-%                0 & 0 & A_{\nrno} & A_{\nrno,\nro} \\
-%                0 & 0 & 0 & A_{\nro}
-%            \end{array} \right] ,
-%        \end{align*}
-%        and
-%        \begin{align*}
-%            \widehat B = PB
-%            &=
-%            \left[ \begin{array}{cccc}
-%                B_{\rno} \\
-%                B_{\ro} \\
-%                0 \\
-%                0 
-%            \end{array} \right] ,
-%        \end{align*}
-%        and
-%        \begin{align*}
-%            \widehat C = CP^{-1}
-%            &=
-%            \left[ \begin{array}{cccc}
-%                0 & C_{\ro} & C_{\nrno} & 0 
-%            \end{array} \right] .
-%        \end{align*}
-%        The transfer function of both systems is given by
-%        \begin{align*}
-%            H(z) = C_{\ro} ( zI - A_{\ro} )^{-1} B_{\ro} .
-%        \end{align*}
-%\end{theorem}
-%
-%In the latter case, we say that the system is \emph{minimal} 
-%-- there is no equivalent system with a smaller number of species.
-%Note that this says that any two equivalent minimal systems
-%are changes of basis of each other.
-%
-%Since any system can be put into this form,
-%and once in this form, its transfer function is determined only by 
-%$C_{\ro}$, $A_{\ro}$, and $B_\ro$,
-%therefore, the set of all equivalent systems are parameterized by
-%the dimension $n$,
-%the choice of basis ($P$),
-%the remaining submatrices in $\widehat A$, $\widehat B$, and $\widehat C$
-%(which are unconstrained),
-%and a invertible transformation of $\spn(P_{\ro})$, which we call $T_\ro$.
-%
-%\begin{theorem}[Parameterization of equivalent systems]
-%    Let $(A,B,C)$ be a minimal system.
-%    \begin{itemize}
-%        \item[(a)]
-%            Every equivalent system is of the form given in Theorem \ref{thm:kalman},
-%            i.e., can be specified by choosing a dimension, $n$;
-%            submatrices in $\widehat A$, $\widehat B$, and $\widehat C$ 
-%            except for $A_\ro=A$, $B_\ro=B$, and $C_\ro=C$;
-%            and choosing an invertible matrix $P$.
-%
-%        \item[(b)]
-%            \plr{conjecture:}
-%            The parameterization is unique
-%            if $P$ is furthermore chosen so that 
-%            each $P_x$ other than $P_\ro$ is a projection matrix,
-%            and that 
-%            \begin{align*}
-%                0
-%                =
-%                P_x^T P_y
-%            \end{align*}
-%            for all $(x,y)$ except $(\ro,\nrno)$.
-%
-%    \end{itemize} 
-%\end{theorem}
-%
-%\plr{Another way of saying it: pick the $\reachable$ and $\unobservable$ subspaces,
-%that must intersect in something of the minimal dimension;
-%then let $P$ be the appropriate basis?}
-%
-%In some situations we may be interested in only ``network rewiring'',
-%where $A$ changes while $B$ and $C$ do not.
-%For instance, 
-%if all non-regulatory functions of each molecule are strongly constrained,
-%then $C$ cannot change.
-%Likewise, if responses of each molecule to the external inputs are not changed by evolution,
-%then $B$ does not change.
-%
-%
-%\subsection{Neutral directions from the Kalman decomposition}
-%
-%The Kalman decomposition above says that any system $(A,B,C)$ can be decomposed into
-%$(P, \hat A, \hat B, \hat C)$ so that
-%$$\begin{aligned}
-%    A &= P^{-1} \hat A P  \\
-%    B &= P^{-1} \hat B  \\
-%    C &= \hat C P ,
-%\end{aligned}$$
-%and we know precisely how we can change these to preserve the transfer function:
-%\begin{enumerate}
-%    \item $P \to P + \epsilon Q$ as long as the result is still invertible,
-%    \item $\hat A \to A + \epsilon X$ as long as $X$ is zero in the correct places,
-%    \item $\hat B \to B + \epsilon Y$ as long as $Y$ is zero in the correct places,
-%    \item $\hat C \to C + \epsilon Y$ as long as $Z$ is zero in the correct places.
-%\end{enumerate}
-%By taking $\epsilon \to 0$, these tell us the local directions we can move a system $(A,B,C)$ in.
-%All statements below are up to first order in $\epsilon$,
-%omitting terms of order $\epsilon^2$.
-%
-%First, since $(P + \epsilon Q)^{-1} = P^{-1} + \epsilon P^{-1} Q P^{-1}$,
-%modifying $P \to P + \epsilon Q$ changes
-%$$\begin{aligned}
-%    A 
-%        &\to A + \epsilon P^{-1} \hat A Q - \epsilon P^{-1} Q P^{-1} \hat A P \\
-%        &= A + \epsilon \left(A P^{-1} Q - P^{-1} Q A\right) , \\
-%    B
-%        &\to B - \epsilon P^{-1} Q B \\
-%    C
-%        &\to C + \epsilon C P^{-1} Q .
-%\end{aligned}$$
-%Since $P$ is invertible and $Q$ can be anything (if $\epsilon$ is small enough),
-%this allows changes in the direction of an arbitrary $W$:
-%$$\begin{aligned}
-%    A 
-%        &= A + \epsilon \left(A W - W A\right) , \\
-%    B
-%        &\to B - \epsilon W B \\
-%    C
-%        &\to C + \epsilon C W .
-%\end{aligned}$$
-%
-%Then, $\hat A \to A + \epsilon X$  does
-%$$\begin{aligned}
-%    A \to A + \epsilon P^{-1} X P 
-%\end{aligned}$$
-%and $\hat B \to B + \epsilon Y$ does
-%$$\begin{aligned}
-%    B \to B + \epsilon P^{-1} Y
-%\end{aligned}$$
-%and $\hat C \to C + \epsilon Z$ does
-%$$\begin{aligned}
-%    C \to C + \epsilon Z P .
-%\end{aligned}$$
-%These degrees of freedom look like they depend on $P$, 
-%which is not unique,
-%but for any two choices of $P$ there are corresponding choices of $X$
-%that give the same actual change in $A$ (and likewise for $Y$ and $Z$).
-%
-%
-%Therefore, this gives us an upper bound on the number of degrees of freedom,
-%in terms of the dimensions of the blocks in the Kalman decomposition ($n_\ro$ etc)
-%and the dimensions of $B$ and $C$ ($n_B$ and $n_C$ respectively):
-%namely, for $W$, $X$, $Y$, and $Z$ respectively:
-%$$\begin{aligned}
-%    n^2 
-%    + (n_{\rno} + n_\ro n_\nro + n_\nrno(n_\nrno + n_\nro) + n_\nro^2)
-%    + n_B n_\rno
-%    + n_C n_\nrno .
-%\end{aligned}$$
-%However, some of these may be redundant.
-%For instance, changing $P$ in the direction of 
-%a $Q$ that satisfies both $A P^{-1} Q = P^{-1} Q A$ and $C P^{-1} Q = 0$
-%is equivalent to changing $B$ by $Y = QB$.
-%
-%\section{Meiotic recombination in linear systems}\label{apx:recombination}
-%
-%  Recombination is performed by taking two analogous system components from $\Sys$ and $\Sys'$ and randomly swapping rows or columns.
-%  \emph{E.g.} gamete systems $(A'', B'', C'')$, produced by the diploid $\{\Sys, \Sys'\}$, are formed by recombining (randomly swapping rows or columns) between two, possibly distinct, systems $\Sys = (A,B,C)$ and $\Sys' = (A', B', C')$ such that,
-%  \begin{align*}
-%    \Sys'' = \left( \begin{array}{ll}
-%    A'' &= MA + (I-M)A' , \\
-%    B'' &= MB + (I-M)B' , \\
-%    C'' &= CM + C'(I-M)
-%    \end{array} \right)
-%  \end{align*}
-%  where $M$ is a diagonal matrix where each diagonal element is a Bernoulli random variable ($M_{ii} = 0$ or $1$ with equal probability, and $M_{ij}=0$ if $i \neq j$).
-If systems are different dimensions the smaller system elements can be augmented with $0$s (\emph{e.g.} $\left[ \begin{smallmatrix} A & 0 \\ 0 & 0 \end{smallmatrix} \right], \left[ \begin{smallmatrix} B \\ 0 \end{smallmatrix} \right], \left[ \begin{smallmatrix} C & 0 \end{smallmatrix}\right]$).
-%
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Genetic drift with a multivariate trait}
-\label{ss:quant_gen}
-
-For completeness, we provide a brief exposition of how a population 
-evolves due to genetic drift
-with a quantitative genetics model,
-as in \citet{lande1981models} or \citet{hansen1996translating}.
-These do not directly model underlying genetic basis,
-but developing a more accurate model is beyond the scope of this paper.
-
-Suppose that the population is distributed in trait space
-as a Gaussian with covariance matrix $\Sigma$ and mean $\mu$,
-whose density we write as $f(\cdot;\Sigma,\mu)$.
-Selection has the effect of multiplying this density by the fitness function and renormalizing,
-so that if expected fitness of $x$ is proportional to $\exp(-\|Lx\|^2/2)$,
-then the distribution post-selection
-has density at $x$ proportional to $f(x;\Sigma,\mu) \exp(-\|Lx\|^2/2)$.
-By the computation below (``Completing the square''),
-the result is a Gaussian distribution
-with covariance matrix $(\Sigma^{-1} + L^T L)^{-1}$ 
-and mean $(\Sigma^{-1}+L^T L)^{-1} \Sigma^{-1} \mu$.
-% Importantly, if $U$ is not invertible, then by $U^{-1}$ 
-% we mean $(U^T U)^{-1} U^T$, Moore-Penrose pseudoinverse of $U$,
-% so that $U U^{-1}$ is equal to the projection matrix
-% onto the span of the columns of $U$.
-
-After selection, we have reproduction:
-suppose this occurs as in the infinitesimal model \citep{infinitesimal},
-so that each offspring of parents with traits $x$ and $y$
-is drawn independently from a Gaussian distribution 
-with mean $(x+y)/2$ and covariance matrix $R$.
-Here, $R$ is the contribution of ``segregation variance''.
-If $\widetilde \Sigma = (\Sigma^{-1} + L^T L)^{-1}$ 
-is the covariance matrix of the parents post-selection,
-then the distribution of offspring will again be Gaussian,
-with mean equal to that of the parents
-and covariance matrix $\widetilde \Sigma/2 + R$.
-
-In summary, a generation under this model modifies the mean ($\mu$)
-and covariance matrix ($\Sigma$) of a population as follows:
-\begin{align*}
-    \mu &\mapsto \mu' = (\Sigma^{-1}+L^T L)^{-1} \Sigma^{-1} \mu \\
-    \Sigma &\mapsto \Sigma' = \frac{1}{2} (\Sigma^{-1} + L^T L)^{-1} + R .
-\end{align*}
-What measures are stable under this transformation?
-The condition $\mu = \mu'$ reduces to $\Sigma L^T L \mu = 0$;
-if we assume $R$ and therefore $\Sigma$ are of full rank,
-then this happens if and only if $\mu$ is in the null space of $L$,
-i.e., if $\mu$ lies in a neutral direction.
-The condition $\Sigma' = \Sigma$ can also be solved,
-at least numerically.
-After rearrangement, it reduces to
-$\Sigma L^T L \Sigma + (I/2 - R L^T L) \Sigma = R$.
-We can find a more explicit description
-if we assume that $x^T L^T L x = \sum_{i=1}^k x_i^2$,
-i.e., that selection only cares about the first $k$ coordinates,
-and then with no interactions between traits.
-If so, the condition $\Sigma' = \Sigma$
-can be written in block form as
-\begin{align*}
-    \begin{bmatrix}
-        \Sigma_{11}^2 + (I/2 - R_{11}) \Sigma_{11}
-        & \Sigma_{11} \Sigma_{12} + (I/2 - R_{11})\Sigma_{12} \\
-        \Sigma_{12}^T \Sigma_{11} + \Sigma_{12}^T/2 - R_{12}^T \Sigma_{11}
-        & \Sigma_{22} - R_{12}^T \Sigma_{12} 
-    \end{bmatrix}
-    &=
-    \begin{bmatrix}
-        R_{11} & R_{12} \\
-        R_{12}^T & R_{22}
-    \end{bmatrix} .
-\end{align*}
-The first equation, $\Sigma_{11}^2 + (I/2 - R_{11}) \Sigma_{11} = R_{11}$,
-can be solved with the quadratic formula: 
-$$\Sigma_{11} = (R_{11} - I/2 + Q)/2$$
-for any $Q$ that commutes with $R_{11}$
-and is a solution to $Q^2 = (R_{11} - I/2)^2 + 4 R_{11}$.
-Since we need $\Sigma$ to be positive definite,
-we take the solution with positive eigenvalues.
-Given $\Sigma_{11}$, the remaining components are
-\begin{align*}
-    \Sigma_{12} 
-    &= 
-    \left( \Sigma_{11} + I/2 - R_{11} \right)^{-1} R_{12} \\
-    \Sigma_{22}
-    &=
-    R_{12}^T \Sigma_{12} + R_{22} .
-\end{align*}
-Importantly, the mean $\mu$ does not affect 
-either how the covariance matrix moves,
-or its stable shape.
-
-Above we have described the \emph{expected} motion of the mean and covariance..
-However, random resampling will introduce noise.
-Suppose that a population of $N$ individuals
-behaves approximately as described above.
-By the above,
-we may expect that the covariance matrix stays close to a constant value $\Sigma$,
-computed from $R$ and $L$ as above,
-so that we need only consider motion of the mean, $\mu$.
-Since we take a sample of size $N$ to construct the next generation,
-the next generation's mean is drawn from a Gaussian distribution with mean $\mu'$
-and covariance matrix $\Sigma/N$.
-Defining $\Gamma = (I - (I + \Sigma L^T L)^{-1})$,
-this can be written as
-\begin{align*}
-    \mu' - \mu = \Gamma \mu + \epsilon/\sqrt{N} ,
-\end{align*}
-where $\epsilon$ is a multivariate Gaussian with mean zero and covariance matrix $\Sigma$.
-Let $\mu(k)$ denote the mean in the $k^\text{th}$ generation,
-and suppose that $\mu$ differs from optimal by something of order $1/N$:
-if $\nu(t) = N \mu(tN)$,
-then the previous equation implies that as $N \to \infty$, 
-in the limit $\nu$ solves the It\^{o} equation
-\begin{align*}
-    d \nu(t) = \Gamma \nu(t) dt + \Sigma^{1/2} dW(t) ,
-\end{align*}
-where now $W(t)$ is a multivariate white noise.
-This has an explicit solution as a multivariate Ornstein-Uhlenbeck process:
-\begin{align*}
-    \nu(t) = e^{-t \Gamma} \nu_0 + \int_0^t e^{-(t-s) \Gamma} \Sigma^{1/2} dW(s).
-\end{align*}
-% Note that after some algebra, $\Gamma = \Sigma L^T L (I - \Sigma L^T L)^{-1}$.
-The asymptotic variance of this process in the direction $z$ is
-\begin{align} \label{eqn:limiting_cov}
-    \lim_{t \to \infty} \var[\nu(t) \cdot z]
-    &=
-    \int_0^\infty z^T e^{-s \Gamma} \Sigma e^{-s \Gamma} z ds ,
-\end{align}
-which is infinite iff $\Gamma z = 0$,
-which occurs iff $Lz=0$.
-In other words, population mean trait values lie away from the optimal set
-by a Gaussian displacement of order $1/N$
-with a covariance matrix given by equation \eqref{eqn:limiting_cov}.
-
-\plri{Now write what this means intuitively.}
-
-\paragraph{Completing the square}
-First note that if $A$ is symmetric,
-\begin{align*}
-    (x-y)^T A (x-y)
-    &=
-    x^T A \left( x - 2y \right) + y^T A y ,
-\end{align*}
-and so if $B$ is also symmetric and $A+B$ is invertible,
-\begin{align*}
-    (x-y)^T A (x-y) + x^T B x
-    &=
-    x^T (A + B) \left( x - 2 (A + B)^{-1} A y \right) + y^T A y \\
-    &=
-    \left( x - (A + B)^{-1} A y \right)^T
-    (A + B)
-    \left( x - (A + B)^{-1} A y \right) \\
-    &\qquad{}
-    + y^T A y - y^T A^T (A+B)^{-1} A y .
-\end{align*}
-Therefore, 
-by substituting $A=\Sigma^{-1}$ and $B=L^T L$,
-\begin{align*}
-    \frac{ f(x;\Sigma,y) \exp(-x^T L^T L x / 2)}{\int f(z;\Sigma,y) \exp(-z^T L^T L z / 2) dz}
-    &=
-    f(x; (\Sigma^{-1} + L^T L)^{-1}, (\Sigma^{-1}+L^T L)^{-1} \Sigma^{-1} y) .
-\end{align*}
-
-%%%%%%%%%%%
-\subsection*{Evolution of segregation covariance}
-\label{apx:seg_cov}
-
-The description above does not completely describe how two diverging populations interact,
-since the amount of \emph{segregation variance}, quantified by $R$,
-will not stay constant.
-To get an idea of how this might change,
-suppose that a multivariate trait is determined by $L$ unlinked, biallelic loci,
-and that the $i^\mathrm{th}$ locus has two alleles with additive effects $\pm x_i$,
-so that begin homozygous for the $+$ allele contributes $+2x_i$ to the trait.
-For simplicity, we will neglect the effects of selection.
-\plri{fixup the below to be actually multivariate}
-If the $+$ allele at locus $i$ is at frequency $p_i$ in a population,
-then the mean and genetic variance of the trait in a diploid population with random mating is
-\plr{the mean should be $\sum_i x_i (2p_i-1)$}
-\begin{align*}
-    m &= 2 \sum_i p_i x_i \\
-    s^2 &= 2 \sum_i p_i (1-p_i) x_i^2 .
-\end{align*}
-
-Segregation variance between two parents depends on the loci at which either are heterozygous,
-and each locus contributes independently
-since alleles are additive.
-If the alleles are at Hardy-Weinberg proportions,
-then since segregation is a fair coin flip, a heterozygous locus contributes $x_i^2/4$ to the variance,
-and so the \emph{mean} segregation variance, averaging across parents, is
-% two parents, prob 2p(1-p) for prob the locus is heterozygous, variance 0.5(1-0.5)x^2.
-\begin{align*}
-    R_0(p) = \frac{1}{2} \sum_i p_i (1-p_i) x_i^2 .
-\end{align*}
-
-On the other hand,
-if the second parent came from a distinct population
-with frequencies $q_i$
-(an $F_1$ hybrid),
-this would be
-\begin{align*}
-    R_1(p,q) &= \frac{1}{4} \sum_i p_i^2 (1-p_i)^2 x_i^2 
-                + \frac{1}{4} \sum_i q_i^2 (1-q_i)^2 x_i^2 \\
-                &= (R_0(p) + R_0(q))/2 .
-\end{align*}
-If we assume that the populations are at equilibrium, $R_0(p) \approx R_0(q)$,
-and so $R_1(p,q) \approx R_0(p)$.
-
-Now consider an $F_2$ hybrid, where both parents are $F_1$
-and so each heterozygous at locus $i$ with probability $p_i (1-q_i) + (1-p_i) q_i$.
-Then
-\begin{align*}
-    R_2(p,q) &= \frac{1}{4} \sum_i \left(p_i (1-q_i) + (1-p_i) q_i \right) x_i^2  .
-\end{align*}
-Suppose that the two populations are slightly drifted from each other,
-with frequency difference $p_i-q_i = 2\epsilon_i$.
-Then,
-\begin{align*}
-    p (1-q) + p (1-q)
-    &=
-    (u+\epsilon) (1-u+\epsilon) + (u-\epsilon) (1-u-\epsilon) \\
-    &=
-    2 u (1-u)
-    % + \epsilon (u + (1-u) - u - (1-u))
-    + 2 \epsilon^2 .
-\end{align*}
-If the frequencies have evolved neutrally in unconnected, Wright-Fisher populations 
-of effective size $N$ for $t$ generations from a common ancestor with allele frequency $u$,
-then $\epsilon$ has mean zero and variance roughly $u (1-u) t/N$.
-Still assuming the populations are at stationarity,
-so that $R_0$ is constant between the two,
-and taking the frequencies $p_i$ as a proxy for the ancestral frequencies $u_i$,
-this implies that we expect
-\begin{align*}
-    R_2 
-    &\approx 
-    R_0 + \frac{1}{2} \sum_i p_i (1-p_i) x_i^2 t/N \\
-    &=
-    \left( 1 + \frac{t}{N} \right) R_0 .
-\end{align*}
-On the other hand, the expected squared difference in trait \emph{means} here is
-\begin{align} \label{eqn:var_dx}
-    4 \sum_i p_i (1-p_i) x_i^2 t / N = 8 R_0 t /N .
-\end{align}
-This implies that under this model,
-segregation variance in $F_2$s between two populations
-is roughly increased by a factor of $1/8$ of the difference between their means.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Away from the optimum}
-\label{apx:away_from_opt}
-
-Let two points on $\allS$ be $x_1$ and $x_2$, let $\bar x = (x_1+x_2)/2$, and let $z=(x_2 - x_1)/2$.
-Then with $\grad D$ and $\grad^2 D$ the first and second derivatives of $D$, respectively,
-Taylor expanding about $x_1$ and $x_2$ finds that
-\begin{align*}
-    D(\bar x) 
-    &= D(x_1) + \grad D(x_1) \cdot z + \frac{1}{2} z^T \grad^2  D(x_1) z + O(\|z\|^3) \\
-    &=  D(x_2) - \grad D(x_2) \cdot z + \frac{1}{2} z^T \grad^2  D(x_2) z + O(\|z\|^3) .
-\end{align*}
-Now, since $ D(x_1) =  D(x_2) = \grad D(x_1) = \grad D(x_2) = 0$ and
-\begin{align*}
-    \grad D(x_2) = \grad D(x_1) + 2 z^T \grad^2  D(x_1) + O(\|z\|^2), \quad \text{and} \\
-    \grad^2 D(x_2) = \grad^2 D(x_1) + O(\|z\|), 
-\end{align*}
-adding together the two equations above and dividing by two gets that
-\begin{align*}
-     D(\bar x) 
-    &= \optph - \frac{3}{2} z^T \grad^2  D(x_1) z + O(\|z\|^3) .
-\end{align*}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Gaussian load}
-\label{apx:gauss_load}
-
-Suppose that a population has a Gaussian distribution in $d$-dimensional trait space
-with mean $\mu$ and covariance matrix $\Sigma$,
-and that fitness of an individual at $x$ is $\exp(-\|Lx\|^2/2)$.
-Then, completing the square as above with $A=\Sigma^{-1}$, $y=\mu$, and $B=L^T L$,
-and defining $Q = (\Sigma^{-1} + L^T L)^{-1}$,
-\begin{align*}
-    &
-    \frac{1}{\sqrt{2 \pi}^n \det(\Sigma)^{1/2}} 
-        \int e^{-\frac{1}{2} x^T \Sigma^{-1} x} e^{-\frac{1}{2} x^T L^T L x} dx \\
-    &\qquad =
-    \frac{1}{\sqrt{2 \pi}^n \det(\Sigma)^{1/2}} 
-        \int e^{-\frac{1}{2}(x-Q \Sigma^{-1} \mu)^T Q^{-1} (x-Q \Sigma^{-1} \mu)} dx 
-        \\ &\qquad \qquad {}
-        \times e^{\mu^T \left( I - \Sigma^{-1} Q \right) \Sigma^{-1} \mu} \\
-    &\qquad =
-    \sqrt{\frac{\det(Q)}{\det(\Sigma)}}
-        \exp\left\{\mu^T \left( I - \Sigma^{-1} Q \right) \Sigma^{-1} \mu\right\} \\
-    &\qquad =
-    \sqrt{\frac{1}{\det(\Sigma)\det(\Sigma^{-1}+L^TL)}}
-        \exp\left\{\mu^T \left( I - (I + L^T L \Sigma)^{-1} \right) \Sigma^{-1} \mu\right\} \\
-\end{align*}
-
-Now suppose that $\Sigma = \sigma^2 I$ and $L = I/\gamma$.
-Then,
-\begin{align*}
-    \sqrt{\frac{1}{\det(\Sigma)\det(\Sigma^{-1}+L^TL)}}
-    &=
-    \sqrt{\frac{1}{\sigma^{2d} (1/\sigma^2 + 1/\gamma^2)^d}} \\
-    &=
-    \frac{1}{(1+ \left(\sigma/\gamma\right)^2)^{d/2}} .
-\end{align*}
-Also, 
-\begin{align*}
-        \left( I - (I + L^T L \Sigma)^{-1} \right) \Sigma^{-1} 
-        &=
-        \frac{1}{\sigma^2}\left( 1 - (1 + (\sigma/\gamma)^2)^{-1} \right) I \\
-        &=
-        \frac{1}{\gamma^2}\frac{1}{(1 + (\sigma/\gamma)^2)} I \\
-\end{align*}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Differentiating the fitness function}
-\label{apx:H_calc}
-
-\plr{Add refs to sensitivity analysis.}
-
-Suppose that $\rho(t) \ge 0$ is a weighting function on $[0,\infty)$
-so that fitness is a function of $L^2(\rho)$ distance of the impulse response from optimal.
-With $h_0(t) = C_0 e^{tA_0} B_0$ a representative of the optimal set:
-\begin{equation}
-    \begin{aligned}
-        D(A, B, C) 
-        &:= 
-        \int_0^\infty \rho(t) \left| h_A(t) - h_0(t) \right|^2 dt \\
-        &:= 
-        \int_0^\infty \rho(t) \left| C e^{At} B - C_0 e^{A_0 t} B_0 \right|^2 dt \\
-        &= 
-        \int_0^\infty \rho(t) \tr\left\{
-            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)^T
-            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)
-        \right\} dt \\
-        &= 
-        \int_0^\infty \rho(t) \tr\left\{
-            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)
-            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)^T
-        \right\} dt  .
-        % &= 
-        % \int_0^\infty \rho(t)  \tr\left\{
-        %         C e^{At} B B^T e^{A^T t} C^T 
-        %         - C_0 e^{A_0t} B_0 B^T e^{A^T t} C^T 
-        %         \right. \\ &\qquad \qquad \left. {}
-        %         - C e^{At} B B_0^T e^{A_0^T t} C_0^T 
-        %         + C_0 e^{A_0t} B_0 B_0^T e^{A_0^T t} C_0^T 
-        % \right\} dt \\
-        % \int_0^\infty \rho(t) \left| C \left( e^{At} - e^{A_0 t} \right) B \right|^2 dt \\
-        % &= 
-        % \int_0^\infty \rho(t) C \left( e^{At} - e^{A_0 t} \right) B B^T \left( e^{At} - e^{A_0 t} \right)^T C^T dt
-    \end{aligned}
-\end{equation}
-How does this change as we perturb about $(A_0, B_0, C_0)$?
-First we differentiate with respect to $A$, keeping $B=B_0$ and $C=C_0$ fixed.
-Since
-\begin{equation}
-  \begin{aligned}
-      \frac{d}{du} e^{(A+uZ)t} \vert_{u=0}
-      &=
-      \int_0^t e^{As} Z e^{A(t-s)} ds, 
-  \end{aligned}
-\end{equation}
-we have that
-\begin{equation}
-  \begin{aligned}
-      \frac{d}{du} D(A+uZ,B_0,C_0)\vert_{u=0}
-      &=
-        2 \int_0^\infty \rho(t) \tr\left\{ C_0 \left( \int_0^t e^{As} Z e^{A(t-s)} ds \right) B_0 B_0^T \left( e^{At} - e^{A_0 t} \right)^T C_0^T \right\} dt \\
-      &=
-        2 \int_0^\infty \rho(t) \tr\left\{ C_0 \left( \int_0^t e^{As} Z e^{A(t-s)} ds \right) B_0 \left( h_A(t) - h_0(t) \right)^T \right\} dt 
-  \end{aligned}
-\end{equation}
-and, by differentiating this and supposing that $A$ is on the optimal set,
-i.e., $h_A(t)=h_0(t)$, (so wolog $A=A_0$):
-\begin{equation}
-  \begin{aligned}
-      \calH^{A,A}(Y,Z) 
-      &:= 
-      \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0+uY+vZ,B_0,C_0)\vert_{u=v=0} \\
-      &=
-        \int_0^\infty \rho(t) \tr\left\{ C_0 
-        \left( \int_0^t e^{A_0 s} Y e^{A_0 (t-s)} ds \right) 
-        B_0 B_0^T 
-        \left( \int_0^t e^{A_0 s} Z e^{A_0 (t-s)} ds \right)^T
-        C_0^T \right\} dt  .
-  \end{aligned}
-\end{equation}
-
-The function $\calH$ will define a quadratic form.
-To illustrate the use of this, suppose that $B$ and $C$ are fixed.
-% Here $\calH$ is the quadratic form underlying the Hamiltonian.
-By defining $\Delta_{ij}$ to be the matrix with a 1 in the $(i,j)$th slot
-and 0 elsewhere,
-the coefficients of the quadratic form is
-\begin{equation}
-    \begin{aligned}
-        H_{ij, k\ell}(A)
-        &:=
-        \calH(\Delta_{ij}, \Delta_{k\ell}) .
-    \end{aligned}
-\end{equation}
-
-We could use this to compute the gradient of $D$,
-or to get the quadratic approximation to $D$ near the optimal set.
-To do so, it'd be nice to have a way to compute the inner integral above.
-Suppose that we can diagonalize $A = U \Lambda U^{-1}$.
-Then
-\begin{equation} \label{eqn:exp_deriv}
-  \begin{aligned}
-      \int_0^t e^{As} Z e^{A(t-s)} ds 
-      &=
-      \int_0^t U e^{\Lambda s} U^{-1} Z U e^{\Lambda (t-s)} U^{-1} ds 
-  \end{aligned}
-\end{equation}
-Now, notice that
-\begin{equation}
-  \begin{aligned}
-      \int_0^t e^{s \lambda_i} e^{(t-s) \lambda_j} ds
-      &=
-      \frac{ e^{t \lambda_i} - e^{t \lambda_j} }{ \lambda_i - \lambda_j } .
-  \end{aligned}
-\end{equation}
-Therefore, 
-defining
-\begin{equation}
-    X_{ij}(t,Z) = \left( U^{-1} Z U \right)_{ij}
-      \frac{ e^{t \lambda_i} - e^{t \lambda_j} }{ \lambda_i - \lambda_j } 
-\end{equation}
-moving the $U$ and $U^{-1}$ outside the integral and integrating we get that
-\begin{equation}
-  \begin{aligned}
-      \int_0^t e^{As} Z e^{A(t-s)} ds 
-      &=
-      U X(t,Z) U^{-1} .
-  \end{aligned}
-\end{equation}
-
-Following on from above, we see that if $Z=\Delta_{k \ell}$, then
-\begin{equation}
-  \begin{aligned}
-      X_{ij}^{k\ell}(t) = 
-      \frac{ e^{t \lambda_i} - e^{t \lambda_j} }{ \lambda_i - \lambda_j } 
-      (U^{-1})_{\cdot k} U_{\ell \cdot},
-  \end{aligned}
-\end{equation}
-where $U_{k \cdot}$ is the $k$th row of $U$,
-and so
-\begin{equation}
-    \begin{aligned}
-        H_{ij, k\ell}(A)
-        &=
-        \int_0^\infty
-            \rho(t) \tr\left\{ C U X^{ij}(t) U^{-1} B B^T (U^{-1})^T X^{k\ell}(t)^T U^T C^T \right\}
-        dt .
-    \end{aligned}
-\end{equation}
-This implies that
-\begin{equation}
-    \begin{aligned}
-        D(A_0+\epsilon Z)
-        &\approx \epsilon^2\sum_{ijk\ell} H^{ij,k\ell} Z_{ij} Z_{k\ell} 
-    \end{aligned}
-\end{equation}
-and so
-\begin{equation}
-    \begin{aligned}
-        D(A_0+\epsilon Z)
-        &\approx \epsilon^2\sum_{ijk\ell} H^{ij,k\ell} Z_{ij} Z_{k\ell} 
-    \end{aligned}
-\end{equation}
-
-By section \ref{ss:quant_gen},
-if we set $\Sigma=\sigma^2 I$ and $U=H$,
-then a population at $A_0+Z$ experiences a restoring force of strength
-$(I + \sigma^2 H^{-1})^{-1} Z$ (treating $Z$ as a vector and $H$ as an operator on these).
-If $\sigma^2$ is small compared to $H^{-1}$
-then this is approximately $-\sigma^2 H^{-1} Z$.
-This suggests that the population mean follows an Ornstein-Uhlenbeck process,
-as described (in different terms) in \citet{hansen1996translating}.
-
-More generally, $B$ and $C$ may also change.
-To extend this we need the remaining second derivatives of $D$.
-First, in $B$:
-\begin{equation}
-    \begin{aligned}
-        \mathcal{H}^{B,B}(Y,Z)
-        &:= 
-        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0,B_0+uY+vZ,C_0)\vert_{u=v=0} \\
-        &=
-        \frac{1}{2} \int_0^\infty \rho(t) \tr\left\{ 
-        C_0 e^{t A_0}
-        \frac{d}{du} \frac{d}{dv} 
-        (uY+vZ)
-        (uY+vZ)^T
-        \vert_{u=v=0} 
-        e^{t A_0^T} C_0^T 
-        \right\} dt  \\
-        &=
-        \frac{1}{2} \int_0^\infty \rho(t) \tr\left\{ 
-        C_0 e^{t A_0}
-        \left( Y Z^T + Z Y^T \right)
-        e^{t A_0^T} C_0^T 
-        \right\} dt  .
-  \end{aligned}
-\end{equation}
-Next, in $C$:
-\begin{equation}
-    \begin{aligned}
-        \mathcal{H}^{B,B}(Y,Z)
-        &:= 
-        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0,B_0,C_0+uY+vZ)\vert_{u=v=0} \\
-        &=
-        \frac{1}{2} \int_0^\infty \rho(t) \tr\left\{ 
-        B_0 e^{t A_0^T}
-        \frac{d}{du} \frac{d}{dv} 
-        (uY+vZ)^T
-        (uY+vZ)
-        \vert_{u=v=0} 
-        e^{t A_0} B_0
-        \right\} dt  \\
-        &=
-        \frac{1}{2} \int_0^\infty \rho(t) \tr\left\{ 
-        B_0 e^{t A_0^T}
-        \left( Y Z^T + Z Y^T \right)
-        e^{t A_0} B_0
-        \right\} dt  .
-  \end{aligned}
-\end{equation}
-Now, the mixed derivatives in $B$ and $C$:
-\begin{equation}
-    \begin{aligned}
-        \mathcal{H}^{B,C}(Y,Z)
-        &:= 
-        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0,B_0+uY,C_0+vZ)\vert_{u=v=0} \\
-        &=
-        \int_0^\infty \rho(t) \tr\left\{ 
-        Y e^{t A_0^T} C_0^T Z e^{t A_0} B_0
-        \right\} dt  .
-  \end{aligned}
-\end{equation}
-In $A$ and $B$
-\begin{equation}
-    \begin{aligned}
-        \mathcal{H}^{A,B}(Y,Z)
-        &:= 
-        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0+uY,B_0+vZ,C_0)\vert_{u=v=0} \\
-        &=
-        \int_0^\infty \rho(t) \tr\left\{ 
-        C_0 \left(\int_0^t e^{s A_0} Y e^{(t-s) A_0} ds \right) B_0
-        Z^T e^{t A_0} C_0
-        \right\} dt  ,
-  \end{aligned}
-\end{equation}
-and finally in $A$ and $C$:
-\begin{equation}
-    \begin{aligned}
-        \mathcal{H}^{A,C}(Y,Z)
-        &:= 
-        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0+uY,B_0,C_0+vZ)\vert_{u=v=0} \\
-        &=
-        \int_0^\infty \rho(t) \tr\left\{ 
-        C_0 \left(\int_0^t e^{s A_0} Y e^{(t-s) A_0} ds \right) B_0
-        B_0 e^{t A_0} Z
-        \right\} dt  .
-  \end{aligned}
-\end{equation}
-
-%\section{Hybrid Vigor: Unit Circle approach}
-%
-%If a fitness optimum is at the origin $(0,0) \in \R^{2}$,
-%%and two populations have drifted away from the optimum by an distance of $r$, then the expected value of their $F_1$ hybrids is,
-%%\begin{align*}
-%%  \mathbb{E}\left[ \left\lVert \frac{1}{2} ((x_1,y_1) + (x_2,y_2))  \right\rVert \right] = \\
-%%  \frac{\sqrt{2}}{8}\left(\sqrt{3\, r^2 - {\sin^{-1}\!\left(r\right)}^2 + r^4 - 2\, r\, \sin^{-1}\!\left(r\right)\, \sqrt{1 - r^2}} + \sqrt{{\sin^{-1}\!\left(r\right)}^2 + 5\, r^2 - r^4 + 2\, r\, \sin^{-1}\!\left(r\right)\, \sqrt{1 - r^2}}\right)
-%%\end{align*} 
-%%If $r < \sim 5.5$ then we will observe hybrid vigor, however if $r$ is sufficiently large, the hybrids will have lower fitness, on average. 
-%and two populations have drifted away from the optimum by a distance of $r=1$,
-%then $F_1$s will have an average fitness of,
-%  \begin{align*}
-%    \mathbb{E} \left[ d(F_1) \right] &= \int_{0}^{2\pi} \frac{1}{4 \pi} \left\lVert (1,0) + (\cos\theta,\sin\theta) \right\rVert d\theta \\
-%    &= \frac{2}{\pi}
-%  \end{align*}
-%  %\begin{align*}
-%  %  \mathbb{E} \left[ d(F_1)] \right] &= \int_{0}^{2\pi} \int_{0}^{2\pi} \frac{1}{8 \pi^{2}}\left\lVert  \left( \left( \cos(\theta_1), \sin(\theta_1) \right) , \left( \cos(\theta_2), \sin(\theta_2) \right) \right) \right\rVert d\theta_1 d\theta_2 \\
-%  %  &= \cos\left(\frac{\pi}{2}\right) = \frac{\sqrt{2}}{2}
-%  %\end{align*}
-%%  \begin{align*}
-%%    \mathbb{E} \left[ d(F_1)] \right] &\approx \sqrt{\int_{0}^{2\pi} \frac{1}{4 \pi}\left\lVert  \left( \cos(\pi), \sin(\pi) \right) , \left( \cos(\theta), \sin(\theta) \right) \right\rVert^2 d\theta} \\
-%%    &= \sqrt{\frac{1}{2}}
-%%  \end{align*}
-%%Haploid $F_2$s will, on average, be a distance of $1$ half of the time, a distance of $0$ a quarter of the time, and a distance of $\sqrt{2}$ a quarter of the time. 
-%
-%  \begin{align*}
-%    \mathbb{E} \left[ d(F_2) \right] &= \int_{0}^{2 \pi} \int_{0}^{2 \pi} \frac{1}{4 \pi^{2}} \sum_{i}^{2} \sum_{j}^{2} \sum_{k}^{2} \sum_{l}^{2} \frac{1}{32} \left\lVert \left( \cos(\theta_i) + \cos(\theta_k), \sin(\theta_j) + \sin(\theta_l) \right) \right\rVert d\theta_1 d\theta_2 \\ &= 0.6705
-%  \end{align*}
-%
-%  \begin{align*}
-%    \mathbb{E}\left[ d(F_{2_{\text{hap}}}) \right] &= \sqrt{\int_{0}^{2 \pi} \int_{0}^{2 \pi} \frac{1}{8 \pi^{2}} \cos(\theta_1)^{2} + \sin(\theta_2)^{2} d\theta_1 d\theta_2} \\
-%    &= \cos \left( \frac{\pi}{2} \right) = \frac{\sqrt{2}}{2}
-%  \end{align*}
-%
-%%  There are $16$ diploid $F_2$s, two have $F_0$ genotypes, two have $F_1$ genotpyes, eight contrain one $F_0$ and one $F_2$ haploid genotype (referred to as $H$; $\fit(H) = \sqrt{\frac{3}{8}}$), two are homozygous for an $F_2$ genotype (referred to as $W$; $\fit(W) = \frac{\sqrt{2}}{2}$), and finally two are heterozygous $F_2$s (referred to as $L$; $\fit(L) = \frac{1}{2}$.)
-%  %The expected average distance of diploid $F_2$s is $0.6705$. 
-%
-%  \section{Hybrid Vigor: Probability}
-%
-%  \begin{align*}
-%    X &\sim \mathcal{N} (x, \sigma^{2}) \\
-%    Y &\sim \mathcal{N} (y, \sigma^{2}) \\
-%    x &\sim \mathcal{N} (0, \eta^{2}) \\
-%    y &\sim \mathcal{N}(0, \eta^{2}) \\
-%    Z &= \frac{X+Y}{2} \rightarrow \mathbb{E} \left[ \lvert z \rvert^{2} \right] \\
-%    Z &= \frac{x+y+U+V}{2} \\
-%    \mathbb{E} \left[ \lvert X \rvert^{2} \right] &= \mathbb{E} \left[ \lvert x + U \rvert^{2} \right] = \mathbb{E} \left[ (x+U)^{T} (x+U) \right] \\
-%    &= \mathbb{E} \left[ x^{T}x + U^{T}x + x^{T}U + U^{T}T \right] \\
-%    &= \mathbb{E} \left[x^{T}x\right] + \mathbb{E} \left[U^{T}U\right] \\
-%    &= d\eta^{2} + d\sigma^{2} \qquad d=\text{dimension} \\
-%    X &\sim \mathcal{N}(0, \sigma^{2} + \eta^{2}) \\
-%    Z &\sim \mathcal{N}\left(0, \frac{1}{4}(2 \eta^{2} + 2\sigma^{2})\right) \\
-%    &\rightarrow \mathbb{E}\left[\lvert Z \rvert^{2} \right] \\
-%    &= \frac{d}{2} \left(\eta^{2} + \sigma^{2} \right) \\
-%    &= \frac{1}{2} \mathbb{E} \left[ \lvert X \rvert^{2} \right]
-%  \end{align*}
-
-\end{document}

### 2017-12-05 speciation_rates.R

@@ -14,14 +14,14 @@ f1 <- function (t, a, b, c_mu=1/20) {
     exp(-(c_mu * t)^2 / (1 + b + 4 * b * t))
 }
 
-plotf <- function (main, Ne=1.0, tmax=10,
+plotf <- function (main, Ne=1.0, tmax=5,
         ...) {
     tt <- seq(0, tmax/Ne, length.out=200)
     plot(tt*Ne, f(tt, ...), type='l', ylim=c(0,1), main=main,
             xlab=if (Ne==1) { expression(T/N[e])} else { expression(T) }, 
             ylab="F2 fitness")
     lines(tt*Ne, f1(tt, ...), col='red', lty=3)
-    lines(tt*Ne, f2(tt, ...), col='green', lty=3)
+    # lines(tt*Ne, f2(tt, ...), col='green', lty=3)
 }
 
 pdf(file="speciation_rates.pdf", width=7, height=3, pointsize=10)
@@ -31,8 +31,11 @@ plotf(a=1, b=1, main="large population")
 plotf(a=.1, b=.1, main="small population")
 plotf(a=10, b=1, main="metapopulation")
 
-legend("topright", lty=c(1,3,3), col=c('black', 'red', 'green'),
-        legend=c("total fitness", "shift of mean", "increase of variance"))
+legend("topright", lty=c(1,3), col=c('black', 'red', 'green'), bg='white',
+        legend=c("F2 fitness", "F1 fitness"))
+
+# legend("topright", lty=c(1,3,3), col=c('black', 'red', 'green'),
+#         legend=c("total fitness", "shift of mean", "increase of variance"))
 
 dev.off()
 

### 2017-12-06 kalman_walk_paper.tex
added paragraphs on quant gen and islands

@@
 This evolutionary exploration, or \emph{system drift}, 
-proceeds at a rate inversely proportional the amount of intrapopulation genetic variation
+proceeds at a rate proportional to the amount of intrapopulation genetic variation
 divided by the effective population size ($N_e$).
@@
 \end{abstract}
-
-\plr{I removed the $0.1 N_e$ because that's only in the "metapopulation" case,
-which is where you've probably already got subpopulations acting like species already.}
 
@@
 \citep{kimura1981possibility, true2001developmental, tanay2005conservation, tsong2006evolution, hare2008sepsid, lavoie2010evolutionary, vierstra2014mouse, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription}.
+It has also been shown theoretically that substantial genetic variation for a polygenic trait can be maintained
+by migration between geographical regions having different genetic bases for the same, optimal trait value \citep{phillips1996maintenance}.
+\plr{the citation is "maintenance of polygenic variation via a migration-selection balance..."}
 
@@
 and that this may lead to speciation.
-For instance, \citet{tulchinsky2014hybrid} looked for hybrid incompatibility
+For instance, \citet{tulchinsky2014hybrid} (and similarly \citet{khatari2015simple}) looked for hybrid incompatibility
 following regulatory sequence drift,
@@
 but only equivocal support for speciation under models of purely neutral drift.
+Additionally, \citet{fierst} observed speciation due to an increase of segregation variance under balancing selection, in models of weak to moderate, but not strong epistasis.
 Our model differs from previous work in that we use linear systems theory
@@
 that would be inaccessible to population simulations.
-\plr{missing any important references here?}
 
@@
     dotted lines give phenotypic distances to the optimum.
-    Two pairs of parental populations are shown in black, along the optimum;
-    a hypothetical population of $F_1$s are shown for each in red,
+    A pair of parental populations are shown in black, along the optimum;
+    a hypothetical population of $F_1$s are shown in red,
     and the distribution of one type of $F_2$ is shown in purple
-    (other types of $F_2$ are not shown).
-    Solid lines depict the distance of the $F_2$ to optimum.
+    (other types of $F_2$ are not shown; 
+    some would be a similar distance to the other side of the optimal set).
+    The distribution of $F_2$s is appropriate for mixed homozygotes
+    if both traits have a simple, one-locus genetic basis,
+    but there is variation within each population at that locus.
 }
@@
 Is there evidence that this is actually occurring?
-System drift and network rewiring has been inferred across the tree of life \citep{dalal2017transcription, johnson2017rewiring},
+System drift and network rewiring has been inferred across the tree of life \citep{wotton2015quantitative, crombach2016gap, dalal2017transcription, johnson2017rewiring},
 and there is often significant regulatory variation segregating within populations.
 Transcription in hybrids between closely related species with conserved transcriptional patterns can also be divergent 
-\citep{haerty2006gene, maheshwari2012cis, coolon2014tempo, michalak2004association, mack2016gene}.
+\citep{haerty2006gene, maheshwari2012cis, coolon2014tempo, michalak2004association, mack2016gene}, and hybrid incompatibilities have been attributed to cryptic molecular divergence underlying conserved body plans \citep{gavin2013embryonic}.
 Furthermore, in cryptic species complexes (e.g., sun skinks \citep{barley2013challenge}),
@@
 
+\paragraph{Population isolates and genetic load}
+Above we have mostly discussed the case of two separated populations of equal size.
+In contrast, consider a small subset of a large population that is suddenly isolated.
+It will have a large amount of genetic variance, thanks to its history,
+but a small population size -- and so can drift quite quickly until it exhausts its genetic variation.
+By this point, it could have accumulated substantial incompatibility with the parental population.
+Similarly, in a species with substantial isolation by distance, geographically local systems drift
+might lead to genetic load due only to recombination, much as in \citep{phillips1996maintenance}.
+
 \paragraph{Nonlinearity and model assumptions}
@@
 Furthermore, nonidentifiability (which implies the existence of neutral directions)
-is often found in practice in moderately complex models of biological systems. 
-\plr{Josh, add citation to sensitivity analysis}
+is often found in practice in moderately complex models of biological systems \citep{gutenkunst2007universally, piazza2008diverse}.
+
+This simple quantitative genetics model we use above
+has been shown to produce good predictions in many situations,
+even when the substantial number of simplifying assumptions are violated \citep{burger1994distribution,turelli1994genetic}.
+\plr{the citations are to "on the distribution of the mean and variance" by Burger and Lande; and the "what, me normal" paper of turelli and barton}
+The calculations above should be fairly robust even to substantial deviations from normality.
+A larger effect on these predictions seems likely due to 
+correlations due to molecular constraint, genetic linkage, population structure, historical contingency and so forth.
+Although such considerations would not change the qualitative predictions of this model,
+their combined effects could substantially change the predicted rate of accumulation of incompatibilities.
 
@@
 
-\paragraph{Peter to add quant gen discussion} about non-normality.
-
 %Maybe compare to \citet{fierst}?
-%
-%Do we talk about the $G$ matrix and mutational correlation somewhere?
-%
-%Note that islands or bottlenecks leave large $\sigma_N$ but small $N_e$ so may go fast.
-%
-%What is Ne?  In a metapopulation depends on migr rate: classic Turelli paper.
-%
-%This has implications for genetic load also!!
 %

### 2017-12-06 kalman_walk_paper.tex
added requested citations, other minor edits

--- 2017-12-05 kalman_walk_paper.tex
@@
 Furthermore, nonidentifiability (which implies the existence of neutral directions)
-is often found in practice in moderately complex models of biological systems. 
-\plr{Josh, add citation to sensitivity analysis}
+is often found in practice in moderately complex models of biological systems \citep{gutenkunst2007universally, piazza2008diverse}.
 

### 2017-12-06 kalman_walk_paper.tex
added citations for empirical allele-specific expression data for flycatchers and baboons

--- 2017-12-06 kalman_walk_paper.tex
@@
 and that this may lead to speciation.
-For instance, \citet{tulchinsky2014hybrid} (and similarly \citet{khatari2015simple}) looked for hybrid incompatibility
+For instance, \citet{tulchinsky2014hybrid} (and similarly \citet{khatri2015simple}) looked for hybrid incompatibility
 following regulatory sequence drift,
@@
 explained about 3\% of expression variation.
-%%\plr{Get some data from at least one other species in here!}
+Studying natural flycatcher populations, \citet{wang2017bayesian} found that 7.2--8.5\% of transcripts showed allele-specific expression -- indicitive of standing genetic \emph{cis}-regulatory variation.
+Similarly, in wild baboons, \citet{tung2015genetic} observed allele-specific expression in 23.4\% of genes studied.
 Taken together, this suggests that variation in the entries of $A$
@@
 doubtless varying substantially between species and between genes.
-\plr{JOSH: insert a citation or two to something non-human?}
 
@@
 Is there evidence that this is actually occurring?
-System drift and network rewiring has been inferred across the tree of life \citep{wotton2015quantitative, crombach2016gap, dalal2017transcription, johnson2017rewiring},
+System drift and network rewiring has been inferred across the tree of life \citep{wotton2015quantitative, crombach2016gap, dalal2017transcription, johnson2017rewiring, ali2017quantitative},
 and there is often significant regulatory variation segregating within populations.

### 2017-12-06 kalman_walk_paper.tex

@@
 
+This simple quantitative genetics model we use above
+has been shown to produce good predictions in many situations,
+even when the substantial number of simplifying assumptions are violated \citep{burger1994distribution,turelli1994genetic}.
+\plr{the citations are to "on the distribution of the mean and variance" by Burger and Lande; and the "what, me normal" paper of turelli and barton}
+The calculations above should be fairly robust even to substantial deviations from normality.
+A larger effect on these predictions seems likely due to 
+correlations due to molecular constraint, genetic linkage, population structure, historical contingency and so forth.
+Although such considerations would not change the qualitative predictions of this model,
+their combined effects could substantially change the predicted rate of accumulation of incompatibilities.
+
 Finally, despite our model's precise separation of phenotype and kryptotype, this relationship in nature may be far more complicated as aspects of the kryptotype may be less ``hidden'' than we currently assume.

### 2017-12-06 examples/F2_osc_plotting.R
fixed figure 6 plot

@@ -85,18 +85,18 @@ f2_divergence <- function(tau, eps) {
   progeny <- progeny_gen(tau,eps)
   avg_div <- 0
   for (i in 1:256){
-    avg_div <- avg_div + D(progeny[[i]])/256
+    avg_div <- avg_div + sqrt(D(progeny[[i]]))/256
   }
   return(avg_div)
 }
 
-d <- seq(0,-1,-0.01)
+d <- seq(0,-0.5,-0.01)
 df2 <- sapply(d, function(d) f2_divergence(0,d))
 
 ##########
 
 f1_divergence <- function(tau, eps){
-  f1div <- D((S(tau) + S(tau+eps))/2)
+  f1div <- sqrt(D((S(tau) + S(tau+eps))/2))
   return(f1div)
 }
 
@@ -106,7 +106,7 @@ df1 <- sapply(d, function(d) f1_divergence(0,d))
 gen_d <- sapply(d, function(d) norm(S(0)-S(0+d),"F"))
 
 pdf("~/kalman_walk/examples/F2_vs_F1_divergence_tau0.pdf")
-plot(df1~gen_d, ylim=c(0,15), type = "l", lwd=3, col="red", xlab="Genetic distance", ylab="Phenotypic divergence")
+plot(df1~gen_d, ylim=c(0,2), type = "l", lwd=3, col="red", xlab="Genetic distance", ylab="Phenotypic divergence")
 lines(df2~gen_d, lwd=3, col="blue")
-legend(0.2,10, legend=c("F1 hybrids", "F2 hybrids"), col=c("red", "blue"), lty=1, lwd=3)
+legend(0.2,1.5, legend=c("F1 hybrids", "F2 hybrids"), col=c("red", "blue"), lty=1, lwd=3)
 dev.off()

### 2017-12-06 kalman_walk_paper.tex
fixed figure 6 plot
added missing citations to bib doc

@@
 by migration between geographical regions having different genetic bases for the same, optimal trait value \citep{phillips1996maintenance}.
-\plr{the citation is "maintenance of polygenic variation via a migration-selection balance..."}
 
@@
     \caption{Phenotype distance to optimum, $D$, using $\rho(t) = \exp(-t/4\pi)$
-    for $F_1$ (red) and $F_2$ (blue) hybrids between an $A(0)$ and an $A(\epsilon)$ parent. Genetic distance is computed as the average coefficient difference between two systems.}
+    for $F_1$ (red) and $F_2$ (blue) hybrids between an $A(0)$ and an $A(\epsilon)$ parent. Genetic distance is computed as the Frobenius norm between two systems.}
     %\textbf{(right)} Same, but for $F_2$ offspring (one line per type of $F_2$).
@@
 the amount of genetic variance ($\sigma_N^2$) is proportional to $N_e$,
-\plr{that citation is Lynch \& Hill 1986, "phenotypic evolution by neutral mutation"}
 which would compensate for this difference, 
@@
 It is generally thought that regulatory sequence change contributes much more to inter- and intraspecific variation
-than does coding sequence change affecting molecular structure \citep{schmidt2010vertebrate}.
+than does coding sequence change affecting molecular structure \citep{schmidt2010fivevertebrate}.
 In the context of transcription factor networks this may be affected 
@@
 even when the substantial number of simplifying assumptions are violated \citep{burger1994distribution,turelli1994genetic}.
-\plr{the citations are to "on the distribution of the mean and variance" by Burger and Lande; and the "what, me normal" paper of turelli and barton}
 The calculations above should be fairly robust even to substantial deviations from normality.

### 2017-12-06 kalman_walk_paper.tex
reparameterized and some formatting edits

@@
 \begin{document}
-\linenumbers
+%\linenumbers
 
 {\centering
-{\Large \bf Rapid speciation despite conservation of phenotype} \\
-Joshua S. Schiffman$^{\dagger}$ \qquad Peter L. Ralph$^{\dagger \ddagger}$ \\
-$^{\dagger}$University of Southern California, Los Angeles, California \qquad 
-$^{\ddagger}$University of Oregon, Eugene, Oregon \\
-\texttt{jsschiff@usc.edu} \qquad 
-\texttt{plr@uoregon.edu}
-\\
+{\LARGE \bf Rapid speciation despite conservation of phenotype} \\ \vspace{0.5cm}
+Joshua S. Schiffman$^{\dagger}$ \qquad Peter L. Ralph$^{\dagger \ddagger}$ \\ \vspace{0.5cm}
+$^{\dagger}${\footnotesize {Molecular and Computational Biology, University of Southern California, Los Angeles, California 90089, U.S.A. \\
+$^{\ddagger}$Departments of Mathematics and Biology \& The Institute for Ecology and Evolution, University of Oregon, Eugene, Oregon 97403, U.S.A.}} \\ \vspace{0.5cm}
+{\small \texttt{jsschiff@usc.edu} \qquad 
+\texttt{plr@uoregon.edu}} \\ \vspace{0.5cm}
+\small \today \\
+\vspace{0.25cm}
 }
@@
 and that the evolutionary exploration of these distinct but equivalent mechanisms
-can lead to the reproductive incompatibility of idependently evolving populations.
+can lead to the reproductive incompatibility of independently evolving populations.
 This evolutionary exploration, or \emph{system drift}, 
@@
 isolated for a sufficiently long period of time, 
-even in the absence of any sort of adaptative, selective, or environmental change. 
+even in the absence of any sort of adaptive, selective, or environmental change. 
 We then use quantitative genetic theory to estimate how quickly 
@@
 that would be inaccessible to population simulations.
-
-REFERENCE: Barton's paper (PETER). \citep{barton2010natural}
+%REFERENCE: Barton's paper (PETER). \citep{barton2010natural}
 
@@
 We first describe the class of models we study, motivated by gene regulatory networks. 
-The behavior (i.e., the temporal dynamics) of such a system is determined by a collection of $n$ coregulating moleculues 
+The behavior (i.e., the temporal dynamics) of such a system is determined by a collection of $n$ coregulating molecules 
 -- such as transcription factors -- as well as by external or environmental inputs.
@@
   \centering
-   %  \begin{tabular}{cc}
+     \begin{tabular}{cc}
         \begin{tikzpicture}
@@
             \node[shape=rectangle] (U) at (2,2) {input ($u$)};
-            \node[shape=rectangle] (y) at (2,-2) {\begin{tabular}{cc} output ($\phi$) \\  \includegraphics[width=0.15\textwidth, height=0.04\paperheight]{osc_impulse} \end{tabular}};
+            \node[shape=rectangle] (y) at (2,-2) {output ($\phi$)};
+            %\\  \includegraphics[width=0.15\textwidth, height=0.04\paperheight]{osc_impulse} \end{tabular}};
         \end{scope}
@@
 
-        \end{tikzpicture} %&
-%   \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{osc_impulse}
-%\end{tabular}
+        \end{tikzpicture} \\
+   \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{osc_impulse}
+\end{tabular}
 %  \end{center}
   \caption{
-    %(Left) 
+    (Top) 
     Diagram of the gene network in Example \ref{ex:oscillator}
-  %, and (right) plot of the phenotype $\phi(t)$ against time $t$
+  , and (bottom) plot of the phenotype $\phi(t)$ against time $t$
   .} \label{fig:oscillator}
@@
     \begin{align*}
-      A(\tau) &= \frac{1}{\tau-1} \begin{bmatrix} \tau & -1 \\ 2 \tau(\tau - 1) + 1 &  -\tau \end{bmatrix} \ \text{for} \ \tau \neq 1 .
+      A(\tau) &= \frac{-1}{\tau+1} \begin{bmatrix} -\tau & -1 \\ 2 \tau(\tau + 1) + 1 &  \tau \end{bmatrix} \ \text{for} \ \tau \neq -1 .
     \end{align*}
@@
               every edge/.style={draw=black, thick}]
-    \path [->, sloped] (A) edge[bend left] node {\tiny $2 \tau + (\tau-1)^{-1}$} (B);
-    \path [->, sloped] (B) edge[bend left] node {\tiny $-(\tau-1)^{-1}$} (A); 
+    \path [->, sloped] (A) edge[bend left] node {\tiny $-2 \tau - \frac{1}{\tau+1}$} (B);
+    \path [->, sloped] (B) edge[bend left] node {\tiny $\frac{1}{\tau+1}$} (A); 
     \path[->] (U) edge node {\tiny $1$} (A);
@@
               every edge/.style={draw=black, thick}]
-    \path [->] (A) edge[loop left] node[sloped, anchor=center, above] {\tiny $1 + (\tau-1)^{-1}$} (A);
-    \path [->] (B) edge[loop right] node[sloped, anchor=center, above] {\tiny $-1 - (\tau-1)^{-1}$} (B);
+    \path [->] (A) edge[loop left] node[left] {\tiny $\frac{\tau}{\tau+1}$} (A);
+    \path [->] (B) edge[loop right] node[right] {\tiny $\frac{-\tau}{\tau+1}$} (B);
 \end{scope}
@@
       (Left) $A(\tau)$, the set of all phenotype-equivalent cell cycle control networks.
-      (Right) Gene-1 dynamics (blue) for both systems $A(0)$ and $A(2)$ are identical, however, $A(0)$ gene-2 dynamics (orange) differ from $A(2)$ (green).
+      (Right) Gene-1 dynamics (blue) for both systems $A(0)$ and $A(-2)$ are identical, however, $A(0)$ gene-2 dynamics (orange) differ from $A(-2)$ (green).
       % Both gene-1 dynamics are given by $\kappa_{1}(t) = \sin t + \cos t$, and gene-2 by $\kappa_{2}(t) = \cos t - \sin t$ ($A(0)$) and $\kappa'_{2}(t) = \cos t + 3 \sin t$ ($A(2)$).
@@
               every edge/.style={draw=black, thick}]
-    \path [->] (A) edge[loop below] node[sloped, anchor=center, below] {\tiny $1$} (A);
-    \path [->] (B) edge[loop above] node[sloped, anchor=center, above] {\tiny $1$} (B);
-    \path [->, >=Rectangle] (C) edge[loop right] node[sloped, anchor=center, above] {\tiny $-1$} (C);
+    \path [->] (A) edge[loop below] node[sloped, below] {\tiny $1$} (A);
+    \path [->] (B) edge[loop above] node[sloped, above] {\tiny $1$} (B);
+    \path [->, >=Rectangle] (C) edge[loop right] node[sloped, above] {\tiny $-1$} (C);
 \end{scope}
@@
 How quickly do hybrid phenotypes break down as genetic distance between parents increases?
-We will quanitfy how far a system's phenotype is from optimal
+We will quantify how far a system's phenotype is from optimal
 using a weighted difference between impulse response functions.
@@
 can easily fail to oscillate.
-For instance, the $F_1$ offspring between homozygous parents at $\tau=0$ and $\tau=2$
+For instance, the $F_1$ offspring between homozygous parents at $\tau=0$ and $\tau=-2$
 has phenotype $\phi_{F_1}(t) = e^t$, rather than $\phi(t) = \sin t + \cos t$.
@@
   \caption{
-    \textbf{(left)} Phenotypes of $F_1$ hybrids between an $A(0)$ parent and, top-to-bottom, an $A(-\sfrac{1}{100})$, an $A(-\sfrac{1}{10})$, and $A(-\sfrac{1}{2})$ parent;
+    \textbf{(left)} Phenotypes of $F_1$ hybrids between an $A(0)$ parent and, top-to-bottom, an $A(\sfrac{1}{100})$, an $A(\sfrac{1}{10})$, and $A(\sfrac{1}{2})$ parent;
     parental coefficients differ by around 0.5\%, 5\%, and 25\% respectively.
-    \plr{These figures are fine; but reparameterize so that the values of $\tau$ are positive.}
+    %\plr{These figures are fine; but reparameterize so that the values of $\tau$ are positive.}
     Parental phenotypes ($\sin t + \cos t$) are shown in solid black, and hybrid phenotypes in dashed blue.
@@
 the population mean coefficient will move a random distance of size $\sigma/\sqrt{N_e}$ per generation,
-simply because this is the the standard deviation of the mean of a random sample \citep{lande1981models}.
+simply because this is the standard deviation of the mean of a random sample \citep{lande1981models}.
 % (This could be taken as a definition of $N_e$.)
@@
 explained about 3\% of expression variation.
-Studying natural flycatcher populations, \citet{wang2017bayesian} found that 7.2--8.5\% of transcripts showed allele-specific expression -- indicitive of standing genetic \emph{cis}-regulatory variation.
+Studying natural flycatcher populations, \citet{wang2017bayesian} found that 7.2--8.5\% of transcripts showed allele-specific expression -- indicative of standing genetic \emph{cis}-regulatory variation.
 Similarly, in wild baboons, \citet{tung2015genetic} observed allele-specific expression in 23.4\% of genes studied.

### 2017-12-06 kalman_walk_paper.tex
reparameterized and some formatting edits

@@
 
-        \end{tikzpicture} \\
+     \end{tikzpicture} &
    \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{osc_impulse}
@@
   \caption{
-    (Top) 
+    (Left) 
     Diagram of the gene network in Example \ref{ex:oscillator}
-  , and (bottom) plot of the phenotype $\phi(t)$ against time $t$
+  , and (right) plot of the phenotype $\phi(t)$ against time $t$
   .} \label{fig:oscillator}

### 2017-12-06 kalman_walk_paper.tex
reparameterized and some formatting edits

@@
     \caption{Phenotype distance to optimum, $D$, using $\rho(t) = \exp(-t/4\pi)$
-    for $F_1$ (red) and $F_2$ (blue) hybrids between an $A(0)$ and an $A(\epsilon)$ parent. Genetic distance is computed as the Frobenius norm between two systems.}
+    for $F_1$ (black) and $F_2$ (blue) hybrids between an $A(0)$ and an $A(\epsilon)$ parent. Genetic distance is computed as the Frobenius norm between two systems.}
     %\textbf{(right)} Same, but for $F_2$ offspring (one line per type of $F_2$).

### 2017-12-06 sims/perturb_Kalman.r

@@ -138,9 +138,9 @@ sys_gen6D <- function (s=sigma, eps=0.05, nreps=100)
     
     new_KalA_list <- perturb_oscillator(KalA_list, eps=eps)
     new_KalA <- assemble_oscillator(new_KalA_list)
-    #dist <- norm(KalA$A-new_KalA$A,"F")
+    dist <- norm(KalA$A-new_KalA$A,"F")
     ##dist <- sum(abs(Kal$A - new_KalA$A))/36
-    dist <- sqrt(sum((Kal$A - new_KalA$A)^2))/36
+    #dist <- sqrt(sum((Kal$A - new_KalA$A)^2))/36
     
     avg_div <- 0
     jj <- 0
@@ -149,7 +149,7 @@ sys_gen6D <- function (s=sigma, eps=0.05, nreps=100)
       Z1 <- recombine(KalA$A, new_KalA$A)
       Z2 <- recombine(KalA$A, new_KalA$A)
       Z3 <- (Z1 + Z2)/2
-      DZ <- tryCatch(D(Z3, BK=KalA$B, CK=KalA$C), error=function(e) NULL)
+      DZ <- tryCatch(sqrt(D(Z3, BK=KalA$B, CK=KalA$C)), error=function(e) NULL)
       if (is.null(DZ)==1)
       {
         j = j-1
@@ -182,9 +182,9 @@ sys_gen4D <- function (s=sigma, eps=0.05, nreps=100)
     new_KalA_list <- perturb_oscillator(KalA_list, eps=eps)
     #new_KalA <- assemble_oscillator(new_KalA_list)
     new_KalA <- assemble_oscillator4d(new_KalA_list)
-    #dist <- norm(KalA$A-new_KalA$A,"F")
+    dist <- norm(KalA$A-new_KalA$A,"F")
     ##dist <- sum(abs(Kal$A - new_KalA$A))/16
-    dist <- sqrt(sum((Kal$A - new_KalA$A)^2))/16
+    #dist <- sqrt(sum((Kal$A - new_KalA$A)^2))/16
     
     avg_div <- 0
     jj <- 0
@@ -193,7 +193,7 @@ sys_gen4D <- function (s=sigma, eps=0.05, nreps=100)
       Z1 <- recombine(KalA$A, new_KalA$A)
       Z2 <- recombine(KalA$A, new_KalA$A)
       Z3 <- (Z1 + Z2)/2
[…]

### 2017-12-06 kalman_walk_paper.tex
de-emphed hybrid breakdown in discussion

@@
 
-Consistent with empirical observation of \emph{hybrid breakdown} \citep[e.g.][]{plotner2017chlorosis}, 
+Consistent with empirical observation of hybrid breakdown \citep[e.g.][]{plotner2017chlorosis}, 
 we see that the fitnesses of $F_2$ hybrids drop at a much faster rate than those of $F_1$s.

### 2017-12-06 kalman_walk_paper.tex
changed Frobenius to Euclidean distance
changed Euclidean distance to equation
changed constant

@@
     \caption{Phenotype distance to optimum, $D$, using $\rho(t) = \exp(-t/4\pi)$
-    for $F_1$ (black) and $F_2$ (blue) hybrids between an $A(0)$ and an $A(\epsilon)$ parent. Genetic distance is computed as the Frobenius norm between two systems.}
+    for $F_1$ (black) and $F_2$ (blue) hybrids between an $A(0)$ and an $A(\epsilon)$ parent. Genetic distance is computed as 
+    %Frobenius norm
+    %Euclidean distance
+    $\sqrt{\sum_{ij} (A_{ij}(0) - A_{ij}(\epsilon))^2}$
+    %between two systems
+    .}
     %\textbf{(right)} Same, but for $F_2$ offspring (one line per type of $F_2$).
@@
 then the rate that $F_2$ fitness drops is, roughly,
-$(1 + 4 (p-q) C T/N_e)^{-q/2} \propto q (p-q)$, where $C$ is a constant.
+$(1 + 4 (p-q) K T/N_e)^{-q/2} \propto q (p-q)$, where $K$ is a constant.
 Both degree of constraint and number of available neutral directions

### 2017-12-06 kalman_walk_paper.tex
edited apx

@@
     \begin{aligned}
-        D(A, B, C) 
+        D(A, B, C)^2
         &:= 
@@
             \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)^T
-        \right\} dt  .
+        \right\} dt  ,
         % &= 
@@
 \end{equation}
+where $\tr X$ denotes the trace of a square matrix $X$.
 How does this change as we perturb about $(A_0, B_0, C_0)$?
@@
   \begin{aligned}
-      \frac{d}{du} D(A+uZ,B_0,C_0)\vert_{u=0}
+      \frac{d}{du} D(A+uZ,B_0,C_0)^2 \vert_{u=0}
       &=
@@
 and, by differentiating this and supposing that $A$ is on the optimal set,
-i.e., $h_A(t)=h_0(t)$, (so wolog $A=A_0$):
+i.e., $h_A(t)=h_0(t)$, (so without loss of generality, $A=A_0$):
 \begin{equation}
@@
       &:= 
-      \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0+uY+vZ,B_0,C_0)\vert_{u=v=0} \\
+      \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0+uY+vZ,B_0,C_0)^2 \vert_{u=v=0} \\
       &=
@@
 
-We could use this to compute the gradient of $D$,
-or to get the quadratic approximation to $D$ near the optimal set.
+We could use this to get the quadratic approximation to $D$ near the optimal set.
 To do so, it'd be nice to have a way to compute the inner integral above.
-Suppose that we can diagonalize $A = U \Lambda U^{-1}$.
+Suppose that we diagonalize $A = U \Lambda U^{-1}$.
 Then
@@
     \begin{aligned}
-        D(A_0+\epsilon Z)
+        D(A_0+\epsilon Z)^2
         &\approx \epsilon^2\sum_{ijk\ell} H^{ij,k\ell} Z_{ij} Z_{k\ell} 
@@
 More generally, $B$ and $C$ may also change.
-To extend this we need the remaining second derivatives of $D$.
+To extend this we need the remaining second derivatives of $D^2$.
 First, in $B$:
@@
 \end{equation}
+
+Together, numerical computation of these expressions,
+along with estimates of genetic covariance within a population,
+allow precise predictions of evolutionary dynamics of a particular system.
+The approximation should be good as long as the second-order Taylor approximation holds.
 
@@
 %%%%%%%%%%%
-\section*{Evolution of segregation covariance}
+\section{Evolution of segregation covariance}
 \label{apx:seg_cov}

### 2017-12-07 kalman_walk_paper.tex
edits

@@
 \citep{kimura1981possibility, true2001developmental, tanay2005conservation, tsong2006evolution, hare2008sepsid, lavoie2010evolutionary, vierstra2014mouse, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription}.
-It has also been shown theoretically that substantial genetic variation for a polygenic trait can be maintained
-by migration between geographical regions having different genetic bases for the same, optimal trait value \citep{phillips1996maintenance}.
+%It has also been shown theoretically that substantial genetic variation for a polygenic trait can be maintained
+%by migration between geographical regions having different genetic bases for the same, optimal trait value \citep{phillips1996maintenance}.
 
@@
 and that this may lead to speciation.
-For instance, \citet{tulchinsky2014hybrid} (and similarly \citet{khatri2015simple}) looked for hybrid incompatibility
+For instance, \citet{tulchinsky2014hybrid}
+%(and similarly \citet{khatri2015simple})
+looked for hybrid incompatibility
 following regulatory sequence drift,
@@
 but only equivocal support for speciation under models of purely neutral drift.
-Additionally, \citet{fierst} observed speciation due to an increase of segregation variance under balancing selection, in models of weak to moderate, but not strong epistasis.
+%Additionally, \citet{fierst} observed speciation due to an increase of segregation variance under balancing selection, in models of weak to moderate, but not strong epistasis.
 Our model differs from previous work in that we use linear systems theory
@@
 We would like to thank Sergey Nuzhdin, Stevan Arnold, Erik Lundgren, and Hossein Asgharian for valuable discussion.
-
-\plr{Josh: please fix capitalization in the references.  To do this, anything that needs to be capitalized gets put in curly braces, which tells bibtex to render the capitalization 'as-is'.}
+%
+%\plr{Josh: please fix capitalization in the references.  To do this, anything that needs to be capitalized gets put in curly braces, which tells bibtex to render the capitalization 'as-is'.}
 

### 2017-12-07 kalman_walk_paper.tex
minor edit
fixed refs

@@
 \citep{kimura1981possibility, true2001developmental, tanay2005conservation, tsong2006evolution, hare2008sepsid, lavoie2010evolutionary, vierstra2014mouse, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription}.
-%It has also been shown theoretically that substantial genetic variation for a polygenic trait can be maintained
-%by migration between geographical regions having different genetic bases for the same, optimal trait value \citep{phillips1996maintenance}.
+Furthermore, theory shows that distinct genotypes encoding identical phenotypes can even persist stably within a species \citep{phillips1996maintenance}.
 
@@
 \end{figure}
-%
+
 \begin{figure}[H]
+    \begin{center}
 \begin{tikzpicture}
@@
   \label{fig:3D_osc}
+\end{center}
 \end{figure}
@@
     some would be a similar distance to the other side of the optimal set).
-    The distribution of $F_2$s is appropriate for mixed homozygotes
+    The distribution of $F_2$ hybrids is appropriate for mixed homozygotes
     if both traits have a simple, one-locus genetic basis,
@@
     %%   it only matters if the offspring is homo/het/homozygous, so there are 3^4 = 81 possible expression patterns.
-    Different colored lines correspond to different $F_2$s;
+    Different colored lines correspond to different $F_2$ hybrids;
     note that some completely fail to oscillate.
@@
   \begin{figure}[H]
-  \label{fig:osc_incompat}
     \centering
@@
 if the regulatory coefficients for a system are shared between the sex and one or more autosomal chromosomes, 
-$F_1$ males are effectively equivalent to purely autosomal-system $F_2$s. 
-In this setting, Haldane's rule is mechanistically similar to \emph{hybrid breakdown}, the observation that $F_2$s will often be less fit than $F_1$ hybrid crosses, in the early stages of speciation.
+$F_1$ males are effectively equivalent to purely autosomal-system $F_2$ hybrids. 
+In this setting, Haldane's rule is mechanistically similar to \emph{hybrid breakdown}, the observation that $F_2$ hybrids will often be less fit than $F_1$ hybrid crosses, in the early stages of speciation.
 This mechanism relies on the genetic variation of a system being shared across the sex chromosomes and the autosomes,
@@
 and so will have mean system coefficients at the midpoint between their means.
-The distribution of $F_2$s will have mean at the average of the two populations,
+The distribution of $F_2$ hybrids will have mean at the average of the two populations,
 but will have higher variance.
-The variance of $F_2$s can be shown to increase linearly with the square of the distance between
+The variance of $F_2$ hybrids can be shown to increase linearly with the square of the distance between
 parental population means
@@
 (the same as within each parental population),
-but the population of $F_2$s will have variance of order $\sigma^2_S + 4 \omega \sigma^2_N T/N_e$,
+but the population of $F_2$ hybrids will have variance of order $\sigma^2_S + 4 \omega \sigma^2_N T/N_e$,
 where $\omega$ is a factor that depends on the genetic basis of the coefficients.
@@
 What are the fitness consequences?
-A population of $F_2$s will begin to be substantially less fit than the parentals
+A population of $F_2$ hybrids will begin to be substantially less fit than the parentals
 once they differ from the optimum by a distance of order $\gamma$,
 i.e., once $\sqrt{4 \omega T/N_e} \approx \gamma / \sigma_N$.
-This implies that hybrid incompatibility among $F_2$s should appear much slower --
+This implies that hybrid incompatibility among $F_2$ hybrids should appear much slower --
 on a time scale of $N_e (\gamma / \sigma_N)^2 / (4 \omega)$ generations.
@@
 This assumes a single trait, for simplicity. %; the multivariate case is done in appendix \ref{apx:gauss_load}.
-A population of $F_2$s will have, as above, variance $\sigma^2 = \sigma^2_S + 4 \omega \sigma^2_N T/N_e$.
+A population of $F_2$ hybrids will have, as above, variance $\sigma^2 = \sigma^2_S + 4 \omega \sigma^2_N T/N_e$.
 The mean diverges with the square of the distance between the parentals, so we set $\mu = c_\mu \gamma T/N_e$,
@@
 depicting how hybrid fitness drops as a function of $T/N_e$ in Figure \ref{fig:speciation_rates}.
-In all cases, the fitness drop for $F_1$ hybrids is much smaller than that of $F_2$s,
+In all cases, the fitness drop for $F_1$ hybrids is much smaller than that of $F_2$ hybrids,
 so we work only with the first (square-root) term in equation \eqref{eqn:fit2}.
@@
 Setting $a = (\sigma_N/\gamma)^2$ to be small,
-the fitness of $F_2$s is $\fit_2 \le 1/\sqrt{1 + 4 \omega a T/N_e} \approx 1 - 2 \omega a T/N_e$.
+the fitness of $F_2$ hybrids is $\fit_2 \le 1/\sqrt{1 + 4 \omega a T/N_e} \approx 1 - 2 \omega a T/N_e$.
 Hybrid fitness seems to drop more slowly in this case in figure \ref{fig:speciation_rates},
@@
 so that $\sigma_S \approx \gamma$ but $\sigma_N/\gamma$ is large.
-Then the fitness of $F_2$s is $\fit_2 \le 1/\sqrt{1 + 2 \omega a T/N_e} \approx 1 - \omega a T/N_e$,
+Then the fitness of $F_2$ hybrids is $\fit_2 \le 1/\sqrt{1 + 2 \omega a T/N_e} \approx 1 - \omega a T/N_e$,
 and could be extremely rapid if $a$ is large.
@@
 \begin{figure}[H]
-\label{fig:speciation_rates}
 \begin{center}
@@
 explained about 3\% of expression variation.
-Studying natural flycatcher populations, \citet{wang2017bayesian} found that 7.2--8.5\% of transcripts showed allele-specific expression -- indicative of standing genetic \emph{cis}-regulatory variation.
-Similarly, in wild baboons, \citet{tung2015genetic} observed allele-specific expression in 23.4\% of genes studied.
+Allele-specific expression is indicative of standing genetic \emph{cis}-regulatory variation;
+\citet{wang2017bayesian} found allele-specific expression in 7.2--8.5\% of transcripts of a flycatcher species,
+while \citet{tung2015genetic} observed allele-specific expression in 23.4\% of genes studied in a baboon species. 
 Taken together, this suggests that variation in the entries of $A$
@@
 This occurs because if the genes underlying a regulatory network are distributed among both autosomes and the sex chromosome, 
-then heterogametic $F_1$s show variation similar to that seen in $F_2$s.
+then heterogametic $F_1$s show variation similar to that seen in $F_2$ hybrids.
 
@@
 This implies that under this model,
-segregation variance in $F_2$s between two populations
+segregation variance in $F_2$ hybrids between two populations
 is roughly increased by a factor of $1/8$ of the difference between their means.

### 2017-12-08 kalman_walk_paper.tex
minor fixing
more minor fixing
.

@@
 Let two points on $\allS$ be $x_1$ and $x_2$, let $\bar x = (x_1+x_2)/2$, and let $z=(x_2 - x_1)/2$.
-Then with $\grad D$ and $\grad^2 D$ the first and second derivatives of $D$, respectively,
+Then with $\grad D^2$ and $\grad^2 D^2$ the first and second derivatives of $D^2$, respectively,
 Taylor expanding about $x_1$ and $x_2$ finds that
 \begin{align*}
-    D(\bar x) 
-    &= D(x_1) + \grad D(x_1) \cdot z + \frac{1}{2} z^T \grad^2  D(x_1) z + O(\|z\|^3) \\
-    &=  D(x_2) - \grad D(x_2) \cdot z + \frac{1}{2} z^T \grad^2  D(x_2) z + O(\|z\|^3) .
-\end{align*}
-Now, since $ D(x_1) =  D(x_2) = \grad D(x_1) = \grad D(x_2) = 0$ and
-\begin{align*}
-    \grad D(x_2) = \grad D(x_1) + 2 z^T \grad^2  D(x_1) + O(\|z\|^2), \quad \text{and} \\
-    \grad^2 D(x_2) = \grad^2 D(x_1) + O(\|z\|), 
+    D(\bar x)^2 
+    &= D(x_1)^2 + \grad D(x_1)^2 \cdot z + \frac{1}{2} z^T \grad^2  D(x_1)^2 z + O(\|z\|^3) \\
+    &=  D(x_2)^2 - \grad D(x_2)^2 \cdot z + \frac{1}{2} z^T \grad^2  D(x_2)^2 z + O(\|z\|^3) .
+\end{align*}
+Now, since $ D(x_1)^2 =  D(x_2)^2 = \grad D(x_1)^2 = \grad D(x_2)^2 = 0$ and
+\begin{align*}
+    \grad D(x_2)^2 = \grad D(x_1)^2 + 2 z^T \grad^2  D(x_1)^2 + O(\|z\|^2), \quad \text{and} \\
+    \grad^2 D(x_2)^2 = \grad^2 D(x_1)^2 + O(\|z\|), 
 \end{align*}
@@
 \begin{align*}
-     D(\bar x) 
-    &= \optph - \frac{3}{2} z^T \grad^2  D(x_1) z + O(\|z\|^3) .
+     D(\bar x)^2
+    &= - \frac{3}{2} z^T \grad^2  D(x_1)^2 z + O(\|z\|^3) .
 \end{align*}
@@
       &=
-      \frac{ e^{t \lambda_i} - e^{t \lambda_j} }{ \lambda_i - \lambda_j } .
+      \frac{ e^{t \lambda_i} - e^{t \lambda_j} }{ \lambda_i - \lambda_j } 
+          \qquad & \text{if} \quad i \neq j \\
+      &=
+          t e^{t \lambda_i} 
+          \qquad & \text{if} \quad i = j 
   \end{aligned}
@@
 \begin{equation}
-    X_{ij}(t,Z) = \left( U^{-1} Z U \right)_{ij}
+    \begin{aligned}
+    X_{ij}(t,Z) 
+       &= 
+        \left( U^{-1} Z U \right)_{ij}
       \frac{ e^{t \lambda_i} - e^{t \lambda_j} }{ \lambda_i - \lambda_j } 
+          \qquad & \text{if} \quad i \neq j \\
+      &=
+          \left( U^{-1} Z U \right)_{ii}
+          t e^{t \lambda_i} 
+          \qquad & \text{if} \quad i = j 
+    \end{aligned}
 \end{equation}
@@
 \end{equation}
-
-Following on from above, we see that if $Z=\Delta_{k \ell}$, then
+This implies that
 \begin{equation}
     \begin{aligned}
-      X_{ij}^{k\ell}(t) = 
+        D(A_0+\epsilon Z)^2
+        &\approx \frac{1}{2} \epsilon^2 
+        \int_0^\infty
+            \rho(t) \tr\left\{ C U X(t,Z) U^{-1} B B^T (U^{-1})^T X(t,Z)^T U^T C^T \right\}
+        dt .
+    \end{aligned}
+\end{equation}
+
+To compute the $n^2 \times n^2$ matrix $H$,
+we see that if $Z=\Delta_{k \ell}$, then
+\begin{equation}
+  \begin{aligned}
+      X_{ij}^{k\ell}(t) 
+      &= 
+      (U^{-1})_{\cdot k} U_{\ell \cdot}
       \frac{ e^{t \lambda_i} - e^{t \lambda_j} }{ \lambda_i - \lambda_j } 
-      (U^{-1})_{\cdot k} U_{\ell \cdot},
+          \qquad & \text{if} \quad i \neq j \\
+      &=
+      (U^{-1})_{\cdot k} U_{\ell \cdot}
+      t e^{t \lambda_i} 
+          \qquad & \text{if} \quad i = j 
   \end{aligned}
@@
         D(A_0+\epsilon Z)^2
-        &\approx \epsilon^2\sum_{ijk\ell} H^{ij,k\ell} Z_{ij} Z_{k\ell} 
-    \end{aligned}
-\end{equation}
-and so
-\begin{equation}
-    \begin{aligned}
-        D(A_0+\epsilon Z)^2
-        &\approx \epsilon^2\sum_{ijk\ell} H^{ij,k\ell} Z_{ij} Z_{k\ell} 
+        &\approx \frac{1}{2} \epsilon^2 \sum_{ijk\ell} H_{ij,k\ell}(A_0) Z_{ij} Z_{k\ell}  .
     \end{aligned}

### 2017-12-08 sims/theoretical_pheno_dist.R
added partial code for theory computation

@@ -1,0 +1,52 @@
+gamete_gen <- function(tau, eps) {
+  gametes <- list()
+  count <- 0
+  for (i in 0:1)
+  {
+    for (j in 0:1)
+    {
+      for (k in 0:1)
+      {
+        for (l in 0:1)
+        {
+          count <- count + 1
+          M <- matrix(c(i,k,j,l), ncol=2, nrow=2)
+          name <- paste("G", count, sep = ""); 
+          assign(name, S(tau)*M + S(tau + eps)*(1 - M))
+          gametes[[name]] <- get(name)
+        }
+      }
+    }
+  }
+  return(gametes)
+}
+
+progeny_gen <- function(tau, eps){
+  gametes <- gamete_gen(tau, eps)
+  progeny <- list()
+  count2 <- 0
+  for (i in 1:16)
+  {
+    for (j in 1:16)
+    {
+      count2 <- count2 + 1
+      progeny[[count2]] <- (gametes[[i]] + gametes[[j]])/2
+    }
+  }
+  return(progeny)
+}
+
+U = matrix(c(sqrt(1/2), 0+1i/sqrt(2), sqrt(1/2), 0-1i/sqrt(2)), ncol=2)
+invU = matrix(c(sqrt(1/2), sqrt(1/2),0-1i/sqrt(2), 0+1i/sqrt(2)), nrow=2)
+
+theory_comp <- function(eps) {
+  progeny <- progeny_gen(0, eps)
+  mean_dist <- 0
+  for (k in k:256) 
+  {
+    X <- invU%*%progeny[[k]]%*%U * matrix( c( sin(t), t*exp(t*(0+1i)), sin(t), t*exp(t*(0-1i)) ), nrow=2)
+    ingd <- C%*%U%*%X%*%invU%*%B%*%t(B)%*%t(invU)%*%t(X)%*%t(U)%*%t(C)
+    mean_dist <- mean_dist + 0.5*(eps^2)*integrate(ingd, upper = 10, lower = 0)
+  }
+  return(mean_dist)
+}

### 2017-12-08 sims/theoretical_pheno_dist.R
fixed code

@@ -1,3 +1,15 @@
+A <- matrix(c(0,-1,1,0), nrow=2);
+B <- matrix(c(1,1), ncol=1);
+C <- matrix(c(1,0), nrow=1);
+
+V <- function (tau) {
+  matrix(c(1,tau,0,1-tau), nrow=2) 
+}
+
+S <- function(tau) {
+  V(tau)%*%A%*%solve(V(tau))
+}
+
 gamete_gen <- function(tau, eps) {
   gametes <- list()
   count <- 0

### 2017-12-08 sims/theoretical_pheno_dist.R
code edits

@@ -23,9 +23,7 @@ gamete_gen <- function(tau, eps) {
         {
           count <- count + 1
           M <- matrix(c(i,k,j,l), ncol=2, nrow=2)
-          name <- paste("G", count, sep = ""); 
-          assign(name, S(tau)*M + S(tau + eps)*(1 - M))
-          gametes[[name]] <- get(name)
+          gametes[[count]] <- S(tau)*M + S(tau + eps)*(1 - M)
         }
       }
     }
@@ -54,11 +52,16 @@ invU = matrix(c(sqrt(1/2), sqrt(1/2),0-1i/sqrt(2), 0+1i/sqrt(2)), nrow=2)
 theory_comp <- function(eps) {
   progeny <- progeny_gen(0, eps)
   mean_dist <- 0
-  for (k in k:256) 
+  for (k in 1:256) 
   {
-    X <- invU%*%progeny[[k]]%*%U * matrix( c( sin(t), t*exp(t*(0+1i)), sin(t), t*exp(t*(0-1i)) ), nrow=2)
-    ingd <- C%*%U%*%X%*%invU%*%B%*%t(B)%*%t(invU)%*%t(X)%*%t(U)%*%t(C)
-    mean_dist <- mean_dist + 0.5*(eps^2)*integrate(ingd, upper = 10, lower = 0)
+    X0 <- invU%*%progeny[[k]]%*%U
+    f <- function (t) {
+        sapply(t, function (tt) {
+              X <- UZU * matrix( c( sin(tt), tt*exp(tt*(0+1i)), sin(tt), tt*exp(tt*(0-1i)) ), nrow=2)
+              return( (C%*%U%*%X%*%invU%*%B)^2 )
+           })
+    }
+    mean_dist <- mean_dist + 0.5*(eps^2)*integrate(f, upper = Inf, lower = 0)
   }
   return(mean_dist)
 }

### 2017-12-08 kalman_walk_paper.tex
remove apx A1 for wrongness

@@
 The $F_1$s will not suffer fitness consequences until the hybrid mean is further than $\gamma$ from the optimum;
-as shown in appendix \ref{apx:away_from_opt} (and suggested by figure \ref{fig:conceptual_fig}),
+% as shown in appendix \ref{apx:away_from_opt} 
+as suggested by figure \ref{fig:conceptual_fig}, Taylor expanding $D^2$ along the optimal set implies that
 this deviation of the mean from optimum grows with the square of the distance between the parental populations,
@@
 \section{Local expansion of the fitness surface}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Away from the optimum}
-\label{apx:away_from_opt}
-
-Let two points on $\allS$ be $x_1$ and $x_2$, let $\bar x = (x_1+x_2)/2$, and let $z=(x_2 - x_1)/2$.
-Then with $\grad D^2$ and $\grad^2 D^2$ the first and second derivatives of $D^2$, respectively,
-Taylor expanding about $x_1$ and $x_2$ finds that
-\begin{align*}
-    D(\bar x)^2 
-    &= D(x_1)^2 + \grad D(x_1)^2 \cdot z + \frac{1}{2} z^T \grad^2  D(x_1)^2 z + O(\|z\|^3) \\
-    &=  D(x_2)^2 - \grad D(x_2)^2 \cdot z + \frac{1}{2} z^T \grad^2  D(x_2)^2 z + O(\|z\|^3) .
-\end{align*}
-Now, since $ D(x_1)^2 =  D(x_2)^2 = \grad D(x_1)^2 = \grad D(x_2)^2 = 0$ and
-\begin{align*}
-    \grad D(x_2)^2 = \grad D(x_1)^2 + 2 z^T \grad^2  D(x_1)^2 + O(\|z\|^2), \quad \text{and} \\
-    \grad^2 D(x_2)^2 = \grad^2 D(x_1)^2 + O(\|z\|), 
-\end{align*}
-adding together the two equations above and dividing by two gets that
-\begin{align*}
-     D(\bar x)^2
-    &= - \frac{3}{2} z^T \grad^2  D(x_1)^2 z + O(\|z\|^3) .
-\end{align*}
-
-\subsection{Differentiating the fitness function}
 \label{apx:H_calc}

### 2017-12-08 kalman_walk_paper.tex
removed other apx

@@
 and their offspring has impulse response function $h_\epsilon(t)$.
-A Taylor expansion of $D(h_\epsilon)$ in $\epsilon$ is explicitly worked out in Appendix \ref{apx:H_calc},
-and shows that the phenotype of an $F_1$ hybrid between these two is at distance proportional to $\epsilon^2$ from optimal,
+A Taylor expansion of $D(h_\epsilon)$ in $\epsilon$ % is explicitly worked out in Appendix \ref{apx:H_calc}, and
+shows that the phenotype of an $F_1$ hybrid between these two is at distance proportional to $\epsilon^2$ from optimal,
 while $F_2$ hybrids are at distance proportional to $\epsilon$.
@@
 \appendix
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Local expansion of the fitness surface}
-\label{apx:H_calc}
-
-Suppose that $\rho(t) \ge 0$ is a weighting function on $[0,\infty)$
-so that fitness is a function of $L^2(\rho)$ distance of the impulse response from optimal.
-With $h_0(t) = C_0 e^{tA_0} B_0$ a representative of the optimal set:
-\begin{equation}
-    \begin{aligned}
-        D(A, B, C)^2
-        &:= 
-        \int_0^\infty \rho(t) \left| h_A(t) - h_0(t) \right|^2 dt \\
-        &:= 
-        \int_0^\infty \rho(t) \left| C e^{At} B - C_0 e^{A_0 t} B_0 \right|^2 dt \\
-        &= 
-        \int_0^\infty \rho(t) \tr\left\{
-            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)^T
-            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)
-        \right\} dt \\
-        &= 
-        \int_0^\infty \rho(t) \tr\left\{
-            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)
-            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)^T
-        \right\} dt  ,
-        % &= 
-        % \int_0^\infty \rho(t)  \tr\left\{
-        %         C e^{At} B B^T e^{A^T t} C^T 
-        %         - C_0 e^{A_0t} B_0 B^T e^{A^T t} C^T 
-        %         \right. \\ &\qquad \qquad \left. {}
-        %         - C e^{At} B B_0^T e^{A_0^T t} C_0^T 
-        %         + C_0 e^{A_0t} B_0 B_0^T e^{A_0^T t} C_0^T 
-        % \right\} dt \\
-        % \int_0^\infty \rho(t) \left| C \left( e^{At} - e^{A_0 t} \right) B \right|^2 dt \\
-        % &= 
-        % \int_0^\infty \rho(t) C \left( e^{At} - e^{A_0 t} \right) B B^T \left( e^{At} - e^{A_0 t} \right)^T C^T dt
-    \end{aligned}
-\end{equation}
-where $\tr X$ denotes the trace of a square matrix $X$.
-How does this change as we perturb about $(A_0, B_0, C_0)$?
-First we differentiate with respect to $A$, keeping $B=B_0$ and $C=C_0$ fixed.
-Since
-\begin{equation}
-  \begin{aligned}
-      \frac{d}{du} e^{(A+uZ)t} \vert_{u=0}
-      &=
-      \int_0^t e^{As} Z e^{A(t-s)} ds, 
-  \end{aligned}
-\end{equation}
-we have that
-\begin{equation}
-  \begin{aligned}
-      \frac{d}{du} D(A+uZ,B_0,C_0)^2 \vert_{u=0}
-      &=
-        2 \int_0^\infty \rho(t) \tr\left\{ C_0 \left( \int_0^t e^{As} Z e^{A(t-s)} ds \right) B_0 B_0^T \left( e^{At} - e^{A_0 t} \right)^T C_0^T \right\} dt \\
-      &=
-        2 \int_0^\infty \rho(t) \tr\left\{ C_0 \left( \int_0^t e^{As} Z e^{A(t-s)} ds \right) B_0 \left( h_A(t) - h_0(t) \right)^T \right\} dt 
-  \end{aligned}
-\end{equation}
-and, by differentiating this and supposing that $A$ is on the optimal set,
-i.e., $h_A(t)=h_0(t)$, (so without loss of generality, $A=A_0$):
-\begin{equation}
-  \begin{aligned}
-      \calH^{A,A}(Y,Z) 
-      &:= 
-      \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0+uY+vZ,B_0,C_0)^2 \vert_{u=v=0} \\
-      &=
-        \int_0^\infty \rho(t) \tr\left\{ C_0 
-        \left( \int_0^t e^{A_0 s} Y e^{A_0 (t-s)} ds \right) 
-        B_0 B_0^T 
-        \left( \int_0^t e^{A_0 s} Z e^{A_0 (t-s)} ds \right)^T
-        C_0^T \right\} dt  .
-  \end{aligned}
-\end{equation}
-
-The function $\calH$ will define a quadratic form.
-To illustrate the use of this, suppose that $B$ and $C$ are fixed.
-% Here $\calH$ is the quadratic form underlying the Hamiltonian.
-By defining $\Delta_{ij}$ to be the matrix with a 1 in the $(i,j)$th slot
-and 0 elsewhere,
-the coefficients of the quadratic form is
-\begin{equation}
-    \begin{aligned}
-        H_{ij, k\ell}(A)
-        &:=
-        \calH(\Delta_{ij}, \Delta_{k\ell}) .
-    \end{aligned}
-\end{equation}
-
-We could use this to get the quadratic approximation to $D$ near the optimal set.
-To do so, it'd be nice to have a way to compute the inner integral above.
-Suppose that we diagonalize $A = U \Lambda U^{-1}$.
-Then
-\begin{equation} \label{eqn:exp_deriv}
-  \begin{aligned}
-      \int_0^t e^{As} Z e^{A(t-s)} ds 
-      &=
-      \int_0^t U e^{\Lambda s} U^{-1} Z U e^{\Lambda (t-s)} U^{-1} ds 
-  \end{aligned}
-\end{equation}
-Now, notice that
-\begin{equation}
-  \begin{aligned}
-      \int_0^t e^{s \lambda_i} e^{(t-s) \lambda_j} ds
-      &=
-      \frac{ e^{t \lambda_i} - e^{t \lambda_j} }{ \lambda_i - \lambda_j } 
-          \qquad & \text{if} \quad i \neq j \\
-      &=
-          t e^{t \lambda_i} 
-          \qquad & \text{if} \quad i = j 
-  \end{aligned}
-\end{equation}
-Therefore, 
-defining
-\begin{equation}
-    \begin{aligned}
-    X_{ij}(t,Z) 
-       &= 
-        \left( U^{-1} Z U \right)_{ij}
-      \frac{ e^{t \lambda_i} - e^{t \lambda_j} }{ \lambda_i - \lambda_j } 
-          \qquad & \text{if} \quad i \neq j \\
-      &=
-          \left( U^{-1} Z U \right)_{ii}
-          t e^{t \lambda_i} 
-          \qquad & \text{if} \quad i = j 
-    \end{aligned}
-\end{equation}
-moving the $U$ and $U^{-1}$ outside the integral and integrating we get that
-\begin{equation}
-  \begin{aligned}
-      \int_0^t e^{As} Z e^{A(t-s)} ds 
-      &=
-      U X(t,Z) U^{-1} .
-  \end{aligned}
-\end{equation}
-This implies that
-\begin{equation}
-    \begin{aligned}
-        D(A_0+\epsilon Z)^2
-        &\approx \frac{1}{2} \epsilon^2 
-        \int_0^\infty
-            \rho(t) \tr\left\{ C U X(t,Z) U^{-1} B B^T (U^{-1})^T X(t,Z)^T U^T C^T \right\}
-        dt .
-    \end{aligned}
-\end{equation}
-
-To compute the $n^2 \times n^2$ matrix $H$,
-we see that if $Z=\Delta_{k \ell}$, then
-\begin{equation}
-  \begin{aligned}
-      X_{ij}^{k\ell}(t) 
-      &= 
-      (U^{-1})_{\cdot k} U_{\ell \cdot}
-      \frac{ e^{t \lambda_i} - e^{t \lambda_j} }{ \lambda_i - \lambda_j } 
-          \qquad & \text{if} \quad i \neq j \\
-      &=
-      (U^{-1})_{\cdot k} U_{\ell \cdot}
-      t e^{t \lambda_i} 
-          \qquad & \text{if} \quad i = j 
-  \end{aligned}
-\end{equation}
-where $U_{k \cdot}$ is the $k$th row of $U$,
-and so
-\begin{equation}
-    \begin{aligned}
-        H_{ij, k\ell}(A)
-        &=
-        \int_0^\infty
-            \rho(t) \tr\left\{ C U X^{ij}(t) U^{-1} B B^T (U^{-1})^T X^{k\ell}(t)^T U^T C^T \right\}
-        dt .
-    \end{aligned}
-\end{equation}
-This implies that
-\begin{equation}
-    \begin{aligned}
-        D(A_0+\epsilon Z)^2
-        &\approx \frac{1}{2} \epsilon^2 \sum_{ijk\ell} H_{ij,k\ell}(A_0) Z_{ij} Z_{k\ell}  .
-    \end{aligned}
-\end{equation}
-
-By section \ref{ss:quant_gen},
-if we set $\Sigma=\sigma^2 I$ and $U=H$,
-then a population at $A_0+Z$ experiences a restoring force of strength
-$(I + \sigma^2 H^{-1})^{-1} Z$ (treating $Z$ as a vector and $H$ as an operator on these).
-If $\sigma^2$ is small compared to $H^{-1}$
-then this is approximately $-\sigma^2 H^{-1} Z$.
-This suggests that the population mean follows an Ornstein-Uhlenbeck process,
-as described (in different terms) in \citet{hansen1996translating}.
-
-More generally, $B$ and $C$ may also change.
-To extend this we need the remaining second derivatives of $D^2$.
-First, in $B$:
-\begin{equation}
-    \begin{aligned}
-        \mathcal{H}^{B,B}(Y,Z)
-        &:= 
-        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0,B_0+uY+vZ,C_0)\vert_{u=v=0} \\
-        &=
-        \frac{1}{2} \int_0^\infty \rho(t) \tr\left\{ 
-        C_0 e^{t A_0}
-        \frac{d}{du} \frac{d}{dv} 
-        (uY+vZ)
-        (uY+vZ)^T
-        \vert_{u=v=0} 
-        e^{t A_0^T} C_0^T 
-        \right\} dt  \\
-        &=
-        \frac{1}{2} \int_0^\infty \rho(t) \tr\left\{ 
-        C_0 e^{t A_0}
-        \left( Y Z^T + Z Y^T \right)
-        e^{t A_0^T} C_0^T 
-        \right\} dt  .
-  \end{aligned}
-\end{equation}
-Next, in $C$:
-\begin{equation}
-    \begin{aligned}
-        \mathcal{H}^{B,B}(Y,Z)
-        &:= 
-        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0,B_0,C_0+uY+vZ)\vert_{u=v=0} \\
-        &=
-        \frac{1}{2} \int_0^\infty \rho(t) \tr\left\{ 
-        B_0 e^{t A_0^T}
-        \frac{d}{du} \frac{d}{dv} 
-        (uY+vZ)^T
-        (uY+vZ)
-        \vert_{u=v=0} 
-        e^{t A_0} B_0
-        \right\} dt  \\
-        &=
-        \frac{1}{2} \int_0^\infty \rho(t) \tr\left\{ 
-        B_0 e^{t A_0^T}
-        \left( Y Z^T + Z Y^T \right)
-        e^{t A_0} B_0
-        \right\} dt  .
-  \end{aligned}
-\end{equation}
-Now, the mixed derivatives in $B$ and $C$:
-\begin{equation}
-    \begin{aligned}
-        \mathcal{H}^{B,C}(Y,Z)
-        &:= 
-        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0,B_0+uY,C_0+vZ)\vert_{u=v=0} \\
-        &=
-        \int_0^\infty \rho(t) \tr\left\{ 
-        Y e^{t A_0^T} C_0^T Z e^{t A_0} B_0
-        \right\} dt  .
-  \end{aligned}
-\end{equation}
-In $A$ and $B$
-\begin{equation}
-    \begin{aligned}
-        \mathcal{H}^{A,B}(Y,Z)
-        &:= 
-        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0+uY,B_0+vZ,C_0)\vert_{u=v=0} \\
-        &=
-        \int_0^\infty \rho(t) \tr\left\{ 
-        C_0 \left(\int_0^t e^{s A_0} Y e^{(t-s) A_0} ds \right) B_0
-        Z^T e^{t A_0} C_0
-        \right\} dt  ,
-  \end{aligned}
-\end{equation}
-and finally in $A$ and $C$:
-\begin{equation}
-    \begin{aligned}
-        \mathcal{H}^{A,C}(Y,Z)
-        &:= 
-        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0+uY,B_0,C_0+vZ)\vert_{u=v=0} \\
-        &=
-        \int_0^\infty \rho(t) \tr\left\{ 
-        C_0 \left(\int_0^t e^{s A_0} Y e^{(t-s) A_0} ds \right) B_0
-        B_0 e^{t A_0} Z
-        \right\} dt  .
-  \end{aligned}
-\end{equation}
-
-Together, numerical computation of these expressions,
-along with estimates of genetic covariance within a population,
-allow precise predictions of evolutionary dynamics of a particular system.
-The approximation should be good as long as the second-order Taylor approximation holds.
 

### 2017-12-08 kalman_walk_paper.tex
changed title
changed title, fixed figure six Y-axis label and caption

@@
 {\centering
-{\LARGE \bf Rapid speciation despite conservation of phenotype} \\ \vspace{0.5cm}
+{\Huge \bf System drift and speciation} \\ \vspace{0.75cm}
 Joshua S. Schiffman$^{\dagger}$ \qquad Peter L. Ralph$^{\dagger \ddagger}$ \\ \vspace{0.5cm}
@@
     %\textbf{(left)} 
-    \caption{Phenotype distance to optimum, $D$, using $\rho(t) = \exp(-t/4\pi)$
-    for $F_1$ (black) and $F_2$ (blue) hybrids between an $A(0)$ and an $A(\epsilon)$ parent. Genetic distance is computed as 
+    \caption{Mean hybrid phenotypic distance from optimum computed with equation \ref{eqn:distance}, using $\rho(t) = \exp(-t/4\pi)$
+    for $F_1$ (black) and $F_2$ (blue) hybrids between $A(0)$ and $A(\epsilon)$ parent oscillators. Genetic distance is computed as 
     %Frobenius norm

### 2017-12-08 kalman_walk_paper.tex
changed title, fixed figure six Y-axis label and caption

@@
 and that these mechanisms may fail to produce viable hybrids.
-This begs the question: does evolution shift molecular mechanisms
+This leads to the question: does evolution shift molecular mechanisms
 fast enough to be a significant driver of speciation?
@@
 
-\paragraph{The origin of species not by means of natural selection?}
+\paragraph{On the origin of species not by means of natural selection?}
 % this is neutral, in contrast to other models. although some folks advocate nonadaptive speciation.

### 2017-12-08 kalman_walk_paper.tex
a few more bibtex fixes

@@
 As above, an individual's coefficients are given by averaging its parentally inherited coefficients and adding random noise due to segregation and possibly new mutation.
-Concretely, we use the \emph{infinitesimal model} for reproduction \citep{barton2016infinitesimal} --
+Concretely, we use the \emph{infinitesimal model} for reproduction \citep{barton2017infinitesimal} --
 the offspring of parents at $x$ and $x'$ will have coefficients $(x+x')/2 + \varepsilon$,

### 2018-01-16 kalman_walk_paper.tex
added citations to intro, stated Haldane's rule consistent with dominance theory, changed the first sentence of the abstract to make it punchier.

@@
 \begin{abstract}
-%It is known that even if a species' phenotype has remained unchanged over evolutionary time,
-%the underlying mechanism may have changed,
-%since distinct molecular pathways can realize identical phenotypes.
-Even if a species' phenotype remains unchanged over evolutionary time, the underlying mechanism may have changed, as distinct molecular pathways can realize identical phenotypes.
+The evolutionary conservation of phenotype under selective and environmental stasis does not necessitate conservation of the underlying mechanism, as distinct molecular pathways can realize identical phenotypes.
+%Even if a species' phenotype remains unchanged over evolutionary time, the underlying mechanism may have changed, as distinct molecular pathways can realize identical phenotypes.
 Here we use quantitative genetics and linear system theory to study
@@
 \citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. 
-The fundamental nonidentifiability of chemical reaction networks is sometimes referred to as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}. 
+The fundamental nonidentifiability of chemical reaction networks is sometimes referred to as the ``dogma of chemical kinetics'' \citep{craciun2008identifiability}. 
 In computer science, this is framed as the relationship among processes that simulate one another \citep{van2004equivalence}.
@@
 It is not a new observations that there is often more than one way to do the same thing,
-and that this may lead to speciation.
+and that this may lead to speciation or that speciation can be the result of neutral and quasi-neutral processes \citep{wright1932roles, nei1983models, barton1989divergence, barton2001role, gavrilets2004fitness}.
 For instance, \citet{tulchinsky2014hybrid}
@@
 
-\begin{example}[All phenotypically equivalent oscillators] \label{ex:all_osc}
+\begin{example}[All phenotypically equivalent gene network wirings] \label{ex:all_osc}
 The oscillator of example \ref{ex:oscillator} is minimal, and so any equivalent system is a change of coordinates
@@
 
-\begin{example}[Hybrid Incompatibility in the Oscillator] \label{ex:hybrid_osc}
+\begin{example}[Hybrid incompatibility: misregulation due to system drift] \label{ex:hybrid_osc}
 Offspring of two equivalent systems from Example \ref{ex:all_osc}
@@
 This mechanism relies on the genetic variation of a system being shared across the sex chromosomes and the autosomes,
-and appears to be distinct from the ``dominance'', ``faster-X'', and ``faster-male'' theories \citep{orr1997haldane, coyne1998evolutionary, turelli1995dominance} usually suggested to explain Haldane's rule.
+and appears to be consistent with the ``dominance'' theory \citep{orr1997haldane, coyne1998evolutionary, turelli1995dominance} which is often invoked to explain Haldane's rule.
 %and makes several different predictions than the dominance theory as formulated by \citet{turelli1995dominance}.

### 2018-01-17 kalman_walk_paper.tex
discussed edits

@@
 \begin{abstract}
-The evolutionary conservation of phenotype under selective and environmental stasis does not necessitate conservation of the underlying mechanism, as distinct molecular pathways can realize identical phenotypes.
-%Even if a species' phenotype remains unchanged over evolutionary time, the underlying mechanism may have changed, as distinct molecular pathways can realize identical phenotypes.
+%The evolutionary conservation of phenotype under selective and environmental stasis does not necessitate conservation of the underlying mechanism, as distinct molecular pathways can realize identical phenotypes.
+Even if a species' phenotype does not change over evolutionary time, the underlying mechanism may change, as distinct molecular pathways can realize identical phenotypes.
 Here we use quantitative genetics and linear system theory to study
@@
 as well as provide concrete numerical predictions \citep{servedio2014not}.
-
-The dynamics of the molecular machinery and its interactions with the environment
-can be mathematically described as a dynamical system \citep{jaeger2015comet}.
-Movement in this direction is ongoing, as researchers have begun to study 
-the evolution of both abstract \citep{wagner1994evolution, wagner1996does, siegal2002waddington, bergman2003evolutionary, draghi2015robustness} 
-and empirically inspired computational and mathematical models of gene regulatory networks,
-\citep[e.g.][]{mjolsness1991connectionist, jaeger2004dynamic, maria1, vitaly1, vitaly2, crombach2016gap, wotton2015quantitative, chertkova2017insilico}.
+%The dynamics of the molecular machinery and its interactions with the environment
+%can be mathematically described as a dynamical system \citep{jaeger2015comet}.
+%Movement in this direction is ongoing, as 
+There is a substantial amount of work studying the evolution of gene regulatory networks, in both abstract \citep{wagner1994evolution, wagner1996does, siegal2002waddington, bergman2003evolutionary, draghi2015robustness}
+and empirically inspired computational
+\citep{mjolsness1991connectionist, jaeger2004dynamic, vitaly1, crombach2016gap, wotton2015quantitative, chertkova2017insilico} frameworks.
 % If we allow the reasonable assumption that the genotype-phenotype map 
@@
 % we can immediately discuss its evolution and function in a much more mechanistic, yet general, manner.
+
 It is well known that in many contexts mathematical models
@@
 \citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. 
-The fundamental nonidentifiability of chemical reaction networks is sometimes referred to as the ``dogma of chemical kinetics'' \citep{craciun2008identifiability}. 
+The inherent nonidentifiability of chemical reaction networks is sometimes referred to as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}. 
 In computer science, this is framed as the relationship among processes that simulate one another \citep{van2004equivalence}.
@@
 \citep{kimura1981possibility, true2001developmental, tanay2005conservation, tsong2006evolution, hare2008sepsid, lavoie2010evolutionary, vierstra2014mouse, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription}.
-Furthermore, theory shows that distinct genotypes encoding identical phenotypes can even persist stably within a species \citep{phillips1996maintenance}.
 
@@
 
-It is not a new observations that there is often more than one way to do the same thing,
+It is not a new observation that there is often more than one way to do the same thing,
 and that this may lead to speciation or that speciation can be the result of neutral and quasi-neutral processes \citep{wright1932roles, nei1983models, barton1989divergence, barton2001role, gavrilets2004fitness}.
+Incompatibilities between distinct genotypes encoding identical phenotypes can
+even persist stably within a species \citep{phillips1996maintenance}.
 For instance, \citet{tulchinsky2014hybrid}
@@
 %REFERENCE: Barton's paper (PETER). \citep{barton2010natural}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-We first describe the class of models we study, motivated by gene regulatory networks. 
-The behavior (i.e., the temporal dynamics) of such a system is determined by a collection of $n$ coregulating molecules 
--- such as transcription factors -- as well as by external or environmental inputs.
+Using gene regulatory network dynamics as the basis for our models, we begin with a collection of $n$ coregulating molecules
+%We first describe the class of models we study, motivated by gene regulatory networks. 
+%The behavior (i.e., the temporal dynamics) of such a system is determined by a collection of $n$ coregulating molecules 
+-- such as transcription factors -- as well as external or environmental inputs.
 % We model the temporal dynamics of the concentrations of a collection of $n$ coregulating molecules
@@
 The rate at which the $i^\text{th}$ concentration changes
-is a weighted sum of the concentrations of the other concentrations
+is a weighted sum of the concentrations
 as well as the input:
@@
 so that the kryptotype measures deviations from initial concentrations. 
-Here $A$ can be any $n \times n$ matrix, $B$ an $n \times m$, and $C$ any $\ell \times n$ dimensional matrix,
+Here $A$ can be any $n \times n$ matrix, $B$ any $n \times m$, and $C$ any $\ell \times n$ dimensional matrix,
 with usually $\ell$ and $m$ less than $n$.
@@
 Although many different biological systems can be modeled with this approach, for clarity, we focus on gene regulatory networks.
-In this interpretation, $A_{ij}$ determines how the $j^\text{th}$ transcription factor regulates the $i^\text{th}$ transcription factor.
+In terms of gene regulatory networks, $A_{ij}$ determines how the $j^\text{th}$ transcription factor regulates the $i^\text{th}$ transcription factor.
 If $A_{ij} > 0$, then $\kappa_j$ upregulates $\kappa_i$, while if $A_{ij} < 0$, then $\kappa_j$ downregulates $\kappa_i$.
@@
 
-\begin{example}[All phenotypically equivalent gene network wirings] \label{ex:all_osc}
+\begin{example}[All equivalent rewirings of the oscillator] \label{ex:all_osc}
 The oscillator of example \ref{ex:oscillator} is minimal, and so any equivalent system is a change of coordinates
@@
 \section*{Acknowledgements}
-We would like to thank Sergey Nuzhdin, Stevan Arnold, Erik Lundgren, and Hossein Asgharian for valuable discussion.
+We would like to thank Sergey Nuzhdin, Stevan Arnold, Michael Turelli, Patrick Phillips, Erik Lundgren, and Hossein Asgharian for valuable discussion.
+We would also like to thank Nick Barton, Sarah Signor and Todd Parsonsfor helpful comments on the manuscript.
 %

### 2018-01-18 kalman_walk_paper.tex
minor edits

@@
 
-It is not a new observation that there is often more than one way to do the same thing,
-and that this may lead to speciation or that speciation can be the result of neutral and quasi-neutral processes \citep{wright1932roles, nei1983models, barton1989divergence, barton2001role, gavrilets2004fitness}.
-Incompatibilities between distinct genotypes encoding identical phenotypes can
-even persist stably within a species \citep{phillips1996maintenance}.
+It is not a new observation that there is often more than one way to do the same thing, or that speciation can be the result of neutral and quasi-neutral processes \citep{wright1932roles, nei1983models, barton1989divergence, barton2001role, gavrilets2004fitness}.
+%Incompatibilities between distinct genotypes encoding identical phenotypes can
+%even persist stably within a species \citep{phillips1996maintenance}.
 For instance, \citet{tulchinsky2014hybrid}

### 2018-01-20 kalman_walk_paper.tex
minor edits
some rewrite of refs

@@
 
-%\plr{Additional ideas to consider adding:}
-%\begin{itemize}
-%    \item speciation literature
-    %\item Haldane's rule (easy point in discussion: makes F1s look like F2s)
-    %\item add point about F1: quartic and F2: quadratic to results and maybe abstract/discussion
-    %\item hybrid vigor (need to do calculation)
- %   \item discuss linearity, linearization, and canalization in introduction
-    %\item obtain estimates of variation in $A$ from thermodynamic occupancy model
- %   \item ``On the origin of species not by means of natural selection''
-%\end{itemize}
-
-%\plri{Note: in the \LaTeX source I'm putting in semantic linebreaks, so it's easy to edit and move around phrases and ideas.}
-
-%\plri{Need to come up with a consistent term for ``the $A_{ij}$''s. -- ``regulatory coefficients''? ``genotype''?}
-
 %%%%%%%%%%%%%%%%%%%%%%
@@
 
-It is not a new observation that there is often more than one way to do the same thing, or that speciation can be the result of neutral and quasi-neutral processes \citep{wright1932roles, nei1983models, barton1989divergence, barton2001role, gavrilets2004fitness}.
-%Incompatibilities between distinct genotypes encoding identical phenotypes can
-%even persist stably within a species \citep{phillips1996maintenance}.
-For instance, \citet{tulchinsky2014hybrid}
+% wright1932roles mentions possibility of equivalent combinations
+% wright1934evolution looks at allele freq dynamics under stabilizing selection on a polygenic quantitative trait and discusses possibility for speciation this way
+% barton1989divergence not available but likely following on from barton1986maintenance analyzing multiple optima in a similar model
+% nei1983models simulates from simple stepwise mut model on one or two loci (without interactions) where incompat betw diff alleles increases with number of steps betw them
+% sved1981twosex mathematically analyzes and (in another paper) simulates speciation in a model of male + female quantitative traits indep'ly evolving with attraction between matching traits
+% barton2010natural has a good historical review of natural selection and speciation; then analyzes a two-niche levene model
+% gavrilets2004fitness talks a lot about quasineutral paths in holey fitness landscapes
+% phillips1996maintenance shows that subpop structure can stably maintain equivalent but incompat genotypes
+% tulchinsky2014hybrid simulates a single promoter region with a bioenergentic model with directional and stabilizing selection, found incompat in both cases
+% porter2002speciation simulated speciation under a simple model of a regulatory cascade
+%%% Fisher's geom model: (incompat due to segregation variance)
+% barton2001role looks at hybrids betw pops adapting to shifting optima in Fisher's geometric model 
+% fraisse2016genetics looks at stabilizing selection using Fisher's geometric model, showing it predicts lots of observations
+% chevin2010niche looks at incompats after adaptation in Fisher's geom model and how it depends on dimension
+% simon2017coadapted follows up on the Fisher's geom model predictions, also comparing to some real data
+
+It is not a new observation that there is often more than one way to do the same thing, 
+or that speciation can be the result of (nearly) neutral processes.
+This has been analyzed for traits under stabilizing selection determined additively by alleles at many loci 
+\citep{wright1934evolution,barton1986maintenance,barton2001role}
+and for pairs of traits that must match but whose value is unconstrained \citep{sved1981twosex}.
+\citet{phillips1996maintenance} showed that population structure
+can allow long-term stable coexistence of incompatible genotypes encoding identical phenotypes.
+Recent work has shown that Fisher's geometric model of stabilizing selection
+predicts many general empirical patterns \citep{fraisse2016genetics,simon2017coadapted}.
+We use similar tools in the second half of this paper,
+differing in that we have a submanifold of optima rather than a single point.
+Using simulations, \citet{tulchinsky2014hybrid}
 %(and similarly \citet{khatri2015simple})
@@
 that would be inaccessible to population simulations.
-%REFERENCE: Barton's paper (PETER). \citep{barton2010natural}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 and appears to be consistent with the ``dominance'' theory \citep{orr1997haldane, coyne1998evolutionary, turelli1995dominance} which is often invoked to explain Haldane's rule.
+\plr{Explain that the effects on fitness are dominant.}
 %and makes several different predictions than the dominance theory as formulated by \citet{turelli1995dominance}.
@@
 then heterogametic $F_1$s show variation similar to that seen in $F_2$ hybrids.
+\plr{cite fisher's model stuff}
 
@@
 % this is neutral, in contrast to other models. although some folks advocate nonadaptive speciation.
-% other mechanisms also probly important
+% other mechanisms also probably important
 % but this one may be strong since happens in all networks at once
@@
 and plausible mechanisms for the origin of Dobzhansky--Muller incompatibilities by neutral genetic drift have been proposed \citep{lynch2000origin} 
+\plr{what}
 and under stabilizing selection \citep{fierst},
@@
 We would like to thank Sergey Nuzhdin, Stevan Arnold, Michael Turelli, Patrick Phillips, Erik Lundgren, and Hossein Asgharian for valuable discussion.
-We would also like to thank Nick Barton, Sarah Signor and Todd Parsonsfor helpful comments on the manuscript.
+We would also like to thank Nick Barton, Sarah Signor and Todd Parsons for helpful comments on the manuscript.
 %

### 2018-01-20 kalman_walk_paper.tex
edits

@@
 % wright1932roles mentions possibility of equivalent combinations
-% wright1934evolution looks at allele freq dynamics under stabilizing selection on a polygenic quantitative trait and discusses possibility for speciation this way
+% wright1935evolution looks at allele freq dynamics under stabilizing selection on a polygenic quantitative trait and discusses possibility for speciation this way
 % barton1989divergence not available but likely following on from barton1986maintenance analyzing multiple optima in a similar model
@@
 This has been analyzed for traits under stabilizing selection determined additively by alleles at many loci 
-\citep{wright1934evolution,barton1986maintenance,barton2001role}
+\citep{wright1935evolution,barton1986maintenance,barton2001role}
 and for pairs of traits that must match but whose value is unconstrained \citep{sved1981twosex}.
@@
 Recent work has shown that Fisher's geometric model of stabilizing selection
-predicts many general empirical patterns \citep{fraisse2016genetics,simon2017coadapted}.
+predicts many general patterns from the speciation literature \citep{fraisse2016genetics}.
 We use similar tools in the second half of this paper,
@@
 Haldane's rule, the observation that if only one hybrid sex is sterile or inviable 
-it is likely the heterogametic sex (e.g., the male in XY sex determination systems) \citep{haldane1922sex}.
+it is likely the heterogametic sex (e.g., the male in XY sex determination systems) \citep{haldane1922sex, orr1997haldane}.
 For example, consider an XY species with a two-gene network where the first gene resides on an autosome and the second gene on the X chromosome.
@@
 $F_1$ males are effectively equivalent to purely autosomal-system $F_2$ hybrids. 
-In this setting, Haldane's rule is mechanistically similar to \emph{hybrid breakdown}, the observation that $F_2$ hybrids will often be less fit than $F_1$ hybrid crosses, in the early stages of speciation.
-This mechanism relies on the genetic variation of a system being shared across the sex chromosomes and the autosomes,
-and appears to be consistent with the ``dominance'' theory \citep{orr1997haldane, coyne1998evolutionary, turelli1995dominance} which is often invoked to explain Haldane's rule.
-\plr{Explain that the effects on fitness are dominant.}
+Although many alleles will be dominant if the phenotype--fitness relationship is convex,
+the underlying mechanism does not depend on the \emph{dominance theory} \citet{turelli1995dominance} to explain Haldane's rule:
+instead, it derives from the nature of segregation load.
+% In this setting, Haldane's rule is mechanistically similar to \emph{hybrid breakdown}, 
+% the observation that $F_2$ hybrids will often be less fit than $F_1$ hybrid crosses, in the early stages of speciation.
+% This mechanism relies on the genetic variation of a system being shared across the sex chromosomes and the autosomes,
+% and appears to be consistent with the ``dominance'' theory \citep{orr1997haldane, coyne1998evolutionary, turelli1995dominance} which is often invoked to explain Haldane's rule.
+% \plr{Explain that the effects on fitness are dominant.}
 %and makes several different predictions than the dominance theory as formulated by \citet{turelli1995dominance}.
@@
 
-Above we synthesize concepts and tools from molecular and evolutionary biology with tools from control theory
-to study the evolution of a mechanistic model of the molecular genotype-phenotype map under stabilizing selection.
-This model allows us to analytically describe the processes discussed verbally as
+In this paper, we have used tools from quantitative genetics and control theory
+to study the evolution of a mechanistic model of the molecular genotype-phenotype map under stabilizing selection,
+providing an explicit model of
 phenogenetic drift \citep{weiss2000phenogenetic} and developmental system drift \citep{true2001developmental}.
@@
 and hence the general existence of axes of genetic variation that are not constrained by selection.
-The independent movement of separated populations along these axes
+The independent movement of separated populations along these axes by genetic drift
 can lead to reduction in hybrid viability, and hence speciation,
@@
 This occurs because if the genes underlying a regulatory network are distributed among both autosomes and the sex chromosome, 
-then heterogametic $F_1$s show variation similar to that seen in $F_2$ hybrids.
-\plr{cite fisher's model stuff}
+then heterogametic $F_1$s show variation similar to that seen in $F_2$ hybrids,
+as a result of segregation load \citep[similarly to][]{fraisse2016genetics}.
 
@@
 
-\paragraph{On the origin of species not by means of natural selection?}
+\paragraph{The origin of species not by means of natural selection?}
 % this is neutral, in contrast to other models. although some folks advocate nonadaptive speciation.
@@
 % but this one may be strong since happens in all networks at once
-Although, as classically formulated, 
+As classically formulated, 
 the Dobzhansky-Muller model of hybrid incompatibility is agnostic to the relative importance of neutral versus selective genetic substitutions \citep{coyne1998evolutionary},
-and plausible mechanisms for the origin of Dobzhansky--Muller incompatibilities by neutral genetic drift have been proposed \citep{lynch2000origin} 
-\plr{what}
-and under stabilizing selection \citep{fierst},
-previous authors have argued that neutral processes are likely too slow to be a significant driver of speciation \citep{nei1983models,seehausen2014genomics}.
-(Also note that the mechanism underlying hybrid inviability here 
-resembles the Dobzhansky-Muller model,
-but is more similar to the ``pathway model'' of \citet{lindtke2015genetic}.)
-Using simulations, \citet{porter2002speciation} demonstrated the accumulation of hybrid incompatibilities under directional, but not stabilizing selection, and \citet{palmer2009dynamics} observed the appearance of incompatibilities in populations evolving in a constant environment, only if the populations were suboptimally adapted.
-This, in light of the few known incompatibilities, has lead some to conclude that hybrid incompatibility is typically a byproduct of positive selection \citep{orr2004speciation, schluter2009evidence} 
-or a consequence of genetic conflict \citep{presgraves2010molecular, crespi2013conflictual}.
-In contrast, the model we develop here suggests that even under strictly neutral conditions, 
-system drift can lead to speciation, at a rate fast enough 
-to play a substantial role in species formation across the tree of life. 
-Our results show that hybrids, under neutral processes, break down as a function of genetic distance,
-and so may, in part, explain broad patterns
+and plausible mechanisms have been proposed
+whereby Dobzhansky--Muller incompatibilities could originate
+under neutral genetic drift \citep{lynch2000origin}
+or stabilizing selection \citep{fierst}.
+The same holds for the ``pathway model'' of \citet{lindtke2015genetic},
+which is closer to the situation here.
+However, previous authors have argued that neutral processes are likely too slow to be a significant driver of speciation \citep{nei1983models,seehausen2014genomics}.
+% Using simulations, \citet{porter2002speciation} demonstrated the accumulation of hybrid incompatibilities under directional, but not stabilizing selection, and \citet{palmer2009dynamics} observed the appearance of incompatibilities in populations evolving in a constant environment, only if the populations were suboptimally adapted.
+% In light of the few known incompatibilities, 
+This has led some to conclude that hybrid incompatibility is typically a byproduct of positive selection \citep{orr2004speciation, schluter2009evidence} 
+or a consequence of genetic conflict \citep{presgraves2010molecular, crespi2013conflictual},
+two processes that typically act much more rapidly than genetic drift.
+% In contrast, the model we develop here suggests that even under strictly neutral conditions, 
+% system drift can lead to speciation, at a rate fast enough 
+% to play a substantial role in species formation across the tree of life. 
+However, our calculations suggest that even under strictly neutral processes,
+hybrid fitness breaks down as a function of genetic distance
+rapidly enough to play a substantial role in species formation across the tree of life. 
+This is consistent with broad patterns
 such as the relationship between molecular divergence and genetic isolation 
@@
 seen by \citet{roux2016shedding}, 
-and the clocklike speciation patterns observed by \citet{hedges2015tree}.
+and the clocklike speciation rates observed by \citet{hedges2015tree}.
 % (as diversification is not observed to be accelerated following mass extinctions).
@@
 are plausible drivers of speciation, and may even interact.
-For instance, reinforcement due to local adaptation could in some situations provide a selective pressure
+Many of our observations carry over to models of directional selection --
+for instance, rapid drift along the set of equivalent systems
+could be driven by adaptation in a different, pleiotropically coupled system.
+Or, reinforcement due to local adaptation might provide a selective pressure
 that speeds up system drift.
-It remains to be seen how the relative strengths of these forces compare.
 Furthermore,
@@
 the cumulative impact of system drift across the many different networks an organism relies on may be substantial. 
-
-Of course the basic assumptions of this model will be violated in practice (constant selective pressures, etc.), however, this model can function as a ``neutral null'' description of gene network evolution (a strategy advocated for by \citet{lynch2007frailty, fay2008evaluating, koonin2016splendor}).
+It remains to be seen how the relative strengths of these forces compare.
+
+% Of course the basic assumptions of this model will be violated in practice (constant selective pressures, etc.), however, this model can function as a ``neutral null'' description of gene network evolution (a strategy advocated for by \citet{lynch2007frailty, fay2008evaluating, koonin2016splendor}).
 If this model succeeds in describing actual system evolution, it can be inferred that the mechanism underlying species formation does not require the inclusion of adaptation or changes in selection to explain the emergence of hybrid incompatibility.
 
-\paragraph{Population isolates and genetic load}
-Above we have mostly discussed the case of two separated populations of equal size.
-In contrast, consider a small subset of a large population that is suddenly isolated.
-It will have a large amount of genetic variance, thanks to its history,
-but a small population size -- and so can drift quite quickly until it exhausts its genetic variation.
-By this point, it could have accumulated substantial incompatibility with the parental population.
-Similarly, in a species with substantial isolation by distance, geographically local systems drift
-might lead to genetic load due only to recombination, much as in \citep{phillips1996maintenance}.
+% \paragraph{Population isolates and genetic load}
+% Above we have mostly discussed the case of two separated populations of equal size.
+% In contrast, consider a small subset of a large population that is suddenly isolated.
+% It will have a large amount of genetic variance, thanks to its history,
+% but a small population size -- and so can drift quite quickly until it exhausts its genetic variation.
+% By this point, it could have accumulated substantial incompatibility with the parental population.
+% Similarly, in a species with substantial isolation by distance, geographically local systems drift
+% might lead to genetic load due only to recombination, much as in \citep{phillips1996maintenance}.
+
+\paragraph{Fisher's geometric model}
+There has been substantial work on models of quadratic stabilizing selection
+around a single optimum -- i.e., Fisher's geometric model \citep[e.g.][]{simon2017coadapted}.
+Our model would have this form if viewed only in phenotype space, 
+where there is a single optimum --
+but differs by explicitly considering evolutionary dynamics of underlying genotype-derived regulatory matrices.
+These make the quantitative genetic model we are led to
+a generalization of Fisher's geometric model,
+with distinct behavior,
+due to the existence of neutral directions in genotype space,
+and curvature of the optimal set,
+However, there are many shared aspects:
+for instance, \citet{barton1989divergence} also found incompatibilities accumulating
+with genetic drift.
 
@@
 After selection, we have reproduction:
-suppose this occurs as in the infinitesimal model \citep{barton2016infinitesimal},
+suppose this occurs as in the infinitesimal model \citep{barton2017infinitesimal},
 so that each offspring of parents with traits $x$ and $y$
@@
 suppose that a trait is determined by $L$ unlinked, biallelic loci,
-and that the $i^\mathrm{th}$ locus has two alleles with additive effects $\pm x_i$,
-so that being homozygous for the $+$ allele contributes $+2x_i$ to the trait.
+and that the $i^\mathrm{th}$ locus has two alleles with additive effects $\pm \alpha_i$,
+so that being homozygous for the $+$ allele contributes $+2 \alpha_i$ to the trait.
 For simplicity, we will neglect the effects of selection.
@@
 \begin{align*}
-    m &= 2 \sum_i (2 p_i - 1) x_i \\
-    s^2 &= 4 \sum_i p_i (1-p_i) x_i^2 .
+    m &= 2 \sum_i (2 p_i - 1) \alpha_i \\
+    s^2 &= 4 \sum_i p_i (1-p_i) \alpha_i^2 .
 \end{align*}
@@
 If the alleles are at Hardy-Weinberg proportions,
-then since segregation acts like a fair coin flip, a heterozygous locus contributes $x_i^2/4$ to the variance,
+then since segregation acts like a fair coin flip, a heterozygous locus contributes $\alpha_i^2/4$ to the variance,
 and so the \emph{mean} segregation variance, averaging across parents, is
-% two parents, prob 2p(1-p) for prob the locus is heterozygous, variance 0.5(1-0.5)x^2.
-\begin{align*}
-    R_0(p) = \sum_i p_i (1-p_i) x_i^2 .
+% two parents, prob 2p(1-p) for prob the locus is heterozygous, variance 0.5(1-0.5)\alpha^2.
+\begin{align*}
+    R_0(p) = \sum_i p_i (1-p_i) \alpha_i^2 .
 \end{align*}
@@
 \begin{align*}
-    R_1(p,q) &= \frac{1}{2} \sum_i p_i^2 (1-p_i)^2 x_i^2 
-                + \frac{1}{2} \sum_i q_i^2 (1-q_i)^2 x_i^2 \\
+    R_1(p,q) &= \frac{1}{2} \sum_i p_i^2 (1-p_i)^2 \alpha_i^2 
+                + \frac{1}{2} \sum_i q_i^2 (1-q_i)^2 \alpha_i^2 \\
                 &= (R_0(p) + R_0(q))/2 .
@@
 \begin{align*}
-    R_2(p,q) &= \frac{1}{2} \sum_i \left(p_i (1-q_i) + (1-p_i) q_i \right) x_i^2  .
+    R_2(p,q) &= \frac{1}{2} \sum_i \left(p_i (1-q_i) + (1-p_i) q_i \right) \alpha_i^2  .
 \end{align*}
@@
     &\approx 
-    R_0 + 2 \sum_i p_i (1-p_i) x_i^2 t/N \\
+    R_0 + 2 \sum_i p_i (1-p_i) \alpha_i^2 t/N \\
     &=
@@
 \begin{align} \label{eqn:var_dx}
-    8 \sum_i p_i (1-p_i) x_i^2 t / N = 8 R_0 t /N .
+    8 \sum_i p_i (1-p_i) \alpha_i^2 t / N = 8 R_0 t /N .
 \end{align}

### 2018-01-26 kalman_walk_paper.tex
final edits

@@
 how a gene network underlying a conserved phenotype evolves,
-as the genetic drift of small mutational changes to these molecular pathways
+as the genetic drift of small changes to these molecular pathways
 cause a population to explore the set of mechanisms with identical phenotypes.
@@
 Finally,
-the field of \emph{inverse problems} studies those cases where,
-even if a one-to-one mapping between model and behavior is possible in theory,
-even tiny amounts of noise can make inference problems nonidentifiable in practice \citep{petrov2005well}.
+the field of \emph{inverse problems} studies those cases in which,
+despite the existence of a theoretical one-to-one mapping between a model and behavior,
+tiny amounts of noise make inference problems nonidentifiable in practice \citep{petrov2005well}.
+%even if a one-to-one mapping between model and behavior is possible in theory,
+%even tiny amounts of noise can make inference problems nonidentifiable in practice \citep{petrov2005well}.
 
@@
 In this paper we outline a theoretical framework to study the evolution of biological systems, such as gene regulatory networks.
-We study the evolution of an optimally adapted population subject to stabilizing selection for phenotype.
+We study the evolution of an optimally adapted population under stable conditions, that is where phenotype is under stabilizing selection.
 Even if the phenotype remains stable over evolutionary time, the underlying mechanism might not remain so,
@@
 Since these phenotypically equivalent gene networks are not necessarily compatible with one another, 
-system drift may result in reproductive incompatibility between populations 
-isolated for a sufficiently long period of time, 
+system drift may result in reproductive incompatibility between sufficiently isolated populations,
+%isolated for a sufficiently long period of time, 
 even in the absence of any sort of adaptive, selective, or environmental change. 
 We then use quantitative genetic theory to estimate how quickly 
-reproductive incompatibility due to system drift will manifest
+reproductive incompatibility due to system drift manifests.
 
@@
 or that speciation can be the result of (nearly) neutral processes.
-This has been analyzed for traits under stabilizing selection determined additively by alleles at many loci 
-\citep{wright1935evolution,barton1986maintenance,barton2001role}
+The potential for speciation has been analyzed in models of 
+traits under stabilizing selection determined additively by alleles at many loci 
+\citep{wright1935evolution,barton1986maintenance,barton2001role},
+in related fitness landscape models \citep{fraisse2016genetics},
 and for pairs of traits that must match but whose value is unconstrained \citep{sved1981twosex}.
-\citet{phillips1996maintenance} showed that population structure
-can allow long-term stable coexistence of incompatible genotypes encoding identical phenotypes.
-Recent work has shown that Fisher's geometric model of stabilizing selection
-predicts many general patterns from the speciation literature \citep{fraisse2016genetics}.
-We use similar tools in the second half of this paper,
-differing in that we have a submanifold of optima rather than a single point.
-Using simulations, \citet{tulchinsky2014hybrid}
-%(and similarly \citet{khatri2015simple})
-looked for hybrid incompatibility
-following regulatory sequence drift,
-and \citet{porter2002speciation} studied speciation in populations expressing a simple regulatory cascade.
-Both found rapid speciation following directional selection,
+% Recent work has shown that Fisher's geometric model of stabilizing selection
+% predicts many general patterns from the speciation literature \citep{fraisse2016genetics}.
+It has also been shown that population structure
+can allow long-term stable coexistence of incompatible genotypes encoding identical phenotypes \citep{phillips1996maintenance}.
+However, previous simulations of system drift in regulatory sequences \citep{tulchinsky2014hybrid}
+and a regulatory cascade \citet{porter2002speciation}
+found that speciation could be rapid under directional selection
 but only equivocal support for speciation under models of purely neutral drift.
@@
 
-Although many different biological systems can be modeled with this approach, for clarity, we focus on gene regulatory networks.
+% Although many different biological systems can be modeled with this approach, for clarity, we focus on gene regulatory networks.
 In terms of gene regulatory networks, $A_{ij}$ determines how the $j^\text{th}$ transcription factor regulates the $i^\text{th}$ transcription factor.
@@
 and $C$ might determine the rate of production of downstream enzymes.
-
 %% {Linearity} also ``zero-state equivalence''
-Here we have assumed that the system is linear,
-and begins from the ``zero'' state ($\kappa(0)=0$).
-Of course, neither of these are necessarily true for real systems,
-but the dynamics of most nonlinear systems can be approximated locally by a linear systems near most points. 
-Furthermore, the ease of analyzing linear systems makes this an attractive place to start.
+% Here we have assumed that the system is linear,
+% and begins from the ``zero'' state ($\kappa(0)=0$).
+% Of course, neither of these are necessarily true for real systems,
+% but the dynamics of most nonlinear systems can be approximated locally by a linear systems near most points. 
+% Furthermore, the ease of analyzing linear systems makes this an attractive place to start.
 To demonstrate this approach, we apply it to construct a simple gene network in Example \ref{ex:oscillator} below.
@@
 Since gene networks can grow or shrink following gene duplications and deletions, 
-these additional degrees of freedom can apply in principle to any system.
+these additional degrees of freedom can apply, in principle, to any system.
 
@@
 Exactly how the coefficients 
-(i.e., entries of $A$, $B$ or $C$)
+(i.e., entries of $A$, $B$ and $C$)
 of a haploid system inherited by an offspring from her diploid parent
@@
 and hence is determined by the promoter region of gene $i$,
-the elements of a row of $A$ to tend to be inherited together,
+the elements of a row of $A$ tend to be inherited together,
 which will create covariance between entries of the same row.
@@
 is proportional to the difference between the two parental systems.
-This is certainly true if each coefficient is determined by a single nonrecombining locus,
-so that each coefficient in the system produced by meiosis is an independent random choice
-between the two parental coefficients.
-On the other hand, if a coefficient is determined by additive contributions of independently segregating loci,
-then the law of total variance applied after conditioning on the parental origin of each allele
-implies that the variance is equal to one-quarter of the squared difference between parental coefficients.
-
-Offspring formed from two phenotypically identical systems do not necessarily exhibit the same phenotype as both of its parents -- in other words $\allS$, the set of all systems phenotypically equivalent to a given one, is not, in general, closed under averaging or recombination.
-If sexual recombination among systems drawn from $\allS$ yields systems with divergent phenotypes, populations containing significant diversity in $\allS$ can carry genetic load, and isolated populations may fail to produce hybrids with viable phenotypes.
+% This is certainly true if each coefficient is determined by a single nonrecombining locus,
+% so that each coefficient in the system produced by meiosis is an independent random choice
+% between the two parental coefficients.
+% On the other hand, if a coefficient is determined by additive contributions of independently segregating loci,
+% then the law of total variance applied after conditioning on the parental origin of each allele
+% implies that the variance is equal to one-quarter of the squared difference between parental coefficients.
+
+Offspring formed from two phenotypically identical systems do not necessarily exhibit the same phenotype as both of its parents 
+-- in other words $\allS$, the set of all systems phenotypically equivalent to a given one, is not, in general, closed under averaging or recombination.
+If sexual recombination among systems drawn from $\allS$ yields systems with divergent phenotypes, 
+populations containing significant diversity in $\allS$ can carry genetic load, and isolated populations may fail to produce hybrids with viable phenotypes.
 
@@
     %\textbf{(left)} 
-    \caption{Mean hybrid phenotypic distance from optimum computed with equation \ref{eqn:distance}, using $\rho(t) = \exp(-t/4\pi)$
+    \caption{Mean hybrid phenotypic distance from optimum computed with equation \eqref{eqn:distance}, using $\rho(t) = \exp(-t/4\pi)$
     for $F_1$ (black) and $F_2$ (blue) hybrids between $A(0)$ and $A(\epsilon)$ parent oscillators. Genetic distance is computed as 
-    %Frobenius norm
-    %Euclidean distance
-    $\sqrt{\sum_{ij} (A_{ij}(0) - A_{ij}(\epsilon))^2}$
-    %between two systems
-    .}
-    %\textbf{(right)} Same, but for $F_2$ offspring (one line per type of $F_2$).
-    %\plri{make this $D$ not $D^2$ and say which line is the overall average $F_2$.
-    %}
+    $\left(\sum_{ij} (A_{ij}(0) - A_{ij}(\epsilon))^2\right)^{1/2}$.
+    }
     \label{fig:osc_incompat}
@@
 has phenotype determined by $A = \left[ \begin{smallmatrix} A_1 & A_2 \\ X_1 & X_2 \end{smallmatrix} \right]$, 
-if dosage compensation upregulates heterogametes by a factor of 2 relative to homogametes,
-% as in \emph{Drosophila}
+  if dosage compensation upregulates heterogametes by a factor of two relative to homogametes (as with \emph{Drosophila}),
 while a female homozygous for the haplotype
@@
 $\left[\begin{smallmatrix} (A_1 + \bar A_1)/2 & (A_2 + \bar A_2)/2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$.
-If both genes resided on the autosomes this system would only be possible in an $F_2$ cross.
+If both genes resided on the autosomes, this system would only be possible in an $F_2$ cross.
 More generally, 
 if the regulatory coefficients for a system are shared between the sex and one or more autosomal chromosomes, 
-$F_1$ males are effectively equivalent to purely autosomal-system $F_2$ hybrids. 
+$F_1$ males are effectively equivalent to purely autosomal-system $F_2$ hybrids, and recall that $F_2$s are significantly less fit on average than $F_1$s (see Figure \ref{fig:osc_incompat}).
 Although many alleles will be dominant if the phenotype--fitness relationship is convex,
@@
 and that these mechanisms may fail to produce viable hybrids.
-This leads to the question: does evolution shift molecular mechanisms
+Does evolution shift molecular mechanisms
 fast enough to be a significant driver of speciation?
@@
 The other relevant scale we denote by $\gamma$,
-which the scale on which distance to phenotypic optimum changes as $x$ moves away from the optimal set, $\allS$.
+which is the scale on which distance to phenotypic optimum changes as $x$ moves away from the optimal set, $\allS$.
 Concretely, $\gamma$ is
@@
 The $F_2$ fitness is from equation \eqref{eqn:fit2},
-and the $F_1$ fitness is determined only by the exponential term of that equation.
-is small relative to that of increased variance.
+and the $F_1$ fitness is determined only by the exponential term of that equation,
+which is small relative to that of increased variance.
 \label{fig:speciation_rates}}
@@
 Nonetheless, these measures likely give us an idea of the scale of variability.
-\citet{kasowski2010variation} found differential occupancy in 7.5\%
-of binding sites of a transcription factor (p65) between human individuals.
-\citet{verlaan2009targeted} showed that cis-regulatory variation
-accounts for around 2--6\% of expression variation in human blood-derived primary cells,
-while \citet{lappalainen2013transcriptome} found that human population variation 
-explained about 3\% of expression variation.
+%\citet{kasowski2010variation} found 
+It has been shown that between human individuals, there is differential occupancy in 7.5\%
+of binding sites of a transcription factor (p65) \cite{kasowski2010variation}.
+%\citet{verlaan2009targeted} showed that 
+It has also been inferred that cis-regulatory variation
+accounts for around 2--6\% of expression variation in human blood-derived primary cells \cite{verlaan2009targeted}, and
+%while \citet{lappalainen2013transcriptome} found 
+that human population variation 
+explained about 3\% of expression variation \cite{lappalainen2013transcriptome}.
 Allele-specific expression is indicative of standing genetic \emph{cis}-regulatory variation;
-\citet{wang2017bayesian} found allele-specific expression in 7.2--8.5\% of transcripts of a flycatcher species,
-while \citet{tung2015genetic} observed allele-specific expression in 23.4\% of genes studied in a baboon species. 
+%\citet{wang2017bayesian} found 
+allele-specific expression in 7.2--8.5\% of transcripts of a flycatcher species has been observed \cite{wang2017bayesian}, as well as
+%while \citet{tung2015genetic} observed 
+allele-specific expression in 23.4\% of genes studied in a baboon species \cite{tung2015genetic}. 
 Taken together, this suggests that variation in the entries of $A$
@@
 
-In this paper, we have used tools from quantitative genetics and control theory
-to study the evolution of a mechanistic model of the molecular genotype-phenotype map under stabilizing selection,
-providing an explicit model of
+In this paper, we use tools from quantitative genetics and control theory
+to study the evolution of a mechanistic model of the genotype-phenotype map, 
+in which the phenotype is subject to stabilizing selection.
+In so doing, we provide an explicit model of
 phenogenetic drift \citep{weiss2000phenogenetic} and developmental system drift \citep{true2001developmental}.
@@
 It also implies that nearly all systems are nonidentifiable,
-and hence the general existence of axes of genetic variation that are not constrained by selection.
+and that, in general, there exist axes of genetic variation unconstrained by natural selection.
 The independent movement of separated populations along these axes by genetic drift
-can lead to reduction in hybrid viability, and hence speciation,
-at a speed that depends on effective population size and amount of genetic variation. 
+can lead to a significant reduction in hybrid viability, and thus precipitate speciation,
+at a speed dependent on the effective population size and the amount of genetic variation. 
 In this model, at biologically reasonable parameter values,
 system drift is a significant -- and possibly rapid -- driver of speciation.
-This may at first be surprising because
-hybrid inviability appears as a consequence of recombining different, yet functionally equivalent, mechanisms.
-
-Consistent with empirical observation of hybrid breakdown \citep[e.g.][]{plotner2017chlorosis}, 
+This may be surprising because
+hybrid inviability appears as a consequence of recombining different, yet functionally equivalent, mechanisms, and since species are often defined by their unique adaptations or morphologies.
+
+Consistent with empirical observation of hybrid breakdown (e.g., \citet{plotner2017chlorosis}), 
 we see that the fitnesses of $F_2$ hybrids drop at a much faster rate than those of $F_1$s.
@@
 This occurs because if the genes underlying a regulatory network are distributed among both autosomes and the sex chromosome, 
-then heterogametic $F_1$s show variation similar to that seen in $F_2$ hybrids,
-as a result of segregation load \citep[similarly to][]{fraisse2016genetics}.
+then heterogametic $F_1$s show variation (and fitnesses) similar to that seen in $F_2$ hybrids.
 
@@
 or stabilizing selection \citep{fierst}.
-The same holds for the ``pathway model'' of \citet{lindtke2015genetic},
+The same holds for the ``pathway model'' \citep{lindtke2015genetic},
 which is closer to the situation here.
@@
 with distinct behavior,
-due to the existence of neutral directions in genotype space,
-and curvature of the optimal set,
+due to the existence of neutral directions in genotype space.
 However, there are many shared aspects:
@@
 with genetic drift.
+Fisher's geometric model also predicts Haldane's rule
+and many other empirically observed patterns \citep{fraisse2016genetics},
+but for different underlying reasons.
 
@@
 \section*{Acknowledgements}
-We would like to thank Sergey Nuzhdin, Stevan Arnold, Michael Turelli, Patrick Phillips, Erik Lundgren, and Hossein Asgharian for valuable discussion.
-We would also like to thank Nick Barton, Sarah Signor and Todd Parsons for helpful comments on the manuscript.
-%
-%\plr{Josh: please fix capitalization in the references.  To do this, anything that needs to be capitalized gets put in curly braces, which tells bibtex to render the capitalization 'as-is'.}
+
+    We would like to thank Sergey Nuzhdin, Stevan Arnold, Michael Turelli, Patrick Phillips, Erik Lundgren and Hossein Asgharian for valuable discussion. 
+    We would also like to thank Nick Barton, Sarah Signor and Todd Parsons for very helpful comments on the manuscript.
+    Work on this project was supported by funds from
+    the Sloan Foundation and the NSF (under DBI-1262645) to PR.
 
@@
 \end{document}
-
-%\section*{Examples}
-%\begin{example}[Metabolic network] \label{ex:metabolic}
-%Consider an organism that can metabolize two different sugars (present at logarithmic concentrations $u_{1}$ and $u_{2}$), 
-%with two enzymes (at log concentrations $\phi_{1}$ and $\phi_{2}$).
-%Further suppose that the second sugar is prefered,
-%%and that the organism has a regulatory network that can deploy situation specific metabolic strategy.
-%and that
-%depending on $u_{1}$ and $u_{2}$ the organism will synthesize an appropriate $\phi_{1}$ and $\phi_{2}$. 
-%Furthermore, assume this system contains at least two transcription factors, whose log concentrations are $\kappa_{1}$, $\kappa_{2}$, $\dots \kappa_{n}$.
-%Minimally such a system may have the architecture,
-%%\begin{align*}
-%%    \mathcal{S}_{\min} = \left\{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 0 & -1 \\ 0 & 0 \end{bmatrix} \begin{bmatrix} \kappa_{1} \\ \kappa_{2} \end{bmatrix} (t) + \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \begin{bmatrix} u_{1} \\ u_{2} \end{bmatrix} (t) \\[11pt]
-%%    \phi(t) &= \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \begin{bmatrix} \kappa_{1} \\ \kappa_{2} \end{bmatrix} (t) \end{array} \right.
-%%\end{align*}
-%  $\Sys_{\min}(U) = (U \left[ \begin{smallmatrix} 0 & -1 \\ 0 & 0 \end{smallmatrix} \right] U^{-1}, U 
-%    %\left[\begin{smallmatrix} 1 & 0 \\ 0 & 1 \end{smallmatrix}\right]
-%      ,
-%     % \left[\begin{smallmatrix} 1 & 0 \\ 0 & 1 \end{smallmatrix}\right]
-%        U^{-1})$. 
-%%
-%%The impulse response of the system is,
-%%$h(t) = \left[\begin{smallmatrix} 1 & - t \\ 0 & 1 \end{smallmatrix}\right]$.
-%    We can find alternative equivalent systems by changning coordinates ($U \rightarrow U'$) or, more generally, by applying the Kalman decomposition \ref{apx:kalman}.
-To illustrate that phenotypic invariance does not require dimensional invariance, we apply the Kalman decomposition to $\Sys_{\min}$ to find
-%    %This may happen if a system is co-opted from another, or may be a consequence of gene duplication and deletion. 
-%  $\Sys(\small D_{1-3},V) \! =  \! \left( \! \begin{array}{ll}
-%     V \!
-\left[\begin{smallmatrix} D_{1} & D_{2} \\ 0 & A \end{smallmatrix}\right] \!
-V^{-1}, V \!
-\left[\begin{smallmatrix} D_{3} \\ B \end{smallmatrix}\right], \left[\begin{smallmatrix} 0 & C \end{smallmatrix}\right] \!
-V^{-1} \end{array} \!
-\right)
-%$, both of which are in $\allS$
-%%  \begin{align*}
-%%    \Sys(X_{1-3},V) = \left( \begin{array}{ll}
-%%     V \left[\begin{smallmatrix} X_{1} & X_{2} \\ 0 & A_{\min} \end{smallmatrix}\right] V^{-1}, V \left[\begin{smallmatrix} X_{3} \\ B_{\min} \end{smallmatrix}\right], \left[\begin{smallmatrix} 0 & C_{\min} \end{smallmatrix}\right] V^{-1} \end{array} \right)
-%%    \end{align*}
-%  ($V$ can be any $4$-dimensional and invertible matrix, $D_{1-3} \in \R^{2 \times 2}$, and $(A,B,C) \in \Sys_{\min}$).
-%%So for example, some system can be wired as follows, and still be input-output equivalent to the minimal metabolic system $\mathcal{S}_{\min}$:
-%% \begin{align*}
-%%   A' &= \begin{bmatrix}
-%%     1.6923  &  1.5385 &  -2.6154 &  -2 \\
-%%     0.8462  &  0.7692 &  -1.3077 &  -1 \\
-%%     1.2692  &  1.1538 &  -1.9615 &  -1.5 \\
-%%     0.4231  &  0.3846 &  -0.6538 &  -0.5
-%%   \end{bmatrix} \\
-%%   B' &= \begin{bmatrix} 4 & 4 \\ 2 & 2 \\ 3 & 3 \\ 1 & 3 \end{bmatrix} \\
-%%     C' &= \begin{bmatrix}  0.2692  &  0.1538  &  0.0385 &  -0.5 \\
-%%     -0.4231 &  -0.3846 &   0.6538  &  0.5 \end{bmatrix}
-%% \end{align*}
-%%
-%%Despite the present example consisting of a minimal $2 \times 2$ and a non-minimal $4 \times 4$ system, 
-%%any $m$-dimensional system can be constructed using this method 
-%%-- applying a change of coordinates to the Kalman decomposition 
-%%-- to construct a mechanistically different system with identical phenotypic dynamics. 
-%%
-%%When applying the Kalman decomposition to real biological networks, one may have to restrict the free parameters, as physiology and biochemistry might not allow all mathematically possible coefficients.
-We note that as there are so many degrees of freedom, it may be difficult to infer system function solely from structure.
-%\end{example}

### 2018-03-06 kalman_walk_paper.tex
edits

@@
 can allow long-term stable coexistence of incompatible genotypes encoding identical phenotypes \citep{phillips1996maintenance}.
-However, previous simulations of system drift in regulatory sequences \citep{tulchinsky2014hybrid}
-and a regulatory cascade \citet{porter2002speciation}
+Despite the findings of these quantitative genetics models, previous simulations of system drift in regulatory sequences \citep{tulchinsky2014hybrid}
+and a regulatory cascade \citep{porter2002speciation}
 found that speciation could be rapid under directional selection
@@
 %Additionally, \citet{fierst} observed speciation due to an increase of segregation variance under balancing selection, in models of weak to moderate, but not strong epistasis.
+It is known that hybrid incompatibility can result from the increased segregation variance of a quantitative trait \citep{slatkin1994segregation, chevin2014niche}.
+Here, by focusing on the explicit biological function or physiology of a phenotype, we show that the evolution of even extremely simple phenotypes can appropriately be approximated by quantitative genetics.
+This is because, like the genetic code, the genotype-phenotype ``code'' is fundamentally degenerate -- as we make explicit -- and thus ammendable to quantitative genetic analysis.
+
 Our model differs from previous work in that we use linear systems theory

### 2018-03-07 attic/biosystem_examples.tex
tidying

--- 2017-03-23 biosystem_examples.tex
@@
 coli} that over or under express lactase, governed by the \emph{lac} operon, in a given sugar environment, their fitness (growth rate) suffers.
+
+    We are interested in network rewiring rates. This can be modelled as a Brownian motion of the free parameters,
+    \begin{align*}
+      A_{t} &= A(f(W_{t}), g(W_{t})) \\
+      \mathbb{E} [ | A_{\epsilon} - A_{0} | ] &= \epsilon + \mathcal{O}(\epsilon^{2})
+    \end{align*}
+
+    This realization is minimal as the rank of both the reachability and observability matrices is full ($4$).
+
+    Note: check that $V$ is the \emph{only} invertible matrix that preserves $C$ and $B$. 
     \subsection{Notes}

### 2018-03-07 kalman_walk_paper.tex
refs and abstract

@@
 %The evolutionary conservation of phenotype under selective and environmental stasis does not necessitate conservation of the underlying mechanism, as distinct molecular pathways can realize identical phenotypes.
-Even if a species' phenotype does not change over evolutionary time, the underlying mechanism may change, as distinct molecular pathways can realize identical phenotypes.
-Here we use quantitative genetics and linear system theory to study
-how a gene network underlying a conserved phenotype evolves,
+Even if a species' phenotype does not change over evolutionary time, 
+the underlying mechanism may change, 
+as distinct molecular pathways can realize identical phenotypes.
+Here we use linear system theory to explore the consequences of this idea,
+describing how a gene network underlying a conserved phenotype evolves,
 as the genetic drift of small changes to these molecular pathways
 cause a population to explore the set of mechanisms with identical phenotypes.
-% Treating an organism as a ``black box''
 To do this, we model an organism's internal state as a linear system of differential equations
 for which the environment provides input and the phenotype is the output,
-%which translates some environmental input
-%to an output that is subject to the pressures of natural selection-- the phenotype.
 in which context there exists 
 an exact characterization of the set of all mechanisms that give the same input--output relationship.
-%There exists an exact expression for the set of all linear mechanisms with identical phenotypes, 
-%Due to the genetic drift of small mutational tweaks to these molecular pathways, we expect an evolving population to explore this set.
 This characterization implies that selectively neutral directions in genotype space should be common
-% there is never a unique network architecture for any phenotype
 and that the evolutionary exploration of these distinct but equivalent mechanisms
@@
 and thus speciation, in fewer than $N_e$ generations.
-%over a number of generations proportional to the effective population size.
 This model also naturally predicts Haldane's rule, 
 thus providing another possible explanation
-of why heterogametic hybrids tend to be disrupted more often than homogametes during the early stages of speciation.
+of why heterogametic hybrids tend to be disrupted more often than homogametes 
+during the early stages of speciation.
 \end{abstract}
@@
 %%% Fisher's geom model: (incompat due to segregation variance)
-% barton2001role looks at hybrids betw pops adapting to shifting optima in Fisher's geometric model 
-% fraisse2016genetics looks at stabilizing selection using Fisher's geometric model, showing it predicts lots of observations
-% chevin2010niche looks at incompats after adaptation in Fisher's geom model and how it depends on dimension
+% poon2000compensating also uses SSSWM with Fisher's model and derives some analytical results for load in small populations; early citation for this process
+% slatkin1994segregation shows that segregation variance increases linearly with divergence if variation is controlled by intermediate freq alleles;
+%   but if it is determined by rare alleles (at mut-sel balance), then it plateaus (at twice the within-pop value?)
+% barton2001role looks at hybrids betw pops adapting to shifting optima in Fisher's geometric model : under fluctuating optima
+%       hybrids will suffer because of cryptic segregating variation
+% fraisse2016genetics uses SSWM; looks at stabilizing selection using Fisher's geometric model, showing it predicts lots of observations;
+%        but mostly only if fitness is exp(-|x|^k) for k > 2
+% chevin2014niche uses SSWM; looks at incompats after adaptation in Fisher's geom model and how it depends on dimension;
+%        uses generalized Fisher's model with neutral directions; terms hybrid unfitness due to segregation variance "transgressive incompats";
+%        finds rate of RI accumulation decreases with Ne (like Ne^(-m/2), with m the trait dimension?)
 % simon2017coadapted follows up on the Fisher's geom model predictions, also comparing to some real data
+% martin2014fishers shows that you can get Fisher's model quite generally if (a) the kryptotype dimension is large compared to the phenotype dimension,
+%   and (b) mutational effects fall under the CLT
+% rosas2010cryptic looks at segregation variance resulting in deviations from optimum phenotype including off of curvey ridges
+%%%% General notes
+% citation for segregation variance around a stationary optimum -> load is Poon & Otto 2000; rosas2010cryptic
 

### 2018-03-07 kalman_walk_paper.tex
reordered statment of main purpose

@@
 %Movement in this direction is ongoing, as 
-There is a substantial amount of work studying the evolution of gene regulatory networks, in both abstract \citep{wagner1994evolution, wagner1996does, siegal2002waddington, bergman2003evolutionary, draghi2015robustness}
-and empirically inspired computational
-\citep{mjolsness1991connectionist, jaeger2004dynamic, vitaly1, crombach2016gap, wotton2015quantitative, chertkova2017insilico} frameworks.
+There is a substantial amount of work studying the evolution of gene regulatory networks, in frameworks 
+both abstract \citep{wagner1994evolution, wagner1996does, siegal2002waddington, bergman2003evolutionary, draghi2015robustness} 
+and empirically inspired
+\citep{mjolsness1991connectionist, jaeger2004dynamic, vitaly1, crombach2016gap, wotton2015quantitative, chertkova2017insilico}.
 % If we allow the reasonable assumption that the genotype-phenotype map 
@@
 or ways to reverse engineer the dynamics of a ``black box'',
-even if all possible input and output experiments on the ``black box'' are performed 
+even if all possible input and output experiments are performed 
 \citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. 
 The inherent nonidentifiability of chemical reaction networks is sometimes referred to as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}. 
-In computer science, this is framed as the relationship among processes that simulate one another \citep{van2004equivalence}.
+In computer science, this is framed as the relationship among processes that \emph{simulate} one another \citep{van2004equivalence}.
 Finally,
@@
 
-In this paper we outline a theoretical framework to study the evolution of biological systems, such as gene regulatory networks.
-We study the evolution of an optimally adapted population under stable conditions, that is where phenotype is under stabilizing selection.
-Even if the phenotype remains stable over evolutionary time, the underlying mechanism might not remain so,
-as many distinct (and mutationally connected) molecular pathways can realize identical phenotypes.
-Below, we first apply results from system theory which give
+The main purpose of this paper is to explore this general idea
+in the concrete framework of linear systems theory,
+thought of as modeling gene regulatory networks.
+First, we apply results from system theory which give
 an analytical description of the set of 
 all linear gene network architectures that yield identical phenotypes.
+This suggests that, quite generally, there are many directions in which a gene network can drift
+without changing the phenotype.
 Since these phenotypically equivalent gene networks are not necessarily compatible with one another, 
-system drift may result in reproductive incompatibility between sufficiently isolated populations,
-%isolated for a sufficiently long period of time, 
-even in the absence of any sort of adaptive, selective, or environmental change. 
-We then use quantitative genetic theory to estimate how quickly 
+system drift may result in reproductive incompatibility between isolated populations,
+even in the absence of any sort of adaptive, or environmental change. 
+Does this neutral system drift matter, i.e., lead to measurable consequences?
+To address this,
+we use quantitative genetic theory to estimate how quickly 
 reproductive incompatibility due to system drift manifests.

### 2018-03-07 kalman_walk_paper.tex
intro

@@
 can allow long-term stable coexistence of incompatible genotypes encoding identical phenotypes \citep{phillips1996maintenance}.
-Despite the findings of these quantitative genetics models, previous simulations of system drift in regulatory sequences \citep{tulchinsky2014hybrid}
+However, previous simulations of system drift in regulatory sequences \citep{tulchinsky2014hybrid}
 and a regulatory cascade \citep{porter2002speciation}
-found that speciation could be rapid under directional selection
+found rapid speciation under directional selection
 but only equivocal support for speciation under models of purely neutral drift.
 %Additionally, \citet{fierst} observed speciation due to an increase of segregation variance under balancing selection, in models of weak to moderate, but not strong epistasis.
-It is known that hybrid incompatibility can result from the increased segregation variance of a quantitative trait \citep{slatkin1994segregation, chevin2014niche}.
-Here, by focusing on the explicit biological function or physiology of a phenotype, we show that the evolution of even extremely simple phenotypes can appropriately be approximated by quantitative genetics.
-This is because, like the genetic code, the genotype-phenotype ``code'' is fundamentally degenerate -- as we make explicit -- and thus ammendable to quantitative genetic analysis.
-
-Our model differs from previous work in that we use linear systems theory
-to both explore a much richer class of regulatory network models,
-and provide analytical expectations in large populations with complex phenotypes
+The rate at which hybrid incompatibility accumulates due to genetic drift creating
+segregation variance between isolated populations is fairly well understood
+\citep{slatkin1994segregation,rosas2010cryptic,chevin2014niche},
+but model assumptions can strongly affect predictions,
+including whether variation is due to rare or common alleles \citep{slatkin1994segregation},
+and the shape of the fitness landscape \citep{fraisse2016genetics}.
+Our main aim is to provide a concrete framework 
+that can provide natural predictions of these model parameters
+across a general class of models.
+Furthermore, tools from system theory allow analytical predictions to be made
+for large populations with complex phenotypes
 that would be inaccessible to population simulations.
@@
 
-Using gene regulatory network dynamics as the basis for our models, we begin with a collection of $n$ coregulating molecules
-%We first describe the class of models we study, motivated by gene regulatory networks. 
-%The behavior (i.e., the temporal dynamics) of such a system is determined by a collection of $n$ coregulating molecules 
+We use a model of gene regulatory networks
+that describes the temporal dynamics of
+a collection of $n$ coregulating molecules
 -- such as transcription factors -- as well as external or environmental inputs.

### 2018-03-07 kalman_walk_paper.tex
related geom model work

@@
 
+%% REVIEW OF RELATED RESULTS
+This fitness landscape is locally Gaussian with a rank-deficient covariance matrix.
+Since we allow for substantial genetic variation within populations,
+this model falls in the same class as \citet{lande1981models} and \citet{lande1983measurement},
+which did not consider reproductive incompatibility.
+A great deal of analytical work on the question of speciation
+has been done under the assumption of a monomorphic population
+whose trait is shifted by sequential fixation of alleles,
+i.e., Fisher's ``geometric model'' \citep{fisher1930genetical,poon2000compensating}.
+\citet{martin2014fishers} derived this model from a few general assumptions,
+and used random matrix theory to calculate the distribution of fitness effects.
+In particular, 
+\citet{chevin2014niche} studied a general version with neutral directions,
+and found that the rate of accumulation of reproductive isolation 
+decreases with $N_e$ with a form that depends on trait dimension.
+\citet{fraisse2016genetics} showed that most recognized empirical patterns in the speciation literature
+could be explained by this model 
+(although best if fitness took the form $\exp(-d^k)$
+for some $k > 2$, where $d$ is distance to the optimum),
+and \citet{simon2017coadapted} further compared predictions from the model to empirical data.
+In the context of a stationary optimum, 
+the primary contribution to hybrid unfitness is segregation variance,
+i.e., greater phenotypic variance in $F_2$ hybrids than in parentals
+due to drift having changed the genetic basis of the trait separately in the parental populations
+(\citet{chevin2014niche} refers to these as ``transgressive incompatibilities'').
+This turns out to be the main source of incompatibilities in the model we study as well,
+even though in principle curvature of the optimal set also contributes
+(as also seen in \citet{rosas2010cryptic}).
+Our analysis relies on local approximations;
+on longer time scales the appropriate model may share properties with
+the ``holey'' landscapes of \citet{gavrilets2004fitness}.
+
 \paragraph{System drift}

### 2018-03-07 kalman_walk_paper.tex
added to disco

@@
 
+As our goal here is to sketch out the rough implications of our main results,
+keeping the assumptions clearly visible,
+we provide often heuristic arguments,
+rather than presenting calculations in full multivariate generality.
+
 \paragraph{System drift}
@@
 The distribution of $F_2$ hybrids will have mean at the average of the two populations,
-but will have higher variance.
-The variance of $F_2$ hybrids can be shown to increase linearly with the square of the distance between
+but will have higher variance \citep{barton2001role,rosas2010cryptic,chevin2014niche}.
+This ``segregation variance'' of $F_2$ hybrids 
+can be shown to increase linearly with the square of the distance between
 parental population means
 under models of both simple and polygenic traits.
-This is suggested by figure \ref{fig:conceptual_fig} and shown in Appendix \ref{apx:seg_cov}.
+This is suggested by figure \ref{fig:conceptual_fig} and shown by \citet{slatkin1994segregation}
+(also see Appendix \ref{apx:seg_cov} for a different derivation).
 Concretely, we expect the population of $F_1$s to have variance $\sigma^2_S$ in the selected direction
@@
 which would compensate for this difference, 
-perhaps even predicting the rate of decrease of hybrid fitness to be \emph{independent} of population size
+perhaps even predicting the rate of decrease of hybrid fitness to be independent of population size
 for small populations.
@@
 
-In this paper, we use tools from quantitative genetics and control theory
+In this paper, we use tools from linear system theory and quantitative genetics
 to study the evolution of a mechanistic model of the genotype-phenotype map, 
@@
 may be nearly morphologically indistinguishable.
+
+Substantial analytical work has made similar observations
+using Fisher's geometric model, 
+including Haldane's rule and the $F_2$ ``speciation clock'' 
+\citep{barton2001role,fraisse2016genetics,simon2017coadapted}.
+These calculations are similar in spirit, 
+but differ from our own in that they primarily assume a monomorphic population.
+Indeed, our model may provide an additional mechanistic context 
+in which Fisher's geometric model is appropriate,
+although the typical degree of pleiotropy
+and stability of the $G$-matrix are unknown in practice.
+Our observations argue for the importance of including neutral directions in these models,
+which is not usually done (but see \citet{chevin2014niche}).
 
@@
 
+Neutral processes are certainly not the only drivers of speciation.
 All of these forces -- adaptive shifts, conflict, and network drift --

### 2018-03-09 kalman_walk_paper.tex
added some details

@@
 but the population of $F_2$ hybrids will have variance of order $\sigma^2_S + 4 \omega \sigma^2_N T/N_e$,
-where $\omega$ is a factor that depends on the genetic basis of the coefficients.
+where $\omega$ is a factor that depends on the genetic basis of the coefficients
+\citep[the ``segregation variance''][]{slatkin1994segregation}.
 If the optimal set $\allS$ has dimension $q$,
@@
 
-Substantial analytical work has made similar observations
-using Fisher's geometric model, 
-including Haldane's rule and the $F_2$ ``speciation clock'' 
+%% Fisher's geometric model
+% Differences:
+% - they look at shifting environments too
+% - they just model substitutions, so only hybrids have genetic variance
+% chevin2014niche:
+% - m is accessible, constrained dimensionlity (allows for a ridge)
+% - predicts rate of RI accumulation (due to substitutions) is \propto N^{-(1+m/2)}
+%   since this is mean log fitnees; says hybrids will have fitness
+%     exp( - m C T N^{-(1+m/2)} ) ~ 1 - m C T N^(-(1+m/2))
+% - says RI accum is *faster* at higher m but this is not obv frm eqns
+% - we say it should be ( exp(-C (T/N)^2) / sqrt(T/N) )^m :
+%     ~ 1 - m C (T/N)
+
+\paragraph{Fisher's geometric model}
+Substantial analytical work on speciation has been done using Fisher's geometric model, 
+which can also predict Haldane's rule and the $F_2$ ``speciation clock'' 
 \citep{barton2001role,fraisse2016genetics,simon2017coadapted}.
-These calculations are similar in spirit, 
-but differ from our own in that they primarily assume a monomorphic population.
+The model is similar to the quantitative genetics model we use 
+to predict the rate of speciation, 
+although work on Fisher's geometric model generally only models substitutions
+between populations, i.e., neglecting within-species polymorphism.
 Indeed, our model may provide an additional mechanistic context 
@@
 which is not usually done (but see \citet{chevin2014niche}).
+
+Although our main focus is on the implications of system theory,
+rather than on the quantitative genetics modeling,
+it is interesting to compare analytic results.
+\citet{chevin2014niche} finds that mean fitness decrease of hybrids
+after $T$ generations grows proportionally to $m N_e^{-1-m/2} T$, for large $N_e$;
+while our corresponding rate is $m T / N_e$.
+This difference in time scaling is substantial, especially for large populations,
+and seems to appear because \citet{chevin2014niche} uses the results of \citet{sella2005application}
+to describe how the steady-state fitness and substitution process
+depends on trait dimensionality
+(the fitness peak is harder to find in higher dimensions).
+We do not incorporate this effect
+because genetic variation may be provided by other sources,
+but if we did, it would enter through our $\sigma_S$.
 
@@
 % might lead to genetic load due only to recombination, much as in \citep{phillips1996maintenance}.
-
-\paragraph{Fisher's geometric model}
-There has been substantial work on models of quadratic stabilizing selection
-around a single optimum -- i.e., Fisher's geometric model \citep[e.g.][]{simon2017coadapted}.
-Our model would have this form if viewed only in phenotype space, 
-where there is a single optimum --
-but differs by explicitly considering evolutionary dynamics of underlying genotype-derived regulatory matrices.
-These make the quantitative genetic model we are led to
-a generalization of Fisher's geometric model,
-with distinct behavior,
-due to the existence of neutral directions in genotype space.
-However, there are many shared aspects:
-for instance, \citet{barton1989divergence} also found incompatibilities accumulating
-with genetic drift.
-Fisher's geometric model also predicts Haldane's rule
-and many other empirically observed patterns \citep{fraisse2016genetics},
-but for different underlying reasons.
 
@@
 \end{align*}
+This is the product of two terms:
+the first gives the drop in fitness due to segregation variance,
+and the second the drop due to a shift in mean away from the optimum.
+A similar decomposition appears in \citet{barton2001role} and \citet{chevin2014niche},
+but for mean log fitness.
 
@@
 will not stay constant.
-To get an idea of how this might change,
+This has been quantified in various models before
+\citep[e.g.,][]{slatkin1994segregation,barton2001role,chevin2014niche};
+we provide the following calculations for completeness.
+
+To get an idea of how segregation variance might evolve,
 suppose that a trait is determined by $L$ unlinked, biallelic loci,

### 2018-03-09 kalman_walk_paper.tex
a few more tweaks

@@
 % wright1935evolution looks at allele freq dynamics under stabilizing selection on a polygenic quantitative trait and discusses possibility for speciation this way
-% barton1989divergence not available but likely following on from barton1986maintenance analyzing multiple optima in a similar model
+% barton1989divergence following on from barton1986maintenance analyzing multiple optima in a similar model, looking at RI due to shifts between equilibria
 % nei1983models simulates from simple stepwise mut model on one or two loci (without interactions) where incompat betw diff alleles increases with number of steps betw them
@@
 traits under stabilizing selection determined additively by alleles at many loci 
-\citep{wright1935evolution,barton1986maintenance,barton2001role},
+\citep{wright1935evolution,barton1986maintenance,barton1989divergence,barton2001role},
 in related fitness landscape models \citep{fraisse2016genetics},
@@
 the underlying mechanism does not depend on the \emph{dominance theory} \citet{turelli1995dominance} to explain Haldane's rule:
-instead, it derives from the nature of segregation load.
+instead, it derives from the nature of segregation variance.
 % In this setting, Haldane's rule is mechanistically similar to \emph{hybrid breakdown}, 
@@
 which did not consider reproductive incompatibility.
-A great deal of analytical work on the question of speciation
-has been done under the assumption of a monomorphic population
+Substantial work on speciation has been done under the assumption of a monomorphic population
 whose trait is shifted by sequential fixation of alleles,
 i.e., Fisher's ``geometric model'' \citep{fisher1930genetical,poon2000compensating}.
+This work has included both stationary optima (like we study)
+and moving optima (i.e., adaptation) \citep[e.g.,][]{barton2001role,chevin2014niche}.
 \citet{martin2014fishers} derived this model from a few general assumptions,
 and used random matrix theory to calculate the distribution of fitness effects.
-In particular, 
 \citet{chevin2014niche} studied a general version with neutral directions,
@@
 keeping the assumptions clearly visible,
-we provide often heuristic arguments,
+we provide relatively rough arguments,
 rather than presenting calculations in full multivariate generality.
@@
 The distribution of $F_2$ hybrids will have mean at the average of the two populations,
-but will have higher variance \citep{barton2001role,rosas2010cryptic,chevin2014niche}.
+but will have higher variance \citep{wright1968evolution,barton2001role,rosas2010cryptic}.
 This ``segregation variance'' of $F_2$ hybrids 
@@
 (the same as within each parental population),
-but the population of $F_2$ hybrids will have variance of order $\sigma^2_S + 4 \omega \sigma^2_N T/N_e$,
-where $\omega$ is a factor that depends on the genetic basis of the coefficients
-\citep[the ``segregation variance''][]{slatkin1994segregation}.
+but the population of $F_2$ hybrids will have variance $\sigma^2_S + 4 \omega \sigma^2_N T/N_e$,
+where $\omega$ is a factor that depends on the genetic basis of the coefficients.
 If the optimal set $\allS$ has dimension $q$,
@@
 Substantial analytical work on speciation has been done using Fisher's geometric model, 
-which can also predict Haldane's rule and the $F_2$ ``speciation clock'' 
+which can also predict Haldane's rule and the regular increase of incompatibility with genetic distance
 \citep{barton2001role,fraisse2016genetics,simon2017coadapted}.
@@
 although work on Fisher's geometric model generally only models substitutions
-between populations, i.e., neglecting within-species polymorphism.
+between populations, neglecting within-species polymorphism.
 Indeed, our model may provide an additional mechanistic context 

### 2018-08-16 review-responses.tex
review reponses added

%%%%%%
%%
%%  Don't reorder the reviewer points; that'll mess up the automatic referencing!
%%
%%%%%

\begin{minipage}[b]{2.5in}
  Resubmission Cover Letter \\
  {\it Evolution}
\end{minipage}
\hfill
\begin{minipage}[b]{2.5in}
    Joshua Schiffman \\
    \emph{and} Peter Ralph \\
  \today
\end{minipage}

\vskip 2em

\noindent
{\bf To the Editor(s) -- }

\vskip 1em

We are pleased to submit a revision of our manuscript,

\noindent \hspace{4em}
\begin{minipage}{3in}
\noindent
{\bf Sincerely,}

\vskip 2em

{\bf 
Joshua Schiffman and
Peter Ralph
}\\
\end{minipage}

\vskip 4em

\pagebreak

%%%%%%%%%%%%%%
\reviewersection{AE}

\begin{quote}
I have received the evaluations of two reviewers, and I have read the paper
myself. First, sorry for the long time it took to review your manuscript; it
was reviewed on the timescale of mathematicians (for Reviewer2), but both
reviewers provided thorough evaluations of your work (which in my opinion is
better than a quick but superficial review).

Your manuscript is quite long already, and the reviewers' suggestions of
modifications and clarification, which need to be implemented, will make the
manuscript even longer. I however share the reviewers' opinion that the
manuscript's two parts are only loosely related. My suggestion is therefore to
publish them separately. Although connecting the systems biology and the
quantitative genetics parts is indeed an exciting endeavor, this is not really
achieved in the current version of the manuscript, and it would be more
profitable to first better describe each part separately. Regarding the
connexion between the two parts, please also pay particular attention to R2's
first specific comment about dimensionality.

This may be because I am more familiar with Fisher's geometric model than
systems biology models, but it seems to me that the first part is more novel,
and should therefore deserve your attention first, should you follow my
suggestion of publishing the two parts separately.
\end{quote}

\begin{point}{Format:}
    please add line numbers to your manuscript; it is straightforward
with the lineno package in LaTeX.  Please also do not increase text width too
much, as this decreases legibility (LaTeX's default settings already optimize
the number of characters per line).  Finally, be careful about not confusing
citep and citet (a lot of missing parentheses around citations in the second part).
\end{point}

\reply{
}

\begin{point}{all figures:}
    please ensure that all axes legends and labels are big enough
and not distorted (same of arrow labels on the diagrams). Please also make sure
that the lines are thick enough to be visible.
\end{point}

\reply{
}

\begin{point}{Above eq (2):}
    ``Under quite general assumptions'': quickly mention them?
\end{point}

\reply{
}

\begin{point}{eq (2):} 
    Do you need the first term on the rhs given that you assume that $\kappa(0)=0$?
\end{point}

\reply{
}

\begin{point}{Example 1:}
    ``and so its phenotype'' whose phenotype? (the subject of the sentence is ``the input'')
\end{point}

\reply{
}

\begin{point}{p5, Note on implementation}
    ``the closure of span(B)'' may not be
    understandable by most of Evolution's readers; please explain what this means.
\end{point}

\reply{
}

\begin{point}{Figures 1-3:}
    In figures 1 and 3, there are two types of arrow heads (+ and -
effects), but not in Figure 2, probably because the sign of the effect depends
on the value of $\tau$. Maybe use a  different type of arrow heads for
interactions that may change signs, to avoid the confusion with positive
effects?
\end{point}

\reply{
}

\begin{point}{p7, Hybrid incompatibility:}
    Please provide a rationale for choosing a particular weight $\rho$.
\end{point}

\reply{
}

\begin{point}{p11, 2nd paragraph}
    Try to better relate these different studies to yours
(e.g., what you add, what they do and you do not). In addition to R1's
suggestions, I add a few more below.
\end{point}

\reply{
}

\begin{point}{p12 Hybridization:}
    What is $\mathcal{X}$?
\end{point}

\reply{
}

\begin{point}{p15}
    ``The importance of including neutral directions in these models, which
is not usually done'' -- Do not line of isofitness correspond to neutral
directions, and aren't these neutral directions already included in those
models?
\end{point}

\reply{
}

\begin{point}{Additional references:}
Weinreich, D. M. and Knies, J. L. (2013), FISHER'S GEOMETRIC MODEL OF ADAPTATION MEETS THE FUNCTIONAL SYNTHESIS: DATA ON PAIRWISE EPISTASIS FOR FITNESS YIELDS INSIGHTS INTO THE SHAPE AND SIZE OF PHENOTYPE SPACE. Evolution, 67: 2957-2972. doi:10.1111/evo.12156

François Blanquart and Thomas Bataillon
Epistasis and the Structure of Fitness Landscapes: Are Experimental Fitness Landscapes Compatible with Fisher’s Geometric Model?
GENETICS Early online April 6, 2016
\end{point}

\reply{
}

%%%%%%%%%%%%%%
\reviewersection{1}

\begin{quote}
The ms combines two parts that are only loosely linked: one ``systems biology''
part and a ``popgen'' part. The combination of both parts is appealing, primarily
because we lack such combined approaches and it is not easy to come up with an
adequate framework. Below, I will discuss both parts in turn.
\end{quote}

\begin{quote}
General remark: line numbers are very helpful for reviewing.  
\end{quote}

\begin{quote}
    \textbf{(I)} The first (systems biology) part makes use of the concept of
``nonidentifiability'' of chemical reaction networks and develops these concepts
in the context of a genotype to phenotype map. In particular, this map is
analyzed for the case where it can be modeled as a linear system.
\end{quote}

\begin{point}{}
    First note that the concept of ``nonidentifiability'' is closely related to
themes that have been discussed in evolutionary biology for a long time, under
names like ``mutational robustness'', ``network neutrality'', ``canalization'',
``redundancy'' etc. In particular, if you characterize your model by ``many
distinct (and mutationally connected) molecular pathways can realize identical
phenotypes'', note that this essentially describes what has been called a
``neutral network'' and studied under this name in many articles. Also concrete
developmental networks have been studied in this context (eg, von Dassow etal
2000, Nature 406:188–192). Discussion of this literature - and more importantly
discussion of terms and notions used here relative to terms that have been used
elsewhere - is largely missing. I won't be able to summarize all this here, but
an older review is the Evolution Perspective piece ``Evolution and detection of
genetic robustness'' (de Visser et al 2003). You can work work backward and
forward from there.  
\end{point}

\reply{
}

\begin{point}{}
    Related to this: I find it non-intuitive to start the introduction with the
notion of ``nonidentifiability'', which is a consequence rather than a cause, the
cause being what you later call ``phenotypic equivalence'' and what had been
discussed under other names in other articles.
\end{point}

\reply{
}

\begin{point}{}
    Still related: You give long lists of references in the introduction, but do
not provide the reader with any information about specific contributions. A bit
more would be helpful (starting with True and Haag coined the term
``developmental system drift'', etc).
\end{point}

\reply{
}

\begin{point}{4.}
    Your ms does not simply assume that there is a large ``neutral network''
underlying a phenotype, but suggests a mechanistic underpinning to create this
neutral space. Since much of the appeal of the first part of the paper is
connected to this fact, I'd like to see some more in-depth discussion.
\textit{(suggestions to follow)}
\end{point}

\reply{
}

\begin{point}{(mechanistic models)}
    you should probably mention somewhere that mechanistic models of neutral
  spaces exist in the (quite different) context of RNA folding (papers by
Hofacker, Fontana, etal)
\end{point}

\reply{
}

\begin{point}{(mechanistic models)}
    For development, Andreas Wagner once suggested a discrete version of a linear
  model (iterated matrix multiplication). Variants of this have later been used
by others (eg Draghi and GP Wagner, I think). How is your model related to
these approaches?
\end{point}

\reply{
}

\begin{point}{(mechanistic models)}
    You model development as a linear system and this assumption is essential
   for all further steps (the explicit solution of the dynamics and the Kalman
decomposition). For me, the only justification of a linear model (other than
mathematical convenience) is local approximation. Do you agree? This will be
important for the second part of the ms, see below.
\end{point}

\reply{
}

\begin{point}{(mechanistic models)}
    You gain a lot of ``neutrality'' by the assumption that the dimension of the
  ``krypotype'' is larger than the dimension of the trait. If the trait is the
expression of a gene at the top of some pathway, this is certainly true.
However, if we include further traits that are affected by the same genes
(i.e., pleiotropy) this is no longer clear. The space of ``all traits under
selection'' of an organism is awfully high-dimensional and pleiotropy is
wide-spread (even if we do not believe in ``omnigenic'' models). It is not clear
to me from what kind of data we could learn more about these dimensions and I
do not expect an answer in your ms, but the issue deserves more discussion.
Currently, ``pleiotropy'' is not even mentioned in the ms.   
\end{point}

\reply{
}

\begin{point}{Introduction:}
    ``Genotypes encoding identical phenotypes can even persist stably within a species'' - if there is population structure, I suppose.
\end{point}

\reply{
}

\begin{point}{Introduction:}
    ``It is not a new observations that there is often more than one way to do the same thing, and that this may lead to speciation'' - Isn't this what people would call a neutral Dobzhansky-Muller incompatibility?
\end{point}

\reply{
}

\begin{point}{Results:}
    (Eq. 1) development as a linear system: has this been done before (probably yes?) - If so, references?
\end{point}

\reply{
}

\begin{point}{Results:}
    ``Of course, neither of these are necessarily true for real systems'' - delete ``necessarily''?
\end{point}

\reply{
}

\begin{point}{Eq. 3:}
    What does a coordinate change mean biologically? In our JC, we discussed for a while whether this is just a different parametrization of the exact same biological thing (like transformation to principle components). I think I now understand that it is more, but this should be explained better.
\end{point}

\reply{
}

\begin{point}{}
    ``Since gene networks can grow or shrink following gene duplications and deletions,
these additional degrees of freedom can apply in principle to any system.'' - are there examples of equivalent gene networks with a different size due to gene duplication or loss?
\end{point}

\reply{
}

\begin{point}{Kalman decomposition:}
    Define the sub-matrices directly when first used.
\end{point}

\reply{
}

\begin{point}{Figure 2:}
    Colors are hardly visible
\end{point}

\reply{
}

\begin{quote}
    \textbf{(II)} The popgen part. This part models the ``system drift'', which is nothing else
but neutral drift of a population along a high-fitness ridge in an epistatic
landscape (conceptual figure 4), right? If two populations in alloparty drift
in different directions, this can lead to hybrid incompatibilities, which are
uncovered upon secondary contact. Fitness is modeled by the weighted distance
from the optimal impulse response function (eq 5). This is a natural
assumption.
\end{quote}

\reply{
}

\begin{point}{}
    Your analysis of the fitness loss in F1 and F2 hybrids rests on a local
Taylor expansion of the fitness landscape. This is adequate given that the
underlying linear network is also only locally valid (see above). However, you
then apply this to a discussion of hybrid incompatibility (Haldane's rule etc).
We are thus interested in *very large* fitness costs for hybrids. This does not
seems to be compatible: to get hybrid incompatibility or sterility, you need to
have epsilon sufficiently large (below equation 5). But then $\epsilon^2$ is no
longer smaller than epsilon. In other words: it seems to me that for the
discussion of hybrid incompatibilities you apply your model of the first part
beyond the local range where it is valid.
\end{point}

\reply{
}

\begin{point}{}
    For system drift, you assume that ``Selection will tend to restrain this
motion, but movement along the optimal set N is unconstrained, and so we expect
the population mean to drift along the optimal set like a particle diffusing.''
I see two major problems with this view. Both lead to slower divergence and
therefore run against your conclusions.
    \textit{(further points to follow)}
\end{point}

\reply{
}

\begin{point}{(drift along a ridge)}
    In the presence of epistasis, evolution on a neutral network (a high-fitness
  ridge) is *not* due to drift alone, but also affected by (weak second order)
selection in favor of mutational robustness / genetic canalization. In contrast
to what you write, diffusion on the set of network coefficients corresponding
to the optimal phenotype is not unbiased - even if the optimal phenotypes do
indeed all have the same fitness. Instead, selection drives the population to
``thicker'' parts of the network where the mean fitness of the population
(including a cloud of mutants) is higher that on a narrow ridge. This is the
basis of the evolution of robustness/canalization. It is possible to account
for this effect, see Hermisson et al 2003. Amer. Nat. 161, 708-734;
Alvarez-Castro et al, TPB 75 (2009) 109-122, or also Rice, 1998, Evolution 52,
647-656; van Nimwegen, etal 1999, PNAS 96:9716–9720. Note that epistasis is
necessary for the neutral evolution of incompatibilities, which is what you are
aiming for.
\end{point}

\reply{
}

\begin{point}{(drift along a ridge)}
    a second problem results from the fact that evolution on a high-fitness ridge
  often requires coordinated changes at many loci. Take the oscillator system
that you use as an example: simultaneous changes at two genes are required to
maintain the phenotype. This leads to a phenomenon called ``adaptive inertia''
(see Baatz and Wagner, TPB 51, 49-66 and Alvarez-Castro etal . TPB 75, 109-122),
which effectively slows down the movement along the ridge considerably. This
problem applies, in particular, in a ``house-of-cards'' mutation regime when
where rarely two mutations occur together on the same haplotype. In small
populations, it typically requires that populations drift through shallow
fitness valleys. While this is possible, it slows down the process.  
\end{point}

\reply{
}

\begin{point}{(drift along a ridge)}
    The relevance of both effects could be studied by simulations in a simple
example (eg the oscillator that is used as an illustration in the ms anyway).
\end{point}

\reply{
}

\begin{point}{}
    If speciation due to accumulation of incompatibilities does not occur in
allopatry, but under (even weak) gene flow, some degree of positive selection
is always needed (Bank et al, Genetics 191, 845–863 for a proof). This also
means that even weak gene flow will counteract the process described in the ms.
\end{point}

\reply{
}

\begin{point}{}
    Population isolates and genetic load: Isn't this exactly ``founder effect
speciation''?
\end{point}

\reply{
}

%%%%%%%%%%%%%%
\reviewersection{2}

\begin{quote}
The present paper is divided into two parts. In the first part, the authors propose a framework
in which distinct genetic architectures can produce the same phenotype. (In the fitness landscape
terminology, this would correspond to the existence of an evolutionary ridge in a fitness landscape.
)The starting point is a nice analogy with linear systems theory. The authors highlight the fact
that in general, two distinct linear (differential) systems can respond identically for any input; i.e.,
different systems (genotypes) can always produce the same output (phenotype) given the same input
(environment). Further, the set of equivalent systems can be nicely characterized through Kalman
decomposition, thus providing a nice characterization of level sets in the underlying fitness landscape.
All along the first part of the paper, this analogy is well illustrated using a simple (yet quite nice)
example of oscillating gene transcription. I especially like Fig. 5 where it is shown that even if $F_1$
hybrids have a phenotypic response close to their parents, $F_2$ hybrids can behave in a drastically
different manner.

In the second part of the paper, motivated by the previous results, the authors explore a general
quantitative genetics model in which populations can drift stochastically near a set of equivalent
and optimal systems. Since the optimal set (or evolutionary ridge) is not closed under averaging or
recombination, two isolated populations can drift apart and accumulate enough genetic differences
so that they do not produce any viable offspring. Using some heuristics, several expressions are
derived to quantify the accumulation of genetic incompatibilities.

Overall, I think the paper is well written. The framework developed in the first part of the paper
is very interesting and that the analogy with system theory is quite enlightening. I am somehow
less convinced by the second part. First, I find the arguments a bit sketchy (see below) and not so
easy to follow. Secondly, it is not entirely clear to me what is the main contribution of this part
compared to previous works. It seems to me that the main result is somehow contains in the fact
that the variance (or ``segregation variance'') of an $F_2$ population is given by
$$\sigma^2_S + 4 \omega \sigma^2_N T / N_e$$
which was already derived Slatkin and Lande according to the authors (except for the explicit
expression of $\omega$, but again I am a little bit confused by the arguments derived in the appendix).

In summary, I think this paper could be a nice contribution to Evolution. However, I am also
convinced that the second part of the paper would require more work, or more explicit reference
to previous works (equation, section etc.) if the authors do not want to re-derive already existing
formula.
\end{quote}

\begin{point}{end of p6:} 
    ``we assume ... $((A + A' )/2, (B + B' )/2, (C + C' )/2)$'' . It is claimed in p4 (right
before the last paragraph) that two kryptotypes need not have the same dimension. In this
case, the previous sum does not make sense, right?
\end{point}

\reply{
}

\begin{point}{p7: fig 4.}
    ``The distribution of $F_2$ ... homozygotes''. I do not really grasp the meaning of this
sentence. More importantly, I find the purple cloud of dots ($F_2$ population) quite confusing.
It seems to me that the purple populations should be a cloud of points concentrated around
a point sitting on the red dotted line, i.e., a Gaussian distribution with the mean at the
average of the two parental populations. According to the authors, the distribution is bi-
modal with a peak below the optimal set (as displayed on the figure) and another distinct
peak sitting on the other side of the optimal. In fact, if I understand the computations of
the second part correctly, the red and purple means should coincide, but the red variance
should simply be greater.
\end{point}

\reply{
}

\begin{point}{p7: end of the page.}
    Why do you need $\rho$ to be square integrable?
\end{point}

\reply{
}

\begin{point}{p8:}
    ``A Taylor expansion of $D(h_\epsilon)$ .. ''. It is not clear to me at all. Could you provide some
extra explanation (e.g., in the appendix)?
\end{point}

\reply{
}

\begin{point}{p9. Fig 5.} 
    In the left panel, it seems to me that the two parents are homozygotes since there
is a single $F_1$ possible offspring (dashed blue curve). It would be worth being more explicit.
For the right panels, could you be more explicit on the number of curves. If I understand
correctly, there are 16 possibilities due to recombination (2 per entries of the matrix) since
$F_1$ individuals are heterozygotes. Is that correct?
Finally, the labels on the y axis are not easy to read.
\end{point}

\reply{
}

\begin{point}{p11. Paragraph system drift.}
    ``move a random distance $\sigma$''. What is $\sigma$? I think it should
be $\sigma_N$ (the std deviation in the direction of the evolutionary ridge) to be consistent with
the assumption that the population drifts along the optimal set. I believe that this is what
is assumed thereafter. Also, the sentence ``It therefore seems .... as cloud of points of width
$\sigma$'' is not very accurate, since the covariance matrix is not the identity.
\end{point}

\reply{
}

\begin{point}{p 11.}
    Approximating the optimal set $\mathcal{N}$ by a quadratic surface should only be accurate if
we look at the genetic divergence at small time scales. This should be at least mentioned.
\end{point}

\reply{
}

\begin{point}{end of p11.}
    $1/( \frac{d}{du} D(x + uz))$ should be evaluated at $u = 0$.
\end{point}

\reply{
}

\begin{point}{p12. Third paragraph}
    $\sqrt{4 \omega T/N_e} \sim \gamma/\sigma_N$.
    I guess the underlying assumption here is that
    $\sigma_S \ll \sqrt{4 \omega \sigma^2_N T/N_e}$?
\end{point}

\reply{
}

\begin{point}{p12. before eq. 7.}
    $\mu = c_\mu \gamma T /N_e$. Why is $\mu$ proportional to $\gamma$?
\end{point}

\reply{
}

\begin{point}{Fig 7.}
    I have one important issue with this figure (and the assumptions of the underly-
ing quantitative genetic model). If one wants to be consistent with the assumption that
parental populations drift along the evolutionary ridge, I think one would need to assume
that selection is strong enough to constraint the mean of the population on the surface. This
would presumably require that $\sigma S /\gamma \ll 1$. First, I think this assumption (or something
alternative to that) should be made explicit in the text. Secondly, it seems to me that this
assumption is not satisfied for panel A and C: under the range of parameters proposed by
the authors, the parental populations could easily drift away from the optimal set, and in
particular, the heuristics derived in the main text would not be satisfied.
\end{point}

\reply{
}

\begin{point}{}
    Finally, it would be worth mentioning several old and recent works relating genetic drift
to speciation : Yamagushi and Iwasa, several articles by Gavrilets et al. (I think several
citations are missing here), and Mirò Pina and Schertzer.
\end{point}

\reply{
}

### 2018-08-16 kalman_walk_paper.tex
review reponses added
first pass at splitting the papers

@@
 \usepackage{authblk}
-\usepackage{lineno}
 \usepackage{multicol}
@@
     \newcommand{\plri}[1]{\plr{#1}}
+\fi
+
+\newif\ifsubmission
+\submissiontrue
+% \submissionfalse
+\ifsubmission
+    % % stuff for submission
+    \usepackage{lineno}
+    % allow line numbers around math environments
+    % from https://tex.stackexchange.com/questions/43648/why-doesnt-lineno-number-a-paragraph-when-it-is-followed-by-an-align-equation
+        \newcommand*\patchAmsMathEnvironmentForLineno[1]{%
+          \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
+          \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
+          \renewenvironment{#1}%
+             {\linenomath\csname old#1\endcsname}%
+             {\csname oldend#1\endcsname\endlinenomath}}% 
+        \newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
+          \patchAmsMathEnvironmentForLineno{#1}%
+          \patchAmsMathEnvironmentForLineno{#1*}}%
+        \AtBeginDocument{%
+        \patchBothAmsMathEnvironmentsForLineno{equation}%
+        \patchBothAmsMathEnvironmentsForLineno{align}%
+        \patchBothAmsMathEnvironmentsForLineno{flalign}%
+        \patchBothAmsMathEnvironmentsForLineno{alignat}%
+        \patchBothAmsMathEnvironmentsForLineno{gather}%
+        \patchBothAmsMathEnvironmentsForLineno{multline}%
+        }
+\else
+    % for the nice version
 \fi
@@
 
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}{Lemma}
-\newtheorem{definition}{Definition}
 \newtheorem{example}{Example}
+
+% Responses to reviews
+\input{review-response-commands}
+% set this to show line numbers and include responses to reviews or not
+\newif\ifreviewresponses
+\reviewresponsestrue  % include them
+% \reviewresponsesfalse  % don't include them
+\newcommand{\responsefile}{review-responses}  % name of the review responses file
 
@@
 This evolutionary exploration, or \emph{system drift}, 
-proceeds at a rate proportional to the amount of intrapopulation genetic variation
+is expected to proceed at a rate proportional to the amount of intrapopulation genetic variation
 divided by the effective population size ($N_e$).
 At biologically reasonable parameter values
-this process can lead to substantial interpopulation incompatibility,
+this could lead to substantial interpopulation incompatibility,
 and thus speciation, in fewer than $N_e$ generations.
@@
 even in the absence of any sort of adaptive, or environmental change.
-Does this neutral system drift matter, i.e., lead to measurable consequences?
-To address this,
-we use quantitative genetic theory to estimate how quickly 
-reproductive incompatibility due to system drift manifests.
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{System drift and the accumulation of incompatibilities}
-
-Thus far we have seen that many distinct molecular mechanisms can realize identical phenotypes
-and that these mechanisms may fail to produce viable hybrids.
-Does evolution shift molecular mechanisms
-fast enough to be a significant driver of speciation?
-To approach this question,
-we explore a general quantitative genetic model in which a population drifts stochastically
-near a set of equivalent and optimal systems
-due to the action of recombination, mutation, and demographic noise.
-Although this is motivated by the results on linear systems above,
-the quantitative genetics calculations are more general,
-and only depend on the presence of genetic variation and a continuous set of phenotypically equivalent systems.
-
-We will suppose that each organism's phenotype is determined by its vector of coefficients, denoted by $x=(x_1, x_2, \ldots, x_p)$,
-and that the corresponding fitness is determined by the distance of its phenotype to optimum.
-The optimum phenotype is unique, but is realized by many distinct $x$ -- those falling in the ``optimal set'' $\allS$.
-The phenotypic distance to optimum of an organism with coefficients $x$ is denoted $D(x)$.
-In the results above, $x = (A,B,C)$ and $D(x)$ is given by equation \eqref{eqn:distance}.
-The fitness of an organism with coefficients $x$ will be $\exp(-D(x)^2)$.
-% this assumption does not entail a loss of generality since $D$ is arbitrary.
-We assume that in the region of interest, the map $D$ is smooth
-and that we can locally approximate the optimal set $\allS$ as a quadratic surface.
-As above, an individual's coefficients are given by averaging its parentally inherited coefficients and adding random noise due to segregation and possibly new mutation.
-Concretely, we use the \emph{infinitesimal model} for reproduction \citep{barton2017infinitesimal} --
-the offspring of parents at $x$ and $x'$ will have coefficients $(x+x')/2 + \varepsilon$,
-where $\varepsilon$ is a random Gaussian displacement due to random assortment of parental alleles.
-
-%% REVIEW OF RELATED RESULTS
-This fitness landscape is locally Gaussian with a rank-deficient covariance matrix.
-Since we allow for substantial genetic variation within populations,
-this model falls in the same class as \citet{lande1981models} and \citet{lande1983measurement},
-which did not consider reproductive incompatibility.
-Substantial work on speciation has been done under the assumption of a monomorphic population
-whose trait is shifted by sequential fixation of alleles,
-i.e., Fisher's ``geometric model'' \citep{fisher1930genetical,poon2000compensating}.
-This work has included both stationary optima (like we study)
-and moving optima (i.e., adaptation) \citep[e.g.,][]{barton2001role,chevin2014niche}.
-\citet{martin2014fishers} derived this model from a few general assumptions,
-and used random matrix theory to calculate the distribution of fitness effects.
-\citet{chevin2014niche} studied a general version with neutral directions,
-and found that the rate of accumulation of reproductive isolation 
-decreases with $N_e$ with a form that depends on trait dimension.
-\citet{fraisse2016genetics} showed that most recognized empirical patterns in the speciation literature
-could be explained by this model 
-(although best if fitness took the form $\exp(-d^k)$
-for some $k > 2$, where $d$ is distance to the optimum),
-and \citet{simon2017coadapted} further compared predictions from the model to empirical data.
-In the context of a stationary optimum, 
-the primary contribution to hybrid unfitness is segregation variance,
-i.e., greater phenotypic variance in $F_2$ hybrids than in parentals
-due to drift having changed the genetic basis of the trait separately in the parental populations
-(\citet{chevin2014niche} refers to these as ``transgressive incompatibilities'').
-This turns out to be the main source of incompatibilities in the model we study as well,
-even though in principle curvature of the optimal set also contributes
-(as also seen in \citet{rosas2010cryptic}).
-Our analysis relies on local approximations;
-on longer time scales the appropriate model may share properties with
-the ``holey'' landscapes of \citet{gavrilets2004fitness}.
-
-As our goal here is to sketch out the rough implications of our main results,
-keeping the assumptions clearly visible,
-we provide relatively rough arguments,
-rather than presenting calculations in full multivariate generality.
-
-\paragraph{System drift}
-% plr how fast does it drift
-We work with a randomly mating population of effective size $N_e$. 
-If the population variation has standard deviation $\sigma$ in a particular direction,
-since subsequent generations resample from this diversity,
-the population mean coefficient will move a random distance of size $\sigma/\sqrt{N_e}$ per generation,
-simply because this is the standard deviation of the mean of a random sample \citep{lande1981models}.
-% (This could be taken as a definition of $N_e$.)
-Selection will tend to restrain this motion,
-but movement along the optimal set $\allS$ is unconstrained,
-and so we expect the population mean to drift along the optimal set like a particle diffusing.
-% (although perhaps complicated by recombination load \citep{recomb_load})
-The amount of variance in particular directions in coefficient space 
-depends on constraints imposed by selection and 
-correlations between the genetic variation underlying different coefficients 
-(the $G$ matrix \citep{arnold2008understanding}).
-It therefore seems reasonable to coarsely model the time evolution of population variation in regulatory coefficients as 
-a ``cloud'' of width $\sigma$ about the population mean, 
-which moves as an unbiased Brownian motion through the set of network coefficients that give the optimal phenotype.
-
-% covariance which may arise due to functional constraints and/or statistical linkage.
-% There may well be functional constraints -- but these are not sufficiently well-known to say anything general about.
-%For instance,
-%if the variation is due to \textit{cis}-regulatory variants,
-%the genetic basis of each \emph{row} of $A$ likely lies within a few kilobases of tightly linked sequence,
-%across which a population may carry only a few common haplotypes.
-%However, covariance due to transiently assembled haplotypes is not expected to be stable over long periods of time --
-%a common \textit{cis}-regulatory haplotype of transcription factor $k$ with particularly strong binding to both $i$ and $j$
-%(leading to positive covariance between $A_{ik}$ and $A_{jk}$)
-%is no more likely to appear than one with strong binding to $i$ but particularly weak binding to $j$ (negative covariance).
-%(Such transient covariances may well increase the variance of the per-generation change in network mean, however \citep{barton_linkage}.)
-
-Next, we calculate with some simplifying assumptions to give the general idea;
-multivariate derivations appear in Appendix \ref{ss:quant_gen}.
-There will in general be different amounts of variation in different directions;
-to keep the discussion intuitive, we only discuss $\sigma_N$, the amount of variation in ``neutral'' directions
-(i.e., directions along $\allS$),
-and $\sigma_S$, the amount of variation in ``selected'' directions (perpendicular to $\allS$).
-The other relevant scale we denote by $\gamma$,
-which is the scale on which distance to phenotypic optimum changes as $x$ moves away from the optimal set, $\allS$.
-Concretely, $\gamma$ is
-%the inverse of the derivative of 
-$1/(\frac{d}{du}D(x+uz))$ 
-where $x$ is optimal and $z$ is a ``selected'' direction perpendicular to $\allS$.
-With these parameters, a typical individual will have a fitness of around $\exp(-(\sigma_S/\gamma)^2)$.
-Of course, there are in general many possible neutral and selected directions;
-we take $\gamma$ to be an appropriate average over possible directions.
-
-\paragraph{Hybridization}
-The means of two allopatric populations each of effective size $N_e$ separated for $T$ generations
-will be a distance roughly of order $2\sigma_N \sqrt{T/N_e}$ apart along $\optx$.
-(Consult figure \ref{fig:conceptual_fig} for a conceptual diagram.)
-A population of $F_1$ hybrids has one haploid genome from each,
-whose coefficients are averaged,
-and so will have mean system coefficients at the midpoint between their means.
-The distribution of $F_2$ hybrids will have mean at the average of the two populations,
-but will have higher variance \citep{wright1968evolution,barton2001role,rosas2010cryptic}.
-This ``segregation variance'' of $F_2$ hybrids 
-can be shown to increase linearly with the square of the distance between
-parental population means
-under models of both simple and polygenic traits.
-This is suggested by figure \ref{fig:conceptual_fig} and shown by \citet{slatkin1994segregation}
-(also see Appendix \ref{apx:seg_cov} for a different derivation).
-Concretely, we expect the population of $F_1$s to have variance $\sigma^2_S$ in the selected direction
-(the same as within each parental population),
-but the population of $F_2$ hybrids will have variance $\sigma^2_S + 4 \omega \sigma^2_N T/N_e$,
-where $\omega$ is a factor that depends on the genetic basis of the coefficients.
-If the optimal set $\allS$ has dimension $q$,
-using the polygenic model of appendix \ref{apx:seg_cov}, 
-$\omega$ is proportional to the number of degrees of freedom: $\omega = (p-q)/8$.
-If each trait is controlled by a single locus, as in figure \ref{fig:conceptual_fig},
-the value is similar.
-
-%%% CITE SOME OF THESE
-%% empirical data on sigma
-% The amount and structure of this standing variation is established over long time scales
-% by many factors, including mutation-selection balance, 
-% shifts in the phenotypic optimum, and/or spatial variation in the optimum \citep{hansen1996translating}.
-% Quantitative genetics models of mutation-selection balance 
-% predict precise levels and structure of standing variation \citep{kimura_mutsel,lande_mutsel,lande1981models},
-% but it is unclear how well these predictions match reality \citep{johnson_barton}
-% and how much they are expected to change over time \citep{arnold_changing_G}.
-
-What are the fitness consequences?
-A population of $F_2$ hybrids will begin to be substantially less fit than the parentals
-once they differ from the optimum by a distance of order $\gamma$,
-i.e., once $\sqrt{4 \omega T/N_e} \approx \gamma / \sigma_N$.
-This implies that hybrid incompatibility among $F_2$ hybrids should appear much slower --
-on a time scale of $N_e (\gamma / \sigma_N)^2 / (4 \omega)$ generations.
-The $F_1$s will not suffer fitness consequences until the hybrid mean is further than $\gamma$ from the optimum;
-% as shown in appendix \ref{apx:away_from_opt} 
-as suggested by figure \ref{fig:conceptual_fig}, Taylor expanding $D^2$ along the optimal set implies that
-this deviation of the mean from optimum grows with the square of the distance between the parental populations,
-and so we expect fitness costs in $F_1$s to appear on a time scale of $N_e^2$ generations.
-
-% If a population has a Gaussian distribution of trait values with variance $\sigma^2$
-% and fitness decays as $\exp(-x^2/2\gamma)$ away from the population mean,
-% then the population mean fitness is 
-% \begin{align*}
-%    \int \frac{1}{\sigma \sqrt{2\pi}} e^{-\frac{x^2}{2\sigma^2}} e^{-\frac{x^2}{2\gamma^2}} dx
-%    &=
-%    \sqrt{\frac{1}{1 + (\sigma/\gamma)^2}}  \le \frac{1}{\sigma/\gamma} .
-% \end{align*}
-% If $\sigma$ is much smaller than $\gamma$, the fitness consequences of variation
-% (i.e., genetic load) are small,
-% but if $\sigma$ is of order $\gamma$, then mean fitness is inversely proportional to trait variance
-% (i.e., $\gamma/\sigma$).
-% This implies that we expect $\sigma_S < \gamma$
-% but that once $T$ is of order $N_e (\gamma / \sigma_H)^2$, there will be loss of fitness
-% in $F_2$ hybrids.
-% Said another way,
-% if we write $\mathcal{F}(T)$ as the mean fitness of an $F_2$ hybrid between two populations separated by $T$ generations,
-% then fitness drops with the square root of $T$, measured in units of $N_e$ generations:
-% \begin{align*}
-%    \mathcal{F}(T) \le \frac{1}{\sigma_H/\gamma}\sqrt{\frac{N_e}{T}} .
-% \end{align*}
-
-For a more concrete prediction, suppose that the distribution among hybrids is Gaussian.
-A population whose trait distribution is Gaussian with mean $\mu$ and variance $\sigma$,
-has mean fitness 
-\begin{align} \label{eqn:gauss_fit}
-\int_{-\infty}^\infty \frac{1}{\sigma\sqrt{2\pi}} e^{-\frac{(x-\mu)^2}{2\sigma^2}} e^{-\frac{x^2}{2\gamma^2}} dx
-&=
-\sqrt{\frac{1}{1 + \sigma^2/\gamma^2}} \exp\left\{-\frac{\mu^2}{\gamma^2}\left(\frac{1}{1 + \sigma^2/\gamma^2}\right)\right\} .
-% &\approx
-% \frac{1}{\sigma/\gamma}\left(1 - \frac{\mu^2}{\gamma^2}\right) .
-\end{align}
-This assumes a single trait, for simplicity. %; the multivariate case is done in appendix \ref{apx:gauss_load}.
-A population of $F_2$ hybrids will have, as above, variance $\sigma^2 = \sigma^2_S + 4 \omega \sigma^2_N T/N_e$.
-The mean diverges with the square of the distance between the parentals, so we set $\mu = c_\mu \gamma T/N_e$,
-where $c_\mu$ is a constant depending on the local geometry of the optimal set.
-The mean fitness in parental populations is as in equation \ref{eqn:gauss_fit} with $\mu=0$ and $\sigma = \sigma_S$.
-This implies that if we define $\fit_2(T)$ to be the mean relative fitness among $F_2$ hybrids
-between two populations separated by $T$ generations, 
-(i.e., the mean fitness divided by the mean fitness of the parents)
-then
-\begin{align} \label{eqn:fit2}
-\fit_2(T) = 
-  \left(1 + \frac{4 \omega (\sigma_N/\gamma)^2}{(1 + (\sigma_S/\gamma)^2)} \frac{T}{N_e} \right)^{-1/2} 
-     \exp\left\{-\left(c_\mu \frac{T}{N_e}\right)^2 \left(\frac{1}{1 + (\sigma_S/\gamma)^2 + 4 \omega (\sigma_N/\gamma)^2 T/N_e}\right)\right\} .
-\end{align}
-If each of the $q$ selected directions acts independently,
-the drop in fitness will be $\fit_2(T)^q$;
-the expression for the correlated, multivariate case is given in Appendix \ref{apx:gauss_load}.
-We discuss the implications of this expression in the next section.
-
-%%%%%%%%%%
-\paragraph{Speciation rates under neutrality}
-Equation \eqref{eqn:fit2} describes how fast hybrids become inviable 
-as the time that the parental populations are isolated increases;
-what does this tell us about speciation rates under neutrality?
-From equation \eqref{eqn:fit2} we observe that
-time is always scaled in units of $N_e$ generations,
-the population standard deviations are always scaled by $\gamma$,
-and the most important term is the rate of accumulation of segregation variance,
-$4 \omega (\sigma_N/\gamma)^2$.
-All else being equal, this process will lead to speciation more quickly in smaller populations
-and in populations with more neutral genetic variation (larger $\sigma_N$).
-These parameters are related -- larger populations generally have more genetic variation --
-but since these details depend on the situation, we leave these separate.
-
-How does this prediction depend on the system size and constraint?
-If there are $p$ trait dimensions, constrained in $q$ dimensions,
-and if $\omega$ is proportional to $p-q$,
-% the rate that hybrid variance increases with separation between parentals, $\omega$,
-% is in some cases proportional to $p-q$, 
-then the rate that $F_2$ fitness drops is, roughly,
-$(1 + 4 (p-q) K T/N_e)^{-q/2} \propto q (p-q)$, where $K$ is a constant.
-Both degree of constraint and number of available neutral directions
-affect the speed of accumulation of incompatibilities --
-more unconstrained directions allows faster system drift,
-but more constrained directions implies greater fitness consequences of hybridization.
-However, note that in real systems, it is likely that $\gamma$ also depends on $p$ and $q$.
-
-Now we will interpret equation \eqref{eqn:fit2} in three situations plausible for different species,
-depicting how hybrid fitness drops as a function of $T/N_e$ in Figure \ref{fig:speciation_rates}.
-In all cases, the fitness drop for $F_1$ hybrids is much smaller than that of $F_2$ hybrids,
-so we work only with the first (square-root) term in equation \eqref{eqn:fit2}.
-
-% simgaN = sigmaS = gamma, large population
-Suppose in a large, genetically diverse population,
-the amount of heritable variation in the neutral and selected directions are roughly equal 
-($\sigma_N \approx \sigma_S$)
-but the overall amount of variation is (weakly) constrained by selection
-($\sigma_N \approx \gamma$).
-If so, then the first term of equation \eqref{eqn:fit2} is
-$1/\sqrt{1 + 2 \omega T/N_e} \approx 1 - \omega T/N_e$.
-If also $\omega=1$, then, for instance, after $0.1 N_e$ generations
-the average $F_2$ fitness has dropped by 10\% relative to the parentals.
-
-%  $\sigma_N \approx \sigma_S \ll \gamma$, ``isolated population''
-Consider instead a much smaller, isolated population
-whose genetic variation is primarily constrained by genetic drift,
-so that $\sigma_N \approx \sigma_S \ll \gamma$.
-Setting $a = (\sigma_N/\gamma)^2$ to be small,
-the fitness of $F_2$ hybrids is $\fit_2 \le 1/\sqrt{1 + 4 \omega a T/N_e} \approx 1 - 2 \omega a T/N_e$.
-Hybrid fitness seems to drop more slowly in this case in figure \ref{fig:speciation_rates},
-but since time is scaled by $N_e$, so speciation may occur \emph{faster} than in a large population.
-However, at least in some models \citep{lynch1986phenotypic}, in small populations at mutation-drift equilibrium
-the amount of genetic variance ($\sigma_N^2$) is proportional to $N_e$,
-which would compensate for this difference, 
-perhaps even predicting the rate of decrease of hybrid fitness to be independent of population size
-for small populations.
-
-% $\sigma_N \gg \sigma_S \approx \gamma$, if $N_e$ is big and there's a lot of recombination load (``species complex'')
-In the other direction, consider large metapopulations (or a ``species complex'')
-among which heritable variation is strongly constrained by selection
-(i.e., there is substantial recombination load),
-so that $\sigma_S \approx \gamma$ but $\sigma_N/\gamma$ is large.
-Then the fitness of $F_2$ hybrids is $\fit_2 \le 1/\sqrt{1 + 2 \omega a T/N_e} \approx 1 - \omega a T/N_e$,
-and could be extremely rapid if $a$ is large.
-
-For instance, between two populations of one million organisms that has 10 generations per year (a drosophilid species, perhaps)
-under the ``large population'' scenario of Figure \ref{fig:speciation_rates}A,
-system drift would lead to a substantial fitness drop of around 10\% in $F_2$ hybrids in only 10,000 years.
-This drop may be enough to induce evolutionary reinforcement of reproductive isolation.
-If one thousand of these organisms is isolated (perhaps on an island, as in Figure~\ref{fig:speciation_rates}B),
-then a similar drop could occur in around 120 years.
-On the other hand, if the population is one of several of similar size
-that have recently come into secondary contact after population re-expansion,
-the situation may be similar to that of Figure~\ref{fig:speciation_rates}C with $N_e = 10^6$,
-and so the same drop could occur after 1,100 years.
-(However, hyperdiverse populations of this last type may not be stable on these time scales.)
-
-\begin{figure}[H]
-\begin{center}
-\includegraphics{speciation_rates}
-\caption{
-Mean drop of $F_1$ and $F_2$ fitness relative to parental species,
-with $\omega=1$ and 
-\textbf{(A)} $\sigma_N^2 = \sigma_S^2 = \gamma^2$
-\textbf{(B)} $\sigma_N^2 = \sigma_S^2 = 0.1 \gamma^2$
-\textbf{(C)} $0.1 \sigma_N^2 = \sigma_S^2 = \gamma^2$.
-The $F_2$ fitness is from equation \eqref{eqn:fit2},
-and the $F_1$ fitness is determined only by the exponential term of that equation,
-which is small relative to that of increased variance.
-\label{fig:speciation_rates}}
-\end{center}
-\end{figure}
+\subsection*{The speed of speciation}
+
+XXX Talk about how drift goes like variance over $N_e$.
 
@@
 may be nearly morphologically indistinguishable.
-
-%% Fisher's geometric model
-% Differences:
-% - they look at shifting environments too
-% - they just model substitutions, so only hybrids have genetic variance
-% chevin2014niche:
-% - m is accessible, constrained dimensionlity (allows for a ridge)
-% - predicts rate of RI accumulation (due to substitutions) is \propto N^{-(1+m/2)}
-%   since this is mean log fitnees; says hybrids will have fitness
-%     exp( - m C T N^{-(1+m/2)} ) ~ 1 - m C T N^(-(1+m/2))
-% - says RI accum is *faster* at higher m but this is not obv frm eqns
-% - we say it should be ( exp(-C (T/N)^2) / sqrt(T/N) )^m :
-%     ~ 1 - m C (T/N)
-
-\paragraph{Fisher's geometric model}
-Substantial analytical work on speciation has been done using Fisher's geometric model, 
-which can also predict Haldane's rule and the regular increase of incompatibility with genetic distance
-\citep{barton2001role,fraisse2016genetics,simon2017coadapted}.
-The model is similar to the quantitative genetics model we use 
-to predict the rate of speciation, 
-although work on Fisher's geometric model generally only models substitutions
-between populations, neglecting within-species polymorphism.
-Indeed, our model may provide an additional mechanistic context 
-in which Fisher's geometric model is appropriate,
-although the typical degree of pleiotropy
-and stability of the $G$-matrix are unknown in practice.
-Our observations argue for the importance of including neutral directions in these models,
-which is not usually done (but see \citet{chevin2014niche}).
-
-Although our main focus is on the implications of system theory,
-rather than on the quantitative genetics modeling,
-it is interesting to compare analytic results.
-\citet{chevin2014niche} finds that mean fitness decrease of hybrids
-after $T$ generations grows proportionally to $m N_e^{-1-m/2} T$, for large $N_e$;
-while our corresponding rate is $m T / N_e$.
-This difference in time scaling is substantial, especially for large populations,
-and seems to appear because \citet{chevin2014niche} uses the results of \citet{sella2005application}
-to describe how the steady-state fitness and substitution process
-depends on trait dimensionality
-(the fitness peak is harder to find in higher dimensions).
-We do not incorporate this effect
-because genetic variation may be provided by other sources,
-but if we did, it would enter through our $\sigma_S$.
 
@@
 
-This simple quantitative genetics model we use above
-has been shown to produce good predictions in many situations,
-even when the substantial number of simplifying assumptions are violated \citep{burger1994distribution,turelli1994genetic}.
-The calculations above should be fairly robust even to substantial deviations from normality.
-A larger effect on these predictions seems likely due to 
-correlations due to molecular constraint, genetic linkage, population structure, historical contingency and so forth.
-Although such considerations would not change the qualitative predictions of this model,
-their combined effects could substantially change the predicted rate of accumulation of incompatibilities.
-
 Finally, despite our model's precise separation of phenotype and kryptotype, this relationship in nature may be far more complicated as aspects of the kryptotype may be less ``hidden'' than we currently assume.
@@
     We would like to thank Sergey Nuzhdin, Stevan Arnold, Michael Turelli, Patrick Phillips, Erik Lundgren and Hossein Asgharian for valuable discussion. 
-    We would also like to thank Nick Barton, Sarah Signor and Todd Parsons for very helpful comments on the manuscript.
+    We would also like to thank Nick Barton, Sarah Signor, Todd Parsons, Joachim Hermisson for very helpful comments on the manuscript.
     Work on this project was supported by funds from
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Genetic drift with a multivariate trait}
-\label{ss:quant_gen}
-
-For completeness, we provide a brief exposition of how a population 
-evolves due to genetic drift
-with a quantitative genetics model,
-as in \citet{lande1981models} or \citet{hansen1996translating}.
-These do not directly model underlying genetic basis,
-but developing a more accurate model is beyond the scope of this paper.
-
-Suppose that the population is distributed in trait space
-as a Gaussian with covariance matrix $\Sigma$ and mean $\mu$,
-whose density we write as $f(\cdot;\Sigma,\mu)$.
-Selection has the effect of multiplying this density by the fitness function and renormalizing,
-so that if expected fitness of $x$ is proportional to $\exp(-\|Lx\|^2/2)$,
-then the distribution post-selection
-has density at $x$ proportional to $f(x;\Sigma,\mu) \exp(-\|Lx\|^2/2)$.
-By the computation below (``Completing the square''),
-the result is a Gaussian distribution
-with covariance matrix $(\Sigma^{-1} + L^T L)^{-1}$ 
-and mean $(\Sigma^{-1}+L^T L)^{-1} \Sigma^{-1} \mu$.
-% Importantly, if $U$ is not invertible, then by $U^{-1}$ 
-% we mean $(U^T U)^{-1} U^T$, Moore-Penrose pseudoinverse of $U$,
-% so that $U U^{-1}$ is equal to the projection matrix
-% onto the span of the columns of $U$.
-
-After selection, we have reproduction:
-suppose this occurs as in the infinitesimal model \citep{barton2017infinitesimal},
-so that each offspring of parents with traits $x$ and $y$
-is drawn independently from a Gaussian distribution 
-with mean $(x+y)/2$ and covariance matrix $R$.
-Here, $R$ is the contribution of ``segregation variance'',
-i.e., random choices of parental alleles.
-If $\widetilde \Sigma = (\Sigma^{-1} + L^T L)^{-1}$ 
-is the covariance matrix of the parents post-selection,
-then the distribution of offspring will again be Gaussian,
-with mean equal to that of the parents
-and covariance matrix $\widetilde \Sigma/2 + R$.
-
-In summary, a generation under this model modifies the mean ($\mu$)
-and covariance matrix ($\Sigma$) of a population as follows:
-\begin{align*}
-    \mu &\mapsto \mu' = (\Sigma^{-1}+L^T L)^{-1} \Sigma^{-1} \mu \\
-    \Sigma &\mapsto \Sigma' = \frac{1}{2} (\Sigma^{-1} + L^T L)^{-1} + R .
-\end{align*}
-What measures are stable under this transformation?
-The condition $\mu = \mu'$ reduces to $\Sigma L^T L \mu = 0$;
-if we assume $R$ and therefore $\Sigma$ are of full rank,
-then this happens if and only if $\mu$ is in the null space of $L$,
-i.e., if $\mu$ lies in a neutral direction.
-The condition $\Sigma' = \Sigma$ can also be solved,
-at least numerically.
-After rearrangement, it reduces to
-$\Sigma L^T L \Sigma + (I/2 - R L^T L) \Sigma = R$.
-%% I DON'T THINK WE USE THE FOLLOWING:
-% We can find a more explicit description
-% if we assume that $x^T L^T L x = \sum_{i=1}^k x_i^2$,
-% i.e., that selection only cares about the first $k$ coordinates,
-% and then with no interactions between traits.
-% If so, the condition $\Sigma' = \Sigma$
-% can be written in block form as
-% \begin{align*}
-%     \begin{bmatrix}
-%         \Sigma_{11}^2 + (I/2 - R_{11}) \Sigma_{11}
-%         & \Sigma_{11} \Sigma_{12} + (I/2 - R_{11})\Sigma_{12} \\
-%         \Sigma_{12}^T \Sigma_{11} + \Sigma_{12}^T/2 - R_{12}^T \Sigma_{11}
-%         & \Sigma_{22} - R_{12}^T \Sigma_{12} 
-%     \end{bmatrix}
-%     &=
-%     \begin{bmatrix}
-%         R_{11} & R_{12} \\
-%         R_{12}^T & R_{22}
-%     \end{bmatrix} .
-% \end{align*}
-% The first equation, $\Sigma_{11}^2 + (I/2 - R_{11}) \Sigma_{11} = R_{11}$,
-% can be solved with the quadratic formula: 
-% $$\Sigma_{11} = (R_{11} - I/2 + Q)/2$$
-% for any $Q$ that commutes with $R_{11}$
-% and is a solution to $Q^2 = (R_{11} - I/2)^2 + 4 R_{11}$.
-% Since we need $\Sigma$ to be positive definite,
-% we take the solution with positive eigenvalues.
-% Given $\Sigma_{11}$, the remaining components are
-% \begin{align*}
-%     \Sigma_{12} 
-%     &= 
-%     \left( \Sigma_{11} + I/2 - R_{11} \right)^{-1} R_{12} \\
-%     \Sigma_{22}
-%     &=
-%     R_{12}^T \Sigma_{12} + R_{22} .
-% \end{align*}
-Importantly, the mean $\mu$ does not affect 
-either how the covariance matrix moves,
-or its stable shape.
-
-Above we have described the \emph{expected} motion of the mean and covariance.
-However, random resampling will introduce noise.
-Suppose that a population of $N$ individuals
-behaves approximately as described above.
-By the above,
-we may expect that the covariance matrix stays close to a constant value $\Sigma$,
-computed from $R$ and $L$ as above,
-so that we need only consider motion of the mean, $\mu$.
-Since we take a sample of size $N$ to construct the next generation,
-the next generation's mean is drawn from a Gaussian distribution with mean $\mu'$
-and covariance matrix $\Sigma/N$.
-Defining $\Gamma = (I - (I + \Sigma L^T L)^{-1})$,
-this can be written as
-\begin{align*}
-    \mu' - \mu = \Gamma \mu + \epsilon/\sqrt{N} ,
-\end{align*}
-where $\epsilon$ is a multivariate Gaussian with mean zero and covariance matrix $\Sigma$.
-Let $\mu(k)$ denote the mean in the $k^\text{th}$ generation,
-and suppose that $\mu$ differs from optimal by something of order $1/\sqrt{N}$:
-if $\nu(t) = \sqrt{N} \mu(t\sqrt{N})$ is the rescaled process,
-then the previous equation implies that as $N \to \infty$, 
-in the limit $\nu$ solves the It\^{o} equation
-\begin{align*}
-    d \nu(t) = \Gamma \nu(t) dt + \Sigma^{1/2} dW(t) ,
-\end{align*}
-where now $W(t)$ is a multivariate white noise.
-This has an explicit solution as a multivariate Ornstein-Uhlenbeck process:
-\begin{align*}
-    \nu(t) = e^{-t \Gamma} \nu(0) + \int_0^t e^{-(t-s) \Gamma} \Sigma^{1/2} dW(s).
-\end{align*}
-% Note that after some algebra, $\Gamma = \Sigma L^T L (I - \Sigma L^T L)^{-1}$.
-The asymptotic variance of this process in the direction $z$ is
-\begin{align} \label{eqn:limiting_cov}
-    \lim_{t \to \infty} \var[\nu(t) \cdot z]
-    &=
-    \int_0^\infty z^T e^{-s \Gamma} \Sigma e^{-s \Gamma} z ds ,
-\end{align}
-which is infinite if and only if $\Gamma z = 0$,
-which occurs if and only if $Lz=0$.
-This implies that at equilibrium, 
-population mean trait values lie away from the optimal set
-by a Gaussian displacement of order $1/\sqrt{N}$
-with a covariance matrix given by equation \eqref{eqn:limiting_cov}.
-
-\paragraph{Completing the square}
-First note that if $A$ is symmetric,
-\begin{align*}
-    (x-y)^T A (x-y)
-    &=
-    x^T A \left( x - 2y \right) + y^T A y ,
-\end{align*}
-and so if $B$ is also symmetric and $A+B$ is invertible,
-\begin{align*}
-    (x-y)^T A (x-y) + x^T B x
-    &=
-    x^T (A + B) \left( x - 2 (A + B)^{-1} A y \right) + y^T A y \\
-    &=
-    \left( x - (A + B)^{-1} A y \right)^T
-    (A + B)
-    \left( x - (A + B)^{-1} A y \right) \\
-    &\qquad{}
-    + y^T A y - y^T A^T (A+B)^{-1} A y .
-\end{align*}
-Therefore, 
-by substituting $A=\Sigma^{-1}$ and $B=L^T L$,
-\begin{align*}
-    \frac{ f(x;\Sigma,y) \exp(-x^T L^T L x / 2)}{\int f(z;\Sigma,y) \exp(-z^T L^T L z / 2) dz}
-    &=
-    f(x; (\Sigma^{-1} + L^T L)^{-1}, (\Sigma^{-1}+L^T L)^{-1} \Sigma^{-1} y) .
-\end{align*}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Gaussian load}
-\label{apx:gauss_load}
-
-Suppose that a population has a Gaussian distribution in $d$-dimensional trait space
-with mean $\mu$ and covariance matrix $\Sigma$,
-and that fitness of an individual at $x$ is $\exp(-\|Lx\|^2/2)$.
-Then, completing the square as above with $A=\Sigma^{-1}$, $y=\mu$, and $B=L^T L$,
-and defining $Q = (\Sigma^{-1} + L^T L)^{-1}$,
-\begin{align*}
-    &
-    \frac{1}{\sqrt{2 \pi}^n \det(\Sigma)^{1/2}} 
-        \int e^{-\frac{1}{2} x^T \Sigma^{-1} x} e^{-\frac{1}{2} x^T L^T L x} dx \\
-    &\qquad =
-    \frac{1}{\sqrt{2 \pi}^n \det(\Sigma)^{1/2}} 
-        \int e^{-\frac{1}{2}(x-Q \Sigma^{-1} \mu)^T Q^{-1} (x-Q \Sigma^{-1} \mu)} dx 
-        \\ &\qquad \qquad {}
-        \times e^{\mu^T \left( I - \Sigma^{-1} Q \right) \Sigma^{-1} \mu} \\
-    &\qquad =
-    \sqrt{\frac{\det(Q)}{\det(\Sigma)}}
-        \exp\left\{\mu^T \left( I - \Sigma^{-1} Q \right) \Sigma^{-1} \mu\right\} \\
-    &\qquad =
-    \sqrt{\frac{1}{\det(\Sigma)\det(\Sigma^{-1}+L^TL)}}
-        \exp\left\{\mu^T \left( I - (I + L^T L \Sigma)^{-1} \right) \Sigma^{-1} \mu\right\} \\
-\end{align*}
-This is the product of two terms:
-the first gives the drop in fitness due to segregation variance,
-and the second the drop due to a shift in mean away from the optimum.
-A similar decomposition appears in \citet{barton2001role} and \citet{chevin2014niche},
-but for mean log fitness.
-
-Now suppose that $\Sigma = \sigma^2 I$ and $L = I/\gamma$.
-Then,
-\begin{align*}
-    \sqrt{\frac{1}{\det(\Sigma)\det(\Sigma^{-1}+L^TL)}}
-    &=
-    \sqrt{\frac{1}{\sigma^{2d} (1/\sigma^2 + 1/\gamma^2)^d}} \\
-    &=
-    \frac{1}{(1+ \left(\sigma/\gamma\right)^2)^{d/2}} .
-\end{align*}
-Also, 
-\begin{align*}
-        \left( I - (I + L^T L \Sigma)^{-1} \right) \Sigma^{-1} 
-        &=
-        \frac{1}{\sigma^2}\left( 1 - (1 + (\sigma/\gamma)^2)^{-1} \right) I \\
-        &=
-        \frac{1}{\gamma^2}\frac{1}{(1 + (\sigma/\gamma)^2)} I \\
-\end{align*}
-
-%%%%%%%%%%%
-\section{Evolution of segregation covariance}
-\label{apx:seg_cov}
-
-The description above does not describe how \emph{two} diverging populations interact,
-since the amount of \emph{segregation variance}, quantified by $R$,
-will not stay constant.
-This has been quantified in various models before
-\citep[e.g.,][]{slatkin1994segregation,barton2001role,chevin2014niche};
-we provide the following calculations for completeness.
-
-To get an idea of how segregation variance might evolve,
-suppose that a trait is determined by $L$ unlinked, biallelic loci,
-and that the $i^\mathrm{th}$ locus has two alleles with additive effects $\pm \alpha_i$,
-so that being homozygous for the $+$ allele contributes $+2 \alpha_i$ to the trait.
-For simplicity, we will neglect the effects of selection.
-If the $+$ allele at locus $i$ is at frequency $p_i$ in a population,
-then the mean and genetic variance of the trait in a diploid population with random mating is
-\begin{align*}
-    m &= 2 \sum_i (2 p_i - 1) \alpha_i \\
-    s^2 &= 4 \sum_i p_i (1-p_i) \alpha_i^2 .
-\end{align*}
-
-Segregation variance between two parents depends on the loci at which either are heterozygous,
-and each locus contributes independently
-since alleles are additive.
-If the alleles are at Hardy-Weinberg proportions,
-then since segregation acts like a fair coin flip, a heterozygous locus contributes $\alpha_i^2/4$ to the variance,
-and so the \emph{mean} segregation variance, averaging across parents, is
-% two parents, prob 2p(1-p) for prob the locus is heterozygous, variance 0.5(1-0.5)\alpha^2.
-\begin{align*}
-    R_0(p) = \sum_i p_i (1-p_i) \alpha_i^2 .
-\end{align*}
-
-On the other hand,
-if the second parent came from a distinct population
-with frequencies $q_i$
-(an $F_1$ hybrid),
-this would be
-\begin{align*}
-    R_1(p,q) &= \frac{1}{2} \sum_i p_i^2 (1-p_i)^2 \alpha_i^2 
-                + \frac{1}{2} \sum_i q_i^2 (1-q_i)^2 \alpha_i^2 \\
-                &= (R_0(p) + R_0(q))/2 .
-\end{align*}
-If we assume that the populations are at equilibrium, $R_0(p) \approx R_0(q)$,
-and so $R_1(p,q) \approx R_0(p)$.
-
-Now consider an $F_2$ hybrid, where both parents are $F_1$
-and so each heterozygous at locus $i$ with probability $p_i (1-q_i) + (1-p_i) q_i$.
-Then
-\begin{align*}
-    R_2(p,q) &= \frac{1}{2} \sum_i \left(p_i (1-q_i) + (1-p_i) q_i \right) \alpha_i^2  .
-\end{align*}
-Suppose that the two populations are slightly drifted from each other,
-with frequency difference $p_i-q_i = 2\epsilon_i$.
-Then,
-\begin{align*}
-    p (1-q) + p (1-q)
-    &=
-    (u+\epsilon) (1-u+\epsilon) + (u-\epsilon) (1-u-\epsilon) \\
-    &=
-    2 u (1-u)
-    % + \epsilon (u + (1-u) - u - (1-u))
-    + 2 \epsilon^2 .
-\end{align*}
-If the frequencies have evolved neutrally in unconnected, Wright-Fisher populations 
-of effective size $N$ for $t$ generations from a common ancestor with allele frequency $u$,
-then $\epsilon$ has mean zero and variance roughly $2 u (1-u) t/N$.
-Still assuming the populations are at stationarity,
-so that $R_0$ is constant between the two,
-and taking the frequencies $p_i$ as a proxy for the ancestral frequencies $u_i$,
-this implies that we expect
-\begin{align*}
-    R_2 
-    &\approx 
-    R_0 + 2 \sum_i p_i (1-p_i) \alpha_i^2 t/N \\
-    &=
-    \left( 1 + \frac{2t}{N} \right) R_0 .
-\end{align*}
-On the other hand, the expected squared difference in trait \emph{means} between two such populations is
-\begin{align} \label{eqn:var_dx}
-    8 \sum_i p_i (1-p_i) \alpha_i^2 t / N = 8 R_0 t /N .
-\end{align}
-This implies that under this model,
-segregation variance in $F_2$ hybrids between two populations
-is roughly increased by a factor of $1/8$ of the difference between their means.
+% reviews here if toggled above
+% \ifsubmission\processdelayedfloats\fi
+
+\includereviews
 

### 2018-08-16 ridge_evolution_paper.tex
first pass at splitting the papers

\documentclass{article}
\usepackage{amsmath, amssymb, color, xcolor, amsthm}
\usepackage{graphicx, wrapfig, float, caption, dsfont, bbm, xfrac}
\usepackage{fullpage}
\usepackage[backref=page, hidelinks, colorlinks=true, citecolor=blue!60!black!100]{hyperref}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, shapes}
\usepackage{caption, subcaption}
\usepackage{natbib} % gives us \citet: Author (year) and \citep: (Author; year)
\usepackage{authblk}
\usepackage{multicol}
\usepackage[bottom]{footmisc}

% for comments in margins or not
\newif\ifmargincomments
\margincommentsfalse
% \margincommentstrue
\ifmargincomments
    \usepackage{todonotes}
    \usepackage[left=1cm,right=6.5cm,top=3cm,bottom=3cm,nohead,nofoot,marginparwidth=6cm]{geometry}
    \newcommand{\plr}[1]{\todo[color=blue!25]{#1}}
    \newcommand{\js}[1]{\todo[color=green!25]{#1}}
    % inline comment version
    \newcommand{\plri}[1]{{\color{blue}\it #1}}
\else
    \newcommand{\plr}[1]{{\color{blue}\it #1}}
    \newcommand{\js}[1]{{\color{green}\it #1}}
    \newcommand{\plri}[1]{\plr{#1}}
\fi

\newif\ifsubmission
\submissiontrue
% \submissionfalse
\ifsubmission
    % % stuff for submission
    \usepackage{lineno}
    % allow line numbers around math environments
    % from https://tex.stackexchange.com/questions/43648/why-doesnt-lineno-number-a-paragraph-when-it-is-followed-by-an-align-equation
        \newcommand*\patchAmsMathEnvironmentForLineno[1]{%
          \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
          \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
          \renewenvironment{#1}%
             {\linenomath\csname old#1\endcsname}%
             {\csname oldend#1\endcsname\endlinenomath}}% 
        \newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
          \patchAmsMathEnvironmentForLineno{#1}%
          \patchAmsMathEnvironmentForLineno{#1*}}%
        \AtBeginDocument{%
        \patchBothAmsMathEnvironmentsForLineno{equation}%
        \patchBothAmsMathEnvironmentsForLineno{align}%
        \patchBothAmsMathEnvironmentsForLineno{flalign}%
        \patchBothAmsMathEnvironmentsForLineno{alignat}%
        \patchBothAmsMathEnvironmentsForLineno{gather}%
        \patchBothAmsMathEnvironmentsForLineno{multline}%
        }
\else
    % for the nice version
\fi

\newcommand{\jss}[1]{{\color{olive}\it #1}}
% \newcommand{\ddt}{\frac{d}{dt}}
\newcommand{\ddt}{\dot}
\newcommand{\ro}{{ro}}
\newcommand{\nro}{{\bar{r}o}}
\newcommand{\rno}{{r\bar{o}}}
\newcommand{\nrno}{{\bar{r}\bar{o}}}
\newcommand{\reachable}{\mathcal{R}}
\newcommand{\unobservable}{\bar{\mathcal{O}}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\var}{\mathop{\mbox{Var}}}
\newcommand{\cov}{\mathop{\mbox{Cov}}}
\newcommand{\tr}{\mathop{\mbox{tr}}} % trace
\newcommand{\pda}{\frac{\partial}{\partial A_{ij}}}
\newcommand{\ind}{\mathds{1}}
\newcommand{\grad}{\nabla}

\newcommand{\calH}{\mathcal{H}}
\newcommand{\diag}{\text{diag}}
\newcommand{\1}{\mathbbm{1}}

% the triple (A,B,C) as 'system' (not using)
\newcommand{\Sys}{\mathcal{S}}
% neutral set of all systems
\newcommand{\allS}{\mathcal{N}}

% fitness as a fn of distance
\newcommand{\fit}{\mathcal{F}}
% fitness as a fn of A
\newcommand{\fitx}{\mathcal{F}}
% set of optimal coefficients
\newcommand{\optx}{\mathcal{X}}
% optimal phenotype
\newcommand{\optph}{\Phi_0}
% distance in phenotype space
\newcommand{\dph}{d}
% incompatibility
\newcommand{\Incompat}{\mathcal{I}}

\DeclareMathOperator{\spn}{span}

\newtheorem{example}{Example}

% Responses to reviews
\input{review-response-commands}
% set this to show line numbers and include responses to reviews or not
\newif\ifreviewresponses
\reviewresponsestrue  % include them
% \reviewresponsesfalse  % don't include them
\newcommand{\responsefile}{review-responses}  % name of the review responses file

\begin{document}
%\linenumbers

{\centering
{\Huge \bf Genetic drift along a fitness ridge} \\ \vspace{0.75cm}
Joshua S. Schiffman$^{\dagger}$ \qquad Peter L. Ralph$^{\dagger \ddagger}$ \\ \vspace{0.5cm}
$^{\dagger}${\footnotesize {Molecular and Computational Biology, University of Southern California, Los Angeles, California 90089, U.S.A. \\
$^{\ddagger}$Departments of Mathematics and Biology \& The Institute for Ecology and Evolution, University of Oregon, Eugene, Oregon 97403, U.S.A.}} \\ \vspace{0.5cm}
{\small \texttt{jsschiff@usc.edu} \qquad 
\texttt{plr@uoregon.edu}} \\ \vspace{0.5cm}
\small \today \\
\vspace{0.25cm}
}

\begin{abstract}
Drift along a ridge.
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Introduction}

%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Results}

%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{System drift and the accumulation of incompatibilities}

Thus far we have seen that many distinct molecular mechanisms can realize identical phenotypes
and that these mechanisms may fail to produce viable hybrids.
Does evolution shift molecular mechanisms
fast enough to be a significant driver of speciation?
To approach this question,
we explore a general quantitative genetic model in which a population drifts stochastically
near a set of equivalent and optimal systems
due to the action of recombination, mutation, and demographic noise.
Although this is motivated by the results on linear systems above,
the quantitative genetics calculations are more general,
and only depend on the presence of genetic variation and a continuous set of phenotypically equivalent systems.

We will suppose that each organism's phenotype is determined by its vector of coefficients, denoted by $x=(x_1, x_2, \ldots, x_p)$,
and that the corresponding fitness is determined by the distance of its phenotype to optimum.
The optimum phenotype is unique, but is realized by many distinct $x$ -- those falling in the ``optimal set'' $\allS$.
The phenotypic distance to optimum of an organism with coefficients $x$ is denoted $D(x)$.
In the results above, $x = (A,B,C)$ and $D(x)$ is given by equation \eqref{eqn:distance}.
The fitness of an organism with coefficients $x$ will be $\exp(-D(x)^2)$.
% this assumption does not entail a loss of generality since $D$ is arbitrary.
We assume that in the region of interest, the map $D$ is smooth
and that we can locally approximate the optimal set $\allS$ as a quadratic surface.
As above, an individual's coefficients are given by averaging its parentally inherited coefficients and adding random noise due to segregation and possibly new mutation.
Concretely, we use the \emph{infinitesimal model} for reproduction \citep{barton2017infinitesimal} --
the offspring of parents at $x$ and $x'$ will have coefficients $(x+x')/2 + \varepsilon$,
where $\varepsilon$ is a random Gaussian displacement due to random assortment of parental alleles.

%% REVIEW OF RELATED RESULTS
This fitness landscape is locally Gaussian with a rank-deficient covariance matrix.
Since we allow for substantial genetic variation within populations,
this model falls in the same class as \citet{lande1981models} and \citet{lande1983measurement},
which did not consider reproductive incompatibility.
Substantial work on speciation has been done under the assumption of a monomorphic population
whose trait is shifted by sequential fixation of alleles,
i.e., Fisher's ``geometric model'' \citep{fisher1930genetical,poon2000compensating}.
This work has included both stationary optima (like we study)
and moving optima (i.e., adaptation) \citep[e.g.,][]{barton2001role,chevin2014niche}.
\citet{martin2014fishers} derived this model from a few general assumptions,
and used random matrix theory to calculate the distribution of fitness effects.
\citet{chevin2014niche} studied a general version with neutral directions,
and found that the rate of accumulation of reproductive isolation 
decreases with $N_e$ with a form that depends on trait dimension.
\citet{fraisse2016genetics} showed that most recognized empirical patterns in the speciation literature
could be explained by this model 
(although best if fitness took the form $\exp(-d^k)$
for some $k > 2$, where $d$ is distance to the optimum),
and \citet{simon2017coadapted} further compared predictions from the model to empirical data.
In the context of a stationary optimum, 
the primary contribution to hybrid unfitness is segregation variance,
i.e., greater phenotypic variance in $F_2$ hybrids than in parentals
due to drift having changed the genetic basis of the trait separately in the parental populations
(\citet{chevin2014niche} refers to these as ``transgressive incompatibilities'').
This turns out to be the main source of incompatibilities in the model we study as well,
even though in principle curvature of the optimal set also contributes
(as also seen in \citet{rosas2010cryptic}).
Our analysis relies on local approximations;
on longer time scales the appropriate model may share properties with
the ``holey'' landscapes of \citet{gavrilets2004fitness}.

As our goal here is to sketch out the rough implications of our main results,
keeping the assumptions clearly visible,
we provide relatively rough arguments,
rather than presenting calculations in full multivariate generality.

\paragraph{System drift}
% plr how fast does it drift
We work with a randomly mating population of effective size $N_e$. 
If the population variation has standard deviation $\sigma$ in a particular direction,
since subsequent generations resample from this diversity,
the population mean coefficient will move a random distance of size $\sigma/\sqrt{N_e}$ per generation,
simply because this is the standard deviation of the mean of a random sample \citep{lande1981models}.
% (This could be taken as a definition of $N_e$.)
Selection will tend to restrain this motion,
but movement along the optimal set $\allS$ is unconstrained,
and so we expect the population mean to drift along the optimal set like a particle diffusing.
% (although perhaps complicated by recombination load \citep{recomb_load})
The amount of variance in particular directions in coefficient space 
depends on constraints imposed by selection and 
correlations between the genetic variation underlying different coefficients 
(the $G$ matrix \citep{arnold2008understanding}).
It therefore seems reasonable to coarsely model the time evolution of population variation in regulatory coefficients as 
a ``cloud'' of width $\sigma$ about the population mean, 
which moves as an unbiased Brownian motion through the set of network coefficients that give the optimal phenotype.

% covariance which may arise due to functional constraints and/or statistical linkage.
% There may well be functional constraints -- but these are not sufficiently well-known to say anything general about.
%For instance,
%if the variation is due to \textit{cis}-regulatory variants,
%the genetic basis of each \emph{row} of $A$ likely lies within a few kilobases of tightly linked sequence,
%across which a population may carry only a few common haplotypes.
%However, covariance due to transiently assembled haplotypes is not expected to be stable over long periods of time --
%a common \textit{cis}-regulatory haplotype of transcription factor $k$ with particularly strong binding to both $i$ and $j$
%(leading to positive covariance between $A_{ik}$ and $A_{jk}$)
%is no more likely to appear than one with strong binding to $i$ but particularly weak binding to $j$ (negative covariance).
%(Such transient covariances may well increase the variance of the per-generation change in network mean, however \citep{barton_linkage}.)

Next, we calculate with some simplifying assumptions to give the general idea;
multivariate derivations appear in Appendix \ref{ss:quant_gen}.
There will in general be different amounts of variation in different directions;
to keep the discussion intuitive, we only discuss $\sigma_N$, the amount of variation in ``neutral'' directions
(i.e., directions along $\allS$),
and $\sigma_S$, the amount of variation in ``selected'' directions (perpendicular to $\allS$).
The other relevant scale we denote by $\gamma$,
which is the scale on which distance to phenotypic optimum changes as $x$ moves away from the optimal set, $\allS$.
Concretely, $\gamma$ is
%the inverse of the derivative of 
$1/(\frac{d}{du}D(x+uz))$ 
where $x$ is optimal and $z$ is a ``selected'' direction perpendicular to $\allS$.
With these parameters, a typical individual will have a fitness of around $\exp(-(\sigma_S/\gamma)^2)$.
Of course, there are in general many possible neutral and selected directions;
we take $\gamma$ to be an appropriate average over possible directions.

\paragraph{Hybridization}
The means of two allopatric populations each of effective size $N_e$ separated for $T$ generations
will be a distance roughly of order $2\sigma_N \sqrt{T/N_e}$ apart along $\optx$.
(Consult figure \ref{fig:conceptual_fig} for a conceptual diagram.)
A population of $F_1$ hybrids has one haploid genome from each,
whose coefficients are averaged,
and so will have mean system coefficients at the midpoint between their means.
The distribution of $F_2$ hybrids will have mean at the average of the two populations,
but will have higher variance \citep{wright1968evolution,barton2001role,rosas2010cryptic}.
This ``segregation variance'' of $F_2$ hybrids 
can be shown to increase linearly with the square of the distance between
parental population means
under models of both simple and polygenic traits.
This is suggested by figure \ref{fig:conceptual_fig} and shown by \citet{slatkin1994segregation}
(also see Appendix \ref{apx:seg_cov} for a different derivation).
Concretely, we expect the population of $F_1$s to have variance $\sigma^2_S$ in the selected direction
(the same as within each parental population),
but the population of $F_2$ hybrids will have variance $\sigma^2_S + 4 \omega \sigma^2_N T/N_e$,
where $\omega$ is a factor that depends on the genetic basis of the coefficients.
If the optimal set $\allS$ has dimension $q$,
using the polygenic model of appendix \ref{apx:seg_cov}, 
$\omega$ is proportional to the number of degrees of freedom: $\omega = (p-q)/8$.
If each trait is controlled by a single locus, as in figure \ref{fig:conceptual_fig},
the value is similar.

%%% CITE SOME OF THESE
%% empirical data on sigma
% The amount and structure of this standing variation is established over long time scales
% by many factors, including mutation-selection balance, 
% shifts in the phenotypic optimum, and/or spatial variation in the optimum \citep{hansen1996translating}.
% Quantitative genetics models of mutation-selection balance 
% predict precise levels and structure of standing variation \citep{kimura_mutsel,lande_mutsel,lande1981models},
% but it is unclear how well these predictions match reality \citep{johnson_barton}
% and how much they are expected to change over time \citep{arnold_changing_G}.

What are the fitness consequences?
A population of $F_2$ hybrids will begin to be substantially less fit than the parentals
once they differ from the optimum by a distance of order $\gamma$,
i.e., once $\sqrt{4 \omega T/N_e} \approx \gamma / \sigma_N$.
This implies that hybrid incompatibility among $F_2$ hybrids should appear much slower --
on a time scale of $N_e (\gamma / \sigma_N)^2 / (4 \omega)$ generations.
The $F_1$s will not suffer fitness consequences until the hybrid mean is further than $\gamma$ from the optimum;
% as shown in appendix \ref{apx:away_from_opt} 
as suggested by figure \ref{fig:conceptual_fig}, Taylor expanding $D^2$ along the optimal set implies that
this deviation of the mean from optimum grows with the square of the distance between the parental populations,
and so we expect fitness costs in $F_1$s to appear on a time scale of $N_e^2$ generations.

% If a population has a Gaussian distribution of trait values with variance $\sigma^2$
% and fitness decays as $\exp(-x^2/2\gamma)$ away from the population mean,
% then the population mean fitness is 
% \begin{align*}
%    \int \frac{1}{\sigma \sqrt{2\pi}} e^{-\frac{x^2}{2\sigma^2}} e^{-\frac{x^2}{2\gamma^2}} dx
%    &=
%    \sqrt{\frac{1}{1 + (\sigma/\gamma)^2}}  \le \frac{1}{\sigma/\gamma} .
% \end{align*}
% If $\sigma$ is much smaller than $\gamma$, the fitness consequences of variation
% (i.e., genetic load) are small,
% but if $\sigma$ is of order $\gamma$, then mean fitness is inversely proportional to trait variance
% (i.e., $\gamma/\sigma$).
% This implies that we expect $\sigma_S < \gamma$
% but that once $T$ is of order $N_e (\gamma / \sigma_H)^2$, there will be loss of fitness
% in $F_2$ hybrids.
% Said another way,
% if we write $\mathcal{F}(T)$ as the mean fitness of an $F_2$ hybrid between two populations separated by $T$ generations,
% then fitness drops with the square root of $T$, measured in units of $N_e$ generations:
% \begin{align*}
%    \mathcal{F}(T) \le \frac{1}{\sigma_H/\gamma}\sqrt{\frac{N_e}{T}} .
% \end{align*}

For a more concrete prediction, suppose that the distribution among hybrids is Gaussian.
A population whose trait distribution is Gaussian with mean $\mu$ and variance $\sigma$,
has mean fitness 
\begin{align} \label{eqn:gauss_fit}
\int_{-\infty}^\infty \frac{1}{\sigma\sqrt{2\pi}} e^{-\frac{(x-\mu)^2}{2\sigma^2}} e^{-\frac{x^2}{2\gamma^2}} dx
&=
\sqrt{\frac{1}{1 + \sigma^2/\gamma^2}} \exp\left\{-\frac{\mu^2}{\gamma^2}\left(\frac{1}{1 + \sigma^2/\gamma^2}\right)\right\} .
% &\approx
% \frac{1}{\sigma/\gamma}\left(1 - \frac{\mu^2}{\gamma^2}\right) .
\end{align}
This assumes a single trait, for simplicity. %; the multivariate case is done in appendix \ref{apx:gauss_load}.
A population of $F_2$ hybrids will have, as above, variance $\sigma^2 = \sigma^2_S + 4 \omega \sigma^2_N T/N_e$.
The mean diverges with the square of the distance between the parentals, so we set $\mu = c_\mu \gamma T/N_e$,
where $c_\mu$ is a constant depending on the local geometry of the optimal set.
The mean fitness in parental populations is as in equation \ref{eqn:gauss_fit} with $\mu=0$ and $\sigma = \sigma_S$.
This implies that if we define $\fit_2(T)$ to be the mean relative fitness among $F_2$ hybrids
between two populations separated by $T$ generations, 
(i.e., the mean fitness divided by the mean fitness of the parents)
then
\begin{align} \label{eqn:fit2}
\fit_2(T) = 
  \left(1 + \frac{4 \omega (\sigma_N/\gamma)^2}{(1 + (\sigma_S/\gamma)^2)} \frac{T}{N_e} \right)^{-1/2} 
     \exp\left\{-\left(c_\mu \frac{T}{N_e}\right)^2 \left(\frac{1}{1 + (\sigma_S/\gamma)^2 + 4 \omega (\sigma_N/\gamma)^2 T/N_e}\right)\right\} .
\end{align}
If each of the $q$ selected directions acts independently,
the drop in fitness will be $\fit_2(T)^q$;
the expression for the correlated, multivariate case is given in Appendix \ref{apx:gauss_load}.
We discuss the implications of this expression in the next section.

%%%%%%%%%%
\paragraph{Speciation rates under neutrality}
Equation \eqref{eqn:fit2} describes how fast hybrids become inviable 
as the time that the parental populations are isolated increases;
what does this tell us about speciation rates under neutrality?
From equation \eqref{eqn:fit2} we observe that
time is always scaled in units of $N_e$ generations,
the population standard deviations are always scaled by $\gamma$,
and the most important term is the rate of accumulation of segregation variance,
$4 \omega (\sigma_N/\gamma)^2$.
All else being equal, this process will lead to speciation more quickly in smaller populations
and in populations with more neutral genetic variation (larger $\sigma_N$).
These parameters are related -- larger populations generally have more genetic variation --
but since these details depend on the situation, we leave these separate.

How does this prediction depend on the system size and constraint?
If there are $p$ trait dimensions, constrained in $q$ dimensions,
and if $\omega$ is proportional to $p-q$,
% the rate that hybrid variance increases with separation between parentals, $\omega$,
% is in some cases proportional to $p-q$, 
then the rate that $F_2$ fitness drops is, roughly,
$(1 + 4 (p-q) K T/N_e)^{-q/2} \propto q (p-q)$, where $K$ is a constant.
Both degree of constraint and number of available neutral directions
affect the speed of accumulation of incompatibilities --
more unconstrained directions allows faster system drift,
but more constrained directions implies greater fitness consequences of hybridization.
However, note that in real systems, it is likely that $\gamma$ also depends on $p$ and $q$.

Now we will interpret equation \eqref{eqn:fit2} in three situations plausible for different species,
depicting how hybrid fitness drops as a function of $T/N_e$ in Figure \ref{fig:speciation_rates}.
In all cases, the fitness drop for $F_1$ hybrids is much smaller than that of $F_2$ hybrids,
so we work only with the first (square-root) term in equation \eqref{eqn:fit2}.

% simgaN = sigmaS = gamma, large population
Suppose in a large, genetically diverse population,
the amount of heritable variation in the neutral and selected directions are roughly equal 
($\sigma_N \approx \sigma_S$)
but the overall amount of variation is (weakly) constrained by selection
($\sigma_N \approx \gamma$).
If so, then the first term of equation \eqref{eqn:fit2} is
$1/\sqrt{1 + 2 \omega T/N_e} \approx 1 - \omega T/N_e$.
If also $\omega=1$, then, for instance, after $0.1 N_e$ generations
the average $F_2$ fitness has dropped by 10\% relative to the parentals.

%  $\sigma_N \approx \sigma_S \ll \gamma$, ``isolated population''
Consider instead a much smaller, isolated population
whose genetic variation is primarily constrained by genetic drift,
so that $\sigma_N \approx \sigma_S \ll \gamma$.
Setting $a = (\sigma_N/\gamma)^2$ to be small,
the fitness of $F_2$ hybrids is $\fit_2 \le 1/\sqrt{1 + 4 \omega a T/N_e} \approx 1 - 2 \omega a T/N_e$.
Hybrid fitness seems to drop more slowly in this case in figure \ref{fig:speciation_rates},
but since time is scaled by $N_e$, so speciation may occur \emph{faster} than in a large population.
However, at least in some models \citep{lynch1986phenotypic}, in small populations at mutation-drift equilibrium
the amount of genetic variance ($\sigma_N^2$) is proportional to $N_e$,
which would compensate for this difference, 
perhaps even predicting the rate of decrease of hybrid fitness to be independent of population size
for small populations.

% $\sigma_N \gg \sigma_S \approx \gamma$, if $N_e$ is big and there's a lot of recombination load (``species complex'')
In the other direction, consider large metapopulations (or a ``species complex'')
among which heritable variation is strongly constrained by selection
(i.e., there is substantial recombination load),
so that $\sigma_S \approx \gamma$ but $\sigma_N/\gamma$ is large.
Then the fitness of $F_2$ hybrids is $\fit_2 \le 1/\sqrt{1 + 2 \omega a T/N_e} \approx 1 - \omega a T/N_e$,
and could be extremely rapid if $a$ is large.

For instance, between two populations of one million organisms that has 10 generations per year (a drosophilid species, perhaps)
under the ``large population'' scenario of Figure \ref{fig:speciation_rates}A,
system drift would lead to a substantial fitness drop of around 10\% in $F_2$ hybrids in only 10,000 years.
This drop may be enough to induce evolutionary reinforcement of reproductive isolation.
If one thousand of these organisms is isolated (perhaps on an island, as in Figure~\ref{fig:speciation_rates}B),
then a similar drop could occur in around 120 years.
On the other hand, if the population is one of several of similar size
that have recently come into secondary contact after population re-expansion,
the situation may be similar to that of Figure~\ref{fig:speciation_rates}C with $N_e = 10^6$,
and so the same drop could occur after 1,100 years.
(However, hyperdiverse populations of this last type may not be stable on these time scales.)

\begin{figure}[H]
\begin{center}
\includegraphics{speciation_rates}
\caption{
Mean drop of $F_1$ and $F_2$ fitness relative to parental species,
with $\omega=1$ and 
\textbf{(A)} $\sigma_N^2 = \sigma_S^2 = \gamma^2$
\textbf{(B)} $\sigma_N^2 = \sigma_S^2 = 0.1 \gamma^2$
\textbf{(C)} $0.1 \sigma_N^2 = \sigma_S^2 = \gamma^2$.
The $F_2$ fitness is from equation \eqref{eqn:fit2},
and the $F_1$ fitness is determined only by the exponential term of that equation,
which is small relative to that of increased variance.
\label{fig:speciation_rates}}
\end{center}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Discussion}

This simple quantitative genetics model we use above
has been shown to produce good predictions in many situations,
even when the substantial number of simplifying assumptions are violated \citep{burger1994distribution,turelli1994genetic}.
The calculations above should be fairly robust even to substantial deviations from normality.
A larger effect on these predictions seems likely due to 
correlations due to molecular constraint, genetic linkage, population structure, historical contingency and so forth.
Although such considerations would not change the qualitative predictions of this model,
their combined effects could substantially change the predicted rate of accumulation of incompatibilities.

%% Fisher's geometric model
% Differences:
% - they look at shifting environments too
% - they just model substitutions, so only hybrids have genetic variance
% chevin2014niche:
% - m is accessible, constrained dimensionlity (allows for a ridge)
% - predicts rate of RI accumulation (due to substitutions) is \propto N^{-(1+m/2)}
%   since this is mean log fitnees; says hybrids will have fitness
%     exp( - m C T N^{-(1+m/2)} ) ~ 1 - m C T N^(-(1+m/2))
% - says RI accum is *faster* at higher m but this is not obv frm eqns
% - we say it should be ( exp(-C (T/N)^2) / sqrt(T/N) )^m :
%     ~ 1 - m C (T/N)

\paragraph{Fisher's geometric model}
Substantial analytical work on speciation has been done using Fisher's geometric model, 
which can also predict Haldane's rule and the regular increase of incompatibility with genetic distance
\citep{barton2001role,fraisse2016genetics,simon2017coadapted}.
The model is similar to the quantitative genetics model we use 
to predict the rate of speciation, 
although work on Fisher's geometric model generally only models substitutions
between populations, neglecting within-species polymorphism.
Indeed, our model may provide an additional mechanistic context 
in which Fisher's geometric model is appropriate,
although the typical degree of pleiotropy
and stability of the $G$-matrix are unknown in practice.
Our observations argue for the importance of including neutral directions in these models,
which is not usually done (but see \citet{chevin2014niche}).

Although our main focus is on the implications of system theory,
rather than on the quantitative genetics modeling,
it is interesting to compare analytic results.
\citet{chevin2014niche} finds that mean fitness decrease of hybrids
after $T$ generations grows proportionally to $m N_e^{-1-m/2} T$, for large $N_e$;
while our corresponding rate is $m T / N_e$.
This difference in time scaling is substantial, especially for large populations,
and seems to appear because \citet{chevin2014niche} uses the results of \citet{sella2005application}
to describe how the steady-state fitness and substitution process
depends on trait dimensionality
(the fitness peak is harder to find in higher dimensions).
We do not incorporate this effect
because genetic variation may be provided by other sources,
but if we did, it would enter through our $\sigma_S$.

\section*{Acknowledgements}

    We would like to thank Sergey Nuzhdin, Stevan Arnold, Michael Turelli, Patrick Phillips, Erik Lundgren and Hossein Asgharian for valuable discussion. 
    We would also like to thank Nick Barton, Sarah Signor, Todd Parsons, Joachim Hermisson for very helpful comments on the manuscript.
    Work on this project was supported by funds from
    the Sloan Foundation and the NSF (under DBI-1262645) to PR.

\bibliographystyle{plainnat}
\bibliography{krefs}
%\end{multicols}

\normalsize
\appendix

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Genetic drift with a multivariate trait}
\label{ss:quant_gen}

For completeness, we provide a brief exposition of how a population 
evolves due to genetic drift
with a quantitative genetics model,
as in \citet{lande1981models} or \citet{hansen1996translating}.
These do not directly model underlying genetic basis,
but developing a more accurate model is beyond the scope of this paper.

Suppose that the population is distributed in trait space
as a Gaussian with covariance matrix $\Sigma$ and mean $\mu$,
whose density we write as $f(\cdot;\Sigma,\mu)$.
Selection has the effect of multiplying this density by the fitness function and renormalizing,
so that if expected fitness of $x$ is proportional to $\exp(-\|Lx\|^2/2)$,
then the distribution post-selection
has density at $x$ proportional to $f(x;\Sigma,\mu) \exp(-\|Lx\|^2/2)$.
By the computation below (``Completing the square''),
the result is a Gaussian distribution
with covariance matrix $(\Sigma^{-1} + L^T L)^{-1}$ 
and mean $(\Sigma^{-1}+L^T L)^{-1} \Sigma^{-1} \mu$.
% Importantly, if $U$ is not invertible, then by $U^{-1}$ 
% we mean $(U^T U)^{-1} U^T$, Moore-Penrose pseudoinverse of $U$,
% so that $U U^{-1}$ is equal to the projection matrix
% onto the span of the columns of $U$.

After selection, we have reproduction:
suppose this occurs as in the infinitesimal model \citep{barton2017infinitesimal},
so that each offspring of parents with traits $x$ and $y$
is drawn independently from a Gaussian distribution 
with mean $(x+y)/2$ and covariance matrix $R$.
Here, $R$ is the contribution of ``segregation variance'',
i.e., random choices of parental alleles.
If $\widetilde \Sigma = (\Sigma^{-1} + L^T L)^{-1}$ 
is the covariance matrix of the parents post-selection,
then the distribution of offspring will again be Gaussian,
with mean equal to that of the parents
and covariance matrix $\widetilde \Sigma/2 + R$.

In summary, a generation under this model modifies the mean ($\mu$)
and covariance matrix ($\Sigma$) of a population as follows:
\begin{align*}
    \mu &\mapsto \mu' = (\Sigma^{-1}+L^T L)^{-1} \Sigma^{-1} \mu \\
    \Sigma &\mapsto \Sigma' = \frac{1}{2} (\Sigma^{-1} + L^T L)^{-1} + R .
\end{align*}
What measures are stable under this transformation?
The condition $\mu = \mu'$ reduces to $\Sigma L^T L \mu = 0$;
if we assume $R$ and therefore $\Sigma$ are of full rank,
then this happens if and only if $\mu$ is in the null space of $L$,
i.e., if $\mu$ lies in a neutral direction.
The condition $\Sigma' = \Sigma$ can also be solved,
at least numerically.
After rearrangement, it reduces to
$\Sigma L^T L \Sigma + (I/2 - R L^T L) \Sigma = R$.
%% I DON'T THINK WE USE THE FOLLOWING:
% We can find a more explicit description
% if we assume that $x^T L^T L x = \sum_{i=1}^k x_i^2$,
% i.e., that selection only cares about the first $k$ coordinates,
% and then with no interactions between traits.
% If so, the condition $\Sigma' = \Sigma$
% can be written in block form as
% \begin{align*}
%     \begin{bmatrix}
%         \Sigma_{11}^2 + (I/2 - R_{11}) \Sigma_{11}
%         & \Sigma_{11} \Sigma_{12} + (I/2 - R_{11})\Sigma_{12} \\
%         \Sigma_{12}^T \Sigma_{11} + \Sigma_{12}^T/2 - R_{12}^T \Sigma_{11}
%         & \Sigma_{22} - R_{12}^T \Sigma_{12} 
%     \end{bmatrix}
%     &=
%     \begin{bmatrix}
%         R_{11} & R_{12} \\
%         R_{12}^T & R_{22}
%     \end{bmatrix} .
% \end{align*}
% The first equation, $\Sigma_{11}^2 + (I/2 - R_{11}) \Sigma_{11} = R_{11}$,
% can be solved with the quadratic formula: 
% $$\Sigma_{11} = (R_{11} - I/2 + Q)/2$$
% for any $Q$ that commutes with $R_{11}$
% and is a solution to $Q^2 = (R_{11} - I/2)^2 + 4 R_{11}$.
% Since we need $\Sigma$ to be positive definite,
% we take the solution with positive eigenvalues.
% Given $\Sigma_{11}$, the remaining components are
% \begin{align*}
%     \Sigma_{12} 
%     &= 
%     \left( \Sigma_{11} + I/2 - R_{11} \right)^{-1} R_{12} \\
%     \Sigma_{22}
%     &=
%     R_{12}^T \Sigma_{12} + R_{22} .
% \end{align*}
Importantly, the mean $\mu$ does not affect 
either how the covariance matrix moves,
or its stable shape.

Above we have described the \emph{expected} motion of the mean and covariance.
However, random resampling will introduce noise.
Suppose that a population of $N$ individuals
behaves approximately as described above.
By the above,
we may expect that the covariance matrix stays close to a constant value $\Sigma$,
computed from $R$ and $L$ as above,
so that we need only consider motion of the mean, $\mu$.
Since we take a sample of size $N$ to construct the next generation,
the next generation's mean is drawn from a Gaussian distribution with mean $\mu'$
and covariance matrix $\Sigma/N$.
Defining $\Gamma = (I - (I + \Sigma L^T L)^{-1})$,
this can be written as
\begin{align*}
    \mu' - \mu = \Gamma \mu + \epsilon/\sqrt{N} ,
\end{align*}
where $\epsilon$ is a multivariate Gaussian with mean zero and covariance matrix $\Sigma$.
Let $\mu(k)$ denote the mean in the $k^\text{th}$ generation,
and suppose that $\mu$ differs from optimal by something of order $1/\sqrt{N}$:
if $\nu(t) = \sqrt{N} \mu(t\sqrt{N})$ is the rescaled process,
then the previous equation implies that as $N \to \infty$, 
in the limit $\nu$ solves the It\^{o} equation
\begin{align*}
    d \nu(t) = \Gamma \nu(t) dt + \Sigma^{1/2} dW(t) ,
\end{align*}
where now $W(t)$ is a multivariate white noise.
This has an explicit solution as a multivariate Ornstein-Uhlenbeck process:
\begin{align*}
    \nu(t) = e^{-t \Gamma} \nu(0) + \int_0^t e^{-(t-s) \Gamma} \Sigma^{1/2} dW(s).
\end{align*}
% Note that after some algebra, $\Gamma = \Sigma L^T L (I - \Sigma L^T L)^{-1}$.
The asymptotic variance of this process in the direction $z$ is
\begin{align} \label{eqn:limiting_cov}
    \lim_{t \to \infty} \var[\nu(t) \cdot z]
    &=
    \int_0^\infty z^T e^{-s \Gamma} \Sigma e^{-s \Gamma} z ds ,
\end{align}
which is infinite if and only if $\Gamma z = 0$,
which occurs if and only if $Lz=0$.
This implies that at equilibrium, 
population mean trait values lie away from the optimal set
by a Gaussian displacement of order $1/\sqrt{N}$
with a covariance matrix given by equation \eqref{eqn:limiting_cov}.

\paragraph{Completing the square}
First note that if $A$ is symmetric,
\begin{align*}
    (x-y)^T A (x-y)
    &=
    x^T A \left( x - 2y \right) + y^T A y ,
\end{align*}
and so if $B$ is also symmetric and $A+B$ is invertible,
\begin{align*}
    (x-y)^T A (x-y) + x^T B x
    &=
    x^T (A + B) \left( x - 2 (A + B)^{-1} A y \right) + y^T A y \\
    &=
    \left( x - (A + B)^{-1} A y \right)^T
    (A + B)
    \left( x - (A + B)^{-1} A y \right) \\
    &\qquad{}
    + y^T A y - y^T A^T (A+B)^{-1} A y .
\end{align*}
Therefore, 
by substituting $A=\Sigma^{-1}$ and $B=L^T L$,
\begin{align*}
    \frac{ f(x;\Sigma,y) \exp(-x^T L^T L x / 2)}{\int f(z;\Sigma,y) \exp(-z^T L^T L z / 2) dz}
    &=
    f(x; (\Sigma^{-1} + L^T L)^{-1}, (\Sigma^{-1}+L^T L)^{-1} \Sigma^{-1} y) .
\end{align*}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Gaussian load}
\label{apx:gauss_load}

Suppose that a population has a Gaussian distribution in $d$-dimensional trait space
with mean $\mu$ and covariance matrix $\Sigma$,
and that fitness of an individual at $x$ is $\exp(-\|Lx\|^2/2)$.
Then, completing the square as above with $A=\Sigma^{-1}$, $y=\mu$, and $B=L^T L$,
and defining $Q = (\Sigma^{-1} + L^T L)^{-1}$,
\begin{align*}
    &
    \frac{1}{\sqrt{2 \pi}^n \det(\Sigma)^{1/2}} 
        \int e^{-\frac{1}{2} x^T \Sigma^{-1} x} e^{-\frac{1}{2} x^T L^T L x} dx \\
    &\qquad =
    \frac{1}{\sqrt{2 \pi}^n \det(\Sigma)^{1/2}} 
        \int e^{-\frac{1}{2}(x-Q \Sigma^{-1} \mu)^T Q^{-1} (x-Q \Sigma^{-1} \mu)} dx 
        \\ &\qquad \qquad {}
        \times e^{\mu^T \left( I - \Sigma^{-1} Q \right) \Sigma^{-1} \mu} \\
    &\qquad =
    \sqrt{\frac{\det(Q)}{\det(\Sigma)}}
        \exp\left\{\mu^T \left( I - \Sigma^{-1} Q \right) \Sigma^{-1} \mu\right\} \\
    &\qquad =
    \sqrt{\frac{1}{\det(\Sigma)\det(\Sigma^{-1}+L^TL)}}
        \exp\left\{\mu^T \left( I - (I + L^T L \Sigma)^{-1} \right) \Sigma^{-1} \mu\right\} \\
\end{align*}
This is the product of two terms:
the first gives the drop in fitness due to segregation variance,
and the second the drop due to a shift in mean away from the optimum.
A similar decomposition appears in \citet{barton2001role} and \citet{chevin2014niche},
but for mean log fitness.

Now suppose that $\Sigma = \sigma^2 I$ and $L = I/\gamma$.
Then,
\begin{align*}
    \sqrt{\frac{1}{\det(\Sigma)\det(\Sigma^{-1}+L^TL)}}
    &=
    \sqrt{\frac{1}{\sigma^{2d} (1/\sigma^2 + 1/\gamma^2)^d}} \\
    &=
    \frac{1}{(1+ \left(\sigma/\gamma\right)^2)^{d/2}} .
\end{align*}
Also, 
\begin{align*}
        \left( I - (I + L^T L \Sigma)^{-1} \right) \Sigma^{-1} 
        &=
        \frac{1}{\sigma^2}\left( 1 - (1 + (\sigma/\gamma)^2)^{-1} \right) I \\
        &=
        \frac{1}{\gamma^2}\frac{1}{(1 + (\sigma/\gamma)^2)} I \\
\end{align*}

%%%%%%%%%%%
\section{Evolution of segregation covariance}
\label{apx:seg_cov}

The description above does not describe how \emph{two} diverging populations interact,
since the amount of \emph{segregation variance}, quantified by $R$,
will not stay constant.
This has been quantified in various models before
\citep[e.g.,][]{slatkin1994segregation,barton2001role,chevin2014niche};
we provide the following calculations for completeness.

To get an idea of how segregation variance might evolve,
suppose that a trait is determined by $L$ unlinked, biallelic loci,
and that the $i^\mathrm{th}$ locus has two alleles with additive effects $\pm \alpha_i$,
so that being homozygous for the $+$ allele contributes $+2 \alpha_i$ to the trait.
For simplicity, we will neglect the effects of selection.
If the $+$ allele at locus $i$ is at frequency $p_i$ in a population,
then the mean and genetic variance of the trait in a diploid population with random mating is
\begin{align*}
    m &= 2 \sum_i (2 p_i - 1) \alpha_i \\
    s^2 &= 4 \sum_i p_i (1-p_i) \alpha_i^2 .
\end{align*}

Segregation variance between two parents depends on the loci at which either are heterozygous,
and each locus contributes independently
since alleles are additive.
If the alleles are at Hardy-Weinberg proportions,
then since segregation acts like a fair coin flip, a heterozygous locus contributes $\alpha_i^2/4$ to the variance,
and so the \emph{mean} segregation variance, averaging across parents, is
% two parents, prob 2p(1-p) for prob the locus is heterozygous, variance 0.5(1-0.5)\alpha^2.
\begin{align*}
    R_0(p) = \sum_i p_i (1-p_i) \alpha_i^2 .
\end{align*}

On the other hand,
if the second parent came from a distinct population
with frequencies $q_i$
(an $F_1$ hybrid),
this would be
\begin{align*}
    R_1(p,q) &= \frac{1}{2} \sum_i p_i^2 (1-p_i)^2 \alpha_i^2 
                + \frac{1}{2} \sum_i q_i^2 (1-q_i)^2 \alpha_i^2 \\
                &= (R_0(p) + R_0(q))/2 .
\end{align*}
If we assume that the populations are at equilibrium, $R_0(p) \approx R_0(q)$,
and so $R_1(p,q) \approx R_0(p)$.

Now consider an $F_2$ hybrid, where both parents are $F_1$
and so each heterozygous at locus $i$ with probability $p_i (1-q_i) + (1-p_i) q_i$.
Then
\begin{align*}
    R_2(p,q) &= \frac{1}{2} \sum_i \left(p_i (1-q_i) + (1-p_i) q_i \right) \alpha_i^2  .
\end{align*}
Suppose that the two populations are slightly drifted from each other,
with frequency difference $p_i-q_i = 2\epsilon_i$.
Then,
\begin{align*}
    p (1-q) + p (1-q)
    &=
    (u+\epsilon) (1-u+\epsilon) + (u-\epsilon) (1-u-\epsilon) \\
    &=
    2 u (1-u)
    % + \epsilon (u + (1-u) - u - (1-u))
    + 2 \epsilon^2 .
\end{align*}
If the frequencies have evolved neutrally in unconnected, Wright-Fisher populations 
of effective size $N$ for $t$ generations from a common ancestor with allele frequency $u$,
then $\epsilon$ has mean zero and variance roughly $2 u (1-u) t/N$.
Still assuming the populations are at stationarity,
so that $R_0$ is constant between the two,
and taking the frequencies $p_i$ as a proxy for the ancestral frequencies $u_i$,
this implies that we expect
\begin{align*}
    R_2 
    &\approx 
    R_0 + 2 \sum_i p_i (1-p_i) \alpha_i^2 t/N \\
    &=
    \left( 1 + \frac{2t}{N} \right) R_0 .
\end{align*}
On the other hand, the expected squared difference in trait \emph{means} between two such populations is
\begin{align} \label{eqn:var_dx}
    8 \sum_i p_i (1-p_i) \alpha_i^2 t / N = 8 R_0 t /N .
\end{align}
This implies that under this model,
segregation variance in $F_2$ hybrids between two populations
is roughly increased by a factor of $1/8$ of the difference between their means.

% reviews here if toggled above
% \ifsubmission\processdelayedfloats\fi

\includereviews

\end{document}

### 2018-08-16 review-responses.tex
started on responses

@@
 \begin{point}{Format:}
-    please add line numbers to your manuscript; it is straightforward
+    Please add line numbers to your manuscript; it is straightforward
 with the lineno package in LaTeX.  Please also do not increase text width too
@@
 \reply{
+    Ah, good point. We've removed it. \revref
 }
@@
 \reply{
+    Fixed. \revref
 }
@@
 \reply{
+    We've added a clarifying note \revref
+    but we're hoping that readers not familiar with linear algebra will realize
+    that they can skip this ``note on implementation'' without loss of continuity.
 }
@@
 \reply{
+    Good suggestion; added. \revref
 }
@@
 (e.g., what you add, what they do and you do not). In addition to R1's
-suggestions, I add a few more below.
+suggestions, consider \citet{weinreich2013fishers} and \citet{blanquart2016epistasis}.
+% Weinreich, D.
+M.
+and Knies, J.
+L.
+(2013), FISHER'S GEOMETRIC MODEL OF ADAPTATION MEETS THE FUNCTIONAL SYNTHESIS: DATA ON PAIRWISE EPISTASIS FOR FITNESS YIELDS INSIGHTS INTO THE SHAPE AND SIZE OF PHENOTYPE SPACE.
+Evolution, 67: 2957-2972.
+doi:10.1111/evo.12156
+% François Blanquart and Thomas Bataillon Epistasis and the Structure of Fitness Landscapes: Are Experimental Fitness Landscapes Compatible with Fisher’s Geometric Model?
+GENETICS Early online April 6, 2016
 \end{point}
@@
 \reply{
+    This part has been removed.
 }
@@
 \reply{
-}
-
-\begin{point}{Additional references:}
-Weinreich, D.
-M.
-and Knies, J.
-L.
-(2013), FISHER'S GEOMETRIC MODEL OF ADAPTATION MEETS THE FUNCTIONAL SYNTHESIS: DATA ON PAIRWISE EPISTASIS FOR FITNESS YIELDS INSIGHTS INTO THE SHAPE AND SIZE OF PHENOTYPE SPACE.
-Evolution, 67: 2957-2972.
-doi:10.1111/evo.12156
-
-François Blanquart and Thomas Bataillon
-Epistasis and the Structure of Fitness Landscapes: Are Experimental Fitness Landscapes Compatible with Fisher’s Geometric Model?
-GENETICS Early online April 6, 2016
-\end{point}
-
-\reply{
+    This part has been removed.
 }
@@
 but neutral drift of a population along a high-fitness ridge in an epistatic
-landscape (conceptual figure 4), right? If two populations in alloparty drift
+landscape (conceptual figure 4), right? If two populations in allopatry drift
 in different directions, this can lead to hybrid incompatibilities, which are
@@
 \reply{
+    \plr{I'm leaving these points because they might apply to whatever handwaving 
+    we replace the quant gen stuff with.}
 }
@@
 \reply{
+    Good point -- clearly some assumptions jointly on the long-time behavior of $h(t)$ and $\rho$ 
+    are required for $D$ to be finite, but just saying that $\rho$ is square integrable doesn't suffice;
+    we've removed the caveat. 
+    \revref
+    (It is always well-defined, anyhow, since the integrand is nonnegative.)
 }
@@
     Finally, it would be worth mentioning several old and recent works relating genetic drift
-to speciation : Yamagushi and Iwasa, several articles by Gavrilets et al. (I think several
+to speciation: Yamagushi and Iwasa, several articles by Gavrilets et al. (I think several
 citations are missing here), and Mirò Pina and Schertzer.

### 2018-08-16 kalman_walk_paper.tex
started on responses

@@
 Under quite general assumptions,
-we can write the phenotype as
+we can write the phenotype as \revpoint{AE}{4}
   \begin{align}
-    \phi(t) = C e^{A t} \kappa(0) + \int_{0}^{t} C e^{A (t-s)} B u(s) ds ,
+    \phi(t) = \int_{0}^{t} C e^{A (t-s)} B u(s) ds ,
   \end{align}
@@
  %    \end{align*}
-Suppose the input is an impulse at time zero (a delta function), and so its phenotype is equal to its impulse response,
+If the input is an impulse at time zero (a delta function), 
+then the phenotype is equal to the impulse response: \revpoint{AE}{5}
 \begin{align*}
@@
 which we denote by $\reachable$,
-is defined to be the closure of $\spn(B)$ under applying $A$,
-% (or equivalently, the span of $B, AB, A^2B, \ldots A^{n-1}B$).
+is defined to be the closure of $\spn(B)$ under applying $A$
+(or equivalently, the span of $B, AB, A^2B, \ldots A^{n-1}B$), \revpoint{AE}{6}
 % Analogously to this, we define
@@
 if the parents have different underlying systems.
-%This leads to the question: 
 How quickly do hybrid phenotypes break down as genetic distance between parents increases?
@@
 using a weighted difference between impulse response functions.
-Suppose that $\rho(t)$ is a nonnegative, smooth, square-integrable weighting function,
+Suppose that $\rho(t)$ is a nonnegative weighting function, \revpoint{2}{3}
 $h_0(t)$ is the \emph{optimal} impulse response function
@@
 \end{align}
+In practice, we take $\rho(t) = \exp(-t/4\pi)$,
+so that fitness is determined by the dynamics of the system over a few multiples of $2 \pi$,
+but not longer. \revpoint{AE}{8}
 Consider reproduction between a parent with system $(A, B, C)$ 

### 2018-08-18 review-responses.tex
minor changes

@@
 \begin{point}{all figures:}
-    please ensure that all axes legends and labels are big enough
+    Please ensure that all axes legends and labels are big enough
 and not distorted (same of arrow labels on the diagrams). Please also make sure
@@
 \begin{point}{(mechanistic models)}
-    you should probably mention somewhere that mechanistic models of neutral
-  spaces exist in the (quite different) context of RNA folding (papers by
-Hofacker, Fontana, etal)
+    You should probably mention somewhere that mechanistic models of neutral
+  spaces exist in the (quite different) context of RNA folding (papers by Hofacker, Fontana, etal)
 \end{point}
@@
 \reply{
+    We have added discussion of this in the Discussion. \revref
 }
@@
 \reply{
+    Yes, that's right. \revref
 }
@@
 \begin{point}{Introduction:}
-    ``It is not a new observations that there is often more than one way to do the same thing, and that this may lead to speciation'' - Isn't this what people would call a neutral Dobzhansky-Muller incompatibility?
+    ``It is not a new observation that there is often more than one way to do the same thing, and that this may lead to speciation'' - Isn't this what people would call a neutral Dobzhansky-Muller incompatibility?
 \end{point}
@@
 \reply{
+    This sentence has been removed.
 }
@@
 \reply{
+    XXX add discussion of equivalent oscillator examples XXX
 }

### 2018-08-18 kalman_walk_paper.tex
minor changes

@@
 It has also been shown that population structure
-can allow long-term stable coexistence of incompatible genotypes encoding identical phenotypes \citep{phillips1996maintenance}.
+can allow long-term stable coexistence of incompatible genotypes encoding identical phenotypes \citep{phillips1996maintenance}. \revpoint{1}{9}
 However, previous simulations of system drift in regulatory sequences \citep{tulchinsky2014hybrid}
@@
 We think of the system as the triple $(A,B,C)$,
-which translates (time-varying) $m$-dimensional input $u(t)$ into the $\ell$-dimensional output $\phi(t)$.
+which translates (time-varying) $m$-dimensional input $u(t)$
+into the $\ell$-dimensional output $\phi(t)$.
 Under quite general assumptions,
@@
 
+\paragraph{The dimensionality of trait space} \revpoint{1}{8}
+We have focused in examples on single traits (where the phenotype is one-dimensional),
+but phenotypes under selection are often high-dimensional,
+and different traits often share a genetic basis.
+If the phenotype and kryptotype are of the same dimension,
+then we do not in general have degrees of freedom, i.e., neutral directions.
+The true ``dimensionality of trait space'' has been debated at some length in the literature
+\citep{trait_space}.
+Whatever it is, however, it seems unlikely 
+that it would approach the dimensionality of kryptotype space,
+which is something related to the number of genes.
+Even under universal pleiotropy \citep{universal_pleiotropy,omnigenic},
+XXX
+
 \paragraph{Nonlinearity and model assumptions}

### 2018-08-19 review-responses.tex
some more reviews; start at the handwaving

@@
 \reply{
-    XXX add discussion of equivalent oscillator examples XXX
+    Biologically, a coordinate change may produce an unrecognizably different system
+    (repression changed to activation or the like);
+    we've added a hopefully clarifying sentence. \revref
 }
@@
 \reply{
+    Good point; we now say ``in block matrix form'' and add more explanation. \revref
 }
@@
 \reply{
+    Good point; added that caveat. \revref
 }
@@
 \reply{
+    XXX Suggestion: change this figure to have two figures of the same type, side-by-side,
+    one with a simple genetic basis (like this one, but with two purple clouds)
+    and another with a more quantitative trait.
 }
@@
 \reply{
+    We have put back in the somewhat lengthy Appendix on this subject 
+    that we'd removed before submission. \revref
+    XXX TODO EDIT THAT APPENDIX XXX
 }
@@
 \reply{
+    We've added ``homozygous'' to the caption,
+    and mentioned that there are $3^4 = 81$ possible $F_2$s -- 
+    this is because it only matters if, at each of the four matrix entries, 
+    the offspring is heterozygous, homozygous for parent 1, or homozygous for parent 2.
 }

### 2018-08-19 kalman_walk_paper.tex
some more reviews; start at the handwaving

@@
   \end{equation}
+These ``changes of coordinates'' are not simply different ways of looking at the same system --
+if each dimension of the kryptotype 
+corresponds to the concentration of a particular transcription factor,
+changing $A$ corresponds to changing the strengths of regulatory interactions.
+We will even see below that interactions may change sign. \revpoint{1}{13}
 However, not all phenotypically equivalent systems are of this form:
@@
 gives a change of basis $P$ such that
-the transformed system $(PAP^{-1},PB,CP^{-1})$  has the following form:
+the transformed system $(PAP^{-1},PB,CP^{-1})$
+can be written in block matrix form:
 \begin{align*}
@@
 \end{align*}
-% and
-% \begin{align*}
-%    CP^{-1}
-%    &=
-%    \left[ \begin{array}{cccc}
-%        0 & C_{\ro} & C_{\nrno} & 0 
-%    \end{array} \right] .
-%\end{align*}
+The $n$-dimensional system has been divided into subspaces
+of dimensions $n_\rno + n_\ro + n_\nrno + n_\nro = n$,
+and so, for instance, $A_{\rno}$ is the $n_\rno \times n_\rno$ square matrix 
+in the top left corner of $P A P^{-1}$. \revpoint{1}{15}
 The impulse response of the system is given by
@@
 First, we must specify how sexual reproduction acts on these systems.
-Suppose that each of a diploid organisms' two genomes encodes a set of system coefficients.
+Suppose that each of a diploid organisms' two genomes encodes a set of system coefficients
+with the same kryptotype dimension. \revpoint{2}{1}
 We assume that a diploid which has inherited systems $(A', B', C')$ and $(A'', B'', C'')$ from its two parents
@@
 and their offspring has impulse response function $h_\epsilon(t)$.
-A Taylor expansion of $D(h_\epsilon)$ in $\epsilon$ % is explicitly worked out in Appendix \ref{apx:H_calc}, and
+A Taylor expansion of $D(h_\epsilon)$ in $\epsilon$ 
+is explicitly worked out in Appendix \ref{apx:H_calc}, and \revpoint{2}{4}
 shows that the phenotype of an $F_1$ hybrid between these two is at distance proportional to $\epsilon^2$ from optimal,
@@
 
-%% PLR: it says this above already.
-% The coefficients in $A$ -- i.e., the regulatory coefficients --
-%   differ between parents by only a few percent (around 0.5\% for $\epsilon = - \sfrac{1}{100}$ and 5\% for $\epsilon= - \sfrac{1}{10}$).
-% This is well within the amount of regulatory coefficient variation we expect to find segregating within real populations
-% (discussed further below). 
-% For these small values of $\epsilon$, hybrid phenotypes remain relatively stable, 
-%   consistent with the idea that natural selection will allow such intrapopulation variation.
-For larger values of $\left| \epsilon \right|$ (here $\sfrac{1}{2}$), hybrid phenotypes can become dysfunctional, and depending on context, manifest as hybrid inviability.
-
-%%% figure of tau=0 and tau=2 hybrid:
-%  \begin{figure}[H] 
-%    \centering
-%    \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{expF1}
-%    \caption{\textbf{Hybrid phenotypic breakdown} $F_1$ hybrid (orange) and parental (blue) phenotypic cell-cycle control (oscillator) dynamics %for an $\Sys(0)$ by $\Sys(2)$ cross from Example \ref{ex:all_osc}.
-The $F_1$ hybrid
-%$\left(A_{F_1}\!\!=\!\!\tfrac{A\!(\!0\!) \! + \! A\!(\!2\!)}{2}\right)$
-%gene expression fails to oscillate and instead grows exponentialy
-%($\phi_{F_0}\! = \! \sin t \! + \! \cos t$, however $\phi_{F_1} \! = \! e^t$).
-%    }
-%    \label{fig:expF1}
-%  \end{figure}
-
 \begin{figure}[H]
@@
   \caption{
-    \textbf{(left)} Phenotypes of $F_1$ hybrids between an $A(0)$ parent and, top-to-bottom, an $A(\sfrac{1}{100})$, an $A(\sfrac{1}{10})$, and $A(\sfrac{1}{2})$ parent;
+    \textbf{(left)} Phenotypes of $F_1$ hybrids between a homozygous $A(0)$ parent and, 
+    top-to-bottom, homozygous $A(\sfrac{1}{100})$, $A(\sfrac{1}{10})$, and $A(\sfrac{1}{2})$ parents;
     parental coefficients differ by around 0.5\%, 5\%, and 25\% respectively.
-    %\plr{These figures are fine; but reparameterize so that the values of $\tau$ are positive.}
     Parental phenotypes ($\sin t + \cos t$) are shown in solid black, and hybrid phenotypes in dashed blue.
-    \textbf{(right)} Phenotypes of all possible $F_2$ hybrids between the same set of parents,
+    \textbf{(right)} Phenotypes of all $3^4 = 81$ possible $F_2$ hybrids between the same set of parents,
     with parental phenotype again in black.
@@
     \centering
-    %\begin{tabular}{cc}
-    %\includegraphics[width=0.4\textwidth]{figures/f1_quartic2} &
-    %\includegraphics[width=0.4\textwidth]{figures/f2_quad2}
-    %\end{tabular}
     \includegraphics[width=0.5\textwidth]{examples/F2_vs_F1_divergence_tau0}
-   % \caption{
-    %\textbf{(left)} 
     \caption{Mean hybrid phenotypic distance from optimum computed with equation \eqref{eqn:distance}, using $\rho(t) = \exp(-t/4\pi)$
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection*{The speed of speciation}
 
-XXX Talk about how drift goes like variance over $N_e$.
-
+We have shown that system drift can lead to speciation in principle,
+but is it rapid enough to be an important factor in practice?
+In other words, after what period of time would we expect the fitness of hybrids
+between two allopatric populations to be substantially lower than the parentals?
+The population mean of an unconstrained quantitative trait
+with genetic variance $V_G$ in a population with effective size $N_e$
+will move in $t$ generations a random amount whose variance is $t V_G / N_e$ \citep{quant_gen}.
+The mean difference between two such populations has twice the variance.
+Although this mean difference is along neutral directions,
+we would in many cases expect the range of variation among $F_2$s
+in \emph{all} directions to be of the same order
+as the differences between the populations, 
+as depicted in Figure \ref{fig:conceptual_fig}.
+This suggests that, naively, two such populations that have been separated for $t$ generations
+will produce $F_2$ offspring that differ from optimal by an amount proportional to $\sqrt{t V_G / N_e}$.
+Since we assume they are at a local figness optimum, without much loss of generality we can assume
+that fitness is locally quadratic, and so $F_2$ fitness decays linearly in time:
+proportionally to $t V_G / N_e$ -- fastest in small, diverse populations.
+
+It is useful to think in more detail about the assumptions in the rough argument above.
+The key aspect is how population differences in neutral directions
+(along the fitness ridge)
+translate to segregation variance in $F_2$s in selectively constrained directions.
+To move the the system (the $A$ matrix) a given distance
+generally involves moving many individual interaction coefficients (the entries $A_{ij}$).
+The movements must be coordinated, for the population to stay near the fitness ridge.
+However, mixing elements between systems that have made independent sets of coordinated changes
+to remain on the fitness ridge
+is unlikely to produce a a set of coordinated changes;
+and the resulting system could move away from the ridge in almost any direction.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection*{Genetic variation in empirical regulatory systems}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Local expansion of the fitness surface}
+\label{apx:H_calc}
+
+Suppose that $\rho(t) \ge 0$ is a weighting function on $[0,\infty)$
+so that fitness is a function of $L^2(\rho)$ distance of the impulse response from optimal.
+With $h_0(t) = C_0 e^{tA_0} B_0$ a representative of the optimal set:
+\begin{equation}
+    \begin{aligned}
+        D(A, B, C)^2
+        &:= 
+        \int_0^\infty \rho(t) \left| h_A(t) - h_0(t) \right|^2 dt \\
+        &:= 
+        \int_0^\infty \rho(t) \left| C e^{At} B - C_0 e^{A_0 t} B_0 \right|^2 dt \\
+        &= 
+        \int_0^\infty \rho(t) \tr\left\{
+            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)^T
+            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)
+        \right\} dt \\
+        &= 
+        \int_0^\infty \rho(t) \tr\left\{
+            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)
+            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)^T
+        \right\} dt  ,
+        % &= 
+        % \int_0^\infty \rho(t)  \tr\left\{
+        %         C e^{At} B B^T e^{A^T t} C^T 
+        %         - C_0 e^{A_0t} B_0 B^T e^{A^T t} C^T 
+        %         \right. \\ &\qquad \qquad \left. {}
+        %         - C e^{At} B B_0^T e^{A_0^T t} C_0^T 
+        %         + C_0 e^{A_0t} B_0 B_0^T e^{A_0^T t} C_0^T 
+        % \right\} dt \\
+        % \int_0^\infty \rho(t) \left| C \left( e^{At} - e^{A_0 t} \right) B \right|^2 dt \\
+        % &= 
+        % \int_0^\infty \rho(t) C \left( e^{At} - e^{A_0 t} \right) B B^T \left( e^{At} - e^{A_0 t} \right)^T C^T dt
+    \end{aligned}
+\end{equation}
+where $\tr X$ denotes the trace of a square matrix $X$.
+How does this change as we perturb about $(A_0, B_0, C_0)$?
+First we differentiate with respect to $A$, keeping $B=B_0$ and $C=C_0$ fixed.
+Since
+\begin{equation}
+  \begin{aligned}
+      \frac{d}{du} e^{(A+uZ)t} \vert_{u=0}
+      &=
+      \int_0^t e^{As} Z e^{A(t-s)} ds, 
+  \end{aligned}
+\end{equation}
+we have that
+\begin{equation}
+  \begin{aligned}
+      \frac{d}{du} D(A+uZ,B_0,C_0)^2 \vert_{u=0}
+      &=
+        2 \int_0^\infty \rho(t) \tr\left\{ C_0 \left( \int_0^t e^{As} Z e^{A(t-s)} ds \right) B_0 B_0^T \left( e^{At} - e^{A_0 t} \right)^T C_0^T \right\} dt \\
+      &=
+        2 \int_0^\infty \rho(t) \tr\left\{ C_0 \left( \int_0^t e^{As} Z e^{A(t-s)} ds \right) B_0 \left( h_A(t) - h_0(t) \right)^T \right\} dt 
+  \end{aligned}
+\end{equation}
+and, by differentiating this and supposing that $A$ is on the optimal set,
+i.e., $h_A(t)=h_0(t)$, (so without loss of generality, $A=A_0$):
+\begin{equation}
+  \begin{aligned}
+      \calH^{A,A}(Y,Z) 
+      &:= 
+      \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0+uY+vZ,B_0,C_0)^2 \vert_{u=v=0} \\
+      &=
+        \int_0^\infty \rho(t) \tr\left\{ C_0 
+        \left( \int_0^t e^{A_0 s} Y e^{A_0 (t-s)} ds \right) 
+        B_0 B_0^T 
+        \left( \int_0^t e^{A_0 s} Z e^{A_0 (t-s)} ds \right)^T
+        C_0^T \right\} dt  .
+  \end{aligned}
+\end{equation}
+
+The function $\calH$ will define a quadratic form.
+To illustrate the use of this, suppose that $B$ and $C$ are fixed.
+% Here $\calH$ is the quadratic form underlying the Hamiltonian.
+By defining $\Delta_{ij}$ to be the matrix with a 1 in the $(i,j)$th slot
+and 0 elsewhere,
+the coefficients of the quadratic form is
+\begin{equation}
+    \begin{aligned}
+        H_{ij, k\ell}(A)
+        &:=
+        \calH(\Delta_{ij}, \Delta_{k\ell}) .
+    \end{aligned}
+\end{equation}
+
+We could use this to get the quadratic approximation to $D$ near the optimal set.
+To do so, it'd be nice to have a way to compute the inner integral above.
+Suppose that we diagonalize $A = U \Lambda U^{-1}$.
+Then
+\begin{equation} \label{eqn:exp_deriv}
+  \begin{aligned}
+      \int_0^t e^{As} Z e^{A(t-s)} ds 
+      &=
+      \int_0^t U e^{\Lambda s} U^{-1} Z U e^{\Lambda (t-s)} U^{-1} ds 
+  \end{aligned}
+\end{equation}
+Now, notice that
+\begin{equation}
+  \begin{aligned}
+      \int_0^t e^{s \lambda_i} e^{(t-s) \lambda_j} ds
+      &=
+      \frac{ e^{t \lambda_i} - e^{t \lambda_j} }{ \lambda_i - \lambda_j } 
+          \qquad & \text{if} \quad i \neq j \\
+      &=
+          t e^{t \lambda_i} 
+          \qquad & \text{if} \quad i = j 
+  \end{aligned}
+\end{equation}
+Therefore, 
+defining
+\begin{equation}
+    \begin{aligned}
+    X_{ij}(t,Z) 
+       &= 
+        \left( U^{-1} Z U \right)_{ij}
+      \frac{ e^{t \lambda_i} - e^{t \lambda_j} }{ \lambda_i - \lambda_j } 
+          \qquad & \text{if} \quad i \neq j \\
+      &=
+          \left( U^{-1} Z U \right)_{ii}
+          t e^{t \lambda_i} 
+          \qquad & \text{if} \quad i = j 
+    \end{aligned}
+\end{equation}
+moving the $U$ and $U^{-1}$ outside the integral and integrating we get that
+\begin{equation}
+  \begin{aligned}
+      \int_0^t e^{As} Z e^{A(t-s)} ds 
+      &=
+      U X(t,Z) U^{-1} .
+  \end{aligned}
+\end{equation}
+This implies that
+\begin{equation}
+    \begin{aligned}
+        D(A_0+\epsilon Z)^2
+        &\approx \frac{1}{2} \epsilon^2 
+        \int_0^\infty
+            \rho(t) \tr\left\{ C U X(t,Z) U^{-1} B B^T (U^{-1})^T X(t,Z)^T U^T C^T \right\}
+        dt .
+    \end{aligned}
+\end{equation}
+
+To compute the $n^2 \times n^2$ matrix $H$,
+we see that if $Z=\Delta_{k \ell}$, then
+\begin{equation}
+  \begin{aligned}
+      X_{ij}^{k\ell}(t) 
+      &= 
+      (U^{-1})_{\cdot k} U_{\ell \cdot}
+      \frac{ e^{t \lambda_i} - e^{t \lambda_j} }{ \lambda_i - \lambda_j } 
+          \qquad & \text{if} \quad i \neq j \\
+      &=
+      (U^{-1})_{\cdot k} U_{\ell \cdot}
+      t e^{t \lambda_i} 
+          \qquad & \text{if} \quad i = j 
+  \end{aligned}
+\end{equation}
+where $U_{k \cdot}$ is the $k$th row of $U$,
+and so
+\begin{equation}
+    \begin{aligned}
+        H_{ij, k\ell}(A)
+        &=
+        \int_0^\infty
+            \rho(t) \tr\left\{ C U X^{ij}(t) U^{-1} B B^T (U^{-1})^T X^{k\ell}(t)^T U^T C^T \right\}
+        dt .
+    \end{aligned}
+\end{equation}
+This implies that
+\begin{equation}
+    \begin{aligned}
+        D(A_0+\epsilon Z)^2
+        &\approx \frac{1}{2} \epsilon^2 \sum_{ijk\ell} H_{ij,k\ell}(A_0) Z_{ij} Z_{k\ell}  .
+    \end{aligned}
+\end{equation}
+
+By section \ref{ss:quant_gen},
+if we set $\Sigma=\sigma^2 I$ and $U=H$,
+then a population at $A_0+Z$ experiences a restoring force of strength
+$(I + \sigma^2 H^{-1})^{-1} Z$ (treating $Z$ as a vector and $H$ as an operator on these).
+If $\sigma^2$ is small compared to $H^{-1}$
+then this is approximately $-\sigma^2 H^{-1} Z$.
+This suggests that the population mean follows an Ornstein-Uhlenbeck process,
+as described (in different terms) in \citet{hansen1996translating}.
+
+More generally, $B$ and $C$ may also change.
+To extend this we need the remaining second derivatives of $D^2$.
+First, in $B$:
+\begin{equation}
+    \begin{aligned}
+        \mathcal{H}^{B,B}(Y,Z)
+        &:= 
+        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0,B_0+uY+vZ,C_0)\vert_{u=v=0} \\
+        &=
+        \frac{1}{2} \int_0^\infty \rho(t) \tr\left\{ 
+        C_0 e^{t A_0}
+        \frac{d}{du} \frac{d}{dv} 
+        (uY+vZ)
+        (uY+vZ)^T
+        \vert_{u=v=0} 
+        e^{t A_0^T} C_0^T 
+        \right\} dt  \\
+        &=
+        \frac{1}{2} \int_0^\infty \rho(t) \tr\left\{ 
+        C_0 e^{t A_0}
+        \left( Y Z^T + Z Y^T \right)
+        e^{t A_0^T} C_0^T 
+        \right\} dt  .
+  \end{aligned}
+\end{equation}
+Next, in $C$:
+\begin{equation}
+    \begin{aligned}
+        \mathcal{H}^{B,B}(Y,Z)
+        &:= 
+        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0,B_0,C_0+uY+vZ)\vert_{u=v=0} \\
+        &=
+        \frac{1}{2} \int_0^\infty \rho(t) \tr\left\{ 
+        B_0 e^{t A_0^T}
+        \frac{d}{du} \frac{d}{dv} 
+        (uY+vZ)^T
+        (uY+vZ)
+        \vert_{u=v=0} 
+        e^{t A_0} B_0
+        \right\} dt  \\
+        &=
+        \frac{1}{2} \int_0^\infty \rho(t) \tr\left\{ 
+        B_0 e^{t A_0^T}
+        \left( Y Z^T + Z Y^T \right)
+        e^{t A_0} B_0
+        \right\} dt  .
+  \end{aligned}
+\end{equation}
+Now, the mixed derivatives in $B$ and $C$:
+\begin{equation}
+    \begin{aligned}
+        \mathcal{H}^{B,C}(Y,Z)
+        &:= 
+        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0,B_0+uY,C_0+vZ)\vert_{u=v=0} \\
+        &=
+        \int_0^\infty \rho(t) \tr\left\{ 
+        Y e^{t A_0^T} C_0^T Z e^{t A_0} B_0
+        \right\} dt  .
+  \end{aligned}
+\end{equation}
+In $A$ and $B$
+\begin{equation}
+    \begin{aligned}
+        \mathcal{H}^{A,B}(Y,Z)
+        &:= 
+        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0+uY,B_0+vZ,C_0)\vert_{u=v=0} \\
+        &=
+        \int_0^\infty \rho(t) \tr\left\{ 
+        C_0 \left(\int_0^t e^{s A_0} Y e^{(t-s) A_0} ds \right) B_0
+        Z^T e^{t A_0} C_0
+        \right\} dt  ,
+  \end{aligned}
+\end{equation}
+and finally in $A$ and $C$:
+\begin{equation}
+    \begin{aligned}
+        \mathcal{H}^{A,C}(Y,Z)
+        &:= 
+        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0+uY,B_0,C_0+vZ)\vert_{u=v=0} \\
+        &=
+        \int_0^\infty \rho(t) \tr\left\{ 
+        C_0 \left(\int_0^t e^{s A_0} Y e^{(t-s) A_0} ds \right) B_0
+        B_0 e^{t A_0} Z
+        \right\} dt  .
+  \end{aligned}
+\end{equation}
+
+Together, numerical computation of these expressions,
+along with estimates of genetic covariance within a population,
+allow precise predictions of evolutionary dynamics of a particular system.
+The approximation should be good as long as the second-order Taylor approximation holds.
+
 % reviews here if toggled above

### 2018-11-09 ridge-review-responses.tex
a bit more

--- 2018-08-16 review-responses.tex
@@
 \begin{minipage}[b]{2.5in}
-  Resubmission Cover Letter \\
+  Cover Letter \\
   {\it Evolution}
@@
 
-We are pleased to submit a revision of our manuscript,
+Explain this was split from the other paper.
 
@@
 \begin{quote}
-I have received the evaluations of two reviewers, and I have read the paper
-myself. First, sorry for the long time it took to review your manuscript; it
-was reviewed on the timescale of mathematicians (for Reviewer2), but both
-reviewers provided thorough evaluations of your work (which in my opinion is
-better than a quick but superficial review).
-
 Your manuscript is quite long already, and the reviewers' suggestions of
@@
 first specific comment about dimensionality.
-
-This may be because I am more familiar with Fisher's geometric model than
-systems biology models, but it seems to me that the first part is more novel,
-and should therefore deserve your attention first, should you follow my
-suggestion of publishing the two parts separately.
 \end{quote}
-
-\begin{point}{Format:}
-    Please add line numbers to your manuscript; it is straightforward
-with the lineno package in LaTeX.  Please also do not increase text width too
-much, as this decreases legibility (LaTeX's default settings already optimize
-the number of characters per line).  Finally, be careful about not confusing
-citep and citet (a lot of missing parentheses around citations in the second part).
-\end{point}
-
-\reply{
-}
-
-\begin{point}{all figures:}
-    please ensure that all axes legends and labels are big enough
-and not distorted (same of arrow labels on the diagrams). Please also make sure
-that the lines are thick enough to be visible.
-\end{point}
-
-\reply{
-}
-
-\begin{point}{Above eq (2):}
-    ``Under quite general assumptions'': quickly mention them?
-\end{point}
-
-\reply{
-}
-
-\begin{point}{eq (2):} 
-    Do you need the first term on the rhs given that you assume that $\kappa(0)=0$?
-\end{point}
-
-\reply{
-    Ah, good point. We've removed it. \revref
-}
-
-\begin{point}{Example 1:}
-    ``and so its phenotype'' whose phenotype? (the subject of the sentence is ``the input'')
-\end{point}
-
-\reply{
-    Fixed. \revref
-}
-
-\begin{point}{p5, Note on implementation}
-    ``the closure of span(B)'' may not be
-    understandable by most of Evolution's readers; please explain what this means.
-\end{point}
-
-\reply{
-    We've added a clarifying note \revref
-    but we're hoping that readers not familiar with linear algebra will realize
-    that they can skip this ``note on implementation'' without loss of continuity.
-}
-
-\begin{point}{Figures 1-3:}
-    In figures 1 and 3, there are two types of arrow heads (+ and -
-effects), but not in Figure 2, probably because the sign of the effect depends
-on the value of $\tau$. Maybe use a  different type of arrow heads for
-interactions that may change signs, to avoid the confusion with positive
-effects?
-\end{point}
-
-\reply{
-}
-
-\begin{point}{p7, Hybrid incompatibility:}
-    Please provide a rationale for choosing a particular weight $\rho$.
-\end{point}
-
-\reply{
-    Good suggestion; added. \revref
-}
 
@@
 \reply{
-    This part has been removed.
 }
 
-\begin{point}{p12 Hybridization:}
+\begin{point}{Hybridization:}
     What is $\mathcal{X}$?
@@
 \reply{
-    This part has been removed.
+    A holdover from previous notation; it is $\allS$ (fixed). \revref
 }
@@
     ``The importance of including neutral directions in these models, which
-is not usually done'' -- Do not line of isofitness correspond to neutral
-directions, and aren't these neutral directions already included in those
-models?
+    is not usually done'' -- Do not lines of isofitness correspond to neutral
+    directions, and aren't these neutral directions already included in those models?
 \end{point}
@@
 \reply{
-    This part has been removed.
+    Good point! We meant neutral directions \emph{within the optimal set},
+    and have reworded this.
 }
@@
 \reviewersection{1}
-
-\begin{quote}
-The ms combines two parts that are only loosely linked: one ``systems biology''
-part and a ``popgen'' part. The combination of both parts is appealing, primarily
-because we lack such combined approaches and it is not easy to come up with an
-adequate framework. Below, I will discuss both parts in turn.
-\end{quote}
-
-\begin{quote}
-General remark: line numbers are very helpful for reviewing.  
-\end{quote}
-
-\begin{quote}
-    \textbf{(I)} The first (systems biology) part makes use of the concept of
-``nonidentifiability'' of chemical reaction networks and develops these concepts
-in the context of a genotype to phenotype map. In particular, this map is
-analyzed for the case where it can be modeled as a linear system.
-\end{quote}
-
-\begin{point}{}
-    First note that the concept of ``nonidentifiability'' is closely related to
-themes that have been discussed in evolutionary biology for a long time, under
-names like ``mutational robustness'', ``network neutrality'', ``canalization'',
-``redundancy'' etc. In particular, if you characterize your model by ``many
-distinct (and mutationally connected) molecular pathways can realize identical
-phenotypes'', note that this essentially describes what has been called a
-``neutral network'' and studied under this name in many articles. Also concrete
-developmental networks have been studied in this context (eg, von Dassow etal
-2000, Nature 406:188–192). Discussion of this literature - and more importantly
-discussion of terms and notions used here relative to terms that have been used
-elsewhere - is largely missing. I won't be able to summarize all this here, but
-an older review is the Evolution Perspective piece ``Evolution and detection of
-genetic robustness'' (de Visser et al 2003). You can work work backward and
-forward from there.  
-\end{point}
-
-\reply{
-}
-
-\begin{point}{}
-    Related to this: I find it non-intuitive to start the introduction with the
-notion of ``nonidentifiability'', which is a consequence rather than a cause, the
-cause being what you later call ``phenotypic equivalence'' and what had been
-discussed under other names in other articles.
-\end{point}
-
-\reply{
-}
-
-\begin{point}{}
-    Still related: You give long lists of references in the introduction, but do
-not provide the reader with any information about specific contributions. A bit
-more would be helpful (starting with True and Haag coined the term
-``developmental system drift'', etc).
-\end{point}
-
-\reply{
-}
-
-\begin{point}{4.}
-    Your ms does not simply assume that there is a large ``neutral network''
-underlying a phenotype, but suggests a mechanistic underpinning to create this
-neutral space. Since much of the appeal of the first part of the paper is
-connected to this fact, I'd like to see some more in-depth discussion.
-\textit{(suggestions to follow)}
-\end{point}
-
-\reply{
-}
-
-\begin{point}{(mechanistic models)}
-    you should probably mention somewhere that mechanistic models of neutral
-  spaces exist in the (quite different) context of RNA folding (papers by
-Hofacker, Fontana, etal)
-\end{point}
-
-\reply{
-}
-
-\begin{point}{(mechanistic models)}
-    For development, Andreas Wagner once suggested a discrete version of a linear
-  model (iterated matrix multiplication). Variants of this have later been used
-by others (eg Draghi and GP Wagner, I think). How is your model related to
-these approaches?
-\end{point}
-
-\reply{
-}
-
-\begin{point}{(mechanistic models)}
-    You model development as a linear system and this assumption is essential
-   for all further steps (the explicit solution of the dynamics and the Kalman
-decomposition). For me, the only justification of a linear model (other than
-mathematical convenience) is local approximation. Do you agree? This will be
-important for the second part of the ms, see below.
-\end{point}
-
-\reply{
-}
-
-\begin{point}{(mechanistic models)}
-    You gain a lot of ``neutrality'' by the assumption that the dimension of the
-  ``krypotype'' is larger than the dimension of the trait. If the trait is the
-expression of a gene at the top of some pathway, this is certainly true.
-However, if we include further traits that are affected by the same genes
-(i.e., pleiotropy) this is no longer clear. The space of ``all traits under
-selection'' of an organism is awfully high-dimensional and pleiotropy is
-wide-spread (even if we do not believe in ``omnigenic'' models). It is not clear
-to me from what kind of data we could learn more about these dimensions and I
-do not expect an answer in your ms, but the issue deserves more discussion.
-Currently, ``pleiotropy'' is not even mentioned in the ms.   
-\end{point}
-
-\reply{
-}
-
-\begin{point}{Introduction:}
-    ``Genotypes encoding identical phenotypes can even persist stably within a species'' - if there is population structure, I suppose.
-\end{point}
-
-\reply{
-}
-
-\begin{point}{Introduction:}
-    ``It is not a new observations that there is often more than one way to do the same thing, and that this may lead to speciation'' - Isn't this what people would call a neutral Dobzhansky-Muller incompatibility?
-\end{point}
-
-\reply{
-}
-
-\begin{point}{Results:}
-    (Eq. 1) development as a linear system: has this been done before (probably yes?) - If so, references?
-\end{point}
-
-\reply{
-}
-
-\begin{point}{Results:}
-    ``Of course, neither of these are necessarily true for real systems'' - delete ``necessarily''?
-\end{point}
-
-\reply{
-}
-
-\begin{point}{Eq. 3:}
-    What does a coordinate change mean biologically?
-In our JC, we discussed for a while whether this is just a different parametrization of the exact same biological thing (like transformation to principle components).
-I think I now understand that it is more, but this should be explained better.
-\end{point}
-
-\reply{
-}
-
-\begin{point}{}
-    ``Since gene networks can grow or shrink following gene duplications and deletions,
-these additional degrees of freedom can apply in principle to any system.'' - are there examples of equivalent gene networks with a different size due to gene duplication or loss?
-\end{point}
-
-\reply{
-}
-
-\begin{point}{Kalman decomposition:}
-    Define the sub-matrices directly when first used.
-\end{point}
-
-\reply{
-}
-
-\begin{point}{Figure 2:}
-    Colors are hardly visible
-\end{point}
-
-\reply{
-}
 
@@
 
-\reply{
-    \plr{I'm leaving these points because they might apply to whatever handwaving 
-    we replace the quant gen stuff with.}
-}
-
-\begin{point}{}
-    Your analysis of the fitness loss in F1 and F2 hybrids rests on a local
-Taylor expansion of the fitness landscape. This is adequate given that the
-underlying linear network is also only locally valid (see above). However, you
-then apply this to a discussion of hybrid incompatibility (Haldane's rule etc).
-We are thus interested in *very large* fitness costs for hybrids. This does not
-seems to be compatible: to get hybrid incompatibility or sterility, you need to
-have epsilon sufficiently large (below equation 5). But then $\epsilon^2$ is no
-longer smaller than epsilon. In other words: it seems to me that for the
-discussion of hybrid incompatibilities you apply your model of the first part
-beyond the local range where it is valid.
-\end{point}
-
-\reply{
-}
+Yes, exactly. This is a natural model,
+so we were surprised to not find an analytic treatment of it
+in the literature.
 
@@
     For system drift, you assume that ``Selection will tend to restrain this
-motion, but movement along the optimal set N is unconstrained, and so we expect
+motion, but movement along the optimal set $\allS$ is unconstrained, and so we expect
 the population mean to drift along the optimal set like a particle diffusing.''
@@
     In the presence of epistasis, evolution on a neutral network (a high-fitness
-  ridge) is *not* due to drift alone, but also affected by (weak second order)
+    ridge) is \emph{not} due to drift alone, but also affected by (weak second order)
     selection in favor of mutational robustness / genetic canalization. In contrast
     to what you write, diffusion on the set of network coefficients corresponding
-to the optimal phenotype is not unbiased - even if the optimal phenotypes do
+    to the optimal phenotype is not unbiased -- even if the optimal phenotypes do
     indeed all have the same fitness. Instead, selection drives the population to
@@
 \begin{point}{(drift along a ridge)}
-    a second problem results from the fact that evolution on a high-fitness ridge
+    A second problem results from the fact that evolution on a high-fitness ridge
     often requires coordinated changes at many loci. Take the oscillator system
@@
 \begin{quote}
-The present paper is divided into two parts. In the first part, the authors propose a framework
-in which distinct genetic architectures can produce the same phenotype. (In the fitness landscape
-terminology, this would correspond to the existence of an evolutionary ridge in a fitness landscape.
-)The starting point is a nice analogy with linear systems theory. The authors highlight the fact
-that in general, two distinct linear (differential) systems can respond identically for any input; i.e.,
-different systems (genotypes) can always produce the same output (phenotype) given the same input
-(environment). Further, the set of equivalent systems can be nicely characterized through Kalman
-decomposition, thus providing a nice characterization of level sets in the underlying fitness landscape.
-All along the first part of the paper, this analogy is well illustrated using a simple (yet quite nice)
-example of oscillating gene transcription. I especially like Fig. 5 where it is shown that even if $F_1$
-hybrids have a phenotypic response close to their parents, $F_2$ hybrids can behave in a drastically
-different manner.
-
 In the second part of the paper, motivated by the previous results, the authors explore a general
@@
 
-\begin{point}{end of p6:} 
-    ``we assume ... $((A + A' )/2, (B + B' )/2, (C + C' )/2)$'' . It is claimed in p4 (right
-before the last paragraph) that two kryptotypes need not have the same dimension. In this
-case, the previous sum does not make sense, right?
-\end{point}
-
-\reply{
-}
-
-\begin{point}{p7: fig 4.}
-    ``The distribution of $F_2$ ... homozygotes''. I do not really grasp the meaning of this
-sentence. More importantly, I find the purple cloud of dots ($F_2$ population) quite confusing.
-It seems to me that the purple populations should be a cloud of points concentrated around
-a point sitting on the red dotted line, i.e., a Gaussian distribution with the mean at the
-average of the two parental populations. According to the authors, the distribution is bi-
-modal with a peak below the optimal set (as displayed on the figure) and another distinct
-peak sitting on the other side of the optimal. In fact, if I understand the computations of
-the second part correctly, the red and purple means should coincide, but the red variance
-should simply be greater.
-\end{point}
-
-\reply{
-}
-
-\begin{point}{p7: end of the page.}
-    Why do you need $\rho$ to be square integrable?
-\end{point}
-
-\reply{
-    Good point -- clearly some assumptions jointly on the long-time behavior of $h(t)$ and $\rho$ 
-    are required for $D$ to be finite, but just saying that $\rho$ is square integrable doesn't suffice;
-    we've removed the caveat. 
-    \revref
-    (It is always well-defined, anyhow, since the integrand is nonnegative.)
-}
-
 \begin{point}{p8:}
@@
 extra explanation (e.g., in the appendix)?
-\end{point}
-
-\reply{
-}
-
-\begin{point}{p9. Fig 5.} 
-    In the left panel, it seems to me that the two parents are homozygotes since there
-is a single $F_1$ possible offspring (dashed blue curve). It would be worth being more explicit.
-For the right panels, could you be more explicit on the number of curves. If I understand
-correctly, there are 16 possibilities due to recombination (2 per entries of the matrix) since
-$F_1$ individuals are heterozygotes. Is that correct?
-Finally, the labels on the y axis are not easy to read.
 \end{point}
@@
 \reply{
-    This part has been removed.
 }
@@
     Finally, it would be worth mentioning several old and recent works relating genetic drift
-to speciation: Yamagushi and Iwasa, several articles by Gavrilets et al. (I think several
+to speciation : Yamagushi and Iwasa, several articles by Gavrilets et al. (I think several
 citations are missing here), and Mirò Pina and Schertzer.

### 2018-11-09 ridge_evolution_paper.tex
started on responses
a bit more

@@
 \newcommand{\fitx}{\mathcal{F}}
-% set of optimal coefficients
-\newcommand{\optx}{\mathcal{X}}
 % optimal phenotype
@@
 % \reviewresponsesfalse  % don't include them
-\newcommand{\responsefile}{review-responses}  % name of the review responses file
+\newcommand{\responsefile}{ridge-review-responses}  % name of the review responses file
 
@@
 \begin{abstract}
-Drift along a ridge.
+    It has been often argued that a common feature of biological evolution
+    is that there are many distinct combinations of low-level traits
+    that combine to produce the same high-level trait
+    -- i.e., there is more than one way to do the same thing.
+    If so, then the fitness landscape induced on the low-level traits has a ridge
+    along which populations will, in principle,
+    drift due to random demographic fluctuations.
+    Here, we use one such model of a fitness ridge and the infinitesimal model
+    to ask how fast populations are expected to drift,
+    and find that this process is expected to reproducive incompatibilities
+    between isolated populations over a number of generations proportional
+    to the effective population size.
 \end{abstract}
@@
 \section*{Introduction}
+
+Ridges in a fitness landscape
+could be due to a variety of causes.
+For instance, lower-level traits could be strictly redundant,
+or could produce a higher-level trait additively.
+There could be a trade-off between two traits under selection,
+such as XXX
+We were motivated to write this paper
+by results from system theory
+that explicitly describe a large parameter space
+of linear systems with identical input--output relationships \citep{shiffman2018system}.
 
@@
 The optimum phenotype is unique, but is realized by many distinct $x$ -- those falling in the ``optimal set'' $\allS$.
-The phenotypic distance to optimum of an organism with coefficients $x$ is denoted $D(x)$.
-In the results above, $x = (A,B,C)$ and $D(x)$ is given by equation \eqref{eqn:distance}.
-The fitness of an organism with coefficients $x$ will be $\exp(-D(x)^2)$.
+The phenotypic distance to optimum of an organism with coefficients $x$ is denoted $D(x)$,
+% In the results above, $x = (A,B,C)$ and $D(x)$ is given by equation \eqref{eqn:distance}.
+and the fitness of an organism with coefficients $x$ will be $\exp(-D(x)^2)$.
 % this assumption does not entail a loss of generality since $D$ is arbitrary.
 We assume that in the region of interest, the map $D$ is smooth
-and that we can locally approximate the optimal set $\allS$ as a quadratic surface.
+and that we can locally approximate $D(x)^2$ near the optimal set $\allS$ 
+as a quadratic surface.
 As above, an individual's coefficients are given by averaging its parentally inherited coefficients and adding random noise due to segregation and possibly new mutation.
@@
 on longer time scales the appropriate model may share properties with
-the ``holey'' landscapes of \citet{gavrilets2004fitness}.
+the ``holey'' fitness landscapes of \citet{gavrilets2004fitness}.
 
@@
 simply because this is the standard deviation of the mean of a random sample \citep{lande1981models}.
-% (This could be taken as a definition of $N_e$.)
-Selection will tend to restrain this motion,
-but movement along the optimal set $\allS$ is unconstrained,
-and so we expect the population mean to drift along the optimal set like a particle diffusing.
+Selection will restrain motion away from $\allS$,
+but movement along the optimal set $\allS$ is relatively unconstrained,
+and so as a first approximation, 
+the population mean will drift along the optimal set like a particle diffusing.
+The speed of this random walk
+will be modified by two factors.
+First, epistasis may require coordinated substitutions at several sites,
+which may be nearly impossible if the individual substitutions are selected against strongly enough
+in isolation.
+This  XXX
+
 % (although perhaps complicated by recombination load \citep{recomb_load})
@@
 
+\begin{figure}[H]
+\centering
+\includegraphics{figures/conceptual_fig}
+\caption{
+    \label{fig:conceptual_fig}
+    A conceptual figure of the fitness consequences of hybridization:
+    axes represent system coefficients (i.e., entries of $A$);
+    the line of optimal system coefficients is down in black;
+    dotted lines give phenotypic distances to the optimum.
+    A pair of parental populations are shown in black, along the optimum;
+    a hypothetical population of $F_1$s are shown in red,
+    and the distribution of one type of $F_2$ is shown in purple
+    (other types of $F_2$ are not shown; 
+    some would be a similar distance to the other side of the optimal set).
+    The distribution of $F_2$ hybrids is appropriate for mixed homozygotes
+    if both traits have a simple, one-locus genetic basis,
+    but there is variation within each population at that locus.
+}
+\end{figure}
+
 \paragraph{Hybridization}
 The means of two allopatric populations each of effective size $N_e$ separated for $T$ generations
-will be a distance roughly of order $2\sigma_N \sqrt{T/N_e}$ apart along $\optx$.
+will be a distance roughly of order $2\sigma_N \sqrt{T/N_e}$ apart along $\allS$. \revpoint{AE}{2}
 (Consult figure \ref{fig:conceptual_fig} for a conceptual diagram.)
@@
 and stability of the $G$-matrix are unknown in practice.
-Our observations argue for the importance of including neutral directions in these models,
-which is not usually done (but see \citet{chevin2014niche}).
+Our observations argue that for questions related to divergence between populations 
+it may be important to model a non-point optimum,
+as is in \citet{chevin2014niche}. \revpoint{AE}{3}
 

### 2020-06-01 review-responses.tex
a bit more
first pass 2020

@@
 \reply{
+  We've added line numbers and corrected improperly formatted citations. 
 }
@@
 \reply{
+  Thank you for the suggestion. We now mention some \revref.
+  \jss{I think the general assumptions are time invariance and linearity? Any others we should mention?}
 }
@@
 \reply{
+  We think that adding a third type of arrowhead may only add to the confusion, so now we are using only the standard arrowhead \revref.
+  \jss{we could add a cirlce arrowhead for the unknown direction, but I think its simpler to just get rid of the rectangular negative arrowhead originally in Figure 3}
 }
@@
 \reply{
+  We've reworked the introduction and included a paragraph on the phenonmenon of biological degeneracy, which unites many of these concepts. \revref
 }
@@
 \reply{
+  Good point. In the introdcution we now introduce nonidentifiability after discussing general cases of phenotypic equivalence. \revref
 }
@@
 \reply{
+  We've reworked the introduction and included more detail about these references. \revref
 }
@@
 \reply{
+  Added \revref.
 }
@@
 
+\begin{point}{}
+    Your analysis of the fitness loss in F1 and F2 hybrids rests on a local
+Taylor expansion of the fitness landscape. This is adequate given that the
+underlying linear network is also only locally valid (see above). However, you
+then apply this to a discussion of hybrid incompatibility (Haldane's rule etc).
+We are thus interested in \emph{very large} fitness costs for hybrids. This does not
+seems to be compatible: to get hybrid incompatibility or sterility, you need to
+have epsilon sufficiently large (below equation 5). But then $\epsilon^2$ is no
+longer smaller than $\epsilon$. In other words: it seems to me that for the
+discussion of hybrid incompatibilities you apply your model of the first part
+beyond the local range where it is valid.
+\end{point}
+
+\reply{
+    That's a good point; we resort to a Taylor expansion 
+    because we couldn't think of any other clear way to get quantitative estimates.
+    However, our simulations XXX.
+}
+
 %%%%%%%%%%%%%%

### 2020-06-01 kalman_walk_paper.tex
first pass 2020

@@
 \usepackage{fullpage}
-\usepackage[backref=page, hidelinks, colorlinks=true, citecolor=blue!60!black!100]{hyperref}
+\usepackage[backref=page, hidelinks, colorlinks=true, allcolors=blue!60!black!100]{hyperref}
 \usepackage{tikz}
@@
 \begin{document}
-%\linenumbers
+\linenumbers
 
@@
 
-It is an overarching goal of many biological subdisciplines 
-to attain a general understanding of the function and evolution of the 
-complex molecular machinery that translates an organism's genome 
-into the characteristics on which natural selection acts,
-the phenotype.
-% Bridging the gulf between an organism's genome and phenotype is a poorly understood and complex molecular machinery. 
-%Attaining a general understanding of the functioning and evolution of this molecular machinery
-For example, 
-there is a growing body of data on the evolutionary histories and molecular characterizations of particular gene regulatory networks
-\citep{jaeger2011gap, davidson2006gene, israel2016comparative}, 
-as well as thoughtful verbal and conceptual models \citep{true2001developmental, weiss2000phenogenetic, edelman2001degeneracy, pavlicev2012model}. 
-Mathematical models of both particular regulatory networks
-and the evolution of such systems in general
-can provide guidance where intuition fails,
-and thus has the potential to discover general principles in the organization of biological systems 
-as well as provide concrete numerical predictions \citep{servedio2014not}.
-%The dynamics of the molecular machinery and its interactions with the environment
-%can be mathematically described as a dynamical system \citep{jaeger2015comet}.
-%Movement in this direction is ongoing, as 
-There is a substantial amount of work studying the evolution of gene regulatory networks, in frameworks 
-both abstract \citep{wagner1994evolution, wagner1996does, siegal2002waddington, bergman2003evolutionary, draghi2015robustness} 
-and empirically inspired
-\citep{mjolsness1991connectionist, jaeger2004dynamic, vitaly1, crombach2016gap, wotton2015quantitative, chertkova2017insilico}.
-% If we allow the reasonable assumption that the genotype-phenotype map 
-% can be represented as a system of differential equations \citep{jaeger2015comet}, 
-% we can immediately discuss its evolution and function in a much more mechanistic, yet general, manner.
-
-It is well known that in many contexts mathematical models
+ Biological systems, at nearly all levels of organization, display considerable \emph{degeneracy} -- 
+that is systems can be structurally different yet remain functionally equivalent \citep{edelman2001degeneracy}.
+Examples can be found across nearly all levels of biological organization from the level of the genetic code itself \citep{crick, woesse}
+all the way up to the convergent evolution of adaptive traits. Between these levels, degeneracy can be found in the \emph{neutral networks}
+  of nucleic acid sequences that map to an RNA secondary structure \citep{gruner, andmore}\revpoint{1}{5} and amino acid sequences that fold similarly \citep{babajide},
+as well as in the thermodynamic stability of proteins \citep{hart2014thermodynamic}. Further examples include the vast space of functionally equivalent
+potential regulatory sequences \citep{hare2008sepsid}, 
+in the logic of transcriptional \citep{tsong2006evolution, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription} 
+and neural circuits \citep{trojanowski}, and in developmental systems \citep{true2001developmental}. 
+\revpoint{1}{1} \revpoint{1}{3}
+
+Likewise, it is well known that in many contexts mathematical models
 can fundamentally be \emph{nonidentifiable} and/or \emph{indistinguishable} -- meaning that 
@@
 can make equally accurate predictions,
-but still not actually reflect the internal dynamics of the system being modelled.
+but still not actually reflect the internal dynamics of the system being modeled.
 In control theory, where electrical circuits and mechanical systems are often the focus, 
@@
 \citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. 
-The inherent nonidentifiability of chemical reaction networks is sometimes referred to as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}. 
-In computer science, this is framed as the relationship among processes that \emph{simulate} one another \citep{van2004equivalence}.
+The inherent nonidentifiability of chemical reaction networks is sometimes
+referred to as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}.
 Finally,
@@
 despite the existence of a theoretical one-to-one mapping between a model and behavior,
-tiny amounts of noise make inference problems nonidentifiable in practice \citep{petrov2005well}.
-%even if a one-to-one mapping between model and behavior is possible in theory,
-%even tiny amounts of noise can make inference problems nonidentifiable in practice \citep{petrov2005well}.
-
-Nonidentifiability is a major barrier to mechanistic understanding of real systems, 
-but viewed from another angle,
-this concept can provide a starting point for thinking about externally equivalent systems
--- systems that evolution can explore, so long as the parameters and structures can be realized biologically.
-These functional symmetries manifest in convergent and parallel evolution, 
-as well as \emph{developmental system drift}: the observation that
-macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level 
-\citep{kimura1981possibility, true2001developmental, tanay2005conservation, tsong2006evolution, hare2008sepsid, lavoie2010evolutionary, vierstra2014mouse, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription}.
-
-The main purpose of this paper is to explore this general idea
-in the concrete framework of linear systems theory,
-thought of as modeling gene regulatory networks.
+  tiny amounts of noise make inference problems nonidentifiable in practice \citep{petrov2005well}. \revpoint{1}{2}
+
+In this paper we use linear systems theory to model gene regulatory networks, linking biological degeneracy 
+with mathematical nonidentifiability. Although our focus is on linear systems, nonidentifability is by no 
+means unique to such frameworks, and is a general property of many nonlinear models \citep{nonlinear}.
+
+System degeneracy has been argued to be necessary for natural selection \citep{edleman2001degeneracy}, 
+as it may function as a mechanism for biological robustness and evolvability \citep{robustness}, or manifest as 
+\emph{canalization} \citep{canalization}. Degeneracy of the genetic code, and of genetic networks, 
+for instance, can make sequences more fault-tolerant to mutations and open up pathways to adaptation (cite). 
+Biological degeneracy can also enable the evolution of neutral and nearly neutral molecular changes \citep{kimura},
+and perhaps even contribute to the formation of new species \citep{gavrilets}.
+
+We are interested in the nature and evolutionary implications of degeneracy and nonidentifiability in the context of gene regulatory networks.
+Specifically, we would like to understand the degree of which gene regulatory network architectures are functionally equivalent and how this relates to network size and efficiency.
+If systems architectures are not functionally unique, can this open up neutral evolutionary paths, and do
+  these paths manifest as \emph{developmental system drift} \citep{true2001developmental} and contribute to speciation?
 First, we apply results from system theory which give
@@
 all linear gene network architectures that yield identical phenotypes.
-This suggests that, quite generally, there are many directions in which a gene network can drift
-without changing the phenotype.
-Since these phenotypically equivalent gene networks are not necessarily compatible with one another, 
-system drift may result in reproductive incompatibility between isolated populations,
-even in the absence of any sort of adaptive, or environmental change.
-
- % \jss{should hare2008sepsid be here? It's sequence but not ``system'' drift}
-% are these good here? {taylor2016diverse -- maybe not, matsui2015regulatory -- I think this relevant, stergachis2014conservation}.
-%
-%AND THESE?
-  %% Over the last several years, different computational models have been applied to study reproductive incompatibility and speciation in models of gene regulation \citep{porter2002speciation, tulchinsky, palmer2009dynamics, tulchinsky2014hybrid, khatari2015simple}.
-  %\citet{tulchinsky} simulated the evolution of a transcription factor and its binding site using a thermodynamic model.
+Finally, we briefly outline and apply a quantitative genetics framework to estimate the speed at which system drift can lead to speciation via reproductive incompatibility,
+which we develop in more detail in a companion paper \citep{schiffman2020genetic}.
+
+%It is an overarching goal of many biological subdisciplines 
+%to attain a general understanding of the function and evolution of the 
+%complex molecular machinery that translates an organism's genome 
+%into the characteristics on which natural selection acts,
+%the phenotype.
+%For example, 
+%there is a growing body of data on the evolutionary histories and molecular characterizations of particular gene regulatory networks
+%\citep{jaeger2011gap, davidson2006gene, israel2016comparative}, 
+%as well as thoughtful verbal and conceptual models \citep{true2001developmental, weiss2000phenogenetic, edelman2001degeneracy, pavlicev2012model}. 
+%Mathematical models of both particular regulatory networks
+%and the evolution of such systems in general
+%can provide guidance where intuition fails,
+%and thus has the potential to discover general principles in the organization of biological systems 
+%as well as provide concrete numerical predictions \citep{servedio2014not}.
+%There is a substantial amount of work studying the evolution of gene regulatory networks, in frameworks 
+%both abstract \citep{wagner1994evolution, wagner1996does, siegal2002waddington, bergman2003evolutionary, draghi2015robustness} 
+%and empirically inspired
+%\citep{mjolsness1991connectionist, jaeger2004dynamic, vitaly1, crombach2016gap, wotton2015quantitative, chertkova2017insilico}.
+%
+%It is well known that in many contexts mathematical models
+%can fundamentally be \emph{nonidentifiable} and/or \emph{indistinguishable} -- meaning that 
+%there can be uncertainty about an inferred model's parameters or even its claims about
+%causal structure, despite access to complete and perfect data \citep{bellman1970structural, grewal1976identifiability, walter1984structural}. 
+%Models with different parameter schemes, or even different mechanics 
+%can make equally accurate predictions,
+%but still not actually reflect the internal dynamics of the system being modelled.
+%In control theory, where electrical circuits and mechanical systems are often the focus, 
+%it is understood that there can be an infinite number of ``realizations'', 
+%or ways to reverse engineer the dynamics of a ``black box'',
+%even if all possible input and output experiments are performed 
+%\citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. 
+%The inherent nonidentifiability of chemical reaction networks is sometimes referred to as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}. 
+%In computer science, this is framed as the relationship among processes that \emph{simulate} one another \citep{van2004equivalence}.
+%Finally,
+%the field of \emph{inverse problems} studies those cases in which,
+%despite the existence of a theoretical one-to-one mapping between a model and behavior,
+%tiny amounts of noise make inference problems nonidentifiable in practice \citep{petrov2005well}.
+%
+%Nonidentifiability is a major barrier to mechanistic understanding of real systems, 
+%but viewed from another angle,
+%this concept can provide a starting point for thinking about externally equivalent systems
+%-- systems that evolution can explore, so long as the parameters and structures can be realized biologically.
+%These functional symmetries manifest in convergent and parallel evolution, 
+%as well as \emph{developmental system drift}: the observation that
+%macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level 
+%\citep{kimura1981possibility, true2001developmental, tanay2005conservation, tsong2006evolution, hare2008sepsid, lavoie2010evolutionary, vierstra2014mouse, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription}.
+%
+%The main purpose of this paper is to explore this general idea
+%in the concrete framework of linear systems theory,
+%thought of as modeling gene regulatory networks.
+%First, we apply results from system theory which give
+%an analytical description of the set of 
+%all linear gene network architectures that yield identical phenotypes.
+%This suggests that, quite generally, there are many directions in which a gene network can drift
+%without changing the phenotype.
+%Since these phenotypically equivalent gene networks are not necessarily compatible with one another, 
+%system drift may result in reproductive incompatibility between isolated populations,
+%even in the absence of any sort of adaptive, or environmental change.
+%We then briefly outline a quantitative genetics framework,
+%which we further develop in a companion paper,
+%which estimates how quickly system drift (and more generally, the presence of a fitness ridge)
+%leads to reproductive incompatibility \citep{schiffman2020genetic}.
+%
+% % \jss{should hare2008sepsid be here? It's sequence but not ``system'' drift}
+%% are these good here? {taylor2016diverse -- maybe not, matsui2015regulatory -- I think this relevant, stergachis2014conservation}.
+%%
+%%AND THESE?
+%  %% Over the last several years, different computational models have been applied to study reproductive incompatibility and speciation in models of gene regulation \citep{porter2002speciation, tulchinsky, palmer2009dynamics, tulchinsky2014hybrid, khatari2015simple}.
+%  %\citet{tulchinsky} simulated the evolution of a transcription factor and its binding site using a thermodynamic model.
 Their simulations suggest that the language by which a transcription factor recognizes its binding site can change, and potentially lead to hybrid incompatibility when allopatric populations employ divergent readout languages.
@@
 Johnson and Porter 2000 did not observe any hybrid fitness declines under stabilizing selection -- only under directional selection.
-%Khatari et al, Tulchinsky et al, and Porter et al, all study hybrid incompatibility from a transcription factor/binding site interaction perspective, not from an overall network architecture perspective.
+%%Khatari et al, Tulchinsky et al, and Porter et al, all study hybrid incompatibility from a transcription factor/binding site interaction perspective, not from an overall network architecture perspective.
 Palmer and Feldman only see hybrid incompatibility in constant environments if the parental populations are relatively poorly adapted initially.
 Otherwise hybrids between two allopatric populations have fairly high fitnesses.
-
-% wright1932roles mentions possibility of equivalent combinations
-% wright1935evolution looks at allele freq dynamics under stabilizing selection on a polygenic quantitative trait and discusses possibility for speciation this way
-% barton1989divergence following on from barton1986maintenance analyzing multiple optima in a similar model, looking at RI due to shifts between equilibria
-% nei1983models simulates from simple stepwise mut model on one or two loci (without interactions) where incompat betw diff alleles increases with number of steps betw them
-% sved1981twosex mathematically analyzes and (in another paper) simulates speciation in a model of male + female quantitative traits indep'ly evolving with attraction between matching traits
-% barton2010natural has a good historical review of natural selection and speciation; then analyzes a two-niche levene model
-% gavrilets2004fitness talks a lot about quasineutral paths in holey fitness landscapes
-% phillips1996maintenance shows that subpop structure can stably maintain equivalent but incompat genotypes
-% tulchinsky2014hybrid simulates a single promoter region with a bioenergentic model with directional and stabilizing selection, found incompat in both cases
-% porter2002speciation simulated speciation under a simple model of a regulatory cascade
-%%% Fisher's geom model: (incompat due to segregation variance)
-% poon2000compensating also uses SSSWM with Fisher's model and derives some analytical results for load in small populations; early citation for this process
-% slatkin1994segregation shows that segregation variance increases linearly with divergence if variation is controlled by intermediate freq alleles;
-%   but if it is determined by rare alleles (at mut-sel balance), then it plateaus (at twice the within-pop value?)
-% barton2001role looks at hybrids betw pops adapting to shifting optima in Fisher's geometric model : under fluctuating optima
-%       hybrids will suffer because of cryptic segregating variation
-% fraisse2016genetics uses SSWM; looks at stabilizing selection using Fisher's geometric model, showing it predicts lots of observations;
-%        but mostly only if fitness is exp(-|x|^k) for k > 2
-% chevin2014niche uses SSWM; looks at incompats after adaptation in Fisher's geom model and how it depends on dimension;
-%        uses generalized Fisher's model with neutral directions; terms hybrid unfitness due to segregation variance "transgressive incompats";
-%        finds rate of RI accumulation decreases with Ne (like Ne^(-m/2), with m the trait dimension?)
-% simon2017coadapted follows up on the Fisher's geom model predictions, also comparing to some real data
-% martin2014fishers shows that you can get Fisher's model quite generally if (a) the kryptotype dimension is large compared to the phenotype dimension,
-%   and (b) mutational effects fall under the CLT
-% rosas2010cryptic looks at segregation variance resulting in deviations from optimum phenotype including off of curvey ridges
-%%%% General notes
-% citation for segregation variance around a stationary optimum -> load is Poon & Otto 2000; rosas2010cryptic
-
-It is not a new observation that there is often more than one way to do the same thing, 
-or that speciation can be the result of (nearly) neutral processes.
-The potential for speciation has been analyzed in models of 
-traits under stabilizing selection determined additively by alleles at many loci 
-\citep{wright1935evolution,barton1986maintenance,barton1989divergence,barton2001role},
-in related fitness landscape models \citep{fraisse2016genetics},
-and for pairs of traits that must match but whose value is unconstrained \citep{sved1981twosex}.
-% Recent work has shown that Fisher's geometric model of stabilizing selection
-% predicts many general patterns from the speciation literature \citep{fraisse2016genetics}.
-It has also been shown that population structure
-can allow long-term stable coexistence of incompatible genotypes encoding identical phenotypes \citep{phillips1996maintenance}. \revpoint{1}{9}
-However, previous simulations of system drift in regulatory sequences \citep{tulchinsky2014hybrid}
-and a regulatory cascade \citep{porter2002speciation}
-found rapid speciation under directional selection
-but only equivocal support for speciation under models of purely neutral drift.
-%Additionally, \citet{fierst} observed speciation due to an increase of segregation variance under balancing selection, in models of weak to moderate, but not strong epistasis.
-The rate at which hybrid incompatibility accumulates due to genetic drift creating
-segregation variance between isolated populations is fairly well understood
-\citep{slatkin1994segregation,rosas2010cryptic,chevin2014niche},
-but model assumptions can strongly affect predictions,
-including whether variation is due to rare or common alleles \citep{slatkin1994segregation},
-and the shape of the fitness landscape \citep{fraisse2016genetics}.
-Our main aim is to provide a concrete framework 
-that can provide natural predictions of these model parameters
-across a general class of models.
-Furthermore, tools from system theory allow analytical predictions to be made
-for large populations with complex phenotypes
-that would be inaccessible to population simulations.
+%
+%% wright1932roles mentions possibility of equivalent combinations
+%% wright1935evolution looks at allele freq dynamics under stabilizing selection on a polygenic quantitative trait and discusses possibility for speciation this way
+%% barton1989divergence following on from barton1986maintenance analyzing multiple optima in a similar model, looking at RI due to shifts between equilibria
+%% nei1983models simulates from simple stepwise mut model on one or two loci (without interactions) where incompat betw diff alleles increases with number of steps betw them
+%% sved1981twosex mathematically analyzes and (in another paper) simulates speciation in a model of male + female quantitative traits indep'ly evolving with attraction between matching traits
+%% barton2010natural has a good historical review of natural selection and speciation; then analyzes a two-niche levene model
+%% gavrilets2004fitness talks a lot about quasineutral paths in holey fitness landscapes
+%% phillips1996maintenance shows that subpop structure can stably maintain equivalent but incompat genotypes
+%% tulchinsky2014hybrid simulates a single promoter region with a bioenergentic model with directional and stabilizing selection, found incompat in both cases
+%% porter2002speciation simulated speciation under a simple model of a regulatory cascade
+%%%% Fisher's geom model: (incompat due to segregation variance)
+%% poon2000compensating also uses SSSWM with Fisher's model and derives some analytical results for load in small populations; early citation for this process
+%% slatkin1994segregation shows that segregation variance increases linearly with divergence if variation is controlled by intermediate freq alleles;
+%%   but if it is determined by rare alleles (at mut-sel balance), then it plateaus (at twice the within-pop value?)
+%% barton2001role looks at hybrids betw pops adapting to shifting optima in Fisher's geometric model : under fluctuating optima
+%%       hybrids will suffer because of cryptic segregating variation
+%% fraisse2016genetics uses SSWM; looks at stabilizing selection using Fisher's geometric model, showing it predicts lots of observations;
+%%        but mostly only if fitness is exp(-|x|^k) for k > 2
+%% chevin2014niche uses SSWM; looks at incompats after adaptation in Fisher's geom model and how it depends on dimension;
+%%        uses generalized Fisher's model with neutral directions; terms hybrid unfitness due to segregation variance "transgressive incompats";
+%%        finds rate of RI accumulation decreases with Ne (like Ne^(-m/2), with m the trait dimension?)
+%% simon2017coadapted follows up on the Fisher's geom model predictions, also comparing to some real data
+%% martin2014fishers shows that you can get Fisher's model quite generally if (a) the kryptotype dimension is large compared to the phenotype dimension,
+%%   and (b) mutational effects fall under the CLT
+%% rosas2010cryptic looks at segregation variance resulting in deviations from optimum phenotype including off of curvey ridges
+%%%%% General notes
+%% citation for segregation variance around a stationary optimum -> load is Poon & Otto 2000; rosas2010cryptic
+%
+%It is not a new observation that there is often more than one way to do the same thing, 
+%or that speciation can be the result of (nearly) neutral processes.
+%The potential for speciation has been analyzed in models of 
+%traits under stabilizing selection determined additively by alleles at many loci 
+%\citep{wright1935evolution,barton1986maintenance,barton1989divergence,barton2001role},
+%in related fitness landscape models \citep{fraisse2016genetics},
+%and for pairs of traits that must match but whose value is unconstrained \citep{sved1981twosex}.
+%% Recent work has shown that Fisher's geometric model of stabilizing selection
+%% predicts many general patterns from the speciation literature \citep{fraisse2016genetics}.
+%It has also been shown that population structure
+%can allow long-term stable coexistence of incompatible genotypes encoding identical phenotypes \citep{phillips1996maintenance}. \revpoint{1}{9}
+%However, previous simulations of system drift in regulatory sequences \citep{tulchinsky2014hybrid}
+%and a regulatory cascade \citep{porter2002speciation}
+%found rapid speciation under directional selection
+%but only equivocal support for speciation under models of purely neutral drift.
+%%Additionally, \citet{fierst} observed speciation due to an increase of segregation variance under balancing selection, in models of weak to moderate, but not strong epistasis.
+%The rate at which hybrid incompatibility accumulates due to genetic drift creating
+%segregation variance between isolated populations is fairly well understood
+%\citep{slatkin1994segregation,rosas2010cryptic,chevin2014niche},
+%but model assumptions can strongly affect predictions,
+%including whether variation is due to rare or common alleles \citep{slatkin1994segregation},
+%and the shape of the fitness landscape \citep{fraisse2016genetics}.
+%Our main aim is to provide a concrete framework 
+%that can provide natural predictions of these model parameters
+%across a general class of models.
+%Furthermore, tools from system theory allow analytical predictions to be made
+%for large populations with complex phenotypes
+%that would be inaccessible to population simulations.
+%
 
@@
 into the $\ell$-dimensional output $\phi(t)$.
-Under quite general assumptions,
+Under quite general assumptions (\emph{e.g.} the system is linear, time-invariant and in the zero-state), \revpoint{AE}{3}
 we can write the phenotype as \revpoint{AE}{4}
@@
 
+\revpoint{AE}{7}
+
 \begin{figure}[H]
@@
     %\path [->, sloped] (B) edge[bend right] node {\tiny ?} (C); 
-    \path [->, >=Rectangle, sloped] (C) edge[bend right] node {\tiny $-1$} (A); 
-    \path [->, >=Rectangle, sloped] (C) edge[bend right] node {\tiny $-2.2$} (B); 
+    \path [->, sloped] (C) edge[bend right] node {\tiny $-1$} (A); 
+    \path [->, sloped] (C) edge[bend right] node {\tiny $-2.2$} (B); 
     \path[->] (U) edge node {\tiny $1$} (A);
@@
     \path [->] (B) edge[loop above] node[sloped, above] {\tiny $1$} (B);
-    \path [->, >=Rectangle] (C) edge[loop right] node[sloped, above] {\tiny $-1$} (C);
+    \path [->] (C) edge[loop right] node[sloped, above] {\tiny $-1$} (C);
 \end{scope}
@@
 Although many alleles will be dominant if the phenotype--fitness relationship is convex,
-the underlying mechanism does not depend on the \emph{dominance theory} \citet{turelli1995dominance} to explain Haldane's rule:
+the underlying mechanism does not depend on the \emph{dominance theory} \citep{turelli1995dominance} to explain Haldane's rule:
 instead, it derives from the nature of segregation variance.
@@
 It has been shown that between human individuals, there is differential occupancy in 7.5\%
-of binding sites of a transcription factor (p65) \cite{kasowski2010variation}.
+of binding sites of a transcription factor (p65) \citep{kasowski2010variation}.
 %\citet{verlaan2009targeted} showed that 
 It has also been inferred that cis-regulatory variation
-accounts for around 2--6\% of expression variation in human blood-derived primary cells \cite{verlaan2009targeted}, and
+accounts for around 2--6\% of expression variation in human blood-derived primary cells \citep{verlaan2009targeted}, and
 %while \citet{lappalainen2013transcriptome} found 
 that human population variation 
-explained about 3\% of expression variation \cite{lappalainen2013transcriptome}.
+explained about 3\% of expression variation \citep{lappalainen2013transcriptome}.
 Allele-specific expression is indicative of standing genetic \emph{cis}-regulatory variation;
 %\citet{wang2017bayesian} found 
-allele-specific expression in 7.2--8.5\% of transcripts of a flycatcher species has been observed \cite{wang2017bayesian}, as well as
+allele-specific expression in 7.2--8.5\% of transcripts of a flycatcher species has been observed \citep{wang2017bayesian}, as well as
 %while \citet{tung2015genetic} observed 
-allele-specific expression in 23.4\% of genes studied in a baboon species \cite{tung2015genetic}. 
+allele-specific expression in 23.4\% of genes studied in a baboon species \citep{tung2015genetic}. 
 Taken together, this suggests that variation in the entries of $A$
@@
 gives an analytical description of all phenotypically equivalent gene networks.
-It also implies that nearly all systems are nonidentifiable,
-and that, in general, there exist axes of genetic variation unconstrained by natural selection.
+%It also implies that nearly all systems are nonidentifiable,
+%and that, in general, there exist axes of genetic variation unconstrained by natural selection.
+This description shows that the space of functionally equivalent network architectures increases with the square of a network's size,
+and that this space increases further if networks grow larger than absolutely necessary -- that is use more interacting components than the most efficient potential architectures.
+In this framework, even minimal gene network architectures -- efficient architectures that contain only the requisite number of interacting parts,
+are not structurally unique with respect to function.
+Functionally equivalent architectures are often related by continuous parameter changes, suggesting that equivalent networks might be mutationally connected, 
+and that there exist axes of genetic variation unconstrained by natural selection.
 The independent movement of separated populations along these axes by genetic drift
@@
 
-Consistent with empirical observation of hybrid breakdown (e.g., \citet{plotner2017chlorosis}), 
+Consistent with empirical observation of hybrid breakdown (e.g., \citet{plotner2017chlorosis}) and transgressive segregation \citep{transgressive}, 
 we see that the fitnesses of $F_2$ hybrids drop at a much faster rate than those of $F_1$s.
@@
 \paragraph{The dimensionality of trait space} \revpoint{1}{8}
-We have focused in examples on single traits (where the phenotype is one-dimensional),
+\jss{I didn't think that our results depend on the kryptotype being in a higher dimensional space than the phenotype.
+Any coordinate change on (A,B,C) will rewire the system. So as long as none of the system matrices are constrained, 
+any minimal system should have an infinite number of realizations. Unless my system theory is rusty?}
+
+We have focused on examples of single traits (where the phenotype is one-dimensional),
 but phenotypes under selection are often high-dimensional,
 and different traits often share a genetic basis.
-If the phenotype and kryptotype are of the same dimension,
-then we do not in general have degrees of freedom, i.e., neutral directions.
+Even in networks where the phenotype and kryptotype are of the same dimension,
+system theory shows us that there will always be available degrees of freedom
+as specific system realizations are only unique up to a change of coordinates.
+Some phenotypes, however, require kryptotypic dimensions to be larger than that of the phenotype.
+For instance, there exists minimal realizations (\emph{e.g.} the oscillator in Example 1) where the dimension of the kryptotpye is larger
+than that of the phenotype, implying that for these phenotypic dynamics to be realized, the kryptotype dimension \emph{has} to be larger than the dimension of the phenotype.
+Of course we see many more degrees of freedom in nonminimal networks, but it is not yet clear if evolution commonly arrives at such structures.
+There are some hints that this may be occuring: compared to the molecular circuitry of the cyanobacteria circadian rhythm,
+the human clock appears to be highly and unneccessarily complex \citep{sancar2008intelligent}. 
 The true ``dimensionality of trait space'' has been debated at some length in the literature
 \citep{trait_space}.
-Whatever it is, however, it seems unlikely 
-that it would approach the dimensionality of kryptotype space,
-which is something related to the number of genes.
-Even under universal pleiotropy \citep{universal_pleiotropy,omnigenic},
+Whatever it is, however, it seems likely to be highly degenerate,
+%Whatever it is, however, it seems unlikely 
+%that it would approach the dimensionality of kryptotype space,
+%which is something related to the number of genes.
+even under universal pleiotropy \citep{universal_pleiotropy,omnigenic},
 XXX
@@
 
-Finally, despite our model's precise separation of phenotype and kryptotype, this relationship in nature may be far more complicated as aspects of the kryptotype may be less ``hidden'' than we currently assume.
-For instance, attributes excluded from the phenotype as modelled here, ignore the potential energy costs associated with excessively large (non-minimal) kryptotypes, as well as the relationship between a specific network architecture and robustness to mutational, transcriptional, or environmental noise.
+Finally, despite our model's precise separation of phenotype and kryptotype, this relationship in nature may be far more complicated as aspects of the kryptotype may be less ``hidden'' than we currently assume, and the neutral network changes we describe here may only be nearly neutral.
+For instance, attributes excluded from the phenotype as modeled here, ignore the potential energy costs associated with excessively large (non-minimal) kryptotypes, as well as the relationship between a specific network architecture and robustness to mutational, transcriptional, or environmental noise.
 % Despite the model's simplifications and potentially nontrivial oversights, it may still roughly reflect and shed light upon speciation patterns as they occur in nature.

### 2020-06-17 kalman_walk_paper.tex
second pass 2020

@@
 \section*{Introduction}
-
- Biological systems, at nearly all levels of organization, display considerable \emph{degeneracy} -- 
+  \jss{I'm not sure we still need this first paragraph -- I don't feel like it says that much?}
+It is an overarching goal of many biological subdisciplines 
+to attain a general understanding of the function and evolution of the 
+complex molecular machinery that translates an organism's genome 
+into the characteristics on which natural selection acts,
+the phenotype.
+For example, 
+there is a growing body of data on the evolutionary histories and molecular characterizations of particular gene regulatory networks
+\citep{jaeger2011gap, davidson2006gene, israel2016comparative}, 
+as well as thoughtful verbal and conceptual models \citep{true2001developmental, weiss2000phenogenetic, edelman2001degeneracy, pavlicev2012model}. 
+Mathematical models of both particular regulatory networks
+and the evolution of such systems in general
+can provide guidance where intuition fails,
+and thus has the potential to discover general principles in the organization of biological systems 
+as well as provide concrete numerical predictions \citep{servedio2014not}.
+There is a substantial amount of work studying the evolution of gene regulatory networks, in frameworks 
+both abstract \citep{wagner1994evolution, wagner1996does, siegal2002waddington, bergman2003evolutionary, draghi2015robustness} 
+and empirically inspired
+\citep{mjolsness1991connectionist, jaeger2004dynamic, vitaly1, crombach2016gap, wotton2015quantitative, chertkova2017insilico}.
+
+ % In biological systems, the relationship between structure and function is often not invertible -- that is information about function is not usually sufficient to determine structure.
+ % Non-invertiblility thus entails a diversity of functionally equivalent structures.
+  Biological systems, at nearly all levels of organization, display considerable degeneracy, or (perhaps more descriptively) a capacity for isofunctional \emph{structural plasticity} -- 
   that is systems can be structurally different yet remain functionally equivalent \citep{edelman2001degeneracy}. 
+  This is a consequence of the often non-invertibility (i.e. degeneracy) of the mapping between a system's structure and function.
 Examples can be found across nearly all levels of biological organization from the level of the genetic code itself \citep{crick, woesse}
-all the way up to the convergent evolution of adaptive traits. Between these levels, degeneracy can be found in the \emph{neutral networks}
+all the way up to the convergent evolution of adaptive traits. Between these levels, structural plasticity can be found in the \emph{neutral networks}
   of nucleic acid sequences that map to an RNA secondary structure \citep{gruner, andmore}\revpoint{1}{5} and amino acid sequences that fold similarly \citep{babajide},
@@
 \revpoint{1}{1} \revpoint{1}{3}
+  \jss{Other ideas instead of "degeneracy" or "isofunctional structural plasticity": non-invertibility, isofunctional plasticity, structural plasticity, phenotypic equivalency, phenotypic invariance, functional invariance, isofunctionality, functional equivalency, genotypic plasticity, conjugacy, functional symmetry, isologous, isomorphic.}
 
@@
 referred to as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}.
+In computer science, this is framed as the relationship among processes that \emph{simulate} one another \citep{van2004equivalence}.
 Finally,
@@
 
-In this paper we use linear systems theory to model gene regulatory networks, linking biological degeneracy 
+In this paper we use linear systems theory to model gene regulatory networks, linking structural plasticity 
 with mathematical nonidentifiability. Although our focus is on linear systems, nonidentifability is by no 
@@
 
-System degeneracy has been argued to be necessary for natural selection \citep{edleman2001degeneracy}, 
+Structural plasticity has been argued to be necessary for natural selection \citep{edleman2001degeneracy}, 
 as it may function as a mechanism for biological robustness and evolvability \citep{robustness}, or manifest as 
@@
 for instance, can make sequences more fault-tolerant to mutations and open up pathways to adaptation (cite). 
-Biological degeneracy can also enable the evolution of neutral and nearly neutral molecular changes \citep{kimura},
+Non-invertibility can also enable the evolution of neutral and nearly neutral molecular changes \citep{kimura},
 and perhaps even contribute to the formation of new species \citep{gavrilets}.
 
-We are interested in the nature and evolutionary implications of degeneracy and nonidentifiability in the context of gene regulatory networks.
+We are interested in the nature and evolutionary implications of structural plasticity and nonidentifiability in the context of gene regulatory networks.
 Specifically, we would like to understand the degree of which gene regulatory network architectures are functionally equivalent and how this relates to network size and efficiency.
@@
 
-%It is an overarching goal of many biological subdisciplines 
-%to attain a general understanding of the function and evolution of the 
-%complex molecular machinery that translates an organism's genome 
-%into the characteristics on which natural selection acts,
-%the phenotype.
-%For example, 
-%there is a growing body of data on the evolutionary histories and molecular characterizations of particular gene regulatory networks
-%\citep{jaeger2011gap, davidson2006gene, israel2016comparative}, 
-%as well as thoughtful verbal and conceptual models \citep{true2001developmental, weiss2000phenogenetic, edelman2001degeneracy, pavlicev2012model}. 
-%Mathematical models of both particular regulatory networks
-%and the evolution of such systems in general
-%can provide guidance where intuition fails,
-%and thus has the potential to discover general principles in the organization of biological systems 
-%as well as provide concrete numerical predictions \citep{servedio2014not}.
-%There is a substantial amount of work studying the evolution of gene regulatory networks, in frameworks 
-%both abstract \citep{wagner1994evolution, wagner1996does, siegal2002waddington, bergman2003evolutionary, draghi2015robustness} 
-%and empirically inspired
-%\citep{mjolsness1991connectionist, jaeger2004dynamic, vitaly1, crombach2016gap, wotton2015quantitative, chertkova2017insilico}.
 %
@@
 %or that speciation can be the result of (nearly) neutral processes.
-%The potential for speciation has been analyzed in models of 
-%traits under stabilizing selection determined additively by alleles at many loci 
-%\citep{wright1935evolution,barton1986maintenance,barton1989divergence,barton2001role},
-%in related fitness landscape models \citep{fraisse2016genetics},
-%and for pairs of traits that must match but whose value is unconstrained \citep{sved1981twosex}.
-%% Recent work has shown that Fisher's geometric model of stabilizing selection
-%% predicts many general patterns from the speciation literature \citep{fraisse2016genetics}.
-%It has also been shown that population structure
-%can allow long-term stable coexistence of incompatible genotypes encoding identical phenotypes \citep{phillips1996maintenance}. \revpoint{1}{9}
-%However, previous simulations of system drift in regulatory sequences \citep{tulchinsky2014hybrid}
-%and a regulatory cascade \citep{porter2002speciation}
-%found rapid speciation under directional selection
-%but only equivocal support for speciation under models of purely neutral drift.
-%%Additionally, \citet{fierst} observed speciation due to an increase of segregation variance under balancing selection, in models of weak to moderate, but not strong epistasis.
-%The rate at which hybrid incompatibility accumulates due to genetic drift creating
-%segregation variance between isolated populations is fairly well understood
-%\citep{slatkin1994segregation,rosas2010cryptic,chevin2014niche},
-%but model assumptions can strongly affect predictions,
-%including whether variation is due to rare or common alleles \citep{slatkin1994segregation},
-%and the shape of the fitness landscape \citep{fraisse2016genetics}.
-%Our main aim is to provide a concrete framework 
-%that can provide natural predictions of these model parameters
-%across a general class of models.
-%Furthermore, tools from system theory allow analytical predictions to be made
-%for large populations with complex phenotypes
-%that would be inaccessible to population simulations.
-%
+The potential for speciation has been analyzed in models of 
+traits under stabilizing selection determined additively by alleles at many loci 
+\citep{wright1935evolution,barton1986maintenance,barton1989divergence,barton2001role},
+in related fitness landscape models \citep{fraisse2016genetics},
+and for pairs of traits that must match but whose value is unconstrained \citep{sved1981twosex}.
+% Recent work has shown that Fisher's geometric model of stabilizing selection
+% predicts many general patterns from the speciation literature \citep{fraisse2016genetics}.
+It has also been shown that population structure
+can allow long-term stable coexistence of incompatible genotypes encoding identical phenotypes \citep{phillips1996maintenance}. \revpoint{1}{9}
+However, previous simulations of system drift in regulatory sequences \citep{tulchinsky2014hybrid}
+and a regulatory cascade \citep{porter2002speciation}
+found rapid speciation under directional selection
+but only equivocal support for speciation under models of purely neutral drift.
+%Additionally, \citet{fierst} observed speciation due to an increase of segregation variance under balancing selection, in models of weak to moderate, but not strong epistasis.
+The rate at which hybrid incompatibility accumulates due to genetic drift creating
+segregation variance between isolated populations is fairly well understood
+\citep{slatkin1994segregation,rosas2010cryptic,chevin2014niche},
+but model assumptions can strongly affect predictions,
+including whether variation is due to rare or common alleles \citep{slatkin1994segregation},
+and the shape of the fitness landscape \citep{fraisse2016genetics}.
+Our main aim is to provide a concrete framework 
+that can provide natural predictions of these model parameters
+across a general class of models.
+Furthermore, tools from system theory allow analytical predictions to be made
+for large populations with complex phenotypes
+that would be inaccessible to population simulations.
 

### 2020-06-18 kalman_walk_paper.tex
added refs

@@
 \section*{Introduction}
-  \jss{I'm not sure we still need this first paragraph -- I don't feel like it says that much?}
 It is an overarching goal of many biological subdisciplines 
@@
 and empirically inspired
-\citep{mjolsness1991connectionist, jaeger2004dynamic, vitaly1, crombach2016gap, wotton2015quantitative, chertkova2017insilico}.
+  \citep{mjolsness1991connectionist, jaeger2004dynamic, vitaly1, crombach2016gap, wotton2015quantitative, chertkova2017insilico}. \jss{Something, something Wagner model here?}
 
@@
   This is a consequence of the often non-invertibility (i.e. degeneracy) of the mapping between a system's structure and function.
-Examples can be found across nearly all levels of biological organization from the level of the genetic code itself \citep{crick, woesse}
+Examples can be found across nearly all levels of biological organization from the level of the genetic code itself \citep{crick1963recent}
 all the way up to the convergent evolution of adaptive traits. Between these levels, structural plasticity can be found in the \emph{neutral networks}
-  of nucleic acid sequences that map to an RNA secondary structure \citep{gruner, andmore}\revpoint{1}{5} and amino acid sequences that fold similarly \citep{babajide},
+  of nucleic acid sequences that map to an RNA secondary structure \citep{gruner1996analysis}\revpoint{1}{5} and amino acid sequences that fold similarly \citep{babajide1997neutral},
 as well as in the thermodynamic stability of proteins \citep{hart2014thermodynamic}. Further examples include the vast space of functionally equivalent
@@
 in the logic of transcriptional \citep{tsong2006evolution, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription} 
-and neural circuits \citep{trojanowski}, and in developmental systems \citep{true2001developmental}. 
+and neural circuits \citep{trojanowski2014neural}, and in developmental systems \citep{true2001developmental}. 
 \revpoint{1}{1} \revpoint{1}{3}
@@
 
-Structural plasticity has been argued to be necessary for natural selection \citep{edleman2001degeneracy}, 
-as it may function as a mechanism for biological robustness and evolvability \citep{robustness}, or manifest as 
-\emph{canalization} \citep{canalization}. Degeneracy of the genetic code, and of genetic networks, 
-for instance, can make sequences more fault-tolerant to mutations and open up pathways to adaptation (cite). 
-Non-invertibility can also enable the evolution of neutral and nearly neutral molecular changes \citep{kimura},
-and perhaps even contribute to the formation of new species \citep{gavrilets}.
+Structural plasticity has been argued to be necessary for natural selection \citep{edelman2001degeneracy}, 
+as it may function as a mechanism for biological robustness and evolvability, or manifest as 
+\emph{canalization} \citep{whitacre2010degeneracy}. Degeneracy of the genetic code, and of genetic networks, 
+for instance, can make sequences more fault-tolerant to mutations \citep{sonneborn1965degeneracy} and open up pathways to adaptation \citep{wagner2008robustness}.
+Codon degeneracy can also enable the evolution of neutral and nearly neutral molecular changes \citep{kimura1968genetic},
+and perhaps even contribute to the formation of new species \citep{gavrilets2014models}.
 
@@
 %which is something related to the number of genes.
-even under universal pleiotropy \citep{universal_pleiotropy,omnigenic},
+even under universal pleiotropy \citep{universal_pleiotropy, boyle2017expanded},
 XXX

### 2020-06-18 kalman_walk_paper.tex
added refs

@@
 for instance, can make sequences more fault-tolerant to mutations \citep{sonneborn1965degeneracy} and open up pathways to adaptation \citep{wagner2008robustness}.
-Codon degeneracy can also enable the evolution of neutral and nearly neutral molecular changes \citep{kimura1968genetic},
+Structural plasticity can also enable the evolution of neutral and nearly neutral molecular changes \citep{kimura1968genetic},
 and perhaps even contribute to the formation of new species \citep{gavrilets2014models}.

### 2020-06-18 kalman_walk_paper.tex
added other refs

--- 2020-06-18 kalman_walk_paper.tex
@@
 \revpoint{1}{1} \revpoint{1}{3}
-  \jss{Other ideas instead of "degeneracy" or "isofunctional structural plasticity": non-invertibility, isofunctional plasticity, structural plasticity, phenotypic equivalency, phenotypic invariance, functional invariance, isofunctionality, functional equivalency, genotypic plasticity, conjugacy, functional symmetry, isologous, isomorphic.}
+  \jss{Other ideas instead of ``degeneracy'' or ``isofunctional structural plasticity'': non-invertibility, isofunctional plasticity, structural plasticity, phenotypic equivalency, phenotypic invariance, functional invariance, isofunctionality, functional equivalency, genotypic plasticity, conjugacy, functional symmetry, isologous, isomorphic.}
 
@@
 as it may function as a mechanism for biological robustness and evolvability, or manifest as 
-\emph{canalization} \citep{whitacre2010degeneracy}. Degeneracy of the genetic code, and of genetic networks, 
-for instance, can make sequences more fault-tolerant to mutations \citep{sonneborn1965degeneracy} and open up pathways to adaptation \citep{wagner2008robustness}.
-Codon degeneracy can also enable the evolution of neutral and nearly neutral molecular changes \citep{kimura1968genetic},
+\emph{canalization} \citep{whitacre2010degeneracy}. Structural plasticity of the genetic code, and of genetic networks, 
+for instance, can make sequences more fault-tolerant to mutations \citep{sonneborn1965degeneracy} and open up pathways to adaptation \citep{wagner2008robustness}.\jss{Is it clear what ``structural plasticity of the genetic code'', as opposed to ``degeneracy of the genetic code'', means here?}
+Structural plasticity can also enable the evolution of neutral and nearly neutral molecular changes \citep{kimura1968genetic},
 and perhaps even contribute to the formation of new species \citep{gavrilets2014models}.
@@
 all linear gene network architectures that yield identical phenotypes.
-Finally, we briefly outline and apply a quantitative genetics framework to estimate the speed at which system drift can lead to speciation via reproductive incompatibility,
-which we develop in more detail in a companion paper \citep{schiffman2020genetic}.
+Finally, we briefly outline and apply a quantitative genetics framework to estimate the speed at which system drift can lead to speciation via reproductive incompatibility.
+In this approach, a population diffuses along the neutral ridges of a high-dimensional space of possible system parameters, in a similar vein as \emph{holey landscape} models \citep{gavrilets1997evolution, yamaguchi2013first, pina2019does}\revpoint{2}{12}.
+Our model is developed in more detail in a companion paper \citep{schiffman2020genetic}.
 

### 2020-06-19 kalman_walk_paper.tex
intro edit

--- 2020-06-18 kalman_walk_paper.tex
@@
 \section*{Introduction}
+
 It is an overarching goal of many biological subdisciplines 
@@
 There is a substantial amount of work studying the evolution of gene regulatory networks, in frameworks 
-both abstract \citep{wagner1994evolution, wagner1996does, siegal2002waddington, bergman2003evolutionary, draghi2015robustness} 
+both abstract \citep{wagner1994evolution, wagner1996does, siegal2002waddington, bergman2003evolutionary, draghi2015robustness} \revpoint{1}{6}
 and empirically inspired
-  \citep{mjolsness1991connectionist, jaeger2004dynamic, vitaly1, crombach2016gap, wotton2015quantitative, chertkova2017insilico}. \jss{Something, something Wagner model here?}
-
- % In biological systems, the relationship between structure and function is often not invertible -- that is information about function is not usually sufficient to determine structure.
- % Non-invertiblility thus entails a diversity of functionally equivalent structures.
-  Biological systems, at nearly all levels of organization, display considerable degeneracy, or (perhaps more descriptively) a capacity for isofunctional \emph{structural plasticity} -- 
-  that is systems can be structurally different yet remain functionally equivalent \citep{edelman2001degeneracy}. 
-  This is a consequence of the often non-invertibility (i.e. degeneracy) of the mapping between a system's structure and function.
-Examples can be found across nearly all levels of biological organization from the level of the genetic code itself \citep{crick1963recent}
-all the way up to the convergent evolution of adaptive traits. Between these levels, structural plasticity can be found in the \emph{neutral networks}
-  of nucleic acid sequences that map to an RNA secondary structure \citep{gruner1996analysis}\revpoint{1}{5} and amino acid sequences that fold similarly \citep{babajide1997neutral},
-as well as in the thermodynamic stability of proteins \citep{hart2014thermodynamic}. Further examples include the vast space of functionally equivalent
+\citep{mjolsness1991connectionist, jaeger2004dynamic, vitaly1, crombach2016gap, wotton2015quantitative, chertkova2017insilico}.
+\jss{Something, something Wagner model here?}
+
+At all levels of biological organization,
+the problems that biological systems have evolved to solve
+often do not have single solutions --
+systems can be structurally different yet remain functionally equivalent \citep{edelman2001degeneracy}. 
+Examples can be found across nearly all levels of biological organization from the level of the genetic code itself % \citep{crick1963recent}
+all the way up to the convergent evolution of adaptive traits.
+In many cases, these functionally equivalent structures
+can be explored through small, local changes to the structure that leave the function unchanged.
+For instance, there are ``neutral networks''
+of nucleic acid sequences that produce the same RNA secondary structure \citep{gruner1996analysis}\revpoint{1}{5},
+amino acid sequences that fold similarly \citep{babajide1997neutral},
+or proteins with equivalent thermodynamic stability \citep{hart2014thermodynamic}.
+Further examples are found in the vast space of functionally equivalent
 potential regulatory sequences \citep{hare2008sepsid}, 
@@
 \revpoint{1}{1} \revpoint{1}{3}
-  \jss{Other ideas instead of "degeneracy" or "isofunctional structural plasticity": non-invertibility, isofunctional plasticity, structural plasticity, phenotypic equivalency, phenotypic invariance, functional invariance, isofunctionality, functional equivalency, genotypic plasticity, conjugacy, functional symmetry, isologous, isomorphic.}
-
-Likewise, it is well known that in many contexts mathematical models
-can fundamentally be \emph{nonidentifiable} and/or \emph{indistinguishable} -- meaning that 
+% \jss{Other ideas instead of ``degeneracy'' or ``isofunctional structural plasticity'': non-invertibility, isofunctional plasticity, structural plasticity, phenotypic equivalency, phenotypic invariance, functional invariance, isofunctionality, functional equivalency, genotypic plasticity, conjugacy, functional symmetry, isologous, isomorphic.}
+% \plr{'structural plasticity' means something specific in neuroscience}
+
+This capacity for isofunctional yet distinct mechanisms, sometimes called \emph{degeneracy},
+is a consequence of a many-to-one mapping between a system's structure and function,
+a concept that has been explored in many fields beyond biology.
+For instance, in many contexts mathematical models
+are fundamentally \emph{nonidentifiable} and/or \emph{indistinguishable} -- meaning that 
 there can be uncertainty about an inferred model's parameters or even its claims about
-causal structure, despite access to complete and perfect data \citep{bellman1970structural, grewal1976identifiability, walter1984structural}. 
+causal structure, despite access to complete and perfect data \citep[e.g.,][]{bellman1970structural, grewal1976identifiability, walter1984structural}. 
 Models with different parameter schemes, or even different mechanics 
@@
 referred to as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}.
-In computer science, this is framed as the relationship among processes that \emph{simulate} one another \citep{van2004equivalence}.
+In computer science, this has been framed as the relationship among processes that \emph{simulate} one another \citep{van2004equivalence}.
 Finally,
@@
 
-In this paper we use linear systems theory to model gene regulatory networks, linking structural plasticity 
-with mathematical nonidentifiability. Although our focus is on linear systems, nonidentifability is by no 
-means unique to such frameworks, and is a general property of many nonlinear models \citep{nonlinear}.
-
-Structural plasticity has been argued to be necessary for natural selection \citep{edelman2001degeneracy}, 
+It has been argued that the ability to modify structure without affecting function
+is necessary for natural selection \citep{edelman2001degeneracy}, 
 as it may function as a mechanism for biological robustness and evolvability, or manifest as 
-\emph{canalization} \citep{whitacre2010degeneracy}. Degeneracy of the genetic code, and of genetic networks, 
-for instance, can make sequences more fault-tolerant to mutations \citep{sonneborn1965degeneracy} and open up pathways to adaptation \citep{wagner2008robustness}.
-Codon degeneracy can also enable the evolution of neutral and nearly neutral molecular changes \citep{kimura1968genetic},
-and perhaps even contribute to the formation of new species \citep{gavrilets2014models}.
-
-We are interested in the nature and evolutionary implications of structural plasticity and nonidentifiability in the context of gene regulatory networks.
-Specifically, we would like to understand the degree of which gene regulatory network architectures are functionally equivalent and how this relates to network size and efficiency.
-If systems architectures are not functionally unique, can this open up neutral evolutionary paths, and do
-  these paths manifest as \emph{developmental system drift} \citep{true2001developmental} and contribute to speciation?
-First, we apply results from system theory which give
+\emph{canalization} \citep{whitacre2010degeneracy}.
+It may even contribute to the formation of new species \citep{gavrilets2014models}. 
+Redundancy of the genetic code, for instance,
+can make sequences more fault-tolerant to mutations \citep{sonneborn1965degeneracy},
+and robustness to modification of genetic networks can allow adaptation 
+without passing through a fitness valley \citep{wagner2008robustness}.
+
+In this paper we use results on mathematical nonidentifiability from
+linear systems theory to study how gene regulatory networks can be modified
+while retaining the same function, and what implications does this have for speciation.
+% Specifically, we would like to understand the degree of which gene regulatory network architectures are functionally equivalent and how this relates to network size and efficiency.
+If system architectures are not functionally unique,
+can this open up neutral evolutionary paths,
+and do these paths manifest as \emph{developmental system drift} \citep{true2001developmental}?
+Is this fast enough to contribute meaningfully to speciation?
+To do this, we describe results on linear dynamical systems which give
 an analytical description of the set of 
-all linear gene network architectures that yield identical phenotypes.
-Finally, we briefly outline and apply a quantitative genetics framework to estimate the speed at which system drift can lead to speciation via reproductive incompatibility,
-which we develop in more detail in a companion paper \citep{schiffman2020genetic}.
-
-%
-%It is well known that in many contexts mathematical models
-%can fundamentally be \emph{nonidentifiable} and/or \emph{indistinguishable} -- meaning that 
-%there can be uncertainty about an inferred model's parameters or even its claims about
-%causal structure, despite access to complete and perfect data \citep{bellman1970structural, grewal1976identifiability, walter1984structural}. 
-%Models with different parameter schemes, or even different mechanics 
-%can make equally accurate predictions,
-%but still not actually reflect the internal dynamics of the system being modelled.
-%In control theory, where electrical circuits and mechanical systems are often the focus, 
-%it is understood that there can be an infinite number of ``realizations'', 
-%or ways to reverse engineer the dynamics of a ``black box'',
-%even if all possible input and output experiments are performed 
-%\citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. 
-%The inherent nonidentifiability of chemical reaction networks is sometimes referred to as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}. 
-%In computer science, this is framed as the relationship among processes that \emph{simulate} one another \citep{van2004equivalence}.
-%Finally,
-%the field of \emph{inverse problems} studies those cases in which,
-%despite the existence of a theoretical one-to-one mapping between a model and behavior,
-%tiny amounts of noise make inference problems nonidentifiable in practice \citep{petrov2005well}.
-%
-%Nonidentifiability is a major barrier to mechanistic understanding of real systems, 
-%but viewed from another angle,
-%this concept can provide a starting point for thinking about externally equivalent systems
-%-- systems that evolution can explore, so long as the parameters and structures can be realized biologically.
-%These functional symmetries manifest in convergent and parallel evolution, 
-%as well as \emph{developmental system drift}: the observation that
-%macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level 
-%\citep{kimura1981possibility, true2001developmental, tanay2005conservation, tsong2006evolution, hare2008sepsid, lavoie2010evolutionary, vierstra2014mouse, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription}.
-%
-%The main purpose of this paper is to explore this general idea
-%in the concrete framework of linear systems theory,
-%thought of as modeling gene regulatory networks.
-%First, we apply results from system theory which give
-%an analytical description of the set of 
-%all linear gene network architectures that yield identical phenotypes.
-%This suggests that, quite generally, there are many directions in which a gene network can drift
-%without changing the phenotype.
-%Since these phenotypically equivalent gene networks are not necessarily compatible with one another, 
-%system drift may result in reproductive incompatibility between isolated populations,
-%even in the absence of any sort of adaptive, or environmental change.
-%We then briefly outline a quantitative genetics framework,
-%which we further develop in a companion paper,
-%which estimates how quickly system drift (and more generally, the presence of a fitness ridge)
-%leads to reproductive incompatibility \citep{schiffman2020genetic}.
-%
-% % \jss{should hare2008sepsid be here? It's sequence but not ``system'' drift}
-%% are these good here? {taylor2016diverse -- maybe not, matsui2015regulatory -- I think this relevant, stergachis2014conservation}.
-%%
-%%AND THESE?
-%  %% Over the last several years, different computational models have been applied to study reproductive incompatibility and speciation in models of gene regulation \citep{porter2002speciation, tulchinsky, palmer2009dynamics, tulchinsky2014hybrid, khatari2015simple}.
-%  %\citet{tulchinsky} simulated the evolution of a transcription factor and its binding site using a thermodynamic model.
-Their simulations suggest that the language by which a transcription factor recognizes its binding site can change, and potentially lead to hybrid incompatibility when allopatric populations employ divergent readout languages.
-This study, despite looking at gene regulation, does not analyze overall gene network architecture -- as we do here -- it only looks at the expression level of a single gene.
-Furthermore, they report reproductive isolation primarily following directional selection for a change in expression levels in each allopatric population; the evidence for reproductive isolation following balancing selection is much weaker.
-Johnson and Porter 2000 did not observe any hybrid fitness declines under stabilizing selection -- only under directional selection.
-%%Khatari et al, Tulchinsky et al, and Porter et al, all study hybrid incompatibility from a transcription factor/binding site interaction perspective, not from an overall network architecture perspective.
-Palmer and Feldman only see hybrid incompatibility in constant environments if the parental populations are relatively poorly adapted initially.
-Otherwise hybrids between two allopatric populations have fairly high fitnesses.
-%
-%% wright1932roles mentions possibility of equivalent combinations
-%% wright1935evolution looks at allele freq dynamics under stabilizing selection on a polygenic quantitative trait and discusses possibility for speciation this way
-%% barton1989divergence following on from barton1986maintenance analyzing multiple optima in a similar model, looking at RI due to shifts between equilibria
-%% nei1983models simulates from simple stepwise mut model on one or two loci (without interactions) where incompat betw diff alleles increases with number of steps betw them
-%% sved1981twosex mathematically analyzes and (in another paper) simulates speciation in a model of male + female quantitative traits indep'ly evolving with attraction between matching traits
-%% barton2010natural has a good historical review of natural selection and speciation; then analyzes a two-niche levene model
-%% gavrilets2004fitness talks a lot about quasineutral paths in holey fitness landscapes
-%% phillips1996maintenance shows that subpop structure can stably maintain equivalent but incompat genotypes
-%% tulchinsky2014hybrid simulates a single promoter region with a bioenergentic model with directional and stabilizing selection, found incompat in both cases
-%% porter2002speciation simulated speciation under a simple model of a regulatory cascade
-%%%% Fisher's geom model: (incompat due to segregation variance)
-%% poon2000compensating also uses SSSWM with Fisher's model and derives some analytical results for load in small populations; early citation for this process
-%% slatkin1994segregation shows that segregation variance increases linearly with divergence if variation is controlled by intermediate freq alleles;
-%%   but if it is determined by rare alleles (at mut-sel balance), then it plateaus (at twice the within-pop value?)
-%% barton2001role looks at hybrids betw pops adapting to shifting optima in Fisher's geometric model : under fluctuating optima
-%%       hybrids will suffer because of cryptic segregating variation
-%% fraisse2016genetics uses SSWM; looks at stabilizing selection using Fisher's geometric model, showing it predicts lots of observations;
-%%        but mostly only if fitness is exp(-|x|^k) for k > 2
-%% chevin2014niche uses SSWM; looks at incompats after adaptation in Fisher's geom model and how it depends on dimension;
-%%        uses generalized Fisher's model with neutral directions; terms hybrid unfitness due to segregation variance "transgressive incompats";
-%%        finds rate of RI accumulation decreases with Ne (like Ne^(-m/2), with m the trait dimension?)
-%% simon2017coadapted follows up on the Fisher's geom model predictions, also comparing to some real data
-%% martin2014fishers shows that you can get Fisher's model quite generally if (a) the kryptotype dimension is large compared to the phenotype dimension,
-%%   and (b) mutational effects fall under the CLT
-%% rosas2010cryptic looks at segregation variance resulting in deviations from optimum phenotype including off of curvey ridges
-%%%%% General notes
-%% citation for segregation variance around a stationary optimum -> load is Poon & Otto 2000; rosas2010cryptic
-%
+all linear gene network architectures that yield identical phenotypes,
+and use quantitative genetics theory to estimate the speed at which system drift can lead to reproductive incompatibility and hence speciation.
+In this model,
+a population diffuses along the neutral ridges of a high-dimensional space of possible system parameters,
+in a similar vein as \emph{holey landscape} models \citep{gavrilets1997evolution, yamaguchi2013first, pina2019does}\revpoint{2}{12}.
+% Our model is developed in more detail in a companion paper \citep{schiffman2020genetic}.
+
 %It is not a new observation that there is often more than one way to do the same thing, 

### 2020-06-26 kalman_walk_paper.tex
edits

@@
 this could lead to substantial interpopulation incompatibility,
-and thus speciation, in fewer than $N_e$ generations.
+and thus speciation, on a time scale of $N_e$ generations.
 This model also naturally predicts Haldane's rule, 
@@
 linear systems theory to study how gene regulatory networks can be modified
-while retaining the same function, and what implications does this have for speciation.
+while retaining the same function, and the possible implications for speciation.
 % Specifically, we would like to understand the degree of which gene regulatory network architectures are functionally equivalent and how this relates to network size and efficiency.
@@
 
-%It is not a new observation that there is often more than one way to do the same thing, 
-%or that speciation can be the result of (nearly) neutral processes.
+Besides those already cited, many other authors have observed
+% It is not a new observation
+that there is often more than one way to do the same thing, 
+and that speciation can be the result of (nearly) neutral processes.
 The potential for speciation has been analyzed in models of 
@@
 into the $\ell$-dimensional output $\phi(t)$.
-Under quite general assumptions (\emph{e.g.} the system is linear, time-invariant and in the zero-state), \revpoint{AE}{3}
+Under quite general assumptions on the input
+(e.g., $|u(t)|$ is integrable) \revpoint{AE}{3}
 we can write the phenotype as \revpoint{AE}{4}
   \begin{align}
-    \phi(t) = \int_{0}^{t} C e^{A (t-s)} B u(s) ds ,
+    \phi(t) = \int_{0}^{t} C e^{(t-s) A} B u(s) ds ,
   \end{align}
@@
     the line of optimal system coefficients is down in black;
-    dotted lines give phenotypic distances to the optimum.
+    solid lines give phenotypic distances to the optimum.
     A pair of parental populations are shown in black, along the optimum;
     a hypothetical population of $F_1$s are shown in red,
-    and the distribution of one type of $F_2$ is shown in purple
-    (other types of $F_2$ are not shown; 
-    some would be a similar distance to the other side of the optimal set).
-    The distribution of $F_2$ hybrids is appropriate for mixed homozygotes
-    if both traits have a simple, one-locus genetic basis,
-    but there is variation within each population at that locus.
+    and the distribution of $F_2$s is shown in purple.
+    The two figures differ in the genetic basis, and hence,
+    the distribution of $F_2$ phenotypes:
+    \textbf{(left)}
+    $F_2$s compose all four mixed homozygotes if variation at both
+    traits has a simple, one-locus genetic basis in both popualtions; and
+    \textbf{(right)}
+    $F_2$ show a much wider distribution of phenotypes
+    if the genetic basis of variation in each population is polygenic.
 }
@@
 The population mean of an unconstrained quantitative trait
-with genetic variance $V_G$ in a population with effective size $N_e$
-will move in $t$ generations a random amount whose variance is $t V_G / N_e$ \citep{quant_gen}.
+with additive genetic variance $V_G$ in a population with effective size $N_e$
+will move in $t$ generations a random amount whose variance is $t V_G / N_e$ \citep{lande1976natural}.
 The mean difference between two such populations has twice the variance.
@@
 will produce $F_2$ offspring that differ from optimal by an amount proportional to $\sqrt{t V_G / N_e}$.
-Since we assume they are at a local figness optimum, without much loss of generality we can assume
+Since we assume they are at a local fitness optimum, without much loss of generality we can assume
 that fitness is locally quadratic, and so $F_2$ fitness decays linearly in time:
-proportionally to $t V_G / N_e$ -- fastest in small, diverse populations.
+proportionally to $t V_G / N_e$ -- fastest in small, diverse populations,
+This predicts that we need only wait some multiple of $N_e$ generations
+until substantial incompatibility has been accumulated.
 
@@
 translate to segregation variance in $F_2$s in selectively constrained directions.
-To move the the system (the $A$ matrix) a given distance
+To move the system (the $A$ matrix) a given distance
 generally involves moving many individual interaction coefficients (the entries $A_{ij}$).
@@
 to remain on the fitness ridge
-is unlikely to produce a a set of coordinated changes;
+is unlikely to produce a set of coordinated changes;
 and the resulting system could move away from the ridge in almost any direction.
@@
 
-Neutral processes are certainly not the only drivers of speciation.
+These explanations are not mutually exclusive.
 All of these forces -- adaptive shifts, conflict, and network drift --
@@
 \paragraph{The dimensionality of trait space} \revpoint{1}{8}
-\jss{I didn't think that our results depend on the kryptotype being in a higher dimensional space than the phenotype.
-Any coordinate change on (A,B,C) will rewire the system. So as long as none of the system matrices are constrained, 
-any minimal system should have an infinite number of realizations. Unless my system theory is rusty?}
-
 We have focused on examples of single traits (where the phenotype is one-dimensional),
 but phenotypes under selection are often high-dimensional,
-and different traits often share a genetic basis.
+and variation in different traits often share a genetic basis.
+However, we still expect many degrees of freedom
+as long as there are components of the system not directly and individually constrained by selection
+(i.e., a kryptotype). % since if C = I then there's no nontrivial transformations that preserve C
 Even in networks where the phenotype and kryptotype are of the same dimension,
@@
 Some phenotypes, however, require kryptotypic dimensions to be larger than that of the phenotype.
-For instance, there exists minimal realizations (\emph{e.g.} the oscillator in Example 1) where the dimension of the kryptotpye is larger
+For instance, many systems have minimal realizations (\emph{e.g.}, the oscillator in Example \ref{ex:all_osc}) where the dimension of the kryptotype is larger
 than that of the phenotype, implying that for these phenotypic dynamics to be realized, the kryptotype dimension \emph{has} to be larger than the dimension of the phenotype.
-Of course we see many more degrees of freedom in nonminimal networks, but it is not yet clear if evolution commonly arrives at such structures.
-There are some hints that this may be occuring: compared to the molecular circuitry of the cyanobacteria circadian rhythm,
-the human clock appears to be highly and unneccessarily complex \citep{sancar2008intelligent}. 
-The true ``dimensionality of trait space'' has been debated at some length in the literature
-\citep{trait_space}.
-Whatever it is, however, it seems likely to be highly degenerate,
-%Whatever it is, however, it seems unlikely 
-%that it would approach the dimensionality of kryptotype space,
-%which is something related to the number of genes.
-even under universal pleiotropy \citep{universal_pleiotropy, boyle2017expanded},
-XXX
+Even if components of the system's internal state are directly subject to selection
+and the mode of action of the environment on the internal state is constrained
+(so, the input and output matrices $B$ and $C$ are fixed)
+then one could still perturb $A$ as described above by $ZA - AZ$
+if $ZB$ and $CZ$ are both zero, implying a number of degrees of freedom that still grows with $n^2$
+for fixed $\ell$ and $m$.
+Generically, the number of degrees of freedom is $n(n - \ell - m)$,
+so that in a system of $n$ components, if even one component is not directly constrained,
+this leads to $n$ degrees of freedom.
+Whatever the true ``dimensionality'' of phenotype space of a typical organism,
+there are undoubtedly aspects of its underlying molecular machinery that are not directly constrained,
+suggesting large numbers of degrees of freedom.
+Note that pleiotropy does not directly affect this argument at all --
+indeed, many phenotypically equivalent changes will lead to denser $A$ matrices and hence more pleiotropy.
+However, more pleiotropic genes may be more strongly constrained,
+making it more difficult for systems to make the required compensatory changes for system drift.
+% Whatever it is, however, it seems unlikely 
+% that it would approach the dimensionality of kryptotype space,
+% which is something related to the number of genes.
+% even under universal pleiotropy \citep{universal_pleiotropy, boyle2017expanded},
 
@@
 implying saturating response curves.
-It remains to be seen how well these results carry over into real systems,
+It remains to be seen how well our results carry over into real systems,
 but the fact that most nonlinear systems can be locally approximated by a linear one
@@
 Finally, despite our model's precise separation of phenotype and kryptotype, this relationship in nature may be far more complicated as aspects of the kryptotype may be less ``hidden'' than we currently assume, and the neutral network changes we describe here may only be nearly neutral.
-For instance, attributes excluded from the phenotype as modeled here, ignore the potential energy costs associated with excessively large (non-minimal) kryptotypes, as well as the relationship between a specific network architecture and robustness to mutational, transcriptional, or environmental noise.
+For instance, attributes excluded from the phenotype as modeled here ignore the potential energy costs associated with excessively large (non-minimal) kryptotypes, as well as the relationship between a specific network architecture and robustness to mutational, transcriptional, or environmental noise.
 % Despite the model's simplifications and potentially nontrivial oversights, it may still roughly reflect and shed light upon speciation patterns as they occur in nature.

### 2020-06-26 kalman_walk_paper.tex
remade figure 5 hybrid plot

@@
   \centering
-  \begin{tabular}{ccc}
-    %\includegraphics[width=0.5\textwidth, height=0.25\paperheight]{F1_comparison} &
-    %\includegraphics[width=0.5\textwidth, height=0.25\paperheight]{F2s_comparison2}
-    \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F1_hundreth_tau0} &
-      \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F2_hundrethtau0} \\
-      \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F1_tenth_tau0.pdf} &
-      \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F2_tenthtau0.pdf} \\
-      \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F1_half_tau0.pdf} &
-      \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F2_halftau0.pdf}
-  \end{tabular}
+ % \begin{tabular}{ccc}
+ %   \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F1_hundreth_tau0} &
+ %     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F2_hundrethtau0} \\
+ %     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F1_tenth_tau0.pdf} &
+ %     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F2_tenthtau0.pdf} \\
+ %     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F1_half_tau0.pdf} &
+ %     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F2_halftau0.pdf}
+ % \end{tabular}
+  \includegraphics[scale=0.5]{examples/f1f2_newplot}
   \caption{

### 2020-06-26 examples/sd_fig5.Rmd
code for fig5 plots

---
title: "sd_fig5"
author: "Josh Schiffman"
date: "6/26/2020"
output: html_document
---

```{r, cache=TRUE}
library(expm)
# System drift Figure 5 plot

A0 <- matrix(c(0,-1,1,0), nrow=2)
B <- matrix(c(1,1), ncol=1)
C <- matrix(c(1,0), nrow=1)

A_fn <- function (tau) { 
    matrix(c(tau/(tau+1), -(2*tau*(tau+1)+1)/(tau+1), 
             1/(tau+1), -tau/(tau+1)), nrow=2) 
}

tt <- seq(0,6*pi, length.out=500)

h <- function (t,tau=0,A=A_fn(tau)) { sapply(t, function (tt) C %*% expm(tt*A) %*% B) }

F1 <- function (tau1,tau2) { (A_fn(tau1)+A_fn(tau2))/2 }

get_F2s <- function(tau1, tau2) 
    {
        vc <- function(tau1, tau2, x, y) 
            {
                out <- c(A_fn(tau1)[x,y], A_fn(tau2)[x,y])
                return(out)
            }
        mlist <- list()
        count <- 1
        for (i in 1:2) 
            {
                for (j in 1:2) 
                    {
                        for (k in 1:2) 
                            {
                                for (l in 1:2)
                                    {
                                        mlist[[count]] <- matrix(c(vc(tau1, tau2, 1, 1)[i], vc(tau1, tau2, 2, 1)[j], vc(tau1, tau2, 1, 2)[k], vc(tau1, tau2, 2, 2)[l]),2,2)
count <- count + 1
                                    }
                            }
                }
            }
        f2list <- list()
        count2 <- 1
        for(i in 1:length(mlist)) 
            {
                for (j in 1:length(mlist))
                    {
                        f2list[[count2]] <- (mlist[[i]] + mlist[[j]])/2
                        count2 <- count2 + 1
                    }
            }
    return(unique(f2list))
}

hhf0 <- sapply( list(A_fn(0), A_fn(0)), function (A) { sapply(tt, h, A=A) } )

hhf1_1 <- sapply( list(A_fn(0), A_fn(0.01), F1(0, 0.01)), function (A) { sapply(tt, h, A=A) } )
hhf1_2 <- sapply( list(A_fn(0), A_fn(0.1), F1(0, 0.1)), function (A) { sapply(tt, h, A=A) } )
hhf1_3 <- sapply( list(A_fn(0), A_fn(0.5), F1(0, 0.5)), function (A) { sapply(tt, h, A=A) } )

hhf2_1 <- sapply( get_F2s(0, 0.01), function (A) { sapply(tt, h, A=A) } )
hhf2_2 <- sapply( get_F2s(0, 0.1), function (A) { sapply(tt, h, A=A) } )
hhf2_3 <- sapply( get_F2s(0, 0.5), function (A) { sapply(tt, h, A=A) } )

```

```{r, fig.align="center", fig.width=14, fig.height=10}

par(mfrow=c(3,2))
matplot(tt, hhf1_1, type='l', xlab="time", ylab="phenotype h(t)", lwd=2, ylim=c(-6,6), col=c("blue", "black"), cex.lab=1.5)
matplot(tt, hhf2_1, type='l', xlab="time", ylab="phenotype h(t)", lty=1, lwd=2, ylim=c(-6,6), cex.lab=1.5)
matlines(tt, hhf0, type='l', xlab="time", ylab="phenotype h(t)", lty=1, ylim=c(-6,6), cex.lab=1.5, col="black", lwd=4)
matplot(tt, hhf1_2, type='l', xlab="time", ylab="phenotype h(t)", lwd=2, ylim=c(-6,6), col=c("blue", "black"), cex.lab=1.5)
matplot(tt, hhf2_2, type='l', xlab="time", ylab="phenotype h(t)", lty=1, lwd=2, ylim=c(-6,6), cex.lab=1.5)
matlines(tt, hhf0, type='l', xlab="time", ylab="phenotype h(t)", lty=1, ylim=c(-6,6), cex.lab=1.5, col="black", lwd=4)
matplot(tt, hhf1_3, type='l', xlab="time", ylab="phenotype h(t)", lwd=2, ylim=c(-6,6), col=c("blue", "black"), cex.lab=1.5)
matplot(tt, hhf2_3, type='l', xlab="time", ylab="phenotype h(t)", lty=1, lwd=2, ylim=c(-6,6), cex.lab=1.5)
matlines(tt, hhf0, type='l', xlab="time", ylab="phenotype h(t)", lty=1, ylim=c(-6,6), cex.lab=1.5, col="black", lwd=4)

```

### 2020-07-02 review-responses.tex
added refs
added other refs
intro edit
edited responses

--- 2018-08-19 review-responses.tex
@@
 
+Thanks for the careful attention to the paper.
+We have indeed decided to go with this suggestion,
+and have split the paper in two;
+this paper is the ``systems biology'' part of it.
+We've left in reviewer comments regarding the ``Fisher's geometric model'' portion below,
+commenting that the relevant bit ``has been removed''.
+
 \begin{point}{Format:}
@@
 \reply{
+    Apologies!
+  We've added line numbers and corrected improperly formatted citations. 
 }
@@
 \reply{
+  Good idea.
+  The formula is quite general,
+  and a more precise statement of the assumptions would veer too far afield
+  (perhaps needing definitions of function spaces),
+  but we've added a hopefully clarifying comment.
+  \revref.
 }
@@
 \reply{
-    We've added a clarifying note \revref
+    We've added a clarifying note \revref{}
     but we're hoping that readers not familiar with linear algebra will realize
@@
 \reply{
+  Thanks for the additional suggestions!
+  We've reworked the introduction, adding these (and many other) additional references. \revref
 }
@@
 \reply{
+    We've tried to explain things better in the revised Introduction. \revref
 }
@@
 \reply{
+    Yes, that's right! We've followed the suggestions (see below),
+    and have added additional discussion to the Discussion.
 }
@@
   model (iterated matrix multiplication). Variants of this have later been used
-by others (eg Draghi and GP Wagner, I think). How is your model related to
-these approaches?
+by others (eg Draghi and GP Wagner, I think). How is your model related to these approaches?
 \end{point}
@@
 \reply{
+    This is a natural quesiton. We've addressed it after introducting the model.
+    \revref
 }
@@
 \reply{
+    We do agree, and have clarified this in the Discussion. \revref
 }
@@
 \reply{
+    Counterintuitively, it is not necessary to have the kryptotype have a higher dimension
+    than the phenotype to have degrees of freedom.
+    The precise details depend on what you want to treat as fixed or evolvable --
+    for instance, should the output matrix $C$ be subject to drift? --
+    but we don't think it's controversial to suppose that there's a bunch of genes
+    whose expression levels aren't the direct target of selection.
     We have added discussion of this in the Discussion. \revref
@@
 \reply{
+    That'd be one way to do it --
+    and, it turns out, Bateson thought BDMIs would probably be neutral;
+    we've added this. \revref{}
+    (The quote from \citet{bateson1909heredity} is
+    \textit{``Now if the sterility
+of the cross-bred be really the consequence of the meeting of two
+complementary factors, we see that the phenomenon could only be
+produced among the divergent offspring of one species by the acquisition of at least two new factors; for if the acquisition of a single
+factor caused sterility the line would then end. Moreover each factor
+must be separately acquired by distinct individuals, for if both were
+present together, the possessors would by hypothesis be sterile. And
+in order to imitate the case of species each of these factors must be
+acquired by distinct breeds. The factors need not, and probably would
+not, produce any other perceptible effects; they might, like the colour-
+factors present in white flowers, make no difference in the form or
+other characters.``})
 }
@@
 \reply{
+    We wouldn't be surprised if it had, but we're not aware of it --
+    other papers all seem to take the discrete-time approach as in \citet{wagner1994evolution}.
 }
@@
 \reply{
+    We're not aware of any biological examples of ``equivalent'' networks
+    in more than one species at all,
+    but we think it's not clear whether we understand the dynamics of \emph{any} gene regulatory network
+    well enough to determine equivalence,
+    so this is an absence of evidence, not evidence of absence.
+    (There are plenty of examples of homologous networks with missing genes,
+    but there's so many uncertainties we don't think bringing this up will clarify things.)
+    \plr{JOSH: how's this?}
 }
@@
 \reply{
+    \plr{FIXME}
 }
@@
 \reply{
-    \plr{I'm leaving these points because they might apply to whatever handwaving 
-    we replace the quant gen stuff with.}
-}
-
-\begin{point}{}
-    Your analysis of the fitness loss in F1 and F2 hybrids rests on a local
-Taylor expansion of the fitness landscape. This is adequate given that the
-underlying linear network is also only locally valid (see above). However, you
-then apply this to a discussion of hybrid incompatibility (Haldane's rule etc).
-We are thus interested in *very large* fitness costs for hybrids. This does not
-seems to be compatible: to get hybrid incompatibility or sterility, you need to
-have epsilon sufficiently large (below equation 5). But then $\epsilon^2$ is no
-longer smaller than epsilon. In other words: it seems to me that for the
-discussion of hybrid incompatibilities you apply your model of the first part
-beyond the local range where it is valid.
-\end{point}
-
-\reply{
+    We've removed the ``popgen'' part, but some of the comments below still apply.
 }
@@
 \reply{
+    We have removed this part.
 }
@@
 \reply{
+    \plr{should put somethign about this in the discussion?}
 }
@@
 \reply{
+    \plr{... together with this?}
 }
@@
 \reply{
+    We aim to do that in the subsequent paper.
 }
@@
 \reply{
+    \plr{... and also this?}
 }
@@
 \reply{
+    We have removed this part.
+}
+
+\begin{point}{}
+    Your analysis of the fitness loss in F1 and F2 hybrids rests on a local
+Taylor expansion of the fitness landscape. This is adequate given that the
+underlying linear network is also only locally valid (see above). However, you
+then apply this to a discussion of hybrid incompatibility (Haldane's rule etc).
+We are thus interested in \emph{very large} fitness costs for hybrids. This does not
+seems to be compatible: to get hybrid incompatibility or sterility, you need to
+have epsilon sufficiently large (below equation 5). But then $\epsilon^2$ is no
+longer smaller than $\epsilon$. In other words: it seems to me that for the
+discussion of hybrid incompatibilities you apply your model of the first part
+beyond the local range where it is valid.
+\end{point}
+
+\reply{
+    This is a good point, but tricky since it's not clear what ``small'' is here.
+    A fitness difference of 0.2 is quite strong selection, for instance,
+    although it is a fairly small number in some ways.
+    Furthermore, the same picture holds in the opposing cases of a single locus
+    or many loci, illustrated in Figure~\ref{fig:conceptual_fig}.
+    In any case, we now rely only on the conceptual idea rather than the numerics of this estimate.
 }
@@
 
+\plr{TODO: respond/edit}
+
 \begin{point}{end of p6:} 
@@
 \reply{
-    XXX Suggestion: change this figure to have two figures of the same type, side-by-side,
-    one with a simple genetic basis (like this one, but with two purple clouds)
-    and another with a more quantitative trait.
+    We have improved Figure~\ref{fig:conceptual_fig} to hopefully better demonstrate the point.
 }
@@
     that we'd removed before submission. \revref
-    XXX TODO EDIT THAT APPENDIX XXX
 }
@@
 \reply{
+  Thank you for the references. We've added citations to the Introduction. \revref
 }

### 2020-07-02 kalman_walk_paper.tex
fixed an arrow in Fig1
edited responses

@@
 There is a substantial amount of work studying the evolution of gene regulatory networks, in frameworks 
-both abstract \citep{wagner1994evolution, wagner1996does, siegal2002waddington, bergman2003evolutionary, draghi2015robustness} \revpoint{1}{6}
+both abstract \citep{wagner1994evolution, wagner1996does, siegal2002waddington, bergman2003evolutionary, draghi2015robustness}
 and empirically inspired
 \citep{mjolsness1991connectionist, jaeger2004dynamic, vitaly1, crombach2016gap, wotton2015quantitative, chertkova2017insilico}.
-\jss{Something, something Wagner model here?}
 
@@
 and that speciation can be the result of (nearly) neutral processes.
+Indeed, \citet{bateson1909heredity} first proposed that what today we call a
+Bateson-Dobzhansky-Muller incompatibility would likely arise through neutral changes. \revpoint{1}{10}
 The potential for speciation has been analyzed in models of 
@@
 
-% Although many different biological systems can be modeled with this approach, for clarity, we focus on gene regulatory networks.
 In terms of gene regulatory networks, $A_{ij}$ determines how the $j^\text{th}$ transcription factor regulates the $i^\text{th}$ transcription factor.
@@
 % Furthermore, the ease of analyzing linear systems makes this an attractive place to start.
-To demonstrate this approach, we apply it to construct a simple gene network in Example \ref{ex:oscillator} below.
+
+\citet{wagner1994evolution} and others have used a similar discrete-time model \revpoint{1}{6}
+(that might be written $\phi_{t+1} = f(A \phi_t)$, where $f$ is a sigmoid).
+Our choice of continuous time does not affect the points we make here,
+but our restriction to \emph{linear} systems is a stronger assumption
+(see the Discussion).
+
+To demonstrate the model,
+we construct a simple gene network in Example \ref{ex:oscillator} below.
 
@@
                       every edge/.style={draw=black, thick}]
-            \path [->, >=Rectangle] (A) edge[bend left] node {\tiny $-1$} (B);
+            \path [->] (A) edge[bend left] node {\tiny $-1$} (B);
             \path [->] (B) edge[bend left] node {\tiny $1$} (A); 
@@
 but the fact that most nonlinear systems can be locally approximated by a linear one
-suggests our qualitative results may hold more generally.
+suggests our qualitative results may hold more generally. \revpoint{1}{7}
 Furthermore, nonidentifiability (which implies the existence of neutral directions)
@@
         \int_0^\infty \rho(t) \left| h_A(t) - h_0(t) \right|^2 dt \\
-        &:= 
+        &= 
         \int_0^\infty \rho(t) \left| C e^{At} B - C_0 e^{A_0 t} B_0 \right|^2 dt \\
@@
 
-By section \ref{ss:quant_gen},
-if we set $\Sigma=\sigma^2 I$ and $U=H$,
-then a population at $A_0+Z$ experiences a restoring force of strength
-$(I + \sigma^2 H^{-1})^{-1} Z$ (treating $Z$ as a vector and $H$ as an operator on these).
-If $\sigma^2$ is small compared to $H^{-1}$
-then this is approximately $-\sigma^2 H^{-1} Z$.
-This suggests that the population mean follows an Ornstein-Uhlenbeck process,
-as described (in different terms) in \citet{hansen1996translating}.
+% By section \ref{ss:quant_gen},
+% if we set $\Sigma=\sigma^2 I$ and $U=H$,
+% then a population at $A_0+Z$ experiences a restoring force of strength
+% $(I + \sigma^2 H^{-1})^{-1} Z$ (treating $Z$ as a vector and $H$ as an operator on these).
+% If $\sigma^2$ is small compared to $H^{-1}$
+% then this is approximately $-\sigma^2 H^{-1} Z$.
+% This suggests that the population mean follows an Ornstein-Uhlenbeck process,
+% as described (in different terms) in \citet{hansen1996translating}.
 

### 2020-07-03 kalman_walk_paper.tex
updated citations and discussion

@@
      \end{tikzpicture} &
-   \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{osc_impulse}
+     \raisebox{0.25\height}{\includegraphics[scale=0.25]{examples/fig1b}}
 \end{tabular}
@@
 \end{scope}
-\end{tikzpicture} &	\includegraphics[width=0.5\textwidth, height=0.125\paperheight]{figures/osc_kryp_compare}
+\end{tikzpicture} &	\raisebox{0.25\height}{\includegraphics[scale=0.25]{examples/fig2b}}
     \end{tabular}
@@
       (Left) $A(\tau)$, the set of all phenotype-equivalent cell cycle control networks.
-      (Right) Gene-1 dynamics (blue) for both systems $A(0)$ and $A(-2)$ are identical, however, $A(0)$ gene-2 dynamics (orange) differ from $A(-2)$ (green).
+      (Right) Gene-1 dynamics (blue) for both systems $A(0)$ and $A(-2)$ are identical, however, $A(0)$ gene-2 dynamics (red) differ from $A(-2)$ (green).
       % Both gene-1 dynamics are given by $\kappa_{1}(t) = \sin t + \cos t$, and gene-2 by $\kappa_{2}(t) = \cos t - \sin t$ ($A(0)$) and $\kappa'_{2}(t) = \cos t + 3 \sin t$ ($A(2)$).
@@
 
-Consistent with empirical observation of hybrid breakdown (e.g., \citet{plotner2017chlorosis}) and transgressive segregation \citep{transgressive}, 
+Consistent with empirical observation of hybrid breakdown (e.g., \citet{plotner2017chlorosis}), 
 we see that the fitnesses of $F_2$ hybrids drop at a much faster rate than those of $F_1$s.
@@
 then heterogametic $F_1$s show variation (and fitnesses) similar to that seen in $F_2$ hybrids.
+This observation appears to be similar to the extreme hybrid phenotypes produced by transgressive segregation \citep{rieseberg1999transgressive}, 
+which can manifest in $F_1$s when only one (dominant) parental allele is expressed at heterozygous loci; this was observed in hybrid gene expression patterns, and increased as a function of parental genetic distance \citep{stelkens2009genetic}.
 

### 2020-07-06 examples/sd_fig5.Rmd
code for fig5 plots
fixed figures

@@
 
-```{r, cache=TRUE}
+```{r}
 library(expm)
@@
 
-tt <- seq(0,6*pi, length.out=500)
+tt <- seq(0,6*2*pi, length.out=500)
 
@@
 ```
+
+```{r, fig.width=2.5, fig.height=1.5, fig.align="center"}
+
+C2 <- matrix(c(0,1), nrow=1)
+h.2 <- function (t,tau=0,A=A_fn(tau)) { sapply(t, function (tt) C2 %*% expm(tt*A) %*% B) }
+
+hhf00 <- sapply( list(A_fn(0)), function (A) { sapply(tt, h, A=A) } )
+hh02 <- sapply( list(A_fn(0)), function (A) { sapply(tt, h.2, A=A) } )
+hhn22 <- sapply( list(A_fn(-2)), function (A) { sapply(tt, h.2, A=A) } )
+
+par(mar=c(4.5, 4, 1, 1)+.1)
+plot(tt, hhf00, type="l", lwd=2, ylab= expression("phenotype" ~ phi(t)), xlab="time t", col="steelblue4", ylim=c(-3,3), cex.lab=1)
+
+par(mar=c(4.5, 4, 1, 1)+.1)
+plot(tt, hhf00, type="l", lwd=2, ylab= expression("kryptotype" ~ kappa(t)), xlab="time t", col="steelblue4", ylim=c(-3,3), cex.lab=1)
+lines(tt, hhn22, lwd=2, col="seagreen")
+lines(tt, hh02, lwd=2, col="red4")
+```

### 2020-07-06 examples/sd_fig5.Rmd
added code for plots

@@
 ```
+
+```{r fig6, fig.align="center", fig.width=5, fig.height=3, cache=TRUE}
+library(parallel)
+
+optimal_h <- function (t) { sin(t)+cos(t) }
+
+Df <- function (A,t) {
+    exp(-t/(4*pi)) * ( h(t,A=A) - optimal_h(t) )^2
+}
+
+D <- function (A, upper=10, ...) {
+    f <- function (t) { Df(A,t) }
+    integrate(f, lower=0, upper=upper, ...)$value
+}
+
+get_F2s_nonunique <- function(tau1, tau2) {
+    vc <- function(tau1, tau2, x, y) {
+    out <- c(A_fn(tau1)[x,y], A_fn(tau2)[x,y])
+    return(out)
+}
+mlist <- list()
+count <- 1
+for (i in 1:2) {
+    for (j in 1:2) {
+        for (k in 1:2)
+            for (l in 1:2) {
+                mlist[[count]] <- matrix(c(vc(tau1, tau2, 1, 1)[i], vc(tau1, tau2, 2, 1)[j], vc(tau1, tau2, 1, 2)[k], vc(tau1, tau2, 2, 2)[l]),2,2)
+                count <- count + 1
+            }
+    }
+}
+
+f2list <- list()
+count2 <- 1
+for(i in 1:length(mlist)) {
+    for (j in 1:length(mlist)) {
+        f2list[[count2]] <- (mlist[[i]] + mlist[[j]])/2
+        count2 <- count2 + 1
+    }
+}
+return(f2list)
+}
+
+sys_gen_dist <- function(d) {
+    sgd <- sqrt( sum( (A_fn(0) - A_fn(0+d))^2 ) )
+    return(sgd)
+}
+
+dvec <- seq(0, 0.5, length.out=20)
+
+D_F1 <- sapply(dvec, function (dtau) {
+D(A=F1(tau1=0, tau2=0+dtau), upper = 10 )
+} )
+
+D_F2 <- do.call(cbind, mclapply(dvec, function (dtau) {
+F2s <- get_F2s_nonunique(0, 0+dtau)
+sapply(F2s, D, upper=10 )
+}, mc.cores=detectCores() ) )
+
+plot( sapply(dvec, sys_gen_dist), sqrt(D_F1), type='l', ylab="phenotype distance (D)", xlab="genetic distance", ylim=c(0, 2), lwd=3)
+lines(sapply(dvec, sys_gen_dist), colMeans(sqrt(D_F2)), lty=1, type="l", ylim=c(0,2), col="steelblue4", lwd=3)
+legend(0, 2, legend = c( expression(F[1]~"hybrids"), expression(F[2]~"hybrids")), lty=1, lwd=3, col=c("black", "steelblue4"))
+
+```

### 2020-07-06 review-responses.tex
minor edits to responses

@@
 \reply{
+  We have regenerated figures to address this.
 }
@@
   We think that adding a third type of arrowhead may only add to the confusion, so now we are using only the standard arrowhead \revref.
-  \jss{we could add a cirlce arrowhead for the unknown direction, but I think its simpler to just get rid of the rectangular negative arrowhead originally in Figure 3}
 }
@@
     \plr{JOSH: how's this?}
+    \jss{Sounds great! We could also add:}
+    Over larger evolutionary distances there is some evidence that structurally distinct
+    isofunctional networks of different size do indeed exist. \emph{E.g.}, the networks underlying circadian
+    rhythm in cyanobacteria and mammals, as discussed by \citet{sancar2008intelligent}.
+    It seems possible, in principle, that, on a smaller scale, such differences could exist between more closely related populations.
+
 }
@@
 \reply{
-    \plr{FIXME}
+    We've regenerated figures to improve clarity. \revref
 }
@@
     \plr{should put somethign about this in the discussion?}
+    \jss{I think we should definitely address it in the discussion -- and in the main text of the ridge paper.
+    Do you think we can say something about the ridge for the minimal oscillator being constant across parameter space?} 
 }

### 2020-11-15 review-responses.tex
added a discussion paragraph on drift constraits due to canalization, inertia and gene flow.

@@
     Do you think we can say something about the ridge for the minimal oscillator being constant across parameter space?}
+    We have added a paragraph to the discussion to address these points. \revref
 }
@@
     \plr{... together with this?}
+    We have added a paragraph to the discussion to address these points. \revref
 }
@@
     \plr{... and also this?}
+    We have added a paragraph to the discussion to address this point. \revref
 }

### 2020-11-15 kalman_walk_paper.tex
fixed figures
fig6
added reference to Example 3 in main text
minor edits to responses
added a discussion paragraph on drift constraits due to canalization, inertia and gene flow.

@@
      \end{tikzpicture} &
-     \raisebox{0.25\height}{\includegraphics[scale=0.25]{examples/fig1b}}
+     \includegraphics{examples/fig1b_small}
 \end{tabular}
@@
     \end{align*}
-The resulting set of systems are depicted in Figure \ref{fig:all_osc}.
+  The resulting set of systems are depicted in Figure \ref{fig:all_osc}. \revpoint{1}{16}
 \end{example}
@@
 \end{scope}
-\end{tikzpicture} &	\raisebox{0.25\height}{\includegraphics[scale=0.25]{examples/fig2b}}
+\end{tikzpicture} &	\includegraphics{examples/fig2b_small}
     \end{tabular}
@@
 if the parents have different underlying systems.
+Hybrid phenotypic break down, as a function of genetic distance between
+phenotypically equivalent parental oscillators (described in Example \ref{ex:all_osc})
+is illustrated in Example \ref{ex:hybrid_osc}.
 How quickly do hybrid phenotypes break down as genetic distance between parents increases?
@@
  % \end{tabular}
-  \includegraphics[scale=0.5]{examples/f1f2_newplot}
+  %\includegraphics[scale=0.5]{examples/f1f2_newplot}
+  \includegraphics{examples/f1f2_small}
   \caption{
@@
     \centering
-    \includegraphics[width=0.5\textwidth]{examples/F2_vs_F1_divergence_tau0}
+    %\includegraphics[width=0.5\textwidth]{examples/F2_vs_F1_divergence_tau0}
+    \includegraphics{examples/fig6_small_maxt10}
     \caption{Mean hybrid phenotypic distance from optimum computed with equation \eqref{eqn:distance}, using $\rho(t) = \exp(-t/4\pi)$
@@
 
+Phenotypically equivalent system evolution may not solely be driven by drift.
+Network coefficient evolution may be biased due to epistasis and driven by second-order selective effects;
+such processes can lead an evolving population's exploration of network space to regions more densely populated by
+and more closely connected to high fitness system architectures \citep{neutral_network_physics}. Our results, however, do not rely on the ``thickness''
+of the phenotypically equivalent set, but on the set's \emph{openess} under sexual recombination.
+So long as the phenotypically equivalent set is not closed under sexual recombination,
+opportunities for incompatibility remain.
+It is plausible, however, that the speed at which system drift can generate incompatibilities might diminish
+in circumstances where selection for robustness is strong enough to constrain a population to a small section of system space.
+Likewise, as the speed of system drift relies on segregating genetic variation, any constraints on such variation,
+possibly due to epistasis, genetic architecture \citep{hermisson2003epistasis}, adaptive inertia \citep{alvarez2009effects} or weak gene flow could plausibly slow it down.
+\revpoint{1}{18} \revpoint{1}{19} \revpoint{1}{21}
+
 \paragraph{Nonlinearity and model assumptions}

### 2021-04-14 review-responses.tex
added cover letter

@@
 
-We are pleased to submit a revision of our manuscript,
+We are pleased to submit a revision of our manuscript, ``{System drift and speciation},'' for your review. 
+We apologize for the delayed resubmission but hope that you will find the
+revised manuscript significantly improved.
+
+We thank you and the two reviewers for providing constructive suggestions 
+and feedback that has helped to improve the manuscript. We have followed 
+the suggestions closely and hope to have fully addressed all the concerns and 
+suggestions as detailed in our pont-by-point response below.
+
+In our original submission, we studied the effect of gene regulatory network evolution on speciation,
+first by applying linear system theory to characterize the large space of phenotypically equivalent network organizations,
+and second by using quantitative genetics to show that neutral genetic drift on this network space can lead to speciation over plausible timescales.
+
+We agree with the reviewers that the combination of these two approaches, while interesting,
+was not seamless and that the quantitative genetics section required further development, 
+and thus will follow the recommendation of both the Editor and Associate Editor, to break the manuscript into
+two papers. One paper will focus on quantitative genetics, and the other -- 
+the focus of this present revision -- will focus on system theory. 
+A revised manuscript focusing on quantitative genetics will be submitted at a later date.
+
+We hope that by focusing this paper system theory, we have clarified our results, which include a mathematical description of phenotypically equivalent network space 
+and its implications for speciation under neutral genetic drift. These results suggest that the space of phenotypically equivalent network organizations is substantial,
+and often not reproductively compatible, leaving many oportunities for independelty evolving populations to become reproductively incompatible.
+The linear system framework applied in this manuscript explicitly describes possible moleuclar pathways and naturally predicts Haldane's rule.
+
+We would like to note that the previous version of this paper as a preprint has since recieved 10 citations as well as attracted
+substantial positive attention as a preprint (e.g., see tweets at
+\url{https://www.biorxiv.org/content/10.1101/231209v2}). Furthermore, this paper has been downloaded more than 2,229 times \url{https://rxivist.org/papers/24911},
+making it the 156$^{\text{th}}$ most downloaded evolutionary biology paper available on the bioRxiv, as of this writing.
 

### 2021-04-14 review-responses.tex
spelling

@@
 the suggestions closely and hope to have fully addressed all the concerns and 
-suggestions as detailed in our pont-by-point response below.
+suggestions as detailed in our point-by-point response below.
 
@@
 
-We hope that by focusing this paper system theory, we have clarified our results, which include a mathematical description of phenotypically equivalent network space 
+We hope that by focusing this paper on system theory, we have clarified our results, which include a mathematical description of phenotypically equivalent network space 
 and its implications for speciation under neutral genetic drift. These results suggest that the space of phenotypically equivalent network organizations is substantial,
-and often not reproductively compatible, leaving many oportunities for independelty evolving populations to become reproductively incompatible.
-The linear system framework applied in this manuscript explicitly describes possible moleuclar pathways and naturally predicts Haldane's rule.
+and often not reproductively compatible, leaving many opportunities for independently evolving populations to become reproductively incompatible.
+The linear system framework applied in this manuscript explicitly describes possible molecular pathways and naturally predicts Haldane's rule.
 
-We would like to note that the previous version of this paper as a preprint has since recieved 10 citations as well as attracted
+We would like to note that the previous version of this paper as a preprint has since received 10 citations as well as attracted
 substantial positive attention as a preprint (e.g., see tweets at
-\url{https://www.biorxiv.org/content/10.1101/231209v2}). Furthermore, this paper has been downloaded more than 2,229 times \url{https://rxivist.org/papers/24911},
-making it the 156$^{\text{th}}$ most downloaded evolutionary biology paper available on the bioRxiv, as of this writing.
+\url{https://www.biorxiv.org/content/10.1101/231209v2}). Furthermore, this paper has been downloaded more than 2,229 times (\url{https://rxivist.org/papers/24911}),
+making it the 156$^{\text{th}}$ most downloaded evolutionary biology paper available on the \emph{bioRxiv}, as of this writing.
 

### 2021-04-23 review-responses.tex
edits to cover letter

@@
 
-We are pleased to submit a revision of our manuscript, ``{System drift and speciation},'' for your review. 
-We apologize for the delayed resubmission but hope that you will find the
+We are writing to submit a revision of our manuscript, ``{System drift and speciation},'' for your review. 
+We sincerely apologize for the delayed resubmission but hope that you will find the
 revised manuscript significantly improved.
@@
 was not seamless and that the quantitative genetics section required further development, 
-and thus will follow the recommendation of both the Editor and Associate Editor, to break the manuscript into
-two papers. One paper will focus on quantitative genetics, and the other -- 
-the focus of this present revision -- will focus on system theory. 
-A revised manuscript focusing on quantitative genetics will be submitted at a later date.
+and thus have followed the recommendation of both the Editor and Associate Editor,
+by breaking the manuscript into two papers.
+One paper -- the present revision -- covers the application of system theory to regulatory networks,
+and the other (which we plan to resubmit soon) will cover the quantitative genetics.
+In addition to detailed responses to the reviews,
+we are submitting a color-coded diff to make it easy to see what has been removed.
+We are happy with the result -- as predicted by the reviewers, the resulting paper stands alone
+without the quantitative genetics results.
 

### 2021-04-27 review-responses.tex
edits through rev 1

@@
 better than a quick but superficial review).
+\end{quote}
 
+Thanks for the apology,
+but given the delay in our resubmission, we can hardly complain, at this point.
+
+\begin{quote}
 Your manuscript is quite long already, and the reviewers' suggestions of
@@
 Thanks for the careful attention to the paper.
-We have indeed decided to go with this suggestion,
+We have decided to go with this suggestion,
 and have split the paper in two;
@@
 \reply{
-  We think that adding a third type of arrowhead may only add to the confusion, so now we are using only the standard arrowhead \revref.
+  We think that adding a third type of arrowhead may only add to the confusion, so now we are using only the standard arrowhead.
 }
@@
 
-\begin{quote}
-General remark: line numbers are very helpful for reviewing.  
-\end{quote}
+This manuscript is mostly the ``systems biology'' part,
+but we think that the potential applications to ``popgen'' are still reasonably visible.
+(And, we plan to resubmit the ``popgen'' part separately.)
 
-\begin{quote}
-    \textbf{(I)} The first (systems biology) part makes use of the concept of
-``nonidentifiability'' of chemical reaction networks and develops these concepts
-in the context of a genotype to phenotype map. In particular, this map is
-analyzed for the case where it can be modeled as a linear system.
-\end{quote}
+% \begin{quote}
+%     \textbf{(I)} The first (systems biology) part makes use of the concept of
+% ``nonidentifiability'' of chemical reaction networks and develops these concepts
+% in the context of a genotype to phenotype map. In particular, this map is
+% analyzed for the case where it can be modeled as a linear system.
+% \end{quote}
 
@@
   Thanks for the additional suggestions!
-  We've reworked the introduction, adding these (and many other) additional references. \revref
+  We've reworked the introduction, adding these (and many other) additional references
+  into the (large and growing) literature. \revref
+  We hope it's more representative now.
 }
@@
     We've tried to explain things better in the revised Introduction. \revref
+    However, we find assigning specific contributions to be difficult
+    because similar ideas appear in many disparate parts of the literature.
 }
@@
 \reply{
-    Yes, that's right. \revref
+    Yes, that's right; we've clarified this. \revref
 }
@@
     but there's so many uncertainties we don't think bringing this up will clarify things.)
-    \plr{JOSH: how's this?}
-    \jss{Sounds great! We could also add:}
-    Over larger evolutionary distances there is some evidence that structurally distinct
-    isofunctional networks of different size do indeed exist. \emph{E.g.}, the networks underlying circadian
-    rhythm in cyanobacteria and mammals, as discussed by \citet{sancar2008intelligent}.
-    It seems possible, in principle, that, on a smaller scale, such differences could exist between more closely related populations.
+    % PLR: I don't think we need this, as the reviewers are aware of these examples.
+    % Over larger evolutionary distances there is some evidence that structurally distinct
+    % isofunctional networks of different size do indeed exist. \emph{E.g.}, the networks underlying circadian
+    % rhythm in cyanobacteria and mammals, as discussed by \citet{sancar2008intelligent}.
+    % It seems possible, in principle, that, on a smaller scale, such differences could exist between more closely related populations.
 
@@
     basis of the evolution of robustness/canalization. It is possible to account
-for this effect, see Hermisson et al 2003. Amer. Nat. 161, 708-734;
-Alvarez-Castro et al, TPB 75 (2009) 109-122, or also Rice, 1998, Evolution 52,
-647-656; van Nimwegen, etal 1999, PNAS 96:9716–9720. Note that epistasis is
-necessary for the neutral evolution of incompatibilities, which is what you are
-aiming for.
+    for this effect, see \citet{hermisson2003epistasis,alvarez2009effects}, 
+    or also \citet{rice1998evolution,neutral_network_physics}.
+    Note that epistasis is necessary for the neutral evolution of incompatibilities,
+    which is what you are aiming for.
 \end{point}
@@
 \reply{
-    \plr{should put somethign about this in the discussion?}
-    \jss{I think we should definitely address it in the discussion -- and in the main text of the ridge paper.
-    Do you think we can say something about the ridge for the minimal oscillator being constant across parameter space?}
-    We have added a paragraph to the discussion to address these points. \revref
+    Thanks for pointing out this important omission.
+    We have added mention of this (and these citations),
+    including a paragraph to the discussion. \revref
 }
@@
 maintain the phenotype. This leads to a phenomenon called ``adaptive inertia''
-(see Baatz and Wagner, TPB 51, 49-66 and Alvarez-Castro etal . TPB 75, 109-122),
+(see \citet{baatz1997adaptive} and \citet{alvarez2009effects}),
 which effectively slows down the movement along the ridge considerably. This
@@
 \reply{
-    \plr{... together with this?}
-    We have added a paragraph to the discussion to address these points. \revref
+    Again, this is an important point covered in additional discussion.
+    \revref
 }
@@
 \reply{
-    We aim to do that in the subsequent paper.
+    We agree, and hope to do this in our follow-up for this part of the paper.
 }
@@
 allopatry, but under (even weak) gene flow, some degree of positive selection
-is always needed (Bank et al, Genetics 191, 845–863 for a proof). This also
+is always needed (\citet{bank2012limits} for a proof). This also
 means that even weak gene flow will counteract the process described in the ms.
@@
 \reply{
-    \plr{... and also this?}
-    We have added a paragraph to the discussion to address this point. \revref
+    Also a good point - we've added mention of this as well. \revref
 }

### 2021-04-27 kalman_walk_paper.tex
edits through rev 1

--- 2020-07-03 kalman_walk_paper.tex
@@
 can provide guidance where intuition fails,
-and thus has the potential to discover general principles in the organization of biological systems 
+and thus have the potential to discover general principles in the organization of biological systems 
 as well as provide concrete numerical predictions \citep{servedio2014not}.
@@
 potential regulatory sequences \citep{hare2008sepsid}, 
-in the logic of transcriptional \citep{tsong2006evolution, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription} 
-and neural circuits \citep{trojanowski2014neural}, and in developmental systems \citep{true2001developmental}. 
+in the logic of transcriptional \citep{tsong2006evolution, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription,jimenez2017spectrum} 
+and neural circuits \citep{trojanowski2014neural}, and in developmental systems \citep{vondassow2000segment,true2001developmental}. 
 \revpoint{1}{1} \revpoint{1}{3}
@@
 For instance, in many contexts mathematical models
-are fundamentally \emph{nonidentifiable} and/or \emph{indistinguishable} -- meaning that 
-there can be uncertainty about an inferred model's parameters or even its claims about
+are fundamentally \emph{nonidentifiable} and/or \emph{indistinguishable} -- \revpoint{1}{2}
+meaning that there can be uncertainty about an inferred model's parameters or even its claims about
 causal structure, despite access to complete and perfect data \citep[e.g.,][]{bellman1970structural, grewal1976identifiability, walter1984structural}. 
@@
 despite the existence of a theoretical one-to-one mapping between a model and behavior,
-  tiny amounts of noise make inference problems nonidentifiable in practice \citep{petrov2005well}. \revpoint{1}{2}
+tiny amounts of noise make inference problems nonidentifiable in practice \citep{petrov2005well}.
 
@@
 is necessary for natural selection \citep{edelman2001degeneracy}, 
-as it may function as a mechanism for biological robustness and evolvability, or manifest as 
+as it may function as a mechanism for biological robustness and evolvability
+\citep[reviewed in][]{devisser2003perspective}, or manifest as 
 \emph{canalization} \citep{whitacre2010degeneracy}.
@@
 can this open up neutral evolutionary paths,
-and do these paths manifest as \emph{developmental system drift} \citep{true2001developmental}?
+and do species explore these paths
+through the process termed \emph{developmental system drift} \citep[by][]{true2001developmental}?
 Is this fast enough to contribute meaningfully to speciation?
@@
 
-Besides those already cited, many other authors have observed
 % It is not a new observation
+The field of population genetics has also explored the consequences of the fact
 that there is often more than one way to do the same thing, 
-and that speciation can be the result of (nearly) neutral processes.
+and observed that speciation might be the result of changes that are themselves neutral.
 Indeed, \citet{bateson1909heredity} first proposed that what today we call a
@@
 \emph{Note on implementation:}
-The \emph{reachable subspace},
-which we denote by $\reachable$,
+The \emph{reachable subspace}
+% which we denote by $\reachable$,
 is defined to be the closure of $\spn(B)$ under applying $A$
@@
 % Analogously to this, we define
-and the \emph{unobservable subspace}, 
+and the \emph{unobservable subspace}
 %$\mathcal{O}$, is the closure of $\spn(C^T)$ under applying $A$.
-denoted $\unobservable$, is the largest $A$-invariant subspace
+% denoted $\unobservable$,
+is the largest $A$-invariant subspace
 contained in the null space of $C$.
@@
 along which a large population will drift under environmental and selective stasis. 
+This drift need not be purely neutral --
+for instance, second-order selection on robustness
+will push the species towards ``flatter'' areas of genotype space \citep{rice1998evolution,hermisson2003epistasis}.
 Even if the phenotype is constrained and remains constant through evolutionary time, 
@@
 (i.e., entries of $A$, $B$ and $C$)
-of a haploid system inherited by an offspring from her diploid parent
+of a haploid system inherited by an offspring from a diploid parent
 are determined by the parent's two systems
@@
     $F_2$s compose all four mixed homozygotes if variation at both
-    traits has a simple, one-locus genetic basis in both popualtions; and
+    traits has a simple, one-locus genetic basis in both populations; and
     \textbf{(right)}
@@
 increases with $\epsilon^2$ and $\epsilon$, respectively.
+\end{example}
 
@@
  % \end{tabular}
-  \includegraphics[scale=0.5]{examples/f1f2_newplot}
+  %\includegraphics[scale=0.5]{examples/f1f2_newplot}
+  \includegraphics{examples/f1f2_small}
   \caption{
     \textbf{(left)} Phenotypes of $F_1$ hybrids between a homozygous $A(0)$ parent and, 
-    top-to-bottom, homozygous $A(\sfrac{1}{100})$, $A(\sfrac{1}{10})$, and $A(\sfrac{1}{2})$ parents;
+    top-to-bottom, homozygous $A(\sfrac{1}{100})$, $A(\sfrac{1}{10})$, and $A(\sfrac{1}{2})$ parents,
+    where $A(\epsilon)$ is defined in figure \ref{fig:all_osc};
     parental coefficients differ by around 0.5\%, 5\%, and 25\% respectively.
@@
     %%   it only matters if the offspring is homo/het/homozygous, so there are 3^4 = 81 possible expression patterns.
-    Different colored lines correspond to different $F_2$ hybrids;
-    note that some completely fail to oscillate.
+    Different colored lines correspond to different $F_2$ hybrids,
+    many of which show complete breakdown.
   } \label{fig:hybs}
@@
 \end{figure}
-\end{example}
 
@@
 between two allopatric populations to be substantially lower than the parentals?
+Selection -- on pleiotropic traits or on robustness --
+may actively push even a strongly constrained system along neutral directions,
+but even the calculations under purely neutral drift are informative.
 The population mean of an unconstrained quantitative trait
@@
 % since $F_1$ and $F_2$ phenotypes diverge linearly and at the square root of time relative to parentals, respectively.
-%\plr{refer to Turelli here}
 % In this model, this is due to recombination shuffling system architectures, leaving the $F_2$ system homozygous for one parent at one particular locus and homozygous for the other parent at another possibly incompatible locus.
@@
 
+Phenotypically equivalent system evolution is probably not only driven by neutral genetic drift.
+\revpoint{1}{18} \revpoint{1}{19} \revpoint{1}{21}
+For one thing, movement along the set of equivalent networks
+is not expected to be completely neutral,
+since second-order selection pushes populations towards
+``flatter'' regions of the fitness landscape
+in which a population centered on the optimal set has lower genetic load
+\citep[as described in different contexts by][]{rice1998evolution,neutral_network_physics}.
+If this bias towards more robust networks is strong enough,
+it may even prevent drift, but it is unclear how strong this effect would be in practice.
+Our results, on the other hand, do not rely on the flatness of the fitness surface
+around the phenotypically equivalent set, but rather on the curvature of the equivalent set itself.
+So long as the phenotypically equivalent set is not closed under sexual recombination,
+opportunities for incompatibility remain.
+However the speed at which system drift can generate incompatibilities might diminish
+if selection for robustness is strong enough to constrain a population
+to a small section of system space,
+although the strength of such effects in practice are not known.
+Likewise, as the speed of system drift relies on segregating genetic variation,
+any constraints on such variation,
+possibly due to epistasis, genetic architecture \citep{hermisson2003epistasis}, adaptive inertia \citep{baatz1997adaptive,alvarez2009effects} or weak gene flow could plausibly slow it down.
+More work on specific systems,
+likely coupled with simulations,
+will be necessary to identify the biologically relevant parameter regimes.
+
 \paragraph{Nonlinearity and model assumptions}
@@
 Furthermore, nonidentifiability (which implies the existence of neutral directions)
-is often found in practice in moderately complex models of biological systems \citep{gutenkunst2007universally, piazza2008diverse}.
+is often found in practice in moderately complex models of biological systems \citep[e.g.,][]{gutenkunst2007universally,piazza2008diverse,jimenez2017spectrum}.
 
@@
     We would like to thank Sergey Nuzhdin, Stevan Arnold, Michael Turelli, Patrick Phillips, Erik Lundgren and Hossein Asgharian for valuable discussion. 
-    We would also like to thank Nick Barton, Sarah Signor, Todd Parsons, Joachim Hermisson for very helpful comments on the manuscript.
+    We would also like to thank Nick Barton, Sarah Signor, Todd Parsons, and Joachim Hermisson for very helpful comments on the manuscript.
     Work on this project was supported by funds from
@@
 and 0 elsewhere,
-the coefficients of the quadratic form is
+the coefficients of the quadratic form are
 \begin{equation}
@@
       \int_0^t e^{s \lambda_i} e^{(t-s) \lambda_j} ds
-      &=
+      &= \begin{cases}
       \frac{ e^{t \lambda_i} - e^{t \lambda_j} }{ \lambda_i - \lambda_j } 
           \qquad & \text{if} \quad i \neq j \\
-      &=
           t e^{t \lambda_i} 
-          \qquad & \text{if} \quad i = j 
+          \qquad & \text{if} \quad i = j .
+      \end{cases}
   \end{aligned}
@@
     X_{ij}(t,Z) 
-       &= 
+        &= \begin{cases}
         \left( U^{-1} Z U \right)_{ij}
@@
           \qquad & \text{if} \quad i \neq j \\
-      &=
       \left( U^{-1} Z U \right)_{ii}
           t e^{t \lambda_i} 
-          \qquad & \text{if} \quad i = j 
+          \qquad & \text{if} \quad i = j ,
+        \end{cases}
     \end{aligned}
@@
     X_{ij}^{k\ell}(t) 
-      &= 
+    &= \begin{cases}
       (U^{-1})_{\cdot k} U_{\ell \cdot}
@@
           \qquad & \text{if} \quad i \neq j \\
-      &=
       (U^{-1})_{\cdot k} U_{\ell \cdot}
       t e^{t \lambda_i} 
-          \qquad & \text{if} \quad i = j 
+          \qquad & \text{if} \quad i = j ,
+    \end{cases}
   \end{aligned}

### 2021-04-27 review-responses.tex
rev 2 edits

@@
 in which distinct genetic architectures can produce the same phenotype. (In the fitness landscape
-terminology, this would correspond to the existence of an evolutionary ridge in a fitness landscape.
-)The starting point is a nice analogy with linear systems theory. The authors highlight the fact
+terminology, this would correspond to the existence of an evolutionary ridge in a fitness landscape.)
+The starting point is a nice analogy with linear systems theory. The authors highlight the fact
 that in general, two distinct linear (differential) systems can respond identically for any input; i.e.,
@@
 
-\plr{TODO: respond/edit}
+Thanks for the encouraging words.
+As discssed above, we've separated out the second (less convincing) part,
+and the manuscript now only covers the first part.
 
 \begin{point}{end of p6:} 
-    ``we assume ... $((A + A' )/2, (B + B' )/2, (C + C' )/2)$'' . It is claimed in p4 (right
-before the last paragraph) that two kryptotypes need not have the same dimension. In this
+    ``we assume ... $((A + A' )/2, (B + B' )/2, (C + C' )/2)$'' . It is claimed \revreffull{2}{1}
+    that two kryptotypes need not have the same dimension. In this
     case, the previous sum does not make sense, right?
@@
 a point sitting on the red dotted line, i.e., a Gaussian distribution with the mean at the
-average of the two parental populations. According to the authors, the distribution is bi-
-modal with a peak below the optimal set (as displayed on the figure) and another distinct
+average of the two parental populations. According to the authors, the distribution is
+bimodal with a peak below the optimal set (as displayed on the figure) and another distinct
 peak sitting on the other side of the optimal. In fact, if I understand the computations of
@@
 \reply{
-    We have improved Figure~\ref{fig:conceptual_fig} to hopefully better demonstrate the point.
+    We have improved Figure~\ref{fig:conceptual_fig} and made the accompanying discussion
+    more concrete to hopefully better demonstrate the point.
 }
@@
     Finally, it would be worth mentioning several old and recent works relating genetic drift
-to speciation: Yamagushi and Iwasa, several articles by Gavrilets et al. (I think several
-citations are missing here), and Mirò Pina and Schertzer.
+    to speciation: \citet{yamaguchi2013first}, several articles by Gavrilets et al. (I think several
+    citations are missing here), and \citet{pina2019does}.
 \end{point}

### 2021-04-27 kalman_walk_paper.tex
rev 2 edits

@@
 a population diffuses along the neutral ridges of a high-dimensional space of possible system parameters,
-in a similar vein as \emph{holey landscape} models \citep{gavrilets1997evolution, yamaguchi2013first, pina2019does}\revpoint{2}{12}.
+in a similar vein as \emph{holey landscape} models \citep{gavrilets1997evolution, yamaguchi2013first, pina2019does}.  \revpoint{2}{12}
 % Our model is developed in more detail in a companion paper \citep{schiffman2020genetic}.

### 2021-04-29 kalman_walk_paper.tex
added current affiliations

@@
 \usepackage{multicol}
-\usepackage[bottom]{footmisc}
+\usepackage[bottom, symbol]{footmisc}
 
@@
 {\Huge \bf System drift and speciation} \\ \vspace{0.75cm}
-Joshua S. Schiffman$^{\dagger}$ \qquad Peter L. Ralph$^{\dagger \ddagger}$ \\ \vspace{0.5cm}
+Joshua S.
+Schiffman$^{\dagger}$ \footnote{Current affiliations: New York Genome Center, New York, New York 10013, U.S.A.
+\& Weill Cornell Medicine, New York, New York 10065, U.S.A.} \qquad Peter L.
+Ralph$^{\dagger \ddagger}$ \\ \vspace{0.5cm}
 $^{\dagger}${\footnotesize {Molecular and Computational Biology, University of Southern California, Los Angeles, California 90089, U.S.A. \\
 $^{\ddagger}$Departments of Mathematics and Biology \& The Institute for Ecology and Evolution, University of Oregon, Eugene, Oregon 97403, U.S.A.}} \\ \vspace{0.5cm}
-{\small \texttt{jsschiff@usc.edu} \qquad 
+{\small \texttt{jschiffman@nygenome.org} \qquad 
 \texttt{plr@uoregon.edu}} \\ \vspace{0.5cm}

### 2021-05-03 title_page_revision.tex
made title page

--- 2021-04-27 kalman_walk_paper.tex
@@
 \usepackage{fullpage}
-\usepackage[backref=page, hidelinks, colorlinks=true, allcolors=blue!60!black!100]{hyperref}
+\usepackage[backref=page, hidelinks, colorlinks=true, citecolor=blue!60!black!100]{hyperref}
 \usepackage{tikz}
@@
 \usepackage{authblk}
+\usepackage{lineno}
 \usepackage{multicol}
@@
     \newcommand{\plri}[1]{\plr{#1}}
-\fi
-
-\newif\ifsubmission
-\submissiontrue
-% \submissionfalse
-\ifsubmission
-    % % stuff for submission
-    \usepackage{lineno}
-    % allow line numbers around math environments
-    % from https://tex.stackexchange.com/questions/43648/why-doesnt-lineno-number-a-paragraph-when-it-is-followed-by-an-align-equation
-        \newcommand*\patchAmsMathEnvironmentForLineno[1]{%
-          \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
-          \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
-          \renewenvironment{#1}%
-             {\linenomath\csname old#1\endcsname}%
-             {\csname oldend#1\endcsname\endlinenomath}}% 
-        \newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
-          \patchAmsMathEnvironmentForLineno{#1}%
-          \patchAmsMathEnvironmentForLineno{#1*}}%
-        \AtBeginDocument{%
-        \patchBothAmsMathEnvironmentsForLineno{equation}%
-        \patchBothAmsMathEnvironmentsForLineno{align}%
-        \patchBothAmsMathEnvironmentsForLineno{flalign}%
-        \patchBothAmsMathEnvironmentsForLineno{alignat}%
-        \patchBothAmsMathEnvironmentsForLineno{gather}%
-        \patchBothAmsMathEnvironmentsForLineno{multline}%
-        }
-\else
-    % for the nice version
 \fi
@@
 
+\newtheorem{theorem}{Theorem}
+\newtheorem{lemma}{Lemma}
+\newtheorem{definition}{Definition}
 \newtheorem{example}{Example}
 
-% Responses to reviews
-\input{review-response-commands}
-% set this to show line numbers and include responses to reviews or not
-\newif\ifreviewresponses
-\reviewresponsestrue  % include them
-% \reviewresponsesfalse  % don't include them
-\newcommand{\responsefile}{review-responses}  % name of the review responses file
-
 \begin{document}
-\linenumbers
+%\linenumbers
 
@@
 {\Huge \bf System drift and speciation} \\ \vspace{0.75cm}
-Joshua S. Schiffman$^{\dagger}$ \qquad Peter L. Ralph$^{\dagger \ddagger}$ \\ \vspace{0.5cm}
-$^{\dagger}${\footnotesize {Molecular and Computational Biology, University of Southern California, Los Angeles, California 90089, U.S.A. \\
-$^{\ddagger}$Departments of Mathematics and Biology \& The Institute for Ecology and Evolution, University of Oregon, Eugene, Oregon 97403, U.S.A.}} \\ \vspace{0.5cm}
-{\small \texttt{jsschiff@usc.edu} \qquad 
+Joshua S. Schiffman$^{\dagger}$ \qquad Peter L. Ralph$^{\ddagger}$ \\ \vspace{0.5cm}
+$^{\dagger}${\footnotesize {New York Genome Center, New York, New York 10013, U.S.A \& Weill Cornell Medicine, New York, New York 10065, USA  \\
+$^{\ddagger}$Departments of Mathematics and Biology \& The Institute for Ecology and Evolution, University of Oregon, Eugene, Oregon 97403, USA}} \\ \vspace{0.5cm}
+{\small \texttt{jschiffman@nygenome.org} \qquad 
 \texttt{plr@uoregon.edu}} \\ \vspace{0.5cm}
-\small \today \\
+%\small \today \\
 \vspace{0.25cm}
@@
 
-% Other title ideas:
-%
-% Rapid speciation despite conservation of phenotype
-%
-% How fast does network drift create incompatibilities?
-%
-% More than one way to grow a cat: speciation despite conservation of phenotype
-%
-% More than one way to grow a cat: regulatory network drift and speciation
-%
-% More than one way to grow a cat: regulatory network redundancy and speciation
-%
-% Neutral network drift leads to incompatibilities on the same time scale as genetic drift
-%
-% More than one way to grow a cat: can network drift lead to speciation?
-%
-% Beyond the snowball: a quantitative model of incompatibility accumulation
-%
-% System drift and speciation: more than one way to grow a cat
-%
-% An explicit model of neutral regulatory network evolution, with applications
-% to the rate of accumulation of hybrid incompatibility
-%
-% Evolutionary conservation of phenotype does not entail conservation of the underlying
-% molecular mechanism leading to rapid speciation
-%
-% The evolution of phenotype-invariant gene networks rapidly leads to hybrid incompatibiliy
-%
-% Evolutionary network rewiring can rapidly lead to hybrid incompatibility despite
-% phenotypic conservation
-%
-% Hybrid incompatibilities can evolve rapidly due to developmental system drift
-%
-% Evolutionary systems theory and speciation
-
 \begin{abstract}
-%The evolutionary conservation of phenotype under selective and environmental stasis does not necessitate conservation of the underlying mechanism, as distinct molecular pathways can realize identical phenotypes.
 Even if a species' phenotype does not change over evolutionary time, 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%
-%\begin{multicols}{2}
-\section*{Introduction}
-
-It is an overarching goal of many biological subdisciplines 
-to attain a general understanding of the function and evolution of the 
-complex molecular machinery that translates an organism's genome 
-into the characteristics on which natural selection acts,
-the phenotype.
-For example, 
-there is a growing body of data on the evolutionary histories and molecular characterizations of particular gene regulatory networks
-\citep{jaeger2011gap, davidson2006gene, israel2016comparative}, 
-as well as thoughtful verbal and conceptual models \citep{true2001developmental, weiss2000phenogenetic, edelman2001degeneracy, pavlicev2012model}. 
-Mathematical models of both particular regulatory networks
-and the evolution of such systems in general
-can provide guidance where intuition fails,
-and thus have the potential to discover general principles in the organization of biological systems 
-as well as provide concrete numerical predictions \citep{servedio2014not}.
-There is a substantial amount of work studying the evolution of gene regulatory networks, in frameworks 
-both abstract \citep{wagner1994evolution, wagner1996does, siegal2002waddington, bergman2003evolutionary, draghi2015robustness}
-and empirically inspired
-\citep{mjolsness1991connectionist, jaeger2004dynamic, vitaly1, crombach2016gap, wotton2015quantitative, chertkova2017insilico}.
-
-At all levels of biological organization,
-the problems that biological systems have evolved to solve
-often do not have single solutions --
-systems can be structurally different yet remain functionally equivalent \citep{edelman2001degeneracy}. 
-Examples can be found across nearly all levels of biological organization from the level of the genetic code itself % \citep{crick1963recent}
-all the way up to the convergent evolution of adaptive traits.
-In many cases, these functionally equivalent structures
-can be explored through small, local changes to the structure that leave the function unchanged.
-For instance, there are ``neutral networks''
-of nucleic acid sequences that produce the same RNA secondary structure \citep{gruner1996analysis}\revpoint{1}{5},
-amino acid sequences that fold similarly \citep{babajide1997neutral},
-or proteins with equivalent thermodynamic stability \citep{hart2014thermodynamic}.
-Further examples are found in the vast space of functionally equivalent
-potential regulatory sequences \citep{hare2008sepsid}, 
-in the logic of transcriptional \citep{tsong2006evolution, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription,jimenez2017spectrum} 
-and neural circuits \citep{trojanowski2014neural}, and in developmental systems \citep{vondassow2000segment,true2001developmental}. 
-\revpoint{1}{1} \revpoint{1}{3}
-% \jss{Other ideas instead of ``degeneracy'' or ``isofunctional structural plasticity'': non-invertibility, isofunctional plasticity, structural plasticity, phenotypic equivalency, phenotypic invariance, functional invariance, isofunctionality, functional equivalency, genotypic plasticity, conjugacy, functional symmetry, isologous, isomorphic.}
-% \plr{'structural plasticity' means something specific in neuroscience}
-
-This capacity for isofunctional yet distinct mechanisms, sometimes called \emph{degeneracy},
-is a consequence of a many-to-one mapping between a system's structure and function,
-a concept that has been explored in many fields beyond biology.
-For instance, in many contexts mathematical models
-are fundamentally \emph{nonidentifiable} and/or \emph{indistinguishable} -- \revpoint{1}{2}
-meaning that there can be uncertainty about an inferred model's parameters or even its claims about
-causal structure, despite access to complete and perfect data \citep[e.g.,][]{bellman1970structural, grewal1976identifiability, walter1984structural}. 
-Models with different parameter schemes, or even different mechanics 
-can make equally accurate predictions,
-but still not actually reflect the internal dynamics of the system being modeled.
-In control theory, where electrical circuits and mechanical systems are often the focus, 
-it is understood that there can be an infinite number of ``realizations'', 
-or ways to reverse engineer the dynamics of a ``black box'',
-even if all possible input and output experiments are performed 
-\citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. 
-The inherent nonidentifiability of chemical reaction networks is sometimes
-referred to as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}.
-In computer science, this has been framed as the relationship among processes that \emph{simulate} one another \citep{van2004equivalence}.
-Finally,
-the field of \emph{inverse problems} studies those cases in which,
-despite the existence of a theoretical one-to-one mapping between a model and behavior,
-tiny amounts of noise make inference problems nonidentifiable in practice \citep{petrov2005well}.
-
-It has been argued that the ability to modify structure without affecting function
-is necessary for natural selection \citep{edelman2001degeneracy}, 
-as it may function as a mechanism for biological robustness and evolvability
-\citep[reviewed in][]{devisser2003perspective}, or manifest as 
-\emph{canalization} \citep{whitacre2010degeneracy}.
-It may even contribute to the formation of new species \citep{gavrilets2014models}. 
-Redundancy of the genetic code, for instance,
-can make sequences more fault-tolerant to mutations \citep{sonneborn1965degeneracy},
-and robustness to modification of genetic networks can allow adaptation 
-without passing through a fitness valley \citep{wagner2008robustness}.
-
-In this paper we use results on mathematical nonidentifiability from
-linear systems theory to study how gene regulatory networks can be modified
-while retaining the same function, and the possible implications for speciation.
-% Specifically, we would like to understand the degree of which gene regulatory network architectures are functionally equivalent and how this relates to network size and efficiency.
-If system architectures are not functionally unique,
-can this open up neutral evolutionary paths,
-and do species explore these paths
-through the process termed \emph{developmental system drift} \citep[by][]{true2001developmental}?
-Is this fast enough to contribute meaningfully to speciation?
-To do this, we describe results on linear dynamical systems which give
-an analytical description of the set of 
-all linear gene network architectures that yield identical phenotypes,
-and use quantitative genetics theory to estimate the speed at which system drift can lead to reproductive incompatibility and hence speciation.
-In this model,
-a population diffuses along the neutral ridges of a high-dimensional space of possible system parameters,
-in a similar vein as \emph{holey landscape} models \citep{gavrilets1997evolution, yamaguchi2013first, pina2019does}.  \revpoint{2}{12}
-% Our model is developed in more detail in a companion paper \citep{schiffman2020genetic}.
-
-% It is not a new observation
-The field of population genetics has also explored the consequences of the fact
-that there is often more than one way to do the same thing, 
-and observed that speciation might be the result of changes that are themselves neutral.
-Indeed, \citet{bateson1909heredity} first proposed that what today we call a
-Bateson-Dobzhansky-Muller incompatibility would likely arise through neutral changes. \revpoint{1}{10}
-The potential for speciation has been analyzed in models of 
-traits under stabilizing selection determined additively by alleles at many loci 
-\citep{wright1935evolution,barton1986maintenance,barton1989divergence,barton2001role},
-in related fitness landscape models \citep{fraisse2016genetics},
-and for pairs of traits that must match but whose value is unconstrained \citep{sved1981twosex}.
-% Recent work has shown that Fisher's geometric model of stabilizing selection
-% predicts many general patterns from the speciation literature \citep{fraisse2016genetics}.
-It has also been shown that population structure
-can allow long-term stable coexistence of incompatible genotypes encoding identical phenotypes \citep{phillips1996maintenance}. \revpoint{1}{9}
-However, previous simulations of system drift in regulatory sequences \citep{tulchinsky2014hybrid}
-and a regulatory cascade \citep{porter2002speciation}
-found rapid speciation under directional selection
-but only equivocal support for speciation under models of purely neutral drift.
-%Additionally, \citet{fierst} observed speciation due to an increase of segregation variance under balancing selection, in models of weak to moderate, but not strong epistasis.
-The rate at which hybrid incompatibility accumulates due to genetic drift creating
-segregation variance between isolated populations is fairly well understood
-\citep{slatkin1994segregation,rosas2010cryptic,chevin2014niche},
-but model assumptions can strongly affect predictions,
-including whether variation is due to rare or common alleles \citep{slatkin1994segregation},
-and the shape of the fitness landscape \citep{fraisse2016genetics}.
-Our main aim is to provide a concrete framework 
-that can provide natural predictions of these model parameters
-across a general class of models.
-Furthermore, tools from system theory allow analytical predictions to be made
-for large populations with complex phenotypes
-that would be inaccessible to population simulations.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Results}
-
-We use a model of gene regulatory networks
-that describes the temporal dynamics of
-a collection of $n$ coregulating molecules
--- such as transcription factors -- as well as external or environmental inputs.
-% We model the temporal dynamics of the concentrations of a collection of $n$ coregulating molecules
-% within an organism, that may also be affected by temporally varying signals from the environment.
-%% Organisms' phenotypes are constructed by gene by gene and gene by environment interactions. 
-%%Here we simply define the \emph{phenotype} to be those aspects of the dynamics directly under natural selection
-%%-- the \emph{what}, \emph{when}, and \emph{how much}, of an organism's molecules that are physiologically or otherwise relevant to survival.
-We write $\kappa(t)$ for the vector of $n$ molecular concentrations at time $t$.
-The vector of $m$ ``inputs'' determined exogenously to the system is denoted $u(t)$,
-and the vector of $\ell$ ``outputs'' is denoted $\phi(t)$.
-The output is merely a linear function of the internal state:
-$\phi_i(t) = \sum_j C_{ij} \kappa_j(t)$
-for some matrix $C$.
-Since $\phi$ is what natural selection acts on, we refer to it as the \emph{phenotype}
-(meaning the ``visible'' aspects of the organism),
-and in contrast refer to $\kappa$ as the \emph{kryptotype},
-as it is ``hidden'' from direct selection.
-Although $\phi$ may depend on all entries of $\kappa$,
-it is usually of lower dimension than $\kappa$,
-and we tend to think of it as the subset of molecules relevant for survival.
-The dynamics are determined by
-the matrix of regulatory coefficients, $A$,
-a time-varying vector of inputs $u(t)$,
-and a matrix $B$ that encodes the effect of each entry of $u$ on the elements of the kryptotype.
-%Thus an organism's phenotype $\phi(t)$ -- a vector of molecular concentrations at time $t$ -- is determined both by the structure and organization of a biological system (\emph{e.g.} a gene regulatory network), given by the triple $(A,B,C)$,
-%and by its environment $u(t)$.
-The rate at which the $i^\text{th}$ concentration changes
-is a weighted sum of the concentrations
-as well as the input:
-\begin{equation}\label{eqn:system}
-   \begin{aligned}
-    \dot{\kappa}(t) &= A \kappa(t) + B u(t) \\
-    \phi(t) &= C \kappa(t) .
-  \end{aligned} 
-\end{equation}
-Furthermore, we always assume that $\kappa(0) = 0$,
-so that the kryptotype measures deviations from initial concentrations. 
-Here $A$ can be any $n \times n$ matrix, $B$ any $n \times m$, and $C$ any $\ell \times n$ dimensional matrix,
-with usually $\ell$ and $m$ less than $n$.
-We think of the system as the triple $(A,B,C)$,
-which translates (time-varying) $m$-dimensional input $u(t)$
-into the $\ell$-dimensional output $\phi(t)$.
-Under quite general assumptions on the input
-(e.g., $|u(t)|$ is integrable) \revpoint{AE}{3}
-we can write the phenotype as \revpoint{AE}{4}
-  \begin{align}
-    \phi(t) = \int_{0}^{t} C e^{(t-s) A} B u(s) ds ,
-  \end{align}
-which is a convolution of the input $u(t)$ with the system's \emph{impulse response},
-which we denote as $h(t) := Ce^{A t}B$.
-
-In terms of gene regulatory networks, $A_{ij}$ determines how the $j^\text{th}$ transcription factor regulates the $i^\text{th}$ transcription factor.
-If $A_{ij} > 0$, then $\kappa_j$ upregulates $\kappa_i$, while if $A_{ij} < 0$, then $\kappa_j$ downregulates $\kappa_i$.
-The $i^\text{th}$ row of $A$ is therefore determined by genetic features such as
-the strength of $j$-binding sites in the promoter of gene $i$,
-factors affecting chromatin accessibility near gene $i$,
-or basal transcription machinery activity.
-The form of $B$ determines how the environment influences transcription factor expression levels,
-and $C$ might determine the rate of production of downstream enzymes.
-%% {Linearity} also ``zero-state equivalence''
-% Here we have assumed that the system is linear,
-% and begins from the ``zero'' state ($\kappa(0)=0$).
-% Of course, neither of these are necessarily true for real systems,
-% but the dynamics of most nonlinear systems can be approximated locally by a linear systems near most points. 
-% Furthermore, the ease of analyzing linear systems makes this an attractive place to start.
-
-\citet{wagner1994evolution} and others have used a similar discrete-time model \revpoint{1}{6}
-(that might be written $\phi_{t+1} = f(A \phi_t)$, where $f$ is a sigmoid).
-Our choice of continuous time does not affect the points we make here,
-but our restriction to \emph{linear} systems is a stronger assumption
-(see the Discussion).
-
-To demonstrate the model,
-we construct a simple gene network in Example \ref{ex:oscillator} below.
-
-\begin{example}[An oscillator]\label{ex:oscillator}
-For illustration, we consider an extremely simplified model of oscillating gene transcription,
-as for instance is found in cell cycle control or the circadian rhythm.
-% Cellular division is governed by many different processes, however it is thought that its rhythm is partially controlled by oscillating gene transcription \citep{orlando2008global}.
-There are two genes, 
-whose transcript concentrations are given by $\kappa_1(t)$ and $\kappa_2(t)$, 
-and gene-2 upregulates gene-1, while gene-1 downregulates gene-2 with equal strength.
-Only the dynamics of gene-1 are consequential to the oscillator 
-(perhaps the amount of gene-1 activates another downstream gene network). 
-Lastly, both genes are equally upregulated by an exogenous signal.
-The dynamics of the system are described by
-    \begin{align*}
-      \dot{\kappa_{1}}(t) &= \kappa_{2}(t) + u(t) \\
-        \dot{\kappa_{2}}(t) &= - \kappa_{1}(t) + u(t) \\
-        \phi(t) &= \kappa_{1}(t)   .
-    \end{align*}
-In matrix form the system regulatory coefficients are given as,
-$A \!=\! \left[\begin{smallmatrix} 0 & 1 \\ -1 & 0 \end{smallmatrix}\right]$, 
-$B \! =\! \left[\begin{smallmatrix} 1 \\ 1 \end{smallmatrix}\right]$,
-and $C \!=\! \left[\begin{smallmatrix} 1 & 0 \end{smallmatrix}\right]$.
-  %  \begin{align*}
-  %    A = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} , \qquad B = \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix}
-  %  \end{align*}
- %       The oscillatory system $\Sigma$ is thus given as
- %   \begin{align*}
- %     \Sigma = \left \{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 
- %       0 & 1 \\ 
- %      -1 & 0 
- %       \end{bmatrix} \kappa(t) + \begin{bmatrix} 1 \\ 1 \end{bmatrix} u(t) \\ 
- %         \phi(t) &= \begin{bmatrix} 1 & 0 \end{bmatrix} \kappa(t) \end{array} \right.
- %    \end{align*}
-If the input is an impulse at time zero (a delta function), 
-then the phenotype is equal to the impulse response: \revpoint{AE}{5}
-\begin{align*}
-\phi(t) = h(t) = \sin t + \cos t  .
-\end{align*}
-The system and its dynamics are referred to in Figure \ref{fig:oscillator}. We return to the evolution of such a system below.   
-\end{example}
-
-\begin{figure}[H]
-   % \begin{center} 
-  \centering
-     \begin{tabular}{cc}
-        \begin{tikzpicture}
-        \begin{scope}[every node/.style={circle,thick,draw}]
-            \node (A) at (0,0) {$\kappa_{1}$};
-            \node (B) at (4,0) {$\kappa_{2}$};
-            \node[shape=rectangle] (U) at (2,2) {input ($u$)};
-            \node[shape=rectangle] (y) at (2,-2) {output ($\phi$)};
-            %\\  \includegraphics[width=0.15\textwidth, height=0.04\paperheight]{osc_impulse} \end{tabular}};
-        \end{scope}
-
-        \begin{scope}[>={Stealth[black]},
-                      every node/.style={fill=white,circle},
-                      every edge/.style={draw=black, thick}]
-            \path [->] (A) edge[bend left] node {\tiny $-1$} (B);
-            \path [->] (B) edge[bend left] node {\tiny $1$} (A); 
-            \path[->] (U) edge node {\tiny $1$} (A);
-            \path[->] (U) edge node {\tiny $1$} (B);
-            \path[->] (A) edge[bend right] node {\tiny $1$} (y);
-        \end{scope}
-        \begin{scope}[>={Stealth[black]},
-                      every edge/.style={draw=black, thick}]
-            %\path [->] (A) edge[loop left] node {\tiny $\lambda_{1}$} (A);
-            %\path [->] (B) edge[loop left] node {\tiny $\lambda_{2}$} (B);
-        \end{scope}
-
-     \end{tikzpicture} &
-     \includegraphics{examples/fig1b_small}
-\end{tabular}
-%  \end{center}
-  \caption{
-    (Left) 
-    Diagram of the gene network in Example \ref{ex:oscillator}
-  , and (right) plot of the phenotype $\phi(t)$ against time $t$
-  .} \label{fig:oscillator}
-\end{figure}
-
-\subsection*{Equivalent gene networks}
-
-As reviewed above,
-some systems with identical phenotypes are known to differ, sometimes substantially, at the molecular level; 
-systems with identical phenotypes do not necessarily have identical kryptotypes.
-How many different mechanisms perform the same function?
-
-Two systems are equivalent if they produce the same phenotype given the same input,
-i.e., have the same input--output relationship.
-We say that
-the systems defined by $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ are
-\textbf{phenotypically equivalent} 
-if their impulse response functions are the same:
-$h(t) = \bar h(t)$ for all $t \ge 0$.
-This implies that for any acceptable input $u(t)$,
-if $(\kappa_u(t),\phi_u(t))$ and $(\bar \kappa_u(t),\bar \phi_u(t))$ are the solutions to equation \eqref{eqn:system}
-of these two systems, respectively, then
-\begin{align*}
-      \phi_u(t) = \bar \phi_u(t) \qquad \text{for all} \; t \ge 0.
-\end{align*}
-% (recall that $\kappa(0) = \bar \kappa(0) = 0$). 
-In other words, phenotypically equivalent systems respond identically for \emph{any} input.
-
-One way to find other systems phenotypically equivalent to a given one
-is by change of coordinates:
-if $V$ is an invertible matrix, then the systems $(A,B,C)$ and $(VAV^{-1},VB,CV^{-1})$
-are phenotypically equivalent because their impulse response functions are equal:
-  \begin{equation}
-    \begin{aligned}
-      h(t) &= C e^{A t} B 
-      = C V^{-1} V e^{A t} V^{-1} V B \\
-      &= C V^{-1} e^{V A V^{-1} t} V B 
-      = \bar{C} e^{\bar{A} t} \bar{B} = \bar h(t).
-    \end{aligned}
-  \end{equation}
-These ``changes of coordinates'' are not simply different ways of looking at the same system --
-if each dimension of the kryptotype 
-corresponds to the concentration of a particular transcription factor,
-changing $A$ corresponds to changing the strengths of regulatory interactions.
-We will even see below that interactions may change sign. \revpoint{1}{13}
-However, not all phenotypically equivalent systems are of this form:
-systems can have identical impulse responses without being coordinate changes of each other.
-In fact, systems with identical impulse responses can involve interactions between different
-numbers of molecules, and thus 
-have kryptotypes in
-different dimensions altogether.
-
-This implies that most systems have at least $n^2$ degrees of freedom,
-where recall $n$ is the number of components of the kryptotype vector.
-This is because for an arbitrary $n \times n$ matrix $Z$,
-taking $V$ to be the identity matrix plus a small perturbation in the direction of $Z$
-above implies that
-moving $A$ in the direction of $ZA - AZ$
-% (e.g., $\bar A = A + \epsilon(ZA-AZ)$)
-while also moving $B$ in the direction of $ZB$ 
-and $C$ in the direction of $-CZ$
-will leave the phenotype unchanged to second order in the size of the perturbation.
-%% I believe the following is necessary and sufficient:
-If the columns of $B$ and the rows of $C$ are not all eigenvectors of $A$,
-then any such $Z$ will result in a different system.
-
-It turns out that in general, there are more degrees of freedom,
-except if the system is \emph{minimal} -- meaning, informally, that it uses the smallest possible number of components
-to achieve the desired dynamics.
-Results in system theory show that any system can be realized in a particular minimal dimension
-(the dimension of the kryptotype, $n_\text{min}$),
-and that any two phenotypically equivalent systems of dimension $n_\text{min}$ are related by a change of coordinates.
-Since gene networks can grow or shrink following gene duplications and deletions, 
-these additional degrees of freedom can apply, in principle, to any system.
-
-Even if the system is not minimal, results from systems theory % -- the Kalman decomposition --
-explicitly describe the set of all phenotypically equivalent systems.
-% Could define script S here for 'system' if we want: $\Sys$
-We refer to $\allS(A_0,B_0,C_0)$ as the set of all systems phenotypically equivalent
-to the system defined by $(A_0, B_0, C_0)$:
-\begin{equation} \label{eqn:equivalence}
-  \begin{aligned}
-    \allS(A_0, B_0, C_0) 
-      &= \left\{
-        (A,B,C) : C e^{At} B = C_0 e^{A_0 t} B_0 \; \text{for}\; t \ge 0 
-      \right\}  .
-  \end{aligned}
-\end{equation}
-These systems need not have the same kryptotypic dimension $n$,
-but must have the same input and output dimensions ($\ell$ and $m$, respectively).
-
-The Kalman decomposition, which we now describe informally, elegantly characterizes this set
-\citep{kalman1963mathematical,kalman1969topics,anderson1966equivalence}.
-To motivate this, first note that the input $u(t)$ only directly pushes the system
-in certain directions (those lying in the span of the columns of $B$).
-As a result, different combinations of input can 
-move the system in any direction that lies in what is known as the \emph{reachable subspace}.
-Analogously, we can only observe motion of the system in certain directions
-(those lying in the span of the rows of $C$),
-and so can only infer motion in what is known as the \emph{observable subspace}.
-The Kalman decomposition then classifies each direction in kryptotype space
-as either reachable or unreachable, and as either observable or unobservable.
-Only the components that are both reachable and observable determine the system's phenotype --
-that is, components that both respond to an input and produce an observable output. 
-% it is possible to add components that respond to an input, 
-% but do not influence the output, or components that, in principle, influence the output, but do not respond to any inputs.
-
-Concretely, the \textbf{Kalman decomposition} of a system $(A,B,C)$  
-gives a change of basis $P$ such that
-the transformed system $(PAP^{-1},PB,CP^{-1})$
-can be written in block matrix form:
-\begin{align*}
-       PAP^{-1}
-       &=
-       \left[ \begin{array}{cccc}
-           A_{\rno} & A_{\rno,\ro} & A_{\rno,\nrno} & A_{\rno,\nro} \\
-           0 & A_{\ro} & 0 & A_{\ro,\nro} \\
-           0 & 0 & A_{\nrno} & A_{\nrno,\nro} \\
-          0 & 0 & 0 & A_{\nro}
-       \end{array} \right] ,
-\end{align*}
-and
-\begin{align*}
-     PB
-     &=
-    \left[ \begin{array}{cccc}
-         B_{\rno} \\
-         B_{\ro} \\
-         0 \\
-         0 
-   \end{array} \right] 
-   &
-   (CP^{-1})^T
-   &=
-   \left[ \begin{array}{cccc}
-       0 \\
-       C_{\ro}^T \\
-       0 \\
-       C_{\nro}^T
-   \end{array} \right] .
-\end{align*}
-The $n$-dimensional system has been divided into subspaces
-of dimensions $n_\rno + n_\ro + n_\nrno + n_\nro = n$,
-and so, for instance, $A_{\rno}$ is the $n_\rno \times n_\rno$ square matrix 
-in the top left corner of $P A P^{-1}$. \revpoint{1}{15}
-The impulse response of the system is given by
-\begin{align*}
-      h(t) = C_{\ro} e^{A_{\ro} t} B_{\ro},
-\end{align*}
-and therefore, the system is phenotypically equivalent to the \emph{minimal} system $(A_{\ro}, B_{\ro}, C_{\ro})$.
-
-This decomposition is unique up to a change of basis that preserves the block structure.
-In particular, 
-the minimal subsystem obtained by the Kalman decomposition
-is unique up to a change of coordinates.
-This implies that there is no equivalent system with a smaller number of kryptotypic dimensions
-than the dimension of the minimal system.
-It is remarkable that the gene regulatory network architecture to achieve a given input--output map is never unique --
-both the change of basis used to obtain the decomposition
-and, once in this form, all submatrices other than $A_{\ro}$, $B_{\ro}$, and $C_{\ro}$ can be changed without affecting the phenotype,
-and so represent degrees of freedom.
-(However, some of these subspaces may affect how the system deals with noise.)
-
-\emph{Note on implementation:}
-The \emph{reachable subspace}
-% which we denote by $\reachable$,
-is defined to be the closure of $\spn(B)$ under applying $A$
-(or equivalently, the span of $B, AB, A^2B, \ldots A^{n-1}B$), \revpoint{AE}{6}
-% Analogously to this, we define
-and the \emph{unobservable subspace}
-%$\mathcal{O}$, is the closure of $\spn(C^T)$ under applying $A$.
-% denoted $\unobservable$,
-is the largest $A$-invariant subspace
-contained in the null space of $C$.
-% and $\reachable$ is the largest $A$-invariant subspace contained in the image of $B$.)
-The four subspaces, $\rno$, $\ro$, $\nrno$, and $\nro$
-are defined from these by intersections and orthogonal complements --
-$\ro$ refers to the both \emph{reachable and observable} subspace,
-while $\nrno$ refers to the \emph{unreachable and unobservable} subspace,
-and similarly for $\nro$ and $\rno$.
-
-%Usually, the dimension $n$ and the reference system $\Sys_0$ is implicit and we write only $\calA$.
-%Further, we refer to the set of all linear systems with impulse response $h(t)$, regardless of dimension, to be $\allS$,
-%  \begin{align}
-%    \allS := \bigcup_{n=\min}^{\infty} \calA_n(\Sys_0)
-%  \end{align}
-
-For the remainder of the paper, we interpret $\allS$ as the neutral set in the fitness landscape, 
-along which a large population will drift under environmental and selective stasis. 
-This drift need not be purely neutral --
-for instance, second-order selection on robustness
-will push the species towards ``flatter'' areas of genotype space \citep{rice1998evolution,hermisson2003epistasis}.
-Even if the phenotype is constrained and remains constant through evolutionary time, 
-the molecular mechanism underpinning it is not constrained and likely will not be conserved.
-
-Finally, note that if $B$ and $C$ are held constant --
-i.e., if the relationships between environment, kryptotype, and phenotype do not change --
-there are \emph{still} usually degrees of freedom. 
-The following example \ref{ex:all_osc} gives the set of minimal systems equivalent to the oscillator of Example \ref{ex:oscillator},
-that all share common $B$ and $C$ matrices.
-The oscillator can also be equivalently realized by a three-gene (or larger) network, and will have even more evolutionary degrees of freedom available, 
-as in Figure \ref{fig:3D_osc}.
-
-\begin{example}[All equivalent rewirings of the oscillator] \label{ex:all_osc}
-The oscillator of example \ref{ex:oscillator} is minimal, and so any equivalent system is a change of coordinates
-by an invertible matrix $V$.
-If we further require $B$ and $C$ to be invariant then we need $VB=B$ and $CV=C$.
-Therefore
-the following one-parameter family $(A(\tau), B, C)$ describes the set of all two-gene systems
-phenotypically equivalent to the oscillator:
-    \begin{align*}
-      A(\tau) &= \frac{-1}{\tau+1} \begin{bmatrix} -\tau & -1 \\ 2 \tau(\tau + 1) + 1 &  \tau \end{bmatrix} \ \text{for} \ \tau \neq -1 .
-    \end{align*}
-  The resulting set of systems are depicted in Figure \ref{fig:all_osc}. \revpoint{1}{16}
-\end{example}
-
-\begin{figure}[H]
-    \centering
-\begin{tabular}{cc}
-\begin{tikzpicture}
-\begin{scope}[every node/.style={circle,thick,draw}]
-  \node (A) at (0,0) {$\kappa_{1}$};
-    \node (B) at (4,0) {$\kappa_{2}$};
-    \node[shape=rectangle] (U) at (2,2) {input ($u$)};
-    \node[shape=rectangle] (y) at (2,-2) {output ($\phi$)};
-\end{scope}
-
-\begin{scope}[>={Stealth[black]},
-              every node/.style={fill=white,circle},
-              every edge/.style={draw=black, thick}]
-    \path [->, sloped] (A) edge[bend left] node {\tiny $-2 \tau - \frac{1}{\tau+1}$} (B);
-    \path [->, sloped] (B) edge[bend left] node {\tiny $\frac{1}{\tau+1}$} (A); 
-    \path[->] (U) edge node {\tiny $1$} (A);
-    \path[->] (U) edge node {\tiny $1$} (B);
-    \path[->] (A) edge[bend right] node {\tiny $1$} (y);
-\end{scope}
-\begin{scope}[>={Stealth[black]},
-              every edge/.style={draw=black, thick}]
-    \path [->] (A) edge[loop left] node[left] {\tiny $\frac{\tau}{\tau+1}$} (A);
-    \path [->] (B) edge[loop right] node[right] {\tiny $\frac{-\tau}{\tau+1}$} (B);
-\end{scope}
-\end{tikzpicture} &	\includegraphics{examples/fig2b_small}
-    \end{tabular}
-% \caption{}
-      \caption{
-      (Left) $A(\tau)$, the set of all phenotype-equivalent cell cycle control networks.
-      (Right) Gene-1 dynamics (blue) for both systems $A(0)$ and $A(-2)$ are identical, however, $A(0)$ gene-2 dynamics (red) differ from $A(-2)$ (green).
-      % Both gene-1 dynamics are given by $\kappa_{1}(t) = \sin t + \cos t$, and gene-2 by $\kappa_{2}(t) = \cos t - \sin t$ ($A(0)$) and $\kappa'_{2}(t) = \cos t + 3 \sin t$ ($A(2)$).
-      } 
-    \label{fig:all_osc}
-\end{figure}
-
-\revpoint{AE}{7}
-
-\begin{figure}[H]
-    \begin{center}
-\begin{tikzpicture}
-\begin{scope}[every node/.style={circle,thick,draw}]
-  \node (A) at (0,0) {$\kappa_{1}$};
-    \node (B) at (0,4) {$\kappa_{2}$};
-    \node (C) at (3.464,2) {$\kappa_{3}$};
-    \node[shape=rectangle, rotate=0] (U) at (-2.5,2) {input ($u$)};
-    \node[shape=rectangle] (y) at (3.464,0) {output ($\phi$)};
-\end{scope}
-
-\begin{scope}[>={Stealth[black]},
-              every node/.style={fill=white,circle},
-              every edge/.style={draw=black, thick}]
-    \path [->, sloped] (A) edge[bend left] node {\tiny $2.2$} (B);
-   % \path [->, sloped] (B) edge[bend left] node {\tiny ?} (A); 
-    \path [->, sloped] (A) edge[bend right] node {\tiny $2$} (C); 
-    %\path [->, sloped] (B) edge[bend right] node {\tiny ?} (C); 
-    \path [->, sloped] (C) edge[bend right] node {\tiny $-1$} (A); 
-    \path [->, sloped] (C) edge[bend right] node {\tiny $-2.2$} (B); 
-    \path[->] (U) edge node {\tiny $1$} (A);
-    \path[->] (U) edge node {\tiny $1$} (B);
-    \path[->] (A) edge node {\tiny $1$} (y);
-\end{scope}
-\begin{scope}[>={Stealth[black]},
-              every edge/.style={draw=black, thick}]
-    \path [->] (A) edge[loop below] node[sloped, below] {\tiny $1$} (A);
-    \path [->] (B) edge[loop above] node[sloped, above] {\tiny $1$} (B);
-    \path [->] (C) edge[loop right] node[sloped, above] {\tiny $-1$} (C);
-\end{scope}
-\end{tikzpicture}
-  \caption{A possible non-minimal three-gene oscillator, phenotypically equivalent to $A(\tau)$, the systems in Examples \ref{ex:oscillator} and \ref{ex:all_osc}.}
-  \label{fig:3D_osc}
-\end{center}
-\end{figure}
-
-\paragraph{Sexual reproduction and recombination}
-Parents with phenotypically equivalent yet differently wired gene networks may produce offspring with dramatically different phenotypes. 
-If the phenotypes are significantly divergent then the offspring may be inviable or otherwise dysfunctional, 
-despite both parents being well adapted. 
-If this is consistent for the entire population, we would consider them to be separate species, in accord with the biological species concept \citep{mayr2000biological}.
-
-First, we must specify how sexual reproduction acts on these systems.
-Suppose that each of a diploid organisms' two genomes encodes a set of system coefficients
-with the same kryptotype dimension. \revpoint{2}{1}
-We assume that a diploid which has inherited systems $(A', B', C')$ and $(A'', B'', C'')$ from its two parents
-has phenotype determined by the system that averages these two,
-$((A'+A'')/2, (B'+B'')/2, (C'+C'')/2)$.
-
-Each genome an organism inherits is generated by meiosis,
-in which both of its diploid parents recombine their two genomes,
-and so an $F_1$ offspring carries one system copy from each parent,
-and an $F_2$ is an offspring of two independently formed $F_1$s.
-If the parents are from distinct populations,
-these are simply first-- and second--generation hybrids, respectively.
-
-Exactly how the coefficients 
-(i.e., entries of $A$, $B$ and $C$)
-of a haploid system inherited by an offspring from a diploid parent
-are determined by the parent's two systems
-depends on the genetic basis of any variation in the coefficients.
-Thanks to the randomness of meiotic segregation,
-the result is random to the extent that each parent is heterozygous
-for alleles that affect the coefficients.
-Since the $i^\text{th}$ row of $A$ summarizes how each gene regulates gene $i$,
-and hence is determined by the promoter region of gene $i$,
-the elements of a row of $A$ tend to be inherited together,
-which will create covariance between entries of the same row.
-It is, however, a quite general observation that the variation seen among recombinant systems
-is proportional to the difference between the two parental systems.
-% This is certainly true if each coefficient is determined by a single nonrecombining locus,
-% so that each coefficient in the system produced by meiosis is an independent random choice
-% between the two parental coefficients.
-% On the other hand, if a coefficient is determined by additive contributions of independently segregating loci,
-% then the law of total variance applied after conditioning on the parental origin of each allele
-% implies that the variance is equal to one-quarter of the squared difference between parental coefficients.
-
-Offspring formed from two phenotypically identical systems do not necessarily exhibit the same phenotype as both of its parents 
--- in other words $\allS$, the set of all systems phenotypically equivalent to a given one, is not, in general, closed under averaging or recombination.
-If sexual recombination among systems drawn from $\allS$ yields systems with divergent phenotypes, 
-populations containing significant diversity in $\allS$ can carry genetic load, and isolated populations may fail to produce hybrids with viable phenotypes.
-
-\begin{figure}[H]
-\centering
-\includegraphics{figures/conceptual_fig}
-\caption{
-    \label{fig:conceptual_fig}
-    A conceptual figure of the fitness consequences of hybridization:
-    axes represent system coefficients (i.e., entries of $A$);
-    the line of optimal system coefficients is down in black;
-    solid lines give phenotypic distances to the optimum.
-    A pair of parental populations are shown in black, along the optimum;
-    a hypothetical population of $F_1$s are shown in red,
-    and the distribution of $F_2$s is shown in purple.
-    The two figures differ in the genetic basis, and hence,
-    the distribution of $F_2$ phenotypes:
-    \textbf{(left)}
-    $F_2$s compose all four mixed homozygotes if variation at both
-    traits has a simple, one-locus genetic basis in both populations; and
-    \textbf{(right)}
-    $F_2$ show a much wider distribution of phenotypes
-    if the genetic basis of variation in each population is polygenic.
-}
-\end{figure}
-
-\subsection*{Hybrid incompatibility}
-Two parents with the optimal phenotype can produce offspring whose phenotype is suboptimal
-if the parents have different underlying systems.
-Hybrid phenotypic break down, as a function of genetic distance between
-phenotypically equivalent parental oscillators (described in Example \ref{ex:all_osc})
-is illustrated in Example \ref{ex:hybrid_osc}.
-How quickly do hybrid phenotypes break down as genetic distance between parents increases?
-We will quantify how far a system's phenotype is from optimal
-using a weighted difference between impulse response functions.
-Suppose that $\rho(t)$ is a nonnegative weighting function, \revpoint{2}{3}
-$h_0(t)$ is the \emph{optimal} impulse response function
-and define the ``distance to optimum'' of another impulse response function
-to be
-\begin{align}
-\label{eqn:distance}
-	D(h) = \left( \int_0^\infty \rho(t) \|h(t) - h_0(t)\|^2 dt \right)^{1/2} .
-\end{align}
-In practice, we take $\rho(t) = \exp(-t/4\pi)$,
-so that fitness is determined by the dynamics of the system over a few multiples of $2 \pi$,
-but not longer. \revpoint{AE}{8}
-Consider reproduction between a parent with system $(A, B, C)$ 
-and another displaced by distance $\epsilon$ in the direction $(X,Y,Z)$,
-i.e., having  system $(A + \epsilon X, B + \epsilon Y, C + \epsilon Z)$.
-We assume both are ``perfectly adapted'' systems, 
-i.e., having impulse response function $h_0(t)$,
-and their offspring has impulse response function $h_\epsilon(t)$.
-A Taylor expansion of $D(h_\epsilon)$ in $\epsilon$ 
-is explicitly worked out in Appendix \ref{apx:H_calc}, and \revpoint{2}{4}
-shows that the phenotype of an $F_1$ hybrid between these two is at distance proportional to $\epsilon^2$ from optimal,
-while $F_2$ hybrids are at distance proportional to $\epsilon$.
-This is because an $F_1$ hybrid has one copy of each parental system,
-and therefore lies directly between the parental systems (see Figure \ref{fig:conceptual_fig}) --
-the parents both lie in $\allS$, which is the valley defined by $D$,
-and so their midpoint only differs from optimal due to curvature of $\allS$.
-In contrast, an $F_2$ hybrid may be homozygous for one parental type in some coefficients
-and homozygous for the other parental type in others;
-this means that each coefficient of an $F_2$ may be equal to either one of the parents,
-or intermediate between the two;
-this means that possible $F_2$ systems may be as far from the optimal set, $\allS$,
-as the distance between the parents.
-The precise rate at which the phenotype of a hybrid diverges depends on the geometry
-of the optimal set $\allS$ relative to segregating genetic variation.
-% in Figure \ref{fig:conceptual_fig}, this is depicted as the angle of the black line (the optimal set) with respect to the coordinates.
-
-\begin{example}[Hybrid incompatibility: misregulation due to system drift] \label{ex:hybrid_osc}
-Offspring of two equivalent systems from Example \ref{ex:all_osc}
-can easily fail to oscillate.
-For instance, the $F_1$ offspring between homozygous parents at $\tau=0$ and $\tau=-2$
-has phenotype $\phi_{F_1}(t) = e^t$, rather than $\phi(t) = \sin t + \cos t$.
-However, the coefficients of these two parental systems differ substantially,
-probably more than would be observed between diverging populations.
-In figure \ref{fig:hybs} we compare the phenotypes for $F_1$ and $F_2$ hybrids between more similar parents,
-and see increasingly divergent phenotypes as the difference between the parental systems increases.
-(In this example, the coefficients of $A(\epsilon)$ differ from those of $A(0)$ by an average factor of $1+\epsilon/2$;
-such small differences could plausibly be caused by changes to promoter sequences.)
-This divergence is quantified in Figure \ref{fig:osc_incompat},
-which shows that mean distance to optimum phenotype of the $F_1$ and $F_2$ hybrid offspring between $A(0)$ and $A(\epsilon)$
-increases with $\epsilon^2$ and $\epsilon$, respectively.
-\end{example}
-
-\begin{figure}[H]
-  \centering
- % \begin{tabular}{ccc}
- %   \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F1_hundreth_tau0} &
- %     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F2_hundrethtau0} \\
- %     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F1_tenth_tau0.pdf} &
- %     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F2_tenthtau0.pdf} \\
- %     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F1_half_tau0.pdf} &
- %     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F2_halftau0.pdf}
- % \end{tabular}
-  %\includegraphics[scale=0.5]{examples/f1f2_newplot}
-  \includegraphics{examples/f1f2_small}
-  \caption{
-    \textbf{(left)} Phenotypes of $F_1$ hybrids between a homozygous $A(0)$ parent and, 
-    top-to-bottom, homozygous $A(\sfrac{1}{100})$, $A(\sfrac{1}{10})$, and $A(\sfrac{1}{2})$ parents,
-    where $A(\epsilon)$ is defined in figure \ref{fig:all_osc};
-    parental coefficients differ by around 0.5\%, 5\%, and 25\% respectively.
-    Parental phenotypes ($\sin t + \cos t$) are shown in solid black, and hybrid phenotypes in dashed blue.
-    \textbf{(right)} Phenotypes of all $3^4 = 81$ possible $F_2$ hybrids between the same set of parents,
-    with parental phenotype again in black.
-    %% PLR: There are at most 128 possible F2s, not 256; but many are equivalent; 
-    %%   it only matters if the offspring is homo/het/homozygous, so there are 3^4 = 81 possible expression patterns.
-    Different colored lines correspond to different $F_2$ hybrids,
-    many of which show complete breakdown.
-  } \label{fig:hybs}
-\end{figure}
-
-\begin{figure}[H]
-    \centering
-    %\includegraphics[width=0.5\textwidth]{examples/F2_vs_F1_divergence_tau0}
-    \includegraphics{examples/fig6_small_maxt10}
-    \caption{Mean hybrid phenotypic distance from optimum computed with equation \eqref{eqn:distance}, using $\rho(t) = \exp(-t/4\pi)$
-    for $F_1$ (black) and $F_2$ (blue) hybrids between $A(0)$ and $A(\epsilon)$ parent oscillators. Genetic distance is computed as 
-    $\left(\sum_{ij} (A_{ij}(0) - A_{ij}(\epsilon))^2\right)^{1/2}$.
-    }
-    \label{fig:osc_incompat}
-\end{figure}
-
-\paragraph{Haldane's rule}
-This model naturally predicts
-Haldane's rule, the observation that if only one hybrid sex is sterile or inviable 
-it is likely the heterogametic sex (e.g., the male in XY sex determination systems) \citep{haldane1922sex, orr1997haldane}.
-For example, consider an XY species with a two-gene network where the first gene resides on an autosome and the second gene on the X chromosome.
-A male whose pair of haplotypes is 
-$\left( \left[ \begin{smallmatrix} A_1 & A_2 \\ \cdot & \cdot \end{smallmatrix} \right], \left[ \begin{smallmatrix} A_1 & A_2 \\ X_1 & X_2 \end{smallmatrix} \right] \right)$
-has phenotype determined by $A = \left[ \begin{smallmatrix} A_1 & A_2 \\ X_1 & X_2 \end{smallmatrix} \right]$, 
-  if dosage compensation upregulates heterogametes by a factor of two relative to homogametes (as with \emph{Drosophila}),
-while a female homozygous for the haplotype
-$\left[\begin{smallmatrix} \bar A_1 & \bar A_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$, has phenotype determined by
-$A = \left[\begin{smallmatrix} \bar A_1 & \bar A_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$.
-An $F_1$ male offspring of these two will have its phenotype determined by
-$\left[\begin{smallmatrix} (A_1 + \bar A_1)/2 & (A_2 + \bar A_2)/2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$.
-If both genes resided on the autosomes, this system would only be possible in an $F_2$ cross.
-More generally, 
-if the regulatory coefficients for a system are shared between the sex and one or more autosomal chromosomes, 
-$F_1$ males are effectively equivalent to purely autosomal-system $F_2$ hybrids, and recall that $F_2$s are significantly less fit on average than $F_1$s (see Figure \ref{fig:osc_incompat}).
-Although many alleles will be dominant if the phenotype--fitness relationship is convex,
-the underlying mechanism does not depend on the \emph{dominance theory} \citep{turelli1995dominance} to explain Haldane's rule:
-instead, it derives from the nature of segregation variance.
-% In this setting, Haldane's rule is mechanistically similar to \emph{hybrid breakdown}, 
-% the observation that $F_2$ hybrids will often be less fit than $F_1$ hybrid crosses, in the early stages of speciation.
-% This mechanism relies on the genetic variation of a system being shared across the sex chromosomes and the autosomes,
-% and appears to be consistent with the ``dominance'' theory \citep{orr1997haldane, coyne1998evolutionary, turelli1995dominance} which is often invoked to explain Haldane's rule.
-% \plr{Explain that the effects on fitness are dominant.}
-%and makes several different predictions than the dominance theory as formulated by \citet{turelli1995dominance}.
-%\plr{what are they?}
-%\citet{turelli1995dominance} employ a population genetic model which strictly requires alleles causing hybrid incompatibility on the X chromosome not to exhibit additive effects, that is the \emph{dominance coefficient} (typically $h$ in classic population genetics) must not be equal to $\frac{1}{2}$.
-In their model, if alleles are strictly additive, that is neither dominant nor recessive, then male and female $F_1$ hybrids will drop in fitness at the exact same rates, and therefore defy Haldane's rule.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{The speed of speciation}
-
-We have shown that system drift can lead to speciation in principle,
-but is it rapid enough to be an important factor in practice?
-In other words, after what period of time would we expect the fitness of hybrids
-between two allopatric populations to be substantially lower than the parentals?
-Selection -- on pleiotropic traits or on robustness --
-may actively push even a strongly constrained system along neutral directions,
-but even the calculations under purely neutral drift are informative.
-The population mean of an unconstrained quantitative trait
-with additive genetic variance $V_G$ in a population with effective size $N_e$
-will move in $t$ generations a random amount whose variance is $t V_G / N_e$ \citep{lande1976natural}.
-The mean difference between two such populations has twice the variance.
-Although this mean difference is along neutral directions,
-we would in many cases expect the range of variation among $F_2$s
-in \emph{all} directions to be of the same order
-as the differences between the populations, 
-as depicted in Figure \ref{fig:conceptual_fig}.
-This suggests that, naively, two such populations that have been separated for $t$ generations
-will produce $F_2$ offspring that differ from optimal by an amount proportional to $\sqrt{t V_G / N_e}$.
-Since we assume they are at a local fitness optimum, without much loss of generality we can assume
-that fitness is locally quadratic, and so $F_2$ fitness decays linearly in time:
-proportionally to $t V_G / N_e$ -- fastest in small, diverse populations,
-This predicts that we need only wait some multiple of $N_e$ generations
-until substantial incompatibility has been accumulated.
-
-It is useful to think in more detail about the assumptions in the rough argument above.
-The key aspect is how population differences in neutral directions
-(along the fitness ridge)
-translate to segregation variance in $F_2$s in selectively constrained directions.
-To move the system (the $A$ matrix) a given distance
-generally involves moving many individual interaction coefficients (the entries $A_{ij}$).
-The movements must be coordinated, for the population to stay near the fitness ridge.
-However, mixing elements between systems that have made independent sets of coordinated changes
-to remain on the fitness ridge
-is unlikely to produce a set of coordinated changes;
-and the resulting system could move away from the ridge in almost any direction.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Genetic variation in empirical regulatory systems}
-
-What is known about the key quantity above, the amount of heritable variation in real regulatory networks?
-%These are known at best only roughly \citep{felsenstein1988phylogenies},
-%so we aim for order-of-magnitude estimates.
-The coefficient $A_{ij}$ from the system \eqref{eqn:system} measures how much the rate of net production of $i$ changes
-per change in concentration of $j$.
-It is generally thought that regulatory sequence change contributes much more to inter- and intraspecific variation
-than does coding sequence change affecting molecular structure \citep{schmidt2010fivevertebrate}.
-In the context of transcription factor networks this may be affected 
-not only by the binding strength of molecule $j$ to the promoter region of gene $i$
-but also the effects of other transcription factors (e.g., cooperativity)
-and local chromatin accessibility \citep{stefflova2013cooperativity}.
-For this reason, 
-the mutational target size for variation in $A_{ij}$ may be much larger than the dozens of base pairs
-typically implicated in the handful of binding sites for transcription factor $j$ of a typical promoter region,
-and single variants may affect many entries of $\allS$ simultaneously.
-% (Recall that although these are best modeled through nonlinear terms,
-% by linearizing we essentially consider first-order effects.)
-% On the other hand, a diverse set of buffering mechanisms are thought to contribute to phenotypic stability
-% in the presence of substantial molecular noise \citep{canalization,buffering},
-% suggesting that substantial variation in the micro-scale dynamics we consider here
-% may be necessary to produce relevant phenotypic effects downstream.
-
-Variation in binding site occupancy may overestimate variation in $A$,
-since it does not capture buffering effects (if for instance only one site of many needs to be occupied for transcription to begin),
-and variation in expression level measures changes in steady-state concentration (our $\kappa_i$) rather than the \emph{rate} of change.
-Nonetheless, these measures likely give us an idea of the scale of variability.
-%\citet{kasowski2010variation} found 
-It has been shown that between human individuals, there is differential occupancy in 7.5\%
-of binding sites of a transcription factor (p65) \citep{kasowski2010variation}.
-%\citet{verlaan2009targeted} showed that 
-It has also been inferred that cis-regulatory variation
-accounts for around 2--6\% of expression variation in human blood-derived primary cells \citep{verlaan2009targeted}, and
-%while \citet{lappalainen2013transcriptome} found 
-that human population variation 
-explained about 3\% of expression variation \citep{lappalainen2013transcriptome}.
-Allele-specific expression is indicative of standing genetic \emph{cis}-regulatory variation;
-%\citet{wang2017bayesian} found 
-allele-specific expression in 7.2--8.5\% of transcripts of a flycatcher species has been observed \citep{wang2017bayesian}, as well as
-%while \citet{tung2015genetic} observed 
-allele-specific expression in 23.4\% of genes studied in a baboon species \citep{tung2015genetic}. 
-Taken together, this suggests that variation in the entries of $A$
-may be on the order of at least a few percent between individuals of a population --
-doubtless varying substantially between species and between genes.
-
-%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%
-\section*{Discussion}
-
-%%% OUTLINE
-% - Summarize methods and main result on rate of speciation.
-% - Consistent with F1/F2 divergence and Haldane's rule
-% - What empirical evidence is there that this is actually happening? 
-%   * incompats due to misregulation (Noor, Witkopp...)
-%   * numerous species that are visually indistinguishable (skinks, ferns)
-%   * maybe also ref phylogenetic lit
-% - What other evidence could be collected?  Note could be just phenotype not fitness. (cite Matute)
-% - Is linearity an important assumption?
-% - We know that divergent selection and genetic conflict can easily lead to speciation; we've shown neutral drift can maybe surprisingly easily do so also; but it remains unclear the relative contributions of these.
-Note that although drift may be weaker in any one network there are lots of networks.
-Could motivate this with Orr's statement.
-Perhaps tie in spandrels.
-
-In this paper, we use tools from linear system theory and quantitative genetics
-to study the evolution of a mechanistic model of the genotype-phenotype map, 
-in which the phenotype is subject to stabilizing selection.
-In so doing, we provide an explicit model of
-phenogenetic drift \citep{weiss2000phenogenetic} and developmental system drift \citep{true2001developmental}.
-In this context, the Kalman decomposition \citep{kalman1963mathematical}
-gives an analytical description of all phenotypically equivalent gene networks.
-%It also implies that nearly all systems are nonidentifiable,
-%and that, in general, there exist axes of genetic variation unconstrained by natural selection.
-This description shows that the space of functionally equivalent network architectures increases with the square of a network's size,
-and that this space increases further if networks grow larger than absolutely necessary -- that is use more interacting components than the most efficient potential architectures.
-In this framework, even minimal gene network architectures -- efficient architectures that contain only the requisite number of interacting parts,
-are not structurally unique with respect to function.
-Functionally equivalent architectures are often related by continuous parameter changes, suggesting that equivalent networks might be mutationally connected, 
-and that there exist axes of genetic variation unconstrained by natural selection.
-The independent movement of separated populations along these axes by genetic drift
-can lead to a significant reduction in hybrid viability, and thus precipitate speciation,
-at a speed dependent on the effective population size and the amount of genetic variation. 
-In this model, at biologically reasonable parameter values,
-system drift is a significant -- and possibly rapid -- driver of speciation.
-This may be surprising because
-hybrid inviability appears as a consequence of recombining different, yet functionally equivalent, mechanisms, and since species are often defined by their unique adaptations or morphologies.
-
-Consistent with empirical observation of hybrid breakdown (e.g., \citet{plotner2017chlorosis}), 
-we see that the fitnesses of $F_2$ hybrids drop at a much faster rate than those of $F_1$s.
-% since $F_1$ and $F_2$ phenotypes diverge linearly and at the square root of time relative to parentals, respectively.
-% In this model, this is due to recombination shuffling system architectures, leaving the $F_2$ system homozygous for one parent at one particular locus and homozygous for the other parent at another possibly incompatible locus.
-Another natural consequence of the model is Haldane's rule,
-that if only one $F_1$ hybrid sex is inviable or sterile it is likely to be the heterogametic sex.
-This occurs because if the genes underlying a regulatory network are distributed among both autosomes and the sex chromosome, 
-then heterogametic $F_1$s show variation (and fitnesses) similar to that seen in $F_2$ hybrids.
-This observation appears to be similar to the extreme hybrid phenotypes produced by transgressive segregation \citep{rieseberg1999transgressive}, 
-which can manifest in $F_1$s when only one (dominant) parental allele is expressed at heterozygous loci; this was observed in hybrid gene expression patterns, and increased as a function of parental genetic distance \citep{stelkens2009genetic}.
-
-Is there evidence that this is actually occurring?
-System drift and network rewiring has been inferred across the tree of life \citep{wotton2015quantitative, crombach2016gap, dalal2017transcription, johnson2017rewiring, ali2017quantitative},
-and there is often significant regulatory variation segregating within populations.
-Transcription in hybrids between closely related species with conserved transcriptional patterns can also be divergent 
-\citep{haerty2006gene, maheshwari2012cis, coolon2014tempo, michalak2004association, mack2016gene}, and hybrid incompatibilities have been attributed to cryptic molecular divergence underlying conserved body plans \citep{gavin2013embryonic}.
-Furthermore, in cryptic species complexes (e.g., sun skinks \citep{barley2013challenge}),
-%strongly reproductively incompatible species
-genetically distinct species
-may be nearly morphologically indistinguishable.
-
-\paragraph{The origin of species not by means of natural selection?}
-% this is neutral, in contrast to other models. although some folks advocate nonadaptive speciation.
-% other mechanisms also probably important
-% but this one may be strong since happens in all networks at once
-As classically formulated, 
-the Dobzhansky-Muller model of hybrid incompatibility is agnostic to the relative importance of neutral versus selective genetic substitutions \citep{coyne1998evolutionary},
-and plausible mechanisms have been proposed
-whereby Dobzhansky--Muller incompatibilities could originate
-under neutral genetic drift \citep{lynch2000origin}
-or stabilizing selection \citep{fierst}.
-The same holds for the ``pathway model'' \citep{lindtke2015genetic},
-which is closer to the situation here.
-However, previous authors have argued that neutral processes are likely too slow to be a significant driver of speciation \citep{nei1983models,seehausen2014genomics}.
-% Using simulations, \citet{porter2002speciation} demonstrated the accumulation of hybrid incompatibilities under directional, but not stabilizing selection, and \citet{palmer2009dynamics} observed the appearance of incompatibilities in populations evolving in a constant environment, only if the populations were suboptimally adapted.
-% In light of the few known incompatibilities, 
-This has led some to conclude that hybrid incompatibility is typically a byproduct of positive selection \citep{orr2004speciation, schluter2009evidence} 
-or a consequence of genetic conflict \citep{presgraves2010molecular, crespi2013conflictual},
-two processes that typically act much more rapidly than genetic drift.
-% In contrast, the model we develop here suggests that even under strictly neutral conditions, 
-% system drift can lead to speciation, at a rate fast enough 
-% to play a substantial role in species formation across the tree of life. 
-However, our calculations suggest that even under strictly neutral processes,
-hybrid fitness breaks down as a function of genetic distance
-rapidly enough to play a substantial role in species formation across the tree of life. 
-This is consistent with broad patterns
-such as the relationship between molecular divergence and genetic isolation 
-% across diverse taxa and ecologies 
-seen by \citet{roux2016shedding}, 
-and the clocklike speciation rates observed by \citet{hedges2015tree}.
-% (as diversification is not observed to be accelerated following mass extinctions).
-
-These explanations are not mutually exclusive.
-All of these forces -- adaptive shifts, conflict, and network drift --
-are plausible drivers of speciation, and may even interact.
-Many of our observations carry over to models of directional selection --
-for instance, rapid drift along the set of equivalent systems
-could be driven by adaptation in a different, pleiotropically coupled system.
-Or, reinforcement due to local adaptation might provide a selective pressure
-that speeds up system drift.
-Furthermore,
-while the fitness consequences of incompatibility in any one given network may be small, 
-the cumulative impact of system drift across the many different networks an organism relies on may be substantial. 
-It remains to be seen how the relative strengths of these forces compare.
-
-% Of course the basic assumptions of this model will be violated in practice (constant selective pressures, etc.), however, this model can function as a ``neutral null'' description of gene network evolution (a strategy advocated for by \citet{lynch2007frailty, fay2008evaluating, koonin2016splendor}).
-If this model succeeds in describing actual system evolution, it can be inferred that the mechanism underlying species formation does not require the inclusion of adaptation or changes in selection to explain the emergence of hybrid incompatibility.
-
-% \paragraph{Population isolates and genetic load}
-% Above we have mostly discussed the case of two separated populations of equal size.
-% In contrast, consider a small subset of a large population that is suddenly isolated.
-% It will have a large amount of genetic variance, thanks to its history,
-% but a small population size -- and so can drift quite quickly until it exhausts its genetic variation.
-% By this point, it could have accumulated substantial incompatibility with the parental population.
-% Similarly, in a species with substantial isolation by distance, geographically local systems drift
-% might lead to genetic load due only to recombination, much as in \citep{phillips1996maintenance}.
-
-\paragraph{The dimensionality of trait space} \revpoint{1}{8}
-We have focused on examples of single traits (where the phenotype is one-dimensional),
-but phenotypes under selection are often high-dimensional,
-and variation in different traits often share a genetic basis.
-However, we still expect many degrees of freedom
-as long as there are components of the system not directly and individually constrained by selection
-(i.e., a kryptotype). % since if C = I then there's no nontrivial transformations that preserve C
-Even in networks where the phenotype and kryptotype are of the same dimension,
-system theory shows us that there will always be available degrees of freedom
-as specific system realizations are only unique up to a change of coordinates.
-Some phenotypes, however, require kryptotypic dimensions to be larger than that of the phenotype.
-For instance, many systems have minimal realizations (\emph{e.g.}, the oscillator in Example \ref{ex:all_osc}) where the dimension of the kryptotype is larger
-than that of the phenotype, implying that for these phenotypic dynamics to be realized, the kryptotype dimension \emph{has} to be larger than the dimension of the phenotype.
-Even if components of the system's internal state are directly subject to selection
-and the mode of action of the environment on the internal state is constrained
-(so, the input and output matrices $B$ and $C$ are fixed)
-then one could still perturb $A$ as described above by $ZA - AZ$
-if $ZB$ and $CZ$ are both zero, implying a number of degrees of freedom that still grows with $n^2$
-for fixed $\ell$ and $m$.
-Generically, the number of degrees of freedom is $n(n - \ell - m)$,
-so that in a system of $n$ components, if even one component is not directly constrained,
-this leads to $n$ degrees of freedom.
-Whatever the true ``dimensionality'' of phenotype space of a typical organism,
-there are undoubtedly aspects of its underlying molecular machinery that are not directly constrained,
-suggesting large numbers of degrees of freedom.
-Note that pleiotropy does not directly affect this argument at all --
-indeed, many phenotypically equivalent changes will lead to denser $A$ matrices and hence more pleiotropy.
-However, more pleiotropic genes may be more strongly constrained,
-making it more difficult for systems to make the required compensatory changes for system drift.
-% Whatever it is, however, it seems unlikely 
-% that it would approach the dimensionality of kryptotype space,
-% which is something related to the number of genes.
-% even under universal pleiotropy \citep{universal_pleiotropy, boyle2017expanded},
-
-Phenotypically equivalent system evolution is probably not only driven by neutral genetic drift.
-\revpoint{1}{18} \revpoint{1}{19} \revpoint{1}{21}
-For one thing, movement along the set of equivalent networks
-is not expected to be completely neutral,
-since second-order selection pushes populations towards
-``flatter'' regions of the fitness landscape
-in which a population centered on the optimal set has lower genetic load
-\citep[as described in different contexts by][]{rice1998evolution,neutral_network_physics}.
-If this bias towards more robust networks is strong enough,
-it may even prevent drift, but it is unclear how strong this effect would be in practice.
-Our results, on the other hand, do not rely on the flatness of the fitness surface
-around the phenotypically equivalent set, but rather on the curvature of the equivalent set itself.
-So long as the phenotypically equivalent set is not closed under sexual recombination,
-opportunities for incompatibility remain.
-However the speed at which system drift can generate incompatibilities might diminish
-if selection for robustness is strong enough to constrain a population
-to a small section of system space,
-although the strength of such effects in practice are not known.
-Likewise, as the speed of system drift relies on segregating genetic variation,
-any constraints on such variation,
-possibly due to epistasis, genetic architecture \citep{hermisson2003epistasis}, adaptive inertia \citep{baatz1997adaptive,alvarez2009effects} or weak gene flow could plausibly slow it down.
-More work on specific systems,
-likely coupled with simulations,
-will be necessary to identify the biologically relevant parameter regimes.
-
-\paragraph{Nonlinearity and model assumptions}
-Of course, real regulatory networks are not linear dynamical systems.
-Most notably, physiological limits put upper bounds on expression levels,
-implying saturating response curves.
-It remains to be seen how well our results carry over into real systems,
-but the fact that most nonlinear systems can be locally approximated by a linear one
-suggests our qualitative results may hold more generally. \revpoint{1}{7}
-Furthermore, nonidentifiability (which implies the existence of neutral directions)
-is often found in practice in moderately complex models of biological systems \citep[e.g.,][]{gutenkunst2007universally,piazza2008diverse,jimenez2017spectrum}.
-
-Finally, despite our model's precise separation of phenotype and kryptotype, this relationship in nature may be far more complicated as aspects of the kryptotype may be less ``hidden'' than we currently assume, and the neutral network changes we describe here may only be nearly neutral.
-For instance, attributes excluded from the phenotype as modeled here ignore the potential energy costs associated with excessively large (non-minimal) kryptotypes, as well as the relationship between a specific network architecture and robustness to mutational, transcriptional, or environmental noise.
-% Despite the model's simplifications and potentially nontrivial oversights, it may still roughly reflect and shed light upon speciation patterns as they occur in nature.
-More precise modeling will require better mechanistic understanding not only of biological systems,
-but also the nature of selective pressures
-and genetic variation in these systems.
-
-%Maybe compare to \citet{fierst}?
-%
-%hominids; neanderthal load.
+\textbf{Key words:} Speciation, Models/Simulations, Genetic Drift
 
@@
 
-\bibliographystyle{plainnat}
-\bibliography{krefs}
-%\end{multicols}
-
-\normalsize
-\appendix
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Local expansion of the fitness surface}
-\label{apx:H_calc}
-
-Suppose that $\rho(t) \ge 0$ is a weighting function on $[0,\infty)$
-so that fitness is a function of $L^2(\rho)$ distance of the impulse response from optimal.
-With $h_0(t) = C_0 e^{tA_0} B_0$ a representative of the optimal set:
-\begin{equation}
-    \begin{aligned}
-        D(A, B, C)^2
-        &:= 
-        \int_0^\infty \rho(t) \left| h_A(t) - h_0(t) \right|^2 dt \\
-        &= 
-        \int_0^\infty \rho(t) \left| C e^{At} B - C_0 e^{A_0 t} B_0 \right|^2 dt \\
-        &= 
-        \int_0^\infty \rho(t) \tr\left\{
-            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)^T
-            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)
-        \right\} dt \\
-        &= 
-        \int_0^\infty \rho(t) \tr\left\{
-            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)
-            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)^T
-        \right\} dt  ,
-        % &= 
-        % \int_0^\infty \rho(t)  \tr\left\{
-        %         C e^{At} B B^T e^{A^T t} C^T 
-        %         - C_0 e^{A_0t} B_0 B^T e^{A^T t} C^T 
-        %         \right. \\ &\qquad \qquad \left. {}
-        %         - C e^{At} B B_0^T e^{A_0^T t} C_0^T 
-        %         + C_0 e^{A_0t} B_0 B_0^T e^{A_0^T t} C_0^T 
-        % \right\} dt \\
-        % \int_0^\infty \rho(t) \left| C \left( e^{At} - e^{A_0 t} \right) B \right|^2 dt \\
-        % &= 
-        % \int_0^\infty \rho(t) C \left( e^{At} - e^{A_0 t} \right) B B^T \left( e^{At} - e^{A_0 t} \right)^T C^T dt
-    \end{aligned}
-\end{equation}
-where $\tr X$ denotes the trace of a square matrix $X$.
-How does this change as we perturb about $(A_0, B_0, C_0)$?
-First we differentiate with respect to $A$, keeping $B=B_0$ and $C=C_0$ fixed.
-Since
-\begin{equation}
-  \begin{aligned}
-      \frac{d}{du} e^{(A+uZ)t} \vert_{u=0}
-      &=
-      \int_0^t e^{As} Z e^{A(t-s)} ds, 
-  \end{aligned}
-\end{equation}
-we have that
-\begin{equation}
-  \begin{aligned}
-      \frac{d}{du} D(A+uZ,B_0,C_0)^2 \vert_{u=0}
-      &=
-        2 \int_0^\infty \rho(t) \tr\left\{ C_0 \left( \int_0^t e^{As} Z e^{A(t-s)} ds \right) B_0 B_0^T \left( e^{At} - e^{A_0 t} \right)^T C_0^T \right\} dt \\
-      &=
-        2 \int_0^\infty \rho(t) \tr\left\{ C_0 \left( \int_0^t e^{As} Z e^{A(t-s)} ds \right) B_0 \left( h_A(t) - h_0(t) \right)^T \right\} dt 
-  \end{aligned}
-\end{equation}
-and, by differentiating this and supposing that $A$ is on the optimal set,
-i.e., $h_A(t)=h_0(t)$, (so without loss of generality, $A=A_0$):
-\begin{equation}
-  \begin{aligned}
-      \calH^{A,A}(Y,Z) 
-      &:= 
-      \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0+uY+vZ,B_0,C_0)^2 \vert_{u=v=0} \\
-      &=
-        \int_0^\infty \rho(t) \tr\left\{ C_0 
-        \left( \int_0^t e^{A_0 s} Y e^{A_0 (t-s)} ds \right) 
-        B_0 B_0^T 
-        \left( \int_0^t e^{A_0 s} Z e^{A_0 (t-s)} ds \right)^T
-        C_0^T \right\} dt  .
-  \end{aligned}
-\end{equation}
-
-The function $\calH$ will define a quadratic form.
-To illustrate the use of this, suppose that $B$ and $C$ are fixed.
-% Here $\calH$ is the quadratic form underlying the Hamiltonian.
-By defining $\Delta_{ij}$ to be the matrix with a 1 in the $(i,j)$th slot
-and 0 elsewhere,
-the coefficients of the quadratic form are
-\begin{equation}
-    \begin{aligned}
-        H_{ij, k\ell}(A)
-        &:=
-        \calH(\Delta_{ij}, \Delta_{k\ell}) .
-    \end{aligned}
-\end{equation}
-
-We could use this to get the quadratic approximation to $D$ near the optimal set.
-To do so, it'd be nice to have a way to compute the inner integral above.
-Suppose that we diagonalize $A = U \Lambda U^{-1}$.
-Then
-\begin{equation} \label{eqn:exp_deriv}
-  \begin{aligned}
-      \int_0^t e^{As} Z e^{A(t-s)} ds 
-      &=
-      \int_0^t U e^{\Lambda s} U^{-1} Z U e^{\Lambda (t-s)} U^{-1} ds 
-  \end{aligned}
-\end{equation}
-Now, notice that
-\begin{equation}
-  \begin{aligned}
-      \int_0^t e^{s \lambda_i} e^{(t-s) \lambda_j} ds
-      &= \begin{cases}
-      \frac{ e^{t \lambda_i} - e^{t \lambda_j} }{ \lambda_i - \lambda_j } 
-          \qquad & \text{if} \quad i \neq j \\
-          t e^{t \lambda_i} 
-          \qquad & \text{if} \quad i = j .
-      \end{cases}
-  \end{aligned}
-\end{equation}
-Therefore, 
-defining
-\begin{equation}
-    \begin{aligned}
-    X_{ij}(t,Z) 
-        &= \begin{cases}
-        \left( U^{-1} Z U \right)_{ij}
-      \frac{ e^{t \lambda_i} - e^{t \lambda_j} }{ \lambda_i - \lambda_j } 
-          \qquad & \text{if} \quad i \neq j \\
-      \left( U^{-1} Z U \right)_{ii}
-          t e^{t \lambda_i} 
-          \qquad & \text{if} \quad i = j ,
-        \end{cases}
-    \end{aligned}
-\end{equation}
-moving the $U$ and $U^{-1}$ outside the integral and integrating we get that
-\begin{equation}
-  \begin{aligned}
-      \int_0^t e^{As} Z e^{A(t-s)} ds 
-      &=
-      U X(t,Z) U^{-1} .
-  \end{aligned}
-\end{equation}
-This implies that
-\begin{equation}
-    \begin{aligned}
-        D(A_0+\epsilon Z)^2
-        &\approx \frac{1}{2} \epsilon^2 
-        \int_0^\infty
-            \rho(t) \tr\left\{ C U X(t,Z) U^{-1} B B^T (U^{-1})^T X(t,Z)^T U^T C^T \right\}
-        dt .
-    \end{aligned}
-\end{equation}
-
-To compute the $n^2 \times n^2$ matrix $H$,
-we see that if $Z=\Delta_{k \ell}$, then
-\begin{equation}
-  \begin{aligned}
-    X_{ij}^{k\ell}(t) 
-    &= \begin{cases}
-      (U^{-1})_{\cdot k} U_{\ell \cdot}
-      \frac{ e^{t \lambda_i} - e^{t \lambda_j} }{ \lambda_i - \lambda_j } 
-          \qquad & \text{if} \quad i \neq j \\
-      (U^{-1})_{\cdot k} U_{\ell \cdot}
-      t e^{t \lambda_i} 
-          \qquad & \text{if} \quad i = j ,
-    \end{cases}
-  \end{aligned}
-\end{equation}
-where $U_{k \cdot}$ is the $k$th row of $U$,
-and so
-\begin{equation}
-    \begin{aligned}
-        H_{ij, k\ell}(A)
-        &=
-        \int_0^\infty
-            \rho(t) \tr\left\{ C U X^{ij}(t) U^{-1} B B^T (U^{-1})^T X^{k\ell}(t)^T U^T C^T \right\}
-        dt .
-    \end{aligned}
-\end{equation}
-This implies that
-\begin{equation}
-    \begin{aligned}
-        D(A_0+\epsilon Z)^2
-        &\approx \frac{1}{2} \epsilon^2 \sum_{ijk\ell} H_{ij,k\ell}(A_0) Z_{ij} Z_{k\ell}  .
-    \end{aligned}
-\end{equation}
-
-% By section \ref{ss:quant_gen},
-% if we set $\Sigma=\sigma^2 I$ and $U=H$,
-% then a population at $A_0+Z$ experiences a restoring force of strength
-% $(I + \sigma^2 H^{-1})^{-1} Z$ (treating $Z$ as a vector and $H$ as an operator on these).
-% If $\sigma^2$ is small compared to $H^{-1}$
-% then this is approximately $-\sigma^2 H^{-1} Z$.
-% This suggests that the population mean follows an Ornstein-Uhlenbeck process,
-% as described (in different terms) in \citet{hansen1996translating}.
-
-More generally, $B$ and $C$ may also change.
-To extend this we need the remaining second derivatives of $D^2$.
-First, in $B$:
-\begin{equation}
-    \begin{aligned}
-        \mathcal{H}^{B,B}(Y,Z)
-        &:= 
-        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0,B_0+uY+vZ,C_0)\vert_{u=v=0} \\
-        &=
-        \frac{1}{2} \int_0^\infty \rho(t) \tr\left\{ 
-        C_0 e^{t A_0}
-        \frac{d}{du} \frac{d}{dv} 
-        (uY+vZ)
-        (uY+vZ)^T
-        \vert_{u=v=0} 
-        e^{t A_0^T} C_0^T 
-        \right\} dt  \\
-        &=
-        \frac{1}{2} \int_0^\infty \rho(t) \tr\left\{ 
-        C_0 e^{t A_0}
-        \left( Y Z^T + Z Y^T \right)
-        e^{t A_0^T} C_0^T 
-        \right\} dt  .
-  \end{aligned}
-\end{equation}
-Next, in $C$:
-\begin{equation}
-    \begin{aligned}
-        \mathcal{H}^{B,B}(Y,Z)
-        &:= 
-        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0,B_0,C_0+uY+vZ)\vert_{u=v=0} \\
-        &=
-        \frac{1}{2} \int_0^\infty \rho(t) \tr\left\{ 
-        B_0 e^{t A_0^T}
-        \frac{d}{du} \frac{d}{dv} 
-        (uY+vZ)^T
-        (uY+vZ)
-        \vert_{u=v=0} 
-        e^{t A_0} B_0
-        \right\} dt  \\
-        &=
-        \frac{1}{2} \int_0^\infty \rho(t) \tr\left\{ 
-        B_0 e^{t A_0^T}
-        \left( Y Z^T + Z Y^T \right)
-        e^{t A_0} B_0
-        \right\} dt  .
-  \end{aligned}
-\end{equation}
-Now, the mixed derivatives in $B$ and $C$:
-\begin{equation}
-    \begin{aligned}
-        \mathcal{H}^{B,C}(Y,Z)
-        &:= 
-        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0,B_0+uY,C_0+vZ)\vert_{u=v=0} \\
-        &=
-        \int_0^\infty \rho(t) \tr\left\{ 
-        Y e^{t A_0^T} C_0^T Z e^{t A_0} B_0
-        \right\} dt  .
-  \end{aligned}
-\end{equation}
-In $A$ and $B$
-\begin{equation}
-    \begin{aligned}
-        \mathcal{H}^{A,B}(Y,Z)
-        &:= 
-        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0+uY,B_0+vZ,C_0)\vert_{u=v=0} \\
-        &=
-        \int_0^\infty \rho(t) \tr\left\{ 
-        C_0 \left(\int_0^t e^{s A_0} Y e^{(t-s) A_0} ds \right) B_0
-        Z^T e^{t A_0} C_0
-        \right\} dt  ,
-  \end{aligned}
-\end{equation}
-and finally in $A$ and $C$:
-\begin{equation}
-    \begin{aligned}
-        \mathcal{H}^{A,C}(Y,Z)
-        &:= 
-        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0+uY,B_0,C_0+vZ)\vert_{u=v=0} \\
-        &=
-        \int_0^\infty \rho(t) \tr\left\{ 
-        C_0 \left(\int_0^t e^{s A_0} Y e^{(t-s) A_0} ds \right) B_0
-        B_0 e^{t A_0} Z
-        \right\} dt  .
-  \end{aligned}
-\end{equation}
-
-Together, numerical computation of these expressions,
-along with estimates of genetic covariance within a population,
-allow precise predictions of evolutionary dynamics of a particular system.
-The approximation should be good as long as the second-order Taylor approximation holds.
-
-% reviews here if toggled above
-% \ifsubmission\processdelayedfloats\fi
-
-\includereviews
-
 \end{document}

### 2021-05-03 title_page_revision.tex
made title page

@@
 {\Huge \bf System drift and speciation} \\ \vspace{0.75cm}
-Joshua S. Schiffman$^{\dagger}$ \qquad Peter L. Ralph$^{\ddagger}$ \\ \vspace{0.5cm}
-$^{\dagger}${\footnotesize {New York Genome Center, New York, New York 10013, U.S.A \& Weill Cornell Medicine, New York, New York 10065, USA  \\
-$^{\ddagger}$Departments of Mathematics and Biology \& The Institute for Ecology and Evolution, University of Oregon, Eugene, Oregon 97403, USA}} \\ \vspace{0.5cm}
+Joshua S. Schiffman$^{\dagger \ast}$ \qquad Peter L. Ralph$^{\ddagger \ast}$ \\ \vspace{0.5cm}
+$^{\dagger}${\footnotesize {Current affiliations: New York Genome Center, NY, NY 10013 \& Weill Cornell Medicine, NY, NY 10065  \\
+$^{\ddagger}$Departments of Mathematics and Biology \& The Institute for Ecology and Evolution, University of Oregon, Eugene, OR 97403 \\ 
+$^{\ast}$Work conducted at: Molecular and Computational Biology, University of Southern California, LA, CA 90089} } \\ \vspace{0.5cm}
 {\small \texttt{jschiffman@nygenome.org} \qquad 

### 2021-05-06 kalman_walk_paper.tex
appendix

@@
 
-Suppose that $\rho(t) \ge 0$ is a weighting function on $[0,\infty)$
-so that fitness is a function of $L^2(\rho)$ distance of the impulse response from optimal.
-With $h_0(t) = C_0 e^{tA_0} B_0$ a representative of the optimal set:
+The fitness of a system 
+depends on the difference between the system's impulse response and the optimal impulse response,
+measured  as a weighted sum of
+the distance between the impulse response from optimal.
+With $\rho(t) \ge 0$ a weighting function on $[0,\infty)$,
+and $h_0(t) = C_0 e^{tA_0} B_0$ a representative of the optimal set,
+the distance is equation \eqref{eqn:distance}:
+\begin{align}
+	D(h) = \left( \int_0^\infty \rho(t) \|h(t) - h_0(t)\|^2 dt \right)^{1/2} .
+\end{align}
+To see how this is affected by small changes to the system,
+first note that if $\|\cdot\|_\rho$ is the $L^2(\rho)$ norm,
+\begin{align}
+	\|g\|_\rho^2 := \int_0^\infty \rho(t) \|g(t)\|^2 dt
+\end{align}
+then the distance to optimum of a perturbed system, $h(t) = h_0(t) + \epsilon g(t)$, is
+\begin{align}
+	D(h_0 + \epsilon g) = \epsilon \|g\|_\rho .
+\end{align}
+
+Now consider the perturbed systems $(A(\epsilon), B(\epsilon), C(\epsilon)) = (A(0), B(0), C(0)) + \epsilon (U, V, W)$,
+and let $\h_\epsilon(t) = C(\epsilon) e^{A(\epsilon) t} B(\epsilon)$.
+If two populations with systems $(A(0), B(0), C(0))$ and $(A(\epsilon), B(\epsilon), C(\epsilon))$ interbreed,
+then the mean resulting system is $(A({\epsilon/2}), B({\epsilon/2}), C({\epsilon/2}))$,
+but with greater variance in the $F_2$ than the $F_1$ offspring (see Figure \ref{fig:conceptual_fig}).
+Indeed, if the difference between (say) $A(0)_{ij}$ and $A(\epsilon)_{ij}$
+is due to $n$ fixed differences with additive effect size $\epsilon/(2n)$ each,
+then the standard deviation of $A_{ij}$ among the $F_1$ offspring is equal to that
+within each parent population, while among the $F_2$ offspring it is $\epsilon / \sqrt{8 n}$.
+This implies that the impulse response, $h(t)$, of an $F_2$ will also differ from optimum
+by something of order $\epsilon$.
+However, since the displacement between the populations lies along a ridge,
+the $F_1$ offspring are closer:
+the Taylor expansions for $h$ about both 0 and $\epsilon$ are
+\begin{align}
+    h_{\epsilon/2}(t)
+    &= h_0(t)
+        + \frac{\epsilon}{2} \partial_\epsilon h_0(t)
+        + \frac{\epsilon^2}{8} \partial_\epsilon^2 h_0(t)
+        + O(\epsilon^3) \\
+    \text{and} \quad
+    &= h_\epsilon(t)
+        - \frac{\epsilon}{2} \partial_\epsilon h_\epsilon(t)
+        + \frac{\epsilon^2}{8} \partial_\epsilon^2 h_\epsilon(t)
+        + O(\epsilon^3)
+\end{align}
+Since $h_\epsilon(t) = h_0(t)$ by assumption,
+and $\partial_\epsilon h_\epsilon(t) = \partial_\epsilon h_0(t) + \epsilon \partial_\epsilon^2 h_0(t) + O(\epsilon^2)$,
+combining these expressions we get that
+\begin{align}
+    h_{\epsilon/2}(t)
+    &= h_0(t)
+        - \frac{\epsilon^2}{8} \partial_\epsilon^2 h_0(t)
+        + O(\epsilon^3) .
+\end{align}
+In other words, if the parental populations are both optimal but differ by $\epsilon$,
+then the $F_1$ offspring differ from optimal by something of order $\epsilon^2$,
+but the $F_2$ by something of order $\epsilon$.
+
+Of course, when moving away from the ridge, the fitness landscape is flatter in some directions than others,
+and how much so depends on the details.
+For completeness, we describe here how to compute the sensitivity of distance to optimum
+with respect to each of $A$, $B$, and $C$.
+Changing notation slightly, let's rewrite the squared distance to optimum
+of a system $(A, B, C)$:
 \begin{equation}

### 2021-05-06 kalman_walk_paper.tex
.

@@
 Indeed, if the difference between (say) $A(0)_{ij}$ and $A(\epsilon)_{ij}$
-is due to $n$ fixed differences with additive effect size $\epsilon/(2n)$ each,
+is due to $n$ fixed differences with absolute additive effect size $\epsilon/\sqrt{n}$ each,
 then the standard deviation of $A_{ij}$ among the $F_1$ offspring is equal to that
-within each parent population, while among the $F_2$ offspring it is $\epsilon / \sqrt{8 n}$.
+within each parent population, while among the $F_2$ offspring it is $\epsilon/2$.
+(This many differences of this size would be expected if the difference was due to drift, for instance.)
 This implies that the impulse response, $h(t)$, of an $F_2$ will also differ from optimum

### 2021-05-07 kalman_walk_paper.tex
.

@@
 then the standard deviation of $A_{ij}$ among the $F_1$ offspring is equal to that
-within each parent population, while among the $F_2$ offspring it is $\epsilon/2$.
+within each parent population, while a simple binomial calculation
+shows that among the $F_2$ offspring it is $\epsilon/2$
+(plus a contribution for within-population genetic variance).
 (This many differences of this size would be expected if the difference was due to drift, for instance.)

### 2021-05-07 kalman_walk_paper.tex
.

--- 2021-05-06 kalman_walk_paper.tex
@@
 is due to $n$ fixed differences with absolute additive effect size $\epsilon/\sqrt{n}$ each,
-then the standard deviation of $A_{ij}$ among the $F_1$ offspring is equal to that
-within each parent population, while among the $F_2$ offspring it is $\epsilon/2$.
+then the standard deviation of $A_{ij}$ among the $F_1$ offspring is still equal to that
+within each parent population, as each locus is heterozygous.
+However, if loci are unlinked then an $F_2$ offspring is homozygous with probability 1/2,
+and takes either homozygote with equal probability,
+so the variance of the contribution of each locus is $\epsilon/4\sqrt{n}$,
+and so the standard deviation across $F_2$ offspring is $\epsilon/4$
+plus a contribution for within-population genetic variance.
 (This many differences of this size would be expected if the difference was due to drift, for instance.)

### 2021-05-07 kalman_walk_paper.tex
adjustments for resubmission

@@
 {\Huge \bf System drift and speciation} \\ \vspace{0.75cm}
-Joshua S.
+%Joshua S.
 Schiffman$^{\dagger}$ \footnote{Current affiliations: New York Genome Center, New York, New York 10013, U.S.A.
@@
 Ralph$^{\dagger \ddagger}$ \\ \vspace{0.5cm}
-$^{\dagger}${\footnotesize {Molecular and Computational Biology, University of Southern California, Los Angeles, California 90089, U.S.A. \\
-$^{\ddagger}$Departments of Mathematics and Biology \& The Institute for Ecology and Evolution, University of Oregon, Eugene, Oregon 97403, U.S.A.}} \\ \vspace{0.5cm}
-{\small \texttt{jschiffman@nygenome.org} \qquad 
-\texttt{plr@uoregon.edu}} \\ \vspace{0.5cm}
+%$^{\dagger}${\footnotesize {Molecular and Computational Biology, University of Southern California, Los Angeles, California 90089, U.S.A. \\
+%$^{\ddagger}$Departments of Mathematics and Biology \& The Institute for Ecology and Evolution, University of Oregon, Eugene, Oregon 97403, U.S.A.}} \\ \vspace{0.5cm}
+%{\small \texttt{jschiffman@nygenome.org} \qquad 
+%\texttt{plr@uoregon.edu}} \\ \vspace{0.5cm}
 \small \today \\
@@
 \end{abstract}
+
+\textbf{Key words:} Speciation, Models/Simulations, Genetic Drift
 
@@
 Now consider the perturbed systems $(A(\epsilon), B(\epsilon), C(\epsilon)) = (A(0), B(0), C(0)) + \epsilon (U, V, W)$,
-and let $\h_\epsilon(t) = C(\epsilon) e^{A(\epsilon) t} B(\epsilon)$.
+and let $h_\epsilon(t) = C(\epsilon) e^{A(\epsilon) t} B(\epsilon)$.
 If two populations with systems $(A(0), B(0), C(0))$ and $(A(\epsilon), B(\epsilon), C(\epsilon))$ interbreed,

### 2021-05-07 kalman_walk_paper.tex
removed ack

@@
 
-\section*{Acknowledgements}
-
-    We would like to thank Sergey Nuzhdin, Stevan Arnold, Michael Turelli, Patrick Phillips, Erik Lundgren and Hossein Asgharian for valuable discussion. 
-    We would also like to thank Nick Barton, Sarah Signor, Todd Parsons, and Joachim Hermisson for very helpful comments on the manuscript.
-    Work on this project was supported by funds from
-    the Sloan Foundation and the NSF (under DBI-1262645) to PR.
+%\section*{Acknowledgements}
+%
+%    We would like to thank Sergey Nuzhdin, Stevan Arnold, Michael Turelli, Patrick Phillips, Erik Lundgren and Hossein Asgharian for valuable discussion. 
+%    We would also like to thank Nick Barton, Sarah Signor, Todd Parsons, and Joachim Hermisson for very helpful comments on the manuscript.
+%    Work on this project was supported by funds from
+%    the Sloan Foundation and the NSF (under DBI-1262645) to PR.
 

### 2021-07-06 review-responses2.tex
started response section

%%%%%%
%%
%%  Don't reorder the reviewer points; that'll mess up the automatic referencing!
%%
%%%%%

\begin{minipage}[b]{2.5in}
  Revision Cover Letter \\
  {\it Evolution}
\end{minipage}
\hfill
\begin{minipage}[b]{2.5in}
    Joshua Schiffman \\
    \emph{and} Peter Ralph \\
  \today
\end{minipage}

\vskip 2em

\noindent
{\bf To the Editor(s) -- }

\vskip 1em

 Cover letter.

\noindent \hspace{4em}
\begin{minipage}{3in}
\noindent
{\bf Sincerely,}

\vskip 2em

{\bf 
Joshua Schiffman and
Peter Ralph
}\\
\end{minipage}

\vskip 4em

\pagebreak

%%%%%%%%%%%%%%
\reviewersection{AE}
\begin{quote}
This is a very interesting theory paper that connects the modeling and evolution of regulatory networks with speciation. While quite theoretical in nature, the authors made an effort to provide examples, and I consider this paper of very high relevance for the field.

The authors have addressed most of the previous comments from the first round of reviews, and the two new reviewers only raise minor concerns and suggestions, especially regarding the paragraph on Haldane’s rule and regarding a better placement of the paper in the context of speciation research. I agree with these concerns; regarding the latter point I have three additional questions/comments, one of which was also raised in the previous round of reviews:
\end{quote}

Thank you for the kind words.

\begin{point}{Gene flow}
As it is described, the model only holds for allopatric populations; I would expect even a small amount of gene flow to make neutral evolution of hybrid incompatibility impossible (Gavrilets 1997, Bank et al. 2012). This is not discussed in the current version of the paper; even more interestingly, would the authors expect conditions in which emergence of incompatibility could be possible in this model also in the presence of gene flow?
\end{point}

\reply{
}

\begin{point}{Orr's Snowball}
The authors state that rapid emergence of incompatibility is possible in their model. How would the accumulation of incompatibility here compare with Orr’s ``snownball'' model, and how is this expected to depend on the complexity of the underlying networks?
\end{point}

\reply{
}

\begin{point}{Genetic architecture}
The genetic architecture (e.g., interactions between cis-regulatory and coding regions) is important to determine the barrier to gene flow that an incompatibiltiy builds. If there is a lot of short-distance interactions (physically in the genome; i.e. low recombination), they might not play a role for speciation per se because the incompatibility will often not be exposed to selection even in hybrids. In the light of our current knowledge of the genetic archtiecture of regulatory networks, how would this affect the conclusions from the model?
\end{point}

\reply{
}

\reviewersection{1}
\begin{quote}
I am basically enthusiastic about this paper, which has been influential for some time on bioRxiv. The central sections, especially e.g. lines 127-219, are very clear and novel. I do think, though that the work could be better placed in context.
\end{quote}

Thank you. We were also excited that the paper had many readers on the bioRxiv.

\begin{point}{Empirical evidence for the (minor?) role of drift in speciation:}
%1. Empirical evidence for the (minor?) role of drift in speciation:
%
Lines 382-389 discuss previous discussion about the role of drift in speciation, but they focus on theory. I think that some strong arguments for the primacy of selection in speciation have been empirical (bottlenecking Drosophila populations and testing for RI, MK-tests on ``speciation genes'' etc.). Some of the earlier work was summarized in the last chapter of Coyne and Orr 2004. Line 396 seems to ignore a lot of this work, and so I would rephrase.
\end{point}

\reply{
}

\begin{point}{Additive quantitative trait models}
%2. Additive quantitative trait models
%
The previous version of this manuscript contained a standard quantitative trait model. I think it was a good idea to remove this material, but I still think that this modelling approach (including Fisher’s geometric model) needs to be discussed a little more here. In the introduction, for example, it seems important to state clearly that ``kryptotypes'' (many-to-one genotype-to-phenotype maps) are also a feature of these standard models, and that previous authors have shown how phenotypically identical populations can evolve RI, in some cases via drift alone (e.g., Mani and Clarke 1990; Rosas et al. 2010; Fraisse et al. 2016 and others).

Martin 2014 argued that the results of these models and ``systems-biology'' type models are likely to approximate one another under some conditions, and that could be mentioned.
\end{point}

\reply{
}

\begin{point}{RI due to kryptotypes not involving drift}
%3. RI due to kryptotypes not involving drift
%
The section beginning on line 419 seems quite important to me, but it seems to downplay an important point. Populations might diverge genetically, or in their kryptotypes, but not in their phenotypes, if they experience fluctuating selection. This goes back the Wright-Fisher debates, and has been discussed in the context of speciation by, e.g., Mani and Clarke 1990, Barton 2001 and later authors. I think that the empirical evidence for fluctuating selection is also quite strong (e.g Bell 2010), and so it seems more important to discuss that, e.g., selection for robustness.
\end{point}

\reply{
}

\begin{point}{Interpretation of the Haldane’s Rule result}
%4. Interpretation of the Haldane’s Rule result
%
The abstract refers to ``another possible explanation'' of Haldane's Rule, and the explanation given is said to be quite different from the ``dominance theory'', because ``it derives from the nature of segregation variance''.  I wonder whether this is not misleading. My view is that the sort of model explored here (like additive quantitative trait model used by Barton 2001) naturally generates the sort of dominance relations required for the ``dominance theory'' to hold. Unless the authors strongly disagree, I would modify the abstract and lines 278-291, or at least provide a fuller clarification for why the expanation is different from that discussed by Barton 2001.
\end{point}

\reply{
}

\begin{point}{}
5. (very minor) Line 358: As you say, F2 hybrid breakdown has been known about for a very long time. As such, the single citation used feels a bit arbitrary.
\end{point}

\reply{
}

\reviewersection{2}
\begin{quote}
The manuscript explores how regulatory changes yielding the same phenotype may lead to hybrid incompatibilities and speciation. The authors model gene regulatory networks as sets of linear equations, where environmental inputs are integrated, giving rise to specific phenotypic outputs. Using Kalman decomposition, the authors identify the subset of such systems that produce the same phenotype and thus, characterise how a given system may be modified without affecting its phenotypic response. Resorting to a simple oscillatory network as an example, the authors show that phenotypically identical parents can generate offspring with considerably different phenotypes, resulting in hybrid incompatibilities if the set of equivalent phenotypes is not closed under the average. The phenotypic distance between parents and offspring is proportional to the square of the genetic distance between parents for F1 hybrids and directly proportional to the genetic distance between parents for F2 hybrids.

This is an interesting paper that nicely bridges core concepts of systems biology and evolution. The examples provided are very ``visual'', making quite clear the idea of ``equivalent systems'' and will greatly assist readers not familiar with Kalman decomposition. The figures are both informative and elegant, and the discussion aptly covers model validity and makes an effort to relate this work to standard speciation models. In addition, the authors have adequately addressed the comments of previous reviewers. Overall, this manuscript constitutes an appropriate addition to Evolution.

We recommend this paper for publication, but we have some suggestions that we consider would improve the clarity of this paper for a broader biological audience.
\end{quote}

Thank you.

\begin{point}{Haldane's rule}
Our primary concern with the text regards Haldane's rule. The authors state that a male F1 offspring has a phenotype that could usually only be obtained in an F2 individual if all the involved genes were autosomal. Although this is true, it does not look like an arbitrary F2 individual but one with a very particular structure. This type of individual is like F1 in all autosomal genes but equal to the parental genotype in the genes present in the sex chromosome. It is not clear to us that this should necessarily generate a worse outcome than a female individual.

The F2 individuals only perform worse than F1 on average, and looking at figure 4 (left), we can see that some F2 individuals actually perform better than the F1 individuals.
\end{point}

\reply{
  The fitness of heterogametic F1s and F2s will always depend on the genetic distance between parents and on the particular gene network. Thus it is not guaranteed that the heterogametic F1s, will always have lower fitnesses than homogametic F1s. This is clear when the genetic distance is zero between parents. The model does show, however, that the particular combination of genes found in F1 heterogametes would only be possible in F2 autosomal crosses. Our model shows that phenotypes generated by networks contained on the autosomes have lower fitnesses in F2s than in F1s, and by extension, lower fitness in the heterogametic sex when networks are contained, at least in part, on the sex chromosomes.

  It is also true that some F2 individuals will perform as well as F1s and F0s -- for instance it is possible to recover an F0-like genotype (and therefore phenotype) after two generations of hybridization. In this model, hybrids have lower fitness because of increased genetic variance. However, there will also be F2 individuals with genes from both parental types -- meaning that there will be more low fitness F2s than low fitness F1s and F0s. \jss{(The F2 fitness is lower on average?)}
}

\begin{point}{}
(p. 2, l. 73): This paragraph summarises the main body of work performed in this paper. Thus, consider moving it to the end of the Introduction (i.e. switching places with the last paragraph).
\end{point}
\reply{
}
\begin{point}{}
(p. 2, l. 92): ``...more than one way to do the same thing...'' sounds a bit too colloquial. Consider changing it to ``...more than one way to accomplish the same outcome...'' or some other alternative phrasing.
\end{point}

\reply{
}

\begin{point}{}
(p. 5, l. 192): ``The Kalman decomposition then classifies each direction in kryptotype space as either reachable or unreachable''. Is there a biological motivation for considering unreachable directions in kryptotype space as degrees of freedom of the neutral space?
\end{point}

\reply{
}

\begin{point}{}
  (p. 6, l. 208): ``both the change of basis used to obtain the decomposition and, once in this form, all submatrices other than $A_{ro}$, $B_{ro}$, and $C_{ro}$ can be changed without affecting the phenotype, and so represent degrees of freedom''. A minor question on generality - does Kalman decomposition require that the coefficient matrices are time-invariant?
\end{point}

\reply{
  As far as we understand, and as applied here, the Kalman decomposition is for linear time-invariant systems. Generalizations of the Kalman decomposition for time-varying systems have been discussed in...
  \jss{Could cite some papers? Looks like Silverman and Meadows, EQUIVALENT REALIZATIONS OF LINEAR SYSTEMS discuss generalizations to the time-variable case.}
}

\begin{point}{Recombination}
(p. 7) In the section ``Sexual reproduction and recombination'', we think that the recombination process is not clearly explained, making it more challenging to read the text from this point. We suggest that the authors clarify how exactly recombination is implemented in the model. Perhaps an example would help the reader to have a better picture of the process. This example could be inline, like the one introduced in the section on Haldane's rule.
\end{point}

\reply{
}

\begin{point}{}
(p. 8, Figure 4 and p. 10, Figure 5): We suggest dividing the figure into sub-figures (e.g. A and B). Furthermore, we suggest adding ``F1'' and ``F2'' labels to the left and right plots in Figure 5, respectively, as well as the values of epsilon. This way, the reader would be immediately informed of what is changing.
\end{point}

\reply{
}

### 2021-07-06 review-responses2.tex
added links to reviewer commented lines

@@
 The central sections, especially e.g.
-lines 127-219, are very clear and novel.
+\hyperlink{rev1:1}{lines 127-219}, are very clear and novel.
 I do think, though that the work could be better placed in context.
@@
 %
-Lines 382-389 discuss previous discussion about the role of drift in speciation, but they focus on theory.
+  \hyperlink{rev1:2}{Lines 382-389} discuss previous discussion about the role of drift in speciation, but they focus on theory.
 I think that some strong arguments for the primacy of selection in speciation have been empirical (bottlenecking Drosophila populations and testing for RI, MK-tests on ``speciation genes'' etc.).
 Some of the earlier work was summarized in the last chapter of Coyne and Orr 2004.
-Line 396 seems to ignore a lot of this work, and so I would rephrase.
+\hyperlink{rev1:3}{Line 396} seems to ignore a lot of this work, and so I would rephrase.
 \end{point}
@@
 %
-The section beginning on line 419 seems quite important to me, but it seems to downplay an important point.
+  The section beginning on \hyperlink{rev1:4}{line 419} seems quite important to me, but it seems to downplay an important point.
 Populations might diverge genetically, or in their kryptotypes, but not in their phenotypes, if they experience fluctuating selection.
@@
 My view is that the sort of model explored here (like additive quantitative trait model used by Barton 2001) naturally generates the sort of dominance relations required for the ``dominance theory'' to hold.
-Unless the authors strongly disagree, I would modify the abstract and lines 278-291, or at least provide a fuller clarification for why the expanation is different from that discussed by Barton 2001.
+Unless the authors strongly disagree, I would modify the abstract and \hyperlink{rev1:5}{lines 278-291}, or at least provide a fuller clarification for why the expanation is different from that discussed by Barton 2001.
 \end{point}
@@
 \begin{point}{}
-5. (very minor) Line 358: As you say, F2 hybrid breakdown has been known about for a very long time. As such, the single citation used feels a bit arbitrary.
+  5. (very minor) \hyperlink{rev1:6}{Line 358}: As you say, F2 hybrid breakdown has been known about for a very long time. As such, the single citation used feels a bit arbitrary.
 \end{point}
@@
 \begin{point}{}
-(p. 2, l. 73): This paragraph summarises the main body of work performed in this paper. Thus, consider moving it to the end of the Introduction (i.e. switching places with the last paragraph).
+  \hyperlink{rev2:1}{(p.
+2, l.
+73)} \jss{Note: R2 is using line numbers from the diff}: This paragraph summarises the main body of work performed in this paper.
+Thus, consider moving it to the end of the Introduction (i.e.
+switching places with the last paragraph).
 \end{point}
@@
 \begin{point}{}
-(p.
+  \hyperlink{rev2:2}{(p.
 2, l.
-92): ``...more than one way to do the same thing...'' sounds a bit too colloquial.
+92)}: ``...more than one way to do the same thing...'' sounds a bit too colloquial.
 Consider changing it to ``...more than one way to accomplish the same outcome...'' or some other alternative phrasing.
@@
 \begin{point}{}
-(p.
+  \hyperlink{rev2:3}{(p.
 5, l.
-192): ``The Kalman decomposition then classifies each direction in kryptotype space as either reachable or unreachable''.
+192)}: ``The Kalman decomposition then classifies each direction in kryptotype space as either reachable or unreachable''.
 Is there a biological motivation for considering unreachable directions in kryptotype space as degrees of freedom of the neutral space?
@@
 \begin{point}{}
-  (p.
+\hyperlink{rev2:4}{(p.
 6, l.
-208): ``both the change of basis used to obtain the decomposition and, once in this form, all submatrices other than $A_{ro}$, $B_{ro}$, and $C_{ro}$ can be changed without affecting the phenotype, and so represent degrees of freedom''.
+208)}: ``both the change of basis used to obtain the decomposition and, once in this form, all submatrices other than $A_{ro}$, $B_{ro}$, and $C_{ro}$ can be changed without affecting the phenotype, and so represent degrees of freedom''.
 A minor question on generality - does Kalman decomposition require that the coefficient matrices are time-invariant?

### 2021-07-06 kalman_walk_paper.tex
started response section
added links to reviewer commented lines

@@
 % \reviewresponsesfalse  % don't include them
-\newcommand{\responsefile}{review-responses}  % name of the review responses file
+\newcommand{\responsefile}{review-responses2}  % name of the review responses file
 
@@
 For instance, there are ``neutral networks''
-of nucleic acid sequences that produce the same RNA secondary structure \citep{gruner1996analysis}\revpoint{1}{5},
+of nucleic acid sequences that produce the same RNA secondary structure \citep{gruner1996analysis}%\revpoint{1}{5},
 amino acid sequences that fold similarly \citep{babajide1997neutral},
@@
 and neural circuits \citep{trojanowski2014neural}, and in developmental systems \citep{vondassow2000segment,true2001developmental}. 
-\revpoint{1}{1} \revpoint{1}{3}
+%\revpoint{1}{1} %\revpoint{1}{3}
 % \jss{Other ideas instead of ``degeneracy'' or ``isofunctional structural plasticity'': non-invertibility, isofunctional plasticity, structural plasticity, phenotypic equivalency, phenotypic invariance, functional invariance, isofunctionality, functional equivalency, genotypic plasticity, conjugacy, functional symmetry, isologous, isomorphic.}
@@
 For instance, in many contexts mathematical models
-are fundamentally \emph{nonidentifiable} and/or \emph{indistinguishable} -- \revpoint{1}{2}
+are fundamentally \emph{nonidentifiable} and/or \emph{indistinguishable} -- %\revpoint{1}{2}
 meaning that there can be uncertainty about an inferred model's parameters or even its claims about
@@
 
-In this paper we use results on mathematical nonidentifiability from
+In this paper we use results on mathematical nonidentifiability from\hypertarget{rev2:1}
 linear systems theory to study how gene regulatory networks can be modified
@@
 a population diffuses along the neutral ridges of a high-dimensional space of possible system parameters,
-in a similar vein as \emph{holey landscape} models \citep{gavrilets1997evolution, yamaguchi2013first, pina2019does}.  \revpoint{2}{12}
+in a similar vein as \emph{holey landscape} models \citep{gavrilets1997evolution, yamaguchi2013first, pina2019does}.  %\revpoint{2}{12}
 % Our model is developed in more detail in a companion paper \citep{schiffman2020genetic}.
@@
 The field of population genetics has also explored the consequences of the fact
-that there is often more than one way to do the same thing, 
+that there is often more than one way to do the same thing,\hypertarget{rev2:2}e
 and observed that speciation might be the result of changes that are themselves neutral.
 Indeed, \citet{bateson1909heredity} first proposed that what today we call a
-Bateson-Dobzhansky-Muller incompatibility would likely arise through neutral changes. \revpoint{1}{10}
+Bateson-Dobzhansky-Muller incompatibility would likely arise through neutral changes. %\revpoint{1}{10}
 The potential for speciation has been analyzed in models of 
@@
 It has also been shown that population structure
-can allow long-term stable coexistence of incompatible genotypes encoding identical phenotypes \citep{phillips1996maintenance}. \revpoint{1}{9}
+can allow long-term stable coexistence of incompatible genotypes encoding identical phenotypes \citep{phillips1996maintenance}. %\revpoint{1}{9}
 However, previous simulations of system drift in regulatory sequences \citep{tulchinsky2014hybrid}
@@
 Under quite general assumptions on the input
-(e.g., $|u(t)|$ is integrable) \revpoint{AE}{3}
-we can write the phenotype as \revpoint{AE}{4}
+(e.g., $|u(t)|$ is integrable) %\revpoint{AE}{3}
+we can write the phenotype as %\revpoint{AE}{4}
   \begin{align}
@@
 
-\citet{wagner1994evolution} and others have used a similar discrete-time model \revpoint{1}{6}
+\citet{wagner1994evolution} and others have used a similar discrete-time model %\revpoint{1}{6}
 (that might be written $\phi_{t+1} = f(A \phi_t)$, where $f$ is a sigmoid).
@@
 
-\begin{example}[An oscillator]\label{ex:oscillator}
+\begin{example}[An oscillator]\label{ex:oscillator}\hypertarget{rev1:1}
   For illustration, we consider an extremely simplified model of oscillating gene transcription,
@@
 If the input is an impulse at time zero (a delta function), 
-then the phenotype is equal to the impulse response: \revpoint{AE}{5}
+then the phenotype is equal to the impulse response: %\revpoint{AE}{5}
 \begin{align*}
@@
 changing $A$ corresponds to changing the strengths of regulatory interactions.
-We will even see below that interactions may change sign. \revpoint{1}{13}
+We will even see below that interactions may change sign. %\revpoint{1}{13}
 However, not all phenotypically equivalent systems are of this form:
@@
 and so can only infer motion in what is known as the \emph{observable subspace}.
-The Kalman decomposition then classifies each direction in kryptotype space
+The Kalman decomposition then classifies each direction in kryptotype space\hypertarget{rev2:3}
 as either reachable or unreachable, and as either observable or unobservable.
@@
 and so, for instance, $A_{\rno}$ is the $n_\rno \times n_\rno$ square matrix 
-in the top left corner of $P A P^{-1}$. \revpoint{1}{15}
+in the top left corner of $P A P^{-1}$. %\revpoint{1}{15}
 The impulse response of the system is given by
@@
 It is remarkable that the gene regulatory network architecture to achieve a given input--output map is never unique --
-both the change of basis used to obtain the decomposition
+both the change of basis used to obtain the decomposition\hypertarget{rev2:4}
 and, once in this form, all submatrices other than $A_{\ro}$, $B_{\ro}$, and $C_{\ro}$ can be changed without affecting the phenotype,
@@
 is defined to be the closure of $\spn(B)$ under applying $A$
-(or equivalently, the span of $B, AB, A^2B, \ldots A^{n-1}B$), \revpoint{AE}{6}
+(or equivalently, the span of $B, AB, A^2B, \ldots A^{n-1}B$), %\revpoint{AE}{6}
 % Analogously to this, we define
@@
     \end{align*}
-  The resulting set of systems are depicted in Figure \ref{fig:all_osc}. \revpoint{1}{16}
+  The resulting set of systems are depicted in Figure \ref{fig:all_osc}. %\revpoint{1}{16}
 \end{example}
@@
 
-\revpoint{AE}{7}
+%\revpoint{AE}{7}
 
@@
 Suppose that each of a diploid organisms' two genomes encodes a set of system coefficients
-with the same kryptotype dimension. \revpoint{2}{1}
+with the same kryptotype dimension. %\revpoint{2}{1}
 We assume that a diploid which has inherited systems $(A', B', C')$ and $(A'', B'', C'')$ from its two parents
@@
 using a weighted difference between impulse response functions.
-Suppose that $\rho(t)$ is a nonnegative weighting function, \revpoint{2}{3}
+Suppose that $\rho(t)$ is a nonnegative weighting function, %\revpoint{2}{3}
 $h_0(t)$ is the \emph{optimal} impulse response function
@@
 so that fitness is determined by the dynamics of the system over a few multiples of $2 \pi$,
-but not longer. \revpoint{AE}{8}
+but not longer. %\revpoint{AE}{8}
 Consider reproduction between a parent with system $(A, B, C)$ 
@@
 A Taylor expansion of $D(h_\epsilon)$ in $\epsilon$ 
-is explicitly worked out in Appendix \ref{apx:H_calc}, and \revpoint{2}{4}
+is explicitly worked out in Appendix \ref{apx:H_calc}, and %\revpoint{2}{4}
 shows that the phenotype of an $F_1$ hybrid between these two is at distance proportional to $\epsilon^2$ from optimal,
@@
 
-\paragraph{Haldane's rule}
+\paragraph{Haldane's rule}\hypertarget{rev1:5}
 This model naturally predicts
@@
 
-Consistent with empirical observation of hybrid breakdown (e.g., \citet{plotner2017chlorosis}), 
+Consistent with empirical observation of hybrid breakdown (e.g., \citet{plotner2017chlorosis}),\hypertarget{rev1:6} 
 we see that the fitnesses of $F_2$ hybrids drop at a much faster rate than those of $F_1$s.
@@
 % In light of the few known incompatibilities, 
-This has led some to conclude that hybrid incompatibility is typically a byproduct of positive selection \citep{orr2004speciation, schluter2009evidence} 
+This has led some to conclude that hybrid incompatibility is typically a byproduct of positive selection \citep{orr2004speciation, schluter2009evidence}\hypertarget{rev1:2} 
 or a consequence of genetic conflict \citep{presgraves2010molecular, crespi2013conflictual},
@@
 while the fitness consequences of incompatibility in any one given network may be small, 
-the cumulative impact of system drift across the many different networks an organism relies on may be substantial. 
+the cumulative impact of system drift across the many different networks an organism relies on may be substantial.\hypertarget{rev1:3} 
 It remains to be seen how the relative strengths of these forces compare.
@@
 
-\paragraph{The dimensionality of trait space} \revpoint{1}{8}
+\paragraph{The dimensionality of trait space} %\revpoint{1}{8}
 We have focused on examples of single traits (where the phenotype is one-dimensional),
@@
 
-Phenotypically equivalent system evolution is probably not only driven by neutral genetic drift.
-\revpoint{1}{18} \revpoint{1}{19} \revpoint{1}{21}
+Phenotypically equivalent system evolution is probably not only driven by neutral genetic drift.\hypertarget{rev1:4}
+%\revpoint{1}{18} %\revpoint{1}{19} %\revpoint{1}{21}
 For one thing, movement along the set of equivalent networks
@@
 but the fact that most nonlinear systems can be locally approximated by a linear one
-suggests our qualitative results may hold more generally. \revpoint{1}{7}
+suggests our qualitative results may hold more generally. %\revpoint{1}{7}
 Furthermore, nonidentifiability (which implies the existence of neutral directions)

### 2021-07-07 review-responses2.tex
more responses

@@
 
-\reply{ 
+\reply{This is an important point. We have added references to the experimental speciation literature \revref. 
 }
@@
 
-\reply{
+\reply{ \jss{I think this might be a question you will be better at answering?}
 }
@@
 
-\reply{
+\reply{This is also a very interesting point worth discussing. We have added discussion on fluctuating selection \revref.
+\jss{Not sure where to put Mani and Clarke yet}.
 }
@@
 
-\reply{
+\reply{Removed the citation. 
 }
@@
 \end{point}
+
 \reply{
 }
+
 \begin{point}{}
@@
 
-\reply{
+\reply{ Fixed \revref. \jss{Although I like the original phrasing better...}
 }
@@
 
-\reply{
+\reply{\jss{No?}
 }

### 2021-07-07 kalman_walk_paper.tex
more responses

@@
 The field of population genetics has also explored the consequences of the fact
-that there is often more than one way to do the same thing,\hypertarget{rev2:2}e
+that phenotype does not prescribe a unique genotype\revpoint{2}{3}
+%there is often more than one way to do the same thing,
+\hypertarget{rev2:2}
 and observed that speciation might be the result of changes that are themselves neutral.
@@
 
-Consistent with empirical observation of hybrid breakdown (e.g., \citet{plotner2017chlorosis}),\hypertarget{rev1:6} 
+Consistent with empirical observation of hybrid breakdown %(e.g., \citet{plotner2017chlorosis}),\hypertarget{rev1:6} 
 we see that the fitnesses of $F_2$ hybrids drop at a much faster rate than those of $F_1$s.
@@
 two processes that typically act much more rapidly than genetic drift.
+Experiments in which population bottlenecks were induced to study the effect of drift on speciation found no reproductive isolation in the majority tested pairs. 
+In contrast, experiment has provided stronger support for incompatibilities evolving as a byproduct of divergent selection\citep{fry2009laboratory, white2020past}, 
+and incompatibilities identified between \emph{Drosophila simulans} and \emph{D. melanogaster} exhibit evidence of adaptative evolution \citep{tang2009evolution}.\revpoint{1}{1} 
 % In contrast, the model we develop here suggests that even under strictly neutral conditions, 
@@
 % (as diversification is not observed to be accelerated following mass extinctions).
+More recently, \citet{sanchez2021widespread} observed widespread transcriptional misregulation in \emph{Drosophila} hybrids, implicating the action of system drift.
 
@@
 
-Phenotypically equivalent system evolution is probably not only driven by neutral genetic drift.\hypertarget{rev1:4}
+Phenotypically equivalent system evolution is probably not only driven by neutral genetic drift. \hypertarget{rev1:4}
 %\revpoint{1}{18} %\revpoint{1}{19} %\revpoint{1}{21}
@@
 possibly due to epistasis, genetic architecture \citep{hermisson2003epistasis}, adaptive inertia \citep{baatz1997adaptive,alvarez2009effects} or weak gene flow could plausibly slow it down.
+Fluctuating selection \citep{bell2010fluctuating} could further accelerate speciation via system drift. 
+Under our model, populations evolving towards different optimal phenotypes will always produce less fit hybrids,
+as the hybrid phenotypes will be displaced from both parental optimums, suggesting that selection fluctuation -- if not synchronized between populations -- will diminish hybrid fitnesses.
+In the case where fluctuations are synchronized, due to the size of the neutral system space, 
+populations may arive at different system organizations when returning to previous phenotypes at a rate faster than by drift alone.\revpoint{1}{3}
 More work on specific systems,

### 2021-07-07 review-responses2.tex
a few more rough responses and notes

@@
 \reply{
+  \jss{This is probably a question you can answer better, but I put some thoughts down to start.} This is correct.
+As the model is currently described, we show that since different mutationally connected systems are not usually reproductively compatible, populations in different parts of system space will produce unfit hybrids.
+We rely on results from quantitative genetics to estimate the rate at which systems drift through the neutral space and thus to estimate the effects of gene flow.
+Without further modeling we do not yet know how a quantitative genetics model incorporating system theory would differ, if it all, from these classic results.
+We are planning to submit a paper where we further develop the removed quantitative genetics model and may be able to address this interesting question there.
 }
@@
 
-\reply{
+\reply{\jss{I think Orr's estimate of the number of incompatible sites should hold?
+Maybe we should clarify that what we calculate is the rate at which phenotypes breakdown/fitness decays, and not the number of incompatible genes?
+I'm not exactly sure how to approach this question.
+This seems related to one of our previous discussions about whether or not this model produces ``DMIs'' or something rather different.}
 }
@@
 
-\reply{
+\reply{This is an interesting question.
+In our model system coefficients are treated as independent heritable units, and thus would represent genes spread out sufficiently far apart along the genome.
+Since the major source of F2 hybrid breakdown in this model is from recombining regulatory coefficients, if the genetic architecture of these coefficients was such that the genes were close enough to not recombine, we would not see the stark pattern of hybrid breakdown that we observe in F2 oscillators.
+We suspect that genetic architectures that allow for recombination, albeit at a lower rate, than for architectures in which genes are very far apart, will diminish the speed of speciation.
+This also leads to interesting questions about the evolution of genetic architecture; perhaps arhitectures in which co-regulating genes are neighbors are more robust.
+\jss{Need to think of where to add discussion like this to the paper.}
 }
@@
 
-\reply{ \jss{I think this might be a question you will be better at answering?}
+\reply{ \jss{I think this might be a question you will be better at answering? FYI, we already cite Rosas 2010 and Fraisse 2016 on page 2.}
 }
@@
 
-\reply{
+\reply{\jss{To do.
+I think our options are to either add an appendix where we show that this model does or does not generate ``dominance relations'', add a sentence explaining why it does or does not, or just say it is consistent with Muller's ``balance argument'' (Barton 2001) and punt on whether or not it is or is not consistent with the dominance theory.
+What do you think?
+
+There also might be a nice solution when discussing the Martin 2014 paper about how systems biology and quantitative genetics models approximate one another (as R1 mentions above).
+Some concepts might be lost in translation.
+}
 }
@@
 \begin{point}{}
-  5. (very minor) \hyperlink{rev1:6}{Line 358}: As you say, F2 hybrid breakdown has been known about for a very long time. As such, the single citation used feels a bit arbitrary.
+  (very minor) \hyperlink{rev1:6}{Line 358}: As you say, F2 hybrid breakdown has been known about for a very long time. As such, the single citation used feels a bit arbitrary.
 \end{point}
@@
 
-\reply{
+\reply{\jss{What do you think?}
 }
@@
 
-\reply{
+\reply{\jss{I can add this soon}
 }
@@
 
-\reply{
+\reply{\jss{To do.}
 }

### 2021-07-21 review-responses2.tex
some  response edits

@@
 \reply{
-  \jss{This is probably a question you can answer better, but I put some thoughts down to start.} This is correct.
+This is correct.
 As the model is currently described, we show that since different mutationally connected systems are not usually reproductively compatible, populations in different parts of system space will produce unfit hybrids.
@@
 We are planning to submit a paper where we further develop the removed quantitative genetics model and may be able to address this interesting question there.
+
+We also briefly mention that gene flow will likely complicate matters in the discussion \revref.
 }
@@
 This also leads to interesting questions about the evolution of genetic architecture; perhaps arhitectures in which co-regulating genes are neighbors are more robust.
-\jss{Need to think of where to add discussion like this to the paper.}
+
+We briefly mention complications to the model in the discussion \revref.
 }
@@
 
-\reply{ \jss{I think this might be a question you will be better at answering? FYI, we already cite Rosas 2010 and Fraisse 2016 on page 2.}
+\reply{In the last paragraph of the introduction we discuss exactly this and the corresponding literature, including Rosas et al 2010 and Fraisse et al 2016.
+\jss{I don't really think Mani and Clarke says what R1 says it does.
+It states that mutational order matters.
+Also I don't thik the introduction is a good place for Martin 2014 -- maybe system and quant gen models have some sort of correspondence, but we don't show it.}
 }
@@
 
-\reply{\jss{To do.
-I think our options are to either add an appendix where we show that this model does or does not generate ``dominance relations'', add a sentence explaining why it does or does not, or just say it is consistent with Muller's ``balance argument'' (Barton 2001) and punt on whether or not it is or is not consistent with the dominance theory.
-What do you think?
+\reply{
+  The connection between the ``dominance theory'' and the situation presented here does not seem that straight forward to us.
+Haldane's rule should occur here even if the alleles are all strictly additive.
+Depending on the specific genetic architecture of a system, the dominance coefficient may not even be defined.
+For instance, if we consider a mutation that changes the $\tau$ parameter in the oscillator, $h$ as typically formulated in undefined.
+If we consider mutation to the particular coefficients themselves, $h$ varies sigmoidally from recessive to dominant, depending on the magnitude of the change (and the fitness function).
+Without an explicit model of mutation it is not clear whether these changes are dominant or not.
 
-There also might be a nice solution when discussing the Martin 2014 paper about how systems biology and quantitative genetics models approximate one another (as R1 mentions above).
-Some concepts might be lost in translation.
-}
+In this model we are also thinking about the effect of many small/quantitative mutations -- the effect of which on fitness, the cumulative effect of which could either be ``dominant'' or ``recessive''.
+
+Figure 5 in Barton 2001 shows that F2 hybrid breakdown under \emph{additive} inheritence.
+Our model links F2 hybrid breakdown with heterogametic inviability -- in our model F1 heterogametes have lower fitness than homogametes because their system looks like an autosomal F2 system.
 }
@@
 
-\reply{\jss{What do you think?}
+\reply{Thank you for this suggestion. We've decided to keep the paragraph order the same though. 
 }
@@
 
-\reply{ Fixed \revref. \jss{Although I like the original phrasing better...}
+\reply{Thank you for the suggestion, but we would prefer to keep the sentence as written. 
+We understand your concern, but in order to make this paper easy to read,
+we have avoided using acronyms and tried to use simple language wherever possible,
+so long as the meaning remains clear.
 }
@@
 
-\reply{\jss{No?}
+\reply{This is a good question.
+The unreachable subspace would correspond to system structures that do not have an effect on the phenotype.
+However, biologically, it could be understood as a form of cryptic genetic variation -- coefficients in the unreachable subspace could have an effect when a system is perturbed (e.g.
+mutationally or in a hybrid).
+\jss{Not sure if/where to add to the paper?}
 }
@@
 \reply{
-  As far as we understand, and as applied here, the Kalman decomposition is for linear time-invariant systems.
-Generalizations of the Kalman decomposition for time-varying systems have been discussed in...
-  \jss{Could cite some papers? Looks like Silverman and Meadows, EQUIVALENT REALIZATIONS OF LINEAR SYSTEMS discuss generalizations to the time-variable case.}
+  Yes, the Kalman decomposition is for linear time-invariant systems. We now note this \revref{}. 
 }
@@
 
-\reply{\jss{I can add this soon}
+\reply{We do not think we can add an example of how recombination generally works -- the mechanism will depend on the details of the particular system under study.
+In the oscillator example, recombination independently draws system coefficients from each parent at random.
 }

### 2021-07-21 kalman_walk_paper.tex
some  response edits

--- 2021-07-06 kalman_walk_paper.tex
@@
 The field of population genetics has also explored the consequences of the fact
-that there is often more than one way to do the same thing,\hypertarget{rev2:2}e
+%that phenotype does not prescribe a unique genotype\revpoint{2}{3}
+that there is often more than one way to do the same thing,
+\hypertarget{rev2:2}
 and observed that speciation might be the result of changes that are themselves neutral.
@@
 to achieve the desired dynamics.
-Results in system theory show that any system can be realized in a particular minimal dimension
+Results in system theory show that any time-invariant\revpoint{2}{5} system can be realized in a particular minimal dimension
 (the dimension of the kryptotype, $n_\text{min}$),
@@
 $F_1$ males are effectively equivalent to purely autosomal-system $F_2$ hybrids, and recall that $F_2$s are significantly less fit on average than $F_1$s (see Figure \ref{fig:osc_incompat}).
+
 Although many alleles will be dominant if the phenotype--fitness relationship is convex,
@@
 In their model, if alleles are strictly additive, that is neither dominant nor recessive, then male and female $F_1$ hybrids will drop in fitness at the exact same rates, and therefore defy Haldane's rule.
+
+A key difference from the dominance theory is that this model treats system coefficients as quantitative traits.
+The classification of alleles as either ``dominant'' or ``recessive'' fits most naturally for Mendellian traits.
+Along these lines, if we think of individual loci of large effect then the effect of mutations on fitness (if convex) will usually be recessive, congruent with the dominance theory's predictions.
+However, the dominance or recessiveness of even large effect mutations will depend on both the magnitude of their effect and precisely how they adjust system coefficients.
+For such mutations that keep systems on the neutral manifold (imagine mutations to the oscillator that only move $\tau$), the dominance coefficient (usually notated as $h$) will be mathematically undefined.
+Mutations that driectly perturb coefficiens of $A$ can be either dominant or recessive, depending on their size.
+
+\citet{barton2001role} explored Haldane's rule in the context of a quantitative trait and concluded that it was a consequence of the dominance theory.
+However, Barton also demonstrated $F_2$ hybrid breakdown in a model with strictly \emph{additive} inheritence.
+The model suggested here explains Haldane's rule as a special case of hybrid breakdown.
+As hybrid breakdown can occur for a quantitative trait under additivity, so can heterogametic inviability.
+This suggests that the model presented here is distinct from the dominance theory.
 
@@
 two processes that typically act much more rapidly than genetic drift.
+Experimental approaches designed to study the effect of drift on speciation which induced population bottlenecks found no reproductive isolation in the majority tested pairs\citep{fry2009laboratory, white2020past}.
+These bottlenecks might not maintain enough genetic variation for systems to drift at an appreciable rate, however.\revpoint{1}{1} \jss{Did you also want to add something about bottlenecking large versus small popoulations?}
+%and incompatibilities identified between \emph{Drosophila simulans} and \emph{D. melanogaster} exhibit evidence of adaptative evolution \citep{tang2009evolution}.\revpoint{1}{1}
+%
 % In contrast, the model we develop here suggests that even under strictly neutral conditions, 
@@
 % (as diversification is not observed to be accelerated following mass extinctions).
+More recently, \citet{sanchez2021widespread} observed widespread transcriptional misregulation in \emph{Drosophila} hybrids, implicating the action of system drift.
+
+Fluctuating selection \citep{bell2010fluctuating, barton2001role} could further accelerate speciation via system drift. 
+Under our model, populations evolving towards different optimal phenotypes will always produce less fit hybrids,
+as the hybrid phenotypes will be displaced from both parental optimums -- that is maladapted to both parental environments.
+This suggests that fluctuating selection -- if not synchronized between populations -- should diminish hybrid fitnesses.
+In the case where fluctuations are synchronized, due to the size of the neutral system space, 
+populations may arive at different system organizations when returning to previous phenotypes at a rate faster than by drift alone.\revpoint{1}{3}
 
 These explanations are not mutually exclusive.
-All of these forces -- adaptive shifts, conflict, and network drift --
+All of these forces -- adaptive shifts, conflict and network drift --
 are plausible drivers of speciation, and may even interact.
@@
 any constraints on such variation,
-possibly due to epistasis, genetic architecture \citep{hermisson2003epistasis}, adaptive inertia \citep{baatz1997adaptive,alvarez2009effects} or weak gene flow could plausibly slow it down.
+possibly due to epistasis, genetic architecture \citep{hermisson2003epistasis}, adaptive inertia \citep{baatz1997adaptive,alvarez2009effects} or weak gene flow could plausibly slow it down.\revpoint{AE}{1}\revpoint{AE}{3}
 More work on specific systems,

### 2021-07-22 review-responses2.tex
some more response edits

@@
 
-\reply{\jss{I think Orr's estimate of the number of incompatible sites should hold?
-Maybe we should clarify that what we calculate is the rate at which phenotypes breakdown/fitness decays, and not the number of incompatible genes?
-I'm not exactly sure how to approach this question.
-This seems related to one of our previous discussions about whether or not this model produces ``DMIs'' or something rather different.}
-}
+\reply{\jss{quantitative model....}}
 
@@
 However, there will also be F2 individuals with genes from both parental types -- meaning that there will be more low fitness F2s than low fitness F1s and F0s.
-\jss{(The F2 fitness is lower on average?)}
 }

### 2021-07-22 kalman_walk_paper.tex
some more response edits

@@
 This model also naturally predicts Haldane's rule, 
-thus providing another possible explanation
+%thus providing another possible explanation
+thus providing a concrete explanation
 of why heterogametic hybrids tend to be disrupted more often than homogametes 
@@
 
-Although many alleles will be dominant if the phenotype--fitness relationship is convex,
-the underlying mechanism does not depend on the \emph{dominance theory} \citep{turelli1995dominance} to explain Haldane's rule:
-instead, it derives from the nature of segregation variance.
+%Although many alleles will be dominant if the phenotype--fitness relationship is convex,
+%the underlying mechanism does not depend on the \emph{dominance theory} \citep{turelli1995dominance} to explain Haldane's rule:
+%instead, it derives from the nature of segregation variance.
+%
 % In this setting, Haldane's rule is mechanistically similar to \emph{hybrid breakdown}, 
@@
 
-A key difference from the dominance theory is that this model treats system coefficients as quantitative traits.
-The classification of alleles as either ``dominant'' or ``recessive'' fits most naturally for Mendellian traits.
-Along these lines, if we think of individual loci of large effect then the effect of mutations on fitness (if convex) will usually be recessive, congruent with the dominance theory's predictions.
+This model differs from the \emph{dominance theory} \citep{turelli1995dominance} in that it treats system coefficients as quantitative traits.
+%The classification of alleles as either ``dominant'' or ``recessive'' fits most naturally for Mendellian traits. 
+If we think of individual loci of large effect then the effect of mutations on fitness (if convex) will usually be recessive, congruent with the dominance theory.
 However, the dominance or recessiveness of even large effect mutations will depend on both the magnitude of their effect and precisely how they adjust system coefficients.
-For such mutations that keep systems on the neutral manifold (imagine mutations to the oscillator that only move $\tau$), the dominance coefficient (usually notated as $h$) will be mathematically undefined.
+%For such mutations that keep systems on the neutral manifold (imagine mutations to the oscillator that only move $\tau$), the dominance coefficient (usually notated as $h$) will be mathematically undefined.
 Mutations that driectly perturb coefficiens of $A$ can be either dominant or recessive, depending on their size.
@@
 \citet{barton2001role} explored Haldane's rule in the context of a quantitative trait and concluded that it was a consequence of the dominance theory.
-However, Barton also demonstrated $F_2$ hybrid breakdown in a model with strictly \emph{additive} inheritence.
+However, Barton also demonstrated $F_2$ hybrid breakdown in a model with strictly \emph{additive} inheritence -- a consequence of segregation variance.
 The model suggested here explains Haldane's rule as a special case of hybrid breakdown.
 As hybrid breakdown can occur for a quantitative trait under additivity, so can heterogametic inviability.
-This suggests that the model presented here is distinct from the dominance theory.
+This suggests that the model presented here is distinct from the dominance theory; like hybrid breakdown, Haldane's rule derives from the nature of segregation variance.
 
@@
 Is there evidence that this is actually occurring?
-System drift and network rewiring has been inferred across the tree of life \citep{wotton2015quantitative, crombach2016gap, dalal2017transcription, johnson2017rewiring, ali2017quantitative},
+System drift and network rewiring has been inferred across the tree of life \citep{wotton2015quantitative, crombach2016gap, dalal2017transcription, johnson2017rewiring, ali2019novel},
 and there is often significant regulatory variation segregating within populations.

### 2021-08-10 review-responses2.tex
edits
.

@@
 
-\reply{\jss{quantitative model....}}
+\reply{
+    Good question:
+    on face value, the similarity to the snowball depends entirely on the relationship between
+    ``distance to optimum'' and fitness: since distance to optimum increases linearly in F2s,
+    if this relation is quadratic then our model will also show a quadratic increase in incompatibility.
+    However, this seems somewhat arbitrary.
+    It is more interesting to consider if our model displays the same underlying behavior that leads to the snowball,
+    in which each pair of lineage-specific substitutions contributes to incompatibility.
+    This may be an empirical question (that could be examined in simulation),
+    but at this level of analysis it seems the answer is ``no'':
+    Consider two lineages moving away from each other along the neutral manifold.
+    Each substitution moves them further apart, but the rate of increase of the distance between them
+    is linear, and each subsequent substitution does not increase their rate of separation.
+    We considered adding this point to the paper,
+    but decided that it is sufficiently confusing that it would not add much.
+}
 
@@
 %
-  \hyperlink{rev1:2}{Lines 382-389} discuss previous discussion about the role of drift in speciation, but they focus on theory.
+    \revref~discusses previous discussion about the role of drift in speciation, but they focus on theory.
 I think that some strong arguments for the primacy of selection in speciation have been empirical (bottlenecking Drosophila populations and testing for RI, MK-tests on ``speciation genes'' etc.).
 Some of the earlier work was summarized in the last chapter of Coyne and Orr 2004.
-\hyperlink{rev1:3}{Line 396} seems to ignore a lot of this work, and so I would rephrase.
+\llname{rp11}~seems to ignore a lot of this work, and so I would rephrase.
 \end{point}
@@
 
-\reply{In the last paragraph of the introduction we discuss exactly this and the corresponding literature, including Rosas et al 2010 and Fraisse et al 2016.
+\reply{
+    In the last paragraph of the introduction we discuss exactly this and the corresponding literature, including Rosas et al 2010 and Fraisse et al 2016. \revref
 \jss{I don't really think Mani and Clarke says what R1 says it does.
@@
 
-\reply{Thank you for this suggestion. We've decided to keep the paragraph order the same though. 
+\reply{Thank you for this suggestion. We've decided to keep the paragraph order the same though.\jss{Having second thoughts on this... rearranging could be clearer.}
 }

### 2021-08-10 kalman_walk_paper.tex
.

@@
 For instance, there are ``neutral networks''
-of nucleic acid sequences that produce the same RNA secondary structure \citep{gruner1996analysis}%\revpoint{1}{5},
+of nucleic acid sequences that produce the same RNA secondary structure \citep{gruner1996analysis}
 amino acid sequences that fold similarly \citep{babajide1997neutral},
@@
 and neural circuits \citep{trojanowski2014neural}, and in developmental systems \citep{vondassow2000segment,true2001developmental}. 
-%\revpoint{1}{1} %\revpoint{1}{3}
 % \jss{Other ideas instead of ``degeneracy'' or ``isofunctional structural plasticity'': non-invertibility, isofunctional plasticity, structural plasticity, phenotypic equivalency, phenotypic invariance, functional invariance, isofunctionality, functional equivalency, genotypic plasticity, conjugacy, functional symmetry, isologous, isomorphic.}
@@
 For instance, in many contexts mathematical models
-are fundamentally \emph{nonidentifiable} and/or \emph{indistinguishable} -- %\revpoint{1}{2}
+are fundamentally \emph{nonidentifiable} and/or \emph{indistinguishable} -- 
 meaning that there can be uncertainty about an inferred model's parameters or even its claims about
@@
 a population diffuses along the neutral ridges of a high-dimensional space of possible system parameters,
-in a similar vein as \emph{holey landscape} models \citep{gavrilets1997evolution, yamaguchi2013first, pina2019does}.  %\revpoint{2}{12}
+in a similar vein as \emph{holey landscape} models \citep{gavrilets1997evolution, yamaguchi2013first, pina2019does}. 
 % Our model is developed in more detail in a companion paper \citep{schiffman2020genetic}.
@@
 The field of population genetics has also explored the consequences of the fact
-%that phenotype does not prescribe a unique genotype\revpoint{2}{3}
+%that phenotype does not prescribe a unique genotype
 that there is often more than one way to do the same thing,
@@
 Indeed, \citet{bateson1909heredity} first proposed that what today we call a
-Bateson-Dobzhansky-Muller incompatibility would likely arise through neutral changes. %\revpoint{1}{10}
+Bateson-Dobzhansky-Muller incompatibility would likely arise through neutral changes. 
 The potential for speciation has been analyzed in models of 
@@
 It has also been shown that population structure
-can allow long-term stable coexistence of incompatible genotypes encoding identical phenotypes \citep{phillips1996maintenance}. %\revpoint{1}{9}
+can allow long-term stable coexistence of incompatible genotypes encoding identical phenotypes \citep{phillips1996maintenance}. 
 However, previous simulations of system drift in regulatory sequences \citep{tulchinsky2014hybrid}
@@
 segregation variance between isolated populations is fairly well understood
-\citep{slatkin1994segregation,rosas2010cryptic,chevin2014niche},
+\citep{slatkin1994segregation,rosas2010cryptic,chevin2014niche}, \revpoint{1}{2}
 but model assumptions can strongly affect predictions,
@@
 Under quite general assumptions on the input
-(e.g., $|u(t)|$ is integrable) %\revpoint{AE}{3}
-we can write the phenotype as %\revpoint{AE}{4}
+(e.g., $|u(t)|$ is integrable) 
+we can write the phenotype as 
   \begin{align}
@@
 
-\citet{wagner1994evolution} and others have used a similar discrete-time model %\revpoint{1}{6}
+\citet{wagner1994evolution} and others have used a similar discrete-time model 
 (that might be written $\phi_{t+1} = f(A \phi_t)$, where $f$ is a sigmoid).
@@
 If the input is an impulse at time zero (a delta function), 
-then the phenotype is equal to the impulse response: %\revpoint{AE}{5}
+then the phenotype is equal to the impulse response: 
 \begin{align*}
@@
 changing $A$ corresponds to changing the strengths of regulatory interactions.
-We will even see below that interactions may change sign. %\revpoint{1}{13}
+We will even see below that interactions may change sign. 
 However, not all phenotypically equivalent systems are of this form:
@@
 to achieve the desired dynamics.
-Results in system theory show that any time-invariant\revpoint{2}{5} system can be realized in a particular minimal dimension
+Results in system theory show that any time-invariant \revpoint{2}{5}
+system can be realized in a particular minimal dimension
 (the dimension of the kryptotype, $n_\text{min}$),
@@
 and so, for instance, $A_{\rno}$ is the $n_\rno \times n_\rno$ square matrix 
-in the top left corner of $P A P^{-1}$. %\revpoint{1}{15}
+in the top left corner of $P A P^{-1}$. 
 The impulse response of the system is given by
@@
 is defined to be the closure of $\spn(B)$ under applying $A$
-(or equivalently, the span of $B, AB, A^2B, \ldots A^{n-1}B$), %\revpoint{AE}{6}
+(or equivalently, the span of $B, AB, A^2B, \ldots A^{n-1}B$), 
 % Analogously to this, we define
@@
     \end{align*}
-  The resulting set of systems are depicted in Figure \ref{fig:all_osc}. %\revpoint{1}{16}
+  The resulting set of systems are depicted in Figure \ref{fig:all_osc}. 
 \end{example}
@@
 
-%\revpoint{AE}{7}
-
 \begin{figure}[H]
@@
 Suppose that each of a diploid organisms' two genomes encodes a set of system coefficients
-with the same kryptotype dimension. %\revpoint{2}{1}
+with the same kryptotype dimension. 
 We assume that a diploid which has inherited systems $(A', B', C')$ and $(A'', B'', C'')$ from its two parents
@@
 using a weighted difference between impulse response functions.
-Suppose that $\rho(t)$ is a nonnegative weighting function, %\revpoint{2}{3}
+Suppose that $\rho(t)$ is a nonnegative weighting function, 
 $h_0(t)$ is the \emph{optimal} impulse response function
@@
 so that fitness is determined by the dynamics of the system over a few multiples of $2 \pi$,
-but not longer. %\revpoint{AE}{8}
+but not longer. 
 Consider reproduction between a parent with system $(A, B, C)$ 
@@
 A Taylor expansion of $D(h_\epsilon)$ in $\epsilon$ 
-is explicitly worked out in Appendix \ref{apx:H_calc}, and %\revpoint{2}{4}
+is explicitly worked out in Appendix \ref{apx:H_calc}, and 
 shows that the phenotype of an $F_1$ hybrid between these two is at distance proportional to $\epsilon^2$ from optimal,
@@
 % In light of the few known incompatibilities, 
-This has led some to conclude that hybrid incompatibility is typically a byproduct of positive selection \citep{orr2004speciation, schluter2009evidence}\hypertarget{rev1:2} 
+This has led some to conclude that hybrid incompatibility is typically a byproduct of positive selection \citep{orr2004speciation, schluter2009evidence}
 or a consequence of genetic conflict \citep{presgraves2010molecular, crespi2013conflictual},
@@
 Experimental approaches designed to study the effect of drift on speciation which induced population bottlenecks found no reproductive isolation in the majority tested pairs\citep{fry2009laboratory, white2020past}.
-These bottlenecks might not maintain enough genetic variation for systems to drift at an appreciable rate, however.\revpoint{1}{1} \jss{Did you also want to add something about bottlenecking large versus small popoulations?}
-%and incompatibilities identified between \emph{Drosophila simulans} and \emph{D. melanogaster} exhibit evidence of adaptative evolution \citep{tang2009evolution}.\revpoint{1}{1}
+These bottlenecks might not maintain enough genetic variation for systems to drift at an appreciable rate, however.
+\revpoint{1}{1}
+\jss{Did you also want to add something about bottlenecking large versus small popoulations?}
+%and incompatibilities identified between \emph{Drosophila simulans} and \emph{D. melanogaster} exhibit evidence of adaptative evolution \citep{tang2009evolution}.
 %
@@
 In the case where fluctuations are synchronized, due to the size of the neutral system space, 
-populations may arive at different system organizations when returning to previous phenotypes at a rate faster than by drift alone.\revpoint{1}{3}
+populations may arive at different system organizations when returning to previous phenotypes at a rate faster than by drift alone. \revpoint{1}{3}
 
@@
 while the fitness consequences of incompatibility in any one given network may be small, 
-the cumulative impact of system drift across the many different networks an organism relies on may be substantial.\hypertarget{rev1:3} 
+the cumulative impact of system drift across the many different networks an organism relies on may be substantial. \linelabel{rp11}
 It remains to be seen how the relative strengths of these forces compare.
@@
 
-\paragraph{The dimensionality of trait space} %\revpoint{1}{8}
+\paragraph{The dimensionality of trait space} 
 We have focused on examples of single traits (where the phenotype is one-dimensional),
@@
 Phenotypically equivalent system evolution is probably not only driven by neutral genetic drift. \hypertarget{rev1:4}
-%\revpoint{1}{18} %\revpoint{1}{19} %\revpoint{1}{21}
 For one thing, movement along the set of equivalent networks
@@
 any constraints on such variation,
-possibly due to epistasis, genetic architecture \citep{hermisson2003epistasis}, adaptive inertia \citep{baatz1997adaptive,alvarez2009effects} or weak gene flow could plausibly slow it down.\revpoint{AE}{1}\revpoint{AE}{3}
+possibly due to epistasis, genetic architecture \citep{hermisson2003epistasis}, adaptive inertia \citep{baatz1997adaptive,alvarez2009effects} or weak gene flow could plausibly slow it down.
+\revpoint{AE}{1}\revpoint{AE}{3}
 More work on specific systems,
@@
 but the fact that most nonlinear systems can be locally approximated by a linear one
-suggests our qualitative results may hold more generally. %\revpoint{1}{7}
+suggests our qualitative results may hold more generally. 
 Furthermore, nonidentifiability (which implies the existence of neutral directions)

### 2021-08-10 review-responses2.tex
edits together

@@
     It is more interesting to consider if our model displays the same underlying behavior that leads to the snowball,
-    in which each pair of lineage-specific substitutions contributes to incompatibility.
+    in which each pair of lineage-specific substitutions possibly contributes to incompatibility.
     This may be an empirical question (that could be examined in simulation),
-    but at this level of analysis it seems the answer is ``no'':
+    but at the level of analysis in our paper it seems the answer is ``no'':
     Consider two lineages moving away from each other along the neutral manifold.
@@
 \reply{
-    In the last paragraph of the introduction we discuss exactly this and the corresponding literature, including Rosas et al 2010 and Fraisse et al 2016. \revref
-\jss{I don't really think Mani and Clarke says what R1 says it does.
-It states that mutational order matters.
-Also I don't thik the introduction is a good place for Martin 2014 -- maybe system and quant gen models have some sort of correspondence, but we don't show it.}
+    We agree that this is an important thing to discuss.
+    We think that this is fairly well covered in the Introduction (paragraph at \revref),
+    and cite \citet{rosas2010cryptic} and \citet{fraisse2016genetics} in that section --
+    we have tried to make the connection to previous literature complete without being overwhelming.
+    We have also added a reference to \citet{mani1990mutational} at \llname{citemani},
+    and added further reference to Fisher's geometric model in the Discussion \llname{quantnote}.
 }
@@
 \reply{This is also a very interesting point worth discussing. We have added discussion on fluctuating selection \revref.
-\jss{Not sure where to put Mani and Clarke yet}.
 }
@@
 \reply{
-  The connection between the ``dominance theory'' and the situation presented here does not seem that straight forward to us.
-Haldane's rule should occur here even if the alleles are all strictly additive.
-Depending on the specific genetic architecture of a system, the dominance coefficient may not even be defined.
-For instance, if we consider a mutation that changes the $\tau$ parameter in the oscillator, $h$ as typically formulated in undefined.
-If we consider mutation to the particular coefficients themselves, $h$ varies sigmoidally from recessive to dominant, depending on the magnitude of the change (and the fitness function).
-Without an explicit model of mutation it is not clear whether these changes are dominant or not.
-
-In this model we are also thinking about the effect of many small/quantitative mutations -- the effect of which on fitness, the cumulative effect of which could either be ``dominant'' or ``recessive''.
-
-Figure 5 in Barton 2001 shows that F2 hybrid breakdown under \emph{additive} inheritence.
-Our model links F2 hybrid breakdown with heterogametic inviability -- in our model F1 heterogametes have lower fitness than homogametes because their system looks like an autosomal F2 system.
+    Good point; on closer consideration we have decided to remove the reference to the dominance theory \revref,
+    and have modified the introduction to say ``a concrete explanation'' rather than ``another explanation''.
+    More generally, we have added a paragraph to the discussion about
+    the connection between our (quantitative and essentially continuous) model and
+    predictions based on the effects of single mutations on fitness. \llname{quantnote}
 }
@@
 
-\reply{This is a good question.
-The unreachable subspace would correspond to system structures that do not have an effect on the phenotype.
-However, biologically, it could be understood as a form of cryptic genetic variation -- coefficients in the unreachable subspace could have an effect when a system is perturbed (e.g.
-mutationally or in a hybrid).
-\jss{Not sure if/where to add to the paper?}
+\reply{
+    Good point; we've added some additional discussion of what the subspaces represent. \revref
 }
@@
 
-\reply{We do not think we can add an example of how recombination generally works -- the mechanism will depend on the details of the particular system under study.
-In the oscillator example, recombination independently draws system coefficients from each parent at random.
+\reply{
+    Good point; this was unclear. We have clarified that we don't pick a particular model for recombination in this seciton. \revref
 }

### 2021-08-10 kalman_walk_paper.tex
edits together

@@
 %that phenotype does not prescribe a unique genotype
-that there is often more than one way to do the same thing,
+that there is often more than one way to do the same thing, \revpoint{1}{2}
 \hypertarget{rev2:2}
@@
 segregation variance between isolated populations is fairly well understood
-\citep{slatkin1994segregation,rosas2010cryptic,chevin2014niche}, \revpoint{1}{2}
+\citep{slatkin1994segregation,rosas2010cryptic,chevin2014niche},
 but model assumptions can strongly affect predictions,
@@
 and so represent degrees of freedom.
-(However, some of these subspaces may affect how the system deals with noise.)
+
+Although the other submatrices do not affect the phenotype
+through the impulse response function $h(t)$,
+they may affect individual fitness in other ways. \revpoint{2}{4}
+For instance,
+the ``unreachable'' subspaces cannot be affected by any of the possible variations
+in input produced by varying $u(t)$.
+However, if the nature of the input changed
+(for instance, by perturbing a previously constant environmental variable),
+differences in the unreachable components might lead to differences in phenotypes
+and thus represent cryptic genetic variation.
+Similarly, ``unobservable'' subspaces might represent aspects of phenotype
+that are not constrained by selection,
+but might become important if conditions change.
+Additionally, all subspaces may affect how the system deals with noise.
 
@@
 $((A'+A'')/2, (B'+B'')/2, (C'+C'')/2)$.
+(We do not, however, specify a model for recombination to produce the inherited systems,
+as the details will depend on the genetic basis of the system coefficients;
+two extremes are depicted in the two panels of Figure~\ref{fig:conceptual_fig}.) \revpoint{2}{6}
 
@@
 $F_1$ males are effectively equivalent to purely autosomal-system $F_2$ hybrids, and recall that $F_2$s are significantly less fit on average than $F_1$s (see Figure \ref{fig:osc_incompat}).
-
-%Although many alleles will be dominant if the phenotype--fitness relationship is convex,
-%the underlying mechanism does not depend on the \emph{dominance theory} \citep{turelli1995dominance} to explain Haldane's rule:
-%instead, it derives from the nature of segregation variance.
-%
-% In this setting, Haldane's rule is mechanistically similar to \emph{hybrid breakdown}, 
-% the observation that $F_2$ hybrids will often be less fit than $F_1$ hybrid crosses, in the early stages of speciation.
-% This mechanism relies on the genetic variation of a system being shared across the sex chromosomes and the autosomes,
-% and appears to be consistent with the ``dominance'' theory \citep{orr1997haldane, coyne1998evolutionary, turelli1995dominance} which is often invoked to explain Haldane's rule.
-% \plr{Explain that the effects on fitness are dominant.}
-%and makes several different predictions than the dominance theory as formulated by \citet{turelli1995dominance}.
-%\plr{what are they?}
-%\citet{turelli1995dominance} employ a population genetic model which strictly requires alleles causing hybrid incompatibility on the X chromosome not to exhibit additive effects, that is the \emph{dominance coefficient} (typically $h$ in classic population genetics) must not be equal to $\frac{1}{2}$.
-In their model, if alleles are strictly additive, that is neither dominant nor recessive, then male and female $F_1$ hybrids will drop in fitness at the exact same rates, and therefore defy Haldane's rule.
-
-This model differs from the \emph{dominance theory} \citep{turelli1995dominance} in that it treats system coefficients as quantitative traits.
-%The classification of alleles as either ``dominant'' or ``recessive'' fits most naturally for Mendellian traits. 
-If we think of individual loci of large effect then the effect of mutations on fitness (if convex) will usually be recessive, congruent with the dominance theory.
-However, the dominance or recessiveness of even large effect mutations will depend on both the magnitude of their effect and precisely how they adjust system coefficients.
-%For such mutations that keep systems on the neutral manifold (imagine mutations to the oscillator that only move $\tau$), the dominance coefficient (usually notated as $h$) will be mathematically undefined.
-Mutations that driectly perturb coefficiens of $A$ can be either dominant or recessive, depending on their size.
-
-\citet{barton2001role} explored Haldane's rule in the context of a quantitative trait and concluded that it was a consequence of the dominance theory.
-However, Barton also demonstrated $F_2$ hybrid breakdown in a model with strictly \emph{additive} inheritence -- a consequence of segregation variance.
-The model suggested here explains Haldane's rule as a special case of hybrid breakdown.
-As hybrid breakdown can occur for a quantitative trait under additivity, so can heterogametic inviability.
-This suggests that the model presented here is distinct from the dominance theory; like hybrid breakdown, Haldane's rule derives from the nature of segregation variance.
+\revpoint{1}{4}
 
@@
 
-Fluctuating selection \citep{bell2010fluctuating, barton2001role} could further accelerate speciation via system drift. 
+Selection could dramatically accelerate speciation by essentially the same process --
+for instance, \citet{mani1990mutational} showed that populations
+developed incompatibilities as they independently evolved towards the same quantitative trait optimum,
+as different alleles fix in different populations. \linelabel{citemani}
+Fluctuating selection \citep{bell2010fluctuating, barton2001role} is ongoing, so could further accelerate speciation.
 Under our model, populations evolving towards different optimal phenotypes will always produce less fit hybrids,
@@
 will be necessary to identify the biologically relevant parameter regimes.
+
+\paragraph{Geometric models} \linelabel{quantnote}
+Our results describe a continuum of possible systems,
+described by their continuous coefficient matrices.
+The arguments we make (e.g., Figure~\ref{fig:conceptual_fig}) are often geometric in nature,
+and so it would be interesting to ask whether Fisher's geometric model
+\citep[as in, e.g.,][]{martin2014,fraisse2016genetics,simon2017coadapted}
+might provide more quantitative, generalizable predictions.
+It is also natural to ask about the distribution of effect sizes of individual mutations
+in a population evolving under this model,
+and what the idstribituion of selection coefficients would be
+(at equilibrium, perhaps).
+These are questions that might be best answered by simulation,
+and could provide much additional information -- for instance,
+are alleles typically dominant in their effects on system coefficients?
+or, fitness?
 

### 2021-08-11 review-responses2.tex
edits

@@
     on face value, the similarity to the snowball depends entirely on the relationship between
-    ``distance to optimum'' and fitness: since distance to optimum increases linearly in F2s,
+    ``distance to optimum'' and fitness: since distance to optimum increases linearly in $F_2$s,
     if this relation is quadratic then our model will also show a quadratic increase in incompatibility.
@@
 In our model system coefficients are treated as independent heritable units, and thus would represent genes spread out sufficiently far apart along the genome.
-Since the major source of F2 hybrid breakdown in this model is from recombining regulatory coefficients, if the genetic architecture of these coefficients was such that the genes were close enough to not recombine, we would not see the stark pattern of hybrid breakdown that we observe in F2 oscillators.
+Since the major source of $F_2$ hybrid breakdown in this model is from recombining regulatory coefficients, if the genetic architecture of these coefficients was such that the genes were close enough to not recombine, we would not see the stark pattern of hybrid breakdown that we observe in $F_2$ oscillators.
 We suspect that genetic architectures that allow for recombination, albeit at a lower rate, than for architectures in which genes are very far apart, will diminish the speed of speciation.
@@
 
-Thank you.
+Thank you. We are happy that you liked the examples and their accompanying visuals.
 
@@
 \reply{
-  The fitness of heterogametic F1s and F2s will always depend on the genetic distance between parents and on the particular gene network.
-Thus it is not guaranteed that the heterogametic F1s, will always have lower fitnesses than homogametic F1s.
-This is clear when the genetic distance is zero between parents.
-The model does show, however, that the particular combination of genes found in F1 heterogametes would only be possible in F2 autosomal crosses.
-Our model shows that phenotypes generated by networks contained on the autosomes have lower fitnesses in F2s than in F1s, and by extension, lower fitness in the heterogametic sex when networks are contained, at least in part, on the sex chromosomes.
+  Yes, that is correct. On average, $F_2$s will have lower fitness than $F_1$s. As the reviewer points out, some $F_2$s could have higher fitnesses than $F_1$s.
+  This will be true when the optimal set is curved.
+In general, hybrid fitnesses will depend on the specific geometry of the optimal set, and in Appendix A we show that $F_2$s will generally be farther than $F_1$s from this set.
+As such, depending on the particular geometry, as well as the parental distances, Haldane's rule might not always occur.
+The key point we would like to communicate is that when $F_2$ hybrid breakdown \emph{can} occur (i.e.
+due to geometry and distance), and system coefficients are distributed across the autosomes and sex chromosomes, Haldane's rule will manifest.
+We have added a clarifying paragraph \revref.
 
-  It is also true that some F2 individuals will perform as well as F1s and F0s -- for instance it is possible to recover an F0-like genotype (and therefore phenotype) after two generations of hybridization.
-In this model, hybrids have lower fitness because of increased genetic variance.
-However, there will also be F2 individuals with genes from both parental types -- meaning that there will be more low fitness F2s than low fitness F1s and F0s.
 }
@@
 
-\reply{Thank you for this suggestion. We've decided to keep the paragraph order the same though.\jss{Having second thoughts on this... rearranging could be clearer.}
+\reply{Thank you for this suggestion, but we would prefer to keep the current paragraph order. 
+We agree that the second-to-last paragraph does provide a nice summary of the work, however,
+we think the current order which first discusses system theory, followed by quantiative genetics, is more consistent with the ordering of the rest of text.
+%We've rearranged the last part of the introduction to improve clarity.
+\jss{Changed it, then changed it back again.
+I think, as I write above, that the original order fits better with the order of the text where the quant part comes second.
+If you think you'd prefer the alternative ordering I'd suggest leaving the part starting with ``Our main aim...'' as the last paragraph and then swapping the rest of the quant paragraph with the system one -- leaving in the holey landscape reference.}
 }

### 2021-08-11 kalman_walk_paper.tex
edits

@@
 
-In this paper we use results on mathematical nonidentifiability from\hypertarget{rev2:1}
+In this paper we use results on mathematical nonidentifiability from \hypertarget{rev2:1}
 linear systems theory to study how gene regulatory networks can be modified
@@
 and use quantitative genetics theory to estimate the speed at which system drift can lead to reproductive incompatibility and hence speciation.
+%
 In this model,
@@
 in a similar vein as \emph{holey landscape} models \citep{gavrilets1997evolution, yamaguchi2013first, pina2019does}.
-% Our model is developed in more detail in a companion paper \citep{schiffman2020genetic}.
 
@@
 and the shape of the fitness landscape \citep{fraisse2016genetics}.
+%
 Our main aim is to provide a concrete framework 
-that can provide natural predictions of these model parameters
+that can make natural predictions of these model parameters
 across a general class of models.
@@
 there are \emph{still} usually degrees of freedom. 
-The following example \ref{ex:all_osc} gives the set of minimal systems equivalent to the oscillator of Example \ref{ex:oscillator},
+Example \ref{ex:all_osc} gives the set of minimal systems equivalent to the oscillator of Example \ref{ex:oscillator},
 that all share common $B$ and $C$ matrices.
@@
 \begin{example}[All equivalent rewirings of the oscillator] \label{ex:all_osc}
-The oscillator of example \ref{ex:oscillator} is minimal, and so any equivalent system is a change of coordinates
+The oscillator of Example \ref{ex:oscillator} is minimal, and so any equivalent system is a change of coordinates
 by an invertible matrix $V$.
@@
 probably more than would be observed between diverging populations.
-In figure \ref{fig:hybs} we compare the phenotypes for $F_1$ and $F_2$ hybrids between more similar parents,
+In Figure \ref{fig:hybs} we compare the phenotypes for $F_1$ and $F_2$ hybrids between more similar parents,
 and see increasingly divergent phenotypes as the difference between the parental systems increases.
@@
     top-to-bottom, homozygous $A(\sfrac{1}{100})$, $A(\sfrac{1}{10})$, and $A(\sfrac{1}{2})$ parents,
-    where $A(\epsilon)$ is defined in figure \ref{fig:all_osc};
+    where $A(\epsilon)$ is defined in Figure \ref{fig:all_osc};
     parental coefficients differ by around 0.5\%, 5\%, and 25\% respectively.
@@
 \revpoint{1}{4}
+
+Distribution of system coefficients across the autosomes and sex chromosomes does not alone guarantee Haldane's rule.
+Under this model, Haldane's rule will only occur when there is sufficient genetic distance seperating parentals
+and when the geometry of the optimal set produces $F_2$ hybrid breakdown. \revpoint{2}{1}
 
@@
 \revpoint{1}{1}
-\jss{Did you also want to add something about bottlenecking large versus small popoulations?}
+%\jss{Did you also want to add something about bottlenecking large versus small popoulations?}
 %and incompatibilities identified between \emph{Drosophila simulans} and \emph{D. melanogaster} exhibit evidence of adaptative evolution \citep{tang2009evolution}.
@@
 and so it would be interesting to ask whether Fisher's geometric model
-\citep[as in, e.g.,][]{martin2014,fraisse2016genetics,simon2017coadapted}
+\citep[as in, e.g.,][]{martin2014fishers,fraisse2016genetics,simon2017coadapted}
 might provide more quantitative, generalizable predictions.
@@
 in a population evolving under this model,
-and what the idstribituion of selection coefficients would be
+and what the distribituion of selection coefficients would be
 (at equilibrium, perhaps).
@@
 and could provide much additional information -- for instance,
-are alleles typically dominant in their effects on system coefficients?
-or, fitness?
+are alleles typically dominant in their effects on system coefficients or fitness?
 

### 2021-08-11 kalman_walk_paper.tex
minor edits

@@
 through the impulse response function $h(t)$,
-they may affect individual fitness in other ways. \revpoint{2}{4}
+they may affect individual fitness in other ways \revpoint{2}{4}
+(which we don't consider here).
 For instance,
@@
 If both genes resided on the autosomes, this system would only be possible in an $F_2$ cross.
-More generally, 
-if the regulatory coefficients for a system are shared between the sex and one or more autosomal chromosomes, 
-$F_1$ males are effectively equivalent to purely autosomal-system $F_2$ hybrids, and recall that $F_2$s are significantly less fit on average than $F_1$s (see Figure \ref{fig:osc_incompat}).
+More generally, the contribution of sex chromosomes to the system coefficients of $F_1$ males
+are effectively equivalent to those of purely autosomal $F_2$ hybrids,
+which are less fit on average than $F_1$s (see Figure \ref{fig:osc_incompat}).
 \revpoint{1}{4}
-
-Distribution of system coefficients across the autosomes and sex chromosomes does not alone guarantee Haldane's rule.
-Under this model, Haldane's rule will only occur when there is sufficient genetic distance seperating parentals
-and when the geometry of the optimal set produces $F_2$ hybrid breakdown. \revpoint{2}{1}
+Note, however, that presence of genetic variation on sex chromosomes
+does not alone guarantee Haldane's rule:
+under this model, Haldane's rule will only occur when there is sufficient genetic distance separating parentals
+and when the nature of genetic variation produces $F_2$ hybrid breakdown. \revpoint{2}{1}
 
@@
 As classically formulated, 
-the Dobzhansky-Muller model of hybrid incompatibility is agnostic to the relative importance of neutral versus selective genetic substitutions \citep{coyne1998evolutionary},
+the Dobzhansky-Muller model of hybrid incompatibility is agnostic to the relative importance of neutral versus selected genetic substitutions \citep{coyne1998evolutionary},
 and plausible mechanisms have been proposed
@@
 two processes that typically act much more rapidly than genetic drift.
-Experimental approaches designed to study the effect of drift on speciation which induced population bottlenecks found no reproductive isolation in the majority tested pairs\citep{fry2009laboratory, white2020past}.
-These bottlenecks might not maintain enough genetic variation for systems to drift at an appreciable rate, however.
-\revpoint{1}{1}
-%\jss{Did you also want to add something about bottlenecking large versus small popoulations?}
-%and incompatibilities identified between \emph{Drosophila simulans} and \emph{D. melanogaster} exhibit evidence of adaptative evolution \citep{tang2009evolution}.
-%
-% In contrast, the model we develop here suggests that even under strictly neutral conditions, 
-% system drift can lead to speciation, at a rate fast enough 
-% to play a substantial role in species formation across the tree of life. 
+Supporting this view, experimental approaches designed to study the effect of drift on speciation
+through induced population bottlenecks
+found no reproductive isolation in the majority of tested pairs \citep{fry2009laboratory, white2020past}. \revpoint{1}{1}
+%and incompatibilities identified between \emph{Drosophila simulans} and \emph{D.~melanogaster} exhibit evidence of adaptive evolution \citep{tang2009evolution}.
 However, our calculations suggest that even under strictly neutral processes,
@@
 % (as diversification is not observed to be accelerated following mass extinctions).
-More recently, \citet{sanchez2021widespread} observed widespread transcriptional misregulation in \emph{Drosophila} hybrids, implicating the action of system drift.
-
-Selection could dramatically accelerate speciation by essentially the same process --
-for instance, \citet{mani1990mutational} showed that populations
+More recently, \citet{sanchez2021widespread} observed widespread transcriptional misregulation in \emph{Drosophila} hybrids,
+which might be due to the action of system drift.
+
+Although the phrase ``system drift'' as we use it implies neutral changes,
+it does not mean that these changes are induced purely by neutral genetic drift.
+Selection could dramatically accelerate speciation by the same process --
+i.e., \emph{genetic draft} \citep{gillespie2000genetic}.
+For instance, \citet{mani1990mutational} showed that populations
 developed incompatibilities as they independently evolved towards the same quantitative trait optimum,
 as different alleles fix in different populations. \linelabel{citemani}
-Fluctuating selection \citep{bell2010fluctuating, barton2001role} is ongoing, so could further accelerate speciation.
+Fluctuating selection has similarly been shown to accelerate speciation
+in some models \citep{bell2010fluctuating,barton2001role}.
 Under our model, populations evolving towards different optimal phenotypes will always produce less fit hybrids,
@@
 In the case where fluctuations are synchronized, due to the size of the neutral system space, 
-populations may arive at different system organizations when returning to previous phenotypes at a rate faster than by drift alone. \revpoint{1}{3}
+populations may arrive at different system organizations when returning to previous phenotypes at a rate faster than by drift alone. \revpoint{1}{3}
 
@@
 any constraints on such variation,
-possibly due to epistasis, genetic architecture \citep{hermisson2003epistasis}, adaptive inertia \citep{baatz1997adaptive,alvarez2009effects} or weak gene flow could plausibly slow it down.
-\revpoint{AE}{1}\revpoint{AE}{3}
+possibly due to epistasis, genetic architecture \citep{hermisson2003epistasis}, adaptive inertia \citep{baatz1997adaptive,alvarez2009effects}
+or weak gene flow could plausibly slow it down. \revpoint{AE}{1}\revpoint{AE}{3}
 More work on specific systems,
@@
 \paragraph{Geometric models} \linelabel{quantnote}
-Our results describe a continuum of possible systems,
-described by their continuous coefficient matrices.
+The model we work with describes a continuum of possible systems,
+parameterized by continuous coefficient matrices.
 The arguments we make (e.g., Figure~\ref{fig:conceptual_fig}) are often geometric in nature,
 and so it would be interesting to ask whether Fisher's geometric model
-\citep[as in, e.g.,][]{martin2014fishers,fraisse2016genetics,simon2017coadapted}
-might provide more quantitative, generalizable predictions.
-It is also natural to ask about the distribution of effect sizes of individual mutations
-in a population evolving under this model,
-and what the distribituion of selection coefficients would be
-(at equilibrium, perhaps).
+might provide more quantitative, generalizable predictions,
+as in for instance \citet{fraisse2016genetics,simon2017coadapted}.
+\citet{martin2014fishers} argued that such an approximation would be often fruitful.
+In the other direction,
+it is also natural to ask about the distribution of effect sizes of segregating mutations
+in a population evolving under this model with a realistic \emph{de novo} effect size distribution,
+as well as what the distribution of fitness effects would be (both within and between populations).
 These are questions that might be best answered by simulation,
 and could provide much additional information -- for instance,
-are alleles typically dominant in their effects on system coefficients or fitness?
+are alleles typically dominant in their effects on either system coefficients or fitness?
 

### 2021-08-11 review-responses2.tex
changed out hyperlinks

@@
 The central sections, especially e.g.
-\hyperlink{rev1:1}{lines 127-219}, are very clear and novel.
+\llname{r11b} to \llname{r11e}, are very clear and novel.
 I do think, though that the work could be better placed in context.
@@
 %
-  The section beginning on \hyperlink{rev1:4}{line 419} seems quite important to me, but it seems to downplay an important point.
+  The section beginning on \llname{r14}~seems quite important to me, but it seems to downplay an important point.
 Populations might diverge genetically, or in their kryptotypes, but not in their phenotypes, if they experience fluctuating selection.
@@
 My view is that the sort of model explored here (like additive quantitative trait model used by Barton 2001) naturally generates the sort of dominance relations required for the ``dominance theory'' to hold.
-Unless the authors strongly disagree, I would modify the abstract and \hyperlink{rev1:5}{lines 278-291}, or at least provide a fuller clarification for why the expanation is different from that discussed by Barton 2001.
+Unless the authors strongly disagree, I would modify the abstract and \llname{r15b} to \llname{r15e}, or at least provide a fuller clarification for why the expanation is different from that discussed by Barton 2001.
 \end{point}
@@
 \begin{point}{}
-  (very minor) \hyperlink{rev1:6}{Line 358}: As you say, F2 hybrid breakdown has been known about for a very long time. As such, the single citation used feels a bit arbitrary.
+  (very minor) \llname{r16}: As you say, F2 hybrid breakdown has been known about for a very long time. As such, the single citation used feels a bit arbitrary.
 \end{point}
@@
 \begin{point}{}
-  \hyperlink{rev2:1}{(p.
-2, l.
-73)} \jss{Note: R2 is using line numbers from the diff}: This paragraph summarises the main body of work performed in this paper.
-Thus, consider moving it to the end of the Introduction (i.e.
-switching places with the last paragraph).
+  \llname{r21}: This paragraph summarises the main body of work performed in this paper. Thus, consider moving it to the end of the Introduction (i.e. switching places with the last paragraph).
 \end{point}
@@
 \begin{point}{}
-  \hyperlink{rev2:2}{(p.
-2, l.
-92)}: ``...more than one way to do the same thing...'' sounds a bit too colloquial.
+  \llname{r22} : ``...more than one way to do the same thing...'' sounds a bit too colloquial.
 Consider changing it to ``...more than one way to accomplish the same outcome...'' or some other alternative phrasing.
@@
 \begin{point}{}
-  \hyperlink{rev2:3}{(p.
-5, l.
-192)}: ``The Kalman decomposition then classifies each direction in kryptotype space as either reachable or unreachable''.
+  \llname{r23}: ``The Kalman decomposition then classifies each direction in kryptotype space as either reachable or unreachable''.
 Is there a biological motivation for considering unreachable directions in kryptotype space as degrees of freedom of the neutral space?
@@
 \begin{point}{}
-\hyperlink{rev2:4}{(p.
-6, l.
-208)}: ``both the change of basis used to obtain the decomposition and, once in this form, all submatrices other than $A_{ro}$, $B_{ro}$, and $C_{ro}$ can be changed without affecting the phenotype, and so represent degrees of freedom''.
+  \llname{r24}: ``both the change of basis used to obtain the decomposition and, once in this form, all submatrices other than $A_{ro}$, $B_{ro}$, and $C_{ro}$ can be changed without affecting the phenotype, and so represent degrees of freedom''.
 A minor question on generality - does Kalman decomposition require that the coefficient matrices are time-invariant?
@@
 \begin{point}{Recombination}
-(p.
-7) In the section ``Sexual reproduction and recombination'', we think that the recombination process is not clearly explained, making it more challenging to read the text from this point.
+ \llname{r26}: In the section ``Sexual reproduction and recombination'', we think that the recombination process is not clearly explained, making it more challenging to read the text from this point.
 We suggest that the authors clarify how exactly recombination is implemented in the model.
@@
 \begin{point}{}
-(p.
-8, Figure 4 and p.
-10, Figure 5): We suggest dividing the figure into sub-figures (e.g.
+  (Figure \ref{fig:conceptual_fig} and Figure \ref{fig:hybs}): We suggest dividing the figure into sub-figures (e.g.
 A and B).

### 2021-08-11 kalman_walk_paper.tex
changed out hyperlinks

@@
 
-In this paper we use results on mathematical nonidentifiability from \hypertarget{rev2:1}
+In this paper we use results on mathematical nonidentifiability from \linelabel{r21}
 linear systems theory to study how gene regulatory networks can be modified
@@
 that there is often more than one way to do the same thing, \revpoint{1}{2}
-\hypertarget{rev2:2}
+\linelabel{r22}
 and observed that speciation might be the result of changes that are themselves neutral.
@@
 
-\begin{example}[An oscillator]\label{ex:oscillator}\hypertarget{rev1:1}
+\begin{example}[An oscillator]\label{ex:oscillator} \linelabel{r11b}
   For illustration, we consider an extremely simplified model of oscillating gene transcription,
@@
 and so can only infer motion in what is known as the \emph{observable subspace}.
-The Kalman decomposition then classifies each direction in kryptotype space\hypertarget{rev2:3}
+The Kalman decomposition then classifies each direction in kryptotype space\linelabel{r23}
 as either reachable or unreachable, and as either observable or unobservable.
@@
 It is remarkable that the gene regulatory network architecture to achieve a given input--output map is never unique --
-both the change of basis used to obtain the decomposition\hypertarget{rev2:4}
+both the change of basis used to obtain the decomposition\linelabel{r24}
 and, once in this form, all submatrices other than $A_{\ro}$, $B_{\ro}$, and $C_{\ro}$ can be changed without affecting the phenotype,
@@
     \label{fig:all_osc}
-\end{figure}
+\end{figure} \linelabel{r11e}
 
@@
 
-\paragraph{Sexual reproduction and recombination}
+\paragraph{Sexual reproduction and recombination}\linelabel{r26}
 Parents with phenotypically equivalent yet differently wired gene networks may produce offspring with dramatically different phenotypes. 
@@
 
-\paragraph{Haldane's rule}\hypertarget{rev1:5}
+\paragraph{Haldane's rule}\linelabel{r15b}
 This model naturally predicts
@@
 are effectively equivalent to those of purely autosomal $F_2$ hybrids,
-which are less fit on average than $F_1$s (see Figure \ref{fig:osc_incompat}).
+which are less fit on average than $F_1$s (see Figure \ref{fig:osc_incompat}). \linelabel{r15e}
 \revpoint{1}{4}
@@
 
-Consistent with empirical observation of hybrid breakdown %(e.g., \citet{plotner2017chlorosis}),\hypertarget{rev1:6} 
+Consistent with empirical observation of hybrid breakdown \linelabel{r16} %(e.g., \citet{plotner2017chlorosis}),\hypertarget{rev1:6} 
 we see that the fitnesses of $F_2$ hybrids drop at a much faster rate than those of $F_1$s.
@@
 
-Phenotypically equivalent system evolution is probably not only driven by neutral genetic drift. \hypertarget{rev1:4}
+Phenotypically equivalent system evolution is probably not only driven by neutral genetic drift. \linelabel{r14}
 For one thing, movement along the set of equivalent networks

### 2021-08-11 figures/landscape_figure.R
subplot labels

@@ -69,7 +69,9 @@ layout(t(1:2))
 par(mar=c(1,1,1,1))
 
 do.plot(p1, p2, sigma=.02, legend=FALSE)
+title(main="(A)", adj=0)
 quant.plot(p1, p2, sigma=.02)
+title(main="(B)", adj=0)
 
 dev.off()
 

### 2021-08-11 review-responses2.tex
edits to responses

@@
 
-Thank you for the kind words.
+Thank you for the kind words and the helpful input.
+We have revised various sections of the paper,
+including the section on Haldane's rule,
+and have added a paragraph to the discussion about the distinction between
+the analysis we conduct here -- essentially that of a continuous-trait quantitative model --
+and predictions of various models that have to do with discrete loci.
 
@@
 \reply{
-This is correct.
-As the model is currently described, we show that since different mutationally connected systems are not usually reproductively compatible, populations in different parts of system space will produce unfit hybrids.
-We rely on results from quantitative genetics to estimate the rate at which systems drift through the neutral space and thus to estimate the effects of gene flow.
-Without further modeling we do not yet know how a quantitative genetics model incorporating system theory would differ, if it all, from these classic results.
-We are planning to submit a paper where we further develop the removed quantitative genetics model and may be able to address this interesting question there.
+This is a very interesting question that we haven't attempted to analyze.
+% As the model is currently described, we show that since different mutationally connected systems are not usually reproductively compatible, populations in different parts of system space will produce unfit hybrids.
+Currently, the paper
+relies on general results from quantitative genetics
+to estimate the rate at which systems drift through the neutral space
+that don't depend on the details of the genetic architecture.
+However, the effects of gene flow depend on more things;
+in particular the effect sizes of mutations,
+something we've stayed away from.
+What effect sizes we might expect to be segregating is a good question
+(and we've added some discussion of this; \llname{quantnote}),
+but we feel that would be a topic for a separate paper
+(perhaps the former other half of this paper).
 
-We also briefly mention that gene flow will likely complicate matters in the discussion \revref.
+We have also added a brief mention
+that gene flow will likely complicate matters in the discussion \revref.
 }
@@
 The authors state that rapid emergence of incompatibility is possible in their model.
-How would the accumulation of incompatibility here compare with Orr’s ``snownball'' model, and how is this expected to depend on the complexity of the underlying networks?
+How would the accumulation of incompatibility here compare with Orr’s ``snowball'' model, and how is this expected to depend on the complexity of the underlying networks?
 \end{point}
@@
     in which each pair of lineage-specific substitutions possibly contributes to incompatibility.
-    This may be an empirical question (that could be examined in simulation),
+    This may be an empirical question (that could be examined in simulation, as noted in \llname{quantnote}),
     but at the level of analysis in our paper it seems the answer is ``no'':
@@
 \begin{point}{Genetic architecture}
-The genetic architecture (e.g., interactions between cis-regulatory and coding regions) is important to determine the barrier to gene flow that an incompatibiltiy builds.
+The genetic architecture (e.g., interactions between cis-regulatory and coding regions) is important to determine the barrier to gene flow that an incompatibility builds.
 If there is a lot of short-distance interactions (physically in the genome; i.e.
 low recombination), they might not play a role for speciation per se because the incompatibility will often not be exposed to selection even in hybrids.
-In the light of our current knowledge of the genetic archtiecture of regulatory networks, how would this affect the conclusions from the model?
+In the light of our current knowledge of the genetic architecture of regulatory networks, how would this affect the conclusions from the model?
 \end{point}
@@
 We suspect that genetic architectures that allow for recombination, albeit at a lower rate, than for architectures in which genes are very far apart, will diminish the speed of speciation.
-This also leads to interesting questions about the evolution of genetic architecture; perhaps arhitectures in which co-regulating genes are neighbors are more robust.
+This also leads to interesting questions about the evolution of genetic architecture; perhaps architectures in which co-regulating genes are neighbors are more robust.
 
-We briefly mention complications to the model in the discussion \revref.
+We briefly mention such complications to the model in the discussion \revref.
 }
@@
 
-\reply{This is an important point. We have added references to the experimental speciation literature \revref. 
+\reply{
+    This is an important point.
+    We have added references to the experimental speciation literature, around \revref. 
 }
@@
     We have also added a reference to \citet{mani1990mutational} at \llname{citemani},
-    and added further reference to Fisher's geometric model in the Discussion \llname{quantnote}.
+    where we think it's better put in context,
+    and added further discussion of the possible connections to Fisher's geometric model in the Discussion \llname{quantnote},
+    including a citation of \citet{martin2014fishers} (which is very relevant; thanks!).
 }
@@
 
-\reply{This is also a very interesting point worth discussing. We have added discussion on fluctuating selection \revref.
+\reply{
+    We agree that the earlier draft of the paper downplayed this point
+    more than it should have;
+    we have added discussion on the role of selection:
+    see the paragraph at \revref.
 }
@@
 \begin{point}{Interpretation of the Haldane’s Rule result}
-%4. Interpretation of the Haldane’s Rule result
-%
   The abstract refers to ``another possible explanation'' of Haldane's Rule, and the explanation given is said to be quite different from the ``dominance theory'', because ``it derives from the nature of segregation variance''.
@@
 My view is that the sort of model explored here (like additive quantitative trait model used by Barton 2001) naturally generates the sort of dominance relations required for the ``dominance theory'' to hold.
-Unless the authors strongly disagree, I would modify the abstract and \llname{r15b} to \llname{r15e}, or at least provide a fuller clarification for why the expanation is different from that discussed by Barton 2001.
+Unless the authors strongly disagree, I would modify the abstract and \llname{r15b} to \llname{r15e}, or at least provide a fuller clarification for why the explanation is different from that discussed by Barton 2001.
 \end{point}
@@
 The authors model gene regulatory networks as sets of linear equations, where environmental inputs are integrated, giving rise to specific phenotypic outputs.
-Using Kalman decomposition, the authors identify the subset of such systems that produce the same phenotype and thus, characterise how a given system may be modified without affecting its phenotypic response.
+Using Kalman decomposition, the authors identify the subset of such systems that produce the same phenotype and thus, characterize how a given system may be modified without affecting its phenotypic response.
 Resorting to a simple oscillatory network as an example, the authors show that phenotypically identical parents can generate offspring with considerably different phenotypes, resulting in hybrid incompatibilities if the set of equivalent phenotypes is not closed under the average.
@@
   Yes, that is correct. On average, $F_2$s will have lower fitness than $F_1$s. As the reviewer points out, some $F_2$s could have higher fitnesses than $F_1$s.
-  This will be true when the optimal set is curved.
-In general, hybrid fitnesses will depend on the specific geometry of the optimal set, and in Appendix A we show that $F_2$s will generally be farther than $F_1$s from this set.
-As such, depending on the particular geometry, as well as the parental distances, Haldane's rule might not always occur.
-The key point we would like to communicate is that when $F_2$ hybrid breakdown \emph{can} occur (i.e.
-due to geometry and distance), and system coefficients are distributed across the autosomes and sex chromosomes, Haldane's rule will manifest.
-We have added a clarifying paragraph \revref.
+In general, hybrid fitnesses will depend on the specific geometry of the optimal set.
+As such, depending on the particular geometry, as well as the parental distances,
+Haldane's rule might not always occur.
+The key point we would like to communicate is that when $F_2$ hybrid breakdown \emph{can} occur
+(i.e. due to geometry and distance),
+and system coefficients are distributed across the autosomes and sex chromosomes,
+Haldane's rule will manifest.
+We have added some clarifying text \revref.
 
@@
 We agree that the second-to-last paragraph does provide a nice summary of the work, however,
-we think the current order which first discusses system theory, followed by quantiative genetics, is more consistent with the ordering of the rest of text.
-%We've rearranged the last part of the introduction to improve clarity.
-\jss{Changed it, then changed it back again.
-I think, as I write above, that the original order fits better with the order of the text where the quant part comes second.
-If you think you'd prefer the alternative ordering I'd suggest leaving the part starting with ``Our main aim...'' as the last paragraph and then swapping the rest of the quant paragraph with the system one -- leaving in the holey landscape reference.}
+after trying out the alternate ordering,
+we decided that the current order
+(which first discusses system theory, followed by quantitative genetics)
+is more consistent with the ordering of the rest of text.
 }
@@
 \reply{
-    Good point; we've added some additional discussion of what the subspaces represent. \revref
+    A natural question; we've added some additional discussion of what the subspaces represent. \revref
 }
@@
 \reply{
-    Good point; this was unclear. We have clarified that we don't pick a particular model for recombination in this seciton. \revref
+    Good point; this was unclear. We have clarified that we don't pick a particular model for recombination in this section. \revref
 }

### 2021-08-11 kalman_walk_paper.tex
edits to responses

--- 2021-08-11 kalman_walk_paper.tex
@@
 
-In this paper we use results on mathematical nonidentifiability from \hypertarget{rev2:1}
+In this paper we use results on mathematical nonidentifiability from \llabel{r21}
 linear systems theory to study how gene regulatory networks can be modified
@@
 that there is often more than one way to do the same thing, \revpoint{1}{2}
-\hypertarget{rev2:2}
+\llabel{r22}
 and observed that speciation might be the result of changes that are themselves neutral.
@@
 
-\begin{example}[An oscillator]\label{ex:oscillator}\hypertarget{rev1:1}
+\begin{example}[An oscillator]\label{ex:oscillator} \llabel{r11b}
   For illustration, we consider an extremely simplified model of oscillating gene transcription,
@@
 and so can only infer motion in what is known as the \emph{observable subspace}.
-The Kalman decomposition then classifies each direction in kryptotype space\hypertarget{rev2:3}
+The Kalman decomposition then classifies each direction in kryptotype space\llabel{r23}
 as either reachable or unreachable, and as either observable or unobservable.
@@
 It is remarkable that the gene regulatory network architecture to achieve a given input--output map is never unique --
-both the change of basis used to obtain the decomposition\hypertarget{rev2:4}
+both the change of basis used to obtain the decomposition\llabel{r24}
 and, once in this form, all submatrices other than $A_{\ro}$, $B_{\ro}$, and $C_{\ro}$ can be changed without affecting the phenotype,
@@
     \label{fig:all_osc}
-\end{figure}
+\end{figure} \llabel{r11e}
 
@@
 
-\paragraph{Sexual reproduction and recombination}
+\paragraph{Sexual reproduction and recombination}\llabel{r26}
 Parents with phenotypically equivalent yet differently wired gene networks may produce offspring with dramatically different phenotypes. 
@@
     the distribution of $F_2$ phenotypes:
-    \textbf{(left)}
+    \textbf{(A)}
     $F_2$s compose all four mixed homozygotes if variation at both
     traits has a simple, one-locus genetic basis in both populations; and
-    \textbf{(right)}
+    \textbf{(B)}
     $F_2$ show a much wider distribution of phenotypes
@@
 
-\paragraph{Haldane's rule}\hypertarget{rev1:5}
+\paragraph{Haldane's rule}\llabel{r15b}
 This model naturally predicts
@@
 are effectively equivalent to those of purely autosomal $F_2$ hybrids,
-which are less fit on average than $F_1$s (see Figure \ref{fig:osc_incompat}).
-\revpoint{1}{4}
+which are less fit on average than $F_1$s (see Figure \ref{fig:osc_incompat}). \llabel{r15e}
+\revpoint{1}{4} \revpoint{2}{1}
 Note, however, that presence of genetic variation on sex chromosomes
@@
 under this model, Haldane's rule will only occur when there is sufficient genetic distance separating parentals
-and when the nature of genetic variation produces $F_2$ hybrid breakdown. \revpoint{2}{1}
+and when the nature of genetic variation produces $F_2$ hybrid breakdown.
 
@@
 
-Consistent with empirical observation of hybrid breakdown %(e.g., \citet{plotner2017chlorosis}),\hypertarget{rev1:6} 
+Consistent with empirical observation of hybrid breakdown \llabel{r16}
 we see that the fitnesses of $F_2$ hybrids drop at a much faster rate than those of $F_1$s.
@@
 developed incompatibilities as they independently evolved towards the same quantitative trait optimum,
-as different alleles fix in different populations. \linelabel{citemani}
+as different alleles fix in different populations. \llabel{citemani}
 Fluctuating selection has similarly been shown to accelerate speciation
@@
 while the fitness consequences of incompatibility in any one given network may be small, 
-the cumulative impact of system drift across the many different networks an organism relies on may be substantial. \linelabel{rp11}
+the cumulative impact of system drift across the many different networks an organism relies on may be substantial. \llabel{rp11}
 It remains to be seen how the relative strengths of these forces compare.
@@
 
-Phenotypically equivalent system evolution is probably not only driven by neutral genetic drift. \hypertarget{rev1:4}
+Phenotypically equivalent system evolution is probably not only driven by neutral genetic drift. \llabel{r14}
 For one thing, movement along the set of equivalent networks
@@
 
-\paragraph{Geometric models} \linelabel{quantnote}
+\paragraph{Geometric models} \llabel{quantnote}
 The model we work with describes a continuum of possible systems,

### 2021-08-11 kalman_walk_paper.tex
minor adjustment

@@
 Selection could dramatically accelerate speciation by the same process --
-i.e., \emph{genetic draft} \citep{gillespie2000genetic}.
+e.g., by genetic draft \citep{gillespie2000genetic} or pleiotropic selection.
 For instance, \citet{mani1990mutational} showed that populations
@@
 Under our model, populations evolving towards different optimal phenotypes will always produce less fit hybrids,
-as the hybrid phenotypes will be displaced from both parental optimums -- that is maladapted to both parental environments.
+as the hybrid phenotypes will be displaced from both parental optima,
+and so maladapted to both parental environments.
 This suggests that fluctuating selection -- if not synchronized between populations -- should diminish hybrid fitnesses.

### 2021-08-11 kalman_walk_paper.tex
bib fixes

@@
 a population diffuses along the neutral ridges of a high-dimensional space of possible system parameters,
-in a similar vein as \emph{holey landscape} models \citep{gavrilets1997evolution, yamaguchi2013first, pina2019does}.
+in a similar vein as \emph{holey landscape} models \citep{gavrilets1997evolution, yamaguchi2013first}.
 
@@
 
-%\section*{Acknowledgements}
-%
-%    We would like to thank Sergey Nuzhdin, Stevan Arnold, Michael Turelli, Patrick Phillips, Erik Lundgren and Hossein Asgharian for valuable discussion. 
-%    We would also like to thank Nick Barton, Sarah Signor, Todd Parsons, and Joachim Hermisson for very helpful comments on the manuscript.
-%    Work on this project was supported by funds from
-%    the Sloan Foundation and the NSF (under DBI-1262645) to PR.
+\section*{Acknowledgements}
+
+We would like to thank Sergey Nuzhdin, Stevan Arnold, Michael Turelli, Patrick Phillips, Erik Lundgren and Hossein Asgharian for valuable discussion. 
+We would also like to thank Nick Barton, Sarah Signor, Todd Parsons, and Joachim Hermisson for very helpful comments on the manuscript.
+Work on this project was supported by funds from
+the Sloan Foundation and the NSF (under DBI-1262645) to PR.
 

### 2021-08-11 kalman_walk_paper.tex
updated figure 5

@@
 \begin{figure}[H]
-  \centering
+  %\centering
  % \begin{tabular}{ccc}
@@
   %\includegraphics[scale=0.5]{examples/f1f2_newplot}
-  \includegraphics{examples/f1f2_small}
+  \includegraphics[scale=0.5]{examples/blue_hybrids}
   \caption{
@@
     parental coefficients differ by around 0.5\%, 5\%, and 25\% respectively.
-    Parental phenotypes ($\sin t + \cos t$) are shown in solid black, and hybrid phenotypes in dashed blue.
+    Parental phenotypes ($\sin t + \cos t$) are shown in solid black, and hybrid phenotypes in blue.
     \textbf{(right)} Phenotypes of all $3^4 = 81$ possible $F_2$ hybrids between the same set of parents,
@@
     %%   it only matters if the offspring is homo/het/homozygous, so there are 3^4 = 81 possible expression patterns.
-    Different colored lines correspond to different $F_2$ hybrids,
+    Blue lines correspond to different $F_2$ hybrids,
     many of which show complete breakdown.

### 2021-08-11 examples/Fig5_updated.R
code for new Fig5

@@ -1,0 +1,143 @@
+library(expm)
+library(grid)
+library(gridExtra)
+library(ggplot2)
+library(ggpubr)
+
+A0 <- matrix(c(0,-1,1,0), nrow=2)
+B <- matrix(c(1,1), ncol=1)
+C <- matrix(c(1,0), nrow=1)
+
+A_fn <- function (tau) { 
+    matrix(c(tau/(tau+1), -(2*tau*(tau+1)+1)/(tau+1), 
+             1/(tau+1), -tau/(tau+1)), nrow=2) 
+}
+
+tt <- seq(0,6*2*pi, length.out=500)
+
+h <- function (t,tau=0,A=A_fn(tau)) { sapply(t, function (tt) C %*% expm(tt*A) %*% B) }
+
+
+F1 <- function (tau1,tau2) { (A_fn(tau1)+A_fn(tau2))/2 }
+
+get_F2s <- function(tau1, tau2) 
+{
+    vc <- function(tau1, tau2, x, y) 
+    {
+        out <- c(A_fn(tau1)[x,y], A_fn(tau2)[x,y])
+        return(out)
+    }
+    mlist <- list()
+    count <- 1
+    for (i in 1:2) 
+    {
+        for (j in 1:2) 
+        {
+            for (k in 1:2) 
+            {
+                for (l in 1:2)
+                {
+                    mlist[[count]] <- matrix(c(vc(tau1, tau2, 1, 1)[i], vc(tau1, tau2, 2, 1)[j], vc(tau1, tau2, 1, 2)[k], vc(tau1, tau2, 2, 2)[l]),2,2)
+                    count <- count + 1
+                }
+            }
+        }
+    }
+    f2list <- list()
+    count2 <- 1
+    for(i in 1:length(mlist)) 
+    {
+        for (j in 1:length(mlist))
+        {
+            f2list[[count2]] <- (mlist[[i]] + mlist[[j]])/2
+            count2 <- count2 + 1
+        }
+    }
+    return(unique(f2list))
+}
+
+
+hhf0 <- sapply( list(A_fn(0), A_fn(0)), function (A) { sapply(tt, h, A=A) } )
+
+hhf1_1 <- sapply( list(A_fn(0), A_fn(0.01), F1(0, 0.01)), function (A) { sapply(tt, h, A=A) } )
+hhf1_2 <- sapply( list(A_fn(0), A_fn(0.1), F1(0, 0.1)), function (A) { sapply(tt, h, A=A) } )
[…]

### 2021-08-11 examples/Fig5_updated.R
code for new Fig5

@@ -3,6 +3,7 @@ library(grid)
 library(gridExtra)
 library(ggplot2)
 library(ggpubr)
+library(reshape2)
 
 A0 <- matrix(c(0,-1,1,0), nrow=2)
 B <- matrix(c(1,1), ncol=1)

### 2021-08-11 review-responses2.tex
cover letter

@@
 
- Cover letter.
+We are writing to submit a revision of our manuscript, ``{System drift and speciation}'' (ID 21-0262), for your review,
+and would like to thank you and the reviewers for your suggestions, as they have helped us improve our manuscript's clairty.
+We also would like to thank you and the reviewers for the enthusiasm and positive feedback; we are happy that our manuscript was seen as interesting and relevant.
+
+Both reviewers made comments about the section on Haldane's rule, of which we have revised to address their concerns. 
+We also made an effort to provide more context for our manuscript by citing the relevant literature, as requested by Reviewer 1. 
+To address some of Reviewer 2's concerns, we added more discussion and detail pertaining to system theory and its biological interpretation.
+
+We hope that by making these revisions you will find this manuscript fit for publication.
+
+%We are writing to submit a revision of our manuscript, ``{System drift and speciation}'' (ID 18-0316), for your review. 
+%We sincerely apologize for the delayed resubmission but hope that you will find the
+%revised manuscript significantly improved.
+%
+%We thank you and the two reviewers for providing constructive suggestions 
+%and feedback that has helped to improve the manuscript. We have followed 
+%the suggestions closely and hope to have fully addressed all the concerns and 
+%suggestions as detailed in our point-by-point response below. 
+%
+%In our original submission, we studied the effect of gene regulatory network evolution on speciation,
+%first by applying linear system theory to characterize the large space of phenotypically equivalent network organizations,
+%and second by using quantitative genetics to show that neutral genetic drift on this network space can lead to speciation over plausible timescales. 
+%
+%We agree with the reviewers that the combination of these two approaches, while interesting,
+%was not seamless and that the quantitative genetics section required further development, 
+%and thus have followed the recommendation of both the Editor and Associate Editor,
+%by breaking the manuscript into two papers.
+%One paper -- the present revision -- covers the application of system theory to regulatory networks,
+%and the other (which we plan to resubmit soon) will cover the quantitative genetics.
+%In addition to the revised text, we are submitting as supplementary files,
+%a detailed response to the reviews and a color-coded diff to make it easy to see what has been removed.
+%We are happy with the result -- as predicted by the reviewers, the resulting paper stands alone
+%without the quantitative genetics results.
+%
+%We hope that by focusing this paper on system theory, we have clarified our results, which include a mathematical description of phenotypically equivalent network space 
+%and its implications for speciation under neutral genetic drift. These results suggest that the space of phenotypically equivalent network organizations is substantial,
+%and often not reproductively compatible, leaving many opportunities for independently evolving populations to become reproductively incompatible.
+%The linear system framework applied in this manuscript explicitly describes possible molecular pathways and naturally predicts Haldane's rule.
+%
+%We would like to note that the previous version of this paper as a preprint has since received 10 citations as well as attracted
+%substantial positive attention as a preprint (e.g., see tweets at
+%\url{https://www.biorxiv.org/content/10.1101/231209v2}). Furthermore, this paper has been downloaded more than 2,326 times (\url{https://rxivist.org/papers/24911}),
+%making it the 155$^{\text{th}}$ most downloaded evolutionary biology paper available on the \emph{bioRxiv}, as of this writing.
+%
 
@@
 
-\reply{\jss{To do.}
+\reply{Done. 
 }

### 2021-08-11 review-responses2.tex
corrected type-o in response to rev2

@@
 (which first discusses system theory, followed by quantitative genetics)
-is more consistent with the ordering of the rest of text.
+is more consistent with the ordering of the rest of the text.
 }

### 2021-08-11 review-responses2.tex
edits

--- 2021-08-11 review-responses2.tex
@@
 
- Cover letter.
+We are writing to submit a revision of our manuscript, ``{System drift and speciation}'' (ID 21-0262), for your review.
+We'd would like to thank you and the reviewers for your enthusiasm
+and your suggestions -- they have helped us improve our manuscript's accuracy and clarity.
+
+Both reviewers made comments about the section on Haldane's rule, which we have revised accordingly.
+We also made an effort to provide more context for our manuscript by citing the additional literature, as requested by Reviewer 1. 
+To address some of Reviewer 2's concerns, we added more discussion and detail pertaining to system theory and its biological interpretation.
+
+Line numbers in the response to reviewers refer to the revised document;
+in addition we are submitting a color-coded pdf showing differences to the previous submission
+that has its own copy of the review responses at the end (and line number references are consistent within this file as well).
+
+We hope that you enjoy the resulting manuscript as much as we do.
+
+\vskip 5em
 

### 2021-08-11 examples/Fig5_updated.R

@@ -4,6 +4,9 @@ library(gridExtra)
 library(ggplot2)
 library(ggpubr)
 library(reshape2)
+library(cowplot)
+
+pdf(file="blue_hybrids.pdf", width=6, height=4, pointsize=10)
 
 A0 <- matrix(c(0,-1,1,0), nrow=2)
 B <- matrix(c(1,1), ncol=1)
@@ -142,3 +145,5 @@ k22 <- grid.arrange(k2, top=tr.grob)
 
 k0 <- plot_grid(k11, k22, nrow = 1)
 k00 <- grid.arrange(k0, bottom=x.grob, left=y.grob)
+
+dev.off()

### 2021-08-12 examples/Fig5_updated.R
fixed fig5 script

@@ -5,8 +5,8 @@ library(ggplot2)
 library(ggpubr)
 library(reshape2)
 library(cowplot)
+library(grDevices)
 
-pdf(file="blue_hybrids.pdf", width=6, height=4, pointsize=10)
 
 A0 <- matrix(c(0,-1,1,0), nrow=2)
 B <- matrix(c(1,1), ncol=1)
@@ -82,60 +82,60 @@ hx <- data.frame("value"=hhf0[,1], "time"=tt, "variable"="parent")
 k <- data.frame(hhf1_1)
 k$time <- tt
 pl11 <- (melt(k, id.vars = "time")) %>% ggplot(aes(x=time, y=value, group=variable, col=variable)) + 
-    geom_line() + theme_classic(12) + theme(legend.position = "none") + 
+    geom_line(size=0.25) + theme_classic(12) + theme(legend.position = "none") + 
     scale_color_manual(values = cf(3)) + ylim(c(-6,6)) + xlab("") + ylab("") +
-    geom_line(data=hx, aes(x=time, y=value), color="black", size=1) + 
-    annotate(geom = "text", label=expression(paste(epsilon, " = 0.01")), x=0, y=6, hjust=0, vjust=1, size=6)
+    geom_line(data=hx, aes(x=time, y=value), color="black", size=0.5) + 
+    annotate(geom = "text", label=expression(paste(epsilon, " = 0.01")), x=0, y=6, hjust=0, vjust=1, size=3)
 
 k <- data.frame(hhf1_2)
 k$time <- tt
 pl12 <- (melt(k, id.vars = "time")) %>% ggplot(aes(x=time, y=value, group=variable, col=variable)) + 
-    geom_line() + theme_classic(12) + theme(legend.position = "none") + 
+    geom_line(size=0.25) + theme_classic(12) + theme(legend.position = "none") + 
     scale_color_manual(values = cf(3)) + ylim(c(-6,6)) + xlab("") + ylab("") + 
-    geom_line(data=hx, aes(x=time, y=value), color="black", size=1) + 
-    annotate(geom = "text", label=expression(paste(epsilon, " = 0.1")), x=0, y=6, hjust=0, vjust=1, size=6)
+    geom_line(data=hx, aes(x=time, y=value), color="black", size=0.5) + 
+    annotate(geom = "text", label=expression(paste(epsilon, " = 0.1")), x=0, y=6, hjust=0, vjust=1, size=3)
 
 k <- data.frame(hhf1_3)
 k$time <- tt
[…]

### 2021-08-12 kalman_walk_paper.tex
cover letter
edits
fixed fig5 script
maybe add an inline recombine mech?

@@
 two extremes are depicted in the two panels of Figure~\ref{fig:conceptual_fig}.) \revpoint{2}{6}
+
+\jss{Could add something like this?}
+
+In Example \ref{ex:hybrid_osc} and Figure \ref{fig:hybs}, however, meiosis is modeled such that it recombines system coefficients independently,
+to form haploid gametes.
+E.g., a possible oscillator gamete could be formed by shuffling the diploid's haploid genomes -- with one genome's coefficients notated with Greek and the other with Latin, as follows,
+$\left[ \begin{smallmatrix} \alpha & \beta \\ \gamma & \delta \end{smallmatrix} \right] \times \left[ \begin{smallmatrix} a & b \\ c & d \end{smallmatrix} \right] \to \left[ \begin{smallmatrix} \alpha & b \\ c & \delta \end{smallmatrix} \right]$.
 
@@
 \begin{figure}[H]
-  %\centering
- % \begin{tabular}{ccc}
- %   \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F1_hundreth_tau0} &
- %     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F2_hundrethtau0} \\
- %     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F1_tenth_tau0.pdf} &
- %     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F2_tenthtau0.pdf} \\
- %     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F1_half_tau0.pdf} &
- %     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F2_halftau0.pdf}
- % \end{tabular}
-  %\includegraphics[scale=0.5]{examples/f1f2_newplot}
-  \includegraphics[scale=0.5]{examples/blue_hybrids}
+  \includegraphics{examples/blue_hybrids2}
   \caption{
@@
     %%   it only matters if the offspring is homo/het/homozygous, so there are 3^4 = 81 possible expression patterns.
-    Blue lines correspond to different $F_2$ hybrids,
-    many of which show complete breakdown.
+    $F_2$ hybrids, many of which show complete breakdown, are shown in shades of blue and grey. 
   } \label{fig:hybs}

### 2021-08-12 review-responses2.tex
recomb clarification

@@
 We are writing to submit a revision of our manuscript, ``{System drift and speciation}'' (ID 21-0262), for your review.
-We'd would like to thank you and the reviewers for your enthusiasm
+We would like to thank you and the reviewers for your enthusiasm
 and your suggestions -- they have helped us improve our manuscript's accuracy and clarity.
@@
 \reply{
-    Good point; this was unclear. We have clarified that we don't pick a particular model for recombination in this section. \revref
+    Good point; this was unclear. We have clarified. \revref
 }

### 2021-08-12 kalman_walk_paper.tex
recomb clarification

@@
 $((A'+A'')/2, (B'+B'')/2, (C'+C'')/2)$.
-(We do not, however, specify a model for recombination to produce the inherited systems,
-as the details will depend on the genetic basis of the system coefficients;
-two extremes are depicted in the two panels of Figure~\ref{fig:conceptual_fig}.) \revpoint{2}{6}
-
-\jss{Could add something like this?}
-
-In Example \ref{ex:hybrid_osc} and Figure \ref{fig:hybs}, however, meiosis is modeled such that it recombines system coefficients independently,
-to form haploid gametes.
-E.g., a possible oscillator gamete could be formed by shuffling the diploid's haploid genomes -- with one genome's coefficients notated with Greek and the other with Latin, as follows,
+In general, how parental systems are recombined and inherited
+depends on the genetic architecture of the segregating variation.
+\revpoint{2}{6}
+For the purposes of examples in this paper
+(Example \ref{ex:hybrid_osc} and Figure \ref{fig:hybs}),
+we take the simplest assumption that each system coefficient is encoded by a nonrecombining locus,
+and that these loci are unlinked.
+For instance, if
+one genome's coefficients are written with Greek and the other with Latin,
+a possible oscillator gamete might be produced as follows:
 $\left[ \begin{smallmatrix} \alpha & \beta \\ \gamma & \delta \end{smallmatrix} \right] \times \left[ \begin{smallmatrix} a & b \\ c & d \end{smallmatrix} \right] \to \left[ \begin{smallmatrix} \alpha & b \\ c & \delta \end{smallmatrix} \right]$.

### 2021-08-16 review-responses2.tex
type-o in cover letter
edits for submission

@@
 
-\begin{minipage}[b]{2.5in}
-  Revision Cover Letter \\
-  {\it Evolution}
-\end{minipage}
-\hfill
-\begin{minipage}[b]{2.5in}
-    Joshua Schiffman \\
-    \emph{and} Peter Ralph \\
-  \today
-\end{minipage}
-
-\vskip 2em
-
-\noindent
-{\bf To the Editor(s) -- }
-
-\vskip 1em
-
-We are writing to submit a revision of our manuscript, ``{System drift and speciation}'' (ID 21-0262), for your review.
-We would like to thank you and the reviewers for your enthusiasm
-and your suggestions -- they have helped us improve our manuscript's accuracy and clarity.
-
-Both reviewers made comments about the section on Haldane's rule, which we have revised accordingly.
-We also made an effort to provide more context for our manuscript by citing the additional literature, as requested by Reviewer 1. 
-To address some of Reviewer 2's concerns, we added more discussion and detail pertaining to system theory and its biological interpretation.
-
-Line numbers in the response to reviewers refer to the revised document;
-in addition we are submitting a color-coded pdf showing differences to the previous submission
-that has its own copy of the review responses at the end (and line number references are consistent within this file as well).
-
-We hope that you enjoy the resulting manuscript as much as we do.
-
-\vskip 5em
-
-\noindent \hspace{4em}
-\begin{minipage}{3in}
-\noindent
-{\bf Sincerely,}
-
-\vskip 2em
-
-{\bf 
-Joshua Schiffman and
-Peter Ralph
-}\\
-\end{minipage}
-
-\vskip 4em
-
-\pagebreak
-
+%\begin{minipage}[b]{2.5in}
+%  Revision Cover Letter \\
+%  {\it Evolution}
+%\end{minipage}
+%\hfill
+%\begin{minipage}[b]{2.5in}
+%    Joshua Schiffman \\
+%    \emph{and} Peter Ralph \\
+%  \today
+%\end{minipage}
+% 
+%\vskip 2em
+% 
+%\noindent
+%{\bf To the Editor(s) -- }
+% 
+%\vskip 1em
+%
+%We are writing to submit a revision of our manuscript, ``{System drift and speciation}'' (ID 21-0262), for your review.
+%We would like to thank you and the reviewers for your enthusiasm
+%and your suggestions -- they have helped us improve our manuscript's accuracy and clarity.
+%
+%Both reviewers made comments about the section on Haldane's rule, which we have revised accordingly.
+%We also made an effort to provide more context for our manuscript by citing the additional literature, as requested by Reviewer 1. 
+%To address some of Reviewer 2's concerns, we added more discussion and detail pertaining to system theory and its biological interpretation.
+%
+%Line numbers in the response to reviewers refer to the revised document;
+%in addition we are submitting a color-coded pdf showing differences to the previous submission
+%that has its own copy of the review responses at the end (and line number references are consistent within this file as well).
+%
+%We hope that you enjoy the resulting manuscript as much as we do.
+% 
+%
+%\vskip 5em
+%
+%\noindent \hspace{4em}
+%\begin{minipage}{3in}
+%\noindent
+%{\bf Sincerely,}
+%
+%\vskip 2em
+%
+%{\bf 
+%Joshua Schiffman and
+%Peter Ralph
+%}\\
+%\end{minipage}
+%
+%\vskip 4em
+%
+%\pagebreak
+%
 %%%%%%%%%%%%%%

### 2021-08-16 kalman_walk_paper.tex
edits for submission

@@
 
-\section*{Acknowledgements}
-
-We would like to thank Sergey Nuzhdin, Stevan Arnold, Michael Turelli, Patrick Phillips, Erik Lundgren and Hossein Asgharian for valuable discussion. 
-We would also like to thank Nick Barton, Sarah Signor, Todd Parsons, and Joachim Hermisson for very helpful comments on the manuscript.
-Work on this project was supported by funds from
-the Sloan Foundation and the NSF (under DBI-1262645) to PR.
+%\section*{Acknowledgements}
+%
+%We would like to thank Sergey Nuzhdin, Stevan Arnold, Michael Turelli, Patrick Phillips, Erik Lundgren and Hossein Asgharian for valuable discussion. 
+%We would also like to thank Nick Barton, Sarah Signor, Todd Parsons, and Joachim Hermisson for very helpful comments on the manuscript.
+%Work on this project was supported by funds from
+%the Sloan Foundation and the NSF (under DBI-1262645) to PR.
+%
 

### 2021-08-16 title_page_revision.tex
title page

@@
 This model also naturally predicts Haldane's rule, 
-thus providing another possible explanation
+thus providing a concrete explanation
 of why heterogametic hybrids tend to be disrupted more often than homogametes 

### 2021-09-02 code/oscillator_plots.Rmd
edits for final version

---
title: "System drift and speciation in an oscillator"
author: "Joshua S. Schiffman and Peter L. Ralph"
date: "9/1/2021"
output: html_document
---

This code reproduces plots for panels in Figures 1, 2, 5 and 6 in _System drift and speciation_.

```{r, echo=FALSE}
library(expm)
library(parallel)
library(grid)
library(gridExtra)
library(ggplot2)
library(ggpubr)
library(reshape2)
library(cowplot)
library(grDevices)
```

```{r}
# Oscillator code

# A0, B and C correspond to the matrices of system coefficients described in Example 1.
A0 <- matrix(c(0,-1,1,0), nrow=2)
B <- matrix(c(1,1), ncol=1)
C <- matrix(c(1,0), nrow=1)

# A_fn corresponds to the parameterization of phenotypically equivalent oscillators described in Example 2.
A_fn <- function (tau) { 
    matrix(c(tau/(tau+1), -(2*tau*(tau+1)+1)/(tau+1), 
             1/(tau+1), -tau/(tau+1)), nrow=2) 
}

tt <- seq(0,6*2*pi, length.out=500)

# h is the impulse response of A_fn(tau)
h <- function (t,tau=0,A=A_fn(tau)) {
    sapply(t, function (tt) C %*% expm(tt*A) %*% B) 
    }

# The 'F1' function averages two haploid systems to form a diploid F1 system.
F1 <- function (tau1,tau2) { (A_fn(tau1)+A_fn(tau2))/2 }

# Generates F2 hybrid systems.
get_F2s <- function(tau1, tau2) 
    {
        vc <- function(tau1, tau2, x, y) 
            {
                out <- c(A_fn(tau1)[x,y], A_fn(tau2)[x,y])
                return(out)
            }
        mlist <- list()
        count <- 1
        for (i in 1:2) 
            {
                for (j in 1:2) 
                    {
                        for (k in 1:2) 
                            {
                                for (l in 1:2)
                                    {
                                        mlist[[count]] <- matrix(c(vc(tau1, tau2, 1, 1)[i], vc(tau1, tau2, 2, 1)[j], vc(tau1, tau2, 1, 2)[k], vc(tau1, tau2, 2, 2)[l]),2,2)
count <- count + 1
                                    }
                            }
                }
            }
        f2list <- list()
        count2 <- 1
        for(i in 1:length(mlist)) 
            {
                for (j in 1:length(mlist))
                    {
                        f2list[[count2]] <- (mlist[[i]] + mlist[[j]])/2
                        count2 <- count2 + 1
                    }
            }
    return(unique(f2list))
}
```

```{r fig1and2, fig.width=2.5, fig.height=1.5, fig.align="center"}

# Plots shown in Figures 1 and 2.

# h.2 function used to visualize kryptotypic dynamics.
C2 <- matrix(c(0,1), nrow=1)
h.2 <- function (t,tau=0,A=A_fn(tau)) { 
    sapply(t, function (tt) C2 %*% expm(tt*A) %*% B)
    }

# Phenotype
hhf00 <- sapply( list(A_fn(0)), function (A) { sapply(tt, h, A=A) } )

# Kryptotypes
hh02 <- sapply( list(A_fn(0)), function (A) { sapply(tt, h.2, A=A) } )
hhn22 <- sapply( list(A_fn(-2)), function (A) { sapply(tt, h.2, A=A) } )

# Phenotypes shown in Figure 1.
#pdf(file="Figure1_phenotype_plot.pdf", width=2.5, height=1.5, pointsize=10)

par(mar=c(4.5, 4, 1, 1)+.1)
plot(tt, hhf00, type="l", lwd=2, ylab= expression("phenotype" ~ phi(t)), xlab="time t", col="steelblue4", ylim=c(-3,3), cex.lab=1)

#dev.off()

# Kryptotypes shown in Figure 2.
#pdf(file="Figure2_kryptotype_plot.pdf", width=2.5, height=1.5, pointsize=10)

par(mar=c(4.5, 4, 1, 1)+.1)
plot(tt, hhf00, type="l", lwd=2, ylab= expression("kryptotype" ~ kappa(t)), xlab="time t", col="steelblue4", ylim=c(-3,3), cex.lab=1)
lines(tt, hhn22, lwd=2, col="seagreen")
lines(tt, hh02, lwd=2, col="red4")

#dev.off()
```

```{r fig5, fig.align="center", fig.width=14, fig.height=10}

# Parental phenotypes shown in Figure 5.
hhf0 <- sapply( list(A_fn(0), A_fn(0)), function (A) { sapply(tt, h, A=A) } )

# F1 phenotypes between A_fn(tau=0) and A_fn(tau=epsilon) shown in Figure 5.
hhf1_1 <- sapply( list(A_fn(0), A_fn(0.01), F1(0, 0.01)), function (A) { sapply(tt, h, A=A) } )
hhf1_2 <- sapply( list(A_fn(0), A_fn(0.1), F1(0, 0.1)), function (A) { sapply(tt, h, A=A) } )
hhf1_3 <- sapply( list(A_fn(0), A_fn(0.5), F1(0, 0.5)), function (A) { sapply(tt, h, A=A) } )

# F2 phenotypes between A_fn(tau=0) and A_fn(tau=epsilon) shown in Figure 5.
hhf2_1 <- sapply( get_F2s(0, 0.01), function (A) { sapply(tt, h, A=A) } )
hhf2_2 <- sapply( get_F2s(0, 0.1), function (A) { sapply(tt, h, A=A) } )
hhf2_3 <- sapply( get_F2s(0, 0.5), function (A) { sapply(tt, h, A=A) } )

```

```{r fig5plots, fig.width=14, fig.height=10}

# Figure 5 plots

cf <- colorRampPalette(colors = c("grey", "steelblue"))
hx <- data.frame("value"=hhf0[,1], "time"=tt, "variable"="parent")

# F1s

k <- data.frame(hhf1_1)
k$time <- tt
pl11 <- (melt(k, id.vars = "time")) %>% ggplot(aes(x=time, y=value, group=variable, col=variable)) + 
    geom_line(size=0.25) + theme_classic(12) + theme(legend.position = "none") + 
    scale_color_manual(values = cf(3)) + ylim(c(-6,6)) + xlab("") + ylab("") +
    geom_line(data=hx, aes(x=time, y=value), color="black", size=0.5) + 
    annotate(geom = "text", label=expression(paste(epsilon, " = 0.01")), x=0, y=6, hjust=0, vjust=1, size=3)

k <- data.frame(hhf1_2)
k$time <- tt
pl12 <- (melt(k, id.vars = "time")) %>% ggplot(aes(x=time, y=value, group=variable, col=variable)) + 
    geom_line(size=0.25) + theme_classic(12) + theme(legend.position = "none") + 
    scale_color_manual(values = cf(3)) + ylim(c(-6,6)) + xlab("") + ylab("") + 
    geom_line(data=hx, aes(x=time, y=value), color="black", size=0.5) + 
    annotate(geom = "text", label=expression(paste(epsilon, " = 0.1")), x=0, y=6, hjust=0, vjust=1, size=3)

k <- data.frame(hhf1_3)
k$time <- tt
pl13 <- (melt(k, id.vars = "time")) %>% ggplot(aes(x=time, y=value, group=variable, col=variable)) + 
    geom_line(size=0.25) + theme_classic(12) + theme(legend.position = "none") + 
    scale_color_manual(values = cf(3)) + ylim(c(-6,6)) + xlab("") + ylab("") +
    geom_line(data=hx, aes(x=time, y=value), color="black", size=0.5) + 
    annotate(geom = "text", label=expression(paste(epsilon, " = 0.5")), x=0, y=6, hjust=0, vjust=1, size=3)

# F2s 
k <- data.frame(hhf2_1)
k$time <- tt
pl21 <- (melt(k, id.vars = "time")) %>% ggplot(aes(x=time, y=value, group=variable, col=variable)) + 
    geom_line(size=0.25) + theme_classic(12) + theme(legend.position = "none") + 
    scale_color_manual(values = cf(81)) + ylim(c(-6,6)) + xlab("") + ylab("") +
    geom_line(data=hx, aes(x=time, y=value), color="black", size=0.5) + 
    annotate(geom = "text", label=expression(paste(epsilon, " = 0.01")), x=0, y=6, hjust=0, vjust=1, size=3)

k <- data.frame(hhf2_2)
k$time <- tt
pl22 <- (melt(k, id.vars = "time")) %>% ggplot(aes(x=time, y=value, group=variable, col=variable)) + 
    geom_line(size=0.25) + theme_classic(12) + theme(legend.position = "none") + 
    scale_color_manual(values = cf(81)) + ylim(c(-6,6)) + xlab("") + ylab("") +
    geom_line(data=hx, aes(x=time, y=value), color="black", size=0.5) + 
    annotate(geom = "text", label=expression(paste(epsilon, " = 0.1")), x=0, y=6, hjust=0, vjust=1, size=3)

k <- data.frame(hhf2_3)
k$time <- tt
pl23 <- (melt(k, id.vars = "time")) %>% ggplot(aes(x=time, y=value, group=variable, col=variable)) + 
    geom_line(size=0.25) + theme_classic(12) + theme(legend.position = "none") + 
    scale_color_manual(values = cf(81)) + ylim(c(-6,6)) + xlab("") + ylab("") +
    geom_line(data=hx, aes(x=time, y=value), color="black", size=0.5) + 
    annotate(geom = "text", label=expression(paste(epsilon, " = 0.5")), x=0, y=6, hjust=0, vjust=1, size=3)

tr.grob <- textGrob(bquote(F[2]), 
                    gp=gpar(fontface="bold", col="black", fontsize=16))
tl.grob <- textGrob(bquote(F[1]), 
                    gp=gpar(fontface="bold", col="black", fontsize=16))
x.grob <- textGrob("time", 
                   gp=gpar(col="black", fontsize=12))
y.grob <- textGrob("phenotype", 
                   gp=gpar(col="black", fontsize=12), rot = 90)

k1 <- plot_grid(pl11, pl12, pl13, ncol=1)
k2 <- plot_grid(pl21, pl22, pl23, ncol=1)

k11 <- arrangeGrob(k1, top=tl.grob)
k22 <- arrangeGrob(k2, top=tr.grob)

k0 <- plot_grid(k11, k22, nrow = 1)

k00 <- arrangeGrob(k0, bottom=x.grob, left=y.grob)
out5 <- ggdraw(k00)
out5

#ggsave(out5, width=14, height=10, dpi=1000, filename="Figure5_hybrid_oscillators.pdf", scale=0.5)

```

```{r fig6, fig.align="center", fig.width=5, fig.height=3, cache=TRUE}

# Code for Figure 6

# Optimal impulse response function for the oscillator
optimal_h <- function (t) { 
    sin(t) + cos(t) 
    }

# Weighted difference between h(t,A) and optimal; square of equation (5)
Df <- function (A,t) {
    exp(-t/(4*pi)) * ( h(t,A=A) - optimal_h(t) )^2
}

D <- function (A, upper=10, ...) {
    f <- function (t) { Df(A,t) }
    integrate(f, lower=0, upper=upper, ...)$value
}

# Get all F2s.
get_F2s_nonunique <- function(tau1, tau2) {
    vc <- function(tau1, tau2, x, y) {
    out <- c(A_fn(tau1)[x,y], A_fn(tau2)[x,y])
    return(out)
    }

    mlist <- list()
    count <- 1
    for (i in 1:2) {
        for (j in 1:2) {
            for (k in 1:2)
                for (l in 1:2) {
                    mlist[[count]] <- matrix(c(vc(tau1, tau2, 1, 1)[i], 
                                               vc(tau1, tau2, 2, 1)[j], 
                                               vc(tau1, tau2, 1, 2)[k], 
                                               vc(tau1, tau2, 2, 2)[l]),
                                             2,2)
                    count <- count + 1
                }
        }
    }

    f2list <- list()
    count2 <- 1

    for(i in 1:length(mlist)) {
        for (j in 1:length(mlist)) {
            f2list[[count2]] <- (mlist[[i]] + mlist[[j]])/2
            count2 <- count2 + 1
        }
    }
    return(f2list)
}

# Compute genetic distance as described in the caption of Figure 6.
sys_gen_dist <- function(d) {
    sgd <- sqrt( sum( (A_fn(0) - A_fn(0+d))^2 ) )
    return(sgd)
}

dvec <- seq(0, 0.5, length.out=20)

D_F1 <- sapply(dvec, function (dtau) {D(A=F1(tau1=0, tau2=0+dtau), upper = 10 )} )

D_F2 <- do.call(cbind, mclapply(dvec, function (dtau) {
F2s <- get_F2s_nonunique(0, 0+dtau)
sapply(F2s, D, upper=10 )
}, mc.cores=detectCores() ) )

#pdf(file="Figure6_plot.pdf", width=5, height=3, pointsize=10)

plot( sapply(dvec, sys_gen_dist), sqrt(D_F1), type='l', ylab="phenotype distance (D)", xlab="genetic distance", ylim=c(0, 2), lwd=3)
lines(sapply(dvec, sys_gen_dist), colMeans(sqrt(D_F2)), lty=1, type="l", ylim=c(0,2), col="steelblue4", lwd=3)
legend(0, 2, legend = c( expression(F[1]~"hybrids"), expression(F[2]~"hybrids")), lty=1, lwd=3, col=c("black", "steelblue4"))

#dev.off()

```

### 2021-09-02 final_system_drift.tex
edits for final version

--- 2021-08-12 kalman_walk_paper.tex
@@
 % Responses to reviews
-\input{review-response-commands}
+%\input{review-response-commands}
 % set this to show line numbers and include responses to reviews or not
-\newif\ifreviewresponses
-\reviewresponsestrue  % include them
+%\newif\ifreviewresponses
+%\reviewresponsestrue  % include them
 % \reviewresponsesfalse  % don't include them
-\newcommand{\responsefile}{review-responses2}  % name of the review responses file
+%\newcommand{\responsefile}{review-responses2}  % name of the review responses file
 
@@
 {\Huge \bf System drift and speciation} \\ \vspace{0.75cm}
-%Joshua S.
+Joshua S.
 Schiffman$^{\dagger}$ \footnote{Current affiliations: New York Genome Center, New York, New York 10013, U.S.A.
@@
 Ralph$^{\dagger \ddagger}$ \\ \vspace{0.5cm}
-%$^{\dagger}${\footnotesize {Molecular and Computational Biology, University of Southern California, Los Angeles, California 90089, U.S.A. \\
-%$^{\ddagger}$Departments of Mathematics and Biology \& The Institute for Ecology and Evolution, University of Oregon, Eugene, Oregon 97403, U.S.A.}} \\ \vspace{0.5cm}
-%{\small \texttt{jschiffman@nygenome.org} \qquad 
-%\texttt{plr@uoregon.edu}} \\ \vspace{0.5cm}
+$^{\dagger}${\footnotesize {Molecular and Computational Biology, University of Southern California, Los Angeles, California 90089, U.S.A. \\
+$^{\ddagger}$Departments of Mathematics and Biology \& The Institute for Ecology and Evolution, University of Oregon, Eugene, Oregon 97403, U.S.A.}} \\ \vspace{0.5cm}
+{\small \texttt{jschiffman@nygenome.org} \qquad 
+\texttt{plr@uoregon.edu}} \\ \vspace{0.5cm}
 \small \today \\
@@
 as the genetic drift of small changes to these molecular pathways
-cause a population to explore the set of mechanisms with identical phenotypes.
+causes a population to explore the set of mechanisms with identical phenotypes.
 To do this, we model an organism's internal state as a linear system of differential equations
@@
 This model also naturally predicts Haldane's rule, 
-%thus providing another possible explanation
 thus providing a concrete explanation
@@
 %%%%%%%%%%%%%%%%%%%%%%
-%\begin{multicols}{2}
 \section*{Introduction}
@@
 systems can be structurally different yet remain functionally equivalent \citep{edelman2001degeneracy}. 
-Examples can be found across nearly all levels of biological organization from the level of the genetic code itself % \citep{crick1963recent}
+Examples can be found across nearly all levels of biological organization from the level of the genetic code itself 
 all the way up to the convergent evolution of adaptive traits.
@@
 and neural circuits \citep{trojanowski2014neural}, and in developmental systems \citep{vondassow2000segment,true2001developmental}.
-% \jss{Other ideas instead of ``degeneracy'' or ``isofunctional structural plasticity'': non-invertibility, isofunctional plasticity, structural plasticity, phenotypic equivalency, phenotypic invariance, functional invariance, isofunctionality, functional equivalency, genotypic plasticity, conjugacy, functional symmetry, isologous, isomorphic.}
-% \plr{'structural plasticity' means something specific in neuroscience}
 
@@
 
-In this paper we use results on mathematical nonidentifiability from \llabel{r21}
+In this paper we use results on mathematical nonidentifiability from %\llabel{r21}
 linear systems theory to study how gene regulatory networks can be modified
 while retaining the same function, and the possible implications for speciation.
-% Specifically, we would like to understand the degree of which gene regulatory network architectures are functionally equivalent and how this relates to network size and efficiency.
 If system architectures are not functionally unique,
@@
 
-% It is not a new observation
 The field of population genetics has also explored the consequences of the fact
-%that phenotype does not prescribe a unique genotype
-that there is often more than one way to do the same thing, \revpoint{1}{2}
-\llabel{r22}
+that there is often more than one way to do the same thing, %\revpoint{1}{2}
+%\llabel{r22}
 and observed that speciation might be the result of changes that are themselves neutral.
@@
 and for pairs of traits that must match but whose value is unconstrained \citep{sved1981twosex}.
-% Recent work has shown that Fisher's geometric model of stabilizing selection
-% predicts many general patterns from the speciation literature \citep{fraisse2016genetics}.
 It has also been shown that population structure
@@
 but only equivocal support for speciation under models of purely neutral drift.
-%Additionally, \citet{fierst} observed speciation due to an increase of segregation variance under balancing selection, in models of weak to moderate, but not strong epistasis.
 The rate at which hybrid incompatibility accumulates due to genetic drift creating
@@
 -- such as transcription factors -- as well as external or environmental inputs.
-% We model the temporal dynamics of the concentrations of a collection of $n$ coregulating molecules
-% within an organism, that may also be affected by temporally varying signals from the environment.
-%% Organisms' phenotypes are constructed by gene by gene and gene by environment interactions. 
-%%Here we simply define the \emph{phenotype} to be those aspects of the dynamics directly under natural selection
-%%-- the \emph{what}, \emph{when}, and \emph{how much}, of an organism's molecules that are physiologically or otherwise relevant to survival.
 We write $\kappa(t)$ for the vector of $n$ molecular concentrations at time $t$.
@@
 and a matrix $B$ that encodes the effect of each entry of $u$ on the elements of the kryptotype.
-%Thus an organism's phenotype $\phi(t)$ -- a vector of molecular concentrations at time $t$ -- is determined both by the structure and organization of a biological system (\emph{e.g.} a gene regulatory network), given by the triple $(A,B,C)$,
-%and by its environment $u(t)$.
 The rate at which the $i^\text{th}$ concentration changes
@@
 and $C$ might determine the rate of production of downstream enzymes.
-%% {Linearity} also ``zero-state equivalence''
-% Here we have assumed that the system is linear,
-% and begins from the ``zero'' state ($\kappa(0)=0$).
-% Of course, neither of these are necessarily true for real systems,
-% but the dynamics of most nonlinear systems can be approximated locally by a linear systems near most points. 
-% Furthermore, the ease of analyzing linear systems makes this an attractive place to start.
 
@@
 
-\begin{example}[An oscillator]\label{ex:oscillator} \llabel{r11b}
+\begin{example}[An oscillator]\label{ex:oscillator} %\llabel{r11b}
   For illustration, we consider an extremely simplified model of oscillating gene transcription,
 as for instance is found in cell cycle control or the circadian rhythm.
-% Cellular division is governed by many different processes, however it is thought that its rhythm is partially controlled by oscillating gene transcription \citep{orlando2008global}.
 There are two genes, 
@@
 and $C \!=\! \left[\begin{smallmatrix} 1 & 0 \end{smallmatrix}\right]$.
-  %  \begin{align*}
-  %    A = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} , \qquad B = \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix}
-  %  \end{align*}
- %       The oscillatory system $\Sigma$ is thus given as
- %   \begin{align*}
- %     \Sigma = \left \{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 
- %       0 & 1 \\ 
- %      -1 & 0 
- %       \end{bmatrix} \kappa(t) + \begin{bmatrix} 1 \\ 1 \end{bmatrix} u(t) \\ 
- %         \phi(t) &= \begin{bmatrix} 1 & 0 \end{bmatrix} \kappa(t) \end{array} \right.
- %    \end{align*}
 If the input is an impulse at time zero (a delta function), 
@@
 \begin{figure}[H]
-   % \begin{center} 
   \centering
@@
             \node[shape=rectangle] (y) at (2,-2) {output ($\phi$)};
-            %\\  \includegraphics[width=0.15\textwidth, height=0.04\paperheight]{osc_impulse} \end{tabular}};
         \end{scope}
@@
                       every edge/.style={draw=black, thick}]
-            %\path [->] (A) edge[loop left] node {\tiny $\lambda_{1}$} (A);
-            %\path [->] (B) edge[loop left] node {\tiny $\lambda_{2}$} (B);
         \end{scope}
@@
 \end{tabular}
-%  \end{center}
   \caption{
@@
 \end{align*}
-% (recall that $\kappa(0) = \bar \kappa(0) = 0$). 
 In other words, phenotypically equivalent systems respond identically for \emph{any} input.
@@
 moving $A$ in the direction of $ZA - AZ$
-% (e.g., $\bar A = A + \epsilon(ZA-AZ)$)
 while also moving $B$ in the direction of $ZB$ 
@@
 will leave the phenotype unchanged to second order in the size of the perturbation.
-%% I believe the following is necessary and sufficient:
 If the columns of $B$ and the rows of $C$ are not all eigenvectors of $A$,
@@
 to achieve the desired dynamics.
-Results in system theory show that any time-invariant \revpoint{2}{5}
+Results in system theory show that any time-invariant %\revpoint{2}{5}
 system can be realized in a particular minimal dimension
@@
 explicitly describe the set of all phenotypically equivalent systems.
-% Could define script S here for 'system' if we want: $\Sys$
 We refer to $\allS(A_0,B_0,C_0)$ as the set of all systems phenotypically equivalent
@@
 and so can only infer motion in what is known as the \emph{observable subspace}.
-The Kalman decomposition then classifies each direction in kryptotype space\llabel{r23}
+The Kalman decomposition then classifies each direction in kryptotype space %\llabel{r23}
 as either reachable or unreachable, and as either observable or unobservable.
@@
 that is, components that both respond to an input and produce an observable output.
-% it is possible to add components that respond to an input, 
-% but do not influence the output, or components that, in principle, influence the output, but do not respond to any inputs.
 
@@
 It is remarkable that the gene regulatory network architecture to achieve a given input--output map is never unique --
-both the change of basis used to obtain the decomposition\llabel{r24}
+both the change of basis used to obtain the decomposition %\llabel{r24}
 and, once in this form, all submatrices other than $A_{\ro}$, $B_{\ro}$, and $C_{\ro}$ can be changed without affecting the phenotype,
@@
 through the impulse response function $h(t)$,
-they may affect individual fitness in other ways \revpoint{2}{4}
+they may affect individual fitness in other ways %\revpoint{2}{4}
 (which we don't consider here).
@@
 The \emph{reachable subspace}
-% which we denote by $\reachable$,
 is defined to be the closure of $\spn(B)$ under applying $A$
 (or equivalently, the span of $B, AB, A^2B, \ldots A^{n-1}B$), 
-% Analogously to this, we define
 and the \emph{unobservable subspace}
-%$\mathcal{O}$, is the closure of $\spn(C^T)$ under applying $A$.
-% denoted $\unobservable$,
 is the largest $A$-invariant subspace
 contained in the null space of $C$.
-% and $\reachable$ is the largest $A$-invariant subspace contained in the image of $B$.)
 The four subspaces, $\rno$, $\ro$, $\nrno$, and $\nro$
@@
 and similarly for $\nro$ and $\rno$.
-
-%Usually, the dimension $n$ and the reference system $\Sys_0$ is implicit and we write only $\calA$.
-%Further, we refer to the set of all linear systems with impulse response $h(t)$, regardless of dimension, to be $\allS$,
-%  \begin{align}
-%    \allS := \bigcup_{n=\min}^{\infty} \calA_n(\Sys_0)
-%  \end{align}
 
@@
       (Right) Gene-1 dynamics (blue) for both systems $A(0)$ and $A(-2)$ are identical, however, $A(0)$ gene-2 dynamics (red) differ from $A(-2)$ (green).
-      % Both gene-1 dynamics are given by $\kappa_{1}(t) = \sin t + \cos t$, and gene-2 by $\kappa_{2}(t) = \cos t - \sin t$ ($A(0)$) and $\kappa'_{2}(t) = \cos t + 3 \sin t$ ($A(2)$).
       } 
     \label{fig:all_osc}
-\end{figure} \llabel{r11e}
+\end{figure} %\llabel{r11e}
 
@@
     \path [->, sloped] (A) edge[bend left] node {\tiny $2.2$} (B);
-   % \path [->, sloped] (B) edge[bend left] node {\tiny ?} (A); 
     \path [->, sloped] (A) edge[bend right] node {\tiny $2$} (C); 
-    %\path [->, sloped] (B) edge[bend right] node {\tiny ?} (C); 
     \path [->, sloped] (C) edge[bend right] node {\tiny $-1$} (A); 
@@
 
-\paragraph{Sexual reproduction and recombination}\llabel{r26}
+\paragraph{Sexual reproduction and recombination} %\llabel{r26}
 Parents with phenotypically equivalent yet differently wired gene networks may produce offspring with dramatically different phenotypes. 
@@
 depends on the genetic architecture of the segregating variation.
-\revpoint{2}{6}
+%\revpoint{2}{6}
 For the purposes of examples in this paper
@@
 is proportional to the difference between the two parental systems.
-% This is certainly true if each coefficient is determined by a single nonrecombining locus,
-% so that each coefficient in the system produced by meiosis is an independent random choice
-% between the two parental coefficients.
-% On the other hand, if a coefficient is determined by additive contributions of independently segregating loci,
-% then the law of total variance applied after conditioning on the parental origin of each allele
-% implies that the variance is equal to one-quarter of the squared difference between parental coefficients.
 
@@
 of the optimal set $\allS$ relative to segregating genetic variation.
-% in Figure \ref{fig:conceptual_fig}, this is depicted as the angle of the black line (the optimal set) with respect to the coordinates.
 
@@
     with parental phenotype again in black.
-    %% PLR: There are at most 128 possible F2s, not 256; but many are equivalent; 
-    %%   it only matters if the offspring is homo/het/homozygous, so there are 3^4 = 81 possible expression patterns.
     $F_2$ hybrids, many of which show complete breakdown, are shown in shades of blue and grey. 
@@
     \centering
-    %\includegraphics[width=0.5\textwidth]{examples/F2_vs_F1_divergence_tau0}
     \includegraphics{examples/fig6_small_maxt10}
@@
 
-\paragraph{Haldane's rule}\llabel{r15b}
+\paragraph{Haldane's rule} %\llabel{r15b}
 This model naturally predicts
@@
 are effectively equivalent to those of purely autosomal $F_2$ hybrids,
-which are less fit on average than $F_1$s (see Figure \ref{fig:osc_incompat}). \llabel{r15e}
-\revpoint{1}{4} \revpoint{2}{1}
+which are less fit on average than $F_1$s (see Figure \ref{fig:osc_incompat}). %\llabel{r15e}
+%\revpoint{1}{4} \revpoint{2}{1}
 Note, however, that presence of genetic variation on sex chromosomes
@@
 What is known about the key quantity above, the amount of heritable variation in real regulatory networks?
-%These are known at best only roughly \citep{felsenstein1988phylogenies},
-%so we aim for order-of-magnitude estimates.
 The coefficient $A_{ij}$ from the system \eqref{eqn:system} measures how much the rate of net production of $i$ changes
@@
 and single variants may affect many entries of $\allS$ simultaneously.
-% (Recall that although these are best modeled through nonlinear terms,
-% by linearizing we essentially consider first-order effects.)
-% On the other hand, a diverse set of buffering mechanisms are thought to contribute to phenotypic stability
-% in the presence of substantial molecular noise \citep{canalization,buffering},
-% suggesting that substantial variation in the micro-scale dynamics we consider here
-% may be necessary to produce relevant phenotypic effects downstream.
 
@@
 Nonetheless, these measures likely give us an idea of the scale of variability.
-%\citet{kasowski2010variation} found 
 It has been shown that between human individuals, there is differential occupancy in 7.5\%
 of binding sites of a transcription factor (p65) \citep{kasowski2010variation}.
-%\citet{verlaan2009targeted} showed that 
 It has also been inferred that cis-regulatory variation
 accounts for around 2--6\% of expression variation in human blood-derived primary cells \citep{verlaan2009targeted}, and
-%while \citet{lappalainen2013transcriptome} found 
 that human population variation 
@@
 Allele-specific expression is indicative of standing genetic \emph{cis}-regulatory variation;
-%\citet{wang2017bayesian} found 
 allele-specific expression in 7.2--8.5\% of transcripts of a flycatcher species has been observed \citep{wang2017bayesian}, as well as
-%while \citet{tung2015genetic} observed 
 allele-specific expression in 23.4\% of genes studied in a baboon species \citep{tung2015genetic}. 
@@
 \section*{Discussion}
-
-%%% OUTLINE
-% - Summarize methods and main result on rate of speciation.
-% - Consistent with F1/F2 divergence and Haldane's rule
-% - What empirical evidence is there that this is actually happening? 
-%   * incompats due to misregulation (Noor, Witkopp...)
-%   * numerous species that are visually indistinguishable (skinks, ferns)
-%   * maybe also ref phylogenetic lit
-% - What other evidence could be collected?  Note could be just phenotype not fitness. (cite Matute)
-% - Is linearity an important assumption?
-% - We know that divergent selection and genetic conflict can easily lead to speciation; we've shown neutral drift can maybe surprisingly easily do so also; but it remains unclear the relative contributions of these.
-Note that although drift may be weaker in any one network there are lots of networks.
-Could motivate this with Orr's statement.
-Perhaps tie in spandrels.
 
@@
 gives an analytical description of all phenotypically equivalent gene networks.
-%It also implies that nearly all systems are nonidentifiable,
-%and that, in general, there exist axes of genetic variation unconstrained by natural selection.
 This description shows that the space of functionally equivalent network architectures increases with the square of a network's size,
@@
 
-Consistent with empirical observation of hybrid breakdown \llabel{r16}
+Consistent with empirical observation of hybrid breakdown %\llabel{r16}
 we see that the fitnesses of $F_2$ hybrids drop at a much faster rate than those of $F_1$s.
-% since $F_1$ and $F_2$ phenotypes diverge linearly and at the square root of time relative to parentals, respectively.
-% In this model, this is due to recombination shuffling system architectures, leaving the $F_2$ system homozygous for one parent at one particular locus and homozygous for the other parent at another possibly incompatible locus.
 Another natural consequence of the model is Haldane's rule,
@@
 Furthermore, in cryptic species complexes (e.g., sun skinks \citep{barley2013challenge}),
-%strongly reproductively incompatible species
 genetically distinct species
@@
 \paragraph{The origin of species not by means of natural selection?}
-% this is neutral, in contrast to other models. although some folks advocate nonadaptive speciation.
-% other mechanisms also probably important
-% but this one may be strong since happens in all networks at once
 As classically formulated, 
@@
 However, previous authors have argued that neutral processes are likely too slow to be a significant driver of speciation \citep{nei1983models,seehausen2014genomics}.
-% Using simulations, \citet{porter2002speciation} demonstrated the accumulation of hybrid incompatibilities under directional, but not stabilizing selection, and \citet{palmer2009dynamics} observed the appearance of incompatibilities in populations evolving in a constant environment, only if the populations were suboptimally adapted.
-% In light of the few known incompatibilities, 
 This has led some to conclude that hybrid incompatibility is typically a byproduct of positive selection \citep{orr2004speciation, schluter2009evidence}
@@
 through induced population bottlenecks
-found no reproductive isolation in the majority of tested pairs \citep{fry2009laboratory, white2020past}. \revpoint{1}{1}
-%and incompatibilities identified between \emph{Drosophila simulans} and \emph{D.~melanogaster} exhibit evidence of adaptive evolution \citep{tang2009evolution}.
+found no reproductive isolation in the majority of tested pairs \citep{fry2009laboratory, white2020past}. %\revpoint{1}{1}
 However, our calculations suggest that even under strictly neutral processes,
@@
 such as the relationship between molecular divergence and genetic isolation 
-% across diverse taxa and ecologies 
 seen by \citet{roux2016shedding}, 
 and the clocklike speciation rates observed by \citet{hedges2015tree}.
-% (as diversification is not observed to be accelerated following mass extinctions).
 More recently, \citet{sanchez2021widespread} observed widespread transcriptional misregulation in \emph{Drosophila} hybrids,
@@
 developed incompatibilities as they independently evolved towards the same quantitative trait optimum,
-as different alleles fix in different populations. \llabel{citemani}
+as different alleles fix in different populations. %\llabel{citemani}
 Fluctuating selection has similarly been shown to accelerate speciation
@@
 In the case where fluctuations are synchronized, due to the size of the neutral system space, 
-populations may arrive at different system organizations when returning to previous phenotypes at a rate faster than by drift alone. \revpoint{1}{3}
+populations may arrive at different system organizations when returning to previous phenotypes at a rate faster than by drift alone. %\revpoint{1}{3}
 
@@
 while the fitness consequences of incompatibility in any one given network may be small, 
-the cumulative impact of system drift across the many different networks an organism relies on may be substantial. \llabel{rp11}
+the cumulative impact of system drift across the many different networks an organism relies on may be substantial. %\llabel{rp11}
 It remains to be seen how the relative strengths of these forces compare.
-
-% Of course the basic assumptions of this model will be violated in practice (constant selective pressures, etc.), however, this model can function as a ``neutral null'' description of gene network evolution (a strategy advocated for by \citet{lynch2007frailty, fay2008evaluating, koonin2016splendor}).
-If this model succeeds in describing actual system evolution, it can be inferred that the mechanism underlying species formation does not require the inclusion of adaptation or changes in selection to explain the emergence of hybrid incompatibility.
-
-% \paragraph{Population isolates and genetic load}
-% Above we have mostly discussed the case of two separated populations of equal size.
-% In contrast, consider a small subset of a large population that is suddenly isolated.
-% It will have a large amount of genetic variance, thanks to its history,
-% but a small population size -- and so can drift quite quickly until it exhausts its genetic variation.
-% By this point, it could have accumulated substantial incompatibility with the parental population.
-% Similarly, in a species with substantial isolation by distance, geographically local systems drift
-% might lead to genetic load due only to recombination, much as in \citep{phillips1996maintenance}.
 
@@
 making it more difficult for systems to make the required compensatory changes for system drift.
-% Whatever it is, however, it seems unlikely 
-% that it would approach the dimensionality of kryptotype space,
-% which is something related to the number of genes.
-% even under universal pleiotropy \citep{universal_pleiotropy, boyle2017expanded},
-
-Phenotypically equivalent system evolution is probably not only driven by neutral genetic drift. \llabel{r14}
+
+Phenotypically equivalent system evolution is probably not only driven by neutral genetic drift. %\llabel{r14}
 For one thing, movement along the set of equivalent networks
@@
 possibly due to epistasis, genetic architecture \citep{hermisson2003epistasis}, adaptive inertia \citep{baatz1997adaptive,alvarez2009effects}
-or weak gene flow could plausibly slow it down. \revpoint{AE}{1}\revpoint{AE}{3}
+or weak gene flow could plausibly slow it down. %\revpoint{AE}{1}\revpoint{AE}{3}
 More work on specific systems,
@@
 
-\paragraph{Geometric models} \llabel{quantnote}
+\paragraph{Geometric models} %\llabel{quantnote}
 The model we work with describes a continuum of possible systems,
@@
 For instance, attributes excluded from the phenotype as modeled here ignore the potential energy costs associated with excessively large (non-minimal) kryptotypes, as well as the relationship between a specific network architecture and robustness to mutational, transcriptional, or environmental noise.
-% Despite the model's simplifications and potentially nontrivial oversights, it may still roughly reflect and shed light upon speciation patterns as they occur in nature.
 More precise modeling will require better mechanistic understanding not only of biological systems,
@@
 and genetic variation in these systems.
-
-%Maybe compare to \citet{fierst}?
-%
-%hominids; neanderthal load.
 
@@
 \bibliography{krefs}
-%\end{multicols}
 
@@
 
-\includereviews
+%\includereviews
 

### 2021-09-14 final_system_drift.tex
added running title, corresponding author, data/code availability

@@
 \usepackage[bottom, symbol]{footmisc}
+
+\usepackage{bbding}
 
@@
 {\centering
-{\Huge \bf System drift and speciation} \\ \vspace{0.75cm}
+{\Huge \bf Title: System drift and speciation} \\ \vspace{0.5cm}
+{\huge \bf Runnning title: System drift and speciation} \\ \vspace{0.75cm}
 Joshua S.
@@
 \& Weill Cornell Medicine, New York, New York 10065, U.S.A.} \qquad Peter L.
-Ralph$^{\dagger \ddagger}$ \\ \vspace{0.5cm}
+Ralph$^{\dagger \ddagger}$ \Envelope \\ \vspace{0.5cm}
 $^{\dagger}${\footnotesize {Molecular and Computational Biology, University of Southern California, Los Angeles, California 90089, U.S.A. \\
-$^{\ddagger}$Departments of Mathematics and Biology \& The Institute for Ecology and Evolution, University of Oregon, Eugene, Oregon 97403, U.S.A.}} \\ \vspace{0.5cm}
+$^{\ddagger}$Departments of Mathematics and Biology \& The Institute for Ecology and Evolution, University of Oregon, Eugene, Oregon 97403, U.S.A. \\
+\Envelope \ Corresponding author}} \\ \vspace{0.5cm}
 {\small \texttt{jschiffman@nygenome.org} \qquad 
@@
 the Sloan Foundation and the NSF (under DBI-1262645) to PR.
+
+\section*{Code and data availability} 
+Code for the oscillator examples will be archived at \url{https://doi.org/10.5281/zenodo.5372992}. We will not be archiving data as this paper does not use any.
 

### 2021-09-14 title_page_revision.tex
added running title, corresponding author, data/code availability

@@
 \usepackage{authblk}
-\usepackage{lineno}
 \usepackage{multicol}
-\usepackage[bottom]{footmisc}
+\usepackage[bottom, symbol]{footmisc}
+
+\usepackage{bbding}
 
@@
 
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}{Lemma}
-\newtheorem{definition}{Definition}
 \newtheorem{example}{Example}
@@
 {\centering
-{\Huge \bf System drift and speciation} \\ \vspace{0.75cm}
-Joshua S. Schiffman$^{\dagger \ast}$ \qquad Peter L. Ralph$^{\ddagger \ast}$ \\ \vspace{0.5cm}
-$^{\dagger}${\footnotesize {Current affiliations: New York Genome Center, NY, NY 10013 \& Weill Cornell Medicine, NY, NY 10065  \\
-$^{\ddagger}$Departments of Mathematics and Biology \& The Institute for Ecology and Evolution, University of Oregon, Eugene, OR 97403 \\ 
-$^{\ast}$Work conducted at: Molecular and Computational Biology, University of Southern California, LA, CA 90089} } \\ \vspace{0.5cm}
+{\Huge \bf Title: System drift and speciation} \\ \vspace{0.5cm}
+{\huge \bf Runnning title: System drift and speciation} \\ \vspace{0.75cm}
+Joshua S.
+Schiffman$^{\dagger}$ \footnote{Current affiliations: New York Genome Center, New York, New York 10013, U.S.A.
+\& Weill Cornell Medicine, New York, New York 10065, U.S.A.} \qquad Peter L.
+Ralph$^{\dagger \ddagger}$ \Envelope \\ \vspace{0.5cm}
+$^{\dagger}${\footnotesize {Molecular and Computational Biology, University of Southern California, Los Angeles, California 90089, U.S.A. \\
+$^{\ddagger}$Departments of Mathematics and Biology \& The Institute for Ecology and Evolution, University of Oregon, Eugene, Oregon 97403, U.S.A. \\
+\Envelope \ Corresponding author}} \\ \vspace{0.5cm}
 {\small \texttt{jschiffman@nygenome.org} \qquad 
 \texttt{plr@uoregon.edu}} \\ \vspace{0.5cm}
-%\small \today \\
+\small \today \\
 \vspace{0.25cm}
